### 2020-05-14 Report/Sections/HHMM.tex
Restructure repository and add gitignore

% !TeX root = ../main.tex

\section{Background}

Hidden Markov models are useful when inferring a single unobserved process, but biological processes often involve multiple simultaneous hidden processes which can occur and at different time scales. For example, a preliminary observation of the killer whale dive data shown in figure (\ref{data}) shows that the behavior of this killer whale changes between approximately hour-long periods of predominately short, shallow dives and long, deep dives. Leos-Barajas et al. encounter a similar issue when modeling the movement of a harbor porpoise in the North Sea, and use it as a motivating example when they introduce hierarchical hidden Markov models.

\subsection{Hidden Markov Models}

\textit{Hidden Markov models} (or HMMs) are comprised of an unobserved Markov chain $X = (X_1, \ldots, X_T)$ and a sequence of observations $Y = (Y_1, \ldots, Y_T)$. Each random variable in the unobserved chain $X_t$ can take one of $N$ possible values, and $X$ has corresponeding probability transition matrix $\Gamma \in \bbR^{N \times N}$ and initial distribution $\delta \in \bbR^N$:

$$\delta_i = P(X_1 = i)$$

$$ \Gamma_{ij} = P(X_{t+1} = j | X_t = i) \qquad \forall t \in \{ 1, \ldots, T-1\} $$

Further, each random varaible $X_t$ emits an observation $Y_t$ that depends only on the value of $X_t$: $p(y_t|x_t, x_{t-1}, \ldots , x_1, y_{t-1}, \ldots , y_1) = p(y_t|x_t)$. A visualization of this structure can be seen if figure (\ref{HMM}). In the field of animal movement, the unobserved chain $X$ usually represents the latent behaviour of an animal (e.g. foraging, resting, migrating, etc.), while the observations $Y$ are often a series of step lengths and turning angles for land animals or depth data for marine animals.

The probability transition matrix $\Gamma$ and the parameters of the emission distributions, $\theta$, can be estimated by maximizing the likelihood of the observed data $y$, $\calL_{\text{HMM}}(y)$, with respect to the $\Gamma$ and $\theta$. In addition, $\calL_{\text{HMM}}(y)$ can be calculated using the \textit{forward algorithm}:

$$\calL_{\text{HMM}}(y) = \delta P(y_1) \prod_{t=2}^T \Gamma P(y_t) \mathbf{1}$$

where:

$$P(y_t) = \text{diag}(p(y_t|x_t = x_1), . . . , p(y_t|x_t = x_N ))$$

and $\mathbf{1}$ is an $N$-dimensional column vector of ones.

In order to enusre identifiability and right-stochasticity after optimizing $\calL_x(y)$, $\Gamma$ is parameterized using $\eta \in \bbR^{N \times N}$ and the following link function:

$$\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, \qquad \eta_{ii} = 0 \quad \forall i \in \{1, \ldots, N\}$$

This allows for unconstrained optimization over $\eta$ and removes the constraint that $\Gamma$ be right-stochastic. $\calL_x(y)$ can be maximized using any numerical optimizer.

\subsection{Hierarchical Hidden Markov Models}

A hierarchical hidden Markov model (or HHMM) is a variation of a hidden Markov model in which each hidden state of the original HMM $X_t$ emits both an observation $Y_t$ \textit{and} another, fine-scale hidden markov model of length $S^*_t$. This fine-scale HMM is comprised of a Markov chain $X^*_t = (X^*_{1,t}, \ldots, X^*_{S_t,t})$ and observations $Y^*_t = (Y^*_{1,t}, \ldots, Y^*_{S_t,t})$. As before, each fine-scale observation $Y^*_{s,t}$ depends only on the value of its corresponding hidden state, $X^*_{s,t}$. $X^*_{s,t}$ can take one of $N^{*(X_t)}$ values and is characterized by an initial distribtuion $\delta^{*(X_t)} \in \bbR^{N^{*(X_t)}}$ and probability transition matrix $\Gamma^{*(X_t)} \in \bbR^ {N^{*(X_t)} \times N^{*(X_t)}}$:

$$\delta^{*(x_t)}_i = P(X^*_{1,t} = i | X_t = x_t)$$

$$\Gamma^{*(x_t)}_{ij} = P(X^*_{s+1,t} = j | X^*_{s,t} = i, X_t = x_t) \qquad \forall s \in \{ 1, \ldots, S^*_t-1\}$$

Finally, the emission probabilities $p(y^*_{s,t} | x^*_{s,t})$ are parameterized by $\theta^{*(X_t)}$. Note the parameters of the fine-scale hidden markov model, $N^{*(X_t)}$, $\Gamma^{*(X_t)}$, $\delta^{*(X_t)}$, and $\theta^{*(X_t)}$ all depend upon the hidden state of the \textit{crude-scale} hidden markov model $X_t$. A visualization of this  structure can be seen in figure (\ref{HHMM}).

Due to the nested structure of a hierarchical hidden Markov model, the likelihood of an HHMM is still easy to calculate using the forward algorithm:

$$\calL_{\text{HHMM}}(y,y^*) = \delta P^*(y_1,y_1^*) \prod_{t=2}^T \Gamma P^*(y_t,y_t^*) \mathbf{1}$$

where:

$$P^*(y_t,y^*_t) = \text{diag}(p(y_t|x_t = x_1)\calL_{\text{HMM}}(y_t^*|X_t = x_1), . . . , p(y_t|x_t = x_N )\calL_{\text{HMM}}(y_t^*|X_t = x_N))$$.

Note that this formualtion assume that the crude-scale observations at a given time $Y_t$ and the fine-scale observation time series $Y_t^*$ are independent conditioned on $X_t$.

For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, model selection and model checking, see Adam et al. \cite{Adam:2019}

\subsection{State decoding}

Once a HMM or HHMM model is fit using the process described above, it is common to find the most likely sequence of hidden states $\hat X$ conditioned on the learned parameters by using the a dynamic programing algorithm called the Viterbi algorithm \cite{Viterbi:1967}. In the case of HHMMs, this can be followed by running the Viterbi algorithm again on each sub-dive state to find the mostly likely sequence of fine-scale hidden states $\hat X_t^*$ conditioned on the learned parameters and the estimated crude-state value $\hat X_t$. Note that while $\hat X$ is a maximum likelihood estimate of $X$, $\hat X^*_t$ is \textit{not} necessarily a maximum likelihood estimate of $\hat X^*_t$ because it is conditioned on the value of $\hat X_t$.

While the Viterbi algorithm is the de-facto standard in the current ecology literature, we suggest to instead find the \textit{probability} of each crude-level state (conditioned on the learned parameters) using the \textit{forward-backward algorithm}. Although not commonly used in the ecology literature, the forward-backward algorithm has the same time complexity as the forward algorithm and is also used to find the \textit{psuedoresiduals} of a given model, which is an important tool for model validation. In addition, For HHMMs in particular, the forward-backward algorithm can be used recursively to find the probability of the fine-level states $X^*_{s,t}$ exactly by marginalizing out $X_t$:

$$P(X^*_{s,t} = x^*_{s,t}) = \sum_{n=1}^N P(X_t = x_n)P(X^*_{s,t} = x^*_{s,t} | X_t = x_n)$$

Where $P(X_t = x_n)$ can be found using the forward-backward algorithm on the crude-level markov chain and $P(X^*_{s,t} = x^*_{s,t} | X_t = x_n)$ can be found by running the forward-backward algorithm on the fine-level HMM for every possible value of $X_t$.

%\vspace{1cm}

%Model selection commonly involves the use of an information criterion such as the Akaike information criterion (AIC) or Bayesian information criterion (BIC). However, these metrics tend to favor complex models with more hidden states than can reasonably be interpreted in a biological setting [11]. While future studies may use non-parametric clustering techniques to find the number of crude-level and dive-level hidden states, Leos-Barajas et al. use exploratory analysis to determine their parameterization, and this work does the same.
%The timescale of the crude-scale Markov chain ideally should be short enough to ensure adequate resolution in the crude scale, but long enough so that the fine-scale Markov chains emitted by the crude-scale hidden Markov model are long enough to be informative. Because crude-scale behavioral changes occur on the time scale of approximately ten minutes, and there are a approximately thirteen dives per ten minutes of observation, this study uses a crude time scale of ten minutes. This differs from Leos-Barajas et al., who set the crude time scale to one hour.
%The killer whale in this study appears to be alternating between periods of mostly pre- dominately short, shallow, smooth dives and periods interspersed with long, deep dives. As a result, we set the number of crude states as K = 2. The number of possible dive types was set as N = 3 to allow for additional flexibility beyond long, deep dives and short, shallow dives. Gamma distributions were used for the emission probability density function because of their flexibility and that the observations are all non-negative. All of these decision follow the lead of Leos-Barajas et al.
%One additional assumption made both here and in the Barajas et al. paper is that
%the distribution of observations given a dive type is constant across all K crude-scale dive-
%level Markov chains. Mathematically this means that α(1) = ... = α(K) and β(1) = ... =
%3.3 Parameter Estimation and Likelihood Maximization
%Like the classical hidden Markov model, it is possible to use the forward algorithm to find the likelihood of an observed sequence of dives, Lw(y). The likelihood of the hierarchical model differs from the traditional HMM only in that the emission probability, p(y(t)|x(t)), is replaced by the likelihood of the fine-scale Markov chain at step m, Lx(y(m)|w(m)). Note that
%9
%(K)
%maximization algorithm.
%β
%. This assumption simplifies the model and facilitates convergence in the likelihood

%this likelihood depends upon the state of the crude-scale Markov chain, since the parameters
%dictating the dynamics of X
%K
%(m)
%(m)
%depend upon the value of W Lw(y) = 􏰚p(y(1:M),W(M) = wi)
%i=1 K
%. Formally:
%= 􏰚Lx(y(M)|W(M) = wi)p(y(1:M−1),W(M) = wi) i=1
%KK
%= 􏰚Lx(y(M)|W(M) = wi)􏰚P(wj → wi)p(y(1:M−1),w(M−1) = wj)
%i=1 j=1
%y(m) is a T(m)-by-R matrix of observations, Lx(y(m)|W(m) = wi) is the likelihood of one dive- level Markov chain given the crude behavioral state wi, and Lw(y) is the likelihood of the entire hierarchical model. Like in the traditional HMM model, this leads to the following recurrence relation:
%K
%p(y(1:m),W(m) =wi)=Lx(y(m)|W(m) =wi)􏰚P(wj →wi)p(y(1:m−1),W(m−1) =wj)
%j=1
%p(y(1), W (1) = wi) = δWwi ∗ Lx(y(1)|W (1) = wi)
%where δW is the initial distribution of crude-scale behavioral states. Again, Leos-Barajas et al. express this algorithm concisely as a matrix product:
%􏰑M􏰒 Lw(y) = δW PW (y(1)) 􏰛 ΓW PW (y(m)) 1
%m=2
%where Pw(y(m)) = diag(Lx(y(m)|W(m) = w1),...,Lx(y(m)|W(m) = wK)). We again copy Leos-Barajas et al. and map the entries of all probability transition matrices to the real line using the same transformation from earlier:
%(m)
%= wj|P(W
%(t−1) exp(ηWij )
%= wi) = 􏰗Nk=1 exp(ηWik), ηWii = 0 ∀i
%γWij = P(W
%γ(k) =P(X(m,t) =x |P(X(m,t−1) =x,W(m) =w )= Xij
%exp(η(k) ) Xij j i k 􏰗N exp(η(k))
%with respect to ΓW , Γ(1:K), α, and β. The Viterbi algorithm is then used to find the most X
%likely sequence of values for the crude-scale behavioral state Markov chain. Once estimates for each internal state are found, we repeat the Viterbi algorithm on each dive-level Markov chain to estimate dive types.

### 2020-05-14 Report/Sections/appendix.tex
Restructure repository and add gitignore

% !TeX root = ../main.tex

\addcontentsline{toc}{section}{Appendices}
\renewcommand*{\thesubsection}{\Alph{subsection}}

\section*{Appendix}

\setcounter{subsection}{0}

\subsection{Equivalency of CarHMM and one-dimensional state-switching Ornstein-Uhlenbeck process}

A one-dimensional state-switching Ornstein-Uhlenbeck process \textbf{x} is the solution to the following stochastic differential equation:
%
$$dx_t = \beta_{b_t}(\gamma_{b_t} - x_t)dt + \omega_{b_t} dW_t$$
%
where $b_t$ is the behavior of the animal at time $t$, $\beta_{b_t}$ relates to rate at which the process returns to its mean value, $\gamma_{b_t}$ is the long-term mean value of the process, $\omega_{b_t}$ is related to short-term variance, and $W$ is a Brownian motion process. $b_t$ is described by an unobserved Markov process. The solution to this equation is known to be the following \cite{Michelot:2019}:
\begin{align*}
	x_{t+\delta} \sim \mathcal{N}\left((1-e^{-\beta_{b_t}\delta})\gamma_{b_t} + e^{-\beta_{b_t}\delta} x_t,\quad \frac{\omega_{b_t}^2}{2\beta_{b_t}} (1-e^{-2\beta_{b_t}\delta})\right)
\end{align*}
suppose that $\delta$ is constant for all observations, and introduce the following transformations:
\begin{align*}
	d_t = x_t, \qquad \mu_{RL,b_t} = \gamma_{b_t}, \qquad \phi_{b_t} = e^{-\beta_{b_t}\delta}, \qquad \sigma^2_{b_t} = \frac{\omega_{b_t}^2}{2\beta_{b_t}} (1-e^{-2\beta_{b_t}\delta})
\end{align*}
Then, we have the following:
\begin{align*}
	d_{t+\delta} \sim \mathcal{N}\left((1-\phi_{b_t})\gamma_{b_t} + \phi_{b_t} d_t,\quad \sigma_{b_t}^2 \right)
\end{align*}
%
If $\delta$ is fixed and $b_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities for the step length sequence. Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again.

%\subsection{Proof of correlation between $d_{t-1}$ and $d_t$}

%THIS IS WRONG
%The correlation between $d_{t-1}$ and $d_t$ is defined as $\left(\mathbbm{E}(d_t d_{t-1}) - \mu_{RL,b}^2 \right) / \sigma_b^2$. Let the probability distribution function of $d_t$ and $d_{t-1}$ be denoted as $p$. Then:

%\begin{align*}
%	\mathbbm{E}(d_t d_{t-1}) &= \int_0^\infty \int_0^\infty x_t x_{t-1} p(x_{t-1}) p(x_t|x_{t-1}) dx_t dx_{t-1} \\
	%
%	&= \int_0^\infty x_{t-1} p(x_{t-1}) \left( \int_0^\infty x_t p(x_t|x_{t-1}) dx_t \right) dx_{t-1} \\
	%
%	&= \int_0^\infty x_{t-1} p(x_{t-1}) \left( (1-\phi_b)\mu_{RL,b} + \phi_b x_{t-1} \right) dx_{t-1} \\
%	&= (1-\phi_b)\mu_{RL,b} \int_0^\infty x_{t-1} p(x_{t-1}) dx_{t-1} + \phi_b \int_0^\infty x_{t-1}^2 p(x_{t-1}) dx_{t-1} \\
%	&= (1-\phi_b)\mu_{RL,b}^2 + \phi_b (\mu_{RL,b}^2 + \sigma_b^2) \\
%	&= \mu_{RL,b}^2 + \phi_b\sigma_b^2
%\end{align*}

%So:

%\begin{align*}
%	corr(d_t,d_{t-1}) &= \frac{\mathbbm{E}(d_t d_{t-1}) - \mu_{RL,b}^2}{\sigma_b^2} \\
%	&= \frac{\mu_{RL,b}^2 + \phi_b\sigma_b^2 - \mu_{RL,b}^2}{\sigma_b^2} \\
%	&= \phi_b
%\end{align*}

### 2020-05-14 Report/Sections/carHMM.tex
Restructure repository and add gitignore

% !TeX root = ../main.tex

\section{Autocorrelation within Fine Scale Behaviors}

One of the key assumption of both HMMs and HHMM is \textit{conditional indpendence} between observations. In particular, given the state $X_t$ or $X^*_{s,t}$, $Y_t$ or $Y^*_{s,t}$ (respectively) is assumed to be independent from all other observations. Therefore,traditional HMMs can fail is when the observations $Y$ exhibit significant auto-correlation. Unfortunately this is often the case for fine-scale animal behaviors. Examples include fluking in marine mammals in Vancouver, BC (see the results section) and swiming behavior in horn sharks in southern California \cite{Adam:2019}.

One way to deal with autocorrelation in fine-scale behavioral processes is to use a state-switching continuous model such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying state of the animal. Continuous time models are advantagous because of their flexibiliy: they can be built up from aribitrarily complex stochastic differential equations and they allow for uneven step lengths in the observations sequence $\mathbf{Y}$. However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.

Another option is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}, in which autocorrelation is explicity modeled into the emission distributions of the HMM while maintaining the strucutre needed to run the forward algorithm for fast direct likelihood maximization. In particular, the CarHMM introduces autocorrleation into the HMM by assuming that an observation $Y_t$ has mean $(1-\phi_{x_t}) \cdot \mu_{x_t} + \phi_{x_t} \cdot y_{t-1}$, instead of $\mu_{x_t}$ in the case of a classic HMM. Note that the autocorrelation term $\phi_{x_t}$ depends upon the behavioral state of the animal. This model easily fits into the HHMM structure, but initially it seems to lack the flexibility and natural interpretation of continuous-time models.

\subsection{Equivalency of CarHMM and one-dimensional Ornstein-Uhlenbeck  state-switching process}

In fact, under certain conditions, the CarHMM and the state-switching continuous model are equivalent. In particular, the conditions are that (1) the underlying behavioral state of the continuous-time model must follow a Markov chain rather than a Markov process, and (2) the emission distributions of the CarHMM must be gaussian.

This allows the theoretically grounded continuous-time state-switching model to be used in the computational convieneint HMM (and therefore HHMM) framework. In addition, it gives new intpretation to the learned parameters of the CarHMM in the context of an Ornstein-Uhlenbeck process.

A one-dimensional state-switching Ornstein-Uhlenbeck process $y^*$ is the solution to the following stochastic differential equation:
%
$$dy^*_t = \beta_{x^*_t}(\gamma_{x^*_t} - y^*_t)dt + \omega_{x^*_t} dW_t$$
%
where $x^*_t$ is the fine-scale behavior of the animal at time $t$, $\beta_{x^*_t}$ relates to rate at which the process returns to its mean value, $\gamma_{x^*_t}$ is the long-term mean value of the process, $\omega_{x^*_t}$ is related to short-term variance, and $W$ is a Wiener process. As before, $x^*_t$ is described by an unobserved Markov process. The solution to this equation is known to be the following \cite{Michelot:2019}:
\begin{align*}
	y^*_{t+\delta} \sim \mathcal{N}\left((1-e^{-\beta_{x^*_t}\delta})\gamma_{x^*_t} + e^{-\beta_{x^*_t}\delta} y^*_t,\quad \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})\right)
\end{align*}
Now, suppose that $\delta$ is constant for all observations, as is the case for hidden Markov models. In addition, introduce the following transformations:
\begin{align*}
	\mu_{x^*_t} = \gamma_{x^*_t}, \qquad \phi_{x^*_t} = e^{-\beta_{x^*_t}\delta}, \qquad \sigma^2_{x^*_t} = \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})
\end{align*}
Then, we have the following:
\begin{align*}
	y^*_{t+\delta} \sim \mathcal{N}\left((1-\phi_{x^*_t})\mu_{x^*_t} + \phi_{x^*_t} y^*_t,\quad \sigma_{x^*_t}^2 \right)
\end{align*}
%
If $\delta$ is fixed and $x^*_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities. Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again. This allows for the pricipled construction of the continuous-time model to be combined with the computational convienence of the CarHMM.

\iffalse

\subsection{Transition Parameters}

The probability transition matrix $\bfA$ must be right stochastic, meaning that its entries must be between 0 and 1 and its rows must add to one. Lawler et al do not go into detail about how to parameterize $\bfA$ such that it is both identifiable and constraint free, but the following parameterization does the trick \cite{Barajas:2017}:
%
\begin{align*}
	a_{ij} = \frac{\exp(\eta_{ij})}{\sum_{\ell = 1}^k \exp(\eta_{i\ell})}, \qquad \eta_{ii} = 0 \quad \forall i \in \{1,\ldots,k\}
\end{align*}
%
This parameterization simplifies estimation of $\bfA$, since it eliminates constraints when maximizing the likelihood.

In total, there are $5k$ parameters from the emission distributions and $k^2 - k$ parameters from the probability transition matrix, leaving a total of $k^2 + 4k$ parameters to estimate.

\subsection{Likelihood Evaluation and State Decoding}

While it is possible to use the EM algorithm to estimate these parameters, Lawler et al use direct likelihood maximization instead. In order to do this, it is necessary to evaluate the likelihood of the observations $\bfY$ given parameters $\mu_{RL,b}, \phi_b, \sigma_b^2, c_b, \rho_b$, and $\bfA$. Let the emission probability density of $\theta_t$ and $d_t$ given $d_{t-1}$ and $b_t$ be denoted as $f(y_t|b_t,y_{t-1})$. Additionally, recall that the stationary distribution of $\bfA$ is $\delta$ such that $\delta \bfA = \delta$ and $\sum_i \delta_i = 1$. Assuming that the initial behavior of the animal is drawn from this stationary distribution, the likelihood of $\bfY$ can be calculated sequentially using the following recurrence relation:
%
\begin{align*}
	\alpha_1(b_1) = p(y_1,b_1) &= f(y_1|b_1)\delta_{b_1} \\\\\\
	%
	\alpha_t(b_t) = p(y_{1:t},b_t) &= f(y_t|b_t,y_{t-1})\sum_{b_{t-1} \in \{1,\ldots,k\}}a_{b_{t-1},b_t}p(y_{1:t-1},b_{t-1}) \\
	%
	&= f(y_t|b_t,y_{t-1})\sum_{b_{t-1} \in \{1,\ldots,k\}}a_{b_{t-1},b_t} \alpha_{t-1}(b_{t-1}) \\\\\\
	%
	L(\bfA,\mu_{RL,b},\phi_b,\sigma_b^2,c_b,\rho_b; \bfY) = p(y_{1:T-1}) &= \sum_{b_{T-1} \in \{1,\ldots,k\}} \alpha_{T-1}(b_{T-1}) 
\end{align*}	
%
Note that $\alpha_t$ is a $k$-dimensional vector, where each dimension corresponds to a distinct behavioral state. The process of evaluating this recurrence relationship sequentially is called the \textbf{forward algorithm} and is vital to estimating the parameters of HMMs via likelihood maximization. One particularly convenient feature of the forward algorithm is that its time complexity is linear in both the number of time steps and behavioral states. The forward algorithm can also be expressed more concisely as a matrix product:
%
$$L(\bfA,\mu_{RL,b},\phi_b,\sigma_b^2,c_b,\rho_b; \bfY) = \delta \bfP(y_1) \prod_{i=1}^{T-1} \bfA \bfP(y_t,y_{t-1}) \mathbf{1}$$
$$\bfP(y_1) = \text{Diag}\left( f(y_1|B_1 = 1), \ldots, f(y_1|B_1 = k) \right)$$
$$\bfP(y_t,y_{t-1}) = \text{Diag}\left( f(y_t|B_1 = 1, y_{t-1}), \ldots, f(y_t|B_1 = k, y_{t-1}) \right)$$
%
$\mathbf{1}$ is a column vector of ones. Once the emission and transition parameters have been estimated, the most likely sequence of underlying behavioral states $\bfB$ can be found using the Viterbi algorithm \cite{Viterbi:1967}.

\subsection{Similarity to Previous Work}

Lawler et al are not the first to incorporate autocorrelation within a hidden Markov model in an ecological setting. Woriskey et al \cite{Whoriskey:2016} developed a hidden Markov model called the HMMM, which is formulated as the following:
%
$$\left(\bfY_t - \bfY_{t-1} |\bfY_{t-1} - \bfY_{t-2}, b_{t-1}\right) \sim \mathcal{N}\left( \phi_{b_{t-1}} T(\theta_{b_{t-1}})(\bfY_{t-1} - \bfY_{t-2}),\Sigma\right)$$
%
where $\phi_b$ is the state-dependent auto-correlation term, $\theta_b$ is the state-dependent mean turning angle, $T(\theta)$ is a standard rotation matrix, and $\Sigma$ is a state-independent two-dimensional covariance matrix. Like Lawler, Woriskey fits $\phi_b$, $\theta_b$, and $\Sigma$ using likelihood maximization. The details of the two approaches are different, but both are similar in that they use auto-correlation within the emission probabilities of an HMM to model animal movement in two dimensions.

In addition, many continuous-time approaches model auto-correlation to account for non-uniform time intervals between time steps. For example, Michelot et al. \cite{Michelot:2019} model $\bfY$ as an integrated Ornstein-Uhlenbeck process, where the position $\bfY_t$ and velocity $\bfV_t$ are modeled as draws from a normal distribution which depends upon $\bfV_{t-1}$, $\bfY_{t-1}$, the behavioral state $b_{t-1}$, and the time step between observations $\Delta_{t-1}$. It is straightforward to adapt this continuous-time approach to an HMM by setting $\Delta_{t}$ to be constant for all values of $t$ and assuming that the behavior of the animal does not change between time steps. In fact, if the CarHMM is altered so that the emission distribution of $d_t$ is normal, then the resulting sequence of step-lengths is equivalent to a one-dimensional state-switching Ornstein-Uhlenbeck process (see the appendix for details).

While the idea of auto-correlation within HMMs is not new, Lawler et al are novel in their specific implementation of an auto-correlated HMM. In addition, they detail several additional tools for data preprocessing and interpretation which are useful when applying HMMs in an ecological setting.

\fi

### 2020-05-14 Report/Sections/conclusion.tex
Restructure repository and add gitignore

% !TeX root = ../main.tex

\section{Conclusion}

While incorporating autocorrelation within a hidden Markov model to analyze animal movement is not new, Lawler et al. introduce a new formulation of this model in the CarHMM. They also review several useful preprocessing tools such as the lag plot and a method for interpolation that includes dividing the data into separate groups if observations are far apart.

This paper summarizes the CarHMM and provides intuition behind the interpolation scheme laid out by Lawler et al. In addition, it shows that if the emission distributions are normal for the sequence of step-sizes $\bfD$, then the CarHMM models $\bfD$ as a one dimensional Ornstein-Uhlenbeck process.

Finally, the CarHMM is adapted to dive data and used to model the behavior of a killer whale off the coast of British Columbia, Canada. In particular, that whale exhibited auto-correlated velocities in the $z$-direction as well as sinusoidal behavior in dynamic body acceleration. The adjusted CarHMM is able to capture within-dive behaviors of active swimming and passive gliding, but more analysis is necessary to determine the exact number of within-dive behaviors and generalize the method to a larger number of dive types.

### 2020-05-14 Report/Sections/fourier.tex
Restructure repository and add gitignore

% !TeX root = ../main.tex

\section{Capturing Non-Markovian Behavior via the Fourier Transform}

Although the CarHMM and continuous-time model can effectively capture autocorrelation within a process, both still assume Markovian dynamics, i.e. that the observation $Y_t$ depends only the value of the behaviorial state $X_t$ and the previous observation $Y_{t-1}$. However, there are many animal movement process which violate the markov propertey. In particular, on a fine scale, swimming behavior of marine mammals can exhibit periodic behavior as the animal repeatedly flukes to propel itself forward. Work has been done in the past to model non-markovian dynamics in the \textit{behavioral} process \cite{Langrock:2012}, but addressing non-markovian dynamics within observations $Y*$ is still a relatively unstudied area. With improvements in tagging technology allowing for data collection at very high frequencies, noisy and non-markovian fine scale behavior appears to be on the rise.

To address this issue, we recommend borrowing techniques from the signal processing literature to compress the data and summarize its essential elements. In particular, we suggest a method which can capture periodic behavior by decompsing a signal into its fourier components.

Suppose a one-dimensional fine-scale process of length $T^*$, $y^*$ exhibits significant periodic and structured behavior that is clearly non-markovian. It is possible to transform this fine-scale process $y^*$ to a shorter, but higher dimensional process $\hat y^*$ by selecting a window of size $w$, dividing $y^*$ into $\lfloor T^* / w \rfloor$ seperate intervals of length $w$ (truncating $y^*$ appropriately), and taking the discrete fourier transform (DFT) of each interval. The resulting time series $\hat y^*$ will be a complex-valued, $w$-dimensional time series of length $\lfloor T^* / w \rfloor$.

Picking the window length $w$ should be done with care. $w$ should be long enough to capture the periodic behavior of the underlying process (at least twice as long as the length of a period), but short enough so that the resolution of the process remains high and behavioral changes can be captured via the HMM.

A visualization of transforming a one-dimensional sequence $y^*$ to a $w$-dimensional complex sequence $\hat y^*$ can be seen in figure (\ref{fig:fourier_example}).

### 2020-05-14 Report/Sections/intro.tex
Restructure repository and add gitignore

% !TeX root = ../main.tex

\section{Introduction}

The field of animal movement is in the midst of a ``data renaissance" in which advancements in tagging technology have given rise to an explosion of data available for statistical modeling. In response, researchers have introduced a variety of new statistical techniques to infer animal behavior from movement data \cite{Hooten:2017}.

These techniques usually fall into one of two different categories: continuous-time models and discrete-time models. Continuous-time models often model the dynamics of an animal as the solution to some stochastic differential equation and can incorporate observations taken at irregular time intervals. These methods are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference. On the other hand, discrete-time techniques use summary statistics taken at discrete time steps to model the animal's behavior and require that observations be taken at regular time intervals. The primary advantage of discrete-time methods is that the associated likelihood is relatively easy to compute, facilitating fast maximum likelihood estimates for parameters.

One of the most prevalent discrete-time techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \cite{Patterson:2017}. Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another conditioned on the underlying behavioral process. However, this assumption is often violated in the real world processes. For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) by Whoriskey et al \cite{Whoriskey:2016}.

\textit{The conditionally autoregressive hidden Markov model (CarHMM): Inferring Behavioral States from animal tracking data exhibiting conditional autocorrelation} by Ethan Lawler, Kim Whoriskey, William H. Aeberhard, Chris Fields, and Joanna Mills Flemming introduces a new framework to model autocorrelation in hidden Markov models. The paper also recommends several best practices for data exploration, model selection, and data preprocessing. This paper summarizes the findings of Lawler et al, elaborates on some of the findings and recommendations within, and draws a connection between the CarHMM and a continuous-time method presented by Michelot et al \cite{Michelot:2019}. The CarHMM is also adapted and applied to real-world dive data to analyze the intra-dive behavior of a killer whale off the coast of British Columbia, Canada.

### 2020-05-14 Report/Sections/results.tex
Restructure repository and add gitignore

% !TeX root = ../main.tex

\section{Case study- Fluking in Killer Whales}

\subsection{Data Collection and Preprocessing}

To demonstrate this methodology as applied to real-world data, the CarHMM was used to analyze dive data from a killer whale off the coast of British Columbia, Canada. The data was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 hertz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time range were ignored. In addition, the tagging technology dropped data between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed data within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 3.1 meters in depth and lasts for at least 10 seconds. Finally, for computational purposes, only dives that reached a depth of at least 20 meters were included in the analysis, resulting a total of 7 dives. Data preprocessing was done in part with the \textit{divebomb} package in Python \cite{Nunes:2018}.

The accelerometer recorded both the acceleration due to animal movement and due to gravity. The magnitude of acceleration due only to animal movement is the overall dynamic body acceleration (ODBA) of the killer whale and is known to be a good proxy of energy expenditure \cite{Gleiss:2011}. However, because accelerometer's orientation is unclear, the average acceleration for each component over the entire data collection window was subtracted out (to account for acceleration due to gravity) and the magnitude of the resulting vector was taken as an estimate for the ODBA. This ad-hoc method of removing the effect of gravity is highly prone to errors, especially because the killer whale changes its orientation often. Future work should use alternative methods to find the overall dynamic body acceleration of the whale.

Velocity in the vertical direction was estimated using a central finite difference of the depth data:
%
$$v_t = \frac{d_{t+1} - d_{t-1}}{2\Delta t} = \frac{d_{t+1} - d_{t-1}}{1/25} $$
%
where $v_t$ is measured in meters per second. Due to the high frequency of data collection, small observation errors in depth readings resulted in relatively large errors in the velocity data. To deal with this, the velocity data was smoothed using a Gaussian kernel function with a width of 51 data points ($\approx$ 1 second) ranging from -3 to 3 standard deviations. In particular, if $\phi$ is the pdf of a standard normal distribution, then:
%
\begin{align*}
v^{(smoothed)}_t &= \sum_{i=-25}^{25} \omega_i v_{t+i} \\
\omega_i &= \frac{\phi\left(\frac{3i}{25}\right)}{\sum_{j=-25}^{25}\phi\left(\frac{3j}{25}\right)}
\end{align*}
%
Figure (\ref{fig:smoothing}) shows the velocity data before and after smoothing.

\begin{figure}[h!]
	\centering
	\includegraphics[height=7.5in]{../Plots/smoothed_v.png}
	\caption{From top to bottom: dive depth, unsmoothed vertical velocity, smoothed vertical velocity, ODBA, and the sum of low-frequency Fourier coefficeints of the ODBA for a randomly selected dive of a killer whale.}
	\label{fig:smoothing}
\end{figure}

Figure (\ref{fig:smoothing}) also shows the calculated ODBA as a function of time for one a specific dive. The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs straightforwardly. As a result, the data was split into two-second intervals (100 data points each), and each interval was summarized by the Fourier transform of the ODBA and the velocity at the \textit{end} of the time interval. Velocity was taken at the end of the time interval because the behavior of the whale during the time interval has no impact on the velocity of the whale at the beginning of the time interval. Because velocity is highly correlated with the average acceleration within an interval, the mean was subtracted out of each ODBA time series before the Fourier transform was taken. To reduce the dimension of the ODBA Fourier transform, Fourier coefficients corresponding to frequencies less than or equal to 5 hertz were summed and frequencies above 5 hertz were ignored. One nice property of this process is that it reduces the number of data points by a factor of 100, which drastically speeds up parameter estimation. The optimal length of the time interval over which to take the Fourier transform is a tuning parameter that is difficult to find. An ideal time interval should be long enough give useful Fourier coefficients, but short enough to preserve as much information about the vertical velocity of the animal as possible.

\subsection{Lag Plot}

In order to find the appropriate HMM to model whale behavior, a lag plot was made for both velocity and the ODBA fourier sum. The results of doing so are shown in figure (\ref{fig:lag}). Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that the velocity exhibits a large degree of autocorrelation. While the ODBA Fourier sum also exhibits some autocorrelation, the relationship is less strong, so autocorrelation was not incorporated in the ODBA Fourier sums emission distribution.

\begin{figure}[h!]
	\centering
	\includegraphics[height=5in]{../Plots/lagplot.png}
	\caption{Lag plot of vertical velocity and $\tilde a$ (left) and a the associated normal kernel density estimates (right)}
	\label{fig:lag}
\end{figure}

\subsection{CarHMM adapted to dive data}

Let $v_t$ represent the velocity in the $z-$direction (derived from depth data) at the end of time interval $t$ and $\tilde a_t$ represent the ODBA Fourier sum of interval $t$. The CarHMM was adapted to the dive data using the following model:

\begin{align*}
	v_{t} &\sim \mathcal{N}\left(\phi_{b_t} v_{t-1} + (1-\phi_{b_t}) \mu_{RL,b_t}, \sigma_{b_t}^2\right) \\
	%
	\tilde a_t &\sim \Gamma\left(\tilde \mu_{b_t}, \tilde \sigma^2_{b_t}\right)
\end{align*}

Note that the arguments in the gamma distribution above correspond to the mean and variance of the distribution, \textit{not} the standard scale and shape parameters. In total, 3 behavioral states were chosen for exploratory analysis as an attempt to capture behavioral states with very negative velocities, very positive velocities, and velocities close to zero. This results in a total of $3^2 + 4*3 = 21$ parameters, all of which were estimated using direct likelihood maximization.

\subsection{Results}

The parameters of the estimated emission distributions for each behavioral state are shown in table (\ref{table:emis_dist}) below. Each distribution is also plotted in figure (\ref{fig:emis_dist}). Note that the autocorrelation within the velocity sequence is not captured in figure (\ref{fig:emis_dist}), so it is important to refer to the estimated auto-correlation parameter $\hat \phi$ from table (\ref{table:emis_dist}) when considering the emission distributions shown in figure (\ref{fig:emis_dist}).
%
\begin{figure}[h!]
	\centering
	\includegraphics[height=2in]{../Plots/emis_dist.png}
	\caption{Estimated probability distributions for the long-term vertical velocity and $\tilde a$ in each behavioral state. Note that the distribution over vertical velocity does not take autocorrelation into account.}
	\label{fig:emis_dist}
\end{figure}
%
\begin{table}[h!]
	\centering
	\begin{tabular}{l|lll}
		& State 1 & State 2 & State 3 \\ \hline
		$\hat \mu_{RL}$        & 0.66    & 0.68    & 0.22    \\
		$\hat \sigma^2$        & 0.095   & 0.070   & 1.30    \\
		$\hat \phi$            & 0.88    & 0.99    & 0.03    \\ \hline
		$\hat \tilde \mu$      & 0.69    & 0.15    & 3.70    \\
		$\hat \tilde{\sigma^2}$ & 0.41    & 0.067   & 3.16   
	\end{tabular}
	\caption{Table of estimated parameters for probability distributions over $v$ and $\tilde a$ for each behavioral state}
	\label{table:emis_dist}
\end{table}
%
The estimated probability transition matrix and associated stationary distribution is shown below:

$$\hat \Gamma = \begin{pmatrix} 
0.7983 & 0.1618 & 0.0398 \\
0.0120 & 0.8796 & 0.0003 \\
0.2425 & 0.0049 & 0.7526
\end{pmatrix} \qquad \hat \delta = \begin{pmatrix} 0.4141 & 0.5020 & 0.0839 \end{pmatrix}$$

Finally, the Viterbi-decoded dive behavior within a randomly selected dive (deeper than 20 meters) is shown in figure (\ref{fig:viterbi}).

\newpage

\begin{figure}[h!]
	\centering
	\includegraphics[height=8in]{../Plots/viterbi.png}
	\caption{Features of a particular killer whale dive and Viterbi-decoded estimates for the intra-dive behavioral states.}
	\label{fig:viterbi}
\end{figure}

\newpage

While the ecological meaning of these behavioral states is tenuous, we hypothesize the following interpretations. Roughly speaking, behavioral state 1 corresponds to mildly active swimming. The mean of $\tilde a$ in this state is larger than behavioral state 2, indicating more activity by the whale, but the autocorrleation term $\hat \phi$ is still very high, indicating that velocities don't change very much every two seconds. Behavioral state 2 corresponds to gliding or turning, where no active swimming is taking place. It is characterized by a low mean $\tilde a$ value and a very high autocorrelation term $\hat \phi$, both of which indicate little activity. Finally, behavioral state 3 represents sudden jerking or more vigorous and active swimming than behavioral state 1. This behavior is so rigorous that $\hat \phi$ drops almost to zero and $\hat \sigma^2$ subsequently rises significantly. In behavioral state 3, the velocity readings of the killer whale are essentially uncorrelated when recorded every two seconds.

\subsection{Future Work}

This analysis is largely exploratory with a purely heuristic approach to decide the number of behavioral states. Future work should repeat the analysis using several different values for the number of behavioral states. It is also important to validate these results by plotting psuedoresiduals and performing goodness-of-fit tests on the emission distributions.

Parameter estimation via likelihood maximization can be slow, especially with an observation rate of 50 hertz. If dives are modeled as uncorrelated batches of data, then using stochastic gradient ascent to maximize the likelihood may considerably speed up parameter estimation.

Finally, this work was run exclusively on dives deeper than 20 meters so that within-dive behaviors would be similar between dives. Future analysis could find a way to include a more heterogeneous set of dives using a framework such as the hierarchical hidden Markov model (HHMM) introduced by Leos-Barajas et al \cite{Barajas:2017}.

### 2020-05-14 Report/Sections/simulation.tex
Restructure repository and add gitignore

--- 2020-05-14 Report/Sections/results.tex
@@
 
-\section{Case study- Fluking in Killer Whales}
-
-\subsection{Data Collection and Preprocessing}
-
-To demonstrate this methodology as applied to real-world data, the CarHMM was used to analyze dive data from a killer whale off the coast of British Columbia, Canada.
-The data was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions.
-Observations were collected at a rate of 50 hertz.
-Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time range were ignored.
-In addition, the tagging technology dropped data between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed data within this time range were ignored as well.
-A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 3.1 meters in depth and lasts for at least 10 seconds.
-Finally, for computational purposes, only dives that reached a depth of at least 20 meters were included in the analysis, resulting a total of 7 dives.
-Data preprocessing was done in part with the \textit{divebomb} package in Python \cite{Nunes:2018}.
-
-The accelerometer recorded both the acceleration due to animal movement and due to gravity.
-The magnitude of acceleration due only to animal movement is the overall dynamic body acceleration (ODBA) of the killer whale and is known to be a good proxy of energy expenditure \cite{Gleiss:2011}.
-However, because accelerometer's orientation is unclear, the average acceleration for each component over the entire data collection window was subtracted out (to account for acceleration due to gravity) and the magnitude of the resulting vector was taken as an estimate for the ODBA.
-This ad-hoc method of removing the effect of gravity is highly prone to errors, especially because the killer whale changes its orientation often.
-Future work should use alternative methods to find the overall dynamic body acceleration of the whale.
-
-Velocity in the vertical direction was estimated using a central finite difference of the depth data:
-%
-$$v_t = \frac{d_{t+1} - d_{t-1}}{2\Delta t} = \frac{d_{t+1} - d_{t-1}}{1/25} $$
-%
-where $v_t$ is measured in meters per second.
-Due to the high frequency of data collection, small observation errors in depth readings resulted in relatively large errors in the velocity data.
-To deal with this, the velocity data was smoothed using a Gaussian kernel function with a width of 51 data points ($\approx$ 1 second) ranging from -3 to 3 standard deviations.
-In particular, if $\phi$ is the pdf of a standard normal distribution, then:
-%
-\begin{align*}
-v^{(smoothed)}_t &= \sum_{i=-25}^{25} \omega_i v_{t+i} \\
-\omega_i &= \frac{\phi\left(\frac{3i}{25}\right)}{\sum_{j=-25}^{25}\phi\left(\frac{3j}{25}\right)}
-\end{align*}
-%
-Figure (\ref{fig:smoothing}) shows the velocity data before and after smoothing.
+\section{Simulation Study}
 
@@
 While the ODBA Fourier sum also exhibits some autocorrelation, the relationship is less strong, so autocorrelation was not incorporated in the ODBA Fourier sums emission distribution.
-
-\begin{figure}[h!]
-	\centering
-	\includegraphics[height=5in]{../Plots/lagplot.png}
-	\caption{Lag plot of vertical velocity and $\tilde a$ (left) and a the associated normal kernel density estimates (right)}
-	\label{fig:lag}
-\end{figure}
-
-\subsection{CarHMM adapted to dive data}
-
-Let $v_t$ represent the velocity in the $z-$direction (derived from depth data) at the end of time interval $t$ and $\tilde a_t$ represent the ODBA Fourier sum of interval $t$.
-The CarHMM was adapted to the dive data using the following model:
-
-\begin{align*}
-	v_{t} &\sim \mathcal{N}\left(\phi_{b_t} v_{t-1} + (1-\phi_{b_t}) \mu_{RL,b_t}, \sigma_{b_t}^2\right) \\
-	%
-	\tilde a_t &\sim \Gamma\left(\tilde \mu_{b_t}, \tilde \sigma^2_{b_t}\right)
-\end{align*}
-
-Note that the arguments in the gamma distribution above correspond to the mean and variance of the distribution, \textit{not} the standard scale and shape parameters.
-In total, 3 behavioral states were chosen for exploratory analysis as an attempt to capture behavioral states with very negative velocities, very positive velocities, and velocities close to zero.
-This results in a total of $3^2 + 4*3 = 21$ parameters, all of which were estimated using direct likelihood maximization.
-
-\subsection{Results}
-
-The parameters of the estimated emission distributions for each behavioral state are shown in table (\ref{table:emis_dist}) below.
-Each distribution is also plotted in figure (\ref{fig:emis_dist}).
-Note that the autocorrelation within the velocity sequence is not captured in figure (\ref{fig:emis_dist}), so it is important to refer to the estimated auto-correlation parameter $\hat \phi$ from table (\ref{table:emis_dist}) when considering the emission distributions shown in figure (\ref{fig:emis_dist}).
-%
-\begin{figure}[h!]
-	\centering
-	\includegraphics[height=2in]{../Plots/emis_dist.png}
-	\caption{Estimated probability distributions for the long-term vertical velocity and $\tilde a$ in each behavioral state.
-Note that the distribution over vertical velocity does not take autocorrelation into account.}
-	\label{fig:emis_dist}
-\end{figure}
-%
-\begin{table}[h!]
-	\centering
-	\begin{tabular}{l|lll}
-		& State 1 & State 2 & State 3 \\ \hline
-		$\hat \mu_{RL}$        & 0.66    & 0.68    & 0.22    \\
-		$\hat \sigma^2$        & 0.095   & 0.070   & 1.30    \\
-		$\hat \phi$            & 0.88    & 0.99    & 0.03    \\ \hline
-		$\hat \tilde \mu$      & 0.69    & 0.15    & 3.70    \\
-		$\hat \tilde{\sigma^2}$ & 0.41    & 0.067   & 3.16   
-	\end{tabular}
-	\caption{Table of estimated parameters for probability distributions over $v$ and $\tilde a$ for each behavioral state}
-	\label{table:emis_dist}
-\end{table}
-%
-The estimated probability transition matrix and associated stationary distribution is shown below:
-
-$$\hat \Gamma = \begin{pmatrix} 
-0.7983 & 0.1618 & 0.0398 \\
-0.0120 & 0.8796 & 0.0003 \\
-0.2425 & 0.0049 & 0.7526
-\end{pmatrix} \qquad \hat \delta = \begin{pmatrix} 0.4141 & 0.5020 & 0.0839 \end{pmatrix}$$
-
-Finally, the Viterbi-decoded dive behavior within a randomly selected dive (deeper than 20 meters) is shown in figure (\ref{fig:viterbi}).
-
-\newpage
-
-\begin{figure}[h!]
-	\centering
-	\includegraphics[height=8in]{../Plots/viterbi.png}
-	\caption{Features of a particular killer whale dive and Viterbi-decoded estimates for the intra-dive behavioral states.}
-	\label{fig:viterbi}
-\end{figure}
-
-\newpage
-
-While the ecological meaning of these behavioral states is tenuous, we hypothesize the following interpretations.
-Roughly speaking, behavioral state 1 corresponds to mildly active swimming.
-The mean of $\tilde a$ in this state is larger than behavioral state 2, indicating more activity by the whale, but the autocorrleation term $\hat \phi$ is still very high, indicating that velocities don't change very much every two seconds.
-Behavioral state 2 corresponds to gliding or turning, where no active swimming is taking place.
-It is characterized by a low mean $\tilde a$ value and a very high autocorrelation term $\hat \phi$, both of which indicate little activity.
-Finally, behavioral state 3 represents sudden jerking or more vigorous and active swimming than behavioral state 1.
-This behavior is so rigorous that $\hat \phi$ drops almost to zero and $\hat \sigma^2$ subsequently rises significantly.
-In behavioral state 3, the velocity readings of the killer whale are essentially uncorrelated when recorded every two seconds.
-
-\subsection{Future Work}
-
-This analysis is largely exploratory with a purely heuristic approach to decide the number of behavioral states.
-Future work should repeat the analysis using several different values for the number of behavioral states.
-It is also important to validate these results by plotting psuedoresiduals and performing goodness-of-fit tests on the emission distributions.
-
-Parameter estimation via likelihood maximization can be slow, especially with an observation rate of 50 hertz.
-If dives are modeled as uncorrelated batches of data, then using stochastic gradient ascent to maximize the likelihood may considerably speed up parameter estimation.
-
-Finally, this work was run exclusively on dives deeper than 20 meters so that within-dive behaviors would be similar between dives.
-Future analysis could find a way to include a more heterogeneous set of dives using a framework such as the hierarchical hidden Markov model (HHMM) introduced by Leos-Barajas et al \cite{Barajas:2017}.

### 2020-05-14 Report/Sections/tools.tex
Restructure repository and add gitignore

% !TeX root = ../main.tex

\section{Additional Tools}

\subsection{Lag Plot}

One tool that Lawler et al use for data interpretation and model selection is a lag plot. A lag plot is a special case of a scatter plot where the $x$ coordinate of each point corresponds to a datum in a time series (except for the last one), and the $y$ coordinate corresponds to the datum directly after it. In the case of the CarHMM, $(d_t,d_{t+1})$ is plotted for every $t \in \{1, \ldots T-2\}$. If there is no auto-correlation within a given behavioral state, the lag plot should look like a group of spherically symmetric droplet patterns along the line $y=x$, and the number of droplets should correspond to the number of behavioral states. However, if there is significant auto-correlation within states, the droplets will be ``smeared" along the $y=x$ line and the number of behavioral states will be less clear. The lag plot is a good diagnostic tool to visually determine the number of behavioral states in an HMM and whether it is necessary to model within-state autocorrelation using the CarHMM. A lag plot for the killer whale case study is shown in figure (\ref{fig:lag}).

\subsection{Data Interpolation}

As mentioned earlier, the use of an HMM or CarHMM requires observations that are equispaced in time. However, this requirement in not always satisfied for real-world data, especially in the case of marine mammals, where the location of an animal is often only known when that animal comes to the surface. Lawler et al suggest linear interpolation as a solution to this issue, but warn that the uniform time step used for interpolation must be chosen with care. If the time step is too long, then the data will be over-smoothed and important information will be lost from the animal's dynamics. The number of effective observations will also be significantly less than the original data set. If the uniform time step is too short, however, then many interpolated observations will occur on the same line segment, resulting in repetitive step lengths and turning angles equal to zero. See figure (\ref{fig:interpolation}) for a visualization of both issues. To deal with this, Lawler et al. suggest using the following procedure:

\begin{enumerate}
	\item Pick a uniform time step $\Delta t_u$ and interpolate the path of the animal accordingly. Lawler et al suggest that $\Delta t_u$ be somewhere between the median and $3^{rd}$ quartile of all time steps in the raw data set. 
	\item Pick a ``group cutoff level" $\Delta t_g$. Divide observations into groups that are separated by time steps larger than $\Delta t_g$, and treat each group as an independent time series. Lawler et al suggest that $\Delta t_g$ be larger than $\Delta t_u$ to prevent fragmenting the observations too much, but less than $2*\Delta t_u$ to avoid having many interpolated points lie on the same line segment between true observations. Figure (\ref{fig:interpolation_group}) visualizes the issues of selecting $\Delta t_g$ to be too small or too large.
\end{enumerate}

\begin{figure}[h!]
	\centering
	\includegraphics[height=3in]{../Plots/Interpolation.png}
	\caption{Consequences of picking $\Delta t_u$ too large (left) or too small (right). The observed track is shown in black and the interpolated track is shown in red. If $\Delta t_u$ is too large, then the true path will be over-smoothed and information about the animal's path will be lost. If $\Delta t_u$ is too small, then multiple interpolated points will appear on the same line segment, resulting in repeated step sizes and zero turning angles.}
	\label{fig:interpolation}
\end{figure}

\begin{figure}[h!]
	\centering
	\includegraphics[height=3in]{../Plots/Interpolation_Group.png}
	\caption{Consequences of picking $\Delta t_g$ too large (above) or too small (below). The observed track is shown in black and the interpolated track is shown in red. If $\Delta t_g$ is too large, then many interpolated points will appear on some line segments, resulting in repeated step lengths and turning angles of zero degrees. If $\Delta t_g$ is too small, then many groups will appear and interpolated points not be able to create paths with turning angles.}
	\label{fig:interpolation_group}
\end{figure}

Even with the above restrictions, there is a considerable amount of freedom in choosing $\Delta t_u$ and $\Delta t_g$. Therefore, Lawler et al. introduce the two statistics to measure the quality of the choice of $\Delta t_u$ and $\Delta t_g$:

$$n_{prop} = \frac{\text{\# of interpolated points}}{\text{\# of observed locations}}$$

$$n_{adj} = \frac{\text{\# of interpolated points} - 2 \cdot \text{\# of groups}}{\text{\# of observed locations} - 2}$$

$n_{prop}$ measures the ratio of interpolated points to observed data points, while $n_{adj}$ measures the ratio of interpolated points that appear in the likelihood calculation to the number of observations that \textit{would} appear in the likelihood calculation had the data not been interpolated. Twice the number of groups is subtracted from the numerator and denominator of $n_{adj}$ because the first and last points of a group have no turning angle associated with them. Lawler et al claim that $n_{prop}$ and $n_{adj}$ should both be as close to one as possible. If these values are much larger than one, the uniform step size may be too small, resulting in replicated (or at least highly correlated) step lengths and turning angles close to zero. If these statistics are much less than one, the uniform step size may be too large, resulting in over-smoothing of the data and lost information.

Lawler et al comment that auto-correlation within the step length sequence is induced by interpolation but claim that further investigation is needed to discover the cause. To establish intuition behind this auto-correlation, suppose an interpolated point $\bfY'_s$ is equal to $\psi_s \bfY_t + (1-\psi_s) \bfY_{t+1}$, another interpolated point $\bfY'_{s+1}$ is equal to $\psi_{s+1} \bfY_{t+1} + (1-\psi_{s+1}) \bfY_{t+2}$, and a third interpolated point $\bfY'_{s+2}$ is equal to $\psi_{s+2} \bfY_{t+2} + (1-\psi_{s+2}) \bfY_{t+3}$, where $0 \leq \psi_s,\psi_{s+1},\psi_{s+2} \leq 1$. This must be the case since a linearly interpolated point is simply a convex combination of two observations. Then, we have:
%
\begin{align*}
	d'_s &= ||\bfY'_{s+1} - \bfY'_s|| \\
	%
	&= ||\psi_{s+1} \bfY_{t+1} + (1-\psi_{s+1}) \bfY_{t+2} - \psi_s \bfY_t - (1-\psi_s) \bfY_{t+1} || \\
	&= ||\psi_{s+1}(\bfY_{t+1} - \bfY_{t+2}) + \psi_s(\bfY_{t+1} - \bfY_t) + \bfY_{t+2} - \bfY_{t+1} ||  \\
	%
	&= ||(1-\psi_{s+1})(\bfY_{t+2} - \bfY_{t+1}) + \psi_s(\bfY_{t+1} - \bfY_t)||
\end{align*}
%
Similarly:
%
$$d'_{s+1} = ||(1-\psi_{s+2})(\bfY_{t+3} - \bfY_{t+2}) + \psi_{s+1}(\bfY_{t+2} - \bfY_{t+1})||$$
%
Both $(1-\psi_{s+1})$ and $\psi_{s+1}$ are positive, and the term $\bfY_{t+2} - \bfY_{t+1}$ shows up in both $d'_s$ and $d'_{s+1}$, so intuitively $d'_s$ and $d'_{s+1}$ should co-vary with one another in most cases. This provides evidence that, in general, the sequence of interpolated step-lengths $\bfD'$ exhibits more auto-correlation than the observed step lengths $\bfD$. It is therefore useful to model auto-correlation using the CarHMM if a uniform time grid is used to interpolate observations.

\subsection{Step-Size Normalization}

Lawler et al recommend dividing all step lengths by the mean observed step length to eliminate units and allow for easy comparison across data sources and animals. This process should be done with care, however, since scaling a random variable can sometimes change its distribution. Luckily, step length in the CarHMM is modeled using a gamma distribution, and gamma random variables still follow a gamma distribution when scaled by a constant. Step-size standardization can only be done if the analysis is done offline, since the mean step length is in flux for online analysis.

### 2020-05-14 Report/main.tex
Restructure repository and add gitignore

\documentclass[10pt]{article}
\input{header}
\input{defs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Inferring Fine Scale Behaviors Within Hierarchical Hidden Markov Models}
\author{Evan Sidrow}
\date{\today}

\begin{document}

\maketitle

\abstract

In the field of animal movement, recent advances in high-frequency tagging technology have made available a vast amount of data which can exhibit simultaneous behavioral processes occurring at different time scales. One way to model this data is to use a hierarchical hidden Markov model (HHMM), where the system is modeled as a nested structure of hidden Markov models (HMMs). At very short time scales, however, observations can exhibit complicated dependence structures which cannot be easily captured by traditional HMMs. This work investigates how to incorporate fine-scale processes into the larger structure of HHMMs while maintaining computational efficiency. We apply our method to dive data collected from a northern resident killer whale off the coast of British Columbia, Canada.

The field of animal movement is in the midst of a ``data renaissance" where advancements in tagging technology have given rise to an explosion of data available for statistical modeling. In particular, tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours. This results in a vast amount of data which often exhibits many different simultaneous behavioral processes occurring at different time scales.

One solution to this issue is to use a hierarchical hidden Markov model (HHMM). HHMMs model the entire system as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process. One nice property of HHMMs is that its likelihood is relatively easy to compute, facilitating fast maximum likelihood estimates for its associated parameters.

At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by a traditional HMMs. To address this issue, it is possible to model small-scale animal behavior as the solution to some stochastic differential equation, but these methods tend to be computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC).

This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. We bridge the gap between discrete hidden Markov models and continuous-time stochastic process models by showing that the two are equivalent under certain conditions. In addition, we extract features from highly structured sub-dive behaviors that otherwise could not be modeled with a simple HMM. Finally, we apply our method to dive data collected from a Northern resident killer whale off the coast of British Columbia, Canada.

%\input Sections/intro.tex
\input Sections/HHMM.tex
\input Sections/carHMM.tex
\input Sections/fourier.tex
%\input Sections/simulation.tex
%\input Sections/results.tex
%\input Sections/conclusion.tex

\newpage

\bibliographystyle{plain}
\bibliography{references}

\newpage

\input Sections/appendix.tex

\end{document}

### 2020-05-14 Code/HHMM.py
Add K=3,3 different substates

@@ -424,7 +424,7 @@ class HHMM:
                 ind += self.pars.K[1]**2
 
             # update fine theta
-            ind_start = ind
+            ind_start = int(ind)
             for k0 in range(self.pars.K[0]):
                 for k1 in range(self.pars.K[1]):
                     for feature in self.pars.features[1]:
@@ -501,7 +501,7 @@ class HHMM:
             ind += self.pars.K[1]**2
 
         # update fine theta
-        ind_start = ind
+        ind_start = int(ind)
         for k0 in range(self.pars.K[0]):
             for k1 in range(self.pars.K[1]):
                 for feature in self.pars.features[1]:

### 2020-05-14 Code/Parameters.py

@@ -26,8 +26,8 @@ class Parameters:
     def __init__(self):
 
         # loading parameters
-        self.cvc_file = '../Data/20190902-182840-CATs_OB_1.cvc'
-        self.csv_file = '../Data/20190902-182840-CATs_OB_1_001.csv'
+        self.cvc_file = '../../Data/20190902-182840-CATs_OB_1.cvc'
+        self.csv_file = '../../Data/20190902-182840-CATs_OB_1_001.csv'
 
         self.cvc_cols = ['date','time',
                          'GPS lat','GPS long','GPS alt','GPS speed',

### 2020-05-14 Code/Visualisor.py
add 3,3, same fine states

@@ -223,7 +223,7 @@ class Visualisor:
             else:
                 nrows += 2
 
-        plt.subplots(nrows,1,figsize=(7.5,nrows*7.5))
+        plt.subplots(nrows,1,figsize=(20,nrows*7.5))
 
         # get df state-by-state
         subdive = df[(df['dive_num'] >= sdive) & (df['dive_num'] <= edive)].copy()

### 2020-05-14 Code/Parameters.py
---

@@ -67,7 +67,8 @@ class Parameters:
                           'bottom_variance':{'corr':False,'f':'gamma'},
                           'dive_duration':{'corr':False,'f':'gamma'}},
 
-                         {'FoVeDBA_low':{'thresh':5,'corr':True,'f':'gamma'},
+                         {'VeDBA':{'corr':True,'f':'gamma'},
+                          'FoVeDBA_low':{'thresh':5,'corr':True,'f':'gamma'},
                           'FoVeDBA_high': {'thresh':5,'corr':True,'f':'gamma'},
                           'Vz': {'corr':True,'f':'normal'},
                           'peak_jerk': {'corr':False,'f':'normal'}, # in seconds

### 2020-05-14 Code/Preprocessor.py

@@ -279,6 +279,10 @@ class Preprocessor:
             dive_seg_features = {}
             dive_seg = subdive_df.iloc[ind_start:ind_start+nperseg]
 
+            # find average VeDBA
+            if 'VeDBA' in self.pars.features[1]:
+                dive_seg_features['VeDBA'] = dive_seg['VeDBA'].mean()
+
             # find FoVeDBA
             if ('FoVeDBA_low' in self.pars.features[1]) or ('FoVeDBA_high' in self.pars.features[1]):
 

### 2020-05-15 Code/HHMM.py
Add k 2,3 same fine states

@@ -149,6 +149,18 @@ class HHMM:
     def find_log_p_yt_given_xt(self,level,feature,data,data_tm1,mu,sig,corr,sample=0):
 
         # find log density of feature
+        if self.pars.features[level][feature]['f'] == 'multivariate_normal':
+
+            # find new mean if there is autocorrelation
+            if self.pars.features[level][feature]['corr'] and data_tm1 is not None:
+                mu = (1.0-corr)*mu + corr*data_tm1[feature]
+
+            if sample > 0:
+                return multivariate_normal.rvs(mu,sig,sample)
+            else:
+                return multivariate_normal.logpdf(data[feature],mu,sig)
+
+        # find log density of feature
         if self.pars.features[level][feature]['f'] == 'normal':
 
             # find new mean if there is autocorrelation

### 2020-05-15 Code/Parameters.py

@@ -41,13 +41,13 @@ class Parameters:
                          'roll','pitch','yaw']
 
         self.renamed_cvc_cols = {'dyn acc x': 'Ax',
-                                'dyn acc y': 'Ay',
-                                'dyn acc z': 'Az',
-                                'mag cal x': 'Mx',
-                                'mag cal y': 'My',
-                                'mag cal z': 'Mz',
-                                'Depth (100bar) [m]' : 'depth',
-                                'Temperature (depth) [°C]': 'temp'}
+                                 'dyn acc y': 'Ay',
+                                 'dyn acc z': 'Az',
+                                 'mag cal x': 'Mx',
+                                 'mag cal y': 'My',
+                                 'mag cal z': 'Mz',
+                                 'Depth (100bar) [m]' : 'depth',
+                                 'Temperature (depth) [°C]': 'temp'}
 
         # preprocessing parameters
         self.freq = 50 # in Hz
@@ -67,7 +67,7 @@ class Parameters:
                           'bottom_variance':{'corr':False,'f':'gamma'},
                           'dive_duration':{'corr':False,'f':'gamma'}},
 
-                         {'VeDBA':{'corr':True,'f':'gamma'},
+                         {'A':{'corr':True,'f':'multivariate_normal'},
                           'FoVeDBA_low':{'thresh':5,'corr':True,'f':'gamma'},
                           'FoVeDBA_high': {'thresh':5,'corr':True,'f':'gamma'},
                           'Vz': {'corr':True,'f':'normal'},

### 2020-05-15 Code/Preprocessor.py

@@ -61,7 +61,7 @@ class Preprocessor:
         only include needed columns in the dataset
         '''
 
-        cols_to_keep = set(['time','depth'])
+        cols_to_keep = set(['time','depth','Ax','Ay','Az'])
         acc_cols = ['FoVeDBA_low','FoVeDBA_high','peak_jerk','roll_at_pj']
         roll_cols = ['roll_at_pj']
         head_cols = ['heading_var']
@@ -280,8 +280,8 @@ class Preprocessor:
             dive_seg = subdive_df.iloc[ind_start:ind_start+nperseg]
 
             # find average VeDBA
-            if 'VeDBA' in self.pars.features[1]:
-                dive_seg_features['VeDBA'] = dive_seg['VeDBA'].mean()
+            if 'A' in self.pars.features[1]:
+                dive_seg_features['VeDBA'] = np.array([dive_seg['Ax'],dive_seg['Ay'],dive_seg['Az']])
 
             # find FoVeDBA
             if ('FoVeDBA_low' in self.pars.features[1]) or ('FoVeDBA_high' in self.pars.features[1]):

### 2020-05-15 Code/HHMM.py
Add "A" and try with K=2;2 130-200

@@ -87,7 +87,10 @@ class HHMM:
                 quantiles = np.linspace(1/(2*K),1-(1/(2*K)),K)
 
                 # first find mu
-                theta[0][feature]['mu'] = np.quantile(feature_data,quantiles)
+                if feature in ['Ax','Ay','Az']:
+                    theta[0][feature]['mu'] = np.mean(feature_data)*np.ones(K)
+                else:
+                    theta[0][feature]['mu'] = np.quantile(feature_data,quantiles)
 
                 # then get varaince of each quantile set of data
                 data_sorted = np.sort(feature_data)
@@ -124,7 +127,10 @@ class HHMM:
                     quantiles = np.linspace(1/(2*K),1-(1/(2*K)),K)
 
                     # first find mu
-                    theta[1][k0][feature]['mu'] = np.quantile(feature_data,quantiles)
+                    if feature in ['Ax','Ay','Az']:
+                        theta[1][k0][feature]['mu'] = np.mean(feature_data)*np.ones(K)
+                    else:
+                        theta[1][k0][feature]['mu'] = np.quantile(feature_data,quantiles)
                     theta[1][k0][feature]['mu'] *= norm.rvs(1,0.01)
 
                     # then get varaince of each quantile set of data
@@ -139,7 +145,7 @@ class HHMM:
                         theta[1][k0][feature]['sig'][k1]*= norm.rvs(1,0.01)
 
                     # finally update correlations randomly
-                    theta[1][k0][feature]['corr'] = 0.1*np.random.random(size=K) - 2.0
+                    theta[1][k0][feature]['corr'] = 0.1*np.random.random(size=K) - 1.0
 
         self.theta = theta
 
@@ -158,7 +164,7 @@ class HHMM:
             if sample > 0:
                 return multivariate_normal.rvs(mu,sig,sample)
             else:
-                return multivariate_normal.logpdf(data[feature],mu,sig)
+                return multivariate_normal.logpdf(mu,data[feature],sig)
 
         # find log density of feature
         if self.pars.features[level][feature]['f'] == 'normal':
@@ -424,10 +430,16 @@ class HHMM:
             # update crude theta
             for k0 in range(self.pars.K[0]):
                 for feature in self.pars.features[0]:
-                    self.theta[0][feature]['mu'][k0] = x[ind]
-                    self.theta[0][feature]['sig'][k0] = max(x[ind+1],eps)
-                    self.theta[0][feature]['corr'][k0] = x[ind+2]
[…]

### 2020-05-15 Code/Parameters.py

@@ -61,7 +61,7 @@ class Parameters:
         # parameters for HHMM features
         self.timescales = [(10,'min'),(2,'sec')]
         self.K = [2,2] # number of states per level
-        self.share_fine_states = False # are fine states same forall crude states?
+        self.share_fine_states = False # are fine states same for all crude states?
 
         self.features = [{'max_depth':{'corr':False,'f':'gamma'},
                           'bottom_variance':{'corr':False,'f':'gamma'},

### 2020-05-15 Code/Preprocessor.py

@@ -62,7 +62,8 @@ class Preprocessor:
         '''
 
         cols_to_keep = set(['time','depth','Ax','Ay','Az'])
-        acc_cols = ['FoVeDBA_low','FoVeDBA_high','peak_jerk','roll_at_pj']
+        acc_cols = ['A','Ahat_low','Ahat_high','FoVeDBA_low','FoVeDBA_high',
+                    'peak_jerk','roll_at_pj']
         roll_cols = ['roll_at_pj']
         head_cols = ['heading_var']
 
@@ -279,9 +280,42 @@ class Preprocessor:
             dive_seg_features = {}
             dive_seg = subdive_df.iloc[ind_start:ind_start+nperseg]
 
-            # find average VeDBA
+            # find average Acceleration
             if 'A' in self.pars.features[1]:
-                dive_seg_features['VeDBA'] = np.array([dive_seg['Ax'],dive_seg['Ay'],dive_seg['Az']])
+                Ax = np.mean(dive_seg['Ax'])
+                Ay = np.mean(dive_seg['Ay'])
+                Az = np.mean(dive_seg['Az'])
+                dive_seg_features['A'] = np.array([Ax,Ay,Az])
+
+            # find average Acceleration
+            if 'Ax' in self.pars.features[1]:
+                dive_seg_features['Ax'] = np.mean(dive_seg['Ax'])
+            if 'Ay' in self.pars.features[1]:
+                dive_seg_features['Ay'] = np.mean(dive_seg['Ay'])
+            if 'Az' in self.pars.features[1]:
+                dive_seg_features['Az'] = np.mean(dive_seg['Az'])
+
+            # find FoVeDBA
+            if ('Ahat_low' in self.pars.features[1]) or ('Ahat_high' in self.pars.features[1]):
+
+                freqs = np.fft.rfftfreq(nperseg, d=1/self.pars.freq)
+                fftx = np.absolute(np.fft.rfft(dive_seg['Ax']))
+                ffty = np.absolute(np.fft.rfft(dive_seg['Ay']))
+                fftz = np.absolute(np.fft.rfft(dive_seg['Az']))
+
+                if 'Ahat_low' in self.pars.features[1]:
+                    thresh = self.pars.features[1]['Ahat_low']['thresh']
+                    thresh_ind = max(np.where(freqs <= thresh)[0]) + 1
+                    dive_seg_features['Ahat_low'] = np.sum(fftx[1:thresh_ind]**2)
[…]

### 2020-05-18 Report/Sections/simulation.tex
Update simulation study

@@
 
-\begin{figure}[h!]
-	\centering
-	\includegraphics[height=7.5in]{../Plots/smoothed_v.png}
-	\caption{From top to bottom: dive depth, unsmoothed vertical velocity, smoothed vertical velocity, ODBA, and the sum of low-frequency Fourier coefficeints of the ODBA for a randomly selected dive of a killer whale.}
-	\label{fig:smoothing}
-\end{figure}
+To test this method, dive and acceleration data was simulated using the following procedure:
+
+\begin{enumerate}
+	\item 100 dive durations were simulated using a gamma distribution using an HMM generative model and the following parameters:
+
+	\begin{align*}
+		\Gamma &= \begin{pmatrix} 0.4 & 0.6 \\ 0.6 & 0.4 \end{pmatrix} \\
+		% 
+		&Y_t|X_t \sim \text{Gamma} \\
+		\bbE(Y_t|X_t = 1) &= 15, \quad \bbE(Y_t|X_t = 2) = 60 \\
+		%
+		\bbV(Y_t|X_t = 1) &= 25, \quad \bbV(Y_t|X_t = 2) = 100
+	\end{align*}
+\end{enumerate}
 

### 2020-05-18 Code/HHMM.py

@@ -33,7 +33,7 @@ class HHMM:
         self.eta = []
         self.ptm = []
 
-        eta_crude = -1.0 + np.random.normal(size=(self.pars.K[0],self.pars.K[0]))
+        eta_crude = -1.0 + 0.1*np.random.normal(size=(self.pars.K[0],self.pars.K[0]))
         ptm_crude = np.exp(eta_crude)
         ptm_crude = (ptm_crude.T/np.sum(ptm_crude,1)).T
 
@@ -43,7 +43,7 @@ class HHMM:
         eta_fine = []
         ptm_fine = []
         for _ in range(self.pars.K[0]):
-            eta_fine_k = -1.0 + np.random.normal(size=(self.pars.K[1],self.pars.K[1]))
+            eta_fine_k = -1.0 + 0.1*np.random.normal(size=(self.pars.K[1],self.pars.K[1]))
             ptm_fine_k = np.exp(eta_fine_k)
             ptm_fine_k = (ptm_fine_k.T/np.sum(ptm_fine_k,1)).T
             eta_fine.append(eta_fine_k)
@@ -103,7 +103,7 @@ class HHMM:
                         theta[0][feature]['sig'][k0] = 1.0
 
                 # finally update correlations randomly
-                theta[0][feature]['corr'] = np.random.random(size=K) - 3
+                theta[0][feature]['corr'] = np.random.random(size=K) - 2.0
 
 
 

### 2020-06-05 Report/Sections/carHMM.tex
Update simulation study
Many Updates

@@
 
-\section{Autocorrelation within Fine Scale Behaviors}
+\section{Results regarding CarHMMs and the Ornstein-Uhlenbeck process}
 
-One of the key assumption of both HMMs and HHMM is \textit{conditional indpendence} between observations.
-In particular, given the state $X_t$ or $X^*_{s,t}$, $Y_t$ or $Y^*_{s,t}$ (respectively) is assumed to be independent from all other observations.
-Therefore,traditional HMMs can fail is when the observations $Y$ exhibit significant auto-correlation.
-Unfortunately this is often the case for fine-scale animal behaviors.
-Examples include fluking in marine mammals in Vancouver, BC (see the results section) and swiming behavior in horn sharks in southern California \cite{Adam:2019}.
+\subsection{Equivalence of CarHMM and OU process}
 
-One way to deal with autocorrelation in fine-scale behavioral processes is to use a state-switching continuous model such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying state of the animal.
-Continuous time models are advantagous because of their flexibiliy: they can be built up from aribitrarily complex stochastic differential equations and they allow for uneven step lengths in the observations sequence $\mathbf{Y}$.
-However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.
+Once the CarHMM is fit by maximizing the likelihood in equation (\ref{CarHMM_likelihood}), the emission distributions can be interpretted as the solution of a state-switching Ornstein-Uhlenbeck process similar to the one introduced by Michelot et al \cite{Michelot:2019}.
+In particular, the following two conditions must be met:
+\begin{enumerate}
+	\item The underlying behavioral state of the continuous-time model must follow a Markov chain rather than a Markov process.
+	\item The emission distributions of the CarHMM must be normal.
+\end{enumerate}
+If both conditions (1) and (2) are met, then the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process.
+This gives new intpretation to the learned parameters of the CarHMM in the context of continuous-time model.
 
-Another option is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}, in which autocorrelation is explicity modeled into the emission distributions of the HMM while maintaining the strucutre needed to run the forward algorithm for fast direct likelihood maximization.
-In particular, the CarHMM introduces autocorrleation into the HMM by assuming that an observation $Y_t$ has mean $(1-\phi_{x_t}) \cdot \mu_{x_t} + \phi_{x_t} \cdot y_{t-1}$, instead of $\mu_{x_t}$ in the case of a classic HMM.
-Note that the autocorrelation term $\phi_{x_t}$ depends upon the behavioral state of the animal.
-This model easily fits into the HHMM structure, but initially it seems to lack the flexibility and natural interpretation of continuous-time models.
+\subsubsection{Proof}
 
-\subsection{Equivalency of CarHMM and one-dimensional Ornstein-Uhlenbeck  state-switching process}
-
-In fact, under certain conditions, the CarHMM and the state-switching continuous model are equivalent.
-In particular, the conditions are that (1) the underlying behavioral state of the continuous-time model must follow a Markov chain rather than a Markov process, and (2) the emission distributions of the CarHMM must be gaussian.
-
-This allows the theoretically grounded continuous-time state-switching model to be used in the computational convieneint HMM (and therefore HHMM) framework.
-In addition, it gives new intpretation to the learned parameters of the CarHMM in the context of an Ornstein-Uhlenbeck process.
-
-A one-dimensional state-switching Ornstein-Uhlenbeck process $y^*$ is the solution to the following stochastic differential equation:
+A one-dimensional state-switching Ornstein-Uhlenbeck process $y$ is the solution to the following stochastic differential equation:
 %
-$$dy^*_t = \beta_{x^*_t}(\gamma_{x^*_t} - y^*_t)dt + \omega_{x^*_t} dW_t$$
+$$dy_t = \beta_{x_t}(\gamma_{x_t} - y_t)dt + \omega_{x_t} dW_t$$
 %
-where $x^*_t$ is the fine-scale behavior of the animal at time $t$, $\beta_{x^*_t}$ relates to rate at which the process returns to its mean value, $\gamma_{x^*_t}$ is the long-term mean value of the process, $\omega_{x^*_t}$ is related to short-term variance, and $W$ is a Wiener process.
-As before, $x^*_t$ is described by an unobserved Markov process.
-The solution to this equation is known to be the following \cite{Michelot:2019}:
+where $x_t$ is the fine-scale behavior of the animal at time $t$, $\beta_{x_t}$ relates to rate at which the process returns to its mean value, $\gamma_{x_t}$ is the long-term mean value of the process, $\omega_{x_t}$ is related to short-term variance, and $W$ is a Wiener process.
+$x_t$ is described by an unobserved Markov process.
+If the behavioral state $x_t$ does not change between observations (i.e.
+$x_t$ follows a Markov chain), the solution to this equation is known to be the following \cite{Michelot:2019}:
 \begin{align*}
-	y^*_{t+\delta} \sim \mathcal{N}\left((1-e^{-\beta_{x^*_t}\delta})\gamma_{x^*_t} + e^{-\beta_{x^*_t}\delta} y^*_t,\quad \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})\right)
+y_{t+\Delta t} | x_{t:t+\Delta t} \sim \mathcal{N}\left((1-e^{-\beta_{x_t}\Delta t})\gamma_{x_t} + e^{-\beta_{x_t}\Delta t} y_t,\quad \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t})\right)
 \end{align*}
-Now, suppose that $\delta$ is constant for all observations, as is the case for hidden Markov models. In addition, introduce the following transformations:
-\begin{align*}
-	\mu_{x^*_t} = \gamma_{x^*_t}, \qquad \phi_{x^*_t} = e^{-\beta_{x^*_t}\delta}, \qquad \sigma^2_{x^*_t} = \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})
-\end{align*}
+Now, suppose that $\Delta t$ is constant for all observations, as is the case for hidden Markov models. In addition, introduce the following transformations:
+\begin{align}
+\mu_{x_t} = \gamma_{x_t}, \qquad \phi_{x_t} = e^{-\beta_{x_t}\Delta t}, \qquad \sigma^2_{x_t} = \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t})
+\label{CarHMM_to_OU}
+\end{align}
 Then, we have the following:
 \begin{align*}
-	y^*_{t+\delta} \sim \mathcal{N}\left((1-\phi_{x^*_t})\mu_{x^*_t} + \phi_{x^*_t} y^*_t,\quad \sigma_{x^*_t}^2 \right)
+y_{t+\Delta t} | x_{t:t+\Delta t} \sim \mathcal{N}\left((1-\phi_{x_t})\mu_{x_t} + \phi_{x_t} y_t,\quad \sigma_{x_t}^2 \right)
 \end{align*}
 %
-If $\delta$ is fixed and $x^*_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities.
+If $\Delta t$ is fixed and $x_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities.
 Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again.
 This allows for the pricipled construction of the continuous-time model to be combined with the computational convienence of the CarHMM.
+
+\subsection{Generalization to unequal time steps}
+
+Using this intuition, we generalize the CarHMM to data with unequal time steps: i.e.
+$\Delta t$ becomes a function of $t$.
+First, it is necessary to assume that the behavioral state $x_t$ only changes once between observations, which is a fair assumption if each time step $\Delta t$ is sufficiently small.
+Next, the following definitions are introduced:
+%
+$$\Gamma_{\Delta t,\Lambda} = \begin{pmatrix} 
+1-e^{-\lambda_1 \Delta t} & p_{12} e^{-\lambda_1 \Delta t} & \dots & p_{1N} e^{-\lambda_1 \Delta t} \\
+p_{21} e^{-\lambda_2 \Delta t} & 1-e^{-\lambda_2 \Delta t} & \dots & p_{2N} e^{-\lambda_2 \Delta t} \\
+\vdots & \vdots & \ddots & \vdots \\
+p_{N1} e^{-\lambda_N \Delta t} & p_{N2} e^{-\lambda_N \Delta t} & \dots  & 1-e^{-\lambda_N \Delta t}
+\end{pmatrix}$$
+%
+where $\Lambda = \{p,\lambda\}$, $p_{ij}$ is the probability that the animal moves from state $i$ to state $j$ given that it left state $i$, and $\lambda_i$ is the rate at which the animal leaves behavioral state $i$.
+Finally, we just note that the parameters in equation (\ref{CarHMM_to_OU}) become functions of the time step between observations $\Delta t$:
+%
+\begin{align*}
+\mu_{x_t}= \gamma_{x_t}, \qquad \phi_{x_t}(\Delta t) = e^{-\beta_{x_t}\Delta t}, \qquad \sigma^2_{x_t}(\Delta t) = \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t})
+\end{align*}
+%
+Then the likelihood of the CarHMM with unequal time steps is just the following:
+%
+$$\calL_{\text{CarHMM}}(y;\theta,\Lambda,\delta) = \delta \prod_{t=2}^{T} \Gamma_{\Delta t,\Lambda} P(y_t;\theta) \mathbf{1}$$
+%
+where:
+%
+$$P(y_t;\theta) = \text{diag}(p_\theta(y_t|y_{t-1}, \Delta t, X_t = x_1), . . . , p_\theta(y_t|y_{t-1}, \Delta t, X_t = x_N )), \qquad t > 1$$
+%
+and:
+%
+$$p_\theta(y_t|y_{t-1}, \Delta t, X_t = x_t) = \mathcal{N}\left((1-\phi_{x_t}(\Delta t))\mu_{x_t} + \phi_{x_t}(\Delta t) y_{t-1},\quad \sigma_{x_t}^2(\Delta t) \right).$$
+%
+now $\calL_{\text{CarHMM}}(y;\theta,\Lambda,\delta)$ is simply maximized with respect to $\theta = \{\beta,\omega,\gamma\}$, $\Lambda = \{p,\lambda\}$, and $\delta$ to find the maximum likelihood estimate of the CarHMM generative model with uneven time steps.
 

### 2020-06-05 Report/Sections/fourier.tex
Many Updates

@@
 
-\section{Capturing Non-Markovian Behavior via the Fourier Transform}
+\section{Model Formulation}
 
-Although the CarHMM and continuous-time model can effectively capture autocorrelation within a process, both still assume Markovian dynamics, i.e.
-that the observation $Y_t$ depends only the value of the behaviorial state $X_t$ and the previous observation $Y_{t-1}$.
-However, there are many animal movement process which violate the markov propertey.
-In particular, on a fine scale, swimming behavior of marine mammals can exhibit periodic behavior as the animal repeatedly flukes to propel itself forward.
-Work has been done in the past to model non-markovian dynamics in the \textit{behavioral} process \cite{Langrock:2012}, but addressing non-markovian dynamics within observations $Y*$ is still a relatively unstudied area.
-With improvements in tagging technology allowing for data collection at very high frequencies, noisy and non-markovian fine scale behavior appears to be on the rise.
+\subsection{Short-time Fourier Transform on Fine-scale process}
+
+One issue with HMMs is that they assume markovian dynamics conditioned on the hidden state, i.e.
+that any observation $Y^*_t$ depends only on the behaviorial state $X^*_t$ (here we focus on the fine-scale process).
+However, there are many animal movement processes which violate this markov propertey on very fine scales.
+For example swimming behaviour of marine mammals can be periodic since the animal repeatedly flukes to propel itself forward.
+Work has been done in the past to model non-markovian dynamics in the \textit{behavioural} process $X^*_t$ \cite{Langrock:2012}, but addressing non-markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area.
+With improvements in tagging technology allowing for data collection at very high frequencies, data exhibiting noisy and non-markovian fine scale behavior is likely to persist in future studies.
 
 To address this issue, we recommend borrowing techniques from the signal processing literature to compress the data and summarize its essential elements.
-In particular, we suggest a method which can capture periodic behavior by decompsing a signal into its fourier components.
+In particular, we suggest performing the discrete-time Short-time Fourier Transform (STFT) over each observed fine-scale process $Y^*_t$:
+%
+$$STFT\{Y^*_{t,t^*:t^*+w-1}\}(n) := \hat{Y}^{*(n)}_{t,t^*} = \sum_{n = 0}^{w-1} Y^*_{t,t^*+n}e^{-i \frac{2\pi k}{w} n} \qquad \forall n \in \{0, \ldots, w-1\}, \quad t^* \in \{1,w+1,2w+1, \ldots, w \left(\lfloor T^*_t / w \rfloor -1 \right) + 1\}.$$
+%
+The STFT slides a moving window of length $w$ accross the time series $Y_t^*$ with a step size of $w$ and transforms the domain of each window from time to frequency.
+This allows the spectrum of $Y_t^*$ at time $t^*$ to be summerized by a $w$-dimensional vector of fourier coeficients.
+While other step sizes can be used for the sliding window, we select $w$ to avoid serial dependence between windows.
 
-Suppose a one-dimensional fine-scale process of length $T^*$, $y^*$ exhibits significant periodic and structured behavior that is clearly non-markovian.
-It is possible to transform this fine-scale process $y^*$ to a shorter, but higher dimensional process $\hat y^*$ by selecting a window of size $w$, dividing $y^*$ into $\lfloor T^* / w \rfloor$ seperate intervals of length $w$ (truncating $y^*$ appropriately), and taking the discrete fourier transform (DFT) of each interval.
-The resulting time series $\hat y^*$ will be a complex-valued, $w$-dimensional time series of length $\lfloor T^* / w \rfloor$.
+If $Y^*_t \in \mathbb{R}^{T^*_t}$, then $\hat{Y}_t^* \in \mathbb{C}^{\lfloor T^*_t / w \rfloor \times w}$.
+While this allows $Y^*_t$ to be represneted in a way that eliminates obvious periodic behavior, the dataset itself is still approximately as large as $Y^*_t$ itself.
+To reduce the size of $\hat{Y}^*_t$, we propose taking summary statistics of each window as follows:
+%
+$$Z_{t,t^*}^{*(1)} = \mathcal{R}\left(\hat{Y}^{*(0)}_{t,t^*}\right) \qquad Z_{t,t^*}^{*(2)} = \frac{1}{w}\sum_{n=1}^{\tilde{w}}|\hat{Y}^{*(n)}_{t,t^*}|^2$$
+%
+$Z_{t,t^*}^{*(1)}$ is equal to the average value of $Y_{t,t^*:t^*+w-1}^*$ while $Z_{t,t^*}^{*(2)}$ is equal to the squared 2-norm of $Y_{t,t^*:t^*+w-1}^*$ that can be attributed to frequencies in the signal between $1$ and $\tilde{w}$ periods per window length.
+Both the window length $w$ and the max frequency $\tilde{w}$ are tuning parameters that should be tuned in a problem-specific way.
+$w$ should be long enough to capture the periodic behavior of the underlying process (at least as long as the length of a period), but short enough to avoid over-smoothing of the data and to maintain high resultion in the behavioral process $X^*$.
+$\tilde{w}$ should be selected such that the maximum frequency of $Y_t^*$ that makes biological sense is $\tilde{w}$ per window length.
+Note that these summary statistics are just one possible choice, and future studies can adjust the definitions as needed.
+A visualization of transforming a one-dimensional sequence $Y^*$ to $Z^*$ can be seen in figure (\ref{fig:fourier_example}).
 
-Picking the window length $w$ should be done with care.
-$w$ should be long enough to capture the periodic behavior of the underlying process (at least twice as long as the length of a period), but short enough so that the resolution of the process remains high and behavioral changes can be captured via the HMM.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=7.5in]{../../Plots/fourier_transform.png}
+	\caption{Visualization of transforming $Y^*_t$ into $Z^*_t$ using a sliding window and fourier transform.}
+	\label{fig:fourier_example}
+\end{figure}
 
-A visualization of transforming a one-dimensional sequence $y^*$ to a $w$-dimensional complex sequence $\hat y^*$ can be seen in figure (\ref{fig:fourier_example}).
+Finally, note that it is possible to accomodate for unequal time steps within each window by using the \textbf{non-uniform discrete Fourier transform (NDFT)}.
+We do not describe the details of this method in this work, but the generalization is straightforward.
+Refer to Bagchi et al \cite{Bagchi:2001} for details.
+
+\subsection{Model Structure: combining the HHMM and CarHMM}
+
+Hierarchical hidden Markov models can be used to jointly model simultaneous coarse-scale and fine-scale processes taking place simultaneously.
+However, as mentioned before, the fine-scale process $Y^*$ can often exhibit autocorrelation and intricate structure.
+Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit autocorrelation, especially in the window averge, $Z_{t,t^*}^{*(1)}$.
+Therefore, we replace the fine-scale HMM within the Hierarchical HMM with a CarHMM according to figure (\ref{fig:CarHHMM}).
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=7.5in]{../../Plots/CarHHMM.png}
+	\caption{Graphical representation of a CarHHMM. The additional arrows representing autocorrelation between observations are shown in red for emphasis.}
+	\label{fig:CarHHMM}
+\end{figure}
+
+The likelihood of this model is still easy to calculate using the forward algorithm:
+%
+$$\calL_{\text{CarHHMM}}(y,z^*;\theta,\theta^*,\Gamma,\Gamma^*,\delta,\delta^*) = \delta P(y_1,z_1^*;\theta,\theta^*,\Gamma^*,\delta^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\theta,\theta^*,\Gamma^*,\delta^*) \mathbf{1}$$
+%
+where:
+%
+\begin{align*}
+P(y_t,z_t^*;\theta,\theta^*,\Gamma^*,\delta^*)  = \text{diag}\left[p_{\theta}(y_t|X_t = x_1)\calL_{\text{CarHMM}}\left(z_t^*;\theta^{*(x_1)},\Gamma^{*(x_1)},\delta^{*(x_1)}\right), . . . , \right.\\
+\left. p_{\theta}(y_t|x_t = x_N )\calL_{\text{CarHMM}}\left(z_t^*;\theta^{*(x_N)},\Gamma^{*(x_N)},\delta^{*(x_N)}\right) \right]
+\end{align*}
+%

### 2020-06-05 Report/Sections/simulation.tex
Many Updates

@@
 
-To test this method, dive and acceleration data was simulated using the following procedure:
+\subsection{Data Simulation}
+
+To test the CarHHMM with STFT, a sequence of 100 marine-animal dives were simulated.
+The coarse-scale observations $Y$ were set as the duration of each dive, and the fine-scale observations $Y^*$ were set as one dimensional acceleration readings simulated at 50 hertz.
+Specifically, the following procedure was followed:
 
 \begin{enumerate}
-	\item 100 dive durations were simulated using a gamma distribution using an HMM generative model and the following parameters:
+	\item 100 dive durations were simulated using an HMM generative model with the following parameters:
 
@@
 		% 
-		&Y_t|X_t \sim \text{Gamma} \\
-		\bbE(Y_t|X_t = 1) &= 15, \quad \bbE(Y_t|X_t = 2) = 60 \\
+		&Y_t|X_t \sim Gamma \\
+		\bbE(Y_t|X_t = 1) &= 15 s, \quad \bbE(Y_t|X_t = 2) = 60 s \\
 		%
-		\bbV(Y_t|X_t = 1) &= 25, \quad \bbV(Y_t|X_t = 2) = 100
+		\bbV(Y_t|X_t = 1) &= 25 s^2, \quad \bbV(Y_t|X_t = 2) = 100 s^2
 	\end{align*}
+
+	\item Once the dive durations were calculated, for each $t \in \{1, \ldots, 100\}$, dive $t$ was broken into $\lfloor Y_t/2 \rfloor$ 2-second segments (the end of the dive sequence was discarded).
+Further, each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where $X^*_{t,t^*} \in \{1,2\}$ and $t^* \in \{1,101,201,\ldots,100*\lfloor Y_t/2 \rfloor + 1\}$.
+The parameters of the fine-scale behaviour Makov chain were set to be as follows:
+	%
+	$$\Gamma^{*(1)} = \begin{pmatrix} 0.25 & 0.75 \\ 0.75 & 025 \end{pmatrix} \qquad 	\Gamma^{*(2)} = \begin{pmatrix} 0.75 & 0.25 \\ 0.25 & 0.75 \end{pmatrix}$$
+	%
+	where $\Gamma^{*(1)}$ was used for dives where $X_t = 1$ and $\Gamma^{*(2)}$ was used for dives where $X_t = 2$
+
+	\item For each 2-second segment, the Fourier modes $\hat{Y}^*_{t,t^*}$ were simulated using the following procedure.
+Note that the $n^{th}$ Fourier mode of $\hat{Y}^*_{t,t^*}$ is denoted as $\hat{Y}^{*(n)}_{t,t^*}$ :
+
+	\begin{align*}
+	\hat{Y}^{*(0)}_{t,1} &\sim \mathcal{N} \left(\mu = 1, \sigma^2 = 0.01 \right) & \\
+	%
+	\hat{Y}^{*(0)}_{t,t^*} &\sim \mathcal{N} \left(\mu = 0.9 Y^{*(0)}_{t,t^*-100} + 0.1, \sigma^2 = 0.01 \right), & t^* \in \{101,201,\ldots, 100*\lfloor Y_t/2 \rfloor + 1\} \\
+	%
+	\hat{Y}^{*(n)}_{t,t^*} &= a_{t,t^*}^{(n)} i\sqrt{b^{(n)}_{t,t^*}}, & n \in \{1,\ldots,49\} \\\\
+	%
+	%
+	a_{t,t^*}^{(n)} &\sim  \left\{\begin{array}{lr}
+	-1 & w.p. \enspace 1/2 \\
+	1  & w.p. \enspace 1/2
+	\end{array}\right. \\
+	%
+	(b^{(n)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim Gamma(1/n^2, 1) & \\
+	%
+	(b^{(n)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim  \left\{\begin{array}{lr}
+	Gamma(1/n^2, 1) & \text{for } n \neq 2\\
+	Gamma(100,1) & \text{for } n = 2
+	\end{array}\right. & \\\\
+	%
+	%
+	\hat{Y}^{*(50)}_{t,t^*} &= 0 & \\
+	%
+	\hat{Y}^{*(n)}_{t,t^*}  &= -\hat{Y}^{*(100-n)}_{t,t^*}, & \qquad n \in \{51,\ldots,99\}
+	\end{align*}
+
+	Finally, $Y^*_{t,t^*:t^*+99}$ is set using the inverse discrete fourier transform of $\hat{Y}^*_{t,t^*}$:
+
+	$$Y^*_{t,t^*:t^*+99} = IDFT\left(\hat{Y}^*_{t,t^*}\right)$$
+
+	This gives a strong periodic component to acceleration when the subdive state $X^*_{t,t^*}= 2$. There are several practical reasons behind this construction:
+
+	\begin{enumerate}
+		\item $\hat{Y}^*_{t,t^*}$ is anti-symetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse fourier transform is real-valued.
+		\item $\hat{Y}^{*(n)}_{t,t^*}$ decays like $1/n^2$ to facilitate continutity.
 	\end{enumerate}
 
-Figure (\ref{fig:smoothing}) also shows the calculated ODBA as a function of time for one a specific dive.
+\end{enumerate}
+
+Note that this process idoes not result in a continuous sequence $Y^*_t$ since the average values of consecutive 2-second segments jump.
+However, note that these average values are highly autocorrelated, so the jumps are not too severe.
+See figure (\ref{fig:sim_data}) for details.
+In addition, $Y^*_t$ has the following desirable properties:
+
+\begin{enumerate}
+	\item $Z^{*(1)}_{t,t^*+1} | Z^{*(1)}_{t,t^*} \sim \mathcal{N} \left(\mu = 0.9 Z^{*(1)}_{t,t^*} + 0.1, \sigma^2 = 0.01 \right)$
+	\item $$Z^{*(2)}_{t,t^*} \sim \left\{\begin{array}{lr} 
+	\text{Gamma}(\sum_{n=1}^{\tilde{w}} \frac{1}{n^2},1) & \text{for } X^*_{t,t^*} = 1 \\
+	\text{Gamma}(20 + \sum_{n=2}^{\tilde{w}} \frac{1}{n^2},1) & \text{for } X^*_{t,t^*} = 2
+	\end{array}\right. $$
+\end{enumerate}
+
+So we can directly compare the results of the CarHHMM with the ground truth.
+
+\begin{figure}[!ht]
+	\centering
+	\includegraphics[width=5in]{../../Plots/sim_data.png}
+	\caption{Simulated Acceleration Data.}
+	\label{fig:sim_data}
+\end{figure}
+I still need to put in the actually results from the simulation study- I had to rethink how to do the simulation in the first place since enforcing continuity was REALLY hard to do in a principled way.
+
+\iffalse
+\subsection{Model Misspecification}
+
+There are two ways to simulate the fourier modes of the acceleration vector $\bfA_{s,t}$:
+
+\begin{enumerate}
+	\item $$\bbE_1\left(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}\right) = \phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} $$
+
+	\item $$\bbE_2(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}, X^*_{s-1,t}) = \left\{\begin{array}{lr}
+	\phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} & \text{for } X^*_{s,t} = X^*_{s-1,t}\\
+	\mu_{X^*_{s,t}} & \text{for } X^*_{s,t} \neq X^*_{s-1,t}
+	\end{array}\right\}$$
+\end{enumerate}
+
+\subsubsection{$\bbE_1$}
+
+First, look at $\bbE_1$.
+This comes with an issue in that when $X^*_{s,t}$ changes values, it will take a while for the process to ``converge" long-run mean $\mu_{X^*_{s,t}}$.
+In particular, if the behavioral process $X^*$ had been in state 1 for a while and then switches to state 2 for a while, then:
+
+\begin{align*}
+	\bbE_1\left(\bfA_{s+1,t}|X^*_{s:s+1,t} = 2, X^*_{s,t} = 1\right) &=  \phi \mu_1 + (1-\phi) \mu_2 \\
+	%
+	\bbE_1\left(\bfA_{s+2,t}|X^*_{s:s+2,t} = 2, X^*_{s,t} = 1\right) &= \phi\left( \phi \mu_1 + (1-\phi) \mu_2\right) + (1-\phi)\mu_2 \\
+	&= \phi^2 \mu_1 + (1-\phi^2) \mu_2\\
+	%
+	\bbE_1\left(\bfA_{s+n,t}|X^*_{s:s+n,t} = 2, X^*_{s,t} = 1\right) &=  \phi^n \mu_1 + (1-\phi^n) \mu_2 \\
+\end{align*}
+
+So, if using $\bbE_1$, there is an assumption that the fourier modes of the acceleration vector $\bfA$ linearly ``converge" to the new value $\mu_i$ from $\mu_j$ when $X^*_{s,t}$ changes from $i$ to $j$.
+An example of this can be seen in figure (\ref{fig:FOVEDBA}).
+
+ \begin{figure}[h!]
+ 	\centering
+ 	\includegraphics[height=3in]{../Code/test.png}
+ 	\caption{The FoVeDBA of simulated data using $\bbE_1$. Note that the FoVeDBA approaches $\approx$ 300 in state 2 from $\approx$ 10 in state 1 with roughly linear convergence.}
+ 	\label{fig:FOVEDBA}
+ \end{figure}
+
+ Note that if the generative model is correlated, but the inference model isn't, then quick jumps between states 1 and 2 can be missed because the chain never ``converges" to the new mean $\mu_i$, as shown in figure (\ref{fig:STATE}).
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[height=3in]{../Code/test2.png}
+	\caption{The probability of being in each behavioral state $X^*_{s,t}$ as estimated by the HHMM. The colors represent the true behavioral state.}
+	\label{fig:STATE}
+\end{figure}
+
+ \subsubsection{$\bbE_2$}
+
+Clearly, the model has issues decodeing the behavioral state when the model is misspecified.
+Therefore, if the true generative process does not have correlation when jumping between states, then it is better to use $\bbE_2$:
+
+$$\bbE_2(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}, X^*_{s-1,t}) = \left\{\begin{array}{lr}
+\phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} & \text{for } X^*_{s,t} = X^*_{s-1,t}\\
+\mu_{X^*_{s,t}} & \text{for } X^*_{s,t} \neq X^*_{s-1,t}
+\end{array}\right\}$$
+
+This fixes the issue of having to ``converge" to the new value of $\mu_i$ when jumping to state $i$.
+However, note that $\bbE_2$ depends upon both $X^*_{s,t}$ and $X^*_{s-1,t}$.
+This breaks the assumption in the forward algorithm that $\bfA_{s,t}$ depends only on $X^*_{s,t}$.
+One way to deal with this issues is to introduce a new hidden state $Z^*_{s,t} = (X^*_{s-1,t},X^*_{s,t})$.
+This fixes the issue, but now $Z^*_{s,t}$ can take one of $s^2$ values instead of $s$.
+This grows the state space of our model and can make inference harder.
+
+ \fi
+
+%Figure (\ref{fig:smoothing}) also shows the calculated ODBA as a function of time for one a specific dive.
 The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs straightforwardly.
@@
 
-\subsection{Lag Plot}
+%\subsection{Lag Plot}
 
-In order to find the appropriate HMM to model whale behavior, a lag plot was made for both velocity and the ODBA fourier sum.
+%In order to find the appropriate HMM to model whale behavior, a lag plot was made for both velocity and the ODBA fourier sum.
 The results of doing so are shown in figure (\ref{fig:lag}).

### 2020-06-05 Code/Preprocessor.py

@@ -250,6 +250,7 @@ class Preprocessor:
         del df_temp
 
         df = df[df['dive_num'] != -1]
+        df = df[df['depth'] > 0.2]
 
         return df,dives
 

### 2020-06-05 Code/Visualisor.py

@@ -109,8 +109,10 @@ class Visualisor:
             ncols = len(self.hhmm.theta[0])
         else:
             features = list(self.hhmm.theta[1][0].keys())
+            print(features)
+            features = ['Ahat_low']
             nrows = self.pars.K[0]
-            ncols = len(self.hhmm.theta[1][0])
+            ncols = len(features)
 
         fig,ax = plt.subplots(nrows,ncols,figsize=(7.5*ncols,7.5*nrows))
         ax = np.reshape(ax,(nrows,ncols))
@@ -124,7 +126,7 @@ class Visualisor:
 
                     dist = self.pars.features[0][feature]['f']
                     K = self.pars.K[0]
-                    colors = [cm.get_cmap('tab10')(i) for i in range(K)]
+                    colors = [cm.get_cmap('tab20')(i) for i in [1,0]]
                     legend = ['Dive Type %d'%(x+1) for x in range(K)]
                 else:
                     mu = self.hhmm.theta[1][row_num][feature]['mu']
@@ -132,18 +134,18 @@ class Visualisor:
 
                     dist = self.pars.features[1][feature]['f']
                     K = self.pars.K[1]
-                    colors = [cm.get_cmap('tab10')(i+self.pars.K[0]) for i in range(K)]
+                    colors = [cm.get_cmap('tab20')(i+self.pars.K[0]) for i in [2,1,0]]
                     legend = ['Subdive Behavior %d'%(x+1) for x in range(K)]
 
                 for state in range(K):
                     if dist == 'gamma':
                         shape = np.square(mu)/np.square(sig)
                         scale = np.square(sig)/np.array(mu)
-                        x = np.linspace(0,max(mu)+3*max(sig),1000)
+                        x = np.linspace(0.1,1e5,100000)
                         y = gamma.pdf(x,shape[state],0,scale[state])
                     elif dist == 'normal':
                         x = np.linspace(min(mu)-3*max(sig),max(mu)+3*max(sig),1000)
-                        y = norm.pdf(x,mu[state],sig[state])
+                        y = norm.pdf(x,0,sig[state])
                     elif dist == 'vonmises':
                         x = np.linspace(-np.pi,np.pi,1000)
                         y = vonmises.pdf(x,sig[state],loc=mu[state])
@@ -153,9 +155,16 @@ class Visualisor:
                     ax[row_num,col_num].set_xlabel(feature,fontsize=14)
                     ax[row_num,col_num].set_ylabel('Probability Density',fontsize=14)
[…]

### 2020-06-05 Report/Sections/HHMM.tex
Add plots to repo for overleaf

@@
 Hidden Markov models are useful when inferring a single unobserved process, but biological processes often involve multiple simultaneous hidden processes which can occur and at different time scales.
-For example, a preliminary observation of the killer whale dive data shown in figure (\ref{data}) shows that the behavior of this killer whale changes between approximately hour-long periods of predominately short, shallow dives and long, deep dives.
+For example, a preliminary observation of the killer whale dive data shown in figure (\ref{fig:data}) shows that the behavior of this killer whale changes between approximately hour-long periods of predominately short, shallow dives and long, deep dives.
 Leos-Barajas et al.
@@
 
-\subsection{Hidden Markov Models}
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=6.5in]{../Plots/raw_data.png}
+	\caption{Raw depth data of a Killer Whale off the coast of British Columbia, Canada.}
+	\label{fig:data}
+\end{figure}
 
-\textit{Hidden Markov models} (or HMMs) are comprised of an unobserved Markov chain $X = (X_1, \ldots, X_T)$ and a sequence of observations $Y = (Y_1, \ldots, Y_T)$.
+\subsection{Hidden Markov models}
+
+\textit{Hidden Markov models} (or HMMs) are comprised of an unobserved Markov chain $X = (X_1, \ldots, X_T)$ and a sequence of (possibly high-dimensional) observations $Y = (Y_1, \ldots, Y_T)$, each of length $T$.
 Each random variable in the unobserved chain $X_t$ can take one of $N$ possible values, and $X$ has corresponeding probability transition matrix $\Gamma \in \bbR^{N \times N}$ and initial distribution $\delta \in \bbR^N$:
 
-$$\delta_i = P(X_1 = i)$$
+$$\delta_i = Pr(X_1 = i)$$
 
-$$ \Gamma_{ij} = P(X_{t+1} = j | X_t = i) \qquad \forall t \in \{ 1, \ldots, T-1\} $$
+$$ \Gamma_{ij} = Pr(X_{t+1} = j | X_t = i) \qquad \forall t \in \{ 1, \ldots, T-1\} $$
 
-Further, each random varaible $X_t$ emits an observation $Y_t$ that depends only on the value of $X_t$: $p(y_t|x_t, x_{t-1}, \ldots , x_1, y_{t-1}, \ldots , y_1) = p(y_t|x_t)$.
-A visualization of this structure can be seen if figure (\ref{HMM}).
+Further, each random varaible $X_t$ emits an observation $Y_t$ whose distribution depends only on the value of $X_t$ and none of the preceding observations or behavioral states: $p_{\theta}(y_t|x_t, x_{t-1}, \ldots , x_1, y_{t-1}, \ldots , y_1) = p_{\theta}(y_t|x_t)$.
+Note that the emission distribution depends upon parameters $\theta$.
+A visualization of this dependence structure can be seen in figure (\ref{fig:HMM}).
 In the field of animal movement, the unobserved chain $X$ usually represents the latent behaviour of an animal (e.g.
-foraging, resting, migrating, etc.), while the observations $Y$ are often a series of step lengths and turning angles for land animals or depth data for marine animals.
+foraging, resting, migrating, etc.), while the observations $Y$ are often a series of step lengths and turning angles for land animals and either depth or accelerometer data (or both) for marine animals.
 
@@
 In addition, $\calL_{\text{HMM}}(y)$ can be calculated using the \textit{forward algorithm}:
-
-$$\calL_{\text{HMM}}(y) = \delta P(y_1) \prod_{t=2}^T \Gamma P(y_t) \mathbf{1}$$
-
+%
+$$\calL_{\text{HMM}}(y;\theta,\Gamma,\delta) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}$$
+%
 where:
-
-$$P(y_t) = \text{diag}(p(y_t|x_t = x_1), . . . , p(y_t|x_t = x_N ))$$
-
+%
+$$P(y_t;\theta) = \text{diag}(p_{\theta}(y_t|X_t = x_1), . . . , p_{\theta}(y_t|X_t = x_N ))$$
+%
 and $\mathbf{1}$ is an $N$-dimensional column vector of ones.
 
-In order to enusre identifiability and right-stochasticity after optimizing $\calL_x(y)$, $\Gamma$ is parameterized using $\eta \in \bbR^{N \times N}$ and the following link function:
+In order to enusre identifiability and right-stochasticity after optimizing $\calL_{\text{HMM}}(y)$, $\Gamma$ is parameterized using $\eta \in \bbR^{N \times N}$ and the following link function:
 
@@
 
-This allows for unconstrained optimization over $\eta$ and removes the constraint that $\Gamma$ be right-stochastic. $\calL_x(y)$ can be maximized using any numerical optimizer.
+This allows for unconstrained optimization over $\eta$ and removes the constraint that $\Gamma$ be right-stochastic.
+$\calL_{\text{HMM}}(y;\theta,\Gamma,\delta)$ can be maximized using any numerical optimizer.
 
-\subsection{Hierarchical Hidden Markov Models}
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=3in]{../Plots/HMM.png}
+	\caption{Graphical representation of a traditional HMM.}
+	\label{fig:HMM}
+\end{figure}
 
-A hierarchical hidden Markov model (or HHMM) is a variation of a hidden Markov model in which each hidden state of the original HMM $X_t$ emits both an observation $Y_t$ \textit{and} another, fine-scale hidden markov model of length $S^*_t$.
-This fine-scale HMM is comprised of a Markov chain $X^*_t = (X^*_{1,t}, \ldots, X^*_{S_t,t})$ and observations $Y^*_t = (Y^*_{1,t}, \ldots, Y^*_{S_t,t})$.
-As before, each fine-scale observation $Y^*_{s,t}$ depends only on the value of its corresponding hidden state, $X^*_{s,t}$.
-$X^*_{s,t}$ can take one of $N^{*(X_t)}$ values and is characterized by an initial distribtuion $\delta^{*(X_t)} \in \bbR^{N^{*(X_t)}}$ and probability transition matrix $\Gamma^{*(X_t)} \in \bbR^ {N^{*(X_t)} \times N^{*(X_t)}}$:
+\subsection{Hierarchical HMMs}
 
-$$\delta^{*(x_t)}_i = P(X^*_{1,t} = i | X_t = x_t)$$
+A hierarchical hidden Markov model (or HHMM) is a variation of a hidden Markov model in which each hidden state of the original HMM $X_t$ emits both an observation $Y_t$ as well as another fine-scale hidden markov model of length $T^*_t$.
+This fine-scale HMM is comprised of a Markov chain $X^*_t = (X^*_{t,1}, \ldots, X^*_{t,T^*_t})$ and observations $Y^*_t = (Y^*_{t,1}, \ldots, Y^*_{t,T^*_t})$.
+As before, each fine-scale observation $Y^*_{t,t^*}$ depends only on the value of its corresponding hidden state, $X^*_{t,t^*}$.
+$X^*_{t,t^*}$ can take one of $N^*$ values and is characterized by an initial distribtuion $\delta^{*(X_t)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(X_t)} \in \bbR^{N^* \times N^*}$:
 
-$$\Gamma^{*(x_t)}_{ij} = P(X^*_{s+1,t} = j | X^*_{s,t} = i, X_t = x_t) \qquad \forall s \in \{ 1, \ldots, S^*_t-1\}$$
+$$\delta^{*(x_t)}_i = Pr(X^*_{t,1} = i | X_t = x_t)$$
 
-Finally, the emission probabilities $p(y^*_{s,t} | x^*_{s,t})$ are parameterized by $\theta^{*(X_t)}$.
-Note the parameters of the fine-scale hidden markov model, $N^{*(X_t)}$, $\Gamma^{*(X_t)}$, $\delta^{*(X_t)}$, and $\theta^{*(X_t)}$ all depend upon the hidden state of the \textit{crude-scale} hidden markov model $X_t$.
-A visualization of this  structure can be seen in figure (\ref{HHMM}).
+$$\Gamma^{*(x_t)}_{ij} = Pr(X^*_{t,t^*+1} = j | X^*_{t,t^*} = i, X_t = x_t) \qquad \forall t^* \in \{ 1, \ldots, T^*_t-1\}$$
+
+Finally, the fine-scale emission probabilities $p_{\theta^{*(X_t)}}(y^*_{s,t} | x^*_{s,t})$ are parameterized by $\theta^{*(X_t)}$.
+Note the parameters of the fine-scale hidden Markov model, $\Gamma^{*(X_t)}$, $\delta^{*(X_t)}$, and $\theta^{*(X_t)}$ all depend upon the hidden state of the \textit{crude-scale} hidden markov model.
+However, depending upon the discretion of the researcher, it is possible to force any of these parameters to be independent of the crude-scale hidden state $X_t$.
+A visualization of the full structure of the HHMM can be seen in figure (\ref{fig:HHMM}).
 
 Due to the nested structure of a hierarchical hidden Markov model, the likelihood of an HHMM is still easy to calculate using the forward algorithm:
+%
+$$\calL_{\text{HHMM}}(y,y^*;\theta,\theta^*,\Gamma,\Gamma^*,\delta,\delta^*) = \delta P(y_1,y_1^*;\theta,\theta^*,\Gamma^*,\delta^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\theta,\theta^*,\Gamma^*,\delta^*) \mathbf{1}$$
+%
+where:
+%
+\begin{align*}
+	P(y_t,y_t^*;\theta,\theta^*,\Gamma^*,\delta^*)  = \text{diag}\left[p_{\theta}(y_t|x_t = x_1)\calL_{\text{HMM}}\left(y_t^*;\theta^{*(x_1)},\Gamma^{*(x_1)},\delta^{*(x_1)}\right), . . . , \right.\\
+	\left. p_{\theta}(y_t|x_t = x_N )\calL_{\text{HMM}}\left(y_t^*;\theta^{*(x_N)},\Gamma^{*(x_N)},\delta^{*(x_N)}\right) \right]
+\end{align*}
+%
+Note that this formualtion assumes that the crude-scale observations at a given time $Y_t$ and the fine-scale observation time series $Y_t^*$ are independent of one another when conditioned on $X_t$.
 
-$$\calL_{\text{HHMM}}(y,y^*) = \delta P^*(y_1,y_1^*) \prod_{t=2}^T \Gamma P^*(y_t,y_t^*) \mathbf{1}$$
+For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, model selection and model checking, see Adam et al \cite{Adam:2019}.
 
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=6.5in]{../Plots/HHMM.png}
+	\caption{Graphical representation of a traditional HHMM.}
+	\label{fig:HHMM}
+\end{figure}
+
+\subsection{Conditionally autoregressive HMMs}
+
+One of the key assumptions of both HMMs and HHMMs is \textit{conditional indpendence} between observations at both the crude and fine scale.
+Namely, given the state $X_t$ or $X^*_{t,t^*}$, $Y_t$ or $Y^*_{t,t^*}$ (respectively) is assumed to be independent from all other observations.
+Therefore, traditional HMMs and HHMMs can fail when the observation sequence $Y$ exhibits significant auto-correlation in time.
+Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swiming behavior of horn sharks off the coast of Southern California \cite{Adam:2019}.
+
+One way to deal with autocorrelation in fine-scale behavioral processes is to use a state-switching continuous model such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behaviour state of the animal.
+Continuous time models are advantagous because of their flexibiliy: they can be built up from aribitrarily complex stochastic differential equations and they allow for uneven step lengths in the observations sequence $Y$.
+However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.
+
+Another option is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}, in which autocorrelation is explicity modeled into the emission distributions of the HMM while maintaining the strucutre needed to run the forward algorithm for fast direct likelihood maximization.
+In particular, if the emission distribution of observation $Y_t$ is parameterized by its mean and variance, i.e.
+$\theta_{x_t} = \{\mu_{x_t},\sigma^2_{x_t}\}$, the CarHMM introduces autocorrleation into the HMM by assuming that an observation $Y_t$ has mean $(1-\phi_{x_t}) \cdot \mu_{x_t} + \phi_{x_t} \cdot y_{t-1}$ rather than $\mu_{x_t}$.
+Note that the autocorrelation term $\phi_{x_t}$ depends upon the behavioral state of the animal.
+This model easily fits into the HHMM structure, but seems to lack the flexibility and natural interpretation of continuous-time models.
+However, we prove in the following section that under certain conditions these two models are in fact equivalent.
+
+The likelihood of CarHMM is still compatible with the forward algorithm:
+\begin{equation}
+\calL_{\text{CarHMM}}(y) = \delta \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}
+\label{CarHMM_likelihood}
+\end{equation}
 where:
+%
+$$P(y_t;\theta) = \text{diag}(p_\theta(y_t|y_{t-1},X_t = x_1), . . . , p_\theta(y_t|y_{t-1},X_t = x_N )), \qquad t > 1$$
+%
+and the graphical model associated with the strcture of a CarHMM is shown in figure (\ref{fig:CarHMM}). Note that the first observation $y_1$ is assumed to be fixed as an initial value.
 
-$$P^*(y_t,y^*_t) = \text{diag}(p(y_t|x_t = x_1)\calL_{\text{HMM}}(y_t^*|X_t = x_1), . . . , p(y_t|x_t = x_N )\calL_{\text{HMM}}(y_t^*|X_t = x_N))$$.
-
-Note that this formualtion assume that the crude-scale observations at a given time $Y_t$ and the fine-scale observation time series $Y_t^*$ are independent conditioned on $X_t$.
-
-For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, model selection and model checking, see Adam et al. \cite{Adam:2019}
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=3in]{../Plots/CarHMM.png}
+	\caption{Graphical representation of a traditional CarHMM. The additional arrows representing autocorrelation between observations are shown in red for emphasis.}
+	\label{fig:CarHMM}
+\end{figure}
 
@@
 
-Once a HMM or HHMM model is fit using the process described above, it is common to find the most likely sequence of hidden states $\hat X$ conditioned on the learned parameters by using the a dynamic programing algorithm called the Viterbi algorithm \cite{Viterbi:1967}.
+Once an HMM, HHMM, or CarHMM model is fit using the process described above, it is common to find the most likely sequence of hidden states $\hat X$ conditioned on the learned parameters by using a dynamic programing algorithm called the Viterbi algorithm \cite{Viterbi:1967}.
 In the case of HHMMs, this can be followed by running the Viterbi algorithm again on each sub-dive state to find the mostly likely sequence of fine-scale hidden states $\hat X_t^*$ conditioned on the learned parameters and the estimated crude-state value $\hat X_t$.
@@
 While the Viterbi algorithm is the de-facto standard in the current ecology literature, we suggest to instead find the \textit{probability} of each crude-level state (conditioned on the learned parameters) using the \textit{forward-backward algorithm}.
-Although not commonly used in the ecology literature, the forward-backward algorithm has the same time complexity as the forward algorithm and is also used to find the \textit{psuedoresiduals} of a given model, which is an important tool for model validation.
+The forward-backward algorithm has the same time complexity as the forward algorithm and is also used to find the \textit{psuedoresiduals} of a given model, which is an important tool for model validation.
 In addition, For HHMMs in particular, the forward-backward algorithm can be used recursively to find the probability of the fine-level states $X^*_{s,t}$ exactly by marginalizing out $X_t$:

### 2020-06-05 Report/Sections/appendix.tex

@@
 
+\subsection{Dimension Reduction on $\hat{Y_t^*}$}
+
+First, we simply use down-sampling and only record every $w^{th}$ time step of $\hat{Y}_t^*$.
+This both reduces the space of $\hat{Y}_t^*$ to $\mathbb{C}^{\lfloor S^*_t / w \rfloor \times w}$ and ensures that none of the sliding windows overlap when taking the STFT.
+This is important because HMMs assume temporal independence between observations.
+Next, the dimension of $\hat{Y}_t^*$ can be cut in half by recognizing that the $Y_t^*$ is real-valued, and therefore $\hat{Y}_{t,k}^*$ is equal to the complex-congugate of $\hat{Y}_{t,w-k-1}^*$.
+Finally, by Parseval's Thereom we have that:
+
+$$\sum_{n = 0}^{w-1} |Y^*_{t+n}|^2 = \sum_{n = 0}^{w-1} |\hat{Y}^*_{t,n}|^2$$
+
 \subsection{Equivalency of CarHMM and one-dimensional state-switching Ornstein-Uhlenbeck process}
 
-A one-dimensional state-switching Ornstein-Uhlenbeck process \textbf{x} is the solution to the following stochastic differential equation:
+If it is the case that (1) the underlying behavioral state of the continuous-time model must follow a Markov chain rather than a Markov process, and (2) the emission distributions of the CarHMM are gaussian, then the CarHMM and the state-switching continuous model are equivalent.
+This allows the theoretically grounded continuous-time state-switching model to be used in the computational convieneint HMM (and therefore HHMM) framework.
+In addition, it gives new intpretation to the learned parameters of the CarHMM in the context of an Ornstein-Uhlenbeck process.
+
+A one-dimensional state-switching Ornstein-Uhlenbeck process $y^*$ is the solution to the following stochastic differential equation:
 %
-$$dx_t = \beta_{b_t}(\gamma_{b_t} - x_t)dt + \omega_{b_t} dW_t$$
+$$dy^*_t = \beta_{x^*_t}(\gamma_{x^*_t} - y^*_t)dt + \omega_{x^*_t} dW_t$$
 %
-where $b_t$ is the behavior of the animal at time $t$, $\beta_{b_t}$ relates to rate at which the process returns to its mean value, $\gamma_{b_t}$ is the long-term mean value of the process, $\omega_{b_t}$ is related to short-term variance, and $W$ is a Brownian motion process.
-$b_t$ is described by an unobserved Markov process.
+where $x^*_t$ is the fine-scale behavior of the animal at time $t$, $\beta_{x^*_t}$ relates to rate at which the process returns to its mean value, $\gamma_{x^*_t}$ is the long-term mean value of the process, $\omega_{x^*_t}$ is related to short-term variance, and $W$ is a Wiener process.
+As before, $x^*_t$ is described by an unobserved Markov process.
 The solution to this equation is known to be the following \cite{Michelot:2019}:
 \begin{align*}
-	x_{t+\delta} \sim \mathcal{N}\left((1-e^{-\beta_{b_t}\delta})\gamma_{b_t} + e^{-\beta_{b_t}\delta} x_t,\quad \frac{\omega_{b_t}^2}{2\beta_{b_t}} (1-e^{-2\beta_{b_t}\delta})\right)
+y^*_{t+\delta} \sim \mathcal{N}\left((1-e^{-\beta_{x^*_t}\delta})\gamma_{x^*_t} + e^{-\beta_{x^*_t}\delta} y^*_t,\quad \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})\right)
 \end{align*}
-suppose that $\delta$ is constant for all observations, and introduce the following transformations:
+Now, suppose that $\delta$ is constant for all observations, as is the case for hidden Markov models. In addition, introduce the following transformations:
 \begin{align*}
-	d_t = x_t, \qquad \mu_{RL,b_t} = \gamma_{b_t}, \qquad \phi_{b_t} = e^{-\beta_{b_t}\delta}, \qquad \sigma^2_{b_t} = \frac{\omega_{b_t}^2}{2\beta_{b_t}} (1-e^{-2\beta_{b_t}\delta})
+\mu_{x^*_t} = \gamma_{x^*_t}, \qquad \phi_{x^*_t} = e^{-\beta_{x^*_t}\delta}, \qquad \sigma^2_{x^*_t} = \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})
 \end{align*}
@@
 \begin{align*}
-	d_{t+\delta} \sim \mathcal{N}\left((1-\phi_{b_t})\gamma_{b_t} + \phi_{b_t} d_t,\quad \sigma_{b_t}^2 \right)
+y^*_{t+\delta} \sim \mathcal{N}\left((1-\phi_{x^*_t})\mu_{x^*_t} + \phi_{x^*_t} y^*_t,\quad \sigma_{x^*_t}^2 \right)
 \end{align*}
 %
-If $\delta$ is fixed and $b_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities for the step length sequence.
+If $\delta$ is fixed and $x^*_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities.
 Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again.
+This allows for the pricipled construction of the continuous-time model to be combined with the computational convienence of the CarHMM.
 

### 2020-06-05 Report/main.tex
Update simulation study
Many Updates
Updates from Overleaf

@@
 
-\title{Inferring Fine Scale Behaviors Within Hierarchical Hidden Markov Models}
+\title{Inferring Fine-Scale Animal Behaviors using Hidden Markov Models}
 \author{Evan Sidrow}
@@
 \abstract
-
-In the field of animal movement, recent advances in high-frequency tagging technology have made available a vast amount of data which can exhibit simultaneous behavioral processes occurring at different time scales.
-One way to model this data is to use a hierarchical hidden Markov model (HHMM), where the system is modeled as a nested structure of hidden Markov models (HMMs).
-At very short time scales, however, observations can exhibit complicated dependence structures which cannot be easily captured by traditional HMMs.
-This work investigates how to incorporate fine-scale processes into the larger structure of HHMMs while maintaining computational efficiency.
-We apply our method to dive data collected from a northern resident killer whale off the coast of British Columbia, Canada.
 
@@
 We bridge the gap between discrete hidden Markov models and continuous-time stochastic process models by showing that the two are equivalent under certain conditions.
-In addition, we extract features from highly structured sub-dive behaviors that otherwise could not be modeled with a simple HMM.
-Finally, we apply our method to dive data collected from a Northern resident killer whale off the coast of British Columbia, Canada.
+In addition, we extract features from highly structured sub-dive behaviors using signal processing techniques.
+These features otherwise could not be modeled with a simple HMM.
+Finally, we apply our method to dive data collected from two Northern resident killer whales off the coast of British Columbia, Canada.
 
@@
 \input Sections/fourier.tex
-%\input Sections/simulation.tex
-%\input Sections/results.tex
-%\input Sections/conclusion.tex
+\input Sections/simulation.tex
+\input Sections/results.tex
+\input Sections/conclusion.tex
 

### 2020-06-08 Code/Parameters.py
Add new plot

@@ -61,7 +61,7 @@ class Parameters:
         # parameters for HHMM features
         self.timescales = [(10,'min'),(2,'sec')]
         self.K = [2,2] # number of states per level
-        self.share_fine_states = False # are fine states same for all crude states?
+        self.share_fine_states = True # are fine states same for all crude states?
 
         self.features = [{'max_depth':{'corr':False,'f':'gamma'},
                           'bottom_variance':{'corr':False,'f':'gamma'},

### 2020-06-08 Report/Sections/background.tex
Updates from Overleaf

--- 2020-06-05 Report/Sections/HHMM.tex
@@
 \textit{Hidden Markov models} (or HMMs) are comprised of an unobserved Markov chain $X = (X_1, \ldots, X_T)$ and a sequence of (possibly high-dimensional) observations $Y = (Y_1, \ldots, Y_T)$, each of length $T$.
-Each random variable in the unobserved chain $X_t$ can take one of $N$ possible values, and $X$ has corresponeding probability transition matrix $\Gamma \in \bbR^{N \times N}$ and initial distribution $\delta \in \bbR^N$:
+In the field of animal movement, the unobserved chain $X$ usually represents the latent behaviour of an animal (e.g.
+foraging, resting, migrating, etc.), while the observations $Y$ are often a series of step lengths and turning angles for land animals and either depth or accelerometer data (or both) for marine animals.
+Each random variable in the unobserved chain $X_t$ can take one of $N$ possible values, and $X$ has corresponding probability transition matrix $\Gamma \in \bbR^{N \times N}$ and initial distribution $\delta \in \bbR^N$:
 
@@
 
-Further, each random varaible $X_t$ emits an observation $Y_t$ whose distribution depends only on the value of $X_t$ and none of the preceding observations or behavioral states: $p_{\theta}(y_t|x_t, x_{t-1}, \ldots , x_1, y_{t-1}, \ldots , y_1) = p_{\theta}(y_t|x_t)$.
+Further, each random variable $X_t$ emits an observation $Y_t$ whose distribution depends only on the value of $X_t$ and none of the preceding observations or behavioral states: $p_{\theta}(y_t|x_t, x_{t-1}, \ldots , x_1, y_{t-1}, \ldots , y_1) = p_{\theta}(y_t|x_t)$.
 Note that the emission distribution depends upon parameters $\theta$.
 A visualization of this dependence structure can be seen in figure (\ref{fig:HMM}).
-In the field of animal movement, the unobserved chain $X$ usually represents the latent behaviour of an animal (e.g.
-foraging, resting, migrating, etc.), while the observations $Y$ are often a series of step lengths and turning angles for land animals and either depth or accelerometer data (or both) for marine animals.
-
-The probability transition matrix $\Gamma$ and the parameters of the emission distributions, $\theta$, can be estimated by maximizing the likelihood of the observed data $y$, $\calL_{\text{HMM}}(y)$, with respect to the $\Gamma$ and $\theta$.
-In addition, $\calL_{\text{HMM}}(y)$ can be calculated using the \textit{forward algorithm}:
+The probability transition matrix $\Gamma$ and the parameters of the emission distributions $\theta$ can be estimated by maximizing the likelihood of the observed data $y$, $\calL_{\text{HMM}}(y;\theta,\Gamma,\delta)$, with respect to $\Gamma$, $\delta$, and $\theta$.
+In addition, $\calL_{\text{HMM}}(y;\theta,\Gamma,\delta)$ can be calculated using the \textit{forward algorithm} as follows:
 %
@@
 and $\mathbf{1}$ is an $N$-dimensional column vector of ones.
-
-In order to enusre identifiability and right-stochasticity after optimizing $\calL_{\text{HMM}}(y)$, $\Gamma$ is parameterized using $\eta \in \bbR^{N \times N}$ and the following link function:
-
-$$\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, \qquad \eta_{ii} = 0 \quad \forall i \in \{1, \ldots, N\}$$
-
+In order to ensure identifiability and right-stochasticity after optimizing $\calL_{\text{HMM}}(y)$, $\Gamma$ is often parameterized using $\eta \in \bbR^{N \times N}$ and the following link function:
+%
+$$\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, \qquad \eta_{ii} = 0 \quad \forall i \in \{1, \ldots, N\}.$$
+%
 This allows for unconstrained optimization over $\eta$ and removes the constraint that $\Gamma$ be right-stochastic.
@@
 
-A hierarchical hidden Markov model (or HHMM) is a variation of a hidden Markov model in which each hidden state of the original HMM $X_t$ emits both an observation $Y_t$ as well as another fine-scale hidden markov model of length $T^*_t$.
-This fine-scale HMM is comprised of a Markov chain $X^*_t = (X^*_{t,1}, \ldots, X^*_{t,T^*_t})$ and observations $Y^*_t = (Y^*_{t,1}, \ldots, Y^*_{t,T^*_t})$.
-As before, each fine-scale observation $Y^*_{t,t^*}$ depends only on the value of its corresponding hidden state, $X^*_{t,t^*}$.
-$X^*_{t,t^*}$ can take one of $N^*$ values and is characterized by an initial distribtuion $\delta^{*(X_t)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(X_t)} \in \bbR^{N^* \times N^*}$:
-
+A hierarchical hidden Markov model (or HHMM) is a variation on a hidden Markov model in which each hidden state of the original HMM $X_t$ emits both an observation $Y_t$ as well as another fine-scale hidden Markov model of length $T^*_t$.
+This fine-scale HMM is comprised of a fine-scale Markov chain $X^*_t = (X^*_{t,1}, \ldots, X^*_{t,T^*_t})$ and fine-scale observations $Y^*_t = (Y^*_{t,1}, \ldots, Y^*_{t,T^*_t})$.
+As before, each fine-scale observation $Y^*_{t,t^*}$ depends only on the value of its corresponding hidden state, $X^*_{t,t^*}$, which can take one of $N^*$ values.
+$X^*_{t}$ is characterized by an initial distribution $\delta^{*(X_t)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(X_t)} \in \bbR^{N^* \times N^*}$:
+%
 $$\delta^{*(x_t)}_i = Pr(X^*_{t,1} = i | X_t = x_t)$$
-
+%
 $$\Gamma^{*(x_t)}_{ij} = Pr(X^*_{t,t^*+1} = j | X^*_{t,t^*} = i, X_t = x_t) \qquad \forall t^* \in \{ 1, \ldots, T^*_t-1\}$$
-
-Finally, the fine-scale emission probabilities $p_{\theta^{*(X_t)}}(y^*_{s,t} | x^*_{s,t})$ are parameterized by $\theta^{*(X_t)}$.
-Note the parameters of the fine-scale hidden Markov model, $\Gamma^{*(X_t)}$, $\delta^{*(X_t)}$, and $\theta^{*(X_t)}$ all depend upon the hidden state of the \textit{crude-scale} hidden markov model.
+%
+The fine-scale emission probabilities $p_{\theta^{*(X_t)}}(y^*_{s,t} | x^*_{s,t})$ are parameterized by $\theta^{*(X_t)}$.
+Note the parameters of the fine-scale hidden Markov model ($\Gamma^{*(X_t)}$, $\delta^{*(X_t)}$, and $\theta^{*(X_t)}$) all depend upon the hidden state of the \textit{crude-scale} hidden Markov model.
 However, depending upon the discretion of the researcher, it is possible to force any of these parameters to be independent of the crude-scale hidden state $X_t$.
 A visualization of the full structure of the HHMM can be seen in figure (\ref{fig:HHMM}).
-
 Due to the nested structure of a hierarchical hidden Markov model, the likelihood of an HHMM is still easy to calculate using the forward algorithm:
@@
 %
-Note that this formualtion assumes that the crude-scale observations at a given time $Y_t$ and the fine-scale observation time series $Y_t^*$ are independent of one another when conditioned on $X_t$.
-
+Note that this formulation assumes that the crude-scale observations at a given time $Y_t$ and the fine-scale observation time series $Y_t^*$ are independent of one another when conditioned on $X_t$.
 For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, model selection and model checking, see Adam et al \cite{Adam:2019}.
@@
 
-\subsection{Conditionally autoregressive HMMs}
+\subsection{Conditionally auto-regressive HMMs}
 
-One of the key assumptions of both HMMs and HHMMs is \textit{conditional indpendence} between observations at both the crude and fine scale.
+One of the key assumptions of both HMMs and HHMMs is \textit{conditional Independence} between observations at both the crude and fine scale.
 Namely, given the state $X_t$ or $X^*_{t,t^*}$, $Y_t$ or $Y^*_{t,t^*}$ (respectively) is assumed to be independent from all other observations.
 Therefore, traditional HMMs and HHMMs can fail when the observation sequence $Y$ exhibits significant auto-correlation in time.
-Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swiming behavior of horn sharks off the coast of Southern California \cite{Adam:2019}.
+Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swimming behavior of horn sharks off the coast of Southern California \cite{Adam:2019}.
 
-One way to deal with autocorrelation in fine-scale behavioral processes is to use a state-switching continuous model such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behaviour state of the animal.
-Continuous time models are advantagous because of their flexibiliy: they can be built up from aribitrarily complex stochastic differential equations and they allow for uneven step lengths in the observations sequence $Y$.
-However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.
+One way to deal with auto-correlation in fine-scale behavioral processes is to use a state-switching continuous model such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal.
+Continuous time models are advantageous because of their flexibility: they can be built up from arbitrarily complex stochastic differential equations and they allow for uneven step lengths in the observations sequence $Y$.
+However, most continuous time models require MCMC algorithms to perform inference over and as a result are not easily incorporated into the HHMM structure.
 
-Another option is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}, in which autocorrelation is explicity modeled into the emission distributions of the HMM while maintaining the strucutre needed to run the forward algorithm for fast direct likelihood maximization.
+Another option is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}.
+The CarHMM explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm.
 In particular, if the emission distribution of observation $Y_t$ is parameterized by its mean and variance, i.e.
-$\theta_{x_t} = \{\mu_{x_t},\sigma^2_{x_t}\}$, the CarHMM introduces autocorrleation into the HMM by assuming that an observation $Y_t$ has mean $(1-\phi_{x_t}) \cdot \mu_{x_t} + \phi_{x_t} \cdot y_{t-1}$ rather than $\mu_{x_t}$.
-Note that the autocorrelation term $\phi_{x_t}$ depends upon the behavioral state of the animal.
-This model easily fits into the HHMM structure, but seems to lack the flexibility and natural interpretation of continuous-time models.
+$\theta_{x_t} = \{\mu_{x_t},\sigma^2_{x_t}\}$, the CarHMM assumes that an observation $Y_t$ has mean $(1-\phi_{x_t}) \cdot \mu_{x_t} + \phi_{x_t} \cdot y_{t-1}$ rather than $\mu_{x_t}$.
+This introduces one new parameter, $\phi_{x_t}$, which represents the auto-correlation of $y_t$ and depends upon the behavioral state of the animal.
+This model easily fits into the HHMM structure, but seems to lack the natural interpretation of continuous-time models.
 However, we prove in the following section that under certain conditions these two models are in fact equivalent.
@@
 %
-and the graphical model associated with the strcture of a CarHMM is shown in figure (\ref{fig:CarHMM}). Note that the first observation $y_1$ is assumed to be fixed as an initial value.
+and the first observation $y_1$ is assumed to be fixed as an initial value. The graphical model associated with the structure of a CarHMM is shown in figure (\ref{fig:CarHMM}).
 
@@
 	\includegraphics[width=3in]{../Plots/CarHMM.png}
-	\caption{Graphical representation of a traditional CarHMM. The additional arrows representing autocorrelation between observations are shown in red for emphasis.}
+	\caption{Graphical representation of a traditional CarHMM. The additional arrows representing auto-correlation between observations are shown in red for emphasis.}
 	\label{fig:CarHMM}
@@
 
-Once an HMM, HHMM, or CarHMM model is fit using the process described above, it is common to find the most likely sequence of hidden states $\hat X$ conditioned on the learned parameters by using a dynamic programing algorithm called the Viterbi algorithm \cite{Viterbi:1967}.
-In the case of HHMMs, this can be followed by running the Viterbi algorithm again on each sub-dive state to find the mostly likely sequence of fine-scale hidden states $\hat X_t^*$ conditioned on the learned parameters and the estimated crude-state value $\hat X_t$.
+Once an HMM, HHMM, or CarHMM model is fit using by maximizing their respective likelihoods with respect to their parameters, it is common to find the most likely sequence of hidden states $\hat X$.
+This is often done using a dynamic programming algorithm called the Viterbi algorithm \cite{Viterbi:1967}.
+In the case of HHMMs, after running the Viterbi algorithm on the coarse-scale process $X$, it can be run again on each fine-scale process $X^*_t$ to find the mostly likely sequence of fine-scale hidden states $\hat X_t^*$ conditioned on the estimated crude-state value $\hat X_t$.
 Note that while $\hat X$ is a maximum likelihood estimate of $X$, $\hat X^*_t$ is \textit{not} necessarily a maximum likelihood estimate of $\hat X^*_t$ because it is conditioned on the value of $\hat X_t$.
@@
 The forward-backward algorithm has the same time complexity as the forward algorithm and is also used to find the \textit{psuedoresiduals} of a given model, which is an important tool for model validation.
-In addition, For HHMMs in particular, the forward-backward algorithm can be used recursively to find the probability of the fine-level states $X^*_{s,t}$ exactly by marginalizing out $X_t$:
-
+In addition, for HHMMs the forward-backward algorithm can be used recursively to find the probability of the fine-level states $X^*_{s,t}$ exactly by marginalizing out $X_t$:
+%
 $$P(X^*_{s,t} = x^*_{s,t}) = \sum_{n=1}^N P(X_t = x_n)P(X^*_{s,t} = x^*_{s,t} | X_t = x_n)$$
-
-Where $P(X_t = x_n)$ can be found using the forward-backward algorithm on the crude-level markov chain and $P(X^*_{s,t} = x^*_{s,t} | X_t = x_n)$ can be found by running the forward-backward algorithm on the fine-level HMM for every possible value of $X_t$.
-
-%\vspace{1cm}
-
-%Model selection commonly involves the use of an information criterion such as the Akaike information criterion (AIC) or Bayesian information criterion (BIC).
-However, these metrics tend to favor complex models with more hidden states than can reasonably be interpreted in a biological setting [11].
-While future studies may use non-parametric clustering techniques to find the number of crude-level and dive-level hidden states, Leos-Barajas et al.
-use exploratory analysis to determine their parameterization, and this work does the same.
-%The timescale of the crude-scale Markov chain ideally should be short enough to ensure adequate resolution in the crude scale, but long enough so that the fine-scale Markov chains emitted by the crude-scale hidden Markov model are long enough to be informative.
-Because crude-scale behavioral changes occur on the time scale of approximately ten minutes, and there are a approximately thirteen dives per ten minutes of observation, this study uses a crude time scale of ten minutes.
-This differs from Leos-Barajas et al., who set the crude time scale to one hour.
-%The killer whale in this study appears to be alternating between periods of mostly pre- dominately short, shallow, smooth dives and periods interspersed with long, deep dives.
-As a result, we set the number of crude states as K = 2.
-The number of possible dive types was set as N = 3 to allow for additional flexibility beyond long, deep dives and short, shallow dives.
-Gamma distributions were used for the emission probability density function because of their flexibility and that the observations are all non-negative.
-All of these decision follow the lead of Leos-Barajas et al.
-%One additional assumption made both here and in the Barajas et al. paper is that
-%the distribution of observations given a dive type is constant across all K crude-scale dive-
-%level Markov chains. Mathematically this means that α(1) = ... = α(K) and β(1) = ... =
-%3.3 Parameter Estimation and Likelihood Maximization
-%Like the classical hidden Markov model, it is possible to use the forward algorithm to find the likelihood of an observed sequence of dives, Lw(y).
-The likelihood of the hierarchical model differs from the traditional HMM only in that the emission probability, p(y(t)|x(t)), is replaced by the likelihood of the fine-scale Markov chain at step m, Lx(y(m)|w(m)).
-Note that
-%9
-%(K)
-%maximization algorithm.
-%β
-%. This assumption simplifies the model and facilitates convergence in the likelihood
-
-%this likelihood depends upon the state of the crude-scale Markov chain, since the parameters
-%dictating the dynamics of X
-%K
-%(m)
-%(m)
-%depend upon the value of W Lw(y) = 􏰚p(y(1:M),W(M) = wi)
-%i=1 K
-%. Formally:
-%= 􏰚Lx(y(M)|W(M) = wi)p(y(1:M−1),W(M) = wi) i=1
-%KK
-%= 􏰚Lx(y(M)|W(M) = wi)􏰚P(wj → wi)p(y(1:M−1),w(M−1) = wj)
-%i=1 j=1
-%y(m) is a T(m)-by-R matrix of observations, Lx(y(m)|W(m) = wi) is the likelihood of one dive- level Markov chain given the crude behavioral state wi, and Lw(y) is the likelihood of the entire hierarchical model.
-Like in the traditional HMM model, this leads to the following recurrence relation:
-%K
-%p(y(1:m),W(m) =wi)=Lx(y(m)|W(m) =wi)􏰚P(wj →wi)p(y(1:m−1),W(m−1) =wj)
-%j=1
-%p(y(1), W (1) = wi) = δWwi ∗ Lx(y(1)|W (1) = wi)
-%where δW is the initial distribution of crude-scale behavioral states. Again, Leos-Barajas et al. express this algorithm concisely as a matrix product:
-%􏰑M􏰒 Lw(y) = δW PW (y(1)) 􏰛 ΓW PW (y(m)) 1
-%m=2
-%where Pw(y(m)) = diag(Lx(y(m)|W(m) = w1),...,Lx(y(m)|W(m) = wK)).
-We again copy Leos-Barajas et al.
-and map the entries of all probability transition matrices to the real line using the same transformation from earlier:
-%(m)
-%= wj|P(W
-%(t−1) exp(ηWij )
-%= wi) = 􏰗Nk=1 exp(ηWik), ηWii = 0 ∀i
-%γWij = P(W
-%γ(k) =P(X(m,t) =x |P(X(m,t−1) =x,W(m) =w )= Xij
-%exp(η(k) ) Xij j i k 􏰗N exp(η(k))
-%with respect to ΓW , Γ(1:K), α, and β. The Viterbi algorithm is then used to find the most X
-%likely sequence of values for the crude-scale behavioral state Markov chain.
-Once estimates for each internal state are found, we repeat the Viterbi algorithm on each dive-level Markov chain to estimate dive types.
+%
+where $P(X_t = x_n)$ can be found using the forward-backward algorithm on the crude-level Markov chain and $P(X^*_{s,t} = x^*_{s,t} | X_t = x_n)$ can be found by running the forward-backward algorithm on the fine-level HMM for every possible value of $X_t$.

### 2020-06-08 Report/Sections/carhmm_ou.tex
Updates from Overleaf

--- 2020-06-05 Report/Sections/carHMM.tex
@@
 Once the CarHMM is fit by maximizing the likelihood in equation (\ref{CarHMM_likelihood}), the emission distributions can be interpretted as the solution of a state-switching Ornstein-Uhlenbeck process similar to the one introduced by Michelot et al \cite{Michelot:2019}.
-In particular, the following two conditions must be met:
+In order for this to be the case, however, the following two conditions must be met:
 \begin{enumerate}
@@
 \end{enumerate}
-If both conditions (1) and (2) are met, then the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process.
-This gives new intpretation to the learned parameters of the CarHMM in the context of continuous-time model.
+If this is the case, then the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process.
+This gives new interpretation to the learned parameters of the CarHMM in the context of continuous-time model.
 
@@
 $x_t$ is described by an unobserved Markov process.
+Here, $t \in \mathbb{R}$ is a continuous time stamp rather than an index.
 If the behavioral state $x_t$ does not change between observations (i.e.
 $x_t$ follows a Markov chain), the solution to this equation is known to be the following \cite{Michelot:2019}:
+%
 \begin{align*}
@@
 \end{align*}
+%
 Now, suppose that $\Delta t$ is constant for all observations, as is the case for hidden Markov models. In addition, introduce the following transformations:
+%
 \begin{align}
-\mu_{x_t} = \gamma_{x_t}, \qquad \phi_{x_t} = e^{-\beta_{x_t}\Delta t}, \qquad \sigma^2_{x_t} = \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t})
+\mu_{x_t} = \gamma_{x_t}, \qquad \phi_{x_t} = e^{-\beta_{x_t}\Delta t}, \qquad \sigma^2_{x_t} = \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t}) \\
+\gamma_{x_t} = \mu_{x_t}, \qquad \beta_{x_t} = -\frac{\log(\phi_{x_t})}{\Delta t}, \qquad \omega_{x_t} = \sqrt{\frac{2\sigma^{2}_{x_t}\log(\phi_{x_t})}{\Delta t (\phi_{x_t}^2-1)}}
 \label{CarHMM_to_OU}
 \end{align}
-Then, we have the following:
+%
+Then, almost immediately we can see the following:
+%
 \begin{align*}
@@
 Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again.
-This allows for the pricipled construction of the continuous-time model to be combined with the computational convienence of the CarHMM.
+This allows for the principled construction of the continuous-time model to be combined with the computational convenience of the CarHMM.
 
@@
 
-Using this intuition, we generalize the CarHMM to data with unequal time steps: i.e.
-$\Delta t$ becomes a function of $t$.
+Using this intuition, we generalize the CarHMM to data with unequal time steps $\Delta t$.
 First, it is necessary to assume that the behavioral state $x_t$ only changes once between observations, which is a fair assumption if each time step $\Delta t$ is sufficiently small.
@@
 %
-Then the likelihood of the CarHMM with unequal time steps is just the following:
+Then the likelihood of the CarHMM with unequal time steps is the following:
 %
@@
 %
-now $\calL_{\text{CarHMM}}(y;\theta,\Lambda,\delta)$ is simply maximized with respect to $\theta = \{\beta,\omega,\gamma\}$, $\Lambda = \{p,\lambda\}$, and $\delta$ to find the maximum likelihood estimate of the CarHMM generative model with uneven time steps.
+now $\calL_{\text{CarHMM}}(y;\theta,\Lambda,\delta)$ is simply maximized with respect to $\theta = \{\beta,\omega,\gamma\}$, $\Lambda = \{p,\lambda\}$, and $\delta$ to find the maximum likelihood estimate of the CarHMM with uneven time steps.
 
 \iffalse
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Lawler et al are not the first to incorporate autocorrelation within a hidden Markov model in an ecological setting.
+Lawler et al are not the first to incorporate auto correlation within a hidden Markov model in an ecological setting.
 Woriskey et al \cite{Whoriskey:2016} developed a hidden Markov model called the HMMM, which is formulated as the following:

### 2020-06-08 Report/Sections/model.tex
update section names
Updates from Overleaf

--- 2020-06-05 Report/Sections/fourier.tex
@@
 
-One issue with HMMs is that they assume markovian dynamics conditioned on the hidden state, i.e.
-that any observation $Y^*_t$ depends only on the behaviorial state $X^*_t$ (here we focus on the fine-scale process).
-However, there are many animal movement processes which violate this markov propertey on very fine scales.
-For example swimming behaviour of marine mammals can be periodic since the animal repeatedly flukes to propel itself forward.
-Work has been done in the past to model non-markovian dynamics in the \textit{behavioural} process $X^*_t$ \cite{Langrock:2012}, but addressing non-markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area.
-With improvements in tagging technology allowing for data collection at very high frequencies, data exhibiting noisy and non-markovian fine scale behavior is likely to persist in future studies.
+One issue with the CarHMM is that it assumes Markovian dynamics conditioned on the hidden state, i.e.
+that any observation $Y^*_{t,t^*}$ depends only on the behavioral state $X^*_{t,t^*}$ and $Y^*_{t,t^*-1}$ (here we focus on the fine-scale process).
+However, there are many animal movement processes which violate this Markov property on very fine scales.
+For example, swimming behaviour of marine mammals can be periodic since the animal repeatedly flukes to propel itself forward.
+Work has been done in the past to model non-Markovian dynamics in the \textit{behavioural} process $X^*_t$ \cite{Langrock:2012}, but addressing non-Markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area (MAYBE?).
+With improvements in tagging technology allowing for data collection at very high frequencies, data exhibiting noisy and non-Markovian fine scale behavior is likely to persist.
 
@@
 %
-The STFT slides a moving window of length $w$ accross the time series $Y_t^*$ with a step size of $w$ and transforms the domain of each window from time to frequency.
-This allows the spectrum of $Y_t^*$ at time $t^*$ to be summerized by a $w$-dimensional vector of fourier coeficients.
+The STFT slides a moving window of length $w$ across the time series $Y_t^*$ with a step size of $w$ and transforms the domain of each window from time to frequency.
+This allows the spectrum of $Y_t^*$ at time $t^*$ to be summarized by a $w$-dimensional vector of Fourier coefficients.
 While other step sizes can be used for the sliding window, we select $w$ to avoid serial dependence between windows.
@@
 If $Y^*_t \in \mathbb{R}^{T^*_t}$, then $\hat{Y}_t^* \in \mathbb{C}^{\lfloor T^*_t / w \rfloor \times w}$.
-While this allows $Y^*_t$ to be represneted in a way that eliminates obvious periodic behavior, the dataset itself is still approximately as large as $Y^*_t$ itself.
+Although this allows $Y^*_t$ to be represented in a way that eliminates obvious periodic behavior, the data set itself is still approximately as large as $Y^*_t$ itself.
 To reduce the size of $\hat{Y}^*_t$, we propose taking summary statistics of each window as follows:
 %
-$$Z_{t,t^*}^{*(1)} = \mathcal{R}\left(\hat{Y}^{*(0)}_{t,t^*}\right) \qquad Z_{t,t^*}^{*(2)} = \frac{1}{w}\sum_{n=1}^{\tilde{w}}|\hat{Y}^{*(n)}_{t,t^*}|^2$$
+$$Z_{t,t^*}^{*(1)} = \mathcal{R}\left(\hat{Y}^{*(0)}_{t,t^*}\right) \qquad Z_{t,t^*}^{*(2)} = \frac{1}{w}\sum_{n=1}^{\tilde{f}}|\hat{Y}^{*(n)}_{t,t^*}|^2$$
 %
-$Z_{t,t^*}^{*(1)}$ is equal to the average value of $Y_{t,t^*:t^*+w-1}^*$ while $Z_{t,t^*}^{*(2)}$ is equal to the squared 2-norm of $Y_{t,t^*:t^*+w-1}^*$ that can be attributed to frequencies in the signal between $1$ and $\tilde{w}$ periods per window length.
-Both the window length $w$ and the max frequency $\tilde{w}$ are tuning parameters that should be tuned in a problem-specific way.
-$w$ should be long enough to capture the periodic behavior of the underlying process (at least as long as the length of a period), but short enough to avoid over-smoothing of the data and to maintain high resultion in the behavioral process $X^*$.
-$\tilde{w}$ should be selected such that the maximum frequency of $Y_t^*$ that makes biological sense is $\tilde{w}$ per window length.
-Note that these summary statistics are just one possible choice, and future studies can adjust the definitions as needed.
+$Z_{t,t^*}^{*(1)}$ is equal to the average value of $Y_{t,t^*:t^*+w-1}^*$, and $Z_{t,t^*}^{*(2)}$ is equal to the squared 2-norm of the component of $Y_{t,t^*:t^*+w-1}^*$ that can be attributed to frequencies in the signal between $1$ and $\tilde{f}$ periods per window length.
+Both the window length $w$ and the max frequency $\tilde{f}$ are tuning parameters that should be tuned in a problem-specific way.
+$w$ should be long enough to capture the periodic behavior of the underlying process (at least as long as the length of a period), but short enough to avoid over-smoothing of the data and to maintain high resolution in the behavioral process $X^*$.
+$\tilde{f}$ should be selected such that the maximum frequency of $Y_t^*$ that makes biological sense is $\tilde{f}$ per window length.
+Note that these summary statistics are just one possible choice of summary statistics within each window, and future studies can adjust these definitions as needed.
 A visualization of transforming a one-dimensional sequence $Y^*$ to $Z^*$ can be seen in figure (\ref{fig:fourier_example}).
@@
 	\centering
-	\includegraphics[width=7.5in]{../../Plots/fourier_transform.png}
+	\includegraphics[width=7.5in]{../Plots/fourier_transform.png}
 	\caption{Visualization of transforming $Y^*_t$ into $Z^*_t$ using a sliding window and fourier transform.}
@@
 
-Finally, note that it is possible to accomodate for unequal time steps within each window by using the \textbf{non-uniform discrete Fourier transform (NDFT)}.
+Finally, note that it is possible to accommodate for unequal time steps within each window by using the \textbf{non-uniform discrete Fourier transform (NDFT)}.
 We do not describe the details of this method in this work, but the generalization is straightforward.
@@
 Hierarchical hidden Markov models can be used to jointly model simultaneous coarse-scale and fine-scale processes taking place simultaneously.
-However, as mentioned before, the fine-scale process $Y^*$ can often exhibit autocorrelation and intricate structure.
-Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit autocorrelation, especially in the window averge, $Z_{t,t^*}^{*(1)}$.
+However, as mentioned before, the fine-scale process $Y^*$ can often exhibit auto-correlation and intricate structure.
+Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit auto-correlation, especially in the window average, $Z_{t,t^*}^{*(1)}$.
 Therefore, we replace the fine-scale HMM within the Hierarchical HMM with a CarHMM according to figure (\ref{fig:CarHHMM}).
@@
 	\centering
-	\includegraphics[width=7.5in]{../../Plots/CarHHMM.png}
-	\caption{Graphical representation of a CarHHMM. The additional arrows representing autocorrelation between observations are shown in red for emphasis.}
+	\includegraphics[width=7.5in]{../Plots/CarHHMM.png}
+	\caption{Graphical representation of a CarHHMM. The additional arrows representing auto-correlation between observations are shown in red for emphasis.}
 	\label{fig:CarHHMM}

### 2020-06-08 Report/Sections/simulation.tex
Add plots to repo for overleaf

@@
 Further, each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where $X^*_{t,t^*} \in \{1,2\}$ and $t^* \in \{1,101,201,\ldots,100*\lfloor Y_t/2 \rfloor + 1\}$.
-The parameters of the fine-scale behaviour Makov chain were set to be as follows:
+The parameters of the fine-scale behaviour Markov chain were set to be as follows:
 	%
@@
 	\begin{align*}
-	\hat{Y}^{*(0)}_{t,1} &\sim \mathcal{N} \left(\mu = 1, \sigma^2 = 0.01 \right) & \\
+    	\hat{Y}^{*(0)}_{t,1} &\sim \mathcal{N} \left(\mu = 1, \sigma^2 = 0.0025 \right) & \\
     	%
-	\hat{Y}^{*(0)}_{t,t^*} &\sim \mathcal{N} \left(\mu = 0.9 Y^{*(0)}_{t,t^*-100} + 0.1, \sigma^2 = 0.01 \right), & t^* \in \{101,201,\ldots, 100*\lfloor Y_t/2 \rfloor + 1\} \\
+    	\hat{Y}^{*(0)}_{t,t^*} &\sim \mathcal{N} \left(\mu = 0.9 Y^{*(0)}_{t,t^*-100} + 0.1, \sigma^2 = 0.0025 \right), & t^* \in \{101,201,\ldots, 100*\lfloor Y_t/2 \rfloor + 1\} \\
     	%
-	\hat{Y}^{*(n)}_{t,t^*} &= a_{t,t^*}^{(n)} i\sqrt{b^{(n)}_{t,t^*}}, & n \in \{1,\ldots,49\} \\\\
-	%
-	%
+    	\hat{Y}^{*(n)}_{t,t^*} &= a_{t,t^*}^{(n)} i\sqrt{b^{(n)}_{t,t^*}}, & n \in \{1,\ldots,49\} \\
+	\end{align*}
+	\begin{align*}
 	    a_{t,t^*}^{(n)} &\sim  \left\{\begin{array}{lr}
@@
     	%
-	(b^{(n)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim Gamma(1/n^2, 1) & \\
+    	(b^{(n)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim Gamma(1/n^2, 1) \\
     	%
@@
     	Gamma(1/n^2, 1) & \text{for } n \neq 2\\
-	Gamma(100,1) & \text{for } n = 2
-	\end{array}\right. & \\\\
-	%
-	%
+    	Gamma(20,1) & \text{for } n = 2
+    	\end{array}\right. 
+	\end{align*}
+
+    \begin{align*}
         \hat{Y}^{*(50)}_{t,t^*} &= 0 & \\
@@
 
-	Finally, $Y^*_{t,t^*:t^*+99}$ is set using the inverse discrete fourier transform of $\hat{Y}^*_{t,t^*}$:
-
+	\item Finally, $Y^*_{t,t^*:t^*+99}$ is set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
 	$$Y^*_{t,t^*:t^*+99} = IDFT\left(\hat{Y}^*_{t,t^*}\right)$$
-
-	This gives a strong periodic component to acceleration when the subdive state $X^*_{t,t^*}= 2$. There are several practical reasons behind this construction:
-
-	\begin{enumerate}
-		\item $\hat{Y}^*_{t,t^*}$ is anti-symetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse fourier transform is real-valued.
-		\item $\hat{Y}^{*(n)}_{t,t^*}$ decays like $1/n^2$ to facilitate continutity.
-	\end{enumerate}
 
@@
 
-Note that this process idoes not result in a continuous sequence $Y^*_t$ since the average values of consecutive 2-second segments jump.
-However, note that these average values are highly autocorrelated, so the jumps are not too severe.
+There are several practical reasons behind this construction.
+First, $\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued.
+$\hat{Y}^{*(n)}_{t,t^*}$ also decays like $1/n^2$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window.
+Note, however, that $Y^*_{t,t^*}$ is not continuous in $t^*$ \textit{between} windows, as can be seen in figure (\ref{fig:fourier_example}).
+However, note the jumps are not too severe since $\hat{Y}^{*(0)}_{t,t^*}$ and $\hat{Y}^{*(0)}_{t,t^*+100}$ are highly correlated.
 See figure (\ref{fig:sim_data}) for details.
-In addition, $Y^*_t$ has the following desirable properties:
 
-\begin{enumerate}
-	\item $Z^{*(1)}_{t,t^*+1} | Z^{*(1)}_{t,t^*} \sim \mathcal{N} \left(\mu = 0.9 Z^{*(1)}_{t,t^*} + 0.1, \sigma^2 = 0.01 \right)$
-	\item $$Z^{*(2)}_{t,t^*} \sim \left\{\begin{array}{lr} 
-	\text{Gamma}(\sum_{n=1}^{\tilde{w}} \frac{1}{n^2},1) & \text{for } X^*_{t,t^*} = 1 \\
-	\text{Gamma}(20 + \sum_{n=2}^{\tilde{w}} \frac{1}{n^2},1) & \text{for } X^*_{t,t^*} = 2
-	\end{array}\right. $$
-\end{enumerate}
+\subsection{Ground-Truth parameters}
 
-So we can directly compare the results of the CarHHMM with the ground truth.
+This generative model gives simulated data where the ground-truth of the CarHHMM is known. For the coarse scale, the ground truth is as follows:
+
+\begin{align*}
+    \delta &= \begin{pmatrix} 0.5 & 0.5 \end{pmatrix}\\
+	\Gamma &= \begin{pmatrix} 0.4 & 0.6 \\ 0.6 & 0.4 \end{pmatrix} \\
+	% 
+	Y_t|X_t = 1 &\sim Gamma\left(\mu = 15 s, \sigma = 5 s\right) \\
+	Y_t|X_t = 2 &\sim Gamma\left(\mu = 60 s, \sigma = 10 s\right) \\
+\end{align*}
+
+For the fine-scale, we have:
+
+\begin{align*}
+    \delta^{*(1)} &= \begin{pmatrix} 0.5 & 0.5 \end{pmatrix} & \delta^{*(2)} &= \begin{pmatrix} 0.5 & 0.5 \end{pmatrix} \\
+    %
+    \Gamma^{*(1)} &= \begin{pmatrix} 0.25 & 0.75 \\ 0.75 & 025 \end{pmatrix} & 	\Gamma^{*(2)} &= \begin{pmatrix} 0.75 & 0.25 \\ 0.25 & 0.75 \end{pmatrix} \\
+\end{align*}
+
+\begin{align*}
+    \left(Z^{*(1)}_{t,t^*+100} | Z^{*(1)}_{t,t^*} \right) &\sim \mathcal{N} \left(\phi^* * Z^{*(1)}_{t,t^*} + \phi^* * \mu^*, \sigma^{*2} \right), & \mu^* = 1, \enspace \phi^* = 0.9, \enspace \sigma^{*2} = 0.025 \\
+    %
+    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 1\right) &\sim Gamma\left(\alpha^{*(1)}, \beta^* \right), & \alpha^{*(1)} = \sum_{n=1}^{\tilde{f}} \frac{1}{n^2}, \enspace \beta^* = 1 \\
+    %
+    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 2\right) &\sim Gamma\left(\alpha^{*(2)}, \beta^* \right), & \alpha^{*(2)} = 20 + \sum_{n=2}^{\tilde{f}} \frac{1}{n^2}, \enspace \beta^* = 1 \\
+\end{align*}
+%
+The emission distribution of the 2-second average acceleration $Z_{t,t^*}^{*(1)}$ in particular can be reparameterized as an OU process:
+%
+$$dZ^{*(1)}_{t,t^*} = \beta^*(\gamma^* - Z^{*(1)}_{t,t^*})dt^* + \omega^* dW_{t,t^*}$$
+
+$$\gamma^* = \mu^* = 1, \qquad \beta^* = -\frac{\log(\phi^*)}{\Delta t^*} = 0.0229, \qquad \omega^{*2} = \frac{2\sigma^{*2}\log(\phi)}{\Delta t (\phi^{*2}-1)} = 0.006$$
+
+Note that the two fine-scale states $X^*_{t,t^*} = 1$ and $X^*_{t,t^*} = 2$ are identical except for the fact that the distribution of $Z^{*(2)}_{t,t^*}$ is shifted up in its shape parameter in fine-scale state 2.
+This corresponds to more periodic behavior in acceleration, but \textit{not} a higher mean acceleration.
 
@@
 	\centering
-	\includegraphics[width=5in]{../../Plots/sim_data.png}
+	\includegraphics[width=5in]{../Plots/sim_data.png}
 	\caption{Simulated Acceleration Data.}
@@
 \end{figure}
-I still need to put in the actually results from the simulation study- I had to rethink how to do the simulation in the first place since enforcing continuity was REALLY hard to do in a principled way.
+
+\subsection{Simulation Results}
+
+Four different models were fit to the generated data shown in figure (\ref{fig:sim_data}):
+
+\begin{enumerate}
+    \item A CarHMM with no hierarchical component (CarHMM) using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations
+    \item A Hierarchical hidden Markov model with no auto-correlation (HHMM) using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations.
+    \item A CarHHMM using $Z^{(1)}$, but not $Z^{(2)}$, as fine-scale observations.
+    \item A CarHHMM using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations
+\end{enumerate}
+
+Table (\ref{table:emis_dist}) gives the MLE estimates of the parameters for each model above.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2020-06-14 Report/Sections/results.tex
Many Updates
include CIs

@@
 
-\section{Case study- Fluking in Killer Whales}
+\section{Results}
 
@@
 
-To demonstrate this methodology as applied to real-world data, the CarHMM was used to analyze dive data from a killer whale off the coast of British Columbia, Canada.
+To demonstrate this methodology as applied to real-world data, the CarHHMM was used to analyze dive data from a killer whale off the coast of British Columbia, Canada.
 The data was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions.
@@
 In addition, the tagging technology dropped data between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed data within this time range were ignored as well.
-A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 3.1 meters in depth and lasts for at least 10 seconds.
-Finally, for computational purposes, only dives that reached a depth of at least 20 meters were included in the analysis, resulting a total of 7 dives.
+A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
+Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second.
 Data preprocessing was done in part with the \textit{divebomb} package in Python \cite{Nunes:2018}.
-
-The accelerometer recorded both the acceleration due to animal movement and due to gravity.
-The magnitude of acceleration due only to animal movement is the overall dynamic body acceleration (ODBA) of the killer whale and is known to be a good proxy of energy expenditure \cite{Gleiss:2011}.
-However, because accelerometer's orientation is unclear, the average acceleration for each component over the entire data collection window was subtracted out (to account for acceleration due to gravity) and the magnitude of the resulting vector was taken as an estimate for the ODBA.
-This ad-hoc method of removing the effect of gravity is highly prone to errors, especially because the killer whale changes its orientation often.
-Future work should use alternative methods to find the overall dynamic body acceleration of the whale.
-
-Velocity in the vertical direction was estimated using a central finite difference of the depth data:
-%
-$$v_t = \frac{d_{t+1} - d_{t-1}}{2\Delta t} = \frac{d_{t+1} - d_{t-1}}{1/25} $$
-%
-where $v_t$ is measured in meters per second.
-Due to the high frequency of data collection, small observation errors in depth readings resulted in relatively large errors in the velocity data.
-To deal with this, the velocity data was smoothed using a Gaussian kernel function with a width of 51 data points ($\approx$ 1 second) ranging from -3 to 3 standard deviations.
-In particular, if $\phi$ is the pdf of a standard normal distribution, then:
-%
-\begin{align*}
-v^{(smoothed)}_t &= \sum_{i=-25}^{25} \omega_i v_{t+i} \\
-\omega_i &= \frac{\phi\left(\frac{3i}{25}\right)}{\sum_{j=-25}^{25}\phi\left(\frac{3j}{25}\right)}
-\end{align*}
-%
-Figure (\ref{fig:smoothing}) shows the velocity data before and after smoothing.
-
-\begin{figure}[h!]
-	\centering
-	\includegraphics[height=7.5in]{../Plots/smoothed_v.png}
-	\caption{From top to bottom: dive depth, unsmoothed vertical velocity, smoothed vertical velocity, ODBA, and the sum of low-frequency Fourier coefficeints of the ODBA for a randomly selected dive of a killer whale.}
-	\label{fig:smoothing}
-\end{figure}
 
@@
 The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs straightforwardly.
-As a result, the data was split into two-second intervals (100 data points each), and each interval was summarized by the Fourier transform of the ODBA and the velocity at the \textit{end} of the time interval.
-Velocity was taken at the end of the time interval because the behavior of the whale during the time interval has no impact on the velocity of the whale at the beginning of the time interval.
-Because velocity is highly correlated with the average acceleration within an interval, the mean was subtracted out of each ODBA time series before the Fourier transform was taken.
-To reduce the dimension of the ODBA Fourier transform, Fourier coefficients corresponding to frequencies less than or equal to 5 hertz were summed and frequencies above 5 hertz were ignored.
-One nice property of this process is that it reduces the number of data points by a factor of 100, which drastically speeds up parameter estimation.
-The optimal length of the time interval over which to take the Fourier transform is a tuning parameter that is difficult to find.
-An ideal time interval should be long enough give useful Fourier coefficients, but short enough to preserve as much information about the vertical velocity of the animal as possible.
+We set $\tilde{f} = 5$ hertz and $w = 2$ seconds to calculate $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$.
+This reduces the dimension of each window from $50 1/s * 2 s = 100$ to 2.
 
@@
 
-In order to find the appropriate HMM to model whale behavior, a lag plot was made for both velocity and the ODBA fourier sum.
+In order to find if the CarHHMM was appropriate for this data, a lag plot was made for both $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$.
 The results of doing so are shown in figure (\ref{fig:lag}).
-Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that the velocity exhibits a large degree of autocorrelation.
-While the ODBA Fourier sum also exhibits some autocorrelation, the relationship is less strong, so autocorrelation was not incorporated in the ODBA Fourier sums emission distribution.
+Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that $Z^{*(1)}_{t,t^*}$ exhibits a large degree of autocorrelation.
+While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, so autocorrelation was not incorporated in the emission distribution of $Z^{*(2)}_{t,t^*}$ for computational purposes.
 
@@
 \end{figure}
-
-\subsection{CarHMM adapted to dive data}
-
-Let $v_t$ represent the velocity in the $z-$direction (derived from depth data) at the end of time interval $t$ and $\tilde a_t$ represent the ODBA Fourier sum of interval $t$.
-The CarHMM was adapted to the dive data using the following model:
-
-\begin{align*}
-	v_{t} &\sim \mathcal{N}\left(\phi_{b_t} v_{t-1} + (1-\phi_{b_t}) \mu_{RL,b_t}, \sigma_{b_t}^2\right) \\
-	%
-	\tilde a_t &\sim \Gamma\left(\tilde \mu_{b_t}, \tilde \sigma^2_{b_t}\right)
-\end{align*}
-
-Note that the arguments in the gamma distribution above correspond to the mean and variance of the distribution, \textit{not} the standard scale and shape parameters.
-In total, 3 behavioral states were chosen for exploratory analysis as an attempt to capture behavioral states with very negative velocities, very positive velocities, and velocities close to zero.
-This results in a total of $3^2 + 4*3 = 21$ parameters, all of which were estimated using direct likelihood maximization.
 

### 2020-06-14 Code/HHMM.py

@@ -32,6 +32,7 @@ class HHMM:
         self.theta = []
         self.eta = []
         self.ptm = []
+        self.data = data
 
         eta_crude = -1.0 + 0.1*np.random.normal(size=(self.pars.K[0],self.pars.K[0]))
         ptm_crude = np.exp(eta_crude)
@@ -53,6 +54,9 @@ class HHMM:
         self.ptm.append(ptm_fine)
 
         self.initialize_theta(data)
+
+        self.true_theta = None
+        self.true_eta = None
 
         return
 
@@ -148,6 +152,13 @@ class HHMM:
                     theta[1][k0][feature]['corr'] = 0.1*np.random.random(size=K) - 1.0
 
         self.theta = theta
+
+        return
+
+
+    def initalize_theta_sim(self):
+
+
 
         return
 

### 2020-06-16 Report/main.tex
update section names
Update Report

@@
-\documentclass[10pt]{article}
+% Initial and revised submissions should be 12 point; this will be removed in the final version.
+\documentclass[12pt]{TD-CJS}
+
+% Initial and revised submissions should also be double spaced.  This command will be removed in the final version.
+\renewcommand{\baselinestretch}{2}
+
+\usepackage{latexsym}
+\usepackage{amsmath}
+\usepackage{amsfonts}
+\usepackage{amssymb}
+\usepackage{psfrag}
+\usepackage{graphicx}
+
 \input{header}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{document}
+\firstpage{1}
+\lastpage{25}
+\jvol{xx}
+\issue{yy}
+\jyear{2020}
+\jid{CJS}
+\aid{???}
+% The running head contains the author names
+\rhauthor{BlindedA and BlindedB}
+\copyrightline{Statistical Society of Canada}
+\Frenchcopyrightline{Soci\'et\'e statistique du Canada}
+% History: received and accepted dates
+\received{\rec{9}{July}{2009}}
+\accepted{\acc{8}{July}{2010}}
 
-\title{Inferring Fine-Scale Animal Behaviors using Hidden Markov Models}
-\author{Evan Sidrow}
-\date{\today}
+% User-defined commands go here
+\renewcommand{\eqref}[1]{(\ref{#1})}
+\newcommand{\mb}[1]{\mathbf{#1}}
+\newcommand{\mbb}[1]{\mathbb{#1}}
+\newcommand{\mt}[1]{\mathrm{#1}}
+\newcommand{\rv}{random variable}
 
-\begin{document}
+% Title, authors, affiliations
+\title[]{Title on one or two lines without capitals, except after a colon}%\query{Q1}
+\author{BlindedA\authorref{1}\thanksref{*}}
+\author{BlindedB\authorref{2}}
+\affiliation[1]{Author affiliations will go here in the accepted manuscript, 
+but do NOT include them in your initial submission because it must be anonymous.}
+\affiliation[2]{Second Affiliation}
 
-\maketitle
-
-\abstract
+% Abstract, keywords, and classification codes
+\startabstract{%
+\keywords{
+\KWDtitle{Key words and phrases}
+Association parameters\sep clustered data\sep mean parameters\sep missing data\sep pairwise
+likelihood\sep repeated measurements.
+% MSC 2010 subject classification codes
+\KWDtitle{MSC 2010}Primary 62???\sep secondary 62???}
+\begin{abstract}
+\abstractsection{}
+\abstractsection{Abstract}
+Insert your abstract here; it should
+typically be up to ten lines long. Avoid symbols as much as possible. Formulas
+are strongly discouraged, and citations should be avoided. 
+The title and the abstract should be concise and descriptive;
+list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
+http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.
 
@@
 
+\abscopyright
+\fabstractsection{}
+\fabstractsection{R\'{e}sum\'{e}}
+Ins\'{e}rer votre r\'{e}sum\'{e}
+ici. We will supply a French abstract for those authors who
+can't prepare it themselves.\Frenchabscopyright
+\end{abstract}}
+\makechaptertitle
+
+% Email address for corresponding author
+\correspondingauthor[*]{\\\email{Insert your email address here only after your paper has been accepted}}
+
 %\input Sections/intro.tex
-\input Sections/HHMM.tex
-\input Sections/carHMM.tex
-\input Sections/fourier.tex
+\input Sections/background.tex
+\input Sections/carhmm_ou.tex
+\input Sections/model.tex
 \input Sections/simulation.tex

### 2020-06-16 Code/HHMM.py
Make code compute canada-ready

@@ -17,6 +17,8 @@ from scipy.optimize import LinearConstraint
 from scipy.signal import convolve
 from scipy.interpolate import interp1d
 
+from copy import deepcopy
+
 import divebomb
 
 import time
@@ -427,148 +429,165 @@ class HHMM:
         return log_alpha, log_beta, gamma, xi
 
 
-    def train_DM(self,data,eps=0.001,max_iters=None):
+    def train_DM(self,data,max_iters=10,tol=0.01,eps=10e-6):
 
-        # define loss function
-        def loss_fn(x):
-            ind = 0
+        options = {'maxiter':10,'disp':False}
+        prev_l = self.likelihood(data)
 
-            # update crude eta
-            self.eta[0] = x[0:self.pars.K[0]**2].reshape((self.pars.K[0],
-                                                          self.pars.K[0]))
-            ind += self.pars.K[0]**2
+        for _ in range(max_iters):
 
-            # update crude theta
+            print(prev_l)
+
+            ### start with crude eta ###
+            def loss_fn(x):
+                ind = 0
+
+                # update crude eta
+                self.eta[0] = x[0:self.pars.K[0]**2].reshape((self.pars.K[0],
+                                                              self.pars.K[0]))
+                l = -self.likelihood(data)
+                return l
+
+            # define inital value
+            x0 = []
+            x0.extend(self.eta[0].flatten())
+
+            # optimize
+            res = minimize(loss_fn, x0, method='Nelder-Mead',options=options)
+
+            # update final values
+            x = np.copy(res['x'])
+            self.eta[0] = x.reshape((self.pars.K[0],self.pars.K[0]))
+
+            #### then do crude theta ###
             for k0 in range(self.pars.K[0]):
                 for feature in self.pars.features[0]:
-                    if feature in ['Ay','Az']:
-                        self.theta[0][feature]['mu'][k0] = x[ind]
-                        self.theta[0][feature]['sig'][k0] = self.theta[0]['Ax']['sig'][k0]
-                        self.theta[0][feature]['corr'][k0] = self.theta[0]['Ax']['corr'][k0]
-                        ind += 1
-                    else:
-                        self.theta[0][feature]['mu'][k0] = x[ind]
-                        self.theta[0][feature]['sig'][k0] = max(x[ind+1],eps)
-                        self.theta[0][feature]['corr'][k0] = x[ind+2]
-                        ind += 3
 
-            # update fine eta
-            for k0 in range(self.pars.K[0]):
[…]

### 2020-06-16 Code/Run_Sim.py

@@ -1,0 +1,472 @@
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.cm as cm
+import random
+import sys
+
+from copy import deepcopy
+
+from scipy.stats import gamma
+from scipy.stats import norm
+from scipy.stats import multivariate_normal
+from scipy.stats import gaussian_kde
+from scipy.special import expit
+from scipy.special import logit
+from scipy.special import logsumexp
+from scipy.optimize import minimize
+from scipy.optimize import LinearConstraint
+from scipy.signal import convolve
+from scipy.interpolate import interp1d
+
+import divebomb
+
+import time
+import pickle
+
+import importlib
+
+import Preprocessor
+import Parameters
+import HHMM
+import Visualisor
+
+rand_seed = int(sys.argv[1])
+
+random.seed(rand_seed)
+np.random.seed(rand_seed)
+
+
+### initialize parameters ####
+
+ndives = 100
+ndatasets = 1
+
+# dive duration parameters
+dd_mu = np.array([20.0,80.0])
+dd_sig = np.array([5.0,50.0])
+
+dd_shape = np.square(dd_mu)/np.square(dd_sig)
+dd_scale = np.square(dd_sig)/np.array(dd_mu)
+
+# FoVeDBA parameters (this is for the SQUARE of the FoVeDBA)
+FoVeDBA_sin_shape = np.array([[(5.0/n) for n in range(1,52)],[(5.0/n) for n in range(1,52)]])
+FoVeDBA_sin_scale = np.ones((2,51))
+FoVeDBA_sin_shape[1,1] = 250
+FoVeDBA_sin_shape[1,2] = 50
+
+# average acceleration parameters
+acc_mu = np.array([0.0,0.0])
+acc_sig = np.array([0.05,0.1])
+
+# number of states for each substate
+K0 = 2
+K1 = 2
+
+# randomly initialize a probablity transition matrix
+ptm_crude = np.array([[0.5,0.5],
+                      [0.5,0.5]])
+
+ptm_fine = [np.array([[0.5,0.5],
+                      [0.1,0.9]]),
+            np.array([[0.8,0.2],
+                      [0.3,0.7]])]
+
+# randomly intialize a correlation within states
+corr_crude = [0.0,0.0]
+corr_fine = [0.99,0.95]
+
+# train new models?
+train_new = True
+
+# initialize the initial states
[…]

### 2020-06-16 prep_python.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+#SBATCH --account=def-nheckman
+cd projects/nheckman-def/evsi8432
+module load python/3.7
+source ~/Whale_Paper_1/bin/activate

### 2020-06-16 run_all_sims.sh

@@ -1,0 +1,15 @@
+#!/bin/bash
+#SBATCH --account=def-nheckman
+
+#SBATCH --mail-user=evan.sidrow@gmail.com
+#SBATCH --mail-type=BEGIN
+#SBATCH --mail-type=END
+#SBATCH --mail-type=FAIL
+#SBATCH --mail-type=REQUEUE
+#SBATCH --mail-type=ALL
+
+#SBATCH --time=24:00:00
+#SBATCH --mem-per-cpu=4G
+#SBATCH --array=1-100
+
+python Run_Sim.py $SLURM_ARRAY_TASK_ID

### 2020-06-16 Report/CJS_Sample.tex

% Initial and revised submissions should be 12 point; this will be removed in the final version.
\documentclass[12pt]{TD-CJS}

% Initial and revised submissions should also be double spaced.  This command will be removed in the final version.
\renewcommand{\baselinestretch}{2}

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{psfrag}
\usepackage{graphicx}

\begin{document}
\firstpage{1}
\lastpage{25}
\jvol{xx}
\issue{yy}
\jyear{20??}
\jid{CJS}
\aid{???}
% The running head contains the author names
\rhauthor{BlindedA and BlindedB}
\copyrightline{Statistical Society of Canada}
\Frenchcopyrightline{Soci\'et\'e statistique du Canada}
% History: received and accepted dates
\received{\rec{9}{July}{2009}}
\accepted{\acc{8}{July}{2010}}

% User-defined commands go here
\renewcommand{\eqref}[1]{(\ref{#1})}
\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mt}[1]{\mathrm{#1}}
\newcommand{\rv}{random variable}

% Title, authors, affiliations
\title[]{Title on one or two lines without capitals, except after a colon}%\query{Q1}
\author{BlindedA\authorref{1}\thanksref{*}}
\author{BlindedB\authorref{2}}
\affiliation[1]{Author affiliations will go here in the accepted manuscript, 
but do NOT include them in your initial submission because it must be anonymous.}
\affiliation[2]{Second Affiliation}

% Abstract, keywords, and classification codes
\startabstract{%
\keywords{
\KWDtitle{Key words and phrases}
Association parameters\sep clustered data\sep mean parameters\sep missing data\sep pairwise
likelihood\sep repeated measurements.
% MSC 2010 subject classification codes
\KWDtitle{MSC 2010}Primary 62???\sep secondary 62???}
\begin{abstract}
\abstractsection{}
\abstractsection{Abstract}
Insert your abstract here; it should
typically be up to ten lines long. Avoid symbols as much as possible. Formulas
are strongly discouraged, and citations should be avoided. 
The title and the abstract should be concise and descriptive;
list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.
\abscopyright
\fabstractsection{}
\fabstractsection{R\'{e}sum\'{e}}
Ins\'{e}rer votre r\'{e}sum\'{e}
ici. We will supply a French abstract for those authors who
can't prepare it themselves.\Frenchabscopyright
\end{abstract}}
\makechaptertitle

% Email address for corresponding author
\correspondingauthor[*]{\\\email{Insert your email address here only after your paper has been accepted}}

\section{INTRODUCTION}
Your text starts here. For English spelling,
we follow the style of the Canadian Oxford Dictionary (Barber, 2004).  
If the dictionary lists more than one acceptable spelling, choose the main entry.

Manuscripts may be submitted in either English or French.  We do accept submissions in Microsoft Word.  We do not provide a Word template, and we ask you to follow the general instructions in this document.

\subsection{Subsections start this way}
Refer to papers by authors (date) throughout. For example,
you might quote Author1 \& Author2 (1986a,~b) or Author (1987, 1992) or 
Author1, Author2, \& Author3 (2011). If there
are four authors or more, refer to (for example) Author1 et al.\ (1990).

When references are in parentheses, separate them by a semicolon and do not
put brackets around the year (e.g., Author1 \& Author2, 1994; Author3 \&
Author4, 1999).

\section{MATHEMATICAL TYPESETTING}
Symbols should not be used at the start of a sentence, and footnotes are not allowed. Try to avoid double subscripts, and never use triple subscripts. Unless central or essential to the flow of the discussion, mathematical arguments should be deferred to the Appendix.
Note that equations should be numbered consecutively, i.e., (1), (2), etc. Number
\textit{only} those equations that are referred to in the text.  Punctuation should be given after equations. Equations should be cited as, for example, Equation (6) or Equations (6)--(8). Within parentheses use an abbreviation: (Eq. 7).

Likewise, number consecutively your definitions, lemmas, propositions,
theorems, corollaries and the like. For example:

\begin{theorem}{Theorem 1.}{}%
Here is the statement of our theorem.
\end{theorem}
\begin{proof}{Proof}{}%
We prove our theorem using Equation (\ref{Einstn}) below:
\begin{equation}
e = mc^2.
\label{Einstn}
\end{equation}
\end{proof}

\section{FIGURES AND TABLES}
When you submit the final version of your manuscript in
\LaTeX \, form, please include postscript files (.ps or .eps) for the figures,
labelling them fig1.ps, fig2.ps, etc. When referring to a figure, spell out the word (e.g., Figure 1) whether or not it is in parentheses.

% Instructions for a figure
%\begin{figure}
%\centerline{\epsfbox{gr1.eps}}
%\caption{This is a full width figure}
%\end{figure}

See Table 1 for an example table layout; note that we do not use
vertical lines between columns.  The data in this table are from Genest (1999).

\begin{table}
% Next comes the table caption
\tbl{Top 10 countries for gross national publication (\textsc{gnp}) of
research in statistics. The ranks are based on variable \textsc{pag}$^{\star}%
$.}{%
\begin{tabular*}{30pc}{@{\hskip5pt}@{\extracolsep{\fill}}r@{}c@{}r@{}r@{}r@{}r@{}r@{}r@{}@{\hskip5pt}}
\toprule
Rank & Country & \textsc{pag}$^{\star}$ & \textsc{pag} & \textsc{art}$^{\star
}$ & \textsc{art} & $\frac{\mbox{{\sc pag}}} {\mbox{{\sc art}}}$ &
$\frac{\mbox{{\sc aut}}}{\mbox{{\sc art}}}$\\
\colrule
1 & \textsc{usa} & 109338 & 60369 & 7240 & 4061 & 14.9 & 1.83\\
2 & United Kingdom\tref{a} & 12597 & 7504 & 884 & 538 & 14.1 & 1.81\\
3 & Canada & 12407 & 6837 & 909 & 516 & 13.6 & 1.89\\
4 & Australia & 7872 & 4261 & 578 & 323 & 13.5 & 1.95\\
5 & Germany & 6782 & 4500 & 456 & 306 & 14.9 & 1.63\\
6 & France & 3647 & 1843 & 261 & 129 & 14.5 & 2.18\\
7 & Japan & 2865 & 1880 & 241 & 163 & 11.6 & 1.60\\
8 & Netherlands & 2864 & 1702 & 191 & 116 & 15.1 & 1.80\\
9 & India & 2559 & 1395 & 275 & 151 & 9.5 & 1.91\\
10 & Israel & 2097 & 1160 & 148 & 83 & 14.5 & 1.99\\
\botrule
\end{tabular*}}
\begin{tabnote}
\tblfno{a}This is a footnote to the table.
\end{tabnote}
\end{table}

\begin{ack}{ACKNOWLEDGEMENTS}
Place all acknowledgements here.  In your initial and revised submission, ensure that any acknowledgements are anonymous; include the full acknowledgements only after your paper has been accepted. Granting agencies should not be abbreviated, and do not include grant numbers.  We are grateful for your assistance with our publication process.\\
\\
A note on references: they should be listed alphabetically. Initials should be separated by blank characters and
journal names should be {\it in italics}. Don't use boldface for volume
number and separate pages by two dashes.
Abbreviations should not be used for authors' or journal's names, nor for the titles of the articles.
\end{ack}

\begin{thebibliography}{}
% For English spelling, we follow the style of this dictionary
\bibitem[]{bib1}
Barber, K. (Ed.) (2004). {\it The Oxford Canadian Dictionary}, 2nd ed., 
Oxford University Press, Toronto.

% Single-author journal article
\bibitem[]{bib2}
Cleveland, W. (1979). Robust locally-weighted regression and smoothing scatterplots. 
{\it Journal of the American Statistical Association}, 74, 829-836.

% Genest provided the data for our sample table
\bibitem[]{bib3}
Genest, C. (1999). Probability and statistics: A tale of two worlds? 
{\it The Canadian Journal of Statistics}, 27(2), 421--444.

% Multiple-author journal article.  All the author names are separated by commas, and 
% note the uppercase letter after the colon in the title.
\bibitem[]{bib4}
Johnes, J., Taylor, J., \& Francis, B. (1993). The research
performance of \textsc{uk} universities: A statistical analysis of the results
of the 1989 research selectivity exercise. {\it Journal of the Royal
Statistical Society Series A}, 156, 271--286.

% Book in French.  Be careful with accents.
\bibitem[]{bib5}
Kidron, M. \& Segal, R. (1992). {\it Atlas du Nouvel \'Etat
du Monde}. \'Editions Autrement, Paris.

% Book, two authors, second edition
\bibitem[]{bib6}
Little, R.~J.~A. \& Rubin, D.~B. (2002). {\it Statistical Analysis with Missing Data}, 2nd ed., 
John Wiley \& Sons, New York.

% Unpublished paper with website link
\bibitem[]{bib7}
Molina, I. \& Rao, J.~N.~K. (2009). Small area estimation of poverty indicators. Working Paper 09-15, Statistics and Econometric Series 05, Universidad Carlos III de Madrid, \\http://hdl.handle.net/10016/5644.

% Edited book
\bibitem[]{bib8}
Roberts, G., Ren, Q., \& Rao, J.~N.~K. (2009). 
Using marginal mean models for data from longitudinal surveys
with a complex design: Some advances in methods. In Lynn, R. (Ed.) {\it Methodology of Longitudinal Surveys}, 
John Wiley \& Sons, New York, 351-366.
\end{thebibliography}

\begin{appendix}
There should be just one appendix, for proofs and longer
mathematical arguments. These proofs are in the following form:

\begin{proof}{Proof of Theorem 1}{}%
We now prove the two parts of Theorem 1.
\end{proof}
\end{appendix}

\CJShistory
\end{document}

### 2020-06-16 Report/Sections/background.tex

@@
 \subsection{Hidden Markov models}
+\label{subsection:HMMs}
 
-\textit{Hidden Markov models} (or HMMs) are comprised of an unobserved Markov chain $X = (X_1, \ldots, X_T)$ and a sequence of (possibly high-dimensional) observations $Y = (Y_1, \ldots, Y_T)$, each of length $T$.
-In the field of animal movement, the unobserved chain $X$ usually represents the latent behaviour of an animal (e.g.
-foraging, resting, migrating, etc.), while the observations $Y$ are often a series of step lengths and turning angles for land animals and either depth or accelerometer data (or both) for marine animals.
-Each random variable in the unobserved chain $X_t$ can take one of $N$ possible values, and $X$ has corresponding probability transition matrix $\Gamma \in \bbR^{N \times N}$ and initial distribution $\delta \in \bbR^N$:
+\textit{Hidden Markov models} (or HMMs) are comprised of a sequence of unobserved states, $X_t$ $(t = 1, \ldots, T)$ which form a Markov chain and a sequence of (possibly high-dimensional) observations $Y_t$ $(t = 1, \ldots, T)$ which are often referred to as ``emissions".
+The index $t$ often represents equi-spaced jumps in time.
+In the field of animal movement, the unobserved chain usually represents the latent behaviour of an animal (e.g.
+foraging, resting, migrating, etc.), while the observations are often a series of step lengths and turning angles for land animals and either depth or accelerometer data (or both) for marine animals.
+Each random variable in the unobserved chain $X_t$ can take one of $N$ possible values, and $X$ has corresponding initial distribution $\delta \in \bbR^N$ and probability transition matrix $\Gamma \in \bbR^{N \times N}$:
 
-$$\delta_i = Pr(X_1 = i)$$
+$$\delta_i = Pr(X_1 = i); \quad i = 1,\ldots,N$$
 
-$$ \Gamma_{ij} = Pr(X_{t+1} = j | X_t = i) \qquad \forall t \in \{ 1, \ldots, T-1\} $$
+$$\Gamma_{ij} = Pr(X_{t+1} = j | X_t = i); \quad t = 1, \ldots, T-1; \enspace i = 1,\ldots,N; \enspace j=1,\ldots,N $$
 
-Further, each random variable $X_t$ emits an observation $Y_t$ whose distribution depends only on the value of $X_t$ and none of the preceding observations or behavioral states: $p_{\theta}(y_t|x_t, x_{t-1}, \ldots , x_1, y_{t-1}, \ldots , y_1) = p_{\theta}(y_t|x_t)$.
-Note that the emission distribution depends upon parameters $\theta$.
-A visualization of this dependence structure can be seen in figure (\ref{fig:HMM}).
-The probability transition matrix $\Gamma$ and the parameters of the emission distributions $\theta$ can be estimated by maximizing the likelihood of the observed data $y$, $\calL_{\text{HMM}}(y;\theta,\Gamma,\delta)$, with respect to $\Gamma$, $\delta$, and $\theta$.
-In addition, $\calL_{\text{HMM}}(y;\theta,\Gamma,\delta)$ can be calculated using the \textit{forward algorithm} as follows:
+$\delta$ is often assumed to be the stationary distribution of the $\Gamma$- an assumption that we make in this work.
+Further, the distribution of each observation $Y_t$ depends only on the value of $X_t$ and none of the preceding observations or behavioral states, i.e.
+$p(y_t|x_1,\ldots, x_N, y_1,\ldots,y_{t-1},y_{t+1},\ldots,y_N) = p(y_t|x_t)$.
+In particular, we denote the distribution of $Y_t$ given $X_t = s$ as $f_s(y_t)$.
+$f_1,\ldots,f_N$ in turn depend upon a vector of parameters $\Theta = (\theta_1,\ldots,\theta_N)$ and hereafter are referred to as ``emission distributions".
+A visualization of the dependence structure can be seen in figure (\ref{fig:HMM}).
+In order to estimate the parameters of an HMM, suppose data $y = (y_1,\ldots,y_T)$ is observed as a realization of an unknown HMM model.
+The probability transition matrix $\Gamma$, and parameters of the emission distributions $\Theta$ can be estimated by maximizing the likelihood of $y$, which we denote as $\calL_{\text{HMM}}(y;\Theta,\Gamma)$, with respect to $\Gamma$, $\delta$, and $\Theta$.
+In addition, $\calL_{\text{HMM}}(y;\theta,\Gamma,\delta)$ can be calculated using the \textit{forward algorithm} \cite{Zucchini:2016} as follows:
 %
-$$\calL_{\text{HMM}}(y;\theta,\Gamma,\delta) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}$$
+$$\calL_{\text{HMM}}(y;\theta,\Gamma) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}$$
 %
-where:
+where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones, and:
 %
-$$P(y_t;\theta) = \text{diag}(p_{\theta}(y_t|X_t = x_1), . . . , p_{\theta}(y_t|X_t = x_N ))$$
+$$P(y_t;\theta) = \text{diag}(f_1(y_t),\ldots , f_N(y_t)).$$
 %
-and $\mathbf{1}$ is an $N$-dimensional column vector of ones.
-In order to ensure identifiability and right-stochasticity after optimizing $\calL_{\text{HMM}}(y)$, $\Gamma$ is often parameterized using $\eta \in \bbR^{N \times N}$ and the following link function:
+In order to ensure remove constraints when maximizing $\calL_{\text{HMM}}(y)$, $\Gamma$ is often parameterized using $\eta \in \bbR^{N \times N}$ and the following link function:
 %
@@
 %
-This allows for unconstrained optimization over $\eta$ and removes the constraint that $\Gamma$ be right-stochastic.
-$\calL_{\text{HMM}}(y;\theta,\Gamma,\delta)$ can be maximized using any numerical optimizer.
+$\eta_{ii}$ is set to zero for identifiability. $\calL_{\text{HMM}}(y;\theta,\Gamma)$ can be maximized using any numerical optimizer.
 
@@
 	\centering
-	\includegraphics[width=3in]{../Plots/HMM.png}
+	\includegraphics[width=5in]{../Plots/HMM.png}
 	\caption{Graphical representation of a traditional HMM.}
@@
 
-A hierarchical hidden Markov model (or HHMM) is a variation on a hidden Markov model in which each hidden state of the original HMM $X_t$ emits both an observation $Y_t$ as well as another fine-scale hidden Markov model of length $T^*_t$.
-This fine-scale HMM is comprised of a fine-scale Markov chain $X^*_t = (X^*_{t,1}, \ldots, X^*_{t,T^*_t})$ and fine-scale observations $Y^*_t = (Y^*_{t,1}, \ldots, Y^*_{t,T^*_t})$.
-As before, each fine-scale observation $Y^*_{t,t^*}$ depends only on the value of its corresponding hidden state, $X^*_{t,t^*}$, which can take one of $N^*$ values.
-$X^*_{t}$ is characterized by an initial distribution $\delta^{*(X_t)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(X_t)} \in \bbR^{N^* \times N^*}$:
+A hierarchical hidden Markov model (or HHMM) is a variation of a hidden Markov model in which $X_t$ satisfies the conditions of section \ref{subsection:HMMs}, but in addition it produces yet another fine-scale hidden Markov model of length $T_t^*$.
+This fine-scale hidden Markov model can be either instead of \cite{Barajas:2017} or in addition to \cite{Adam:2019} a coarse-scale observation $Y_t$ which must satisfy the conditions of section \ref{subsection:HMMs}.
+The fine-scale HMM is comprised of a fine-scale Markov chain $X^*_t = (X^*_{t,1}, \ldots, X^*_{t,T^*_t})$ and fine-scale observations $Y^*_t = (Y^*_{t,1}, \ldots, Y^*_{t,T^*_t})$, each of length $T^*_t$.
+$X^*_t$ can take one of $N^*$ values.
+Note that while $N^*$ can be set to depend upon $X_t$, we assume here that it does not for simplicity in both the model and the notation.
+$X^*_t$ is characterized by an initial distribution $\delta^{*(X_t)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(X_t)} \in \bbR^{N^* \times N^*}$:
 %
-$$\delta^{*(x_t)}_i = Pr(X^*_{t,1} = i | X_t = x_t)$$
+$$\delta^{*(s)}_i = Pr(X^*_{t,1} = i | X_t = s); \quad i = 1,\ldots,N^*$$
 %
-$$\Gamma^{*(x_t)}_{ij} = Pr(X^*_{t,t^*+1} = j | X^*_{t,t^*} = i, X_t = x_t) \qquad \forall t^* \in \{ 1, \ldots, T^*_t-1\}$$
+$$\Gamma^{*(s)}_{ij} = Pr(X^*_{t,t^*+1} = j | X^*_{t,t^*} = i, X_t = s); \quad t^* = 1, \ldots, T^*_t-1; \enspace i,j = 1,\ldots,N^*$$
 %
-The fine-scale emission probabilities $p_{\theta^{*(X_t)}}(y^*_{s,t} | x^*_{s,t})$ are parameterized by $\theta^{*(X_t)}$.
+Again, we assume that the initial distribution $\delta^{*(s)}$ is the stationary distribution of $\Gamma^{*(s)}$.
+Each fine-scale observation $Y^*_{t,t^*}$ depends only on the value of its corresponding fine-scale hidden state $X^*_{t,t^*}$ and coarse-scale hidden state, $X_t$.
+We denote the distribution of $Y^*_{t,t^*}$ given that $X_t = s$ and $X^*_{t,t^*} = s^*$ as $f_{s,s^*}$.
+Therefore, in the hierarchical setting $\Theta$ is a $N$-by-$N^*$ matrix of parameters where $\theta_{s,s^*}$ describes the distribution $f_{s,s^*}$.
 Note the parameters of the fine-scale hidden Markov model ($\Gamma^{*(X_t)}$, $\delta^{*(X_t)}$, and $\theta^{*(X_t)}$) all depend upon the hidden state of the \textit{crude-scale} hidden Markov model.
@@
 
-One way to deal with auto-correlation in fine-scale behavioral processes is to use a state-switching continuous model such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal.
-Continuous time models are advantageous because of their flexibility: they can be built up from arbitrarily complex stochastic differential equations and they allow for uneven step lengths in the observations sequence $Y$.
-However, most continuous time models require MCMC algorithms to perform inference over and as a result are not easily incorporated into the HHMM structure.
-
-Another option is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}.
-The CarHMM explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm.
-In particular, if the emission distribution of observation $Y_t$ is parameterized by its mean and variance, i.e.
-$\theta_{x_t} = \{\mu_{x_t},\sigma^2_{x_t}\}$, the CarHMM assumes that an observation $Y_t$ has mean $(1-\phi_{x_t}) \cdot \mu_{x_t} + \phi_{x_t} \cdot y_{t-1}$ rather than $\mu_{x_t}$.
-This introduces one new parameter, $\phi_{x_t}$, which represents the auto-correlation of $y_t$ and depends upon the behavioral state of the animal.
-This model easily fits into the HHMM structure, but seems to lack the natural interpretation of continuous-time models.
-However, we prove in the following section that under certain conditions these two models are in fact equivalent.
+One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}.
+If the emission distribution of observation $Y_t$ is parameterized by its mean and variance, i.e.
+$\theta_{x_t} = \{\mu_{x_t},\sigma^2_{x_t}\}$, then the CarHMM models the distribution of $Y_t|X_t,Y_{t-1}$ as the following:
+%
+$$\mathbb{E}(Y_t) = \phi_{x_t}*Y_{t-1} + (1-\phi_{x_t}) * \mu_{x_t}$$
+$$\mathbb{V}(Y_t) = \sigma^2_{x_t}$$
+%
+The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm and only adding one additional parameter per observation and hidden state.
 
@@
 
-Once an HMM, HHMM, or CarHMM model is fit using by maximizing their respective likelihoods with respect to their parameters, it is common to find the most likely sequence of hidden states $\hat X$.
-This is often done using a dynamic programming algorithm called the Viterbi algorithm \cite{Viterbi:1967}.
-In the case of HHMMs, after running the Viterbi algorithm on the coarse-scale process $X$, it can be run again on each fine-scale process $X^*_t$ to find the mostly likely sequence of fine-scale hidden states $\hat X_t^*$ conditioned on the estimated crude-state value $\hat X_t$.
-Note that while $\hat X$ is a maximum likelihood estimate of $X$, $\hat X^*_t$ is \textit{not} necessarily a maximum likelihood estimate of $\hat X^*_t$ because it is conditioned on the value of $\hat X_t$.
+Once an HMM, HHMM, or CarHMM model is fit by maximizing their respective likelihoods with respect to their parameters, it is common to find the most likely sequence of hidden states $\hat X$.
 
-While the Viterbi algorithm is the de-facto standard in the current ecology literature, we suggest to instead find the \textit{probability} of each crude-level state (conditioned on the learned parameters) using the \textit{forward-backward algorithm}.
-The forward-backward algorithm has the same time complexity as the forward algorithm and is also used to find the \textit{psuedoresiduals} of a given model, which is an important tool for model validation.
-In addition, for HHMMs the forward-backward algorithm can be used recursively to find the probability of the fine-level states $X^*_{s,t}$ exactly by marginalizing out $X_t$:
+This can be done in one of two ways: first, the \textit{most likely sequence} of \text{all} hidden states $\hat X$ is often found using a dynamic programming algorithm called the Viterbi algorithm \cite{Viterbi:1967}.
+Second, the \textit{probability} of \textit{each} crude-level state (conditioned on the learned parameters) can be found using the \textit{forward-backward algorithm}.
+In the case of HHMMs, both the Viterbi algorithm and the forward-backward algorithm are easy to extend to HHMMs by recursively running the algorithm on the fine-scale HMMs after decoding the states of the coarse-scale HMM.
+
+While both the Viterbi algorithm and the forward-backward algorithm are used in the current ecology literature, we use the \textit{forward-backward algorithm} in this work for several reasons.
+In particular, the forward-backward algorithm is used to find the \textit{psuedoresiduals} of a given model, which is an important tool for model validation.
+In addition, for HHMMs the forward-backward algorithm can be used recursively to find the probability of the fine-level states $X^*_{t^*,t}$ exactly by marginalizing out $X_t$:
 %
-$$P(X^*_{s,t} = x^*_{s,t}) = \sum_{n=1}^N P(X_t = x_n)P(X^*_{s,t} = x^*_{s,t} | X_t = x_n)$$
+$$P(X^*_{t^*,t} = x^*_{t^*,t}) = \sum_{n=1}^N P(X_t = x_n)P(X^*_{t^*,t} = x^*_{t^*,t} | X_t = x_n)$$
 %

### 2020-06-16 Report/Sections/carhmm_ou.tex

@@
 
-\subsection{Equivalence of CarHMM and OU process}
+\subsection{Continuous Time models}
 
-Once the CarHMM is fit by maximizing the likelihood in equation (\ref{CarHMM_likelihood}), the emission distributions can be interpretted as the solution of a state-switching Ornstein-Uhlenbeck process similar to the one introduced by Michelot et al \cite{Michelot:2019}.
-In order for this to be the case, however, the following two conditions must be met:
-\begin{enumerate}
-	\item The underlying behavioral state of the continuous-time model must follow a Markov chain rather than a Markov process.
-	\item The emission distributions of the CarHMM must be normal.
-\end{enumerate}
-If this is the case, then the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process.
-This gives new interpretation to the learned parameters of the CarHMM in the context of continuous-time model.
-
-\subsubsection{Proof}
-
-A one-dimensional state-switching Ornstein-Uhlenbeck process $y$ is the solution to the following stochastic differential equation:
+Although the CarHMM can incorporate auto-correlation into the structure of an HMM, they break down when observations are taken at irregular time intervals since the probability transition matrix $\Gamma$ would then depend upon the time step between observations.
+A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal.
+Namely, a state-switching Ornstein-Uhlenbeck process $y$ is the solution to the following stochastic differential equation:
 %
@@
 where $x_t$ is the fine-scale behavior of the animal at time $t$, $\beta_{x_t}$ relates to rate at which the process returns to its mean value, $\gamma_{x_t}$ is the long-term mean value of the process, $\omega_{x_t}$ is related to short-term variance, and $W$ is a Wiener process.
-$x_t$ is described by an unobserved Markov process.
+$x_t$ is assumed to follow an unobserved continuous-time Markov process.
 Here, $t \in \mathbb{R}$ is a continuous time stamp rather than an index.
-If the behavioral state $x_t$ does not change between observations (i.e.
-$x_t$ follows a Markov chain), the solution to this equation is known to be the following \cite{Michelot:2019}:
-%
-\begin{align*}
-y_{t+\Delta t} | x_{t:t+\Delta t} \sim \mathcal{N}\left((1-e^{-\beta_{x_t}\Delta t})\gamma_{x_t} + e^{-\beta_{x_t}\Delta t} y_t,\quad \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t})\right)
-\end{align*}
-%
-Now, suppose that $\Delta t$ is constant for all observations, as is the case for hidden Markov models. In addition, introduce the following transformations:
+If the behavioral state $x_t$ is known and does not change between observations, the solution is known to be the following \cite{Michelot:2019}:
 %
 \begin{align}
-\mu_{x_t} = \gamma_{x_t}, \qquad \phi_{x_t} = e^{-\beta_{x_t}\Delta t}, \qquad \sigma^2_{x_t} = \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t}) \\
-\gamma_{x_t} = \mu_{x_t}, \qquad \beta_{x_t} = -\frac{\log(\phi_{x_t})}{\Delta t}, \qquad \omega_{x_t} = \sqrt{\frac{2\sigma^{2}_{x_t}\log(\phi_{x_t})}{\Delta t (\phi_{x_t}^2-1)}}
-\label{CarHMM_to_OU}
+y_{t+\Delta t} | x_{t:t+\Delta t} \sim \mathcal{N}\left((1-e^{-\beta_{x_t}\Delta t})\gamma_{x_t} + e^{-\beta_{x_t}\Delta t} y_t,\quad \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t})\right)
+\label{eqn:OU_sol}
 \end{align}
 %
-Then, almost immediately we can see the following:
+where $\Delta t$ is the distance between any two observations $Y_t$ and $Y_{t+\delta t}$.
+Continuous time models can be built up from arbitrarily complex stochastic differential equations and allow for uneven step lengths in the observations sequence $Y$.
+However, most continuous time models require MCMC algorithms to perform inference over and as a result are not easily incorporated into the HHMM structure.
+
+\subsection{Equivalence of CarHMM and OU process}
+
+Once the CarHMM is fit by maximizing the likelihood in equation (\ref{CarHMM_likelihood}), the emission distributions can be interpreted as the solution of a state-switching Ornstein-Uhlenbeck process similar to the one introduced by Michelot et al \cite{Michelot:2019}.
+In order for this to be the case, however, the following two conditions must be met:
+\begin{enumerate}
+	\item The underlying behavioral state of the continuous-time model follows a Markov chain rather than a Markov process.
+	\item The emission distributions of the CarHMM are normal.
+\end{enumerate}
+If this is the case, then the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process.
+This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.
+
+\subsubsection{Proof}
+
+Suppose that $\Delta t$ is constant for all observations. In addition, introduce the following transformations:
+%
+\begin{align}
+\mu_{x_t} = \gamma_{x_t}, \qquad \phi_{x_t} = e^{-\beta_{x_t}\Delta t}, \qquad \sigma^2_{x_t} = \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t}) \label{eqn:CarHMM_to_OU} \\
+\gamma_{x_t} = \mu_{x_t}, \qquad \beta_{x_t} = -\frac{\log(\phi_{x_t})}{\Delta t}, \qquad \omega_{x_t} = \sqrt{\frac{2\sigma^{2}_{x_t}\log(\phi_{x_t})}{\Delta t (\phi_{x_t}^2-1)}}
+\label{eqn:OU_to_CarHMM}
+\end{align}
+%
+Combining equation (\ref{eqn:CarHMM_to_OU}) with equation (\ref{eqn:OU_sol}) gives the following:
 %
@@
 %
-where $\Lambda = \{p,\lambda\}$, $p_{ij}$ is the probability that the animal moves from state $i$ to state $j$ given that it left state $i$, and $\lambda_i$ is the rate at which the animal leaves behavioral state $i$.
+where $\Lambda = \{p,\lambda\}$, $p_{ij}$ is the probability that the animal moves from state $i$ to state $j$ given that it left state $i$ such that $\sum_{j \neq i} p_ij = 1$, and $\lambda_i$ is the rate at which the animal leaves behavioral state $i$.
 Finally, we just note that the parameters in equation (\ref{CarHMM_to_OU}) become functions of the time step between observations $\Delta t$:

### 2020-06-16 Report/Sections/simulation.tex

@@
 
-To test the CarHHMM with STFT, a sequence of 100 marine-animal dives were simulated.
-The coarse-scale observations $Y$ were set as the duration of each dive, and the fine-scale observations $Y^*$ were set as one dimensional acceleration readings simulated at 50 hertz.
-Specifically, the following procedure was followed:
+%!% Have you simulated this only once? I think you would want to do this multiple time and present results across simulations, no?
+
+To test the CarHHMM with STFT, 100 separate sequences of 100 marine-animal dives were simulated.
+The coarse-scale observations $Y$ were set as the duration of each dive, and the fine-scale observations $Y^*$ were set as one-dimensional acceleration readings simulated at 50 hertz.
+the specific procedure used to generate data was as follows:
 
@@
 	\begin{align*}
-		\Gamma &= \begin{pmatrix} 0.4 & 0.6 \\ 0.6 & 0.4 \end{pmatrix} \\
+		&\Gamma = \begin{pmatrix} 0.4 & 0.6 \\ 0.6 & 0.4 \end{pmatrix} \\
 		% 
 		&Y_t|X_t \sim Gamma \\
-		\bbE(Y_t|X_t = 1) &= 15 s, \quad \bbE(Y_t|X_t = 2) = 60 s \\
+		&\bbE(Y_t|X_t = 1) = 15 s, \quad \quad \bbE(Y_t|X_t = 2) = 60 s \\
 		%
-		\bbV(Y_t|X_t = 1) &= 25 s^2, \quad \bbV(Y_t|X_t = 2) = 100 s^2
+		&\bbV(Y_t|X_t = 1) = 25 s^2, \quad \quad  \bbV(Y_t|X_t = 2) = 100 s^2
 	\end{align*}
@@
 	\item Once the dive durations were calculated, for each $t \in \{1, \ldots, 100\}$, dive $t$ was broken into $\lfloor Y_t/2 \rfloor$ 2-second segments (the end of the dive sequence was discarded).
-Further, each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where $X^*_{t,t^*} \in \{1,2\}$ and $t^* \in \{1,101,201,\ldots,100*\lfloor Y_t/2 \rfloor + 1\}$.
+Further, each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where $X^*_{t,t^*} \in \{1,2\}$ and $t^* \in \{1,101,201,\ldots,100\lfloor Y_t/2 \rfloor - 99\}$.
 The parameters of the fine-scale behaviour Markov chain were set to be as follows:
@@
 
-	\item Finally, $Y^*_{t,t^*:t^*+99}$ is set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
+	\item Finally, $Y^*_{t,t^*:t^*+99}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
 	$$Y^*_{t,t^*:t^*+99} = IDFT\left(\hat{Y}^*_{t,t^*}\right)$$
@@
 
-Four different models were fit to the generated data shown in figure (\ref{fig:sim_data}):
+Four different models were fit to the 100 simulated data sets:
 
 \begin{enumerate}
-    \item A CarHMM with no hierarchical component (CarHMM) using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations
-    \item A Hierarchical hidden Markov model with no auto-correlation (HHMM) using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations.
-    \item A CarHHMM using $Z^{(1)}$, but not $Z^{(2)}$, as fine-scale observations.
-    \item A CarHHMM using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations
+    \item A CarHMM with no hierarchical component using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations (CarHMM).
+    \item A Hierarchical hidden Markov model with no auto-correlation using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations (HHMM).
+    \item A CarHHMM using $Z^{(1)}$, but not $Z^{(2)}$, as fine-scale observations (CarHHMM1).
+    \item A CarHHMM using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations (CarHHMM2).
 \end{enumerate}
-
-Table (\ref{table:emis_dist}) gives the MLE estimates of the parameters for each model above.
+%
+For each model, the following were recorded:
+%
+\begin{enumerate}
+    \item an empirical distribution of the MLE estimates of $\Gamma$ and $\theta$ using the 100 generated data sets generated from true parameters $\Gamma_0$ and $\theta_0$.
+    \item a confusion matrix showing the state decoding capabilities of each model across all data sets.
+    \item The true coverage of 95\% confidence intervals estimated using the observed information matrix (to see how over- or under- confident our parameter estimates are).
+\end{enumerate}
 

### 2020-06-16 Report/Sections/background.tex
Revise Report

@@
 $p(y_t|x_1,\ldots, x_N, y_1,\ldots,y_{t-1},y_{t+1},\ldots,y_N) = p(y_t|x_t)$.
-In particular, we denote the distribution of $Y_t$ given $X_t = s$ as $f_s(y_t)$.
-$f_1,\ldots,f_N$ in turn depend upon a vector of parameters $\Theta = (\theta_1,\ldots,\theta_N)$ and hereafter are referred to as ``emission distributions".
+In particular, we denote the distribution of $Y_t$ given $X_t = s$ as $f^{(s)}(y_t)$.
+$f^{(1)},\ldots,f^{(N)}$ in turn depend upon a vector of parameters $\Theta = (\theta^{(1)},\ldots,\theta^{(N)})$ and hereafter are referred to as ``emission distributions".
 A visualization of the dependence structure can be seen in figure (\ref{fig:HMM}).
@@
 %
-$$P(y_t;\theta) = \text{diag}(f_1(y_t),\ldots , f_N(y_t)).$$
+$$P(y_t;\theta) = \text{diag}(f^{(1)}(y_t),\ldots , f^{(N)}(y_t)).$$
 %
@@
 Note that while $N^*$ can be set to depend upon $X_t$, we assume here that it does not for simplicity in both the model and the notation.
-$X^*_t$ is characterized by an initial distribution $\delta^{*(X_t)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(X_t)} \in \bbR^{N^* \times N^*}$:
+Conditioned on the coarse-scale hidden state $X_t = s$, $X^*_t$ is characterized by an initial distribution $\delta^{*(s)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(s)} \in \bbR^{N^* \times N^*}$:
 %
-$$\delta^{*(s)}_i = Pr(X^*_{t,1} = i | X_t = s); \quad i = 1,\ldots,N^*$$
+$$\delta^{*(s)}_{i^*} = Pr(X^*_{t,1} = i^* | X_t = s); \quad i^* = 1,\ldots,N^*$$
 %
-$$\Gamma^{*(s)}_{ij} = Pr(X^*_{t,t^*+1} = j | X^*_{t,t^*} = i, X_t = s); \quad t^* = 1, \ldots, T^*_t-1; \enspace i,j = 1,\ldots,N^*$$
+$$\Gamma^{*(s)}_{i^*j^*} = Pr(X^*_{t,t^*+1} = j^* | X^*_{t,t^*} = i^*, X_t = s); \quad t^* = 1, \ldots, T^*_t-1; \enspace i^*,j^* = 1,\ldots,N^*$$
 %
-Again, we assume that the initial distribution $\delta^{*(s)}$ is the stationary distribution of $\Gamma^{*(s)}$.
+Again, although not necessary, we assume that the initial distribution $\delta^{*(s)}$ is the stationary distribution of $\Gamma^{*(s)}$.
+In total, there are $N$ different values that $\Gamma^{*(s)}$ and $\delta^{*(s)}$ can take- one for each possible value of $X_t$.
+We denote the sets of all fine-scale initial distributions and probability transition matrices as $\delta^*$ and $\Gamma^*$, respectively, where $|\delta^*| = |\Gamma^*| = N$.
+
 Each fine-scale observation $Y^*_{t,t^*}$ depends only on the value of its corresponding fine-scale hidden state $X^*_{t,t^*}$ and coarse-scale hidden state, $X_t$.
-We denote the distribution of $Y^*_{t,t^*}$ given that $X_t = s$ and $X^*_{t,t^*} = s^*$ as $f_{s,s^*}$.
-Therefore, in the hierarchical setting $\Theta$ is a $N$-by-$N^*$ matrix of parameters where $\theta_{s,s^*}$ describes the distribution $f_{s,s^*}$.
-Note the parameters of the fine-scale hidden Markov model ($\Gamma^{*(X_t)}$, $\delta^{*(X_t)}$, and $\theta^{*(X_t)}$) all depend upon the hidden state of the \textit{crude-scale} hidden Markov model.
-However, depending upon the discretion of the researcher, it is possible to force any of these parameters to be independent of the crude-scale hidden state $X_t$.
+In particular, $p(y^*_{t,t^*} | y^*_{t,1}, \ldots y^*_{t,t^*-1},y^*_{t,t^*+1},\ldots,y^*_{t,T^*_t},x^*_t,x,y) = p(y^*_{t,t^*}|x^*_{t,t^*},x_t)$.
+We denote the distribution of $Y^*_{t,t^*}$ given that $X_t = s$ and $X^*_{t,t^*} = s^*$ as $f^{*(s,s^*)}(y^*_{t,t^*})$.
+Conditioned on the fact that $X_t = s$, the fine-scale distributions $f^{*(s,1)}, \ldots, f^{*(s,N^*)}$ are parameterized by an $N^*$-dimensional vector of parameters $\Theta^{*(s)} = (\theta^{*(s,1)},\ldots,\theta^{*(s,N^*)})$, where $\theta^{*(s,s^*)}$ describes the distribution $f^{*(s,s^*)}$.
+Depending upon the behaviour and responses being modelled, it is possible to force the fine-scale emission distributions to be equal across all values of $s \in \{1, \ldots, N\}: \enspace \theta^{*(1,s^*)} = \theta^{*(2,s^*)} = \ldots = \theta^{*(N,s^*)}$.
 A visualization of the full structure of the HHMM can be seen in figure (\ref{fig:HHMM}).
-Due to the nested structure of a hierarchical hidden Markov model, the likelihood of an HHMM is still easy to calculate using the forward algorithm:
+
+Due to the nested structure of a hierarchical hidden Markov model, the likelihood of an HHMM is still easy to calculate using the forward algorithm.
+Suppose that both coarse-scale and fine scale observations ($y$ and $y^*$, respectively) are observed from an HHMM with unknown parameters.
+Then the likelihood of the observed data is:
 %
-$$\calL_{\text{HHMM}}(y,y^*;\theta,\theta^*,\Gamma,\Gamma^*,\delta,\delta^*) = \delta P(y_1,y_1^*;\theta,\theta^*,\Gamma^*,\delta^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\theta,\theta^*,\Gamma^*,\delta^*) \mathbf{1}$$
+$$\calL_{\text{HHMM}}(y,y^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,y_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N$$
 %
@@
 \begin{align*}
-	P(y_t,y_t^*;\theta,\theta^*,\Gamma^*,\delta^*)  = \text{diag}\left[p_{\theta}(y_t|x_t = x_1)\calL_{\text{HMM}}\left(y_t^*;\theta^{*(x_1)},\Gamma^{*(x_1)},\delta^{*(x_1)}\right), . . . , \right.\\
-	\left. p_{\theta}(y_t|x_t = x_N )\calL_{\text{HMM}}\left(y_t^*;\theta^{*(x_N)},\Gamma^{*(x_N)},\delta^{*(x_N)}\right) \right]
+	P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) = \text{diag}\left[&f^{(1)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right),\ldots, \right. \\
+	\left. &f^{(N)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \right]
 \end{align*}
 %
-Note that this formulation assumes that the crude-scale observations at a given time $Y_t$ and the fine-scale observation time series $Y_t^*$ are independent of one another when conditioned on $X_t$.
+Note that this formulation assumes that the coarse-scale observations at a given time $Y_t$ and the fine-scale observation time series $Y_t^*$ are independent of one another when conditioned on $X_t$.
 For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, model selection and model checking, see Adam et al \cite{Adam:2019}.
@@
 
-\subsection{Conditionally auto-regressive HMMs}
+\subsection{conditionally auto-regressive HMMs}
 
-One of the key assumptions of both HMMs and HHMMs is \textit{conditional Independence} between observations at both the crude and fine scale.
-Namely, given the state $X_t$ or $X^*_{t,t^*}$, $Y_t$ or $Y^*_{t,t^*}$ (respectively) is assumed to be independent from all other observations.
-Therefore, traditional HMMs and HHMMs can fail when the observation sequence $Y$ exhibits significant auto-correlation in time.
+One of the key assumptions of both HMMs and HHMMs is \textit{conditional independence} between observations at both the coarse and fine scale.
+Namely, given the state $X_t$, $Y_t$ is assumed to be independent from all other observations.
+Therefore, traditional HMMs and HHMMs can fail when the observations $Y$ exhibit significant correlation in time.
 Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swimming behavior of horn sharks off the coast of Southern California \cite{Adam:2019}.
@@
 One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}.
-If the emission distribution of observation $Y_t$ is parameterized by its mean and variance, i.e.
-$\theta_{x_t} = \{\mu_{x_t},\sigma^2_{x_t}\}$, then the CarHMM models the distribution of $Y_t|X_t,Y_{t-1}$ as the following:
+In particular, the CarHMM describes the distribution of $(Y_t|Y_{t-1}, X_t=s)$ with parameters $\theta^{(s)} = \{\mu^{(s)},\sigma^{2(s)},\phi^{(s)}\}$, where:
 %
-$$\mathbb{E}(Y_t) = \phi_{x_t}*Y_{t-1} + (1-\phi_{x_t}) * \mu_{x_t}$$
-$$\mathbb{V}(Y_t) = \sigma^2_{x_t}$$
+$$\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=s) = \phi^{(s)}*y_{t-1} + (1-\phi^{(s)}) * \mu^{(s)}$$
+$$\mathbb{V}(Y_t| X_t = s) = \sigma^{2(s)}$$
 %
-The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm and only adding one additional parameter per observation and hidden state.
+The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm and only adding one additional parameter per hidden state $(\phi^{(s)})$.
 
@@
 \begin{equation}
-\calL_{\text{CarHMM}}(y) = \delta \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}
+\calL_{\text{CarHMM}}(y) = \delta \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N
 \label{CarHMM_likelihood}
@@
 %
-$$P(y_t;\theta) = \text{diag}(p_\theta(y_t|y_{t-1},X_t = x_1), . . . , p_\theta(y_t|y_{t-1},X_t = x_N )), \qquad t > 1$$
+$$P(y_t;\Theta) = \text{diag}\left(f^{(1)}(y_t|y_{t-1}), \ldots , f^{(N)}(y_t|y_{t-1}) \right)$$
 %
@@
 
-Once an HMM, HHMM, or CarHMM model is fit by maximizing their respective likelihoods with respect to their parameters, it is common to find the most likely sequence of hidden states $\hat X$.
+Once an HMM, HHMM, or CarHMM model is fit by maximizing its respective likelihood with respect to its parameters, it is common to find the most likely sequence of hidden states $\hat X$.
 
-This can be done in one of two ways: first, the \textit{most likely sequence} of \text{all} hidden states $\hat X$ is often found using a dynamic programming algorithm called the Viterbi algorithm \cite{Viterbi:1967}.
-Second, the \textit{probability} of \textit{each} crude-level state (conditioned on the learned parameters) can be found using the \textit{forward-backward algorithm}.
+This can be done in one of two ways: first, the most likely \textit{sequence} of \text{all} hidden states can be found using a dynamic programming algorithm called the Viterbi algorithm \cite{Viterbi:1967}.
+Alternatively, the probability of \textit{each} coarse-level state (conditioned on the learned parameters) can be found using the \textit{forward-backward algorithm}.
 In the case of HHMMs, both the Viterbi algorithm and the forward-backward algorithm are easy to extend to HHMMs by recursively running the algorithm on the fine-scale HMMs after decoding the states of the coarse-scale HMM.
@@
 While both the Viterbi algorithm and the forward-backward algorithm are used in the current ecology literature, we use the \textit{forward-backward algorithm} in this work for several reasons.
-In particular, the forward-backward algorithm is used to find the \textit{psuedoresiduals} of a given model, which is an important tool for model validation.
-In addition, for HHMMs the forward-backward algorithm can be used recursively to find the probability of the fine-level states $X^*_{t^*,t}$ exactly by marginalizing out $X_t$:
+In particular, the forward-backward algorithm is used to find the \textit{psuedoresiduals} of a given model, which are an important tool for model validation.
+In addition, the forward-backward algorithm can be implemented recursively to find the probability of the fine-level states $X^*_{t^*,t}$ exactly by marginalizing out $X_t$:
 %
-$$P(X^*_{t^*,t} = x^*_{t^*,t}) = \sum_{n=1}^N P(X_t = x_n)P(X^*_{t^*,t} = x^*_{t^*,t} | X_t = x_n)$$
+$$P(X^*_{t^*,t} = s^*) = \sum_{s=1}^N P(X_t = s)P(X^*_{t^*,t} = s^* | X_t = s)$$
 %
-where $P(X_t = x_n)$ can be found using the forward-backward algorithm on the crude-level Markov chain and $P(X^*_{s,t} = x^*_{s,t} | X_t = x_n)$ can be found by running the forward-backward algorithm on the fine-level HMM for every possible value of $X_t$.
+where $P(X_t = s)$ can be found using the forward-backward algorithm on the coarse-level Markov chain and $P(X^*_{s,t} = s^* | X_t = s)$ can be found by running the forward-backward algorithm on the fine-level HMM for every possible value of $s = 1, \ldots, N$.

### 2020-06-16 Report/Sections/carhmm_ou.tex
Revise Report

@@
 A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal.
-Namely, a state-switching Ornstein-Uhlenbeck process $y$ is the solution to the following stochastic differential equation:
+Namely, a state-switching Ornstein-Uhlenbeck process $Y$ is the solution to the following stochastic differential equation:
 %
-$$dy_t = \beta_{x_t}(\gamma_{x_t} - y_t)dt + \omega_{x_t} dW_t$$
+\begin{equation}
+    \label{eqn:SDE}
+    dY_t = \beta_{X_t}(\gamma_{X_t} - Y_t)dt + \omega_{X_t} dW_t
+\end{equation}
 %
-where $x_t$ is the fine-scale behavior of the animal at time $t$, $\beta_{x_t}$ relates to rate at which the process returns to its mean value, $\gamma_{x_t}$ is the long-term mean value of the process, $\omega_{x_t}$ is related to short-term variance, and $W$ is a Wiener process.
-$x_t$ is assumed to follow an unobserved continuous-time Markov process.
-Here, $t \in \mathbb{R}$ is a continuous time stamp rather than an index.
-If the behavioral state $x_t$ is known and does not change between observations, the solution is known to be the following \cite{Michelot:2019}:
+where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta_{X_t}$ relates to rate at which the process returns to its mean value, $\gamma_{X_t}$ is the long-term mean value of the process, $\omega_{X_t}$ is related to short-term variance, and $W$ is a Wiener process.
+One notable difference between the state-switching OU process and the CarHMM is that the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer.
+If the behavioral state $X_t$ is known and does not change between observations, the solution to Equation (\ref{eqn:SDE}) is known to be the following \cite{Michelot:2019}:
 %
 \begin{align}
-y_{t+\Delta t} | x_{t:t+\Delta t} \sim \mathcal{N}\left((1-e^{-\beta_{x_t}\Delta t})\gamma_{x_t} + e^{-\beta_{x_t}\Delta t} y_t,\quad \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t})\right)
+Y_{t+\Delta t} | X_{t:t+\Delta t} \sim \mathcal{N}\left((1-e^{-\beta_{X_t}\Delta t})\gamma_{X_t} + e^{-\beta_{X_t}\Delta t} Y_t,\quad \frac{\omega_{X_t}^2}{2\beta_{X_t}} (1-e^{-2\beta_{X_t}\Delta t})\right)
 \label{eqn:OU_sol}
@@
 %
-where $\Delta t$ is the distance between any two observations $Y_t$ and $Y_{t+\delta t}$.
+where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$.
 Continuous time models can be built up from arbitrarily complex stochastic differential equations and allow for uneven step lengths in the observations sequence $Y$.
@@
 
-Once the CarHMM is fit by maximizing the likelihood in equation (\ref{CarHMM_likelihood}), the emission distributions can be interpreted as the solution of a state-switching Ornstein-Uhlenbeck process similar to the one introduced by Michelot et al \cite{Michelot:2019}.
-In order for this to be the case, however, the following two conditions must be met:
-\begin{enumerate}
-	\item The underlying behavioral state of the continuous-time model follows a Markov chain rather than a Markov process.
-	\item The emission distributions of the CarHMM are normal.
-\end{enumerate}
-If this is the case, then the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process.
+Although most continuous time models are difficult to incorporate into an HHMM, it can be shown that under certain conditions, the CarHMM is equivalent to the state-switching Ornstein-Uhlenbeck process shown in equation (\ref{eqn:OU_sol}).
 This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.
 
-\subsubsection{Proof}
+\begin{theorem}{Theorem 1.}{}%
+If:
+\begin{enumerate}
+    \item The hidden behavioural process $X$ from (Eq.
+\ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occurring at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
+    \item Observations of the SDE from (Eq. \ref{eqn:SDE}) $y_t$ are taken concurrently with every transition of the Markov chain as well as at time $t = 0$, then:
+\end{enumerate}
+The observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission probabilities and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(s)} = \{\mu^{(s)},\sigma^{2(s)},\phi^{(s)}\}$, where:
 
-Suppose that $\Delta t$ is constant for all observations. In addition, introduce the following transformations:
-%
 \begin{align}
-\mu_{x_t} = \gamma_{x_t}, \qquad \phi_{x_t} = e^{-\beta_{x_t}\Delta t}, \qquad \sigma^2_{x_t} = \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t}) \label{eqn:CarHMM_to_OU} \\
-\gamma_{x_t} = \mu_{x_t}, \qquad \beta_{x_t} = -\frac{\log(\phi_{x_t})}{\Delta t}, \qquad \omega_{x_t} = \sqrt{\frac{2\sigma^{2}_{x_t}\log(\phi_{x_t})}{\Delta t (\phi_{x_t}^2-1)}}
-\label{eqn:OU_to_CarHMM}
+\mu^{(s)} = \gamma_{s}, \qquad \sigma^{2(s)} = \frac{\omega_{s}^2}{2\beta_{s}} (1-e^{-2\beta_{s}\Delta t}), \qquad \phi^{(s)} = e^{-\beta_{s}\Delta t} \label{eqn:CarHMM_to_OU}
 \end{align}
-%
-Combining equation (\ref{eqn:CarHMM_to_OU}) with equation (\ref{eqn:OU_sol}) gives the following:
+
+\end{theorem}
+\begin{proof}{Proof}{}%
+Combining equation (\ref{eqn:CarHMM_to_OU}) with equation (\ref{eqn:OU_sol}) gives:
 %
 \begin{align*}
-y_{t+\Delta t} | x_{t:t+\Delta t} \sim \mathcal{N}\left((1-\phi_{x_t})\mu_{x_t} + \phi_{x_t} y_t,\quad \sigma_{x_t}^2 \right)
+\left(Y_{i \Delta t} | X_{(i-1)\Delta t: i \Delta t} = s \right) \sim \mathcal{N}\left((1-\phi^{(s)}) \mu^{(s)} + \phi^{(s)} Y_{(i-1) \Delta t}, \enspace \sigma^{2(s)} \right),\\
+i = 1, \ldots, T-1
 \end{align*}
 %
-If $\Delta t$ is fixed and $x_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities.
-Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again.
-This allows for the principled construction of the continuous-time model to be combined with the computational convenience of the CarHMM.
+If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(i-1) \Delta t} = Y'_{i}$ and $X_{(i-2)\Delta t: (i-1) \Delta t} = X'_i$ for $i = 2,\ldots T$, yielding the desired result:
+
+\begin{align*}
+\left(Y'_{i} | X'_{i} = s \right) \sim \mathcal{N}\left((1-\phi^{(s)}) \mu^{(s)} + \phi^{(s)} Y'_{i-1}, \enspace \sigma^{2(s)} \right)\\
+i = 2, \ldots, T
+\end{align*}
+%
+$X'$ follows a Markov chain, and the distribution of $(Y'_i|X'_i,Y'_{i+1})$ is consistent with that of a CarHMM with normal emission distributions.
+\end{proof}
 

### 2020-06-23 Report/Sections/background.tex
Update Report

@@
 	\centering
-	\includegraphics[width=6.5in]{../Plots/raw_data.png}
+	\includegraphics[width=5in]{../Plots/raw_data.png}
 	\caption{Raw depth data of a Killer Whale off the coast of British Columbia, Canada.}
@@
 Each random variable in the unobserved chain $X_t$ can take one of $N$ possible values, and $X$ has corresponding initial distribution $\delta \in \bbR^N$ and probability transition matrix $\Gamma \in \bbR^{N \times N}$:
-
+%
 $$\delta_i = Pr(X_1 = i); \quad i = 1,\ldots,N$$
-
-$$\Gamma_{ij} = Pr(X_{t+1} = j | X_t = i); \quad t = 1, \ldots, T-1; \enspace i = 1,\ldots,N; \enspace j=1,\ldots,N $$
-
-$\delta$ is often assumed to be the stationary distribution of the $\Gamma$- an assumption that we make in this work.
+%
+$$\Gamma_{ij} = Pr(X_{t+1} = j | X_t = i); \quad t = 1, \ldots, T-1; \enspace i,j = 1,\ldots,N $$
+%
+We assume that $\delta$ is the stationary distribution of the $\Gamma$, implying that the hidden process runs infinitely before observations are taken.
 Further, the distribution of each observation $Y_t$ depends only on the value of $X_t$ and none of the preceding observations or behavioral states, i.e.
 $p(y_t|x_1,\ldots, x_N, y_1,\ldots,y_{t-1},y_{t+1},\ldots,y_N) = p(y_t|x_t)$.
-In particular, we denote the distribution of $Y_t$ given $X_t = s$ as $f^{(s)}(y_t)$.
+In particular, we denote the distribution of $Y_t$ given $X_t = i$ as $f^{(i)}(y_t)$.
 $f^{(1)},\ldots,f^{(N)}$ in turn depend upon a vector of parameters $\Theta = (\theta^{(1)},\ldots,\theta^{(N)})$ and hereafter are referred to as ``emission distributions".
@@
 %
-In order to ensure remove constraints when maximizing $\calL_{\text{HMM}}(y)$, $\Gamma$ is often parameterized using $\eta \in \bbR^{N \times N}$ and the following link function:
+To remove constraints when maximizing $\calL_{\text{HMM}}(y)$, we parameterize $\Gamma$ using $\eta \in \bbR^{N \times N}$ and the following link function \cite{Barajas:2017}:
 %
-$$\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, \qquad \eta_{ii} = 0 \quad \forall i \in \{1, \ldots, N\}.$$
+$$\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{j'=1}^N \exp(\eta_{i(j')})}, \qquad \eta_{ii} = 0 \quad \forall i \in \{1, \ldots, N\}.$$
 %
@@
 Note that while $N^*$ can be set to depend upon $X_t$, we assume here that it does not for simplicity in both the model and the notation.
-Conditioned on the coarse-scale hidden state $X_t = s$, $X^*_t$ is characterized by an initial distribution $\delta^{*(s)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(s)} \in \bbR^{N^* \times N^*}$:
+Conditioned on the coarse-scale hidden state $X_t = i$, $X^*_t$ is characterized by an initial distribution $\delta^{*(i)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(i)} \in \bbR^{N^* \times N^*}$:
 %
-$$\delta^{*(s)}_{i^*} = Pr(X^*_{t,1} = i^* | X_t = s); \quad i^* = 1,\ldots,N^*$$
+$$\delta^{*(i)}_{i^*} = Pr(X^*_{t,1} = i^* | X_t = i); \quad i^* = 1,\ldots,N^*$$
 %
-$$\Gamma^{*(s)}_{i^*j^*} = Pr(X^*_{t,t^*+1} = j^* | X^*_{t,t^*} = i^*, X_t = s); \quad t^* = 1, \ldots, T^*_t-1; \enspace i^*,j^* = 1,\ldots,N^*$$
+$$\Gamma^{*(i)}_{i^*j^*} = Pr(X^*_{t,t^*+1} = j^* | X^*_{t,t^*} = i^*, X_t = i); \quad t^* = 1, \ldots, T^*_t-1; \enspace i^*,j^* = 1,\ldots,N^*$$
 %
-Again, although not necessary, we assume that the initial distribution $\delta^{*(s)}$ is the stationary distribution of $\Gamma^{*(s)}$.
-In total, there are $N$ different values that $\Gamma^{*(s)}$ and $\delta^{*(s)}$ can take- one for each possible value of $X_t$.
+Again, although not necessary, we assume that the initial distribution $\delta^{*(i)}$ is the stationary distribution of $\Gamma^{*(i)}$.
+In total, there are $N$ different values that $\Gamma^{*(i)}$ and $\delta^{*(i)}$ can take- one for each possible value of $X_t$.
 We denote the sets of all fine-scale initial distributions and probability transition matrices as $\delta^*$ and $\Gamma^*$, respectively, where $|\delta^*| = |\Gamma^*| = N$.
@@
 In particular, $p(y^*_{t,t^*} | y^*_{t,1}, \ldots y^*_{t,t^*-1},y^*_{t,t^*+1},\ldots,y^*_{t,T^*_t},x^*_t,x,y) = p(y^*_{t,t^*}|x^*_{t,t^*},x_t)$.
-We denote the distribution of $Y^*_{t,t^*}$ given that $X_t = s$ and $X^*_{t,t^*} = s^*$ as $f^{*(s,s^*)}(y^*_{t,t^*})$.
-Conditioned on the fact that $X_t = s$, the fine-scale distributions $f^{*(s,1)}, \ldots, f^{*(s,N^*)}$ are parameterized by an $N^*$-dimensional vector of parameters $\Theta^{*(s)} = (\theta^{*(s,1)},\ldots,\theta^{*(s,N^*)})$, where $\theta^{*(s,s^*)}$ describes the distribution $f^{*(s,s^*)}$.
-Depending upon the behaviour and responses being modelled, it is possible to force the fine-scale emission distributions to be equal across all values of $s \in \{1, \ldots, N\}: \enspace \theta^{*(1,s^*)} = \theta^{*(2,s^*)} = \ldots = \theta^{*(N,s^*)}$.
+We denote the distribution of $Y^*_{t,t^*}$ given that $X_t = i$ and $X^*_{t,t^*} = i^*$ as $f^{*(i,i^*)}(y^*_{t,t^*})$.
+Conditioned on the fact that $X_t = i$, the fine-scale distributions $f^{*(i,1)}, \ldots, f^{*(i,N^*)}$ are parameterized by an $N^*$-dimensional vector of parameters $\Theta^{*(i)} = (\theta^{*(i,1)},\ldots,\theta^{*(i,N^*)})$, where $\theta^{*(i,i^*)}$ describes the distribution $f^{*(i,i^*)}$.
+Depending upon the behaviour and responses being modelled, it is possible to force the fine-scale emission distributions to be equal across all values of $i \in \{1, \ldots, N\}: \enspace \theta^{*(1,i^*)} = \theta^{*(2,i^*)} = \ldots = \theta^{*(N,i^*)}$.
 A visualization of the full structure of the HHMM can be seen in figure (\ref{fig:HHMM}).
@@
 \begin{align*}
-	P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) = \text{diag}\left[&f^{(1)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right),\ldots, \right. \\
-	\left. &f^{(N)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \right]
+	P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) = \text{diag}\Big[&f^{(1)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right),\ldots, \\
+	&f^{(N)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big]
 \end{align*}
@@
 %
-$$\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=s) = \phi^{(s)}*y_{t-1} + (1-\phi^{(s)}) * \mu^{(s)}$$
-$$\mathbb{V}(Y_t| X_t = s) = \sigma^{2(s)}$$
+$$\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)}*y_{t-1} + (1-\phi^{(i)}) * \mu^{(i)}$$
+$$\mathbb{V}(Y_t| X_t = i) = \sigma^{2(i)}$$
 %
-The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm and only adding one additional parameter per hidden state $(\phi^{(s)})$.
+The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm and only adding one additional parameter per hidden state $(\phi^{(i)})$.
 
@@
 \begin{equation}
-\calL_{\text{CarHMM}}(y) = \delta \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N
+\calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N
 \label{CarHMM_likelihood}
@@
 Alternatively, the probability of \textit{each} coarse-level state (conditioned on the learned parameters) can be found using the \textit{forward-backward algorithm}.
-In the case of HHMMs, both the Viterbi algorithm and the forward-backward algorithm are easy to extend to HHMMs by recursively running the algorithm on the fine-scale HMMs after decoding the states of the coarse-scale HMM.
+Both the Viterbi algorithm and the forward-backward algorithm are easy to extend to Hierarchical HMMs by recursively running the algorithm on the fine-scale HMMs after decoding the hidden states of the coarse-scale process.
 
@@
 %
-$$P(X^*_{t^*,t} = s^*) = \sum_{s=1}^N P(X_t = s)P(X^*_{t^*,t} = s^* | X_t = s)$$
+$$P(X^*_{t^*,t} = i^*) = \sum_{i=1}^N P(X_t = i)P(X^*_{t^*,t} = i^* | X_t = i)$$
 %
-where $P(X_t = s)$ can be found using the forward-backward algorithm on the coarse-level Markov chain and $P(X^*_{s,t} = s^* | X_t = s)$ can be found by running the forward-backward algorithm on the fine-level HMM for every possible value of $s = 1, \ldots, N$.
+where $P(X_t = i)$ can be found using the forward-backward algorithm on the coarse-level Markov chain and $P(X^*_{t^*,t} = i^* | X_t = i)$ can be found by running the forward-backward algorithm on the fine-level HMM for every possible value of $i = 1, \ldots, N$.

### 2020-06-23 Report/Sections/carhmm_ou.tex
Update Report

@@
     \label{eqn:SDE}
-    dY_t = \beta_{X_t}(\gamma_{X_t} - Y_t)dt + \omega_{X_t} dW_t
+    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
 \end{equation}
 %
-where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta_{X_t}$ relates to rate at which the process returns to its mean value, $\gamma_{X_t}$ is the long-term mean value of the process, $\omega_{X_t}$ is related to short-term variance, and $W$ is a Wiener process.
+where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
 One notable difference between the state-switching OU process and the CarHMM is that the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer.
@@
 \begin{align}
-Y_{t+\Delta t} | X_{t:t+\Delta t} \sim \mathcal{N}\left((1-e^{-\beta_{X_t}\Delta t})\gamma_{X_t} + e^{-\beta_{X_t}\Delta t} Y_t,\quad \frac{\omega_{X_t}^2}{2\beta_{X_t}} (1-e^{-2\beta_{X_t}\Delta t})\right)
+Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)}^2}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
 \label{eqn:OU_sol}
@@
 
-Although most continuous time models are difficult to incorporate into an HHMM, it can be shown that under certain conditions, the CarHMM is equivalent to the state-switching Ornstein-Uhlenbeck process shown in equation (\ref{eqn:OU_sol}).
+Although most continuous time models are difficult to incorporate into an HHMM, we show in theorem (1) that under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process shown in equation (\ref{eqn:OU_sol}).
 This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.
@@
     \item The hidden behavioural process $X$ from (Eq.
-\ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occurring at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
-    \item Observations of the SDE from (Eq. \ref{eqn:SDE}) $y_t$ are taken concurrently with every transition of the Markov chain as well as at time $t = 0$, then:
+\ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
+    \item Observations of the SDE from (Eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
 \end{enumerate}
-The observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission probabilities and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(s)} = \{\mu^{(s)},\sigma^{2(s)},\phi^{(s)}\}$, where:
+then the observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{2(i)},\phi^{(i)}\}$, where:
 
 \begin{align}
-\mu^{(s)} = \gamma_{s}, \qquad \sigma^{2(s)} = \frac{\omega_{s}^2}{2\beta_{s}} (1-e^{-2\beta_{s}\Delta t}), \qquad \phi^{(s)} = e^{-\beta_{s}\Delta t} \label{eqn:CarHMM_to_OU}
+\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{2(i)} = \frac{\omega^{(i)}^2}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t}), \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
 \end{align}
@@
 \end{theorem}
+
 \begin{proof}{Proof}{}%
@@
 \begin{align*}
-\left(Y_{i \Delta t} | X_{(i-1)\Delta t: i \Delta t} = s \right) \sim \mathcal{N}\left((1-\phi^{(s)}) \mu^{(s)} + \phi^{(s)} Y_{(i-1) \Delta t}, \enspace \sigma^{2(s)} \right),\\
-i = 1, \ldots, T-1
+\left(Y_{s \Delta t} | X_{(s-1)\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{(s-1) \Delta t}, \enspace \sigma^{2(i)} \right),\\
+s = 1, \ldots, T-1
 \end{align*}
 %
-If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(i-1) \Delta t} = Y'_{i}$ and $X_{(i-2)\Delta t: (i-1) \Delta t} = X'_i$ for $i = 2,\ldots T$, yielding the desired result:
+If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t: (s-1) \Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:
 
 \begin{align*}
-\left(Y'_{i} | X'_{i} = s \right) \sim \mathcal{N}\left((1-\phi^{(s)}) \mu^{(s)} + \phi^{(s)} Y'_{i-1}, \enspace \sigma^{2(s)} \right)\\
-i = 2, \ldots, T
+\left(Y'_s| X'_s = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y'_{s-1}, \enspace \sigma^{2(i)} \right)\\
+s = 2, \ldots, T
 \end{align*}
 %
-$X'$ follows a Markov chain, and the distribution of $(Y'_i|X'_i,Y'_{i+1})$ is consistent with that of a CarHMM with normal emission distributions.
+$X'$ follows a Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s+1})$ is consistent with that of a CarHMM with normal emission distributions.
 \end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%  End of Section  %%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\iffalse
 
@@
 now $\calL_{\text{CarHMM}}(y;\theta,\Lambda,\delta)$ is simply maximized with respect to $\theta = \{\beta,\omega,\gamma\}$, $\Lambda = \{p,\lambda\}$, and $\delta$ to find the maximum likelihood estimate of the CarHMM with uneven time steps.
-
-\iffalse
 

### 2020-06-23 Report/Sections/model.tex
Update Report

@@
 
-One issue with the CarHMM is that it assumes Markovian dynamics conditioned on the hidden state, i.e.
-that any observation $Y^*_{t,t^*}$ depends only on the behavioral state $X^*_{t,t^*}$ and $Y^*_{t,t^*-1}$ (here we focus on the fine-scale process).
+One issue with the CarHMM is that it assumes Markovian dynamics when conditioned on the hidden state, i.e.
+that any observation $Y^*_{t,t^*}$ depends only on the behavioral state $X^*_{t,t^*}$ and $Y^*_{t,t^*-1}$.
 However, there are many animal movement processes which violate this Markov property on very fine scales.
-For example, swimming behaviour of marine mammals can be periodic since the animal repeatedly flukes to propel itself forward.
+For example, swimming behaviour of marine mammals often exhibits periodic behavior since the animal repeatedly flukes to propel itself forward.
 Work has been done in the past to model non-Markovian dynamics in the \textit{behavioural} process $X^*_t$ \cite{Langrock:2012}, but addressing non-Markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area (MAYBE?).
-With improvements in tagging technology allowing for data collection at very high frequencies, data exhibiting noisy and non-Markovian fine scale behavior is likely to persist.
+With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine scale behavior is likely to persist.
 
-To address this issue, we recommend borrowing techniques from the signal processing literature to compress the data and summarize its essential elements.
-In particular, we suggest performing the discrete-time Short-time Fourier Transform (STFT) over each observed fine-scale process $Y^*_t$:
+To address this issue, we recommend borrowing techniques from signal processing literature to compress the data and summarize its essential elements.
+In particular, we suggest performing the discrete-time short-time Fourier transform (STFT) over the observed fine-scale process $Y^*_t$:
 %
-$$STFT\{Y^*_{t,t^*:t^*+w-1}\}(n) := \hat{Y}^{*(n)}_{t,t^*} = \sum_{n = 0}^{w-1} Y^*_{t,t^*+n}e^{-i \frac{2\pi k}{w} n} \qquad \forall n \in \{0, \ldots, w-1\}, \quad t^* \in \{1,w+1,2w+1, \ldots, w \left(\lfloor T^*_t / w \rfloor -1 \right) + 1\}.$$
+\begin{align*}
+    STFT\{Y^*_{t,(s^*w)+1:(s^*+1)w) }\}(k) := \hat{Y}^{*(k)}_{t,s^*} = \sum_{n = 1}^{w} Y^*_{t,(s^*w)+n}e^{-i \frac{2\pi k}{w} (n-1)} \\ k = 0, 1, \ldots, w-1; \quad s^* = 0,1, \ldots, \lfloor T^*_t / w \rfloor - 1
+\end{align*}
 %
-The STFT slides a moving window of length $w$ across the time series $Y_t^*$ with a step size of $w$ and transforms the domain of each window from time to frequency.
+Where $i$ in the equation above refers the $\sqrt{-1}$. Note here that $s^*$ indexes the \textit{windows} moving across $Y^*_t$ while $t^*$ indexes the \textit{observations} of $Y^*_t$.
+
+The STFT slides a moving window of length $w$ across the time series $Y_t^*$ and transforms the domain of each window from time to frequency.
 This allows the spectrum of $Y_t^*$ at time $t^*$ to be summarized by a $w$-dimensional vector of Fourier coefficients.
@@
 Although this allows $Y^*_t$ to be represented in a way that eliminates obvious periodic behavior, the data set itself is still approximately as large as $Y^*_t$ itself.
-To reduce the size of $\hat{Y}^*_t$, we propose taking summary statistics of each window as follows:
+To reduce the size of $\hat{Y}^*_t$, we take two summary statistics of each window:
 %
-$$Z_{t,t^*}^{*(1)} = \mathcal{R}\left(\hat{Y}^{*(0)}_{t,t^*}\right) \qquad Z_{t,t^*}^{*(2)} = \frac{1}{w}\sum_{n=1}^{\tilde{f}}|\hat{Y}^{*(n)}_{t,t^*}|^2$$
+$$Z_{t,s^*}^{*(1)} = \mathcal{R}\left(\hat{Y}^{*(0)}_{t,s^*}\right) \qquad Z_{t,s^*}^{*(2)} = \frac{1}{w}\sum_{k=1}^{\tilde{f}}|\hat{Y}^{*(k)}_{t,s^*}|^2$$
 %
-$Z_{t,t^*}^{*(1)}$ is equal to the average value of $Y_{t,t^*:t^*+w-1}^*$, and $Z_{t,t^*}^{*(2)}$ is equal to the squared 2-norm of the component of $Y_{t,t^*:t^*+w-1}^*$ that can be attributed to frequencies in the signal between $1$ and $\tilde{f}$ periods per window length.
-Both the window length $w$ and the max frequency $\tilde{f}$ are tuning parameters that should be tuned in a problem-specific way.
+$Z_{t,s^*}^{*(1)}$ is equal to the average value of $Y_{t,s^*}^*$, and $Z_{t,s^*}^{*(2)}$ is equal to the squared 2-norm of the component of $Y_{t,s^*}^*$ that can be attributed to frequencies in the signal between $1$ and $\tilde{f}$ periods per window length.
+Both the window length $w$ and the max frequency $\tilde{f}$ are problem-specific tuning parameters.
 $w$ should be long enough to capture the periodic behavior of the underlying process (at least as long as the length of a period), but short enough to avoid over-smoothing of the data and to maintain high resolution in the behavioral process $X^*$.
 $\tilde{f}$ should be selected such that the maximum frequency of $Y_t^*$ that makes biological sense is $\tilde{f}$ per window length.
-Note that these summary statistics are just one possible choice of summary statistics within each window, and future studies can adjust these definitions as needed.
+Note that these summary statistics are just one possible choice to describe each window.
+Other choices of summary statistics include the maximum and argmax of $\hat Y^*_{s^*,t}$.
+Future studies can adjust these definitions as needed.
 A visualization of transforming a one-dimensional sequence $Y^*$ to $Z^*$ can be seen in figure (\ref{fig:fourier_example}).
@@
 	\centering
-	\includegraphics[width=7.5in]{../Plots/fourier_transform.png}
+	\includegraphics[width=5in]{../Plots/fourier_transform.png}
 	\caption{Visualization of transforming $Y^*_t$ into $Z^*_t$ using a sliding window and fourier transform.}
@@
 
-Finally, note that it is possible to accommodate for unequal time steps within each window by using the \textbf{non-uniform discrete Fourier transform (NDFT)}.
+Note that it is possible to accommodate for unequal time steps within each window by using the \textbf{non-uniform discrete Fourier transform (NDFT)}.
 We do not describe the details of this method in this work, but the generalization is straightforward.
 Refer to Bagchi et al \cite{Bagchi:2001} for details.
+
+Fourier analysis has previously been used in the field of animal movement to explain animal behavior \cite{Fehlmann:2017} and specifically fluking \cite{Shorter:2017} from accelerometer data, and Fourier analysis within an HMM has been used to describe daily behavioral cycles of marine mammals  \cite{Heerah:2017}, but we are currently unaware of any studies which have incorporated Fourier analysis of accelerometer data within the structure of an HMM.
 
@@
 However, as mentioned before, the fine-scale process $Y^*$ can often exhibit auto-correlation and intricate structure.
-Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit auto-correlation, especially in the window average, $Z_{t,t^*}^{*(1)}$.
-Therefore, we replace the fine-scale HMM within the Hierarchical HMM with a CarHMM according to figure (\ref{fig:CarHHMM}).
+Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit auto-correlation, especially in $Z_{t,s^*}^{*(1)}$.
+Therefore, we replace the fine-scale HMM within the Hierarchical HMM with a CarHMM, as shown in figure (\ref{fig:CarHHMM}).
 
@@
 	\centering
-	\includegraphics[width=7.5in]{../Plots/CarHHMM.png}
+	\includegraphics[width=5in]{../Plots/CarHHMM.png}
 	\caption{Graphical representation of a CarHHMM. The additional arrows representing auto-correlation between observations are shown in red for emphasis.}
@@
 %
-$$\calL_{\text{CarHHMM}}(y,z^*;\theta,\theta^*,\Gamma,\Gamma^*,\delta,\delta^*) = \delta P(y_1,z_1^*;\theta,\theta^*,\Gamma^*,\delta^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\theta,\theta^*,\Gamma^*,\delta^*) \mathbf{1}$$
+$$\calL_{\text{CarHHMM}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N$$
 %
@@
 \begin{align*}
-P(y_t,z_t^*;\theta,\theta^*,\Gamma^*,\delta^*)  = \text{diag}\left[p_{\theta}(y_t|X_t = x_1)\calL_{\text{CarHMM}}\left(z_t^*;\theta^{*(x_1)},\Gamma^{*(x_1)},\delta^{*(x_1)}\right), . . . , \right.\\
-\left. p_{\theta}(y_t|x_t = x_N )\calL_{\text{CarHMM}}\left(z_t^*;\theta^{*(x_N)},\Gamma^{*(x_N)},\delta^{*(x_N)}\right) \right]
+P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
+&f^{(N)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big]
 \end{align*}

### 2020-06-23 Report/Sections/simulation.tex
Revise Report
Update Report

@@
 
-To test the CarHHMM with STFT, 100 separate sequences of 100 marine-animal dives were simulated.
+To test the CarHHMM with STFT, 500 separate sequences of 100 marine-mammal dives were simulated.
 The coarse-scale observations $Y$ were set as the duration of each dive, and the fine-scale observations $Y^*$ were set as one-dimensional acceleration readings simulated at 50 hertz.
@@
 	\begin{align*}
-		&\Gamma = \begin{pmatrix} 0.4 & 0.6 \\ 0.6 & 0.4 \end{pmatrix} \\
+		&\Gamma = \begin{pmatrix} 0.4 & 0.6 \\ 0.6 & 0.4 \end{pmatrix} & \\
 		% 
-		&Y_t|X_t \sim Gamma \\
-		&\bbE(Y_t|X_t = 1) = 15 s, \quad \quad \bbE(Y_t|X_t = 2) = 60 s \\
+		&Y_t|X_t \sim \rm{Gamma} & \\
+		&\bbE(Y_t|X_t = 1) = 15 s &\bbE(Y_t|X_t = 2) = 60 s \\
 		%
-		&\bbV(Y_t|X_t = 1) = 25 s^2, \quad \quad  \bbV(Y_t|X_t = 2) = 100 s^2
+		&\bbV(Y_t|X_t = 1) = 25 s^2 &\bbV(Y_t|X_t = 2) = 100 s^2
 	\end{align*}
@@
 	\item Once the dive durations were calculated, for each $t \in \{1, \ldots, 100\}$, dive $t$ was broken into $\lfloor Y_t/2 \rfloor$ 2-second segments (the end of the dive sequence was discarded).
-Further, each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where $X^*_{t,t^*} \in \{1,2\}$ and $t^* \in \{1,101,201,\ldots,100\lfloor Y_t/2 \rfloor - 99\}$.
+Further, each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where $X^*_{t,s^*} \in \{1,2\}$ and $s^* \in \{0,1,\ldots,\lfloor Y_t/2 \rfloor\}$.
 The parameters of the fine-scale behaviour Markov chain were set to be as follows:
@@
 
-	\item For each 2-second segment, the Fourier modes $\hat{Y}^*_{t,t^*}$ were simulated using the following procedure.
-Note that the $n^{th}$ Fourier mode of $\hat{Y}^*_{t,t^*}$ is denoted as $\hat{Y}^{*(n)}_{t,t^*}$ :
+	\item For each 2-second segment, the Fourier modes $\hat{Y}^*_{t,s^*}$ were simulated using the following procedure.
+Note that the $k^{th}$ Fourier mode of $\hat{Y}^*_{t,s^*}$ is denoted as $\hat{Y}^{*(k)}_{t,s^*}$ :
 
 	\begin{align*}
-    	\hat{Y}^{*(0)}_{t,1} &\sim \mathcal{N} \left(\mu = 1, \sigma^2 = 0.0025 \right) & \\
+    	\hat{Y}^{*(0)}_{t,0} &\sim \mathcal{N} \left(\mu = 0, \sigma^2 = 0.0025 \right) & \\
     	%
-    	\hat{Y}^{*(0)}_{t,t^*} &\sim \mathcal{N} \left(\mu = 0.9 Y^{*(0)}_{t,t^*-100} + 0.1, \sigma^2 = 0.0025 \right), & t^* \in \{101,201,\ldots, 100*\lfloor Y_t/2 \rfloor + 1\} \\
+    	\hat{Y}^{*(0)}_{t,s^*} &\sim \mathcal{N} \left(\mu = 0.9Y^{*(0)}_{t,s^*-1}, \sigma^2 = 0.0025 \right), & s^* = 1,2,\ldots, \lfloor Y_t/2 \rfloor \\
     	%
-    	\hat{Y}^{*(n)}_{t,t^*} &= a_{t,t^*}^{(n)} i\sqrt{b^{(n)}_{t,t^*}}, & n \in \{1,\ldots,49\} \\
+    	\hat{Y}^{*(k)}_{t,t^*} &= a_{t,s^*}^{(k)} i\sqrt{b^{(k)}_{t,s^*}}, & k = 1,\ldots,49 \\
 	\end{align*}
 	\begin{align*}
-	    a_{t,t^*}^{(n)} &\sim  \left\{\begin{array}{lr}
-    	-1 & w.p. \enspace 1/2 \\
-    	1  & w.p. \enspace 1/2
+	    a_{t,s^*}^{(k)} &\sim  \left\{\begin{array}{lr}
+    	-1 & \quad w.p. \enspace 1/2 \\
+    	1  & \quad w.p. \enspace 1/2
     	\end{array}\right. \\
     	%
-    	(b^{(n)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim Gamma(1/n^2, 1) \\
+    	(b^{(k)}_{t,s^*}|X^*_{t,s^*}  = 1) &\sim {\rm{Gamma}}(1/k^2, 1) \\
     	%
-    	(b^{(n)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
-    	Gamma(1/n^2, 1) & \text{for } n \neq 2\\
-    	Gamma(20,1) & \text{for } n = 2
+    	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
+    	{\rm{Gamma}}(1/k^2, 1), \quad & k \notin \{1,2\} \\
+    	{\rm{Gamma}}(100,1), & k = 1 \\
+    	{\rm{Gamma}}(50,1), & k = 2
     	\end{array}\right. 
@@
     \begin{align*}
-        \hat{Y}^{*(50)}_{t,t^*} &= 0 & \\
+        \hat{Y}^{*(50)}_{t,s^*} &= 0 & \\
     	%
-    	\hat{Y}^{*(n)}_{t,t^*}  &= -\hat{Y}^{*(100-n)}_{t,t^*}, & \qquad n \in \{51,\ldots,99\}
+    	\hat{Y}^{*(k)}_{t,s^*}  &= -\hat{Y}^{*(100-n)}_{t,s^*}, & \qquad k = 51,\ldots,99
     \end{align*}
 
-	\item Finally, $Y^*_{t,t^*:t^*+99}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
-	$$Y^*_{t,t^*:t^*+99} = IDFT\left(\hat{Y}^*_{t,t^*}\right)$$
+	\item Finally, $Y^*_{t,(100s^*+1):(100s^*+100)}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,s^*}$:
+	$$Y^*_{t,(100s^*+1):(100s^*+100)} = IDFT\left(\hat{Y}^*_{t,s^*}\right)$$
 
@@
 There are several practical reasons behind this construction.
-First, $\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued.
-$\hat{Y}^{*(n)}_{t,t^*}$ also decays like $1/n^2$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window.
-Note, however, that $Y^*_{t,t^*}$ is not continuous in $t^*$ \textit{between} windows, as can be seen in figure (\ref{fig:fourier_example}).
-However, note the jumps are not too severe since $\hat{Y}^{*(0)}_{t,t^*}$ and $\hat{Y}^{*(0)}_{t,t^*+100}$ are highly correlated.
+First, $\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,s^*}$ so that its inverse Fourier transform is real-valued.
+$\hat{Y}^{*(k)}_{t,s^*}$ also decays like $1/k^2$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window.
+Note, however, that $Y^*_{t,s^*}$ is not continuous \textit{between} windows, as can be seen in figure (\ref{fig:fourier_example}).
+However, the jumps are not too severe since $\hat{Y}^{*(0)}_{t,s^*}$ and $\hat{Y}^{*(0)}_{t,s^*+1}$ are highly correlated.
 See figure (\ref{fig:sim_data}) for details.
@@
 	% 
-	Y_t|X_t = 1 &\sim Gamma\left(\mu = 15 s, \sigma = 5 s\right) \\
-	Y_t|X_t = 2 &\sim Gamma\left(\mu = 60 s, \sigma = 10 s\right) \\
+	Y_t|X_t = 1 &\sim {\rm{Gamma}}\left(\mu = 15 s, \sigma = 5 s\right) \\
+	Y_t|X_t = 2 &\sim {\rm{Gamma}}\left(\mu = 60 s, \sigma = 10 s\right) \\
 \end{align*}
@@
 \begin{align*}
-    \left(Z^{*(1)}_{t,t^*+100} | Z^{*(1)}_{t,t^*} \right) &\sim \mathcal{N} \left(\phi^* * Z^{*(1)}_{t,t^*} + \phi^* * \mu^*, \sigma^{*2} \right), & \mu^* = 1, \enspace \phi^* = 0.9, \enspace \sigma^{*2} = 0.025 \\
+    \left(Z^{*(1)}_{t,s^*} | Z^{*(1)}_{t,s^*-1} \right) &\sim \mathcal{N} \left(\phi^*Z^{*(1)}_{t,s^*-1} + \phi^*\mu^*, \sigma^{*2} \right), & \mu^* = 0, \enspace \phi^* = 0.9, \enspace \sigma^{*2} = 0.025 \\
     %
-    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 1\right) &\sim Gamma\left(\alpha^{*(1)}, \beta^* \right), & \alpha^{*(1)} = \sum_{n=1}^{\tilde{f}} \frac{1}{n^2}, \enspace \beta^* = 1 \\
+    \left(Z^{*(2)}_{t,s^*} | X^*_{t,s^*} = 1\right) &\sim {\rm{Gamma}}\left(\alpha^{*(1)}, \beta^* \right), & \alpha^{*(1)} = \sum_{k=1}^{\tilde{f}} \frac{1}{k^2}, \enspace \beta^* = 1 \\
     %
-    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 2\right) &\sim Gamma\left(\alpha^{*(2)}, \beta^* \right), & \alpha^{*(2)} = 20 + \sum_{n=2}^{\tilde{f}} \frac{1}{n^2}, \enspace \beta^* = 1 \\
+    \left(Z^{*(2)}_{t,t^*} | X^*_{t,s^*} = 2\right) &\sim {\rm{Gamma}}\left(\alpha^{*(2)}, \beta^* \right), & \alpha^{*(2)} = 20 + \sum_{k=2}^{\tilde{f}} \frac{1}{k^2}, \enspace \beta^* = 1 \\
 \end{align*}
 %
-The emission distribution of the 2-second average acceleration $Z_{t,t^*}^{*(1)}$ in particular can be reparameterized as an OU process:
+The emission distribution of the 2-second average acceleration $Z_{t,s^*}^{*(1)}$ can be reparameterized as an OU process according to Theorem 1:
 %
@@
 
-$$\gamma^* = \mu^* = 1, \qquad \beta^* = -\frac{\log(\phi^*)}{\Delta t^*} = 0.0229, \qquad \omega^{*2} = \frac{2\sigma^{*2}\log(\phi)}{\Delta t (\phi^{*2}-1)} = 0.006$$
-
-Note that the two fine-scale states $X^*_{t,t^*} = 1$ and $X^*_{t,t^*} = 2$ are identical except for the fact that the distribution of $Z^{*(2)}_{t,t^*}$ is shifted up in its shape parameter in fine-scale state 2.
-This corresponds to more periodic behavior in acceleration, but \textit{not} a higher mean acceleration.
+$$\gamma^* = \mu^* = 0, \qquad \beta^* = -\frac{\log(\phi^*)}{\Delta t^*} = 0.0229, \qquad \omega^{*2} = \frac{2\sigma^{*2}\log(\phi)}{\Delta t (\phi^{*2}-1)} = 0.006$$
 
@@
 
-Four different models were fit to the 100 simulated data sets:
+Four different models were fit to each simulated data set:
 
@@
 \begin{enumerate}
-    \item an empirical distribution of the MLE estimates of $\Gamma$ and $\theta$ using the 100 generated data sets generated from true parameters $\Gamma_0$ and $\theta_0$.
-    \item a confusion matrix showing the state decoding capabilities of each model across all data sets.
-    \item The true coverage of 95\% confidence intervals estimated using the observed information matrix (to see how over- or under- confident our parameter estimates are).
+    \item An empirical distribution of the MLE estimates of $\Gamma$ and $\Theta$ using the 100 generated data sets generated from true parameters $\Gamma_0$ and $\Theta_0$.
+    \item The estimated spread of these distributions using the observed information matrix.
+    \item A confusion matrix showing the state decoding capabilities of each model across all data sets.
 \end{enumerate}
+
+The empirical distribution of
 

### 2020-06-23 Report/CJS_Sample.tex
Preliminary Simulation Study Results DONE

@@
 Cleveland, W. (1979). Robust locally-weighted regression and smoothing scatterplots. 
-{\it Journal of the American Statistical Association}, 74, 829-836.
+{\it Journal of the American Statistical Association}, 74, 829-836.
 
@@
 with a complex design: Some advances in methods. In Lynn, R. (Ed.) {\it Methodology of Longitudinal Surveys}, 
-John Wiley \& Sons, New York, 351-366.
+John Wiley \& Sons, New York, 351-366.
 \end{thebibliography}

### 2020-06-23 Code/HHMM.py

@@ -11,6 +11,7 @@ from scipy.stats import gaussian_kde
 from scipy.stats import circstd
 from scipy.special import iv
 from scipy.special import expit
+from scipy.special import logit
 from scipy.special import logsumexp
 from scipy.optimize import minimize
 from scipy.optimize import LinearConstraint
@@ -25,6 +26,29 @@ import time
 import pickle
 
 
+def corr_2_rho(corr):
+    return logit(corr)
+
+def ptm_2_eta(gamma):
+
+    eta = np.zeros_like(gamma)
+    N = len(gamma)
+
+    for _ in range(100):
+        for i in range(N):
+            A = sum(np.exp(eta[i]))
+            for j in range(N):
+                if i != j:
+                    A0 = A-np.exp(eta[i,j])
+                    eta[i,j] = np.log((A0*gamma[i,j])/(1-gamma[i,j]))
+
+    return eta
+
+def eta_2_ptm(eta):
+    ptm = np.exp(eta)
+    return (ptm.T/np.sum(ptm,1)).T
+
+
 class HHMM:
 
 
@@ -34,7 +58,12 @@ class HHMM:
         self.theta = []
         self.eta = []
         self.ptm = []
+
+
         self.data = data
+        self.SEs = None
+        self.train_time = None
+        self.CM = None
 
         eta_crude = -1.0 + 0.1*np.random.normal(size=(self.pars.K[0],self.pars.K[0]))
         ptm_crude = np.exp(eta_crude)
@@ -158,13 +187,6 @@ class HHMM:
         return
 
 
-    def initalize_theta_sim(self):
-
-
-
-        return
-
-
     def find_log_p_yt_given_xt(self,level,feature,data,data_tm1,mu,sig,corr,sample=0):
 
         # find log density of feature
@@ -229,6 +251,10 @@ class HHMM:
 
 
     def dive_likelihood(self,dive_data,state):
+
+        # deal with dive lengths of 0
+        if len(dive_data) == 0:
+            return 0
 
         # find tpm
         self.eta[1][state][np.diag_indices(self.pars.K[1])] = 0
@@ -431,12 +457,14 @@ class HHMM:
 
     def train_DM(self,data,max_iters=10,tol=0.01,eps=10e-6):
 
+        stime = time.time()
         options = {'maxiter':10,'disp':False}
         prev_l = self.likelihood(data)
 
         for _ in range(max_iters):
 
             print(prev_l)
+            print(self.theta)
 
[…]

### 2020-06-23 Code/Run_Sim.py

@@ -32,6 +32,7 @@ import HHMM
 import Visualisor
 
 rand_seed = int(sys.argv[1])
+model = str(sys.argv[2])
 
 random.seed(rand_seed)
 np.random.seed(rand_seed)
@@ -308,6 +309,8 @@ hhmm_Vs = []
 hhmm_FV_uncorrs = []
 hhmm_FVs = []
 
+h = 0.001
+
 for dataset_num in range(ndatasets):
 
 
@@ -319,154 +322,204 @@ for dataset_num in range(ndatasets):
 
 
     ### CarHMM ###
-    print('')
-    print('STARTING CarHMM')
-    print('')
-    pars = Parameters.Parameters()
-    pars.K = [1,2]
-    pars.features = [{'dive_duration':{'corr':False,'f':'gamma'}},
-                     {'FoVeDBA':{'corr':False,'f':'gamma'},
-                      'A':{'corr':True,'f':'normal'}}]
-    pars.theta = hmm_FV_theta
+    if model == 'CarHMM':
+        print('')
+        print('STARTING CarHMM')
+        print('')
+        pars = Parameters.Parameters()
+        pars.K = [1,2]
+        pars.features = [{'dive_duration':{'corr':False,'f':'gamma'}},
+                         {'FoVeDBA':{'corr':False,'f':'gamma'},
+                          'A':{'corr':True,'f':'normal'}}]
+        pars.theta = hmm_FV_theta
 
-    hmm_FV = HHMM.HHMM(pars,data_FV)
-    hmm_FV.theta = hmm_FV_theta
-    hmm_FV.eta = hmm_FV_eta
-    hmm_FV.true_theta = deepcopy(hmm_FV_theta)
-    hmm_FV.true_eta = deepcopy(hmm_FV_eta)
+        hmm_FV = HHMM.HHMM(pars,data_FV)
+        hmm_FV.theta = hmm_FV_theta
+        hmm_FV.eta = hmm_FV_eta
+        hmm_FV.true_theta = deepcopy(hmm_FV_theta)
+        hmm_FV.true_eta = deepcopy(hmm_FV_eta)
 
-    if train_new:
         hmm_FV.train_DM(data_FV)
-        hmm_FV.save('../Params/hmm_FV_%d_%d'%(dataset_num,rand_seed))
-    else:
-        hmm_FV = hmm_FV.load('../Params/hmm_FV_%d_%d'%(dataset_num,rand_seed))
[…]

### 2020-06-23 prep_python.sh

@@ -1,5 +1,4 @@
 #!/bin/bash
 #SBATCH --account=def-nheckman
-cd projects/nheckman-def/evsi8432
 module load python/3.7
 source ~/Whale_Paper_1/bin/activate

### 2020-06-23 Code/run_all_sims.sh
add plots (for real this time)

@@ -8,8 +8,11 @@
 #SBATCH --mail-type=REQUEUE
 #SBATCH --mail-type=ALL
 
-#SBATCH --time=24:00:00
+#SBATCH --time=20:00:00
 #SBATCH --mem-per-cpu=4G
-#SBATCH --array=1-100
+#SBATCH --array=1-500
 
-python Run_Sim.py $SLURM_ARRAY_TASK_ID
+python Run_Sim.py $SLURM_ARRAY_TASK_ID CarHMM
+python Run_Sim.py $SLURM_ARRAY_TASK_ID HHMM
+python Run_Sim.py $SLURM_ARRAY_TASK_ID CarHHMM1
+python Run_Sim.py $SLURM_ARRAY_TASK_ID CarHHMM2

### 2020-06-25 Code/HHMM.py
update fisher information calculation

@@ -781,22 +781,28 @@ class HHMM:
                     # get plus value
                     if param == 'corr':
                         self.theta[1][0][feature][param][state_num] = corr_2_rho(theta+h0)
+                        self.theta[1][1][feature][param][state_num] = self.theta[1][0][feature][param][state_num]
                     else:
                         self.theta[1][0][feature][param][state_num] += h
+                        self.theta[1][1][feature][param][state_num] = self.theta[1][0][feature][param][state_num]
                     th_tp1 = self.likelihood(data)
 
                     # get minus value
                     if param == 'corr':
                         self.theta[1][0][feature][param][state_num] = corr_2_rho(theta-h0)
+                        self.theta[1][1][feature][param][state_num] = self.theta[1][0][feature][param][state_num]
                     else:
                         self.theta[1][0][feature][param][state_num] += -2*h
+                        self.theta[1][1][feature][param][state_num] = self.theta[1][0][feature][param][state_num]
                     th_tm1 = self.likelihood(data)
 
                     # return theta
                     if param == 'corr':
                         self.theta[1][0][feature][param][state_num] = corr_2_rho(theta)
+                        self.theta[1][1][feature][param][state_num] = self.theta[1][0][feature][param][state_num]
                     else:
                         self.theta[1][0][feature][param][state_num] += h
+                        self.theta[1][1][feature][param][state_num] = self.theta[1][0][feature][param][state_num]
 
                     # get estimate
                     if param =='corr':

### 2020-06-25 Report/Sections/appendix.tex
Update Report

@@
 \setcounter{subsection}{0}
+
+\subsection{Detailed Description of Data Simulation}
+
+$\hat{Y}^*_{t,s^*}$ was simulated using the following procedure. Note that the $k^{th}$ Fourier mode of $\hat{Y}^*_{t,s^*}$ is denoted as $\hat{Y}^{*(k)}_{t,s^*}$:
+
+\begin{align*}
+	(\hat{Y}^{*(0)}_{t,0}|X^*_{t,0} = i) &\sim \mathcal{N} \left(0, \sigma^{*(i)} \right) & \\
+	%
+	(\hat{Y}^{*(0)}_{t,s^*}|X^*_{t,s^*} = i) &\sim \mathcal{N} \left(\phi^{*(i)} * \hat{Y}^{*(0)}_{t,s^*-1}, \sigma^{*(i)} \right), & s^* = 1,2,\ldots, \lfloor Y_t/2 \rfloor \\
+	%
+	\hat{Y}^{*(k)}_{t,s^*} &= a_{t,s^*}^{(k)} i\sqrt{b^{(k)}_{t,s^*}}, & k = 1,\ldots,49 \\
+\end{align*}
+\begin{align*}
+    a_{t,s^*}^{(k)} &\sim  \left\{\begin{array}{lr}
+	-1 & \quad w.p. \enspace 1/2 \\
+	1  & \quad w.p. \enspace 1/2
+	\end{array}\right. \\
+	%
+	(b^{(k)}_{t,s^*}|X^*_{t,s^*}  = 1) &\sim {\rm{Gamma}}(1/k^2, 1) \\
+	%
+	(b^{(k)}_{t,s^*}|X^*_{t,s^*} = 2) &\sim \left\{\begin{array}{lr}
+	{\rm{Gamma}}(1/k^2, 1), \quad & k \notin \{1,2\} \\
+	{\rm{Gamma}}(100,1), & k = 1 \\
+	{\rm{Gamma}}(50,1), & k = 2
+	\end{array}\right. 
+\end{align*}
+
+\begin{align*}
+    \hat{Y}^{*(50)}_{t,s^*} &= 0 & \\
+	%
+	\hat{Y}^{*(k)}_{t,s^*}  &= -\hat{Y}^{*(100-n)}_{t,s^*}, & \qquad k = 51,\ldots,99
+\end{align*}
+
+$Y^*_{t,(100s^*+1):(100s^*+100)}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,s^*}$:
+
+$$Y^*_{t,(100s^*+1):(100s^*+100)} = IDFT\left(\hat{Y}^*_{t,s^*}\right), \qquad s^* = 0,\ldots,\lfloor Y_t/2 \rfloor - 1$$
+
+There are several practical reasons behind this construction.
+First, $\hat{Y}^*_{t,s^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,s^*}$ so that its inverse Fourier transform is real-valued.
+$\hat{Y}^{*(k)}_{t,s^*}$ also decays like $1/k^2$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window.
+Note, however, that $Y^*_{t,s^*}$ is not continuous \textit{between} windows, as can be seen in (fig \ref{fig:fourier_example}).
+However, the jumps are not too severe since $\hat{Y}^{*(0)}_{t,s^*}$ and $\hat{Y}^{*(0)}_{t,s^*+1}$ are highly correlated.
+See (fig.
+\ref{fig:sim_data}) for details.
+
+From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$ after setting $\tilde{f}$.
+We pick $\tilde{f} = 10$ periods per window, or 5 hertz.
+Note that $Z^{*(1)}_{t,s^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,s^*}\right)$, so selecting $\sigma^{*(i)}$ and $\phi^{*(1)}$ appropriately will ensure the distributions from (eqn \ref{}).
+
+Note that $Z^{*(2)}$, is the sum of gamma distributions with the same scale parameter, so the distribution of $Z^{*(2)}$ is also a Gamma distribution:
+
+$$Z^{*(2)} = \sum_{k=1}^{10} b^{(k)}_{t,s^*}$$
+$$Z^{*(2)} \sim {\rm{Gamma}}
+
+\iffalse
 
@@
 Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again.
-This allows for the pricipled construction of the continuous-time model to be combined with the computational convienence of the CarHMM.
+This allows for the principled construction of the continuous-time model to be combined with the computational convenience of the CarHMM.
 
-%\subsection{Proof of correlation between $d_{t-1}$ and $d_t$}
-
-%THIS IS WRONG
-%The correlation between $d_{t-1}$ and $d_t$ is defined as $\left(\mathbbm{E}(d_t d_{t-1}) - \mu_{RL,b}^2 \right) / \sigma_b^2$.
-Let the probability distribution function of $d_t$ and $d_{t-1}$ be denoted as $p$.
-Then:
-
-%\begin{align*}
-%	\mathbbm{E}(d_t d_{t-1}) &= \int_0^\infty \int_0^\infty x_t x_{t-1} p(x_{t-1}) p(x_t|x_{t-1}) dx_t dx_{t-1} \\
-	%
-%	&= \int_0^\infty x_{t-1} p(x_{t-1}) \left( \int_0^\infty x_t p(x_t|x_{t-1}) dx_t \right) dx_{t-1} \\
-	%
-%	&= \int_0^\infty x_{t-1} p(x_{t-1}) \left( (1-\phi_b)\mu_{RL,b} + \phi_b x_{t-1} \right) dx_{t-1} \\
-%	&= (1-\phi_b)\mu_{RL,b} \int_0^\infty x_{t-1} p(x_{t-1}) dx_{t-1} + \phi_b \int_0^\infty x_{t-1}^2 p(x_{t-1}) dx_{t-1} \\
-%	&= (1-\phi_b)\mu_{RL,b}^2 + \phi_b (\mu_{RL,b}^2 + \sigma_b^2) \\
-%	&= \mu_{RL,b}^2 + \phi_b\sigma_b^2
-%\end{align*}
-
-%So:
-
-%\begin{align*}
-%	corr(d_t,d_{t-1}) &= \frac{\mathbbm{E}(d_t d_{t-1}) - \mu_{RL,b}^2}{\sigma_b^2} \\
-%	&= \frac{\mu_{RL,b}^2 + \phi_b\sigma_b^2 - \mu_{RL,b}^2}{\sigma_b^2} \\
-%	&= \phi_b
-%\end{align*}
+\fi

### 2020-06-25 Report/Sections/background.tex
Update Report

@@
 
-Hidden Markov models are useful when inferring a single unobserved process, but biological processes often involve multiple simultaneous hidden processes which can occur and at different time scales.
-For example, a preliminary observation of the killer whale dive data shown in figure (\ref{fig:data}) shows that the behavior of this killer whale changes between approximately hour-long periods of predominately short, shallow dives and long, deep dives.
-Leos-Barajas et al.
-encounter a similar issue when modeling the movement of a harbor porpoise in the North Sea, and use it as a motivating example when they introduce hierarchical hidden Markov models.
-
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=5in]{../Plots/raw_data.png}
-	\caption{Raw depth data of a Killer Whale off the coast of British Columbia, Canada.}
-	\label{fig:data}
-\end{figure}
-
 \subsection{Hidden Markov models}
@@
 
-\textit{Hidden Markov models} (or HMMs) are comprised of a sequence of unobserved states, $X_t$ $(t = 1, \ldots, T)$ which form a Markov chain and a sequence of (possibly high-dimensional) observations $Y_t$ $(t = 1, \ldots, T)$ which are often referred to as ``emissions".
+\textit{Hidden Markov models} (or HMMs) are comprised of a sequence of $T$ unobserved states $X_t$ $(t = 1, \ldots, T)$ which forms a Markov chain and an associated sequence of $T$ (possibly high-dimensional) observations $Y_t$ $(t = 1, \ldots, T)$.
+These observations are often referred to as ``emissions".
 The index $t$ often represents equi-spaced jumps in time.
@@
 foraging, resting, migrating, etc.), while the observations are often a series of step lengths and turning angles for land animals and either depth or accelerometer data (or both) for marine animals.
-Each random variable in the unobserved chain $X_t$ can take one of $N$ possible values, and $X$ has corresponding initial distribution $\delta \in \bbR^N$ and probability transition matrix $\Gamma \in \bbR^{N \times N}$:
+Each random variable in the unobserved chain $X_t$ can take one of $N$ possible values, and $X$ has initial distribution $\delta \in \bbR^N$ and probability transition matrix $\Gamma \in \bbR^{N \times N}$.
+In particular:
 %
@@
 %
-We assume that $\delta$ is the stationary distribution of the $\Gamma$, implying that the hidden process runs infinitely before observations are taken.
+We assume that $\delta$ is the stationary distribution of $\Gamma$, implying that the hidden process runs infinitely before observations are recorded.
 Further, the distribution of each observation $Y_t$ depends only on the value of $X_t$ and none of the preceding observations or behavioral states, i.e.
 $p(y_t|x_1,\ldots, x_N, y_1,\ldots,y_{t-1},y_{t+1},\ldots,y_N) = p(y_t|x_t)$.
-In particular, we denote the distribution of $Y_t$ given $X_t = i$ as $f^{(i)}(y_t)$.
-$f^{(1)},\ldots,f^{(N)}$ in turn depend upon a vector of parameters $\Theta = (\theta^{(1)},\ldots,\theta^{(N)})$ and hereafter are referred to as ``emission distributions".
-A visualization of the dependence structure can be seen in figure (\ref{fig:HMM}).
-In order to estimate the parameters of an HMM, suppose data $y = (y_1,\ldots,y_T)$ is observed as a realization of an unknown HMM model.
-The probability transition matrix $\Gamma$, and parameters of the emission distributions $\Theta$ can be estimated by maximizing the likelihood of $y$, which we denote as $\calL_{\text{HMM}}(y;\Theta,\Gamma)$, with respect to $\Gamma$, $\delta$, and $\Theta$.
-In addition, $\calL_{\text{HMM}}(y;\theta,\Gamma,\delta)$ can be calculated using the \textit{forward algorithm} \cite{Zucchini:2016} as follows:
+In particular, we denote the distribution of $Y_t$ given that $X_t = i$ as $f^{(i)}(y_t)$.
+$f^{(1)},\ldots,f^{(N)}$ are described by a vector of parameters $\Theta = (\theta^{(1)},\ldots,\theta^{(N)})$ and hereafter are referred to as ``emission distributions".
+A visualization of the dependence structure can be seen in (fig \ref{fig:HMM}).
+In order to estimate the parameters of an HMM, suppose data $y = (y_1,\ldots,y_T)$ are observed as a realization of an unknown HMM model.
+The probability transition matrix $\Gamma$ and parameters of the emission distributions $\Theta$ can be estimated by maximizing the likelihood of $y$, which we denote as $\calL_{\text{HMM}}(y;\Theta,\Gamma)$, with respect to $\Gamma$ and $\Theta$.
+In addition, $\calL_{\text{HMM}}(y;\theta,\Gamma)$ can be calculated using the well-known \textit{forward algorithm} \cite{Zucchini:2016} as follows:
 %
-$$\calL_{\text{HMM}}(y;\theta,\Gamma) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}$$
+$$\calL_{\text{HMM}}(y;\theta,\Gamma) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N$$
 %
@@
 %
-$$\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{j'=1}^N \exp(\eta_{i(j')})}, \qquad \eta_{ii} = 0 \quad \forall i \in \{1, \ldots, N\}.$$
+$$\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{j'=1}^N \exp(\eta_{i(j')})}, \qquad \eta_{ii} = 0 \quad i = 1, \ldots, N$$
 %
-$\eta_{ii}$ is set to zero for identifiability. $\calL_{\text{HMM}}(y;\theta,\Gamma)$ can be maximized using any numerical optimizer.
+$\eta_{ii}$ is set to zero for identifiability, and $\calL_{\text{HMM}}(y;\theta,\Gamma)$ can be maximized using any numerical optimizer.
 
@@
 
-A hierarchical hidden Markov model (or HHMM) is a variation of a hidden Markov model in which $X_t$ satisfies the conditions of section \ref{subsection:HMMs}, but in addition it produces yet another fine-scale hidden Markov model of length $T_t^*$.
-This fine-scale hidden Markov model can be either instead of \cite{Barajas:2017} or in addition to \cite{Adam:2019} a coarse-scale observation $Y_t$ which must satisfy the conditions of section \ref{subsection:HMMs}.
+Hidden Markov models are useful when inferring a single unobserved process, but biological processes often involve multiple simultaneous hidden processes which can occur and at different time scales.
+For example, a preliminary observation of the killer whale dive data shown in (fig \ref{fig:data}) shows that the behavior of the killer whale changes between approximately hour-long periods of predominately short, shallow dives and long, deep dives.
+Leos-Barajas et al.
+\cite{Barajas:2017} encounter a similar issue when modeling the movement of a harbor porpoise in the North Sea, and use it as a motivating example when they introduce hierarchical hidden Markov models.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=5.5in]{../Plots/raw_data.png}
+	\caption{Raw depth data of a Killer Whale off the coast of British Columbia, Canada. \textcolor{red}{I WILL REFORMAT THIS PLOT LATER.}}
+	\label{fig:data}
+\end{figure}
+
+A hierarchical hidden Markov model (or HHMM) is a variation of a hidden Markov model in which $X_t$ satisfies the conditions of section \ref{subsection:HMMs}, but in addition it ``emits" yet another fine-scale hidden Markov model of length $T_t^*$.
+This fine-scale hidden Markov model can be either instead of \cite{Barajas:2017} or in addition to \cite{Adam:2019} a coarse-scale observation $Y_t$, which itself must satisfy the conditions from section \ref{subsection:HMMs}.
 The fine-scale HMM is comprised of a fine-scale Markov chain $X^*_t = (X^*_{t,1}, \ldots, X^*_{t,T^*_t})$ and fine-scale observations $Y^*_t = (Y^*_{t,1}, \ldots, Y^*_{t,T^*_t})$, each of length $T^*_t$.
 $X^*_t$ can take one of $N^*$ values.
-Note that while $N^*$ can be set to depend upon $X_t$, we assume here that it does not for simplicity in both the model and the notation.
+Note that while $N^*$ can be set to depend upon $X_t$, we assume here that it does not for simplicity in both the model presented here and the notation.
 Conditioned on the coarse-scale hidden state $X_t = i$, $X^*_t$ is characterized by an initial distribution $\delta^{*(i)} \in \bbR^{N^*}$ and probability transition matrix $\Gamma^{*(i)} \in \bbR^{N^* \times N^*}$:
@@
 Again, although not necessary, we assume that the initial distribution $\delta^{*(i)}$ is the stationary distribution of $\Gamma^{*(i)}$.
-In total, there are $N$ different values that $\Gamma^{*(i)}$ and $\delta^{*(i)}$ can take- one for each possible value of $X_t$.
+In total, there are $N$ different values that $\Gamma^{*(i)}$ and $\delta^{*(i)}$ can take- one for each possible value of the coarse-scale process $X_t$.
 We denote the sets of all fine-scale initial distributions and probability transition matrices as $\delta^*$ and $\Gamma^*$, respectively, where $|\delta^*| = |\Gamma^*| = N$.
@@
 Conditioned on the fact that $X_t = i$, the fine-scale distributions $f^{*(i,1)}, \ldots, f^{*(i,N^*)}$ are parameterized by an $N^*$-dimensional vector of parameters $\Theta^{*(i)} = (\theta^{*(i,1)},\ldots,\theta^{*(i,N^*)})$, where $\theta^{*(i,i^*)}$ describes the distribution $f^{*(i,i^*)}$.
-Depending upon the behaviour and responses being modelled, it is possible to force the fine-scale emission distributions to be equal across all values of $i \in \{1, \ldots, N\}: \enspace \theta^{*(1,i^*)} = \theta^{*(2,i^*)} = \ldots = \theta^{*(N,i^*)}$.
-A visualization of the full structure of the HHMM can be seen in figure (\ref{fig:HHMM}).
+Depending upon the behaviour and responses being modelled, it may be desirable to force the fine-scale emission distributions to be equal across all values of $i \in \{1, \ldots, N\}: \enspace \theta^{*(1,i^*)} = \theta^{*(2,i^*)} = \ldots = \theta^{*(N,i^*)}$.
+A visualization of the full structure of the HHMM can be seen in (fig \ref{fig:HHMM}).
 
@@
 
-One of the key assumptions of both HMMs and HHMMs is \textit{conditional independence} between observations at both the coarse and fine scale.
-Namely, given the state $X_t$, $Y_t$ is assumed to be independent from all other observations.
+One of the key assumptions of both HMMs and HHMMs is \textit{conditional independence} between observations on both the coarse and fine scale.
+Namely, when given the state $X_t$, $Y_t$ is assumed to be independent from all other observations, including $Y_{t-1}$.
 Therefore, traditional HMMs and HHMMs can fail when the observations $Y$ exhibit significant correlation in time.
@@
 One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}.
-In particular, the CarHMM describes the distribution of $(Y_t|Y_{t-1}, X_t=s)$ with parameters $\theta^{(s)} = \{\mu^{(s)},\sigma^{2(s)},\phi^{(s)}\}$, where:
+In particular, the CarHMM describes the distribution of $(Y_t|Y_{t-1}, X_t=i)$ with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{2(i)},\phi^{(i)}\}$, where:
 %
@@
 %
-The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm and only adding one additional parameter per hidden state $(\phi^{(i)})$.
+The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm.
+It also only adds one additional parameter per possible hidden state $(\phi^{(i)})$.
 
-The likelihood of CarHMM is still compatible with the forward algorithm:
+The likelihood of CarHMM is still compatible with the forward algorithm, and is therefore easy to calculate:
 \begin{equation}
@@
 While both the Viterbi algorithm and the forward-backward algorithm are used in the current ecology literature, we use the \textit{forward-backward algorithm} in this work for several reasons.
-In particular, the forward-backward algorithm is used to find the \textit{psuedoresiduals} of a given model, which are an important tool for model validation.
+First, the forward-backward algorithm is used to find the \textit{psuedoresiduals} of a given model, which are an important tool for model validation.
 In addition, the forward-backward algorithm can be implemented recursively to find the probability of the fine-level states $X^*_{t^*,t}$ exactly by marginalizing out $X_t$:
@@
 %
-where $P(X_t = i)$ can be found using the forward-backward algorithm on the coarse-level Markov chain and $P(X^*_{t^*,t} = i^* | X_t = i)$ can be found by running the forward-backward algorithm on the fine-level HMM for every possible value of $i = 1, \ldots, N$.
+where $P(X_t = i)$ can be found using the forward-backward algorithm on the coarse-level Markov chain and $P(X^*_{t^*,t} = i^* | X_t = i)$ can be found by running the forward-backward algorithm on the fine-scale HMM.

### 2020-06-25 Report/Sections/carhmm_ou.tex
Update Report

@@
 
-Although the CarHMM can incorporate auto-correlation into the structure of an HMM, they break down when observations are taken at irregular time intervals since the probability transition matrix $\Gamma$ would then depend upon the time step between observations.
+Although the CarHMM can incorporate auto-correlation into the structure of an HMM, they break down when observations are taken at irregular time intervals since the probability transition matrix $\Gamma$ would then have to depend upon the time step between observations.
 A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal.
@@
 One notable difference between the state-switching OU process and the CarHMM is that the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer.
-If the behavioral state $X_t$ is known and does not change between observations, the solution to Equation (\ref{eqn:SDE}) is known to be the following \cite{Michelot:2019}:
+If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \cite{Michelot:2019}:
 %
@@
 Continuous time models can be built up from arbitrarily complex stochastic differential equations and allow for uneven step lengths in the observations sequence $Y$.
-However, most continuous time models require MCMC algorithms to perform inference over and as a result are not easily incorporated into the HHMM structure.
+However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.
 
@@
 
-Although most continuous time models are difficult to incorporate into an HHMM, we show in theorem (1) that under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process shown in equation (\ref{eqn:OU_sol}).
+Although most continuous time models are difficult to incorporate into an HHMM, we show in (thrm 1) that under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process shown in (eqn \ref{eqn:OU_sol}).
 This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.
@@
 \end{enumerate}
-then the observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{2(i)},\phi^{(i)}\}$, where:
+then the observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
 
 \begin{align}
-\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{2(i)} = \frac{\omega^{(i)}^2}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t}), \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
+\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{(i)} = \sqrt{\frac{\omega^{(i)^2}}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t})}, \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
 \end{align}

### 2020-06-25 Report/Sections/intro.tex
Update Report

@@
 \section{Introduction}
+
+The field of animal movement is in the midst of a ``data renaissance" where advancements in tagging technology have given rise to an explosion of data available for statistical modeling.
+In particular, tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
+This results in a vast amount of data which often exhibits many different simultaneous behavioral processes occurring at different time scales.
+
+One solution to this issue is to use a hierarchical hidden Markov model (HHMM).
+HHMMs model the entire system as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.
+One nice property of HHMMs is that its likelihood is relatively easy to compute, facilitating fast maximum likelihood estimates for its associated parameters.
+
+At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by a traditional HMMs.
+To address this issue, it is possible to model small-scale animal behavior as the solution to some stochastic differential equation, but these methods tend to be computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC).
+
+This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency.
+We bridge the gap between discrete hidden Markov models and continuous-time stochastic process models by showing that the two are equivalent under certain conditions.
+In addition, we extract features from highly structured sub-dive behaviors using signal processing techniques.
+These features otherwise could not be modeled with a simple HMM.
+Finally, we apply our method to dive data collected from two Northern resident killer whales off the coast of British Columbia, Canada.
 

### 2020-06-25 Report/Sections/model.tex
Update Report

@@
 
-\subsection{Short-time Fourier Transform on Fine-scale process}
+\subsection{Short-time Fourier Transform on Fine-scale processes}
 
@@
 However, there are many animal movement processes which violate this Markov property on very fine scales.
-For example, swimming behaviour of marine mammals often exhibits periodic behavior since the animal repeatedly flukes to propel itself forward.
-Work has been done in the past to model non-Markovian dynamics in the \textit{behavioural} process $X^*_t$ \cite{Langrock:2012}, but addressing non-Markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area (MAYBE?).
+For example, swimming behaviour of marine mammals is often exhibits periodic since the animal repeatedly flukes (swims up and down) to propel itself forward.
+Work has been done in the past to model non-Markovian dynamics in the \textit{behavioural} process $X^*_t$ \cite{Langrock:2012}, but to our knowledge addressing non-Markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area \textcolor{red}{(MAYBE?)}.
 With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine scale behavior is likely to persist.
 
-To address this issue, we recommend borrowing techniques from signal processing literature to compress the data and summarize its essential elements.
+We borrow techniques from signal processing literature to compress the data and summarize its essential elements.
 In particular, we suggest performing the discrete-time short-time Fourier transform (STFT) over the observed fine-scale process $Y^*_t$:
@@
 \begin{align*}
-    STFT\{Y^*_{t,(s^*w)+1:(s^*+1)w) }\}(k) := \hat{Y}^{*(k)}_{t,s^*} = \sum_{n = 1}^{w} Y^*_{t,(s^*w)+n}e^{-i \frac{2\pi k}{w} (n-1)} \\ k = 0, 1, \ldots, w-1; \quad s^* = 0,1, \ldots, \lfloor T^*_t / w \rfloor - 1
+    STFT\{Y^*_{t,(s^*h+1):(s^*h+h) }\}(k) := \hat{Y}^{*(k)}_{t,s^*} = \sum_{n = 1}^{h} Y^*_{t,(s^*h)+n}e^{-i \frac{2\pi k}{h} (n-1)} \\ k = 0, 1, \ldots, h-1; \quad s^* = 0,1, \ldots, \lfloor T^*_t / h \rfloor - 1
 \end{align*}
 %
-Where $i$ in the equation above refers the $\sqrt{-1}$. Note here that $s^*$ indexes the \textit{windows} moving across $Y^*_t$ while $t^*$ indexes the \textit{observations} of $Y^*_t$.
+Where $i$ in the equation above refers the $\sqrt{-1}$.
+The STFT slides a moving window of length $h$ across the time series $Y_t^*$ and transforms the domain of each window from time to frequency.
+This allows the spectrum of $Y_t^*$ at time $t^*$ to be summarized by a $h$-dimensional vector of Fourier coefficients.
+While other step sizes can be used for the sliding window, we select $h$ to avoid serial dependence between windows.
+Note here that $s^*$ indexes the \textit{windows} moving across $Y^*_t$ while $t^*$ indexes the \textit{observations} of $Y^*_t$.
 
-The STFT slides a moving window of length $w$ across the time series $Y_t^*$ and transforms the domain of each window from time to frequency.
-This allows the spectrum of $Y_t^*$ at time $t^*$ to be summarized by a $w$-dimensional vector of Fourier coefficients.
-While other step sizes can be used for the sliding window, we select $w$ to avoid serial dependence between windows.
-
-If $Y^*_t \in \mathbb{R}^{T^*_t}$, then $\hat{Y}_t^* \in \mathbb{C}^{\lfloor T^*_t / w \rfloor \times w}$.
+If $Y^*_t \in \mathbb{R}^{T^*_t}$, then $\hat{Y}_t^* \in \mathbb{C}^{\lfloor T^*_t / h \rfloor \times h}$.
 Although this allows $Y^*_t$ to be represented in a way that eliminates obvious periodic behavior, the data set itself is still approximately as large as $Y^*_t$ itself.
@@
 %
-$$Z_{t,s^*}^{*(1)} = \mathcal{R}\left(\hat{Y}^{*(0)}_{t,s^*}\right) \qquad Z_{t,s^*}^{*(2)} = \frac{1}{w}\sum_{k=1}^{\tilde{f}}|\hat{Y}^{*(k)}_{t,s^*}|^2$$
+$$Z_{t,s^*}^{*(1)} = \mathcal{R}\left(\hat{Y}^{*(0)}_{t,s^*}\right) \qquad Z_{t,s^*}^{*(2)} = \frac{1}{h}\sum_{k=1}^{\tilde{f}}|\hat{Y}^{*(k)}_{t,s^*}|^2$$
 %
 $Z_{t,s^*}^{*(1)}$ is equal to the average value of $Y_{t,s^*}^*$, and $Z_{t,s^*}^{*(2)}$ is equal to the squared 2-norm of the component of $Y_{t,s^*}^*$ that can be attributed to frequencies in the signal between $1$ and $\tilde{f}$ periods per window length.
-Both the window length $w$ and the max frequency $\tilde{f}$ are problem-specific tuning parameters.
-$w$ should be long enough to capture the periodic behavior of the underlying process (at least as long as the length of a period), but short enough to avoid over-smoothing of the data and to maintain high resolution in the behavioral process $X^*$.
-$\tilde{f}$ should be selected such that the maximum frequency of $Y_t^*$ that makes biological sense is $\tilde{f}$ per window length.
-Note that these summary statistics are just one possible choice to describe each window.
-Other choices of summary statistics include the maximum and argmax of $\hat Y^*_{s^*,t}$.
+Both the window length $h$ and the max frequency $\tilde{f}$ are problem-specific tuning parameters.
+$h$ should be long enough to capture the periodic behavior of the underlying process (at least as long as the length of a period), but short enough to avoid over-smoothing of the data and to maintain high resolution in the behavioral process $X^*$.
+$\tilde{f}$ should be selected such that $\tilde{f}$ periods per window length is the maximum frequency of $Y_t^*$ that makes biological sense.
+Note that these summary statistics are just one possible choice to describe each window, and other choices include the maximum and argmax of $\hat Y^*_{t,s^*}$.
 Future studies can adjust these definitions as needed.
@@
 	\centering
-	\includegraphics[width=5in]{../Plots/fourier_transform.png}
-	\caption{Visualization of transforming $Y^*_t$ into $Z^*_t$ using a sliding window and fourier transform.}
+	\includegraphics[width=5.5in]{../Plots/fourier_transform.png}
+	\caption{Visualization of transforming $Y^*_t$ into $Z^*_t$ using a sliding window and Fourier transform.}
 	\label{fig:fourier_example}
@@
 
-Fourier analysis has previously been used in the field of animal movement to explain animal behavior \cite{Fehlmann:2017} and specifically fluking \cite{Shorter:2017} from accelerometer data, and Fourier analysis within an HMM has been used to describe daily behavioral cycles of marine mammals  \cite{Heerah:2017}, but we are currently unaware of any studies which have incorporated Fourier analysis of accelerometer data within the structure of an HMM.
+Fourier analysis has previously been used in the field of animal movement to explain animal behavior \cite{Fehlmann:2017} and specifically fluking \cite{Shorter:2017} from accelerometer data, and Fourier analysis within an HMM has been used to describe daily behavioral cycles of marine mammals  \cite{Heerah:2017}.
+However, we are currently unaware of any studies which have incorporated Fourier analysis of accelerometer data within the structure of an HMM.
 
@@
 
-Hierarchical hidden Markov models can be used to jointly model simultaneous coarse-scale and fine-scale processes taking place simultaneously.
+Hierarchical hidden Markov models can be used to jointly model simultaneous coarse-scale and fine-scale processes.
 However, as mentioned before, the fine-scale process $Y^*$ can often exhibit auto-correlation and intricate structure.
 Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit auto-correlation, especially in $Z_{t,s^*}^{*(1)}$.
-Therefore, we replace the fine-scale HMM within the Hierarchical HMM with a CarHMM, as shown in figure (\ref{fig:CarHHMM}).
-
+Therefore, we replace the fine-scale HMM within the hierarchical HMM with a CarHMM as shown in (fig \ref{fig:CarHHMM}).
+%
 \begin{figure}[h!]
@@
 \end{figure}
-
+%
+Naturally, we refer to this model as the \textbf{CarHHMM}, and it can be implemented both with and without the short-time Fourier-transform (STFT).
 The likelihood of this model is still easy to calculate using the forward algorithm:

### 2020-06-25 Report/Sections/simulation.tex
Update Report

@@
 
-%!% Have you simulated this only once? I think you would want to do this multiple time and present results across simulations, no?
-
 To test the CarHHMM with STFT, 500 separate sequences of 100 marine-mammal dives were simulated.
 The coarse-scale observations $Y$ were set as the duration of each dive, and the fine-scale observations $Y^*$ were set as one-dimensional acceleration readings simulated at 50 hertz.
-the specific procedure used to generate data was as follows:
+The number of possible states for both the coarse- and fine- scale Markov chains was set to 2, i.e.
+$N = N^* = 2$.
 
-\begin{enumerate}
-	\item 100 dive durations were simulated using an HMM generative model with the following parameters:
-
+On the coarse scale, the following parameters were used:
+$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix} $$
+$$Y_t|X_t \sim \rm{Gamma} $$
 \begin{align*}
-		&\Gamma = \begin{pmatrix} 0.4 & 0.6 \\ 0.6 & 0.4 \end{pmatrix} & \\
-		% 
-		&Y_t|X_t \sim \rm{Gamma} & \\
 	&\bbE(Y_t|X_t = 1) = 15 s &\bbE(Y_t|X_t = 2) = 60 s \\
@@
 \end{align*}
-
-	\item Once the dive durations were calculated, for each $t \in \{1, \ldots, 100\}$, dive $t$ was broken into $\lfloor Y_t/2 \rfloor$ 2-second segments (the end of the dive sequence was discarded).
-Further, each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where $X^*_{t,s^*} \in \{1,2\}$ and $s^* \in \{0,1,\ldots,\lfloor Y_t/2 \rfloor\}$.
-The parameters of the fine-scale behaviour Markov chain were set to be as follows:
+Once the coarse-scale dive durations were calculated for $t = 1, \ldots, 100$, dive $t$ was broken into $\lfloor Y_t/2 \rfloor$ 2-second segments (the end of the dive sequence was discarded).
+Each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where the parameters were set as follows:
 %
-	$$\Gamma^{*(1)} = \begin{pmatrix} 0.25 & 0.75 \\ 0.75 & 025 \end{pmatrix} \qquad 	\Gamma^{*(2)} = \begin{pmatrix} 0.75 & 0.25 \\ 0.25 & 0.75 \end{pmatrix}$$
+$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.9 & 0.1 \end{pmatrix} \qquad 	\Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}$$
 %
-	where $\Gamma^{*(1)}$ was used for dives where $X_t = 1$ and $\Gamma^{*(2)}$ was used for dives where $X_t = 2$
-
-	\item For each 2-second segment, the Fourier modes $\hat{Y}^*_{t,s^*}$ were simulated using the following procedure.
-Note that the $k^{th}$ Fourier mode of $\hat{Y}^*_{t,s^*}$ is denoted as $\hat{Y}^{*(k)}_{t,s^*}$ :
+where $\Gamma^{*(1)}$ was used for dives where $X_t = 1$ and $\Gamma^{*(2)}$ was used for dives where $X_t = 2$.
+For each 2-second segment, the Fourier modes $\hat{Y}^*_{t,s^*} \in \mathbb{C}^{100}$ were simulated such that $Z^{*(1)}_{t,s^*}$ and $Z^{*(2)}_{t,s^*}$ would have the following distributions:
 
 \begin{align*}
-    	\hat{Y}^{*(0)}_{t,0} &\sim \mathcal{N} \left(\mu = 0, \sigma^2 = 0.0025 \right) & \\
+    \left(Z^{*(1)}_{t,s^*} | Z^{*(1)}_{t,s^*-1}, X^*_{t,s^*} = 1 \right) &\sim \mathcal{N} \left(\phi^{*(1)}Z^{*(1)}_{t,s^*-1} + \phi^{*(1)}\mu^{*(1)}, (\sigma^{*(1)})^2 \right) \\
     %
-    	\hat{Y}^{*(0)}_{t,s^*} &\sim \mathcal{N} \left(\mu = 0.9Y^{*(0)}_{t,s^*-1}, \sigma^2 = 0.0025 \right), & s^* = 1,2,\ldots, \lfloor Y_t/2 \rfloor \\
+    \left(Z^{*(1)}_{t,s^*} | Z^{*(1)}_{t,s^*-1}, X^*_{t,s^*} = 2 \right) &\sim \mathcal{N} \left(\phi^{*(2)}Z^{*(1)}_{t,s^*-1} + \phi^{*(2)}\mu^{*(2)}, (\sigma^{*(2)})^2 \right) \\
     %
-    	\hat{Y}^{*(k)}_{t,t^*} &= a_{t,s^*}^{(k)} i\sqrt{b^{(k)}_{t,s^*}}, & k = 1,\ldots,49 \\
-	\end{align*}
-	\begin{align*}
-	    a_{t,s^*}^{(k)} &\sim  \left\{\begin{array}{lr}
-    	-1 & \quad w.p. \enspace 1/2 \\
-    	1  & \quad w.p. \enspace 1/2
-    	\end{array}\right. \\
+    \mu^{*(1)} = 0.0,\enspace \sigma^{*(1)} &= 0.05,\enspace \phi^{*(1)} = 0.99 \\
     %
-    	(b^{(k)}_{t,s^*}|X^*_{t,s^*}  = 1) &\sim {\rm{Gamma}}(1/k^2, 1) \\
+    \mu^{*(2)} = 0.0,\enspace \sigma^{*(2)} &= 0.1,\enspace \phi^{*(2)} = 0.95 \\\\
     %
-    	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
-    	{\rm{Gamma}}(1/k^2, 1), \quad & k \notin \{1,2\} \\
-    	{\rm{Gamma}}(100,1), & k = 1 \\
-    	{\rm{Gamma}}(50,1), & k = 2
-    	\end{array}\right. 
-	\end{align*}
-
-    \begin{align*}
-        \hat{Y}^{*(50)}_{t,s^*} &= 0 & \\
+    \left(Z^{*(2)}_{t,s^*} | X^*_{t,s^*} = 1\right) &\sim {\rm{Gamma}}\left(\alpha^{*(1)}, \beta^{*(1)} \right) \\
     %
-    	\hat{Y}^{*(k)}_{t,s^*}  &= -\hat{Y}^{*(100-n)}_{t,s^*}, & \qquad k = 51,\ldots,99
-    \end{align*}
-
-	\item Finally, $Y^*_{t,(100s^*+1):(100s^*+100)}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,s^*}$:
-	$$Y^*_{t,(100s^*+1):(100s^*+100)} = IDFT\left(\hat{Y}^*_{t,s^*}\right)$$
-
-\end{enumerate}
-
-There are several practical reasons behind this construction.
-First, $\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,s^*}$ so that its inverse Fourier transform is real-valued.
-$\hat{Y}^{*(k)}_{t,s^*}$ also decays like $1/k^2$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window.
-Note, however, that $Y^*_{t,s^*}$ is not continuous \textit{between} windows, as can be seen in figure (\ref{fig:fourier_example}).
-However, the jumps are not too severe since $\hat{Y}^{*(0)}_{t,s^*}$ and $\hat{Y}^{*(0)}_{t,s^*+1}$ are highly correlated.
-See figure (\ref{fig:sim_data}) for details.
-
-\subsection{Ground-Truth parameters}
-
-This generative model gives simulated data where the ground-truth of the CarHHMM is known. For the coarse scale, the ground truth is as follows:
-
-\begin{align*}
-    \delta &= \begin{pmatrix} 0.5 & 0.5 \end{pmatrix}\\
-	\Gamma &= \begin{pmatrix} 0.4 & 0.6 \\ 0.6 & 0.4 \end{pmatrix} \\
+    \left(Z^{*(2)}_{t,s^*} | X^*_{t,s^*} = 2\right) &\sim {\rm{Gamma}}\left(\alpha^{*(2)}, \beta^{*(2)} \right) \\
     %
-	Y_t|X_t = 1 &\sim {\rm{Gamma}}\left(\mu = 15 s, \sigma = 5 s\right) \\
-	Y_t|X_t = 2 &\sim {\rm{Gamma}}\left(\mu = 60 s, \sigma = 10 s\right) \\
-\end{align*}
-
-For the fine-scale, we have:
-
-\begin{align*}
-    \delta^{*(1)} &= \begin{pmatrix} 0.5 & 0.5 \end{pmatrix} & \delta^{*(2)} &= \begin{pmatrix} 0.5 & 0.5 \end{pmatrix} \\
+    \alpha^{*(1)} = 10.10, \enspace & \enspace \beta^{*(1)} = 1.00 \\
     %
-    \Gamma^{*(1)} &= \begin{pmatrix} 0.25 & 0.75 \\ 0.75 & 025 \end{pmatrix} & 	\Gamma^{*(2)} &= \begin{pmatrix} 0.75 & 0.25 \\ 0.25 & 0.75 \end{pmatrix} \\
-\end{align*}
-
-\begin{align*}
-    \left(Z^{*(1)}_{t,s^*} | Z^{*(1)}_{t,s^*-1} \right) &\sim \mathcal{N} \left(\phi^*Z^{*(1)}_{t,s^*-1} + \phi^*\mu^*, \sigma^{*2} \right), & \mu^* = 0, \enspace \phi^* = 0.9, \enspace \sigma^{*2} = 0.025 \\
-    %
-    \left(Z^{*(2)}_{t,s^*} | X^*_{t,s^*} = 1\right) &\sim {\rm{Gamma}}\left(\alpha^{*(1)}, \beta^* \right), & \alpha^{*(1)} = \sum_{k=1}^{\tilde{f}} \frac{1}{k^2}, \enspace \beta^* = 1 \\
-    %
-    \left(Z^{*(2)}_{t,t^*} | X^*_{t,s^*} = 2\right) &\sim {\rm{Gamma}}\left(\alpha^{*(2)}, \beta^* \right), & \alpha^{*(2)} = 20 + \sum_{k=2}^{\tilde{f}} \frac{1}{k^2}, \enspace \beta^* = 1 \\
+    \alpha^{*(2)} = 305.94, \enspace & \enspace \beta^{*(2)} = 1.00
 \end{align*}
 %
-The emission distribution of the 2-second average acceleration $Z_{t,s^*}^{*(1)}$ can be reparameterized as an OU process according to Theorem 1:
+See the appendix for more details regarding the simulation procedure for $\hat Y^*$.
+The emission distribution of the 2-second average acceleration $Z_{t,s^*}^{*(1)}$ in particular can be reparameterized as an OU process according to theorem 1:
 %
-$$dZ^{*(1)}_{t,t^*} = \beta^*(\gamma^* - Z^{*(1)}_{t,t^*})dt^* + \omega^* dW_{t,t^*}$$
+$$dZ^{*(1)}_{t,s^*} = \beta^*(\gamma^* - Z^{*(1)}_{t,s^*})dt^* + \omega^* dW_{t,t^*}$$
+%
+$$\gamma^{*(1)} = 0, \enspace \beta^{*(1)} = -\frac{\log(\phi^{*(1)})}{\Delta t^*} = 0.0111, \enspace \omega^{*(1)} = \sqrt{\frac{2(\sigma^{^{*(1)}})^2\log(\phi^{*(1)})}{\Delta t^* ((\phi^{*(1)})^2-1)}} = 0.0355$$
+$$\gamma^{*(2)} = 0, \enspace \beta^{*(2)} = -\frac{\log(\phi^{*(2)})}{\Delta t^*} = 0.0256, \enspace \omega^{*(2)} = \sqrt{\frac{2(\sigma^{*(2)})^2\log(\phi^{*(2)})}{\Delta t^* ((\phi^{*(2)})^2-1)}} = 0.7625$$
 
-$$\gamma^* = \mu^* = 0, \qquad \beta^* = -\frac{\log(\phi^*)}{\Delta t^*} = 0.0229, \qquad \omega^{*2} = \frac{2\sigma^{*2}\log(\phi)}{\Delta t (\phi^{*2}-1)} = 0.006$$
+A visualization of the first 5 dives of one simulated data set can be seen in (fig. \ref{fig:sim_data}).
 
@@
 	\includegraphics[width=5in]{../Plots/sim_data.png}
-	\caption{Simulated Acceleration Data.}
+	\caption{Simulated acceleration data for one dive.
+The color of the line corresponds to the true fine-scale state of the sub-dive process, while the color of the background corresponds to the true dive type of the simulated whale, where blue corresponds to dive type 1 and orange corresponds to dive type 2.}
 	\label{fig:sim_data}
@@
 \begin{enumerate}
-    \item A CarHMM with no hierarchical component using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations (CarHMM).
-    \item A Hierarchical hidden Markov model with no auto-correlation using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations (HHMM).
-    \item A CarHHMM using $Z^{(1)}$, but not $Z^{(2)}$, as fine-scale observations (CarHHMM1).
-    \item A CarHHMM using both $Z^{(1)}$ and $Z^{(2)}$ as fine-scale observations (CarHHMM2).
+    \item A CarHMM with no hierarchical structure using both $Z^{*(1)}$ and $Z^{*(2)}$ as fine-scale observations (\textbf{CarHMM}).
+    \item An HHMM with no auto-correlation using both $Z^{*(1)}$ and $Z^{*(2)}$ as fine-scale observations (\textbf{HHMM}).
+    \item A CarHHMM using $Z^{*(1)}$, but NOT $Z^{*(2)}$ as fine-scale observations (\textbf{CarHHMM1}).
+To ensure that the CarHHMM could still see the ``wigliness" of the process, we halved the window size to 1 second for the CarHHMM1.
+    \item A CarHHMM using both $Z^{*(1)}$ and $Z^{*(2)}$ as fine-scale observations (\textbf{CarHHMM2}).
 \end{enumerate}
 %
-For each model, the following were recorded:
-%
-\begin{enumerate}
-    \item An empirical distribution of the MLE estimates of $\Gamma$ and $\Theta$ using the 100 generated data sets generated from true parameters $\Gamma_0$ and $\Theta_0$.
-    \item The estimated spread of these distributions using the observed information matrix.
-    \item A confusion matrix showing the state decoding capabilities of each model across all data sets.
-\end{enumerate}
+The purpose of including each of these models is to leave one important aspect of the full model (CarHHMM2) out of each of the preceding three and observe the effects on a simulated data set.
+In particular, the CarHMM lacks a hierarchical structure, the HHMM is missing auto-correlation within the fine-scale observations, and the CarHHMM1 does not have $Z^{*(2)}$ as observations.
 
-The empirical distribution of
+%%% Accuracy %%%
+Every model was able to decode the fine-scale hidden states of the process effectively perfectly except for the CarHHMM1, which did not have access to the Fourier modes $Z^{*(2)}$.
+This is intuitively clear because the fine-scale states are much more distinct in their Fourier modes than in their average acceleration.
+For the coarse-scale hidden state, the CarHMM lacked a hierarchical structure and could not make any predictions at all.
+The other three models all achieved an accuracy of approximately 90\%, with the CarHHMM slightly more likely to categorize a dive as dive type 2 than the other models.
+(Fig.
+\ref{fig:acc}) shows the decoded state probabilities of both the fine and coarse hidden states for the first 10 dives of one simulated data set.
+(Tbl.
+\ref{table:accuracy}) also lists more details regarding the accuracy and training times of each model for each true dive and sub-dive state.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\iffalse
-\subsection{Model Misspecification}
-
-There are two ways to simulate the fourier modes of the acceleration vector $\bfA_{s,t}$:
-
-\begin{enumerate}
-	\item $$\bbE_1\left(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}\right) = \phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} $$
-
-	\item $$\bbE_2(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}, X^*_{s-1,t}) = \left\{\begin{array}{lr}
-	\phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} & \text{for } X^*_{s,t} = X^*_{s-1,t}\\
-	\mu_{X^*_{s,t}} & \text{for } X^*_{s,t} \neq X^*_{s-1,t}
-	\end{array}\right\}$$
-\end{enumerate}
-
-\subsubsection{$\bbE_1$}
-
-First, look at $\bbE_1$.
-This comes with an issue in that when $X^*_{s,t}$ changes values, it will take a while for the process to ``converge" long-run mean $\mu_{X^*_{s,t}}$.
-In particular, if the behavioral process $X^*$ had been in state 1 for a while and then switches to state 2 for a while, then:
-
-\begin{align*}
-	\bbE_1\left(\bfA_{s+1,t}|X^*_{s:s+1,t} = 2, X^*_{s,t} = 1\right) &=  \phi \mu_1 + (1-\phi) \mu_2 \\
-	%
-	\bbE_1\left(\bfA_{s+2,t}|X^*_{s:s+2,t} = 2, X^*_{s,t} = 1\right) &= \phi\left( \phi \mu_1 + (1-\phi) \mu_2\right) + (1-\phi)\mu_2 \\
-	&= \phi^2 \mu_1 + (1-\phi^2) \mu_2\\
-	%
-	\bbE_1\left(\bfA_{s+n,t}|X^*_{s:s+n,t} = 2, X^*_{s,t} = 1\right) &=  \phi^n \mu_1 + (1-\phi^n) \mu_2 \\
-\end{align*}
-
-So, if using $\bbE_1$, there is an assumption that the fourier modes of the acceleration vector $\bfA$ linearly ``converge" to the new value $\mu_i$ from $\mu_j$ when $X^*_{s,t}$ changes from $i$ to $j$.
-An example of this can be seen in figure (\ref{fig:FOVEDBA}).
-
- \begin{figure}[h!]
+\begin{figure}[t!]
     \centering
- 	\includegraphics[height=3in]{../Code/test.png}
- 	\caption{The FoVeDBA of simulated data using $\bbE_1$. Note that the FoVeDBA approaches $\approx$ 300 in state 2 from $\approx$ 10 in state 1 with roughly linear convergence.}
- 	\label{fig:FOVEDBA}
+    \begin{subfigure}[t]{1.0\textwidth}
+        \centering
+        \includegraphics[width=5in]{Plots/Posterior_Coarse_States.png}
+        \caption{Coarse-scale hidden process}
+    \end{subfigure}
+    \newline
+    \begin{subfigure}[t]{1.0\textwidth}
+        \centering
+        \includegraphics[width=5in]{Plots/Posterior_Fine_States.png}
+        \caption{Fine-scale hidden process}
+    \end{subfigure}
+	\caption{Decoded state probabilities of each model.
+The color of the line corresponds to the true fine-scale state of the sub-dive process, while the color of the background corresponds to the true dive type of the simulated whale.}
+	\label{fig:acc}
 \end{figure}
 
- Note that if the generative model is correlated, but the inference model isn't, then quick jumps between states 1 and 2 can be missed because the chain never ``converges" to the new mean $\mu_i$, as shown in figure (\ref{fig:STATE}).
+%%% dive duration %%%
+For the emission distributions of dive duration, estimates of standard error using the observed fisher information tended to be underestimates due to correlation between $\hat \mu$ and $\hat \sigma$.
+In addition, $\hat \sigma$ tends to underestimate $\sigma$ for all models, which is a finding consistent with properties of MLEs, especially because the sample size for a particular dive type is approximately 50 for each simulation.
+The CarHMM model in particular severely underestimates $\sigma$.
+This is likely due to the fact that dive duration follows a mixture of gammas rather than a gamma distribution, and the CarHMM does not have the machinary to capture this fact.
+A full table of parameter estimates for all models is shown in (table \ref{table:dive_duration}).
 
-\begin{figure}[h!]
+%%% Acceleration %%%
+For acceleration ($Z^{*(1)}$), both the CarHMM and the CarHHMM2 regularly converged to the correct parameters with very little standard error.
+Note, however, that both have relatively large observed fisher standard error in $\hat \mu$.
+This likely is due to the very large autocorrelation terms ($\phi^{(1)} = 0.99$ and $\phi^{(2)} = 0.95$) since as $\phi \to 1$, the value of $\mu$ does not matter in the likelihood evaluation as $\phi \to 1$ implies that $\mathbb{E}(Z^{*(1)}_{t,s^*}) = \phi*(Z^{*(1)}_{t,s^*-1}) + (1-\phi)\mu \to Z^{*(1)}_{t,s^*-1}$.
+The HHMM overestimates the variance of $Z^{*(1)}$ since it does not incorporate auto-correlation into the emission distribution of $Z^{*(1)}$, and the CarHHMM1 has large biases in many of its parameter estimates, especially in subdive state 2, because it sees the ``wigliness" of the acceleration data as raw acceleration rather than as Fourier modes, which deflates the value of $\hat \phi$ and inflates the value of $\hat \sigma$.
+See (table \ref{table:acceleration}) for a detailed breakdown of parameter values for acceleration emission distributions.
+
+%%% FoVeDBA %%%
+
+Finally, the distribution of parameters for the Fourier sums ($Z^{*(2)}$) for all models (except for the CarHHMM1) converge to the correct values, and the observed fisher information standard errors are in line with the empirical standard error.
+The distribution of $Z^{*(2)}$ is very different between subdive states 1 and 2, so the signal is therefore very easy for each model to pick up on.
+(Table \ref{table:FoVeDBA}) shows a detailed breakdown of the emission distribution of $Z^{*(2)}$ for all models.
+
+\textcolor{red}{I also have similar tables for $\Gamma$ and $\Gamma^*$, and I also have plots corresponding to each table.
+For example, I have included figure \ref{fig:MLE_dist} as an example below for the emission distribution of the MLEs for dive duration of the CarHMM.
+Should I use those instead?
+But there are just SO many plots (4 models * 3 observations * 2 states = 24 plots)}
+
+\begin{figure}[!ht]
 	\centering
-	\includegraphics[height=3in]{../Code/test2.png}
-	\caption{The probability of being in each behavioral state $X^*_{s,t}$ as estimated by the HHMM. The colors represent the true behavioral state.}
-	\label{fig:STATE}
+	\includegraphics[width=5in]{../Plots/hmm_FV_MLE_density_dive_duration_-1.png}
+	\caption{\textcolor{red}{KDE plot of $\hat \mu$ and $\hat \sigma$ for the dive duration emission distribution for the CarHMM. This plot is for Marie and Nancy to look over.}}
+	\label{fig:MLE_dist}
 \end{figure}
 
- \subsubsection{$\bbE_2$}
+%%% Accuracy %%%
 
-Clearly, the model has issues decodeing the behavioral state when the model is misspecified.
-Therefore, if the true generative process does not have correlation when jumping between states, then it is better to use $\bbE_2$:
+\begin{table}[]
+\centering
+\caption{Accuracies and run times for all models. All reported values are averages, and $\pm$ refers to the standard deviation.}
+\scalebox{0.8}{
+\begin{tabular}{ccccccc}
+Model                      & \multicolumn{1}{c}{Training Time (Minutes)} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Dive Accuracy} & \multicolumn{1}{c}{Subdive Accuracy}  \\ \hline
+\multirow{5}{*}{CarHMM}    & \multirow{5}{*}{$15.74 \pm 2.46$}   & Both                          & Both                             & -------------                     & $1.00 \pm 0.00$                       \\
+                           &                                    & 1                             & 1                                & \multirow{2}{*}{-------------}    & $1.00 \pm 0.00$                       \\
+                           &                                    & 1                             & 2                                &                                   & $1.00 \pm 0.00$                       \\
+                           &                                    & 2                             & 1                                & \multirow{2}{*}{-------------}    & $1.00 \pm 0.00$                       \\
+                           &                                    & 2                             & 2                                &                                   & $1.00 \pm 0.00$                       \\ \hline
+\multirow{5}{*}{HHMM}      & \multirow{5}{*}{$82.43 \pm 11.48$}   & Both                          & Both                             & $0.94 \pm 0.02$                   & $1.00 \pm 0.00$                       \\
+                           &                                    & 1                             & 1                                & \multirow{2}{*}{$0.94\pm0.03$}    & $1.00 \pm 0.00$                       \\
+                           &                                    & 1                             & 2                                &                                   & $1.00 \pm 0.00$                       \\
+                           &                                    & 2                             & 1                                & \multirow{2}{*}{$0.94\pm0.03$}    & $1.00 \pm 0.00$                       \\
+                           &                                    & 2                             & 2                                &                                   & $1.00 \pm 0.00$                       \\ \hline
+\multirow{5}{*}{CarHHMM1}  & \multirow{5}{*}{$70.85 \pm 15.89$}   & Both                          & Both                             & $0.91 \pm 0.03$                   & $0.89 \pm 0.02$                       \\
+                           &                                    & 1                             & 1                                & \multirow{2}{*}{$0.87\pm0.04$}    & $0.44 \pm 0.12$                       \\
+                           &                                    & 1                             & 2                                &                                   & $1.00 \pm 0.00$                       \\
+                           &                                    & 2                             & 1                                & \multirow{2}{*}{$0.95\pm0.03$}    & $0.81 \pm 0.04$                       \\
+                           &                                    & 2                             & 2                                &                                   & $1.00 \pm 0.00$                       \\ \hline
+\multirow{5}{*}{CarHHMM2}  & \multirow{5}{*}{$81.22 \pm 16.10$}   & Both                          & Both                             & $0.94 \pm 0.02$                   & $1.00 \pm 0.00$                       \\
+                           &                                    & 1                             & 1                                & \multirow{2}{*}{$0.94\pm0.03$}    & $1.00 \pm 0.00$                       \\
+                           &                                    & 1                             & 2                                &                                   & $1.00 \pm 0.00$                       \\
+                           &                                    & 2                             & 1                                & \multirow{2}{*}{$0.94\pm0.03$}    & $1.00 \pm 0.00$                       \\
+                           &                                    & 2                             & 2                                &                                   & $1.00 \pm 0.00$                       \\ \hline
+\end{tabular}
+}
+\label{table:accuracy}
+\end{table}
 
-$$\bbE_2(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}, X^*_{s-1,t}) = \left\{\begin{array}{lr}
-\phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} & \text{for } X^*_{s,t} = X^*_{s-1,t}\\
-\mu_{X^*_{s,t}} & \text{for } X^*_{s,t} \neq X^*_{s-1,t}
-\end{array}\right\}$$
+%%% dive duration %%%
 
-This fixes the issue of having to ``converge" to the new value of $\mu_i$ when jumping to state $i$.
-However, note that $\bbE_2$ depends upon both $X^*_{s,t}$ and $X^*_{s-1,t}$.
-This breaks the assumption in the forward algorithm that $\bfA_{s,t}$ depends only on $X^*_{s,t}$.
-One way to deal with this issues is to introduce a new hidden state $Z^*_{s,t} = (X^*_{s-1,t},X^*_{s,t})$.
-This fixes the issue, but now $Z^*_{s,t}$ can take one of $s^2$ values instead of $s$.
-This grows the state space of our model and can make inference harder.
+\begin{table}[]
+\centering
+\caption{Estimates and standard errors of parameters for dive duration distribution for all four models.
+All reported values are averages, except for the Fisher observed standard error, which are medians.
+$\pm$ refers to the IQR.}
+\scalebox{0.9}{
+\begin{tabular}{ccccccc}
+Model                      & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}           \\ \hline
+\multirow{4}{*}{CarHMM}    & \multirow{2}{*}{$\mu$}        & 1                             & $49.72$                         & $-0.28$                     & $4.78$                             & $2.47 \pm 0.34$                             \\
+                           &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & $39.00$                         & $-7.51$                     & $5.05$                             & $2.50 \pm 0.40$                             \\
+                           &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\ \hline
+\multirow{4}{*}{HHMM}      & \multirow{2}{*}{$\mu$}        & 1                             & $19.99$                         & $-0.01$                     & $0.75$                             & $0.69 \pm 0.11$                             \\
+                           &                               & 2                             & $79.85$                         & $-0.15$                     & $8.05$                             & $5.85 \pm 1.10$                             \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & $4.90$                         & $-0.10$                     & $0.61$                             & $0.53 \pm 0.10$                             \\
+                           &                               & 2                             & $48.74$                         & $-1.26$                     & $6.50$                             & $5.15 \pm 1.02$                             \\ \hline
+\multirow{4}{*}{CarHHMM 1} & \multirow{2}{*}{$\mu$}        & 1                             & $19.91$                         & $-0.09$                     & $0.77$                             & $0.71 \pm 0.12$                             \\
+                           &                               & 2                             & $75.80$                         & $-4.20$                     & $7.72$                             & $5.32 \pm 0.98$                             \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & $4.73$                         & $-0.27$                     & $0.59$                             & $0.55 \pm 0.10$                             \\
+                           &                               & 2                             & $49.48$                         & $-0.52$                     & $6.26$                             & $4.79 \pm 0.93$                             \\ \hline
+\multirow{4}{*}{CarHHMM 2} & \multirow{2}{*}{$\mu$}        & 1                             & $19.99$                         & $-0.01$                     & $0.75$                             & $0.69 \pm 0.12$                             \\
+                           &                               & 2                             & $79.85$                         & $-0.15$                     & $8.05$                             & $5.85 \pm 1.10$                             \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & $4.90$                         & $-0.10$                     & $0.61$                             & $0.53 \pm 0.10$                             \\
+                           &                               & 2                             & $48.74$                         & $-1.26$                     & $6.50$                             & $5.15 \pm 1.02$
+\end{tabular}
+}
+\label{table:dive_duration}
+\end{table}
 
- \fi
+%%% Acceleration %%%
 
-%Figure (\ref{fig:smoothing}) also shows the calculated ODBA as a function of time for one a specific dive.
-The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs straightforwardly.
-As a result, the data was split into two-second intervals (100 data points each), and each interval was summarized by the Fourier transform of the ODBA and the velocity at the \textit{end} of the time interval.
-Velocity was taken at the end of the time interval because the behavior of the whale during the time interval has no impact on the velocity of the whale at the beginning of the time interval.
-Because velocity is highly correlated with the average acceleration within an interval, the mean was subtracted out of each ODBA time series before the Fourier transform was taken.
-To reduce the dimension of the ODBA Fourier transform, Fourier coefficients corresponding to frequencies less than or equal to 5 hertz were summed and frequencies above 5 hertz were ignored.
-One nice property of this process is that it reduces the number of data points by a factor of 100, which drastically speeds up parameter estimation.
-The optimal length of the time interval over which to take the Fourier transform is a tuning parameter that is difficult to find.
-An ideal time interval should be long enough give useful Fourier coefficients, but short enough to preserve as much information about the vertical velocity of the animal as possible.
+\begin{table}[]
+\centering
+\caption{Estimates and standard errors of parameters for $Z^{*(1)}_{t,s^*}$ for all four models.
+All reported values are averages, except for the Fisher observed standard error, which are medians.
+$\pm$ refers to the IQR.}
+\scalebox{0.8}{
+\begin{tabular}{ccccccc}
+Model                      & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}        \\ \hline
+\multirow{6}{*}{CarHMM}    & \multirow{2}{*}{$\mu$}        & 1                             & $0.00$                         & $0.00$                     & $0.00$                             & $0.14 \pm 0.13$                             \\
+                           &                               & 2                             & $0.00$                         & $0.00$                     & $0.01$                             & $0.06 \pm 0.02$                             \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & $0.05$                         & $-0.00$                     & $0.00$                             & $0.00 \pm 0.00$                             \\
+                           &                               & 2                             & $0.10$                         & $-0.00$                     & $0.00$                             & $0.00 \pm 0.00$                             \\
+                           & \multirow{2}{*}{$\phi$}       & 1                             & $0.99$                         & $-0.00$                     & $0.01$                             & $0.01 \pm 0.00$                             \\
+                           &                               & 2                             & $0.95$                         & $-0.00$                     & $0.01$                             & $0.01 \pm 0.00$                             \\ \hline
+\multirow{6}{*}{HHMM}      & \multirow{2}{*}{$\mu$}        & 1                             & $-0.01$                         & $-0.01$                     & $0.02$                             & $0.01 \pm 0.00$                             \\
+                           &                               & 2                             & $-0.00$                         & $-0.00$                     & $0.02$                             & $0.01 \pm 0.00$                             \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & $0.25$                         & $0.20$                     & $0.04$                             & $0.00 \pm 0.00$                             \\
+                           &                               & 2                             & $0.24$                         & $0.14$                     & $0.03$                             & $0.00 \pm 0.00$                             \\
+                           & \multirow{2}{*}{$\phi$}       & 1                             & ------                         & ------                     & ------                             & ------                                      \\
+                           &                               & 2                             & ------                         & ------                     & ------                             & ------                                      \\ \hline
+\multirow{6}{*}{CarHHMM 1} & \multirow{2}{*}{$\mu$}        & 1                             & $0.00$                         & $0.00$                     & $0.00$                             & $0.08 \pm 0.04$                             \\
+                           &                               & 2                             & $-0.01$                         & $-0.01$                     & $0.01$                             & $0.01 \pm 0.00$                             \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & $0.05$                         & $0.00$                     & $0.04$                             & $0.00 \pm 0.00$                             \\
+                           &                               & 2                             & $0.27$                         & $0.17$                     & $0.01$                             & $0.00 \pm 0.00$                             \\
+                           & \multirow{2}{*}{$\phi$}       & 1                             & $0.97$                         & $-0.02$                     & $0.10$                             & $0.00 \pm 0.00$                             \\
+                           &                               & 2                             & $0.49$                         & $-0.46$                     & $0.05$                             & $0.02 \pm 0.00$                             \\ \hline
+\multirow{6}{*}{CarHHMM 2} & \multirow{2}{*}{$\mu$}        & 1                             & $0.00$                         & $0.00$                     & $0.00$                             & $0.13 \pm 0.12$                             \\
+                           &                               & 2                             & $0.00$                         & $0.00$                     & $0.00$                             & $0.06 \pm 0.02$                             \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & $0.05$                         & $-0.00$                     & $0.00$                             & $0.00 \pm 0.00$                             \\
+                           &                               & 2                             & $0.10$                         & $-0.00$                     & $0.00$                             & $0.00 \pm 0.00$                             \\
+                           & \multirow{2}{*}{$\phi$}       & 1                             & $0.99$                         & $-0.00$                     & $0.01$                             & $0.01 \pm 0.00$                             \\
+                           &                               & 2                             & $0.95$                         & $-0.00$                     & $0.01$                             & $0.01 \pm 0.00$                             \\ \hline
+\end{tabular}
+}
+\label{table:acceleration}
+\end{table}
 
-%\subsection{Lag Plot}
+%%% FoVeDBA %%%
 
-%In order to find the appropriate HMM to model whale behavior, a lag plot was made for both velocity and the ODBA fourier sum.
-The results of doing so are shown in figure (\ref{fig:lag}).
-Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that the velocity exhibits a large degree of autocorrelation.
-While the ODBA Fourier sum also exhibits some autocorrelation, the relationship is less strong, so autocorrelation was not incorporated in the ODBA Fourier sums emission distribution.
+\begin{table}[]
+\centering
+\caption{Estimates and standard errors of parameters for $Z^{*(2)}_{t,s^*}$ for all four models.
+All reported values are averages, except for the Fisher observed standard error, which are medians.
+$\pm$ refers to the IQR.}
+\scalebox{0.8}{
+\begin{tabular}{ccccccc}
+Model                      & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}        \\ \hline
+\multirow{4}{*}{CarHMM}    & \multirow{2}{*}{$\mu$}        & 1                             & $10.10$                         & $-0.00$                     & $0.09$                             & $0.08 \pm 0.01$                             \\
+                           &                               & 2                             & $305.97$                         & $0.03$                     & $0.54$                             & $0.51 \pm 0.03$                             \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & $3.18$                         & $-0.00$                     & $0.07$                             & $0.06 \pm 0.01$                             \\
+                           &                               & 2                             & $17.46$                         & $-0.03$                     & $0.37$                             & $0.36 \pm 0.02$                             \\ \hline
+\multirow{4}{*}{HHMM}      & \multirow{2}{*}{$\mu$}        & 1                             & $10.10$                         & $-0.00$                     & $0.09$                             & $0.08 \pm 0.01$                             \\
+                           &                               & 2                             & $305.97$                         & $0.03$                     & $0.54$                             & $0.51 \pm 0.03$                             \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & $3.18$                         & $-0.00$                     & $0.07$                             & $0.06 \pm 0.01$                             \\
+                           &                               & 2                             & $17.46$                         & $-0.03$                     & $0.37$                             & $0.36 \pm 0.02$                             \\ \hline
+\multirow{4}{*}{CarHHMM 1} & \multirow{2}{*}{$\mu$}        & 1                             & ------                         & ------                     & ------                             & ------                                      \\
+                           &                               & 2                             & ------                         & ------                     & ------                             & ------                                      \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & ------                         & ------                     & ------                             & ------                                      \\
+                           &                               & 2                             & ------                         & ------                     & ------                             & ------                                      \\ \hline
+\multirow{4}{*}{CarHHMM 2} & \multirow{2}{*}{$\mu$}        & 1                             & $10.10$                         & $-0.00$                     & $0.09$                             & $0.08 \pm 0.01$                             \\
+                           &                               & 2                             & $305.97$                         & $0.03$                     & $0.54$                             & $0.51 \pm 0.03$                             \\
+                           & \multirow{2}{*}{$\sigma$}     & 1                             & $3.18$                         & $-0.00$                     & $0.07$                             & $0.06 \pm 0.01$                             \\
+                           &                               & 2                             & $17.46$                         & $-0.03$                     & $0.37$                             & $0.36 \pm 0.02$
+\end{tabular}
+}
+\label{table:FoVeDBA}
+\end{table}

### 2020-06-25 Report/Sections/tools.tex
Update Report

@@
 Step-size standardization can only be done if the analysis is done offline, since the mean step length is in flux for online analysis.
+
+%%%%%%%%%%%% THROWAWAY STUFF FOR LATER HERE %%%%%%%%%%
+
+\iffalse
+
+\section{for simulation results}
+
+\subsection{Model Misspecification}
+
+There are two ways to simulate the fourier modes of the acceleration vector $\bfA_{s,t}$:
+
+\begin{enumerate}
+	\item $$\bbE_1\left(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}\right) = \phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} $$
+
+	\item $$\bbE_2(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}, X^*_{s-1,t}) = \left\{\begin{array}{lr}
+	\phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} & \text{for } X^*_{s,t} = X^*_{s-1,t}\\
+	\mu_{X^*_{s,t}} & \text{for } X^*_{s,t} \neq X^*_{s-1,t}
+	\end{array}\right\}$$
+\end{enumerate}
+
+\subsubsection{$\bbE_1$}
+
+First, look at $\bbE_1$.
+This comes with an issue in that when $X^*_{s,t}$ changes values, it will take a while for the process to ``converge" long-run mean $\mu_{X^*_{s,t}}$.
+In particular, if the behavioral process $X^*$ had been in state 1 for a while and then switches to state 2 for a while, then:
+
+\begin{align*}
+	\bbE_1\left(\bfA_{s+1,t}|X^*_{s:s+1,t} = 2, X^*_{s,t} = 1\right) &=  \phi \mu_1 + (1-\phi) \mu_2 \\
+	%
+	\bbE_1\left(\bfA_{s+2,t}|X^*_{s:s+2,t} = 2, X^*_{s,t} = 1\right) &= \phi\left( \phi \mu_1 + (1-\phi) \mu_2\right) + (1-\phi)\mu_2 \\
+	&= \phi^2 \mu_1 + (1-\phi^2) \mu_2\\
+	%
+	\bbE_1\left(\bfA_{s+n,t}|X^*_{s:s+n,t} = 2, X^*_{s,t} = 1\right) &=  \phi^n \mu_1 + (1-\phi^n) \mu_2 \\
+\end{align*}
+
+So, if using $\bbE_1$, there is an assumption that the fourier modes of the acceleration vector $\bfA$ linearly ``converge" to the new value $\mu_i$ from $\mu_j$ when $X^*_{s,t}$ changes from $i$ to $j$.
+An example of this can be seen in figure (\ref{fig:FOVEDBA}).
+
+ \begin{figure}[h!]
+ 	\centering
+ 	\includegraphics[height=3in]{../Code/test.png}
+ 	\caption{The FoVeDBA of simulated data using $\bbE_1$. Note that the FoVeDBA approaches $\approx$ 300 in state 2 from $\approx$ 10 in state 1 with roughly linear convergence.}
+ 	\label{fig:FOVEDBA}
+ \end{figure}
+
+ Note that if the generative model is correlated, but the inference model isn't, then quick jumps between states 1 and 2 can be missed because the chain never ``converges" to the new mean $\mu_i$, as shown in figure (\ref{fig:STATE}).
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[height=3in]{../Code/test2.png}
+	\caption{The probability of being in each behavioral state $X^*_{s,t}$ as estimated by the HHMM. The colors represent the true behavioral state.}
+	\label{fig:STATE}
+\end{figure}
+
+ \subsubsection{$\bbE_2$}
+
+Clearly, the model has issues decodeing the behavioral state when the model is misspecified.
+Therefore, if the true generative process does not have correlation when jumping between states, then it is better to use $\bbE_2$:
+
+$$\bbE_2(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}, X^*_{s-1,t}) = \left\{\begin{array}{lr}
+\phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} & \text{for } X^*_{s,t} = X^*_{s-1,t}\\
+\mu_{X^*_{s,t}} & \text{for } X^*_{s,t} \neq X^*_{s-1,t}
+\end{array}\right\}$$
+
+This fixes the issue of having to ``converge" to the new value of $\mu_i$ when jumping to state $i$.
+However, note that $\bbE_2$ depends upon both $X^*_{s,t}$ and $X^*_{s-1,t}$.
+This breaks the assumption in the forward algorithm that $\bfA_{s,t}$ depends only on $X^*_{s,t}$.
+One way to deal with this issues is to introduce a new hidden state $Z^*_{s,t} = (X^*_{s-1,t},X^*_{s,t})$.
+This fixes the issue, but now $Z^*_{s,t}$ can take one of $s^2$ values instead of $s$.
+This grows the state space of our model and can make inference harder.
+
+ \fi
+
+%Figure (\ref{fig:smoothing}) also shows the calculated ODBA as a function of time for one a specific dive.
+The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs straightforwardly.
+As a result, the data was split into two-second intervals (100 data points each), and each interval was summarized by the Fourier transform of the ODBA and the velocity at the \textit{end} of the time interval.
+Velocity was taken at the end of the time interval because the behavior of the whale during the time interval has no impact on the velocity of the whale at the beginning of the time interval.
+Because velocity is highly correlated with the average acceleration within an interval, the mean was subtracted out of each ODBA time series before the Fourier transform was taken.
+To reduce the dimension of the ODBA Fourier transform, Fourier coefficients corresponding to frequencies less than or equal to 5 hertz were summed and frequencies above 5 hertz were ignored.
+One nice property of this process is that it reduces the number of data points by a factor of 100, which drastically speeds up parameter estimation.
+The optimal length of the time interval over which to take the Fourier transform is a tuning parameter that is difficult to find.
+An ideal time interval should be long enough give useful Fourier coefficients, but short enough to preserve as much information about the vertical velocity of the animal as possible.
+
+%\subsection{Lag Plot}
+
+%In order to find the appropriate HMM to model whale behavior, a lag plot was made for both velocity and the ODBA fourier sum.
+The results of doing so are shown in figure (\ref{fig:lag}).
+Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that the velocity exhibits a large degree of autocorrelation.
+While the ODBA Fourier sum also exhibits some autocorrelation, the relationship is less strong, so autocorrelation was not incorporated in the ODBA Fourier sums emission distribution.

### 2020-06-25 Report/main.tex
Update Report

@@
 \usepackage{graphicx}
+\usepackage[dvipsnames]{xcolor}
 
@@
 % Title, authors, affiliations
-\title[]{Title on one or two lines without capitals, except after a colon}%\query{Q1}
+\title[]{Modelling fine-scale dive behavior with hidden Markov models }%\query{Q1}
 \author{BlindedA\authorref{1}\thanksref{*}}
@@
 \abstractsection{Abstract}
-Insert your abstract here; it should
-typically be up to ten lines long. Avoid symbols as much as possible. Formulas
-are strongly discouraged, and citations should be avoided. 
-The title and the abstract should be concise and descriptive;
-list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
-http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.
+%Insert your abstract here; it should
+%typically be up to ten lines long. Avoid symbols as much as possible. Formulas
+%are strongly discouraged, and citations should be avoided. 
+%The title and the abstract should be concise and descriptive;
+%list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
+%http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.
 
-The field of animal movement is in the midst of a ``data renaissance" where advancements in tagging technology have given rise to an explosion of data available for statistical modeling.
-In particular, tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
-This results in a vast amount of data which often exhibits many different simultaneous behavioral processes occurring at different time scales.
-
-One solution to this issue is to use a hierarchical hidden Markov model (HHMM).
-HHMMs model the entire system as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.
-One nice property of HHMMs is that its likelihood is relatively easy to compute, facilitating fast maximum likelihood estimates for its associated parameters.
-
-At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by a traditional HMMs.
-To address this issue, it is possible to model small-scale animal behavior as the solution to some stochastic differential equation, but these methods tend to be computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC).
-
-This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency.
-We bridge the gap between discrete hidden Markov models and continuous-time stochastic process models by showing that the two are equivalent under certain conditions.
-In addition, we extract features from highly structured sub-dive behaviors using signal processing techniques.
-These features otherwise could not be modeled with a simple HMM.
-Finally, we apply our method to dive data collected from two Northern resident killer whales off the coast of British Columbia, Canada.
+Recent advances in high-frequency tagging
+technology have made available a vast
+amount of animal movement data. This
+rich new data can exhibit simultaneous behavioural processes occurring at different
+time scales. One can model these processes
+through a hierarchical hidden Markov model
+(HHMM), where the system is modelled as a
+nested structure of hidden Markov models.
+At very short time scales, however, observations can exhibit complicated dependence
+structures that cannot be easily captured by
+traditional HMMs. We demonstrate how
+to incorporate fine-scale processes into the
+larger structure of HHMMs while maintaining computational efficiency. We apply our
+method to dive and accelerometer data collected from northern resident killer whales off
+the coast of British Columbia, Canada.
 
@@
 \input Sections/simulation.tex
-\input Sections/results.tex
-\input Sections/conclusion.tex
+%\input Sections/results.tex
+%\input Sections/conclusion.tex
 
-\newpage
+\clearpage
 
@@
 
-\newpage
+\clearpage
 

### 2020-07-02 Report/Sections/0627NH.tex
Nancy Revisions

I've replaced your Background section with my Models and Parameter Estimation (NH) section, borrowing heavily from your writing.

Here are some comments and explanations.

General things (to always keep in mind - I may add to this list of rules as we go along)
\begin{enumerate}
\item When I recompile in overleaf, I see there are 21 warnings (in the symbol next to the recompile button).  It looks like these are from sections you have written, mostly from problems with newblock.  Please go through these since Latex may be making decisions that you don't like.  It's always good to go through the Latex errors and warnings. 
    \item Don't begin a sentence with a symbol, e.g.~don't write ``$\Theta$ is an important parameter''. 
    It's better to write ``The parameter $\Theta$ is important''.
    \item If you start a section with a subsection, there needs to be some text before the subsection starts.  This text can be a summary of what is in the section.  I see some writers don't follow my rule!  But it bugs me to have a subsection just start up with no explanation of the section.
    \item  Journal articles do not have as many things in math display mode as a thesis or a report.  This is to save space - printing is expensive.  Oh, there is no printing these days, or very little!  But it is the journal style that developed during the print era.  Of course, if you refer to an equation, it is displayed and numbered. A complicated equation (lots of fractions or layers with subscripts, long expressions) should be displayed since it would be hard to read in text mode, where you aren't allowed to use things like frac for fraction. 
    Deciding whether to display or not is always a judgement call.
    In the end, the copy editor will make decisions, but it's good to have the submission somewhat in a journal style.  
    \item Bibliography things (not crucial at this point but must be done before submission):  these issues are common in all stat journals, not just CJS.  Biology journals tend to use numbers and list papers as they appear in the article.  This is not the case in stat journals.
    \begin{enumerate}
        \item 
    Make sure you have caps where you need them - in your bib file,  markov needs to be \{M\}arkov, for instance, to force capitalization.
    \item You have a bib-style that cites in the text via a number, as opposed to via author (data).   CJS uses author (date).  Also,  CJS doesn't use full first name of authors.  See the sample.pdf document that comes with the CJS template.
    \end{enumerate}
    \item From CJS sample.pdf: ``Try to avoid double subscripts, and never use triple subscripts.''  I think you're good -- no triples, but you do have doubles.  We can take a look at those at some point.   This is a common expectation in stat journals.
    \item From CJS sample.pdf: ``Unless central or essential to the ﬂow of the discussion, mathematical arguments should be deferred to the Appendix.''  So we'll see about that theorem you have.  This is getting more common in stat journals - but it's something you need to check.
    \item From CJS sample.pdf: ``When you submit the ﬁnal version of your manuscript in Latex form, please include postscript ﬁles (.ps or .eps) for the ﬁgures, labelling them ﬁg1.ps, ﬁg2.ps, etc.''  That's a direction for the final submission, after review and acceptance. For this submission, you probably just submit one pdf file (I haven't checked).  If so, I'd put each figure and table on a separate page, not merged into text, at the end of the pdf document.
    \item Vector notation - boldface or not?  matrix notation - caligraphy or not?   I find it easier to keep things straight if vectors are in bold, matrices in cal.   It looks like CJS has no specific rules (some journals do) but seems to allow this (I looked at some published articles). I suggest using bold and caligraphy, but using newcommand in case we want to switch the notation.    What do you think?
\end{enumerate}

Things specific to this version.
\begin{enumerate}
    \item I have commented out the section title lines in your subfiles (the ones you input) and placed the section title lines in main.tex.  This way, Marie and I can see what sections you're planning to include.  But we can also see the titles in the input files, so we don't get confused.
    \item You have done a great job with the notation. I see that is not a simple task. 
    \item
    Some of the write-up can be a little cleaner, technically, and I think some of the explanatory material, although nice, can distract from the technical specifications of the model.  It's good to put the explanatory stuff in the introduction.   See my comments in the introductory section (which I've written in main.tex).
    \begin{enumerate}
    \item Typically there is less detail in a journal article than in, say, a comprehensive proposal.  For the journal article, you can assume that the reader either already knows about things like HMMs or can get enough of an idea from what you have said (possibly supplementing knowledge from your references).   But it is useful to lay out the notation, as you have done. So ... there is a balance!
   \item
     In the end, you consider three models: CarHMM, HHMM, CarHHMM.  So these need to be sharply defined, so the reader can glance back and skim. 
     \item fine-scale, finescale, fine scale.  I keep using different ones!  I think if you use it as a noun, then it is fine scale.  ``The Markov chain is on a fine scale.''  I think if you use it as an adjective, it is fine-scale, as in ``that is a fine-scale model''. 
     What do you think?
     \item HMMs or HMM's,  $X_t$s or $X_t$'s??  I don't know, and I am inconsistent.  I think either is acceptable.  Which do you prefer?
\end{enumerate}
\end{enumerate}

### 2020-07-02 Report/main.tex
Nancy Revisions

@@
 
+%%%%%%%%%%%%
+\vfill\eject
+{\centerline{Large{Notes from Nancy}}}
+\input{Sections/0627NH}
+\vfill\eject
+%%%%%%%%%%%%%
+
+%
+\section{Introduction}
+
+{\bf{NH: These are my comments.  I hope they help you.}}
+
+This is where you describe how your method contributes to statistical methodology.
+That is the big goal.
+So there will be literature review of directly relevant work.
+You don't want to leave the non-expert completely lost though, so it is a balance, how much background you give.
+Thank goodness for the Zucchini book, a great place to direct readers for further background.  
+The introduction is a good place to put information about the motivating data set and how researchers have modelled it, what the shortcomings have been.
+I see that you have some great statements about this in your Background section.
+Those statements can be moved to the introduction (with editing -- you can see how I edited).
+
+A hard part about the introduction is that you want to use terminology like CarHMM without getting into too many technical details.
+Sometimes you just have to be technical.
+But I think you can avoid a lot of that.
+
+Here is a great thing you wrote that would go well in the introduction, with some editing so that it doesn't rely on notation that hasn't been introduced.
+You could introduce the basic HMM notation in the intro, if you need to.
+I don't think introducing HMM notation is too technical.
+\begin{quote}
+ One of the key assumptions of both HMMs and HHMMs is \textit{conditional independence} between observations on both the coarse and fine scale.
+Namely, when given the state $X_t$, $Y_t$ is assumed to be independent from all other observations, including $Y_{t-1}$.
+Therefore, traditional HMMs and HHMMs can fail when the observations $Y$ exhibit significant correlation in time.
+ [leave in these examples for intro]  Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swimming behavior of horn sharks off the coast of Southern California \cite{Adam:2019}.
+
+One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}.
+.....
+The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm.
+It also only adds one additional parameter per possible hidden state $(\phi^{(i)})$.
+\end{quote}
+
 %\input Sections/intro.tex
-\input Sections/background.tex
+%
+\section{Models and parameter estimation (NH)}
+\input Sections/model_Nancy.tex
+%
+
+%%
+%\section{Background}
+%\input Sections/background.tex
+%
+\subsection{Continuous Time models}
 \input Sections/carhmm_ou.tex
+%
+\section{Model Formulation}
 \input Sections/model.tex
+%
+\section{Simulation Study}
 \input Sections/simulation.tex
+%
+\subsection{Data Collection and Preprocessing}
 %\input Sections/results.tex
+%
+\section{Discussion}
 %\input Sections/conclusion.tex
+%
+\section{Acknowledgements}
 

### 2020-07-02 Code/HHMM.py
Update visualizer and data location

@@ -260,7 +260,6 @@ class HHMM:
         self.eta[1][state][np.diag_indices(self.pars.K[1])] = 0
         ptm = np.exp(self.eta[1][state])
         ptm = (ptm.T/np.sum(ptm,1)).T
-        #ptm = self.ptm[1][state]
         log_ptm = np.log(ptm)
 
         # find the initial distribution (stationary distribution)
@@ -303,7 +302,6 @@ class HHMM:
         self.eta[0][np.diag_indices(self.pars.K[0])] = 0
         ptm = np.exp(self.eta[0])
         ptm = (ptm.T/np.sum(ptm,1)).T
-        #ptm = self.ptm[0]
         log_ptm = np.log(ptm)
 
         # find the initial distribution (stationary distribution)
@@ -455,96 +453,155 @@ class HHMM:
         return log_alpha, log_beta, gamma, xi
 
 
-    def train_DM(self,data,max_iters=10,tol=0.01,eps=10e-6):
+    def train_DM(self,data,max_iters=10,max_steps=10,tol=0.01,eps=10e-6):
 
         stime = time.time()
-        options = {'maxiter':10,'disp':False}
+        options = {'maxiter':max_iters,'disp':False}
         prev_l = self.likelihood(data)
 
-        for _ in range(max_iters):
+        for n in range(max_steps):
 
+            print('ITERATION %d'%n)
             print(prev_l)
-            print(self.theta)
+            print('')
 
             ### start with crude eta ###
-            def loss_fn(x):
-                ind = 0
+            for i in range(self.pars.K[0]):
+                backup = deepcopy(self.eta[0][i,:])
+                backup_l = deepcopy(self.likelihood(data))
+                def loss_fn(x):
 
-                # update crude eta
-                self.eta[0] = x[0:self.pars.K[0]**2].reshape((self.pars.K[0],
-                                                              self.pars.K[0]))
-                l = -self.likelihood(data)
-                return l
+                    # update crude eta
+                    for j,xj in enumerate(x):
+                        if j < i:
+                            self.eta[0][i,j] = xj
+                        else:
+                            self.eta[0][i,j+1] = xj
 
-            # define inital value
-            x0 = []
-            x0.extend(self.eta[0].flatten())
+                    return -self.likelihood(data)
 
-            # optimize
-            res = minimize(loss_fn, x0, method='Nelder-Mead',options=options)
+                # define inital value
+                x0 = deepcopy(self.eta[0][i,:])
[…]

### 2020-07-02 Code/Parameters.py

@@ -26,8 +26,8 @@ class Parameters:
     def __init__(self):
 
         # loading parameters
-        self.cvc_file = '../../Data/20190902-182840-CATs_OB_1.cvc'
-        self.csv_file = '../../Data/20190902-182840-CATs_OB_1_001.csv'
+        self.cvc_file = '../Data/20190902-182840-CATs_OB_1.cvc'
+        self.csv_file = '../Data/20190902-182840-CATs_OB_1_001.csv'
 
         self.cvc_cols = ['date','time',
                          'GPS lat','GPS long','GPS alt','GPS speed',
@@ -55,7 +55,7 @@ class Parameters:
         self.etime = '2019-09-02 18:00:00'
         self.drop_times = [[1.6,1.8],[3.3,4.3]] # in hr after start
         self.smoother = [0.2]*5
-        self.smooth_cols = ['VeDBA','depth','elevation']#,'Vz','heading','roll',]
+        self.smooth_cols = ['depth','elevation']#,'Vz','heading','roll',]
 
 
         # parameters for HHMM features

### 2020-07-02 Code/Visualisor.py

@@ -109,8 +109,6 @@ class Visualisor:
             ncols = len(self.hhmm.theta[0])
         else:
             features = list(self.hhmm.theta[1][0].keys())
-            print(features)
-            features = ['Ahat_low']
             nrows = self.pars.K[0]
             ncols = len(features)
 
@@ -141,13 +139,13 @@ class Visualisor:
                     if dist == 'gamma':
                         shape = np.square(mu)/np.square(sig)
                         scale = np.square(sig)/np.array(mu)
-                        x = np.linspace(0.1,1e5,100000)
+                        x = np.linspace(0.01,max(mu)+5*max(sig),100000)
                         y = gamma.pdf(x,shape[state],0,scale[state])
                     elif dist == 'normal':
-                        x = np.linspace(min(mu)-3*max(sig),max(mu)+3*max(sig),1000)
-                        y = norm.pdf(x,0,sig[state])
+                        x = np.linspace(min(mu)-3*max(sig),max(mu)+3*max(sig),100000)
+                        y = norm.pdf(x,mu[state],sig[state])
                     elif dist == 'vonmises':
-                        x = np.linspace(-np.pi,np.pi,1000)
+                        x = np.linspace(-np.pi,np.pi,100000)
                         y = vonmises.pdf(x,sig[state],loc=mu[state])
                     else:
                         raise('distribution %s not recognized' % dist)
@@ -158,13 +156,13 @@ class Visualisor:
                         title = 'State Specific Emission Probabilites for Dive Duration'#%s'%feature
                         ax[row_num,col_num].set_xlabel('Dive Duration (seconds)')
                     else:
-                        title = 'State Specific Emission Probabilites for Fourier Component'#%(feature,row_num+1)
-                        ax[row_num,col_num].set_xlabel('$\sum \hat{\mathbf{A}}^2$')
-                        #ax[row_num,col_num].set_xlim([-1,1000])
-                        ax[row_num,col_num].set_xscale('log')
-                        ax[row_num,col_num].set_yscale('log')
-                        ax[row_num,col_num].set_ylim([10e-8,10e-1])
-                        ax[row_num,col_num].set_xlim([10e-1,10e4])
+                        title = 'State Specific Emission Probabilites for %s'%feature
+                        if feature == 'Ahat_low':
+                            ax[row_num,col_num].set_xlabel('$\sum \hat{\mathbf{A}}^2$')
[…]

### 2020-07-02 Report/Sections/model_Nancy.tex

%\section{Models and parameter estimation (NH)}
%\input Sections/model_Nancy.tex

We consider several models for the data, the classical hierarchical hidden Markov model (HHMM) (reference), the CarHMM model of (reference), and our new model that blends the two (CarHHMM).
The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes.
The CarHMM  modifies the usual HMM by modelling additional dependence into the sequence of observed data.
In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.

{\bf{NH: It strikes me that the above might belong in the introduction.  In that case, the what appears above in this section would be modified to something much shorter. 
}} 
We first remind the reader of the definition of a hidden Markov model,   the model of the coarse-scale process.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The hidden Markov model (HMM)}

A \textit{hidden Markov model}  is comprised of a sequence of  unobserved states $X_t$, $t = 1, \ldots, T$, and an associated sequence of  possibly high-dimensional observations $Y_t$, $t = 1, \ldots, T$,
The $Y_t$s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
 The $X_t$s form a Markov chain and take possible values $1, \ldots, N$. Their distribution is governed by the distribution of the initial state $X_1$ and the  $N \times N$ transition probability matrix $\Gamma$ where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$, $t=1,\ldots, T-1$, ~$i, j = 1,\ldots, N$. 
 %
 We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i$th component
$\delta_i = \Pr\{X_1 = i\},~ i = 1,\ldots,N.$
%
Recall that a Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$.

%
The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$, that is, the distribution of  $Y_t$ conditional on $X_1,\ldots, X_T$ and $\{Y_1,\ldots, Y_T\}/ \{Y_t\}$ is simply the distribution of $Y_t$ given $X_t$.  
%
These conditional distributions are governed by state-dependent parameters:  if $X_t = i$, the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes  $f^{(i)}(\cdot)$.
%
A visualization of the dependence structure can be seen in Figure \ref{fig:HMM}.

To find the maximum likelihood estimates of the parameters $\Gamma$ and $\Theta \equiv (\theta^{(1)},\ldots,\theta^{(N)})$, let $y = (y_1,\ldots,y_T)$ be the observed emissions. 
 To facilitate calculations we write the likelihood $\calL_{\text{HMM}}$ in a specific form, using the  well-known \textit{forward algorithm} \cite{Zucchini:2016} as follows:
%
$$\calL_{\text{HMM}}(y;\Theta,\Gamma) = \delta P(y_1;\Theta) \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N$$
%
where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and
%
$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry  $f^{(i)}(y_t; \theta^{(i)})$.
%

We will always parameterize transition probability matrices to remove constraints that the entries of the matrix are non-negative and that  the rows sum to 1.  We do this by parameterizing the $N \times N$ transition probability matrix $\Gamma$ using $\eta \in \bbR^{N \times N}$ \cite{Barajas:2017}:
%
\[
\Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, 
\]
%
setting $\eta_{ii}$  to zero, $i=1,\ldots, N$, for identifiability.  Then $\calL_{\text{HMM}}(y;\Theta,\Gamma)$ can be maximized using any numerical optimizer.  For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of  $\eta$.

\begin{figure}[h!]
	\centering
	\includegraphics[width=5in]{../Plots/HMM.png}
	\caption{Graphical representation of a traditional HMM.}
	\label{fig:HMM}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The hierarchical hidden Markov model (HHMM)}

{\bf{NH:
I wonder if it would be easier on the reader to put the CarHMM first, since the CarHMM deals with just one Markov chain.
}}

A hierarchical hidden Markov model contains two process levels, a coarse-scale process and a fine-scale process.  The coarse-scale process is a hidden Markov model as defined in the previous subsection,  with  $X_1\ldots, X_T$ the unobserved Markov chain of states and $Y_1,\ldots, Y_T$ the corresponding observed responses.  The possible states are $1,\ldots, N$.   
%

For each value of $t$, the state $X_t$ emits not only $Y_t$ but also  a sequence of unobserved states, 
$X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of observed emissions, 
$Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$. 
These form our fine-scale process.
Conditional on the coarse-scale hidden state $X_t$, the joint fine-scale process,  $X_t^*$, $Y_t^*$, is a hidden Markov model with parameters depending on the value of $X_t$.  Specifically, given that $X_t=i$,   $X_t^*$ is a Markov chain with states $1,\ldots, N_t^*$ with 
%
\hfill\break{\bf{NH:  Do we write $1^*$? That's strange.}}
\hfill\break
%
 distribution of $X_t^*$, conditional on $X_t=i$,  given by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial probability, denoted by the $N_t^*$-vector $\delta^{*(i)}$, which we take equal to the stationary distribution of the chain.     For simplicity, we  take $N_t^* \equiv N^*$ although this is not necessary.
%
The distribution of $Y_{t, t^*}$
given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter  $\theta^{(i,i^*)}$, and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or sometimes simply denoted by $f^{*(i,i^*)}(\cdot)$.
Let $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$, the fine-scale emission parameter vector corresponding to $X_t=i$.
%
\hfill\break
{\bf{NH:  will there be some parameters that are common across the fine-scale emission distributions? Do we want to separate them out?\\
Also - I used left ( and right ) above to make the parentheses a little bigger.  I think it looks better.  So we may want to do this throughout.
$f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ versus
$f^{*(i,i^*)}(\cdot; \theta^{(i,i^*)})$.
}}
\hfill\break
%
The dependencies among the processes are defined as follows.
%
Given the coarse-scale states, $X_1,\ldots, X_T$, and the coarse scale emissions,  
$Y_1,\ldots, Y_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs.  
%%
\hfill\break
{\bf{I think that works for the dependencies - might need to add more. Does that look OK to you?}}
%%

Due to the nested structure of the hierarchical hidden Markov model, the likelihood  is easy to calculate via the forward algorithm, just as for HMMs.
%
Let $y$ be the $T$-vector of the observed coarse-scale emissions and
$y^*$ be the $(T_1^* + \cdots T_T^*)$-vector of the observed fine-scale emissions.
%
Let $\Theta^*$ denote the collection of all fine-scale emission parameters,
$\Theta^{*(i)}$, $i=1,\ldots, N$, and let $\Gamma^*$ denote the collection of all fine-scale transition probability matrices, $\Gamma^{*(i)}$, $i=1,\ldots, N$.
%
Then the likelihood of the observed data is:
%
\[
\calL_{\text{HHMM}}(y,y^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,y_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N
\]
%
where 	$P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N^ \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to 
$f^{(i)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(i)},
\Gamma^{*(i)}\right)$.

For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, model selection and model checking, see Adam et al \cite{Adam:2019}.

\begin{figure}[h!]
	\centering
	\includegraphics[width=6.5in]{../Plots/HHMM.png}
	\caption{Graphical representation of a traditional HHMM.}
	\label{fig:HHMM}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Conditionally auto-regressive hidden Markov models (CarHMM)}

{\bf{NH: 
I've taken what you have written and edited.

Here is a comment about a big change I made.  You give a definition of the distribution of the correlated emissions in terms of means and variances, and I think you want to say something in a more general form, that is, that the distribution of $Y_t$ is determined by $X_t$ and $Y_{t-1}$.  Also,  you only define the conditional variance of $Y_t$ given $X_t$,  excluding $Y_{t-1}$.  You should have the left side of that equation for the variance be conditional on both $X_t$ and $Y_{t-1}$ for your math to go through.  It's no problem if you want to assume that the variance on the right side only has $i$ in it.  That is an assumption about the distribution of $Y_t$  conditional on both $X_t$ and $Y_{t-1}$ 
}}

One of the key assumptions of an HMM is \textit{conditional independence} between observations, given the state sequence.  Therefore, traditional HMMs do not hold when the observations exhibit certain forms of significant correlation in time.  The same issue arises in HHMMs.

One way to model extra correlation in an HMM type structure is via a CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}. 
In a CarHMM, we have a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution.
The CarHMM assumes that the distribution of the $t$th emission, $Y_t$, conditional on $X_1,\ldots, X_T$ and $ \{Y_1,\ldots, Y_T\}/\{Y_t\}$, depends only on $X_t$ and $Y_{t-1}$. 
{\bf{NH:  True?  or just conditional on the $X_t$s and $Y_1,\ldots, Y_{t-1}$?}}
The first emission $Y_1$ is assumed to be fixed as an initial value.
%
{\bf{NH: does $Y_1$ depend on $X_1$?}}
%
 The graphical model associated with the structure of a CarHMM is shown in Figure (\ref{fig:CarHMM}).

\begin{figure}[h!]
	\centering
	\includegraphics[width=3in]{../Plots/CarHMM.png}
	\caption{Graphical representation of a traditional CarHMM. The additional arrows representing auto-correlation between observations are shown in red for emphasis.}
	\label{fig:CarHMM}
\end{figure}

%
We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1})$ or as
$f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$.
As an example, we could assume that this conditional distribution is normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{2(i)},\phi^{(i)}\}$, with:
%
\[
\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}
\]
and
\[
\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = \sigma^{2(i)}.
\]
%
{\bf{Ah - now I see the issue about changing states.  It would be great to have something about the distribution of $Y_t$ given $X_t=i, X_{t-1}=j$. You'd want a different conditional distribution, depending on if $i=j$ or not.
}}
%
The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to write the likelihood using the forward algorithm. Note that in our normal model formulation,  we have added only one additional parameter, $\phi^{(i)}$,  per possible hidden state.

The likelihood for the CarHMM can be easily calculated using the forward algorithm.  As previously, let $y$ be the vector of observed emissions.  Then
\begin{equation}
\calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N
\label{CarHMM_likelihood}
\end{equation}
where
%
$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
%

% do I need this subsections
\subsection{Decoding}

{\bf{I'm stopping here.  Perhaps you can take a stab at revisions here?}}

### 2020-07-02 Code/Train_Model.py
Add bash scripts for compute Canada

@@ -1,0 +1,95 @@
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.cm as cm
+import matplotlib
+
+import sys
+
+from scipy.stats import gamma
+from scipy.stats import norm
+from scipy.stats import multivariate_normal
+from scipy.stats import gaussian_kde
+from scipy.special import expit
+from scipy.special import logsumexp
+from scipy.optimize import minimize
+from scipy.optimize import LinearConstraint
+from scipy.signal import convolve
+from scipy.interpolate import interp1d
+
+import divebomb
+
+import time
+import pickle
+
+import Preprocessor
+import Parameters
+import HHMM
+import Visualisor
+
+np.random.seed(0)
+
+model = str(sys.argv[1])
+
+# set parameters
+pars = Parameters.Parameters()
+pars.features = [{'dive_duration':{'corr':False,'f':'gamma'}},
+                 {'Ahat_low':{'thresh':5,'corr':False,'f':'gamma'},
+                  'Ax':{'corr':True,'f':'normal'},
+                  'Ay':{'corr':True,'f':'normal'},
+                  'Az':{'corr':True,'f':'normal'}}]
+pars.K = [2,3]
+pars.share_fine_states = True
+
+if model == 'CarHHM':
+    pars.K = [1,3]
+elif model == 'HHMM':
+    pars.features[1]['Ax']['corr'] = False
+    pars.features[1]['Ay']['corr'] = False
+    pars.features[1]['Az']['corr'] = False
+elif model == 'CarHHMM1':
+    pars.features[1] = {'Ax':{'corr':True,'f':'normal'},
+                        'Ay':{'corr':True,'f':'normal'},
+                        'Az':{'corr':True,'f':'normal'}}
+
+# define files
+HHMM_file = '../Params/%s_k_%s_%s_dives_same_fine_states_long' % (model,pars.K[0],pars.K[1])
+data_outfile = '../Params/Data_%s_k_%s_%s_dives_same_fine_states_long' % (model,pars.K[0],pars.K[1])
+
+# preprocess data
+prep = Preprocessor.Preprocessor(pars)
+print('loading data')
+df = prep.load_data(pars.cvc_file,pars.csv_file,pars.cvc_cols)
+print('pruning cols')
+df = prep.prune_cols(df)
+print('pruning times')
[…]

### 2020-07-03 Report/Sections/results.tex
Revise Report
Updates from Overleaf

@@
 
-\section{Results}
+%\section{Killer Whale Case Study}
 
@@
 A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
+This process resulted in a total of 267 dives.
 Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second.
@@
 
-Figure (\ref{fig:smoothing}) also shows the calculated ODBA as a function of time for one a specific dive.
-The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs straightforwardly.
-We set $\tilde{f} = 5$ hertz and $w = 2$ seconds to calculate $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$.
-This reduces the dimension of each window from $50 1/s * 2 s = 100$ to 2.
+\subsection{Model Selection}
 
-\subsection{Lag Plot}
+The collection of all dive durations were set to be the coarse-scale observations $Y$, and the acceleration data was used to determine the fine-scale observations $Y^*$.
+Figure (\ref{fig:raw_data_one_dive}) shows both the depth profile and the raw accelerometer data for one specific dive.
+The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs without signal processing of the raw acceleration data.
+Therefore, the STFT was used to process the raw acceleration data and calculate both $Y{*(1)}$ and $Y^{*(2)}$.
+We set $\tilde{f} = 5$ hertz and $h = 2$ seconds, which reduced the dimension of each window from $50 s^{-1} * 2 s = 100$ to 2.
 
-In order to find if the CarHHMM was appropriate for this data, a lag plot was made for both $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$.
+In order to find if the CarHHMM was appropriate for this data, a lag plot was made for both $Y^{*(1)}_{t,s^*}$ and $Y^{*(2)}_{t,s^*}$.
 The results of doing so are shown in figure (\ref{fig:lag}).
-Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that $Z^{*(1)}_{t,t^*}$ exhibits a large degree of autocorrelation.
-While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, so autocorrelation was not incorporated in the emission distribution of $Z^{*(2)}_{t,t^*}$ for computational purposes.
+Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that $Y^{*(1)}_{t,s^*}$ exhibits a large degree of auto-correlation.
+While $Y^{*(2)}_{t,s^*}$ also exhibits some auto-correlation, the relationship is less strong, so auto-correlation was not incorporated in the emission distribution of $Y^{*(2)}_{t,s^*}$.
 
@@
 
+Since information criteria tend to overestimate the number of states in biological processes \cite{Pohle:2017}, we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically and admittedly somewhat arbitrarily.
+
 \subsection{Results}
 
-The parameters of the estimated emission distributions for each behavioral state are shown in table (\ref{table:emis_dist}) below.
+The parameters of the estimated emission distributions for each behavioral state are shown in table (\ref{table:emis_dist}).
 Each distribution is also plotted in figure (\ref{fig:emis_dist}).
-Note that the autocorrelation within the velocity sequence is not captured in figure (\ref{fig:emis_dist}), so it is important to refer to the estimated auto-correlation parameter $\hat \phi$ from table (\ref{table:emis_dist}) when considering the emission distributions shown in figure (\ref{fig:emis_dist}).
+Note that the auto-correlation within the velocity sequence is not captured in figure (\ref{fig:emis_dist}), so it is important to refer to the estimated auto-correlation parameter $\hat \phi$ from table (\ref{table:emis_dist}) when considering the emission distributions shown in figure (\ref{fig:emis_dist}).
 %
@@
 Future analysis could find a way to include a more heterogeneous set of dives using a framework such as the hierarchical hidden Markov model (HHMM) introduced by Leos-Barajas et al \cite{Barajas:2017}.
+
+\subsection{Model evaluation}

### 2020-07-05 Report/Sections/results.tex

@@
 
-To demonstrate this methodology as applied to real-world data, the CarHHMM was used to analyze dive data from a killer whale off the coast of British Columbia, Canada.
+To apply this methodology to real-world data, the CarHHMM was used to analyze dive data from a killer whale off the coast of British Columbia, Canada.
 The data was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions.
@@
 A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
-This process resulted in a total of 267 dives.
 Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second.
 Data preprocessing was done in part with the \textit{divebomb} package in Python \cite{Nunes:2018}.
+After preprocessing the raw data, a total of 267 dives were observed.
+A plot of the processed data for all dives and one selected dive can be seen in figures \ref{fig:data} and \ref{fig:data_one_dive}, respectively.
 
@@
 Figure (\ref{fig:raw_data_one_dive}) shows both the depth profile and the raw accelerometer data for one specific dive.
-The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs without signal processing of the raw acceleration data.
-Therefore, the STFT was used to process the raw acceleration data and calculate both $Y{*(1)}$ and $Y^{*(2)}$.
-We set $\tilde{f} = 5$ hertz and $h = 2$ seconds, which reduced the dimension of each window from $50 s^{-1} * 2 s = 100$ to 2.
+The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs without some kind of signal processing.
+Therefore, the STFT was used to calculate both $Y{*(1)}$ and $Y^{*(2)}$ as described in section \ref{subsec:STFT}.
+We set $\tilde{f} = 5$ hertz and $h = 2$ seconds, which reduced the dimension of each window from $50 s^{-1} * 2 s = 100$ to $2$.
 
-In order to find if the CarHHMM was appropriate for this data, a lag plot was made for both $Y^{*(1)}_{t,s^*}$ and $Y^{*(2)}_{t,s^*}$.
-The results of doing so are shown in figure (\ref{fig:lag}).
-Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that $Y^{*(1)}_{t,s^*}$ exhibits a large degree of auto-correlation.
+In order to determine if the CarHHMM was appropriate for this data, a lag plot was made for both $Y^{*(1)}_{t,s^*}$ and $Y^{*(2)}_{t,s^*}$, as shown in figure (\ref{fig:lag}).
+The number of behavioral states is not clear from the lag plot, but it is clear that $Y^{*(1)}_{t,s^*}$ exhibits a large degree of auto-correlation.
 While $Y^{*(2)}_{t,s^*}$ also exhibits some auto-correlation, the relationship is less strong, so auto-correlation was not incorporated in the emission distribution of $Y^{*(2)}_{t,s^*}$.
@@
 
-Since information criteria tend to overestimate the number of states in biological processes \cite{Pohle:2017}, we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically and admittedly somewhat arbitrarily.
+Information criteria tends to overestimate the number of states in biological processes \cite{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically and admittedly somewhat arbitrarily.
+This is a common inssue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria.
+Section \ref{subsec:model_validation} describes our process of validating this model in particular.
 
@@
 The parameters of the estimated emission distributions for each behavioral state are shown in table (\ref{table:emis_dist}).
-Each distribution is also plotted in figure (\ref{fig:emis_dist}).
-Note that the auto-correlation within the velocity sequence is not captured in figure (\ref{fig:emis_dist}), so it is important to refer to the estimated auto-correlation parameter $\hat \phi$ from table (\ref{table:emis_dist}) when considering the emission distributions shown in figure (\ref{fig:emis_dist}).
+Each distribution is also plotted in figures (\ref{fig:coarse_emis}) and (\ref{fig:fine_emis}).
+Note that the auto-correlation within the velocity sequence is not captured in figure (\ref{fig:emis_dist}), so it is important to refer to the estimated auto-correlation parameter $\hat \phi$ from table (\ref{table:emis_dists}) when considering the emission distributions shown in figure (\ref{fig:fine_emis}).
 %
@@
 	\centering
-	\includegraphics[height=2in]{../Plots/emis_dist.png}
+	\includegraphics[height=4in]{../Plots/coarse-emissions.png}
 	\caption{Estimated probability distributions for the long-term vertical velocity and $\tilde a$ in each behavioral state.
-Note that the distribution over vertical velocity does not take autocorrelation into account.}
-	\label{fig:emis_dist}
+Note that the distribution over vertical velocity does not take auto-correlation into account.}
+	\label{fig:coarse_emis}
+\end{figure}
+%
+\begin{figure}[h!]
+	\centering
+	\includegraphics[height=2.5in]{../Plots/fine-emissions.png}
+	\caption{Estimated probability distributions for the long-term vertical velocity and $\tilde a$ in each behavioral state.
+Note that the distribution over vertical velocity does not take auto-correlation into account.}
+	\label{fig:coarse_emis}
 \end{figure}
@@
     \centering
-	\begin{tabular}{l|lll}
-		& State 1 & State 2 & State 3 \\ \hline
-		$\hat \mu_{RL}$        & 0.66    & 0.68    & 0.22    \\
-		$\hat \sigma^2$        & 0.095   & 0.070   & 1.30    \\
-		$\hat \phi$            & 0.88    & 0.99    & 0.03    \\ \hline
-		$\hat \tilde \mu$      & 0.69    & 0.15    & 3.70    \\
-		$\hat \tilde{\sigma^2}$ & 0.41    & 0.067   & 3.16   
+    \label{table:emis_dists}
+    \caption{Estimates and standard errors of emission parameters for killer whale data.}
+    \begin{tabular}{ccccc}
+    \multirow{2}{*}{Feature}                 & \multirow{2}{*}{Dive / Subdive Type} & \multicolumn{3}{c}{Parameter Estimate}              \\
+                                             &                                      & $\hat \mu$      & $\hat \sigma$   & $\hat \phi$     \\ \hline
+    \multirow{2}{*}{Dive Duration (seconds)} & 1                                    & $27.23 \pm 0.63$ & $10.89 \pm 0.56$ & ---             \\
+                                             & 2                                    & $127.96 \pm 11.50$ & $64.13 \pm 9.21$ & ---             \\ \hline
+    \multirow{3}{*}{$Y^{*(1)}_x$}            & 1                                    & $0.98 \pm 0.07$ & $0.04 \pm 0.00$ & $0.99 \pm 0.00$ \\
+                                             & 2                                    & $0.22 \pm 0.01$ & $0.08 \pm 0.00$ & $0.87 \pm 0.01$ \\
+                                             & 3                                    & $0.23 \pm 0.03$ & $0.28 \pm 0.01$ & $0.62 \pm 0.03$ \\ \hline
+    \multirow{3}{*}{$Y^{*(1)}_y$}            & 1                                    & $0.61 \pm 0.09$ & $0.05 \pm 0.00$ & $0.99 \pm 0.00$ \\
+                                             & 2                                    & $0.43 \pm 0.01$ & $0.09 \pm 0.00$ & $0.87 \pm 0.01$ \\
+                                             & 3                                    & $0.38 \pm 0.04$ & $0.35 \pm 0.01$ & $0.62 \pm 0.04$ \\ \hline
+    \multirow{3}{*}{$Y^{*(1)}_z$}            & 1                                    & $-1.00 \pm 0.11$ & $0.05 \pm 0.00$ & $0.99 \pm 0.00$ \\
+                                             & 2                                    & $-0.57 \pm 0.01$ & $0.10 \pm 0.00$ & $0.87 \pm 0.01$ \\
+                                             & 3                                    & $-0.35 \pm 0.04$ & $0.34 \pm 0.01$ & $0.62 \pm 0.04$ \\ \hline
+    \multirow{3}{*}{$Y^{*(2)}$}              & 1                                    & $27.16 \pm 0.32$ & $16.67 \pm 0.32$ & ---             \\
+                                             & 2                                    & $406.98 \pm 4.42$ & $438.09 \pm 5.49$ & ---             \\
+                                             & 3                                    & $9688.54 \pm 221.95$ & $14584.02 \pm 358.40$ & ---             \\ \hline
     \end{tabular}
-	\caption{Table of estimated parameters for probability distributions over $v$ and $\tilde a$ for each behavioral state}
-	\label{table:emis_dist}
 \end{table}
 %
-The estimated probability transition matrix and associated stationary distribution is shown below:
-
+The estimated probability transition matrices and associated stationary distributions are shown below:
+%
 $$\hat \Gamma = \begin{pmatrix} 
-0.7983 & 0.1618 & 0.0398 \\
-0.0120 & 0.8796 & 0.0003 \\
-0.2425 & 0.0049 & 0.7526
-\end{pmatrix} \qquad \hat \delta = \begin{pmatrix} 0.4141 & 0.5020 & 0.0839 \end{pmatrix}$$
-
-Finally, the Viterbi-decoded dive behavior within a randomly selected dive (deeper than 20 meters) is shown in figure (\ref{fig:viterbi}).
-
-\newpage
+0.849 & 0.151 \\
+0.907 & 0.093
+\end{pmatrix}$$
+$$\hat \delta = \begin{pmatrix} 0.857 & 0.143 \end{pmatrix}$$
+%
+$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
+0.724 & 0.276 & 0.000 \\
+0.057 & 0.887 & 0.056 \\
+0.000 & 0.247 & 0.753
+\end{pmatrix} \qquad 
+\hat \Gamma^{*(2)} = \begin{pmatrix} 
+0.871 & 0.129 & 0.000 \\
+0.135 & 0.829 & 0.036 \\
+0.000 & 0.246 & 0.754
+\end{pmatrix}$$
+$$\hat \delta^{*(1)} = \begin{pmatrix} 0.143 & 0.698 & 0.159 \end{pmatrix} \qquad
+\hat \delta^{*(2)} = \begin{pmatrix} 0.476 & 0.456 & 0.067 \end{pmatrix}$$
+%
+Finally, the decoded dive behavior within a selected dive (deeper than 20 meters) is shown in figure (\ref{fig:viterbi}).
 
@@
 \end{figure}
-
-\newpage
 

### 2020-07-05 Report/Sections/results.tex

@@
 %
+While the ecological meaning of these behavioral states is tenuous, we hypothesize the following interpretations.
+For the coarse-scale, dive type 1 corresponds to shorter, shallower dives that the killer whale uses to rest before performing a dive of type 2, which corresponds to a deeper, more sustained dive.
+For the fine-scale, behavioral state 1 corresponds to gliding and passive swimming.
+The mean of $Y^{*(2)}$ in this state is at least an order of magnitude smaller than sub-dive behavior 2, the variance of $Y^{*(1)}$ is smaller, and the autocorrelation of $Y^{*(1)}$ is higher than every other behavioral state.
+This indicates a relatively constant reading of acceleration and therefore low levels of activity.
+On the other hand, sub-dive state 3 corresponds to vigorous swimming activity, as the mean of $Y^{*(2)}$ and variance of $Y^{*(1)}$ are much higher than every other state.
+The auto-correlation of $Y^{*(1)}$ is also much lower in this state, implying more variation in acceleration every 2 seconds.
+Behavioral state 2 is between the other two behavioral states in almost every parameter estimate.
 The estimated probability transition matrices and associated stationary distributions are shown below:
@@
 %
-Finally, the decoded dive behavior within a selected dive (deeper than 20 meters) is shown in figure (\ref{fig:viterbi}).
+About 85\% of the dives observed here are short, as the whale usually rests for many dives in a row before performing a deep dive.
+Interestingly, the probability transition matrix $\hat \Gamma$ shows that the current dive does not effect the dive type of the following dive much.
+In addition, the fine-scale probability transition matrices imply that the killer whale is much more likely to be in a less active sub-dive state when performing deep dives than when performing shallow dives- 14\% of shallow dives are spend in sub-dive state 1 while 48\% of deep dives are spent in sub-dive state 1.
+This is likely because the killer whale needs to conserve energy when diving at depth and holding its breath for longer.
+
+Finally, the decoded dive behavior within a selected dive (deeper than 20 meters) is shown in figure (\ref{fig:labeled_dives}).
+In addition, the probability of each dive type and sub-dive state is shown in (fig \ref{fig:coarse_probs}) and (fig \ref{fine_probs}), respectively.
 
@@
 	\centering
-	\includegraphics[height=8in]{../Plots/viterbi.png}
-	\caption{Features of a particular killer whale dive and Viterbi-decoded estimates for the intra-dive behavioral states.}
-	\label{fig:viterbi}
+	\includegraphics[width=5in]{../Plots/labeled_dives.png}
+	\caption{Features of a particular set of killer whale dives and Viterbi-decoded estimates for the intra-dive behavioral states.}
+	\label{fig:labeled_dives}
+\end{figure}
+%
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=5in]{../Plots/Coarse_state_probs.png}
+	\caption{Probability of dive types for a particular set of killer whale dives.}
+	\label{fig:coarse_probs}
+\end{figure}
+%
+\begin{figure}[h!]
+	\centering
+	\begin{subfigure}[t]{1.0\textwidth}
+        \centering
+        \includegraphics[width=5in]{../Plots/Fine_state_probs_1.png}
+        \caption{Fine-scale state 1 probabilities}
+    \end{subfigure}
+    \newline
+    \begin{subfigure}[t]{1.0\textwidth}
+        \centering
+        \includegraphics[width=5in]{../Plots/Fine_state_probs_3.png}
+        \caption{Fine-scale state 3 probabilities}
+    \end{subfigure}
+	\caption{Probability of sub-dive types for a particular set of killer whale dives.}
+	\label{fig:fine_probs}
 \end{figure}
 
-While the ecological meaning of these behavioral states is tenuous, we hypothesize the following interpretations.
-Roughly speaking, behavioral state 1 corresponds to mildly active swimming.
-The mean of $\tilde a$ in this state is larger than behavioral state 2, indicating more activity by the whale, but the autocorrleation term $\hat \phi$ is still very high, indicating that velocities don't change very much every two seconds.
-Behavioral state 2 corresponds to gliding or turning, where no active swimming is taking place.
-It is characterized by a low mean $\tilde a$ value and a very high autocorrelation term $\hat \phi$, both of which indicate little activity.
-Finally, behavioral state 3 represents sudden jerking or more vigorous and active swimming than behavioral state 1.
-This behavior is so rigorous that $\hat \phi$ drops almost to zero and $\hat \sigma^2$ subsequently rises significantly.
-In behavioral state 3, the velocity readings of the killer whale are essentially uncorrelated when recorded every two seconds.
+\subsection{Model evaluation}
 
-\subsection{Future Work}
+Two visual tools were used to evaluate the model for this data: pseudo-residuals and empirical histograms.
+A pseudo-residual of a particular observation, as described in \cite{Zucchini:2016}, is the marginal cdf of an observation, conditioned on all other observations, and mapped to the quantile function of a standard normal distribution.
+In particular, the pseudo-residual of an observation $y_{t,s^*}$ is $\Phi^{-1} \left(Pr(Y_{t,s^*} < y_{t,s^*}|Y/Y_{t,s^*}) \right)$, where $\Phi$ is the cdf of a standard normal distribution.
+If the model is well-specified, then all pseudoresiduals should be independent and follow a standard normal distribution.
+While histograms of the pseudoresiduals support this for all features, one notable exception is the case of $Y^{*(2)}$, shown in (fig \ref{fig:pseudoresids}).
+These psuedoresiduals are noticeably right-skewed, implying that the true distribution of $Y^{*(2)}$ may follow a heavier-tailed distribution compared to a gamma distribution such as a power law.
 
-This analysis is largely exploratory with a purely heuristic approach to decide the number of behavioral states.
-Future work should repeat the analysis using several different values for the number of behavioral states.
-It is also important to validate these results by plotting psuedoresiduals and performing goodness-of-fit tests on the emission distributions.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=5in]{../Plots/Coarse_state_probs.png}
+	\caption{Psuedoresiduals of $Y^{*(2)}$}
+	\label{fig:pseudoresids}
+\end{figure}
 
-Parameter estimation via likelihood maximization can be slow, especially with an observation rate of 50 hertz.
-If dives are modeled as uncorrelated batches of data, then using stochastic gradient ascent to maximize the likelihood may considerably speed up parameter estimation.
+In addition to psuedoresiduals, we plotted a histogram of observations for each feature where the observations were weighted by the probability that the whale was in a particular hidden state.
+This empirical distribution was then be plotted on top of the estimated true pdf of that feature conditioned on the hidden state.
+This procedure is similar to that for pseudo-residuals, but it produces individual plots for each sub-dive state.
+Again, our results mostly support the model in question with the exception of $Y^{*(2)}$, which is right-skewed.
+In addition, $Y^{*(1)}$ has heavy tails for sub-dive state 3 (see fig \ref{fig:empirical_dist}), indicating the existence of rare events corresponding to very violent trashing of the killer whale.
+These outliers are potential subjects for future studies to better understand rare behavioral events for this killer whale.
 
-Finally, this work was run exclusively on dives deeper than 20 meters so that within-dive behaviors would be similar between dives.
-Future analysis could find a way to include a more heterogeneous set of dives using a framework such as the hierarchical hidden Markov model (HHMM) introduced by Leos-Barajas et al \cite{Barajas:2017}.
-
-\subsection{Model evaluation}
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=5in]{../Plots/Coarse_state_probs.png}
+	\caption{Empirical Distribution of $Y^*(1)_x$ for sub-dive state 3 plotted against its estimated pdf}
+	\label{fig:empirical_dist}
+\end{figure}

### 2020-07-06 Report/Sections/results.tex

@@
 After preprocessing the raw data, a total of 267 dives were observed.
-A plot of the processed data for all dives and one selected dive can be seen in figures \ref{fig:data} and \ref{fig:data_one_dive}, respectively.
+A plot of the processed data for all dives and a collection of five selected dive can be seen in figures \ref{fig:data} and \ref{fig:data_one_dive}, respectively.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[height=5in]{../Plots/lagplot.png}
+	\caption{Lag plot of vertical velocity and $\tilde a$ (left) and a the associated normal kernel density estimates (right)}
+	\label{fig:data}
+\end{figure}
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[height=5in]{../Plots/lagplot.png}
+	\caption{Lag plot of vertical velocity and $\tilde a$ (left) and a the associated normal kernel density estimates (right)}
+	\label{fig:data_one_dive}
+\end{figure}
 
@@
 Information criteria tends to overestimate the number of states in biological processes \cite{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically and admittedly somewhat arbitrarily.
-This is a common inssue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria.
+This is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria.
 Section \ref{subsec:model_validation} describes our process of validating this model in particular.
@@
 
-\subsection{Model evaluation}
+\subsection{Model Validation}
+\label{model_validation}
 
@@
 	\centering
-	\includegraphics[width=5in]{../Plots/Coarse_state_probs.png}
+	\includegraphics[width=5.5in]{../Plots/pseudoresids.png}
 	\caption{Psuedoresiduals of $Y^{*(2)}$}
@@
 	\centering
-	\includegraphics[width=5in]{../Plots/Coarse_state_probs.png}
-	\caption{Empirical Distribution of $Y^*(1)_x$ for sub-dive state 3 plotted against its estimated pdf}
+	\includegraphics[width=5.5in]{../Plots/empirical_dist.png}
+	\caption{Empirical Distribution of $Y^{*(1)}_z$ for sub-dive state 3 plotted against its estimated pdf}
 	\label{fig:empirical_dist}

### 2020-07-07 Code/Visualisor.py
Update Plots

@@ -124,7 +124,7 @@ class Visualisor:
 
                     dist = self.pars.features[0][feature]['f']
                     K = self.pars.K[0]
-                    colors = [cm.get_cmap('tab20')(i) for i in [1,0]]
+                    colors = [cm.get_cmap('tab10')(i) for i in [0,1]]
                     legend = ['Dive Type %d'%(x+1) for x in range(K)]
                 else:
                     mu = self.hhmm.theta[1][row_num][feature]['mu']
@@ -132,7 +132,7 @@ class Visualisor:
 
                     dist = self.pars.features[1][feature]['f']
                     K = self.pars.K[1]
-                    colors = [cm.get_cmap('tab20')(i+self.pars.K[0]) for i in [2,1,0]]
+                    colors = [cm.get_cmap('viridis')(i) for i in [0.,0.5,1.]]
                     legend = ['Subdive Behavior %d'%(x+1) for x in range(K)]
 
                 for state in range(K):
@@ -247,7 +247,7 @@ class Visualisor:
 
             # dive-level coloring
             plt.figure(figsize=(30,5))
-            colors = [cm.get_cmap('tab20')(i) for i in [1,0]]
+            colors = [cm.get_cmap('tab10')(i) for i in [0,1]]
             legend = ['Dive Type %d' % (i+1) for i in range(self.pars.K[0])]
             for state,dive_df in enumerate(dives):
                 plt.plot(dive_df['sec_from_start']/60,dive_df[col],
@@ -270,7 +270,7 @@ class Visualisor:
 
             # subdive-level coloring
             plt.figure(figsize=(30,5))
-            colors = [cm.get_cmap('tab20')(i+self.pars.K[0]) for i in [2,1,0]]
+            colors = [cm.get_cmap('viridis')(i) for i in [0.,0.5,1.]]
             legend = ['Subdive Behavior %d' % (i+1) for i in range(self.pars.K[1])]
             for state,subdive_df in enumerate(subdives):
                 plt.plot(subdive_df['sec_from_start'],subdive_df[col],
@@ -302,7 +302,7 @@ class Visualisor:
 
                 # dive-level columns - color by dive type only
                 plt.figure(figsize=(30,5))
-                colors = [cm.get_cmap('tab20')(i) for i in [1,0]]
+                colors = [cm.get_cmap('tab10')(i) for i in [0,1]]
[…]

### 2020-07-07 Report/Sections/appendix.tex
Updates from Overleaf

@@
 We pick $\tilde{f} = 10$ periods per window, or 5 hertz.
-Note that $Z^{*(1)}_{t,s^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,s^*}\right)$, so selecting $\sigma^{*(i)}$ and $\phi^{*(1)}$ appropriately will ensure the distributions from (eqn \ref{}).
+Note that $Z^{*(1)}_{t,s^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,s^*}\right)$, so selecting $\sigma^{*(i)}$ and $\phi^{*(1)}$ appropriately will ensure correct distribution.
 
@@
 $$Z^{*(2)} = \sum_{k=1}^{10} b^{(k)}_{t,s^*}$$
-$$Z^{*(2)} \sim {\rm{Gamma}}
+$$Z^{*(2)} \sim {\rm{Gamma}}$$
 

### 2020-07-07 Report/Sections/intro.tex
Nancy Revisions
Updates from Overleaf

--- 2020-05-14 Report/Sections/intro.tex
@@
 % !TeX root = ../main.tex
+
+%\section{Introduction}
+
+The field of animal movement is in the midst of a ``data renaissance" where advancements in tagging technology have given rise to an explosion of data available for statistical modeling.
+In particular, tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
+In response, researchers have introduced a variety of new statistical techniques to infer animal behavior from movement data \citep{Hooten:2017}.
+
+One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}.
+Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another conditioned on the underlying behavioral process.
+However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies.
+For example, the location of an animal at a given time is highly correlated with the location of that animal one second later.
+Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) by Whoriskey et al \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model (CarHMM) by Lawer et al.
+\citep{Lawler:2019}.
+The CarHMM in particular explicitly models auto-correlation into an HMM while maintaining the structure needed to run the forward algorithm.
+It also only adds one additional parameter per possible hidden state
+
+Another issue that arises in high-frequency data is that several simultaneous behavioral processes may occur at different time scales.
+In this work we consider the example where killer whales perform a variety of different types of dives at a coarse scale, but also exhibit many different types of swimming behaviors within dives at a fine scale.
+One solution to this issue is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017} \citep{Adam:2019}.
+HHMMs model the entire time series in question as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.
+
+At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs.
+Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}.
+One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation.
+Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals.
+However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
+
+This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency.
+We describe a general procedure that can be used to extract features from highly structured fine-scale behaviors that otherwise could not be modeled with existing HMM models.
+In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions.
+We then perform a simulation study to compare the performance each existing model with ours in a controlled setting.
+Finally, we apply our method to dive data collected from a Northern resident killer whale off the coast of British Columbia, Canada.
+
+We consider several models- the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM).
+The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes.
+The CarHMM  modifies the usual HMM by modelling additional dependence into the sequence of observed data.
+In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.
+
+%%%%%%%%%% Nancy's comments %%%%%%%%%%%%%
+
+\iffalse
 
@@
 
-The field of animal movement is in the midst of a ``data renaissance" in which advancements in tagging technology have given rise to an explosion of data available for statistical modeling.
-In response, researchers have introduced a variety of new statistical techniques to infer animal behavior from movement data \cite{Hooten:2017}.
+{\bf{NH: These are my comments.  I hope they help you.}}
 
-These techniques usually fall into one of two different categories: continuous-time models and discrete-time models.
-Continuous-time models often model the dynamics of an animal as the solution to some stochastic differential equation and can incorporate observations taken at irregular time intervals.
-These methods are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
-On the other hand, discrete-time techniques use summary statistics taken at discrete time steps to model the animal's behavior and require that observations be taken at regular time intervals.
-The primary advantage of discrete-time methods is that the associated likelihood is relatively easy to compute, facilitating fast maximum likelihood estimates for parameters.
+This is where you describe how your method contributes to statistical methodology.
+That is the big goal.
+So there will be literature review of directly relevant work.
+You don't want to leave the non-expert completely lost though, so it is a balance, how much background you give.
+Thank goodness for the Zucchini book, a great place to direct readers for further background.  
+The introduction is a good place to put information about the motivating data set and how researchers have modelled it, what the shortcomings have been.
+I see that you have some great statements about this in your Background section.
+Those statements can be moved to the introduction (with editing -- you can see how I edited).
 
-One of the most prevalent discrete-time techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \cite{Patterson:2017}.
-Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another conditioned on the underlying behavioral process.
-However, this assumption is often violated in the real world processes.
-For example, the location of an animal at a given time is highly correlated with the location of that animal one second later.
-Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) by Whoriskey et al \cite{Whoriskey:2016}.
+A hard part about the introduction is that you want to use terminology like CarHMM without getting into too many technical details.
+Sometimes you just have to be technical.
+But I think you can avoid a lot of that.
 
-\textit{The conditionally autoregressive hidden Markov model (CarHMM): Inferring Behavioral States from animal tracking data exhibiting conditional autocorrelation} by Ethan Lawler, Kim Whoriskey, William H.
-Aeberhard, Chris Fields, and Joanna Mills Flemming introduces a new framework to model autocorrelation in hidden Markov models.
-The paper also recommends several best practices for data exploration, model selection, and data preprocessing.
-This paper summarizes the findings of Lawler et al, elaborates on some of the findings and recommendations within, and draws a connection between the CarHMM and a continuous-time method presented by Michelot et al \cite{Michelot:2019}.
-The CarHMM is also adapted and applied to real-world dive data to analyze the intra-dive behavior of a killer whale off the coast of British Columbia, Canada.
+Here is a great thing you wrote that would go well in the introduction, with some editing so that it doesn't rely on notation that hasn't been introduced.
+You could introduce the basic HMM notation in the intro, if you need to.
+I don't think introducing HMM notation is too technical.
+\begin{quote}
+ One of the key assumptions of both HMMs and HHMMs is \textit{conditional independence} between observations on both the coarse and fine scale.
+Namely, when given the state $X_t$, $Y_t$ is assumed to be independent from all other observations, including $Y_{t-1}$.
+Therefore, traditional HMMs and HHMMs can fail when the observations $Y$ exhibit significant correlation in time.
+ [leave in these examples for intro]  Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swimming behavior of horn sharks off the coast of Southern California \citep{Adam:2019}.
+
+One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \citep{Lawler:2019}.
+.....
+The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm.
+It also only adds one additional parameter per possible hidden state
+$(\phi^{(i)})$. 
+\end{quote}
+
+\fi

### 2020-07-07 Report/Sections/model.tex
Updates from Overleaf

@@
 
-\section{Model Formulation}
+%\section{Model Formulation}
 
-\subsection{Short-time Fourier Transform on Fine-scale processes}
+\subsection{Moving windows and the Short-time Fourier transform}
+\label{subsec:STFT}
 
@@
 For example, swimming behaviour of marine mammals is often exhibits periodic since the animal repeatedly flukes (swims up and down) to propel itself forward.
-Work has been done in the past to model non-Markovian dynamics in the \textit{behavioural} process $X^*_t$ \cite{Langrock:2012}, but to our knowledge addressing non-Markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area \textcolor{red}{(MAYBE?)}.
+Work has been done in the past to model non-Markovian dynamics in the \textit{behavioural} process $X^*_t$ \citep{Langrock:2012}, but to our knowledge addressing non-Markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area \textcolor{red}{(MAYBE?)}.
 With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine scale behavior is likely to persist.
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht]
 	\centering
-	\includegraphics[width=5.5in]{../Plots/fourier_transform.png}
+	\includegraphics[width=5in]{../Plots/fourier_transform.png}
 	\caption{Visualization of transforming $Y^*_t$ into $Z^*_t$ using a sliding window and Fourier transform.}
@@
 
-Note that it is possible to accommodate for unequal time steps within each window by using the \textbf{non-uniform discrete Fourier transform (NDFT)}.
+It is possible to accommodate for unequal time steps within each window by using the \textbf{non-uniform discrete Fourier transform (NDFT)}.
 We do not describe the details of this method in this work, but the generalization is straightforward.
-Refer to Bagchi et al \cite{Bagchi:2001} for details.
+Refer to Bagchi et al \citep{Bagchi:1999} for details.
 
-Fourier analysis has previously been used in the field of animal movement to explain animal behavior \cite{Fehlmann:2017} and specifically fluking \cite{Shorter:2017} from accelerometer data, and Fourier analysis within an HMM has been used to describe daily behavioral cycles of marine mammals  \cite{Heerah:2017}.
-However, we are currently unaware of any studies which have incorporated Fourier analysis of accelerometer data within the structure of an HMM.
+The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}.
+In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data.
+Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appeared a promising simple approach to account for additional correlation in data that is cyclical in nature.
 
 \subsection{Model Structure: combining the HHMM and CarHMM}
+\label{subsec:model_structure}
 
@@
 %
-\begin{figure}[h!]
+\begin{figure}[ht]
 	\centering

### 2020-07-07 Report/Sections/simulation.tex
Updates from Overleaf

@@
 
-\section{Simulation Study}
+%\section{Simulation Study}
 
@@
 
-\begin{figure}[!ht]
+\begin{figure}[ht]
 	\centering
@@
 
-\begin{figure}[t!]
+\begin{figure}[ht]
     \centering
@@
 The CarHMM model in particular severely underestimates $\sigma$.
-This is likely due to the fact that dive duration follows a mixture of gammas rather than a gamma distribution, and the CarHMM does not have the machinary to capture this fact.
+This is likely due to the fact that dive duration follows a mixture of gammas rather than a gamma distribution, and the CarHMM does not have the machinery to capture this fact.
 A full table of parameter estimates for all models is shown in (table \ref{table:dive_duration}).
@@
 
-\begin{table}[]
+\begin{table}
 \centering
@@
 $\pm$ refers to the IQR.}
-\scalebox{0.8}{
+\scalebox{0.7}{
 \begin{tabular}{ccccccc}

### 2020-07-07 Report/Sections/background.tex

@@
 
-\section{Background}
+%\section{Background}
 
@@
 The probability transition matrix $\Gamma$ and parameters of the emission distributions $\Theta$ can be estimated by maximizing the likelihood of $y$, which we denote as $\calL_{\text{HMM}}(y;\Theta,\Gamma)$, with respect to $\Gamma$ and $\Theta$.
-In addition, $\calL_{\text{HMM}}(y;\theta,\Gamma)$ can be calculated using the well-known \textit{forward algorithm} \cite{Zucchini:2016} as follows:
+In addition, $\calL_{\text{HMM}}(y;\theta,\Gamma)$ can be calculated using the well-known \textit{forward algorithm} \citep{Zucchini:2016} as follows:
 %
@@
 %
-To remove constraints when maximizing $\calL_{\text{HMM}}(y)$, we parameterize $\Gamma$ using $\eta \in \bbR^{N \times N}$ and the following link function \cite{Barajas:2017}:
+To remove constraints when maximizing $\calL_{\text{HMM}}(y)$, we parameterize $\Gamma$ using $\eta \in \bbR^{N \times N}$ and the following link function \citep{Barajas:2017}:
 %
@@
 Leos-Barajas et al.
-\cite{Barajas:2017} encounter a similar issue when modeling the movement of a harbor porpoise in the North Sea, and use it as a motivating example when they introduce hierarchical hidden Markov models.
+\citep{Barajas:2017} encounter a similar issue when modeling the movement of a harbor porpoise in the North Sea, and use it as a motivating example when they introduce hierarchical hidden Markov models.
 
@@
 A hierarchical hidden Markov model (or HHMM) is a variation of a hidden Markov model in which $X_t$ satisfies the conditions of section \ref{subsection:HMMs}, but in addition it ``emits" yet another fine-scale hidden Markov model of length $T_t^*$.
-This fine-scale hidden Markov model can be either instead of \cite{Barajas:2017} or in addition to \cite{Adam:2019} a coarse-scale observation $Y_t$, which itself must satisfy the conditions from section \ref{subsection:HMMs}.
+This fine-scale hidden Markov model can be either instead of \citep{Barajas:2017} or in addition to \citep{Adam:2019} a coarse-scale observation $Y_t$, which itself must satisfy the conditions from section \ref{subsection:HMMs}.
 The fine-scale HMM is comprised of a fine-scale Markov chain $X^*_t = (X^*_{t,1}, \ldots, X^*_{t,T^*_t})$ and fine-scale observations $Y^*_t = (Y^*_{t,1}, \ldots, Y^*_{t,T^*_t})$, each of length $T^*_t$.
@@
 Note that this formulation assumes that the coarse-scale observations at a given time $Y_t$ and the fine-scale observation time series $Y_t^*$ are independent of one another when conditioned on $X_t$.
-For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, model selection and model checking, see Adam et al \cite{Adam:2019}.
+For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, model selection and model checking, see Adam et al \citep{Adam:2019}.
 
@@
 Therefore, traditional HMMs and HHMMs can fail when the observations $Y$ exhibit significant correlation in time.
-Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swimming behavior of horn sharks off the coast of Southern California \cite{Adam:2019}.
+Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swimming behavior of horn sharks off the coast of Southern California \citep{Adam:2019}.
 
-One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}.
+One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \citep{Lawler:2019}.
 In particular, the CarHMM describes the distribution of $(Y_t|Y_{t-1}, X_t=i)$ with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{2(i)},\phi^{(i)}\}$, where:
@@
 
-This can be done in one of two ways: first, the most likely \textit{sequence} of \text{all} hidden states can be found using a dynamic programming algorithm called the Viterbi algorithm \cite{Viterbi:1967}.
+This can be done in one of two ways: first, the most likely \textit{sequence} of \text{all} hidden states can be found using a dynamic programming algorithm called the Viterbi algorithm \citep{Viterbi:1967}.
 Alternatively, the probability of \textit{each} coarse-level state (conditioned on the learned parameters) can be found using the \textit{forward-backward algorithm}.

### 2020-07-07 Report/Sections/carhmm_ou.tex
Nancy Revisions

@@
 
-\subsection{Continuous Time models}
+%\subsection{Continuous Time models}
 
 Although the CarHMM can incorporate auto-correlation into the structure of an HMM, they break down when observations are taken at irregular time intervals since the probability transition matrix $\Gamma$ would then have to depend upon the time step between observations.
-A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal.
+A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \citep{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal.
 Namely, a state-switching Ornstein-Uhlenbeck process $Y$ is the solution to the following stochastic differential equation:
@@
 One notable difference between the state-switching OU process and the CarHMM is that the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer.
-If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \cite{Michelot:2019}:
+If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
 %
 \begin{align}
-Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)}^2}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
+Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
 \label{eqn:OU_sol}
@@
 The probability transition matrix $\bfA$ must be right stochastic, meaning that its entries must be between 0 and 1 and its rows must add to one.
-Lawler et al do not go into detail about how to parameterize $\bfA$ such that it is both identifiable and constraint free, but the following parameterization does the trick \cite{Barajas:2017}:
+Lawler et al do not go into detail about how to parameterize $\bfA$ such that it is both identifiable and constraint free, but the following parameterization does the trick \citep{Barajas:2017}:
 %
@@
 $\mathbf{1}$ is a column vector of ones.
-Once the emission and transition parameters have been estimated, the most likely sequence of underlying behavioral states $\bfB$ can be found using the Viterbi algorithm \cite{Viterbi:1967}.
+Once the emission and transition parameters have been estimated, the most likely sequence of underlying behavioral states $\bfB$ can be found using the Viterbi algorithm \citep{Viterbi:1967}.
 
@@
 Lawler et al are not the first to incorporate auto correlation within a hidden Markov model in an ecological setting.
-Woriskey et al \cite{Whoriskey:2016} developed a hidden Markov model called the HMMM, which is formulated as the following:
+Woriskey et al \citep{Whoriskey:2016} developed a hidden Markov model called the HMMM, which is formulated as the following:
 %
@@
 For example, Michelot et al.
-\cite{Michelot:2019} model $\bfY$ as an integrated Ornstein-Uhlenbeck process, where the position $\bfY_t$ and velocity $\bfV_t$ are modeled as draws from a normal distribution which depends upon $\bfV_{t-1}$, $\bfY_{t-1}$, the behavioral state $b_{t-1}$, and the time step between observations $\Delta_{t-1}$.
+\citep{Michelot:2019} model $\bfY$ as an integrated Ornstein-Uhlenbeck process, where the position $\bfY_t$ and velocity $\bfV_t$ are modeled as draws from a normal distribution which depends upon $\bfV_{t-1}$, $\bfY_{t-1}$, the behavioral state $b_{t-1}$, and the time step between observations $\Delta_{t-1}$.
 It is straightforward to adapt this continuous-time approach to an HMM by setting $\Delta_{t}$ to be constant for all values of $t$ and assuming that the behavior of the animal does not change between time steps.

### 2020-07-07 Report/Sections/model_Nancy.tex

@@
 
-We consider several models for the data, the classical hierarchical hidden Markov model (HHMM) (reference), the CarHMM model of (reference), and our new model that blends the two (CarHHMM).
-The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes.
-The CarHMM  modifies the usual HMM by modelling additional dependence into the sequence of observed data.
-In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.
-
-{\bf{NH: It strikes me that the above might belong in the introduction.  In that case, the what appears above in this section would be modified to something much shorter. 
-}} 
-We first remind the reader of the definition of a hidden Markov model,   the model of the coarse-scale process.
+We first remind the reader of the definition of a hidden Markov model and extend this to the CarHMM.
+We then introduce hierarchical structures within HMMs using both the short-time Fourier transform (STFT) and the hierarchical hidden Markov model (HHMM)
 
@@
 The $X_t$s form a Markov chain and take possible values $1, \ldots, N$.
-Their distribution is governed by the distribution of the initial state $X_1$ and the  $N \times N$ transition probability matrix $\Gamma$ where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$, $t=1,\ldots, T-1$, ~$i, j = 1,\ldots, N$.
+Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$ where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$, for $t=1,\ldots, T-1$, and $i, j = 1,\ldots, N$.
 %
@@
 %
-Recall that a Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$.
-
+Recall that a Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
 %
-The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$, that is, the distribution of  $Y_t$ conditional on $X_1,\ldots, X_T$ and $\{Y_1,\ldots, Y_T\}/ \{Y_t\}$ is simply the distribution of $Y_t$ given $X_t$.
+The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
 %
@@
 %
-A visualization of the dependence structure can be seen in Figure \ref{fig:HMM}.
+(Fig \ref{fig:HMM}) represents the dependence structure.
 
 To find the maximum likelihood estimates of the parameters $\Gamma$ and $\Theta \equiv (\theta^{(1)},\ldots,\theta^{(N)})$, let $y = (y_1,\ldots,y_T)$ be the observed emissions. 
- To facilitate calculations we write the likelihood $\calL_{\text{HMM}}$ in a specific form, using the  well-known \textit{forward algorithm} \cite{Zucchini:2016} as follows:
+ To facilitate calculations we write the likelihood $\calL_{\text{HMM}}$ in a specific form, using the  well-known \textit{forward algorithm} \citep{Zucchini:2016} as follows:
 %
@@
 
-We will always parameterize transition probability matrices to remove constraints that the entries of the matrix are non-negative and that  the rows sum to 1.
-We do this by parameterizing the $N \times N$ transition probability matrix $\Gamma$ using $\eta \in \bbR^{N \times N}$ \cite{Barajas:2017}:
+We will always parameterize transition probability matrices to remove the constraints that the entries of the matrix are non-negative and that the rows sum to 1.
+We do this by parameterizing the $N \times N$ transition probability matrix $\Gamma$ using $\eta \in \bbR^{N \times N}$ \citep{Barajas:2017}:
 %
@@
 
-\begin{figure}[h!]
+\begin{figure}[ht]
 	\centering
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{The conditionally auto-regressive hidden Markov model (CarHMM)}
+
+One of the key assumptions of an HMM is \textit{conditional independence} between observations, given the state sequence.
+Therefore, traditional HMMs do not hold when the observations exhibit certain forms of significant correlation in time.
+
+One way to model extra correlation in an HMM type structure is via a CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by \citep{Lawler:2019}.
+Like a traditional HMM, A CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution.
+Unlike a traditional HMM, the CarHMM assumes that the distribution of the $t$th emission, $Y_t$, conditional on $X_1,\ldots, X_T$ and $ \{Y_1,\ldots, Y_{t-1}\}$, depends on both $X_t$ \textit{and} $Y_{t-1}$.
+The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. (Fig \ref{fig:CarHMM}) shows the structure of a CarHMM.
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=3in]{../Plots/CarHMM.png}
+	\caption{Graphical representation of a traditional CarHMM.}
+	\label{fig:CarHMM}
+\end{figure}
+
+%
+We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1})$ or as
+$f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$.
+As an example, we could assume that this conditional distribution is normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{2(i)},\phi^{(i)}\}$, with:
+%
+\[
+\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}
+\]
+and
+\[
+\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = \sigma^{2(i)}.
+\]
+%
+{\bf{Ah - now I see the issue about changing states.
+It would be great to have something about the distribution of $Y_t$ given $X_t=i, X_{t-1}=j$.
+You'd want a different conditional distribution, depending on if $i=j$ or not.
+}}
+%
+The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to write the likelihood using the forward algorithm.
+In our normal model formulation, we have added only one additional parameter, $\phi^{(i)}$,  per possible hidden state.
+
+The likelihood for the CarHMM can be easily calculated using the forward algorithm.  As previously, let $y$ be the vector of observed emissions.  Then
+\begin{equation}
+\calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N
+\label{CarHMM_likelihood}
+\end{equation}
+where
+%
+$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
+%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{The hierarchical hidden Markov model (HHMM)}
-
-{\bf{NH:
-I wonder if it would be easier on the reader to put the CarHMM first, since the CarHMM deals with just one Markov chain.
-}}
 
@@
 
-For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, model selection and model checking, see Adam et al \cite{Adam:2019}.
+For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, model selection and model checking, see Adam et al \citep{Adam:2019}.
 
-\begin{figure}[h!]
+\begin{figure}[ht]
 	\centering
-	\includegraphics[width=6.5in]{../Plots/HHMM.png}
+	\includegraphics[width=5in]{../Plots/HHMM.png}
 	\caption{Graphical representation of a traditional HHMM.}
@@
 \end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Conditionally auto-regressive hidden Markov models (CarHMM)}
-
-{\bf{NH: 
-I've taken what you have written and edited.
-
-Here is a comment about a big change I made.
-You give a definition of the distribution of the correlated emissions in terms of means and variances, and I think you want to say something in a more general form, that is, that the distribution of $Y_t$ is determined by $X_t$ and $Y_{t-1}$.
-Also,  you only define the conditional variance of $Y_t$ given $X_t$,  excluding $Y_{t-1}$.
-You should have the left side of that equation for the variance be conditional on both $X_t$ and $Y_{t-1}$ for your math to go through.
-It's no problem if you want to assume that the variance on the right side only has $i$ in it.
-That is an assumption about the distribution of $Y_t$  conditional on both $X_t$ and $Y_{t-1}$
-}}
-
-One of the key assumptions of an HMM is \textit{conditional independence} between observations, given the state sequence.
-Therefore, traditional HMMs do not hold when the observations exhibit certain forms of significant correlation in time.
-The same issue arises in HHMMs.
-
-One way to model extra correlation in an HMM type structure is via a CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}. 
-In a CarHMM, we have a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution.
-The CarHMM assumes that the distribution of the $t$th emission, $Y_t$, conditional on $X_1,\ldots, X_T$ and $ \{Y_1,\ldots, Y_T\}/\{Y_t\}$, depends only on $X_t$ and $Y_{t-1}$. 
-{\bf{NH:  True?  or just conditional on the $X_t$s and $Y_1,\ldots, Y_{t-1}$?}}
-The first emission $Y_1$ is assumed to be fixed as an initial value.
-%
-{\bf{NH: does $Y_1$ depend on $X_1$?}}
-%
- The graphical model associated with the structure of a CarHMM is shown in Figure (\ref{fig:CarHMM}).
-
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=3in]{../Plots/CarHMM.png}
-	\caption{Graphical representation of a traditional CarHMM. The additional arrows representing auto-correlation between observations are shown in red for emphasis.}
-	\label{fig:CarHMM}
-\end{figure}
-
-%
-We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1})$ or as
-$f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$.
-As an example, we could assume that this conditional distribution is normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{2(i)},\phi^{(i)}\}$, with:
-%
-\[
-\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}
-\]
-and
-\[
-\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = \sigma^{2(i)}.
-\]
-%
-{\bf{Ah - now I see the issue about changing states.
-It would be great to have something about the distribution of $Y_t$ given $X_t=i, X_{t-1}=j$.
-You'd want a different conditional distribution, depending on if $i=j$ or not.
-}}
-%
-The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to write the likelihood using the forward algorithm.
-Note that in our normal model formulation,  we have added only one additional parameter, $\phi^{(i)}$,  per possible hidden state.
-
-The likelihood for the CarHMM can be easily calculated using the forward algorithm.  As previously, let $y$ be the vector of observed emissions.  Then
-\begin{equation}
-\calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N
-\label{CarHMM_likelihood}
-\end{equation}
-where
-%
-$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
-%
 

### 2020-07-07 Report/Sections/results.tex

--- 2020-07-05 Report/Sections/results.tex
@@
 Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second.
-Data preprocessing was done in part with the \textit{divebomb} package in Python \cite{Nunes:2018}.
+Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}.
 After preprocessing the raw data, a total of 267 dives were observed.
-A plot of the processed data for all dives and one selected dive can be seen in figures \ref{fig:data} and \ref{fig:data_one_dive}, respectively.
+A plot of the raw data for all dives and a collection of five selected dives can be seen in figures \ref{fig:data} and \ref{fig:data_one_dive}, respectively.
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=5in]{../Plots/raw_data.png}
+	\caption{Dive profile and Acceleration data of entire data set}
+	\label{fig:data}
+\end{figure}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=5in]{../Plots/raw_data_5_dives.png}
+	\caption{Dive profile and acceleration data for a collection of 5 dives of a killer whale.}
+	\label{fig:data_one_dive}
+\end{figure}
 
@@
 The collection of all dive durations were set to be the coarse-scale observations $Y$, and the acceleration data was used to determine the fine-scale observations $Y^*$.
-Figure (\ref{fig:raw_data_one_dive}) shows both the depth profile and the raw accelerometer data for one specific dive.
-The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs without some kind of signal processing.
-Therefore, the STFT was used to calculate both $Y{*(1)}$ and $Y^{*(2)}$ as described in section \ref{subsec:STFT}.
+The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs without some kind of signal processing (see fig \ref{fig:data_one_dive}).
+Therefore, the STFT was used to calculate both $Y^{*(1)}$ and $Y^{*(2)}$ as described in section \ref{subsec:STFT}.
 We set $\tilde{f} = 5$ hertz and $h = 2$ seconds, which reduced the dimension of each window from $50 s^{-1} * 2 s = 100$ to $2$.
 
-In order to determine if the CarHHMM was appropriate for this data, a lag plot was made for both $Y^{*(1)}_{t,s^*}$ and $Y^{*(2)}_{t,s^*}$, as shown in figure (\ref{fig:lag}).
-The number of behavioral states is not clear from the lag plot, but it is clear that $Y^{*(1)}_{t,s^*}$ exhibits a large degree of auto-correlation.
-While $Y^{*(2)}_{t,s^*}$ also exhibits some auto-correlation, the relationship is less strong, so auto-correlation was not incorporated in the emission distribution of $Y^{*(2)}_{t,s^*}$.
+To determine if the CarHHMM was appropriate for this data, a lag plot was made for both $Y^{*(1)}_{t,s^*}$ and $Y^{*(2)}_{t,s^*}$, as shown in figure (\ref{fig:lag}).
+The number of behavioral states is not clear from the lag plot, but it is clear that $Y^{*(1)}_{t,s^*}$ exhibits a large degree of auto-correlation, so the CarHMM was used for $Y^{*(1)}_{t,s^*}$.
+The auto-correlation parameter $\hat phi$ was set to be shared among all dimensions of $Y^{*(1)}_{t,s^*}$, but both the means and variances were allowed to be different.
+While $Y^{*(2)}_{t,s^*}$ also exhibits some auto-correlation, the relationship is less strong.
+In addition, the biological interpretation of auto-correlation within $Y^{*(2)}_{t,s^*}$ is more difficult than for $Y^{*(1)}_{t,s^*}$, so auto-correlation was not incorporated into the emission distribution of $Y^{*(2)}_{t,s^*}$.
 
-\begin{figure}[h!]
+\begin{figure}[ht]
 	\centering
-	\includegraphics[height=5in]{../Plots/lagplot.png}
-	\caption{Lag plot of vertical velocity and $\tilde a$ (left) and a the associated normal kernel density estimates (right)}
+	\includegraphics[height=7in]{../Plots/lagplot.png}
+	\caption{Lag plot of all features (left) and the associated normal kernel density estimates (right)}
 	\label{fig:lag}
@@
 
-Information criteria tends to overestimate the number of states in biological processes \cite{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically and admittedly somewhat arbitrarily.
-This is a common inssue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria.
+It is known that information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically and admittedly somewhat arbitrarily.
+The absence of principled tool to select the number of hidden states is a common issue in statistical ecology.
+Therefore, it is important to use model validation techniques in lieu of information criteria.
 Section \ref{subsec:model_validation} describes our process of validating this model in particular.
@@
 
-The parameters of the estimated emission distributions for each behavioral state are shown in table (\ref{table:emis_dist}).
-Each distribution is also plotted in figures (\ref{fig:coarse_emis}) and (\ref{fig:fine_emis}).
-Note that the auto-correlation within the velocity sequence is not captured in figure (\ref{fig:emis_dist}), so it is important to refer to the estimated auto-correlation parameter $\hat \phi$ from table (\ref{table:emis_dists}) when considering the emission distributions shown in figure (\ref{fig:fine_emis}).
+The parameters of the estimated emission distributions for each behavioral state are shown in (table \ref{table:emis_dists}).
+Each distribution is also plotted in (fig \ref{fig:coarse_emis}) and (fig \ref{fig:fine_emis}).
+Note that the auto-correlation within the acceleration sequence is not captured in (fig \ref{fig:fine_emis}), so it is important to refer to the estimated auto-correlation parameter $\hat \phi$ from (table \ref{table:emis_dists}) when considering the emission distributions shown in (fig \ref{fig:fine_emis}).
 %
-\begin{figure}[h!]
+\begin{figure}[ht]
 	\centering
 	\includegraphics[height=4in]{../Plots/coarse-emissions.png}
-	\caption{Estimated probability distributions for the long-term vertical velocity and $\tilde a$ in each behavioral state.
-Note that the distribution over vertical velocity does not take auto-correlation into account.}
+	\caption{Estimated probability distributions for each coarse-scale observation in each dive type.}
 	\label{fig:coarse_emis}
@@
 %
-\begin{figure}[h!]
+\begin{figure}[ht]
 	\centering
 	\includegraphics[height=2.5in]{../Plots/fine-emissions.png}
-	\caption{Estimated probability distributions for the long-term vertical velocity and $\tilde a$ in each behavioral state.
-Note that the distribution over vertical velocity does not take auto-correlation into account.}
-	\label{fig:coarse_emis}
+	\caption{Estimated probability distributions for each fine-scale observation in each behavioral state. Note that the distributions of acceleration does not take auto-correlation into account.}
+	\label{fig:fine_emis}
 \end{figure}
 %
-\begin{table}[h!]
+\begin{table}[ht]
     \centering
-    \label{table:emis_dists}
     \caption{Estimates and standard errors of emission parameters for killer whale data.}
+    \scalebox{0.8}{
     \begin{tabular}{ccccc}
-    \multirow{2}{*}{Feature}                 & \multirow{2}{*}{Dive / Subdive Type} & \multicolumn{3}{c}{Parameter Estimate}              \\
+    \multirow{2}{*}{Feature}                 & \multirow{2}{*}{Dive / Sub-dive Type} & \multicolumn{3}{c}{Parameter Estimate}              \\
                                              &                                      & $\hat \mu$      & $\hat \sigma$   & $\hat \phi$     \\ \hline
@@
     \end{tabular}
+    }
+    \label{table:emis_dists}
 \end{table}
 %
+While the ecological meaning of the modeled behavioral states is tenuous, we hypothesize the following interpretations.
+For the coarse scale, dive type 1 corresponds to shorter, shallower dives which serve a variety of purposes, including to rest before performing a dive of type 2, which corresponds to a deeper, more sustained dive.
+For the fine scale, behavioral state 1 corresponds to gliding and less overall activity of the whale compared to the other behavioral states.
+The mean of $Y^{*(2)}$ in this state is at least an order of magnitude smaller than sub-dive behavior 2, the variance of $Y^{*(1)}$ is smaller than sub-dive behavior 2, and the auto-correlation of $Y^{*(1)}$ is higher than every other behavioral state.
+This indicates a relatively constant acceleration and therefore low levels of activity.
+Sub-dive state 3 corresponds to vigorous swimming activity, as the mean of $Y^{*(2)}$ and variance of $Y^{*(1)}$ are much higher than every other state.
+The auto-correlation of $Y^{*(1)}$ is also much lower in this state, implying more variation in acceleration every 2 seconds.
+Sub-dive state 2 corresponds to a medium amount of activity, as almost every parameter estimate is between the other two behavioral states.
+
 The estimated probability transition matrices and associated stationary distributions are shown below:
@@
 %
-Finally, the decoded dive behavior within a selected dive (deeper than 20 meters) is shown in figure (\ref{fig:viterbi}).
+About 86\% of observed dives are short- the whale usually rests for many dives in a row before performing a deep dive.
+Interestingly, the probability transition matrix $\hat \Gamma$ shows that the current dive does not effect the dive type of the following dive much.
+In addition, the fine-scale probability transition matrices imply that the killer whale is much more likely to be in a less active sub-dive state when performing deep dives than when performing shallow dives- 14\% of shallow dives are spent in sub-dive state 1 while 48\% of deep dives are spent in sub-dive state 1.
+Using less active sub-dive states when diving deep could be an energy reduction strategy for these long periods of holding breath.
 
-\begin{figure}[h!]
+The decoded dive behavior within 5 selected dives is shown in (fig \ref{fig:labeled_dives}).
+In addition, the probability of each dive type and sub-dive state is shown in (fig \ref{fig:coarse_probs}) and (fig \ref{fig:fine_probs}), respectively.
+
+\begin{figure}[ht]
 	\centering
-	\includegraphics[height=8in]{../Plots/viterbi.png}
-	\caption{Features of a particular killer whale dive and Viterbi-decoded estimates for the intra-dive behavioral states.}
-	\label{fig:viterbi}
+	\includegraphics[width=5in]{../Plots/labeled_dives.png}
+	\caption{Features of a particular set of killer whale dives and decoded estimates for the intra-dive behavioral states.
+The color of the plot corresponds to behavioral or dive state with the highest probability.}
+	\label{fig:labeled_dives}
+\end{figure}
+%
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=5in]{../Plots/Coarse_state_probs.png}
+	\caption{Probabilities of dive types for the set of killer whale dives from (fig \ref{fig:labeled_dives}).}
+	\label{fig:coarse_probs}
+\end{figure}
+%
+\begin{figure}[ht]
+	\centering
+	\begin{subfigure}[t]{1.0\textwidth}
+        \centering
+        \includegraphics[width=5in]{../Plots/Fine_state_probs_1.png}
+        \caption{Fine-scale state 1 probabilities}
+    \end{subfigure}
+    \newline
+    \begin{subfigure}[t]{1.0\textwidth}
+        \centering
+        \includegraphics[width=5in]{../Plots/Fine_state_probs_3.png}
+        \caption{Fine-scale state 3 probabilities}
+    \end{subfigure}
+	\caption{Probabilities of sub-dive types for the set of killer whale dives from (fig \ref{fig:labeled_dives}).}
+	\label{fig:fine_probs}
 \end{figure}
 
-While the ecological meaning of these behavioral states is tenuous, we hypothesize the following interpretations.
-Roughly speaking, behavioral state 1 corresponds to mildly active swimming.
-The mean of $\tilde a$ in this state is larger than behavioral state 2, indicating more activity by the whale, but the autocorrleation term $\hat \phi$ is still very high, indicating that velocities don't change very much every two seconds.
-Behavioral state 2 corresponds to gliding or turning, where no active swimming is taking place.
-It is characterized by a low mean $\tilde a$ value and a very high autocorrelation term $\hat \phi$, both of which indicate little activity.
-Finally, behavioral state 3 represents sudden jerking or more vigorous and active swimming than behavioral state 1.
-This behavior is so rigorous that $\hat \phi$ drops almost to zero and $\hat \sigma^2$ subsequently rises significantly.
-In behavioral state 3, the velocity readings of the killer whale are essentially uncorrelated when recorded every two seconds.
+\subsection{Model Validation}
+\label{subsec:model_validation}
 
-\subsection{Future Work}
+Two visual tools were used to evaluate the model: pseudo-residuals and empirical histograms.
+A pseudo-residual of a particular observation is the marginal cdf of an observation, conditioned on all other observations, and mapped to the quantile function of a standard normal distribution (see \citep{Zucchini:2016} for details).
+In particular, the pseudo-residual of an observation $y_{t,s^*}$ is $\Phi^{-1} \left(Pr(Y_{t,s^*} < y_{t,s^*}|\{Y_1,\ldots,Y_T\}/\{Y_{t,s^*}\}) \right)$, where $\Phi$ is the cdf of a standard normal distribution.
+If the model is well-specified, then all pseudo-residuals should be independent and follow a standard normal distribution.
+While histograms of the pseudoresiduals support this for most observations, one notable exception is the case of $Y^{*(2)}$, shown in (fig \ref{fig:pseudoresids}), which are noticeably right-skewed.
+This implies that the true distribution of $Y^{*(2)}$ may follow a heavier-tailed distribution compared to a gamma distribution such as a power law.
 
-This analysis is largely exploratory with a purely heuristic approach to decide the number of behavioral states.
-Future work should repeat the analysis using several different values for the number of behavioral states.
-It is also important to validate these results by plotting psuedoresiduals and performing goodness-of-fit tests on the emission distributions.
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=5in]{../Plots/pseudoresids.png}
+	\caption{Psuedoresiduals of $Y^{*(2)}$}
+	\label{fig:pseudoresids}
+\end{figure}
 
-Parameter estimation via likelihood maximization can be slow, especially with an observation rate of 50 hertz.
-If dives are modeled as uncorrelated batches of data, then using stochastic gradient ascent to maximize the likelihood may considerably speed up parameter estimation.
+In addition to psuedoresiduals, we plotted histograms of observations for each feature within each hidden state.
+In particular, observations were weighted by the probability that the whale was in a particular hidden state.
+This empirical distribution was then plotted over the estimated true pdf of that feature for the hidden state.
+Again, our results mostly support the model in question with the exception of $Y^{*(2)}$, which is right-skewed.
+In addition, $Y^{*(1)}$ has heavy tails for sub-dive state 3 (see fig \ref{fig:empirical_dist}), indicating the existence of rare events corresponding to very violent trashing of the killer whale.
+These outliers are potential subjects for future studies to better understand rare behavioral events for this killer whale.
 
-Finally, this work was run exclusively on dives deeper than 20 meters so that within-dive behaviors would be similar between dives.
-Future analysis could find a way to include a more heterogeneous set of dives using a framework such as the hierarchical hidden Markov model (HHMM) introduced by Leos-Barajas et al \cite{Barajas:2017}.
-
-\subsection{Model evaluation}
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=5in]{../Plots/empirical_dist.png}
+	\caption{Empirical distribution of $Y^{*(1)}_z$ for sub-dive state 3 plotted over its estimated pdf.}
+	\label{fig:empirical_dist}
+\end{figure}

### 2020-07-08 Report/Sections/appendix.tex
Commit before changing STFT to Fourier Transform

@@
 \setcounter{subsection}{0}
+
+\section{Results regarding CarHMMs and the Ornstein-Uhlenbeck process}
+
+\subsection{Continuous Time models}
+
+Although the CarHMM can incorporate auto-correlation into the structure of an HMM, they break down when observations are taken at irregular time intervals since the probability transition matrix $\Gamma$ would then have to depend upon the time step between observations.
+A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \citep{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal.
+Namely, a state-switching Ornstein-Uhlenbeck process $Y$ is the solution to the following stochastic differential equation:
+%
+\begin{equation}
+    \label{eqn:SDE}
+    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
+\end{equation}
+%
+where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
+One notable difference between the state-switching OU process and the CarHMM is that the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer.
+If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
+%
+\begin{align}
+Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
+\label{eqn:OU_sol}
+\end{align}
+%
+where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$.
+Continuous time models can be built up from arbitrarily complex stochastic differential equations and allow for uneven step lengths in the observations sequence $Y$.
+However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.
+
+\subsection{Equivalence of CarHMM and OU process}
+
+Although most continuous time models are difficult to incorporate into an HHMM, we show in (thrm 1) that under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process shown in (eqn \ref{eqn:OU_sol}).
+This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.
+
+\begin{theorem}{Theorem 1.}{}%
+If:
+\begin{enumerate}
+    \item The hidden behavioural process $X$ from (Eq.
+\ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
+    \item Observations of the SDE from (Eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
+\end{enumerate}
+then the observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
+
+\begin{align}
+\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{(i)} = \sqrt{\frac{\omega^{(i)^2}}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t})}, \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
+\end{align}
+
+\end{theorem}
+
+\begin{proof}{Proof}{}%
+Combining equation (\ref{eqn:CarHMM_to_OU}) with equation (\ref{eqn:OU_sol}) gives:
+%
+\begin{align*}
+\left(Y_{s \Delta t} | X_{(s-1)\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{(s-1) \Delta t}, \enspace \sigma^{2(i)} \right),\\
+s = 1, \ldots, T-1
+\end{align*}
+%
+If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t: (s-1) \Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:
+
+\begin{align*}
+\left(Y'_s| X'_s = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y'_{s-1}, \enspace \sigma^{2(i)} \right)\\
+s = 2, \ldots, T
+\end{align*}
+%
+$X'$ follows a Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s+1})$ is consistent with that of a CarHMM with normal emission distributions.
+\end{proof}
 

### 2020-07-08 Report/Sections/carhmm_ou.tex
Commit before changing STFT to Fourier Transform

@@
 % !TeX root = ../main.tex
-
-\section{Results regarding CarHMMs and the Ornstein-Uhlenbeck process}
-
-%\subsection{Continuous Time models}
-
-Although the CarHMM can incorporate auto-correlation into the structure of an HMM, they break down when observations are taken at irregular time intervals since the probability transition matrix $\Gamma$ would then have to depend upon the time step between observations.
-A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \citep{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal.
-Namely, a state-switching Ornstein-Uhlenbeck process $Y$ is the solution to the following stochastic differential equation:
-%
-\begin{equation}
-    \label{eqn:SDE}
-    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
-\end{equation}
-%
-where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
-One notable difference between the state-switching OU process and the CarHMM is that the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer.
-If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
-%
-\begin{align}
-Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
-\label{eqn:OU_sol}
-\end{align}
-%
-where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$.
-Continuous time models can be built up from arbitrarily complex stochastic differential equations and allow for uneven step lengths in the observations sequence $Y$.
-However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.
-
-\subsection{Equivalence of CarHMM and OU process}
-
-Although most continuous time models are difficult to incorporate into an HHMM, we show in (thrm 1) that under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process shown in (eqn \ref{eqn:OU_sol}).
-This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.
-
-\begin{theorem}{Theorem 1.}{}%
-If:
-\begin{enumerate}
-    \item The hidden behavioural process $X$ from (Eq.
-\ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
-    \item Observations of the SDE from (Eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
-\end{enumerate}
-then the observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
-
-\begin{align}
-\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{(i)} = \sqrt{\frac{\omega^{(i)^2}}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t})}, \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
-\end{align}
-
-\end{theorem}
-
-\begin{proof}{Proof}{}%
-Combining equation (\ref{eqn:CarHMM_to_OU}) with equation (\ref{eqn:OU_sol}) gives:
-%
-\begin{align*}
-\left(Y_{s \Delta t} | X_{(s-1)\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{(s-1) \Delta t}, \enspace \sigma^{2(i)} \right),\\
-s = 1, \ldots, T-1
-\end{align*}
-%
-If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t: (s-1) \Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:
-
-\begin{align*}
-\left(Y'_s| X'_s = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y'_{s-1}, \enspace \sigma^{2(i)} \right)\\
-s = 2, \ldots, T
-\end{align*}
-%
-$X'$ follows a Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s+1})$ is consistent with that of a CarHMM with normal emission distributions.
-\end{proof}
 
@@
 now $\calL_{\text{CarHMM}}(y;\theta,\Lambda,\delta)$ is simply maximized with respect to $\theta = \{\beta,\omega,\gamma\}$, $\Lambda = \{p,\lambda\}$, and $\delta$ to find the maximum likelihood estimate of the CarHMM with uneven time steps.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2020-07-08 Report/Sections/intro.tex
Commit before changing STFT to Fourier Transform

@@
 For example, the location of an animal at a given time is highly correlated with the location of that animal one second later.
-Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) by Whoriskey et al \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model (CarHMM) by Lawer et al.
-\citep{Lawler:2019}.
+Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model (CarHMM) \citep{Lawler:2019}.
 The CarHMM in particular explicitly models auto-correlation into an HMM while maintaining the structure needed to run the forward algorithm.
-It also only adds one additional parameter per possible hidden state
+It also only adds one additional parameter per possible hidden state.
 
@@
 
+For periodic behavior in particular, One way to avoid the use of continuous time models is to use signal processing techniques such as the Fourier transform on the raw data.
+The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}.
+In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data.
+Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appeared a promising simple approach to account for additional correlation in data that is cyclical in nature.
+
+We consider several models- the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM).
+The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes.
+The CarHMM modifies the usual HMM by modelling additional dependence into the sequence of observed data.
+In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.
+
 This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency.
@@
 Finally, we apply our method to dive data collected from a Northern resident killer whale off the coast of British Columbia, Canada.
-
-We consider several models- the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM).
-The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes.
-The CarHMM  modifies the usual HMM by modelling additional dependence into the sequence of observed data.
-In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.
 
@@
 
-One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \citep{Lawler:2019}.
+One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by \citep{Lawler:2019}.
 .....

### 2020-07-08 Report/Sections/models.tex
Commit before changing STFT to Fourier Transform

--- 2020-07-07 Report/Sections/model_Nancy.tex
@@
 	\centering
-	\includegraphics[width=3in]{../Plots/CarHMM.png}
+	\includegraphics[width=5in]{../Plots/CarHMM.png}
 	\caption{Graphical representation of a traditional CarHMM.}
@@
 $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$.
-As an example, we could assume that this conditional distribution is normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{2(i)},\phi^{(i)}\}$, with:
+As an example, we could assume that this conditional distribution is normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
 %
@@
 \[
-\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = \sigma^{2(i)}.
+\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = (\sigma^{(i)})^2.
 \]
-%
-{\bf{Ah - now I see the issue about changing states.
-It would be great to have something about the distribution of $Y_t$ given $X_t=i, X_{t-1}=j$.
-You'd want a different conditional distribution, depending on if $i=j$ or not.
-}}
 %
@@
 
-The likelihood for the CarHMM can be easily calculated using the forward algorithm.  As previously, let $y$ be the vector of observed emissions.  Then
+The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $\bf y$ be the vector of observed emissions. Then
 \begin{equation}
-\calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N
+\calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
 \label{CarHMM_likelihood}
@@
 %
-$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
+$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
 %
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{The hierarchical hidden Markov model (HHMM)}
 
-A hierarchical hidden Markov model contains two process levels, a coarse-scale process and a fine-scale process.
-The coarse-scale process is a hidden Markov model as defined in the previous subsection,  with  $X_1\ldots, X_T$ the unobserved Markov chain of states and $Y_1,\ldots, Y_T$ the corresponding observed responses.
+A \textit{hierarchical hidden Markov model} contains both a coarse-scale process and a fine-scale process.
+The coarse-scale process is a hidden Markov model as defined in the previous subsection, with $X_1, \ldots, X_T$ the unobserved Markov chain of states and $Y_1,\ldots, Y_T$ the corresponding observed responses.
 The possible states are $1,\ldots, N$.
 %
-
-For each value of $t$, the state $X_t$ emits not only $Y_t$ but also  a sequence of unobserved states, 
-$X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of observed emissions, 
-$Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$. 
+For each value of $t$, the state $X_t$ emits not only $Y_t$ but also a sequence of unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
 These form our fine-scale process.
+The length of each fine-scale process, $T^*_t$, is a problem-specific tuning parameter that is determined by how the fine-scale observations are split.
 Conditional on the coarse-scale hidden state $X_t$, the joint fine-scale process, $X_t^*$, $Y_t^*$, is a hidden Markov model with parameters depending on the value of $X_t$.
-Specifically, given that $X_t=i$,   $X_t^*$ is a Markov chain with states $1,\ldots, N_t^*$ with
-%
-\hfill\break{\bf{NH:  Do we write $1^*$? That's strange.}}
-\hfill\break
-%
- distribution of $X_t^*$, conditional on $X_t=i$,  given by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial probability, denoted by the $N_t^*$-vector $\delta^{*(i)}$, which we take equal to the stationary distribution of the chain.
+Specifically, given that $X_t=i$, $X_t^*$ is a Markov chain with states $1,\ldots, N_t^*$.
+The distribution of $X_t^*$, conditional on $X_t=i$, is given by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial probability, denoted by the $N_t^*$-vector $\delta^{*(i)}$, which we take equal to the stationary distribution of the chain.
 For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary.
 %
-The distribution of $Y_{t, t^*}$
-given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter  $\theta^{(i,i^*)}$, and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or sometimes simply denoted by $f^{*(i,i^*)}(\cdot)$.
+The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$, and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or sometimes simply denoted by $f^{*(i,i^*)}(\cdot)$.
 Let $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$, the fine-scale emission parameter vector corresponding to $X_t=i$.
-%
-\hfill\break
-{\bf{NH:  will there be some parameters that are common across the fine-scale emission distributions? Do we want to separate them out?\\
-Also - I used left ( and right ) above to make the parentheses a little bigger.  I think it looks better.  So we may want to do this throughout.
-$f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ versus
-$f^{*(i,i^*)}(\cdot; \theta^{(i,i^*)})$.
-}}
-\hfill\break
 %
@@
 $Y_1,\ldots, Y_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs.
-%%
-\hfill\break
-{\bf{I think that works for the dependencies - might need to add more. Does that look OK to you?}}
-%%
+It is also possible to force certain parameters to be shared across different coarse or fine states.
+For example, in our killer whale case study we force certain fine-scale parameters to be shared across coarse-scale hidden states, i.e.
+$\theta^{(1,i*)} = \ldots = \theta^{(N,i*)}$.
 
@@
 
-For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, model selection and model checking, see Adam et al \citep{Adam:2019}.
+For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citep{Adam:2019}.
 
@@
 
-% do I need this subsections
-\subsection{Decoding}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{General hierarchical structures withing HMMs}
+\label{subsec:STFT}
 
-{\bf{I'm stopping here.  Perhaps you can take a stab at revisions here?}}
+One issue with all HMMs is that they assume Markovian dynamics when conditioned on the hidden state, i.e.
+that any observation $Y_t$ depends only on the behavioral state $X_t$ and $Y_{t-1}$ when conditioned on all previous time steps.
+However, there are many processes which violate this Markov property on very fine scales.
+Using the field of ecology as an example, swimming behaviour of marine mammals often exhibits periodic behavior since the animal repeatedly flukes (swims up and down) to propel itself forward.
+Work has been done in the past to model non-Markovian dynamics in the \textit{behavioural} process $X^*_t$ \citep{Langrock:2012}, but to our knowledge addressing non-Markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area \textcolor{red}{(MAYBE?)}.
+With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine scale behavior is likely to persist.
+
+To deal with fine-scale structures in the data which violate the Markov property, the fine-scale process $Y^*_t$ can be modeled with \textbf{any} parametric model which admits an easy-to compute likelihood.
+We remind the reader that the size of each chunk, $T^*_t$, is a problem-specific tuning parameter.
+It should be long enough to capture the fine-scale behavior within each chunk, but short enough to avoid over-smoothing of the data and to maintain high resolution in the hidden process $X^*$.
+The state-specific emission distribution $\calL_{\text{HMM}}$ from the HHMM likelihood is then replaced by the likelihood of the fine-scale model, $\calL_{\text{fine}}(\mathbf{y}_t;\Theta^{(i)})$.
+This gives the following likelihood function:
+\[
+\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta P(\mathbf{y}_1;\Theta) \prod_{t=2}^T \Gamma P(y,y_t;\Theta,\Theta^*) \mathbf{1}_N
+\]
+where $P(\mathbf{y}_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^(i)\left(\cdot\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
+This definition is straightforward to extend to the CarHMM on both the coarse- and fine- scale.
+
+We describe one example of a fine-scale model which can simultaneously deal with non-Markovian, periodic behavior and reduce the dimension of the fine-scale process $y^*$.
+
+\subsubsection{The short-time Fourier transform}
+
+Suppose that the fine scale process $y^*_t$ exhibits significant period behavior which cannot be modeled with a Markov chain.
+One alternative is to use the Fourier transform of $y^*_t$: with a fixed window length $h$:
+%
+\begin{align*}
+    STFT\{z_{(th+1):(th+h)}\}(k) := \hat{z}^{(k)}_{t} = \sum_{n = 1}^{h} z_{(th)+n}e^{-i \frac{2\pi k}{h} (n-1)} \\ k = 0, 1, \ldots, h-1; \quad t = 0,1, \ldots, \lfloor S/h \rfloor - 1
+\end{align*}
+%
+where $i$ in the equation above refers the $\sqrt{-1}$.
+The STFT slides a moving window of length $h$ across the time series $\mathbf{z}$ and transforms the domain of each window from time to frequency.
+This allows the spectrum of $\mathbf{z}$ at time $s$ to be summarized by a $h$-dimensional vector of Fourier coefficients.
+We prevent any overlap in sliding windows to avoid serial dependence.
+$t$ indexes the \textit{windows} moving across $\mathbf{z}$ while $s$ indexes the \textit{observations} of $\mathbf{z}$.
+
+If $\mathbf{z} \in \mathbb{R}^{S}$, then $\hat{\mathbf{z}} \in \mathbb{C}^{\lfloor S / h \rfloor \times h}$, which may be very large.
+Summary statistics can drastically reduce the dimension of $\hat{z}_s$.
+One possible example is to use the following:
+%
+$$y_t^{(1)} = \mathcal{R}\left(\hat{z}^{(0)}_t\right) \qquad y_t^{(2)} = \frac{1}{h}\sum_{k=1}^{\tilde{f}}|\hat{z}^{(k)}_s|^2$$
+%
+$y_t^{(1)}$ is equal to the average value of $\mathbf{z}$ within window $t$, and $y_t^{(2)}$ is equal to the squared 2-norm of the component of $\mathbf{z}$ within the window that can be attributed to frequencies between $1$ and $\tilde{f}$ periods per window length.
+Like $h$, the max frequency $\tilde{f}$ is a problem-specific tuning parameter.
+It should be selected such that $\tilde{f}$ periods per window length is the maximum frequency of $\mathbf{z}$ that makes biological sense.
+These summary statistics are just one possible choice to describe each window; other choices include the maximum and argmax of $\hat z_t$.
+Figure (\ref{fig:fourier_example}) visually shows the process of moving from $\mathbf{z}$ to $\mathbf{y}$.
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=5in]{../Plots/fourier_transform.png}
+	\caption{Visualization of transforming $\mathbf{z}_t$ into $\mathbf{y}_t$ using a sliding window and Fourier transform.}
+	\label{fig:fourier_example}
+\end{figure}
+
+It is possible to accommodate for unequal time steps within each window by using the \textbf{non-uniform discrete Fourier transform (NDFT)}.
+We do not describe the details of this method in this work, but the generalization is straightforward.
+Refer to Bagchi et al \citep{Bagchi:1999} for details.

### 2020-07-08 Report/Sections/simulation.tex
Commit before changing STFT to Fourier Transform

@@
 %\section{Simulation Study}
+
+\subsection{Model Formulation}
+\label{subsec:model_structure}
+
+Hierarchical hidden Markov models can be used to jointly model simultaneous coarse-scale and fine-scale processes.
+However, as mentioned before, the fine-scale process $Y^*$ can often exhibit auto-correlation and intricate structure.
+Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit auto-correlation, especially in $Z_{t,s^*}^{*(1)}$.
+Therefore, we replace the fine-scale HMM within the hierarchical HMM with a CarHMM as shown in (fig \ref{fig:CarHHMM}).
+%
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=5in]{../Plots/CarHHMM.png}
+	\caption{Graphical representation of a CarHHMM. The additional arrows representing auto-correlation between observations are shown in red for emphasis.}
+	\label{fig:CarHHMM}
+\end{figure}
+%
+Naturally, we refer to this model as the \textbf{CarHHMM}, and it can be implemented both with and without the short-time Fourier-transform (STFT).
+The likelihood of this model is still easy to calculate using the forward algorithm:
+%
+$$\calL_{\text{CarHHMM}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N$$
+%
+where:
+%
+\begin{align*}
+P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
+&f^{(N)}(y_t)\calL_{\text{CarHMM}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big]
+\end{align*}
 

### 2020-07-12 Report/Sections/appendix.tex
Update appendix - before moving tables and figures

--- 2020-07-07 Report/Sections/appendix.tex
@@
 
-\section*{Appendix}
-
-\setcounter{subsection}{0}
-
-\subsection{Detailed Description of Data Simulation}
-
-$\hat{Y}^*_{t,s^*}$ was simulated using the following procedure. Note that the $k^{th}$ Fourier mode of $\hat{Y}^*_{t,s^*}$ is denoted as $\hat{Y}^{*(k)}_{t,s^*}$:
+\begin{proof}{Proof of Theorem 1}{}
+Combining equation (\ref{eqn:CarHMM_to_OU}) with equation (\ref{eqn:OU_sol}) gives:
+%
+\begin{align*}
+\left(Y_{s \Delta t} | X_{(s-1)\Delta t} = i^* \right) \sim \mathcal{N}\left((1-\phi^{(i^*)}) \mu^{(i^*)} + \phi^{(i^*)} Y_{(s-1) \Delta t}, \enspace \sigma^{2(i^*)} \right),\\
+s = 1, \ldots, T-1
+\end{align*}
+%
+If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t: (s-1) \Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:
 
 \begin{align*}
-	(\hat{Y}^{*(0)}_{t,0}|X^*_{t,0} = i) &\sim \mathcal{N} \left(0, \sigma^{*(i)} \right) & \\
+\left(Y'_s| X'_s = i^* \right) \sim \mathcal{N}\left((1-\phi^{(i^*)}) \mu^{(i^*)} + \phi^{(i^*)} Y'_{s-1}, \enspace \sigma^{2(i^*)} \right)\\
+s = 2, \ldots, T
+\end{align*}
 %
-	(\hat{Y}^{*(0)}_{t,s^*}|X^*_{t,s^*} = i) &\sim \mathcal{N} \left(\phi^{*(i)} * \hat{Y}^{*(0)}_{t,s^*-1}, \sigma^{*(i)} \right), & s^* = 1,2,\ldots, \lfloor Y_t/2 \rfloor \\
+$X'$ follows a Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s+1})$ is consistent with that of a CarHMM with normal emission distributions.
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{proof}{Description of simulated data}{}
+
+$\hat{Y}^*_{t,t^*}$ was simulated using the following procedure. The $k^{th}$ Fourier mode of $\hat{Y}^*_{t,t^*}$ is denoted as $\hat{Y}^{*(k)}_{t,t^*}$:
+
+\begin{align}
+	(\hat{Y}^{*(0)}_{t,0}|X^*_{t,0} = i^*) &\sim \mathcal{N} \left(0, \sigma^{*(i^*)} \right) & \nonumber \\
 	%
-	\hat{Y}^{*(k)}_{t,s^*} &= a_{t,s^*}^{(k)} i\sqrt{b^{(k)}_{t,s^*}}, & k = 1,\ldots,49 \\
-\end{align*}
+	(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*) &\sim \mathcal{N} \left(\phi^{*(i^*)} * \hat{Y}^{*(0)}_{t,t^*-1}, \sigma^{*(i^*)} \right), & t^* = 1,2,\ldots, \lfloor Y_t/2 \rfloor \label{eqn:y_hat0} \\
+	%
+	\hat{Y}^{*(k)}_{t,t^*} &= a_{t,t^*}^{(k)} i\sqrt{b^{(k)}_{t,t^*}}, & k = 1,\ldots,49 \nonumber
+\end{align}
+
 \begin{align*}
-    a_{t,s^*}^{(k)} &\sim  \left\{\begin{array}{lr}
+    a_{t,t^*}^{(k)} &\sim  \left\{\begin{array}{lr}
 	-1 & \quad w.p. \enspace 1/2 \\
@@
 	%
-	(b^{(k)}_{t,s^*}|X^*_{t,s^*}  = 1) &\sim {\rm{Gamma}}(1/k^2, 1) \\
+	(b^{(k)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim {\rm{Gamma}}(5/k^2, 1) \\
 	%
-	(b^{(k)}_{t,s^*}|X^*_{t,s^*} = 2) &\sim \left\{\begin{array}{lr}
-	{\rm{Gamma}}(1/k^2, 1), \quad & k \notin \{1,2\} \\
-	{\rm{Gamma}}(100,1), & k = 1 \\
+	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
+	{\rm{Gamma}}(5/k^2, 1), \quad & k \notin \{1,2\} \\
+	{\rm{Gamma}}(250,1), & k = 1 \\
 	{\rm{Gamma}}(50,1), & k = 2
@@
 \begin{align*}
-    \hat{Y}^{*(50)}_{t,s^*} &= 0 & \\
+    \hat{Y}^{*(50)}_{t,t^*} &= 0 & \\
 	%
-	\hat{Y}^{*(k)}_{t,s^*}  &= -\hat{Y}^{*(100-n)}_{t,s^*}, & \qquad k = 51,\ldots,99
+	\hat{Y}^{*(k)}_{t,t^*}  &= -\hat{Y}^{*(100-k)}_{t,t^*}, & \qquad k = 51,\ldots,99
 \end{align*}
 
-$Y^*_{t,(100s^*+1):(100s^*+100)}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,s^*}$:
+$Y^*_{t,t^*,1:100}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
 
-$$Y^*_{t,(100s^*+1):(100s^*+100)} = IDFT\left(\hat{Y}^*_{t,s^*}\right), \qquad s^* = 0,\ldots,\lfloor Y_t/2 \rfloor - 1$$
+$$Y^*_{t,t^*,1:100} = IDFT\left(\hat{Y}^*_{t,t^*}\right), \qquad t^* = 1,\ldots,\lfloor Y_t/2 \rfloor$$
 
-There are several practical reasons behind this construction.
-First, $\hat{Y}^*_{t,s^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,s^*}$ so that its inverse Fourier transform is real-valued.
-$\hat{Y}^{*(k)}_{t,s^*}$ also decays like $1/k^2$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window.
-Note, however, that $Y^*_{t,s^*}$ is not continuous \textit{between} windows, as can be seen in (fig \ref{fig:fourier_example}).
-However, the jumps are not too severe since $\hat{Y}^{*(0)}_{t,s^*}$ and $\hat{Y}^{*(0)}_{t,s^*+1}$ are highly correlated.
+$\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued.
+$\hat{Y}^{*(k)}_{t,t^*}$ also decays like $1/k$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window.
+$Y^*_{t,t^*}$ is not continuous \textit{between} windows, but the jump discontinuities are not very severe since $\hat{Y}^{*(0)}_{t,t^*}$ and $\hat{Y}^{*(0)}_{t,t^*+1}$ are highly correlated.
 See (fig.
@@
 
-From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$ after setting $\tilde{f}$.
-We pick $\tilde{f} = 10$ periods per window, or 5 hertz.
-Note that $Z^{*(1)}_{t,s^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,s^*}\right)$, so selecting $\sigma^{*(i)}$ and $\phi^{*(1)}$ appropriately will ensure correct distribution.
+From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$.
+We pick $\tilde{\omega} = 10$ periods per window, or 5 hertz.
+To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use (eqn \ref{eqn:y_hat0}):
+%
+$$\left(Z^{*(1)}_{t,t^*} | X^*_{t,t^*} = i^* \right) = \left(\mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right) | X^*_{t,t^*} = i^* \right) \sim \mathcal{N} \left(\phi^{*(i^*)} * Z^{*(1)}_{t,t^*-1}, \sigma^{*(i^*)} \right)$$
+%
+$Z^{*(2)}_{t,t^*}$ is the sum of Gamma-distributed random variables with the same scale parameter, so the distribution of $Z^{*(2)}_{t,t^*}$ is also a Gamma distribution:
+%
+$$Z^{*(2)}_{t,t^*} = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
+$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = \sum_{k=1}^{10} 5/k^2 = 10.10, \beta = 1.00 \right)$$
+$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = 300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , \beta = 1.00 \right)$$
+\end{proof}
 
-Note that $Z^{*(2)}$, is the sum of gamma distributions with the same scale parameter, so the distribution of $Z^{*(2)}$ is also a Gamma distribution:
+\begin{proof}{Likelihood of Simulation study model}{}
 
-$$Z^{*(2)} = \sum_{k=1}^{10} b^{(k)}_{t,s^*}$$
-$$Z^{*(2)} \sim {\rm{Gamma}}$$
+The overall likelihood of the CarHHMM-DFT model is as follows:
+%
+$$\calL_{\text{CarHHMM-DFT}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N$$
+%
+where:
+%
+\begin{align*}
+P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
+&f^{(N)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big]
+\end{align*}
+%
+$f^{(i)}(y_t;\Theta^{(i)})$ is the emission distribution of the dive duration $y_t$ conditioned on the fact that $X_t = i$.
+
+$\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale chain:
+%
+$$\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(i)},\Gamma^{*(i)}\right) = \delta^{*(i)} \prod_{t=2}^T \Gamma^{*(i)} P(z^*_{t,t^*}|z^*_{t,t^*-1};\Theta^{*(i^*)}) \mathbf{1}_N$$
+%
+where $P(z^*_{t,t^*}|z^*_{t,t^*-1};\Theta^{*(i)})$ is an $N^* \times N^*$ diagonal matrix with $i^*i^*$th entry equal to $f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$.
+%
+$f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$ is the emission distribution of $Z^*_{t,t^*}$ conditioned on the fact that $X^*_{t,t^*} = i^*$ and $Z^*_{t,t^*-1}$.
+
+\end{proof}
 
@@
 First, we simply use down-sampling and only record every $w^{th}$ time step of $\hat{Y}_t^*$.
-This both reduces the space of $\hat{Y}_t^*$ to $\mathbb{C}^{\lfloor S^*_t / w \rfloor \times w}$ and ensures that none of the sliding windows overlap when taking the STFT.
+This both reduces the space of $\hat{Y}_t^*$ to $\mathbb{C}^{\lfloor t^*_t / w \rfloor \times w}$ and ensures that none of the sliding windows overlap when taking the STFT.
 This is important because HMMs assume temporal independence between observations.

### 2020-07-12 Report/Sections/discussion.tex
Update appendix - before moving tables and figures

% !TeX root = ../main.tex

%\section{Discussion}

We presented a collection of HMM models which can be combined together in increasingly complex hierarchical models to match the complexity of particular problems faced by practitioners. Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics within the hidden state. However, many real-world processes are more complicated than this.

For observations that do not appear to be conditionally independent in time, the CarHMM is a good choice. The CarHMM explicitly models auto-correlation in the emission distributions of the HMM while maintaining the structure needed to evaluate the likelihood using the forward algorithm. In our normal model formulation, we have added only one additional parameter, $\phi^{(i)}$, per possible hidden state \citep{Lawler:2019}. In addition, we prove in the appendix that under certain circumstances the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process. Several useful model selection tools such as the lag plot can test if there is significant auto-correlation within the observation sequence.

For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs. A coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.

Finally, for processes with very high sampling frequencies and/or with intricate fine-scale structure, the HHMM structure developed by \citep{Barajas:2017,Adam:2019} can be generalized so the fine-scale model can be any model which admits an easy-to-calculate likelihood. For example, if the sampling rate of the fine-scale process is very high, then the fine-scale model can be a simple probability distribution over the summary statistics of a moving window of observations. In addition, researchers can recursively stack hierarchical HMM models together like building blocks to create increasingly complex models. This should be done with care, as it is important to balance the need to effectively capture the process in question with the need to avoiding over-fitting and slow parameter estimation. One way to temper model complexity is to reduce the dimension of the parameter space by forcing fine-scale states to be shared across the coarse-scale states, but model complexity inevitably grows rapidly as hierarchical structures are stack on top of each other.

An example of balancing model complexity with fast model fitting is presented in the simulation study here. While (fine-scale) sub-dive behavioral states were shared across (coarse-scale) dive types to reduce model complexity for the hierarchical models, by far the fastest model to train (~15 minutes) was the simple CarHMM-DFT. This models had no hierarchical component but near-perfect accuracy when decoding sub-dive behavioral states of the simulated whale. This model would be the preferred one if simultaneous modeling of dive types was not required. However, this model is deficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.

The simulation study also shows that the observed fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases. One notable exception is that the standard errors of the probability transition matrix estimates, $\hat \Gamma$ and $\hat \Gamma^*$, tend to be overestimated by the observed fisher information.

Finally, we used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada. The CarHHMM-DFT was able to distinguish three distinct sub-dive behaviors and two dive types simultaneously due to its hierarchical structure. The DFT component proved useful in determining the sub-dive behaviors of the whale, as the mean of the emission distribution of $Z^{*(2)}$ was separated by an order of magnitude between each sub-dive state. Finally, the learned auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi$, was above 0.5 for every dimension and sub-dive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit of the data. The introduction of the parameter $\phi$ allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).

Because traditional information criteria tend to overestimate the number of states in biological processes (\citep{Pohle:2017}), the number of dive types and sub-dive behaviors was selected in an ad-hoc manor. There does appear to be some heterogeneity within dive types, and future work can be done to determine the optimal number of dive types and within-dive behaviors.

### 2020-07-12 Report/Sections/intro.tex
Update appendix - before moving tables and figures

@@
 
+All HMMs presented so far assume Markovian dynamics in the underlying process, i.e.
+that any observation $Y_t$ depends only on the behavioral state $X_t$ and $Y_{t-1}$ when conditioned on all previous time steps.
 At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs.
 Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}.
+With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine-scale processes are likely to persist.
+
 One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation.

### 2020-07-12 Report/Sections/models.tex
Update appendix - before moving tables and figures

--- 2020-07-07 Report/Sections/model_Nancy.tex
@@
-%\section{Models and parameter estimation (NH)}
+%\section{Models and parameter estimation}
 %\input Sections/model_Nancy.tex
 
-We first remind the reader of the definition of a hidden Markov model and extend this to the CarHMM.
-We then introduce hierarchical structures within HMMs using both the short-time Fourier transform (STFT) and the hierarchical hidden Markov model (HHMM)
+In this section we first remind the reader of the definition of a hidden Markov model (HMM).
+We then review three variations on the HMM; the CarHMM, HHMM, and HMM-DFT.
+Finally, we generalize the hierarchical structure of these models and describe how they can be flexibly altered and combined to form new models.
 
@@
 %
-(Fig \ref{fig:HMM}) represents the dependence structure.
+(Fig \ref{fig:HMM}) represents the dependence structure of an HMM.
 
 To find the maximum likelihood estimates of the parameters $\Gamma$ and $\Theta \equiv (\theta^{(1)},\ldots,\theta^{(N)})$, let $y = (y_1,\ldots,y_T)$ be the observed emissions. 
- To facilitate calculations we write the likelihood $\calL_{\text{HMM}}$ in a specific form, using the  well-known \textit{forward algorithm} \citep{Zucchini:2016} as follows:
+We write the likelihood $\calL_{\text{HMM}}$ in a specific form, using the  well-known \textit{forward algorithm} \citep{Zucchini:2016} as follows:
 %
@@
 
-One of the key assumptions of an HMM is \textit{conditional independence} between observations, given the state sequence.
+A key assumptions of an HMM is \textit{conditional independence} between observations given the state sequence.
 Therefore, traditional HMMs do not hold when the observations exhibit certain forms of significant correlation in time.
 
-One way to model extra correlation in an HMM type structure is via a CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by \citep{Lawler:2019}.
-Like a traditional HMM, A CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution.
-Unlike a traditional HMM, the CarHMM assumes that the distribution of the $t$th emission, $Y_t$, conditional on $X_1,\ldots, X_T$ and $ \{Y_1,\ldots, Y_{t-1}\}$, depends on both $X_t$ \textit{and} $Y_{t-1}$.
+The CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by \citep{Lawler:2019}, explicit models auto-correlation within an HMM.
+Like a traditional HMM, A CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
+Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$, conditional on $X_1,\ldots, X_T$ and $ \{Y_1,\ldots, Y_{t-1}\}$, depends on \textit{both} $X_t$ \textit{and} $Y_{t-1}$.
 The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. (Fig \ref{fig:CarHMM}) shows the structure of a CarHMM.
@@
 %
-We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1})$ or as
-$f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$.
-As an example, we could assume that this conditional distribution is normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{2(i)},\phi^{(i)}\}$, with:
+We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
+As an example, we could assume that this conditional distribution is normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
 %
@@
 \[
-\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = \sigma^{2(i)}.
+\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = (\sigma^{(i)})^2.
 \]
 %
-{\bf{Ah - now I see the issue about changing states.
-It would be great to have something about the distribution of $Y_t$ given $X_t=i, X_{t-1}=j$.
-You'd want a different conditional distribution, depending on if $i=j$ or not.
-}}
-%
-The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to write the likelihood using the forward algorithm.
-In our normal model formulation, we have added only one additional parameter, $\phi^{(i)}$,  per possible hidden state.
-
 The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
 \begin{equation}
-\calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N
+\calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
 \label{CarHMM_likelihood}
@@
 %
-$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
+$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
 %
+\subsubsection{Equivalence with continuous-time process}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Although the CarHMM can incorporate auto-correlation into the structure of an HMM, they break down when observations are taken at irregular time intervals since the probability transition matrix $\Gamma$ would then have to depend upon the time step between observations.
+A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \citep{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal.
+Namely, a state-switching Ornstein-Uhlenbeck process $Y$ is the solution to the following stochastic differential equation:
+%
+\begin{equation}
+    \label{eqn:SDE}
+    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
+\end{equation}
+%
+where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
+One notable difference between the state-switching OU process and the CarHMM is that the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer.
+If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
+%
+\begin{align}
+Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
+\label{eqn:OU_sol}
+\end{align}
+%
+where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$.
+Continuous time models can be built up from arbitrarily complex stochastic differential equations and allow for uneven step lengths in the observations sequence $Y$.
+However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.
+
+Although most continuous time models are difficult to incorporate into an HHMM, we show in that under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process shown in (eqn \ref{eqn:OU_sol}).
+This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.
+
+\begin{theorem}{Theorem 1.}{}%
+If:
+\begin{enumerate}
+    \item The hidden behavioural process $X$ from (Eq.
+\ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
+    \item Observations of the SDE from (Eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
+\end{enumerate}
+then the observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
+
+\begin{align}
+\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{(i)} = \sqrt{\frac{\omega^{(i)^2}}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t})}, \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
+\end{align}
+
+\end{theorem}
+
+See the appendix for a proof of Theorem 1.
+
 \subsection{The hierarchical hidden Markov model (HHMM)}
 
-A hierarchical hidden Markov model contains two process levels, a coarse-scale process and a fine-scale process.
-The coarse-scale process is a hidden Markov model as defined in the previous subsection,  with  $X_1\ldots, X_T$ the unobserved Markov chain of states and $Y_1,\ldots, Y_T$ the corresponding observed responses.
+A \textit{hierarchical hidden Markov model} contains both a coarse-scale process and a fine-scale process, both of which are modeled by an HMM.
+The coarse-scale process is a hidden Markov model as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain and $Y_1,\ldots, Y_T$ are the corresponding observed responses.
 The possible states are $1,\ldots, N$.
 %
-
-For each value of $t$, the state $X_t$ emits not only $Y_t$ but also  a sequence of unobserved states, 
-$X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of observed emissions, 
-$Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$. 
-These form our fine-scale process.
-Conditional on the coarse-scale hidden state $X_t$, the joint fine-scale process,  $X_t^*$, $Y_t^*$, is a hidden Markov model with parameters depending on the value of $X_t$.
-Specifically, given that $X_t=i$,   $X_t^*$ is a Markov chain with states $1,\ldots, N_t^*$ with
-%
-\hfill\break{\bf{NH:  Do we write $1^*$? That's strange.}}
-\hfill\break
-%
- distribution of $X_t^*$, conditional on $X_t=i$,  given by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial probability, denoted by the $N_t^*$-vector $\delta^{*(i)}$, which we take equal to the stationary distribution of the chain.
+In the hierarchical setting, each state $X_t$ emits not only $Y_t$ but also another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
+The length of each fine-scale process, $T^*_t$, is a problem-specific tuning parameter.
+Conditional on the coarse-scale hidden state $X_t$, the joint fine-scale process, ($X_t^*, Y_t^*$), is a hidden Markov model with parameters depending on the value of $X_t$.
+Specifically, given that $X_t=i$, $X_t^*$ is a Markov chain with states $1,\ldots, N_t^*$.
+The distribution of $X_t^*$, conditional on $X_t=i$, is given by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial probability, denoted by the $N_t^*$-vector $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
 For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary.
 %
-The distribution of $Y_{t, t^*}$
-given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter  $\theta^{(i,i^*)}$, and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or sometimes simply denoted by $f^{*(i,i^*)}(\cdot)$.
+The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$, and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or sometimes simply denoted by $f^{*(i,i^*)}(\cdot)$.
 Let $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$, the fine-scale emission parameter vector corresponding to $X_t=i$.
 %
-\hfill\break
-{\bf{NH:  will there be some parameters that are common across the fine-scale emission distributions? Do we want to separate them out?\\
-Also - I used left ( and right ) above to make the parentheses a little bigger.  I think it looks better.  So we may want to do this throughout.
-$f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ versus
-$f^{*(i,i^*)}(\cdot; \theta^{(i,i^*)})$.
-}}
-\hfill\break
-%
-The dependencies among the processes are defined as follows.
-%
-Given the coarse-scale states, $X_1,\ldots, X_T$, and the coarse scale emissions,  
-$Y_1,\ldots, Y_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs.  
-%%
-\hfill\break
-{\bf{I think that works for the dependencies - might need to add more. Does that look OK to you?}}
-%%
+In terms of the dependence structure, given the coarse-scale states, $X_1,\ldots, X_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs.
+Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states.
+For example, in the killer whale case study in section \ref{sec:case_study} we force the fine-scale emission parameters to be shared across coarse-scale hidden states, i.e.
+$\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$.
+(Fig \ref{fig:HHMM}) shows the dependence structure for an HHMM.
 
-Due to the nested structure of the hierarchical hidden Markov model, the likelihood  is easy to calculate via the forward algorithm, just as for HMMs.
+Due to the nested structure of the hierarchical hidden Markov model, the likelihood remains easy to calculate via the forward algorithm.
 %
@@
 %
-Then the likelihood of the observed data is:
+Then the likelihood of the observed data is
 %
@@
 %
-where 	$P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N^ \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to 
+where $P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to 
 $f^{(i)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(i)},
@@
 
-% do I need this subsections
-\subsection{Decoding}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{The HMM with discrete Fourier transform (HMM-DFT)}
+\label{subsec:STFT}
 
-{\bf{I'm stopping here.  Perhaps you can take a stab at revisions here?}}
+The HMM with discrete Fourier transform or HMM-DFT, is a variation of the hierarchical hidden Markov model where the fine-scale process is no longer modeled with an HMM and instead summarized using it Fourier transform.
+For simplicity of presentation, we assume that the length of the fine-scale processes is constant, i.e.
+that $T^*_t = T^*$, although this need not be the case in general.
+Suppose that the fine-scale process $y^*_t$ does not switch states hidden states, but it does exhibit significant periodic behavior which cannot be modeled with a Markov chain.
+Then, we suggest using the discrete Fourier transform (DFT) on $y^*_t$:
+%
+\begin{align*}
+    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right) \qquad k = 0, 1, \ldots, T^*-1
+\end{align*}
+%
+Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One example of summary statistics include the following:
+%
+\begin{equation}
+    \label{eqn:z}
+    z_t^{*(1)} := \mathcal{R}\left(\hat{y}^{(0)}_t\right) \qquad z_t^{*(2)} := \frac{1}{T^*}\sum_{k=1}^{\tilde{\omega}}|\hat{y}^{(k)}_t|^2
+\end{equation}
+%
+$z_t^{*(1)}$ is equal to the average value of $y^*_t$ and $z_t^{*(2)}$ is equal to the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$.
+The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care.
+These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$.
+Figure (\ref{fig:fourier_example}) visually displays the process of transforming $y^*$ into $z^*$.
+
+\begin{figure}[t]
+	\centering
+	\includegraphics[width=5in]{../Plots/fourier_transform.png}
+	\caption{Visualization of transforming $\mathbf{y^*}_t$ into $\mathbf{z}_t$ using a discrete Fourier transform.}
+	\label{fig:fourier_example}
+\end{figure}
+
+Once $z^*_t$ is calculated, it can simply be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or simply $f^{(i)}\left(y_t,z^*_t\right)$.
+In total, the likelihood of the hierarchical HMM-DFT is as follows:
+\begin{equation}
+\calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
+\label{HMMDFT_likelihood}
+\end{equation}
+where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.
+
+It is possible to accommodate for unequal time steps within $y_t^*$ by using the non-uniform discrete Fourier transform (NDFT).
+We do not describe this method here, but the generalization is straightforward.
+Refer to \citep{Bagchi:1999} for details.
+
+\subsection{General hierarchical structures}
+
+In addition to the DFT and HMM models described above, the fine-scale process $Y^*_t$ can be modeled with \textbf{any} parametric model which admits an easy-to-compute likelihood.
+The state-specific emission distribution $\calL_{\text{HMM}}$ from the HHMM likelihood is then replaced by the likelihood of the fine-scale model, $\calL_{\text{fine}}(\mathbf{y}^*_t;\Theta^{*(i)})$:
+\[
+\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta P(y_1,y^*_1;\Theta,\Theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\Theta,\Theta^*) \mathbf{1}_N
+\]
+where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
+This definition is straightforward to extend to the CarHMM as well:
+\[
+\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
+\]
+where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$
+
+As previously mentioned, the fine-scale model can be any parametric model with easy-to-compute likelihood, including any of the four models described in the previous four subsections (HMM, CarHMM, HHMM, and HMM-DFT).
+These HMM models then act as building blocks that can be used to build increasingly complex hierarchical models based on the HMMs.
+In the sections that follow, we perform both a simulation study and real-world case study modelling killer whale dive behaviour using models built from these building blocks.

### 2020-07-12 Report/Sections/results.tex
Update appendix - before moving tables and figures

@@
 
+The CarHHMM-DFT was used to analyze dive data from a northern resident killer whale off the coast of British Columbia, Canada.
+Acceleration data can be a good proxy for energy expenditure \citep{Green:2009}, but studies suggest that the behavioral state must be taken into account when using accelerometer data \citep{Dot:2016}.
+Therefore, understanding both the behavioral state of the killer whale as well as the distribution of accelerometer data within each behavioral state is import to understand the energetic requirements of killer whales.
+This in turn can assist conservation efforts.
+
 \subsection{Data Collection and Preprocessing}
 
-To apply this methodology to real-world data, the CarHHMM was used to analyze dive data from a killer whale off the coast of British Columbia, Canada.
-The data was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions.
+The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions.
 Observations were collected at a rate of 50 hertz.
@@
 After preprocessing the raw data, a total of 267 dives were observed.
-A plot of the raw data for all dives and a collection of five selected dives can be seen in figures \ref{fig:data} and \ref{fig:data_one_dive}, respectively.
+A plot of the raw data for all dives and a collection of five selected dives can be seen in (fig \ref{fig:data}) and (fig \ref{fig:data_one_dive}), respectively.
 
@@
 
-The collection of all dive durations were set to be the coarse-scale observations $Y$, and the acceleration data was used to determine the fine-scale observations $Y^*$.
-The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs without some kind of signal processing (see fig \ref{fig:data_one_dive}).
-Therefore, the STFT was used to calculate both $Y^{*(1)}$ and $Y^{*(2)}$ as described in section \ref{subsec:STFT}.
-We set $\tilde{f} = 5$ hertz and $h = 2$ seconds, which reduced the dimension of each window from $50 s^{-1} * 2 s = 100$ to $2$.
+The coarse-scale observations were made up of the collection of dive durations in seconds, and the fine-scale observations were determined from the within-dive acceleration data.
+The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
 
-To determine if the CarHHMM was appropriate for this data, a lag plot was made for both $Y^{*(1)}_{t,s^*}$ and $Y^{*(2)}_{t,s^*}$, as shown in figure (\ref{fig:lag}).
-The number of behavioral states is not clear from the lag plot, but it is clear that $Y^{*(1)}_{t,s^*}$ exhibits a large degree of auto-correlation, so the CarHMM was used for $Y^{*(1)}_{t,s^*}$.
-The auto-correlation parameter $\hat phi$ was set to be shared among all dimensions of $Y^{*(1)}_{t,s^*}$, but both the means and variances were allowed to be different.
-While $Y^{*(2)}_{t,s^*}$ also exhibits some auto-correlation, the relationship is less strong.
-In addition, the biological interpretation of auto-correlation within $Y^{*(2)}_{t,s^*}$ is more difficult than for $Y^{*(1)}_{t,s^*}$, so auto-correlation was not incorporated into the emission distribution of $Y^{*(2)}_{t,s^*}$.
+$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}$$
+$$\mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2$$
+
+The acceleration exhibits significant sinusoidal behavior at several points in time (see fig \ref{fig:data_one_dive}), so the fine-scale observations $Z^*$ were made up of the DFT summary statistics of a two-second sliding window.
+Acceleration data was a 3-dimensional vector, so $Z^*$ was calculated as follows:
+%
+$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2$$
+%
+$\mathbf{z}_{t,t^*}^{*(1)}$ is a 3-dimensional vector while $z_t^{*(2)}$ is a scalar.
+Summing the first 10 Fourier modes to calculate $z^{(2)}_{t,t^*}$ corresponds to a maximum frequency of $\tilde \omega = 5$ hertz.
+(Fig \ref{fig:lag}) displays a lag plot for all major features and reveals strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions.
+Auto-correlation was therefore directly modeled into the the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$, which is assumed to be normally distributed with the following parameters:
+%
+$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i^*) = \phi_1^{*(i^*)} \mathbf{z} + (1-\phi_1^{*(i^*)}) \mathbf{\mu}_1^{*(i^*)}$$
+$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = \left(\mathbf{\sigma}_1^{*(i^*)}\right)^2 I_3$$
+%
+where $I_3$ is the identity matrix, $\phi_1^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i^*)} \in \mathbb{R}^3$.
+
+While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, and the biological interpretation of auto-correlation within $Z^{*(2)}_{t,t^*}$ is unclear.
+For this reason, auto-correlation was not incorporated into the emission distribution of $Z^{*(2)}_{t,t^*}$.
+The distribution of $Z^{*(2)}_{t,t^*}$ was also assumed to be gamma:
+%
+$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = \mu_2^{*(i^*)}$$
+$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = (\sigma_2^{*(i^*)})^2$$
+%
+$Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent when conditioned of the sub-dive state $X_{t,t^*}$.
 
@@
 It is known that information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically and admittedly somewhat arbitrarily.
-The absence of principled tool to select the number of hidden states is a common issue in statistical ecology.
-Therefore, it is important to use model validation techniques in lieu of information criteria.
-Section \ref{subsec:model_validation} describes our process of validating this model in particular.
+The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
+
+The final model is nearly identical to the one from the simulation study, with the exception that the fine scale Markov chain has three sub-dive behaviors instead of two ($N^* = 3$), and that $\mathbf{z}^{*(1)}_{t,t^*}$ is a vector rather than a scalar.
+The model is comprised of two levels: a coarse-scale HMM and a fine-scale CarHHMM-DFT.
+The coarse-scale model is comprised of an HMM with no auto-correlation and no DFT with hidden states corresponding to dive types and observations corresponding to dive durations.
+The fine-scale model is comprised of a CarHMM-DFT where auto-correlation is modeled into $\mathbf{z}^{*(1)}$ (the average acceleration with a 2-second window), but not $z^{*(2)}$ (the "wiggliness" of the 2-second window).
+Refer to (fig \ref{fig:CarHHMM}) for a graphical representation of this model.
 
@@
 
-The parameters of the estimated emission distributions for each behavioral state are shown in (table \ref{table:emis_dists}).
-Each distribution is also plotted in (fig \ref{fig:coarse_emis}) and (fig \ref{fig:fine_emis}).
-Note that the auto-correlation within the acceleration sequence is not captured in (fig \ref{fig:fine_emis}), so it is important to refer to the estimated auto-correlation parameter $\hat \phi$ from (table \ref{table:emis_dists}) when considering the emission distributions shown in (fig \ref{fig:fine_emis}).
+(Table \ref{table:emis_dists}) displays the emission distribution parameter estimates. Each distribution is also plotted in (fig \ref{fig:coarse_emis}) and (fig \ref{fig:fine_emis}).
 %
@@
 	\includegraphics[height=2.5in]{../Plots/fine-emissions.png}
-	\caption{Estimated probability distributions for each fine-scale observation in each behavioral state. Note that the distributions of acceleration does not take auto-correlation into account.}
+	\caption{Estimated probability distributions for each fine-scale observation in each behavioral state.
+Note that the distributions of acceleration do not take auto-correlation into account (see table \ref{table:emis_dists})}
 	\label{fig:fine_emis}
@@
 %
-While the ecological meaning of the modeled behavioral states is tenuous, we hypothesize the following interpretations.
-For the coarse scale, dive type 1 corresponds to shorter, shallower dives which serve a variety of purposes, including to rest before performing a dive of type 2, which corresponds to a deeper, more sustained dive.
-For the fine scale, behavioral state 1 corresponds to gliding and less overall activity of the whale compared to the other behavioral states.
-The mean of $Y^{*(2)}$ in this state is at least an order of magnitude smaller than sub-dive behavior 2, the variance of $Y^{*(1)}$ is smaller than sub-dive behavior 2, and the auto-correlation of $Y^{*(1)}$ is higher than every other behavioral state.
-This indicates a relatively constant acceleration and therefore low levels of activity.
-Sub-dive state 3 corresponds to vigorous swimming activity, as the mean of $Y^{*(2)}$ and variance of $Y^{*(1)}$ are much higher than every other state.
-The auto-correlation of $Y^{*(1)}$ is also much lower in this state, implying more variation in acceleration every 2 seconds.
-Sub-dive state 2 corresponds to a medium amount of activity, as almost every parameter estimate is between the other two behavioral states.
+While the ecological interpretation of behavioral states is tenuous for HMMs, we hypothesize the following meanings.
+Dive type 1 corresponds to shorter, shallower dives which serve a variety of purposes, including as rest before dives of type 2, which are deeper and more sustained.
+Sub-dive behavioural state 1 corresponds to gliding and less overall activity compared to the other behavioral states.
+The mean of $Z^{*(2)}$ in this state is at least an order of magnitude smaller than sub-dive behavior 2, the variance of $\mathbf{Z}^{*(1)}$ is smaller than sub-dive behavior 2 for every component, and the auto-correlation of $\mathbf{Z}^{*(1)}$ is higher than every other behavioral state.
+Sub-dive state 3, on the other hand, corresponds to vigorous swimming activity, as the mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ for every component are much higher than every other state.
+The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this state, implying more variation in acceleration every 2 seconds.
+Finally, sub-dive state 2 corresponds to a moderate amount of activity, as almost every parameter estimate is between the other two behavioral states.
 
-The estimated probability transition matrices and associated stationary distributions are shown below:
+The estimated probability transition matrices and associated stationary distributions are
 %
@@
 $$\hat \delta^{*(1)} = \begin{pmatrix} 0.143 & 0.698 & 0.159 \end{pmatrix} \qquad
-\hat \delta^{*(2)} = \begin{pmatrix} 0.476 & 0.456 & 0.067 \end{pmatrix}$$
+\hat \delta^{*(2)} = \begin{pmatrix} 0.476 & 0.456 & 0.067 \end{pmatrix}$$.
 %
 About 86\% of observed dives are short- the whale usually rests for many dives in a row before performing a deep dive.
-Interestingly, the probability transition matrix $\hat \Gamma$ shows that the current dive does not effect the dive type of the following dive much.
-In addition, the fine-scale probability transition matrices imply that the killer whale is much more likely to be in a less active sub-dive state when performing deep dives than when performing shallow dives- 14\% of shallow dives are spent in sub-dive state 1 while 48\% of deep dives are spent in sub-dive state 1.
-Using less active sub-dive states when diving deep could be an energy reduction strategy for these long periods of holding breath.
-
-The decoded dive behavior within 5 selected dives is shown in (fig \ref{fig:labeled_dives}).
-In addition, the probability of each dive type and sub-dive state is shown in (fig \ref{fig:coarse_probs}) and (fig \ref{fig:fine_probs}), respectively.
+The probability transition matrix $\hat \Gamma$ shows that the distribution of a particular dive's type is approximately equal regardless of the previous dive type.
+The fine-scale probability transition matrices imply that the killer whale is much more likely to be in a less active sub-dive state when performing deep dives than when performing shallow dives- 14\% of shallow dives are spent in sub-dive state 1 while 48\% of deep dives are spent in sub-dive state 1.
+Using less active sub-dive states when diving deep could be an energy reduction strategy for these long periods of holding breath- a phenomenon which has been observed in bottle-nose dolphins \citep{Williams:1999}.
+(Fig \ref{fig:labeled_dives}) shows the decoded dive behavior of 5 selected dives, and (fig \ref{fig:coarse_probs}) and (fig \ref{fig:fine_probs}) display the probability of each dive type and sub-dive state, respectively.
 
@@
 
-Two visual tools were used to evaluate the model: pseudo-residuals and empirical histograms.
-A pseudo-residual of a particular observation is the marginal cdf of an observation, conditioned on all other observations, and mapped to the quantile function of a standard normal distribution (see \citep{Zucchini:2016} for details).
-In particular, the pseudo-residual of an observation $y_{t,s^*}$ is $\Phi^{-1} \left(Pr(Y_{t,s^*} < y_{t,s^*}|\{Y_1,\ldots,Y_T\}/\{Y_{t,s^*}\}) \right)$, where $\Phi$ is the cdf of a standard normal distribution.
-If the model is well-specified, then all pseudo-residuals should be independent and follow a standard normal distribution.
-While histograms of the pseudoresiduals support this for most observations, one notable exception is the case of $Y^{*(2)}$, shown in (fig \ref{fig:pseudoresids}), which are noticeably right-skewed.
-This implies that the true distribution of $Y^{*(2)}$ may follow a heavier-tailed distribution compared to a gamma distribution such as a power law.
+Two visual tools were used to evaluate this model: pseudo-residuals and empirical histograms.
+A pseudo-residual of a particular observation is the marginal cdf of an observation conditioned on all other observations under the learned model \citep{Zucchini:2016}.
+To easily visualize outliers, the pseudo-residual is often passed through the quantile function of the standard normal distribution.
+In particular, the pseudo-residual of an observation $y_t$ is $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard normal distribution.
+If the model is correct, then all pseudo-residuals should be independent and follow a standard normal distribution.
+We find that histograms of the pseudoresiduals mostly support that the model is well-specified.
+$Z^{*(2)}$ is a notable exception, as its pseudo-residuals are noticeably right-skewed (fig \ref{fig:pseudoresids}).
+This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution compared to a gamma distribution such as a power law.
 
@@
 
-In addition to psuedoresiduals, we plotted histograms of observations for each feature within each hidden state.
-In particular, observations were weighted by the probability that the whale was in a particular hidden state.
-This empirical distribution was then plotted over the estimated true pdf of that feature for the hidden state.
-Again, our results mostly support the model in question with the exception of $Y^{*(2)}$, which is right-skewed.
-In addition, $Y^{*(1)}$ has heavy tails for sub-dive state 3 (see fig \ref{fig:empirical_dist}), indicating the existence of rare events corresponding to very violent trashing of the killer whale.
-These outliers are potential subjects for future studies to better understand rare behavioral events for this killer whale.
+In addition to psuedoresiduals, we plot histograms of observations, weighted by the probability that the whale was in a particular hidden state.
+This empirical distribution was then plotted over the fitted probability distribution function of that feature and behavioral state.
+Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is right-skewed.
+In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for sub-dive state 3 (\ref{fig:empirical_dist}), indicating the existence of rare events corresponding to very violent trashing of the killer whale.
+These outliers are potential subjects for future study.
 

### 2020-07-12 Report/Sections/simulation.tex
Update appendix - before moving tables and figures

--- 2020-07-07 Report/Sections/simulation.tex
@@
 
+We base the simulated study on acceleration data of a killer whale dive sequence.
+The parameters used in this simulation study are loosely based on those learned from the case study in section \ref{sec:case_study}.
+
 \subsection{Data Simulation}
 
-To test the CarHHMM with STFT, 500 separate sequences of 100 marine-mammal dives were simulated.
-The coarse-scale observations $Y$ were set as the duration of each dive, and the fine-scale observations $Y^*$ were set as one-dimensional acceleration readings simulated at 50 hertz.
-The number of possible states for both the coarse- and fine- scale Markov chains was set to 2, i.e.
-$N = N^* = 2$.
+500 separate sequences of 100 killer whale dives were simulated according to an HMM, where the hidden Markov chain $X$ was set as a collection of dive types and the observations $Y$ were the corresponding dive durations (in seconds).
+Each dive could be one of $N=2$ dive types, and the duration of dive $t$, $Y_t$, followed a gamma distribution whose parameters $\theta^{(i)}$ depended upon the dive type $X_t = i$.
+We parameterize the gamma distribution by its mean and variance:
 
-On the coarse scale, the following parameters were used:
-$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix} $$
+$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}, \qquad \delta =  \begin{pmatrix} 0.5 & 0.5 \end{pmatrix}$$
 $$Y_t|X_t \sim \rm{Gamma} $$
@@
 \end{align*}
-Once the coarse-scale dive durations were calculated for $t = 1, \ldots, 100$, dive $t$ was broken into $\lfloor Y_t/2 \rfloor$ 2-second segments (the end of the dive sequence was discarded).
-Each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where the parameters were set as follows:
+
+Once the dive durations were calculated for all 100 dives, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments (the end of the dive sequence was discarded) which made up a second fine-scale hidden Markov model.
+Each 2-second segment of the dive was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$, and the probability transition matrices for these fine-scale Markov chains were set as follows:
 %
@@
 where $\Gamma^{*(1)}$ was used for dives where $X_t = 1$ and $\Gamma^{*(2)}$ was used for dives where $X_t = 2$.
-For each 2-second segment, the Fourier modes $\hat{Y}^*_{t,s^*} \in \mathbb{C}^{100}$ were simulated such that $Z^{*(1)}_{t,s^*}$ and $Z^{*(2)}_{t,s^*}$ would have the following distributions:
+
+Each two-second sub-dive window had 100 associated acceleration readings, $Y^*_{t,t^*} \equiv \left(Y^*_{t,t^*,1}, \ldots, Y^*_{t,t^*,100}\right)$.
+To accurately recreate active swimming vs passive gliding on the fine-scale Markov chain, the DFT of each 2-second segment $\hat Y^*_{t,t^*}$ was simulated such that $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$ (see eqn \ref{eqn:z}) would have the following distributions:
 
 \begin{align*}
-    \left(Z^{*(1)}_{t,s^*} | Z^{*(1)}_{t,s^*-1}, X^*_{t,s^*} = 1 \right) &\sim \mathcal{N} \left(\phi^{*(1)}Z^{*(1)}_{t,s^*-1} + \phi^{*(1)}\mu^{*(1)}, (\sigma^{*(1)})^2 \right) \\
+    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 1 \right) &\sim \mathcal{N} \left(\phi^{*(1)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(1)})\mu^{*(1)}, (\sigma^{*(1)})^2 \right) \\
     %
-    \left(Z^{*(1)}_{t,s^*} | Z^{*(1)}_{t,s^*-1}, X^*_{t,s^*} = 2 \right) &\sim \mathcal{N} \left(\phi^{*(2)}Z^{*(1)}_{t,s^*-1} + \phi^{*(2)}\mu^{*(2)}, (\sigma^{*(2)})^2 \right) \\
+    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 2 \right) &\sim \mathcal{N} \left(\phi^{*(2)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(2)})\mu^{*(2)}, (\sigma^{*(2)})^2 \right) \\
     %
@@
     %
-    \left(Z^{*(2)}_{t,s^*} | X^*_{t,s^*} = 1\right) &\sim {\rm{Gamma}}\left(\alpha^{*(1)}, \beta^{*(1)} \right) \\
+    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 1\right) &\sim {\rm{Gamma}}\left(\alpha^{*(1)}, \beta^{*(1)} \right) \\
     %
-    \left(Z^{*(2)}_{t,s^*} | X^*_{t,s^*} = 2\right) &\sim {\rm{Gamma}}\left(\alpha^{*(2)}, \beta^{*(2)} \right) \\
+    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 2\right) &\sim {\rm{Gamma}}\left(\alpha^{*(2)}, \beta^{*(2)} \right) \\
     %
@@
 %
-See the appendix for more details regarding the simulation procedure for $\hat Y^*$.
-The emission distribution of the 2-second average acceleration $Z_{t,s^*}^{*(1)}$ in particular can be reparameterized as an OU process according to theorem 1:
-%
-$$dZ^{*(1)}_{t,s^*} = \beta^*(\gamma^* - Z^{*(1)}_{t,s^*})dt^* + \omega^* dW_{t,t^*}$$
-%
-$$\gamma^{*(1)} = 0, \enspace \beta^{*(1)} = -\frac{\log(\phi^{*(1)})}{\Delta t^*} = 0.0111, \enspace \omega^{*(1)} = \sqrt{\frac{2(\sigma^{^{*(1)}})^2\log(\phi^{*(1)})}{\Delta t^* ((\phi^{*(1)})^2-1)}} = 0.0355$$
-$$\gamma^{*(2)} = 0, \enspace \beta^{*(2)} = -\frac{\log(\phi^{*(2)})}{\Delta t^*} = 0.0256, \enspace \omega^{*(2)} = \sqrt{\frac{2(\sigma^{*(2)})^2\log(\phi^{*(2)})}{\Delta t^* ((\phi^{*(2)})^2-1)}} = 0.7625$$
-
-A visualization of the first 5 dives of one simulated data set can be seen in (fig. \ref{fig:sim_data}).
+In short, sub-dive behavior 1 corresponds to passive gliding while sub-dive behaviour 2 corresponds to active swimming at a frequency of 1 hertz.
+In addition, sub-dive behaviors 1 and 2 are the same for both dive types.
+See the appendix for more details regarding procedure for simulating $\hat Y^*$ and $Y^*$ in addition to $Z^*$.
+(fig.
+\ref{fig:sim_data}) shows the first 5 dives of one simulated data set.
 
@@
 
+\subsection{Model Formulation}
+\label{subsec:model_structure}
+
+The building blocks from the previous section were used to build a well-specified model for this simulated data.
+Specifically, we used a hierarchical HMM where the sequence of dive durations $Y$ was modeled using a simple HMM, and the fine-scale process $Z^*$ was modeled using a HMM-DFT with explicit auto-correlation in $Z^{*(1)}$.
+Naturally, we refer to this model as the \textit{CarHHMM-DFT}, and (Fig.
+\ref{fig:CarHHMM}) shows the corresponding graphical model.
+%
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=5in]{../Plots/CarHHMM.png}
+	\caption{Graphical representation of this model.}
+	\label{fig:CarHHMM}
+\end{figure}
+%
+
+\subsection{Parameters}
+
+On the coarse scale, the dive types follow a Markov chain with $N=2$ possible states and unknown probability transition matrix $\Gamma$.
+Given the dive type, the duration of a dive follows a gamma distribution which depends upon the dive type and unknown parameters $\Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\}$.
+
+On the fine scale, the sub-dive behavior of each two-second window comprises a Markov chain with $N^*=2$ possible states and unknown probability transition matrices $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$, depending upon the dive type.
+Each two-second window is summarized by the observations $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$.
+$Z^{*(1)}_{t,t^*}$ follows a normal distribution whose parameters depend upon the sub-dive behavior $X^*_{t,t^*}$ and $Z^{*(1)}_{t,t^*-1}$.
+In particular:
+
+$$\mathbb{E}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = \phi_1^{*(i^*)}z + (1-\phi_1^{*(i^*)}) \mu_1^{*(i^*)}$$
+$$\mathbb{V}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = \left(\sigma_1^{*(i^*)}\right)^2$$
+%
+$Z^{*(2)}_{t,t^*}$ follows a gamma distribution whose parameters depend upon only $X^*-{t,t^*}$:
+%
+$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = \mu_2^{*(i^*)}$$
+$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = (\sigma_2^{*(i^*)})^2$$
+%
+None of the fine-scale emission distributions depend upon dive type. In total the parameters to learn are:
+
+$$\Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}$$
+$$\Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters)}$$
+$$\Theta^* = \{\Theta^{*(1)},\Theta^{*(2)}\}  \qquad \text{(fine-scale emission parameters)}$$
+$$\Theta^{*(i^*)} =  \{\{\mu_1^{*(i^*)},\sigma_1^{*(i^*)},\phi_1^{*(i^*)}\},\{\mu_2^{*(i^*)},\sigma_2^{*(i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters)}$$
+
+The likelihood of this model is still easy to calculate using the forward algorithm, and can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation.
+
+Including the one described above, four different models were fit to the simulated data sets:
+
+\begin{enumerate}
+    \item A CarHHMM-DFT as described above (\textbf{CarHHMM-DFT}).
+    \item An HHMM-DFT, which is similar to the model above, but with no modeled auto-correlation, i.e. $\phi_1^{*(i^*)} = 0$ for $i^* = 1,2$ (\textbf{HHMM-DFT}).
+    \item A CarHHMM, which is similar to the model above, but which does not have access to $Z^{*(2)}$, i.e $\Theta^{*(i^*)} = \{\mu_1^{(i^*)},\sigma_1^{(i^*)},\phi_1^{(i^*)}\}$, $i^* = 1,2$ (\textbf{CarHHMM}).
+    \item A CarHMM-DFT similar to the CarHHMM-DFT, but with $N=1$ instead of $N=2$, i.e. $\Gamma^* = \{\Gamma^{*(1)}\}$ and $\Theta^* = \{\Theta^{*(1)}\}$ (\textbf{CarHMM-DFT}). This is 
+\end{enumerate}
+%
+Each of the last three models leaves out one important aspect of the full model (CarHHMM-DFT).
+The CarHMM-DFT lacks a hierarchical structure, the HHMM-DFT is missing auto-correlation within the fine-scale observations, and the CarHHMM does not have access to the Fourier transform sums $Z^{*(2)}$ as observations.
+All models were run on the Cedar Compute Canada cluster with 1 GPU and 4 GB of dedicated memory per model.
+
 \subsection{Simulation Results}
 
-Four different models were fit to each simulated data set:
-
-\begin{enumerate}
-    \item A CarHMM with no hierarchical structure using both $Z^{*(1)}$ and $Z^{*(2)}$ as fine-scale observations (\textbf{CarHMM}).
-    \item An HHMM with no auto-correlation using both $Z^{*(1)}$ and $Z^{*(2)}$ as fine-scale observations (\textbf{HHMM}).
-    \item A CarHHMM using $Z^{*(1)}$, but NOT $Z^{*(2)}$ as fine-scale observations (\textbf{CarHHMM1}).
-To ensure that the CarHHMM could still see the ``wigliness" of the process, we halved the window size to 1 second for the CarHHMM1.
-    \item A CarHHMM using both $Z^{*(1)}$ and $Z^{*(2)}$ as fine-scale observations (\textbf{CarHHMM2}).
-\end{enumerate}
-%
-The purpose of including each of these models is to leave one important aspect of the full model (CarHHMM2) out of each of the preceding three and observe the effects on a simulated data set.
-In particular, the CarHMM lacks a hierarchical structure, the HHMM is missing auto-correlation within the fine-scale observations, and the CarHHMM1 does not have $Z^{*(2)}$ as observations.
-
 %%% Accuracy %%%
-Every model was able to decode the fine-scale hidden states of the process effectively perfectly except for the CarHHMM1, which did not have access to the Fourier modes $Z^{*(2)}$.
+Every model was able to decode the fine-scale hidden states of the process almost perfectly except for the \textit{CarHHMM}, which did not have access to the Fourier modes $Z^{*(2)}$.
 This is intuitively clear because the fine-scale states are much more distinct in their Fourier modes than in their average acceleration.
@@
 (Fig.
-\ref{fig:acc}) shows the decoded state probabilities of both the fine and coarse hidden states for the first 10 dives of one simulated data set.
+\ref{fig:acc}) shows the decoded state probabilities of both the fine and coarse hidden states for the first 5 dives of one simulated data set.
 (Tbl.
@@
 %%% Acceleration %%%
-For acceleration ($Z^{*(1)}$), both the CarHMM and the CarHHMM2 regularly converged to the correct parameters with very little standard error.
-Note, however, that both have relatively large observed fisher standard error in $\hat \mu$.
-This likely is due to the very large autocorrelation terms ($\phi^{(1)} = 0.99$ and $\phi^{(2)} = 0.95$) since as $\phi \to 1$, the value of $\mu$ does not matter in the likelihood evaluation as $\phi \to 1$ implies that $\mathbb{E}(Z^{*(1)}_{t,s^*}) = \phi*(Z^{*(1)}_{t,s^*-1}) + (1-\phi)\mu \to Z^{*(1)}_{t,s^*-1}$.
-The HHMM overestimates the variance of $Z^{*(1)}$ since it does not incorporate auto-correlation into the emission distribution of $Z^{*(1)}$, and the CarHHMM1 has large biases in many of its parameter estimates, especially in subdive state 2, because it sees the ``wigliness" of the acceleration data as raw acceleration rather than as Fourier modes, which deflates the value of $\hat \phi$ and inflates the value of $\hat \sigma$.
+For acceleration ($Z^{*(1)}$), both the CarHMM-DFT and the CarHHMM-DFT regularly converged to the correct parameters with very little standard error.
+However, the fisher standard error regularly overestimated the standard error for $\hat \mu$ for both of these models.
+%This likely is due to the very large auto-correlation terms ($\phi^{(1)} = 0.99$ and $\phi^{(2)} = 0.95$) since as $\phi \to 1$, the value of $\mu$ does not matter in the likelihood evaluation as $\phi \to 1$ implies that $\mathbb{E}(Z^{*(1)}_{t,s^*}) = \phi*(Z^{*(1)}_{t,s^*-1}) + (1-\phi)\mu \to Z^{*(1)}_{t,s^*-1}$.
+The HHMM regularly overestimates the variance $Z^{*(1)}$ since it does not incorporate auto-correlation into the emission distribution of $Z^{*(1)}$, and the CarHHMM has large biases in many of its parameter estimates, especially in sub-dive state 2.
+%because it sees the ``wigliness" of the acceleration data as raw acceleration rather than as Fourier modes, which deflates the value of $\hat \phi$ and inflates the value of $\hat \sigma$. 
 See (table \ref{table:acceleration}) for a detailed breakdown of parameter values for acceleration emission distributions.
@@
 %%% FoVeDBA %%%
-
-Finally, the distribution of parameters for the Fourier sums ($Z^{*(2)}$) for all models (except for the CarHHMM1) converge to the correct values, and the observed fisher information standard errors are in line with the empirical standard error.
-The distribution of $Z^{*(2)}$ is very different between subdive states 1 and 2, so the signal is therefore very easy for each model to pick up on.
+The distribution of parameters for the Fourier sums ($Z^{*(2)}$) for all models except for the CarHHMM converge regularly to the correct values, and the observed fisher information standard errors are good approximation of the empirical standard error.
+%The distribution of $Z^{*(2)}$ is very different between subdive states 1 and 2, so the signal is therefore very easy for each model to pick up on.
 (Table \ref{table:FoVeDBA}) shows a detailed breakdown of the emission distribution of $Z^{*(2)}$ for all models.
 
-\textcolor{red}{I also have similar tables for $\Gamma$ and $\Gamma^*$, and I also have plots corresponding to each table.
+%%% Gamma %%%
+
+Estimates for $\Gamma$ for all models (except for the CarHMM-DFT) are very accurate, with empirical standard errors of approximately 0.02.
+This is significantly less than the observed fisher information estimate of standard error of approximately 0.08.
+%likely because the distribution of $\hat \Gamma$ did not converge to a normal distribution in only 100 observed dives. 
+For $\Gamma^*$, the HHMM-DFT and CarHHMM-DFT both showed practically no bias with standard errors on the order of $10^{-2}$.
+One notable exception is the standard error of $\hat \Gamma^{*(1)}_{12}$, whose standard error was on the order of $10^{-4}$, which is much lower than the observed Fischer information predicted.
+The CarHMM-DFT is mis-specified, so its results cannot be interpreted.
+The CarHHMM consistently underestimated the probability of a hidden state change on the sub-dive behavior.
+Again, one notable exception is $\hat \Gamma^{*(1)}_{12}$, which had almost no bias and was remarkably consistent.
+See (Table \ref{table:Gamma}) for a full list of estimates and standard errors.
+
+\textcolor{red}{I also have plots corresponding to each table.
 For example, I have included figure \ref{fig:MLE_dist} as an example below for the emission distribution of the MLEs for dive duration of the CarHMM.
@@
 \centering
-\caption{Estimates and standard errors of parameters for $Z^{*(1)}_{t,s^*}$ for all four models.
+\caption{Estimates and standard errors of parameters for $Z^{*(1)}_{t,t^*}$ for all four models.
 All reported values are averages, except for the Fisher observed standard error, which are medians.
@@
 \centering
-\caption{Estimates and standard errors of parameters for $Z^{*(2)}_{t,s^*}$ for all four models.
+\caption{Estimates and standard errors of parameters for $Z^{*(2)}_{t,t^*}$ for all four models.
 All reported values are averages, except for the Fisher observed standard error, which are medians.
@@
 \end{table}
+
+% Gamma
+\begin{table}[t]
+\centering
+\caption{Estimates and standard errors of $\Gamma$ and $\Gamma^*$ for all four models. All reported values are averages except for the observed fisher SE, which is a median. $\pm$ refers to the IQR.}
+\scalebox{0.7}{
+\begin{tabular}{ccccccc}
+Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
+\multirow{6}{*}{HHMM-DFT}    & $\Gamma_{12}$                 & $0.50$                         & $-0.00$                   & $0.03$                           & $0.08 \pm 0.01$                             \\
+                             & $\Gamma_{21}$                 & $0.50$                         & $-0.00$                   & $0.03$                           & $0.08 \pm 0.01$                             \\
+                             & $\Gamma^{*(1)}_{12}$          & $0.50$                         & $0.00$                   & $0.00$                           & $0.07 \pm 0.01$                             \\
+                             & $\Gamma^{*(1)}_{21}$          & $0.10$                         & $0.00$                   & $0.02$                           & $0.02 \pm 0.00$                             \\
+                             & $\Gamma^{*(2)}_{12}$          & $0.20$                         & $-0.00$                   & $0.01$                           & $0.01 \pm 0.00$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.30$                         & $-0.00$                   & $0.02$                           & $0.02 \pm 0.00$                             \\ \hline
+\multirow{6}{*}{CarHMM-DFT}  & $\Gamma_{12}$                 & ------                         & ------                   & ------                           & ------                                      \\
+                             & $\Gamma_{21}$                 & ------                         & ------                   & ------                           & ------                                      \\
+                             & $\Gamma^{*(1)}_{12}$          & $0.23$                         & ------                   & $0.01$                           & $0.00 \pm 0.00$                             \\
+                             & $\Gamma^{*(1)}_{21}$          & $0.23$                         & ------                   & $0.02$                           & $0.00 \pm 0.00$                             \\
+                             & $\Gamma^{*(2)}_{12}$          & ------                         & ------                   & ------                           & ------                                      \\
+                             & $\Gamma^{*(2)}_{21}$          & ------                         & ------                   & ------                           & ------                                      \\ \hline
+\multirow{6}{*}{CarHHMM}     & $\Gamma_{12}$                 & $0.49$                         & $-0.01$                   & $0.02$                           & $0.08 \pm 0.01$                             \\
+                             & $\Gamma_{21}$                 & $0.50$                         & $0.00$                   & $0.02$                           & $0.08 \pm 0.01$                             \\
+                             & $\Gamma^{*(1)}_{12}$          & $0.50$                         & $-0.00$                   & $0.00$                           & $0.23 \pm 0.19$                             \\
+                             & $\Gamma^{*(1)}_{21}$          & $0.04$                         & $-0.06$                   & $0.02$                           & $0.01 \pm 0.00$                             \\
+                             & $\Gamma^{*(2)}_{12}$          & $0.11$                         & $-0.09$                   & $0.02$                           & $0.01 \pm 0.00$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.11$                         & $-0.19$                   & $0.03$                           & $0.01 \pm 0.00$                             \\ \hline
+\multirow{6}{*}{CarHHMM-DFT} & $\Gamma_{12}$                 & $0.50$                         & $-0.00$                   & $0.03$                           & $0.08 \pm 0.01$                             \\
+                             & $\Gamma_{21}$                 & $0.50$                         & $-0.00$                   & $0.03$                           & $0.08 \pm 0.01$                             \\
+                             & $\Gamma^{*(1)}_{12}$          & $0.50$                         & $0.00$                   & $0.00$                           & $0.07 \pm 0.01$                             \\
+                             & $\Gamma^{*(1)}_{21}$          & $0.10$                         & $0.00$                   & $0.02$                           & $0.02 \pm 0.00$                             \\
+                             & $\Gamma^{*(2)}_{12}$          & $0.20$                         & $-0.00$                   & $0.01$                           & $0.01 \pm 0.00$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.30$                         & $-0.00$                   & $0.02$                           & $0.02 \pm 0.00$                             \\ \hline
+
+\end{tabular}
+}
+\label{table:Gamma}
+\end{table}

### 2020-07-13 Report/Sections/discussion.tex
Small Edits

@@
 We presented a collection of HMM models which can be combined together in increasingly complex hierarchical models to match the complexity of particular problems faced by practitioners.
+This flexible framework can be used to deal with complicated dependence structures within time-series data.
+
 Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics within the hidden state.
-However, many real-world processes are more complicated than this.
+However, many real-world processes are more complicated than this and require more complex models.
 
-For observations that do not appear to be conditionally independent in time, the CarHMM is a good choice.
-The CarHMM explicitly models auto-correlation in the emission distributions of the HMM while maintaining the structure needed to evaluate the likelihood using the forward algorithm.
+The CarHMM generalizes the HMM by explicitly modeling auto-correlation in the emission distributions of the HMM while maintaining the structure needed to evaluate the likelihood using the forward algorithm.
 In our normal model formulation, we have added only one additional parameter, $\phi^{(i)}$, per possible hidden state \citep{Lawler:2019}.
-In addition, we prove in the appendix that under certain circumstances the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process.
-Several useful model selection tools such as the lag plot can test if there is significant auto-correlation within the observation sequence.
+Several useful model selection tools such as the lag plot can test if there is significant auto-correlation within an observation sequence.
 
-For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs.
-A coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.
+Although the CarHMM can incorporate auto-correlation into the structure of an HMM, it can break down when observations are taken at irregular time intervals.
+A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by \citep{Michelot:2019}.
+Most continuous time models require relatively slow MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.
+However, we prove that certain continuous-time methods are equivalent to an CarHMM under certain conditions (see appendix).
+
+For simultaneous observed processes taking place at different time scales, the HHMM as described by \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs.
+In particular, each hidden state of a coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.
 
 Finally, for processes with very high sampling frequencies and/or with intricate fine-scale structure, the HHMM structure developed by \citep{Barajas:2017,Adam:2019} can be generalized so the fine-scale model can be any model which admits an easy-to-calculate likelihood.
-For example, if the sampling rate of the fine-scale process is very high, then the fine-scale model can be a simple probability distribution over the summary statistics of a moving window of observations.
+For example, if the sampling rate of the fine-scale process is very high, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.
 In addition, researchers can recursively stack hierarchical HMM models together like building blocks to create increasingly complex models.
 This should be done with care, as it is important to balance the need to effectively capture the process in question with the need to avoiding over-fitting and slow parameter estimation.
-One way to temper model complexity is to reduce the dimension of the parameter space by forcing fine-scale states to be shared across the coarse-scale states, but model complexity inevitably grows rapidly as hierarchical structures are stack on top of each other.
 
-An example of balancing model complexity with fast model fitting is presented in the simulation study here.
-While (fine-scale) sub-dive behavioral states were shared across (coarse-scale) dive types to reduce model complexity for the hierarchical models, by far the fastest model to train (~15 minutes) was the simple CarHMM-DFT.
-This models had no hierarchical component but near-perfect accuracy when decoding sub-dive behavioral states of the simulated whale.
-This model would be the preferred one if simultaneous modeling of dive types was not required.
-However, this model is deficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.
+One way to temper model complexity is to reduce the dimension of the parameter space by forcing fine-scale states to be shared across the coarse-scale states.
+Even still, model complexity inevitably grows rapidly as hierarchical structures are stack on top of each other.
+
+An example of balancing model complexity with fast model fitting is presented in the simulation study.
+While (fine-scale) sub-dive behavioral states were shared across (coarse-scale) dive types to reduce model complexity for the hierarchical models, by far the fastest model to train (\~ 15 minutes) was the CarHMM-DFT.
+The CarHMM-DFT had no hierarchical component but near-perfect accuracy when decoding sub-dive behavioral states of the simulated whale.
+This model would be preferred if simultaneous modeling of dive types was not required.
+However, this model is not sufficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.
 
 The simulation study also shows that the observed fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases.
-One notable exception is that the standard errors of the probability transition matrix estimates, $\hat \Gamma$ and $\hat \Gamma^*$, tend to be overestimated by the observed fisher information.
+One notable exception is that the standard errors of the probability transition matrix estimates ($\hat \Gamma$ and $\hat \Gamma^*$) tend to be overestimated by the observed fisher information.
 
@@
 The DFT component proved useful in determining the sub-dive behaviors of the whale, as the mean of the emission distribution of $Z^{*(2)}$ was separated by an order of magnitude between each sub-dive state.
-Finally, the learned auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi$, was above 0.5 for every dimension and sub-dive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit of the data.
-The introduction of the parameter $\phi$ allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).
+Finally, the learned auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and sub-dive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data.
+The introduction of the parameter $\phi^*$ also allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).
 

### 2020-07-13 Report/Sections/models.tex
Small Edits

@@
 %\section{Models and parameter estimation}
-%\input Sections/model_Nancy.tex
 
@@
 We then review three variations on the HMM; the CarHMM, HHMM, and HMM-DFT.
+We also briefly describe a continuous-time model and its connection with the CarHMM.
 Finally, we generalize the hierarchical structure of these models and describe how they can be flexibly altered and combined to form new models.
@@
 
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/HMM.png}
-	\caption{Graphical representation of a traditional HMM.}
-	\label{fig:HMM}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-A key assumptions of an HMM is \textit{conditional independence} between observations given the state sequence.
+A key assumption of an HMM is \textit{conditional independence} between observations given the state sequence.
 Therefore, traditional HMMs do not hold when the observations exhibit certain forms of significant correlation in time.
 
-The CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by \citep{Lawler:2019}, explicit models auto-correlation within an HMM.
+The CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by \citep{Lawler:2019}, explicitly models auto-correlation within an HMM.
 Like a traditional HMM, A CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
@@
 The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. (Fig \ref{fig:CarHMM}) shows the structure of a CarHMM.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/CarHMM.png}
-	\caption{Graphical representation of a traditional CarHMM.}
-	\label{fig:CarHMM}
-\end{figure}
 
@@
 %
-\subsubsection{Equivalence with continuous-time process}
+\subsection{Continuous-time processes}
 
-Although the CarHMM can incorporate auto-correlation into the structure of an HMM, they break down when observations are taken at irregular time intervals since the probability transition matrix $\Gamma$ would then have to depend upon the time step between observations.
-A common solution to this issue is to use a continuous-time method such as the one introduced by Michelot et al \citep{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying behavioural state of the animal.
-Namely, a state-switching Ornstein-Uhlenbeck process $Y$ is the solution to the following stochastic differential equation:
+While CarHMMs can model auto-correlation within the observation sequence, they fail when observations are taken at irregular time intervals.
+Another way to model a stochastic process is to use a stochastic differential equation, or SDE.
+As an example, \citep{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
 %
@@
 where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
-One notable difference between the state-switching OU process and the CarHMM is that the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer.
+This SDE is referred to as a state-switching Ornstein-Uhlenbeck process.
+Unlike HMMs, the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer.
 If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
@@
 where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$.
-Continuous time models can be built up from arbitrarily complex stochastic differential equations and allow for uneven step lengths in the observations sequence $Y$.
-However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.
-
-Although most continuous time models are difficult to incorporate into an HHMM, we show in that under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process shown in (eqn \ref{eqn:OU_sol}).
+Most continuous time models are difficult to incorporate into an HHMM and require MCMC methods to fit.
+However, under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process.
 This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.
@@
 \begin{enumerate}
-    \item The hidden behavioural process $X$ from (Eq.
+    \item The hidden behavioural process $X$ from (eq.
 \ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
-    \item Observations of the SDE from (Eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
+    \item Observations of the SDE from (eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
 \end{enumerate}
-then the observations $y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
+then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
 
@@
 
-A \textit{hierarchical hidden Markov model} contains both a coarse-scale process and a fine-scale process, both of which are modeled by an HMM.
-The coarse-scale process is a hidden Markov model as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain and $Y_1,\ldots, Y_T$ are the corresponding observed responses.
-The possible states are $1,\ldots, N$.
+A \textit{hierarchical hidden Markov model} contains both a coarse-scale process and a fine-scale process, both of which are HMMs.
+The coarse-scale process is a hidden Markov model as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observed responses.
 %
-In the hierarchical setting, each state $X_t$ emits not only $Y_t$ but also another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
-The length of each fine-scale process, $T^*_t$, is a problem-specific tuning parameter.
+In the hierarchical setting, each state $X_t$ emits another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
+This fine-scale process can be either instead of or in addition to the coarse-scale observations $Y_t$.
+The length of each fine-scale process is denoted as $T^*_t$ and can often be controlled by the practitioner.
 Conditional on the coarse-scale hidden state $X_t$, the joint fine-scale process, ($X_t^*, Y_t^*$), is a hidden Markov model with parameters depending on the value of $X_t$.
@@
 %
-The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$, and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or sometimes simply denoted by $f^{*(i,i^*)}(\cdot)$.
+The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$, and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
 Let $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$, the fine-scale emission parameter vector corresponding to $X_t=i$.
 %
-In terms of the dependence structure, given the coarse-scale states, $X_1,\ldots, X_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs.
+Given the coarse-scale states, $X_1,\ldots, X_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs.
 Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states.
@@
 $\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$.
-(Fig \ref{fig:HHMM}) shows the dependence structure for an HHMM.
+(Fig \ref{fig:HHMM}) shows a graphical representation of the dependence structure for an HHMM.
 
@@
 
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/HHMM.png}
-	\caption{Graphical representation of a traditional HHMM.}
-	\label{fig:HHMM}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The HMM with discrete Fourier transform or HMM-DFT, is a variation of the hierarchical hidden Markov model where the fine-scale process is no longer modeled with an HMM and instead summarized using it Fourier transform.
+The HMM with discrete Fourier transform, or HMM-DFT, incorporates hierarchical structure into an HMM differently than an HHMM.
+In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform.
 For simplicity of presentation, we assume that the length of the fine-scale processes is constant, i.e.
 that $T^*_t = T^*$, although this need not be the case in general.
-Suppose that the fine-scale process $y^*_t$ does not switch states hidden states, but it does exhibit significant periodic behavior which cannot be modeled with a Markov chain.
+Suppose that the fine-scale process $y^*_t$ does not switch states hidden states, but does exhibit significant periodic behavior.
 Then, we suggest using the discrete Fourier transform (DFT) on $y^*_t$:
@@
 %
-Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One example of summary statistics include the following:
+Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One such example is as follows:
 %
@@
 %
-$z_t^{*(1)}$ is equal to the average value of $y^*_t$ and $z_t^{*(2)}$ is equal to the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$.
+$z_t^{*(1)}$ is the average value of $y^*_t$ and $z_t^{*(2)}$ is the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$.
 The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care.
@@
 
-\begin{figure}[t]
-	\centering
-	\includegraphics[width=5in]{../Plots/fourier_transform.png}
-	\caption{Visualization of transforming $\mathbf{y^*}_t$ into $\mathbf{z}_t$ using a discrete Fourier transform.}
-	\label{fig:fourier_example}
-\end{figure}
-
-Once $z^*_t$ is calculated, it can simply be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or simply $f^{(i)}\left(y_t,z^*_t\right)$.
+Once $z^*_t$ is calculated, it can simply be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$.
 In total, the likelihood of the hierarchical HMM-DFT is as follows:
@@
 
-In addition to the DFT and HMM models described above, the fine-scale process $Y^*_t$ can be modeled with \textbf{any} parametric model which admits an easy-to-compute likelihood.
+In addition to the DFT and HMM models described above, the fine-scale process $Y^*_t$ can be modeled with \textit{any} parametric model which admits an easy-to-compute likelihood.
 The state-specific emission distribution $\calL_{\text{HMM}}$ from the HHMM likelihood is then replaced by the likelihood of the fine-scale model, $\calL_{\text{fine}}(\mathbf{y}^*_t;\Theta^{*(i)})$:
@@
 \]
-where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$
+where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
 
-As previously mentioned, the fine-scale model can be any parametric model with easy-to-compute likelihood, including any of the four models described in the previous four subsections (HMM, CarHMM, HHMM, and HMM-DFT).
-These HMM models then act as building blocks that can be used to build increasingly complex hierarchical models based on the HMMs.
-In the sections that follow, we perform both a simulation study and real-world case study modelling killer whale dive behaviour using models built from these building blocks.
+Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, state-switching OU, HHMM, and HMM-DFT).
+These HMM models can act as building blocks which can be used to construct increasingly complex hierarchical models based on the HMMs.
+In the sections that follow, we perform both a simulation study and real-world case study modelling killer whale dive behaviour using models constructed from these building blocks.

### 2020-07-13 Report/Sections/results.tex
Small Edits

@@
 
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/raw_data.png}
-	\caption{Dive profile and Acceleration data of entire data set}
-	\label{fig:data}
-\end{figure}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/raw_data_5_dives.png}
-	\caption{Dive profile and acceleration data for a collection of 5 dives of a killer whale.}
-	\label{fig:data_one_dive}
-\end{figure}
-
 \subsection{Model Selection}
@@
 %
-$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i^*) = \phi_1^{*(i^*)} \mathbf{z} + (1-\phi_1^{*(i^*)}) \mathbf{\mu}_1^{*(i^*)}$$
-$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = \left(\mathbf{\sigma}_1^{*(i^*)}\right)^2 I_3$$
+$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)}$$
+$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right]$$
 %
-where $I_3$ is the identity matrix, $\phi_1^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i^*)} \in \mathbb{R}^3$.
+where $\phi_1^{*(i)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i)} \in \mathbb{R}^3$.
 
-While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, and the biological interpretation of auto-correlation within $Z^{*(2)}_{t,t^*}$ is unclear.
-For this reason, auto-correlation was not incorporated into the emission distribution of $Z^{*(2)}_{t,t^*}$.
+While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, and the biological interpretation of auto-correlation within $Z^{*(2)}_{t,t^*}$ is less clear.
+Auto-correlation was therefore not incorporated into the emission distribution of $Z^{*(2)}_{t,t^*}$.
 The distribution of $Z^{*(2)}_{t,t^*}$ was also assumed to be gamma:
 %
-$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = \mu_2^{*(i^*)}$$
-$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = (\sigma_2^{*(i^*)})^2$$
+$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
+$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2$$
 %
 $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent when conditioned of the sub-dive state $X_{t,t^*}$.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[height=7in]{../Plots/lagplot.png}
-	\caption{Lag plot of all features (left) and the associated normal kernel density estimates (right)}
-	\label{fig:lag}
-\end{figure}
 
@@
 
-The final model is nearly identical to the one from the simulation study, with the exception that the fine scale Markov chain has three sub-dive behaviors instead of two ($N^* = 3$), and that $\mathbf{z}^{*(1)}_{t,t^*}$ is a vector rather than a scalar.
-The model is comprised of two levels: a coarse-scale HMM and a fine-scale CarHHMM-DFT.
+The final model is nearly identical to the one from the simulation study, with the exception that the fine scale Markov chain has three sub-dive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a vector rather than a scalar.
+The model is comprised of two levels: a coarse-scale HMM and a fine-scale CarHMM-DFT.
 The coarse-scale model is comprised of an HMM with no auto-correlation and no DFT with hidden states corresponding to dive types and observations corresponding to dive durations.
-The fine-scale model is comprised of a CarHMM-DFT where auto-correlation is modeled into $\mathbf{z}^{*(1)}$ (the average acceleration with a 2-second window), but not $z^{*(2)}$ (the "wiggliness" of the 2-second window).
-Refer to (fig \ref{fig:CarHHMM}) for a graphical representation of this model.
+The fine-scale model is comprised of a CarHMM-DFT where auto-correlation is modeled into the distribution of $\mathbf{Z}^{*(1)}$ (the average acceleration with a 2-second window), but not $Z^{*(2)}$ (the ``wiggliness" of the 2-second window).
+Refer to (fig.
+\ref{fig:CarHHMM}) for a full graphical representation of this model.
 
@@
 
-(Table \ref{table:emis_dists}) displays the emission distribution parameter estimates. Each distribution is also plotted in (fig \ref{fig:coarse_emis}) and (fig \ref{fig:fine_emis}).
-%
-\begin{figure}[ht]
-	\centering
-	\includegraphics[height=4in]{../Plots/coarse-emissions.png}
-	\caption{Estimated probability distributions for each coarse-scale observation in each dive type.}
-	\label{fig:coarse_emis}
-\end{figure}
-%
-\begin{figure}[ht]
-	\centering
-	\includegraphics[height=2.5in]{../Plots/fine-emissions.png}
-	\caption{Estimated probability distributions for each fine-scale observation in each behavioral state.
-Note that the distributions of acceleration do not take auto-correlation into account (see table \ref{table:emis_dists})}
-	\label{fig:fine_emis}
-\end{figure}
-%
-\begin{table}[ht]
-    \centering
-    \caption{Estimates and standard errors of emission parameters for killer whale data.}
-    \scalebox{0.8}{
-    \begin{tabular}{ccccc}
-    \multirow{2}{*}{Feature}                 & \multirow{2}{*}{Dive / Sub-dive Type} & \multicolumn{3}{c}{Parameter Estimate}              \\
-                                             &                                      & $\hat \mu$      & $\hat \sigma$   & $\hat \phi$     \\ \hline
-    \multirow{2}{*}{Dive Duration (seconds)} & 1                                    & $27.23 \pm 0.63$ & $10.89 \pm 0.56$ & ---             \\
-                                             & 2                                    & $127.96 \pm 11.50$ & $64.13 \pm 9.21$ & ---             \\ \hline
-    \multirow{3}{*}{$Y^{*(1)}_x$}            & 1                                    & $0.98 \pm 0.07$ & $0.04 \pm 0.00$ & $0.99 \pm 0.00$ \\
-                                             & 2                                    & $0.22 \pm 0.01$ & $0.08 \pm 0.00$ & $0.87 \pm 0.01$ \\
-                                             & 3                                    & $0.23 \pm 0.03$ & $0.28 \pm 0.01$ & $0.62 \pm 0.03$ \\ \hline
-    \multirow{3}{*}{$Y^{*(1)}_y$}            & 1                                    & $0.61 \pm 0.09$ & $0.05 \pm 0.00$ & $0.99 \pm 0.00$ \\
-                                             & 2                                    & $0.43 \pm 0.01$ & $0.09 \pm 0.00$ & $0.87 \pm 0.01$ \\
-                                             & 3                                    & $0.38 \pm 0.04$ & $0.35 \pm 0.01$ & $0.62 \pm 0.04$ \\ \hline
-    \multirow{3}{*}{$Y^{*(1)}_z$}            & 1                                    & $-1.00 \pm 0.11$ & $0.05 \pm 0.00$ & $0.99 \pm 0.00$ \\
-                                             & 2                                    & $-0.57 \pm 0.01$ & $0.10 \pm 0.00$ & $0.87 \pm 0.01$ \\
-                                             & 3                                    & $-0.35 \pm 0.04$ & $0.34 \pm 0.01$ & $0.62 \pm 0.04$ \\ \hline
-    \multirow{3}{*}{$Y^{*(2)}$}              & 1                                    & $27.16 \pm 0.32$ & $16.67 \pm 0.32$ & ---             \\
-                                             & 2                                    & $406.98 \pm 4.42$ & $438.09 \pm 5.49$ & ---             \\
-                                             & 3                                    & $9688.54 \pm 221.95$ & $14584.02 \pm 358.40$ & ---             \\ \hline
-    \end{tabular}
-    }
-    \label{table:emis_dists}
-\end{table}
-%
+(Tbl.
+\ref{table:emis_dists}) displays estimates of the emission distribution parameters.
+Each fitted distribution is also plotted in (fig \ref{fig:coarse_emis}) and (fig \ref{fig:fine_emis}).
 While the ecological interpretation of behavioral states is tenuous for HMMs, we hypothesize the following meanings.
@@
 $$\hat \delta^{*(1)} = \begin{pmatrix} 0.143 & 0.698 & 0.159 \end{pmatrix} \qquad
-\hat \delta^{*(2)} = \begin{pmatrix} 0.476 & 0.456 & 0.067 \end{pmatrix}$$.
+\hat \delta^{*(2)} = \begin{pmatrix} 0.476 & 0.456 & 0.067 \end{pmatrix}.$$
 %
@@
 
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/labeled_dives.png}
-	\caption{Features of a particular set of killer whale dives and decoded estimates for the intra-dive behavioral states.
-The color of the plot corresponds to behavioral or dive state with the highest probability.}
-	\label{fig:labeled_dives}
-\end{figure}
-%
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/Coarse_state_probs.png}
-	\caption{Probabilities of dive types for the set of killer whale dives from (fig \ref{fig:labeled_dives}).}
-	\label{fig:coarse_probs}
-\end{figure}
-%
-\begin{figure}[ht]
-	\centering
-	\begin{subfigure}[t]{1.0\textwidth}
-        \centering
-        \includegraphics[width=5in]{../Plots/Fine_state_probs_1.png}
-        \caption{Fine-scale state 1 probabilities}
-    \end{subfigure}
-    \newline
-    \begin{subfigure}[t]{1.0\textwidth}
-        \centering
-        \includegraphics[width=5in]{../Plots/Fine_state_probs_3.png}
-        \caption{Fine-scale state 3 probabilities}
-    \end{subfigure}
-	\caption{Probabilities of sub-dive types for the set of killer whale dives from (fig \ref{fig:labeled_dives}).}
-	\label{fig:fine_probs}
-\end{figure}
-
 \subsection{Model Validation}
@@
 To easily visualize outliers, the pseudo-residual is often passed through the quantile function of the standard normal distribution.
-In particular, the pseudo-residual of an observation $y_t$ is $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard normal distribution.
-If the model is correct, then all pseudo-residuals should be independent and follow a standard normal distribution.
-We find that histograms of the pseudoresiduals mostly support that the model is well-specified.
-$Z^{*(2)}$ is a notable exception, as its pseudo-residuals are noticeably right-skewed (fig \ref{fig:pseudoresids}).
+Mathematically, the pseudo-residual of an observation $y_t$ is $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard normal distribution.
+If the model is correct, then all pseudo-residuals are independent and follow a standard normal distribution.
+We find that histograms of the pseudoresiduals of this model mostly support that the model is well-specified.
+$Z^{*(2)}$ is an exception, as its pseudo-residuals are noticeably right-skewed (fig \ref{fig:pseudoresids}).
 This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution compared to a gamma distribution such as a power law.
 
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/pseudoresids.png}
-	\caption{Psuedoresiduals of $Y^{*(2)}$}
-	\label{fig:pseudoresids}
-\end{figure}
-
-In addition to psuedoresiduals, we plot histograms of observations, weighted by the probability that the whale was in a particular hidden state.
-This empirical distribution was then plotted over the fitted probability distribution function of that feature and behavioral state.
+In addition to psuedoresiduals, we plotted histograms of each feature where each observation was weighted by the probability that the whale was in a particular hidden state.
+This empirical distribution was then plotted over the fitted probability distribution function of that feature and hidden state.
+If the model is correct, then the histograms of features should closely resemble their respective fitted probability distribution.
 Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is right-skewed.
-In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for sub-dive state 3 (\ref{fig:empirical_dist}), indicating the existence of rare events corresponding to very violent trashing of the killer whale.
+In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for sub-dive state 3 (fig.
+\ref{fig:empirical_dist}), indicating the existence of rare events corresponding to very violent trashing of the killer whale.
 These outliers are potential subjects for future study.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/empirical_dist.png}
-	\caption{Empirical distribution of $Y^{*(1)}_z$ for sub-dive state 3 plotted over its estimated pdf.}
-	\label{fig:empirical_dist}
-\end{figure}

### 2020-07-13 Report/Sections/simulation.tex
Small Edits

@@
 
-We base the simulated study on acceleration data of a killer whale dive sequence.
-The parameters used in this simulation study are loosely based on those learned from the case study in section \ref{sec:case_study}.
+We base the following simulated study on acceleration data of a killer whale dive sequence.
+The parameters used to generate the data are loosely based on those learned from the case study in section \ref{sec:case_study}.
+We fit four separate models to the simulated data and compare their performance.
 
@@
 We parameterize the gamma distribution by its mean and variance:
-
+%
 $$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}, \qquad \delta =  \begin{pmatrix} 0.5 & 0.5 \end{pmatrix}$$
@@
 \end{align*}
-
+%
 Once the dive durations were calculated for all 100 dives, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments (the end of the dive sequence was discarded) which made up a second fine-scale hidden Markov model.
-Each 2-second segment of the dive was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$, and the probability transition matrices for these fine-scale Markov chains were set as follows:
+Each 2-second segment of the dive was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$, and the probability transition matrices for these fine-scale Markov chains were set as
 %
-$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.9 & 0.1 \end{pmatrix} \qquad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}$$
+$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.9 & 0.1 \end{pmatrix}, \qquad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix},$$ 
 %
@@
 
-\begin{align*}
-    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 1 \right) &\sim \mathcal{N} \left(\phi^{*(1)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(1)})\mu^{*(1)}, (\sigma^{*(1)})^2 \right) \\
+\begin{gather*}
+    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 1 \right) \sim \mathcal{N} \left(\phi^{*(1)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(1)})\mu^{*(1)}, (\sigma^{*(1)})^2 \right) \\
     %
-    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 2 \right) &\sim \mathcal{N} \left(\phi^{*(2)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(2)})\mu^{*(2)}, (\sigma^{*(2)})^2 \right) \\
+    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 2 \right) \sim \mathcal{N} \left(\phi^{*(2)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(2)})\mu^{*(2)}, (\sigma^{*(2)})^2 \right) \\
     %
-    \mu^{*(1)} = 0.0,\enspace \sigma^{*(1)} &= 0.05,\enspace \phi^{*(1)} = 0.99 \\
+    \mu^{*(1)} = 0.0,\enspace \sigma^{*(1)} = 0.05,\enspace \phi^{*(1)} = 0.99 \\
     %
-    \mu^{*(2)} = 0.0,\enspace \sigma^{*(2)} &= 0.1,\enspace \phi^{*(2)} = 0.95 \\\\
+    \mu^{*(2)} = 0.0,\enspace \sigma^{*(2)} = 0.1,\enspace \phi^{*(2)} = 0.95 \\\\
     %
-    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 1\right) &\sim {\rm{Gamma}}\left(\alpha^{*(1)}, \beta^{*(1)} \right) \\
+    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha^{*(1)}, \beta^{*(1)} \right) \\
     %
-    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 2\right) &\sim {\rm{Gamma}}\left(\alpha^{*(2)}, \beta^{*(2)} \right) \\
+    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(\alpha^{*(2)}, \beta^{*(2)} \right) \\
     %
-    \alpha^{*(1)} = 10.10, \enspace & \enspace \beta^{*(1)} = 1.00 \\
+    \alpha^{*(1)} = 10.10, \enspace \enspace \beta^{*(1)} = 1.00 \\
     %
-    \alpha^{*(2)} = 305.94, \enspace & \enspace \beta^{*(2)} = 1.00
-\end{align*}
+    \alpha^{*(2)} = 305.94, \enspace \enspace \beta^{*(2)} = 1.00
+\end{gather*}
 %
-In short, sub-dive behavior 1 corresponds to passive gliding while sub-dive behaviour 2 corresponds to active swimming at a frequency of 1 hertz.
-In addition, sub-dive behaviors 1 and 2 are the same for both dive types.
-See the appendix for more details regarding procedure for simulating $\hat Y^*$ and $Y^*$ in addition to $Z^*$.
-(fig.
+Sub-dive behavior 1 corresponds to passive gliding while sub-dive behaviour 2 corresponds to active swimming with a dominant frequency of $\frac{1}{2}s^{-1}$.
+Sub-dive behaviors 1 and 2 are the same for both dive types.
+See the appendix for more details regarding procedure for simulating $\hat Y^*$ and $Y^*$ such that $Z^*$ has the preceding distribution.
+(Fig.
 \ref{fig:sim_data}) shows the first 5 dives of one simulated data set.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/sim_data.png}
-	\caption{Simulated acceleration data for one dive.
-The color of the line corresponds to the true fine-scale state of the sub-dive process, while the color of the background corresponds to the true dive type of the simulated whale, where blue corresponds to dive type 1 and orange corresponds to dive type 2.}
-	\label{fig:sim_data}
-\end{figure}
 
@@
 Specifically, we used a hierarchical HMM where the sequence of dive durations $Y$ was modeled using a simple HMM, and the fine-scale process $Z^*$ was modeled using a HMM-DFT with explicit auto-correlation in $Z^{*(1)}$.
-Naturally, we refer to this model as the \textit{CarHHMM-DFT}, and (Fig.
-\ref{fig:CarHHMM}) shows the corresponding graphical model.
-%
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/CarHHMM.png}
-	\caption{Graphical representation of this model.}
-	\label{fig:CarHHMM}
-\end{figure}
-%
-
-\subsection{Parameters}
+Naturally, we refer to this model as the \textit{CarHHMM-DFT}.
 
@@
 Each two-second window is summarized by the observations $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$.
-$Z^{*(1)}_{t,t^*}$ follows a normal distribution whose parameters depend upon the sub-dive behavior $X^*_{t,t^*}$ and $Z^{*(1)}_{t,t^*-1}$.
+The distribution of $Z^{*(1)}_{t,t^*}$ is normal and its parameters depend upon the sub-dive behavior $X^*_{t,t^*}$ and $Z^{*(1)}_{t,t^*-1}$.
 In particular:
@@
 %
-$Z^{*(2)}_{t,t^*}$ follows a gamma distribution whose parameters depend upon only $X^*-{t,t^*}$:
+The distribution of $Z^{*(2)}_{t,t^*}$ is gamma and its parameters depend upon only $X^*_{t,t^*}$:
 %
@@
 %
-None of the fine-scale emission distributions depend upon dive type. In total the parameters to learn are:
+None of the fine-scale emission distributions depend upon dive type. In total the parameters to learn are
+%
+\begin{gather*}
+    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)} \\
+    %
+    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters)} \\
+    %
+    \Theta^* = \{\Theta^{*(1)},\Theta^{*(2)}\}  \qquad \text{(fine-scale emission parameters)} \\
+    %
+    \Theta^{*(i^*)} =  \{\{\mu_1^{*(i^*)},\sigma_1^{*(i^*)},\phi_1^{*(i^*)}\},\{\mu_2^{*(i^*)},\sigma_2^{*(i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters)}
+\end{gather*}
+%
+The likelihood of this model is still easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
+See the appendix for details of likelihood evaluation, and (fig.
+\ref{fig:CarHHMM}) shows the corresponding graphical model.
 
-$$\Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}$$
-$$\Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters)}$$
-$$\Theta^* = \{\Theta^{*(1)},\Theta^{*(2)}\}  \qquad \text{(fine-scale emission parameters)}$$
-$$\Theta^{*(i^*)} =  \{\{\mu_1^{*(i^*)},\sigma_1^{*(i^*)},\phi_1^{*(i^*)}\},\{\mu_2^{*(i^*)},\sigma_2^{*(i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters)}$$
-
-The likelihood of this model is still easy to calculate using the forward algorithm, and can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation.
-
-Including the one described above, four different models were fit to the simulated data sets:
-
+Including the CarHHMM-DFT described above, four different models were fit to the simulated data sets:
 \begin{enumerate}
-    \item A CarHHMM-DFT as described above (\textbf{CarHHMM-DFT}).
-    \item An HHMM-DFT, which is similar to the model above, but with no modeled auto-correlation, i.e. $\phi_1^{*(i^*)} = 0$ for $i^* = 1,2$ (\textbf{HHMM-DFT}).
-    \item A CarHHMM, which is similar to the model above, but which does not have access to $Z^{*(2)}$, i.e $\Theta^{*(i^*)} = \{\mu_1^{(i^*)},\sigma_1^{(i^*)},\phi_1^{(i^*)}\}$, $i^* = 1,2$ (\textbf{CarHHMM}).
-    \item A CarHMM-DFT similar to the CarHHMM-DFT, but with $N=1$ instead of $N=2$, i.e. $\Gamma^* = \{\Gamma^{*(1)}\}$ and $\Theta^* = \{\Theta^{*(1)}\}$ (\textbf{CarHMM-DFT}). This is 
+    \item A \textbf{CarHHMM-DFT} as described above.
+    \item An \textbf{HHMM-DFT}, which is similar to the model above, but with no modeled auto-correlation, i.e. $\phi_1^{*(i^*)} = 0$ for $i^* = 1,2$.
+    \item A \textbf{CarHHMM}, which is similar to the model above, but without $Z^{*(2)}$ as an observation, i.e $\Theta^{*(i^*)} = \{\mu_1^{(i^*)},\sigma_1^{(i^*)},\phi_1^{(i^*)}\}$, $i^* = 1,2$.
+    \item A \textbf{CarHMM-DFT} similar to the CarHHMM-DFT, but with $N=1$ instead of $N=2$, i.e.
+$\Gamma^* = \{\Gamma^{*(1)}\}$ and $\Theta^* = \{\Theta^{*(1)}\}$.
+This is equivalent to loosing one level of the hierarchical structure.
 \end{enumerate}
 %
-Each of the last three models leaves out one important aspect of the full model (CarHHMM-DFT).
-The CarHMM-DFT lacks a hierarchical structure, the HHMM-DFT is missing auto-correlation within the fine-scale observations, and the CarHHMM does not have access to the Fourier transform sums $Z^{*(2)}$ as observations.
-All models were run on the Cedar Compute Canada cluster with 1 GPU and 4 GB of dedicated memory per model.
+Each of the last three models leaves out one important aspect of the full CarHHMM-DFT.
+The CarHMM-DFT lacks a hierarchical structure, the HHMM-DFT is missing auto-correlation within the fine-scale observations, and the CarHHMM does not have access to the Fourier transform sums $(Z^{*(2)})$ as observations.
+All models were run on the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
 
@@
 %%% Accuracy %%%
-Every model was able to decode the fine-scale hidden states of the process almost perfectly except for the \textit{CarHHMM}, which did not have access to the Fourier modes $Z^{*(2)}$.
-This is intuitively clear because the fine-scale states are much more distinct in their Fourier modes than in their average acceleration.
-For the coarse-scale hidden state, the CarHMM lacked a hierarchical structure and could not make any predictions at all.
+Every model was able to decode the fine-scale hidden states of the process almost perfectly except for the \textit{CarHHMM}, which did not have access to the Fourier modes ($Z^{*(2)}$).
+This is intuitively clear because the distribution of $Z^{*(2)}$ varies between fine-scale states much more than $Z^{*(1)}$.
+For the coarse-scale hidden states, the CarHMM lacked a hierarchical structure and could not make any predictions at all.
 The other three models all achieved an accuracy of approximately 90\%, with the CarHHMM slightly more likely to categorize a dive as dive type 2 than the other models.
 (Fig.
-\ref{fig:acc}) shows the decoded state probabilities of both the fine and coarse hidden states for the first 5 dives of one simulated data set.
+\ref{fig:acc}) shows the decoded state probabilities for both the fine- and coarse- scales.
 (Tbl.
-\ref{table:accuracy}) also lists more details regarding the accuracy and training times of each model for each true dive and sub-dive state.
-
-\begin{figure}[ht]
-    \centering
-    \begin{subfigure}[t]{1.0\textwidth}
-        \centering
-        \includegraphics[width=5in]{Plots/Posterior_Coarse_States.png}
-        \caption{Coarse-scale hidden process}
-    \end{subfigure}
-    \newline
-    \begin{subfigure}[t]{1.0\textwidth}
-        \centering
-        \includegraphics[width=5in]{Plots/Posterior_Fine_States.png}
-        \caption{Fine-scale hidden process}
-    \end{subfigure}
-	\caption{Decoded state probabilities of each model.
-The color of the line corresponds to the true fine-scale state of the sub-dive process, while the color of the background corresponds to the true dive type of the simulated whale.}
-	\label{fig:acc}
-\end{figure}
+\ref{table:accuracy}) also lists more details regarding the accuracy and training times of each model.
 
@@
 The CarHMM model in particular severely underestimates $\sigma$.
-This is likely due to the fact that dive duration follows a mixture of gammas rather than a gamma distribution, and the CarHMM does not have the machinery to capture this fact.
+%This is likely due to the fact that dive duration follows a mixture of gammas rather than a gamma distribution, and the CarHMM does not have the machinery to capture this fact.
 A full table of parameter estimates for all models is shown in (table \ref{table:dive_duration}).
@@
 %because it sees the ``wigliness" of the acceleration data as raw acceleration rather than as Fourier modes, which deflates the value of $\hat \phi$ and inflates the value of $\hat \sigma$. 
-See (table \ref{table:acceleration}) for a detailed breakdown of parameter values for acceleration emission distributions.
+See (tbl. \ref{table:acceleration}) for a detailed breakdown of parameter values for acceleration emission distributions.
 
 %%% FoVeDBA %%%
-The distribution of parameters for the Fourier sums ($Z^{*(2)}$) for all models except for the CarHHMM converge regularly to the correct values, and the observed fisher information standard errors are good approximation of the empirical standard error.
+For all models there is no bias in the parameter estimates for the distribution of the Fourier sums ($Z^{*(2)}$).
+One exception is the CarHHMM, which does not model $Z^{*(2)}$ as an observation.
+The observed fisher information standard errors are good approximations of the empirical standard error.
 %The distribution of $Z^{*(2)}$ is very different between subdive states 1 and 2, so the signal is therefore very easy for each model to pick up on.
-(Table \ref{table:FoVeDBA}) shows a detailed breakdown of the emission distribution of $Z^{*(2)}$ for all models.
+(Tbl. \ref{table:FoVeDBA}) shows a detailed breakdown of the emission distribution of $Z^{*(2)}$ for all models.
 
@@
 
-Estimates for $\Gamma$ for all models (except for the CarHMM-DFT) are very accurate, with empirical standard errors of approximately 0.02.
+Estimates of the probability transition matrix $\Gamma$ for all models (except for the CarHMM-DFT) are very accurate, with empirical standard errors of approximately 0.02.
 This is significantly less than the observed fisher information estimate of standard error of approximately 0.08.
@@
 For $\Gamma^*$, the HHMM-DFT and CarHHMM-DFT both showed practically no bias with standard errors on the order of $10^{-2}$.
-One notable exception is the standard error of $\hat \Gamma^{*(1)}_{12}$, whose standard error was on the order of $10^{-4}$, which is much lower than the observed Fischer information predicted.
-The CarHMM-DFT is mis-specified, so its results cannot be interpreted.
+One notable exception is the standard error of $\hat \Gamma^{*(1)}_{12}$, whose standard error was on the order of $10^{-4}$, which is much lower than the observed fisher information predicted.
+The CarHMM-DFT is mis-specified, so its results cannot be easily interpreted.
 The CarHHMM consistently underestimated the probability of a hidden state change on the sub-dive behavior.
@@
 But there are just SO many plots (4 models * 3 observations * 2 states = 24 plots)}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=5in]{../Plots/hmm_FV_MLE_density_dive_duration_-1.png}
-	\caption{\textcolor{red}{KDE plot of $\hat \mu$ and $\hat \sigma$ for the dive duration emission distribution for the CarHMM. This plot is for Marie and Nancy to look over.}}
-	\label{fig:MLE_dist}
-\end{figure}
-
-%%% Accuracy %%%
-
-\begin{table}
-\centering
-\caption{Accuracies and run times for all models. All reported values are averages, and $\pm$ refers to the standard deviation.}
-\scalebox{0.8}{
-\begin{tabular}{ccccccc}
-Model                      & \multicolumn{1}{c}{Training Time (Minutes)} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Dive Accuracy} & \multicolumn{1}{c}{Subdive Accuracy}  \\ \hline
-\multirow{5}{*}{CarHMM}    & \multirow{5}{*}{$15.74 \pm 2.46$}   & Both                          & Both                             & -------------                     & $1.00 \pm 0.00$                       \\
-                           &                                    & 1                             & 1                                & \multirow{2}{*}{-------------}    & $1.00 \pm 0.00$                       \\
-                           &                                    & 1                             & 2                                &                                   & $1.00 \pm 0.00$                       \\
-                           &                                    & 2                             & 1                                & \multirow{2}{*}{-------------}    & $1.00 \pm 0.00$                       \\
-                           &                                    & 2                             & 2                                &                                   & $1.00 \pm 0.00$                       \\ \hline
-\multirow{5}{*}{HHMM}      & \multirow{5}{*}{$82.43 \pm 11.48$}   & Both                          & Both                             & $0.94 \pm 0.02$                   & $1.00 \pm 0.00$                       \\
-                           &                                    & 1                             & 1                                & \multirow{2}{*}{$0.94\pm0.03$}    & $1.00 \pm 0.00$                       \\
-                           &                                    & 1                             & 2                                &                                   & $1.00 \pm 0.00$                       \\
-                           &                                    & 2                             & 1                                & \multirow{2}{*}{$0.94\pm0.03$}    & $1.00 \pm 0.00$                       \\
-                           &                                    & 2                             & 2                                &                                   & $1.00 \pm 0.00$                       \\ \hline
-\multirow{5}{*}{CarHHMM1}  & \multirow{5}{*}{$70.85 \pm 15.89$}   & Both                          & Both                             & $0.91 \pm 0.03$                   & $0.89 \pm 0.02$                       \\
-                           &                                    & 1                             & 1                                & \multirow{2}{*}{$0.87\pm0.04$}    & $0.44 \pm 0.12$                       \\
-                           &                                    & 1                             & 2                                &                                   & $1.00 \pm 0.00$                       \\
-                           &                                    & 2                             & 1                                & \multirow{2}{*}{$0.95\pm0.03$}    & $0.81 \pm 0.04$                       \\
-                           &                                    & 2                             & 2                                &                                   & $1.00 \pm 0.00$                       \\ \hline
-\multirow{5}{*}{CarHHMM2}  & \multirow{5}{*}{$81.22 \pm 16.10$}   & Both                          & Both                             & $0.94 \pm 0.02$                   & $1.00 \pm 0.00$                       \\
-                           &                                    & 1                             & 1                                & \multirow{2}{*}{$0.94\pm0.03$}    & $1.00 \pm 0.00$                       \\
-                           &                                    & 1                             & 2                                &                                   & $1.00 \pm 0.00$                       \\
-                           &                                    & 2                             & 1                                & \multirow{2}{*}{$0.94\pm0.03$}    & $1.00 \pm 0.00$                       \\
-                           &                                    & 2                             & 2                                &                                   & $1.00 \pm 0.00$                       \\ \hline
-\end{tabular}
-}
-\label{table:accuracy}
-\end{table}
-
-%%% dive duration %%%
-
-\begin{table}
-\centering
-\caption{Estimates and standard errors of parameters for dive duration distribution for all four models.
-All reported values are averages, except for the Fisher observed standard error, which are medians.
-$\pm$ refers to the IQR.}
-\scalebox{0.9}{
-\begin{tabular}{ccccccc}
-Model                      & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}           \\ \hline
-\multirow{4}{*}{CarHMM}    & \multirow{2}{*}{$\mu$}        & 1                             & $49.72$                         & $-0.28$                     & $4.78$                             & $2.47 \pm 0.34$                             \\
-                           &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & $39.00$                         & $-7.51$                     & $5.05$                             & $2.50 \pm 0.40$                             \\
-                           &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\ \hline
-\multirow{4}{*}{HHMM}      & \multirow{2}{*}{$\mu$}        & 1                             & $19.99$                         & $-0.01$                     & $0.75$                             & $0.69 \pm 0.11$                             \\
-                           &                               & 2                             & $79.85$                         & $-0.15$                     & $8.05$                             & $5.85 \pm 1.10$                             \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & $4.90$                         & $-0.10$                     & $0.61$                             & $0.53 \pm 0.10$                             \\
-                           &                               & 2                             & $48.74$                         & $-1.26$                     & $6.50$                             & $5.15 \pm 1.02$                             \\ \hline
-\multirow{4}{*}{CarHHMM 1} & \multirow{2}{*}{$\mu$}        & 1                             & $19.91$                         & $-0.09$                     & $0.77$                             & $0.71 \pm 0.12$                             \\
-                           &                               & 2                             & $75.80$                         & $-4.20$                     & $7.72$                             & $5.32 \pm 0.98$                             \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & $4.73$                         & $-0.27$                     & $0.59$                             & $0.55 \pm 0.10$                             \\
-                           &                               & 2                             & $49.48$                         & $-0.52$                     & $6.26$                             & $4.79 \pm 0.93$                             \\ \hline
-\multirow{4}{*}{CarHHMM 2} & \multirow{2}{*}{$\mu$}        & 1                             & $19.99$                         & $-0.01$                     & $0.75$                             & $0.69 \pm 0.12$                             \\
-                           &                               & 2                             & $79.85$                         & $-0.15$                     & $8.05$                             & $5.85 \pm 1.10$                             \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & $4.90$                         & $-0.10$                     & $0.61$                             & $0.53 \pm 0.10$                             \\
-                           &                               & 2                             & $48.74$                         & $-1.26$                     & $6.50$                             & $5.15 \pm 1.02$
-\end{tabular}
-}
-\label{table:dive_duration}
-\end{table}
-
-%%% Acceleration %%%
-
-\begin{table}
-\centering
-\caption{Estimates and standard errors of parameters for $Z^{*(1)}_{t,t^*}$ for all four models.
-All reported values are averages, except for the Fisher observed standard error, which are medians.
-$\pm$ refers to the IQR.}
-\scalebox{0.7}{
-\begin{tabular}{ccccccc}
-Model                      & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}        \\ \hline
-\multirow{6}{*}{CarHMM}    & \multirow{2}{*}{$\mu$}        & 1                             & $0.00$                         & $0.00$                     & $0.00$                             & $0.14 \pm 0.13$                             \\
-                           &                               & 2                             & $0.00$                         & $0.00$                     & $0.01$                             & $0.06 \pm 0.02$                             \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & $0.05$                         & $-0.00$                     & $0.00$                             & $0.00 \pm 0.00$                             \\
-                           &                               & 2                             & $0.10$                         & $-0.00$                     & $0.00$                             & $0.00 \pm 0.00$                             \\
-                           & \multirow{2}{*}{$\phi$}       & 1                             & $0.99$                         & $-0.00$                     & $0.01$                             & $0.01 \pm 0.00$                             \\
-                           &                               & 2                             & $0.95$                         & $-0.00$                     & $0.01$                             & $0.01 \pm 0.00$                             \\ \hline
-\multirow{6}{*}{HHMM}      & \multirow{2}{*}{$\mu$}        & 1                             & $-0.01$                         & $-0.01$                     & $0.02$                             & $0.01 \pm 0.00$                             \\
-                           &                               & 2                             & $-0.00$                         & $-0.00$                     & $0.02$                             & $0.01 \pm 0.00$                             \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & $0.25$                         & $0.20$                     & $0.04$                             & $0.00 \pm 0.00$                             \\
-                           &                               & 2                             & $0.24$                         & $0.14$                     & $0.03$                             & $0.00 \pm 0.00$                             \\
-                           & \multirow{2}{*}{$\phi$}       & 1                             & ------                         & ------                     & ------                             & ------                                      \\
-                           &                               & 2                             & ------                         & ------                     & ------                             & ------                                      \\ \hline
-\multirow{6}{*}{CarHHMM 1} & \multirow{2}{*}{$\mu$}        & 1                             & $0.00$                         & $0.00$                     & $0.00$                             & $0.08 \pm 0.04$                             \\
-                           &                               & 2                             & $-0.01$                         & $-0.01$                     & $0.01$                             & $0.01 \pm 0.00$                             \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & $0.05$                         & $0.00$                     & $0.04$                             & $0.00 \pm 0.00$                             \\
-                           &                               & 2                             & $0.27$                         & $0.17$                     & $0.01$                             & $0.00 \pm 0.00$                             \\
-                           & \multirow{2}{*}{$\phi$}       & 1                             & $0.97$                         & $-0.02$                     & $0.10$                             & $0.00 \pm 0.00$                             \\
-                           &                               & 2                             & $0.49$                         & $-0.46$                     & $0.05$                             & $0.02 \pm 0.00$                             \\ \hline
-\multirow{6}{*}{CarHHMM 2} & \multirow{2}{*}{$\mu$}        & 1                             & $0.00$                         & $0.00$                     & $0.00$                             & $0.13 \pm 0.12$                             \\
-                           &                               & 2                             & $0.00$                         & $0.00$                     & $0.00$                             & $0.06 \pm 0.02$                             \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & $0.05$                         & $-0.00$                     & $0.00$                             & $0.00 \pm 0.00$                             \\
-                           &                               & 2                             & $0.10$                         & $-0.00$                     & $0.00$                             & $0.00 \pm 0.00$                             \\
-                           & \multirow{2}{*}{$\phi$}       & 1                             & $0.99$                         & $-0.00$                     & $0.01$                             & $0.01 \pm 0.00$                             \\
-                           &                               & 2                             & $0.95$                         & $-0.00$                     & $0.01$                             & $0.01 \pm 0.00$                             \\ \hline
-\end{tabular}
-}
-\label{table:acceleration}
-\end{table}
-
-%%% FoVeDBA %%%
-
-\begin{table}
-\centering
-\caption{Estimates and standard errors of parameters for $Z^{*(2)}_{t,t^*}$ for all four models.
-All reported values are averages, except for the Fisher observed standard error, which are medians.
-$\pm$ refers to the IQR.}
-\scalebox{0.8}{
-\begin{tabular}{ccccccc}
-Model                      & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}        \\ \hline
-\multirow{4}{*}{CarHMM}    & \multirow{2}{*}{$\mu$}        & 1                             & $10.10$                         & $-0.00$                     & $0.09$                             & $0.08 \pm 0.01$                             \\
-                           &                               & 2                             & $305.97$                         & $0.03$                     & $0.54$                             & $0.51 \pm 0.03$                             \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & $3.18$                         & $-0.00$                     & $0.07$                             & $0.06 \pm 0.01$                             \\
-                           &                               & 2                             & $17.46$                         & $-0.03$                     & $0.37$                             & $0.36 \pm 0.02$                             \\ \hline
-\multirow{4}{*}{HHMM}      & \multirow{2}{*}{$\mu$}        & 1                             & $10.10$                         & $-0.00$                     & $0.09$                             & $0.08 \pm 0.01$                             \\
-                           &                               & 2                             & $305.97$                         & $0.03$                     & $0.54$                             & $0.51 \pm 0.03$                             \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & $3.18$                         & $-0.00$                     & $0.07$                             & $0.06 \pm 0.01$                             \\
-                           &                               & 2                             & $17.46$                         & $-0.03$                     & $0.37$                             & $0.36 \pm 0.02$                             \\ \hline
-\multirow{4}{*}{CarHHMM 1} & \multirow{2}{*}{$\mu$}        & 1                             & ------                         & ------                     & ------                             & ------                                      \\
-                           &                               & 2                             & ------                         & ------                     & ------                             & ------                                      \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & ------                         & ------                     & ------                             & ------                                      \\
-                           &                               & 2                             & ------                         & ------                     & ------                             & ------                                      \\ \hline
-\multirow{4}{*}{CarHHMM 2} & \multirow{2}{*}{$\mu$}        & 1                             & $10.10$                         & $-0.00$                     & $0.09$                             & $0.08 \pm 0.01$                             \\
-                           &                               & 2                             & $305.97$                         & $0.03$                     & $0.54$                             & $0.51 \pm 0.03$                             \\
-                           & \multirow{2}{*}{$\sigma$}     & 1                             & $3.18$                         & $-0.00$                     & $0.07$                             & $0.06 \pm 0.01$                             \\
-                           &                               & 2                             & $17.46$                         & $-0.03$                     & $0.37$                             & $0.36 \pm 0.02$
-\end{tabular}
-}
-\label{table:FoVeDBA}
-\end{table}
-
-% Gamma
-\begin{table}[t]
-\centering
-\caption{Estimates and standard errors of $\Gamma$ and $\Gamma^*$ for all four models. All reported values are averages except for the observed fisher SE, which is a median. $\pm$ refers to the IQR.}
-\scalebox{0.7}{
-\begin{tabular}{ccccccc}
-Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
-\multirow{6}{*}{HHMM-DFT}    & $\Gamma_{12}$                 & $0.50$                         & $-0.00$                   & $0.03$                           & $0.08 \pm 0.01$                             \\
-                             & $\Gamma_{21}$                 & $0.50$                         & $-0.00$                   & $0.03$                           & $0.08 \pm 0.01$                             \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.50$                         & $0.00$                   & $0.00$                           & $0.07 \pm 0.01$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.10$                         & $0.00$                   & $0.02$                           & $0.02 \pm 0.00$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.20$                         & $-0.00$                   & $0.01$                           & $0.01 \pm 0.00$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.30$                         & $-0.00$                   & $0.02$                           & $0.02 \pm 0.00$                             \\ \hline
-\multirow{6}{*}{CarHMM-DFT}  & $\Gamma_{12}$                 & ------                         & ------                   & ------                           & ------                                      \\
-                             & $\Gamma_{21}$                 & ------                         & ------                   & ------                           & ------                                      \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.23$                         & ------                   & $0.01$                           & $0.00 \pm 0.00$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.23$                         & ------                   & $0.02$                           & $0.00 \pm 0.00$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & ------                         & ------                   & ------                           & ------                                      \\
-                             & $\Gamma^{*(2)}_{21}$          & ------                         & ------                   & ------                           & ------                                      \\ \hline
-\multirow{6}{*}{CarHHMM}     & $\Gamma_{12}$                 & $0.49$                         & $-0.01$                   & $0.02$                           & $0.08 \pm 0.01$                             \\
-                             & $\Gamma_{21}$                 & $0.50$                         & $0.00$                   & $0.02$                           & $0.08 \pm 0.01$                             \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.50$                         & $-0.00$                   & $0.00$                           & $0.23 \pm 0.19$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.04$                         & $-0.06$                   & $0.02$                           & $0.01 \pm 0.00$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.11$                         & $-0.09$                   & $0.02$                           & $0.01 \pm 0.00$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.11$                         & $-0.19$                   & $0.03$                           & $0.01 \pm 0.00$                             \\ \hline
-\multirow{6}{*}{CarHHMM-DFT} & $\Gamma_{12}$                 & $0.50$                         & $-0.00$                   & $0.03$                           & $0.08 \pm 0.01$                             \\
-                             & $\Gamma_{21}$                 & $0.50$                         & $-0.00$                   & $0.03$                           & $0.08 \pm 0.01$                             \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.50$                         & $0.00$                   & $0.00$                           & $0.07 \pm 0.01$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.10$                         & $0.00$                   & $0.02$                           & $0.02 \pm 0.00$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.20$                         & $-0.00$                   & $0.01$                           & $0.01 \pm 0.00$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.30$                         & $-0.00$                   & $0.02$                           & $0.02 \pm 0.00$                             \\ \hline
-
-\end{tabular}
-}
-\label{table:Gamma}
-\end{table}

### 2020-07-13 Report/main.tex
Commit before changing STFT to Fourier Transform
Update appendix - before moving tables and figures
Small Edits

--- 2020-06-25 Report/main.tex
@@
 \usepackage[dvipsnames]{xcolor}
-
+\usepackage{url}
 \input{header}
@@
 \newcommand{\rv}{random variable}
+\newcommand{\newblock}{}
+\bibliographystyle{abbrvnat}
 
 % Title, authors, affiliations
-\title[]{Modelling fine-scale dive behavior with hidden Markov models }%\query{Q1}
+\title[]{Modelling fine-scale correlation structures using hidden Markov models}%\query{Q1}
 \author{BlindedA\authorref{1}\thanksref{*}}
@@
 technology have made available a vast
-amount of animal movement data. This
+amount of movement data in a variety of fields. This
 rich new data can exhibit simultaneous behavioural processes occurring at different
-time scales. One can model these processes
+time scales, resulting in very complicated dependence structures. One can model these processes
 through a hierarchical hidden Markov model
@@
 nested structure of hidden Markov models.
-At very short time scales, however, observations can exhibit complicated dependence
-structures that cannot be easily captured by
-traditional HMMs. We demonstrate how
+At very short time scales, however, many assumptions of traditional HMMs are violated. We demonstrate how
 to incorporate fine-scale processes into the
 larger structure of HHMMs while maintaining computational efficiency. We apply our
-method to dive and accelerometer data collected from northern resident killer whales off
+method to dive and accelerometer data collected from a northern resident killer whale off
 the coast of British Columbia, Canada.
@@
 % Email address for corresponding author
-\correspondingauthor[*]{\\\email{Insert your email address here only after your paper has been accepted}}
+%\correspondingauthor[*]{\\\email{}} % only put this in after acceptance
 
-%\input Sections/intro.tex
-\input Sections/background.tex
-\input Sections/carhmm_ou.tex
-\input Sections/model.tex
+%%%%%%%%%%%%
+\vfill\eject
+{\centerline{\Large{Notes from Nancy}}}
+\input{Sections/0627NH}
+\vfill\eject
+%%%%%%%%%%%%%
+%
+\section{Introduction}
+\input Sections/intro.tex
+%
+\section{Models and parameter estimation}
+\input Sections/models.tex
+%
+\section{Simulation Study}
 \input Sections/simulation.tex
-%\input Sections/results.tex
-%\input Sections/conclusion.tex
-
-\clearpage
-
-\bibliographystyle{plain}
+%
+\section{Killer Whale Case Study}
+\label{sec:case_study}
+\input Sections/results.tex
+%
+\section{Discussion}
+\input Sections/discussion.tex
+%
+\newpage
 \bibliography{references}
-
-\clearpage
-
+%
+\newpage
+\begin{appendix}
 \input Sections/appendix.tex
+\end{appendix}
+%
+\newpage
+\section{Figures and Tables}
+\input Sections/figures.tex
+%
+%\CJShistory
 

### 2020-07-16 Code/Visualisor.py
Format Plots

@@ -2,6 +2,9 @@ import pandas as pd
 import numpy as np
 import matplotlib.pyplot as plt
 import matplotlib.cm as cm
+from matplotlib.patches import Patch
+from matplotlib.lines import Line2D
+from matplotlib.ticker import FormatStrFormatter
 
 from scipy.stats import gamma
 from scipy.stats import norm
@@ -38,56 +41,58 @@ class Visualisor:
 
     def lagplot(self,lims={},file=None):
 
-        ncols = 2
-        nrows = len(self.pars.features[0]) + len(self.pars.features[1])
-        plt.subplots(nrows,ncols,figsize=(7.5*ncols,7.5*nrows))
+        nrows = 1
+        ncols = len(self.pars.features[0]) + len(self.pars.features[1])
+        fig,ax = plt.subplots(nrows,ncols,figsize=(10*ncols,10*nrows))
         fig_num = 1
 
         features = list(self.pars.features[0].keys()) + list(self.pars.features[1].keys())
+        xlabs = ['Dive Duration $(s)$',
+                 r'$\left(Z^{*(1)}\right)_x$ $(m/s^2)$',
+                 r'$\left(Z^{*(1)}\right)_y$ $(m/s^2)$',
+                 r'$\left(Z^{*(1)}\right)_z$ $(m/s^2)$',
+                 r'$Z^{*(2)}$']
 
         # lag plots of dive-level data
-        for feature in features:
+        for i,feature in enumerate(features):
 
-            # lag plot
-            plt.subplot(nrows,ncols,fig_num)
+            # get data
             if feature in self.pars.features[0]:
                 x = [x0[feature] for x0 in self.data[:-1]]
                 y = [y0[feature] for y0 in self.data[1:]]
-                plt.title('Lag plot of %s, all dives'%feature,fontsize=20)
             else:
                 x = []
                 y = []
                 for dive in self.data:
                     x.extend([x0[feature] for x0 in dive['subdive_features'][:-1]])
                     y.extend([y0[feature] for y0 in dive['subdive_features'][1:]])
-                plt.title('Lag plot of %s, all subdives'%feature,fontsize=20)
-            plt.plot(x,y,'.',markersize=4)
-            if feature in lims:
-                plt.xlim(lims[feature])
-                plt.ylim(lims[feature])
-            plt.xlabel('%s$_{t}$'%feature,fontsize = 16)
[…]

### 2020-07-16 Report/Sections/intro.tex
Updates from Overleaf

--- 2020-07-08 Report/Sections/intro.tex
@@
 The field of animal movement is in the midst of a ``data renaissance" where advancements in tagging technology have given rise to an explosion of data available for statistical modeling.
-In particular, tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
+In particular, some tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
 In response, researchers have introduced a variety of new statistical techniques to infer animal behavior from movement data \citep{Hooten:2017}.
 
-One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}.
-Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another conditioned on the underlying behavioral process.
-However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies.
+One of the most prevalent techniques in recent literature is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}.
+Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another after conditioning on the underlying behavioral process.
+However, this assumption is often violated in real world processes, especially when observations are taken at high frequencies.
 For example, the location of an animal at a given time is highly correlated with the location of that animal one second later.
@@
 Another issue that arises in high-frequency data is that several simultaneous behavioral processes may occur at different time scales.
-In this work we consider the example where killer whales perform a variety of different types of dives at a coarse scale, but also exhibit many different types of swimming behaviors within dives at a fine scale.
+In this work, we consider an example where killer whales exhibit a variety of different types of dives at a coarse scale, but also exhibit many different types of swimming behaviors within dives at a fine scale.
 One solution to this issue is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017} \citep{Adam:2019}.
@@
 
+All HMMs presented so far assume Markovian dynamics in the underlying process (i.e.
+that any observation $Y_t$ depends only on the behavioral state $X_t$ and $Y_{t-1}$ when conditioned on all previous time steps).
 At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs.
 Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}.
-One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation.
+With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine-scale processes are likely to persist.
+
+One solution is to model the fine-scale behavior using a continuous-time model, which involves modelling the dynamics of an animal as the solution to a stochastic differential equation.
 Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals.
-However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
+Moreover, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
 
-For periodic behavior in particular, One way to avoid the use of continuous time models is to use signal processing techniques such as the Fourier transform on the raw data.
+For periodic behavior in particular, one way to avoid the use of continuous time models is to use signal processing techniques such as the Fourier transform on the raw data.
 The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}.
-In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data.
+In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017}, specifically fluking \citep{Shorter:2017} from accelerometer data.
 Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appeared a promising simple approach to account for additional correlation in data that is cyclical in nature.
 
-We consider several models- the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM).
+We consider several models: the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM).
 The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes.

### 2020-07-16 Report/Sections/models.tex
Updates from Overleaf

@@
 In this section we first remind the reader of the definition of a hidden Markov model (HMM).
-We then review three variations on the HMM; the CarHMM, HHMM, and HMM-DFT.
+We then review three variations on the HMM: the CarHMM, HHMM, and HMM-DFT.
 We also briefly describe a continuous-time model and its connection with the CarHMM.
@@
 A \textit{hidden Markov model}  is comprised of a sequence of  unobserved states $X_t$, $t = 1, \ldots, T$, and an associated sequence of  possibly high-dimensional observations $Y_t$, $t = 1, \ldots, T$,
-The $Y_t$s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
-The $X_t$s form a Markov chain and take possible values $1, \ldots, N$.
+The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
+The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$.
 Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$ where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$, for $t=1,\ldots, T-1$, and $i, j = 1,\ldots, N$.
@@
 %
-These conditional distributions are governed by state-dependent parameters: if $X_t = i$, the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes  $f^{(i)}(\cdot)$.
+These conditional distributions are governed by state-dependent parameters.
+If $X_t = i$, then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes  $f^{(i)}(\cdot)$.
 %
@@
 We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
-As an example, we could assume that this conditional distribution is normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
+As an example, we could assume that this conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
 %
@@
 While CarHMMs can model auto-correlation within the observation sequence, they fail when observations are taken at irregular time intervals.
-Another way to model a stochastic process is to use a stochastic differential equation, or SDE.
+Another way to model a stochastic process is to use a stochastic differential equation (SDE).
 As an example, \citep{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
@@
 %
-The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$, and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
+The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
 Let $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$, the fine-scale emission parameter vector corresponding to $X_t=i$.
@@
 Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states.
-For example, in the killer whale case study in section \ref{sec:case_study} we force the fine-scale emission parameters to be shared across coarse-scale hidden states, i.e.
-$\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$.
-(Fig \ref{fig:HHMM}) shows a graphical representation of the dependence structure for an HHMM.
+For example, in the killer whale case study in section \ref{sec:case_study}, we force the fine-scale emission parameters to be shared across coarse-scale hidden states (i.e.
+$\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$).
+Figure \ref{fig:HHMM} shows a graphical representation of the dependence structure for an HHMM.
 
-Due to the nested structure of the hierarchical hidden Markov model, the likelihood remains easy to calculate via the forward algorithm.
+Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easy to calculate via the forward algorithm.
 %
 Let $y$ be the $T$-vector of the observed coarse-scale emissions and
-$y^*$ be the $(T_1^* + \cdots T_T^*)$-vector of the observed fine-scale emissions.
+$y^*$ be the $(T_1^* + \cdots + T_T^*)$-vector of the observed fine-scale emissions.
 %
@@
 %
-Then the likelihood of the observed data is
+The likelihood of the observed data is
 %
@@
 In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform.
-For simplicity of presentation, we assume that the length of the fine-scale processes is constant, i.e.
-that $T^*_t = T^*$, although this need not be the case in general.
-Suppose that the fine-scale process $y^*_t$ does not switch states hidden states, but does exhibit significant periodic behavior.
-Then, we suggest using the discrete Fourier transform (DFT) on $y^*_t$:
+For simplicity, we assume that the length of the fine-scale processes is constant (i.e.
+that $T^*_t = T^*$), although this need not be the case in general.
+Suppose that the fine-scale process $y^*_t$ does not switch hidden states, but does exhibit significant periodic behaviour.
+We then suggest using the discrete Fourier transform (DFT) on $y^*_t$:
 %
@@
 
-Once $z^*_t$ is calculated, it can simply be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$.
+Once $z^*_t$ is calculated, it can be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$.
 In total, the likelihood of the hierarchical HMM-DFT is as follows:

### 2020-07-16 Report/main.tex
Updates from Overleaf

@@
 Recent advances in high-frequency tagging
-technology have made available a vast
-amount of movement data in a variety of fields. This
-rich new data can exhibit simultaneous behavioural processes occurring at different
-time scales, resulting in very complicated dependence structures. One can model these processes
+technology have made a vast
+amount of movement data available in a variety of fields. This
+rich data can exhibit simultaneous behavioural processes occurring at different
+time scales, resulting in very complicated dependence structures. These processes can be modelled 
 through a hierarchical hidden Markov model
 (HHMM), where the system is modelled as a
-nested structure of hidden Markov models.
+nested structure of hidden Markov models (HMMs).
 At very short time scales, however, many assumptions of traditional HMMs are violated. We demonstrate how

### 2020-07-16 Report/Sections/discussion.tex

@@
 A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by \citep{Michelot:2019}.
-Most continuous time models require relatively slow MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.
+Most continuous time models require relatively slow MCMC algorithms to perform inference, and as a result, are not easily incorporated into the HHMM structure.
 However, we prove that certain continuous-time methods are equivalent to an CarHMM under certain conditions (see appendix).
@@
 One way to temper model complexity is to reduce the dimension of the parameter space by forcing fine-scale states to be shared across the coarse-scale states.
-Even still, model complexity inevitably grows rapidly as hierarchical structures are stack on top of each other.
+Even still, model complexity inevitably grows rapidly as hierarchical structures are stacked on top of each other.
 
@@
 
-The simulation study also shows that the observed fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases.
-One notable exception is that the standard errors of the probability transition matrix estimates ($\hat \Gamma$ and $\hat \Gamma^*$) tend to be overestimated by the observed fisher information.
+The simulation study also shows that the observed Fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases.
+One notable exception is that the standard errors of the probability transition matrix estimates ($\hat \Gamma$ and $\hat \Gamma^*$) tend to be overestimated by the observed Fisher information.
 
@@
 
-Because traditional information criteria tend to overestimate the number of states in biological processes (\citep{Pohle:2017}), the number of dive types and sub-dive behaviors was selected in an ad-hoc manor.
+Because traditional information criteria tend to overestimate the number of states in biological processes \citep{Pohle:2017}, the number of dive types and sub-dive behaviors was selected in an ad-hoc manner.
 There does appear to be some heterogeneity within dive types, and future work can be done to determine the optimal number of dive types and within-dive behaviors.

### 2020-07-16 Report/Sections/results.tex

@@
 
-The CarHHMM-DFT was used to analyze dive data from a northern resident killer whale off the coast of British Columbia, Canada.
+The CarHHMM-DFT was used to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada.
 Acceleration data can be a good proxy for energy expenditure \citep{Green:2009}, but studies suggest that the behavioral state must be taken into account when using accelerometer data \citep{Dot:2016}.
 Therefore, understanding both the behavioral state of the killer whale as well as the distribution of accelerometer data within each behavioral state is import to understand the energetic requirements of killer whales.
-This in turn can assist conservation efforts.
+This, in turn, can assist conservation efforts.
 
@@
 The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions.
-Observations were collected at a rate of 50 hertz.
+Observations were collected at a rate of 50 Hz.
 Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time range were ignored.
@@
 After preprocessing the raw data, a total of 267 dives were observed.
-A plot of the raw data for all dives and a collection of five selected dives can be seen in (fig \ref{fig:data}) and (fig \ref{fig:data_one_dive}), respectively.
+A plot of the raw data for all dives and a collection of five selected dives can be seen in Figure \ref{fig:data} and Figure \ref{fig:data_one_dive}, respectively.
 
@@
 $\mathbf{z}_{t,t^*}^{*(1)}$ is a 3-dimensional vector while $z_t^{*(2)}$ is a scalar.
-Summing the first 10 Fourier modes to calculate $z^{(2)}_{t,t^*}$ corresponds to a maximum frequency of $\tilde \omega = 5$ hertz.
-(Fig \ref{fig:lag}) displays a lag plot for all major features and reveals strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions.
-Auto-correlation was therefore directly modeled into the the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$, which is assumed to be normally distributed with the following parameters:
+Summing the first 10 Fourier modes to calculate $z^{(2)}_{t,t^*}$ corresponds to a maximum frequency of $\tilde \omega = 5$ Hz.
+Figure \ref{fig:lag} displays a lag plot for all major features and reveals strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions.
+Auto-correlation was therefore directly modeled into the the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$, which is assumed to be Normally distributed with the following parameters:
 %
@@
 The coarse-scale model is comprised of an HMM with no auto-correlation and no DFT with hidden states corresponding to dive types and observations corresponding to dive durations.
-The fine-scale model is comprised of a CarHMM-DFT where auto-correlation is modeled into the distribution of $\mathbf{Z}^{*(1)}$ (the average acceleration with a 2-second window), but not $Z^{*(2)}$ (the ``wiggliness" of the 2-second window).
-Refer to (fig.
-\ref{fig:CarHHMM}) for a full graphical representation of this model.
+The fine-scale model is comprised of a CarHMM-DFT where auto-correlation is modeled into the distribution of $\mathbf{Z}^{*(1)}$ (the average acceleration with a 2-second window), but not $Z^{*(2)}$ (the ``wiggliness" of the two second window).
+Refer to Figure \ref{fig:CarHHMM} for a full graphical representation of this model.
 
@@
 
-(Tbl.
-\ref{table:emis_dists}) displays estimates of the emission distribution parameters.
-Each fitted distribution is also plotted in (fig \ref{fig:coarse_emis}) and (fig \ref{fig:fine_emis}).
+Table \ref{table:emis_dists} displays estimates of the emission distribution parameters.
+Each fitted distribution is also plotted in Figure \ref{fig:coarse_emis} and Figure \ref{fig:fine_emis}.
 While the ecological interpretation of behavioral states is tenuous for HMMs, we hypothesize the following meanings.
@@
 %
-About 86\% of observed dives are short- the whale usually rests for many dives in a row before performing a deep dive.
+About 86\% of observed dives are short; the whale usually rests for many dives in a row before performing a deep dive.
 The probability transition matrix $\hat \Gamma$ shows that the distribution of a particular dive's type is approximately equal regardless of the previous dive type.
@@
 Two visual tools were used to evaluate this model: pseudo-residuals and empirical histograms.
-A pseudo-residual of a particular observation is the marginal cdf of an observation conditioned on all other observations under the learned model \citep{Zucchini:2016}.
-To easily visualize outliers, the pseudo-residual is often passed through the quantile function of the standard normal distribution.
-Mathematically, the pseudo-residual of an observation $y_t$ is $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard normal distribution.
-If the model is correct, then all pseudo-residuals are independent and follow a standard normal distribution.
+A pseudo-residual of a particular observation is the marginal CDF of an observation conditioned on all other observations under the learned model \citep{Zucchini:2016}.
+To easily visualize outliers, the pseudo-residual is often passed through the quantile function of the standard Normal distribution.
+Mathematically, the pseudo-residual of an observation $y_t$ is $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
+If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution.
 We find that histograms of the pseudoresiduals of this model mostly support that the model is well-specified.
@@
 In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for sub-dive state 3 (fig.
-\ref{fig:empirical_dist}), indicating the existence of rare events corresponding to very violent trashing of the killer whale.
+\ref{fig:empirical_dist}), indicating the existence of rare events corresponding to very violent thrashing of the killer whale.
 These outliers are potential subjects for future study.

### 2020-07-16 Report/Sections/simulation.tex

@@
 
-We base the following simulated study on acceleration data of a killer whale dive sequence.
-The parameters used to generate the data are loosely based on those learned from the case study in section \ref{sec:case_study}.
-We fit four separate models to the simulated data and compare their performance.
+We base the following simulated study on the acceleration data of a killer whale's dive sequence.
+The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}.
+We fit four separate models to the simulated data and compare their performances against one another.
 
@@
 500 separate sequences of 100 killer whale dives were simulated according to an HMM, where the hidden Markov chain $X$ was set as a collection of dive types and the observations $Y$ were the corresponding dive durations (in seconds).
-Each dive could be one of $N=2$ dive types, and the duration of dive $t$, $Y_t$, followed a gamma distribution whose parameters $\theta^{(i)}$ depended upon the dive type $X_t = i$.
+Each dive could be one of $N=2$ dive types, and the duration of dive $t$, $Y_t$, followed a gamma distribution whose parameters $\theta^{(i)}$ were dependent on the dive type $X_t = i$.
 We parameterize the gamma distribution by its mean and variance:
@@
 Once the dive durations were calculated for all 100 dives, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments (the end of the dive sequence was discarded) which made up a second fine-scale hidden Markov model.
-Each 2-second segment of the dive was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$, and the probability transition matrices for these fine-scale Markov chains were set as
+Each two second segment of the dive was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$, and the probability transition matrices for these fine-scale Markov chains were set as
 %
@@
 
-Each two-second sub-dive window had 100 associated acceleration readings, $Y^*_{t,t^*} \equiv \left(Y^*_{t,t^*,1}, \ldots, Y^*_{t,t^*,100}\right)$.
-To accurately recreate active swimming vs passive gliding on the fine-scale Markov chain, the DFT of each 2-second segment $\hat Y^*_{t,t^*}$ was simulated such that $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$ (see eqn \ref{eqn:z}) would have the following distributions:
+Each two second sub-dive window had 100 associated acceleration readings, $Y^*_{t,t^*} \equiv \left(Y^*_{t,t^*,1}, \ldots, Y^*_{t,t^*,100}\right)$.
+To accurately recreate active swimming versus passive gliding on the fine-scale Markov chain, the DFT of each two second segment $\hat Y^*_{t,t^*}$ was simulated such that $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$ (see Eqn \ref{eqn:z}) would have the following distributions:
 
@@
 See the appendix for more details regarding procedure for simulating $\hat Y^*$ and $Y^*$ such that $Z^*$ has the preceding distribution.
-(Fig.
-\ref{fig:sim_data}) shows the first 5 dives of one simulated data set.
+Figure \ref{fig:sim_data} shows the first 5 dives of one simulated data set.
 
@@
 Each two-second window is summarized by the observations $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$.
-The distribution of $Z^{*(1)}_{t,t^*}$ is normal and its parameters depend upon the sub-dive behavior $X^*_{t,t^*}$ and $Z^{*(1)}_{t,t^*-1}$.
+The distribution of $Z^{*(1)}_{t,t^*}$ is Normal and its parameters depend upon the sub-dive behavior $X^*_{t,t^*}$ and $Z^{*(1)}_{t,t^*-1}$.
 In particular:
@@
 The likelihood of this model is still easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
-See the appendix for details of likelihood evaluation, and (fig.
-\ref{fig:CarHHMM}) shows the corresponding graphical model.
+See the appendix for details of likelihood evaluation, and Figure \ref{fig:CarHHMM} shows the corresponding graphical model.
 
@@
 The other three models all achieved an accuracy of approximately 90\%, with the CarHHMM slightly more likely to categorize a dive as dive type 2 than the other models.
-(Fig.
-\ref{fig:acc}) shows the decoded state probabilities for both the fine- and coarse- scales.
+Figure \ref{fig:acc} shows the decoded state probabilities for both the fine- and coarse- scales.
 (Tbl.
@@
 %%% dive duration %%%
-For the emission distributions of dive duration, estimates of standard error using the observed fisher information tended to be underestimates due to correlation between $\hat \mu$ and $\hat \sigma$.
+For the emission distributions of dive duration, estimates of standard error using the observed Fisher information tended to be underestimates due to correlation between $\hat \mu$ and $\hat \sigma$.
 In addition, $\hat \sigma$ tends to underestimate $\sigma$ for all models, which is a finding consistent with properties of MLEs, especially because the sample size for a particular dive type is approximately 50 for each simulation.
@@
 %This is likely due to the fact that dive duration follows a mixture of gammas rather than a gamma distribution, and the CarHMM does not have the machinery to capture this fact.
-A full table of parameter estimates for all models is shown in (table \ref{table:dive_duration}).
+A full table of parameter estimates for all models is shown in Table \ref{table:dive_duration}.
 
@@
 For acceleration ($Z^{*(1)}$), both the CarHMM-DFT and the CarHHMM-DFT regularly converged to the correct parameters with very little standard error.
-However, the fisher standard error regularly overestimated the standard error for $\hat \mu$ for both of these models.
+However, the Fisher standard error regularly overestimated the standard error for $\hat \mu$ for both of these models.
 %This likely is due to the very large auto-correlation terms ($\phi^{(1)} = 0.99$ and $\phi^{(2)} = 0.95$) since as $\phi \to 1$, the value of $\mu$ does not matter in the likelihood evaluation as $\phi \to 1$ implies that $\mathbb{E}(Z^{*(1)}_{t,s^*}) = \phi*(Z^{*(1)}_{t,s^*-1}) + (1-\phi)\mu \to Z^{*(1)}_{t,s^*-1}$.
@@
 %because it sees the ``wigliness" of the acceleration data as raw acceleration rather than as Fourier modes, which deflates the value of $\hat \phi$ and inflates the value of $\hat \sigma$. 
-See (tbl. \ref{table:acceleration}) for a detailed breakdown of parameter values for acceleration emission distributions.
+See Table \ref{table:acceleration} for a detailed breakdown of parameter values for acceleration emission distributions.
 
@@
 For $\Gamma^*$, the HHMM-DFT and CarHHMM-DFT both showed practically no bias with standard errors on the order of $10^{-2}$.
-One notable exception is the standard error of $\hat \Gamma^{*(1)}_{12}$, whose standard error was on the order of $10^{-4}$, which is much lower than the observed fisher information predicted.
+One notable exception is the standard error of $\hat \Gamma^{*(1)}_{12}$, whose standard error was on the order of $10^{-4}$, which is much lower than the observed Fisher information predicted.
 The CarHMM-DFT is mis-specified, so its results cannot be easily interpreted.

### 2020-07-16 Code/Visualisor.py
Add Legend to plot

@@ -160,6 +160,7 @@ class Visualisor:
                         title = 'Emission Distributions, Dive Duration'
                         ax[row_num,col_num].set_xlabel('Dive Duration (seconds)')
                         ax[row_num,col_num].set_title(title,fontsize=24)
+                        plt.legend(['Dive Type 1','Dive Type 2'])
                     else:
                         titles = [r'$\left(Z^{*(1)}\right)_x$ $(m/s^2)$',
                                   r'$\left(Z^{*(1)}\right)_y$ $(m/s^2)$',

### 2020-07-16 Code/Visualisor.py
Update lagplot

@@ -5,6 +5,7 @@ import matplotlib.cm as cm
 from matplotlib.patches import Patch
 from matplotlib.lines import Line2D
 from matplotlib.ticker import FormatStrFormatter
+from mpl_toolkits.axes_grid1 import make_axes_locatable
 
 from scipy.stats import gamma
 from scipy.stats import norm
@@ -41,8 +42,8 @@ class Visualisor:
 
     def lagplot(self,lims={},file=None):
 
-        nrows = 1
-        ncols = len(self.pars.features[0]) + len(self.pars.features[1])
+        ncols = 1
+        nrows = len(self.pars.features[0]) + len(self.pars.features[1])
         fig,ax = plt.subplots(nrows,ncols,figsize=(10*ncols,10*nrows))
         fig_num = 1
 
@@ -85,18 +86,21 @@ class Visualisor:
             im = plt.imshow(np.rot90(Ztemp),extent = xlim + ylim)
             plt.title(xlabs[i],fontsize=36)
             plt.gca().ticklabel_format(style='sci',scilimits = (-3,3))
-            c = fig.colorbar(im, orientation='horizontal',pad=0.125)
+            divider = make_axes_locatable(plt.gca())
+            cax = divider.append_axes("right", size="5%", pad=0.05)
+            c = fig.colorbar(im,cax=cax)#im, orientation='horizontal',pad=0.125)
             c.ax.ticklabel_format(style='sci',scilimits = (-3,3))
             c.set_label('Density',fontsize = 30)
 
             fig_num += 1
 
-        plt.subplots_adjust(wspace=0.05, hspace=0.05)
-        fig.text(0.5, 1.0, 'Lag Plots', ha='center', fontsize=50)
+        plt.subplots_adjust(wspace=-0.6, hspace=-0.6)
+        #fig.text(0.5, 1.0, 'Lag Plots', ha='center', fontsize=50)
 
         if file is None:
             plt.show()
         else:
+            plt.tight_layout()
             plt.savefig(file)
 
         return

### 2020-07-16 Code/Visualisor.py
Update Lagplot

@@ -42,9 +42,10 @@ class Visualisor:
 
     def lagplot(self,lims={},file=None):
 
-        ncols = 1
-        nrows = len(self.pars.features[0]) + len(self.pars.features[1])
+        ncols = 2
+        nrows = int(np.ceil(len(self.pars.features[0]) + len(self.pars.features[1])/2))
         fig,ax = plt.subplots(nrows,ncols,figsize=(10*ncols,10*nrows))
+        fig.delaxes(ax[2,1])
         fig_num = 1
 
         features = list(self.pars.features[0].keys()) + list(self.pars.features[1].keys())

### 2020-07-16 Code/Visualisor.py
Update Fine Emissions

@@ -119,81 +119,83 @@ class Visualisor:
             ncols = len(self.hhmm.theta[0])
         else:
             features = ['Ax','Ay','Az','Ahat_low'] #list(self.hhmm.theta[1][0].keys())
-            nrows = 1 #self.pars.K[0]
-            ncols = len(features)
+            nrows = 2 #self.pars.K[0]
+            ncols = int(len(features)/2)
 
         fig, ax = plt.subplots(nrows,ncols,figsize=(7.5*ncols,7.5*nrows))
         ax = np.reshape(ax,(nrows,ncols))
 
-        for col_num,feature in enumerate(features):
-            for row_num in range(nrows):
+        for feature_num,feature in enumerate(features):
 
-                if level == 0:
-                    mu = self.hhmm.theta[0][feature]['mu']
-                    sig = self.hhmm.theta[0][feature]['sig']
+            row_num = int(feature_num/2)
+            col_num = feature_num%2
 
-                    dist = self.pars.features[0][feature]['f']
-                    K = self.pars.K[0]
-                    colors = [cm.get_cmap('tab10')(i) for i in [0,1]]
-                    legend = ['Dive Type %d'%(x+1) for x in range(K)]
+            if level == 0:
+                mu = self.hhmm.theta[0][feature]['mu']
+                sig = self.hhmm.theta[0][feature]['sig']
+
+                dist = self.pars.features[0][feature]['f']
+                K = self.pars.K[0]
+                colors = [cm.get_cmap('tab10')(i) for i in [0,1]]
+                legend = ['Dive Type %d'%(x+1) for x in range(K)]
+            else:
+                mu = self.hhmm.theta[1][row_num][feature]['mu']
+                sig = self.hhmm.theta[1][row_num][feature]['sig']
+
+                dist = self.pars.features[1][feature]['f']
+                K = self.pars.K[1]
+                colors = [cm.get_cmap('viridis')(i) for i in [0.,0.5,1.]]
+                legend = ['Subdive Behavior %d'%(x+1) for x in range(K)]
+
+            for state in range(K):
+                if dist == 'gamma':
+                    shape = np.square(mu)/np.square(sig)
+                    scale = np.square(sig)/np.array(mu)
+                    x = np.linspace(0.01,max(mu)+5*max(sig),100000)
+                    y = gamma.pdf(x,shape[state],0,scale[state])
+                elif dist == 'normal':
[…]

### 2020-07-16 Code/Visualisor.py
adjust plots

@@ -201,7 +201,7 @@ class Visualisor:
             plt.show()
         else:
             plt.tight_layout()
-            plt.savefig(file)
+            plt.savefig(file,bbox_inches='tight')
 
         return
 

### 2020-07-17 Code/HHMM.py
Update figures

@@ -485,24 +485,27 @@ class HHMM:
                 x0 = np.delete(x0,i)
 
                 # optimize
-                res = minimize(loss_fn, x0, method='BFGS',options=options)
+                if len(x0) > 0:
+                    res = minimize(loss_fn, x0, method='BFGS',options=options)
 
-                # update final values
-                x = np.copy(res['x'])
-                if self.likelihood(data) >= backup_l:
-                    for j,xj in enumerate(x):
-                        if j < i:
-                            self.eta[0][i,j] = xj
-                        else:
-                            self.eta[0][i,j+1] = xj
+                    # update final values
+                    x = np.copy(res['x'])
+                    if self.likelihood(data) >= backup_l:
+                        for j,xj in enumerate(x):
+                            if j < i:
+                                self.eta[0][i,j] = xj
+                            else:
+                                self.eta[0][i,j+1] = xj
+                    else:
+                        self.eta[0][i,:] = backup
+                        print('DANGER')
+
+                    print('updated coarse eta, row %d' % i)
+                    print('original: ', backup)
+                    print('new: ', x)
+                    print(self.likelihood(data))
                 else:
-                    self.eta[0][i,:] = backup
-                    print('DANGER')
-
-                print('updated coarse eta, row %d' % i)
-                print('original: ', backup)
-                print('new: ', x)
-                print(self.likelihood(data))
+                    print('N = 1, no Coarse Gamma')
 
 
             #### then do crude theta ###
@@ -811,28 +814,34 @@ class HHMM:
                     # get plus value
                     if param == 'corr':
                         self.theta[1][0][feature][param][state_num] = corr_2_rho(theta+h0)
-                        self.theta[1][1][feature][param][state_num] = self.theta[1][0][feature][param][state_num]
+                        for i in range(1,self.pars.K[0]):
+                            self.theta[1][i][feature][param][state_num] = self.theta[1][0][feature][param][state_num]
                     else:
                         self.theta[1][0][feature][param][state_num] += h
-                        self.theta[1][1][feature][param][state_num] = self.theta[1][0][feature][param][state_num]
+                        for i in range(1,self.pars.K[0]):
+                            self.theta[1][i][feature][param][state_num] = self.theta[1][0][feature][param][state_num]
                     th_tp1 = self.likelihood(data)
 
                     # get minus value
                     if param == 'corr':
                         self.theta[1][0][feature][param][state_num] = corr_2_rho(theta-h0)
[…]

### 2020-07-17 Code/Run_Sim.py

@@ -490,7 +490,7 @@ for dataset_num in range(ndatasets):
         hhmm_FV.true_theta = deepcopy(hhmm_FV_theta)
         hhmm_FV.true_eta = deepcopy(hhmm_FV_eta)
 
-        hhmm_FV.train_DM(data_FV)
+        hhmm_FV.train_DM(data_FV,max_iters=5)
         hhmm_FV.get_SEs(data_FV,h)
         data,data_V,data_FV = create_data()
 

### 2020-07-17 Report/Sections/appendix.tex
Small Edits
Update Report

@@
 \begin{align*}
-\left(Y_{s \Delta t} | X_{(s-1)\Delta t} = i^* \right) \sim \mathcal{N}\left((1-\phi^{(i^*)}) \mu^{(i^*)} + \phi^{(i^*)} Y_{(s-1) \Delta t}, \enspace \sigma^{2(i^*)} \right),\\
+\left(Y_{s \Delta t} | X_{(s-1)\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{(s-1) \Delta t}, \enspace \left(\sigma^{(i)}\right)^2 \right),\\
 s = 1, \ldots, T-1
@@
 \begin{align*}
-\left(Y'_s| X'_s = i^* \right) \sim \mathcal{N}\left((1-\phi^{(i^*)}) \mu^{(i^*)} + \phi^{(i^*)} Y'_{s-1}, \enspace \sigma^{2(i^*)} \right)\\
+\left(Y'_s| X'_s = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y'_{s-1}, \enspace \left(\sigma^{(i)}\right)^2 \right)\\
 s = 2, \ldots, T
@@
 $\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued.
-$\hat{Y}^{*(k)}_{t,t^*}$ also decays like $1/k$ so that $Y^*_{t,t^*}$ remains continuous within a 2-second window.
+$\hat{Y}^{*(k)}_{t,t^*}$ also decays like $1/k$ so that $Y^*_{t,t^*}$ remains continuous within a two-second window.
 $Y^*_{t,t^*}$ is not continuous \textit{between} windows, but the jump discontinuities are not very severe since $\hat{Y}^{*(0)}_{t,t^*}$ and $\hat{Y}^{*(0)}_{t,t^*+1}$ are highly correlated.
-See (fig.
-\ref{fig:sim_data}) for details.
+See Figure \ref{fig:sim_data} for details.
 
@@
 We pick $\tilde{\omega} = 10$ periods per window, or 5 hertz.
-To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use (eqn \ref{eqn:y_hat0}):
+To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use Equation (\ref{eqn:y_hat0}):
 %
@@
 %
-$Z^{*(2)}_{t,t^*}$ is the sum of Gamma-distributed random variables with the same scale parameter, so the distribution of $Z^{*(2)}_{t,t^*}$ is also a Gamma distribution:
+$Z^{*(2)}_{t,t^*}$ is the sum of gamma-distributed random variables with the same scale parameter, so the distribution of $Z^{*(2)}_{t,t^*}$ is also a gamma distribution:
 %
@@
 %
-where $P(z^*_{t,t^*}|z^*_{t,t^*-1};\Theta^{*(i)})$ is an $N^* \times N^*$ diagonal matrix with $i^*i^*$th entry equal to $f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$.
+where $P(z^*_{t,t^*}|z^*_{t,t^*-1};\Theta^{*(i)})$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$.
 %
-$f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$ is the emission distribution of $Z^*_{t,t^*}$ conditioned on the fact that $X^*_{t,t^*} = i^*$ and $Z^*_{t,t^*-1}$.
+$f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$ is the emission distribution of $Z^*_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $Z^*_{t,t^*-1} = z^*_{t,t^*-1}$.
 
 \end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2020-07-17 Report/Sections/discussion.tex
Update Report

@@
 
-We presented a collection of HMM models which can be combined together in increasingly complex hierarchical models to match the complexity of particular problems faced by practitioners.
+We presented a collection of HMM models which can be combined together to form increasingly complex hierarchical models to match the complexity of particular problems faced by researchers.
 This flexible framework can be used to deal with complicated dependence structures within time-series data.
 
-Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics within the hidden state.
+Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics when conditioned of the hidden state.
 However, many real-world processes are more complicated than this and require more complex models.
 
-The CarHMM generalizes the HMM by explicitly modeling auto-correlation in the emission distributions of the HMM while maintaining the structure needed to evaluate the likelihood using the forward algorithm.
-In our normal model formulation, we have added only one additional parameter, $\phi^{(i)}$, per possible hidden state \citep{Lawler:2019}.
+The CarHMM generalizes the HMM by explicitly modeling auto-correlation in the emission distributions of the HMM \citep{Lawler:2019}.
+CarHMMs also maintain the structure needed to evaluate the likelihood using the forward algorithm.
+In our Normal model formulation, we have added only one additional parameter, $\phi^{(i)}$, per possible hidden state.
 Several useful model selection tools such as the lag plot can test if there is significant auto-correlation within an observation sequence.
@@
 Although the CarHMM can incorporate auto-correlation into the structure of an HMM, it can break down when observations are taken at irregular time intervals.
-A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by \citep{Michelot:2019}.
-Most continuous time models require relatively slow MCMC algorithms to perform inference, and as a result, are not easily incorporated into the HHMM structure.
-However, we prove that certain continuous-time methods are equivalent to an CarHMM under certain conditions (see appendix).
+A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by Michelot \& Blackwell (2019).
+Most continuous time models require relatively slow MCMC algorithms to perform inference, and as a result are not easily incorporated into the HHMM structure.
+However, we show in Theorem 1 that certain continuous-time methods are equivalent to an CarHMM under certain conditions (see appendix for proof).
 
-For simultaneous observed processes taking place at different time scales, the HHMM as described by \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs.
-In particular, each hidden state of a coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.
+For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs.
+In particular, each hidden state of the coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.
 
-Finally, for processes with very high sampling frequencies and/or with intricate fine-scale structure, the HHMM structure developed by \citep{Barajas:2017,Adam:2019} can be generalized so the fine-scale model can be any model which admits an easy-to-calculate likelihood.
+For processes with very high sampling frequencies and/or with intricate fine-scale structure, it is possible to generalize the HHMM such that the fine-scale model can be any model which admits an easy-to-calculate likelihood.
 For example, if the sampling rate of the fine-scale process is very high, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.
-In addition, researchers can recursively stack hierarchical HMM models together like building blocks to create increasingly complex models.
-This should be done with care, as it is important to balance the need to effectively capture the process in question with the need to avoiding over-fitting and slow parameter estimation.
+
+Combining these models together should be done with care, as it is important to balance the need to effectively capture the process in question with the need to avoiding over-fitting and slow parameter estimation.
 
@@
 
-An example of balancing model complexity with fast model fitting is presented in the simulation study.
-While (fine-scale) sub-dive behavioral states were shared across (coarse-scale) dive types to reduce model complexity for the hierarchical models, by far the fastest model to train (\~ 15 minutes) was the CarHMM-DFT.
-The CarHMM-DFT had no hierarchical component but near-perfect accuracy when decoding sub-dive behavioral states of the simulated whale.
-This model would be preferred if simultaneous modeling of dive types was not required.
+An example of balancing model complexity with efficient fitting is presented in our simulation study.
+Sub-dive behavioral states were shared across dive types to reduce model complexity, and by far the fastest model to train ($\approx$ 15 minutes) was the CarHMM-DFT, which had no hierarchical component.
+Even still, the CarHMM-DFT had near-perfect accuracy when decoding sub-dive behavioral states of the simulated whale.
 However, this model is not sufficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.
@@
 Finally, we used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada.
-The CarHHMM-DFT was able to distinguish three distinct sub-dive behaviors and two dive types simultaneously due to its hierarchical structure.
-The DFT component proved useful in determining the sub-dive behaviors of the whale, as the mean of the emission distribution of $Z^{*(2)}$ was separated by an order of magnitude between each sub-dive state.
-Finally, the learned auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and sub-dive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data.
+The CarHHMM-DFT was able to simultaneously distinguish three distinct sub-dive behaviors and two dive types.
+The DFT component proved useful in determining the sub-dive behaviour of the whale, as the mean of the emission distribution of $Z^{*(2)}$ for each sub-dive state was separated by an order of magnitude.
+Finally, the estimated auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and sub-dive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data.
 The introduction of the parameter $\phi^*$ also allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).

### 2020-07-17 Report/Sections/models.tex
Update Report

@@
 
-In this section we first remind the reader of the definition of a hidden Markov model (HMM).
-We then review three variations on the HMM: the CarHMM, HHMM, and HMM-DFT.
-We also briefly describe a continuous-time model and its connection with the CarHMM.
-Finally, we generalize the hierarchical structure of these models and describe how they can be flexibly altered and combined to form new models.
+Traditional hidden Markov models (HMMs) are useful tools to \textit{blah blah blah - quick (5-7words) restatement of the main advantage of HMM}, and will be used as the core structure of our models.
+However, traditional HMMs assume conditional independence between observations given the state sequence, and thus do not hold when the observations exhibit certain forms of significant correlation in time.
+Traditional HMMs model process autocorrelation with a single Markov chain, however many processes have temporal dependencies acting at different scales.
+\textit{blah blah blah -quick restatements (5-15 words to potentially many sentences) of the main issues you are going to fix made a quick try at it, work on smoothing things.
+You can use some of the text I removed from below}.
+To account for each of these common additional dependence structures, we will explore three variations on the traditional HMM: the conditionally auto-regressive hidden Markov model (CarHMM), hierarchical hidden Markov model (HHMM), and HMM-DFT.
+We will show how each of these variations can be altered and combined to form a wide variety of new models.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The hidden Markov model (HMM)}
+\subsection{The base structure of HMMs}
 
-A \textit{hidden Markov model}  is comprised of a sequence of  unobserved states $X_t$, $t = 1, \ldots, T$, and an associated sequence of  possibly high-dimensional observations $Y_t$, $t = 1, \ldots, T$,
+An HMM is comprised of a sequence of unobserved states $X_t$, $t = 1, \ldots, T$, and an associated sequence of  possibly high-dimensional observations $Y_t$, $t = 1, \ldots, T$.
 The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
 The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$.
-Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$ where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$, for $t=1,\ldots, T-1$, and $i, j = 1,\ldots, N$.
+Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$, for $t=1,\ldots, T-1$, and $i, j = 1,\ldots, N$.
 %
-We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i$th component
+We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i^{th}$ component
 $\delta_i = \Pr\{X_1 = i\},~ i = 1,\ldots,N.$
-%
-Recall that a Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
+A Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
 %
@@
 %
-(Fig \ref{fig:HMM}) represents the dependence structure of an HMM.
+Figure \ref{fig:HMM} represents the dependence structure of an HMM.
 
-To find the maximum likelihood estimates of the parameters $\Gamma$ and $\Theta \equiv (\theta^{(1)},\ldots,\theta^{(N)})$, let $y = (y_1,\ldots,y_T)$ be the observed emissions. 
-We write the likelihood $\calL_{\text{HMM}}$ in a specific form, using the  well-known \textit{forward algorithm} \citep{Zucchini:2016} as follows:
+Using observed emissions, here denoted $y = (y_1,\ldots,y_T)$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\Theta \equiv (\theta^{(1)},\ldots,\theta^{(N)})$.
+We write the likelihood $\calL_{\text{HMM}}$ using the  well-known \textit{forward algorithm} \citep{Zucchini:2016}:
 %
@@
 %
-$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry  $f^{(i)}(y_t; \theta^{(i)})$.
+$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry  $f^{(i)}(y_t; \theta^{(i)})$.
 %
 
-We will always parameterize transition probability matrices to remove the constraints that the entries of the matrix are non-negative and that the rows sum to 1.
-We do this by parameterizing the $N \times N$ transition probability matrix $\Gamma$ using $\eta \in \bbR^{N \times N}$ \citep{Barajas:2017}:
+Following \citet{Barajas:2017}, we parameterize the $N \times N$ transition probability matrix $\Gamma$ in a way that forces the entries of the matrix to be non-negative and for the rows to sum to 1:
 %
@@
 %
-setting $\eta_{ii}$  to zero, $i=1,\ldots, N$, for identifiability.
-Then $\calL_{\text{HMM}}(y;\Theta,\Gamma)$ can be maximized using any numerical optimizer.
+where $\eta \in \bbR^{N \times N}$ and $\eta_{ii}$ is set to zero for identifiability.
+Then $\calL_{\text{HMM}}(y;\Theta,\Gamma)$ can be maximized using a wide range of optimizer.
 For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of  $\eta$.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The conditionally auto-regressive hidden Markov model (CarHMM)}
+\subsection{Relaxing the conditional independence assumption with a CarHMM}
 
-A key assumption of an HMM is \textit{conditional independence} between observations given the state sequence.
-Therefore, traditional HMMs do not hold when the observations exhibit certain forms of significant correlation in time.
-
-The CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by \citep{Lawler:2019}, explicitly models auto-correlation within an HMM.
-Like a traditional HMM, A CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
-Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$, conditional on $X_1,\ldots, X_T$ and $ \{Y_1,\ldots, Y_{t-1}\}$, depends on \textit{both} $X_t$ \textit{and} $Y_{t-1}$.
-The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. (Fig \ref{fig:CarHMM}) shows the structure of a CarHMM.
+The CarHMM, introduced by Lawer et al.
+(2019), explicitly models auto-correlation in observation within an HMM.
+Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
+Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$, depends on \textit{both} $X_t$ \textit{and} $Y_{t-1}$ rather than only $X_t$.
+The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. Figure \ref{fig:CarHMM} shows the dependence structure of a CarHMM.
 
@@
 We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
-As an example, we could assume that this conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
+For example, one could assume that this conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
 %
@@
 %
-$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
-%
-\subsection{Continuous-time processes}
+$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
 
@@
 Another way to model a stochastic process is to use a stochastic differential equation (SDE).
-As an example, \citep{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
+As an example, Michelot \& Blackwell (2019) model the movement of an animal as the solution to the following stochastic differential equation, which they refer to as a state-switching Ornstein-Uhlenbeck process:
 %
@@
 %
-where $X_t$ follows some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
-This SDE is referred to as a state-switching Ornstein-Uhlenbeck process.
-Unlike HMMs, the time index $t \in \mathbb{R}$ exists in continuous time and is not necessarily an integer.
-If the behavioral state $X_t$ is known and does not change between observations, the solution to (eqn \ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
+where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
+Unlike HMMs, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer.
+If the behavioral state $X_t$ is known and does not change between observations, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
 %
@@
 \begin{theorem}{Theorem 1.}{}%
-If:
+If the following conditions are met:
 \begin{enumerate}
-    \item The hidden behavioural process $X$ from (eq.
-\ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$, and:
-    \item Observations of the SDE from (eq. \ref{eqn:SDE}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
+    \item The hidden behavioural process $X$ from Equation \ref{eqn:SDE} follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$,
+
+    \item Observations of the SDE from Equation 2 are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
 \end{enumerate}
@@
 
-\subsection{The hierarchical hidden Markov model (HHMM)}
+\subsection{Incorporating different scales of correlations with HHMM}
 
-A \textit{hierarchical hidden Markov model} contains both a coarse-scale process and a fine-scale process, both of which are HMMs.
-The coarse-scale process is a hidden Markov model as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observed responses.
+HHMM accounts for different levels of correlation by modeling a coarse-scale process and a fine-scale process, each of which with an HMM \textit{CITATION(s)}.
+The coarse-scale process is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observed responses.
 %
-In the hierarchical setting, each state $X_t$ emits another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
-This fine-scale process can be either instead of or in addition to the coarse-scale observations $Y_t$.
-The length of each fine-scale process is denoted as $T^*_t$ and can often be controlled by the practitioner.
-Conditional on the coarse-scale hidden state $X_t$, the joint fine-scale process, ($X_t^*, Y_t^*$), is a hidden Markov model with parameters depending on the value of $X_t$.
-Specifically, given that $X_t=i$, $X_t^*$ is a Markov chain with states $1,\ldots, N_t^*$.
-The distribution of $X_t^*$, conditional on $X_t=i$, is given by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial probability, denoted by the $N_t^*$-vector $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
+In the hierarchical setting, each state $X_t$ emits yet another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
+The fine-scale process ($X_t^*, Y_t^*$) makes up another HMM with parameters that depend on the value of $X_t$.
+Specifically, if $X_t=i$, then the distribution of $X_t^*$ is characterized by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
 For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary.
-%
-The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
-Let $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$, the fine-scale emission parameter vector corresponding to $X_t=i$.
-%
+
+The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.We denote the fine-scale emission parameter vector corresponding to $X_t=i$ with $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$.
+
 Given the coarse-scale states, $X_1,\ldots, X_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs.
 Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states.
-For example, in the killer whale case study in section \ref{sec:case_study}, we force the fine-scale emission parameters to be shared across coarse-scale hidden states (i.e.
+For example, in the killer whale case study (Section \ref{sec:case_study}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states (i.e.
 $\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$).
-Figure \ref{fig:HHMM} shows a graphical representation of the dependence structure for an HHMM.
+Figure \ref{fig:HHMM} represents the dependence structure for an HHMM.
 
@@
 %
-Let $\Theta^*$ denote the collection of all fine-scale emission parameters,
-$\Theta^{*(i)}$, $i=1,\ldots, N$, and let $\Gamma^*$ denote the collection of all fine-scale transition probability matrices, $\Gamma^{*(i)}$, $i=1,\ldots, N$.
-%
-The likelihood of the observed data is
+Let $\Theta^* \equiv \{\Theta^{*(1)}, \ldots, \Theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and let $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices.
+The likelihood of the observed data is then
 %
@@
 
-For more information on specific considerations for HHMMs such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citep{Adam:2019}.
+For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see Adam et al.
+(2019).
 
@@
 \begin{align*}
-    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right) \qquad k = 0, 1, \ldots, T^*-1
+    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right), \quad k = 0, 1, \ldots, T^*-1.
 \end{align*}
 %
-Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One such example is as follows:
+Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One example is as follows:
 %
@@
 %
-$z_t^{*(1)}$ is the average value of $y^*_t$ and $z_t^{*(2)}$ is the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$.
+In words, $z_t^{*(1)}$ is the average value of $y^*_t$ and $z_t^{*(2)}$ is the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$.
 The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care.
 These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$.
-Figure (\ref{fig:fourier_example}) visually displays the process of transforming $y^*$ into $z^*$.
+Figure \ref{fig:HMM-DFT} shows a graphical representation of the HMM-DFT.
 
@@
 \end{equation}
-where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $ii$th entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.
+where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.
 
-It is possible to accommodate for unequal time steps within $y_t^*$ by using the non-uniform discrete Fourier transform (NDFT).
-We do not describe this method here, but the generalization is straightforward.
-Refer to \citep{Bagchi:1999} for details.
+It is possible to accommodate unequal time steps within $y_t^*$ by using the non-uniform discrete Fourier transform (NDFT) \citep{Bagchi:1999}.
+We do not describe the method in detail here, but the generalization is straightforward.
 
@@
 
-In addition to the DFT and HMM models described above, the fine-scale process $Y^*_t$ can be modeled with \textit{any} parametric model which admits an easy-to-compute likelihood.
-The state-specific emission distribution $\calL_{\text{HMM}}$ from the HHMM likelihood is then replaced by the likelihood of the fine-scale model, $\calL_{\text{fine}}(\mathbf{y}^*_t;\Theta^{*(i)})$:
+In addition to the models described above, the fine-scale process $Y^*_t$ can be modeled using \textit{any} parametric model which admits an easy-to-compute likelihood.
+The fine-scale likelihood $\calL_{\text{HMM}}$ from the HHMM likelihood is then replaced by the likelihood of the general fine-scale model, $\calL_{\text{fine}}(\mathbf{y}^*_t;\Theta^{*(i)})$:
 \[
@@
 \]
-where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
+where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
 This definition is straightforward to extend to the CarHMM as well:
@@
 \]
-where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
+where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
 
 Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, state-switching OU, HHMM, and HMM-DFT).
-These HMM models can act as building blocks which can be used to construct increasingly complex hierarchical models based on the HMMs.
+These five models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models based on HMMs.
 In the sections that follow, we perform both a simulation study and real-world case study modelling killer whale dive behaviour using models constructed from these building blocks.

### 2020-07-17 Report/Sections/simulation.tex
Update Report

@@
 The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}.
-We fit four separate models to the simulated data and compare their performances against one another.
+We fit four separate models to the simulated data and compare their test the accuracy of each on a simulated test data set.
 
@@
 
-500 separate sequences of 100 killer whale dives were simulated according to an HMM, where the hidden Markov chain $X$ was set as a collection of dive types and the observations $Y$ were the corresponding dive durations (in seconds).
-Each dive could be one of $N=2$ dive types, and the duration of dive $t$, $Y_t$, followed a gamma distribution whose parameters $\theta^{(i)}$ were dependent on the dive type $X_t = i$.
+500 separate sequences of 100 killer whale dives were simulated according to an HMM, where the hidden Markov chain $X$ was a collection of dive types and the observations $Y$ were the corresponding dive durations (in seconds).
+Each dive could be one of $N=2$ dive types, and the duration of the $t^{th}$ dive, $Y_t$, followed a gamma distribution whose parameters $\theta^{(i)}$ were dependent on the dive type $X_t = i$.
 We parameterize the gamma distribution by its mean and variance:
 %
-$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}, \qquad \delta =  \begin{pmatrix} 0.5 & 0.5 \end{pmatrix}$$
-$$Y_t|X_t \sim \rm{Gamma} $$
+$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}, \qquad \delta =  \begin{pmatrix} 0.5 & 0.5 \end{pmatrix},$$
+$$Y_t|X_t \sim \rm{Gamma},$$
 \begin{align*}
-	&\bbE(Y_t|X_t = 1) = 15 s &\bbE(Y_t|X_t = 2) = 60 s \\
+	&\bbE(Y_t|X_t = 1) = 15 s, &\bbE(Y_t|X_t = 2) = 60 s, \\
 	%
-	&\bbV(Y_t|X_t = 1) = 25 s^2 &\bbV(Y_t|X_t = 2) = 100 s^2
+	&\bbV(Y_t|X_t = 1) = 25 s^2, &\bbV(Y_t|X_t = 2) = 100 s^2.
 \end{align*}
@@
 Once the dive durations were calculated for all 100 dives, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments (the end of the dive sequence was discarded) which made up a second fine-scale hidden Markov model.
-Each two second segment of the dive was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$, and the probability transition matrices for these fine-scale Markov chains were set as
+Each two second segment was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$.
+The probability transition matrices for these fine-scale Markov chains were set as
 %
@@
 
-Each two second sub-dive window had 100 associated acceleration readings, $Y^*_{t,t^*} \equiv \left(Y^*_{t,t^*,1}, \ldots, Y^*_{t,t^*,100}\right)$.
-To accurately recreate active swimming versus passive gliding on the fine-scale Markov chain, the DFT of each two second segment $\hat Y^*_{t,t^*}$ was simulated such that $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$ (see Eqn \ref{eqn:z}) would have the following distributions:
+Each two-second sub-dive window had 100 associated acceleration readings, $Y^*_{t,t^*} \equiv \left(Y^*_{t,t^*,1}, \ldots, Y^*_{t,t^*,100}\right)$.
+To accurately recreate active swimming versus passive gliding on the fine-scale Markov chain, the DFT of each two-second segment $\hat Y^*_{t,t^*}$ was simulated such that $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$ (see Eq.
+\ref{eqn:z}) have the following distributions:
 
@@
 Specifically, we used a hierarchical HMM where the sequence of dive durations $Y$ was modeled using a simple HMM, and the fine-scale process $Z^*$ was modeled using a HMM-DFT with explicit auto-correlation in $Z^{*(1)}$.
-Naturally, we refer to this model as the \textit{CarHHMM-DFT}.
+Naturally, we refer to this model as the \textbf{CarHHMM-DFT}.
+Figure \ref{fig:CarHHMM-DFT} shows a graphical representation of this model.
 
 On the coarse scale, the dive types follow a Markov chain with $N=2$ possible states and unknown probability transition matrix $\Gamma$.
-Given the dive type, the duration of a dive follows a gamma distribution which depends upon the dive type and unknown parameters $\Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\}$.
+The duration of a dive follows a gamma distribution which depends upon the dive type and unknown parameters $\Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\}$.
 
@@
 
-$$\mathbb{E}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = \phi_1^{*(i^*)}z + (1-\phi_1^{*(i^*)}) \mu_1^{*(i^*)}$$
-$$\mathbb{V}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = \left(\sigma_1^{*(i^*)}\right)^2$$
+$$\mathbb{E}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \phi_1^{*(i)}z + (1-\phi_1^{*(i)}) \mu_1^{*(i)}$$
+$$\mathbb{V}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_1^{*(i)}\right)^2.$$
 %
-The distribution of $Z^{*(2)}_{t,t^*}$ is gamma and its parameters depend upon only $X^*_{t,t^*}$:
+The distribution of $Z^{*(2)}_{t,t^*}$ is gamma and its parameters depend upon only $X^*_{t,t^*}$ (not $X_t$):
 %
-$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = \mu_2^{*(i^*)}$$
-$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i^*) = (\sigma_2^{*(i^*)})^2$$
+$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
+$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
 %
-None of the fine-scale emission distributions depend upon dive type. In total the parameters to learn are
+None of the fine-scale emission distributions depend upon dive type. In total the parameters to estimate are
 %
 \begin{gather*}
-    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)} \\
+    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
     %
-    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters)} \\
+    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters), and} \\
     %
-    \Theta^* = \{\Theta^{*(1)},\Theta^{*(2)}\}  \qquad \text{(fine-scale emission parameters)} \\
+    \Theta^* = \{\Theta^{*(1)},\Theta^{*(2)}\}  \qquad \text{(fine-scale emission parameters), where} \\
     %
-    \Theta^{*(i^*)} =  \{\{\mu_1^{*(i^*)},\sigma_1^{*(i^*)},\phi_1^{*(i^*)}\},\{\mu_2^{*(i^*)},\sigma_2^{*(i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters)}
+    \Theta^{*(i^*)} =  \{\{\mu_1^{*(i^*)},\sigma_1^{*(i^*)},\phi_1^{*(i^*)}\},\{\mu_2^{*(i^*)},\sigma_2^{*(i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
 \end{gather*}
@@
 The likelihood of this model is still easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
-See the appendix for details of likelihood evaluation, and Figure \ref{fig:CarHHMM} shows the corresponding graphical model.
+See the appendix for details of likelihood evaluation.
+Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model.
 
@@
     \item An \textbf{HHMM-DFT}, which is similar to the model above, but with no modeled auto-correlation, i.e. $\phi_1^{*(i^*)} = 0$ for $i^* = 1,2$.
-    \item A \textbf{CarHHMM}, which is similar to the model above, but without $Z^{*(2)}$ as an observation, i.e $\Theta^{*(i^*)} = \{\mu_1^{(i^*)},\sigma_1^{(i^*)},\phi_1^{(i^*)}\}$, $i^* = 1,2$.
-    \item A \textbf{CarHMM-DFT} similar to the CarHHMM-DFT, but with $N=1$ instead of $N=2$, i.e.
-$\Gamma^* = \{\Gamma^{*(1)}\}$ and $\Theta^* = \{\Theta^{*(1)}\}$.
-This is equivalent to loosing one level of the hierarchical structure.
+    \item A \textbf{CarHHMM}, which is similar to the model above, but without $Z^{*(2)}$ as an observation (i.e $\Theta^{*(i^*)} = \{\mu_1^{(i^*)},\sigma_1^{(i^*)},\phi_1^{(i^*)}\}$, $i^* = 1,2$).
+    \item A \textbf{CarHMM-DFT}, which is similar to the CarHHMM-DFT, but with $N=1$ instead of $N=2$ on the coarse scale.
+This is equivalent to loosing the coarse level of the hierarchical structure, as there is only one dive type.
 \end{enumerate}
@@
 The CarHMM-DFT lacks a hierarchical structure, the HHMM-DFT is missing auto-correlation within the fine-scale observations, and the CarHHMM does not have access to the Fourier transform sums $(Z^{*(2)})$ as observations.
-All models were run on the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
 
@@
 
+All models were run on the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
 %%% Accuracy %%%
-Every model was able to decode the fine-scale hidden states of the process almost perfectly except for the \textit{CarHHMM}, which did not have access to the Fourier modes ($Z^{*(2)}$).
-This is intuitively clear because the distribution of $Z^{*(2)}$ varies between fine-scale states much more than $Z^{*(1)}$.
-For the coarse-scale hidden states, the CarHMM lacked a hierarchical structure and could not make any predictions at all.
+Each model was able to decode the fine-scale hidden states of the process almost perfectly except for the CarHHMM, whose accuracy was 89\%.
+This is intuitively clear because the distribution of $Z^{*(2)}$ varies between fine-scale states much more than that of $Z^{*(1)}$.
+For the coarse-scale hidden states, the CarHMM-DFT lacked a hierarchical structure and could not make any predictions at all.
 The other three models all achieved an accuracy of approximately 90\%, with the CarHHMM slightly more likely to categorize a dive as dive type 2 than the other models.
 Figure \ref{fig:acc} shows the decoded state probabilities for both the fine- and coarse- scales.
-(Tbl.
-\ref{table:accuracy}) also lists more details regarding the accuracy and training times of each model.
+Table \ref{table:accuracy} lists the accuracy and training times of each model.
 
 %%% dive duration %%%
-For the emission distributions of dive duration, estimates of standard error using the observed Fisher information tended to be underestimates due to correlation between $\hat \mu$ and $\hat \sigma$.
-In addition, $\hat \sigma$ tends to underestimate $\sigma$ for all models, which is a finding consistent with properties of MLEs, especially because the sample size for a particular dive type is approximately 50 for each simulation.
-The CarHMM model in particular severely underestimates $\sigma$.
+For the emission distributions of dive duration, Figure \ref{fig:MLE_dist} shows the empirical joint density of $\hat \mu$ and $\hat \sigma$ for the CarHHMM-DFT.
+The supplementary material shows the empirical joint densities for all other models and features.
+
+Estimates of standard error using the observed Fisher information tended to be underestimates due to correlation between $\hat \mu$ and $\hat \sigma$.
+In addition, $\hat \sigma$ tends to be an underestimate of $\sigma$ for all models and dive types.
+This finding is consistent with properties of MLEs for standard deviation, especially with a small sample size.
+The CarHMM-DFT model in particular is not well specified one the coarse-scale and severely underestimates $\sigma$.
 %This is likely due to the fact that dive duration follows a mixture of gammas rather than a gamma distribution, and the CarHMM does not have the machinery to capture this fact.
-A full table of parameter estimates for all models is shown in Table \ref{table:dive_duration}.
+A full table of parameter estimates for all models is shown in the supplementary material.
 
@@
 For acceleration ($Z^{*(1)}$), both the CarHMM-DFT and the CarHHMM-DFT regularly converged to the correct parameters with very little standard error.
-However, the Fisher standard error regularly overestimated the standard error for $\hat \mu$ for both of these models.
+However, the estimated standard error based on the observed Fisher information regularly overestimated the empirical standard error of $\hat \mu$ for both of these models.
 %This likely is due to the very large auto-correlation terms ($\phi^{(1)} = 0.99$ and $\phi^{(2)} = 0.95$) since as $\phi \to 1$, the value of $\mu$ does not matter in the likelihood evaluation as $\phi \to 1$ implies that $\mathbb{E}(Z^{*(1)}_{t,s^*}) = \phi*(Z^{*(1)}_{t,s^*-1}) + (1-\phi)\mu \to Z^{*(1)}_{t,s^*-1}$.
-The HHMM regularly overestimates the variance $Z^{*(1)}$ since it does not incorporate auto-correlation into the emission distribution of $Z^{*(1)}$, and the CarHHMM has large biases in many of its parameter estimates, especially in sub-dive state 2.
+The HHMM-DFT regularly overestimates the variance $Z^{*(1)}$ since it does not incorporate auto-correlation into the emission distribution of $Z^{*(1)}$.
+The CarHHMM has large biases in many of its parameter estimates, especially in sub-dive state 2.
 %because it sees the ``wigliness" of the acceleration data as raw acceleration rather than as Fourier modes, which deflates the value of $\hat \phi$ and inflates the value of $\hat \sigma$. 
-See Table \ref{table:acceleration} for a detailed breakdown of parameter values for acceleration emission distributions.
+See the supplementary material for a detailed breakdown of parameter values for acceleration emission distributions.
 
@@
 For all models there is no bias in the parameter estimates for the distribution of the Fourier sums ($Z^{*(2)}$).
-One exception is the CarHHMM, which does not model $Z^{*(2)}$ as an observation.
-The observed fisher information standard errors are good approximations of the empirical standard error.
+One exception is the CarHHMM, which does not model $Z^{*(2)}$ as an observation at all.
+The standard error estimates based on the observed Fisher information closely approximate the empirical standard error.
 %The distribution of $Z^{*(2)}$ is very different between subdive states 1 and 2, so the signal is therefore very easy for each model to pick up on.
-(Tbl. \ref{table:FoVeDBA}) shows a detailed breakdown of the emission distribution of $Z^{*(2)}$ for all models.
+The supplementary material also shows a detailed breakdown of the emission distribution of $Z^{*(2)}$ for all models.
 
 %%% Gamma %%%
-
-Estimates of the probability transition matrix $\Gamma$ for all models (except for the CarHMM-DFT) are very accurate, with empirical standard errors of approximately 0.02.
-This is significantly less than the observed fisher information estimate of standard error of approximately 0.08.
+Estimates of the coarse-scale probability transition matrix $\Gamma$ for all models (except for the CarHMM-DFT) are very accurate, with empirical standard errors of approximately 0.02 and biases of nearly zero.
+This is significantly less than the estimate of standard error based on the observed Fisher information, which is about 0.08 for all models.
 %likely because the distribution of $\hat \Gamma$ did not converge to a normal distribution in only 100 observed dives. 
-For $\Gamma^*$, the HHMM-DFT and CarHHMM-DFT both showed practically no bias with standard errors on the order of $10^{-2}$.
-One notable exception is the standard error of $\hat \Gamma^{*(1)}_{12}$, whose standard error was on the order of $10^{-4}$, which is much lower than the observed Fisher information predicted.
-The CarHMM-DFT is mis-specified, so its results cannot be easily interpreted.
-The CarHHMM consistently underestimated the probability of a hidden state change on the sub-dive behavior.
+For the fine-scale transition matrices $\Gamma^*$, the HHMM-DFT and CarHHMM-DFT both showed practically no bias with standard errors on the order of $10^{-2}$.
+One notable exception is the standard error of $\hat \Gamma^{*(1)}_{12}$, whose standard error was on the order of $10^{-4}$, much lower than the observed Fisher information would predict.
+The CarHMM-DFT is mis-specified, so its results cannot be easily interpreted, but $\hat \Gamma^*$ regularly converged to a constant value with a standard error on the order of $10^{-2}$.
+The CarHHMM consistently overestimated the trace of $\Gamma^{*(i^*)}$.
 Again, one notable exception is $\hat \Gamma^{*(1)}_{12}$, which had almost no bias and was remarkably consistent.
-See (Table \ref{table:Gamma}) for a full list of estimates and standard errors.
-
-\textcolor{red}{I also have plots corresponding to each table.
-For example, I have included figure \ref{fig:MLE_dist} as an example below for the emission distribution of the MLEs for dive duration of the CarHMM.
-Should I use those instead?
-But there are just SO many plots (4 models * 3 observations * 2 states = 24 plots)}
+See the supplementary material for a full list of estimates and standard errors.

### 2020-07-17 Report/Sections/intro.tex

@@
 In this work, we consider an example where killer whales exhibit a variety of different types of dives at a coarse scale, but also exhibit many different types of swimming behaviors within dives at a fine scale.
-One solution to this issue is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017} \citep{Adam:2019}.
+One solution to this issue is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
 HHMMs model the entire time series in question as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.
 
-All HMMs presented so far assume Markovian dynamics in the underlying process (i.e.
+HMMs assume Markovian dynamics in the underlying process (i.e.
 that any observation $Y_t$ depends only on the behavioral state $X_t$ and $Y_{t-1}$ when conditioned on all previous time steps).
@@
 
-One solution is to model the fine-scale behavior using a continuous-time model, which involves modelling the dynamics of an animal as the solution to a stochastic differential equation.
+One solution is to model the fine-scale behavior using a continuous-time model, which involves modelling the dynamics of an animal as the solution of a stochastic differential equation.
 Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals.
-Moreover, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
+Unfortunately, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to fit.
 
@@
 The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}.
-In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017}, specifically fluking \citep{Shorter:2017} from accelerometer data.
+In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data.
 Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appeared a promising simple approach to account for additional correlation in data that is cyclical in nature.
-
-We consider several models: the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM).
-The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes.
-The CarHMM modifies the usual HMM by modelling additional dependence into the sequence of observed data.
-In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.
 
@@
 We then perform a simulation study to compare the performance each existing model with ours in a controlled setting.
-Finally, we apply our method to dive data collected from a Northern resident killer whale off the coast of British Columbia, Canada.
+Finally, we apply our method to dive data of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
 

### 2020-07-17 Report/Sections/results.tex

@@
 The CarHHMM-DFT was used to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada.
-Acceleration data can be a good proxy for energy expenditure \citep{Green:2009}, but studies suggest that the behavioral state must be taken into account when using accelerometer data \citep{Dot:2016}.
+Acceleration data can be a good proxy for energy expenditure \citep{Green:2009}, but studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
 Therefore, understanding both the behavioral state of the killer whale as well as the distribution of accelerometer data within each behavioral state is import to understand the energetic requirements of killer whales.
-This, in turn, can assist conservation efforts.
+This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.
 
@@
 After preprocessing the raw data, a total of 267 dives were observed.
-A plot of the raw data for all dives and a collection of five selected dives can be seen in Figure \ref{fig:data} and Figure \ref{fig:data_one_dive}, respectively.
+Figure \ref{fig:data} displays the dive profile and accelerometer data.
 
@@
 
-The acceleration exhibits significant sinusoidal behavior at several points in time (see fig \ref{fig:data_one_dive}), so the fine-scale observations $Z^*$ were made up of the DFT summary statistics of a two-second sliding window.
-Acceleration data was a 3-dimensional vector, so $Z^*$ was calculated as follows:
+The acceleration exhibits significant sinusoidal behavior, so the fine-scale observations $Z^*$ were made up of the DFT summary statistics of a two-second sliding window.
+Unlike the simulation study, the acceleration observations here are 3-dimensional vectors rather than scalars, so the observations $z^*$ were calculated as follows:
 %
-$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2$$
+$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2.$$
 %
-$\mathbf{z}_{t,t^*}^{*(1)}$ is a 3-dimensional vector while $z_t^{*(2)}$ is a scalar.
-Summing the first 10 Fourier modes to calculate $z^{(2)}_{t,t^*}$ corresponds to a maximum frequency of $\tilde \omega = 5$ Hz.
-Figure \ref{fig:lag} displays a lag plot for all major features and reveals strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions.
-Auto-correlation was therefore directly modeled into the the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$, which is assumed to be Normally distributed with the following parameters:
+The observations are therefore made up of $\mathbf{z}_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $z_t^{*(2)}$, a scalar.
+We calculate $z_t^{*(2)}$ by summing the first 10 Fourier modes, and corresponds to a maximum recorded frequency of $\tilde \omega = 5$ Hz.
+There is strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions (see the supplementary material for a lag plot).
+Therefore, auto-correlation was directly modeled into the the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$, which is assumed to be Normally distributed with the following parameters:
 %
@@
 Auto-correlation was therefore not incorporated into the emission distribution of $Z^{*(2)}_{t,t^*}$.
-The distribution of $Z^{*(2)}_{t,t^*}$ was also assumed to be gamma:
+In particular, the distribution of $Z^{*(2)}_{t,t^*}$ was assumed to be gamma and parameterized by its mean and variance:
 %
 $$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
-$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2$$
+$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
 %
-$Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent when conditioned of the sub-dive state $X_{t,t^*}$.
+The observations $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent of one another when conditioned of the sub-dive state $X_{t,t^*}$.
 
-It is known that information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically and admittedly somewhat arbitrarily.
+It is known that information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically.
 The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
 
-The final model is nearly identical to the one from the simulation study, with the exception that the fine scale Markov chain has three sub-dive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a vector rather than a scalar.
+The final model is nearly identical to the one from the simulation study, with the exception that the fine-scale Markov chain has three sub-dive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a 3-dimensional vector rather than a scalar.
 The model is comprised of two levels: a coarse-scale HMM and a fine-scale CarHMM-DFT.
-The coarse-scale model is comprised of an HMM with no auto-correlation and no DFT with hidden states corresponding to dive types and observations corresponding to dive durations.
+The coarse-scale model is comprised of an HMM with hidden states corresponding to dive types and observations corresponding to dive durations.
 The fine-scale model is comprised of a CarHMM-DFT where auto-correlation is modeled into the distribution of $\mathbf{Z}^{*(1)}$ (the average acceleration with a 2-second window), but not $Z^{*(2)}$ (the ``wiggliness" of the two second window).
-Refer to Figure \ref{fig:CarHHMM} for a full graphical representation of this model.
+Refer to Figure \ref{fig:CarHHMM-DFT} for a full graphical representation of this model.
 
@@
 
-Table \ref{table:emis_dists} displays estimates of the emission distribution parameters.
-Each fitted distribution is also plotted in Figure \ref{fig:coarse_emis} and Figure \ref{fig:fine_emis}.
-While the ecological interpretation of behavioral states is tenuous for HMMs, we hypothesize the following meanings.
+Table \ref{table:emis_dists} displays estimates of the emission distribution parameters for this case study.
+The fitted emission distributions are also plotted in Figures \ref{fig:coarse_emis} and \ref{fig:fine_emis}.
+While the ecological meaning behind behavioral states is tenuous for HMMs, we hypothesize the following interpretations.
 Dive type 1 corresponds to shorter, shallower dives which serve a variety of purposes, including as rest before dives of type 2, which are deeper and more sustained.
 Sub-dive behavioural state 1 corresponds to gliding and less overall activity compared to the other behavioral states.
-The mean of $Z^{*(2)}$ in this state is at least an order of magnitude smaller than sub-dive behavior 2, the variance of $\mathbf{Z}^{*(1)}$ is smaller than sub-dive behavior 2 for every component, and the auto-correlation of $\mathbf{Z}^{*(1)}$ is higher than every other behavioral state.
-Sub-dive state 3, on the other hand, corresponds to vigorous swimming activity, as the mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ for every component are much higher than every other state.
+The mean of $Z^{*(2)}$ in this state is at least an order of magnitude smaller than sub-dive behavioral state 2, the variance of $\mathbf{Z}^{*(1)}$ is smaller than sub-dive behavior 2 for every component, and the auto-correlation of $\mathbf{Z}^{*(1)}$ is higher than every other behavioral state.
+Sub-dive state 3, on the other hand, corresponds to vigorous swimming activity, as the mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ for every component is much higher than every other state.
 The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this state, implying more variation in acceleration every 2 seconds.
@@
 0.907  &  0.093
-\end{pmatrix}$$
-$$\hat \delta = \begin{pmatrix} 0.857 & 0.143 \end{pmatrix}$$
+\end{pmatrix},$$
+$$\hat \delta = \begin{pmatrix} 0.857 & 0.143 \end{pmatrix},$$
 %
@@
 0.000 & 0.247 & 0.753
-\end{pmatrix} \qquad 
+\end{pmatrix}, \qquad 
 \hat \Gamma^{*(2)} = \begin{pmatrix} 
@@
 0.000 & 0.246 & 0.754
-\end{pmatrix}$$
-$$\hat \delta^{*(1)} = \begin{pmatrix} 0.143 & 0.698 & 0.159 \end{pmatrix} \qquad
+\end{pmatrix},$$
+$$\hat \delta^{*(1)} = \begin{pmatrix} 0.143 & 0.698 & 0.159 \end{pmatrix}, \qquad
 \hat \delta^{*(2)} = \begin{pmatrix} 0.476 & 0.456 & 0.067 \end{pmatrix}.$$
@@
 About 86\% of observed dives are short; the whale usually rests for many dives in a row before performing a deep dive.
-The probability transition matrix $\hat \Gamma$ shows that the distribution of a particular dive's type is approximately equal regardless of the previous dive type.
-The fine-scale probability transition matrices imply that the killer whale is much more likely to be in a less active sub-dive state when performing deep dives than when performing shallow dives- 14\% of shallow dives are spent in sub-dive state 1 while 48\% of deep dives are spent in sub-dive state 1.
-Using less active sub-dive states when diving deep could be an energy reduction strategy for these long periods of holding breath- a phenomenon which has been observed in bottle-nose dolphins \citep{Williams:1999}.
-(Fig \ref{fig:labeled_dives}) shows the decoded dive behavior of 5 selected dives, and (fig \ref{fig:coarse_probs}) and (fig \ref{fig:fine_probs}) display the probability of each dive type and sub-dive state, respectively.
+The probability transition matrix $\hat \Gamma$ shows that the probability of a particular dive type does not depend much on the previous dive type.
+This killer whale is much more likely to be in a less active sub-dive state when performing deep dives than when performing shallow dives:  sub-dive state 1 makes up 14\% of shallow dives, but it makes up 48\% of deep dives.
+Using less active sub-dive states when diving deep could be an energy reduction strategy for these long periods of holding breath.
+This phenomenon has been observed in bottle-nose dolphins \citep{Williams:1999} and may also be the case for killer whales.
+Figure \ref{fig:labeled_dives} shows the decoded dive behavior of 5 selected dives.
+The supplementary material also shows the estimated \textit{probability} of each dive time and sub-dive hidden state.
 
@@
 A pseudo-residual of a particular observation is the marginal CDF of an observation conditioned on all other observations under the learned model \citep{Zucchini:2016}.
-To easily visualize outliers, the pseudo-residual is often passed through the quantile function of the standard Normal distribution.
-Mathematically, the pseudo-residual of an observation $y_t$ is $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
+To easily visualize outliers, this pseudo-residual is often passed through the quantile function of the standard Normal distribution.
+Mathematically, the pseudo-residual of an observation $y_t$ in a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
 If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution.
 We find that histograms of the pseudoresiduals of this model mostly support that the model is well-specified.
-$Z^{*(2)}$ is an exception, as its pseudo-residuals are noticeably right-skewed (fig \ref{fig:pseudoresids}).
+$Z^{*(2)}$ is an exception, as its pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
 This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution compared to a gamma distribution such as a power law.
 
-In addition to psuedoresiduals, we plotted histograms of each feature where each observation was weighted by the probability that the whale was in a particular hidden state.
+We also plotted separate histograms of each feature where each observation was weighted by the probability that the whale was in a particular hidden state.
 This empirical distribution was then plotted over the fitted probability distribution function of that feature and hidden state.
-If the model is correct, then the histograms of features should closely resemble their respective fitted probability distribution.
-Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is right-skewed.
-In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for sub-dive state 3 (fig.
-\ref{fig:empirical_dist}), indicating the existence of rare events corresponding to very violent thrashing of the killer whale.
+If the model is correct, then the histogram of each feature and hidden state should closely resemble its corresponding fitted probability distribution.
+See Figure \ref{fig:empirical_dist} for an example of the dive duration.
+Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed.
+In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for sub-dive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
 These outliers are potential subjects for future study.
+See the supplementary material for empirical distributions of every feature and every hidden state.

### 2020-07-17 Report/main.tex

@@
 technology have made a vast
-amount of movement data available in a variety of fields. This
-rich data can exhibit simultaneous behavioural processes occurring at different
-time scales, resulting in very complicated dependence structures. These processes can be modelled 
+amount of movement data available in a variety of fields. This data can exhibit simultaneous behavioural processes occurring at different
+time scales, resulting in increasingly complicated dependence structures. These processes can be modelled 
 through a hierarchical hidden Markov model
-(HHMM), where the system is modelled as a
+(HHMM), which models the system as a
 nested structure of hidden Markov models (HMMs).
-At very short time scales, however, many assumptions of traditional HMMs are violated. We demonstrate how
+At very short time scales, however, many basic assumptions of traditional HMMs are violated. We demonstrate how
 to incorporate fine-scale processes into the
@@
 %%%%%%%%%%%%
-\vfill\eject
-{\centerline{\Large{Notes from Nancy}}}
-\input{Sections/0627NH}
-\vfill\eject
+%\vfill\eject
+%{\centerline{\Large{Notes from Nancy}}}
+%\input{Sections/0627NH}
+%\vfill\eject
 %%%%%%%%%%%%%

### 2020-07-19 Code/HHMM.py
Update Simulation Plots- better optimizer

@@ -539,7 +539,7 @@ class HHMM:
                         x0 = self.theta[0][feature][param][k0]
 
                         # optimize
-                        res = minimize(loss_fn, x0, method='Nelder-Mead', options=options)
+                        res = minimize(loss_fn, x0, method='BFGS', options=options)
 
                         # update final values
                         x = np.copy(res['x'])
@@ -651,7 +651,7 @@ class HHMM:
                             x0 = self.theta[1][k0][feature][param][k1]
 
                             # optimize
-                            res = minimize(loss_fn, x0, method='Nelder-Mead', options=options)
+                            res = minimize(loss_fn, x0, method='BFGS', options=options)
 
                             # update final values
                             if self.likelihood(data) >= backup_l:

### 2020-07-19 Code/run_all_sims.sh

@@ -8,7 +8,7 @@
 #SBATCH --mail-type=REQUEUE
 #SBATCH --mail-type=ALL
 
-#SBATCH --time=20:00:00
+#SBATCH --time=48:00:00
 #SBATCH --mem-per-cpu=4G
 #SBATCH --array=1-500
 

### 2020-07-19 Report/Sections/models.tex
Updates from Overleaf

@@
 You can use some of the text I removed from below}.
-To account for each of these common additional dependence structures, we will explore three variations on the traditional HMM: the conditionally auto-regressive hidden Markov model (CarHMM), hierarchical hidden Markov model (HHMM), and HMM-DFT.
+To account for each of these common additional dependence structures, we will explore three variations on the traditional HMM: the conditionally auto-regressive hidden Markov model (CarHMM), hierarchical hidden Markov model (HHMM), and HMM with discrete Fourier transform (HMM-DFT).
 We will show how each of these variations can be altered and combined to form a wide variety of new models.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The HMM with discrete Fourier transform (HMM-DFT)}
+\subsection{Accounting for periodicity with HMM-DFT}
 \label{subsec:STFT}
 
-The HMM with discrete Fourier transform, or HMM-DFT, incorporates hierarchical structure into an HMM differently than an HHMM.
+The HMM-DFT incorporates hierarchical structure into an HMM differently than an HHMM (\textit{ref Fig 1}).
 In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform.
 For simplicity, we assume that the length of the fine-scale processes is constant (i.e.
-that $T^*_t = T^*$), although this need not be the case in general.
+that $T^*_t = T^*$), although this need not be the case.
 Suppose that the fine-scale process $y^*_t$ does not switch hidden states, but does exhibit significant periodic behaviour.
@@
 These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$.
-Figure \ref{fig:HMM-DFT} shows a graphical representation of the HMM-DFT.
+Figure \ref{fig:HMM-DFT} represents the structure of the HMM-DFT.
 
-Once $z^*_t$ is calculated, it can be treated as an observation of the coarse-scale HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$.
-In total, the likelihood of the hierarchical HMM-DFT is as follows:
+Once $z^*_t$ is calculated, it can be treated as an observation of the HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$.
+The likelihood of the HMM-DFT is as follows:
 \begin{equation}
@@
 
-\subsection{General hierarchical structures}
+\subsection{General structure for building complexity}
 
@@
 Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, state-switching OU, HHMM, and HMM-DFT).
-These five models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models based on HMMs.
+These alternative models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models based on HMMs.
 In the sections that follow, we perform both a simulation study and real-world case study modelling killer whale dive behaviour using models constructed from these building blocks.

### 2020-07-19 Report/Sections/results.tex
Updates from Overleaf

@@
 The CarHHMM-DFT was used to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada.
-Acceleration data can be a good proxy for energy expenditure \citep{Green:2009}, but studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
-Therefore, understanding both the behavioral state of the killer whale as well as the distribution of accelerometer data within each behavioral state is import to understand the energetic requirements of killer whales.
+Acceleration data can be used to approximate an animal's energy expenditure \citep{Green:2009}, but studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
+Therefore, understanding both the behavioral state of the killer whale, as well as the distribution of accelerometer data within each behavioral state, is important to understand the energetic requirements of killer whales.
 This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.
@@
 Observations were collected at a rate of 50 Hz.
-Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time range were ignored.
-In addition, the tagging technology dropped data between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed data within this time range were ignored as well.
+Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time periods were ignored.
+In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed dives within this time range were ignored as well.
 A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
@@
 
-It is known that information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically.
+Information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically.
 The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

### 2020-07-19 Report/Sections/simulation.tex

@@
 
-We base the following simulated study on the acceleration data of a killer whale's dive sequence.
+We based our simulation study on the acceleration data of a killer whale's dive sequence.
 The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}.
-We fit four separate models to the simulated data and compare their test the accuracy of each on a simulated test data set.
+We fit four separate models to the simulated data and compare their accuracy.
 
@@
 
-500 separate sequences of 100 killer whale dives were simulated according to an HMM, where the hidden Markov chain $X$ was a collection of dive types and the observations $Y$ were the corresponding dive durations (in seconds).
+Five hundred separate sequences of 100 killer whale dives were simulated according to an HMM, where the hidden Markov chain $X$ was a collection of dive types and the observations $Y$ were the corresponding dive durations (in seconds).
 Each dive could be one of $N=2$ dive types, and the duration of the $t^{th}$ dive, $Y_t$, followed a gamma distribution whose parameters $\theta^{(i)}$ were dependent on the dive type $X_t = i$.
@@
 %
-$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
-$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
+$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(2)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
+$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(2)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
 %
@@
 
-Including the CarHHMM-DFT described above, four different models were fit to the simulated data sets:
+Four different models were fit to the simulated data sets:
 \begin{enumerate}
@@
 Each model was able to decode the fine-scale hidden states of the process almost perfectly except for the CarHHMM, whose accuracy was 89\%.
-This is intuitively clear because the distribution of $Z^{*(2)}$ varies between fine-scale states much more than that of $Z^{*(1)}$.
+The lower accuracy of the CarHHMM can be attributed to the higher fine-scale states variation in the distribution of $Z^{*(2)}$ compared to $Z^{*(1)}$.
 For the coarse-scale hidden states, the CarHMM-DFT lacked a hierarchical structure and could not make any predictions at all.
 The other three models all achieved an accuracy of approximately 90\%, with the CarHHMM slightly more likely to categorize a dive as dive type 2 than the other models.
-Figure \ref{fig:acc} shows the decoded state probabilities for both the fine- and coarse- scales.
+Figure \ref{fig:acc} shows the decoded state probabilities for both the fine- and coarse- scales for one of the simulation run.
 Table \ref{table:accuracy} lists the accuracy and training times of each model.
@@
 
-Estimates of standard error using the observed Fisher information tended to be underestimates due to correlation between $\hat \mu$ and $\hat \sigma$.
+Estimates of standard error using the observed Fisher information tended to be negatively biased due to correlation between $\hat \mu$ and $\hat \sigma$.
 In addition, $\hat \sigma$ tends to be an underestimate of $\sigma$ for all models and dive types.
 This finding is consistent with properties of MLEs for standard deviation, especially with a small sample size.
-The CarHMM-DFT model in particular is not well specified one the coarse-scale and severely underestimates $\sigma$.
+In particular, the CarHMM-DFT model is not well specified one the coarse-scale and severely underestimates $\sigma$.
 %This is likely due to the fact that dive duration follows a mixture of gammas rather than a gamma distribution, and the CarHMM does not have the machinery to capture this fact.

### 2020-07-19 Code/Visualisor.py
Add decoded state probability plot

@@ -262,6 +262,8 @@ class Visualisor:
             else:
                 nrows += 2
 
+        nrows=5
+
         plt.subplots(nrows,1,figsize=(30,5*nrows))
         fignum = 1
 
@@ -278,58 +280,69 @@ class Visualisor:
         for state in range(self.pars.K[0]):
             dives.append(dive[dive['ML_dive'] == state])
 
-        ylabs = [r'$\left(Z^{*(1)}\right)_x$ $(m/s^2)$','Depth $(m)$']
+        ylabs = [r'$P\left(X_t = 1\right)$',
+                 r'$P\left(X_t = 2\right)$',
+                 r'$P\left(X^*_{t,t^*} = 1\right)$',
+                 r'$P\left(X^*_{t,t^*} = 2\right)$',
+                 r'$P\left(X^*_{t,t^*} = 3\right)$']
+
+        title = 'Hidden State Probability Estimates'
 
         for i,col in enumerate(df_cols):
 
             # dive-level coloring
-            plt.subplot(nrows,1,fignum)
-            colors = [cm.get_cmap('tab10')(i) for i in [0,1]]
-            legend = ['Dive Type %d' % (i+1) for i in range(self.pars.K[0])]
-            for state,dive_df in enumerate(dives):
-                plt.plot(dive_df['sec_from_start']/60,dive_df[col],
-                         '.',color=colors[state],markersize=10)
-            if 'prob' in col:
-                plt.plot(dive[dive[col] > -0.01]['sec_from_start']/60,dive[dive[col] > -0.01][col],'k-')
-                plt.axhline(0.5,color='k')
-                plt.ylim([-0.05,1.05])
-            else:
-                plt.plot(dive['sec_from_start']/60,dive[col],'k--')
-            plt.yticks(fontsize=30)
-            plt.ylabel(ylabs[i],fontsize=30)
-            plt.xticks([])
-            if col == df_cols[0]:
-                plt.title('Decoded Dive/Accelorometer Data',fontsize=36)
-            if col == 'depth':
-                plt.gca().invert_yaxis()
-            fignum += 2
+            if 'subdive_state_' not in col:
+
+                plt.subplot(nrows,1,fignum)
+                colors = [cm.get_cmap('tab10')(i) for i in [0,1]]
+                legend = ['Dive Type %d' % (i+1) for i in range(self.pars.K[0])]
+                for state,dive_df in enumerate(dives):
[…]

### 2020-07-19 Report/Sections/intro.tex

--- 2020-07-12 Report/Sections/intro.tex
@@
-% !TeX root = ../main.tex
+{\em{This still needs some editing}}
 
-%\section{Introduction}
+\vskip 10pt\noindent
 
-The field of animal movement is in the midst of a ``data renaissance" where advancements in tagging technology have given rise to an explosion of data available for statistical modeling.
-In particular, tagging technologies are capable of recording observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
-In response, researchers have introduced a variety of new statistical techniques to infer animal behavior from movement data \citep{Hooten:2017}.
+Recent advances in technology allow us to gather enormous amounts of data collected almost continuously in time.
+The collection and analysis of movement data from devices such as accelerometers and GPS trackers have brought new understanding to {\bf{more references}}  physical activity levels in children \citep{Morris:2007} ...
+In this article, we focus on the study of animal movement;  tags attached to animals record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours and providing biologists with new tools for understanding animal behaviour.
 
+Many statistical techniques have been developed and applied to these types of longitudinal or functional data.  
+Some techniques, such as time series, are highly parametric. 
+While these parametric techniques do not allow the data to ``speak for itself", the ensuing data analysis results typically are readily interpretable, providing important subject area insights.
+At the other extreme are black box type methods of machine learning.  
+While these methods are incredibly successful at predicting how processes evolve, they provide little structured insight into the driving mechanisms of the process.
+Functional Data Analysis (FDA) techniques lie somewhere in between these extremes: the output of an analysis provides visual tools for understanding processes and, often, some dimension reduction to allow for easier exploration.
+The development of structured FDA techniques incorporating more complex dependence structures has become a topic of recent interest  (see, e.g., Crainiceanu. and references therein).
+
+Our goal is to develop movement analysis methods that are flexible but also use subject area knowledge to provide insight into curve data with complex structure.
+We  model complexities arising  from curve to curve variation, along with the complexities of within-curve fine-scale characteristics and temporal dependence,
+combining existing methods in novel ways.
+We enforce temporal structure via Markov processes.  
+While our methodology has general applicability, [rephrase - awkward] we  showcase our ideas to study the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
+For more information about the data, see Section \ref{sec:case_study}.
+ Figure \ref{fig:data} shows the data we will analyze.
+   The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and maximum depths. 
+The bottom three panels show the acceleration in each of three orthogonal directions,  as recorded by an accelerometer,   
+with axes oriented relative to the whale's forward movement.
+Here, we see spikes in acceleration and, in contrast, periods of fairly constant acceleration.
+Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
+The plot shows the magnitude of the acceleration and from this, we can see fine-scale behaviour within each dive which seems to indicate some fine scale short-term periodicity. 
+We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour.  
+ Such an analysis helps biologists to understand the foraging behaviour and feeding success of killer whales.
+
+The killer whale dives form a sequence of dependent curves.   However, in FDA, curves are typically deemed independent.  
+A noted exception involves multilevel modelling [references] which uses random effects and is appropriate when the set of curves can be partitioned into independent subsets, typically with each subset arising from one individual.
+[NEED REFERENCES]  
+However, this approach is not appropriate here, so we consider a temporal structure for dependence, with the characteristics of each curve determined by a latent state that evolves according to a Markov model.
+This HMM allows us to interpret each curve as arising from a type of behaviour.
+
+From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
+In FDA, within curve structure is modelled by a generic smooth mean function and covariance function.
+Alternatively, this structure is sometimes generated via random regression.
+However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.  
+We model the within curve dependence by combining a hidden Markov model of within-curve behaviour combined with an autoregressive-type dependence.
+Finally, for the fine-scale behaviour, we use summary information that reflects the application area's meaning for the curve.
+Specifically, for the dive data, we use summaries of Fourier transforms to account for sudden and extreme ``wiggliness'' within some segments of a curve.
+The distribution of this ``wiggliness'' depends on the hiddent state.
+
+{\bf{ Make sure to reference Langrock}}
+
+Our work relies on work in the vast field of methodology for the analysis of animal movement
+\citep{Hooten:2017}.
 One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}.
-Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another conditioned on the underlying behavioral process.
-However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies.
-For example, the location of an animal at a given time is highly correlated with the location of that animal one second later.
+%
+%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another conditioned on the underlying behavioral process.
+ %However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
+ %For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
  Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model (CarHMM) \citep{Lawler:2019}.
-The CarHMM in particular explicitly models auto-correlation into an HMM while maintaining the structure needed to run the forward algorithm.
-It also only adds one additional parameter per possible hidden state.
+ The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
+ %while maintaining the structure needed to run the forward algorithm. 
+ %It also only adds one additional parameter per possible hidden state.
 
-Another issue that arises in high-frequency data is that several simultaneous behavioral processes may occur at different time scales.
-In this work we consider the example where killer whales perform a variety of different types of dives at a coarse scale, but also exhibit many different types of swimming behaviors within dives at a fine scale.
-One solution to this issue is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017} \citep{Adam:2019}.
+Animal movement modelers have recognized that several simultaneous behavioral processes may occur at different time scales. 
+One approach to the analysis of this type of data is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017} \citep{Adam:2019}. 
 HHMMs model the entire time series in question as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.
 
-All HMMs presented so far assume Markovian dynamics in the underlying process, i.e.
-that any observation $Y_t$ depends only on the behavioral state $X_t$ and $Y_{t-1}$ when conditioned on all previous time steps.
 At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs.
 Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}.
-With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine-scale processes are likely to persist.
-
 One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation.
@@
 
-For periodic behavior in particular, One way to avoid the use of continuous time models is to use signal processing techniques such as the Fourier transform on the raw data.
+For periodic behavior in particular, one way to avoid the use of continuous time models is to use signal processing techniques such as the Fourier transform on the raw data. 
 The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}.
  In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data.
-Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appeared a promising simple approach to account for additional correlation in data that is cyclical in nature.
+ Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears a promising simple approach to account for additional correlation in data that is cyclical in nature.
 
-We consider several models- the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM).
+We consider several models: the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM). 
 The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes. 
@@
 In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. 
-We then perform a simulation study to compare the performance each existing model with ours in a controlled setting.
-Finally, we apply our method to dive data collected from a Northern resident killer whale off the coast of British Columbia, Canada.
-
-%%%%%%%%%% Nancy's comments %%%%%%%%%%%%%
-
-\iffalse
-
-\section{Introduction}
-
-{\bf{NH: These are my comments.  I hope they help you.}}
-
-This is where you describe how your method contributes to statistical methodology.
-That is the big goal.
-So there will be literature review of directly relevant work.
-You don't want to leave the non-expert completely lost though, so it is a balance, how much background you give.
-Thank goodness for the Zucchini book, a great place to direct readers for further background.  
-The introduction is a good place to put information about the motivating data set and how researchers have modelled it, what the shortcomings have been.
-I see that you have some great statements about this in your Background section.
-Those statements can be moved to the introduction (with editing -- you can see how I edited).
-
-A hard part about the introduction is that you want to use terminology like CarHMM without getting into too many technical details.
-Sometimes you just have to be technical.
-But I think you can avoid a lot of that.
-
-Here is a great thing you wrote that would go well in the introduction, with some editing so that it doesn't rely on notation that hasn't been introduced.
-You could introduce the basic HMM notation in the intro, if you need to.
-I don't think introducing HMM notation is too technical.
-\begin{quote}
- One of the key assumptions of both HMMs and HHMMs is \textit{conditional independence} between observations on both the coarse and fine scale.
-Namely, when given the state $X_t$, $Y_t$ is assumed to be independent from all other observations, including $Y_{t-1}$.
-Therefore, traditional HMMs and HHMMs can fail when the observations $Y$ exhibit significant correlation in time.
- [leave in these examples for intro]  Examples include fluking in marine mammals in Vancouver, BC (see the results section) and the swimming behavior of horn sharks off the coast of Southern California \citep{Adam:2019}.
-
-One way to deal with auto-correlation in fine-scale behavioral processes is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by \citep{Lawler:2019}.
-.....
-The CarHMM therefore explicitly models auto-correlation into the emission distributions of the HMM while maintaining the structure needed to run the forward algorithm.
-It also only adds one additional parameter per possible hidden state
-$(\phi^{(i)})$. 
-\end{quote}
-
-\fi
+We then perform a simulation study to compare the performance of each existing model with ours in a controlled setting. 
+Finally, we apply our method to the dive data of Figure 6.

### 2020-07-19 Code/Visualisor.py
Update Plots

@@ -264,7 +264,7 @@ class Visualisor:
 
         nrows=5
 
-        plt.subplots(nrows,1,figsize=(30,5*nrows))
+        plt.subplots(nrows,1,figsize=(20,5*nrows))
         fignum = 1
 
         # get df state-by-state
@@ -461,7 +461,7 @@ class Visualisor:
                                   markerfacecolor=cm.get_cmap('viridis')(0.5), markersize=30),
                            Line2D([0], [0], marker='o', color='w', label='Subdive Type 3',
                                   markerfacecolor=cm.get_cmap('viridis')(1.0), markersize=30)]
-        plt.gca().legend(handles=legend_elements,prop={'size': 24}, ncol=5, mode="expand", borderaxespad=0.,
+        plt.gca().legend(handles=legend_elements,prop={'size': 20}, ncol=5, mode="expand", borderaxespad=0.,
                          bbox_to_anchor=(0., -0.5, 1., .102), loc='lower left')
         plt.subplots_adjust(wspace=0, hspace=0.1)
 

### 2020-07-22 Code/Visualisor.py
Add model-specific plots

@@ -118,10 +118,11 @@ class Visualisor:
             nrows = 1
             ncols = len(self.hhmm.theta[0])
         else:
-            features = ['Ax','Ay','Az','Ahat_low'] #list(self.hhmm.theta[1][0].keys())
+            features = list(self.hhmm.theta[1][0].keys())
             nrows = 2 #self.pars.K[0]
-            ncols = int(len(features)/2)
+            ncols = int((len(features)+1)/2)
 
+        print(nrows,ncols)
         fig, ax = plt.subplots(nrows,ncols,figsize=(7.5*ncols,7.5*nrows))
         ax = np.reshape(ax,(nrows,ncols))
 
@@ -262,7 +263,8 @@ class Visualisor:
             else:
                 nrows += 2
 
-        nrows=5
+        #nrows=5
+        nrows=4
 
         plt.subplots(nrows,1,figsize=(20,5*nrows))
         fignum = 1
@@ -280,14 +282,17 @@ class Visualisor:
         for state in range(self.pars.K[0]):
             dives.append(dive[dive['ML_dive'] == state])
 
-        ylabs = [r'$P\left(X_t = 1\right)$',
-                 r'$P\left(X_t = 2\right)$',
-                 r'$P\left(X^*_{t,t^*} = 1\right)$',
-                 r'$P\left(X^*_{t,t^*} = 2\right)$',
-                 r'$P\left(X^*_{t,t^*} = 3\right)$']
+        #ylabs = [r'$P\left(X_t = 1\right)$',
+        #         r'$P\left(X_t = 2\right)$',
+        #         r'$P\left(X^*_{t,t^*} = 1\right)$',
+        #         r'$P\left(X^*_{t,t^*} = 2\right)$',
+        #         r'$P\left(X^*_{t,t^*} = 3\right)$']
+        ylabs = [r'$\left(Z^{*(1)}\right)_x$ $(m/s^2)$',
+                 r'Depth $(m)$']
 
-        title = 'Hidden State Probability Estimates'
-
+        #title = 'Hidden State Probability Estimates'
+        title = 'Decoded Dive Profile / Accelerometer Data'
+        print(df_cols)
         for i,col in enumerate(df_cols):
 
             # dive-level coloring
@@ -305,7 +310,7 @@ class Visualisor:
                     plt.ylim([-0.05,1.05])
                     plt.yticks([0,0.5,1.0],fontsize=30)
                 else:
[…]

### 2020-07-22 Report/Sections/appendix.tex
Updates from Overleaf

@@
 \renewcommand*{\thesubsection}{\Alph{subsection}}
+
+Michelot \& Blackwell (2019) model the movement of an animal as the solution to the following stochastic differential equation, which they refer to as a state-switching Ornstein-Uhlenbeck process:
+%
+\begin{equation}
+    \label{eqn:SDE}
+    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
+\end{equation}
+%
+where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
+Unlike HMMs, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer.
+If the behavioral state $X_t$ is known and does not change between observations, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
+%
+\begin{align}
+    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
+    \label{eqn:OU_sol}
+\end{align}
+%
+where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$
+
+\begin{theorem}{Theorem 1.}{}%
+If the following conditions are met:
+\begin{enumerate}
+    \item The hidden behavioural process $X$ from Equation \ref{eqn:SDE} follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$,
+
+    \item Observations of the SDE from Equation 2 are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
+\end{enumerate}
+then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
+
+\begin{align}
+\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{(i)} = \sqrt{\frac{\omega^{(i)^2}}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t})}, \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
+\end{align}
+
+\end{theorem}
 
@@
 We pick $\tilde{\omega} = 10$ periods per window, or 5 hertz.
-To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use Equation (\ref{eqn:y_hat0}):
+To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use Equation \ref{eqn:y_hat0}:
 %

### 2020-07-22 Report/Sections/discussion.tex
Updates from Overleaf

@@
 
-We presented a collection of HMM models which can be combined together to form increasingly complex hierarchical models to match the complexity of particular problems faced by researchers.
-This flexible framework can be used to deal with complicated dependence structures within time-series data.
-
 Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics when conditioned of the hidden state.
-However, many real-world processes are more complicated than this and require more complex models.
+However, many real-world processes require more complex models.
+Here, we presented a flexible framework that can account for complicated dependence structures within time-series data.
+Our framework is a collection of HMM models which can be combined together to form increasingly complex hierarchical models to match the complexity of particular problems faced by researchers.
 
@@
 Although the CarHMM can incorporate auto-correlation into the structure of an HMM, it can break down when observations are taken at irregular time intervals.
-A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by Michelot \& Blackwell (2019).
+A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by \citet{Michelot:2019}.
 Most continuous time models require relatively slow MCMC algorithms to perform inference, and as a result are not easily incorporated into the HHMM structure.
-However, we show in Theorem 1 that certain continuous-time methods are equivalent to an CarHMM under certain conditions (see appendix for proof).
+However, we show in Theorem 1 (see appendix) that the state-switching OU process is equivalent to an CarHMM under certain conditions.
+This implies that the parameter estimates from the CarHMM are interpretable in the context of a stochastic differential equation.
 
@@
 In particular, each hidden state of the coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.
+\citet{Adam:2019} utilize the HHMM to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California.
+However, they run into clear violations of conditional independence and apparent periodic behavior when plotting the auto-correlation of acceleration as a function of lag, implying the need to account for complicated dependence structures within fine-scale processes with high sampling frequencies.
 
-For processes with very high sampling frequencies and/or with intricate fine-scale structure, it is possible to generalize the HHMM such that the fine-scale model can be any model which admits an easy-to-calculate likelihood.
-For example, if the sampling rate of the fine-scale process is very high, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.
+%For processes such as these with very high sampling frequencies and/or intricate fine-scale structure, 
+To deal with these issues, it is possible to generalize the HHMM such that the fine-scale model can be any model which admits an easy-to-calculate likelihood.
+For example, if the sampling rate of the fine-scale process is very high and clearly period, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.
 
-Combining these models together should be done with care, as it is important to balance the need to effectively capture the process in question with the need to avoiding over-fitting and slow parameter estimation.
+In particular, we use $Z^{*(1)}$ and $Z^{*(2)}$ based on the DFT as defined previously.
+These summary statistics prove to be vital when decoding the behavior of a killer whale.
+This is evidenced by both our simulation study, where the sub-dive decoding accuracy is nearly perfect when $Z^{*(2)}$ is accounted for, and the case study, where the CarHHMM without DFT visually fails to pick up fluking behavior in the killer whale.
+
+While we combine these models together effectively for our case study, in general this practice should be done with care.
+It is important to balance the need to effectively capture the process in question with the need to avoid over-fitting and slow parameter estimation.
 
@@
 The simulation study also shows that the observed Fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases.
-One notable exception is that the standard errors of the probability transition matrix estimates ($\hat \Gamma$ and $\hat \Gamma^*$) tend to be overestimated by the observed Fisher information.
+One notable exception is the HHMM-DFT, which ignores auto-correlation and therefore underestimates the standard error of the parameters associated with acceleration.
+This can be dangerous in the field of ecology, as overconfident parameter estimates can result in overconfident conservation decisions.
+While the point estimates are generally good in this case study, this may not be the case in more general settings and for shorter time series in particular.
 
-Finally, we used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada.
+%We used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada.
 The CarHHMM-DFT was able to simultaneously distinguish three distinct sub-dive behaviors and two dive types.
@@
 There does appear to be some heterogeneity within dive types, and future work can be done to determine the optimal number of dive types and within-dive behaviors.
+
+In total, this work provides a flexible framework to model functional time-series data using HMM models which are easy to fit using maximum likelihood methods.
+The model of interest can be easily adjusted to match the complexity of the problem at hand.
+Researchers can use this method to build model and perform inference in any setting which involves state-switching fine-scale processes, including finance and medicine.

### 2020-07-22 Report/Sections/intro.tex
Updates from Overleaf

@@
 Functional Data Analysis (FDA) techniques lie somewhere in between these extremes: the output of an analysis provides visual tools for understanding processes and, often, some dimension reduction to allow for easier exploration.
-The development of structured FDA techniques incorporating more complex dependence structures has become a topic of recent interest  (see, e.g., Crainiceanu. and references therein).
+The development of structured FDA techniques which incorporate more complex dependence structures has become a topic of recent interest  (see, e.g., Crainiceanu. and references therein).
 
@@
 We enforce temporal structure via Markov processes.  
-While our methodology has general applicability, [rephrase - awkward] we  showcase our ideas to study the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
+While our methodology is generally applicable, [rephrase - awkward] we  showcase our ideas to study the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
 For more information about the data, see Section \ref{sec:case_study}.
@@
 with axes oriented relative to the whale's forward movement.
-Here, we see spikes in acceleration and, in contrast, periods of fairly constant acceleration.
+Here, we see spikes in acceleration which contrast with periods of fairly constant acceleration.
 Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
-The plot shows the magnitude of the acceleration and from this, we can see fine-scale behaviour within each dive which seems to indicate some fine scale short-term periodicity. 
+The plot shows one component of the acceleration, and from this we can see fine-scale behaviour within each dive which seems to indicate some fine scale short-term periodicity. 
 We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour. 
@@
 However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
-We model the within curve dependence by combining a hidden Markov model of within-curve behaviour combined with an autoregressive-type dependence.
+We model the within-curve dependence by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence.
 Finally, for the fine-scale behaviour, we use summary information that reflects the application area's meaning for the curve.
 Specifically, for the dive data, we use summaries of Fourier transforms to account for sudden and extreme ``wiggliness'' within some segments of a curve.
-The distribution of this ``wiggliness'' depends on the hiddent state.
+The distribution of this ``wiggliness'' depends on the hidden state.
 
@@
 %
-%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another conditioned on the underlying behavioral process.
+%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another when conditioned on the underlying behavioral process.
 %However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
 %For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
- Several publications have dealt with this issue in the past, including the hidden movement Markov model (HMMM) \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model (CarHMM) \citep{Lawler:2019}.
- The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
+%Several publications have dealt with this issue in the past, including the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
+%The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
 %while maintaining the structure needed to run the forward algorithm. 
@@
 Animal movement modelers have recognized that several simultaneous behavioral processes may occur at different time scales. 
-One approach to the analysis of this type of data is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017} \citep{Adam:2019}. 
-HHMMs model the entire time series in question as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.
+One approach to the analysis of this type of data is to use a hierarchical hidden Markov model, or HHMM \citep{Barajas:2017,Adam:2019}. 
+HHMMs model the entire time series in question as a nested structure of hidden Markov models where each HMM corresponds to one behavioral process.
 
@@
 Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}.
-One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation.
+%One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation.
 Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals.
 However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
-
-For periodic behavior in particular, one way to avoid the use of continuous time models is to use signal processing techniques such as the Fourier transform on the raw data. 
+One way to model periodic behavior in particular is to use signal processing techniques such as the Fourier transform on the raw data. 
 The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}.
@@
 Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears a promising simple approach to account for additional correlation in data that is cyclical in nature.
-
-We consider several models: the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM). 
-The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes. 
-The CarHMM modifies the usual HMM by modelling additional dependence into the sequence of observed data. 
-In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.
 

### 2020-07-22 Report/Sections/models.tex

@@
 
-Traditional hidden Markov models (HMMs) are useful tools to \textit{blah blah blah - quick (5-7words) restatement of the main advantage of HMM}, and will be used as the core structure of our models.
-However, traditional HMMs assume conditional independence between observations given the state sequence, and thus do not hold when the observations exhibit certain forms of significant correlation in time.
-Traditional HMMs model process autocorrelation with a single Markov chain, however many processes have temporal dependencies acting at different scales.
-\textit{blah blah blah -quick restatements (5-15 words to potentially many sentences) of the main issues you are going to fix made a quick try at it, work on smoothing things.
+Hidden Markov models are useful tools to model state-switching Markovian processes in discrete time and will be used as the core structure of our models.
+Traditional HMMs model process auto-correlation by assuming that the underlying behavior of a processes follows a single Markov chain.
+However, when conditioned on the hidden state sequence observations are assumed to be independent.
+Therefore, classical HMMs do not hold when the observations exhibit other forms of significant correlation in time.
+Examples of this include continuous-time processes such as a Wiener process, systems which exhibit temporal dependencies acting at several different time scales \citep{Barajas:2017}, and periodic behaviour including the motion of an ideal spring.
+%\textit{blah blah blah -quick restatements (5-15 words to potentially many sentences) of the main issues you are going to fix made a quick try at it, work on smoothing things.
 You can use some of the text I removed from below}.
-To account for each of these common additional dependence structures, we will explore three variations on the traditional HMM: the conditionally auto-regressive hidden Markov model (CarHMM), hierarchical hidden Markov model (HHMM), and HMM with discrete Fourier transform (HMM-DFT).
+To account for each of these common additional dependence structures, we will explore three variations on the traditional HMM.
+First, we explain the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
+Next, we review the hierarchical hidden Markov model (HHMM), which models simultaneous temporal dependencies at different time scales as two distinct hidden Markov models in a hierarchical model structure.
+Finally, we describe an HMM over a discrete Fourier transform (HMM-DFT), which applies the DFT to a moving window over the raw data to summarize its periodic behavior as a function of time.
 We will show how each of these variations can be altered and combined to form a wide variety of new models.
@@
 These conditional distributions are governed by state-dependent parameters.
-If $X_t = i$, then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes  $f^{(i)}(\cdot)$.
+If $X_t = i$, then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes simply $f^{(i)}(\cdot)$.
 %
-Figure \ref{fig:HMM} represents the dependence structure of an HMM.
+Figure \ref{fig:models} represents the dependence structure of an HMM.
 
@@
 
-Following \citet{Barajas:2017}, we parameterize the $N \times N$ transition probability matrix $\Gamma$ in a way that forces the entries of the matrix to be non-negative and for the rows to sum to 1:
+Following \citet{Barajas:2017}, we parameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows to sum to 1:
 %
@@
 where $\eta \in \bbR^{N \times N}$ and $\eta_{ii}$ is set to zero for identifiability.
-Then $\calL_{\text{HMM}}(y;\Theta,\Gamma)$ can be maximized using a wide range of optimizer.
+Then $\calL_{\text{HMM}}(y;\Theta,\Gamma)$ can be maximized using a wide range of optimizers.
 For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of  $\eta$.
@@
 
-The CarHMM, introduced by Lawer et al.
-(2019), explicitly models auto-correlation in observation within an HMM.
+The CarHMM, introduced by \citet{Lawler:2019}, explicitly models auto-correlation in the observations of an HMM.
 Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
 Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$, depends on \textit{both} $X_t$ \textit{and} $Y_{t-1}$ rather than only $X_t$.
-The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. Figure \ref{fig:CarHMM} shows the dependence structure of a CarHMM.
+The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. Figure \ref{fig:models} shows the dependence structure of a CarHMM.
 
@@
 
-While CarHMMs can model auto-correlation within the observation sequence, they fail when observations are taken at irregular time intervals.
-Another way to model a stochastic process is to use a stochastic differential equation (SDE).
-As an example, Michelot \& Blackwell (2019) model the movement of an animal as the solution to the following stochastic differential equation, which they refer to as a state-switching Ornstein-Uhlenbeck process:
-%
-\begin{equation}
-    \label{eqn:SDE}
-    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
-\end{equation}
-%
-where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
-Unlike HMMs, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer.
-If the behavioral state $X_t$ is known and does not change between observations, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
-%
-\begin{align}
-    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
-    \label{eqn:OU_sol}
-\end{align}
-%
-where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$.
-Most continuous time models are difficult to incorporate into an HHMM and require MCMC methods to fit.
-However, under certain conditions, the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process.
-This gives new interpretation to the learned parameters of the CarHMM in the context of a continuous-time model.
+\subsection{Incorporating correlation on multiple time scales with an HHMM}
 
-\begin{theorem}{Theorem 1.}{}%
-If the following conditions are met:
-\begin{enumerate}
-    \item The hidden behavioural process $X$ from Equation \ref{eqn:SDE} follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$,
-
-    \item Observations of the SDE from Equation 2 are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
-\end{enumerate}
-then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
-
-\begin{align}
-\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{(i)} = \sqrt{\frac{\omega^{(i)^2}}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t})}, \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
-\end{align}
-
-\end{theorem}
-
-See the appendix for a proof of Theorem 1.
-
-\subsection{Incorporating different scales of correlations with HHMM}
-
-HHMM accounts for different levels of correlation by modeling a coarse-scale process and a fine-scale process, each of which with an HMM \textit{CITATION(s)}.
+An HHMM accounts for different levels of correlation by modelling a coarse-scale process and a fine-scale process, each of which with an HMM \citep{Barajas:2017,Adam:2019}.
 The coarse-scale process is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observed responses.
@@
 
-The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.We denote the fine-scale emission parameter vector corresponding to $X_t=i$ with $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$.
+The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
+We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$.
 
@@
 $\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$).
-Figure \ref{fig:HHMM} represents the dependence structure for an HHMM.
+Figure \ref{fig:models} represents the dependence structure for an HHMM.
 
@@
 %
-Let $\Theta^* \equiv \{\Theta^{*(1)}, \ldots, \Theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and let $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices.
+In addition, let $\Theta^* \equiv \{\Theta^{*(1)}, \ldots, \Theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices.
 The likelihood of the observed data is then
@@
 %
-where $P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $ii$th entry corresponding to $X_t=i$ and equal to 
+where $P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to 
 $f^{(i)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(i)},
@@
 
-For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see Adam et al.
-(2019).
+For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citet{Adam:2019}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Accounting for periodicity with HMM-DFT}
+\subsection{Accounting for periodicity with the HMM-DFT}
 \label{subsec:STFT}
 
-The HMM-DFT incorporates hierarchical structure into an HMM differently than an HHMM (\textit{ref Fig 1}).
+The HMM-DFT incorporates hierarchical structure into an HMM differently than an HHMM (Fig \ref{fig:models}).
 In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform.
@@
 These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$.
-Figure \ref{fig:HMM-DFT} represents the structure of the HMM-DFT.
+Figure \ref{fig:models} represents the structure of the HMM-DFT.
 
@@
 
-It is possible to accommodate unequal time steps within $y_t^*$ by using the non-uniform discrete Fourier transform (NDFT) \citep{Bagchi:1999}.
+It is possible to accommodate unequal time steps within $y_t^*$ by using the non-uniform discrete Fourier transform (NDFT).
 We do not describe the method in detail here, but the generalization is straightforward.
+See \citet{Bagchi:1999} for details.
 
-\subsection{General structure for building complexity}
+\subsection{General structure for building complex models}
 
 In addition to the models described above, the fine-scale process $Y^*_t$ can be modeled using \textit{any} parametric model which admits an easy-to-compute likelihood.
-The fine-scale likelihood $\calL_{\text{HMM}}$ from the HHMM likelihood is then replaced by the likelihood of the general fine-scale model, $\calL_{\text{fine}}(\mathbf{y}^*_t;\Theta^{*(i)})$:
+The fine-scale likelihood $\calL_{\text{HMM}}$ from the HHMM likelihood can therefore be replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(y^*_t;\Theta^{*(i)})$:
 \[
@@
 
-Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, state-switching OU, HHMM, and HMM-DFT).
+Possible candidates for the fine-scale model include, but are not limited to, any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HMM-DFT).
 These alternative models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models based on HMMs.

### 2020-07-22 Report/Sections/results.tex

@@
 In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed dives within this time range were ignored as well.
-A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
+A killer whale ``dive" is considered to be any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
 Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second.
@@
 
-The coarse-scale observations were made up of the collection of dive durations in seconds, and the fine-scale observations were determined from the within-dive acceleration data.
+We used a hierarchical HHM, where the coarse-scale observations were a time series of dive durations in seconds, and the fine-scale observations were the within-dive acceleration data (Figure \ref{fig:CarHHMM-DFT}).
+The coarse-scale level is comprised of an HMM with hidden states corresponding to dive types.
 The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
+$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)},$$
+$$\mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
 
-$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}$$
-$$\mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2$$
-
-The acceleration exhibits significant sinusoidal behavior, so the fine-scale observations $Z^*$ were made up of the DFT summary statistics of a two-second sliding window.
-Unlike the simulation study, the acceleration observations here are 3-dimensional vectors rather than scalars, so the observations $z^*$ were calculated as follows:
+The fine-scale level model subdive behavior and is comprised of a CarHMM-DFT where auto-correlation is modeled into the distribution of $\mathbf{Z}^{*(1)}$ (the average acceleration with a 2-second window), but not $Z^{*(2)}$ (the ``wiggliness" of the two second window).
+The acceleration exhibits significant sinusoidal behavior, thus the fine-scale observations $Z^*$ were made up of the DFT summary statistics of a two-second sliding window.
+Unlike the simulation study, where we had one time series of acceleration data, we have here acceleration data in three dimensions.
+These three dimensions represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags (Citation?).
+The following two sets of observations $z^*$ were calculated for modeling:
 %
 $$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2.$$
-%
 The observations are therefore made up of $\mathbf{z}_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $z_t^{*(2)}$, a scalar.
 We calculate $z_t^{*(2)}$ by summing the first 10 Fourier modes, and corresponds to a maximum recorded frequency of $\tilde \omega = 5$ Hz.
-There is strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions (see the supplementary material for a lag plot).
-Therefore, auto-correlation was directly modeled into the the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$, which is assumed to be Normally distributed with the following parameters:
+
+There is strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions (see the supplementary material for a lag plot), thus auto-correlation was directly modeled into the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$ using a CarHMM.
+We assumed the error to be Normally distributed with the following parameters:
 %
@@
 The final model is nearly identical to the one from the simulation study, with the exception that the fine-scale Markov chain has three sub-dive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a 3-dimensional vector rather than a scalar.
-The model is comprised of two levels: a coarse-scale HMM and a fine-scale CarHMM-DFT.
-The coarse-scale model is comprised of an HMM with hidden states corresponding to dive types and observations corresponding to dive durations.
-The fine-scale model is comprised of a CarHMM-DFT where auto-correlation is modeled into the distribution of $\mathbf{Z}^{*(1)}$ (the average acceleration with a 2-second window), but not $Z^{*(2)}$ (the ``wiggliness" of the two second window).
-Refer to Figure \ref{fig:CarHHMM-DFT} for a full graphical representation of this model.
 
@@
 
-Table \ref{table:emis_dists} displays estimates of the emission distribution parameters for this case study.
-The fitted emission distributions are also plotted in Figures \ref{fig:coarse_emis} and \ref{fig:fine_emis}.
-While the ecological meaning behind behavioral states is tenuous for HMMs, we hypothesize the following interpretations.
-Dive type 1 corresponds to shorter, shallower dives which serve a variety of purposes, including as rest before dives of type 2, which are deeper and more sustained.
-Sub-dive behavioural state 1 corresponds to gliding and less overall activity compared to the other behavioral states.
+The estimates for the emission distribution parameters suggest that this whale has two distinct dive behaviors (Table \ref{table:emis_dists},  figure \ref{fig:coarse_emis}).
+Dive type 1 corresponds to shorter, shallower dives.
+Ecologists have attributed a variety of purposes to such short dives, including resting (CITATION, you can ask Ian and Sarah for help here - look at the Tennessen paper with HMM).
+Dive type 2 is longer and deeper.
+These types of deep sustained dives have been associated with foraging behaviors in killer whales (CItation Tennessen maybe HMM and the other Tennesen with foraging data).
+
+The three subdive behaviors also appeared distinct, particularly with respect to the fourier sum (Table \ref{table:emis_dists},  Figure  \ref{fig:fine_emis}).
+Sub-dive type 1 corresponds to gliding and less overall activity compared to the other behavioral states.
 The mean of $Z^{*(2)}$ in this state is at least an order of magnitude smaller than sub-dive behavioral state 2, the variance of $\mathbf{Z}^{*(1)}$ is smaller than sub-dive behavior 2 for every component, and the auto-correlation of $\mathbf{Z}^{*(1)}$ is higher than every other behavioral state.
@@
 %
-About 86\% of observed dives are short; the whale usually rests for many dives in a row before performing a deep dive.
-The probability transition matrix $\hat \Gamma$ shows that the probability of a particular dive type does not depend much on the previous dive type.
-This killer whale is much more likely to be in a less active sub-dive state when performing deep dives than when performing shallow dives:  sub-dive state 1 makes up 14\% of shallow dives, but it makes up 48\% of deep dives.
-Using less active sub-dive states when diving deep could be an energy reduction strategy for these long periods of holding breath.
-This phenomenon has been observed in bottle-nose dolphins \citep{Williams:1999} and may also be the case for killer whales.
+The coarse-scale stationary distribution indicates that most dives are short type 1 dives.
+The coarse-scale transition probability matrix further indicates that the whale makes multiple short type 1 dives before doing only one (or a few) long type 2 dive.
+The fine-scale stationary distributions indicate that this killer whale is more likely to be in the less active sub-dive type 1 when performing long deep dives than when performing short shallow dives.
+Using less active swimming behavior when diving, and holding breath, for long periods of time has been suggested to be an energy reduction strategy in bottle-nose dolphins \citep{Williams:1999}.
+Such strategy could explain why sub-dive state 1 makes up 14\% of shallow dives of our killer whale, but it makes up 48\% of deep dives.
 Figure \ref{fig:labeled_dives} shows the decoded dive behavior of 5 selected dives.
@@
 $Z^{*(2)}$ is an exception, as its pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
-This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution compared to a gamma distribution such as a power law.
+This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution than the gamma distribution.
 

### 2020-07-22 Report/Sections/simulation.tex

@@
 %
-Once the dive durations were calculated for all 100 dives, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments (the end of the dive sequence was discarded) which made up a second fine-scale hidden Markov model.
-Each two second segment was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$.
+Once the dive durations were calculated for all 100 dives, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments, which made up a second fine-scale hidden Markov model.
+The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive were discarded.
+Each two-second segment was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$.
 The probability transition matrices for these fine-scale Markov chains were set as
 %
-$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.9 & 0.1 \end{pmatrix}, \qquad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix},$$ 
+$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix}, \qquad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix},$$ 
 %
@@
 %
-The distribution of $Z^{*(2)}_{t,t^*}$ is gamma and its parameters depend upon only $X^*_{t,t^*}$ (not $X_t$):
+The distribution of $Z^{*(2)}_{t,t^*}$ is gamma and its parameters depend upon only $X^*_{t,t^*}$ (not $Z^{*(2)}_{t,t^*-1}$):
 %
@@
     %
-    \Theta^{*(i^*)} =  \{\{\mu_1^{*(i^*)},\sigma_1^{*(i^*)},\phi_1^{*(i^*)}\},\{\mu_2^{*(i^*)},\sigma_2^{*(i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
+    \Theta^{*(i)} =  \{\{\mu_1^{*(i)},\sigma_1^{*(i)},\phi_1^{*(i)}\},\{\mu_2^{*(i)},\sigma_2^{*(i)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
 \end{gather*}
@@
     \item A \textbf{CarHHMM-DFT} as described above.
-    \item An \textbf{HHMM-DFT}, which is similar to the model above, but with no modeled auto-correlation, i.e. $\phi_1^{*(i^*)} = 0$ for $i^* = 1,2$.
-    \item A \textbf{CarHHMM}, which is similar to the model above, but without $Z^{*(2)}$ as an observation (i.e $\Theta^{*(i^*)} = \{\mu_1^{(i^*)},\sigma_1^{(i^*)},\phi_1^{(i^*)}\}$, $i^* = 1,2$).
-    \item A \textbf{CarHMM-DFT}, which is similar to the CarHHMM-DFT, but with $N=1$ instead of $N=2$ on the coarse scale.
+    \item An \textbf{HHMM-DFT}, which is similar to the CarHHMM-DFT, but with no modeled auto-correlation \big(i.e. $\phi_1^{*(i)} = 0$ for $i = 1,2$\big).
+    \item A \textbf{CarHHMM}, which is also similar to the CarHHMM-DFT, but without the Fourier sums $(Z^{*(2)})$ as an observation \big(i.e $\Theta^{*(i)} = \{\mu_1^{(i)},\sigma_1^{(i)},\phi_1^{(i)}\}$, $i = 1,2$\big).
+    \item A \textbf{CarHMM-DFT}, which is again similar to the CarHHMM-DFT, but with $N=1$ instead of $N=2$ on the coarse scale.
 This is equivalent to loosing the coarse level of the hierarchical structure, as there is only one dive type.
@@
 
+All models were run on the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
+
 \subsection{Simulation Results}
 
-All models were run on the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
+%%%
+
+As expected, the full CarHHMM-DFT was the best-performing model.
+It was able to decode the fine-scale hidden states of the process almost perfectly (mean error $< 10^{-12}$ for all simulations) and decode the hidden dive types with an average accuracy of $94 \pm 3 \%$, where $\pm$ refers to the standard deviation across simulations (see Table \ref{table:accuracy}).
+All parameter estimates have biases which are at least approximately one order of magnitude smaller than the associated standard error, which is consistent with a sample size of 500 (section S-1.2, tables 2-5).
+One exception is for estimates of $\hat \sigma$, which were regularly underestimated.
+For dive duration $(Y_t)$ this finding is consistent with properties of MLEs for standard deviation, especially with a small sample size.
+For acceleration $(Z^{*(1)})$ this is likely due to over-fitting the auto-correlation parameter $\phi$.
+The Fourier sums $(Z^{*(2)})$ show no significant bias in $\hat \sigma^*_2$.
+The empirical standard error of all parameters estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) were well-approximated by the inverse of the observed fisher information matrix (section S-1.2, tables 2-5).
+One exception is $\hat \Theta$, whose standard errors were slightly underestimated using the observed fisher information matrix.
+%The empirical standard error of $\hat \mu_2$ was 8.5, while the estimate of standard error using the observed Fisher information was 5.8. 
+See Section S-1.2, tables 2-5 for more information.
+
+The HHMM-DFT performed almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias. 
+%- it had very low error for the fine-scale hidden states and $\approx 94\%$ accuracy on the dive types (see Table \ref{table:accuracy}).
+In addition, the standard errors of the parameter estimates were comparable to those of the CarHHMM-DFT.
+The estimates of standard error using the observed Fisher information were also comparable to the CarHHMM-DFT.
+The HHMM-DFT \textit{did} differ from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$.
+For both sub-dive states, $\hat \sigma_1^*$ greatly overestimated the true standard deviation.
+In particular, $\bar \hat \sigma_1^{*(1)} = 0.248$ and $\bar \hat \sigma_1^{*(2)} = 0.242$, but $\sigma_1^{*(1)} = 0.05$ and $\sigma_1^{*(2)} = 0.1$.
+in addition, the empirical standard error of both $\hat \mu_1^*$ and $\hat \sigma_1^*$ was much larger than the estimates of standard error using the observed fisher information (see Section S-1.2, Table 3).
+This implies that parameter uncertainty can be underestimated when auto-correlation is ignored.
+
+The CarHHMM was the worst-performing model in terms of decoding accuracy, with an accuracy of $89 \pm 2\%$ when determining sub-dive behavior and an accuracy of $91 \pm 3\%$ when determining dive types.
+It was much more likely to misclassify sub-dive behavior 1 than sub-dive behavior 2 (see Section S-1.2, Table 1).
+It was also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy.
+$\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ in particular were poor estimates of the true parameters (see Section S-1.1.3).
+
+Finally, the CarHMM-DFT was comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations.
+In addition, the time required to fit the CarHHMM was approximately 1/5 that of the other models.
+However, it lacked any hierarchical structure, so it was unable to decode the coarse-scale dive types.
+The CarHMM-DFT still did fit a gamma distribution to the dive duration of \textit{all} dives.
+The parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ is a significant underestimate by $\approx$ 5 seconds.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% EVERYTHING BELOW HERE IS OLD AND I THINK IT IS BAD. THE STUFF ABOVE IS MUCH BETTER.
+
+\iffalse
 %%% Accuracy %%%
-Each model was able to decode the fine-scale hidden states of the process almost perfectly except for the CarHHMM, whose accuracy was 89\%.
-The lower accuracy of the CarHHMM can be attributed to the higher fine-scale states variation in the distribution of $Z^{*(2)}$ compared to $Z^{*(1)}$.
-For the coarse-scale hidden states, the CarHMM-DFT lacked a hierarchical structure and could not make any predictions at all.
-The other three models all achieved an accuracy of approximately 90\%, with the CarHHMM slightly more likely to categorize a dive as dive type 2 than the other models.
+Each model was able to decode the fine-scale hidden states of the process almost perfectly except for the CarHHMM, whose accuracy was 89\% (Table \ref{table:accuracy}).
+The lower accuracy of the CarHHMM (i.e.
+model without $Z^{*(2)}$)  can be attributed to the higher fine-scale states variation in the distribution of $Z^{*(2)}$ compared to $Z^{*(1)}$.
+For the coarse-scale hidden states, the CarHMM-DFT lacked a hierarchical structure and could not make coarse-scale predictions.
+All three other models achieved an accuracy of approximately 90\% (Table \ref{table:accuracy}), with the CarHHMM slightly more likely to categorize a dive as dive type 2 than the other models (Fig ?
+or Sup Table ?
+that give quantitative info on this).
 Figure \ref{fig:acc} shows the decoded state probabilities for both the fine- and coarse- scales for one of the simulation run.
-Table \ref{table:accuracy} lists the accuracy and training times of each model.
 
@@
 For the emission distributions of dive duration, Figure \ref{fig:MLE_dist} shows the empirical joint density of $\hat \mu$ and $\hat \sigma$ for the CarHHMM-DFT.
-The supplementary material shows the empirical joint densities for all other models and features.
 
@@
 %because it sees the ``wigliness" of the acceleration data as raw acceleration rather than as Fourier modes, which deflates the value of $\hat \phi$ and inflates the value of $\hat \sigma$.
-See the supplementary material for a detailed breakdown of parameter values for acceleration emission distributions.
+
+The supplementary material provides a detailed breakdown of parameter values for acceleration emission distributions and the empirical joint densities for all models and features.
 
@@
 See the supplementary material for a full list of estimates and standard errors.
+
+\fi

### 2020-07-22 Code/Visualisor.py
Update Plots

@@ -118,7 +118,7 @@ class Visualisor:
             nrows = 1
             ncols = len(self.hhmm.theta[0])
         else:
-            features = list(self.hhmm.theta[1][0].keys())
+            features = ['Ax','Ay','Az','Ahat_low']#list(self.hhmm.theta[1][0].keys())
             nrows = 2 #self.pars.K[0]
             ncols = int((len(features)+1)/2)
 

### 2020-07-22 Report/Sections/appendix.tex
Updates from Overleaf

@@
 \begin{enumerate}
-    \item The hidden behavioural process $X$ from Equation \ref{eqn:SDE} follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$,
+    \item The hidden behavioural process $X$ from Equation (\ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$,
 
@@
 \begin{proof}{Proof of Theorem 1}{}
-Combining equation (\ref{eqn:CarHMM_to_OU}) with equation (\ref{eqn:OU_sol}) gives:
+Combining Equation (\ref{eqn:OU_sol}) with Equation (\ref{eqn:CarHMM_to_OU}) gives:
 %
@@
 We pick $\tilde{\omega} = 10$ periods per window, or 5 hertz.
-To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use Equation \ref{eqn:y_hat0}:
+To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use Equation (\ref{eqn:y_hat0}):
 %
-$$\left(Z^{*(1)}_{t,t^*} | X^*_{t,t^*} = i^* \right) = \left(\mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right) | X^*_{t,t^*} = i^* \right) \sim \mathcal{N} \left(\phi^{*(i^*)} * Z^{*(1)}_{t,t^*-1}, \sigma^{*(i^*)} \right)$$
+$$\left(Z^{*(1)}_{t,t^*} | X^*_{t,t^*} = i^* \right) = \left(\mathcal{R}\left(\hat{Y}^{*(0)}_{t,t^*}\right) | X^*_{t,t^*} = i^* \right) \sim \mathcal{N} \left(\phi^{*(i^*)} * Z^{*(1)}_{t,t^*-1}, \sigma^{*(i^*)} \right)$$
 %
@@
 $$Z^{*(2)}_{t,t^*} = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
+%
 $$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = \sum_{k=1}^{10} 5/k^2 = 10.10, \beta = 1.00 \right)$$
-$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = 300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , \beta = 1.00 \right)$$
+%
+$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(\alpha = 300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , \beta = 1.00 \right)$$
 \end{proof}
@@
 %
-$f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$ is the emission distribution of $Z^*_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $Z^*_{t,t^*-1} = z^*_{t,t^*-1}$.
+Note that $f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$ is the emission distribution of $Z^*_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $Z^*_{t,t^*-1} = z^*_{t,t^*-1}$.
 

### 2020-07-22 Code/run_all_sims.sh
Update plots

@@ -8,7 +8,7 @@
 #SBATCH --mail-type=REQUEUE
 #SBATCH --mail-type=ALL
 
-#SBATCH --time=48:00:00
+#SBATCH --time=96:00:00
 #SBATCH --mem-per-cpu=4G
 #SBATCH --array=1-500
 

### 2020-07-22 Report/Sections/simulation.tex
Updates from Overleaf

@@
 For both sub-dive states, $\hat \sigma_1^*$ greatly overestimated the true standard deviation.
-In particular, $\bar \hat \sigma_1^{*(1)} = 0.248$ and $\bar \hat \sigma_1^{*(2)} = 0.242$, but $\sigma_1^{*(1)} = 0.05$ and $\sigma_1^{*(2)} = 0.1$.
-in addition, the empirical standard error of both $\hat \mu_1^*$ and $\hat \sigma_1^*$ was much larger than the estimates of standard error using the observed fisher information (see Section S-1.2, Table 3).
+In particular, $\overline{\hat \sigma_1^{*(1)}} = 0.248$ and $\overline{\hat \sigma_1^{*(2)}} = 0.242$, but $\sigma_1^{*(1)} = 0.05$ and $\sigma_1^{*(2)} = 0.1$.
+In addition, the empirical standard error of both $\hat \mu_1^*$ and $\hat \sigma_1^*$ was much larger than the estimates of standard error using the observed fisher information (see Section S-1.2, Table 3).
 This implies that parameter uncertainty can be underestimated when auto-correlation is ignored.
 
-The CarHHMM was the worst-performing model in terms of decoding accuracy, with an accuracy of $89 \pm 2\%$ when determining sub-dive behavior and an accuracy of $91 \pm 3\%$ when determining dive types.
+The CarHHMM was the worst-performing model in terms of decoding accuracy, with an accuracy of $89 \pm 1\%$ when determining sub-dive behavior and an accuracy of $87 \pm 11\%$ when determining dive types.
 It was much more likely to misclassify sub-dive behavior 1 than sub-dive behavior 2 (see Section S-1.2, Table 1).
@@
 Finally, the CarHMM-DFT was comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations.
-In addition, the time required to fit the CarHHMM was approximately 1/5 that of the other models.
+In addition, the time required to fit the CarHHMM was less than half of that of the other models (Tbl \ref{table:accuracy}).
 However, it lacked any hierarchical structure, so it was unable to decode the coarse-scale dive types.
 The CarHMM-DFT still did fit a gamma distribution to the dive duration of \textit{all} dives.
-The parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ is a significant underestimate by $\approx$ 5 seconds.
+The parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds.
 

### 2020-07-22 Report/Sections/discussion.tex

@@
 An example of balancing model complexity with efficient fitting is presented in our simulation study.
-Sub-dive behavioral states were shared across dive types to reduce model complexity, and by far the fastest model to train ($\approx$ 15 minutes) was the CarHMM-DFT, which had no hierarchical component.
+Sub-dive behavioral states were shared across dive types to reduce model complexity, and by far the fastest model to train was the CarHMM-DFT, which had no hierarchical component.
 Even still, the CarHMM-DFT had near-perfect accuracy when decoding sub-dive behavioral states of the simulated whale.

### 2020-07-22 Code/Visualisor.py
Update Plots

@@ -251,7 +251,7 @@ class Visualisor:
         return
 
 
-    def plot_dive_features(self,sdive,edive,df_cols,data_cols,file=None):
+    def plot_dive_features(self,sdive,edive,df_cols,data_cols,file=None,ylabs=None):
 
         df = self.df
         data = self.data
@@ -263,8 +263,8 @@ class Visualisor:
             else:
                 nrows += 2
 
-        #nrows=5
-        nrows=4
+        nrows=5
+        #nrows=4
 
         plt.subplots(nrows,1,figsize=(20,5*nrows))
         fignum = 1
@@ -282,13 +282,14 @@ class Visualisor:
         for state in range(self.pars.K[0]):
             dives.append(dive[dive['ML_dive'] == state])
 
-        #ylabs = [r'$P\left(X_t = 1\right)$',
-        #         r'$P\left(X_t = 2\right)$',
-        #         r'$P\left(X^*_{t,t^*} = 1\right)$',
-        #         r'$P\left(X^*_{t,t^*} = 2\right)$',
-        #         r'$P\left(X^*_{t,t^*} = 3\right)$']
-        ylabs = [r'$\left(Z^{*(1)}\right)_x$ $(m/s^2)$',
-                 r'Depth $(m)$']
+        if ylabs is None:
+            ylabs = [r'$P\left(X_t = 1\right)$',
+                     r'$P\left(X_t = 2\right)$',
+                     r'$P\left(X^*_{t,t^*} = 1\right)$',
+                     r'$P\left(X^*_{t,t^*} = 2\right)$',
+                     r'$P\left(X^*_{t,t^*} = 3\right)$']
+            #ylabs = [r'$\left(Z^{*(1)}\right)_x$ $(m/s^2)$',
+            #         r'Depth $(m)$']
 
         #title = 'Hidden State Probability Estimates'
         title = 'Decoded Dive Profile / Accelerometer Data'

### 2020-07-22 Code/Visualisor.py
Update Plots

@@ -251,20 +251,18 @@ class Visualisor:
         return
 
 
-    def plot_dive_features(self,sdive,edive,df_cols,data_cols,file=None,ylabs=None):
+    def plot_dive_features(self,sdive,edive,df_cols,data_cols,file=None,ylabs=None,nrows=None):
 
         df = self.df
         data = self.data
 
-        nrows = 2*len(df_cols)
-        for col in data_cols:
-            if col in self.pars.features[0]:
-                nrows += 1
-            else:
-                nrows += 2
-
-        nrows=5
-        #nrows=4
+        if nrows is None:
+            nrows = 2*len(df_cols)
+            for col in data_cols:
+                if col in self.pars.features[0]:
+                    nrows += 1
+                else:
+                    nrows += 2
 
         plt.subplots(nrows,1,figsize=(20,5*nrows))
         fignum = 1

### 2020-07-25 Report/Sections/discussion.tex
Commit before restructure!

@@
 These summary statistics prove to be vital when decoding the behavior of a killer whale.
-This is evidenced by both our simulation study, where the sub-dive decoding accuracy is nearly perfect when $Z^{*(2)}$ is accounted for, and the case study, where the CarHHMM without DFT visually fails to pick up fluking behavior in the killer whale.
+This is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when $Z^{*(2)}$ is accounted for, and the case study, where the CarHHMM without DFT visually fails to pick up fluking behavior in the killer whale.
 
@@
 An example of balancing model complexity with efficient fitting is presented in our simulation study.
-Sub-dive behavioral states were shared across dive types to reduce model complexity, and by far the fastest model to train was the CarHMM-DFT, which had no hierarchical component.
-Even still, the CarHMM-DFT had near-perfect accuracy when decoding sub-dive behavioral states of the simulated whale.
+subdive behavioral states were shared across dive types to reduce model complexity, and by far the fastest model to train was the CarHMM-DFT, which had no hierarchical component.
+Even still, the CarHMM-DFT had near-perfect accuracy when decoding subdive behavioral states of the simulated whale.
 However, this model is not sufficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.
@@
 %We used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada.
-The CarHHMM-DFT was able to simultaneously distinguish three distinct sub-dive behaviors and two dive types.
-The DFT component proved useful in determining the sub-dive behaviour of the whale, as the mean of the emission distribution of $Z^{*(2)}$ for each sub-dive state was separated by an order of magnitude.
-Finally, the estimated auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and sub-dive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data.
+The CarHHMM-DFT was able to simultaneously distinguish three distinct subdive behaviors and two dive types.
+The DFT component proved useful in determining the subdive behaviour of the whale, as the mean of the emission distribution of $Z^{*(2)}$ for each subdive state was separated by an order of magnitude.
+Finally, the estimated auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and subdive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data.
 The introduction of the parameter $\phi^*$ also allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).
 
-Because traditional information criteria tend to overestimate the number of states in biological processes \citep{Pohle:2017}, the number of dive types and sub-dive behaviors was selected in an ad-hoc manner.
+Because traditional information criteria tend to overestimate the number of states in biological processes \citep{Pohle:2017}, the number of dive types and subdive behaviors was selected in an ad-hoc manner.
 There does appear to be some heterogeneity within dive types, and future work can be done to determine the optimal number of dive types and within-dive behaviors.

### 2020-07-25 Report/Sections/intro.tex
Commit before restructure!

@@
 
-Many statistical techniques have been developed and applied to these types of longitudinal or functional data.  
-Some techniques, such as time series, are highly parametric. 
-While these parametric techniques do not allow the data to ``speak for itself", the ensuing data analysis results typically are readily interpretable, providing important subject area insights.
-At the other extreme are black box type methods of machine learning.  
-While these methods are incredibly successful at predicting how processes evolve, they provide little structured insight into the driving mechanisms of the process.
-Functional Data Analysis (FDA) techniques lie somewhere in between these extremes: the output of an analysis provides visual tools for understanding processes and, often, some dimension reduction to allow for easier exploration.
-The development of structured FDA techniques which incorporate more complex dependence structures has become a topic of recent interest  (see, e.g., Crainiceanu. and references therein).
+%Many statistical techniques have been developed and applied to these types of longitudinal or functional data.  
+%Some techniques, such as time series, are highly parametric. 
+%While these parametric techniques do not allow the data to ``speak for itself", the ensuing data analysis results typically are readily interpretable, providing important subject area insights.
+%At the other extreme are black box type methods of machine learning.  
+%While these methods are incredibly successful at predicting how processes evolve, they provide little structured insight into the driving mechanisms of the process.
+%Functional Data Analysis (FDA) techniques lie somewhere in between these extremes: the output of an analysis provides visual tools for understanding processes and, often, some dimension reduction to allow for easier exploration.
+%The development of structured FDA techniques which incorporate more complex dependence structures has become a topic of recent interest  (see, e.g., Crainiceanu. and references therein).
+
+Often, movement data can be viewed as a set of many curves.
+In Morris, each child in the data set is associated with two curves from two different days.
+In our data, we will consider curves from a sequence of dives of one killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
+These types of data are amenable to techniques of Functional Data Analysis (FDA).
+FDA was originally developed for curves that can be considered independent replicates  (Ramsay ref, a theoretical ref).  
+However, the structure of the data is often more complicated.
+For instance, in Morris, clearly, the two curves from the same child are dependent.
+And in our dive data, the sequence of days exhibits temporal dependence.
+
+More complicated structures in FDA have been considered in the form of hierarchical ...  typically incorporating random effects.
+Time series of curves  ...  RELATED would be appropriate here.
+Here, we consider Markov modelling of a curve sequence.  Common in animal movement because it gives biological insight.
 
@@
 We enforce temporal structure via Markov processes.  
-While our methodology is generally applicable, [rephrase - awkward] we  showcase our ideas to study the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
+While our methodology is generally applicable, we  showcase our ideas to study the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
 For more information about the data, see Section \ref{sec:case_study}.
@@
 From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
-In FDA, within curve structure is modelled by a generic smooth mean function and covariance function.
+In FDA, within-curve structure is modelled by a generic smooth mean function and covariance function.
 Alternatively, this structure is sometimes generated via random regression.
@@
 Our work relies on work in the vast field of methodology for the analysis of animal movement
-\citep{Hooten:2017}.
+\citep{Hooten:2017,Mcclintock:2020}.
 One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}.

### 2020-07-25 Report/Sections/results.tex
Updates from Overleaf
Commit before restructure!

@@
 
-The CarHHMM-DFT was used to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada.
-Acceleration data can be used to approximate an animal's energy expenditure \citep{Green:2009}, but studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
-Therefore, understanding both the behavioral state of the killer whale, as well as the distribution of accelerometer data within each behavioral state, is important to understand the energetic requirements of killer whales.
+In order to test the advantages of the CarHHM-DFT over the other models on real-world data, we used the four models described in the simulation study to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada.
+Acceleration data can be used to approximate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, which can be helpful for conservation efforts.
+However, studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
+Understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the energetic requirements of killer whales.
 This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.
@@
 Observations were collected at a rate of 50 Hz.
-Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time periods were ignored.
-In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed dives within this time range were ignored as well.
+Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods were ignored.
+In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range were ignored as well.
 A killer whale ``dive" is considered to be any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
@@
 We used a hierarchical HHM, where the coarse-scale observations were a time series of dive durations in seconds, and the fine-scale observations were the within-dive acceleration data (Figure \ref{fig:CarHHMM-DFT}).
-The coarse-scale level is comprised of an HMM with hidden states corresponding to dive types.
+The coarse-scale level was comprised of an HMM with hidden states corresponding to dive types.
 The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
@@
 
-The fine-scale level model subdive behavior and is comprised of a CarHMM-DFT where auto-correlation is modeled into the distribution of $\mathbf{Z}^{*(1)}$ (the average acceleration with a 2-second window), but not $Z^{*(2)}$ (the ``wiggliness" of the two second window).
-The acceleration exhibits significant sinusoidal behavior, thus the fine-scale observations $Z^*$ were made up of the DFT summary statistics of a two-second sliding window.
-Unlike the simulation study, where we had one time series of acceleration data, we have here acceleration data in three dimensions.
-These three dimensions represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags (Citation?).
-The following two sets of observations $z^*$ were calculated for modeling:
+The fine-scale acceleration data exhibits significant sinusoidal behavior, thus the fine-scale observations $Y^*$ were transformed using the DFT summary statistics of a two-second sliding window.
+Unlike the simulation study, where the time series of acceration was one-dimensional, the acceleration data here was collected in three dimensions.
+These three dimensions represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
+The following two sets of observations $z^*$ were calculated:
 %
@@
 The observations are therefore made up of $\mathbf{z}_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $z_t^{*(2)}$, a scalar.
-We calculate $z_t^{*(2)}$ by summing the first 10 Fourier modes, and corresponds to a maximum recorded frequency of $\tilde \omega = 5$ Hz.
+We calculated $z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum recorded frequency of $\tilde \omega = 5$ Hz.
 
-There is strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions (see the supplementary material for a lag plot), thus auto-correlation was directly modeled into the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$ using a CarHMM.
-We assumed the error to be Normally distributed with the following parameters:
+There is strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions (section S-2.1), thus auto-correlation was directly modeled into the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$ using a CarHMM.
+We assumed $\mathbf{Z}^{*(1)}_{t,t^*}$ to be Normally distributed with the following parameters:
 %
@@
 %
-The observations $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent of one another when conditioned of the sub-dive state $X_{t,t^*}$.
+The observations $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent of one another when conditioned of the subdive state $X_{t,t^*}$.
 
-Information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically.
+Information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ subdive behaviors heuristically.
 The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
 
-The final model is nearly identical to the one from the simulation study, with the exception that the fine-scale Markov chain has three sub-dive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a 3-dimensional vector rather than a scalar.
+The final model is nearly identical to the one from the simulation study, with the exception that the fine-scale Markov chain has three subdive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a 3-dimensional vector rather than a scalar.
 
@@
 
-The estimates for the emission distribution parameters suggest that this whale has two distinct dive behaviors (Table \ref{table:emis_dists},  figure \ref{fig:coarse_emis}).
+The estimates for the emission distribution parameters suggest that this whale has at least two distinct dive behaviors (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}).
 Dive type 1 corresponds to shorter, shallower dives.
@@
 Dive type 2 is longer and deeper.
-These types of deep sustained dives have been associated with foraging behaviors in killer whales (CItation Tennessen maybe HMM and the other Tennesen with foraging data).
+These types of deep sustained dives have been associated with foraging behaviors in killer whales (Citation Tennessen maybe HMM and the other Tennesen with foraging data).
 
-The three subdive behaviors also appeared distinct, particularly with respect to the fourier sum (Table \ref{table:emis_dists},  Figure  \ref{fig:fine_emis}).
-Sub-dive type 1 corresponds to gliding and less overall activity compared to the other behavioral states.
-The mean of $Z^{*(2)}$ in this state is at least an order of magnitude smaller than sub-dive behavioral state 2, the variance of $\mathbf{Z}^{*(1)}$ is smaller than sub-dive behavior 2 for every component, and the auto-correlation of $\mathbf{Z}^{*(1)}$ is higher than every other behavioral state.
-Sub-dive state 3, on the other hand, corresponds to vigorous swimming activity, as the mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ for every component is much higher than every other state.
-The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this state, implying more variation in acceleration every 2 seconds.
-Finally, sub-dive state 2 corresponds to a moderate amount of activity, as almost every parameter estimate is between the other two behavioral states.
+The three subdive behaviors also appeared distinct, particularly with respect to their Fourier sums, as the mean of $Z^{*(2)}$ is separated by an order of magnitude between the three subdive types (Table \ref{table:emis_dists_CarHHMM-DFT},  Figure  \ref{fig:fine_emis}).
+Subdive type 1 corresponds to gliding and less overall activity compared to the other subdive types.
+The mean of $Z^{*(2)}$ and the variance of $\mathbf{Z}^{*(1)}$ in this state is smaller than the other subdive types.
+The auto-correlation of $\mathbf{Z}^{*(1)}$ also is higher for subdive type 1 than every other subdive type.
+Subdive type 2 is characterized by a mean Fourier sum ($Z^{*(2)}$) one order of magnitude high than subdive type 1.
+Acceleration has about twice the variance in this subdive type compared to subdive type 2, and slightly lower auto-correlation.
+Subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behavior in acceleration is characteristic of this behavior marine mammals \citep{Simon:2012}.
+Finally, subdive type 3 corresponds to vigorous swimming activity, especially as the killer whale submerges or surfaces.
+The mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ is much higher in subdive type 3 compared to every other state.
+The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this subdive type, implying more variation in acceleration.
 
@@
 $$\hat \Gamma = \begin{pmatrix} 
-0.849  &  0.151 \\
-0.907  &  0.093
+0.788 & 0.212 \\
+0.809 & 0.191
 \end{pmatrix},$$
-$$\hat \delta = \begin{pmatrix} 0.857 & 0.143 \end{pmatrix},$$
+$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix},$$
 %
 $$\hat \Gamma^{*(1)} = \begin{pmatrix} 
-0.724 & 0.276 & 0.000 \\
-0.057 & 0.887 & 0.056 \\
-0.000 & 0.247 & 0.753
+0.679 & 0.321 & 0.000 \\
+0.038 & 0.904 & 0.058 \\
+0.000 & 0.232 & 0.768
 \end{pmatrix}, \qquad 
 \hat \Gamma^{*(2)} = \begin{pmatrix} 
-0.871 & 0.129 & 0.000 \\
-0.135 & 0.829 & 0.036 \\
-0.000 & 0.246 & 0.754
+0.859 & 0.141 & 0.000 \\
+0.114 & 0.841 & 0.045 \\
+0.000 & 0.216 & 0.784
 \end{pmatrix},$$
-$$\hat \delta^{*(1)} = \begin{pmatrix} 0.143 & 0.698 & 0.159 \end{pmatrix}, \qquad
-\hat \delta^{*(2)} = \begin{pmatrix} 0.476 & 0.456 & 0.067 \end{pmatrix}.$$
+$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \qquad
+\hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}.$$
 %
 The coarse-scale stationary distribution indicates that most dives are short type 1 dives.
-The coarse-scale transition probability matrix further indicates that the whale makes multiple short type 1 dives before doing only one (or a few) long type 2 dive.
-The fine-scale stationary distributions indicate that this killer whale is more likely to be in the less active sub-dive type 1 when performing long deep dives than when performing short shallow dives.
+The coarse-scale transition probability matrix further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive.
+The fine-scale stationary distributions indicate that this killer whale is more likely to be in the less active subdive type 1 when performing long deep dives than when performing short shallow dives.
 Using less active swimming behavior when diving, and holding breath, for long periods of time has been suggested to be an energy reduction strategy in bottle-nose dolphins \citep{Williams:1999}.
-Such strategy could explain why sub-dive state 1 makes up 14\% of shallow dives of our killer whale, but it makes up 48\% of deep dives.
-Figure \ref{fig:labeled_dives} shows the decoded dive behavior of 5 selected dives.
-The supplementary material also shows the estimated \textit{probability} of each dive time and sub-dive hidden state.
+Such strategy could explain why subdive type 1 makes up 9\% of shallow dives for this killer whale, but it makes up 40\% of deep dives.
+Figure \ref{fig:labeled_dives} shows the decoded dive behavior of 6 selected dives.
+The supplementary material also shows the estimated \textit{probability} of each dive time and subdive hidden state (section S-2.2).
 
@@
 
-We also plotted separate histograms of each feature where each observation was weighted by the probability that the whale was in a particular hidden state.
-This empirical distribution was then plotted over the fitted probability distribution function of that feature and hidden state.
-If the model is correct, then the histogram of each feature and hidden state should closely resemble its corresponding fitted probability distribution.
-See Figure \ref{fig:empirical_dist} for an example of the dive duration.
+We also plotted histograms of each feature where observations were weighted by the probability of a particular hidden state.
+This empirical distribution was then plotted over its corresponding fitted emission distribution (Figure \ref{fig:empirical_dist}).
 Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed.
-In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for sub-dive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
+In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
 These outliers are potential subjects for future study.

### 2020-07-25 Report/Sections/simulation.tex
Commit before restructure!

@@
 \begin{align*}
-	&\bbE(Y_t|X_t = 1) = 15 s, &\bbE(Y_t|X_t = 2) = 60 s, \\
+	&\mu^{(1)} \equiv \bbE(Y_t|X_t = 1) = 15 s, &\bbE(Y_t|X_t = 2) = 60 s, \\
 	%
@@
 %
-$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix}, \qquad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix},$$ 
+$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix},$$ 
 %
@@
 
-Each two-second sub-dive window had 100 associated acceleration readings, $Y^*_{t,t^*} \equiv \left(Y^*_{t,t^*,1}, \ldots, Y^*_{t,t^*,100}\right)$.
+Each two-second subdive window had 100 associated acceleration readings, $Y^*_{t,t^*} \equiv \left(Y^*_{t,t^*,1}, \ldots, Y^*_{t,t^*,100}\right)$.
 To accurately recreate active swimming versus passive gliding on the fine-scale Markov chain, the DFT of each two-second segment $\hat Y^*_{t,t^*}$ was simulated such that $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$ (see Eq.
@@
 \begin{gather*}
-    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 1 \right) \sim \mathcal{N} \left(\phi^{*(1)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(1)})\mu^{*(1)}, (\sigma^{*(1)})^2 \right) \\
+    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 1 \right) \sim \mathcal{N} \left(\phi^{*(1)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(1)})\mu^{*(1)}, (\sigma^{*(1)})^2 \right), \\
     %
-    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 2 \right) \sim \mathcal{N} \left(\phi^{*(2)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(2)})\mu^{*(2)}, (\sigma^{*(2)})^2 \right) \\
+    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 2 \right) \sim \mathcal{N} \left(\phi^{*(2)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(2)})\mu^{*(2)}, (\sigma^{*(2)})^2 \right), \\
     %
-    \mu^{*(1)} = 0.0,\enspace \sigma^{*(1)} = 0.05,\enspace \phi^{*(1)} = 0.99 \\
+    \mu^{*(1)} = 0.0,\enspace \sigma^{*(1)} = 0.05,\enspace \phi^{*(1)} = 0.99, \\
     %
-    \mu^{*(2)} = 0.0,\enspace \sigma^{*(2)} = 0.1,\enspace \phi^{*(2)} = 0.95 \\\\
+    \mu^{*(2)} = 0.0,\enspace \sigma^{*(2)} = 0.1,\enspace \phi^{*(2)} = 0.95, \\\\
     %
-    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha^{*(1)}, \beta^{*(1)} \right) \\
+    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha^{*(1)}, \beta^{*(1)} \right), \\
     %
-    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(\alpha^{*(2)}, \beta^{*(2)} \right) \\
+    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(\alpha^{*(2)}, \beta^{*(2)} \right), \\
     %
-    \alpha^{*(1)} = 10.10, \enspace \enspace \beta^{*(1)} = 1.00 \\
+    \alpha^{*(1)} = 10.10, \enspace \enspace \beta^{*(1)} = 1.00, \\
     %
-    \alpha^{*(2)} = 305.94, \enspace \enspace \beta^{*(2)} = 1.00
+    \alpha^{*(2)} = 305.94, \enspace \enspace \beta^{*(2)} = 1.00.
 \end{gather*}
 %
-Sub-dive behavior 1 corresponds to passive gliding while sub-dive behaviour 2 corresponds to active swimming with a dominant frequency of $\frac{1}{2}s^{-1}$.
-Sub-dive behaviors 1 and 2 are the same for both dive types.
+subdive behavior 1 corresponds to passive gliding while subdive behaviour 2 corresponds to active swimming with a dominant frequency of $1/2 s^{-1}$.
+subdive behaviors 1 and 2 are the same for both dive types.
 See the appendix for more details regarding procedure for simulating $\hat Y^*$ and $Y^*$ such that $Z^*$ has the preceding distribution.
@@
 
-The building blocks from the previous section were used to build a well-specified model for this simulated data.
+The building blocks from the previous section were used to build a well-specified model for these simulated data.
 Specifically, we used a hierarchical HMM where the sequence of dive durations $Y$ was modeled using a simple HMM, and the fine-scale process $Z^*$ was modeled using a HMM-DFT with explicit auto-correlation in $Z^{*(1)}$.
@@
 
-On the fine scale, the sub-dive behavior of each two-second window comprises a Markov chain with $N^*=2$ possible states and unknown probability transition matrices $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$, depending upon the dive type.
+On the fine scale, the subdive behavior of each two-second window comprises a Markov chain with $N^*=2$ possible states and unknown probability transition matrices $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$, depending upon the dive type.
 Each two-second window is summarized by the observations $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$.
-The distribution of $Z^{*(1)}_{t,t^*}$ is Normal and its parameters depend upon the sub-dive behavior $X^*_{t,t^*}$ and $Z^{*(1)}_{t,t^*-1}$.
+The distribution of $Z^{*(1)}_{t,t^*}$ is Normal and its parameters depend upon the subdive behavior $X^*_{t,t^*}$ and $Z^{*(1)}_{t,t^*-1}$.
 In particular:
@@
 The Fourier sums $(Z^{*(2)})$ show no significant bias in $\hat \sigma^*_2$.
-The empirical standard error of all parameters estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) were well-approximated by the inverse of the observed fisher information matrix (section S-1.2, tables 2-5).
-One exception is $\hat \Theta$, whose standard errors were slightly underestimated using the observed fisher information matrix.
+The empirical standard error of all parameters estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) were well-approximated by the inverse of the observed Fisher information matrix (section S-1.2, tables 2-5).
+One exception is $\hat \Theta$, whose standard errors were slightly underestimated using the observed Fisher information matrix.
 %The empirical standard error of $\hat \mu_2$ was 8.5, while the estimate of standard error using the observed Fisher information was 5.8. 
@@
 The HHMM-DFT \textit{did} differ from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$.
-For both sub-dive states, $\hat \sigma_1^*$ greatly overestimated the true standard deviation.
+For both subdive states, $\hat \sigma_1^*$ greatly overestimated the true standard deviation.
 In particular, $\overline{\hat \sigma_1^{*(1)}} = 0.248$ and $\overline{\hat \sigma_1^{*(2)}} = 0.242$, but $\sigma_1^{*(1)} = 0.05$ and $\sigma_1^{*(2)} = 0.1$.
-In addition, the empirical standard error of both $\hat \mu_1^*$ and $\hat \sigma_1^*$ was much larger than the estimates of standard error using the observed fisher information (see Section S-1.2, Table 3).
+In addition, the empirical standard error of both $\hat \mu_1^*$ and $\hat \sigma_1^*$ was much larger than the estimates of standard error using the observed Fisher information (see Section S-1.2, Table 3).
 This implies that parameter uncertainty can be underestimated when auto-correlation is ignored.
 
-The CarHHMM was the worst-performing model in terms of decoding accuracy, with an accuracy of $89 \pm 1\%$ when determining sub-dive behavior and an accuracy of $87 \pm 11\%$ when determining dive types.
-It was much more likely to misclassify sub-dive behavior 1 than sub-dive behavior 2 (see Section S-1.2, Table 1).
+The CarHHMM was the worst-performing model in terms of decoding accuracy, with an accuracy of $89 \pm 1\%$ when determining subdive behavior and an accuracy of $87 \pm 11\%$ when determining dive types.
+It was much more likely to misclassify subdive behavior 1 than subdive behavior 2 (see Section S-1.2, Table 1).
 It was also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy.
@@
 The HHMM-DFT regularly overestimates the variance $Z^{*(1)}$ since it does not incorporate auto-correlation into the emission distribution of $Z^{*(1)}$.
-The CarHHMM has large biases in many of its parameter estimates, especially in sub-dive state 2.
+The CarHHMM has large biases in many of its parameter estimates, especially in subdive state 2.
 %because it sees the ``wigliness" of the acceleration data as raw acceleration rather than as Fourier modes, which deflates the value of $\hat \phi$ and inflates the value of $\hat \sigma$.

### 2020-07-26 Code/HHMM.py
Add other plots

@@ -195,6 +195,8 @@ class HHMM:
             # find new mean if there is autocorrelation
             if self.pars.features[level][feature]['corr'] and data_tm1 is not None:
                 mu = (1.0-corr)*mu + corr*data_tm1[feature]
+            elif self.pars.features[level][feature]['corr'] and data_tm1 is None:
+                return 0
 
             if sample > 0:
                 return multivariate_normal.rvs(mu,sig,sample)
@@ -207,6 +209,8 @@ class HHMM:
             # find new mean if there is autocorrelation
             if self.pars.features[level][feature]['corr'] and data_tm1 is not None:
                 mu = (1.0-corr)*mu + corr*data_tm1[feature]
+            elif self.pars.features[level][feature]['corr'] and data_tm1 is None:
+                return 0
 
             if sample > 0:
                 return norm.rvs(mu,sig,sample)
@@ -218,6 +222,8 @@ class HHMM:
             # find new mean if there is autocorrelation
             if self.pars.features[level][feature]['corr'] and data_tm1 is not None:
                 mu = (1.0-corr)*mu + corr*data_tm1[feature]
+            elif self.pars.features[level][feature]['corr'] and data_tm1 is None:
+                return 0
 
             shape = np.square(mu)/np.square(sig)
             scale = np.square(sig)/np.array(mu)
@@ -239,6 +245,8 @@ class HHMM:
                 mu = ((mu+np.pi)%(2*np.pi))-np.pi
             elif self.pars.features[1][feature]['corr'] and data_tm1 is not None:
                 mu = (1.0-corr)*mu + corr*data_tm1[feature]
+            elif self.pars.features[level][feature]['corr'] and data_tm1 is None:
+                return 0
             else:
                 pass
 
@@ -270,10 +278,10 @@ class HHMM:
 
         # initialize values
         log_L = 0
-        seg_tm1 = dive_data[0]
+        seg_tm1 = None
 
         # iterate through dive segments
-        for i,seg in enumerate(dive_data):
+        for seg in dive_data:
 
             log_p_yt_given_xt = np.zeros(self.pars.K[1])
 
@@ -539,7 +547,7 @@ class HHMM:
                         x0 = self.theta[0][feature][param][k0]
 
                         # optimize
-                        res = minimize(loss_fn, x0, method='BFGS', options=options)
+                        res = minimize(loss_fn, x0, method='Nelder–Mead', options=options)
 
                         # update final values
[…]

### 2020-07-26 Code/Train_Model.py

@@ -41,7 +41,7 @@ pars.features = [{'dive_duration':{'corr':False,'f':'gamma'}},
 pars.K = [2,3]
 pars.share_fine_states = True
 
-if model == 'CarHHM':
+if model == 'CarHMM':
     pars.K = [1,3]
 elif model == 'HHMM':
     pars.features[1]['Ax']['corr'] = False

### 2020-07-26 Code/Visualisor.py

@@ -107,7 +107,7 @@ class Visualisor:
         return
 
 
-    def plot_emission_probs(self,level,file=None):
+    def plot_emission_probs(self,level,file=None,features=None):
 
         if self.hhmm is None:
             print('No trained model')
@@ -118,7 +118,8 @@ class Visualisor:
             nrows = 1
             ncols = len(self.hhmm.theta[0])
         else:
-            features = ['Ax','Ay','Az','Ahat_low']#list(self.hhmm.theta[1][0].keys())
+            if features is None:
+                features = ['Ax','Ay','Az','Ahat_low']#list(self.hhmm.theta[1][0].keys())
             nrows = 2 #self.pars.K[0]
             ncols = int((len(features)+1)/2)
 

### 2020-07-26 Report/Sections/data.tex
Updates from Overleaf

% !TeX root = ../main.tex

%\section{Data Collection and Model Construction}

To illustrate the process of constructing a model using the HMM model building blocks, we investigate the dive behavior of a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada, and construct several candidate generative models.

Acceleration data can be used to approximate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, which can be helpful for conservation efforts. However, studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the energetic requirements of killer whales. This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.

We begin by describing the data collection process used by ecologists. We then detail the procedure used to build several candidate generative models from this data. Next, we perform a simulation study by selecting a preferred model, simulating data from it, and fit the candidate models to the simulated data to compare their performance in terms of parameter estimates and decoding accuracy. Finally, we fit each candidate model to the original data and analyze the results.

\subsection{Data collection and preprocessing}

The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods were ignored. In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. After preprocessing the raw data, a total of 267 dives were observed. Figure \ref{fig:data} displays the dive profile and accelerometer data.

Each dive is treated as one curve, and the collection of dives make up the coarse-scale process. As such, the coarse-scale observations $Y$ were a time series of dive durations in seconds ($s$) with corresponding hidden dive types $X$. In addition, the fine-scale processes $Y^*$ were the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive types $X^*$.

\subsection{Model Selection}

The coarse-scale level exhibited no significant amount of auto-correlation (See section S-2.1) and no obvious intricate structure. As such, we selected a simple HMM to model this data. The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)},$$
$$\mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$

The fine-scale acceleration data exhibited significant sinusoidal behavior, thus the fine-scale observations $y^*$ were transformed to $z^*$ using the DFT summary statistics of a two-second sliding window. A two-second window was selected heuristically since it appeared to be the smallest window size that could capture the sinusoidal behavior. Because the data was collected at the rate of 50 Hz, each two-second window contained a total of 100 raw acceleration readings. The acceleration data here was collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. As a result, the transformation from the raw accelerometer data $\mathbf{y}^*$ to the observation sequence $z^*$ was calculated as follows:
%
$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2.$$
%
Each observation was therefore made up of $\mathbf{z}_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $z_t^{*(2)}$, a scalar. We calculated $z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum frequency of $\tilde \omega = 5$ Hz.

Even after accounting for fine-scale structure in the raw acceleration data, there was still strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ (section S-2.1), thus a CarHMM was used to model $\mathbf{Z}^{*(1)}_{t,t^*}$. In particular, we assumed $\mathbf{Z}^{*(1)}_{t,t^*}$ to be Normally distributed with unknown parameters $\phi_1^{*(i)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i)} \in \mathbb{R}^3$.
%
%$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)},$$
%$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right].$$
%

While $Z^{*(2)}$ also exhibits some auto-correlation, the relationship is less strong than that of $\mathbf{Z}^{*(1)}$, and the biological interpretation of auto-correlation within $Z^{*(2)}$ is less clear. Therefore, a simple HMM was used to model $Z^{*(2)}$ and the distribution of $Z^{*(2)}_{t,t^*}$ was assumed to be gamma and parameterized by its mean $\mu_2^{*(i)}$ and variance $\sigma_2^{*(i)}$.
%
%$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
%$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%
The observations $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ shared a subdive state $X^*_{t,t^*}$, but were assumed to be independent of one another when conditioned on $X_{t,t^*}$. In addition, the distribution of $Z^*_{t,t^*}$ is assumed to depend only upon the subdive type and the previous observation $Z^*_{t,t^*-1}$, not the dive type $X_t$.

Finally, to select the number of dive types and subdive types, information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}. Instead, we selected $N = 2$ dive types and $N^* = 3$ subdive behaviors heuristically. The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

The final model is a hierarchical hidden Markov model with a traditional HMM at the coarse scale. On the fine scale, $y^*$ was transformed to $z^*$, $z^{*(1)}$ was modeled with a CarHMM, and $z^{*(2)}$ was modeled with an HMM. We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMM models discussed here. The parameters to estimate are
%
\begin{gather*}
    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
    %
    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters), and} \\
    %
    \Theta^* = \{\Theta^{*(1)},\Theta^{*(2)}\}  \qquad \text{(fine-scale emission parameters), where} \\
    %
    \Theta^{*(i)} =  \{\{\mu_1^{*(i)},\sigma_1^{*(i)},\phi_1^{*(i)}\},\{\mu_2^{*(i)},\sigma_2^{*(i)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
\end{gather*}
%
The likelihood of this model is still easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above. See the appendix for details of likelihood evaluation. Figure \ref{fig:CarHHMM-DFT} also shows the corresponding graphical model.

In addition to the preferred CarHHMM-DFT, we consider three similar variations for comparison:
\begin{enumerate}
    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using the DFT, but models \textit{both} $\mathbf{Z}^{*(1)}$ \textit{and} $Z^{*(2)}$ with a simple HMM rather than a CarHMM.
    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations \textit{without} using the DFT, and models $Z^{*(1)}$ using a CarHMM. The Fourier sums $\mathbf{Z}^{*(2)}$ are omitted as observations in this model.
    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $\mathbf{Z}^{*(1)}$ with a CarHMM and $Z^{*(2)}$ with an HMM. This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
\end{enumerate}
All models have the same emission distributions as the CarHHMM-DFT, when applicable. Each of the three models leaves out one important aspect of the full CarHHMM-DFT. The HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the Fourier transform sums $(Z^{*(2)})$ as observations, and the CarHMM-DFT lacks a hierarchical structure by only modelling the fine-scale behavior of the whale.

### 2020-07-26 Report/Sections/intro.tex
Updates from Overleaf

@@
-{\em{This still needs some editing}}
-
-\vskip 10pt\noindent
-
-Recent advances in technology allow us to gather enormous amounts of data collected almost continuously in time.
-The collection and analysis of movement data from devices such as accelerometers and GPS trackers have brought new understanding to {\bf{more references}}  physical activity levels in children \citep{Morris:2007} ...
-In this article, we focus on the study of animal movement;  tags attached to animals record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours and providing biologists with new tools for understanding animal behaviour.
+Recent advances in technology for tracking movement provide researchers with data collected almost continuously in time.
+The collection and analysis of data from devices such as accelerometers have brought new understanding to areas ranging from  {\bf{more references}}  to physical activity levels in children \citep{Morris:2007}.
+In this article, we focus on the study of animal movement;  tags attached to animals record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours and providing biologists with new information about  animal behaviour.
 
@@
 
-Often, movement data can be viewed as a set of many curves.
-In Morris, each child in the data set is associated with two curves from two different days.
-In our data, we will consider curves from a sequence of dives of one killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
-These types of data are amenable to techniques of Functional Data Analysis (FDA).
-FDA was originally developed for curves that can be considered independent replicates  (Ramsay ref, a theoretical ref).  
-However, the structure of the data is often more complicated.
-For instance, in Morris, clearly, the two curves from the same child are dependent.
-And in our dive data, the sequence of days exhibits temporal dependence.
+Movement data are sometimes modelled as a set of curves that can  be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}).
+In \cite{Morris:2007}, a child's activity level throughout the day can be viewed as a curve, specifically, as activity as a function of time.
+In the data  we consider, a single dive of a whale can be viewed as a curve:  acceleration as a function of time.
+FDA was originally developed for curves that can be considered independent replicates.
+However, the structure of the data is often more complex.
 
-More complicated structures in FDA have been considered in the form of hierarchical ...  typically incorporating random effects.
-Time series of curves  ...  RELATED would be appropriate here.
-Here, we consider Markov modelling of a curve sequence.  Common in animal movement because it gives biological insight.
-
-Our goal is to develop movement analysis methods that are flexible but also use subject area knowledge to provide insight into curve data with complex structure.
+Our goal is to develop movement analysis methods that provide subject-area insight  into the complex structure of curve data.
 We  model complexities arising  from curve to curve variation, along with the complexities of within-curve fine-scale characteristics and temporal dependence,
 combining existing methods in novel ways.
-We enforce temporal structure via Markov processes.  
-While our methodology is generally applicable, we  showcase our ideas to study the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
+We enforce temporal structure and dependence from dive to dive via a hidden Markov model (HMM).
+We model within dive variation by combining an HMM and a time series.
+Using HMMs at both the dive level and the fine scale allows us to interpret the data as arising from different types of behaviour.
+While our methodology is generally applicable, we  showcase our methods in a  study of the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
 For more information about the data, see Section \ref{sec:case_study}.
@@
 The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and maximum depths. 
-The bottom three panels show the acceleration in each of three orthogonal directions, as recorded by an accelerometer,   
+The bottom three panels show the acceleration in each of three orthogonal directions as recorded by an accelerometer,   
 with axes oriented relative to the whale's forward movement.
@@
 Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
-The plot shows one component of the acceleration, and from this we can see fine-scale behaviour within each dive which seems to indicate some fine scale short-term periodicity. 
+The plot shows one component of the acceleration, and from this we can see fine-scale behaviour within each dive,  behaviour that seems to indicate some fine-scale short-term periodicity. 
 We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour. 
@@
 
-The killer whale dives form a sequence of dependent curves. However, in FDA, curves are typically deemed independent.  
-A noted exception involves multilevel modelling [references] which uses random effects and is appropriate when the set of curves can be partitioned into independent subsets, typically with each subset arising from one individual.
-[NEED REFERENCES]  
-However, this approach is not appropriate here, so we consider a temporal structure for dependence, with the characteristics of each curve determined by a latent state that evolves according to a Markov model.
-This HMM allows us to interpret each curve as arising from a type of behaviour.
+Particular types of between curve dependence structures have received attention in the FDA literature, 
+notablly multilevel models that incorporate dependence via random effects.
+For instance, in the aforementioned study of children's activity levels in  \cite{Morris:2007}, the authors use a multilevel approach since each child yields not one but two daily activity curves and so the model includes a random effect for ``child''.
+Perhaps the first work with this multilevel structure is \cite{Bromback:1998} in a study of ....
+More recent work on multilevel models includes that of \cite{di:2009}, \cite{chen:2012}} and Crainiceanu [REF].
+For our data, the multilevel models are not appropriate, since they do not account for the temporal dependence.
+
+FDA researchers have modelled temporal dependence in a sequence of curves  via functional time series, extending the ideas of classic time series to model how one curve evolves into the next curve.
+See, for instance, \cite{Kokoszka:2018}.
+While  functional time series methods can be applied to our dive data, they would not account for the different dive types.
+Our HMM models account for both temporal dive-to-dive dependence and dive types.
+
+%A noted exception involves multilevel modelling [references] which uses random effects and is appropriate when the set of curves can be partitioned into independent subsets, typically with each subset arising from one individual.
 
@@
 We model the within-curve dependence by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence.
-Finally, for the fine-scale behaviour, we use summary information that reflects the application area's meaning for the curve.
-Specifically, for the dive data, we use summaries of Fourier transforms to account for sudden and extreme ``wiggliness'' within some segments of a curve.
+Finally, for the fine-scale behaviour, we use summary information that is biologically meaninful.
+Specifically, for the dive data, we use summaries of Fourier transforms to account for the ``wiggliness'' in some segments of a curve.
 The distribution of this ``wiggliness'' depends on the hidden state.
 
-{\bf{ Make sure to reference Langrock}}
+Our work relies on work in the vast field of methodology for the analysis of animal movement
+\citep{Hooten:2017,Mcclintock:2020}, where
+one of the most prevalent techniques in the literature of late is the HMM \citep{Patterson:2017}. 
+However, HMMs have seen little use in FDA.  
+One exception is the work of Roland Langrock (see, for instance \citealt{Langrock:2018}).
+Animal movement modelers have also used  hierarchical hidden Markov models (HHMMs) to model  simultaneous behavioral processes that occur at different time scales \citep{Barajas:2017,Adam:2019}, and we borrow from that work here.
+Part of the purpose of our article is to introduce the FDA community to these very useful approaches to modelling complex curves.
 
-Our work relies on work in the vast field of methodology for the analysis of animal movement
-\citep{Hooten:2017,Mcclintock:2020}.
-One of the most prevalent techniques in the literature of late is the hidden Markov model (HMM), where observations depend upon the state of an associated unobserved behavioral process following Markovian dynamics \citep{Patterson:2017}.
 %
@@
 
-Animal movement modelers have recognized that several simultaneous behavioral processes may occur at different time scales. 
-One approach to the analysis of this type of data is to use a hierarchical hidden Markov model, or HHMM \citep{Barajas:2017,Adam:2019}. 
-HHMMs model the entire time series in question as a nested structure of hidden Markov models where each HMM corresponds to one behavioral process.
+We also borrow techniques from animal movement to model fine-scale behaviour, beyond simply using an HMM.  
+Under the traditional HMM model,  observations are independent given  the underlying hidden state process.
+However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
+%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
+Several solutions have been proposed, such as  the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
+The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
 
-At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs.
+An additional complication is that, at the shortest time scales,  observations often exhibit complicated  structure which cannot easily be captured by traditional HMMs, CarHMMs, or HHMMs.
 Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}.
@@
 However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
-One way to model periodic behavior in particular is to use signal processing techniques such as the Fourier transform on the raw data. 
-The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}.
+Our dive data exhibit  fine-scale periodic behaviour, so we transform the data via
+ the signal processing technique of the Fourier transform. 
+\cite{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM  in the context of describing daily behavioral cycles of marine mammals.
 In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data.
@@
 Finally, we apply our method to the dive data of Figure 6.
+
+In the following section, we describe HMMs, CarHMMs and HHMMs.
+We show how these models can be combined in order to analyze complex data.
+We also discuss transforming data.
+In Section 3, we study various models for complex ...
+Section 4 contains data analysis and Section 5 discussion.

### 2020-07-26 Report/Sections/results.tex
Updates from Overleaf

@@
 
-In order to test the advantages of the CarHHM-DFT over the other models on real-world data, we used the four models described in the simulation study to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada.
-Acceleration data can be used to approximate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, which can be helpful for conservation efforts.
-However, studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
-Understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the energetic requirements of killer whales.
-This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.
+%In order to test the advantages of the CarHHM-DFT over the other models on real-world data, we used the four models described in the simulation study to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada.
 
-\subsection{Data Collection and Preprocessing}
+%\section{Results}
 
-The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions.
-Observations were collected at a rate of 50 Hz.
-Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods were ignored.
-In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range were ignored as well.
-A killer whale ``dive" is considered to be any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
-Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second.
-Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}.
-After preprocessing the raw data, a total of 267 dives were observed.
-Figure \ref{fig:data} displays the dive profile and accelerometer data.
+We report the results from fitting the CarHHMM-DFT in detail, validate the results, and then compare the results of this model with those from the other candidate models.
 
-\subsection{Model Selection}
-
-We used a hierarchical HHM, where the coarse-scale observations were a time series of dive durations in seconds, and the fine-scale observations were the within-dive acceleration data (Figure \ref{fig:CarHHMM-DFT}).
-The coarse-scale level was comprised of an HMM with hidden states corresponding to dive types.
-The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
-$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)},$$
-$$\mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
-
-The fine-scale acceleration data exhibits significant sinusoidal behavior, thus the fine-scale observations $Y^*$ were transformed using the DFT summary statistics of a two-second sliding window.
-Unlike the simulation study, where the time series of acceration was one-dimensional, the acceleration data here was collected in three dimensions.
-These three dimensions represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
-The following two sets of observations $z^*$ were calculated:
-%
-$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2.$$
-The observations are therefore made up of $\mathbf{z}_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $z_t^{*(2)}$, a scalar.
-We calculated $z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum recorded frequency of $\tilde \omega = 5$ Hz.
-
-There is strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions (section S-2.1), thus auto-correlation was directly modeled into the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$ using a CarHMM.
-We assumed $\mathbf{Z}^{*(1)}_{t,t^*}$ to be Normally distributed with the following parameters:
-%
-$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)}$$
-$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right]$$
-%
-where $\phi_1^{*(i)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i)} \in \mathbb{R}^3$.
-
-While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, and the biological interpretation of auto-correlation within $Z^{*(2)}_{t,t^*}$ is less clear.
-Auto-correlation was therefore not incorporated into the emission distribution of $Z^{*(2)}_{t,t^*}$.
-In particular, the distribution of $Z^{*(2)}_{t,t^*}$ was assumed to be gamma and parameterized by its mean and variance:
-%
-$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
-$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
-%
-The observations $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent of one another when conditioned of the subdive state $X_{t,t^*}$.
-
-Information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ subdive behaviors heuristically.
-The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
-
-The final model is nearly identical to the one from the simulation study, with the exception that the fine-scale Markov chain has three subdive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a 3-dimensional vector rather than a scalar.
-
-\subsection{Results}
-
-The estimates for the emission distribution parameters suggest that this whale has at least two distinct dive behaviors (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}).
+The estimates for the emission distribution parameters suggest that the killer whale has at least two distinct dive behaviors (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). 
 Dive type 1 corresponds to shorter, shallower dives.
@@
 
-The three subdive behaviors also appeared distinct, particularly with respect to their Fourier sums, as the mean of $Z^{*(2)}$ is separated by an order of magnitude between the three subdive types (Table \ref{table:emis_dists_CarHHMM-DFT},  Figure  \ref{fig:fine_emis}).
-Subdive type 1 corresponds to gliding and less overall activity compared to the other subdive types.
-The mean of $Z^{*(2)}$ and the variance of $\mathbf{Z}^{*(1)}$ in this state is smaller than the other subdive types.
+The three subdive behaviors also appeared distinct with respect to their Fourier sums, as the mean of $Z^{*(2)}$ is separated by an order of magnitude between the three subdive types (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}).
+For subdive type 1, the mean of $Z^{*(2)}$ and the variance of $\mathbf{Z}^{*(1)}$ in this state is smaller than the other subdive types.
 The auto-correlation of $\mathbf{Z}^{*(1)}$ also is higher for subdive type 1 than every other subdive type.
-Subdive type 2 is characterized by a mean Fourier sum ($Z^{*(2)}$) one order of magnitude high than subdive type 1.
-Acceleration has about twice the variance in this subdive type compared to subdive type 2, and slightly lower auto-correlation.
-Subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behavior in acceleration is characteristic of this behavior marine mammals \citep{Simon:2012}.
-Finally, subdive type 3 corresponds to vigorous swimming activity, especially as the killer whale submerges or surfaces.
-The mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ is much higher in subdive type 3 compared to every other state.
+This corresponds to less overall activity compared to the other subdive types.
+Subdive type 2 is characterized by a mean Fourier sum ($Z^{*(2)}$) one order of magnitude higher than subdive type 1.
+Acceleration has about twice the variance in this subdive type compared to subdive type 1, and slightly lower auto-correlation.
+Subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behavior in acceleration is characteristic of this behavior in marine mammals \citep{Simon:2012}.
+In subdive type 3, the mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ is much higher compared to every other state.
 The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this subdive type, implying more variation in acceleration.
+This corresponds to vigorous swimming activity, especially as the killer whale begins a dive or surfaces from a dive.
 
@@
 \end{pmatrix},$$
-$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \qquad
-\hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}.$$
+$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}.$$
 %
@@
 Using less active swimming behavior when diving, and holding breath, for long periods of time has been suggested to be an energy reduction strategy in bottle-nose dolphins \citep{Williams:1999}.
-Such strategy could explain why subdive type 1 makes up 9\% of shallow dives for this killer whale, but it makes up 40\% of deep dives.
 Figure \ref{fig:labeled_dives} shows the decoded dive behavior of 6 selected dives.
-The supplementary material also shows the estimated \textit{probability} of each dive time and subdive hidden state (section S-2.2).
+The supplementary material also shows the estimated probability of each dive and subdive type (section S-2.2).
 
@@
 See the supplementary material for empirical distributions of every feature and every hidden state.
+
+\subsection{Comparison with candidate models}
+
+The CarHMM-DFT produced very similar to those of the CarHHMM-DFT on the fine scale.
+However, it failed to differentiate between short and long dives, and it therefore did not infer any information relating to the dive level Markov chain or the relationship between the dive level and the subdive level.
+For example, the CarHMM-DFT missed the insight that the whale was much more likely to be in subdive type 1 when engaged in longer dives than when engaged in shorter dives.
+
+The HHMM-DFT decoded the dive and subdive types similarly to the CarHHMM-DFT, but was less likely to categorize the beginning and end of dives as subdive type 3.
+In addition, the HHMM-DFT produced estimates $\hat \sigma_1^*$ that were approximately 50-100 \% larger than those for the CarHHMM-DFT, and the estimated uncertainty of $\hat \mu^*_1$ was less than half of that for the CarHHMM-DFT for all dimensions and subdive states.
+These findings are consistent with the trend to overestimate variance and produce overconfident parameter estimates when auto-correlation is ignored.
+Further, the empirical auto-correlation of $\mathbf{Z}^{*(1)}$ within each decoded subdive type was at least 0.5 for all dimensions and subdive types.
+Finally, the pseudoresiduals of the HHMM-DFT had many more outliers than the pseudoresiduals of the CarHHMM-DFT, providing evidence that the HHMM-DFT was a worse fit to this data than the CarHHMM-DFT.
+
+The CarHHMM did not have access to the Fourier sums of the acceleration data, so it failed to decode the subdive types into biologically interpretable results.
+In addition, the pseudoresiduals corresponding to the CarHHMM appears to be less heavy-tailed than a normal distribution.
+
+See section S-??? for a more complete set of results for each of the candidate models.

### 2020-07-26 Report/Sections/simulation.tex
Updates from Overleaf

@@
 
-We based our simulation study on the acceleration data of a killer whale's dive sequence.
+To compare the performance of the candidate models when the ground-truth is known, we performed a simulation study based on data generated from the full CarHHMM-DFT model.
 The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}.
-We fit four separate models to the simulated data and compare their accuracy.
+The ground-truth parameters differ slightly from those in the case study for primarily pedagogical reasons.
+For example, we set $\Gamma$ such that the simulated data would contain approximately equal numbers of short and long dives.
+This allows meaningful conclusions to be made for both dive types.
+In addition, the number subdive states was set to $N^*=2$ to simplify the analysis of the subdive states, and $\mathbf{Z}^{*(2)}$ was changed from a vector to a scalar for additional simplicity.
+
+We fit all four candidate models to the data generated from the CarHHMM-DFT in order to analyze the impact of model mis-specification and compare them using several different metrics.
+These metrics include the decoding accuracy of the dive and subdive types, bias in parameter estimates, standard error of parameter estimates, and fitting times.
+In addition, we compare the empirical standard error of each model's parameter estimates with the estimated standard error using the observed Fisher information under that model, which allows us to judge how accurate a given model is when estimating its own uncertainty.
 
@@
 
-Five hundred separate sequences of 100 killer whale dives were simulated according to an HMM, where the hidden Markov chain $X$ was a collection of dive types and the observations $Y$ were the corresponding dive durations (in seconds).
-Each dive could be one of $N=2$ dive types, and the duration of the $t^{th}$ dive, $Y_t$, followed a gamma distribution whose parameters $\theta^{(i)}$ were dependent on the dive type $X_t = i$.
-We parameterize the gamma distribution by its mean and variance:
+We generated 500 independent data sets from the CarHHMM-DFT.
+Each data set consisted of a sequence of 100 curves, which we will call a sequence of killer whale dives.
+Each dive could be one of $N=2$ dive types and the coarse-scale probability transition matrix was set as
 %
-$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}, \qquad \delta =  \begin{pmatrix} 0.5 & 0.5 \end{pmatrix},$$
-$$Y_t|X_t \sim \rm{Gamma},$$
-\begin{align*}
-	&\mu^{(1)} \equiv \bbE(Y_t|X_t = 1) = 15 s, &\bbE(Y_t|X_t = 2) = 60 s, \\
+$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
 %
-	&\bbV(Y_t|X_t = 1) = 25 s^2, &\bbV(Y_t|X_t = 2) = 100 s^2.
-\end{align*}
+Dive duration was gamma distributed, and the coarse-scale emission parameters $\Theta$ were set as
+$$
+    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
+    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
+$$
 %
-Once the dive durations were calculated for all 100 dives, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments, which made up a second fine-scale hidden Markov model.
+After the dive durations were calculated for all 100 dives for each data set using $\Gamma$ and $\Theta$, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments, which made up the fine-scale hidden Markov model.
 The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive were discarded.
-Each two-second segment was assigned one of $N^*=2$ behaviours (active swimming or passive gliding) according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$.
-The probability transition matrices for these fine-scale Markov chains were set as
+Each two-second segment was assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
 %
-$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix},$$ 
+$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
 %
-where $\Gamma^{*(1)}$ was used for dives where $X_t = 1$ and $\Gamma^{*(2)}$ was used for dives where $X_t = 2$.
-
-Each two-second subdive window had 100 associated acceleration readings, $Y^*_{t,t^*} \equiv \left(Y^*_{t,t^*,1}, \ldots, Y^*_{t,t^*,100}\right)$.
-To accurately recreate active swimming versus passive gliding on the fine-scale Markov chain, the DFT of each two-second segment $\hat Y^*_{t,t^*}$ was simulated such that $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$ (see Eq.
-\ref{eqn:z}) have the following distributions:
-
+and the observations $Z^*$ were generated using the following parameters:
+%
 \begin{gather*}
-    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 1 \right) \sim \mathcal{N} \left(\phi^{*(1)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(1)})\mu^{*(1)}, (\sigma^{*(1)})^2 \right), \\
+    \mu_1^{*(1)} = 0.0 s, \enspace \sigma_1^{*(1)} = 0.05s, \enspace \phi_1^{*(1)} = 0.99, \\
     %
-    \left(Z^{*(1)}_{t,t^*} | Z^{*(1)}_{t,t^*-1}, X^*_{t,t^*} = 2 \right) \sim \mathcal{N} \left(\phi^{*(2)}Z^{*(1)}_{t,t^*-1} + (1-\phi^{*(2)})\mu^{*(2)}, (\sigma^{*(2)})^2 \right), \\
+    \mu_1^{*(2)} = 0.0 s, \enspace \sigma_1^{*(2)} = 0.1s, \enspace \phi_1^{*(2)} = 0.95, \\
     %
-    \mu^{*(1)} = 0.0,\enspace \sigma^{*(1)} = 0.05,\enspace \phi^{*(1)} = 0.99, \\
+    \mu_2^{*(1)} = 10.10, \quad \sigma_2^{*(1)} = 3.18, \\
     %
-    \mu^{*(2)} = 0.0,\enspace \sigma^{*(2)} = 0.1,\enspace \phi^{*(2)} = 0.95, \\\\
-    %
-    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha^{*(1)}, \beta^{*(1)} \right), \\
-    %
-    \left(Z^{*(2)}_{t,t^*} | X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(\alpha^{*(2)}, \beta^{*(2)} \right), \\
-    %
-    \alpha^{*(1)} = 10.10, \enspace \enspace \beta^{*(1)} = 1.00, \\
-    %
-    \alpha^{*(2)} = 305.94, \enspace \enspace \beta^{*(2)} = 1.00.
+    \mu_2^{*(2)} = 305.94, \quad \sigma_2^{*(2)} = 17.49.
 \end{gather*}
 %
-subdive behavior 1 corresponds to passive gliding while subdive behaviour 2 corresponds to active swimming with a dominant frequency of $1/2 s^{-1}$.
-subdive behaviors 1 and 2 are the same for both dive types.
-See the appendix for more details regarding procedure for simulating $\hat Y^*$ and $Y^*$ such that $Z^*$ has the preceding distribution.
-Figure \ref{fig:sim_data} shows the first 5 dives of one simulated data set.
+Recall that $Z^{*(1)}$ follows a CarHMM and is Normally distributed while $Z^{*(2)}$ follows an HMM and is Gamma distributed.
+It is not possible to uniquely recreate the raw fine-scale data $Y^*$ from $Z^*$ alone, but we describe a method to reconstruct one consistent sequence of observations $Y^*$ in the appendix.
+Figure \ref{fig:sim_data} shows this reconstructed sequence along with $Z^*$ for five dives of one simulated data set.
 
-\subsection{Model Formulation}
-\label{subsec:model_structure}
-
-The building blocks from the previous section were used to build a well-specified model for these simulated data.
-Specifically, we used a hierarchical HMM where the sequence of dive durations $Y$ was modeled using a simple HMM, and the fine-scale process $Z^*$ was modeled using a HMM-DFT with explicit auto-correlation in $Z^{*(1)}$.
-Naturally, we refer to this model as the \textbf{CarHHMM-DFT}.
-Figure \ref{fig:CarHHMM-DFT} shows a graphical representation of this model.
-
-On the coarse scale, the dive types follow a Markov chain with $N=2$ possible states and unknown probability transition matrix $\Gamma$.
-The duration of a dive follows a gamma distribution which depends upon the dive type and unknown parameters $\Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\}$.
-
-On the fine scale, the subdive behavior of each two-second window comprises a Markov chain with $N^*=2$ possible states and unknown probability transition matrices $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$, depending upon the dive type.
-Each two-second window is summarized by the observations $Z^{*(1)}_{t,t^*}$ and $Z^{*(2)}_{t,t^*}$.
-The distribution of $Z^{*(1)}_{t,t^*}$ is Normal and its parameters depend upon the subdive behavior $X^*_{t,t^*}$ and $Z^{*(1)}_{t,t^*-1}$.
-In particular:
-
-$$\mathbb{E}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \phi_1^{*(i)}z + (1-\phi_1^{*(i)}) \mu_1^{*(i)}$$
-$$\mathbb{V}(Z^{*(1)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_1^{*(i)}\right)^2.$$
-%
-The distribution of $Z^{*(2)}_{t,t^*}$ is gamma and its parameters depend upon only $X^*_{t,t^*}$ (not $Z^{*(2)}_{t,t^*-1}$):
-%
-$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(2)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
-$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(2)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
-%
-None of the fine-scale emission distributions depend upon dive type. In total the parameters to estimate are
-%
-\begin{gather*}
-    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
-    %
-    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters), and} \\
-    %
-    \Theta^* = \{\Theta^{*(1)},\Theta^{*(2)}\}  \qquad \text{(fine-scale emission parameters), where} \\
-    %
-    \Theta^{*(i)} =  \{\{\mu_1^{*(i)},\sigma_1^{*(i)},\phi_1^{*(i)}\},\{\mu_2^{*(i)},\sigma_2^{*(i)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
-\end{gather*}
-%
-The likelihood of this model is still easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
-See the appendix for details of likelihood evaluation.
-Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model.
-
-Four different models were fit to the simulated data sets:
-\begin{enumerate}
-    \item A \textbf{CarHHMM-DFT} as described above.
-    \item An \textbf{HHMM-DFT}, which is similar to the CarHHMM-DFT, but with no modeled auto-correlation \big(i.e. $\phi_1^{*(i)} = 0$ for $i = 1,2$\big).
-    \item A \textbf{CarHHMM}, which is also similar to the CarHHMM-DFT, but without the Fourier sums $(Z^{*(2)})$ as an observation \big(i.e $\Theta^{*(i)} = \{\mu_1^{(i)},\sigma_1^{(i)},\phi_1^{(i)}\}$, $i = 1,2$\big).
-    \item A \textbf{CarHMM-DFT}, which is again similar to the CarHHMM-DFT, but with $N=1$ instead of $N=2$ on the coarse scale.
-This is equivalent to loosing the coarse level of the hierarchical structure, as there is only one dive type.
-\end{enumerate}
-%
-Each of the last three models leaves out one important aspect of the full CarHHMM-DFT.
-The CarHMM-DFT lacks a hierarchical structure, the HHMM-DFT is missing auto-correlation within the fine-scale observations, and the CarHHMM does not have access to the Fourier transform sums $(Z^{*(2)})$ as observations.
-
-All models were run on the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
+The two subdive types differ primarily due to $\mu_2^*$ and $\sigma_2^*$, both of which are much higher for subdive type 2.
+This corresponds to much more vigorous and variable periodic behavior in the acceleration data when the simulated whale is in subdive type 2.
 
@@
 
-%%%
+All four candidate models (CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT) were fit to all 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
 
 As expected, the full CarHHMM-DFT was the best-performing model.
-It was able to decode the fine-scale hidden states of the process almost perfectly (mean error $< 10^{-12}$ for all simulations) and decode the hidden dive types with an average accuracy of $94 \pm 3 \%$, where $\pm$ refers to the standard deviation across simulations (see Table \ref{table:accuracy}).
-All parameter estimates have biases which are at least approximately one order of magnitude smaller than the associated standard error, which is consistent with a sample size of 500 (section S-1.2, tables 2-5).
-One exception is for estimates of $\hat \sigma$, which were regularly underestimated.
-For dive duration $(Y_t)$ this finding is consistent with properties of MLEs for standard deviation, especially with a small sample size.
-For acceleration $(Z^{*(1)})$ this is likely due to over-fitting the auto-correlation parameter $\phi$.
-The Fourier sums $(Z^{*(2)})$ show no significant bias in $\hat \sigma^*_2$.
-The empirical standard error of all parameters estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) were well-approximated by the inverse of the observed Fisher information matrix (section S-1.2, tables 2-5).
-One exception is $\hat \Theta$, whose standard errors were slightly underestimated using the observed Fisher information matrix.
-%The empirical standard error of $\hat \mu_2$ was 8.5, while the estimate of standard error using the observed Fisher information was 5.8. 
+Given the true subdive type of a particular segment $x^*_{t,t^*} = i$, the CarHHMM-DFT would predict the correct subdive type with an average confidence of $1 - 10^{-12}$ across all simulations.
+For the dive types, this average confidence was above 0.9 (see Table \ref{table:accuracy}).
+Estimates of hidden states were made using the forward-backward algorithm.
+All parameter estimates had statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased.
+Even still, these biases are the least severe for the CarHHMM-DFT among the four models.
+The empirical standard error of all parameters estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) were well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tended to be slight underestimates of the empirical standard error.
 See Section S-1.2, tables 2-5 for more information.
@@
 The HHMM-DFT performed almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias.
-%- it had very low error for the fine-scale hidden states and $\approx 94\%$ accuracy on the dive types (see Table \ref{table:accuracy}).
-In addition, the standard errors of the parameter estimates were comparable to those of the CarHHMM-DFT.
-The estimates of standard error using the observed Fisher information were also comparable to the CarHHMM-DFT.
-The HHMM-DFT \textit{did} differ from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$.
+The HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$.
 For both subdive states, $\hat \sigma_1^*$ greatly overestimated the true standard deviation.
-In particular, $\overline{\hat \sigma_1^{*(1)}} = 0.248$ and $\overline{\hat \sigma_1^{*(2)}} = 0.242$, but $\sigma_1^{*(1)} = 0.05$ and $\sigma_1^{*(2)} = 0.1$.
-In addition, the empirical standard error of both $\hat \mu_1^*$ and $\hat \sigma_1^*$ was much larger than the estimates of standard error using the observed Fisher information (see Section S-1.2, Table 3).
-This implies that parameter uncertainty can be underestimated when auto-correlation is ignored.
+In addition, the empirical standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ were both much larger than the estimates of standard error using the observed Fisher information (see Section S-1.2, Table 3).
+This implies that parameter estimates can be too confident when auto-correlation is ignored.
 
-The CarHHMM was the worst-performing model in terms of decoding accuracy, with an accuracy of $89 \pm 1\%$ when determining subdive behavior and an accuracy of $87 \pm 11\%$ when determining dive types.
-It was much more likely to misclassify subdive behavior 1 than subdive behavior 2 (see Section S-1.2, Table 1).
+The CarHHMM was the worst-performing model in terms of decoding accuracy, with an average confidence of $89 \pm 1\%$ when determining subdive behavior and an average confidence of $87 \pm 11\%$ when determining dive types.
+It was much more likely to mis-classify subdive behavior 1 than subdive behavior 2 (see Section S-1.2, Table 1).
 It was also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy.
@@
 Finally, the CarHMM-DFT was comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations.
-In addition, the time required to fit the CarHHMM was less than half of that of the other models (Tbl \ref{table:accuracy}).
-However, it lacked any hierarchical structure, so it was unable to decode the coarse-scale dive types.
+In addition, the time required to fit the CarHMM-DFT was less than half of that of the other models (see Table \ref{table:accuracy}).
+However, it lacked any hierarchical structure, so it was unable to estimate the coarse-scale dive types.
 The CarHMM-DFT still did fit a gamma distribution to the dive duration of \textit{all} dives.
-The parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds.
+These parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds.
 

### 2020-07-27 Report/Sections/intro.tex
Update Intro- Nancy

@@
 Movement data are sometimes modelled as a set of curves that can  be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}).
-In \cite{Morris:2007}, a child's activity level throughout the day can be viewed as a curve, specifically, as activity as a function of time.
-In the data  we consider, a single dive of a whale can be viewed as a curve:  acceleration as a function of time.
+In \cite{Morris:2007}, a child's activity level throughout the day is viewed as a curve, specifically, as activity as a function of time.
+In the data  we consider, a single dive of a whale is viewed as several curves:  acceleration and depth as  functions of time.
 FDA was originally developed for curves that can be considered independent replicates.
@@
 Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
-The plot shows one component of the acceleration, and from this we can see fine-scale behaviour within each dive,  behaviour that seems to indicate some fine-scale short-term periodicity. 
+The top panel  shows a summary  of the acceleration, and from this we can see fine-scale behaviour  that seems to indicate some  short-term periodicity. 
 We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour. 
@@
 
-Particular types of between curve dependence structures have received attention in the FDA literature, 
-notablly multilevel models that incorporate dependence via random effects.
+Models for particular types of between curve dependence structures appear in the FDA literature, 
+notably multilevel models that incorporate dependence via random effects.
 For instance, in the aforementioned study of children's activity levels in  \cite{Morris:2007}, the authors use a multilevel approach since each child yields not one but two daily activity curves and so the model includes a random effect for ``child''.
-Perhaps the first work with this multilevel structure is \cite{Bromback:1998} in a study of ....
-More recent work on multilevel models includes that of \cite{di:2009}, \cite{chen:2012}} and Crainiceanu [REF].
-For our data, the multilevel models are not appropriate, since they do not account for the temporal dependence.
+In a classic early paper that introduced this multilevel structure,  \cite{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles.
+More recent work on multilevel models includes that of \cite{di:2009}, \cite{chen:2012} and Crainiceanu [REF].
+For our data, multilevel models are not appropriate, since they do not account for the temporal dependence in the sequence of dives nor do they account for different dive types.
 
@@
 While  functional time series methods can be applied to our dive data, they would not account for the different dive types.
-Our HMM models account for both temporal dive-to-dive dependence and dive types.
+Our HMM models account for both temporal dive-to-dive dependence and different types of dives.
 
@@
 However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
-We model the within-curve dependence by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence.
+We model this behaviour by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence.
 Finally, for the fine-scale behaviour, we use summary information that is biologically meaninful.
@@
 Animal movement modelers have also used  hierarchical hidden Markov models (HHMMs) to model  simultaneous behavioral processes that occur at different time scales \citep{Barajas:2017,Adam:2019}, and we borrow from that work here.
-Part of the purpose of our article is to introduce the FDA community to these very useful approaches to modelling complex curves.
+Part of the purpose of our article is to introduce the FDA community to these useful approaches to modelling complex curves.
 
@@
 Several solutions have been proposed, such as  the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
-The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
+The CarHMM  explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
 

### 2020-07-30 Report/Sections/data.tex
Update Report

@@
 
-Acceleration data can be used to approximate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, which can be helpful for conservation efforts.
-However, studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
-Understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the energetic requirements of killer whales.
-This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.
+Understanding animal behaviour can be important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}.
+NRKW prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus
+tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKW for predation \citep{Williams:2009,Noren:2011}.
+It is therefore of interest to determine the energetic expenditure of NRKW to forage for Chinook, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
 
-We begin by describing the data collection process used by ecologists.
-We then detail the procedure used to build several candidate generative models from this data.
-Next, we perform a simulation study by selecting a preferred model, simulating data from it, and fit the candidate models to the simulated data to compare their performance in terms of parameter estimates and decoding accuracy.
-Finally, we fit each candidate model to the original data and analyze the results.
+Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
+Therefore, understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the true energetic requirements of NRKW.
 
@@
 
-The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions.
+The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions.
 Observations were collected at a rate of 50 Hz.
@@
 Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}.
-After preprocessing the raw data, a total of 267 dives were observed.
+The preprocessed data contains a total of 267 dives.
 Figure \ref{fig:data} displays the dive profile and accelerometer data.
 
-Each dive is treated as one curve, and the collection of dives make up the coarse-scale process.
-As such, the coarse-scale observations $Y$ were a time series of dive durations in seconds ($s$) with corresponding hidden dive types $X$.
-In addition, the fine-scale processes $Y^*$ were the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive types $X^*$.
+Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
+As such, the coarse-scale observations $Y$ are a time series of dive durations in seconds ($s$) with corresponding hidden dive types $X$.
+In addition, the fine-scale processes $Y^*$ are the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive states $X^*$.
 
@@
 
-The coarse-scale level exhibited no significant amount of auto-correlation (See section S-2.1) and no obvious intricate structure.
+A lag plot was constructed to determine where auto-correlation should be incorporated into the model (see Section S-2.1).
+Prior to fitting the model, the coarse-scale level exhibited no significant amount of auto-correlation and no obvious intricate structure.
 As such, we selected a simple HMM to model this data.
 The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
-$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)},$$
-$$\mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
 
-The fine-scale acceleration data exhibited significant sinusoidal behavior, thus the fine-scale observations $y^*$ were transformed to $z^*$ using the DFT summary statistics of a two-second sliding window.
-A two-second window was selected heuristically since it appeared to be the smallest window size that could capture the sinusoidal behavior.
+$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
+
+The fine-scale acceleration data exhibited significant sinusoidal behavior; thus the fine-scale observations $Y^*$ were transformed to $Z^*$ using the DFT summary statistics of a two-second sliding window.
+A two-second window was selected heuristically since it appeared to be the smallest window size that could capture the.
+``wiggles" in their entirety.
 Because the data was collected at the rate of 50 Hz, each two-second window contained a total of 100 raw acceleration readings.
 The acceleration data here was collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
-As a result, the transformation from the raw accelerometer data $\mathbf{y}^*$ to the observation sequence $z^*$ was calculated as follows:
+As a result, the transformation from the raw accelerometer data $\mathbf{Y}^*$ to the observation sequence $Z^*$ is calculated as follows:
 %
-$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2.$$
+\begin{equation*}
+    Z_{t,t^*}^{*(1)} := \text{Re}\left(DFT\{Y^*_{t,t^*}\}(0)\right) \qquad Z_{t,t^*}^{*(2)} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2
+\end{equation*}
 %
-Each observation was therefore made up of $\mathbf{z}_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $z_t^{*(2)}$, a scalar.
-We calculated $z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum frequency of $\tilde \omega = 5$ Hz.
+Each observation is therefore made up of $Z_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $Z_{t,t^*}^{*(2)}$, a scalar.
+We calculated $Z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum frequency of $\tilde \omega = 5$ Hz.
 
-Even after accounting for fine-scale structure in the raw acceleration data, there was still strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ (section S-2.1), thus a CarHMM was used to model $\mathbf{Z}^{*(1)}_{t,t^*}$.
-In particular, we assumed $\mathbf{Z}^{*(1)}_{t,t^*}$ to be Normally distributed with unknown parameters $\phi_1^{*(i)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i)} \in \mathbb{R}^3$.
+Even after accounting for fine-scale structure in the raw acceleration data, there was still strong auto-correlation within $Z^{*(1)}_{t,t^*}$ prior to fitting the model (section S-2.1).
+Therefore, a CarHMM was used to model $Z^{*(1)}_{t,t^*}$.
+In particular, if $X^*_{t,t^*} = i^*$, then we assume $Z^{*(1)}_{t,t^*}$ to be Normally distributed with unknown parameters $\phi_1^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i^*)} \in \mathbb{R}^3$.
 %
@@
 
-While $Z^{*(2)}$ also exhibits some auto-correlation, the relationship is less strong than that of $\mathbf{Z}^{*(1)}$, and the biological interpretation of auto-correlation within $Z^{*(2)}$ is less clear.
-Therefore, a simple HMM was used to model $Z^{*(2)}$ and the distribution of $Z^{*(2)}_{t,t^*}$ was assumed to be gamma and parameterized by its mean $\mu_2^{*(i)}$ and variance $\sigma_2^{*(i)}$.
+While $Z^{*(2)}$ also exhibits some auto-correlation prior to fitting the model, the relationship is less strong than that of $Z^{*(1)}$, and the biological interpretation of auto-correlation within $Z^{*(2)}$ is less clear.
+In addition, much of the auto-correlation that \textit{is} evident from the lag plot may be explained by persistence in the subdive type $X^*_{t,t^*}$.
+Therefore, a simple HMM was used to model $Z^{*(2)}$.
+In addition, given that $X^*_{t,t^*} = i^*$ the distribution of $Z^{*(2)}_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_2^{*(i^*)}$ and variance $\sigma_2^{*(i^*)}$.
 %
@@
 %
-The observations $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ shared a subdive state $X^*_{t,t^*}$, but were assumed to be independent of one another when conditioned on $X_{t,t^*}$.
-In addition, the distribution of $Z^*_{t,t^*}$ is assumed to depend only upon the subdive type and the previous observation $Z^*_{t,t^*-1}$, not the dive type $X_t$.
 
-Finally, to select the number of dive types and subdive types, information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}.
-Instead, we selected $N = 2$ dive types and $N^* = 3$ subdive behaviors heuristically.
-The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
+We do not use information criteria to select the number of dive types and subdive types since it tends to overestimate the number of states in biological processes \citep{Pohle:2017}.
+In the particular case of well-separated hidden states, a lag plot will reveal $N$ circular patterns for each dive feature and $N^*$ circular patterns for each subdive feature.
+This is unfortunately not the case for the killer whale data, but we still consult the lag plots together with a heuristic approach to choose $N = 2$ dive types and $N^* = 3$ subdive behaviors.
+The absence of a more principled method to select the number of hidden states highlights the importance of model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
 
 The final model is a hierarchical hidden Markov model with a traditional HMM at the coarse scale.
-On the fine scale, $y^*$ was transformed to $z^*$, $z^{*(1)}$ was modeled with a CarHMM, and $z^{*(2)}$ was modeled with an HMM.
+On the fine scale, $Y^*$ is transformed to $Z^*$, $Z^{*(1)}$ is modeled with a CarHMM, and $Z^{*(2)}$ is modeled with an HMM.
+In addition, we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity.
+This implies that dive types 1 and 2 share subdive states.
 We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMM models discussed here.
@@
     %
-    \Theta^* = \{\Theta^{*(1)},\Theta^{*(2)}\}  \qquad \text{(fine-scale emission parameters), where} \\
+    \Theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{(fine-scale emission parameters), where} \\
     %
-    \Theta^{*(i)} =  \{\{\mu_1^{*(i)},\sigma_1^{*(i)},\phi_1^{*(i)}\},\{\mu_2^{*(i)},\sigma_2^{*(i)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
+    \theta^{*(\cdot,i^*)} =  \{\{\mu_1^{*(\cdot,i^*)},\sigma_1^{*(\cdot,i^*)},\phi_1^{*(\cdot,i^*)}\},\{\mu_2^{*(\cdot,i^*)},\sigma_2^{*(\cdot,i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
 \end{gather*}
 %
-The likelihood of this model is still easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
+Remember that $\theta_j^{*(\cdot,i^*)}$ are the parameters describing the distribution of $Z^{*(j)}_{t,t^*}$ when conditioning on $X^*_{t,t^*} = i^*$.
+The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
 See the appendix for details of likelihood evaluation.
-Figure \ref{fig:CarHHMM-DFT} also shows the corresponding graphical model.
-
-In addition to the preferred CarHHMM-DFT, we consider three similar variations for comparison:
+Figure \ref{fig:CarHHMM-DFT} also shows the corresponding graphical model for the CarHHMM-DFT.
+Additionally, we consider three similar variations for comparison:
 \begin{enumerate}
-    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using the DFT, but models \textit{both} $\mathbf{Z}^{*(1)}$ \textit{and} $Z^{*(2)}$ with a simple HMM rather than a CarHMM.
-    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations \textit{without} using the DFT, and models $Z^{*(1)}$ using a CarHMM.
-The Fourier sums $\mathbf{Z}^{*(2)}$ are omitted as observations in this model.
+    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using the DFT, but models \textit{both} $Z^{*(1)}$ \textit{and} $Z^{*(2)}$ with a simple HMM rather than a CarHMM.
+    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ using a CarHMM.
+However, the Fourier sums $Z^{*(2)}$ are omitted as observations in this model.
+The Fourier sums $Z^{*(2)}$ describe the ``wiggliness" of the accelerometer data, while $Z^{*(1)}$ is simply the average acceleration over a particular window.
     \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $\mathbf{Z}^{*(1)}$ with a CarHMM and $Z^{*(2)}$ with an HMM.

### 2020-07-30 Report/Sections/discussion.tex
Update Report

@@
 
-Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics when conditioned of the hidden state.
+Current FDA literature largely address sequences of curves either as multilevel models with random effects and no temporal dependence \cite{chen:2012,di:2009} or with functional time series which model the curves as an evolving process and overlook the possibility of a discrete number of curve ``types" \cite{Kokoszka:2018}.
+However, many real-world curve sequences exhibit temporal dependence and a discrete number of ``types".
+One obvious example is the case study examined here, where a sequence of marine mammal dives serve a number of discrete of purposes such as traveling, socializing, resting, and foraging \citep{Tennessen:2019a}.
+One other example can be found when measuring accelerometer data in school children, where bouts of activity can be associated with running, walking, etc.
+\citep{Morris:2007}.
+
+To deal with both temporal dependence between curves and a discrete number of curve types, we suggest modelling the sequence of curves with either an HMM or CarHMM, depending upon the level auto-correlation within the data.
+Several useful model selection tools such as the lag plot can test if there is significant auto-correlation within an sequence of curves.
+One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use another within-curve parametric model as emissions.
+The parameters of this fine-scale model in turn will depend upon the curve type, and together these two levels make up a hierarchical model which can simultaneously account for processes taking place at different time-scales.
+
+%Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics when conditioned of the hidden state.
 However, many real-world processes require more complex models.
@@
 
-The CarHMM generalizes the HMM by explicitly modeling auto-correlation in the emission distributions of the HMM \citep{Lawler:2019}.
+%The CarHMM generalizes the HMM by explicitly modeling auto-correlation in the emission distributions of the HMM \citep{Lawler:2019}.
 CarHMMs also maintain the structure needed to evaluate the likelihood using the forward algorithm.
 In our Normal model formulation, we have added only one additional parameter, $\phi^{(i)}$, per possible hidden state.
-Several useful model selection tools such as the lag plot can test if there is significant auto-correlation within an observation sequence.
 
+Several possible choices exist for the fine-scale model, including yet another HMM.
+\citet{Barajas:2017} introduce this model in the context of statistical ecology as an HHMM, and \citet{Adam:2019} use it to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California.
+However, traditional HMMs prove insufficient at very fine-scales, as \citet{Adam:2019} run into clear violations of conditional independence and apparent periodic behavior in the acceleration data.
+This indicates the need to account for complicated dependence structures within fine-scale processes, especially those with high sampling frequencies.
+
+%For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs.
+In particular, each hidden state of the coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.
+
+%For processes such as these with very high sampling frequencies and/or intricate fine-scale structure,
+
+%To deal with these issues, it is possible to generalize the HHMM such that the fine-scale model can be any model which admits an easy-to-calculate likelihood.
+For example, if the sampling rate of the fine-scale process is very high and clearly period, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.
+
+To deal with these problematic fine-scale processes, we suggest using a moving-window transformation of the fine-scale observations.
+For periodic behaviour in particular we use a DFT-based transformation.
+The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when $Z^{*(2)}$ is accounted for, and our case study, where the CarHHMM without DFT visually fails to pick up a killer whale's clear fluking behaviour.
+
+We also suggest using a CarHMM instead of a traditional HMM for the fine-scale process when expert opinion and careful observation (possibly using a log plot) calls for it.
 Although the CarHMM can incorporate auto-correlation into the structure of an HMM, it can break down when observations are taken at irregular time intervals.
@@
 This implies that the parameter estimates from the CarHMM are interpretable in the context of a stochastic differential equation.
-
-For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs.
-In particular, each hidden state of the coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.
-\citet{Adam:2019} utilize the HHMM to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California.
-However, they run into clear violations of conditional independence and apparent periodic behavior when plotting the auto-correlation of acceleration as a function of lag, implying the need to account for complicated dependence structures within fine-scale processes with high sampling frequencies.
-
-%For processes such as these with very high sampling frequencies and/or intricate fine-scale structure, 
-To deal with these issues, it is possible to generalize the HHMM such that the fine-scale model can be any model which admits an easy-to-calculate likelihood.
-For example, if the sampling rate of the fine-scale process is very high and clearly period, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.
-
-In particular, we use $Z^{*(1)}$ and $Z^{*(2)}$ based on the DFT as defined previously.
-These summary statistics prove to be vital when decoding the behavior of a killer whale.
-This is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when $Z^{*(2)}$ is accounted for, and the case study, where the CarHHMM without DFT visually fails to pick up fluking behavior in the killer whale.
 

### 2020-07-30 Report/Sections/intro.tex
Update Report

@@
 We model this behaviour by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence.
-Finally, for the fine-scale behaviour, we use summary information that is biologically meaninful.
-Specifically, for the dive data, we use summaries of Fourier transforms to account for the ``wiggliness'' in some segments of a curve.
+Finally, we use a biologically meaningful transformation on the fine-scale behaviour to account for the ``wiggliness'' in some segments of a curve.
+The distribution of this ``wiggliness'' depends on the hidden state.
+% Specifically, for the dive data, we use summaries of Fourier transforms to account for the ``wiggliness'' in some segments of a curve.
 The distribution of this ``wiggliness'' depends on the hidden state.
@@
 
-We also borrow techniques from animal movement to model fine-scale behaviour, beyond simply using an HMM.  
+We also borrow techniques from animal movement to model within-dive behaviour, beyond simply using an HMM.  
 Under the traditional HMM model,  observations are independent given  the underlying hidden state process.
@@
 
-An additional complication is that, at the shortest time scales,  observations often exhibit complicated  structure which cannot easily be captured by traditional HMMs, CarHMMs, or HHMMs.
-Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}.
+An additional complication seen in the ecology literature is that, at the shortest time scales, observations often exhibit complicated structure which cannot easily be captured by traditional HMMs, CarHMMs, or HHMMs.
+One example in addition to the case study from this work is the swimming patterns of horn sharks off the coast of Southern California \citep{Adam:2019}, both of which exhibit fine-scale periodic behavior.
 %One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation.
@@
 However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
-Our dive data exhibit  fine-scale periodic behaviour, so we transform the data via
- the signal processing technique of the Fourier transform. 
+%Our dive data exhibit fine-scale periodic behaviour, so we transform the data via the Fourier transform. 
 \cite{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM in the context of describing daily behavioral cycles of marine mammals.
-In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data.
+In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior using accelerometer data \citep{Fehlmann:2017,Shorter:2017}.
 Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears a promising simple approach to account for additional correlation in data that is cyclical in nature.
 
-This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. 
-We describe a general procedure that can be used to extract features from highly structured fine-scale behaviors that otherwise could not be modeled with existing HMM models. 
-In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. 
-We then perform a simulation study to compare the performance of each existing model with ours in a controlled setting. 
-Finally, we apply our method to the dive data of Figure 6.
+%This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. 
+%We describe a general procedure that can be used to extract features from highly structured fine-scale behaviors that otherwise could not be modeled with existing HMM models. 
+%In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. 
+%We then perform a simulation study to compare the performance of each existing model with ours in a controlled setting.
 
-In the following section, we describe HMMs, CarHMMs and HHMMs.
-We show how these models can be combined in order to analyze complex data.
-We also discuss transforming data.
-In Section 3, we study various models for complex ...
-Section 4 contains data analysis and Section 5 discussion.
+In the following section, we describe HMMs, generalize them to CarHMMs, introduce hierarchical structure with HHMMs, and discuss methods to transform intricate fine-scale data.
+We also show how these models can be combined in order to analyze complex data.
+In Section 3, we construct several candidate models based on the data from Figure \ref{fig:data} and Section 4 details a simulation study based on these candidate models.
+In Section 5 we fit each candidate model to the original data and analyze the results, and Section 6 contains our discussion of the results.

### 2020-07-30 Report/Sections/models.tex
Update Report

@@
 
-Hidden Markov models are useful tools to model state-switching Markovian processes in discrete time and will be used as the core structure of our models.
-Traditional HMMs model process auto-correlation by assuming that the underlying behavior of a processes follows a single Markov chain.
+Consider a sequence of $T$ possibly high-dimensional curves (or dives, in our context) with associated curve-level or coarse-scale observations $Y \equiv \left(Y_1, \ldots, Y_T\right)$.
+The coarse-scale observations contain summaries of each curve, such as duration, maximum value, variance, etc.
+In addition, curve $t$ has a total of $T^*_t$ within-curve or fine-scale associated observations: $Y^*_t \equiv \left(Y^*_{t,1}, \ldots Y^*_{t,t^*}, \ldots, Y^*_{t,T^*_t}\right)$.
+The fine-scale observations $Y^*_t$ make up a vector of function values from curve $t$, and we index $Y^*_t$ using $t^*$.
+The set of all fine-scale observations is denoted at $Y^* \equiv \left(Y^*_1,\ldots,Y^*_T\right)$.
+
+Hidden Markov models are useful tools to model state-switching Markovian processes in discrete time and will be used as the core structure of the models we use to describe $Y$ and $Y^*$.
+%Traditional HMMs model process auto-correlation by assuming that the underlying behavior of a processes follows a single Markov chain.
 However, when conditioned on the hidden state sequence observations are assumed to be independent.
@@
 You can use some of the text I removed from below}.
-To account for each of these common additional dependence structures, we will explore three variations on the traditional HMM.
-First, we explain the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
-Next, we review the hierarchical hidden Markov model (HHMM), which models simultaneous temporal dependencies at different time scales as two distinct hidden Markov models in a hierarchical model structure.
-Finally, we describe an HMM over a discrete Fourier transform (HMM-DFT), which applies the DFT to a moving window over the raw data to summarize its periodic behavior as a function of time.
-We will show how each of these variations can be altered and combined to form a wide variety of new models.
+%To account for each of these common additional dependence structures,
+We first describe the structure of a traditional HMM and a variation introduced by \cite{Lawler:2019} called the the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
+Neither the traditional HMM nor the CarHMM have any hierarchical structure, so both can only be used to model $Y$ or as a sub-model for a particular $Y^*_t$.
+In order to simultaneously model both $Y$ and $Y^*$, we describe the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}, which simultaneously models $Y$ and $Y^*$ as two distinct (but related) hidden Markov models.
+Next, we highlight the need to deal with intricate dependence structure by applying transformations to the observations using subject area expertise.
+This procedure is particularly necessary for $Y^*$.
+One particular transformation that we use applies the DFT to a moving window over $Y^*$ to summarize its periodic behavior.
+Finally, we show how each of these models can be altered and combined to form a wide variety of new models.
 
@@
 
-An HMM is comprised of a sequence of unobserved states $X_t$, $t = 1, \ldots, T$, and an associated sequence of  possibly high-dimensional observations $Y_t$, $t = 1, \ldots, T$.
-The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time. 
+An HMM is comprised of a sequence of unobserved states $X_t$, $t = 1, \ldots, T$, which are associated with the observations $Y_t$, $t = 1, \ldots, T$.
+Here we focus on the coarse-scale observations $Y$, but an HMM can also be used as a sub-model for a particular fine-scale process $Y_t^*$.
+In the context of an HMM, the $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time.
 The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$.
@@
 If $X_t = i$, then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes simply $f^{(i)}(\cdot)$.
+If $Y_t$ or $Y^*_{t,t^*}$ is a vector, we assume that the dimensions of the observations are independent $\left(f^{(i)}(\cdot) = \prod_{d} f^{(i)}_d(\cdot)\right)$, but this need not be the case in general.
 %
@@
 %
-We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $ X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
+We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
 For example, one could assume that this conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
@@
 
-\subsection{Incorporating correlation on multiple time scales with an HHMM}
+\subsection{Incorporating both $Y$ and $Y^*$ with an HHMM}
 
-An HHMM accounts for different levels of correlation by modelling a coarse-scale process and a fine-scale process, each of which with an HMM \citep{Barajas:2017,Adam:2019}.
-The coarse-scale process is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observed responses.
+An HHMM accounts for different levels of correlation by modelling both the coarse-scale process and fine-scale process with an HMM \citep{Barajas:2017,Adam:2019}.
+The coarse-scale process is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations.
 %
-In the hierarchical setting, each state $X_t$ emits yet another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
+In the hierarchical setting, however, each state $X_t$ emits yet another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
 The fine-scale process ($X_t^*, Y_t^*$) makes up another HMM with parameters that depend on the value of $X_t$.
@@
 
-The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
-We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\Theta^{*(i)}=\left(\theta^{(i,1)}, \ldots, \theta^{(i,N^*)}\right)$.
+The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
+We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\Theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^*)}\right)$.
 
@@
 Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states.
-For example, in the killer whale case study (Section \ref{sec:case_study}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states (i.e.
-$\theta^{(1,i^*)} = \ldots = \theta^{(N,i^*)}$ for $i^* = 1, \ldots, N^*$).
+For example, in the killer whale case study (Section \ref{sec:case_study}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( \theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1, \ldots, N^* \right)$.
 Figure \ref{fig:models} represents the dependence structure for an HHMM.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Accounting for periodicity with the HMM-DFT}
+\subsection{Transforming $Y_t^*$ to account for fine-scale behaviour}
 \label{subsec:STFT}
 
-The HMM-DFT incorporates hierarchical structure into an HMM differently than an HHMM (Fig \ref{fig:models}).
-In particular, the fine-scale process is no longer modeled with an HMM and instead summarized using its Fourier transform.
-For simplicity, we assume that the length of the fine-scale processes is constant (i.e.
-that $T^*_t = T^*$), although this need not be the case.
-Suppose that the fine-scale process $y^*_t$ does not switch hidden states, but does exhibit significant periodic behaviour.
-We then suggest using the discrete Fourier transform (DFT) on $y^*_t$:
+In many applications where data is collected at high frequencies, intricate dependency structures may arise within the fine-scale process $Y^*$ which cannot be adequately modeled with any of the HMM-based models mentioned thus far.
+For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behavior.
+To deal with this issue, we recommend replacing the fine-scale process $Y_t^*$ with relevant statistics that summarize this non-Markovian behavior.
+In order to maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over $Y_t^*$.
+Subject-area experts are often required to determine the optimal window size and the specific summary statistics used.
+
+%This moving window approach complicates the indexing of $Y_t^*$.
+Suppose a window size of $h$ observations is selected.
+Moving forward, we re-index $Y^*$ such that the fine-scale process associated with curve $t$ is denoted as $Y^*_t$ as before.
+However, the $t^{*th}$ \textit{window} of $Y^*_t$ is now denoted as $Y^*_{t,t^*}$, and the $t^{**th}$ \textit{observation} of $Y^*_{t,t^*}$ is denoted as $Y^*_{t,t^*,t^{**}}$.
+As a result, there are a total of $T^*_t$ windows associated with $Y_t^*$, each of which are of length $h$.
+Note that this relabeling may significantly reduce the length of $Y^*_t$ depending upon the window size $h$ and the step-size between windows.
+
+For our particular application, we transform the fine-scale process using the discrete Fourier transform (DFT) of a two-second moving window across $Y^*$, where the moving window has a step size of two seconds:
 %
 \begin{align*}
-    DFT\{y^*_t\}(k) := \hat{y}^{*(k)}_{t} = \sum_{t^* = 1}^{T^*} y^*_{t,t^*}\exp\left(-i \frac{2\pi k}{T^*} (t^*-1)\right), \quad k = 0, 1, \ldots, T^*-1.
+    DFT\{Y^*_{t,t^*}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right),
 \end{align*}
 %
-Summary statistics can then drastically reduce the dimension of $\hat{y}^*_t$. One example is as follows:
+for $k = 0, 1, \ldots, h-1$.
+The value of $h$ is equal to the window size, which is 100 for a two-second window over 50 Hz observations.
+Summary statistics are then used to drastically reduce the dimension of the DFT:
 %
@@
     \label{eqn:z}
-    z_t^{*(1)} := \mathcal{R}\left(\hat{y}^{(0)}_t\right) \qquad z_t^{*(2)} := \frac{1}{T^*}\sum_{k=1}^{\tilde{\omega}}|\hat{y}^{(k)}_t|^2
+    Z_{t,t^*}^{*(1)} := \text{Re}\left(DFT\{Y^*_{t,t^*}\}(0)\right) \qquad Z_{t,t^*}^{*(2)} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|^2
 \end{equation}
 %
-In words, $z_t^{*(1)}$ is the average value of $y^*_t$ and $z_t^{*(2)}$ is the squared 2-norm of the component of $y^*_t$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window length $T^*$.
+In words, $Z_{t,t^*}^{*(1)}$ is the average value of $\{Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\}$ and $Z_{t,t^*}^{*(2)}$ is the squared 2-norm of the component of $\{Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\}$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window (by the Plancherel theorem).
 The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care.
-These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $y^*_t$.
-Figure \ref{fig:models} represents the structure of the HMM-DFT.
+These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $Y^*_{t,t^*}$.
+Figure \ref{fig:models} represents the process of transforming $Y^*_{t,t^*}$ into $Z^*_{t,t^*}$.
+To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $Z^*_t$ such that $Z^*_t = \left(Z^*_{t,0},Z^*_{t,h},Z^*_{t,2h},\ldots\right)$.
+This both removes residual correlation and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
 
-Once $z^*_t$ is calculated, it can be treated as an observation of the HMM and incorporated into the emission distribution $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$.
-The likelihood of the HMM-DFT is as follows:
-\begin{equation}
-    \calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
-    \label{HMMDFT_likelihood}
-\end{equation}
-where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.
+Once $Z^*_t$ is calculated, it simply replaces the original fine-scale process $Y^*_t$ in the HHMM structure.
+As such, the emission distributions and the likelihood remain identical to the original HHMM when transforming $Y^*$ to $Z^*$.
+We refer to an HHMM which uses $Z^*$ as fine-scale observations as an HHMM-DFT.
 
-It is possible to accommodate unequal time steps within $y_t^*$ by using the non-uniform discrete Fourier transform (NDFT).
+%so that the fine-scale emission distribution becomes $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$. The likelihood of the HMM-DFT is as follows:
+%\begin{equation}
+%    \calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
+%    \label{HMMDFT_likelihood}
+%\end{equation}
+%where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.
+
+%It is possible to accommodate unequal time steps within $y_t^*$ by using the non-uniform discrete Fourier transform (NDFT).
 We do not describe the method in detail here, but the generalization is straightforward.
@@
 
-In addition to the models described above, the fine-scale process $Y^*_t$ can be modeled using \textit{any} parametric model which admits an easy-to-compute likelihood.
-The fine-scale likelihood $\calL_{\text{HMM}}$ from the HHMM likelihood can therefore be replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(y^*_t;\Theta^{*(i)})$:
+The HHMM by definition models the fine-scale process with a traditional HMM.
+However, in principle $Y^*_t$ can be modeled using \textit{any} parametric model which admits an easy-to-compute likelihood.
+The fine-scale likelihood $\calL_{\text{HMM}}$ from the HHMM likelihood would simply be replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(y^*_t;\Theta^{*(i)})$:
 \[
@@
 where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
-This definition is straightforward to extend to the CarHMM as well:
+The coarse-scale HMM can also be generalized by replacing \textit{it} with a CarHMM:
 \[
@@
 
-Possible candidates for the fine-scale model include, but are not limited to, any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HMM-DFT).
+Possible candidates for the fine-scale model include, but are not limited to, any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT).
 These alternative models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models based on HMMs.
-In the sections that follow, we perform both a simulation study and real-world case study modelling killer whale dive behaviour using models constructed from these building blocks.
+One possible model is, say, the \textbf{CarHHMM}, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM.
+Alternatively, one may construct an \textbf{HHHMM}, where three levels of observations are recorded ($Y$, $Y^*$, and $Y^{**}$), and traditional HMMs are used to model all three levels.
+It is clear that infinitely many possible models can be built using this framework.
+However, these models can quickly become complicated and should be constructed achieve an adequate fit of the data while avoiding over-fitting and high computational costs.
+%In the sections that follow, we construct several candidate models from the building blocks above to describe killer whale dive behavior.
+We select one preferred model and compare it to several of the simpler candidate models to ensure its complexity is necessary.

### 2020-07-30 Report/Sections/results.tex
Update Intro- Nancy
Update Report

@@
 
-We report the results from fitting the CarHHMM-DFT in detail, validate the results, and then compare the results of this model with those from the other candidate models.
+After fitting the four candidate models to simulated data, we repeat the same exercises and fit all four models to the data shown in Figure \ref{fig:data}.
+Here we report the results from the full CarHHMM-DFT defined in Section \ref{sec:data} in detail and asses the quality of the fit.
+We also compare the results of the CarHHMM-DFT with the other candidate models.
 
-The estimates for the emission distribution parameters suggest that the killer whale has at least two distinct dive behaviors (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). 
-Dive type 1 corresponds to shorter, shallower dives.
-Ecologists have attributed a variety of purposes to such short dives, including resting (CITATION, you can ask Ian and Sarah for help here - look at the Tennessen paper with HMM).
+The estimates for the emission distribution parameters suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}).
+Dive type 1 corresponds to shorter, shallower dives. Ecologists have attributed a variety of purposes to such short dives, including resting, traveling, and searching for prey \citep{Tennessen:2019a}.
 Dive type 2 is longer and deeper.
-These types of deep sustained dives have been associated with foraging behaviors in killer whales (Citation Tennessen maybe HMM and the other Tennesen with foraging data).
+These types of deep sustained dives have been associated with behaviours such as foraging and socialization in killer whales \citep{Tennessen:2019a,Tennessen:2019b}.
+However, \citet{Wright:2017} note that most prey captures occur at depths deeper than 100m, so it is unlikely that any of the 267 dives used in this study are successful foraging dives.
 
-The three subdive behaviors also appeared distinct with respect to their Fourier sums, as the mean of $Z^{*(2)}$ is separated by an order of magnitude between the three subdive types (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}).
-For subdive type 1, the mean of $Z^{*(2)}$ and the variance of $\mathbf{Z}^{*(1)}$ in this state is smaller than the other subdive types.
-The auto-correlation of $\mathbf{Z}^{*(1)}$ also is higher for subdive type 1 than every other subdive type.
-This corresponds to less overall activity compared to the other subdive types.
-Subdive type 2 is characterized by a mean Fourier sum ($Z^{*(2)}$) one order of magnitude higher than subdive type 1.
-Acceleration has about twice the variance in this subdive type compared to subdive type 1, and slightly lower auto-correlation.
-Subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behavior in acceleration is characteristic of this behavior in marine mammals \citep{Simon:2012}.
-In subdive type 3, the mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ is much higher compared to every other state.
-The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this subdive type, implying more variation in acceleration.
-This corresponds to vigorous swimming activity, especially as the killer whale begins a dive or surfaces from a dive.
+The three subdive behaviours have distinct $Z^{*2}$ values, as the means of each are separated by an order of magnitude. (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}). 
+Subdive state 1 has the smallest mean $Z^{*(2)}$ and the smallest variance of $Z^{*(1)}$.
+It also has the highest auto-correlation in $Z^{*(1)}$.
+This corresponds to less overall activity and more consistent acceleration compared to the other subdive states.
+Subdive type 2 has a mean Fourier sum ($Z^{*(2)}$) one order of magnitude higher than subdive state 1, and its acceleration has about twice the variance compared to subdive state 1.
+The auto-correlation of acceleration is also slightly lower in this state.
+We hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
+The mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ in subdive state 3 is much higher than the other two states.
+The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower, implying more variation in acceleration between consecutive segments.
+This corresponds to vigorous swimming activity, especially as the killer whale begins a dive or surfaces from a dive (see Figure \ref{fig:labeled_dives}).
 
@@
 %
+for the transitions from dive to dive. For dive types 1 and 2,
 $$\hat \Gamma^{*(1)} = \begin{pmatrix} 
@@
 \end{pmatrix},$$
-$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}.$$
+$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
 %
-The coarse-scale stationary distribution indicates that most dives are short type 1 dives.
-The coarse-scale transition probability matrix further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive.
-The fine-scale stationary distributions indicate that this killer whale is more likely to be in the less active subdive type 1 when performing long deep dives than when performing short shallow dives.
-Using less active swimming behavior when diving, and holding breath, for long periods of time has been suggested to be an energy reduction strategy in bottle-nose dolphins \citep{Williams:1999}.
-Figure \ref{fig:labeled_dives} shows the decoded dive behavior of 6 selected dives.
-The supplementary material also shows the estimated probability of each dive and subdive type (section S-2.2).
+for the fine-scale within-dive transitions.
+The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that most dives are short type 1 dives.
+The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive.
+This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short dives before a killer whale performs a longer, deeper dive.
+The fine-scale stationary distributions $\hat{\delta^{*(i^*)}}$ indicate that this killer whale is more likely to be in the less active subdive state 1 when performing long deep dives than when performing short shallow dives.
+Using less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}.
+Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of 6 selected dives.
+The supplementary material also shows the estimated probability of each dive and subdive state given the data (Section S-2.2).
 
@@
 Two visual tools were used to evaluate this model: pseudo-residuals and empirical histograms.
-A pseudo-residual of a particular observation is the marginal CDF of an observation conditioned on all other observations under the learned model \citep{Zucchini:2016}.
+A pseudo-residual of a particular observation is the marginal cumulative distribution function of an observation conditioned on all other observations under the learned model \citep{Zucchini:2016}.
 To easily visualize outliers, this pseudo-residual is often passed through the quantile function of the standard Normal distribution.
@@
 We find that histograms of the pseudoresiduals of this model mostly support that the model is well-specified.
-$Z^{*(2)}$ is an exception, as its pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
+One exception is $Z^{*(2)}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
 This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution than the gamma distribution.
 
-We also plotted histograms of each feature where observations were weighted by the probability of a particular hidden state.
-This empirical distribution was then plotted over its corresponding fitted emission distribution (Figure \ref{fig:empirical_dist}).
+We plot histograms of $Y$ and $Z^*$ separated by dive type or subdive state in Figure \ref{fig:empirical_dist}.
+Observations are weighted by the decoded probability of a particular hidden state, and the histogram is plotted over the emission distribution learned by the CarHHMM-DFT (Figure \ref{fig:empirical_dist}).
 Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed.
-In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
-These outliers are potential subjects for future study.
-See the supplementary material for empirical distributions of every feature and every hidden state.
+In addition, $Z^{*(1)}$ has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
+These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \cite{Tennessen:2019a}.
+See the supplementary material for histograms for every observation and hidden state.
 
@@
 
-The CarHMM-DFT produced very similar to those of the CarHHMM-DFT on the fine scale.
-However, it failed to differentiate between short and long dives, and it therefore did not infer any information relating to the dive level Markov chain or the relationship between the dive level and the subdive level.
-For example, the CarHMM-DFT missed the insight that the whale was much more likely to be in subdive type 1 when engaged in longer dives than when engaged in shorter dives.
+The CarHMM-DFT produces very similar parameter estimates and state estimates compared to those of the CarHHMM-DFT on the fine scale.
+However, it fails to differentiate between short and long dives, and it therefore does not provide any information relating to the dive level Markov chain or the relationship between the dive level and the subdive level.
+For example, the CarHMM-DFT does not indicate that the whale is much more likely to be in subdive state 1 when engaged in longer dives than when engaged in shorter dives.
 
-The HHMM-DFT decoded the dive and subdive types similarly to the CarHHMM-DFT, but was less likely to categorize the beginning and end of dives as subdive type 3.
-In addition, the HHMM-DFT produced estimates $\hat \sigma_1^*$ that were approximately 50-100 \% larger than those for the CarHHMM-DFT, and the estimated uncertainty of $\hat \mu^*_1$ was less than half of that for the CarHHMM-DFT for all dimensions and subdive states.
-These findings are consistent with the trend to overestimate variance and produce overconfident parameter estimates when auto-correlation is ignored.
-Further, the empirical auto-correlation of $\mathbf{Z}^{*(1)}$ within each decoded subdive type was at least 0.5 for all dimensions and subdive types.
-Finally, the pseudoresiduals of the HHMM-DFT had many more outliers than the pseudoresiduals of the CarHHMM-DFT, providing evidence that the HHMM-DFT was a worse fit to this data than the CarHHMM-DFT.
+The HHMM-DFT decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3.
+In addition, the HHMM-DFT produces estimates of $\sigma_1^{*(1)}$, $\sigma_1^{*(2)}$, and $\sigma_1^{*(3)}$ (the standard deviation of acceleration) which are approximately 50-100 \% larger than those for the CarHHMM-DFT across all axes of acceleration.
+The estimated uncertainties of the estimates of $\mu_1^{*(1)}$, $\mu_1^{*(2)}$, and $\mu_1^{*(3)}$ are also less than half of those for the CarHHMM-DFT across all axes of acceleration.
+These findings are consistent with the tendency to overestimate variance and underestimate standard error when ignoring auto-correlation.
+The empirical auto-correlation of $Z^{*(1)}$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT show several outliers.
+Both of these findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.
 
-The CarHHMM did not have access to the Fourier sums of the acceleration data, so it failed to decode the subdive types into biologically interpretable results.
-In addition, the pseudoresiduals corresponding to the CarHHMM appears to be less heavy-tailed than a normal distribution.
-
-See section S-??? for a more complete set of results for each of the candidate models.
+The CarHHMM does not use the Fourier sums of the acceleration data, so it fails to decode the subdive states into biologically interpretable results.
+The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution.
+See section S-2 for a more complete set of results for each of the candidate models.

### 2020-07-30 Report/Sections/simulation.tex
Update Intro- Nancy
Update Report

@@
 The HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$.
-For both subdive states, $\hat \sigma_1^*$ greatly overestimated the true standard deviation.
+For both subdive types, $\hat \sigma_1^*$ greatly overestimated the true standard deviation.
 In addition, the empirical standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ were both much larger than the estimates of standard error using the observed Fisher information (see Section S-1.2, Table 3).
@@
 
-The CarHHMM was the worst-performing model in terms of decoding accuracy, with an average confidence of $89 \pm 1\%$ when determining subdive behavior and an average confidence of $87 \pm 11\%$ when determining dive types.
-It was much more likely to mis-classify subdive behavior 1 than subdive behavior 2 (see Section S-1.2, Table 1).
-It was also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy.
-$\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ in particular were poor estimates of the true parameters (see Section S-1.1.3).
+The CarHHMM was the worst-performing model in terms of decoding accuracy, with an average confidence below $0.9$ when determining both dive and subdive types.
+The CarHHMM did not have access to the Fourier sums, which were the most informative observations in terms of determining subdive type.
+It was much more likely to mis-classify subdive type 1 than subdive type 2 (see Section S-1.2, Table 1).
+The CarHHMM was also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy.
+$\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ were especially poor estimates (see Section S-1.1.3).
 
@@
 However, it lacked any hierarchical structure, so it was unable to estimate the coarse-scale dive types.
-The CarHMM-DFT still did fit a gamma distribution to the dive duration of \textit{all} dives.
+The CarHMM-DFT nonetheless fit a gamma distribution to the dive duration of \textit{all} dives.
 These parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds.
+See Figures \ref{fig:acc_coarse} \& \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive/subdive type probabilities as determined by each model.
 

### 2020-07-30 Report/Sections/data.tex
Commit before Evan full edit

@@
 
-To illustrate the process of constructing a model using the HMM model building blocks, we investigate the dive behavior of a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada, and construct several candidate generative models.
+To illustrate the process of constructing a model using HMM building blocks, we investigate the dive behavior of a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada, and construct several candidate generative models to describe the associated data.
 
@@
 
-Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
+Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
 Therefore, understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the true energetic requirements of NRKW.
@@
 Observations were collected at a rate of 50 Hz.
-Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods were ignored.
-In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range were ignored as well.
-A killer whale ``dive" is considered to be any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
-Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second.
-Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}.
+Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored.
+In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range are ignored as well.
+A killer whale ``dive" is any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
+Accelerometer and depth data are smoothed by taking a moving average with a window of 1/10th of a second.
+Data preprocessing is done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}.
 The preprocessed data contains a total of 267 dives.
@@
 Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
-As such, the coarse-scale observations $Y$ are a time series of dive durations in seconds ($s$) with corresponding hidden dive types $X$.
-In addition, the fine-scale processes $Y^*$ are the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive states $X^*$.
+Specifically, $Y$ is a time series of dive durations in seconds ($s$) with corresponding hidden dive types $X$.
+The fine-scale process $Y^*$ is the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive states $X^*$.
 
@@
 
-A lag plot was constructed to determine where auto-correlation should be incorporated into the model (see Section S-2.1).
-Prior to fitting the model, the coarse-scale level exhibited no significant amount of auto-correlation and no obvious intricate structure.
-As such, we selected a simple HMM to model this data.
-The dive durations $Y_t$ were assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
-
+A lag plot is used to determine where auto-correlation should be incorporated into the model (see Section S-2.1).
+The coarse-scale level exhibits no significant amount of auto-correlation and no obvious intricate structure prior to fitting the model.
+As such, we select a simple HMM to model the coarse-scale process.
+The dive durations $Y_t$ are assumed to follow a gamma distribution with unknown parameters $\{\mu,\sigma\}$:
+%
 $$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
 
-The fine-scale acceleration data exhibited significant sinusoidal behavior; thus the fine-scale observations $Y^*$ were transformed to $Z^*$ using the DFT summary statistics of a two-second sliding window.
-A two-second window was selected heuristically since it appeared to be the smallest window size that could capture the.
-``wiggles" in their entirety.
-Because the data was collected at the rate of 50 Hz, each two-second window contained a total of 100 raw acceleration readings.
-The acceleration data here was collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left) and is often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
-As a result, the transformation from the raw accelerometer data $\mathbf{Y}^*$ to the observation sequence $Z^*$ is calculated as follows:
+The fine-scale acceleration data exhibits significant sinusoidal behavior; thus $Y^*$ is transformed to $Z^*$ using the DFT summary statistics.
+A window size of two-second window is selected since it appears be the smallest window size that captures the ``wiggles" of the acceleration data in their entirety.
+Because the data was collected at the rate of 50 Hz, each two-second window contains a total of 100 raw acceleration readings.
+The acceleration data was also tri-axial, or collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left).
+Tri-axial acceleration readings are often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
+As a result, the transformation from the raw accelerometer data $Y^*$ to $Z^*$ is calculated slightly differently than Equation (\ref{eqn:z}):
 %
 \begin{equation*}
-    Z_{t,t^*}^{*(1)} := \text{Re}\left(DFT\{Y^*_{t,t^*}\}(0)\right) \qquad Z_{t,t^*}^{*(2)} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2
+    Z_{t,t^*}^{*(1)} := \text{Re}\left(DFT\{Y^*_{t,t^*}\}(0)\right), \qquad Z_{t,t^*}^{*(2)} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
 \end{equation*}
 %
-Each observation is therefore made up of $Z_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $Z_{t,t^*}^{*(2)}$, a scalar.
+Each observation is made up of $Z_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $Z_{t,t^*}^{*(2)}$, a scalar.
 We calculated $Z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum frequency of $\tilde \omega = 5$ Hz.
 
-Even after accounting for fine-scale structure in the raw acceleration data, there was still strong auto-correlation within $Z^{*(1)}_{t,t^*}$ prior to fitting the model (section S-2.1).
-Therefore, a CarHMM was used to model $Z^{*(1)}_{t,t^*}$.
-In particular, if $X^*_{t,t^*} = i^*$, then we assume $Z^{*(1)}_{t,t^*}$ to be Normally distributed with unknown parameters $\phi_1^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i^*)} \in \mathbb{R}^3$.
+Even after accounting for fine-scale structure in the raw acceleration data, there is still strong auto-correlation within $Z^{*(1)}_{t,t^*}$ prior to fitting the model (Section S-2.1).
+Therefore, a CarHMM is used to model $Z^{*(1)}_{t,t^*}$.
+In particular, if $X^*_{t,t^*} = i^*$, then we assume $Z^{*(1)}_{t,t^*}$ is Normally distributed with unknown parameters $\phi_1^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i^*)} \in \mathbb{R}^3$.
 %
@@
 While $Z^{*(2)}$ also exhibits some auto-correlation prior to fitting the model, the relationship is less strong than that of $Z^{*(1)}$, and the biological interpretation of auto-correlation within $Z^{*(2)}$ is less clear.
-In addition, much of the auto-correlation that \textit{is} evident from the lag plot may be explained by persistence in the subdive type $X^*_{t,t^*}$.
-Therefore, a simple HMM was used to model $Z^{*(2)}$.
-In addition, given that $X^*_{t,t^*} = i^*$ the distribution of $Z^{*(2)}_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_2^{*(i^*)}$ and variance $\sigma_2^{*(i^*)}$.
+In addition, much of the auto-correlation that \textit{is} evident from the lag plot may be explained because subsequent observations share a subdive state $X^*_{t,t^*}$.
+Therefore, a simple HMM is used to model $Z^{*(2)}$.
+Given that $X^*_{t,t^*} = i^*$ the distribution of $Z^{*(2)}_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_2^{*(i^*)}$ and variance $\sigma_2^{*(i^*)}$.
 %
@@
 
-We do not use information criteria to select the number of dive types and subdive types since it tends to overestimate the number of states in biological processes \citep{Pohle:2017}.
-In the particular case of well-separated hidden states, a lag plot will reveal $N$ circular patterns for each dive feature and $N^*$ circular patterns for each subdive feature.
-This is unfortunately not the case for the killer whale data, but we still consult the lag plots together with a heuristic approach to choose $N = 2$ dive types and $N^* = 3$ subdive behaviors.
-The absence of a more principled method to select the number of hidden states highlights the importance of model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
+We do not use information criteria to select the number of dive types and subdive types since they tend to overestimate the number of states in biological processes \citep{Pohle:2017}.
+If the hidden states are well-separated, a lag plot will reveal $N$ distinct patterns corresponding to each dive type and $N^*$ distinct patterns corresponding to each subdive state.
+This is unfortunately not the case for the killer whale data, but we still consult these lag plots together with a heuristic approach to choose $N = 2$ dive types and $N^* = 3$ subdive behaviors.
+The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
 
-The final model is a hierarchical hidden Markov model with a traditional HMM at the coarse scale.
-On the fine scale, $Y^*$ is transformed to $Z^*$, $Z^{*(1)}$ is modeled with a CarHMM, and $Z^{*(2)}$ is modeled with an HMM.
-In addition, we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity.
+The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale.
+On the fine scale, $Y^*$ is transformed to $Z^*$ using a DFT, $Z^{*(1)}$ is modelled with a CarHMM, and $Z^{*(2)}$ is modeled with an HMM.
+We force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity.
 This implies that dive types 1 and 2 share subdive states.
-We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMM models discussed here.
 The parameters to estimate are
@@
 Remember that $\theta_j^{*(\cdot,i^*)}$ are the parameters describing the distribution of $Z^{*(j)}_{t,t^*}$ when conditioning on $X^*_{t,t^*} = i^*$.
+
+We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMM models discussed so far.
 The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
@@
     \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ using a CarHMM.
-However, the Fourier sums $Z^{*(2)}$ are omitted as observations in this model.
-The Fourier sums $Z^{*(2)}$ describe the ``wiggliness" of the accelerometer data, while $Z^{*(1)}$ is simply the average acceleration over a particular window.
-    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $\mathbf{Z}^{*(1)}$ with a CarHMM and $Z^{*(2)}$ with an HMM.
+However, the Fourier sums $Z^{*(2)}$ are omitted from this model altogether.
+The CarHHMM therefore does not know about the ``wiggliness" of the accelerometer data.
+    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ with a CarHMM and $Z^{*(2)}$ with an HMM.
 This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive.
 \end{enumerate}
-All models have the same emission distributions as the CarHHMM-DFT, when applicable.
-Each of the three models leaves out one important aspect of the full CarHHMM-DFT.
-The HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the Fourier transform sums $(Z^{*(2)})$ as observations, and the CarHMM-DFT lacks a hierarchical structure by only modelling the fine-scale behavior of the whale.
+All models have the same emission distributions as the CarHHMM-DFT when applicable.
+Each of the three models leaves out one important aspect of the full CarHHMM-DFT: the HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the Fourier sums $(Z^{*(2)})$, and the CarHMM-DFT lacks a hierarchical structure.

### 2020-07-30 Report/Sections/discussion.tex
Commit before Evan full edit

@@
 
-Current FDA literature largely address sequences of curves either as multilevel models with random effects and no temporal dependence \cite{chen:2012,di:2009} or with functional time series which model the curves as an evolving process and overlook the possibility of a discrete number of curve ``types" \cite{Kokoszka:2018}.
-However, many real-world curve sequences exhibit temporal dependence and a discrete number of ``types".
-One obvious example is the case study examined here, where a sequence of marine mammal dives serve a number of discrete of purposes such as traveling, socializing, resting, and foraging \citep{Tennessen:2019a}.
-One other example can be found when measuring accelerometer data in school children, where bouts of activity can be associated with running, walking, etc.
+Current FDA literature addresses sequences of curves either as multilevel models with random effects and no temporal dependence \citep{chen:2012,di:2009} or as functional time series which model the curves as an evolving process and overlook the possibility of a discrete number of curve ``types" \citep{Kokoszka:2018}.
+However, many real-world curve sequences exhibit both temporal dependence and a discrete number of types.
+One obvious example is the case study examined here, where a sequence of marine mammal dives serve a number of discrete of purposes such as traveling, socializing, resting, or foraging \citep{Tennessen:2019a}.
+Another example is accelerometer data in school children, where bouts of activity can be associated with running, walking, etc.
 \citep{Morris:2007}.
 
-To deal with both temporal dependence between curves and a discrete number of curve types, we suggest modelling the sequence of curves with either an HMM or CarHMM, depending upon the level auto-correlation within the data.
-Several useful model selection tools such as the lag plot can test if there is significant auto-correlation within an sequence of curves.
-One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use another within-curve parametric model as emissions.
-The parameters of this fine-scale model in turn will depend upon the curve type, and together these two levels make up a hierarchical model which can simultaneously account for processes taking place at different time-scales.
+To deal with both temporal dependence between curves and a discrete number of curve types, we suggest modelling the sequence of curves with either an HMM or CarHMM, depending upon the level auto-correlation between curves.
+Several useful model selection tools such as a lag plot can be used to test if there is significant auto-correlation within a sequence of curves.
+One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use yet another within-curve parametric model as the HMM emissions.
+The parameters of this fine-scale model then in turn depend upon the curve type.
+Together, the two levels make up a hierarchical model which can account for processes taking place simultaneously and at different time-scales.
 
@@
 Several possible choices exist for the fine-scale model, including yet another HMM.
-\citet{Barajas:2017} introduce this model in the context of statistical ecology as an HHMM, and \citet{Adam:2019} use it to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California.
-However, traditional HMMs prove insufficient at very fine-scales, as \citet{Adam:2019} run into clear violations of conditional independence and apparent periodic behavior in the acceleration data.
+\citet{Barajas:2017} introduce this particular situation as an HHMM, and \citet{Adam:2019} use it to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California.
+However, traditional HMMs prove insufficient at very fine-scales, as \citet{Adam:2019} run into clear violations of conditional independence and observe apparent periodic behavior in the acceleration data.
 This indicates the need to account for complicated dependence structures within fine-scale processes, especially those with high sampling frequencies.
@@
 
-To deal with these problematic fine-scale processes, we suggest using a moving-window transformation of the fine-scale observations.
-For periodic behaviour in particular we use a DFT-based transformation.
-The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when $Z^{*(2)}$ is accounted for, and our case study, where the CarHHMM without DFT visually fails to pick up a killer whale's clear fluking behaviour.
+To deal with these problematic fine-scale processes, we suggest using a moving-window transformation over the fine-scale observations.
+For periodic behaviour in particular, we use a DFT-based transformation.
+The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when including $Z^{*(2)}$ in the model, and our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale.
+On top of the DFT transformation, we also suggest using a CarHMM instead of a traditional HMM when expert opinion and careful observation (possibly using a lag plot) calls for it.
+After putting all of these building blocks together, we arrive at our final model: the CarHMM-DFT.
+This model effectively decodes both dive types and subdive states of marine mammals while also taking into account the connection between these two levels.
 
-We also suggest using a CarHMM instead of a traditional HMM for the fine-scale process when expert opinion and careful observation (possibly using a log plot) calls for it.
-Although the CarHMM can incorporate auto-correlation into the structure of an HMM, it can break down when observations are taken at irregular time intervals.
-A common solution to this issue is to use a continuous-time method such as the state-switching OU processes described by \citet{Michelot:2019}.
-Most continuous time models require relatively slow MCMC algorithms to perform inference, and as a result are not easily incorporated into the HHMM structure.
-However, we show in Theorem 1 (see appendix) that the state-switching OU process is equivalent to an CarHMM under certain conditions.
-This implies that the parameter estimates from the CarHMM are interpretable in the context of a stochastic differential equation.
+This solves an important problem, as the energetic requirements of killer whales together with their energetic intake from predation is directly related to their survival \citep{Williams:2009,Noren:2011}.
+Previous studies have looked into the energy consumption of killer whales, but many rely on invasive and expensive methodology \citep{Williams:2009,Noren:2011}.
+Accelerometers are an appealing alternative, but require decoding the behavioral state of the animal in order to effectively estimate energy expenditure \citep{Dot:2016}.
+The CarHHMM-DFT solves this problem by decoding both the subdive behavioral state and the dive type of a killer whale.
+It also describes the distribution of acceleration in the states.
+All of this information meaningfully effects the energy expenditure of killer whales and is therefore valuable information for conservation efforts \citep{Williams:2009,Noren:2011}.
 
 While we combine these models together effectively for our case study, in general this practice should be done with care.
-It is important to balance the need to effectively capture the process in question with the need to avoid over-fitting and slow parameter estimation.
+There are drawbacks to building arbitrarily complex models with HMMs.
+For example, HMMs require observations to be taken at regular intervals.
+Continuous-time methods such as the state-switching OU process \citep{Michelot:2019} are based on stochastic differential equations and resolve this issue, but we do not describe them in detail here.
+This is because most continuous-time models require relatively slow MCMC algorithms to perform inference and are prohibitively difficult to fit.
+In addition, we show in Theorem 1 (see appendix) that the state-switching OU process of \citet{Michelot:2019} is equivalent to an CarHMM under certain conditions.
+This implies that a CarHMM is interpretable like a continuous-time model, but as easy to fit as a traditional HMM.
 
-One way to temper model complexity is to reduce the dimension of the parameter space by forcing fine-scale states to be shared across the coarse-scale states.
+Another consideration when using HMMs to build complex models is the importance of effectively capturing the process being modelled while avoiding over-fitting and slow parameter estimation.
+One way to temper model complexity is to share fine-scale state parameters across the coarse-scale states (i.e.
+subdive state 1 is the same across all dive types).
 Even still, model complexity inevitably grows rapidly as hierarchical structures are stacked on top of each other.
-
-An example of balancing model complexity with efficient fitting is presented in our simulation study.
-subdive behavioral states were shared across dive types to reduce model complexity, and by far the fastest model to train was the CarHMM-DFT, which had no hierarchical component.
-Even still, the CarHMM-DFT had near-perfect accuracy when decoding subdive behavioral states of the simulated whale.
+For example, by far the fastest model to train in the simulation study is the CarHMM-DFT, as it has no hierarchical component and is the simplest of the four candidate models.
+The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioral states.
 However, this model is not sufficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.
 
-The simulation study also shows that the observed Fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases.
+%The simulation study also shows that the observed Fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases.
 One notable exception is the HHMM-DFT, which ignores auto-correlation and therefore underestimates the standard error of the parameters associated with acceleration.
@@
 
-Because traditional information criteria tend to overestimate the number of states in biological processes \citep{Pohle:2017}, the number of dive types and subdive behaviors was selected in an ad-hoc manner.
+%Because traditional information criteria tend to overestimate the number of states in biological processes \citep{Pohle:2017}, the number of dive types and subdive behaviors was selected in an ad-hoc manner.
 There does appear to be some heterogeneity within dive types, and future work can be done to determine the optimal number of dive types and within-dive behaviors.
 
-In total, this work provides a flexible framework to model functional time-series data using HMM models which are easy to fit using maximum likelihood methods.
-The model of interest can be easily adjusted to match the complexity of the problem at hand.
-Researchers can use this method to build model and perform inference in any setting which involves state-switching fine-scale processes, including finance and medicine.
+Despite these drawbacks, this work provides a flexible framework to model functional time-series data using HMM models.
+When model complexity is kept under control, the resulting hierarchical model is both highly flexible and easy to fit using maximum likelihood methods.
+We demonstrate the usefulness of such a structure using an ecological example to determine the energetic requirements of a killer whale.
+It is our hope that researchers adapt the methodology described here to solve problems and perform inference in other settings which involve temporally dependent state-switching processes.

### 2020-07-30 Report/Sections/intro.tex
Commit before Evan full edit

@@
-Recent advances in technology for tracking movement provide researchers with data collected almost continuously in time.
-The collection and analysis of data from devices such as accelerometers have brought new understanding to areas ranging from  {\bf{more references}}  to physical activity levels in children \citep{Morris:2007}.
-In this article, we focus on the study of animal movement;  tags attached to animals record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours and providing biologists with new information about  animal behaviour.
+The study of high-frequency time series has been profoundly affected by recent advances in tracking technology which provide researchers with data collected almost continuously in time \cite{Hooten:2017}.
+The collection and analysis of data from devices such as accelerometers has brought new understanding to areas ranging from monitoring machine health \citep{Getman:2009} to physical activity levels in children \citep{Morris:2007}.
+In this article, we focus on the study of animal movement; tags attached to animals can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
+These data provide biologists with a wealth of new knowledge about animal behaviour, but parsing the vast amount of information is a challenge for statisticians.
 
@@
 In the data  we consider, a single dive of a whale is viewed as several curves:  acceleration and depth as  functions of time.
-FDA was originally developed for curves that can be considered independent replicates.
-However, the structure of the data is often more complex.
+FDA was originally developed for curves that can be considered independent replicates, but the structure of the data is often more complex.
 
@@
 Using HMMs at both the dive level and the fine scale allows us to interpret the data as arising from different types of behaviour.
-While our methodology is generally applicable, we  showcase our methods in a  study of the underlying behaviour indicated by the complex dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
+While our methodology is generally applicable, we  showcase our methods in a study of the dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
 For more information about the data, see Section \ref{sec:case_study}.
 Figure \ref{fig:data} shows the data we will analyze.
-The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and maximum depths. 
-The bottom three panels show the acceleration in each of three orthogonal directions as recorded by an accelerometer,   
-with axes oriented relative to the whale's forward movement.
-Here, we see spikes in acceleration which contrast with periods of fairly constant acceleration.
+The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and depths. 
+The bottom three panels show the acceleration in each of three orthogonal directions as recorded by an accelerometer.   
+%with axes oriented relative to the whale's forward movement.
+%Here, 
+We see spikes in acceleration which contrast with periods in which acceleration is fairly constant.
 Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
 The top panel shows a summary of the acceleration, and from this we can see fine-scale behaviour  that seems to indicate some  short-term periodicity. 
-We seek to understand both the coarse scale behaviour (e.g.~the dive type) and the fine-scale, within-dive behaviour. 
+We seek to understand both the coarse-scale behaviour (e.g. the dive type) and the fine-scale, within-dive behaviour. 
 Such an analysis helps biologists to understand the foraging behaviour and feeding success of killer whales.
 
-Models for particular types of between curve dependence structures appear in the FDA literature, 
+Models for particular types of between-curve dependence structures appear in the FDA literature, 
 notably multilevel models that incorporate dependence via random effects.
-For instance, in the aforementioned study of children's activity levels in  \cite{Morris:2007}, the authors use a multilevel approach since each child yields not one but two daily activity curves and so the model includes a random effect for ``child''.
+For instance, in the aforementioned study of children's activity levels \citep{Morris:2007}, the authors use a multilevel approach since each child yields not one but two daily activity curves and so the model includes a random effect for ``child''.
 In a classic early paper that introduced this multilevel structure,  \cite{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles.
 More recent work on multilevel models includes that of \cite{di:2009}, \cite{chen:2012} and Crainiceanu [REF].
-For our data, multilevel models are not appropriate, since they do not account for the temporal dependence in the sequence of dives nor do they account for different dive types.
+Multilevel models are not appropriate for our data since they do not account for the temporal dependence in the dive sequence or the different dive types.
 
-FDA researchers have modelled temporal dependence in a sequence of curves  via functional time series, extending the ideas of classic time series to model how one curve evolves into the next curve.
+FDA researchers have modelled temporal dependence in sequences of curves via functional time series, extending the ideas of classic time series to model the evolution of one curve into the next.
 See, for instance, \cite{Kokoszka:2018}.
 While functional time series methods can be applied to our dive data, they would not account for the different dive types.
-Our HMM models account for both temporal dive-to-dive dependence and different types of dives.
+HMM models account for both temporal dive-to-dive dependence and different types of dives.
 
@@
 
-From the plotted killer whale data in Figure {\ref{fig:labeled_dives}}, we see complexities within each dive that would be difficult to model as a realization of a simple stochastic process. 
+Figure \ref{fig:labeled_dives} reveals complexities within each dive that is difficult to model as a realization of a simple stochastic process. 
 In FDA, within-curve structure is modelled by a generic smooth mean function and covariance function.
 Alternatively, this structure is sometimes generated via random regression.
-However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in Figure 9.
-We model this behaviour by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence.
-Finally, we use a biologically meaningful transformation on the fine-scale behaviour to account for the ``wiggliness'' in some segments of a curve.
+However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in this dive sequence.
+Instead, we model this behaviour by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence.
+In addition to this, we use a biologically meaningful transformation on the fine-scale behaviour to account for the ``wiggliness'' in some segments of a curve.
 The distribution of this ``wiggliness'' depends on the hidden state.
@@
 one of the most prevalent techniques in the literature of late is the HMM \citep{Patterson:2017}. 
-However, HMMs have seen little use in FDA.  
-One exception is the work of Roland Langrock (see, for instance \citealt{Langrock:2018}).
-Animal movement modelers have also used  hierarchical hidden Markov models (HHMMs) to model  simultaneous behavioral processes that occur at different time scales \citep{Barajas:2017,Adam:2019}, and we borrow from that work here.
+However, HMMs have seen little use in FDA with  
+the exception of the work of Roland Langrock (see, for instance \citealt{Langrock:2018}).
+Animal movement modelers have also used  hierarchical hidden Markov models (HHMMs) to model  simultaneous behavioural processes that occur at different time scales \citep{Barajas:2017,Adam:2019}, and we borrow from that work here.
 Part of the purpose of our article is to introduce the FDA community to these useful approaches to modelling complex curves.
@@
 %
-%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another when conditioned on the underlying behavioral process.
+%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another when conditioned on the underlying behavioural process.
 %However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
@@
 
-We also borrow techniques from animal movement to model within-dive behaviour, beyond simply using an HMM.  
+We also borrow techniques from animal movement to model within-dive behaviour beyond a simple HMM.  
 Under the traditional HMM model, observations are independent given the underlying hidden state process.
-However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
+However, this assumption is often violated in the real-world processes, especially when observations are taken a high frequencies. 
 %For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
 Several solutions have been proposed, such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
-The CarHMM explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
+The CarHMM in particular explicitly models auto-correlation into an HMM while adding minimal complexity to the base structure of an HMM.
 
-An additional complication seen in the ecology literature is that, at the shortest time scales, observations often exhibit complicated structure which cannot easily be captured by traditional HMMs, CarHMMs, or HHMMs.
-One example in addition to the case study from this work is the swimming patterns of horn sharks off the coast of Southern California \citep{Adam:2019}, both of which exhibit fine-scale periodic behavior.
-%One solution is to model the fine-scale behavior using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation.
+An additional complication seen in the ecology literature is that observations at the shortest time scales often exhibit complicated structure which cannot easily be captured by traditional HMMs, CarHMMs, or HHMMs.
+One example is the fine-scale periodic swimming pattern of horn sharks off the coast of Southern California \citep{Adam:2019}.
+%One solution is to model the fine-scale behaviour using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation.
 Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals.
@@
 %Our dive data exhibit fine-scale periodic behaviour, so we transform the data via the Fourier transform. 
-\cite{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM in the context of describing daily behavioral cycles of marine mammals.
-In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior using accelerometer data \citep{Fehlmann:2017,Shorter:2017}.
+\cite{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM in the context of describing daily behavioural cycles of marine mammals.
+In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behaviour using accelerometer data \citep{Fehlmann:2017,Shorter:2017}.
 Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears a promising simple approach to account for additional correlation in data that is cyclical in nature.
@@
 %This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. 
-%We describe a general procedure that can be used to extract features from highly structured fine-scale behaviors that otherwise could not be modeled with existing HMM models. 
+%We describe a general procedure that can be used to extract features from highly structured fine-scale behaviours that otherwise could not be modeled with existing HMM models. 
 %In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. 
@@
 
-In the following section, we describe HMMs, generalize them to CarHMMs, introduce hierarchical structure with HHMMs, and discuss methods to transform intricate fine-scale data.
-We also show how these models can be combined in order to analyze complex data.
-In Section 3, we construct several candidate models based on the data from Figure \ref{fig:data} and Section 4 details a simulation study based on these candidate models.
-In Section 5 we fit each candidate model to the original data and analyze the results, and Section 6 contains our discussion of the results.
+In the following section, we describe HMMs in addition to two variants (CarHMMs and HHMMs) and discuss methods to deal with fine-scale intricate dependence structures.
+We also show how these methods can be combined in order to analyze increasingly complex data.
+In Section 3, we construct several candidate models to describe the data from Figure \ref{fig:data}.
+Section 4 details a simulation study based on these candidate models.
+In Section 5 we fit each candidate model to the original data and analyze the results, and in Section 6 we discuss the results.

### 2020-07-30 Report/Sections/models.tex
Commit before Evan full edit

@@
 
-Consider a sequence of $T$ possibly high-dimensional curves (or dives, in our context) with associated curve-level or coarse-scale observations $Y \equiv \left(Y_1, \ldots, Y_T\right)$.
-The coarse-scale observations contain summaries of each curve, such as duration, maximum value, variance, etc.
-In addition, curve $t$ has a total of $T^*_t$ within-curve or fine-scale associated observations: $Y^*_t \equiv \left(Y^*_{t,1}, \ldots Y^*_{t,t^*}, \ldots, Y^*_{t,T^*_t}\right)$.
+Consider a sequence of $T$ possibly high-dimensional curves (or dives, in our context) with associated curve-level (or coarse-scale) observations $Y \equiv \left(Y_1, \ldots, Y_T\right)$.
+The coarse-scale observations contain summaries of each curve such as duration or maximum value.
+Curve $t$ has a total of $T^*_t$ within-curve (or fine-scale) associated observations: $Y^*_t \equiv \left(Y^*_{t,1}, \ldots Y^*_{t,t^*}, \ldots, Y^*_{t,T^*_t}\right)$.
 The fine-scale observations $Y^*_t$ make up a vector of function values from curve $t$, and we index $Y^*_t$ using $t^*$.
@@
 We first describe the structure of a traditional HMM and a variation introduced by \cite{Lawler:2019} called the the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
-Neither the traditional HMM nor the CarHMM have any hierarchical structure, so both can only be used to model $Y$ or as a sub-model for a particular $Y^*_t$.
-In order to simultaneously model both $Y$ and $Y^*$, we describe the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}, which simultaneously models $Y$ and $Y^*$ as two distinct (but related) hidden Markov models.
-Next, we highlight the need to deal with intricate dependence structure by applying transformations to the observations using subject area expertise.
-This procedure is particularly necessary for $Y^*$.
-One particular transformation that we use applies the DFT to a moving window over $Y^*$ to summarize its periodic behavior.
+Either of these models can be used to model $Y$ or as a sub-model for $Y^*_t$.
+To simultaneously model both $Y$ and $Y^*$, we describe the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}, which simultaneously models $Y$ and $Y^*$ as two distinct (but related) hidden Markov models.
+Next, we describe how to reduce fine-scale correlation by transforming observations using subject area expertise.
+In particular, we use the DFT on a moving window over $Y^*$ to summarize its periodic behavior.
 Finally, we show how each of these models can be altered and combined to form a wide variety of new models.
@@
 
-The HHMM by definition models the fine-scale process with a traditional HMM.
-However, in principle $Y^*_t$ can be modeled using \textit{any} parametric model which admits an easy-to-compute likelihood.
-The fine-scale likelihood $\calL_{\text{HMM}}$ from the HHMM likelihood would simply be replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(y^*_t;\Theta^{*(i)})$:
+By definition, an HHMM models both the coarse-scale and the fine-scale processes with a traditional HMM.
+However, the coarse-scale HMM can be generalized by replacing it with a CarHMM, yielding the following likelihood:
+\[
+\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
+\]
+where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{HMM}}\left(y^*_t;\Theta^{*(i)}\right)$.
+On the fine scale, $Y^*_t$ can be modelled using \textit{any} parametric model which admits an easy-to-compute likelihood.
+The likelihood $\calL_{\text{HMM}}$ is then replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(y^*_t;\Theta^{*(i)})$:
 \[
@@
 where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
-The coarse-scale HMM can also be generalized by replacing \textit{it} with a CarHMM:
-\[
-\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
-\]
-where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
 
 Possible candidates for the fine-scale model include, but are not limited to, any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT).
-These alternative models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models based on HMMs.
-One possible model is, say, the \textbf{CarHHMM}, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM.
-Alternatively, one may construct an \textbf{HHHMM}, where three levels of observations are recorded ($Y$, $Y^*$, and $Y^{**}$), and traditional HMMs are used to model all three levels.
+These alternative fine-scale models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models based on HMMs.
+As an example, consider a \textbf{CarHHMM}, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM.
+Alternatively, one may construct an \textbf{HHHMM}, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$), and traditional HMMs are used to model all three levels.
 It is clear that infinitely many possible models can be built using this framework.

### 2020-07-30 Report/Sections/results.tex
Commit before Evan full edit

@@
 
-After fitting the four candidate models to simulated data, we repeat the same exercises and fit all four models to the data shown in Figure \ref{fig:data}.
-Here we report the results from the full CarHHMM-DFT defined in Section \ref{sec:data} in detail and asses the quality of the fit.
+After fitting the four candidate models to simulated data, we fit the same four models to the data shown in Figure \ref{fig:data}.
+Here we report the results from the full CarHHMM-DFT (defined in Section \ref{sec:data}) in detail and asses the quality of the fit.
 We also compare the results of the CarHHMM-DFT with the other candidate models.
@@
 The estimates for the emission distribution parameters suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}).
-Dive type 1 corresponds to shorter, shallower dives. Ecologists have attributed a variety of purposes to such short dives, including resting, traveling, and searching for prey \citep{Tennessen:2019a}.
+Dive type 1 corresponds to shorter, shallower dives. Ecologists have attributed a variety of purposes to these short dives such as resting, traveling, and searching for prey \citep{Tennessen:2019a}.
 Dive type 2 is longer and deeper.
@@
 
-The three subdive behaviours have distinct $Z^{*2}$ values, as the means of each are separated by an order of magnitude. (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}). 
-Subdive state 1 has the smallest mean $Z^{*(2)}$ and the smallest variance of $Z^{*(1)}$.
+The three subdive behaviours have distinct $Z^{*(2)}$ values, as the means of each subdive state are separated by an order of magnitude (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}).
+Subdive state 1 has the smallest mean value of $Z^{*(2)}$ and the smallest variance corresponding to $Z^{*(1)}$.
 It also has the highest auto-correlation in $Z^{*(1)}$.
-This corresponds to less overall activity and more consistent acceleration compared to the other subdive states.
-Subdive type 2 has a mean Fourier sum ($Z^{*(2)}$) one order of magnitude higher than subdive state 1, and its acceleration has about twice the variance compared to subdive state 1.
+This implies less overall activity and more consistent acceleration compared to the other subdive states.
+Subdive state 2 has a mean Fourier sum ($Z^{*(2)}$) one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
 The auto-correlation of acceleration is also slightly lower in this state.
 We hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
-The mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ in subdive state 3 is much higher than the other two states.
+Finally, the mean of $Z^{*(2)}$ and variance of $Z^{*(1)}$ in subdive state 3 is much higher than the other two states.
 The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower, implying more variation in acceleration between consecutive segments.
-This corresponds to vigorous swimming activity, especially as the killer whale begins a dive or surfaces from a dive (see Figure \ref{fig:labeled_dives}).
+This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).
 
@@
 0.809 & 0.191
-\end{pmatrix},$$
-$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix},$$
+\end{pmatrix} \text{ and }$$
+$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
 %
@@
 This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short dives before a killer whale performs a longer, deeper dive.
-The fine-scale stationary distributions $\hat{\delta^{*(i^*)}}$ indicate that this killer whale is more likely to be in the less active subdive state 1 when performing long deep dives than when performing short shallow dives.
+The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is more likely to be in the less active subdive state 1 when performing long deep dives than when performing short shallow dives.
 Using less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}.
@@
 Two visual tools were used to evaluate this model: pseudo-residuals and empirical histograms.
-A pseudo-residual of a particular observation is the marginal cumulative distribution function of an observation conditioned on all other observations under the learned model \citep{Zucchini:2016}.
-To easily visualize outliers, this pseudo-residual is often passed through the quantile function of the standard Normal distribution.
+A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}.
+To easily visualize outliers, pseudo-residuals are often passed through the quantile function of the standard Normal distribution.
 Mathematically, the pseudo-residual of an observation $y_t$ in a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
 If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution.
-We find that histograms of the pseudoresiduals of this model mostly support that the model is well-specified.
+Histograms of the pseudoresiduals of this model mostly support that the model is well-specified.
 One exception is $Z^{*(2)}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
@@
 
-We plot histograms of $Y$ and $Z^*$ separated by dive type or subdive state in Figure \ref{fig:empirical_dist}.
-Observations are weighted by the decoded probability of a particular hidden state, and the histogram is plotted over the emission distribution learned by the CarHHMM-DFT (Figure \ref{fig:empirical_dist}).
-Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed.
-In addition, $Z^{*(1)}$ has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
-These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \cite{Tennessen:2019a}.
-See the supplementary material for histograms for every observation and hidden state.
+We also plot histograms of $Y$ and $Z^*$ separated by dive type or subdive state in Figure \ref{fig:empirical_dist}.
+Observations are weighted by the decoded probability that they correspond to a particular hidden state, and the histogram is finally plotted over the emission distribution learned by the CarHHMM-DFT (Figure \ref{fig:empirical_dist}).
+Our results again mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed.
+Another exception is $Z^{*(1)}$, which has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
+These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.
+See Section S-2.?
+for histograms for every observation and hidden state.
 
@@
 The CarHMM-DFT produces very similar parameter estimates and state estimates compared to those of the CarHHMM-DFT on the fine scale.
-However, it fails to differentiate between short and long dives, and it therefore does not provide any information relating to the dive level Markov chain or the relationship between the dive level and the subdive level.
-For example, the CarHMM-DFT does not indicate that the whale is much more likely to be in subdive state 1 when engaged in longer dives than when engaged in shorter dives.
+However, its lack of a hierarchical structure means that it fails to differentiate between short and long dives, and it therefore does not provide any information relating to the dive-level Markov chain or the relationship between it and the subdive level.
+For example, the CarHMM-DFT does not indicate that the whale is much more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives.
 
@@
 
-The CarHHMM does not use the Fourier sums of the acceleration data, so it fails to decode the subdive states into biologically interpretable results.
+The CarHHMM does not use the Fourier sums of the acceleration data, so it fails produce biologically interpretable results when decoding the subdive states.
 The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution.

### 2020-07-30 Report/Sections/simulation.tex
Commit before Evan full edit

@@
 
-To compare the performance of the candidate models when the ground-truth is known, we performed a simulation study based on data generated from the full CarHHMM-DFT model.
+To compare the performance of the candidate models when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT model.
 The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}.
 The ground-truth parameters differ slightly from those in the case study for primarily pedagogical reasons.
-For example, we set $\Gamma$ such that the simulated data would contain approximately equal numbers of short and long dives.
+For example, we set $\Gamma$ such that the simulated data contains approximately equal numbers of short and long dives.
 This allows meaningful conclusions to be made for both dive types.
-In addition, the number subdive states was set to $N^*=2$ to simplify the analysis of the subdive states, and $\mathbf{Z}^{*(2)}$ was changed from a vector to a scalar for additional simplicity.
+In addition, the number subdive states is set to $N^*=2$ to simplify the analysis of the subdive states, and $Z^{*(2)}$ is changed from a vector to a scalar for additional simplicity.
 
-We fit all four candidate models to the data generated from the CarHHMM-DFT in order to analyze the impact of model mis-specification and compare them using several different metrics.
-These metrics include the decoding accuracy of the dive and subdive types, bias in parameter estimates, standard error of parameter estimates, and fitting times.
-In addition, we compare the empirical standard error of each model's parameter estimates with the estimated standard error using the observed Fisher information under that model, which allows us to judge how accurate a given model is when estimating its own uncertainty.
+We fit all four candidate models to the data generated from the CarHHMM-DFT to analyze the impact of model mis-specification and compare them using several different metrics.
+These metrics include the decoding accuracy of the dive types and subdive states, bias in parameter estimates, standard error of parameter estimates, and fitting times.
+We also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the observed Fisher information under that model.
+This allows us to judge how accurate each model is when estimating its own uncertainty.
 
@@
 
-We generated 500 independent data sets from the CarHHMM-DFT.
-Each data set consisted of a sequence of 100 curves, which we will call a sequence of killer whale dives.
-Each dive could be one of $N=2$ dive types and the coarse-scale probability transition matrix was set as
+We generate 500 independent data sets from the CarHHMM-DFT.
+Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives.
+Each dive is one of $N=2$ dive types and the coarse-scale probability transition matrix is set as
 %
@@
 %
-Dive duration was gamma distributed, and the coarse-scale emission parameters $\Theta$ were set as
+Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are set as
 $$
@@
 %
-After the dive durations were calculated for all 100 dives for each data set using $\Gamma$ and $\Theta$, dive $t$ was broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments, which made up the fine-scale hidden Markov model.
-The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive were discarded.
-Each two-second segment was assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
+After the dive durations are calculated for all 100 dives for each data set using $\Gamma$ and $\Theta$, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model.
+The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive are discarded.
+Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
 %
@@
 %
-and the observations $Z^*$ were generated using the following parameters:
+The observations $Z^*$ are generated using the following parameters:
 %
@@
 Recall that $Z^{*(1)}$ follows a CarHMM and is Normally distributed while $Z^{*(2)}$ follows an HMM and is Gamma distributed.
-It is not possible to uniquely recreate the raw fine-scale data $Y^*$ from $Z^*$ alone, but we describe a method to reconstruct one consistent sequence of observations $Y^*$ in the appendix.
+It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe a method to reconstruct one possible realization of $Y^*$ in the appendix.
 Figure \ref{fig:sim_data} shows this reconstructed sequence along with $Z^*$ for five dives of one simulated data set.
 
-The two subdive types differ primarily due to $\mu_2^*$ and $\sigma_2^*$, both of which are much higher for subdive type 2.
-This corresponds to much more vigorous and variable periodic behavior in the acceleration data when the simulated whale is in subdive type 2.
+The two subdive types differ primarily due to $\mu_2^*$ and $\sigma_2^*$, both of which are much higher in subdive state 2.
+This corresponds to much more vigorous and variable periodic behavior in the acceleration data.
 
@@
 
-All four candidate models (CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT) were fit to all 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
+All four candidate models (CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT) are fit to all 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
 
-As expected, the full CarHHMM-DFT was the best-performing model.
-Given the true subdive type of a particular segment $x^*_{t,t^*} = i$, the CarHHMM-DFT would predict the correct subdive type with an average confidence of $1 - 10^{-12}$ across all simulations.
+As expected, the full CarHHMM-DFT is the best-performing model.
+Given the true subdive type of a particular segment $X^*_{t,t^*}$, the CarHHMM-DFT predicts the correct subdive type with an average confidence of $1 - 10^{-12}$ across all simulations.
 For the dive types, this average confidence was above 0.9 (see Table \ref{table:accuracy}).
-Estimates of hidden states were made using the forward-backward algorithm.
-All parameter estimates had statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased.
+Estimates of hidden states are made using the forward-backward algorithm.
+All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased.
 Even still, these biases are the least severe for the CarHHMM-DFT among the four models.
-The empirical standard error of all parameters estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) were well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tended to be slight underestimates of the empirical standard error.
+The empirical standard error of all parameter estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slight underestimates of the empirical standard error.
 See Section S-1.2, tables 2-5 for more information.
 
-The HHMM-DFT performed almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias.
+The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias.
 The HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$.
-For both subdive types, $\hat \sigma_1^*$ greatly overestimated the true standard deviation.
-In addition, the empirical standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ were both much larger than the estimates of standard error using the observed Fisher information (see Section S-1.2, Table 3).
+For both subdive types, $\hat \sigma_1^*$ greatly overestimates the true standard deviation.
+The empirical standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ are both much larger than the estimates of standard error using the observed Fisher information (see Section S-1.2, Table 3).
 This implies that parameter estimates can be too confident when auto-correlation is ignored.
 
-The CarHHMM was the worst-performing model in terms of decoding accuracy, with an average confidence below $0.9$ when determining both dive and subdive types.
-The CarHHMM did not have access to the Fourier sums, which were the most informative observations in terms of determining subdive type.
-It was much more likely to mis-classify subdive type 1 than subdive type 2 (see Section S-1.2, Table 1).
-The CarHHMM was also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy.
-$\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ were especially poor estimates (see Section S-1.1.3).
+The CarHHMM is the worst-performing model in terms of decoding accuracy, with an average confidence below $0.9$ when determining both dive and subdive types.
+The CarHHMM does not have access to the Fourier sums, which are the most informative observations in terms of determining subdive type.
+It is much more likely to mis-classify subdive type 1 than subdive type 2 (see Section S-1.2, Table 1).
+The CarHHMM is also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy.
+$\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ are especially poor estimates under the CarHHMM (see Section S-1.1.3).
 
-Finally, the CarHMM-DFT was comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations.
-In addition, the time required to fit the CarHMM-DFT was less than half of that of the other models (see Table \ref{table:accuracy}).
-However, it lacked any hierarchical structure, so it was unable to estimate the coarse-scale dive types.
-The CarHMM-DFT nonetheless fit a gamma distribution to the dive duration of \textit{all} dives.
+Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations.
+In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}).
+However, it lacks any hierarchical structure, so it is unable to estimate dive types.
+The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of \textit{all} dives.
 These parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds.

### 2020-07-31 Report/Sections/data.tex
Stylistic changes to report

@@
 
-To illustrate the process of constructing a model using HMM building blocks, we investigate the dive behavior of a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada, and construct several candidate generative models to describe the associated data.
+To illustrate the process of constructing a model using these HMM building blocks, we investigate the dive behavior of a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada, and construct several candidate generative models to describe the associated data.
 
-Understanding animal behaviour can be important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}.
+Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}.
 NRKW prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus
@@
 It is therefore of interest to determine the energetic expenditure of NRKW to forage for Chinook, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
-
 Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
-Therefore, understanding both the behavioral state of the killer whale and the distribution of accelerometer data within each behavioral state is important to understand the true energetic requirements of NRKW.
+Therefore, understanding both the behavioral state of the killer whale and the distribution of acceleration within each behavioral state is important to understand the true energetic requirements of a NRKW.
 
@@
 
-The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions.
+The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions.
 Observations were collected at a rate of 50 Hz.
@@
 A killer whale ``dive" is any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
-Accelerometer and depth data are smoothed by taking a moving average with a window of 1/10th of a second.
+Accelerometer and depth data are smoothed by taking a moving average within a window of 1/10th of a second.
 Data preprocessing is done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}.
-The preprocessed data contains a total of 267 dives.
-Figure \ref{fig:data} displays the dive profile and accelerometer data.
-
+The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}.
 Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
@@
 The fine-scale acceleration data exhibits significant sinusoidal behavior; thus $Y^*$ is transformed to $Z^*$ using the DFT summary statistics.
-A window size of two-second window is selected since it appears be the smallest window size that captures the ``wiggles" of the acceleration data in their entirety.
-Because the data was collected at the rate of 50 Hz, each two-second window contains a total of 100 raw acceleration readings.
-The acceleration data was also tri-axial, or collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left).
-Tri-axial acceleration readings are often collected by these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
-As a result, the transformation from the raw accelerometer data $Y^*$ to $Z^*$ is calculated slightly differently than Equation (\ref{eqn:z}):
+A window size of two-seconds is selected since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data.
+The acceleration data was also collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left).
+Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
+As a result, the transformation from the raw accelerometer data $Y^*$ to $Z^*$ is calculated slightly differently than in Equation (\ref{eqn:z}):
 %
@@
 Each observation is made up of $Z_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $Z_{t,t^*}^{*(2)}$, a scalar.
-We calculated $Z_t^{*(2)}$ by summing the first 10 Fourier modes, which corresponds to a maximum frequency of $\tilde \omega = 5$ Hz.
+We calculate $Z_t^{*(2)}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to a maximum frequency of 5 Hz.
 
-Even after accounting for fine-scale structure in the raw acceleration data, there is still strong auto-correlation within $Z^{*(1)}_{t,t^*}$ prior to fitting the model (Section S-2.1).
+Even after accounting for fine-scale structure in the raw acceleration data, there is still strong auto-correlation within each axis of $Z^{*(1)}_{t,t^*}$ prior to fitting the model (Section S-2.1).
 Therefore, a CarHMM is used to model $Z^{*(1)}_{t,t^*}$.
 In particular, if $X^*_{t,t^*} = i^*$, then we assume $Z^{*(1)}_{t,t^*}$ is Normally distributed with unknown parameters $\phi_1^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i^*)} \in \mathbb{R}^3$.
+In words, each axis of acceleration has its own mean and variance, but all axes share an auto-correlation parameter.
 %
@@
 While $Z^{*(2)}$ also exhibits some auto-correlation prior to fitting the model, the relationship is less strong than that of $Z^{*(1)}$, and the biological interpretation of auto-correlation within $Z^{*(2)}$ is less clear.
-In addition, much of the auto-correlation that \textit{is} evident from the lag plot may be explained because subsequent observations share a subdive state $X^*_{t,t^*}$.
+In addition, much of the auto-correlation evident from the lag plot may be explained by subsequent observations sharing a subdive state $X^*_{t,t^*}$.
 Therefore, a simple HMM is used to model $Z^{*(2)}$.
-Given that $X^*_{t,t^*} = i^*$ the distribution of $Z^{*(2)}_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_2^{*(i^*)}$ and variance $\sigma_2^{*(i^*)}$.
+Given that $X^*_{t,t^*} = i^*$, the distribution of $Z^{*(2)}_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_2^{*(i^*)}$ and variance $\sigma_2^{*(i^*)}$.
 %
@@
 We do not use information criteria to select the number of dive types and subdive types since they tend to overestimate the number of states in biological processes \citep{Pohle:2017}.
-If the hidden states are well-separated, a lag plot will reveal $N$ distinct patterns corresponding to each dive type and $N^*$ distinct patterns corresponding to each subdive state.
-This is unfortunately not the case for the killer whale data, but we still consult these lag plots together with a heuristic approach to choose $N = 2$ dive types and $N^* = 3$ subdive behaviors.
+If the hidden states are well-separated, a lag plot reveals $N$ distinct patterns corresponding to each dive type and $N^*$ distinct patterns corresponding to each subdive state \cite{Lawler:2019}.
+This is unfortunately not the case for our killer whale data, but we still consult our lag plots together with a heuristic approach to choose $N = 2$ dive types and $N^* = 3$ subdive behaviors.
 The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
@@
 On the fine scale, $Y^*$ is transformed to $Z^*$ using a DFT, $Z^{*(1)}$ is modelled with a CarHMM, and $Z^{*(2)}$ is modeled with an HMM.
-We force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity.
+We also force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity.
 This implies that dive types 1 and 2 share subdive states.
-The parameters to estimate are
+In total, the parameters to estimate are
 %
@@
     %
-    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{(coarse-scale emission parameters), and} \\
+    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{($Y$ emission parameters), and} \\
     %
@@
 %
-Remember that $\theta_j^{*(\cdot,i^*)}$ are the parameters describing the distribution of $Z^{*(j)}_{t,t^*}$ when conditioning on $X^*_{t,t^*} = i^*$.
+Recall that $\theta_j^{*(\cdot,i^*)}$ are the parameters describing the distribution of $Z^{*(j)}_{t,t^*}$ when conditioning on $X^*_{t,t^*} = i^*$.
 
@@
 Figure \ref{fig:CarHHMM-DFT} also shows the corresponding graphical model for the CarHHMM-DFT.
-Additionally, we consider three similar variations for comparison:
+In addition to the CarHHMM-DFT we consider three similar variations for comparison:
 \begin{enumerate}
@@
 \end{enumerate}
-All models have the same emission distributions as the CarHHMM-DFT when applicable.
-Each of the three models leaves out one important aspect of the full CarHHMM-DFT: the HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the Fourier sums $(Z^{*(2)})$, and the CarHMM-DFT lacks a hierarchical structure.
+All models have emission distributions identical to the CarHHMM-DFT when applicable.
+Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the ``wiggliness" $(Z^{*(2)})$, and the CarHMM-DFT lacks a hierarchical structure.

### 2020-07-31 Report/Sections/discussion.tex
Stylistic changes to report

@@
 
-Current FDA literature addresses sequences of curves either as multilevel models with random effects and no temporal dependence \citep{chen:2012,di:2009} or as functional time series which model the curves as an evolving process and overlook the possibility of a discrete number of curve ``types" \citep{Kokoszka:2018}.
+Current FDA literature addresses sequences of curves either with multilevel models with random effects and no temporal dependence \citep{chen:2012,di:2009}, or with functional time series which model the curves as an evolving process and overlook the possibility of a discrete number of curve ``types" \citep{Kokoszka:2018}.
 However, many real-world curve sequences exhibit both temporal dependence and a discrete number of types.
@@
 
-To deal with both temporal dependence between curves and a discrete number of curve types, we suggest modelling the sequence of curves with either an HMM or CarHMM, depending upon the level auto-correlation between curves.
-Several useful model selection tools such as a lag plot can be used to test if there is significant auto-correlation within a sequence of curves.
+In this work, we deal with both temporal dependence between curves and a discrete number of curve types by modelling the sequence of curves with either an HMM or CarHMM, depending upon the level auto-correlation between curves.
+Several useful model selection tools such as a lag plot can be used to test if there is significant auto-correlation between curves.
 One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use yet another within-curve parametric model as the HMM emissions.
 The parameters of this fine-scale model then in turn depend upon the curve type.
-Together, the two levels make up a hierarchical model which can account for processes taking place simultaneously and at different time-scales.
+Together, the two levels make up a hierarchical model which can account for processes taking place simultaneously at different time-scales.
 
@@
 \citet{Barajas:2017} introduce this particular situation as an HHMM, and \citet{Adam:2019} use it to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California.
-However, traditional HMMs prove insufficient at very fine-scales, as \citet{Adam:2019} run into clear violations of conditional independence and observe apparent periodic behavior in the acceleration data.
+Traditional HMMs prove insufficient at these very fine-scales, as \citet{Adam:2019} run into clear violations of conditional independence and observe periodic behavior in the acceleration data.
 This indicates the need to account for complicated dependence structures within fine-scale processes, especially those with high sampling frequencies.
@@
 
-To deal with these problematic fine-scale processes, we suggest using a moving-window transformation over the fine-scale observations.
+To deal with these problematic fine-scale processes, we detail a moving-window transformation over the fine-scale observations.
 For periodic behaviour in particular, we use a DFT-based transformation.
 The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when including $Z^{*(2)}$ in the model, and our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale.
-On top of the DFT transformation, we also suggest using a CarHMM instead of a traditional HMM when expert opinion and careful observation (possibly using a lag plot) calls for it.
+On top of the DFT transformation, we also suggest using a CarHMM instead of a traditional HMM when expert opinion and careful observation (aided by a lag plot) calls for it.
 After putting all of these building blocks together, we arrive at our final model: the CarHMM-DFT.
-This model effectively decodes both dive types and subdive states of marine mammals while also taking into account the connection between these two levels.
+This model effectively decodes both dive types and subdive states of a killer whale while also taking into account the interaction between these two levels.
 
@@
 Previous studies have looked into the energy consumption of killer whales, but many rely on invasive and expensive methodology \citep{Williams:2009,Noren:2011}.
-Accelerometers are an appealing alternative, but require decoding the behavioral state of the animal in order to effectively estimate energy expenditure \citep{Dot:2016}.
-The CarHHMM-DFT solves this problem by decoding both the subdive behavioral state and the dive type of a killer whale.
-It also describes the distribution of acceleration in the states.
+Accelerometers are an appealing alternative, but require decoding both the behavioral state \citep{Dot:2016} and the dive type \citep{Hastie:2006} of the animal in order to effectively estimate energy expenditure.
+The CarHHMM-DFT solves this problem by decoding both behaviors simultaneously at different scales.
+It also describes the distribution of acceleration within each hidden state.
 All of this information meaningfully effects the energy expenditure of killer whales and is therefore valuable information for conservation efforts \citep{Williams:2009,Noren:2011}.
 
-While we combine these models together effectively for our case study, in general this practice should be done with care.
-There are drawbacks to building arbitrarily complex models with HMMs.
+While we combine the HMM models of Section \ref{sec:models} together effectively for our case study, in general this practice should be done with care as there are drawbacks to using exclusively HMMs to build these hierarchical models.
 For example, HMMs require observations to be taken at regular intervals.
@@
 
-Another consideration when using HMMs to build complex models is the importance of effectively capturing the process being modelled while avoiding over-fitting and slow parameter estimation.
-One way to temper model complexity is to share fine-scale state parameters across the coarse-scale states (i.e.
-subdive state 1 is the same across all dive types).
+Another consideration when using HMMs to build complex models is the importance of effectively fit the data in question while avoiding \textit{over}-fitting or computationally expensive parameter estimation.
+One way to temper model complexity is to share fine-scale state parameters across the coarse-scale states.
 Even still, model complexity inevitably grows rapidly as hierarchical structures are stacked on top of each other.
@@
 The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioral states.
-However, this model is not sufficient if ecologists wish to understand to joint relationship between dive type and intra-dive behaviour.
+However, this model is not sufficient if ecologists wish to understand to joint relationship between the dive type and intra-dive behaviour.
 
@@
 Despite these drawbacks, this work provides a flexible framework to model functional time-series data using HMM models.
-When model complexity is kept under control, the resulting hierarchical model is both highly flexible and easy to fit using maximum likelihood methods.
+When model complexity is kept under control, a hierarchical model constructed using this processes can be both highly flexible and easy to fit using maximum likelihood methods.
 We demonstrate the usefulness of such a structure using an ecological example to determine the energetic requirements of a killer whale.
-It is our hope that researchers adapt the methodology described here to solve problems and perform inference in other settings which involve temporally dependent state-switching processes.
+It is our hope that researchers adapt this methodology to solve problems and perform inference in other settings which involve temporally dependent state-switching processes.

### 2020-07-31 Report/Sections/intro.tex
Stylistic changes to report

@@
-The study of high-frequency time series has been profoundly affected by recent advances in tracking technology which provide researchers with data collected almost continuously in time \cite{Hooten:2017}.
+The study of high-frequency time series has been profoundly affected by recent advances in tracking technology which provide researchers with data collected almost continuously in time \citep{Hooten:2017}.
 The collection and analysis of data from devices such as accelerometers has brought new understanding to areas ranging from monitoring machine health \citep{Getman:2009} to physical activity levels in children \citep{Morris:2007}.
 In this article, we focus on the study of animal movement; tags attached to animals can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
-These data provide biologists with a wealth of new knowledge about animal behaviour, but parsing the vast amount of information is a challenge for statisticians.
+These data provide biologists with a wealth of knowledge about animal behaviour, but parsing the vast amount of information is a challenge for statisticians.
 
@@
 Movement data are sometimes modelled as a set of curves that can be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}).
-In \cite{Morris:2007}, a child's activity level throughout the day is viewed as a curve, specifically, as activity as a function of time.
-In the data  we consider, a single dive of a whale is viewed as several curves:  acceleration and depth as  functions of time.
-FDA was originally developed for curves that can be considered independent replicates, but the structure of the data is often more complex.
+In \cite{Morris:2007}, a child's activity level throughout the day is viewed as a curve where metabolic activity is a function of time.
+In our case study, we view a single dive of a killer whale as several curves where acceleration and depth are functions of time.
 
-Our goal is to develop movement analysis methods that provide subject-area insight  into the complex structure of curve data.
-We model complexities arising from curve to curve variation, along with the complexities of within-curve fine-scale characteristics and temporal dependence,
-combining existing methods in novel ways.
-We enforce temporal structure and dependence from dive to dive via a hidden Markov model (HMM).
-We model within dive variation by combining an HMM and a time series.
-Using HMMs at both the dive level and the fine scale allows us to interpret the data as arising from different types of behaviour.
-While our methodology is generally applicable, we  showcase our methods in a study of the dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.
-For more information about the data, see Section \ref{sec:case_study}.
-Figure \ref{fig:data} shows the data we will analyze.
-The top panel shows the depth of the whale below the surface, and we clearly see dives of various time durations and depths. 
-The bottom three panels show the acceleration in each of three orthogonal directions as recorded by an accelerometer.   
-%with axes oriented relative to the whale's forward movement.
-%Here, 
-We see spikes in acceleration which contrast with periods in which acceleration is fairly constant.
-Figure {\ref{fig:labeled_dives}} gives a more detailed view, showing a segment of the data containing six dives of at least two different types.
-The top panel shows a summary of the acceleration, and from this we can see fine-scale behaviour  that seems to indicate some  short-term periodicity. 
-We seek to understand both the coarse-scale behaviour (e.g. the dive type) and the fine-scale, within-dive behaviour. 
-Such an analysis helps biologists to understand the foraging behaviour and feeding success of killer whales.
+FDA was originally developed for curves that can be considered independent replicates, but the structure of the data is often more complex than this.
+In particular, we focus on functional data which display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale).
+While our methodology is applicable in a wide variety of settings, we focus on the dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada (see Section \ref{sec:data} for more information about data collection and preprocessing).
+The dive profile displayed in the top panel of Figure \ref{fig:data} clearly shows temporal dependence \textit{between} dives, as dives of various time durations and depths appear to be clustered together in time.
+Previous studies suggest that each of these dives have one of a discrete number of purposes (or types), including resting, traveling, or foraging \citep{Tennessen:2019b}.
+Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence \textit{within} dives, as the accelerometer data within each dive indicate bouts of short-term periodicity.
+The combination of apparent behavioral changes combined with periodic behavior makes this accelerometer data difficult to model using simple stochastic processes.
 
-Models for particular types of between-curve dependence structures appear in the FDA literature, 
-notably multilevel models that incorporate dependence via random effects.
-For instance, in the aforementioned study of children's activity levels \citep{Morris:2007}, the authors use a multilevel approach since each child yields not one but two daily activity curves and so the model includes a random effect for ``child''.
+Models for particular types of between-curve dependence structures appear in the FDA literature, but these methods fall short when modelling the coarse-scale temporal structure seen here.
+One example of such methods are multilevel models which incorporate dependence via random effects.
+In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors use a multilevel approach since each child yields not one but two daily activity curves; the model includes a random effect for ``child''.
 In a classic early paper that introduced this multilevel structure,  \cite{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles.
 More recent work on multilevel models includes that of \cite{di:2009}, \cite{chen:2012} and Crainiceanu [REF].
-Multilevel models are not appropriate for our data since they do not account for the temporal dependence in the dive sequence or the different dive types.
+However, multilevel models are not appropriate for this killer whale data since they account for neither the temporal dependence in the dive sequence nor the different dive types.
+In addition to multilevel models, FDA researchers have used functional time series to model dependence in sequences of curves.
+Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}.
+While functional time series methods can in theory be applied to our dive data, they do not account for the discrete number of different dive types.
 
-FDA researchers have modelled temporal dependence in sequences of curves via functional time series, extending the ideas of classic time series to model the evolution of one curve into the next.
-See, for instance, \cite{Kokoszka:2018}.
-While functional time series methods can be applied to our dive data, they would not account for the different dive types.
-HMM models account for both temporal dive-to-dive dependence and different types of dives.
+Traditional FDA techniques similarly fall short when modelling the within-curve acceleration data seen here.
+In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function.
+Alternatively, this structure is sometimes generated via random regression.
+These classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in the accelerometer data of this dive sequence.
 
-%A noted exception involves multilevel modelling [references] which uses random effects and is appropriate when the set of curves can be partitioned into independent subsets, typically with each subset arising from one individual.
-
-Figure \ref{fig:labeled_dives} reveals complexities within each dive that is difficult to model as a realization of a simple stochastic process. 
-In FDA, within-curve structure is modelled by a generic smooth mean function and covariance function.
-Alternatively, this structure is sometimes generated via random regression.
-However, these classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in this dive sequence.
-Instead, we model this behaviour by combining a hidden Markov model of within-curve behaviour with an autoregressive-type dependence.
-In addition to this, we use a biologically meaningful transformation on the fine-scale behaviour to account for the ``wiggliness'' in some segments of a curve.
-The distribution of this ``wiggliness'' depends on the hidden state.
-% Specifically, for the dive data, we use summaries of Fourier transforms to account for the ``wiggliness'' in some segments of a curve.
-The distribution of this ``wiggliness'' depends on the hidden state.
-
-Our work relies on work in the vast field of methodology for the analysis of animal movement
-\citep{Hooten:2017,Mcclintock:2020}, where
-one of the most prevalent techniques in the literature of late is the HMM \citep{Patterson:2017}. 
-However, HMMs have seen little use in FDA with  
-the exception of the work of Roland Langrock (see, for instance \citealt{Langrock:2018}).
-Animal movement modelers have also used  hierarchical hidden Markov models (HHMMs) to model  simultaneous behavioural processes that occur at different time scales \citep{Barajas:2017,Adam:2019}, and we borrow from that work here.
-Part of the purpose of our article is to introduce the FDA community to these useful approaches to modelling complex curves.
+In light of these challenges, we turn to vast field of methodology for the analysis of animal movement \citep{Hooten:2017,Mcclintock:2020}, where one of the most prevalent techniques in the literature of late is the hidden Markov model, or HMM \citep{Patterson:2017}.
+HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of the animal from movement observations.
+While ubiquitous in the ecology literature, HMMs have seen little use in FDA, not including the work of Roland Langrock (see, for instance \citealt{Langrock:2018}).
 
@@
 However, this assumption is often violated in the real-world processes, especially when observations are taken a high frequencies.
-%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
-Several solutions have been proposed, such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}. 
+Several solutions have been proposed in the ecology literature, such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
 The CarHMM in particular explicitly models auto-correlation into an HMM while adding minimal complexity to the base structure of an HMM.
+Animal movement modelers have used hierarchical hidden Markov models (HHMMs) to model simultaneous behavioural processes that occur at different time scales \citep{Barajas:2017,Adam:2019}.
+We borrow from ideas from both the CarHMM and the HHMM here.
+Part of the purpose of our article is to introduce the FDA community to these useful approaches to modelling complex curves.
 
-An additional complication seen in the ecology literature is that observations at the shortest time scales often exhibit complicated structure which cannot easily be captured by traditional HMMs, CarHMMs, or HHMMs.
+However, even the ecology literature falls short at the shortest time scales, which often exhibit complicated structure that cannot easily be captured by traditional HMMs, CarHMMs, or HHMMs.
 One example is the fine-scale periodic swimming pattern of horn sharks off the coast of Southern California \citep{Adam:2019}.
-%One solution is to model the fine-scale behaviour using a continuous-time model, which usually involves modelling the dynamics of an animal as the solution to a stochastic differential equation.
-Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals.
-However, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.
-%Our dive data exhibit fine-scale periodic behaviour, so we transform the data via the Fourier transform. 
-\cite{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM in the context of describing daily behavioural cycles of marine mammals.
-In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behaviour using accelerometer data \citep{Fehlmann:2017,Shorter:2017}.
-Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears a promising simple approach to account for additional correlation in data that is cyclical in nature.
+For periodic behavior in particular, \cite{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM in the context of describing daily behavioural cycles of marine mammals.
+In addition, Fourier analysis has previously been used on accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
+Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears to be a promising and simple approach to account for periodic structures in fine-scale data.
+In particular, we slide a moving window over the raw acceleration data and perform a biologically meaningful transformation of the data in this window.
+This transformation accounts for the ``wiggliness'' in some segments of the dive, and the ``wiggliness'' of each window in turn depends on the hidden state of an HMM.
 
+In this paper, we unify and build off of several existing methods from the ecology literature in novel ways to account for complex temporally dependent functional data.
+The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models and solve FDA problems.
 %This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. 
@@
 In the following section, we describe HMMs in addition to two variants (CarHMMs and HHMMs) and discuss methods to deal with fine-scale intricate dependence structures.
-We also show how these methods can be combined in order to analyze increasingly complex data.
+We also show how these methods can be combined to analyze increasingly complex data.
 In Section 3, we construct several candidate models to describe the data from Figure \ref{fig:data}.
 Section 4 details a simulation study based on these candidate models.
-In Section 5 we fit each candidate model to the original data and analyze the results, and in Section 6 we discuss the results.
+In Section 5 we fit each candidate model to the original data, and in Section 6 we discuss the results.

### 2020-07-31 Report/Sections/models.tex
Stylistic changes to report

@@
 
-Consider a sequence of $T$ possibly high-dimensional curves (or dives, in our context) with associated curve-level (or coarse-scale) observations $Y \equiv \left(Y_1, \ldots, Y_T\right)$.
-The coarse-scale observations contain summaries of each curve such as duration or maximum value.
-Curve $t$ has a total of $T^*_t$ within-curve (or fine-scale) associated observations: $Y^*_t \equiv \left(Y^*_{t,1}, \ldots Y^*_{t,t^*}, \ldots, Y^*_{t,T^*_t}\right)$.
+Consider a sequence of $T$ possibly high-dimensional curves (or dives, in our context) with associated curve-level (or coarse-scale) observations $Y \equiv \Big(Y_1, \ldots, Y_T\Big)$.
+The coarse-scale observations contain summaries of each curve such as curve length or maximum value.
+Curve $t$ has a total of $T^*_t$ within-curve (or fine-scale) associated observations: $Y^*_t \equiv \Big(Y^*_{t,1}, \ldots Y^*_{t,t^*}, \ldots, Y^*_{t,T^*_t}\Big)$.
 The fine-scale observations $Y^*_t$ make up a vector of function values from curve $t$, and we index $Y^*_t$ using $t^*$.
-The set of all fine-scale observations is denoted at $Y^* \equiv \left(Y^*_1,\ldots,Y^*_T\right)$.
+The vector of all fine-scale observations is denoted at $Y^* \equiv \Big(Y^*_1,\ldots,Y^*_T\Big)$.
 
-Hidden Markov models are useful tools to model state-switching Markovian processes in discrete time and will be used as the core structure of the models we use to describe $Y$ and $Y^*$.
+Hidden Markov models are useful tools to describe state-switching Markovian processes in discrete time and will be the core structure of the models we use to describe $Y$ and $Y^*$.
 %Traditional HMMs model process auto-correlation by assuming that the underlying behavior of a processes follows a single Markov chain.
@@
 We first describe the structure of a traditional HMM and a variation introduced by \cite{Lawler:2019} called the the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
-Either of these models can be used to model $Y$ or as a sub-model for $Y^*_t$.
-To simultaneously model both $Y$ and $Y^*$, we describe the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}, which simultaneously models $Y$ and $Y^*$ as two distinct (but related) hidden Markov models.
+Either of these models can be used to model $Y$ or $Y^*_t$ for a particular curve.
+To simultaneously model both $Y$ and $Y^*$, we describe the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}, which simultaneously models $Y$ and $Y^*$ as two distinct, but related, HMM.
 Next, we describe how to reduce fine-scale correlation by transforming observations using subject area expertise.
@@
 
-An HMM is comprised of a sequence of unobserved states $X_t$, $t = 1, \ldots, T$, which are associated with the observations $Y_t$, $t = 1, \ldots, T$.
+An HMM is comprised of a sequence of unobserved states $X_t$ $(t = 1, \ldots, T)$ which are associated with the observations $Y_t$.
 Here we focus on the coarse-scale observations $Y$, but an HMM can also be used as a sub-model for a particular fine-scale process $Y_t^*$.
@@
 The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$.
-Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$, for $t=1,\ldots, T-1$, and $i, j = 1,\ldots, N$.
+Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$.
 %
 We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i^{th}$ component
-$\delta_i = \Pr\{X_1 = i\},~ i = 1,\ldots,N.$
+$\delta_i = \Pr\{X_1 = i\}.$
 A Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
@@
 If $X_t = i$, then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes simply $f^{(i)}(\cdot)$.
-If $Y_t$ or $Y^*_{t,t^*}$ is a vector, we assume that the dimensions of the observations are independent $\left(f^{(i)}(\cdot) = \prod_{d} f^{(i)}_d(\cdot)\right)$, but this need not be the case in general.
+If $Y_t$ is a vector, we assume that the dimensions of the observations are independent $\left(f^{(i)}(\cdot) = \prod_{d} f^{(i)}_d(\cdot)\right)$, but this need not be the case in general.
 %
@@
 where $\eta \in \bbR^{N \times N}$ and $\eta_{ii}$ is set to zero for identifiability.
-Then $\calL_{\text{HMM}}(y;\Theta,\Gamma)$ can be maximized using a wide range of optimizers.
+This simplifies likelihood maximization and $\calL_{\text{HMM}}(y;\Theta,\Gamma)$ can be maximized using a wide range of optimizers.
 For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of  $\eta$.
@@
 
-The CarHMM, introduced by \citet{Lawler:2019}, explicitly models auto-correlation in the observations of an HMM.
+The CarHMM, introduced by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence.
 Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
 Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$, depends on \textit{both} $X_t$ \textit{and} $Y_{t-1}$ rather than only $X_t$.
-The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$. Figure \ref{fig:models} shows the dependence structure of a CarHMM.
-
+The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$.
 %
@@
 %
-The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
+Figure \ref{fig:models} shows a graphical representation of the dependence structure of a CarHMM.
+The likelihood for the CarHMM can be easily calculated using the forward algorithm.
+As previously, let $y$ be the vector of observed emissions.
+Then
 \begin{equation}
@@
 
-\subsection{Incorporating both $Y$ and $Y^*$ with an HHMM}
+\subsection{Incorporating multiple scales with an HHMM}
 
@@
 %
-In the hierarchical setting, however, each state $X_t$ emits yet another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observed emissions, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
-The fine-scale process ($X_t^*, Y_t^*$) makes up another HMM with parameters that depend on the value of $X_t$.
+In the hierarchical setting, each state $X_t$ also emits yet another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observations, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
+The fine-scale process ($X_t^*, Y_t^*$) then makes up another HMM.
+Given the coarse-scale hidden states, $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are all independent of one another.
+The parameters governing each fine-scale HMM $(X_t^*,Y_t^*)$ depend on the value of $X_t$.
 Specifically, if $X_t=i$, then the distribution of $X_t^*$ is characterized by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
-For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary.
-
-The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by a parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
+For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary in general.
+The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
 We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\Theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^*)}\right)$.
 
-Given the coarse-scale states, $X_1,\ldots, X_T$, the $T$ fine-scale processes $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are independent HMMs.
-Depending upon the process being modeled, it is possible to force certain parameters to be shared across different coarse or fine states.
-For example, in the killer whale case study (Section \ref{sec:case_study}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( \theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1, \ldots, N^* \right)$.
+Depending upon the process being modeled, it is possible to force certain parameters to be shared to reduce the complexity of the HHMM.
+For example, in the killer whale case study (Section \ref{sec:case_study}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( i.e.
+\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
+
 Figure \ref{fig:models} represents the dependence structure for an HHMM.
-
 Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easy to calculate via the forward algorithm.
@@
 To deal with this issue, we recommend replacing the fine-scale process $Y_t^*$ with relevant statistics that summarize this non-Markovian behavior.
-In order to maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over $Y_t^*$.
+Summary statistics can be calculated from a moving window over $Y_t^*$ in order to maintain the temporal structure of the fine-scale process.
 Subject-area experts are often required to determine the optimal window size and the specific summary statistics used.
-
+%
 %This moving window approach complicates the indexing of $Y_t^*$.
@@
 Note that this relabeling may significantly reduce the length of $Y^*_t$ depending upon the window size $h$ and the step-size between windows.
-
-For our particular application, we transform the fine-scale process using the discrete Fourier transform (DFT) of a two-second moving window across $Y^*$, where the moving window has a step size of two seconds:
+%
+For our particular application, we transform the fine-scale process using the discrete Fourier transform (DFT) of a moving window of width $h$ across $Y^*$:
 %
@@
 %
-for $k = 0, 1, \ldots, h-1$.
-The value of $h$ is equal to the window size, which is 100 for a two-second window over 50 Hz observations.
-Summary statistics are then used to drastically reduce the dimension of the DFT:
+for $k = 0, 1, \ldots, h-1$. Summary statistics are then used to drastically reduce the dimension of the DFT:
 %
@@
 As such, the emission distributions and the likelihood remain identical to the original HHMM when transforming $Y^*$ to $Z^*$.
-We refer to an HHMM which uses $Z^*$ as fine-scale observations as an HHMM-DFT.
+In order to clearly differentiate models, we refer to an HHMM with $Z^*$ as fine-scale observations as an HHMM-DFT.
 
@@
 
-By definition, an HHMM models both the coarse-scale and the fine-scale processes with a traditional HMM.
+By definition, an HHMM treats both the coarse-scale and the fine-scale processes as a traditional HMM.
 However, the coarse-scale HMM can be generalized by replacing it with a CarHMM, yielding the following likelihood:
@@
 On the fine scale, $Y^*_t$ can be modelled using \textit{any} parametric model which admits an easy-to-compute likelihood.
-The likelihood $\calL_{\text{HMM}}$ is then replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(y^*_t;\Theta^{*(i)})$:
+The likelihood $\calL_{\text{HMM}}$ is then replaced by the likelihood of this general fine-scale model, $\calL_{\text{fine}}(y^*_t;\Theta^{*(i)})$:
 \[
@@
 Possible candidates for the fine-scale model include, but are not limited to, any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT).
-These alternative fine-scale models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models based on HMMs.
+These alternative fine-scale models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models.
 As an example, consider a \textbf{CarHHMM}, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM.
-Alternatively, one may construct an \textbf{HHHMM}, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$), and traditional HMMs are used to model all three levels.
+Alternatively, one may construct an \textbf{HHHMM}, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$) and traditional HMMs are used to model all three levels.
 It is clear that infinitely many possible models can be built using this framework.
-However, these models can quickly become complicated and should be constructed achieve an adequate fit of the data while avoiding over-fitting and high computational costs.
+However, these models quickly become complicated and should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.
 %In the sections that follow, we construct several candidate models from the building blocks above to describe killer whale dive behavior.

### 2020-07-31 Report/Sections/results.tex
Stylistic changes to report

@@
 
-After fitting the four candidate models to simulated data, we fit the same four models to the data shown in Figure \ref{fig:data}.
-Here we report the results from the full CarHHMM-DFT (defined in Section \ref{sec:data}) in detail and asses the quality of the fit.
-We also compare the results of the CarHHMM-DFT with the other candidate models.
+We fit all four models from the simulation study to the data shown in Figure \ref{fig:data} to test their performance in a real-world setting.
+We first report the results from the full CarHHMM-DFT (defined in Section \ref{sec:data}) in detail and then asses the quality of the fit.
+Finally, we compare the results of the CarHHMM-DFT with those of the other candidate models.
 
-The estimates for the emission distribution parameters suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}).
-Dive type 1 corresponds to shorter, shallower dives. Ecologists have attributed a variety of purposes to these short dives such as resting, traveling, and searching for prey \citep{Tennessen:2019a}.
-Dive type 2 is longer and deeper.
-These types of deep sustained dives have been associated with behaviours such as foraging and socialization in killer whales \citep{Tennessen:2019a,Tennessen:2019b}.
-However, \citet{Wright:2017} note that most prey captures occur at depths deeper than 100m, so it is unlikely that any of the 267 dives used in this study are successful foraging dives.
+The emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). 
+Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, traveling, and searching for prey.
+On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as foraging and socialization \citep{Tennessen:2019b}.
+No dive in this data set has a maximum depth deeper than $\approx$ 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100m.
+It is therefore unlikely that any of the dives used in this study are successful foraging dives.
 
-The three subdive behaviours have distinct $Z^{*(2)}$ values, as the means of each subdive state are separated by an order of magnitude (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}).
-Subdive state 1 has the smallest mean value of $Z^{*(2)}$ and the smallest variance corresponding to $Z^{*(1)}$.
+The three subdive behaviours have distinct average ``wiggliness" values, as the means of $Z^{*(2)}$ for each state are separated by an order of magnitude (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}).
+Subdive state 1 has the smallest mean of $Z^{*(2)}$ and the smallest variance corresponding to $Z^{*(1)}$.
 It also has the highest auto-correlation in $Z^{*(1)}$.
 This implies less overall activity and more consistent acceleration compared to the other subdive states.
-Subdive state 2 has a mean Fourier sum ($Z^{*(2)}$) one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
-The auto-correlation of acceleration is also slightly lower in this state.
-We hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
+Subdive state 2 has a mean ``wiggliness" ($Z^{*(2)}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
+The auto-correlation of acceleration is also slightly lower in than subdive state 1.
+We therefore hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
 Finally, the mean of $Z^{*(2)}$ and variance of $Z^{*(1)}$ in subdive state 3 is much higher than the other two states.
-The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower, implying more variation in acceleration between consecutive segments.
+The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower, implying more variation in acceleration between segments.
 This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).
@@
 %
-for the transitions from dive to dive. For dive types 1 and 2,
+for the transitions between dives. Within dives, we have
 $$\hat \Gamma^{*(1)} = \begin{pmatrix} 
@@
 %
-for the fine-scale within-dive transitions.
-The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that most dives are short type 1 dives.
+for dive types 1 and 2.
+The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that most dives are short, type 1 dives.
 The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive.
-This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short dives before a killer whale performs a longer, deeper dive.
+This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
 The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is more likely to be in the less active subdive state 1 when performing long deep dives than when performing short shallow dives.
 Using less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}.
-Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of 6 selected dives.
+Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives.
 The supplementary material also shows the estimated probability of each dive and subdive state given the data (Section S-2.2).
@@
 
-Two visual tools were used to evaluate this model: pseudo-residuals and empirical histograms.
+Two visual tools are used to evaluate this model: pseudo-residuals and empirical histograms.
 A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}.
 To easily visualize outliers, pseudo-residuals are often passed through the quantile function of the standard Normal distribution.
-Mathematically, the pseudo-residual of an observation $y_t$ in a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
+Mathematically, the pseudo-residual of an observation $y_t$ of a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
 If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution.
-Histograms of the pseudoresiduals of this model mostly support that the model is well-specified.
+Histograms of the pseudoresiduals mostly support that this model is well-specified.
 One exception is $Z^{*(2)}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
-This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution than the gamma distribution.
+This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution than the gamma distribution used in the case study.
 
-We also plot histograms of $Y$ and $Z^*$ separated by dive type or subdive state in Figure \ref{fig:empirical_dist}.
-Observations are weighted by the decoded probability that they correspond to a particular hidden state, and the histogram is finally plotted over the emission distribution learned by the CarHHMM-DFT (Figure \ref{fig:empirical_dist}).
+We also plot histograms of $Y$ and $Z^*$ corresponding to each dive type or subdive state in Figure \ref{fig:empirical_dist}.
+Observations are weighted by the decoded probability that they correspond to a particular hidden state, and this histogram is plotted over the emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}).
 Our results again mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed.
@@
 The CarHMM-DFT produces very similar parameter estimates and state estimates compared to those of the CarHHMM-DFT on the fine scale.
-However, its lack of a hierarchical structure means that it fails to differentiate between short and long dives, and it therefore does not provide any information relating to the dive-level Markov chain or the relationship between it and the subdive level.
-For example, the CarHMM-DFT does not indicate that the whale is much more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives.
+However, its lack of a hierarchical structure means that it fails to differentiate between short and long dives.
+This model therefore does not give any information regarding the dive-level Markov chain or the relationship between the dive and subdive levels.
+For example, the CarHMM-DFT does not indicate that the whale is much more likely to be in subdive state 1 when engaged in longer dives compared to shorter one.
 
 The HHMM-DFT decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3.
-In addition, the HHMM-DFT produces estimates of $\sigma_1^{*(1)}$, $\sigma_1^{*(2)}$, and $\sigma_1^{*(3)}$ (the standard deviation of acceleration) which are approximately 50-100 \% larger than those for the CarHHMM-DFT across all axes of acceleration.
+In addition, the HHMM-DFT produces estimates of $\sigma_1^{*(1)}$, $\sigma_1^{*(2)}$, and $\sigma_1^{*(3)}$ (the standard deviation of acceleration) which are approximately 50-100 \% larger than those of the CarHHMM-DFT across all axes of acceleration.
 The estimated uncertainties of the estimates of $\mu_1^{*(1)}$, $\mu_1^{*(2)}$, and $\mu_1^{*(3)}$ are also less than half of those for the CarHHMM-DFT across all axes of acceleration.
 These findings are consistent with the tendency to overestimate variance and underestimate standard error when ignoring auto-correlation.
-The empirical auto-correlation of $Z^{*(1)}$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT show several outliers.
+Further, the empirical auto-correlation of $Z^{*(1)}$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers.
 Both of these findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.
 
-The CarHHMM does not use the Fourier sums of the acceleration data, so it fails produce biologically interpretable results when decoding the subdive states.
+Finally, the CarHHMM does not use the ``wiggliness" of the acceleration data as an observation, so it fails produce biologically interpretable results when decoding the subdive states.
+This fact alone essentially disqualifies the CarHHMM as a viable model for this data.
 The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution.

### 2020-07-31 Report/Sections/simulation.tex
Stylistic changes to report

@@
 
-To compare the performance of the candidate models when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT model.
+To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT.
 The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}.
-The ground-truth parameters differ slightly from those in the case study for primarily pedagogical reasons.
-For example, we set $\Gamma$ such that the simulated data contains approximately equal numbers of short and long dives.
+The ground-truth parameters differ slightly from those learned in the case study primarily for simplicity of presentation.
+For example, we set $\Gamma$ such that the simulated data has approximately equal numbers of short and long dives.
 This allows meaningful conclusions to be made for both dive types.
-In addition, the number subdive states is set to $N^*=2$ to simplify the analysis of the subdive states, and $Z^{*(2)}$ is changed from a vector to a scalar for additional simplicity.
-
-We fit all four candidate models to the data generated from the CarHHMM-DFT to analyze the impact of model mis-specification and compare them using several different metrics.
-These metrics include the decoding accuracy of the dive types and subdive states, bias in parameter estimates, standard error of parameter estimates, and fitting times.
-We also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the observed Fisher information under that model.
+In addition, the number subdive states is set to $N^*=2$ and $Z^{*(2)}$ is set as a scalar for additional simplicity.
+Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times.
+We also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.
 This allows us to judge how accurate each model is when estimating its own uncertainty.
@@
 Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives.
-Each dive is one of $N=2$ dive types and the coarse-scale probability transition matrix is set as
+Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
 %
@@
 %
-Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are set as
+Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are
 $$
@@
 %
-After the dive durations are calculated for all 100 dives for each data set using $\Gamma$ and $\Theta$, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model.
+After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model.
 The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive are discarded.
@@
 Recall that $Z^{*(1)}$ follows a CarHMM and is Normally distributed while $Z^{*(2)}$ follows an HMM and is Gamma distributed.
-It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe a method to reconstruct one possible realization of $Y^*$ in the appendix.
+It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix.
 Figure \ref{fig:sim_data} shows this reconstructed sequence along with $Z^*$ for five dives of one simulated data set.
 
+The two dive types differ in that dives of type 1 are much shorter on average than dives of type 2.
 The two subdive types differ primarily due to $\mu_2^*$ and $\sigma_2^*$, both of which are much higher in subdive state 2.
-This corresponds to much more vigorous and variable periodic behavior in the acceleration data.
+This corresponds to much more vigorous and variable periodic behavior in the acceleration data for subdive state 2.
 
@@
 As expected, the full CarHHMM-DFT is the best-performing model.
-Given the true subdive type of a particular segment $X^*_{t,t^*}$, the CarHHMM-DFT predicts the correct subdive type with an average confidence of $1 - 10^{-12}$ across all simulations.
-For the dive types, this average confidence was above 0.9 (see Table \ref{table:accuracy}).
-Estimates of hidden states are made using the forward-backward algorithm.
+Using the forward-backward algorithm, the CarHHMM-DFT decodes the correct subdive type given the fine-scale data with an average confidence of $1 - 10^{-12}$ across all simulations.
+For dive type the average confidence was above 0.9 (see Table \ref{table:accuracy}).
 All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased.
@@
 The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias.
-The HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$.
+However, the HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$.
 For both subdive types, $\hat \sigma_1^*$ greatly overestimates the true standard deviation.
-The empirical standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ are both much larger than the estimates of standard error using the observed Fisher information (see Section S-1.2, Table 3).
-This implies that parameter estimates can be too confident when auto-correlation is ignored.
+The estimated standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ using the observed Fisher information are both much smaller than the empirical standard error (see Section S-1.2, Table 3).
+This implies that variance can be overestimated and parameter estimates can be too confident when auto-correlation is ignored.
 
 The CarHHMM is the worst-performing model in terms of decoding accuracy, with an average confidence below $0.9$ when determining both dive and subdive types.
-The CarHHMM does not have access to the Fourier sums, which are the most informative observations in terms of determining subdive type.
-It is much more likely to mis-classify subdive type 1 than subdive type 2 (see Section S-1.2, Table 1).
-The CarHHMM is also the worst model in terms of parameter estimates, in part due to its relatively poor decoding accuracy.
-$\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ are especially poor estimates under the CarHHMM (see Section S-1.1.3).
+This result is consistent with expectations because the CarHHMM does not have access to the ``wiggles" of the fine-scale process, which are the most informative observations when determining subdive type.
+The CarHHMM is also much more likely to mis-classify subdive type 1 than subdive type 2 (see Section S-1.2, Table 1).
+The CarHHMM is the worst of the four at estimating parameters, and this is in part due to its relatively poor decoding accuracy.
+$\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ are especially poor estimates (see Section S-1.1.3).
 
@@
 The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of \textit{all} dives.
-These parameter estimates $\hat \mu$ and $\hat \sigma$ are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds.
+The parameter estimates for this distribution ($\hat \mu$ and $\hat \sigma$) are highly correlated and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds.
 See Figures \ref{fig:acc_coarse} \& \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive/subdive type probabilities as determined by each model.

### 2020-08-03 Report/Sections/intro.tex
Update report- Marie is updating

@@
-The study of high-frequency time series has been profoundly affected by recent advances in tracking technology which provide researchers with data collected almost continuously in time \citep{Hooten:2017}.
-The collection and analysis of data from devices such as accelerometers has brought new understanding to areas ranging from monitoring machine health \citep{Getman:2009} to physical activity levels in children \citep{Morris:2007}.
-In this article, we focus on the study of animal movement; tags attached to animals can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
-These data provide biologists with a wealth of knowledge about animal behaviour, but parsing the vast amount of information is a challenge for statisticians.
+Tracking technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
+The collection and analysis of data from devices such as accelerometers has brought new understanding to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}.
+In particular, the study of animal movement has been transformed by tracking devices that collects biologging data (CITATION).
+Tags attached to animals can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
+These data provide biologists with a wealth of knowledge about animal behaviour, but parsing this vast amount of information is a challenge for statisticians.
 
@@
 In \cite{Morris:2007}, a child's activity level throughout the day is viewed as a curve where metabolic activity is a function of time.
-In our case study, we view a single dive of a killer whale as several curves where acceleration and depth are functions of time.
+The dive profile of marine mammals can be viewed as a curve, and Fu (CITATION) showed that [methods used by Eric and Nancy] is useful to classify [blah blah blah].
+However, now many such animals are tracked with multiple devices providing information at different scales.
+For example, in our case study, a single dive of a killer whale (\textit{Orcinus orca}) can be viewed as having several curves where acceleration and depth are functions of time.
 
-FDA was originally developed for curves that can be considered independent replicates, but the structure of the data is often more complex than this.
-In particular, we focus on functional data which display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale).
-While our methodology is applicable in a wide variety of settings, we focus on the dive patterns of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada (see Section \ref{sec:data} for more information about data collection and preprocessing).
-The dive profile displayed in the top panel of Figure \ref{fig:data} clearly shows temporal dependence \textit{between} dives, as dives of various time durations and depths appear to be clustered together in time.
-Previous studies suggest that each of these dives have one of a discrete number of purposes (or types), including resting, traveling, or foraging \citep{Tennessen:2019b}.
+FDA was originally developed for curves that can be considered independent replicates, but the structure of time-series data is often complicated by additional dependencies.
+In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) (CITATION would be good - Fu's discussion?
+Our Vianey's paper or Timo Adam's paper?).
+Our case study data clearly demonstrate this common additional dependencies between curves.
+The dive profile (Figure \ref{fig:data}A) of a killer whale from the threatened Northern resident population off the coast of British Columbia, Canada (see Section \ref{sec:data} for more information about data collection and preprocessing) clearly shows temporal dependence \textit{between} dives, as dives of various time durations and depths appear to be clustered together in time.
+Previous studies have categorized these dives into different types, each associated with a different behavior, including resting, traveling, or foraging \citep{Tennessen:2019b}.
 Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence \textit{within} dives, as the accelerometer data within each dive indicate bouts of short-term periodicity.
 The combination of apparent behavioral changes combined with periodic behavior makes this accelerometer data difficult to model using simple stochastic processes.
+[Here if you can say something about how common this combined periodicity and behavioral dependece is in other accelerometer data, e.g.
+in human with a citation, you'd really effectively broaden the generality of the problem.]
 
-Models for particular types of between-curve dependence structures appear in the FDA literature, but these methods fall short when modelling the coarse-scale temporal structure seen here.
-One example of such methods are multilevel models which incorporate dependence via random effects.
-In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors use a multilevel approach since each child yields not one but two daily activity curves; the model includes a random effect for ``child''.
-In a classic early paper that introduced this multilevel structure,  \cite{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles.
+The FDA literature provides models that account for some between-curve dependence structures, but these methods are not adequate for modelling the coarse-scale temporal structure seen here.
+For example, multilevel models incorporate dependence via random effects.
+In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors use a multilevel approach since each child yields two daily activity curves; the model includes a random effect for ``child''.
+In a classic paper that introduced this multilevel structure,  \cite{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles.
 More recent work on multilevel models includes that of \cite{di:2009}, \cite{chen:2012} and Crainiceanu [REF].
@@
 Traditional FDA techniques similarly fall short when modelling the within-curve acceleration data seen here.
-In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function.
+In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function (e.g., CITATION- an example with accelerometer data would be particularly good - just did a quick search and they may not be appropriate but could look at: https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0169576 or https://academic-oup-com.ezproxy.library.ubc.ca/biostatistics/article/16/2/352/266114).
 Alternatively, this structure is sometimes generated via random regression.
-These classic FDA techniques would be hard-pressed to model the type of radically changing behaviour that we see in the accelerometer data of this dive sequence.
+Radically changing behavior that we see in the accelerometer data of similar to our dive sequence would be challenging to modelled with these classic FDA techniques.
 
@@
 HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of the animal from movement observations.
-While ubiquitous in the ecology literature, HMMs have seen little use in FDA, not including the work of Roland Langrock (see, for instance \citealt{Langrock:2018}).
+The ubiquitous applications of HMMs in ecology have seen little use in FDA, with a few notable exceptions.
+In particular,  \citet{Langrock:2018} did [blah blah blah], which does not deal with [blah blah blah].
+We show here how HHMs can provide a new tool to the FDA toolbox.
 
@@
 
-We also borrow techniques from animal movement to model within-dive behaviour beyond a simple HMM.
-Under the traditional HMM model, observations are independent given the underlying hidden state process.
-However, this assumption is often violated in the real-world processes, especially when observations are taken a high frequencies.
+Under the traditional HMM, observations are independent given the underlying hidden state process.
+However, this assumption is often violated in real-world processes, especially when observations are taken at high frequencies.
 Several solutions have been proposed in the ecology literature, such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
@@
 We borrow from ideas from both the CarHMM and the HHMM here.
-Part of the purpose of our article is to introduce the FDA community to these useful approaches to modelling complex curves.
 
-However, even the ecology literature falls short at the shortest time scales, which often exhibit complicated structure that cannot easily be captured by traditional HMMs, CarHMMs, or HHMMs.
-One example is the fine-scale periodic swimming pattern of horn sharks off the coast of Southern California \citep{Adam:2019}.
-For periodic behavior in particular, \cite{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM in the context of describing daily behavioural cycles of marine mammals.
+Traditional HMMs, CarHMMs, or HHMMs cannot easily capture the complicated structure that biologging data often exhibit at short time scales.
+For example, the HHMM of \citet{Adam:2019} did not capture the fine-scale periodic swimming pattern of  horn sharks (LATIN NAME).
+However, \citet{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM to account for periodic behavior in the context of describing daily behavioural cycles of marine mammals.
 In addition, Fourier analysis has previously been used on accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
-Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears to be a promising and simple approach to account for periodic structures in fine-scale data.
-In particular, we slide a moving window over the raw acceleration data and perform a biologically meaningful transformation of the data in this window.
-This transformation accounts for the ``wiggliness'' in some segments of the dive, and the ``wiggliness'' of each window in turn depends on the hidden state of an HMM.
+Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears to be a promising approach to account for periodic structures in fine-scale data.
 
@@
 
-In the following section, we describe HMMs in addition to two variants (CarHMMs and HHMMs) and discuss methods to deal with fine-scale intricate dependence structures.
+In this paper, we first describe HMMs in addition to two variants (CarHMMs and HHMMs) and discuss  how Fourier analysis can handle fine-scale dependence structures.
 We also show how these methods can be combined to analyze increasingly complex data.
@@
 Section 4 details a simulation study based on these candidate models.
-In Section 5 we fit each candidate model to the original data, and in Section 6 we discuss the results.
+In Section 5, we fit the candidate models to the original data, and in Section 6 we discuss the results.

### 2020-08-06 Report/Sections/appendix.tex
Update Report

@@
 
-Michelot \& Blackwell (2019) model the movement of an animal as the solution to the following stochastic differential equation, which they refer to as a state-switching Ornstein-Uhlenbeck process:
+\citet{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation, which they refer to as a state-switching Ornstein-Uhlenbeck process:
 %
@@
     \label{eqn:SDE}
-    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
+    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t,
 \end{equation}
@@
 Unlike HMMs, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer.
-If the behavioral state $X_t$ is known and does not change between observations, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
+If the behavioral state $X_t$ is known and does not change between observations $\Big($i.e $X_s = X_t$ for all $s \in [t,t+\Delta t)\Big)$, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
 %
@@
 %
-where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$
+where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$.
+In fact, under certain conditions, the output of a state-switching OU process is identical to that of a CarHMM, as described by Theorem 1.
 
@@
 %
-If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t: (s-1) \Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:
+If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:
 
@@
 %
-$X'$ follows a Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s+1})$ is consistent with that of a CarHMM with normal emission distributions.
+$X'$ follows a (possibly unobserved) Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s-1})$ is consistent with that of a CarHMM with Normal emission distributions.
 \end{proof}
@@
 
-\begin{proof}{Description of simulated data}{}
+\begin{proof}{Detailed description of simulated data}{}
 
-$\hat{Y}^*_{t,t^*}$ was simulated using the following procedure. The $k^{th}$ Fourier mode of $\hat{Y}^*_{t,t^*}$ is denoted as $\hat{Y}^{*(k)}_{t,t^*}$:
+Assume that $X$, $Y$, and $X^*$ are known and simulated using the procedure detailed in Section \ref{sec:sim_study}.
+Let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and let $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for shorthand.
 
-\begin{align}
-	(\hat{Y}^{*(0)}_{t,0}|X^*_{t,0} = i^*) &\sim \mathcal{N} \left(0, \sigma^{*(i^*)} \right) & \nonumber \\
-	%
-	(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*) &\sim \mathcal{N} \left(\phi^{*(i^*)} * \hat{Y}^{*(0)}_{t,t^*-1}, \sigma^{*(i^*)} \right), & t^* = 1,2,\ldots, \lfloor Y_t/2 \rfloor \label{eqn:y_hat0} \\
-	%
-	\hat{Y}^{*(k)}_{t,t^*} &= a_{t,t^*}^{(k)} i\sqrt{b^{(k)}_{t,t^*}}, & k = 1,\ldots,49 \nonumber
-\end{align}
-
+Then, $\hat{Y}^*_{t}$ is simulated using the following procedure for every dive. First, simulate the average acceleration of the first two-second window of the dive:
+$$
+	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(\sigma^{*(i^*)}\right)^2 \right)
+$$
+Remember that the average acceleration within a window is equal to the real part of the $0^{th}$ Fourier mode, so $\hat{Y}^{*(0)}_{t,1}$ is the average acceleration within the first window.
+Next, simulate the average acceleration for each subsequent two-second window:
+\begin{equation}
+	\Big(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*,\hat{Y}^{*(0)}_{t,t^*-100}\Big) \sim \mathcal{N} \left(\phi^{*(i^*)} \hat{Y}^{*(0)}_{t,t^*-100}, \left(\sigma^{*(i^*)}\right)^2 \right), \quad t^* = 101,201,\ldots
+	\label{eqn:yhat_0}
+\end{equation}
+Note that $t^*$ has increments of 100 because $h=100$ when using two-second window sizes. Finally, simulate the other Fourier modes within each window:
+$$
+	\hat{Y}^{*(k)}_{t,t^*} = a_{t,t^*}^{(k)} i\sqrt{b^{(k)}_{t,t^*}}, \quad k = 1,\ldots,49;
+$$
+where $a$ and $b$ are random variables with the following distributions:
 \begin{align*}
@@
 	{\rm{Gamma}}(250,1), & k = 1 \\
-	{\rm{Gamma}}(50,1), & k = 2
+	{\rm{Gamma}}(50,1), & k = 2.
 	\end{array}\right. 
 \end{align*}
-
-\begin{align*}
-    \hat{Y}^{*(50)}_{t,t^*} &= 0 & \\
+Note that $b_{t,t^*}$ decays like $1/k^2$ to ``smooth out" the raw acceleration. For Fourier modes 50 through 99, we use the following procedure to ensure that the raw acceleration is real-valued:
+$$
+    \hat{Y}^{*(50)}_{t,t^*} = 0; \qquad
+	\hat{Y}^{*(k)}_{t,t^*} = -\hat{Y}^{*(100-k)}_{t,t^*} \enspace \text{for} \enspace k = 51,\ldots,99.
+$$
 %
-	\hat{Y}^{*(k)}_{t,t^*}  &= -\hat{Y}^{*(100-k)}_{t,t^*}, & \qquad k = 51,\ldots,99
-\end{align*}
-
-$Y^*_{t,t^*,1:100}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
-
-$$Y^*_{t,t^*,1:100} = IDFT\left(\hat{Y}^*_{t,t^*}\right), \qquad t^* = 1,\ldots,\lfloor Y_t/2 \rfloor$$
-
-$\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued.
+$\Big(Y^*_{t,t^*},\ldots,Y^*_{t,t^*+99}\Big)$ is finally set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
+%
+$$Y^*_{t,t^*+n} = IDFT\left\{\hat{Y}^*_{t,t^*}\right\}(n); \quad n = 0,\ldots,99; \quad t^* = 1,101,\ldots$$
+%
+%$\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued.
 $\hat{Y}^{*(k)}_{t,t^*}$ also decays like $1/k$ so that $Y^*_{t,t^*}$ remains continuous within a two-second window.
@@
 
-From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$.
-We pick $\tilde{\omega} = 10$ periods per window, or 5 hertz.
-To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use Equation (\ref{eqn:y_hat0}):
+We now prove that this construction of $Y^*$ results in the correct distributions listed in Section \ref{sec:sim_study}.
+
+It is straightforward to calculate both $\Zone$ and $\Ztwo$ from $Y^*$.
+We pick $\tilde{\omega} = 10$ periods per window, or 5 Hz.
+To find the distribution of $\Zone_{t,t^*}$, use Equation (\ref{eqn:yhat_0}) together with the substitution $\Zone_{t,t^*} = \text{Re}\left(\hat{Y}^{(0)}_{t,t^*}\right) = \hat{Y}^{(0)}_{t,t^*}$:
 %
-$$\left(Z^{*(1)}_{t,t^*} | X^*_{t,t^*} = i^* \right) = \left(\mathcal{R}\left(\hat{Y}^{*(0)}_{t,t^*}\right) | X^*_{t,t^*} = i^* \right) \sim \mathcal{N} \left(\phi^{*(i^*)} * Z^{*(1)}_{t,t^*-1}, \sigma^{*(i^*)} \right)$$
+$$\left(\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(\phi^{*(i^*)} \Zone_{t,t^*-100}, \left(\sigma^{*(i^*)}\right)^2 \right).$$
 %
-$Z^{*(2)}_{t,t^*}$ is the sum of gamma-distributed random variables with the same scale parameter, so the distribution of $Z^{*(2)}_{t,t^*}$ is also a gamma distribution:
+This distributions is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(i^*)} = 0$.
+The other parameters $\sigma_A^{*(i^*)}$ and $\phi_A^{*(i^*)}$ are free, and we select $\sigma_A^{*(1)} = 0.05s$, $\phi_A^{*(1)} = 0.99$, $\sigma_A^{*(2)} = 0.1s$, and $\phi_A^{*(2)} = 0.95$.
+
+The wiggliness $\Ztwo_{t,t^*}$ is the sum of gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also gamma:
 %
-$$Z^{*(2)}_{t,t^*} = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
+$$\Ztwo_{t,t^*} = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
 %
-$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = \sum_{k=1}^{10} 5/k^2 = 10.10, \beta = 1.00 \right)$$
+$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = \sum_{k=1}^{10} 5/k^2 = 10.10, \beta = 1.00 \right)$$
 %
-$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(\alpha = 300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , \beta = 1.00 \right)$$
+$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(\alpha = 300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , \beta = 1.00 \right).$$
+%
+It follows that $\mu_W^{*(1)} = 10.10$, $\sigma_W^{*(1)} = 3.18$, $\mu_W^{*(2)} = 305.94$, and $\sigma_W^{*(2)} = 17.49$.
 \end{proof}
@@
 %
-$$\calL_{\text{CarHHMM-DFT}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N$$
+$$\calL_{\text{CarHHMM-DFT}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N.$$
 %
-where:
+In particular,
 %
@@
 P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
-&f^{(N)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big]
+&f^{(N)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big],
 \end{align*}
 %
-$f^{(i)}(y_t;\Theta^{(i)})$ is the emission distribution of the dive duration $y_t$ conditioned on the fact that $X_t = i$.
+where $f^{(i)}(y_t;\Theta^{(i)})$ is the emission distribution of the dive duration $Y_t$ conditioned on the fact that $X_t = i$.
 
-$\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale chain:
+The likelihood $\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale chain and has the following expression:
 %
-$$\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(i)},\Gamma^{*(i)}\right) = \delta^{*(i)} \prod_{t=2}^T \Gamma^{*(i)} P(z^*_{t,t^*}|z^*_{t,t^*-1};\Theta^{*(i^*)}) \mathbf{1}_N$$
+$$\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(i)},\Gamma^{*(i)}\right) = \delta^{*(i)} \prod_{t=2}^T \Gamma^{*(i)} P(z^*_{t,t^*}|z^*_{t,t^*-100};\Theta^{*(i^*)}) \mathbf{1}_N,$$
 %
-where $P(z^*_{t,t^*}|z^*_{t,t^*-1};\Theta^{*(i)})$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$.
+where $P(z^*_{t,t^*}|z^*_{t,t^*-100};\Theta^{*(i^*)})$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{*(\cdot,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$.
 %
-Note that $f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$ is the emission distribution of $Z^*_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $Z^*_{t,t^*-1} = z^*_{t,t^*-1}$.
+Note that $f^{*(\cdot,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$ is the emission distribution of $\Z_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.
 

### 2020-08-06 Report/Sections/data.tex
Update Report

@@
 Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}.
-NRKW prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus
+HMMs have been used to understand the how diving behaviors of various species are affected by disturbances (e.g.
+https://projecteuclid.org/euclid.aoas/1491616885  and https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.2044).
+For NRKWs, we are interested in categorizing different diving behaviors, with a particular focus on differentiating potential foraging dives.
+NRKWs prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus
 tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKW for predation \citep{Williams:2009,Noren:2011}.
@@
 The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions.
-Observations were collected at a rate of 50 Hz.
+Observations were collected at a rate of 50 Hz using tag [NAME OF TAG - ask Sarah].
 Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored.
-In addition, the tagging technology malfunctioned between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so all dives within this time range are ignored as well.
+We also ignore all dives between 2:25pm and 2:37pm and between 4:07 and 5:07, when the tagging technology malfunctioned.
 A killer whale ``dive" is any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
@@
 Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
-Specifically, $Y$ is a time series of dive durations in seconds ($s$) with corresponding hidden dive types $X$.
-The fine-scale process $Y^*$ is the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive states $X^*$.
+Specifically, the observed coarse-scale observation set $y$ is a time series of dive durations in seconds ($s$) with corresponding unobserved dive types $x$.
+The fine-scale observations $y^*$ are the within-dive acceleration data in meters per second squared ($m/s^2$) with corresponding subdive states $x^*$.
 
-\subsection{Model Selection}
+\subsection{Model Development and Selection}
 
-A lag plot is used to determine where auto-correlation should be incorporated into the model (see Section S-2.1).
+To choose how to model the coarse-scale diving data, we first inspect lag plots to determine whether auto-correlation should be incorporated (see Section S-2.1).
 The coarse-scale level exhibits no significant amount of auto-correlation and no obvious intricate structure prior to fitting the model.
@@
 
-The fine-scale acceleration data exhibits significant sinusoidal behavior; thus $Y^*$ is transformed to $Z^*$ using the DFT summary statistics.
-A window size of two-seconds is selected since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data.
-The acceleration data was also collected in three dimensions which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left).
+In contrast, the fine-scale acceleration data exhibits significant sinusoidal behavior.
+Thus, we transform $Y^*$ to $Z^*$ using the DFT for this model.
+We select a window size of two-seconds since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data.
+The acceleration data was collected in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left).
 Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
@@
 \begin{equation*}
-    Z_{t,t^*}^{*(1)} := \text{Re}\left(DFT\{Y^*_{t,t^*}\}(0)\right), \qquad Z_{t,t^*}^{*(2)} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
+    \Zone_{t,t^*} := \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n}, \qquad \Ztwo_{t,t^*} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
 \end{equation*}
 %
-Each observation is made up of $Z_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $Z_{t,t^*}^{*(2)}$, a scalar.
-We calculate $Z_t^{*(2)}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to a maximum frequency of 5 Hz.
+Each observation is made up of $\Zone_{t,t^*}$ (a 3-dimensional vector of average acceleration) and $\Ztwo_{t,t^*}$ (a scalar describing ``wiggliness").
+We calculate $\Ztwo_{t,t^*}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to a maximum frequency of 5 Hz.
 
-Even after accounting for fine-scale structure in the raw acceleration data, there is still strong auto-correlation within each axis of $Z^{*(1)}_{t,t^*}$ prior to fitting the model (Section S-2.1).
-Therefore, a CarHMM is used to model $Z^{*(1)}_{t,t^*}$.
-In particular, if $X^*_{t,t^*} = i^*$, then we assume $Z^{*(1)}_{t,t^*}$ is Normally distributed with unknown parameters $\phi_1^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i^*)} \in \mathbb{R}^3$.
+Even after accounting for fine-scale structure in the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (Section S-2.1).
+Therefore, a CarHMM is used to model $\Zone_{t,t^*}$.
+In particular, if $X^*_{t,t^*} = i^*$, then we assume $\Zone_{t,t^*}$ is Normally distributed with unknown parameters $\phi_A^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_A^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_A^{*(i^*)} \in \mathbb{R}^3$.
 In words, each axis of acceleration has its own mean and variance, but all axes share an auto-correlation parameter.
 %
-%$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)},$$
-%$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right].$$
+%$$\mathbb{E}(\Zone_{t,t^*}|\\Zone_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_A^{*(i)} \mathbf{z} + (1-\phi_A^{*(i)}) \mathbf{\mu}_A^{*(i)},$$
+%$$\mathbb{V}(\Zone_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_A^{*(i)}\right)^2\right].$$
 %
 
-While $Z^{*(2)}$ also exhibits some auto-correlation prior to fitting the model, the relationship is less strong than that of $Z^{*(1)}$, and the biological interpretation of auto-correlation within $Z^{*(2)}$ is less clear.
+While $\Ztwo$ also exhibits some auto-correlation prior to fitting the model, the relationship is less strong than that of $\Zone$, and the biological interpretation of auto-correlation within $\Ztwo$ is less clear.
 In addition, much of the auto-correlation evident from the lag plot may be explained by subsequent observations sharing a subdive state $X^*_{t,t^*}$.
-Therefore, a simple HMM is used to model $Z^{*(2)}$.
-Given that $X^*_{t,t^*} = i^*$, the distribution of $Z^{*(2)}_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_2^{*(i^*)}$ and variance $\sigma_2^{*(i^*)}$.
+Therefore, a simple HMM is used to model $\Ztwo$.
+Given that $X^*_{t,t^*} = i^*$, the distribution of $\Ztwo_{t,t^*}$ is assumed to be gamma and parameterized by its mean $\mu_W^{*(i^*)}$ and variance $\sigma_W^{*(i^*)}$.
 %
-%$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
-%$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
+%$$\mathbb{E}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_W^{*(i)}$$
+%$$\mathbb{V}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_W^{*(i)}\right)^2.$$
 %
@@
 The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale.
-On the fine scale, $Y^*$ is transformed to $Z^*$ using a DFT, $Z^{*(1)}$ is modelled with a CarHMM, and $Z^{*(2)}$ is modeled with an HMM.
-We also force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity.
+On the fine scale, $Y^*$ is transformed to $Z^*$ using a DFT, $\Zone$ is modelled with a CarHMM, and $\Ztwo$ is modeled with an HMM.
+We also force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity.
 This implies that dive types 1 and 2 share subdive states.
@@
     %
-    \theta^{*(\cdot,i^*)} =  \{\{\mu_1^{*(\cdot,i^*)},\sigma_1^{*(\cdot,i^*)},\phi_1^{*(\cdot,i^*)}\},\{\mu_2^{*(\cdot,i^*)},\sigma_2^{*(\cdot,i^*)}\}\} \qquad \text{(}Z^{*(1)} \text{ and } Z^{*(2)} \text{ parameters).}
+    \theta^{*(\cdot,i^*)} =  \{\{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)}\},\{\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}\}.
 \end{gather*}
 %
-Recall that $\theta_j^{*(\cdot,i^*)}$ are the parameters describing the distribution of $Z^{*(j)}_{t,t^*}$ when conditioning on $X^*_{t,t^*} = i^*$.
+Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $Z_{t,t^*}$ when conditioned on $X^*_{t,t^*} = i^*$.
 
-We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMM models discussed so far.
+We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMMs discussed so far.
 The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
@@
 Figure \ref{fig:CarHHMM-DFT} also shows the corresponding graphical model for the CarHHMM-DFT.
-In addition to the CarHHMM-DFT we consider three similar variations for comparison:
+In addition to the CarHHMM-DFT, we consider three similar variations for comparison:
 \begin{enumerate}
-    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using the DFT, but models \textit{both} $Z^{*(1)}$ \textit{and} $Z^{*(2)}$ with a simple HMM rather than a CarHMM.
-    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ using a CarHMM.
-However, the Fourier sums $Z^{*(2)}$ are omitted from this model altogether.
+    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using the DFT, but models \textit{both} $\Zone$ \textit{and} $\Ztwo$ with a simple HMM rather than a CarHMM.
+    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using the DFT, and models $\Zone$ using a CarHMM.
+However, the Fourier sums $\Ztwo$ are omitted from this model altogether.
 The CarHHMM therefore does not know about the ``wiggliness" of the accelerometer data.
-    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $Z^{*(1)}$ with a CarHMM and $Z^{*(2)}$ with an HMM.
+    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations \textit{as an independent and identically distributed sequence of dives}, transforms the fine-scale observations using the DFT, and models $\Zone$ with a CarHMM and $\Ztwo$ with an HMM.
 This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive.
 \end{enumerate}
-All models have emission distributions identical to the CarHHMM-DFT when applicable.
-Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the ``wiggliness" $(Z^{*(2)})$, and the CarHMM-DFT lacks a hierarchical structure.
+
+Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the ``wiggliness" $\Ztwo$, and the CarHMM-DFT lacks a hierarchical structure.

### 2020-08-06 Report/Sections/discussion.tex
Update Report

@@
 
-In this work, we deal with both temporal dependence between curves and a discrete number of curve types by modelling the sequence of curves with either an HMM or CarHMM, depending upon the level auto-correlation between curves.
-Several useful model selection tools such as a lag plot can be used to test if there is significant auto-correlation between curves.
-One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use yet another within-curve parametric model as the HMM emissions.
+In this work, we deal with both temporal dependence between curves and a discrete number of curve types by using either an HMM or CarHMM, depending upon the level auto-correlation between curves.
+To help select the appropriate model, lag plots can be used to test if there is significant auto-correlation between curves.
+One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use another within-curve parametric model as the HMM emissions.
 The parameters of this fine-scale model then in turn depend upon the curve type.
@@
 For periodic behaviour in particular, we use a DFT-based transformation.
-The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when including $Z^{*(2)}$ in the model, and our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale.
-On top of the DFT transformation, we also suggest using a CarHMM instead of a traditional HMM when expert opinion and careful observation (aided by a lag plot) calls for it.
-After putting all of these building blocks together, we arrive at our final model: the CarHMM-DFT.
-This model effectively decodes both dive types and subdive states of a killer whale while also taking into account the interaction between these two levels.
+The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when including $\Ztwo$ in the model, and our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale.
+[MAYBE ADD SOMETHING ABOUT THE CONFIRMATION OF THIS IN THE SIMULATION STUDY.] For simpler transformations that do not directly account for periodicity but for which me may expect temporal correlation (e.g., moving average), we suggest using a CarHMM instead of a traditional HMM when expert opinion and careful observation (aided by a lag plot) calls for it.
+Our simulation study showed how ignoring this auto-correlation can negatively bias standard error estimates.
 
-This solves an important problem, as the energetic requirements of killer whales together with their energetic intake from predation is directly related to their survival \citep{Williams:2009,Noren:2011}.
-Previous studies have looked into the energy consumption of killer whales, but many rely on invasive and expensive methodology \citep{Williams:2009,Noren:2011}.
-Accelerometers are an appealing alternative, but require decoding both the behavioral state \citep{Dot:2016} and the dive type \citep{Hastie:2006} of the animal in order to effectively estimate energy expenditure.
-The CarHHMM-DFT solves this problem by decoding both behaviors simultaneously at different scales.
-It also describes the distribution of acceleration within each hidden state.
-All of this information meaningfully effects the energy expenditure of killer whales and is therefore valuable information for conservation efforts \citep{Williams:2009,Noren:2011}.
+Accelerometers and time depth recorders are powerful tools to understand the diving behavior and energetic requirements of NRKWs, information essential to understand their foraging behaviour and survival \citep{Williams:2009,Noren:2011}.
+But this biologging data require decoding the behavioral state \citep{Dot:2016} or dive type \citep{Hastie:2006} of the animal in order to effectively understand estimate energy expenditure.
+The CarHHMM-DFT model we developed solves this problem by characterising dives into different types.
+It also describes the distribution of acceleration within each hidden state, providing a deeper understanding of the tri-axial movement, and thus fine-scale energy expenditure, of the animal.
+[ADD a ctation that linked up tri axial movement to fluking and energy expenditure.
+Check maybe some of the papers Courtney sent.] [Also add a sentence that this model is likely applicable to the many diving animals such as ..., maybe cite Adams, Vianey or anything that has both dive and accelerometer data.]
 
-While we combine the HMM models of Section \ref{sec:models} together effectively for our case study, in general this practice should be done with care as there are drawbacks to using exclusively HMMs to build these hierarchical models.
+While we combine the HMMs of Section \ref{sec:models} together effectively for our case study, in general this practice should be done with care as there are drawbacks to using exclusively HMMs to build these hierarchical models.
 For example, HMMs require observations to be taken at regular intervals.
-Continuous-time methods such as the state-switching OU process \citep{Michelot:2019} are based on stochastic differential equations and resolve this issue, but we do not describe them in detail here.
-This is because most continuous-time models require relatively slow MCMC algorithms to perform inference and are prohibitively difficult to fit.
-In addition, we show in Theorem 1 (see appendix) that the state-switching OU process of \citet{Michelot:2019} is equivalent to an CarHMM under certain conditions.
-This implies that a CarHMM is interpretable like a continuous-time model, but as easy to fit as a traditional HMM.
+Continuous-time methods such as the state-switching OU process \citep{Michelot:2019} are based on stochastic differential equations and resolve this issue.
+However, most continuous-time models require relatively slow MCMC algorithms to perform inference and are prohibitively difficult to fit.
+We show in Theorem 1 (see appendix) that the state-switching OU process of \citet{Michelot:2019} is equivalent to an CarHMM under certain conditions.
+This equivalence implies that a CarHMM is interpretable as a continuous-time model, but has the added advantage of relative computational efficiency.
 
-Another consideration when using HMMs to build complex models is the importance of effectively fit the data in question while avoiding \textit{over}-fitting or computationally expensive parameter estimation.
+When using HMMs to build complex models, one must consider how to effectively fit the data while avoiding \textit{over}-fitting or computationally expensive parameter estimation.
 One way to temper model complexity is to share fine-scale state parameters across the coarse-scale states.
-Even still, model complexity inevitably grows rapidly as hierarchical structures are stacked on top of each other.
+Regardless, model complexity inevitably grows rapidly as hierarchical structures are stacked on top of each other.
 For example, by far the fastest model to train in the simulation study is the CarHMM-DFT, as it has no hierarchical component and is the simplest of the four candidate models.
 The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioral states.
-However, this model is not sufficient if ecologists wish to understand to joint relationship between the dive type and intra-dive behaviour.
+However, this model is not sufficient if ecologists wish to understand the joint relationship between the dive type and intra-dive behaviour.
 
@@
 
-Despite these drawbacks, this work provides a flexible framework to model functional time-series data using HMM models.
-When model complexity is kept under control, a hierarchical model constructed using this processes can be both highly flexible and easy to fit using maximum likelihood methods.
-We demonstrate the usefulness of such a structure using an ecological example to determine the energetic requirements of a killer whale.
-It is our hope that researchers adapt this methodology to solve problems and perform inference in other settings which involve temporally dependent state-switching processes.
+Our work provides a flexible framework to model functional time-series data using HMMs.
+When model complexity is kept under control, a hierarchical model constructed using the main building blocks proposed can be both highly flexible and easy to fit using maximum likelihood methods.
+We demonstrate the usefulness of such a structure using an ecological example, where we used a complex HMM to classify the coarse and fine scale diving behavior of a killer whale.
+Such model will be applicable to similar datasets.
+In addition, as temporally dependent state-switching processes are common (IF YOU HAD CITATIONS HERE YOU I'D be AMZING-not necessary, but would seal the deal.), we believe that researchers can adapt this methodology to perform inference on a wide range of time-series data in other fields.

### 2020-08-06 Report/Sections/intro.tex
Update Report

@@
 Tracking technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
-The collection and analysis of data from devices such as accelerometers has brought new understanding to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}.
-In particular, the study of animal movement has been transformed by tracking devices that collects biologging data (CITATION).
-Tags attached to animals can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
-These data provide biologists with a wealth of knowledge about animal behaviour, but parsing this vast amount of information is a challenge for statisticians.
+The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}.
+In particular, the study of animal movement has been transformed by tracking devices that collect biologging data \citep{Borger:2020}.
+Tags can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
+These data contain a wealth of knowledge about human and animal behaviour, but parsing this vast amount of information is a challenge for statisticians.
 
@@
 
-Movement data are sometimes modelled as a set of curves that can be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}).
+Biologging data are sometimes modelled as a set of curves that can be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}).
 In \cite{Morris:2007}, a child's activity level throughout the day is viewed as a curve where metabolic activity is a function of time.
-The dive profile of marine mammals can be viewed as a curve, and Fu (CITATION) showed that [methods used by Eric and Nancy] is useful to classify [blah blah blah].
-However, now many such animals are tracked with multiple devices providing information at different scales.
-For example, in our case study, a single dive of a killer whale (\textit{Orcinus orca}) can be viewed as having several curves where acceleration and depth are functions of time.
+The dive profile of a marine mammal can also be viewed as a curve, and \citet{Fu:2017} showed that estimating the amplitude and phase variation of elephant seal curves is useful to classify dive types, although they ignore sequential dependence between dives.
+%However, now many such animals are tracked with high-frequency tags over a period of several dives, providing information at different time scales. 
+In our case study, the dive profile of a killer whale (\textit{Orcinus orca}) can be viewed as a sequence of curves where three orthogonal axes of acceleration are each functions of time.
 
 FDA was originally developed for curves that can be considered independent replicates, but the structure of time-series data is often complicated by additional dependencies.
-In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) (CITATION would be good - Fu's discussion?
-Our Vianey's paper or Timo Adam's paper?).
+In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017,Adam:2019}.
 Our case study data clearly demonstrate this common additional dependencies between curves.
 The dive profile (Figure \ref{fig:data}A) of a killer whale from the threatened Northern resident population off the coast of British Columbia, Canada (see Section \ref{sec:data} for more information about data collection and preprocessing) clearly shows temporal dependence \textit{between} dives, as dives of various time durations and depths appear to be clustered together in time.
-Previous studies have categorized these dives into different types, each associated with a different behavior, including resting, traveling, or foraging \citep{Tennessen:2019b}.
+Previous studies have categorized these dives into different types where each dive type is associated with a different behaviour.
+These behaviours can include resting, travelling, and foraging \citep{Tennessen:2019b}.
 Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence \textit{within} dives, as the accelerometer data within each dive indicate bouts of short-term periodicity.
-The combination of apparent behavioral changes combined with periodic behavior makes this accelerometer data difficult to model using simple stochastic processes.
-[Here if you can say something about how common this combined periodicity and behavioral dependece is in other accelerometer data, e.g.
-in human with a citation, you'd really effectively broaden the generality of the problem.]
+The combination of apparent behavioural changes combined with periodic behaviour makes this accelerometer data difficult to model using simple stochastic processes.
+Another example of behaviourally-dependent periodicity is in the field of machine health, where \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks.
+The raw acceleration data exhibit extreme periodicity (vibration) that depends upon the behaviour of the train as the load and speed of the train change.
 

### 2020-08-06 Report/Sections/simulation.tex
Update Report

@@
 To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT.
-The parameters used to generate the data are loosely based on those learned from the case study in Section \ref{sec:case_study}.
-The ground-truth parameters differ slightly from those learned in the case study primarily for simplicity of presentation.
+The parameters used to generate the data are loosely based on those estimated in the case study in Section \ref{sec:case_study} (see table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made to simplify presentation.
 For example, we set $\Gamma$ such that the simulated data has approximately equal numbers of short and long dives.
 This allows meaningful conclusions to be made for both dive types.
-In addition, the number subdive states is set to $N^*=2$ and $Z^{*(2)}$ is set as a scalar for additional simplicity.
+In addition, the number subdive states is set to $N^*=2$ and $\Ztwo$ is set as a scalar for additional simplicity.
 Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times.
-We also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.
-This allows us to judge how accurate each model is when estimating its own uncertainty.
+To assess how accurate each model is when estimating its own uncertainty, we also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.
 
@@
 After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model.
-The last $Y_t - 2\lfloor Y_t/2 \rfloor$ seconds of the dive are discarded.
 Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
@@
 \begin{gather*}
-    \mu_1^{*(1)} = 0.0 s, \enspace \sigma_1^{*(1)} = 0.05s, \enspace \phi_1^{*(1)} = 0.99, \\
+    \mu_A^{*(1)} = 0.0 s, \enspace \sigma_A^{*(1)} = 0.05s, \enspace \phi_A^{*(1)} = 0.99, \\
     %
-    \mu_1^{*(2)} = 0.0 s, \enspace \sigma_1^{*(2)} = 0.1s, \enspace \phi_1^{*(2)} = 0.95, \\
+    \mu_A^{*(2)} = 0.0 s, \enspace \sigma_A^{*(2)} = 0.1s, \enspace \phi_A^{*(2)} = 0.95, \\
     %
-    \mu_2^{*(1)} = 10.10, \quad \sigma_2^{*(1)} = 3.18, \\
+    \mu_W^{*(1)} = 10.10, \quad \sigma_W^{*(1)} = 3.18, \\
     %
-    \mu_2^{*(2)} = 305.94, \quad \sigma_2^{*(2)} = 17.49.
+    \mu_W^{*(2)} = 305.94, \quad \sigma_W^{*(2)} = 17.49.
 \end{gather*}
 %
-Recall that $Z^{*(1)}$ follows a CarHMM and is Normally distributed while $Z^{*(2)}$ follows an HMM and is Gamma distributed.
+Recall that $\Zone$ follows a CarHMM and is Normally distributed while $\Ztwo$ follows an HMM and is Gamma distributed.
 It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix.
-Figure \ref{fig:sim_data} shows this reconstructed sequence along with $Z^*$ for five dives of one simulated data set.
+Figure \ref{fig:sim_data} shows this reconstructed sequence along with one realization of $Z^*$ for five dives of one simulated data set.
 
 The two dive types differ in that dives of type 1 are much shorter on average than dives of type 2.
-The two subdive types differ primarily due to $\mu_2^*$ and $\sigma_2^*$, both of which are much higher in subdive state 2.
-This corresponds to much more vigorous and variable periodic behavior in the acceleration data for subdive state 2.
+The two subdive types differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher in subdive state 2 than subdive state 1.
+This corresponds to much more vigorous and variable periodic behavior in the acceleration data.
+
+All four candidate models (CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT) are fit to all 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
 
@@
 
-All four candidate models (CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT) are fit to all 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
-
-As expected, the full CarHHMM-DFT is the best-performing model.
+As expected, since it is the generating model, the full CarHHMM-DFT is the best-performing model.
 Using the forward-backward algorithm, the CarHHMM-DFT decodes the correct subdive type given the fine-scale data with an average confidence of $1 - 10^{-12}$ across all simulations.
@@
 The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias.
-However, the HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_1^*$.
-For both subdive types, $\hat \sigma_1^*$ greatly overestimates the true standard deviation.
-The estimated standard errors of $\hat \mu_1^*$ and $\hat \sigma_1^*$ using the observed Fisher information are both much smaller than the empirical standard error (see Section S-1.2, Table 3).
+However, the HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_A^*$.
+For both subdive types, $\hat \sigma_A^*$ greatly overestimates the true standard deviation.
+The estimated standard errors of $\hat \mu_A^*$ and $\hat \sigma_A^*$ using the observed Fisher information are both much smaller than the empirical standard error (see Section S-1.2, Table 3).
 This implies that variance can be overestimated and parameter estimates can be too confident when auto-correlation is ignored.

### 2020-08-06 Code/HHMM.py
Fix typos in code

@@ -547,7 +547,7 @@ class HHMM:
                         x0 = self.theta[0][feature][param][k0]
 
                         # optimize
-                        res = minimize(loss_fn, x0, method='Nelder–Mead', options=options)
+                        res = minimize(loss_fn, x0, method='Nelder-Mead', options=options)
 
                         # update final values
                         x = np.copy(res['x'])

### 2020-08-06 Code/Visualisor.py

@@ -76,6 +76,10 @@ class Visualisor:
                 xlim = [min(x),max(x)]
                 ylim = [min(y),max(y)]
 
+            # print autocorrelation
+            print(feature)
+            print(np.corrcoef(x,y))
+
             # KDE of lag plot
             plt.subplot(nrows,ncols,fig_num)
             kernel = gaussian_kde([x,y])

### 2020-08-06 Report/Sections/models.tex

@@
 The coarse-scale observations contain summaries of each curve such as curve length or maximum value.
-Curve $t$ has a total of $T^*_t$ within-curve (or fine-scale) associated observations: $Y^*_t \equiv \Big(Y^*_{t,1}, \ldots Y^*_{t,t^*}, \ldots, Y^*_{t,T^*_t}\Big)$.
-The fine-scale observations $Y^*_t$ make up a vector of function values from curve $t$, and we index $Y^*_t$ using $t^*$.
-The vector of all fine-scale observations is denoted at $Y^* \equiv \Big(Y^*_1,\ldots,Y^*_T\Big)$.
+In addition, consider an associated sequence of within-curve (or fine-scale) observations $Y^* \equiv \Big(Y^*_1,\ldots,Y^*_T \Big)$, where $Y^*_{t} \equiv \Big(Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\Big)$ is a vector of fine-scale observations corresponding to curve $t$.
+The vector $Y^*_{t}$ is made up of function values from curve $t$, and we index each observation of $Y^*_t$ using $t^*$.
 
-Hidden Markov models are useful tools to describe state-switching Markovian processes in discrete time and will be the core structure of the models we use to describe $Y$ and $Y^*$.
+Hidden Markov models can describe state-switching Markovian processes in discrete time and will be the core structure of the models we use to describe $Y$ and $Y^*$.
 %Traditional HMMs model process auto-correlation by assuming that the underlying behavior of a processes follows a single Markov chain.
@@
 %To account for each of these common additional dependence structures,
-We first describe the structure of a traditional HMM and a variation introduced by \cite{Lawler:2019} called the the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
+We first describe the structure of a traditional HMM.
+We then describe a variation introduced by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
 Either of these models can be used to model $Y$ or $Y^*_t$ for a particular curve.
-To simultaneously model both $Y$ and $Y^*$, we describe the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}, which simultaneously models $Y$ and $Y^*$ as two distinct, but related, HMM.
-Next, we describe how to reduce fine-scale correlation by transforming observations using subject area expertise.
-In particular, we use the DFT on a moving window over $Y^*$ to summarize its periodic behavior.
+To simultaneously model both $Y$ and $Y^*$, we describe the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}, which use two distinct, but related, HMMs to jointly model these time series.
+Next, we describe how to reduce fine-scale correlation by transforming raw observations to meaningful summaries.
+In particular, we use the discrete Fourier transform (DFT) on a moving window over $Y^*$ to summarize its periodic behavior.
 Finally, we show how each of these models can be altered and combined to form a wide variety of new models.
@@
 
-An HMM is comprised of a sequence of unobserved states $X_t$ $(t = 1, \ldots, T)$ which are associated with the observations $Y_t$.
+An HMM is comprised of a sequence of unobserved states $X = \big(X_1, \ldots, X_T\big)$ where $X_t$ is associated with the observation $Y_t$.
 Here we focus on the coarse-scale observations $Y$, but an HMM can also be used as a sub-model for a particular fine-scale process $Y_t^*$.
-In the context of an HMM, the $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time.
+The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time.
 The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$.
@@
 where $\eta \in \bbR^{N \times N}$ and $\eta_{ii}$ is set to zero for identifiability.
-This simplifies likelihood maximization and $\calL_{\text{HMM}}(y;\Theta,\Gamma)$ can be maximized using a wide range of optimizers.
+This simplifies likelihood maximization by removing constraints in the optimization problem.
 For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$.
@@
 
-The CarHMM, introduced by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence.
+The CarHMM, introduced by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence not accounted by the HMM state.
 Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
@@
 Figure \ref{fig:models} shows a graphical representation of the dependence structure of a CarHMM.
-The likelihood for the CarHMM can be easily calculated using the forward algorithm.
-As previously, let $y$ be the vector of observed emissions.
-Then
-\begin{equation}
+
+The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
+\begin{equation*}
     \calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
     \label{CarHMM_likelihood}
-\end{equation}
+\end{equation*}
 where
@@
 
-An HHMM accounts for different levels of correlation by modelling both the coarse-scale process and fine-scale process with an HMM \citep{Barajas:2017,Adam:2019}.
+An HHMM accounts for different levels of correlation by modelling both the coarse-scale process and fine-scale process with distinct HMMs \citep{Barajas:2017,Adam:2019}.
 The coarse-scale process is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations.
 %
-In the hierarchical setting, each state $X_t$ also emits yet another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$ and a sequence of fine-scale observations, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
-The fine-scale process ($X_t^*, Y_t^*$) then makes up another HMM.
-Given the coarse-scale hidden states, $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are all independent of one another.
-The parameters governing each fine-scale HMM $(X_t^*,Y_t^*)$ depend on the value of $X_t$.
+In the hierarchical setting, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$.
+The fine-scale state sequence in turn emits a sequence of fine-scale observations, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
+This fine-scale process ($X_t^*, Y_t^*$) then makes up another HMM whose parameters depend on the value of $X_t$.
 Specifically, if $X_t=i$, then the distribution of $X_t^*$ is characterized by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
@@
 We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\Theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^*)}\right)$.
+Given the coarse-scale hidden states, $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are all assumed to be independent of one another.
 
-Depending upon the process being modeled, it is possible to force certain parameters to be shared to reduce the complexity of the HHMM.
-For example, in the killer whale case study (Section \ref{sec:case_study}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( i.e.
-\theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
+Depending upon the process being modelled, it is possible to force certain parameters to be shared to reduce the complexity of the HHMM.
+For example, in the killer whale case study (Section \ref{sec:case_study}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( \theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
 
@@
 Let $y$ be the $T$-vector of the observed coarse-scale emissions and
-$y^*$ be the $(T_1^* + \cdots + T_T^*)$-vector of the observed fine-scale emissions.
+$y^* \equiv \left(y^*_1, \ldots,y^*_T\right)$ be the collection of $T$ observed fine-scale emission vectors.
 %
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Transforming $Y_t^*$ to account for fine-scale behaviour}
+\subsection{Transforming observations to account for fine-scale behaviour}
 \label{subsec:STFT}
@@
 For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behavior.
-To deal with this issue, we recommend replacing the fine-scale process $Y_t^*$ with relevant statistics that summarize this non-Markovian behavior.
-Summary statistics can be calculated from a moving window over $Y_t^*$ in order to maintain the temporal structure of the fine-scale process.
+To handle these additional fine-scale structures, we recommend replacing the fine-scale observations $Y_t^*$ with relevant statistics that summarize this non-Markovian behavior.
+To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over $Y_t^*$.
 Subject-area experts are often required to determine the optimal window size and the specific summary statistics used.
@@
 %
-For our particular application, we transform the fine-scale process using the discrete Fourier transform (DFT) of a moving window of width $h$ across $Y^*$:
+
+For our case study, we use the discrete Fourier transform (DFT) of a moving window of width $h$ across $Y^*$:
 %
@@
 %
-for $k = 0, 1, \ldots, h-1$. Summary statistics are then used to drastically reduce the dimension of the DFT:
+for $k = 0, 1, \ldots, h-1$. Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
 %
@@
     \label{eqn:z}
-    Z_{t,t^*}^{*(1)} := \text{Re}\left(DFT\{Y^*_{t,t^*}\}(0)\right) \qquad Z_{t,t^*}^{*(2)} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|^2
+    \Zone_{t,t^*} := \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n} \quad \text{and} \quad \Ztwo_{t,t^*} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|^2.
 \end{equation}
 %
-In words, $Z_{t,t^*}^{*(1)}$ is the average value of $\{Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\}$ and $Z_{t,t^*}^{*(2)}$ is the squared 2-norm of the component of $\{Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\}$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window (by the Plancherel theorem).
-The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with care.
-These summary statistics are just one possible choice to describe each window; other choices include the dominant frequency and amplitude of $Y^*_{t,t^*}$.
-Figure \ref{fig:models} represents the process of transforming $Y^*_{t,t^*}$ into $Z^*_{t,t^*}$.
-To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $Z^*_t$ such that $Z^*_t = \left(Z^*_{t,0},Z^*_{t,h},Z^*_{t,2h},\ldots\right)$.
-This both removes residual correlation and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
+In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared 2-norm of the component of $\{Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\}$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window (by the Plancherel theorem).
+We use $\Ztwo$ to capture the ``wiggliness" of the acceleration within a window.
+The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert.
+Other choices for window summary statistics are possible, including the dominant frequency and amplitude.
+Figure \ref{fig:models} represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM.
+To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \left(\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\right)$.
+The down sampling removes the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
+By summarizing the data, the method overall also reduces correlation in the raw observations that would due to the complex dependencies structure.
 
-Once $Z^*_t$ is calculated, it simply replaces the original fine-scale process $Y^*_t$ in the HHMM structure.
-As such, the emission distributions and the likelihood remain identical to the original HHMM when transforming $Y^*$ to $Z^*$.
-In order to clearly differentiate models, we refer to an HHMM with $Z^*$ as fine-scale observations as an HHMM-DFT.
+Once $\Z_t$ is calculated, it simply replaces the original fine-scale observations $Y^*_t$ in the HHMM structure.
+As such, the emission distributions and the likelihood remain identical to the original HHMM when transforming $Y^*$ to $\Z$.
+In order to clearly differentiate models, we refer to an HHMM with $\Z$ as fine-scale observations as an HHMM-DFT.
 
@@
 By definition, an HHMM treats both the coarse-scale and the fine-scale processes as a traditional HMM.
-However, the coarse-scale HMM can be generalized by replacing it with a CarHMM, yielding the following likelihood:
+However, it is straightforward to generalize the coarse-scale HMM by replacing it with a CarHMM, yielding the following likelihood:
 \[

### 2020-08-06 Report/Sections/results.tex

@@
 We first report the results from the full CarHHMM-DFT (defined in Section \ref{sec:data}) in detail and then asses the quality of the fit.
-Finally, we compare the results of the CarHHMM-DFT with those of the other candidate models.
+We also compare the results of the CarHHMM-DFT with those of the other candidate models.
 
@@
 
-The three subdive behaviours have distinct average ``wiggliness" values, as the means of $Z^{*(2)}$ for each state are separated by an order of magnitude (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}).
-Subdive state 1 has the smallest mean of $Z^{*(2)}$ and the smallest variance corresponding to $Z^{*(1)}$.
-It also has the highest auto-correlation in $Z^{*(1)}$.
+The three subdive behaviours have distinct average ``wiggliness" values, as the means of $\Ztwo$ for each state are separated by an order of magnitude (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}).
+Subdive state 1 has the smallest mean of $\Ztwo$ and the smallest variance corresponding to $\Zone$.
+It also has the highest auto-correlation in $\Zone$.
 This implies less overall activity and more consistent acceleration compared to the other subdive states.
-Subdive state 2 has a mean ``wiggliness" ($Z^{*(2)}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
+Subdive state 2 has a mean ``wiggliness" ($\Ztwo$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
 The auto-correlation of acceleration is also slightly lower in than subdive state 1.
 We therefore hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
-Finally, the mean of $Z^{*(2)}$ and variance of $Z^{*(1)}$ in subdive state 3 is much higher than the other two states.
-The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower, implying more variation in acceleration between segments.
+Finally, the mean of $\Ztwo$ and variance of $\Zone$ in subdive state 3 is much higher than the other two states.
+The auto-correlation of $\Zone$ is also much lower, implying more variation in acceleration between segments.
 This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).
@@
 Histograms of the pseudoresiduals mostly support that this model is well-specified.
-One exception is $Z^{*(2)}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
-This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution than the gamma distribution used in the case study.
+One exception is $\Ztwo$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
+This implies that the true distribution of $\Ztwo$ may follow a heavier-tailed distribution than the gamma distribution used in the case study.
 
@@
 Observations are weighted by the decoded probability that they correspond to a particular hidden state, and this histogram is plotted over the emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}).
-Our results again mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed.
-Another exception is $Z^{*(1)}$, which has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
+Our results again mostly support a well-specified model with the exception of $\Ztwo$, which is again right-skewed.
+Another exception is $\Zone$, which has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
 These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.
@@
 The HHMM-DFT decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3.
-In addition, the HHMM-DFT produces estimates of $\sigma_1^{*(1)}$, $\sigma_1^{*(2)}$, and $\sigma_1^{*(3)}$ (the standard deviation of acceleration) which are approximately 50-100 \% larger than those of the CarHHMM-DFT across all axes of acceleration.
-The estimated uncertainties of the estimates of $\mu_1^{*(1)}$, $\mu_1^{*(2)}$, and $\mu_1^{*(3)}$ are also less than half of those for the CarHHMM-DFT across all axes of acceleration.
+In addition, the HHMM-DFT produces estimates of $\sigma_A^{*(1)}$, $\sigma_A^{*(2)}$, and $\sigma_A^{*(3)}$ (the standard deviation of acceleration) which are approximately 50-100 \% larger than those of the CarHHMM-DFT across all axes of acceleration.
+The estimated uncertainties of the estimates of $\mu_A^{*(1)}$, $\mu_A^{*(2)}$, and $\mu_A^{*(3)}$ are also less than half of those for the CarHHMM-DFT across all axes of acceleration.
 These findings are consistent with the tendency to overestimate variance and underestimate standard error when ignoring auto-correlation.
-Further, the empirical auto-correlation of $Z^{*(1)}$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers.
+Further, the empirical auto-correlation of $\Zone$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers.
 Both of these findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.

### 2020-08-06 Report/main.tex
Updates from Overleaf
Update Report
Stylistic changes to report

@@
 % Title, authors, affiliations
-\title[]{Modelling fine-scale correlation structures using hidden Markov models}%\query{Q1}
+\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
 \author{BlindedA\authorref{1}\thanksref{*}}
@@
 
-Recent advances in high-frequency tagging
-technology have made a vast
-amount of movement data available in a variety of fields. This data can exhibit simultaneous behavioural processes occurring at different
-time scales, resulting in increasingly complicated dependence structures. These processes can be modelled 
-through a hierarchical hidden Markov model
-(HHMM), which models the system as a
-nested structure of hidden Markov models (HMMs).
-At very short time scales, however, many basic assumptions of traditional HMMs are violated. We demonstrate how
-to incorporate fine-scale processes into the
-larger structure of HHMMs while maintaining computational efficiency. We apply our
-method to dive and accelerometer data collected from a northern resident killer whale off
-the coast of British Columbia, Canada.
+Advances in tagging technology have made a vast amount of high-frequency time series data available in several fields.
+These data often exhibit complicated temporal dependence structure which cannot be modelled using common methods of Functional Data Analysis (FDA).
+The field of animal movement suggests a hierarchical approach using hidden Markov models (HMMs), but many intricate fine-scale processes violate basic assumptions of HMMs.
+We detail a generalization of the hierarchical HMM approach that incorporates a wider class of parametric models at problematic fine scales. 
+%We also provide several examples of viable fine-scale models.
+As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada. 
+Our model outputs more interpretable results as well as more accurate parameter and standard error estimates compared to existing models.
 
@@
 \section{Models and parameter estimation}
+\label{sec:models}
 \input Sections/models.tex
 %
+\section{Data Collection and Model Construction}
+\label{sec:data}
+\input Sections/data.tex
+%
 \section{Simulation Study}
+\label{sec:sim_study}
 \input Sections/simulation.tex
 %
-\section{Killer Whale Case Study}
+\section{Case Study Results}
 \label{sec:case_study}
 \input Sections/results.tex
-%
+%%
 \section{Discussion}

### 2020-08-14 Report/Sections/discussion.tex
Updates from Overleaf

@@
 
-Current FDA literature addresses sequences of curves either with multilevel models with random effects and no temporal dependence \citep{chen:2012,di:2009}, or with functional time series which model the curves as an evolving process and overlook the possibility of a discrete number of curve ``types" \citep{Kokoszka:2018}.
-However, many real-world curve sequences exhibit both temporal dependence and a discrete number of types.
-One obvious example is the case study examined here, where a sequence of marine mammal dives serve a number of discrete of purposes such as traveling, socializing, resting, or foraging \citep{Tennessen:2019a}.
-Another example is accelerometer data in school children, where bouts of activity can be associated with running, walking, etc.
+Current FDA literature addresses dependence between curves either with multilevel models with random effects and no time component \citep{Chen:2012,Di:2009}, or with functional time series which model the curves as an evolving process \citep{Kokoszka:2018} and overlook the possibility that curves have several distinct ``types".
+However, many curve sequences exhibit both temporal dependence and a discrete number of types.
+One obvious example is the case study examined here, where one marine mammal dive serves a purpose such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}.
+Another example is accelerometer data in school children, where bouts of activity can be associated with a behaviour such as running or walking.
 \citep{Morris:2007}.
 
-In this work, we deal with both temporal dependence between curves and a discrete number of curve types by using either an HMM or CarHMM, depending upon the level auto-correlation between curves.
-To help select the appropriate model, lag plots can be used to test if there is significant auto-correlation between curves.
-One may use summary statistics of the curve as the observations of the HMM, but it is also possible to use another within-curve parametric model as the HMM emissions.
-The parameters of this fine-scale model then in turn depend upon the curve type.
+In this work, we deal with both temporal dependence between curves and distinct curve ``types" by using either an HMM or CarHMM.
+To help select the appropriate model between these two, lag plots can be used to test if there is significant auto-correlation between curves.
+A traditional HMM approach requires that practitioners take summary statistics of curves as emissions.
+However, HMM emissions can also be the entire curve, which is then modelled with a fine-scale parametric model.
+The parameters of this fine-scale model then depend upon the curve type.
+There are a wide range of possible fine-scale models, including another HMM, CarHMM, a spline-based approach similar to that of \citet{Langrock:2018}, and many more.
 Together, the two levels make up a hierarchical model which can account for processes taking place simultaneously at different time-scales.
@@
 
-Several possible choices exist for the fine-scale model, including yet another HMM.
-\citet{Barajas:2017} introduce this particular situation as an HHMM, and \citet{Adam:2019} use it to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California.
-Traditional HMMs prove insufficient at these very fine-scales, as \citet{Adam:2019} run into clear violations of conditional independence and observe periodic behavior in the acceleration data.
+When the fine-scale model is  an HMM, the resulting overall model is the HHMM as introduced by \citet{Barajas:2017}.
+\citet{Adam:2019} in particular use the HHMM to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California.
+This model can be inappropriate at very fine scales, as \citet{Adam:2019} find clear violations of conditional independence given hidden states and also observe periodic behaviour in the acceleration data.
 This indicates the need to account for complicated dependence structures within fine-scale processes, especially those with high sampling frequencies.
@@
 
-To deal with these problematic fine-scale processes, we detail a moving-window transformation over the fine-scale observations.
+To deal with these problematic fine-scale phenomena, we explore a moving-window transformation over the fine-scale observations.
 For periodic behaviour in particular, we use a DFT-based transformation.
-The effectiveness of this approach is evidenced by both our simulation study, where the subdive decoding accuracy is nearly perfect when including $\Ztwo$ in the model, and our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale.
-[MAYBE ADD SOMETHING ABOUT THE CONFIRMATION OF THIS IN THE SIMULATION STUDY.] For simpler transformations that do not directly account for periodicity but for which me may expect temporal correlation (e.g., moving average), we suggest using a CarHMM instead of a traditional HMM when expert opinion and careful observation (aided by a lag plot) calls for it.
-Our simulation study showed how ignoring this auto-correlation can negatively bias standard error estimates.
+The effectiveness of this approach is evidenced by both our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale, and our simulation study, where the subdive decoding accuracy is by far the worst when omitting $\Ztwo$ (``wiggliness'') from the model.
+For simpler transformations, e.g.
+an average such as $\Zone$, either subject area experts or exploratory plots may suggest temporal correlation.
+In this case,  we suggest using a CarHMM instead of a traditional HMM.
+Our simulation study shows how ignoring this auto-correlation can result in underestimating standard errors and overestimating the variance of the observations.
 
-Accelerometers and time depth recorders are powerful tools to understand the diving behavior and energetic requirements of NRKWs, information essential to understand their foraging behaviour and survival \citep{Williams:2009,Noren:2011}.
-But this biologging data require decoding the behavioral state \citep{Dot:2016} or dive type \citep{Hastie:2006} of the animal in order to effectively understand estimate energy expenditure.
-The CarHHMM-DFT model we developed solves this problem by characterising dives into different types.
-It also describes the distribution of acceleration within each hidden state, providing a deeper understanding of the tri-axial movement, and thus fine-scale energy expenditure, of the animal.
-[ADD a ctation that linked up tri axial movement to fluking and energy expenditure.
-Check maybe some of the papers Courtney sent.] [Also add a sentence that this model is likely applicable to the many diving animals such as ..., maybe cite Adams, Vianey or anything that has both dive and accelerometer data.]
+We combine generalizations of HMMs with hierarchical structures and moving-window transformations to model animal behaviour as recorded from accelerometers.
+Time depth recorders and accelerometers are powerful tools to understand the diving behaviour and energetic requirements of NRKWs, and this is essential to understand their foraging behaviour and survival \citep{Williams:2009,Noren:2011}.
+However, biologging data require classifying the underlying activity \citep{Dot:2016} or dive type \citep{Hastie:2006} of the animal in order to effectively understand and estimate energy expenditure.
+The CarHHMM-DFT separates dives into different types on the coarse scale while describing the distribution of acceleration within each subdive state on the fine scale.
+This gives a deeper understanding of the animal's tri-axial movement, and thus its fine-scale energy expenditure \citep{Gleiss:2011,Qasem:2012}.
+Although our case study focuses on Northern resident killer whales, this model is likely applicable to many dive animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.
 
-While we combine the HMMs of Section \ref{sec:models} together effectively for our case study, in general this practice should be done with care as there are drawbacks to using exclusively HMMs to build these hierarchical models.
+While we combine the HMMs of section \ref{sec:models} effectively for our case study, in general this practice should be done with care as there are drawbacks to using exclusively HMMs to build these hierarchical models.
 For example, HMMs require observations to be taken at regular intervals.
-Continuous-time methods such as the state-switching OU process \citep{Michelot:2019} are based on stochastic differential equations and resolve this issue.
-However, most continuous-time models require relatively slow MCMC algorithms to perform inference and are prohibitively difficult to fit.
-We show in Theorem 1 (see appendix) that the state-switching OU process of \citet{Michelot:2019} is equivalent to an CarHMM under certain conditions.
+Continuous-time methods such as the state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019} and the continuous-time HMM (CTHMM) \citep{Liu:2015} are often based on stochastic differential equations and resolve this issue.
+However most continuous-time models require relatively slow MCMC algorithms to perform inference and are prohibitively difficult to fit.
+We show the appendix that the state-switching OU process of \citet{Michelot:2019} is equivalent to a CarHMM under certain conditions.
 This equivalence implies that a CarHMM is interpretable as a continuous-time model, but has the added advantage of relative computational efficiency.
 
-When using HMMs to build complex models, one must consider how to effectively fit the data while avoiding \textit{over}-fitting or computationally expensive parameter estimation.
-One way to temper model complexity is to share fine-scale state parameters across the coarse-scale states.
-Regardless, model complexity inevitably grows rapidly as hierarchical structures are stacked on top of each other.
+When using HMMs to build complex models, one must consider how to effectively fit the data while avoiding over-fitting and at the same time ensuring computationally efficient parameter estimation.
+One way to temper model complexity is to assume that the fine-scale state parameters are the same for all coarse-scale states.
+Regardless, model complexity inevitably grows rapidly as more layers are added to the hierarchical structure.
 For example, by far the fastest model to train in the simulation study is the CarHMM-DFT, as it has no hierarchical component and is the simplest of the four candidate models.
-The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioral states.
-However, this model is not sufficient if ecologists wish to understand the joint relationship between the dive type and intra-dive behaviour.
+The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioural states.
+However, this model does not allow  ecologists  to study the joint relationship between the dive type and intra-dive behaviour.
 
@@
 Our work provides a flexible framework to model functional time-series data using HMMs.
-When model complexity is kept under control, a hierarchical model constructed using the main building blocks proposed can be both highly flexible and easy to fit using maximum likelihood methods.
-We demonstrate the usefulness of such a structure using an ecological example, where we used a complex HMM to classify the coarse and fine scale diving behavior of a killer whale.
-Such model will be applicable to similar datasets.
-In addition, as temporally dependent state-switching processes are common (IF YOU HAD CITATIONS HERE YOU I'D be AMZING-not necessary, but would seal the deal.), we believe that researchers can adapt this methodology to perform inference on a wide range of time-series data in other fields.
+Provided the model is not overly complex, a hierarchical model constructed using the proposed building blocks can be both flexible and easy to fit using maximum likelihood methods.
+We demonstrate the usefulness of such a model using an ecological example, where we use a complex HMM to classify the coarse and fine scale diving behaviour of a killer whale.
+In addition, as complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology to the analysis of a wide range of time-series data in a variety of fields.

### 2020-08-14 Report/Sections/appendix.tex

@@
 
-\addcontentsline{toc}{section}{Appendices}
-\renewcommand*{\thesubsection}{\Alph{subsection}}
+\subsection{Equivalence of State-Switching OU process and CarHMM}
 
@@
 
-\begin{proof}{Detailed description of simulated data}{}
+\subsection{Detailed description of simulated data}
 
@@
 It follows that $\mu_W^{*(1)} = 10.10$, $\sigma_W^{*(1)} = 3.18$, $\mu_W^{*(2)} = 305.94$, and $\sigma_W^{*(2)} = 17.49$.
-\end{proof}
 
-\begin{proof}{Likelihood of Simulation study model}{}
+\subsection{Likelihood of Simulation study model}
 
@@
 %
-$$\calL_{\text{CarHHMM-DFT}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N.$$
+$$\calL_{\text{CarHHMM-DFT}}(\Theta,\Theta^*,\Gamma,\Gamma^*;y,z^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N.$$
 %
@@
 \begin{align*}
-P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
-&f^{(N)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big],
+P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(\Theta^{*(1)},\Gamma^{*(1)};z_t^*\right), \ldots , \\
+&f^{(N)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(\Theta^{*(N)},\Gamma^{*(N)};z_t^*\right) \Big],
 \end{align*}
@@
 %
-$$\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(i)},\Gamma^{*(i)}\right) = \delta^{*(i)} \prod_{t=2}^T \Gamma^{*(i)} P(z^*_{t,t^*}|z^*_{t,t^*-100};\Theta^{*(i^*)}) \mathbf{1}_N,$$
+$$\calL_{\text{CarHMM-DFT}}\left(\Theta^{*(i)},\Gamma^{*(i)};z_t^*\right) = \delta^{*(i)} \prod_{t=2}^T \Gamma^{*(i)} P(z^*_{t,t^*}|z^*_{t,t^*-100};\Theta^{*(i^*)}) \mathbf{1}_N,$$
 %
@@
 Note that $f^{*(\cdot,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$ is the emission distribution of $\Z_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.
-
-\end{proof}
 

### 2020-08-14 Report/Sections/intro.tex

@@
-Tracking technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
+Biologging technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
 The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}.
@@
 In \cite{Morris:2007}, a child's activity level throughout the day is viewed as a curve where metabolic activity is a function of time.
-The dive profile of a marine mammal can also be viewed as a curve, and \citet{Fu:2017} showed that estimating the amplitude and phase variation of elephant seal curves is useful to classify dive types, although they ignore sequential dependence between dives.
+The dive profile of a marine mammal can also be viewed as a curve.
+\citet{Fu:2017} show that estimating the amplitude and phase variation of elephant seal (\textit{Mirounga leonina}) dive curves is useful to classify dive types, although they ignore sequential dependence between dives.
 %However, now many such animals are tracked with high-frequency tags over a period of several dives, providing information at different time scales. 
-In our case study, the dive profile of a killer whale (\textit{Orcinus orca}) can be viewed as a sequence of curves where three orthogonal axes of acceleration are each functions of time.
+Similarly, our case study views the dive profile of a killer whale (\textit{Orcinus orca}) as a sequence of curves where three orthogonal axes of acceleration are each functions of time.
 
 FDA was originally developed for curves that can be considered independent replicates, but the structure of time-series data is often complicated by additional dependencies.
-In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017,Adam:2019}.
-Our case study data clearly demonstrate this common additional dependencies between curves.
-The dive profile (Figure \ref{fig:data}A) of a killer whale from the threatened Northern resident population off the coast of British Columbia, Canada (see Section \ref{sec:data} for more information about data collection and preprocessing) clearly shows temporal dependence \textit{between} dives, as dives of various time durations and depths appear to be clustered together in time.
-Previous studies have categorized these dives into different types where each dive type is associated with a different behaviour.
-These behaviours can include resting, travelling, and foraging \citep{Tennessen:2019b}.
+In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017,Adam:2019}, as is the case for our case study data.
+Namely, the dive profile (Figure \ref{fig:data}a) of a killer whale from the threatened Northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence \textit{between} dives, as dives of various time durations and depths appear to be clustered together in time.
+Previous studies have categorized these dives into different types where each dive type is associated with a different behaviour, including resting, travelling, and foraging \citep{Tennessen:2019b}.
 Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence \textit{within} dives, as the accelerometer data within each dive indicate bouts of short-term periodicity.
@@
 Another example of behaviourally-dependent periodicity is in the field of machine health, where \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks.
-The raw acceleration data exhibit extreme periodicity (vibration) that depends upon the behaviour of the train as the load and speed of the train change.
+The raw acceleration data exhibit extreme periodicity (in the form of vibration) that depends upon the behaviour of the train as the load and speed of the train change.
 
-The FDA literature provides models that account for some between-curve dependence structures, but these methods are not adequate for modelling the coarse-scale temporal structure seen here.
+FDA literature provides models that account for some between-curve dependence, but these methods are not adequate when this dependence is in time.
 For example, multilevel models incorporate dependence via random effects.
@@
 In a classic paper that introduced this multilevel structure,  \cite{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles.
-More recent work on multilevel models includes that of \cite{di:2009}, \cite{chen:2012} and Crainiceanu [REF].
-However, multilevel models are not appropriate for this killer whale data since they account for neither the temporal dependence in the dive sequence nor the different dive types.
+More recent work on multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012} and \citet{Crainiceanu:2009}.
+However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types.
 In addition to multilevel models, FDA researchers have used functional time series to model dependence in sequences of curves.
 Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}.
-While functional time series methods can in theory be applied to our dive data, they do not account for the discrete number of different dive types.
+While functional time series methods can in theory be applied to biologging data such as ours, they do not account for sequences of curves that switch between a discrete number of types over time.
 
-Traditional FDA techniques similarly fall short when modelling the within-curve acceleration data seen here.
-In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function (e.g., CITATION- an example with accelerometer data would be particularly good - just did a quick search and they may not be appropriate but could look at: https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0169576 or https://academic-oup-com.ezproxy.library.ubc.ca/biostatistics/article/16/2/352/266114).
-Alternatively, this structure is sometimes generated via random regression.
-Radically changing behavior that we see in the accelerometer data of similar to our dive sequence would be challenging to modelled with these classic FDA techniques.
+Traditional FDA techniques similarly fall short when modelling complicated within-curve acceleration data.
+In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function.
+\citet{Yao:2005} use this framework to construct a method of sparse longitudinal data analysis using functional principal component analysis.
+%In terms of applications, \citet{Gruen:2017} and \citet{Xiao:2014} investigate degenerative joint disease in cats and circadian rhythms in aging patients, respectively. 
+Alternatively, within-curve structure is sometimes generated via random regression.
+For example, \citet{Rice:2001} use random regression to construct non-parametric mixed effects models for unequally sampled noisy curves.
+However, radically changing behaviour on short time-scales similar to our dive sequence is challenging to model with these classic FDA techniques.
 
-In light of these challenges, we turn to vast field of methodology for the analysis of animal movement \citep{Hooten:2017,Mcclintock:2020}, where one of the most prevalent techniques in the literature of late is the hidden Markov model, or HMM \citep{Patterson:2017}.
+In light of these challenges, we turn to vast field of methodology for the analysis of animal movement \citep{Hooten:2017}, where one of the most prevalent techniques in the literature of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
 HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of the animal from movement observations.
-The ubiquitous applications of HMMs in ecology have seen little use in FDA, with a few notable exceptions.
-In particular,  \citet{Langrock:2018} did [blah blah blah], which does not deal with [blah blah blah].
-We show here how HHMs can provide a new tool to the FDA toolbox.
+While ubiquitous in the ecology literature, HMMs have seen little use in non-parametric functional modelling, with a few notable exceptions.
+In particular, \citet{Langrock:2018} take a non-parametric approach to model the distribution of HMM observations in terms of B-splines.
+This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments.
+However, this approach does not account for certain types of time correlation within or between the time series themselves.
+We focus on the time series themselves as functions here and show how HHMs can provide a new tool to the FDA toolbox.
 
@@
 
-Under the traditional HMM, observations are independent given the underlying hidden state process.
+Under a traditional HMM, observations are independent given the underlying hidden state process.
 However, this assumption is often violated in real-world processes, especially when observations are taken at high frequencies.
@@
 The CarHMM in particular explicitly models auto-correlation into an HMM while adding minimal complexity to the base structure of an HMM.
-Animal movement modelers have used hierarchical hidden Markov models (HHMMs) to model simultaneous behavioural processes that occur at different time scales \citep{Barajas:2017,Adam:2019}.
+Animal movement modellers have also used hierarchical hidden Markov models (HHMMs) to model simultaneous behavioural processes that occur at different time scales \citep{Barajas:2017,Adam:2019}.
 We borrow from ideas from both the CarHMM and the HHMM here.
 
-Traditional HMMs, CarHMMs, or HHMMs cannot easily capture the complicated structure that biologging data often exhibit at short time scales.
-For example, the HHMM of \citet{Adam:2019} did not capture the fine-scale periodic swimming pattern of  horn sharks (LATIN NAME).
-However, \citet{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM to account for periodic behavior in the context of describing daily behavioural cycles of marine mammals.
+Although they are useful in many situations, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated structures at short time scales.
+The HHMM of \citet{Adam:2019} fails to capture the fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}).
+\citet{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM to account for periodic behaviour in the context of describing daily behavioural cycles of marine mammals.
 In addition, Fourier analysis has previously been used on accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
-Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appears to be a promising approach to account for periodic structures in fine-scale data.
+Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for periodic structures in fine-scale accelerometer data.
 
-In this paper, we unify and build off of several existing methods from the ecology literature in novel ways to account for complex temporally dependent functional data.
-The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models and solve FDA problems.
+In this paper, we combine several existing methods from the ecology literature in novel ways to account for complex temporally dependent functional data.
+The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models and thus extend the methods available for the analysis of functional data.
 %This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. 
-%We describe a general procedure that can be used to extract features from highly structured fine-scale behaviours that otherwise could not be modeled with existing HMM models. 
+%We describe a general procedure that can be used to extract features from highly structured fine-scale behaviours that otherwise could not be modelled with existing HMM models. 
 %In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. 
 %We then perform a simulation study to compare the performance of each existing model with ours in a controlled setting.
-
-In this paper, we first describe HMMs in addition to two variants (CarHMMs and HHMMs) and discuss  how Fourier analysis can handle fine-scale dependence structures.
+We first describe HMMs as well as to two variants (CarHMMs and HHMMs) and discuss how Fourier analysis can handle fine-scale dependence structures.
 We also show how these methods can be combined to analyze increasingly complex data.
-In Section 3, we construct several candidate models to describe the data from Figure \ref{fig:data}.
-Section 4 details a simulation study based on these candidate models.
-In Section 5, we fit the candidate models to the original data, and in Section 6 we discuss the results.
+In section 3, we construct several candidate models and fit them to the data from Figure \ref{fig:data}.
+Section 4 details a simulation study based on these candidate models, and in section 5 we discuss our results.

### 2020-08-14 Report/Sections/models.tex

@@
 
-Consider a sequence of $T$ possibly high-dimensional curves (or dives, in our context) with associated curve-level (or coarse-scale) observations $Y \equiv \Big(Y_1, \ldots, Y_T\Big)$.
-The coarse-scale observations contain summaries of each curve such as curve length or maximum value.
-In addition, consider an associated sequence of within-curve (or fine-scale) observations $Y^* \equiv \Big(Y^*_1,\ldots,Y^*_T \Big)$, where $Y^*_{t} \equiv \Big(Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\Big)$ is a vector of fine-scale observations corresponding to curve $t$.
-The vector $Y^*_{t}$ is made up of function values from curve $t$, and we index each observation of $Y^*_t$ using $t^*$.
+Consider a sequence of $T$ curves (or dives, in our context). Any particular dive $t$, $t = 1,
+\ldots,T$, has an associated curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of within-curve (or fine-scale) observations $Y^*_{t} \equiv \Big(Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\Big)$.
+The vector $Y^*_{t}$ is made up of function values from curve $t$.
+We call the vectors of coarse-scale observations $Y \equiv \Big(Y_1, \ldots, Y_T\Big)$ and the collection of all fine-scale observations $Y^* \equiv \Big\{Y^*_1,\ldots,Y^*_T \Big\}$.
 
-Hidden Markov models can describe state-switching Markovian processes in discrete time and will be the core structure of the models we use to describe $Y$ and $Y^*$.
-%Traditional HMMs model process auto-correlation by assuming that the underlying behavior of a processes follows a single Markov chain.
+Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$.
+%Traditional HMMs model process auto-correlation by assuming that the underlying behaviour of a processes follows a single Markov chain.
 However, when conditioned on the hidden state sequence observations are assumed to be independent.
@@
 %To account for each of these common additional dependence structures,
-We first describe the structure of a traditional HMM.
-We then describe a variation introduced by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
-Either of these models can be used to model $Y$ or $Y^*_t$ for a particular curve.
-To simultaneously model both $Y$ and $Y^*$, we describe the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}, which use two distinct, but related, HMMs to jointly model these time series.
-Next, we describe how to reduce fine-scale correlation by transforming raw observations to meaningful summaries.
-In particular, we use the discrete Fourier transform (DFT) on a moving window over $Y^*$ to summarize its periodic behavior.
+We first detail the structure of a traditional HMM.
+We then review a variation introduced by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
+Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$.
+To simultaneously model $Y$ and $Y^*$, we use the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
+%which is comprised of two distinct, but related, HMMs that jointly model these time series. 
+Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics.
+In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour.
 Finally, we show how each of these models can be altered and combined to form a wide variety of new models.
@@
 An HMM is comprised of a sequence of unobserved states $X = \big(X_1, \ldots, X_T\big)$ where $X_t$ is associated with the observation $Y_t$.
-Here we focus on the coarse-scale observations $Y$, but an HMM can also be used as a sub-model for a particular fine-scale process $Y_t^*$.
+Here we focus on the coarse scale sequence of observations $Y_1,\ldots, Y_T$, but an HMM can also be used as a sub-model for the fine-scale sequence for a particular dive.
 The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time.
@@
 We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i^{th}$ component
-$\delta_i = \Pr\{X_1 = i\}.$
+$\delta_i = \Pr(X_1 = i).$
 A Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
@@
 These conditional distributions are governed by state-dependent parameters.
-If $X_t = i$, then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes simply $f^{(i)}(\cdot)$.
-If $Y_t$ is a vector, we assume that the dimensions of the observations are independent $\left(f^{(i)}(\cdot) = \prod_{d} f^{(i)}_d(\cdot)\right)$, but this need not be the case in general.
+Namely, if $X_t = i$ then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes simply $f^{(i)}(\cdot)$.
 %
-Figure \ref{fig:models} represents the dependence structure of an HMM.
+Figure \ref{fig:models}a represents the dependence structure of an HMM.
 
@@
 %
-$$\calL_{\text{HMM}}(y;\Theta,\Gamma) = \delta P(y_1;\Theta) \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N$$
+$$\calL_{\text{HMM}}(\Theta,\Gamma;y) = \delta P(y_1;\Theta) \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N$$
 %
@@
 %
-where $\eta \in \bbR^{N \times N}$ and $\eta_{ii}$ is set to zero for identifiability.
+where $i,j = 1,\ldots,N$ and $\eta_{ii}$ is set to zero for identifiability.
 This simplifies likelihood maximization by removing constraints in the optimization problem.
@@
 
-The CarHMM, introduced by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence not accounted by the HMM state.
+The CarHMM, introduced by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process.
 Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
-Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$, depends on \textit{both} $X_t$ \textit{and} $Y_{t-1}$ rather than only $X_t$.
-The first emission $Y_1$ is assumed to be fixed as an initial value which does not depend upon $X_1$.
+Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$.
+The first emission $Y_1$ is assumed to be fixed and known as an initial value which does not depend upon $X_1$.
 %
-We denote the conditional distribution of $Y_t$ given $Y_{t-1}= y_{t-1}$ and $X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
-For example, one could assume that this conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
+We denote the conditional density of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
+For example, one could assume that the conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
 %
-\[
-\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}
-\]
-and
-\[
-\mathbb{V}(Y_t| Y_{t-1} =y_{t-1}, X_t = i) = (\sigma^{(i)})^2.
-\]
+\begin{align*}
+\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) &= \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}, \text{ and } \\
+\mathbb{V}(Y_t| Y_{t-1} = y_{t-1}, X_t = i) &= (\sigma^{(i)})^2.
+\end{align*}
 %
-Figure \ref{fig:models} shows a graphical representation of the dependence structure of a CarHMM.
+Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.
 
@@
 \begin{equation*}
-    \calL_{\text{CarHMM}}(y;\Theta,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
+    \calL_{\text{CarHMM}}(\Theta,\Gamma;y) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
     \label{CarHMM_likelihood}
@@
 In the hierarchical setting, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$.
-The fine-scale state sequence in turn emits a sequence of fine-scale observations, $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
-This fine-scale process ($X_t^*, Y_t^*$) then makes up another HMM whose parameters depend on the value of $X_t$.
-Specifically, if $X_t=i$, then the distribution of $X_t^*$ is characterized by an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
+The fine-scale state sequence in turn emits a sequence of fine-scale observations $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
+For each $t$, the fine-scale process ($X_t^*, Y_t^*$) then follows another HMM whose parameters depend on the value of $X_t$.
+Specifically, if $X_t=i$, then the components of $X_t^*$ are a Markov chain with an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
 For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary in general.
-The distribution of $Y_{t, t^*}$ given $X_{t, t^*}=i^*$ and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
+The distribution of $Y_{t,t^*}$ given $X_{t,t^*}=i^*$ and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
 We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\Theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^*)}\right)$.
-Given the coarse-scale hidden states, $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are all assumed to be independent of one another.
+Given the coarse-scale hidden states, the $T$ fine-scale HMMs $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are all assumed to be independent of one another.
 
 Depending upon the process being modelled, it is possible to force certain parameters to be shared to reduce the complexity of the HHMM.
-For example, in the killer whale case study (Section \ref{sec:case_study}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( \theta^{*(1,i^*)} = \ldots = \theta^{*(N,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
+For example, in the killer whale case study (section \ref{sec:data}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( \text{i.e.
+} \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
+Figure \ref{fig:models}c represents the dependence structure for an HHMM.
 
-Figure \ref{fig:models} represents the dependence structure for an HHMM.
 Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easy to calculate via the forward algorithm.
@@
 Let $y$ be the $T$-vector of the observed coarse-scale emissions and
-$y^* \equiv \left(y^*_1, \ldots,y^*_T\right)$ be the collection of $T$ observed fine-scale emission vectors.
+$y^* \equiv \left\{y^*_1, \ldots,y^*_T\right\}$ be the collection of $T$ observed fine-scale emission vectors.
 %
@@
 \[
-\calL_{\text{HHMM}}(y,y^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,y_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N
+\calL_{\text{HHMM}}(\Theta,\Theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N
 \]
@@
 where $P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to 
-$f^{(i)}(y_t)\calL_{\text{HMM}}\left(y_t^*;\Theta^{*(i)},
-\Gamma^{*(i)}\right)$.
+$f^{(i)}(y_t)\calL_{\text{HMM}}\left(\Theta^{*(i)},
+\Gamma^{*(i)};y_t^*\right)$.
 
@@
 
-In many applications where data is collected at high frequencies, intricate dependency structures may arise within the fine-scale process $Y^*$ which cannot be adequately modeled with any of the HMM-based models mentioned thus far.
-For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behavior.
-To handle these additional fine-scale structures, we recommend replacing the fine-scale observations $Y_t^*$ with relevant statistics that summarize this non-Markovian behavior.
+In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far.
+For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour.
+To handle these additional fine-scale structures, we recommend replacing the fine-scale observations $Y_t^*$ with relevant statistics that summarize this non-Markovian behaviour.
 To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over $Y_t^*$.
@@
 
-For our case study, we use the discrete Fourier transform (DFT) of a moving window of width $h$ across $Y^*$:
+For our case study, we use the discrete Fourier transform (DFT) of a moving window of width $h$ across $Y^*_t$:
 %
@@
 %
-In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared 2-norm of the component of $\{Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\}$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window (by the Plancherel theorem).
-We use $\Ztwo$ to capture the ``wiggliness" of the acceleration within a window.
+In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared 2-norm of the component of $\left(Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\right)$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window (by the Plancherel theorem).
+More intuitively, $\Ztwo$ corresponds to the ``wiggliness" of the acceleration data within a window.
 The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert.
-Other choices for window summary statistics are possible, including the dominant frequency and amplitude.
-Figure \ref{fig:models} represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM.
+Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude.
 To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \left(\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\right)$.
-The down sampling removes the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
-By summarizing the data, the method overall also reduces correlation in the raw observations that would due to the complex dependencies structure.
+The down-sampling avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
+By summarizing the data using a relevant transformation, the method overall also reduces correlation in the raw observations due to the complex dependence structure.
+Figure \ref{fig:models}d represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM, including down-sampling.
 
-Once $\Z_t$ is calculated, it simply replaces the original fine-scale observations $Y^*_t$ in the HHMM structure.
-As such, the emission distributions and the likelihood remain identical to the original HHMM when transforming $Y^*$ to $\Z$.
-In order to clearly differentiate models, we refer to an HHMM with $\Z$ as fine-scale observations as an HHMM-DFT.
+Once $\Z_t$ is calculated, it simply replaces the original fine-scale observations $Y^*_t$ in the fine-scale HMM structure.
+As such, the emission distributions and the likelihood remain identical to the original HMM when transforming $Y^*$ to $\Z$.
+In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations as an HMM-DFT.
 
@@
 \[
-\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
+\calL_{\text{coarse}}(\Theta,\Theta^*,\Gamma;y,y^*) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
 \]
-where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{HMM}}\left(y^*_t;\Theta^{*(i)}\right)$.
-On the fine scale, $Y^*_t$ can be modelled using \textit{any} parametric model which admits an easy-to-compute likelihood.
-The likelihood $\calL_{\text{HMM}}$ is then replaced by the likelihood of this general fine-scale model, $\calL_{\text{fine}}(y^*_t;\Theta^{*(i)})$:
+where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{HMM}}\left(\Theta^{*(i)};y^*_t\right)$.
+On the fine scale, $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods (or penalized likelihoods).
+The likelihood $\calL_{\text{HMM}}$ is then replaced by the likelihood of this general fine-scale model, $\calL_{\text{fine}}(\Theta^{*(i)};y^*_t)$:
 \[
-\calL_{\text{coarse}}(y,y^*;\Theta,\Theta^*,\Gamma) = \delta P(y_1,y^*_1;\Theta,\Theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\Theta,\Theta^*) \mathbf{1}_N
+\calL_{\text{coarse}}(\Theta,\Theta^*,\Gamma;y,y^*) = \delta P(y_1,y^*_1;\Theta,\Theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\Theta,\Theta^*) \mathbf{1}_N
 \]
-where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(y^*_t;\Theta^{*(i)}\right)$.
+where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(\Theta^{*(i)};y^*_t\right)$.
 
-Possible candidates for the fine-scale model include, but are not limited to, any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT).
-These alternative fine-scale models can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models.
+Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT) in addition to many others not described here. 
+For example, a continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}.
+In particular, \citet{Xu:2018} model biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours.
+However, they take a longitudinal approach whereas we use a coarse-scale HMM in our case study.
+Further, \citet{Bebbington:2007} and \citet{Borchers:2013} model data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model.
+If the fine-scale model is a simple Poisson process, this model reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
+In addition, \citep{Langrock:2018} use B-splines to model several aspects of an HMM, and this modified model admits a penalized likelihood which can be incorporated into the hierarchical structure here.
+%For these and countless other possible fine-scale models, added flexibility comes at a cost of increased model complexity.
+
+These examples are a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models.
 As an example, consider a \textbf{CarHHMM}, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM.
 Alternatively, one may construct an \textbf{HHHMM}, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$) and traditional HMMs are used to model all three levels.
-It is clear that infinitely many possible models can be built using this framework.
+It is clear that a a myriad of possible models can be built using this framework.
 However, these models quickly become complicated and should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.
-%In the sections that follow, we construct several candidate models from the building blocks above to describe killer whale dive behavior.
+%In the sections that follow, we construct several candidate models from the building blocks above to describe killer whale dive behaviour.
 We select one preferred model and compare it to several of the simpler candidate models to ensure its complexity is necessary.

### 2020-08-14 Report/Sections/results.tex

@@
 
-%\section{Killer Whale Case Study}
+%\section{Data Collection and Model Construction}
 
-%In order to test the advantages of the CarHHM-DFT over the other models on real-world data, we used the four models described in the simulation study to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada.
+To illustrate the process of constructing a model using these HMM building blocks, we investigate the dive behaviour of a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada, and construct several candidate generative models to describe the associated data.
 
-%\section{Results}
+Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}.
+HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g.
+\citep{DeRuiter:2017} and \citep{Isojunno:2017}).
+For NRKWs, we are interested in categorizing different diving behaviours, with a particular focus on differentiating potential foraging dives.
+NRKWs prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKW for predation \citep{Williams:2009,Noren:2011}.
+It is therefore of interest to determine the energetic expenditure of NRKW to forage for Chinook, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
+Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the behavioural state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
+Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is important to understand the true energetic requirements of a NRKW.
 
-We fit all four models from the simulation study to the data shown in Figure \ref{fig:data} to test their performance in a real-world setting.
-We first report the results from the full CarHHMM-DFT (defined in Section \ref{sec:data}) in detail and then asses the quality of the fit.
+\subsection{Data collection and preprocessing}
+
+The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions.
+Observations were collected at a rate of 50 Hz using
+% an accelerometer and time-depth recorder.
+[NAME OF TAGS - do you know Sarah?].
+The acceleration data were collected in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left).
+Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
+Tagging the killer whale caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored.
+We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm, when the tagging technology malfunctioned.
+A killer whale ``dive" is any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
+We smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second.
+Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}.
+The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}.
+Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
+Specifically, the vector of observed coarse-scale observations $y = \Big(y_1,\ldots,y_{267}\Big)$ is a time series of dive durations in seconds ($s$) with corresponding vector of unobserved dive types $x = \Big(x_1,\ldots,x_{267}\Big)$.
+For dive $t^*$, the fine-scale observations are $y^*_{t^*}$, the vector of the within-dive acceleration data in meters per second squared ($m/s^2$).
+The corresponding vector of subdive states is $x_{t^*}^*$.
+As usual, $y^* = \Big\{y^*_1,\ldots,y^*_{267}\Big\}$ and $x^* = \Big\{x^*_1,\ldots,x^*_{267}\Big\}$ is the corresponding set of subdive states.
+
+\subsection{Model Development and Selection}
+
+Building an suitable model for these data involves selecting appropriate model structures, emission distributions, and numbers of states on both the coarse and fine scales.
+To choose how to model the coarse-scale observations, we first inspect lag plots to determine whether auto-correlation should be incorporated in the model structure (see section 2.1 in the supplementary material).
+The coarse-scale level exhibits no significant amount of auto-correlation and no obvious intricate structure prior to fitting the model.
+As such, we select a simple HMM to model the coarse-scale process.
+Given that dive $t$ has type $i$, the dive duration $Y_t$ is assumed to follow a gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:
+%
+$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
+
+In contrast, the fine-scale acceleration data exhibits significant sinusoidal behaviour.
+Thus, we transform $Y^*$ to $Z^*$ using the DFT for this model.
+We select a window size of two seconds since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data.
+As a result of the tri-axial acceleration readings, the transformation from the raw accelerometer data $Y^*$ to $Z^*$ is calculated slightly differently than in Equation (\ref{eqn:z}):
+%
+\begin{equation}
+    \label{eqn:z0}
+    \Zone_{t,t^*} := \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n}, \qquad \Ztwo_{t,t^*} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
+\end{equation}
+%
+Each observation is made up of $\Zone_{t,t^*}$ (a 3-dimensional vector of average acceleration) and $\Ztwo_{t,t^*}$ (a scalar describing ``wiggliness").
+We calculate $\Ztwo_{t,t^*}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to a maximum frequency of 5 Hz.
+
+Even after accounting for fine-scale structure in the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (see section 2.1 of the supplementary material).
+Therefore, a CarHMM is used to model $\Zone_{t,t^*}$.
+We assume that $\Zone_{t,t^*}$ is Normally distributed with unknown parameters $\phi_A^{*(i^*)} \in \mathbb{R}$, $\mathbf{\mu}_A^{*(i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_A^{*(i^*)} \in \mathbb{R}^3$, where the $j^{th}$ component of $\Zone_{t,t^*}$ follows
+%
+\begin{align*}
+    \mathbb{E}\left[\left(\Zone_{t,t^*}\right)_j\Big|\left(\Zone_{t,t^*-1}\right)_j ,X^*_t=i^*\right] &= \phi_A^{*(i^*)} ~ \Big(\Zone_{t,t^*-1}\Big)_j ~+ ~(1-\phi_A^{*(i^*)}) ~\left(\mu_A^{*(i^*)}\right)_j,\\
+    %
+    \mathbb{V}\left[\left(\Zone_{t,t^*}\right)_j\Big|\left(\Zone_{t,t^*-1}\right)_j ,X^*_t=i^*\right] &= \left(\sigma^{*(i^*)}\right)^2_j.
+\end{align*}
+%
+Each axis of acceleration has its own mean and variance, but all axes have the same auto-correlation parameter. We assume that all three components of $\Zone$ are independent.
+%
+%$$\mathbb{E}(\Zone_{t,t^*}|\\Zone_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_A^{*(i)} \mathbf{z} + (1-\phi_A^{*(i)}) \mathbf{\mu}_A^{*(i)},$$
+%$$\mathbb{V}(\Zone_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_A^{*(i)}\right)^2\right].$$
+%
+While $\Ztwo$ also exhibits some auto-correlation prior to fitting the model, the relationship is less strong than that of $\Zone$, and the biological interpretation of auto-correlation within $\Ztwo$ is less clear.
+In addition, much of the auto-correlation evident from the lag plot may be explained by subsequent observations sharing a subdive state $X^*_{t,t^*}$.
+Therefore, a simple HMM is used to model $\Ztwo$.
+Given that $X^*_{t,t^*} = i^*$, the distribution of $\Ztwo_{t,t^*}$ is assumed to be gamma distributed and parameterized by its mean $\mu_W^{*(i^*)} \in \mathbb{R}$ and standard deviation $\sigma_W^{*(i^*)} \in \mathbb{R}$.
+%
+%$$\mathbb{E}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_W^{*(i)}$$
+%$$\mathbb{V}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_W^{*(i)}\right)^2.$$
+%
+
+We do not use information criteria to select the number of dive types and subdive types since they tend to overestimate the number of states in biological processes \citep{Pohle:2017}.
+If the hidden states are well-separated, the lag plot should reveal $N$ distinct patterns corresponding to each dive type and $N^*$ distinct patterns corresponding to each subdive state \citep{Lawler:2019}.
+This is unfortunately not the case for our killer whale data, but we still consult our lag plots together with a heuristic approach to choose $N = 2$ dive types and $N^* = 3$ subdive behaviours.
+The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
+
+The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale.
+On the fine scale, $Y^*$ is transformed to $Z^*$ using Equation (\ref{eqn:z0}), $\Zone$ is modelled with a CarHMM, and $\Ztwo$ is modelled with an HMM.
+We assume that all components of $\Zone$ and $\Ztwo$ are independent.
+We also force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity.
+This implies that dive types 1 and 2 share subdive states.
+In total, the parameters to estimate are
+%
+\begin{gather*}
+    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
+    %
+    \Theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{($Y$ emission parameters), and} \\
+    %
+    \Theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{(fine-scale emission parameters), where} \\
+    %
+    \theta^{*(\cdot,i^*)} =  \{\{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)}\},\{\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}\}.
+\end{gather*}
+%
+Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $Z_{t,t^*}$ when conditioned on $X^*_{t,t^*} = i^*$.
+
+We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMMs discussed so far.
+Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model for the CarHHMM-DFT.
+The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
+See the appendix for details of likelihood evaluation.
+In addition to the CarHHMM-DFT, we consider three similar variations for comparison:
+\begin{enumerate}
+    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z0}), but models both $\Zone$ and $\Ztwo$ with a simple HMM rather than a CarHMM.
+    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z0}), and models $\Zone$ using a CarHMM.
+However, the Fourier sums $\Ztwo$ are omitted from this model altogether.
+The CarHHMM therefore does not know about the ``wiggliness" of the accelerometer data.
+    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z0}), and models $\Zone$ with a CarHMM and $\Ztwo$ with an HMM.
+This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive.
+\end{enumerate}
+
+Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT is missing auto-correlation within the fine-scale observations, the CarHHMM does not have access to the ``wiggliness" $\Ztwo$, and the CarHMM-DFT lacks a hierarchical structure.
+
+\subsection{Results}
+
+We fit all four models to the data shown in Figure \ref{fig:data} to test their performance in a real-world setting.
+We first report the results from the full CarHHMM-DFT in detail and then asses the quality of the fit.
 We also compare the results of the CarHHMM-DFT with those of the other candidate models.
@@
 Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives.
-The supplementary material also shows the estimated probability of each dive and subdive state given the data (Section S-2.2).
+The supplementary material also shows the estimated probability of each dive and subdive state given the data (see section 2.2 of the supplementary material).
 
@@
 
-Two visual tools are used to evaluate this model: pseudo-residuals and empirical histograms.
+We use two visual tools to evaluate this model: pseudo-residual plots and empirical histograms.
 A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}.
 To easily visualize outliers, pseudo-residuals are often passed through the quantile function of the standard Normal distribution.
-Mathematically, the pseudo-residual of an observation $y_t$ of a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
+Mathematically, the pseudo-residual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
 If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution.
@@
 This implies that the true distribution of $\Ztwo$ may follow a heavier-tailed distribution than the gamma distribution used in the case study.
+See section 2 of the supplementary material for pseudo-residual plots for all observations.
 
-We also plot histograms of $Y$ and $Z^*$ corresponding to each dive type or subdive state in Figure \ref{fig:empirical_dist}.
-Observations are weighted by the decoded probability that they correspond to a particular hidden state, and this histogram is plotted over the emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}).
-Our results again mostly support a well-specified model with the exception of $\Ztwo$, which is again right-skewed.
-Another exception is $\Zone$, which has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
+We also plot histograms of $Y$ corresponding to each dive type in Figure \ref{fig:empirical_dist}.
+Observations are weighted by the decoded probability that they correspond to a particular dive type.
+This results in 2 histogram corresponding to $Y$ - one for dive type 1 and another for dive type 2.
+These histograms are then plotted over their corresponding emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}).
+See section 2 of the supplementary material for histograms corresponding to the fine-scale observations.
+Our results again mostly support a well-specified model with some exceptions.
+In particular, $\Ztwo$, is again right-skewed while $\Zone$ has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
 These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.
-See Section S-2.?
-for histograms for every observation and hidden state.
 
@@
 
-The CarHMM-DFT produces very similar parameter estimates and state estimates compared to those of the CarHHMM-DFT on the fine scale.
+The CarHMM-DFT, which lacks a hierarchical structure, produces very similar parameter estimates and state estimates compared to those of the CarHHMM-DFT on the fine scale.
 However, its lack of a hierarchical structure means that it fails to differentiate between short and long dives.
@@
 
-The HHMM-DFT decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3.
+The HHMM-DFT, which ignores auto-correlation in $\Zone$, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3.
 In addition, the HHMM-DFT produces estimates of $\sigma_A^{*(1)}$, $\sigma_A^{*(2)}$, and $\sigma_A^{*(3)}$ (the standard deviation of acceleration) which are approximately 50-100 \% larger than those of the CarHHMM-DFT across all axes of acceleration.
 The estimated uncertainties of the estimates of $\mu_A^{*(1)}$, $\mu_A^{*(2)}$, and $\mu_A^{*(3)}$ are also less than half of those for the CarHHMM-DFT across all axes of acceleration.
-These findings are consistent with the tendency to overestimate variance and underestimate standard error when ignoring auto-correlation.
 Further, the empirical auto-correlation of $\Zone$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers.
-Both of these findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.
+These findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.
 
-Finally, the CarHHMM does not use the ``wiggliness" of the acceleration data as an observation, so it fails produce biologically interpretable results when decoding the subdive states.
+Finally, the CarHHMM does not use the ``wiggliness" of the acceleration data as an observation, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives}.
 This fact alone essentially disqualifies the CarHHMM as a viable model for this data.
 The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution.
-See section S-2 for a more complete set of results for each of the candidate models.
+See section 2 of the supplementary material for a more complete set of results for each of the candidate models.

### 2020-08-14 Report/Sections/results_old.tex

--- 2020-08-06 Report/Sections/results.tex
@@
 The emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). 
-Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, traveling, and searching for prey.
+Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, travelling, and searching for prey.
 On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as foraging and socialization \citep{Tennessen:2019b}.
@@
 We also plot histograms of $Y$ and $Z^*$ corresponding to each dive type or subdive state in Figure \ref{fig:empirical_dist}.
-Observations are weighted by the decoded probability that they correspond to a particular hidden state, and this histogram is plotted over the emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}).
+When constructing the histogram corresponding to each dive type or subdive state, observations are weighted by the decoded probability that they correspond to that hidden state.
+This allows each hidden state to be evaluated separately.
+These histograms are plotted over their corresponding emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}).
 Our results again mostly support a well-specified model with the exception of $\Ztwo$, which is again right-skewed.
@@
 The estimated uncertainties of the estimates of $\mu_A^{*(1)}$, $\mu_A^{*(2)}$, and $\mu_A^{*(3)}$ are also less than half of those for the CarHHMM-DFT across all axes of acceleration.
-These findings are consistent with the tendency to overestimate variance and underestimate standard error when ignoring auto-correlation.
 Further, the empirical auto-correlation of $\Zone$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers.
-Both of these findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.
+These findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.
 
-Finally, the CarHHMM does not use the ``wiggliness" of the acceleration data as an observation, so it fails produce biologically interpretable results when decoding the subdive states.
+Finally, the CarHHMM does not use the ``wiggliness" of the acceleration data as an observation, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives}.
 This fact alone essentially disqualifies the CarHHMM as a viable model for this data.

### 2020-08-14 Report/Sections/simulation.tex

@@
 To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT.
-The parameters used to generate the data are loosely based on those estimated in the case study in Section \ref{sec:case_study} (see table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made to simplify presentation.
+The parameters used to generate the data are loosely based on those estimated in the case study (see able \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity.
 For example, we set $\Gamma$ such that the simulated data has approximately equal numbers of short and long dives.
 This allows meaningful conclusions to be made for both dive types.
-In addition, the number subdive states is set to $N^*=2$ and $\Ztwo$ is set as a scalar for additional simplicity.
+In addition, the number subdive states is set to $N^*=2$ and $\Ztwo$ is set to a scalar.
 Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times.
-To assess how accurate each model is when estimating its own uncertainty, we also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.
+To assess the accuracy of uncertainty estimates, we also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.
 
@@
 The two subdive types differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher in subdive state 2 than subdive state 1.
-This corresponds to much more vigorous and variable periodic behavior in the acceleration data.
+This corresponds to much more vigorous and variable periodic behaviour in the acceleration data.
 
@@
 Using the forward-backward algorithm, the CarHHMM-DFT decodes the correct subdive type given the fine-scale data with an average confidence of $1 - 10^{-12}$ across all simulations.
-For dive type the average confidence was above 0.9 (see Table \ref{table:accuracy}).
-All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased.
+In other words, it predicts the \textit{incorrect} subdive state with an average probability of $10^{-12}$.
+For dive type, the average confidence was above 0.9 (see Table \ref{table:accuracy}).
+All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased (see section 1 of the supplementary material).
 Even still, these biases are the least severe for the CarHHMM-DFT among the four models.
 The empirical standard error of all parameter estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slight underestimates of the empirical standard error.
-See Section S-1.2, tables 2-5 for more information.
+See section 1.2 and tables 2-5 of the supplementary material for more information.
 
@@
 For both subdive types, $\hat \sigma_A^*$ greatly overestimates the true standard deviation.
-The estimated standard errors of $\hat \mu_A^*$ and $\hat \sigma_A^*$ using the observed Fisher information are both much smaller than the empirical standard error (see Section S-1.2, Table 3).
-This implies that variance can be overestimated and parameter estimates can be too confident when auto-correlation is ignored.
+The estimated standard errors of $\hat \mu_A^*$ and $\hat \sigma_A^*$ using the observed Fisher information are both much smaller than the empirical standard error (see section 1.2, Table 3 of the supplementary material).
+This implies that the variance of acceleration can be overestimated and the standard error of the parameter estimates can be underestimated when auto-correlation is ignored.
+These findings are consistent with the results from the case study, where the HHMM-DFT produced higher variance estimates and narrower standard error estimates than the CarHHMM-DFT.
 
@@
 This result is consistent with expectations because the CarHHMM does not have access to the ``wiggles" of the fine-scale process, which are the most informative observations when determining subdive type.
-The CarHHMM is also much more likely to mis-classify subdive type 1 than subdive type 2 (see Section S-1.2, Table 1).
-The CarHHMM is the worst of the four at estimating parameters, and this is in part due to its relatively poor decoding accuracy.
-$\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ are especially poor estimates (see Section S-1.1.3).
+The CarHHMM is also much more likely to mis-classify subdive type 1 than subdive type 2 (see section 1.2, Table 1 of the supplementary material).
+The CarHHMM is also the worst of the four at estimating parameters, and this is in part due to its relatively poor decoding accuracy.
+$\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ are especially poor estimates (see section 1 of the supplementary material).
 
@@
 The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of \textit{all} dives.
-The parameter estimates for this distribution ($\hat \mu$ and $\hat \sigma$) are highly correlated and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds.
+The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds.
 See Figures \ref{fig:acc_coarse} \& \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive/subdive type probabilities as determined by each model.

### 2020-08-14 Report/main.tex

@@
 % Title, authors, affiliations
-\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
+%\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
+\title[]{Modelling state-switching behaviour in functional data with hidden Markov models}
 \author{BlindedA\authorref{1}\thanksref{*}}
@@
 
-Advances in tagging technology have made a vast amount of high-frequency time series data available in several fields.
+Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields.
 These data often exhibit complicated temporal dependence structure which cannot be modelled using common methods of Functional Data Analysis (FDA).
-The field of animal movement suggests a hierarchical approach using hidden Markov models (HMMs), but many intricate fine-scale processes violate basic assumptions of HMMs.
-We detail a generalization of the hierarchical HMM approach that incorporates a wider class of parametric models at problematic fine scales. 
+Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs), but many intricate fine-scale processes violate basic assumptions of HMMs.
+We detail a generalization of the hierarchical HMM approach that incorporates a wider class of parametric models at fine scales. 
 %We also provide several examples of viable fine-scale models.
 As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada. 
-Our model outputs more interpretable results as well as more accurate parameter and standard error estimates compared to existing models.
+Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing models.
 
@@
 %
-\section{Data Collection and Model Construction}
+\section{Killer Whale Case Study}
 \label{sec:data}
-\input Sections/data.tex
+\input Sections/results.tex
 %
@@
 %
-\section{Case Study Results}
-\label{sec:case_study}
-\input Sections/results.tex
+%\section{Case Study Results}
+%\label{sec:case_study}
+%\input Sections/results.tex
 %%

### 2020-08-14 Code/Visualisor.py
Update plot notation

@@ -50,10 +50,10 @@ class Visualisor:
 
         features = list(self.pars.features[0].keys()) + list(self.pars.features[1].keys())
         xlabs = ['Dive Duration $(s)$',
-                 r'$\left(Z^{*(1)}\right)_x$ $(m/s^2)$',
-                 r'$\left(Z^{*(1)}\right)_y$ $(m/s^2)$',
-                 r'$\left(Z^{*(1)}\right)_z$ $(m/s^2)$',
-                 r'$Z^{*(2)}$']
+                 r'$A^*_x$ $(m/s^2)$',
+                 r'$A^*_y$ $(m/s^2)$',
+                 r'$A^*_z$ $(m/s^2)$',
+                 r'$W^*$']
 
         # lag plots of dive-level data
         for i,feature in enumerate(features):
@@ -176,10 +176,10 @@ class Visualisor:
                     ax[row_num,col_num].set_title(title,fontsize=24)
                     plt.legend(['Dive Type 1','Dive Type 2'])
                 else:
-                    titles = [r'$\left(Z^{*(1)}\right)_x$ $(m/s^2)$',
-                              r'$\left(Z^{*(1)}\right)_y$ $(m/s^2)$',
-                              r'$\left(Z^{*(1)}\right)_z$ $(m/s^2)$',
-                              r'$Z^{*(2)}$']
+                    titles = [r'$A^*_x$ $(m/s^2)$',
+                              r'$A^*_y$ $(m/s^2)$',
+                              r'$A^*_z$ $(m/s^2)$',
+                              r'$W^*$']
                     title = titles[feature_num]
                     ax[row_num,col_num].set_xlabel(title,fontsize=24)
                     if feature == 'Ahat_low':
@@ -310,11 +310,11 @@ class Visualisor:
                              '.',color=colors[state],markersize=10)
                 if 'prob' in col:
                     plt.plot(dive[dive[col] > -0.01]['sec_from_start']/60,dive[dive[col] > -0.01][col],'k-')
-                    plt.axhline(0.5,color='k',linestyle=':',alpha=0.5)
+                    plt.axhline(0.5,color='k',linestyle=':',alpha=0.5,linewidth=1)
                     plt.ylim([-0.05,1.05])
                     plt.yticks([0,0.5,1.0],fontsize=30)
                 else:
-                    plt.plot(dive['sec_from_start']/60,dive[col],'k--',alpha=0.5)
[…]

### 2020-08-19 Code/Visualisor.py
Standardize plot sizes

@@ -44,12 +44,12 @@ class Visualisor:
 
         ncols = 2
         nrows = int(np.ceil(len(self.pars.features[0]) + len(self.pars.features[1])/2))
-        fig,ax = plt.subplots(nrows,ncols,figsize=(10*ncols,10*nrows))
+        fig,ax = plt.subplots(nrows,ncols,figsize=(2.5*ncols,2.5*nrows))
         fig.delaxes(ax[2,1])
         fig_num = 1
 
         features = list(self.pars.features[0].keys()) + list(self.pars.features[1].keys())
-        xlabs = ['Dive Duration $(s)$',
+        xlabs = ['$Y$ $(s)$',
                  r'$A^*_x$ $(m/s^2)$',
                  r'$A^*_y$ $(m/s^2)$',
                  r'$A^*_z$ $(m/s^2)$',
@@ -89,24 +89,31 @@ class Visualisor:
 
             np.set_printoptions(suppress=True,precision=4)
             im = plt.imshow(np.rot90(Ztemp),extent = xlim + ylim)
-            plt.title(xlabs[i],fontsize=36)
+            plt.xlabel(xlabs[i],fontsize=12)
+            plt.xticks(fontsize=10)
+            plt.yticks(fontsize=10)
             plt.gca().ticklabel_format(style='sci',scilimits = (-3,3))
+            plt.gca().yaxis.get_offset_text().set_fontsize(10)
+            plt.gca().xaxis.get_offset_text().set_fontsize(10)
             divider = make_axes_locatable(plt.gca())
             cax = divider.append_axes("right", size="5%", pad=0.05)
             c = fig.colorbar(im,cax=cax)#im, orientation='horizontal',pad=0.125)
+            c.ax.tick_params(labelsize=10)
             c.ax.ticklabel_format(style='sci',scilimits = (-3,3))
-            c.set_label('Density',fontsize = 30)
+            c.ax.yaxis.get_offset_text().set_fontsize(10)
+            if i%2 == 1 or i==4:
+                c.set_label('Density',fontsize = 10)
 
             fig_num += 1
 
-        plt.subplots_adjust(wspace=-0.6, hspace=-0.6)
+        plt.subplots_adjust(wspace=0.5, hspace=0.5)
         #fig.text(0.5, 1.0, 'Lag Plots', ha='center', fontsize=50)
 
         if file is None:
             plt.show()
         else:
-            plt.tight_layout()
[…]

### 2020-08-19 Code/Visualisor.py
Update Plots again

@@ -3,6 +3,7 @@ import numpy as np
 import matplotlib.pyplot as plt
 import matplotlib.cm as cm
 from matplotlib.patches import Patch
+from matplotlib.transforms import Bbox
 from matplotlib.lines import Line2D
 from matplotlib.ticker import FormatStrFormatter
 from mpl_toolkits.axes_grid1 import make_axes_locatable
@@ -176,7 +177,7 @@ class Visualisor:
                     raise('distribution %s not recognized' % dist)
                 ax[row_num,col_num].plot(x,y,color=colors[state])
                 if col_num == 0:
-                    ax[row_num,col_num].set_ylabel('Probability Density',fontsize=12)
+                    ax[row_num,col_num].set_ylabel('Density',fontsize=12)
                 if level == 0:
                     title = 'Emission Distributions, Dive Duration'
                     ax[row_num,col_num].set_xlabel('Dive Duration (seconds)')
@@ -205,7 +206,7 @@ class Visualisor:
                     fig.legend(handles=legend_elements,
                                prop={'size': 10}, ncol=3,
                                mode='expand',
-                               bbox_to_anchor=(0.0, 1.011, 1., .1),
+                               bbox_to_anchor=(0.0, 0.99, 1., .1),
                                loc='lower left')
 
                     #fig.text(0.5, 1.0, 'Fine Scale Emission Distributions', ha='center', fontsize=50)
@@ -214,7 +215,8 @@ class Visualisor:
             plt.show()
         else:
             plt.tight_layout()
-            plt.savefig(file,bbox_inches='tight')
+            bbox = Bbox([[0,0],[5,5.5]])
+            plt.savefig(file,bbox_inches=bbox,dpi=500)
 
         return
 

### 2020-08-20 Code/Parameters.py
Update supplemental material plots

@@ -26,8 +26,8 @@ class Parameters:
     def __init__(self):
 
         # loading parameters
-        self.cvc_file = '../Data/20190902-182840-CATs_OB_1.cvc'
-        self.csv_file = '../Data/20190902-182840-CATs_OB_1_001.csv'
+        self.cvc_file = '../Data.nosync/20190902-182840-CATs_OB_1.cvc'
+        self.csv_file = '../Data.nosync/20190902-182840-CATs_OB_1_001.csv'
 
         self.cvc_cols = ['date','time',
                          'GPS lat','GPS long','GPS alt','GPS speed',

### 2020-08-20 Code/Visualisor.py

@@ -153,8 +153,8 @@ class Visualisor:
                 colors = [cm.get_cmap('tab10')(i) for i in [0,1]]
                 legend = ['Dive Type %d'%(x+1) for x in range(K)]
             else:
-                mu = self.hhmm.theta[1][row_num][feature]['mu']
-                sig = self.hhmm.theta[1][row_num][feature]['sig']
+                mu = self.hhmm.theta[1][0][feature]['mu']
+                sig = self.hhmm.theta[1][0][feature]['sig']
 
                 dist = self.pars.features[1][feature]['f']
                 K = self.pars.K[1]

### 2020-08-20 Report/Sections/appendix.tex

@@
 
-Assume that $X$, $Y$, and $X^*$ are known and simulated using the procedure detailed in Section \ref{sec:sim_study}.
-Let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and let $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for shorthand.
+Assume that $X$, $Y$, and $X^*$ are known and simulated using the procedure detailed in Section \ref{sec:sim_study}, leaving only $Y^*$ to be simulated.
+Let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and let $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for convenience.
+We simulate $\hat Y^*$ and then take the inverse $DFT$ to find $Y^*$.
 
-Then, $\hat{Y}^*_{t}$ is simulated using the following procedure for every dive. First, simulate the average acceleration of the first two-second window of the dive:
+\begin{enumerate}
+    \item We begin by simulating $\hat{Y}^{*(0)}_{t,1}$:
     $$
-	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(\sigma^{*(i^*)}\right)^2 \right)
+    	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(\sigma_A^{*(i^*)}\right)^2 \right)
     $$
-Remember that the average acceleration within a window is equal to the real part of the $0^{th}$ Fourier mode, so $\hat{Y}^{*(0)}_{t,1}$ is the average acceleration within the first window.
-Next, simulate the average acceleration for each subsequent two-second window:
+    where $\sigma_A^{*(1)} = 0.05s$ and $\sigma_A^{*(2)} = 0.1s$. 
+    \item Next, we simulate $\hat{Y}^{*(0)}_{t,101},\enspace \hat{Y}^{*(0)}_{t,201}, \enspace \ldots$ using Equation (\ref{eqn:yhat_0}):
     \begin{equation}	
-	\Big(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*,\hat{Y}^{*(0)}_{t,t^*-100}\Big) \sim \mathcal{N} \left(\phi^{*(i^*)} \hat{Y}^{*(0)}_{t,t^*-100}, \left(\sigma^{*(i^*)}\right)^2 \right), \quad t^* = 101,201,\ldots
+        \Big(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*,\hat{Y}^{*(0)}_{t,t^*-100}\Big) \sim \mathcal{N} \left(\phi_A^{*(i^*)} \hat{Y}^{*(0)}_{t,t^*-100}, \left(\sigma_A^{*(i^*)}\right)^2 \right), \quad t^* = 101,201,\ldots
     	\label{eqn:yhat_0}
     \end{equation}
-Note that $t^*$ has increments of 100 because $h=100$ when using two-second window sizes. Finally, simulate the other Fourier modes within each window:
+    where $\phi_A^{*(1)} = 0.99$ and $\phi_A^{*(2)} = 0.95$. Note that $t^*$ has increments of 100 because $h=100$ when using two-second window sizes. 
+    %
+    \item We then simulate Fourier modes 1 through 49 within each window:
     $$
@@
     $$
-where $a$ and $b$ are random variables with the following distributions:
+    where $a$ is equals 1 or -1 with probability 1/2 and $b^{(k)}_{t,t^*}$ has the following distribution:
     \begin{align*}
-    a_{t,t^*}^{(k)} &\sim  \left\{\begin{array}{lr}
-	-1 & \quad w.p. \enspace 1/2 \\
-	1  & \quad w.p. \enspace 1/2
-	\end{array}\right. \\
-	%
     	(b^{(k)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim {\rm{Gamma}}(5/k^2, 1) \\
@@
     \end{align*}
-Note that $b_{t,t^*}$ decays like $1/k^2$ to ``smooth out" the raw acceleration. For Fourier modes 50 through 99, we use the following procedure to ensure that the raw acceleration is real-valued:
+    This means that the expected values of Fourier modes 1 and 2 are much higher for subdive state 2.
+Note that $\mathbb{E}\left(|\hat{Y}^{(k)}_{t,t^*}|^2\right)$ is proportional to $1/k^2$.
+This silences the higher Fourier modes, which ``smooths out" the raw acceleration.
+    \item For Fourier modes 50 through 99, we use the following procedure to ensure that the inverse DFT is real-valued:
     $$
@@
     %
-$\Big(Y^*_{t,t^*},\ldots,Y^*_{t,t^*+99}\Big)$ is finally set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
+    \item Finally, $\Big(Y^*_{t,t^*},\ldots,Y^*_{t,t^*+99}\Big)$ is set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
     %
     $$Y^*_{t,t^*+n} = IDFT\left\{\hat{Y}^*_{t,t^*}\right\}(n); \quad n = 0,\ldots,99; \quad t^* = 1,101,\ldots$$
+\end{enumerate}
 %
@@
 We now prove that this construction of $Y^*$ results in the correct distributions listed in Section \ref{sec:sim_study}.
+To find the distribution of $\Zone_{t,t^*}$, use Equation (\ref{eqn:yhat_0}) and note that $\Zone_{t,t^*} = \text{Re}\left(\hat{Y}^{(0)}_{t,t^*}\right) = \hat{Y}^{(0)}_{t,t^*}$, since the 0th Fourier mode is equal to the average value of an interval and $\hat{Y}^{(0)}_{t,t^*}$ is real-valued by construction:
+%
+$$\left(\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(\phi_A^{*(i^*)} \Zone_{t,t^*-100}, \left(\sigma_A^{*(i^*)}\right)^2 \right).$$
+%
+This distribution is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(i^*)} = 0$.
+We also have that $\sigma_A^{*(1)} = 0.05s$, $\phi_A^{*(1)} = 0.99$, $\sigma_A^{*(2)} = 0.1s$, and $\phi_A^{*(2)} = 0.95$.
 
-It is straightforward to calculate both $\Zone$ and $\Ztwo$ from $Y^*$.
-We pick $\tilde{\omega} = 10$ periods per window, or 5 Hz.
-To find the distribution of $\Zone_{t,t^*}$, use Equation (\ref{eqn:yhat_0}) together with the substitution $\Zone_{t,t^*} = \text{Re}\left(\hat{Y}^{(0)}_{t,t^*}\right) = \hat{Y}^{(0)}_{t,t^*}$:
-%
-$$\left(\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(\phi^{*(i^*)} \Zone_{t,t^*-100}, \left(\sigma^{*(i^*)}\right)^2 \right).$$
-%
-This distributions is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(i^*)} = 0$.
-The other parameters $\sigma_A^{*(i^*)}$ and $\phi_A^{*(i^*)}$ are free, and we select $\sigma_A^{*(1)} = 0.05s$, $\phi_A^{*(1)} = 0.99$, $\sigma_A^{*(2)} = 0.1s$, and $\phi_A^{*(2)} = 0.95$.
-
-The wiggliness $\Ztwo_{t,t^*}$ is the sum of gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also gamma:
+To calculate the wiggliness $\Ztwo$, we pick $\tilde{\omega} = 10$ periods per window, or 5 Hz.
+Recall that $\Ztwo_{t,t^*}$ is the sum of gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also gamma:
 %
@@
 
-\subsection{Likelihood of Simulation study model}
+\subsection{Likelihood of simulation study model}
 

### 2020-08-31 Code/Visualisor.py
Fix Plots

@@ -50,11 +50,16 @@ class Visualisor:
         fig_num = 1
 
         features = list(self.pars.features[0].keys()) + list(self.pars.features[1].keys())
-        xlabs = ['$Y$ $(s)$',
-                 r'$A^*_x$ $(m/s^2)$',
-                 r'$A^*_y$ $(m/s^2)$',
-                 r'$A^*_z$ $(m/s^2)$',
-                 r'$W^*$']
+        xlabs = ['$Y_t$',
+                 r'$\left(A^*_{t,t^*}\right)_x$',
+                 r'$\left(A^*_{t,t^*}\right)_y$',
+                 r'$\left(A^*_{t,t^*}\right)_z$',
+                 r'$W^*_{t,t^*}$']
+        ylabs = ['$Y_{t+1}$',
+                 r'$\left(A^*_{t,t^*+1}\right)_x$',
+                 r'$\left(A^*_{t,t^*+1}\right)_y$',
+                 r'$\left(A^*_{t,t^*+1}\right)_z$',
+                 r'$W^*_{t,t^*+1}$']
 
         # lag plots of dive-level data
         for i,feature in enumerate(features):
@@ -91,6 +96,7 @@ class Visualisor:
             np.set_printoptions(suppress=True,precision=4)
             im = plt.imshow(np.rot90(Ztemp),extent = xlim + ylim)
             plt.xlabel(xlabs[i],fontsize=12)
+            plt.ylabel(ylabs[i],fontsize=12)
             plt.xticks(fontsize=10)
             plt.yticks(fontsize=10)
             plt.gca().ticklabel_format(style='sci',scilimits = (-3,3))
@@ -107,7 +113,7 @@ class Visualisor:
 
             fig_num += 1
 
-        plt.subplots_adjust(wspace=0.5, hspace=0.5)
+        plt.subplots_adjust(wspace=0.9, hspace=0.9)
         #fig.text(0.5, 1.0, 'Lag Plots', ha='center', fontsize=50)
 
         if file is None:
@@ -180,14 +186,14 @@ class Visualisor:
                     ax[row_num,col_num].set_ylabel('Density',fontsize=12)
                 if level == 0:
                     title = 'Emission Distributions, Dive Duration'
-                    ax[row_num,col_num].set_xlabel('Dive Duration (seconds)')
+                    ax[row_num,col_num].set_xlabel('$Y_t$ $(s)$')
                     #ax[row_num,col_num].set_title(title,fontsize=12)
[…]

### 2020-08-31 Report/Sections/appendix.tex
Updates from Overleaf

@@
 
-\citet{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation, which they refer to as a state-switching Ornstein-Uhlenbeck process:
+\citet{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
 %
@@
 where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
+They refer to this as a state-switching Ornstein-Uhlenbeck process.
 Unlike HMMs, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer.
-If the behavioral state $X_t$ is known and does not change between observations $\Big($i.e $X_s = X_t$ for all $s \in [t,t+\Delta t)\Big)$, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
+If the behavioural state $X_t$ is known and does not change between observations $\Big($i.e $X_s = X_t$ for all $s \in [t,t+\Delta t)\Big)$, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
 %
@@
 where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$.
-In fact, under certain conditions, the output of a state-switching OU process is identical to that of a CarHMM, as described by Theorem 1.
+In fact, under certain conditions, the output of a state-switching OU process is identical to that of a CarHMM, as described by Theorem 1 below.
 
@@
 
-    \item Observations of the SDE from Equation 2 are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
+    \item Observations of the SDE from Equation (\ref{eqn:OU_sol}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
 \end{enumerate}
-then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\Theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
+then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
 
@@
 Let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and let $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for convenience.
-We simulate $\hat Y^*$ and then take the inverse $DFT$ to find $Y^*$.
+We simulate $\hat Y^*$ and then take the inverse $DFT$ to find $Y^*$ according to the following procedure.
 
@@
     where $\sigma_A^{*(1)} = 0.05s$ and $\sigma_A^{*(2)} = 0.1s$. 
-    \item Next, we simulate $\hat{Y}^{*(0)}_{t,101},\enspace \hat{Y}^{*(0)}_{t,201}, \enspace \ldots$ using Equation (\ref{eqn:yhat_0}):
+    %
+    \item Next, for each dive $t$, we simulate $\hat{Y}^{*(0)}_{t,101},\enspace \hat{Y}^{*(0)}_{t,201}, \enspace \ldots ~ $ using Equation (\ref{eqn:yhat_0}) below:
     \begin{equation}	
@@
     $$
-    where $a$ is equals 1 or -1 with probability 1/2 and $b^{(k)}_{t,t^*}$ has the following distribution:
-    \begin{align*}
+    where $a^{(k)}_{t,t^*}$ is equals either 1 or -1 with probability 1/2 and $b^{(k)}_{t,t^*}$ has the following distribution:
+    \begin{align}
+    \begin{split}
     	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 1) &\sim {\rm{Gamma}}(5/k^2, 1) \\
@@
     	\end{array}\right. 
-    \end{align*}
-    This means that the expected values of Fourier modes 1 and 2 are much higher for subdive state 2.
+    \end{split}
+    \end{align}
+    This means that the expected values of Fourier modes 1 and 2 are much higher in subdive state 2 $(X^*_{t,t^*} = 2)$ Than subdive state 1 $(X^*_{t,t^*} = 1)$.
 Note that $\mathbb{E}\left(|\hat{Y}^{(k)}_{t,t^*}|^2\right)$ is proportional to $1/k^2$.
-This silences the higher Fourier modes, which ``smooths out" the raw acceleration.
-    \item For Fourier modes 50 through 99, we use the following procedure to ensure that the inverse DFT is real-valued:
+This silences the higher Fourier modes and subsequently ``smooths out" the raw acceleration after taking the inverse DFT of $\hat Y$.
+    \item For Fourier modes 50 through 99, we use the following values to ensure that the inverse DFT is real-valued:
     $$
@@
 
-We now prove that this construction of $Y^*$ results in the correct distributions listed in Section \ref{sec:sim_study}.
-To find the distribution of $\Zone_{t,t^*}$, use Equation (\ref{eqn:yhat_0}) and note that $\Zone_{t,t^*} = \text{Re}\left(\hat{Y}^{(0)}_{t,t^*}\right) = \hat{Y}^{(0)}_{t,t^*}$, since the 0th Fourier mode is equal to the average value of an interval and $\hat{Y}^{(0)}_{t,t^*}$ is real-valued by construction:
+We now show that this construction of $Y^*$ results in the correct distributions listed in Section \ref{sec:sim_study}.
+First, note that $\Zone_{t,t^*} = \text{Re}\left(\hat{Y}^{(0)}_{t,t^*}\right) = \hat{Y}^{(0)}_{t,t^*}$ because the real part of the Fourier coefficient corresponding to a frequency of 0 is equal to the average value of an interval and $\hat{Y}^{(0)}_{t,t^*}$ is real-valued by construction.
+Then, combine this fact with Equation (\ref{eqn:yhat_0}) to find the distribution of $\Zone_{t,t^*}$:
 %
@@
 
-To calculate the wiggliness $\Ztwo$, we pick $\tilde{\omega} = 10$ periods per window, or 5 Hz.
+To calculate the wiggliness $\Ztwo$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz).
 Recall that $\Ztwo_{t,t^*}$ is the sum of gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also gamma:
 %
-$$\Ztwo_{t,t^*} = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
+$$\Ztwo_{t,t^*} = \sum_{k=1}^{10} \big|\hat{Y}^{(k)}_{t,t^*}\big|^2 = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
+Combining this with Equation block (5), we have:
 %
-$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = \sum_{k=1}^{10} 5/k^2 = 10.10, \beta = 1.00 \right)$$
+$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 5/k^2 = 10.10, 1.00 \right)$$
 %
-$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(\alpha = 300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , \beta = 1.00 \right).$$
+$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , 1.00 \right).$$
 %
@@
 %
-$$\calL_{\text{CarHHMM-DFT}}(\Theta,\Theta^*,\Gamma,\Gamma^*;y,z^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N.$$
+$$\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,z^*) = \delta P(y_1,z_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\theta,\theta^*,\Gamma^*) \mathbf{1}_N.$$
 %
@@
 \begin{align*}
-P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(\Theta^{*(1)},\Gamma^{*(1)};z_t^*\right), \ldots , \\
-&f^{(N)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(\Theta^{*(N)},\Gamma^{*(N)};z_t^*\right) \Big],
+P(y_t,z_t^*;\theta,\theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\theta^{(1)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(1)};z_t^*\right), \ldots , \\
+&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(N)};z_t^*\right) \Big],
 \end{align*}
 %
-where $f^{(i)}(y_t;\Theta^{(i)})$ is the emission distribution of the dive duration $Y_t$ conditioned on the fact that $X_t = i$.
-
-The likelihood $\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale chain and has the following expression:
+where $f^{(i)}(y_t;\theta^{(i)})$ is the emission distribution of dive duration conditioned on the fact that $X_t = i$.
+The likelihood $\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale chain and is equal to the following:
 %
-$$\calL_{\text{CarHMM-DFT}}\left(\Theta^{*(i)},\Gamma^{*(i)};z_t^*\right) = \delta^{*(i)} \prod_{t=2}^T \Gamma^{*(i)} P(z^*_{t,t^*}|z^*_{t,t^*-100};\Theta^{*(i^*)}) \mathbf{1}_N,$$
+$$\calL_{\text{CarHMM-DFT}}\left(\theta^{*},\Gamma^{*(i)};z_t^*\right) = \delta^{*(i)} \prod_{z^*_{t,t^*} \in z^*_t} \Gamma^{*(i)} P(z^*_{t,t^*}|z^*_{t,t^*-100};\theta^*) \mathbf{1}_N,$$
 %
-where $P(z^*_{t,t^*}|z^*_{t,t^*-100};\Theta^{*(i^*)})$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{*(\cdot,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$.
+where $P(z^*_{t,t^*}|z^*_{t,t^*-100};\theta^*)$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{*(\cdot,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$.
 %
-Note that $f^{*(\cdot,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$ is the emission distribution of $\Z_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.
+Recall that $z^*_t = \Big(z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\Big)$ and that $f^{*(\cdot,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$ is the emission distribution of $\Z_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.
 

### 2020-08-31 Report/Sections/models.tex
Updates from Overleaf

@@
 Consider a sequence of $T$ curves (or dives, in our context). Any particular dive $t$, $t = 1,
-\ldots,T$, has an associated curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of within-curve (or fine-scale) observations $Y^*_{t} \equiv \Big(Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\Big)$.
-The vector $Y^*_{t}$ is made up of function values from curve $t$.
-We call the vectors of coarse-scale observations $Y \equiv \Big(Y_1, \ldots, Y_T\Big)$ and the collection of all fine-scale observations $Y^* \equiv \Big\{Y^*_1,\ldots,Y^*_T \Big\}$.
+\ldots,T$, has a possibly high-dimensional associated curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of within-curve (or fine-scale) observations $Y^*_{t} \equiv \Big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\Big\}$.
+The sequence $Y^*_{t}$ is made up of function values from curve $t$.
+We call the sequence of coarse-scale observations $Y \equiv \Big\{Y_1, \ldots, Y_T\Big\}$ and the collection of all fine-scale observations $Y^* \equiv \Big\{Y^*_1,\ldots,Y^*_T \Big\}$.
 
@@
 We first detail the structure of a traditional HMM.
-We then review a variation introduced by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
+We then review a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
 Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$.
@@
 
-An HMM is comprised of a sequence of unobserved states $X = \big(X_1, \ldots, X_T\big)$ where $X_t$ is associated with the observation $Y_t$.
+An HMM is comprised of a sequence of unobserved states $X = \big\{X_1, \ldots, X_T\big\}$ where $X_t$ is associated with the observation $Y_t$.
 Here we focus on the coarse scale sequence of observations $Y_1,\ldots, Y_T$, but an HMM can also be used as a sub-model for the fine-scale sequence for a particular dive.
@@
 
-Using observed emissions, here denoted $y = (y_1,\ldots,y_T)$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\Theta \equiv (\theta^{(1)},\ldots,\theta^{(N)})$.
+Using observed emissions, here denoted $y = \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$.
 We write the likelihood $\calL_{\text{HMM}}$ using the  well-known \textit{forward algorithm} \citep{Zucchini:2016}:
 %
-$$\calL_{\text{HMM}}(\Theta,\Gamma;y) = \delta P(y_1;\Theta) \prod_{t=2}^T \Gamma P(y_t;\Theta) \mathbf{1}_N$$
+$$\calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N$$
 %
@@
 %
-$P(y_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry  $f^{(i)}(y_t; \theta^{(i)})$.
+$P(y_t;\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry  $f^{(i)}(y_t; \theta^{(i)})$.
 %
@@
 
-The CarHMM, introduced by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process.
+The CarHMM, described by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process.
 Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
@@
 We denote the conditional density of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
-For example, one could assume that the conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
+For example, if $Y_t$ is a scalar, one could assume that its conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
 %
-\begin{align*}
+\begin{align}
+\label{eqn:carhmm}
+\begin{split}
 \mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) &= \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}, \text{ and } \\
 \mathbb{V}(Y_t| Y_{t-1} = y_{t-1}, X_t = i) &= (\sigma^{(i)})^2.
-\end{align*}
+\end{split}
+\end{align}
 %
+If $Y_t$ is a vector, equation (\ref{eqn:carhmm}) above can model either a subset of the dimensions of $Y_t$ if only a subset of those dimensions exhibit significant auto-correlation.
 Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.
@@
 \begin{equation*}
-    \calL_{\text{CarHMM}}(\Theta,\Gamma;y) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\Theta) \mathbf{1}_N
+    \calL_{\text{CarHMM}}(\theta,\Gamma;y) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\theta) \mathbf{1}_N
     \label{CarHMM_likelihood}
@@
 %
-$P(y_t|y_{t-1};\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
+$P(y_t|y_{t-1};\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
 
@@
 The distribution of $Y_{t,t^*}$ given $X_{t,t^*}=i^*$ and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
-We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\Theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^*)}\right)$.
+We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^*)}\right)$.
 Given the coarse-scale hidden states, the $T$ fine-scale HMMs $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are all assumed to be independent of one another.
@@
 %
-In addition, let $\Theta^* \equiv \{\Theta^{*(1)}, \ldots, \Theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices.
+In addition, let $\theta^* \equiv \{\theta^{*(1)}, \ldots, \theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices.
 The likelihood of the observed data is then
@@
 \[
-\calL_{\text{HHMM}}(\Theta,\Theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N
+\calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\theta,\theta^*,\Gamma^*) \mathbf{1}_N
 \]
 %
-where $P(y_t,y_t^*;\Theta,\Theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to 
-$f^{(i)}(y_t)\calL_{\text{HMM}}\left(\Theta^{*(i)},
+where $P(y_t,y_t^*;\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to 
+$f^{(i)}(y_t)\calL_{\text{HMM}}\left(\theta^{*(i)},
 \Gamma^{*(i)};y_t^*\right)$.
@@
 %
-for $k = 0, 1, \ldots, h-1$. Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
+for $k = 0, 1, \ldots, h-1$.
+If $Y^*_{t,t^*}$ is high-dimensional, then the DFT is taken component-wise.
+Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
 %
@@
     \label{eqn:z}
-    \Zone_{t,t^*} := \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n} \quad \text{and} \quad \Ztwo_{t,t^*} := \frac{1}{h}\sum_{k=1}^{\tilde{\omega}}\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|^2.
+    \Zone_{t,t^*} \equiv \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n} \quad \text{and} \quad \Ztwo_{t,t^*} \equiv \sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
 \end{equation}
@@
 In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared 2-norm of the component of $\left(Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\right)$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window (by the Plancherel theorem).
-More intuitively, $\Ztwo$ corresponds to the ``wiggliness" of the acceleration data within a window.
+More intuitively, $\Ztwo_{t,t^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within the window starting at time stamp $t^*$.
 The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert.
@@
 %\begin{equation}
-%    \calL_{\text{HMM-DFT}}(y,z^*;\Theta,\Gamma) = \delta P(y_1,z^*_1;\Theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\Theta) \mathbf{1}_N
+%    \calL_{\text{HMM-DFT}}(y,z^*;\theta,\Gamma) = \delta P(y_1,z^*_1;\theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\theta) \mathbf{1}_N
 %    \label{HMMDFT_likelihood}
 %\end{equation}
-%where $P(y_t,z^*_t;\Theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.
+%where $P(y_t,z^*_t;\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.
 
@@
 \[
-\calL_{\text{coarse}}(\Theta,\Theta^*,\Gamma;y,y^*) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) \mathbf{1}_N
+\calL_{\text{coarse}}(\theta,\theta^*,\Gamma;y,y^*) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\theta,\theta^*) \mathbf{1}_N
 \]
-where $P(y_t,y^*_t|y_{t-1};\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\Theta^{(i)}\right)\calL_{\text{HMM}}\left(\Theta^{*(i)};y^*_t\right)$.
+where $P(y_t,y^*_t|y_{t-1};\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\theta^{(i)}\right)\calL_{\text{HMM}}\left(\theta^{*(i)};y^*_t\right)$.
 On the fine scale, $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods (or penalized likelihoods).
-The likelihood $\calL_{\text{HMM}}$ is then replaced by the likelihood of this general fine-scale model, $\calL_{\text{fine}}(\Theta^{*(i)};y^*_t)$:
+The likelihood $\calL_{\text{HMM}}$ is then replaced by the likelihood of this general fine-scale model, $\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
 \[
-\calL_{\text{coarse}}(\Theta,\Theta^*,\Gamma;y,y^*) = \delta P(y_1,y^*_1;\Theta,\Theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\Theta,\Theta^*) \mathbf{1}_N
+\calL_{\text{coarse}}(\theta,\theta^*,\Gamma;y,y^*) = \delta P(y_1,y^*_1;\theta,\theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\theta,\theta^*) \mathbf{1}_N
 \]
-where $P(y_t,y^*_t;\Theta,\Theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\Theta^{(i)}\right)\calL_{\text{fine}}\left(\Theta^{*(i)};y^*_t\right)$.
+where $P(y_t,y^*_t;\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\theta^{(i)}\right)\calL_{\text{fine}}\left(\theta^{*(i)};y^*_t\right)$.
 

### 2020-08-31 Report/Sections/bibliography.tex

% !TeX root = ../main.tex

\begin{thebibliography}{}
% For English spelling, we follow the style of this dictionary
\bibitem[]{bib1}
Barber, K. (Ed.) (2004). {\it The Oxford Canadian Dictionary}, 2nd ed., 
Oxford University Press, Toronto.

% Single-author journal article
\bibitem[]{bib2}
Cleveland, W. (1979). Robust locally-weighted regression and smoothing scatterplots. 
{\it Journal of the American Statistical Association}, 74, 829-836.

% Genest provided the data for our sample table
\bibitem[]{bib3}
Genest, C. (1999). Probability and statistics: A tale of two worlds? 
{\it The Canadian Journal of Statistics}, 27(2), 421--444.

% Multiple-author journal article.  All the author names are separated by commas, and 
% note the uppercase letter after the colon in the title.
\bibitem[]{bib4}
Johnes, J., Taylor, J., \& Francis, B. (1993). The research
performance of \textsc{uk} universities: A statistical analysis of the results
of the 1989 research selectivity exercise. {\it Journal of the Royal
Statistical Society Series A}, 156, 271--286.

% Book in French.  Be careful with accents.
\bibitem[]{bib5}
Kidron, M. \& Segal, R. (1992). {\it Atlas du Nouvel \'Etat
du Monde}. \'Editions Autrement, Paris.

% Book, two authors, second edition
\bibitem[]{bib6}
Little, R.~J.~A. \& Rubin, D.~B. (2002). {\it Statistical Analysis with Missing Data}, 2nd ed., 
John Wiley \& Sons, New York.

% Unpublished paper with website link
\bibitem[]{bib7}
Molina, I. \& Rao, J.~N.~K. (2009). Small area estimation of poverty indicators. Working Paper 09-15, Statistics and Econometric Series 05, Universidad Carlos III de Madrid, \\http://hdl.handle.net/10016/5644.

% Edited book
\bibitem[]{bib8}
Roberts, G., Ren, Q., \& Rao, J.~N.~K. (2009). 
Using marginal mean models for data from longitudinal surveys
with a complex design: Some advances in methods. In Lynn, R. (Ed.) {\it Methodology of Longitudinal Surveys}, 
John Wiley \& Sons, New York, 351-366.

% !TeX root = main.tex

\bibitem[]{Adam:2019}
Adam, T., Griffiths, C., Leos Barajas, V., Meese, E., Lowe, C., Blackwell, P., Righton, D., \& Langrock, R. (2019).
Joint modelling of multi-scale animal movement data using hierarchical hidden Markov models.
{\it Methods in Ecology and Evolution}, 10(9), 1536–1550.

\iffalse

\bibitem[]{Ailliot:2006,
	author = {Pierre Ailliot},
	year = {2006},
	pages = {271-274},
	title = {Some theoretical results on markov-switching autoregressive models with gamma innovations},
	volume = {343},
	journal = {Comptes Rendus Mathematique},
}

@book{Bagchi:1999,
	author = {Bagchi, Sonali and Mitra, Sanjit},
	year = {1999},
	month = {01},
	pages = {},
	title = {The Nonuniform Discrete Fourier Transform},
	publisher = {Kluwer Academic Publishing},
	doi = {10.1007/978-1-4615-1229-5\_7}
}

\bibitem[]{Barajas:2017,
	author = {Leos-Barajas, Vianey and Gangloff, Eric and Adam, Timo and Langrock, Roland and van Beest, Floris and Nabe-Nielsen, Jacob and Morales, Juan},
	year = {2017},
	month = {02},
	title = {Multi-scale Modeling of Animal Movement and General Behavior Data Using Hidden Markov Models with Hierarchical Structures},
	journal = {Journal of Agricultural Biological and Environmental Statistics},
	doi = {10.1007/s13253-017-0282-9}
}

\bibitem[]{Bebbington:2007,
    author = {Bebbington, Mark},
    year = {2007},
    month = {08},
    pages = {921 - 942},
    title = {Identifying volcanic regimes using Hidden Markov Models},
    volume = {171},
    journal = {Geophysical Journal International},
    doi = {10.1111/j.1365-246X.2007.03559.x}
}

\bibitem[]{Borchers:2013,
    author = {Borchers, David and Zucchini, W and Heide-Jørgensen, Mads Peter and Cañadas, Ana and Langrock, R},
    year = {2013},
    month = {07},
    pages = {},
    title = {Using Hidden Markov Models to Deal with Availability Bias on Line Transect Surveys},
    journal = {Biometrics},
    doi = {10.1111/biom.12049}
}

\bibitem[]{Borger:2020,
    author = {Börger, Luca and Bijleveld, Allert and Fayet, Annette and Machovsky-Capuska, Gabriel and Patrick, Samantha and Street, Garrett and Wal, Eric},
    year = {2020},
    month = {01},
    pages = {6-15},
    title = {Biologging Special Feature},
    volume = {89},
    journal = {Journal of Animal Ecology},
    doi = {10.1111/1365-2656.13163}
}

\bibitem[]{Bulla:2012,
    author = {Bulla, Jan and Lagona, Francesco and Maruotti, Antonello and Picone, Marco},
    year = {2012},
    month = {12},
    pages = {544-567},
    title = {A Multivariate Hidden Markov Model for the Identification of Sea Regimes from Incomplete Skewed and Circular Time Series},
    volume = {17},
    journal = {Journal of Agricultural Biological and Environmental Statistics},
    doi = {10.1007/s13253-012-0110-1}
}

\bibitem[]{Bromback:1998,
    author = { Babette A.   Brumback  and  John A.   Rice },
    title = {Smoothing Spline Models for the Analysis of Nested and Crossed Samples of Curves},
    journal = {Journal of the American Statistical Association},
    volume = {93},
    number = {443},
    pages = {961-976},
    year  = {1998},
    publisher = {Taylor & Francis},
    doi = {10.1080/01621459.1998.10473755}
}

\bibitem[]{Cade:2017,
    author = {Cade, David and Barr, Kelly and Calambokidis, John and Friedlaender, Ari and Goldbogen, Jeremy},
    year = {2017},
    month = {11},
    pages = {jeb.170449},
    title = {Determining forward speed from accelerometer jiggle in aquatic nvironments},
    volume = {221},
    journal = {The Journal of Experimental Biology},
    doi = {10.1242/jeb.170449}
}

\bibitem[]{Chen:2012,
	Author = {Kehui Chen and Hans-Georg M{\"u}ller},
	Date-Modified = {2020-07-26 09:26:32 -0700},
	Issn = {01621459},
	Journal = {Journal of the American Statistical Association},
	Number = {500},
	Pages = {1599--1609},
	Publisher = {[American Statistical Association, Taylor & Francis, Ltd.]},
	Title = {Modeling Repeated Functional Observations},
	Url = {http://www.jstor.org/stable/23427358},
	Volume = {107},
	Year = {2012}
}

\bibitem[]{Crainiceanu:2009,
    author = {Crainiceanu, Ciprian and Staicu, Ana-Maria and Di, Chongzhi},
    year = {2009},
    month = {12},
    pages = {1550-1561},
    title = {Generalized Multilevel Functional Regression},
    volume = {104},
    journal = {Journal of the American Statistical Association},
    doi = {10.1198/jasa.2009.tm08564}
}

\bibitem[]{DeRuiter:2017,
    author = {DeRuiter, Stacy and Langrock, Roland and Skirbutas, Tomas and Goldbogen, Jeremy and Calambokidis, John and Friedlaender, Ari and Southall, Brandon},
    year = {2017},
    month = {03},
    pages = {362-392},
    title = {A multivariate mixed hidden Markov model for blue whale behaviour and responses to sound exposure},
    volume = {11},
    journal = {The Annals of Applied Statistics},
    doi = {10.1214/16-AOAS1008}
}

\bibitem[]{Di:2009,
	Author = {Di, Chong-Zhi and Crainiceanu, Ciprian M. and Caffo, Brian S. and Punjabi, Naresh M.},
	Doi = {10.1214/08-AOAS206},
	Fjournal = {Annals of Applied Statistics},
	Journal = {Ann. Appl. Stat.},
	Month = {03},
	Number = {1},
	Pages = {458--488},
	Publisher = {The Institute of Mathematical Statistics},
	Title = {Multilevel functional principal component analysis},
	Url = {https://doi.org/10.1214/08-AOAS206},
	Volume = {3},
	Year = {2009}
	}

\bibitem[]{Douc:2004,
	author = {Douc, Randal and Moulines, Eric and Rydén, Tobias},
	year = {2004},
	month = {10},
	pages = {2254-2304},
	title = {Asymptotic properties of the maximum likelihood estimator in autoregressive models with Markov regime},
	volume = {32},
	journal = {The Annals of Statistics},
	doi = {10.1214/009053604000000021}
}

\bibitem[]{Dot:2016,
    author = {Jeanniard du Dot, Tiphaine and Trites, Andrew and Arnould, John and Speakman, John and Guinet, Christophe},
    year = {2016},
    month = {10},
    pages = {},
    title = {Activity-specific metabolic rates for diving, transiting, and resting at sea can be estimated from time-activity budgets in free-ranging marine mammals},
    volume = {7},
    journal = {Ecology and Evolution},
    doi = {10.1002/ece3.2546}
}

\bibitem[]{Fehlmann:2017,
    author = {Fehlmann, Gaelle and O'Riain, J.M. and Hopkins, Phil and O'Sullivan, Jack and Holton, Mark and Shepard, Emily and King, Andrew},
    year = {2017},
    month = {12},
    pages = {},
    title = {Identification of behaviours from accelerometer data in a wild social primate},
    volume = {5},
    journal = {Animal Biotelemetry},
    doi = {10.1186/s40317-017-0121-3}
}

\bibitem[]{Fischer:1993,
  title={The Markov-Modulated Poisson Process (MMPP) Cookbook},
  author={Wolfgang Fischer and Kathleen S. Meier-Hellstern},
  journal={Perform. Evaluation},
  year={1993},
  volume={18},
  pages={149-171}
}

\bibitem[]{Ford:2006,
    author = {Ford, John and Ellis, Graeme},
    year = {2006},
    month = {07},
    pages = {185-199},
    title = {Selective foraging by fish-eating killer whales Orcinus orca in British Columbia},
    volume = {316},
    journal = {Marine Ecology-progress Series - MAR ECOL-PROGR SER},
    doi = {10.3354/meps316185}
}

@techreport{Ford:2015,
    author = {Ford, John et al.},
    year = {2015},
    month = {12},
    title = {Status review update for Pacific salmon and steelhead listed under the Endangered Species Act: Pacific Northwest},
    institution = {National Marine Fisheries Service, Northwest Fisheries Science Center},
    url = {http://www.westcoast.fisheries.noaa.gov/publications/status_reviews/salmon_steelhead/2016_status_review.html}
}

\bibitem[]{Fu:2017,
    author = {Fu, Eric and Heckman, Nancy},
    year = {2017},
    month = {12},
    pages = {},
    title = {Model-based curve registration via stochastic approximation EM algorithm},
    journal = {Computational Statistics and Data Analysis},
    doi = {10.1016/j.csda.2018.06.010}
}

\bibitem[]{Gao:2012,
	author = {Gao, Fuchang and Han, Lixing},
	year = {2012},
	month = {05},
	pages = {259-277},
	title = {Implementing the Nelder-Mead simplex algorithm with adaptive parameters},
	volume = {51},
	journal = {Computational Optimization and Applications},
	doi = {10.1007/s10589-010-9329-3}
}

@inproceedings{Getman:2009,
    author = {Getman, Anya and Cooper, Christopher and Key, Gary and Zhou, Heng and Frankle, Nick},
    year = {2009},
    month = {05},
    pages = {101 - 104},
    title = {Detection of mobile machine damage using accelerometer data and prognostic health monitoring techniques},
    doi = {10.1109/CIVVS.2009.4938730}
}

\bibitem[]{Gleiss:2011,
	author = {Gleiss, Adrian and Wilson, Rory and Shepard, Emily},
	year = {2011},
	month = {01},
	pages = {23-33},
	title = {Making overall dynamic body acceleration work: On the theory of acceleration as a proxy for energy expenditure},
	journal = {Methods in Ecology and Evolution},
	doi = {10.1111/j.2041-210X.2010.00057.x}
}

\bibitem[]{Green:2009,
	author = {Green, J. A. and Halsey, L. G. and Wilson, R. P. and Frappell, P. B.},
	title = {Estimating energy expenditure of animals using the accelerometry technique: activity, inactivity and comparison with the heart-rate technique},
	volume = {212},
	number = {5},
	pages = {745--746},
	year = {2009},
	doi = {10.1242/jeb.030049},
	publisher = {The Company of Biologists Ltd},
	issn = {0022-0949},
	URL = {https://jeb.biologists.org/content/212/5/745},
	eprint = {https://jeb.biologists.org/content/212/5/745.full.pdf},
	journal = {Journal of Experimental Biology}
}

\bibitem[]{Gruen:2017,
    author = {Gruen, Margaret and Alfaro Córdoba, Marcela and Thomson, Andrea and Worth, Alicia and Staicu, Ana-Maria and Lascelles, B Duncan X},
    year = {2017},
    month = {01},
    pages = {e0169576},
    title = {The Use of Functional Data Analysis to Evaluate Activity in a Spontaneous Model of Degenerative Joint Disease Associated Pain in Cats},
    volume = {12},
    journal = {PLOS ONE},
    doi = {10.1371/journal.pone.0169576}
}

\bibitem[]{Hastie:2006,
    author = {Hastie, Gordon and Rosen, David and Trites, Andrew},
    year = {2006},
    month = {09},
    pages = {163-170},
    title = {The influence of depth on a breath-hold diver: Predicting the diving metabolism of Steller sea lions (Eumetopias jubatus)},
    journal = {Journal of Experimental Marine Biology and Ecology},
    doi = {10.1016/j.jembe.2006.05.004}
}

\bibitem[]{Heerah:2017,
    author = {Heerah, Karine and Woillez, Mathieu and Fablet, Ronan and Garren, François and Martin, Stéphane and De Pontual, Hélène},
    year = {2017},
    month = {09},
    pages = {},
    title = {Coupling spectral analysis and hidden Markov models for the segmentation of behavioural patterns},
    volume = {5},
    journal = {Movement ecology},
    doi = {10.1186/s40462-017-0111-3}
}

\bibitem[]{Hooten:2017,
	author = {Hooten, Mevin and King, Ruth and Langrock, Roland},
	year = {2017},
	month = {08},
	pages = {},
	title = {Guest Editor’s Introduction to the Special Issue on “Animal Movement Modeling”},
	journal = {Journal of Agricultural, Biological and Environmental Statistics},
	doi = {10.1007/s13253-017-0299-0}
}

\bibitem[]{Isojunno:2017,
    author = {Isojunno, Saana and Sadykova, Dina and DeRuiter, S. and Curé, Charlotte and Visser, Fleur and Thomas, Len and Miller, P. and Harris, C.},
    year = {2017},
    month = {12},
    pages = {},
    title = {Individual, ecological, and anthropogenic influences on activity budgets of long‐finned pilot whales},
    volume = {8},
    journal = {Ecosphere},
    doi = {10.1002/ecs2.2044}
}

\bibitem[]{Juang:1991,
    author = { B. H.   Juang  and  L. R.   Rabiner },
    title = {Hidden Markov Models for Speech Recognition},
    journal = {Technometrics},
    volume = {33},
    number = {3},
    pages = {251-272},
    year  = {1991},
    publisher = {Taylor & Francis},
    doi = {10.1080/00401706.1991.10484833}
}

\bibitem[]{Johnson:2008,
	author = {Johnson, Devin and London, Josh and Lea, Mary-Anne and Durban, John},
	year = {2008},
	month = {06},
	pages = {1208-15},
	title = {Continuous-time correlated random walk model for animal telemetry data},
	volume = {89},
	journal = {Ecology},
	doi = {10.1890/07-1032.1}
}

@inbook{Kokoszka:2018,
	Author = {P Kokoszka and M Reimherr},
	Chapter = {8},
	Date-Added = {2020-07-26 09:36:01 -0700},
	Date-Modified = {2020-07-26 09:37:50 -0700},
	Publisher = {Chapman and Hall/CRC},
	Title = {Introduction to Functional Data Analysis},
	Year = {2018}
}

\bibitem[]{Langrock:2012,
	author = {Roland Langrock and Ruth King and Jason Matthiopoulos and Len Thomas and Daniel Fortin and Juan Morales},
	year = {2012},
	month = {11},
	pages = {2336-42},
	title = {Flexible and practical modeling of animal telemetry data: Hidden Markov models and extensions},
	volume = {93},
	journal = {Ecology},
	doi = {10.2307/41739305}
}

\bibitem[]{Langrock:2013,
    author = {Langrock, Roland and Swihart, Bruce and Caffo, Brian and Punjabi, Naresh and Crainiceanu, Ciprian},
    year = {2013},
    month = {08},
    pages = {},
    title = {Combining hidden Markov models for comparing the dynamics of multiple sleep electroencephalograms},
    volume = {32},
    journal = {Statistics in medicine},
    doi = {10.1002/sim.5747}
}

\bibitem[]{Langrock:2018,
	author = {R Langrock and T Adam and V Leos‐Barajas and S Mews S and DL Miller and YP Papastamatiou YP},
	year = {2018},
	month = {},
	pages = {179-200},
	volume = {72},
	title = {Spline‐based nonparametric inference in general state‐switching models},
	journal = {Statistica Neerlandica},
	doi = {10.1111/stan.12133} 
}

\bibitem[]{Lawler:2019,
	author = {Lawler, Ethan and Whoriskey, Kim and Aeberhard, William and Field, Chris and Mills Flemming, Joanna},
	year = {2019},
	month = {05},
	pages = {},
	title = {The Conditionally Autoregressive Hidden Markov Model (CarHMM): Inferring Behavioural States from Animal Tracking Data Exhibiting Conditional Autocorrelation},
	journal = {Journal of Agricultural, Biological and Environmental Statistics},
	doi = {10.1007/s13253-019-00366-2}
}

\bibitem[]{Lewin:2011,
	author = {Lewin-Koh, Nicholas and Bivand, Roger and Pebesma, Edzer and Archer, E and Baddeley, A and Bibiko, Hans-Jörg and Dray, Stéphane and Forrest, D and Friendly, M and Giraudoux, Patrick},
	year = {2011},
	month = {01},
	pages = {},
	title = {maptools: Tools for reading and handling spatial objects. R package version 0.8-27}
}

\bibitem[]{Liu:2015,
    author = {Liu, Yu-Ying and Li, Shuang and Li, Fuxin and Song, Le and Rehg, James},
    year = {2015},
    month = {01},
    pages = {3599-3607},
    title = {Efficient Learning of Continuous-Time Hidden Markov Models for Disease Progression},
    volume = {28},
    journal = {Advances in neural information processing systems}
}

@inproceedings{Lucero:2019,  
    author={P. {Lucero} and R. {Sánchez} and J. {Macancela} and D. {Cabrera} and M. {Cerrada} and C. {Li} and H. R. {Alonso}},  booktitle={2019 Prognostics and System Health Management Conference (PHM-Paris)},  
    title={Accelerometer Placement Comparison for Crack Detection in Railway Axles Using Vibration Signals and Machine Learning},   
    year={2019},  
    volume={},  
    number={},  
    pages={291-296}
}

\bibitem[]{McClintock:2017,
	author = {McClintock, Brett},
	year = {2017},
	month = {06},
	pages = {1-21},
	title = {Incorporating Telemetry Error into Hidden Markov Models of Animal Movement Using Multiple Imputation},
	journal = {Journal of Agricultural, Biological, and Environmental Statistics},
	doi = {10.1007/s13253-017-0285-6}
}

\bibitem[]{McClintock:2020,
    author = {McClintock, Brett and Langrock, Roland and Gimenez, Olivier and Cam, Emmanuelle and Borchers, David and Glennie, Richard and Patterson, Toby},
    year = {2020},
    month = {02},
    pages = {},
    title = {Uncovering ecological state dynamics with hidden Markov models},
    journal = {Preprint}
}

\bibitem[]{Michelot:2019,
	author = {Michelot, Th\`eo and Blackwell, Paul},
	year = {2019},
	month = {01},
	pages = {},
	title = {State-switching continuous-time correlated random walks},
	journal = {Methods in Ecology and Evolution},
	doi = {10.1111/2041-210X.13154}
}

\bibitem[]{Morales:2004,
	author = {Morales, Juan and Haydon, Daniel and Frair, Jacqui and Holsinger, Kent and Fryxell, John},
	year = {2004},
	month = {09},
	pages = {},
	title = {Extracting More out of Relocation Data: Building Movement Models as Mixtures of Random Walks},
	volume = {85},
	journal = {EEB Articles},
	doi = {10.1890/03-0269}
}

\bibitem[]{Morris:2007,
	Author = {Jeffrey S Morris and Cassandra Arroyo and Brent A Coull and Louise M Ryan and Richard Herrick and Steven L Gortmaker},
	Doi = {10.1198/016214506000000465},
	Journal = {Journal of the American Statistical Association},
	Number = {476},
	Pages = {1352-1364},
	Publisher = {Taylor & Francis},
	Title = {Using Wavelet-Based Functional Mixed Models to Characterize Population Heterogeneity in Accelerometer Profiles},
	Volume = {101},
	Year = {2006}
}

\bibitem[]{Noren:2011,
    author = {Noren, Dawn},
    year = {2011},
    month = {01},
    pages = {60 - 77},
    title = {Estimated field metabolic rates and prey requirements of resident killer whales},
    volume = {27},
    journal = {Marine Mammal Science},
    doi = {10.1111/j.1748-7692.2010.00386.x}
}

@misc{Nunes:2018,
	author = {Alex Nunes},
	title = {Divebomb},
	url = {https://github.com/ocean-tracking-network/divebomb/blob/master/docs/index.rst},
	year = {2018},
	month = {07}
}

\bibitem[]{Patterson:2017,
	author = {Patterson, Toby and Parton, Alison and Langrock, Roland and Blackwell, Paul and Thomas, Len and King, Ruth},
	year = {2017},
	month = {07},
	pages = {},
	title = {Statistical modelling of individual animal movement: an overview of key methods and a discussion of practical challenges},
	journal = {AStA Advances in Statistical Analysis},
	doi = {10.1007/s10182-017-0302-7}
}

\bibitem[]{Pohle:2017,
	author = {Pohle, Jennifer and Langrock, Roland and van Beest, Floris and Schmidt, Niels},
	year = {2017},
	month = {06},
	pages = {1-24},
	title = {Selecting the Number of States in Hidden Markov Models: Pragmatic Solutions Illustrated Using Animal Movement},
	volume = {22},
	journal = {Journal of Agricultural, Biological, and Environmental Statistics},
	doi = {10.1007/s13253-017-0283-8}
}

\bibitem[]{Qasem:2012,
    author = {Qasem, Lama and Cardew, Antonia and Wilson, Alexis and Griffiths, Iwan and Halsey, Lewis and Shepard, Emily and Gleiss, Adrian and Wilson, Rory},
    year = {2012},
    month = {02},
    pages = {e31187},
    title = {Tri-Axial Dynamic Acceleration as a Proxy for Animal Energy Expenditure; Should We Be Summing Values or Calculating the Vector?},
    volume = {7},
    journal = {PloS one},
    doi = {10.1371/journal.pone.0031187}
}

@book{Ramsay:2005,
	Author = {J O Ramsay and Bernard Silverman},
	Date-Added = {2020-07-26 09:30:43 -0700},
	Date-Modified = {2020-07-26 09:32:18 -0700},
	Publisher = {Springer},
	Title = {Functional Data Analysis},
	Year = {2005}}

\bibitem[]{Rice:2001,
    author = {Rice, John and Wu, Colin},
    year = {2001},
    month = {04},
    pages = {253-9},
    title = {Nonparametric Mixed Effects Models for Unequally Sampled Noisy Curves},
    volume = {57},
    journal = {Biometrics},
    doi = {10.1111/j.0006-341X.2001.00253.x}
}

@book{Serfozo:2009,
	title = {Basics of Applied Stochastic Processes},
	author = {Serfozo, Richard},
	year = {2009},
	publisher = {Springer-Verlag Berlin Heidelberg},
	doi = {10.1007/978-3-540-89332-5},
	ISSN = {1431-7028}
}

\bibitem[]{Shorter:2017,
    author = {Shorter, Kenneth and Shao, Yunli and Ojeda, Lauro and Barton, Kira and Rocho-Levine, Julie and Van der Hoop, Julie and Moore, Michael},
    year = {2017},
    month = {06},
    pages = {},
    title = {A day in the life of a dolphin: Using bio-logging tags for improved animal health and well-being},
    journal = {Marine Mammal Science},
    doi = {10.1111/mms.12408}
}

\bibitem[]{Simon:2012,
    author = {Simon, Malene and Johnson, Mark and Madsen, Peter},
    year = {2012},
    month = {11},
    pages = {3786-98},
    title = {Keeping momentum with a mouthful of water: Behavior and kinematics of humpback whale lunge feeding},
    volume = {215},
    journal = {The Journal of experimental biology},
    doi = {10.1242/jeb.071092}
}

\bibitem[]{Sutherland:1998,
    author = {Sutherland, William},
    year = {1998},
    month = {11},
    pages = {801-809},
    title = {The importance of behavioral studies in conservation biology},
    volume = {56},
    journal = {Animal Behaviour},
    doi = {10.1006/anbe.1998.0896}
}

\bibitem[]{Tennessen:2019a,
    author = {Tennessen, Jennifer and Holt, Marla and Hanson, M. and Emmons, Candice and Giles, Deborah and Hogan, Jeffrey},
    year = {2019},
    month = {02},
    pages = {},
    title = {Kinematic signatures of prey capture from archival tags reveal sex differences in killer whale foraging activity},
    volume = {222},
    journal = {The Journal of Experimental Biology},
    doi = {10.1242/jeb.191874}
}

\bibitem[]{Tennessen:2019b,
    author = {Tennessen, Jennifer and Holt, Marla and Ward, Eric and Hanson, M. and Emmons, Candice and Giles, Deborah and Hogan, Jeffrey},
    year = {2019},
    month = {10},
    pages = {14951},
    title = {Hidden Markov models reveal temporal patterns and sex differences in killer whale behavior},
    volume = {9},
    journal = {Scientific Reports},
    doi = {10.1038/s41598-019-50942-2}
}

\bibitem[]{Virtanen:2019,
	author = {{Virtanen}, Pauli and {Gommers}, Ralf and {Oliphant},
	Travis E.  and {Haberland}, Matt and {Reddy}, Tyler and
	{Cournapeau}, David and {Burovski}, Evgeni and {Peterson}, Pearu and {Weckesser}, Warren and {Bright}, Jonathan and {van der Walt},
	St{\'e}fan J.  and {Brett}, Matthew and {Wilson}, Joshua and
	{Jarrod Millman}, K.  and {Mayorov}, Nikolay and {Nelson}, Andrew
	R.~J. and {Jones}, Eric and {Kern}, Robert and {Larson}, Eric and
	{Carey}, CJ and {Polat}, {\.I}lhan and {Feng}, Yu and {Moore},
	Eric W. and {Vand erPlas}, Jake and {Laxalde}, Denis and
	{Perktold}, Josef and {Cimrman}, Robert and {Henriksen}, Ian and
	{Quintero}, E.~A. and {Harris}, Charles R and {Archibald}, Anne M.
	and {Ribeiro}, Ant{\^o}nio H. and {Pedregosa}, Fabian and
	{van Mulbregt}, Paul and {Contributors}, SciPy 1. 0},
	title = "{SciPy 1.0--Fundamental Algorithms for Scientific Computing in Python}",
	journal = {arXiv e-prints},
	year = "2019",
	month = "Jul",
	eid = {arXiv:1907.10121},
	pages = {arXiv:1907.10121},
	archivePrefix = {arXiv},
	eprint = {1907.10121},
	primaryClass = {cs.MS},
	adsurl = {https://ui.adsabs.harvard.edu/abs/2019arXiv190710121V},
	adsnote = {Provided by the SAO/NASA Astrophysics Data System}
}

\bibitem[]{Viterbi:1967,
	author = {Viterbi, A.},
	year = {1967},
	month = {April},
	pages = {260-269},
	title = {Error bounds for convolutional codes and an asymptotically optimum decoding algorithm},
	journal = {IEEE Transactions on Information Theory},
	doi = {10.1109/TIT.1967.1054010}
}

\bibitem[]{Whoriskey:2016,
	author = {Whoriskey, Kim and Auger-M\'eth\'e, Marie and Albertsen, Christoffer and Whoriskey, F. and Binder, Thomas and Krueger, Charles and Flemming, Joanna},
	year = {2016},
	month = {12},
	pages = {},
	title = {A Hidden Markov Movement Model for rapidly identifying behavioral states from animal tracks},
	volume = {7},
	journal = {Ecology and Evolution},
	doi = {10.1002/ece3.2795}
}

\bibitem[]{Williams:1999,
    author = {Williams, Terrie and Haun, J and Friedl, WA},
    year = {1999},
    month = {11},
    pages = {2739-48},
    title = {The diving physiology of bottlenose dolphins (Tursiops truncatus): I. Balancing the demands of exercise for energy conservation at depth},
    volume = {202},
    journal = {The Journal of experimental biology}
}

\bibitem[]{Williams:2009,
    author = {Williams, Rob and Noren, Dawn},
    year = {2009},
    month = {04},
    pages = {327 - 350},
    title = {Swimming speed, respiration rate, and estimated cost of transport in adult killer whales},
    volume = {25},
    journal = {Marine Mammal Science},
    doi = {10.1111/j.1748-7692.2008.00255.x}
}

\bibitem[]{Wilson:2019,
    author = {Wilson, Rory and Börger, Luca and Holton, Mark and Scantlebury, D. and Laich, Agustina and Quintana, Flavio and Rosell, Frank and Graf, Patricia and Williams, Hannah and Gunner, Richard and Hopkins, Lloyd and Marks, Nikki and Geraldi, Nathan and Duarte, Carlos and Scott, Rebecca and Strano, Michael and Robotka, Hermina and Eizaguirre, Christophe and Fahlman, Andreas and Shepard, Emily},
    year = {2019},
    month = {06},
    pages = {},
    title = {Estimates for energy expenditure in free-living animals using acceleration proxies: A reappraisal},
    volume = {89},
    journal = {Journal of Animal Ecology},
    doi = {10.1111/1365-2656.13040}
}

\bibitem[]{Wright:2017,
    author = {Wright, Brianna and Ford, John and Ellis, Graeme and Deecke, Volker and Shapiro, Ari and Battaile, Brian and Trites, Andrew},
    year = {2017},
    month = {12},
    pages = {},
    title = {Fine-scale foraging movements by fish-eating killer whales (Orcinus orca) relate to the vertical distributions and escape responses of salmonid prey (Oncorhynchus spp.)},
    volume = {5},
    journal = {Movement Ecology},
    doi = {10.1186/s40462-017-0094-0}
}

\bibitem[]{Xiao:2014,
    author = {Xiao, Luo and Huang, Lei and Schrack, Jennifer and Ferrucci, Luigi and Zipunnikov, Vadim and Crainiceanu, Ciprian},
    year = {2014},
    month = {10},
    pages = {},
    title = {Quantifying the lifetime circadian rhythm of physical activity: A covariate-dependent functional approach},
    volume = {16},
    journal = {Biostatistics (Oxford, England)},
    doi = {10.1093/biostatistics/kxu045}
}

\bibitem[]{Xu:2018,
    author = {Xu, Zekun and Laber, Eric and Staicu, Ana-Maria},
    year = {2018},
    month = {12},
    pages = {},
    title = {Hierarchical Continuous Time Hidden Markov Model, with Application in Zero-Inflated Accelerometer Data},
    journal = {Preprint}
}

\bibitem[]{Yao:2005,
    author = {Yao, Fang and Muller, Hans-Georg and Wang, Jane-Ling},
    year = {2005},
    month = {06},
    pages = {577-590},
    title = {Functional Data Analysis for Sparse Longitudinal Data},
    volume = {100},
    journal = {Journal of the American Statistical Association},
    doi = {10.1198/016214504000001745}
}

@book{Zucchini:2016,
	author = {Zucchini, Walter and Macdonald, Iain and Langrock, Roland},
	year = {2016},
	month = {06},
	publisher = {CRC Press},
	title = {Hidden Markov Models for Time Series - An Introduction Using R},
	isbn = {9781482253832}
}

\fi

\end{thebibliography}

### 2020-08-31 Report/Sections/discussion.tex

@@
 For periodic behaviour in particular, we use a DFT-based transformation.
-The effectiveness of this approach is evidenced by both our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale, and our simulation study, where the subdive decoding accuracy is by far the worst when omitting $\Ztwo$ (``wiggliness'') from the model.
+The effectiveness of this approach is evidenced by both our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale, and our simulation study, where the subdive decoding accuracy is by far the worst when omitting $\Ztwo_{t,t^*}$ (``wiggliness'') from the model.
 For simpler transformations, e.g.
-an average such as $\Zone$, either subject area experts or exploratory plots may suggest temporal correlation.
+an average such as $\Zone_{t,t^*}$, either subject area experts or exploratory plots may suggest temporal correlation.
 In this case,  we suggest using a CarHMM instead of a traditional HMM.

### 2020-08-31 Report/Sections/results.tex
Commit Before Andrew Access
Updates from Overleaf

--- 2020-08-06 Report/Sections/results.tex
@@
 
-%\section{Killer Whale Case Study}
+%\section{Data Collection and Model Construction}
 
-%In order to test the advantages of the CarHHM-DFT over the other models on real-world data, we used the four models described in the simulation study to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada.
+To illustrate the process of constructing a model using these HMM building blocks, we investigate the dive behaviour of a northern resident killer whale (NRKW) off the coast of British Columbia, Canada, and construct several candidate models to describe the associated data.
 
-%\section{Results}
+Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}.
+HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g.
+\citet{DeRuiter:2017} and \citet{Isojunno:2017}).
+For NRKWs, we are interested in categorizing different diving behaviours, with a particular focus on differentiating potential foraging dives.
+NRKWs prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKWs for predation \citep{Williams:2009,Noren:2011}.
+It is therefore of interest to determine the energetic expenditure of NRKWs to forage for Chinook, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
+Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the behavioural state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
+Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is important to understand the true energetic requirements of a NRKW.
 
-We fit all four models from the simulation study to the data shown in Figure \ref{fig:data} to test their performance in a real-world setting.
-We first report the results from the full CarHHMM-DFT (defined in Section \ref{sec:data}) in detail and then asses the quality of the fit.
-We also compare the results of the CarHHMM-DFT with those of the other candidate models.
+\subsection{Data collection and preprocessing}
 
-The emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}). 
-Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, traveling, and searching for prey.
+The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions.
+Observations were collected at a rate of 50 Hz using
+% an accelerometer and time-depth recorder.
+[NAME OF TAGS - do you know Sarah?].
+The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left).
+Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
+Tagging the killer whale caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored.
+We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm, when the tagging technology malfunctioned.
+We smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second.
+We then define a killer whale ``dive" as any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
+Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}.
+The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}.
+Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
+Specifically, the vector of observed coarse-scale observations $y = \Big\{y_1,\ldots,y_{267}\Big\}$ is a time series of dive durations in seconds ($s$) with corresponding vector of unobserved dive types $x = \Big\{x_1,\ldots,x_{267}\Big\}$.
+For dive $t$, the fine-scale observations are contained in $y^*_{t}$, which is a vector of the within-dive acceleration data in meters per second squared ($m/s^2$).
+The corresponding vector of subdive states is $x^*_{t}$.
+Naturally, $y^* = \Big\{y^*_1,\ldots,y^*_{267}\Big\}$ and $x^* = \Big\{x^*_1,\ldots,x^*_{267}\Big\}$ is the corresponding set of subdive states.
+
+\subsection{Model definition and selection}
+
+Building a suitable model for these data involves selecting appropriate model structures, emission distributions, and numbers of states on both the coarse and fine scales.
+
+To choose how to model the coarse-scale observations (dive duration), we begin by selecting the model structure.
+Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations.
+Visual inspection also shows no obvious intricate structure.
+Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor transformation is called for.
+%
+Next, we select the emission distribution of dive duration conditioned on dive type. Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows
+a gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:
+%
+$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
+%
+This is consistent with previous studies, including \citet{Barajas:2017}. 
+%
+Finally, we select the number of dive types for the coarse scale.
+We do not use information criteria since these metrics tend to overestimate the number of states in biological processes \citep{Pohle:2017}.
+If the hidden states are well-separated, the lag plot should reveal $N$ distinct patterns where each pattern corresponds to one dive type \citep{Lawler:2019}.
+This is unfortunately not the case for our killer whale data, but we still consult lag plots together with a heuristic approach based on visual inspection to choose $N = 2$ dive types.
+The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
+
+To choose how to model the fine-scale observations (acceleration), we again begin with determining model structure.
+In contrast to the coarse-scale, the fine-scale acceleration data exhibit significant sinusoidal behaviour.
+Thus, we transform each fine-scale observation sequence $y_t^*$ into $z_t^*$ using Equation (\ref{eqn:z}).
+We select a window size of two seconds since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data.
+%
+We calculate $\Ztwo_{t,t^*}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to maximum frequencies of 5 Hz for each axis of acceleration.
+We also down-sample each observation $z^*_t$ such that $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\Big\}$ to avoid overlapping windows.
+The resulting sequence of observations is $z^* = \Big\{z^*_1,\ldots,z^*_{267}\Big\}$, where $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},\ldots\Big\}$ and $z^*_{t,t^*} = \{a^*_{t,t^*},w^*_{t,t^*}\}$.
+We have that $a^*_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $w^*_{t,t^*}$ is a scalar describing the ``wiggliness" of a given window.
+%
+After transforming the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (see section 2.1 of the supplementary material).
+Therefore, we assume that all three components of $\Zone_{t,t^*}$, $t^* = 1,101,\ldots$ follow the CarHMM as defined in section 2.2.
+%
+%$$\mathbb{E}(\Zone_{t,t^*}|\\Zone_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_A^{*(i)} \mathbf{z} + (1-\phi_A^{*(i)}) \mathbf{\mu}_A^{*(i)},$$
+%$$\mathbb{V}(\Zone_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_A^{*(i)}\right)^2\right].$$
+%
+While the $\Ztwo_{t,t^*}$ observations also exhibit some auto-correlation prior to fitting the model, it is less strong than that of $\Zone_{t,t^*}$, and 
+%the biological interpretation of auto-correlation between observations of $\Ztwo_{t,t^*}$ is less clear. In addition, 
+much of the auto-correlation evident from the lag plot may be explained by subsequent observations having the same subdive state.
+Therefore, a simple HMM structure is used to model $\Ztwo_{t,t^*}$ for all $t$ and $t^*$.
+%
+%$$\mathbb{E}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_W^{*(i)}$$
+%$$\mathbb{V}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_W^{*(i)}\right)^2.$$
+%
+
+Next, we select the emission distributions for both $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$.
+We assume that the fine-scale emission parameters are shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity.
+This implies that the subdive states in dive type 1 have the same interpretation as those in dive type 2.
+%
+For $\Zone_{t,t^*}$, we assume that each component is Normally distributed with mean and variance defined as in Equation (\ref{eqn:carhmm}).
+Given that $X^*_{t,t^*} = i^*$, the corresponding unknown parameters are $\phi_A^{*(\cdot,i^*)} \in \mathbb{R}$, $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$.
+In other words, each axis of acceleration has its own mean and variance, but all axes have the same auto-correlation parameter.
+We assume that observations of all three axes of acceleration are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$.
+%
+On the other hand, we assume that $\Ztwo_{t,t^*}$ is gamma distributed for all $t$ and $t^*$.
+Further, given that $X^*_{t,t^*} = i^*$, the distribution of $\Ztwo_{t,t^*}$ is parameterized by its mean $\mu_W^{*(\cdot,i^*)} \in \mathbb{R}$ and standard deviation $\sigma_W^{*(\cdot,i^*)} \in \mathbb{R}$.
+We assume that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$.
+
+Finally, to select the number of subdive states $N^*$ we again rely on a heuristic approach based on visual inspection to select $N^* = 3$.
+Although $N^*$ is selected heuristically, we test the validity of this model using techniques in section \ref{subsec:model_validation}.
+
+The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale.
+On the fine scale, each fine-scale sequence $y^*_t$ is transformed to $z^*$ using Equation (\ref{eqn:z}), $\Zone_{t,t^*}$ is modelled with a CarHMM, and $\Ztwo_{t,t^*}$ is modelled with an HMM for all $t$ and $t^*$.
+We assume that $\Ztwo_{t,t^*}$ in addition to the components of $\Zone_{t,t^*}$ are all independent of one another and that the subdive types have identical emission distributions across all dive types.
+In total, the parameters to estimate are
+%
+\begin{gather*}
+    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
+    %
+    \theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{($Y$ emission parameters), and} \\
+    %
+    \theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{(fine-scale emission parameters), where} \\
+    %
+    \theta^{*(\cdot,i^*)} =  \{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)},\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}.
+\end{gather*}
+%
+Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $Z_{t,t^*}$ when conditioned on $X^*_{t,t^*} = i^*$.
+
+We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMMs discussed so far.
+Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model for the CarHHMM-DFT.
+The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
+See the appendix for details of likelihood evaluation.
+In addition to the CarHHMM-DFT, we consider three similar variations for comparison:
+\begin{enumerate}
+    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Zone_{t,t^*}$ and $\Ztwo{t,t^*}$ as emissions of a simple HMM rather than a CarHMM.
+    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emssions of a CarHMM.
+However, the Fourier sums $\Ztwo_{t,t^*}$ are omitted from this model altogether.
+The CarHHMM therefore does not incorporate the ``wiggliness" of the accelerometer data.
+    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emissions of a CarHMM and $\Ztwo_{t,t^*}$ as emissions an HMM.
+This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive.
+\end{enumerate}
+%
+Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes no auto-correlation within the fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.
+
+\subsection{Case study results}
+
+To illustrate an application of this method and compare the candidate models, we fit all four models to the data shown in Figure \ref{fig:data}.
+We first report the results from the full CarHHMM-DFT in detail and then assess the quality of the fit.
+We then compare the results of the CarHHMM-DFT with those of the other candidate models.
+
+The coarse-scale emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}).
+Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, travelling, and searching for prey.
 On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as foraging and socialization \citep{Tennessen:2019b}.
-No dive in this data set has a maximum depth deeper than $\approx$ 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100m.
+However, no dive in this data set has a maximum depth deeper than $\approx$ 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100m.
 It is therefore unlikely that any of the dives used in this study are successful foraging dives.
 
-The three subdive behaviours have distinct average ``wiggliness" values, as the means of $\Ztwo$ for each state are separated by an order of magnitude (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}).
-Subdive state 1 has the smallest mean of $\Ztwo$ and the smallest variance corresponding to $\Zone$.
-It also has the highest auto-correlation in $\Zone$.
+The three subdive behaviours have distinct average ``wiggliness" values, as the means corresponding to $\Ztwo_{t,t^*}$ are separated by an order of magnitude for each subdive state (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}).
+Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,t^*}$ and the smallest variance corresponding to $\Zone_{t,t^*}$.
+It also has the highest auto-correlation in $\Zone_{t,t^*}$.
 This implies less overall activity and more consistent acceleration compared to the other subdive states.
-Subdive state 2 has a mean ``wiggliness" ($\Ztwo$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
-The auto-correlation of acceleration is also slightly lower in than subdive state 1.
+Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,t^*}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
+The auto-correlation of acceleration is also slightly lower than in subdive state 1.
 We therefore hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
-Finally, the mean of $\Ztwo$ and variance of $\Zone$ in subdive state 3 is much higher than the other two states.
-The auto-correlation of $\Zone$ is also much lower, implying more variation in acceleration between segments.
+Finally, the mean of $\Ztwo_{t,t^*}$ and variance of $\Zone_{t,t^*}$ in subdive state 3 are both much higher than in the other two states.
+The auto-correlation of $\Zone_{t,t^*}$ is also much lower.
 This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).
@@
 for dive types 1 and 2.
-The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that most dives are short, type 1 dives.
-The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale usually performs multiple short type 1 dives before doing only one (or a few) long type 2 dive.
+The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that about 79\% of dives are short dives of type 1.
+The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1.
 This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
-The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is more likely to be in the less active subdive state 1 when performing long deep dives than when performing short shallow dives.
-Using less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}.
+The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 much more often when performing long deep dives compared to short shallow dives (40\% of the time for dive type 1 and 9\% of the time for dive type 2).
+Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}.
 Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives.
-The supplementary material also shows the estimated probability of each dive and subdive state given the data (Section S-2.2).
+Section 2.2 of the supplementary material also shows the estimated probability of each dive and subdive state given the data.
 
-\subsection{Model Validation}
+\subsection{Model validation}
 \label{subsec:model_validation}
 
-Two visual tools are used to evaluate this model: pseudo-residuals and empirical histograms.
-A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}.
+We use two visual tools to evaluate the CarHHMM-DFT: pseudo-residual plots and empirical histograms. 
+%A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}.
 To easily visualize outliers, pseudo-residuals are often passed through the quantile function of the standard Normal distribution.
-Mathematically, the pseudo-residual of an observation $y_t$ of a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
+Mathematically, the pseudo-residual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
 If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution.
 Histograms of the pseudoresiduals mostly support that this model is well-specified.
-One exception is $\Ztwo$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
-This implies that the true distribution of $\Ztwo$ may follow a heavier-tailed distribution than the gamma distribution used in the case study.
+One exception is $\Ztwo_{t,t^*}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
+This implies that the true distribution of $\Ztwo_{t,t^*}$ may follow a heavier-tailed distribution than the gamma distribution used in the case study.
+See section 2 of the supplementary material for pseudo-residual plots for all observations.
 
-We also plot histograms of $Y$ and $Z^*$ corresponding to each dive type or subdive state in Figure \ref{fig:empirical_dist}.
-Observations are weighted by the decoded probability that they correspond to a particular hidden state, and this histogram is plotted over the emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}).
-Our results again mostly support a well-specified model with the exception of $\Ztwo$, which is again right-skewed.
-Another exception is $\Zone$, which has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
+We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}.
+Each observed dive duration is weighted by the decoded probability that it corresponds to a particular dive type.
+This results in two histograms - one corresponding to dive type 1 and another corresponding to dive type 2.
+Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}).
+See section 2 of the supplementary material for analogous histograms corresponding to the fine-scale observations $z^*_{t,t^*}$.
+Our results again mostly show that the CarHHMM-DFT fits the data well, but there are some exceptions.
+In particular, $\Ztwo_{t,t^*}$ is right-skewed and $\Zone_{t,t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
 These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.
-See Section S-2.?
-for histograms for every observation and hidden state.
 
@@
 
-The CarHMM-DFT produces very similar parameter estimates and state estimates compared to those of the CarHHMM-DFT on the fine scale.
-However, its lack of a hierarchical structure means that it fails to differentiate between short and long dives.
+The CarHMM-DFT, which lacks a hierarchical structure, produces similar fine-scale parameter estimates and state estimates compared to those of the CarHHMM-DFT However, its lack of a hierarchical structure means that it fails to differentiate between short and long dives.
 This model therefore does not give any information regarding the dive-level Markov chain or the relationship between the dive and subdive levels.
@@
 
-The HHMM-DFT decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3.
-In addition, the HHMM-DFT produces estimates of $\sigma_A^{*(1)}$, $\sigma_A^{*(2)}$, and $\sigma_A^{*(3)}$ (the standard deviation of acceleration) which are approximately 50-100 \% larger than those of the CarHHMM-DFT across all axes of acceleration.
-The estimated uncertainties of the estimates of $\mu_A^{*(1)}$, $\mu_A^{*(2)}$, and $\mu_A^{*(3)}$ are also less than half of those for the CarHHMM-DFT across all axes of acceleration.
-These findings are consistent with the tendency to overestimate variance and underestimate standard error when ignoring auto-correlation.
-Further, the empirical auto-correlation of $\Zone$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers.
-Both of these findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.
+The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3.
+In addition, the HHMM-DFT produces estimates of $\sigma_A^{*(1)}$, $\sigma_A^{*(2)}$, and $\sigma_A^{*(3)}$ which are approximately $50$ to $100$ percent larger than those of the CarHHMM-DFT for all axes of acceleration.
+The estimated uncertainties of the estimates of $\mu_A^{*(1)}$, $\mu_A^{*(2)}$, and $\mu_A^{*(3)}$ are also less than half of those for the CarHHMM-DFT for all axes of acceleration.
+Further, the empirical auto-correlation of acceleration within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers.
+These findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.
 
-Finally, the CarHHMM does not use the ``wiggliness" of the acceleration data as an observation, so it fails produce biologically interpretable results when decoding the subdive states.
-This fact alone essentially disqualifies the CarHHMM as a viable model for this data.
+Finally, the CarHHMM does not use the ``wiggliness" of the acceleration data as an observation, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives}.
+This fact alone essentially disqualifies the CarHHMM as a viable model for these data.
 The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution.
-See section S-2 for a more complete set of results for each of the candidate models.
+See section 2 of the supplementary material for a more complete set of results for each of the candidate models.

### 2020-08-31 Report/Sections/simulation.tex
Commit Before Andrew Access
Updates from Overleaf

@@
 This allows meaningful conclusions to be made for both dive types.
-In addition, the number subdive states is set to $N^*=2$ and $\Ztwo$ is set to a scalar.
+In addition, the number subdive states is set to $N^*=2$ and $\Ztwo_{t,t^*}$ is set to a scalar.
 Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times.
@@
 
-\subsection{Data Simulation}
+\subsection{Data simulation}
 
@@
 %
-Recall that $\Zone$ follows a CarHMM and is Normally distributed while $\Ztwo$ follows an HMM and is Gamma distributed.
+Recall that $\Zone_{t,t^*}$ follows a CarHMM and is Normally distributed while $\Ztwo_{t,t^*}$ follows an HMM and is Gamma distributed.
 It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix.
@@
 
-\subsection{Simulation Results}
+\subsection{Simulation results}
 

### 2020-08-31 Report/Sections/appendix.tex
Updates from Overleaf

@@
     $$
-    	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(\sigma_A^{*(i^*)}\right)^2 \right)
+    	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
     $$
-    where $\sigma_A^{*(1)} = 0.05s$ and $\sigma_A^{*(2)} = 0.1s$. 
+    where $\sigma_A^{*(\cdot,1)} = 0.05s$ and $\sigma_A^{*(\cdot,2)} = 0.1s$. 
     %
@@
     \begin{equation}	
-        \Big(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*,\hat{Y}^{*(0)}_{t,t^*-100}\Big) \sim \mathcal{N} \left(\phi_A^{*(i^*)} \hat{Y}^{*(0)}_{t,t^*-100}, \left(\sigma_A^{*(i^*)}\right)^2 \right), \quad t^* = 101,201,\ldots
+        \Big(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*,\hat{Y}^{*(0)}_{t,t^*-100}\Big) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,t^*-100}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \quad t^* = 101,201,\ldots
     	\label{eqn:yhat_0}
     \end{equation}
-    where $\phi_A^{*(1)} = 0.99$ and $\phi_A^{*(2)} = 0.95$. Note that $t^*$ has increments of 100 because $h=100$ when using two-second window sizes.
+    where $\phi_A^{*(\cdot,1)} = 0.99$ and $\phi_A^{*(\cdot,2)} = 0.95$. Note that $t^*$ has increments of 100 because $h=100$ when using two-second window sizes.
     %
@@
 %
-$$\left(\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(\phi_A^{*(i^*)} \Zone_{t,t^*-100}, \left(\sigma_A^{*(i^*)}\right)^2 \right).$$
+$$\left(\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \Zone_{t,t^*-100}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right).$$
 %
-This distribution is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(i^*)} = 0$.
-We also have that $\sigma_A^{*(1)} = 0.05s$, $\phi_A^{*(1)} = 0.99$, $\sigma_A^{*(2)} = 0.1s$, and $\phi_A^{*(2)} = 0.95$.
+This distribution is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(\cdot,i^*)} = 0$.
+We also have that $\sigma_A^{*(\cdot,1)} = 0.05s$, $\phi_A^{*(\cdot,1)} = 0.99$, $\sigma_A^{*(\cdot,2)} = 0.1s$, and $\phi_A^{*(\cdot,2)} = 0.95$.
 
-To calculate the wiggliness $\Ztwo$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz).
+To calculate the wiggliness $\Ztwo_{t,t^*}$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz).
 Recall that $\Ztwo_{t,t^*}$ is the sum of gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also gamma:
@@
 %
-It follows that $\mu_W^{*(1)} = 10.10$, $\sigma_W^{*(1)} = 3.18$, $\mu_W^{*(2)} = 305.94$, and $\sigma_W^{*(2)} = 17.49$.
+It follows that $\mu_W^{*(\cdot,1)} = 10.10$, $\sigma_W^{*(\cdot,1)} = 3.18$, $\mu_W^{*(\cdot,2)} = 305.94$, and $\sigma_W^{*(\cdot,2)} = 17.49$.
 

### 2020-08-31 Report/Sections/results.tex

@@
 The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3.
-In addition, the HHMM-DFT produces estimates of $\sigma_A^{*(1)}$, $\sigma_A^{*(2)}$, and $\sigma_A^{*(3)}$ which are approximately $50$ to $100$ percent larger than those of the CarHHMM-DFT for all axes of acceleration.
-The estimated uncertainties of the estimates of $\mu_A^{*(1)}$, $\mu_A^{*(2)}$, and $\mu_A^{*(3)}$ are also less than half of those for the CarHHMM-DFT for all axes of acceleration.
+In addition, the HHMM-DFT produces estimates of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$ which are approximately $50$ to $100$ percent larger than those of the CarHHMM-DFT for all axes of acceleration.
+The estimated uncertainties of the estimates of $\mu_A^{*(\cdot,1)}$, $\mu_A^{*(\cdot,2)}$, and $\mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT for all axes of acceleration.
 Further, the empirical auto-correlation of acceleration within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers.

### 2020-09-01 Report/Sections/appendix.tex
Clean up report and delete old files

@@
 
-\subsection{Equivalence of State-Switching OU process and CarHMM}
+\subsection{Equivalence of state-switching OU process and CarHMM}
 
@@
 They refer to this as a state-switching Ornstein-Uhlenbeck process.
-Unlike HMMs, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer.
+Unlike an HMM, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer in this case.
 If the behavioural state $X_t$ is known and does not change between observations $\Big($i.e $X_s = X_t$ for all $s \in [t,t+\Delta t)\Big)$, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
@@
 \begin{proof}{Proof of Theorem 1}{}
-Combining Equation (\ref{eqn:OU_sol}) with Equation (\ref{eqn:CarHMM_to_OU}) gives:
+Combining Equation (\ref{eqn:OU_sol}) with Equation (\ref{eqn:CarHMM_to_OU}) gives
 %
@@
 \left(Y_{s \Delta t} | X_{(s-1)\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{(s-1) \Delta t}, \enspace \left(\sigma^{(i)}\right)^2 \right),\\
-s = 1, \ldots, T-1
+s = 1, \ldots, T-1.
 \end{align*}
@@
 Let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and let $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for convenience.
-We simulate $\hat Y^*$ and then take the inverse $DFT$ to find $Y^*$ according to the following procedure.
+We first simulate $\hat Y^*_{t,t^*}$ for all $t$ and $t^*$ and then take the inverse $DFT$ to find $Y^*$ according to the following procedure.
 
 \begin{enumerate}
-    \item We begin by simulating $\hat{Y}^{*(0)}_{t,1}$:
+    \item We begin by simulating $\hat{Y}^{*(0)}_{t,1}$ for each dive $t = 1,\ldots,100$:
     $$
-    	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
+    	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right),
     $$
@@
     \item Next, for each dive $t$, we simulate $\hat{Y}^{*(0)}_{t,101},\enspace \hat{Y}^{*(0)}_{t,201}, \enspace \ldots ~ $ using Equation (\ref{eqn:yhat_0}) below:
-    \begin{equation}	
-        \Big(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*,\hat{Y}^{*(0)}_{t,t^*-100}\Big) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,t^*-100}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \quad t^* = 101,201,\ldots
+    \begin{align}	
+        \Big(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*,\hat{Y}^{*(0)}_{t,t^*-100}\Big) &\sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,t^*-100}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \nonumber \\
+        %
+        &t^* = 101,201,\ldots
     	\label{eqn:yhat_0}
-    \end{equation}
-    where $\phi_A^{*(\cdot,1)} = 0.99$ and $\phi_A^{*(\cdot,2)} = 0.95$. Note that $t^*$ has increments of 100 because $h=100$ when using two-second window sizes.
+    \end{align}
+    where $\phi_A^{*(\cdot,1)} = 0.99$ and $\phi_A^{*(\cdot,2)} = 0.95$.
+Note that $t^*$ has increments of 100 because $h=100$ when using two-second window sizes.
+After this step we have effectively simulated the average acceleration within each two-second window.
     %
     \item We then simulate Fourier modes 1 through 49 within each window:
-    $$
+    \begin{equation}
         \hat{Y}^{*(k)}_{t,t^*} = a_{t,t^*}^{(k)} i\sqrt{b^{(k)}_{t,t^*}}, \quad k = 1,\ldots,49;
-    $$
+        \label{eqn:abYhat}
+    \end{equation}
     where $a^{(k)}_{t,t^*}$ is equals either 1 or -1 with probability 1/2 and $b^{(k)}_{t,t^*}$ has the following distribution:
@@
     	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
-    	{\rm{Gamma}}(5/k^2, 1), \quad & k \notin \{1,2\} \\
     	{\rm{Gamma}}(250,1), & k = 1 \\
-    	{\rm{Gamma}}(50,1), & k = 2.
+    	{\rm{Gamma}}(50,1), & k = 2 \\
+    	{\rm{Gamma}}(5/k^2, 1), \quad & k \notin \{1,2\}.
     	\end{array}\right. 
     \end{split}
+    \label{eqn:bdist}
     \end{align}
-    This means that the expected values of Fourier modes 1 and 2 are much higher in subdive state 2 $(X^*_{t,t^*} = 2)$ Than subdive state 1 $(X^*_{t,t^*} = 1)$.
-Note that $\mathbb{E}\left(|\hat{Y}^{(k)}_{t,t^*}|^2\right)$ is proportional to $1/k^2$.
-This silences the higher Fourier modes and subsequently ``smooths out" the raw acceleration after taking the inverse DFT of $\hat Y$.
-    \item For Fourier modes 50 through 99, we use the following values to ensure that the inverse DFT is real-valued:
+    This means that the expected values of Fourier modes 1 and 2 are much higher in subdive state 2 $(X^*_{t,t^*} = 2)$ than in subdive state 1 $(X^*_{t,t^*} = 1)$.
+Note that $\mathbb{E}\left(||\hat{Y}^{(k)}_{t,t^*}||^2\right)$ is proportional to $1/k^2$.
+This facilitates lower expected values for higher Fourier modes and subsequently ``smooths out" the raw acceleration after taking the inverse DFT of $\hat Y$.
+    %
+    \item We set Fourier modes 50 through 99 using the following equation to ensure that the inverse DFT is real-valued:
     $$
@@
     %
-    \item Finally, $\Big(Y^*_{t,t^*},\ldots,Y^*_{t,t^*+99}\Big)$ is set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
+    \item Finally, for each dive $t$, $\Big(Y^*_{t,t^*},\ldots,Y^*_{t,t^*+99}\Big)$ is set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
     %
@@
 \end{enumerate}
+
+We now show that this construction of $Y^*$ results in the correct distributions as listed in Section \ref{sec:sim_study}. First, note the following:
 %
-%$\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued.
-$\hat{Y}^{*(k)}_{t,t^*}$ also decays like $1/k$ so that $Y^*_{t,t^*}$ remains continuous within a two-second window.
-$Y^*_{t,t^*}$ is not continuous \textit{between} windows, but the jump discontinuities are not very severe since $\hat{Y}^{*(0)}_{t,t^*}$ and $\hat{Y}^{*(0)}_{t,t^*+1}$ are highly correlated.
-See Figure \ref{fig:sim_data} for details.
-
-We now show that this construction of $Y^*$ results in the correct distributions listed in Section \ref{sec:sim_study}.
-First, note that $\Zone_{t,t^*} = \text{Re}\left(\hat{Y}^{(0)}_{t,t^*}\right) = \hat{Y}^{(0)}_{t,t^*}$ because the real part of the Fourier coefficient corresponding to a frequency of 0 is equal to the average value of an interval and $\hat{Y}^{(0)}_{t,t^*}$ is real-valued by construction.
-Then, combine this fact with Equation (\ref{eqn:yhat_0}) to find the distribution of $\Zone_{t,t^*}$:
+\begin{align}
+    \hat{Y}^{(0)}_{t,t^*} = DFT\{Y^*_{t,t^*}\}(0) &= \sum_{n=0}^{99} Y^*_{t,t^*+n} \exp\left(-\frac{i2\pi}{99}*0*n\right) \nonumber \\
+    &= \sum_{n=0}^{99} Y^*_{t,t^*+n} \nonumber \\
+    &= 100A^*_{t,t^*}.
+    \label{eqn:avg_val}
+\end{align}
+%
+Combining Equation (\ref{eqn:avg_val}) with Equation (\ref{eqn:yhat_0}) gives the distribution of $\Zone_{t,t^*}$: 
+%
+$$\left(100\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(100\phi_A^{*(\cdot,i^*)} \Zone_{t,t^*-100}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \text{ so }$$
 %
@@
 To calculate the wiggliness $\Ztwo_{t,t^*}$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz).
-Recall that $\Ztwo_{t,t^*}$ is the sum of gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also gamma:
+Recall that $\Ztwo_{t,t^*}$ is the sum of gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also gamma.
+From Equation (\ref{eqn:abYhat}), we have:
 %
-$$\Ztwo_{t,t^*} = \sum_{k=1}^{10} \big|\hat{Y}^{(k)}_{t,t^*}\big|^2 = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
-Combining this with Equation block (5), we have:
+$$\Ztwo_{t,t^*} = \sum_{k=1}^{10} \big|\big|\hat{Y}^{(k)}_{t,t^*}\big|\big|^2 = \sum_{k=1}^{10} \big|\big|a^{(k)}_{t,t^*}i\sqrt{b^{(k)}_{t,t^*}}\big|\big|^2 = \sum_{k=1}^{10} \big|b^{(k)}_{t,t^*}\big|$$
+since $a^{(k)}_{t,t^*}$ is equal to either 1 or -1. Combining this with Equation block (\ref{eqn:bdist}) gives
 %
-$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 5/k^2 = 10.10, 1.00 \right)$$
+$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 5/k^2 = 10.10, 1.00 \right) \text { and }$$
 %
@@
 %
-It follows that $\mu_W^{*(\cdot,1)} = 10.10$, $\sigma_W^{*(\cdot,1)} = 3.18$, $\mu_W^{*(\cdot,2)} = 305.94$, and $\sigma_W^{*(\cdot,2)} = 17.49$.
+Where the first argument is the scale parameter and the second is the shape parameter.
+Simple calculation of the mean and variance of a gamma distribution shows that $\mu_W^{*(\cdot,1)} = 10.10$, $\sigma_W^{*(\cdot,1)} = 3.18$, $\mu_W^{*(\cdot,2)} = 305.94$, and $\sigma_W^{*(\cdot,2)} = 17.49$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 %
-where $f^{(i)}(y_t;\theta^{(i)})$ is the emission distribution of dive duration conditioned on the fact that $X_t = i$.
-The likelihood $\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale chain and is equal to the following:
+where $f^{(i)}(y_t;\theta^{(i)})$ is the emission distribution of dive duration given that $X_t = i$.
+The likelihood $\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale model and is equal to the following:
 %
@@
 %
-Recall that $z^*_t = \Big(z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\Big)$ and that $f^{*(\cdot,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$ is the emission distribution of $\Z_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\iffalse
-
-\subsection{Dimension Reduction on $\hat{Y_t^*}$}
-
-First, we simply use down-sampling and only record every $w^{th}$ time step of $\hat{Y}_t^*$.
-This both reduces the space of $\hat{Y}_t^*$ to $\mathbb{C}^{\lfloor t^*_t / w \rfloor \times w}$ and ensures that none of the sliding windows overlap when taking the STFT.
-This is important because HMMs assume temporal independence between observations.
-Next, the dimension of $\hat{Y}_t^*$ can be cut in half by recognizing that the $Y_t^*$ is real-valued, and therefore $\hat{Y}_{t,k}^*$ is equal to the complex-congugate of $\hat{Y}_{t,w-k-1}^*$.
-Finally, by Parseval's Thereom we have that:
-
-$$\sum_{n = 0}^{w-1} |Y^*_{t+n}|^2 = \sum_{n = 0}^{w-1} |\hat{Y}^*_{t,n}|^2$$
-
-\subsection{Equivalency of CarHMM and one-dimensional state-switching Ornstein-Uhlenbeck process}
-
-If it is the case that (1) the underlying behavioral state of the continuous-time model must follow a Markov chain rather than a Markov process, and (2) the emission distributions of the CarHMM are gaussian, then the CarHMM and the state-switching continuous model are equivalent.
-This allows the theoretically grounded continuous-time state-switching model to be used in the computational convieneint HMM (and therefore HHMM) framework.
-In addition, it gives new intpretation to the learned parameters of the CarHMM in the context of an Ornstein-Uhlenbeck process.
-
-A one-dimensional state-switching Ornstein-Uhlenbeck process $y^*$ is the solution to the following stochastic differential equation:
-%
-$$dy^*_t = \beta_{x^*_t}(\gamma_{x^*_t} - y^*_t)dt + \omega_{x^*_t} dW_t$$
-%
-where $x^*_t$ is the fine-scale behavior of the animal at time $t$, $\beta_{x^*_t}$ relates to rate at which the process returns to its mean value, $\gamma_{x^*_t}$ is the long-term mean value of the process, $\omega_{x^*_t}$ is related to short-term variance, and $W$ is a Wiener process.
-As before, $x^*_t$ is described by an unobserved Markov process.
-The solution to this equation is known to be the following \cite{Michelot:2019}:
-\begin{align*}
-y^*_{t+\delta} \sim \mathcal{N}\left((1-e^{-\beta_{x^*_t}\delta})\gamma_{x^*_t} + e^{-\beta_{x^*_t}\delta} y^*_t,\quad \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})\right)
-\end{align*}
-Now, suppose that $\delta$ is constant for all observations, as is the case for hidden Markov models. In addition, introduce the following transformations:
-\begin{align*}
-\mu_{x^*_t} = \gamma_{x^*_t}, \qquad \phi_{x^*_t} = e^{-\beta_{x^*_t}\delta}, \qquad \sigma^2_{x^*_t} = \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})
-\end{align*}
-Then, we have the following:
-\begin{align*}
-y^*_{t+\delta} \sim \mathcal{N}\left((1-\phi_{x^*_t})\mu_{x^*_t} + \phi_{x^*_t} y^*_t,\quad \sigma_{x^*_t}^2 \right)
-\end{align*}
-%
-If $\delta$ is fixed and $x^*_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities.
-Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again.
-This allows for the principled construction of the continuous-time model to be combined with the computational convenience of the CarHMM.
-
-\fi
+Recall that $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\Big\}$ and that $f^{*(\cdot,i^*)}(\cdot|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.

### 2020-09-01 Report/Sections/models.tex
Clean up report and delete old files

@@
 Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$.
-%Traditional HMMs model process auto-correlation by assuming that the underlying behaviour of a processes follows a single Markov chain.
-However, when conditioned on the hidden state sequence observations are assumed to be independent.
-Therefore, classical HMMs do not hold when the observations exhibit other forms of significant correlation in time.
-Examples of this include continuous-time processes such as a Wiener process, systems which exhibit temporal dependencies acting at several different time scales \citep{Barajas:2017}, and periodic behaviour including the motion of an ideal spring.
-%\textit{blah blah blah -quick restatements (5-15 words to potentially many sentences) of the main issues you are going to fix made a quick try at it, work on smoothing things.
-You can use some of the text I removed from below}.
-%To account for each of these common additional dependence structures,
 We first detail the structure of a traditional HMM.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Relaxing the conditional independence assumption with a CarHMM}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Transforming observations to account for fine-scale behaviour}
@@
 Subject-area experts are often required to determine the optimal window size and the specific summary statistics used.
-%
-%This moving window approach complicates the indexing of $Y_t^*$.
-Suppose a window size of $h$ observations is selected.
-Moving forward, we re-index $Y^*$ such that the fine-scale process associated with curve $t$ is denoted as $Y^*_t$ as before.
-However, the $t^{*th}$ \textit{window} of $Y^*_t$ is now denoted as $Y^*_{t,t^*}$, and the $t^{**th}$ \textit{observation} of $Y^*_{t,t^*}$ is denoted as $Y^*_{t,t^*,t^{**}}$.
-As a result, there are a total of $T^*_t$ windows associated with $Y_t^*$, each of which are of length $h$.
-Note that this relabeling may significantly reduce the length of $Y^*_t$ depending upon the window size $h$ and the step-size between windows.
-%
-
 For our case study, we use the discrete Fourier transform (DFT) of a moving window of width $h$ across $Y^*_t$:
@@
 
-%so that the fine-scale emission distribution becomes $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$. The likelihood of the HMM-DFT is as follows:
-%\begin{equation}
-%    \calL_{\text{HMM-DFT}}(y,z^*;\theta,\Gamma) = \delta P(y_1,z^*_1;\theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\theta) \mathbf{1}_N
-%    \label{HMMDFT_likelihood}
-%\end{equation}
-%where $P(y_t,z^*_t;\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.
-
-%It is possible to accommodate unequal time steps within $y_t^*$ by using the non-uniform discrete Fourier transform (NDFT).
-We do not describe the method in detail here, but the generalization is straightforward.
-See \citet{Bagchi:1999} for details.
-
 \subsection{General structure for building complex models}
@@
 In addition, \citep{Langrock:2018} use B-splines to model several aspects of an HMM, and this modified model admits a penalized likelihood which can be incorporated into the hierarchical structure here.
-%For these and countless other possible fine-scale models, added flexibility comes at a cost of increased model complexity.
 
@@
 However, these models quickly become complicated and should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.
-%In the sections that follow, we construct several candidate models from the building blocks above to describe killer whale dive behaviour.
-We select one preferred model and compare it to several of the simpler candidate models to ensure its complexity is necessary.

### 2020-09-01 Report/Sections/results.tex
Clean up report and delete old files

@@
 The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions.
-Observations were collected at a rate of 50 Hz using
-% an accelerometer and time-depth recorder.
-[NAME OF TAGS - do you know Sarah?].
+Observations were collected at a rate of 50 Hz using Customizable Animal Solutions Tracking (CATs) tags (model number ###).
 The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left).
@@
 %
-After transforming the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (see section 2.1 of the supplementary material).
+After transforming the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (see section 1.1 of the supplementary material).
 Therefore, we assume that all three components of $\Zone_{t,t^*}$, $t^* = 1,101,\ldots$ follow the CarHMM as defined in section 2.2.
@@
 Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives.
-Section 2.2 of the supplementary material also shows the estimated probability of each dive and subdive state given the data.
+Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.
 
@@
 
-We use two visual tools to evaluate the CarHHMM-DFT: pseudo-residual plots and empirical histograms. 
-%A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}.
-To easily visualize outliers, pseudo-residuals are often passed through the quantile function of the standard Normal distribution.
-Mathematically, the pseudo-residual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
-If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution.
+We use two visual tools to evaluate the CarHHMM-DFT: pseudoresidual plots and empirical histograms.
+Mathematically, the pseudoresidual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
+If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution.
 Histograms of the pseudoresiduals mostly support that this model is well-specified.
-One exception is $\Ztwo_{t,t^*}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
+One exception is $\Ztwo_{t,t^*}$, whose pseudoresiduals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
 This implies that the true distribution of $\Ztwo_{t,t^*}$ may follow a heavier-tailed distribution than the gamma distribution used in the case study.
-See section 2 of the supplementary material for pseudo-residual plots for all observations.
+See sections 1.4 - 1.6 of the supplementary material for pseudoresidual plots for all observations.
 
@@
 Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}).
-See section 2 of the supplementary material for analogous histograms corresponding to the fine-scale observations $z^*_{t,t^*}$.
+See sections 1.4 - 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $z^*_{t,t^*}$.
 Our results again mostly show that the CarHHMM-DFT fits the data well, but there are some exceptions.
@@
 The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution.
-See section 2 of the supplementary material for a more complete set of results for each of the candidate models.
+See section 1 of the supplementary material for a more complete set of results for each of the candidate models.

### 2020-09-01 Report/Sections/simulation.tex
Clean up report and delete old files

@@
 \begin{gather*}
-    \mu_A^{*(1)} = 0.0 s, \enspace \sigma_A^{*(1)} = 0.05s, \enspace \phi_A^{*(1)} = 0.99, \\
+    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
     %
-    \mu_A^{*(2)} = 0.0 s, \enspace \sigma_A^{*(2)} = 0.1s, \enspace \phi_A^{*(2)} = 0.95, \\
+    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
     %
-    \mu_W^{*(1)} = 10.10, \quad \sigma_W^{*(1)} = 3.18, \\
+    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
     %
-    \mu_W^{*(2)} = 305.94, \quad \sigma_W^{*(2)} = 17.49.
+    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
 \end{gather*}
@@
 For dive type, the average confidence was above 0.9 (see Table \ref{table:accuracy}).
-All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased (see section 1 of the supplementary material).
+All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased (see section 2 of the supplementary material).
 Even still, these biases are the least severe for the CarHHMM-DFT among the four models.
 The empirical standard error of all parameter estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slight underestimates of the empirical standard error.
-See section 1.2 and tables 2-5 of the supplementary material for more information.
+See Tables 6-9 of the supplementary material for more information.
 
@@
 For both subdive types, $\hat \sigma_A^*$ greatly overestimates the true standard deviation.
-The estimated standard errors of $\hat \mu_A^*$ and $\hat \sigma_A^*$ using the observed Fisher information are both much smaller than the empirical standard error (see section 1.2, Table 3 of the supplementary material).
+The estimated standard errors of $\hat \mu_A^*$ and $\hat \sigma_A^*$ using the observed Fisher information are both much smaller than the empirical standard error (see Table 7 of the supplementary material).
 This implies that the variance of acceleration can be overestimated and the standard error of the parameter estimates can be underestimated when auto-correlation is ignored.
@@
 This result is consistent with expectations because the CarHHMM does not have access to the ``wiggles" of the fine-scale process, which are the most informative observations when determining subdive type.
-The CarHHMM is also much more likely to mis-classify subdive type 1 than subdive type 2 (see section 1.2, Table 1 of the supplementary material).
-The CarHHMM is also the worst of the four at estimating parameters, and this is in part due to its relatively poor decoding accuracy.
-$\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$ are especially poor estimates (see section 1 of the supplementary material).
+The CarHHMM is also much more likely to misclassify subdive type 1 than subdive type 2 (see Table \ref{table:accuracy}).
+The CarHHMM is also the worst of the four at estimating parameters in part due to its relatively poor decoding accuracy.
+The fine-scale probability transition matrices associated with dive type 2 ($\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$) are especially poor estimates (see Table 9 and Figure 33 of the supplementary material).
 
@@
 See Figures \ref{fig:acc_coarse} \& \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive/subdive type probabilities as determined by each model.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% EVERYTHING BELOW HERE IS OLD AND I THINK IT IS BAD. THE STUFF ABOVE IS MUCH BETTER.
-
-\iffalse
-%%% Accuracy %%%
-Each model was able to decode the fine-scale hidden states of the process almost perfectly except for the CarHHMM, whose accuracy was 89\% (Table \ref{table:accuracy}).
-The lower accuracy of the CarHHMM (i.e.
-model without $Z^{*(2)}$)  can be attributed to the higher fine-scale states variation in the distribution of $Z^{*(2)}$ compared to $Z^{*(1)}$.
-For the coarse-scale hidden states, the CarHMM-DFT lacked a hierarchical structure and could not make coarse-scale predictions.
-All three other models achieved an accuracy of approximately 90\% (Table \ref{table:accuracy}), with the CarHHMM slightly more likely to categorize a dive as dive type 2 than the other models (Fig ?
-or Sup Table ?
-that give quantitative info on this).
-Figure \ref{fig:acc} shows the decoded state probabilities for both the fine- and coarse- scales for one of the simulation run.
-
-%%% dive duration %%%
-For the emission distributions of dive duration, Figure \ref{fig:MLE_dist} shows the empirical joint density of $\hat \mu$ and $\hat \sigma$ for the CarHHMM-DFT.
-
-Estimates of standard error using the observed Fisher information tended to be negatively biased due to correlation between $\hat \mu$ and $\hat \sigma$.
-In addition, $\hat \sigma$ tends to be an underestimate of $\sigma$ for all models and dive types.
-This finding is consistent with properties of MLEs for standard deviation, especially with a small sample size.
-In particular, the CarHMM-DFT model is not well specified one the coarse-scale and severely underestimates $\sigma$.
-%This is likely due to the fact that dive duration follows a mixture of gammas rather than a gamma distribution, and the CarHMM does not have the machinery to capture this fact.
-A full table of parameter estimates for all models is shown in the supplementary material.
-
-%%% Acceleration %%%
-For acceleration ($Z^{*(1)}$), both the CarHMM-DFT and the CarHHMM-DFT regularly converged to the correct parameters with very little standard error.
-However, the estimated standard error based on the observed Fisher information regularly overestimated the empirical standard error of $\hat \mu$ for both of these models.
-%This likely is due to the very large auto-correlation terms ($\phi^{(1)} = 0.99$ and $\phi^{(2)} = 0.95$) since as $\phi \to 1$, the value of $\mu$ does not matter in the likelihood evaluation as $\phi \to 1$ implies that $\mathbb{E}(Z^{*(1)}_{t,s^*}) = \phi*(Z^{*(1)}_{t,s^*-1}) + (1-\phi)\mu \to Z^{*(1)}_{t,s^*-1}$.
-The HHMM-DFT regularly overestimates the variance $Z^{*(1)}$ since it does not incorporate auto-correlation into the emission distribution of $Z^{*(1)}$.
-The CarHHMM has large biases in many of its parameter estimates, especially in subdive state 2.
-%because it sees the ``wigliness" of the acceleration data as raw acceleration rather than as Fourier modes, which deflates the value of $\hat \phi$ and inflates the value of $\hat \sigma$.
-
-The supplementary material provides a detailed breakdown of parameter values for acceleration emission distributions and the empirical joint densities for all models and features.
-
-%%% FoVeDBA %%%
-For all models there is no bias in the parameter estimates for the distribution of the Fourier sums ($Z^{*(2)}$).
-One exception is the CarHHMM, which does not model $Z^{*(2)}$ as an observation at all.
-The standard error estimates based on the observed Fisher information closely approximate the empirical standard error.
-%The distribution of $Z^{*(2)}$ is very different between subdive states 1 and 2, so the signal is therefore very easy for each model to pick up on.
-The supplementary material also shows a detailed breakdown of the emission distribution of $Z^{*(2)}$ for all models.
-
-%%% Gamma %%%
-Estimates of the coarse-scale probability transition matrix $\Gamma$ for all models (except for the CarHMM-DFT) are very accurate, with empirical standard errors of approximately 0.02 and biases of nearly zero.
-This is significantly less than the estimate of standard error based on the observed Fisher information, which is about 0.08 for all models.
-%likely because the distribution of $\hat \Gamma$ did not converge to a normal distribution in only 100 observed dives. 
-For the fine-scale transition matrices $\Gamma^*$, the HHMM-DFT and CarHHMM-DFT both showed practically no bias with standard errors on the order of $10^{-2}$.
-One notable exception is the standard error of $\hat \Gamma^{*(1)}_{12}$, whose standard error was on the order of $10^{-4}$, much lower than the observed Fisher information would predict.
-The CarHMM-DFT is mis-specified, so its results cannot be easily interpreted, but $\hat \Gamma^*$ regularly converged to a constant value with a standard error on the order of $10^{-2}$.
-The CarHHMM consistently overestimated the trace of $\Gamma^{*(i^*)}$.
-Again, one notable exception is $\hat \Gamma^{*(1)}_{12}$, which had almost no bias and was remarkably consistent.
-See the supplementary material for a full list of estimates and standard errors.
-
-\fi

### 2020-09-04 Report/Sections/appendix.tex
Update report- Evan Edits

@@
 %
-where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
+where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process in state $X_t$, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
 They refer to this as a state-switching Ornstein-Uhlenbeck process.
@@
 \begin{align*}
-\left(Y_{s \Delta t} | X_{(s-1)\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{(s-1) \Delta t}, \enspace \left(\sigma^{(i)}\right)^2 \right),\\
-s = 1, \ldots, T-1.
+    \left(Y_{(s+1) \Delta t} | X_{s\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{s \Delta t}, \enspace \left(\sigma^{(i)}\right)^2 \right),\\
+    s = 0, \ldots, T-2.
 \end{align*}
@@
     \left(Y'_s| X'_s = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y'_{s-1}, \enspace \left(\sigma^{(i)}\right)^2 \right)\\
-s = 2, \ldots, T
+    s = 2, \ldots, T.
 \end{align*}
 %
-$X'$ follows a (possibly unobserved) Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s-1})$ is consistent with that of a CarHMM with Normal emission distributions.
+$X'$ is assumed to follow a (possibly unobserved) Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s-1})$ is consistent with that of a CarHMM with Normal emission distributions.
 \end{proof}
@@
 
-\subsection{Detailed description of simulated data}
+\subsection{Detailed description of data simulation}
 
-Assume that $X$, $Y$, and $X^*$ are known and simulated using the procedure detailed in Section \ref{sec:sim_study}, leaving only $Y^*$ to be simulated.
-Let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and let $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for convenience.
-We first simulate $\hat Y^*_{t,t^*}$ for all $t$ and $t^*$ and then take the inverse $DFT$ to find $Y^*$ according to the following procedure.
+It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
+
+To simulate $X^*_t$ for each dive $t$, note that we simulate accelerometer readings taken at 50 hertz and have two-second window lengths.
+Therefore, each window contains 100 accelerometer readings and $X^*_t = \left\{X^*_{t,1}, X^*_{t,101}, \ldots, X^*_{t,100T^*_t-99}\right\}$.
+In addition, the length of each fine-scale Markov chain is $T^*_t = \lfloor Y_t/2 \rfloor$.
+Then, given the dive types $X_t$, dive durations $Y_t$, and the fine-scale probability transition matrices $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$, it is straightforward to simulate $X^*_t$ for each dive $t = 1,2,\ldots,100$.
+Recall that $\Gamma^{*(i)}_{jk} = Pr(X^*_{t,t^*+100} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1, 101, \ldots$.
+It is much more difficult to simulate the raw acceleration data $Y^*$, and we focus on this task for the remainder of this section.
+
+Assume that $X$, $Y$, and $X^*$ are simulated using the procedure described above and are therefore known.
+Additionally, let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for convenience.
+We first simulate $\hat Y^{*(k)}_{t,t^*}$ for all dives $t = 1,2,\ldots,100$, all subdive windows $t^* = 1,101,\ldots$, and all Fourier coefficients $k = 0,1,\ldots,99$.
+We then take the inverse DFT of $\hat{Y}^*_{t,t^*}$ for all dives $t = 1,2,\ldots,100$ and subdive windows $t^* = 1,101,\ldots$ to find $Y^*$.
+The details of this procedure are described below:
 
 \begin{enumerate}
-    \item We begin by simulating $\hat{Y}^{*(0)}_{t,1}$ for each dive $t = 1,\ldots,100$:
+    \item We begin by simulating $\hat{Y}^{*(0)}_{t,1}$ for each dive $t = 1,2,\ldots,100$:
     $$
@@
     where $\phi_A^{*(\cdot,1)} = 0.99$ and $\phi_A^{*(\cdot,2)} = 0.95$.
-Note that $t^*$ has increments of 100 because $h=100$ when using two-second window sizes.
-After this step we have effectively simulated the average acceleration within each two-second window.
+After this step we have simulated the average acceleration within each two-second window for all dives (see Equation (\ref{eqn:avg_val}) below).
     %
-    \item We then simulate Fourier modes 1 through 49 within each window:
+    \item We then simulate Fourier coefficients 1 through 49 within each window:
     \begin{equation}
@@
     \end{equation}
-    where $a^{(k)}_{t,t^*}$ is equals either 1 or -1 with probability 1/2 and $b^{(k)}_{t,t^*}$ has the following distribution:
+    where $a^{(k)}_{t,t^*}$ is equal to either 1 or -1, each with probability 1/2. In addition, $b^{(k)}_{t,t^*}$ has the following distribution:
     \begin{align}
@@
     \end{align}
-    This means that the expected values of Fourier modes 1 and 2 are much higher in subdive state 2 $(X^*_{t,t^*} = 2)$ than in subdive state 1 $(X^*_{t,t^*} = 1)$.
-Note that $\mathbb{E}\left(||\hat{Y}^{(k)}_{t,t^*}||^2\right)$ is proportional to $1/k^2$.
-This facilitates lower expected values for higher Fourier modes and subsequently ``smooths out" the raw acceleration after taking the inverse DFT of $\hat Y$.
+    The first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the scale parameter and the second is the shape parameter.
+It is important that $\mathbb{E}\left(|b^{(k)}_{t,t^*}|\right)$ is proportional to $1/k^2$, as smaller expected values for larger Fourier coefficients ``smooths out" the raw acceleration data.
     %
-    \item We set Fourier modes 50 through 99 using the following equation to ensure that the inverse DFT is real-valued:
+    \item We set Fourier coefficients 50 through 99 strategically to ensure that the inverse DFT is real-valued:
     $$
@@
     %
-    \item Finally, for each dive $t$, $\Big(Y^*_{t,t^*},\ldots,Y^*_{t,t^*+99}\Big)$ is set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
+    \item Finally, for each dive $t = 1,2,\ldots,100$ and each window $t^*=1,101,\ldots$, the raw acceleration data $\left\{Y^*_{t,t^*},\ldots,Y^*_{t,t^*+99}\right\}$ is set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
     %
@@
 
-We now show that this construction of $Y^*$ results in the correct distributions as listed in Section \ref{sec:sim_study}. First, note the following:
+We now show that this construction of $Y^*$ results in the correct distributions as listed in Section \ref{subsec:data_simulation}.
+First, note that $\hat{Y}^{(0)}_{t,t^*}$ is just the sum of the raw acceleration values within the corresponding window:
 %
@@
 %
-Combining Equation (\ref{eqn:avg_val}) with Equation (\ref{eqn:yhat_0}) gives the distribution of $\Zone_{t,t^*}$: 
+Combining Equation (\ref{eqn:avg_val}) above with Equation (\ref{eqn:yhat_0}) gives the distribution of $\Zone_{t,t^*}$: 
 %
-$$\left(100\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(100\phi_A^{*(\cdot,i^*)} \Zone_{t,t^*-100}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \text{ so }$$
+$$\left(100\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(100\phi_A^{*(\cdot,i^*)} \Zone_{t,t^*-100}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)$$
 %
-$$\left(\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \Zone_{t,t^*-100}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right).$$
+$$\implies \left(\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \Zone_{t,t^*-100}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right).$$
 %
@@
 To calculate the wiggliness $\Ztwo_{t,t^*}$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz).
-Recall that $\Ztwo_{t,t^*}$ is the sum of gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also gamma.
+Recall that $\Ztwo_{t,t^*}$ is the sum of Gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also Gamma-distributed.
 From Equation (\ref{eqn:abYhat}), we have:
 %
-$$\Ztwo_{t,t^*} = \sum_{k=1}^{10} \big|\big|\hat{Y}^{(k)}_{t,t^*}\big|\big|^2 = \sum_{k=1}^{10} \big|\big|a^{(k)}_{t,t^*}i\sqrt{b^{(k)}_{t,t^*}}\big|\big|^2 = \sum_{k=1}^{10} \big|b^{(k)}_{t,t^*}\big|$$
-since $a^{(k)}_{t,t^*}$ is equal to either 1 or -1. Combining this with Equation block (\ref{eqn:bdist}) gives
+\begin{equation}
+    \Ztwo_{t,t^*} = \sum_{k=1}^{10} \big|\big|\hat{Y}^{(k)}_{t,t^*}\big|\big|^2 = \sum_{k=1}^{10} \big|\big|a^{(k)}_{t,t^*}i\sqrt{b^{(k)}_{t,t^*}}\big|\big|^2 = \sum_{k=1}^{10} \big|b^{(k)}_{t,t^*}\big|
+    \label{eqn:Wb}
+\end{equation}
+%
+since $a^{(k)}_{t,t^*} \in \{-1,1\}$. Combining Equation (\ref{eqn:Wb}) above with Equation block (\ref{eqn:bdist}) gives
 %
@@
 %
-$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , 1.00 \right).$$
+$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , 1.00 \right),$$
 %
-Where the first argument is the scale parameter and the second is the shape parameter.
-Simple calculation of the mean and variance of a gamma distribution shows that $\mu_W^{*(\cdot,1)} = 10.10$, $\sigma_W^{*(\cdot,1)} = 3.18$, $\mu_W^{*(\cdot,2)} = 305.94$, and $\sigma_W^{*(\cdot,2)} = 17.49$.
+where the first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the scale parameter and the second is the shape parameter.
+Simple calculation of the mean and variance of a Gamma distribution shows that $\mu_W^{*(\cdot,1)} = 10.10$, $\sigma_W^{*(\cdot,1)} = 3.18$, $\mu_W^{*(\cdot,2)} = 305.94$, and $\sigma_W^{*(\cdot,2)} = 17.49$.
 
@@
 %
-Recall that $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\Big\}$ and that $f^{*(\cdot,i^*)}(\cdot|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.
+Recall that $z^*_t = \left\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\right\}$ and that $f^{*(\cdot,i^*)}(\cdot|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.

### 2020-09-04 Report/Sections/discussion.tex
Clean up report and delete old files
Update report- Evan Edits

--- 2020-08-14 Report/Sections/discussion.tex
@@
 
-Current FDA literature addresses dependence between curves either with multilevel models with random effects and no time component \citep{Chen:2012,Di:2009}, or with functional time series which model the curves as an evolving process \citep{Kokoszka:2018} and overlook the possibility that curves have several distinct ``types".
+Current FDA literature addresses dependence between curves either with multilevel models with random effects and no time component at all \citep{Chen:2012,Di:2009}, or with functional time series which model the curves as an evolving process \citep{Kokoszka:2018} which overlook the possibility that curves have several distinct ``types".
 However, many curve sequences exhibit both temporal dependence and a discrete number of types.
-One obvious example is the case study examined here, where one marine mammal dive serves a purpose such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}.
+%One obvious example is the case study examined here, where one marine mammal dive serves a distinct purpose such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}.
 Another example is accelerometer data in school children, where bouts of activity can be associated with a behaviour such as running or walking.
 \citep{Morris:2007}.
-
 In this work, we deal with both temporal dependence between curves and distinct curve ``types" by using either an HMM or CarHMM.
 To help select the appropriate model between these two, lag plots can be used to test if there is significant auto-correlation between curves.
-A traditional HMM approach requires that practitioners take summary statistics of curves as emissions.
-However, HMM emissions can also be the entire curve, which is then modelled with a fine-scale parametric model.
-The parameters of this fine-scale model then depend upon the curve type.
-There are a wide range of possible fine-scale models, including another HMM, CarHMM, a spline-based approach similar to that of \citet{Langrock:2018}, and many more.
-Together, the two levels make up a hierarchical model which can account for processes taking place simultaneously at different time-scales.
 
-%Traditional HMMs can be used to model a state-switching process with conditionally independent observations and Markovian dynamics when conditioned of the hidden state.
-However, many real-world processes require more complex models.
-Here, we presented a flexible framework that can account for complicated dependence structures within time-series data.
-Our framework is a collection of HMM models which can be combined together to form increasingly complex hierarchical models to match the complexity of particular problems faced by researchers.
+%When used traditionally, each of these models require that practitioners take summary statistics of curves as emissions which may lack important information about the structure of the curve.
+However, HMM emissions can also be viewed as the entire curve, where the emission distribution of that curve is described by yet another fine-scale model.
+There are a wide range of possible fine-scale models, including, but not limited to, another HMM, a CarHMM, a Poisson process, or a spline-based approach similar to that of \citet{Langrock:2018}.
+The parameters of this fine-scale model then depend upon the curve type as determined by the coarse-scale model.
+Together, these coarse- and fine-scale models make up a hierarchical model which can account for simultaneous processes taking place at different time-scales.
 
-%The CarHMM generalizes the HMM by explicitly modeling auto-correlation in the emission distributions of the HMM \citep{Lawler:2019}.
-CarHMMs also maintain the structure needed to evaluate the likelihood using the forward algorithm.
-In our Normal model formulation, we have added only one additional parameter, $\phi^{(i)}$, per possible hidden state.
+%One interesting class of possible fine-scale models is the set of continuous time methods such as the state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019} and the continuous-time HMM (CTHMM) \citep{Liu:2015}, which are often based on stochastic differential equations.
+Unfortunately, most continuous-time models require MCMC algorithms to perform inference and can be slow to fit.
+However, we show the appendix that the state-switching OU process of \citet{Michelot:2019} is equivalent to a CarHMM under certain conditions.
+This equivalence implies that a CarHMM is interpretable as a continuous-time model, but has the added advantage of relative computational efficiency.
 
-When the fine-scale model is  an HMM, the resulting overall model is the HHMM as introduced by \citet{Barajas:2017}.
+%Another fine-scale model that we investigate in detail here is yet another HMM.
+If the fine-scale model is an HMM, then the resulting overall model is just the HHMM as introduced by \citet{Barajas:2017}.
 \citet{Adam:2019} in particular use the HHMM to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California.
-This model can be inappropriate at very fine scales, as \citet{Adam:2019} find clear violations of conditional independence given hidden states and also observe periodic behaviour in the acceleration data.
+This model can be inappropriate at very fine scales, as \citet{Adam:2019} find that subsequent fine-scale observations are not independent when conditioned on the hidden states due to periodic behaviour in the acceleration data.
 This indicates the need to account for complicated dependence structures within fine-scale processes, especially those with high sampling frequencies.
-
-%For simultaneous observed processes taking place at different time scales, the HHMM \citep{Barajas:2017,Adam:2019} utilizes hierarchical structures to jointly model both as HMMs.
-In particular, each hidden state of the coarse-scale HMM is assumed to emit both an observation $Y_t$ as well as another fine-scale HMM with hidden states $X^*_t$ and observations $Y^*_t$.
-
-%For processes such as these with very high sampling frequencies and/or intricate fine-scale structure,
-
-%To deal with these issues, it is possible to generalize the HHMM such that the fine-scale model can be any model which admits an easy-to-calculate likelihood.
-For example, if the sampling rate of the fine-scale process is very high and clearly period, then the fine-scale model can be described by a simple probability distribution over the summary statistics of a moving window of observations.
-
 To deal with these problematic fine-scale phenomena, we explore a moving-window transformation over the fine-scale observations.
-For periodic behaviour in particular, we use a DFT-based transformation.
-The effectiveness of this approach is evidenced by both our case study, where the CarHHMM without DFT visually fails to recognize fluking behaviour in a killer whale, and our simulation study, where the subdive decoding accuracy is by far the worst when omitting $\Ztwo$ (``wiggliness'') from the model.
-For simpler transformations, e.g.
-an average such as $\Zone$, either subject area experts or exploratory plots may suggest temporal correlation.
-In this case,  we suggest using a CarHMM instead of a traditional HMM.
+Since our case study data clearly exhibits periodic behaviour, we specifically use a DFT-based transformation.
+The effectiveness of this approach is evidenced by both our case study, where the CarHHMM without a DFT transformation visually fails to recognize fluking behaviour in a killer whale, and our simulation study, where the average accuracy associated with the subdive state is by far the worst when omitting $\Ztwo_{t,t^*}$ (``wiggliness'') from the model.
+After transforming the fine-scale observations, either subject area experts or exploratory analysis may suggest a large degree of temporal correlation.
+In this case, we suggest using a CarHMM instead of a traditional HMM to account for the auto-correlation in the transformed fine-scale observations.
 Our simulation study shows how ignoring this auto-correlation can result in underestimating standard errors and overestimating the variance of the observations.
 
-We combine generalizations of HMMs with hierarchical structures and moving-window transformations to model animal behaviour as recorded from accelerometers.
-Time depth recorders and accelerometers are powerful tools to understand the diving behaviour and energetic requirements of NRKWs, and this is essential to understand their foraging behaviour and survival \citep{Williams:2009,Noren:2011}.
-However, biologging data require classifying the underlying activity \citep{Dot:2016} or dive type \citep{Hastie:2006} of the animal in order to effectively understand and estimate energy expenditure.
-The CarHHMM-DFT separates dives into different types on the coarse scale while describing the distribution of acceleration within each subdive state on the fine scale.
-This gives a deeper understanding of the animal's tri-axial movement, and thus its fine-scale energy expenditure \citep{Gleiss:2011,Qasem:2012}.
-Although our case study focuses on Northern resident killer whales, this model is likely applicable to many dive animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.
-
-While we combine the HMMs of section \ref{sec:models} effectively for our case study, in general this practice should be done with care as there are drawbacks to using exclusively HMMs to build these hierarchical models.
-For example, HMMs require observations to be taken at regular intervals.
-Continuous-time methods such as the state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019} and the continuous-time HMM (CTHMM) \citep{Liu:2015} are often based on stochastic differential equations and resolve this issue.
-However most continuous-time models require relatively slow MCMC algorithms to perform inference and are prohibitively difficult to fit.
-We show the appendix that the state-switching OU process of \citet{Michelot:2019} is equivalent to a CarHMM under certain conditions.
-This equivalence implies that a CarHMM is interpretable as a continuous-time model, but has the added advantage of relative computational efficiency.
-
-When using HMMs to build complex models, one must consider how to effectively fit the data while avoiding over-fitting and at the same time ensuring computationally efficient parameter estimation.
-One way to temper model complexity is to assume that the fine-scale state parameters are the same for all coarse-scale states.
-Regardless, model complexity inevitably grows rapidly as more layers are added to the hierarchical structure.
+%When using HMMs to build complex hierarchical models, it is important to build a model that is complex enough to fit the data but not so complex that it over-fits the data or is computationally inefficient.
+Model complexity inevitably grows rapidly as more layers are added to the hierarchical structure.
 For example, by far the fastest model to train in the simulation study is the CarHMM-DFT, as it has no hierarchical component and is the simplest of the four candidate models.
@@
 However, this model does not allow ecologists to study the joint relationship between the dive type and intra-dive behaviour.
+One way to temper model complexity is to reduce the number of parameters within a given model structure.
+In particular, we assume that the fine-scale state parameters are the same for all coarse-scale states for all models in both our case study and simulation study.
 
-%The simulation study also shows that the observed Fisher information serves as a suitable approximation for the standard errors of parameter estimates in most cases.
-One notable exception is the HHMM-DFT, which ignores auto-correlation and therefore underestimates the standard error of the parameters associated with acceleration.
-This can be dangerous in the field of ecology, as overconfident parameter estimates can result in overconfident conservation decisions.
-While the point estimates are generally good in this case study, this may not be the case in more general settings and for shorter time series in particular.
+%To illustrate the potential applications of this method, we model the behaviour of a northern resident killer whale animal behaviour using data collected from time depth recorders and accelerometers.
+Understanding the diving behaviour and energetic requirements of NRKWs is essential to understand their foraging behaviour and survival \citep{Williams:2009,Noren:2011}.
+However, it is necessary to classify the underlying activity \citep{Dot:2016} or dive type \citep{Hastie:2006} of an animal in order to effectively understand and estimate its energy expenditure from biologging data.
+The CarHMMM-DFT model uses coarse-scale data to separate dives into different types while also describing the distribution of acceleration within each subdive state on the fine scale.
 
-%We used the CarHHMM-DFT to model the behavior of a killer whale off the coast of British Columbia, Canada.
-The CarHHMM-DFT was able to simultaneously distinguish three distinct subdive behaviors and two dive types.
-The DFT component proved useful in determining the subdive behaviour of the whale, as the mean of the emission distribution of $Z^{*(2)}$ for each subdive state was separated by an order of magnitude.
-Finally, the estimated auto-correlation parameter for $\mathbf{Z}^{*(1)}$, $\phi^*$, was above 0.5 for every dimension and subdive type, providing evidence that the conditionally auto-regressive component of the CarHHMM-DFT resulted in a better fit to the data.
-The introduction of the parameter $\phi^*$ also allows $\mathbf{Z}^{*(1)}$ to be interpreted as a state-switching OU process (see appendix).
-
-%Because traditional information criteria tend to overestimate the number of states in biological processes \citep{Pohle:2017}, the number of dive types and subdive behaviors was selected in an ad-hoc manner.
-There does appear to be some heterogeneity within dive types, and future work can be done to determine the optimal number of dive types and within-dive behaviors.
-
-Our work provides a flexible framework to model functional time-series data using HMMs.
+%Our work provides a flexible framework to model functional time-series data using HMMs.
+We deal with temporal dependence between curves and distinct curve ``types" by using either an HMM or CarHMM.
 Provided the model is not overly complex, a hierarchical model constructed using the proposed building blocks can be both flexible and easy to fit using maximum likelihood methods.
-We demonstrate the usefulness of such a model using an ecological example, where we use a complex HMM to classify the coarse and fine scale diving behaviour of a killer whale.
-In addition, as complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology to the analysis of a wide range of time-series data in a variety of fields.
+We demonstrate the usefulness of such a model using an ecological example, where we use a complex HMM to classify the coarse- and fine-scale diving behaviour of a killer whale.
+This gives a deeper understanding of the animal's tri-axial movement and thus its fine-scale energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both NRKWs and their prey \citep{Noren:2011}.
+Although our case study focuses on northern resident killer whales, this model is likely applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.
+In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time-series data in a variety of fields.

### 2020-09-04 Report/Sections/intro.tex
Clean up report and delete old files
Update report- Evan Edits

@@
 Tags can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
-These data contain a wealth of knowledge about human and animal behaviour, but parsing this vast amount of information is a challenge for statisticians.
-
-%Many statistical techniques have been developed and applied to these types of longitudinal or functional data.  
-%Some techniques, such as time series, are highly parametric. 
-%While these parametric techniques do not allow the data to ``speak for itself", the ensuing data analysis results typically are readily interpretable, providing important subject area insights.
-%At the other extreme are black box type methods of machine learning.  
-%While these methods are incredibly successful at predicting how processes evolve, they provide little structured insight into the driving mechanisms of the process.
-%Functional Data Analysis (FDA) techniques lie somewhere in between these extremes: the output of an analysis provides visual tools for understanding processes and, often, some dimension reduction to allow for easier exploration.
-%The development of structured FDA techniques which incorporate more complex dependence structures has become a topic of recent interest  (see, e.g., Crainiceanu. and references therein).
+These data contain a wealth of knowledge about human and animal behaviour, but parsing these large data sets is a challenge for statisticians.
 
 Biologging data are sometimes modelled as a set of curves that can be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}).
-In \cite{Morris:2007}, a child's activity level throughout the day is viewed as a curve where metabolic activity is a function of time.
-The dive profile of a marine mammal can also be viewed as a curve.
-\citet{Fu:2017} show that estimating the amplitude and phase variation of elephant seal (\textit{Mirounga leonina}) dive curves is useful to classify dive types, although they ignore sequential dependence between dives.
-%However, now many such animals are tracked with high-frequency tags over a period of several dives, providing information at different time scales. 
+For example, \cite{Morris:2007} view a child's activity level throughout the day as a curve where metabolic activity is a function of time.
+\citet{Fu:2017} also show that estimating the amplitude and phase variation of elephant seal (\textit{Mirounga leonina}) dive curves is useful to classify dive types, although they ignore sequential dependence between dives.
 Similarly, our case study views the dive profile of a killer whale (\textit{Orcinus orca}) as a sequence of curves where three orthogonal axes of acceleration are each functions of time.
@@
 In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017,Adam:2019}, as is the case for our case study data.
-Namely, the dive profile (Figure \ref{fig:data}a) of a killer whale from the threatened Northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence \textit{between} dives, as dives of various time durations and depths appear to be clustered together in time.
+Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time.
 Previous studies have categorized these dives into different types where each dive type is associated with a different behaviour, including resting, travelling, and foraging \citep{Tennessen:2019b}.
-Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence \textit{within} dives, as the accelerometer data within each dive indicate bouts of short-term periodicity.
-The combination of apparent behavioural changes combined with periodic behaviour makes this accelerometer data difficult to model using simple stochastic processes.
-Another example of behaviourally-dependent periodicity is in the field of machine health, where \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks.
+Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence within dives, as the accelerometer data within each dive indicate bouts of short-term periodicity.
+Behaviourally dependent periodicity is not unique to the biologging data- in the field of machine health, \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks.
 The raw acceleration data exhibit extreme periodicity (in the form of vibration) that depends upon the behaviour of the train as the load and speed of the train change.
 
-FDA literature provides models that account for some between-curve dependence, but these methods are not adequate when this dependence is in time.
+FDA literature provides models that account for some between-curve dependence, but these methods are not adequate when modelling certain types of time dependencies, including that of Figure \ref{fig:data}.
 For example, multilevel models incorporate dependence via random effects.
-In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors use a multilevel approach since each child yields two daily activity curves; the model includes a random effect for ``child''.
+In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors use a multilevel approach with a random effects to model variation between children.
 In a classic paper that introduced this multilevel structure,  \cite{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles.
@@
 \citet{Yao:2005} use this framework to construct a method of sparse longitudinal data analysis using functional principal component analysis.
-%In terms of applications, \citet{Gruen:2017} and \citet{Xiao:2014} investigate degenerative joint disease in cats and circadian rhythms in aging patients, respectively. 
 Alternatively, within-curve structure is sometimes generated via random regression.
 For example, \citet{Rice:2001} use random regression to construct non-parametric mixed effects models for unequally sampled noisy curves.
-However, radically changing behaviour on short time-scales similar to our dive sequence is challenging to model with these classic FDA techniques.
+However, The combination of apparent behavioural changes combined with periodic behaviour makes accelerometer data such as ours difficult to model with these classic FDA techniques.
 
 In light of these challenges, we turn to vast field of methodology for the analysis of animal movement \citep{Hooten:2017}, where one of the most prevalent techniques in the literature of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
-HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of the animal from movement observations.
+HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from movement observations.
 While ubiquitous in the ecology literature, HMMs have seen little use in non-parametric functional modelling, with a few notable exceptions.
-In particular, \citet{Langrock:2018} take a non-parametric approach to model the distribution of HMM observations in terms of B-splines.
-This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments.
-However, this approach does not account for certain types of time correlation within or between the time series themselves.
-We focus on the time series themselves as functions here and show how HHMs can provide a new tool to the FDA toolbox.
+In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations in terms of B-splines.
+%This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments. 
+However, this approach does not account for certain types of time correlation within or between the time series themselves, which is the focus of our efforts here.
 
-%
-%Importantly, under the traditional HMM model, subsequent observations are assumed to be independent from one another when conditioned on the underlying behavioural process.
-%However, this assumption is often violated in the real world processes, especially when observations are taken a high frequencies. 
-%For example, the location of an animal at a given time is highly correlated with the location of that animal one second later. 
-%Several publications have dealt with this issue in the past, including the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
-%The CarHMM in particular explicitly models auto-correlation into an HMM in a way that is easily analyzed using usual HMM techniques.
-%while maintaining the structure needed to run the forward algorithm. 
-%It also only adds one additional parameter per possible hidden state.
-
-Under a traditional HMM, observations are independent given the underlying hidden state process.
-However, this assumption is often violated in real-world processes, especially when observations are taken at high frequencies.
+While useful, traditional HMMs are also not sufficient to model accelerometer data such as ours for three primary reasons.
+Firstly, subsequent observations are assumed to be independent given the underlying hidden state process, but this is often not the case when observations are taken at high frequencies.
 Several solutions have been proposed in the ecology literature, such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
 The CarHMM in particular explicitly models auto-correlation into an HMM while adding minimal complexity to the base structure of an HMM.
-Animal movement modellers have also used hierarchical hidden Markov models (HHMMs) to model simultaneous behavioural processes that occur at different time scales \citep{Barajas:2017,Adam:2019}.
-We borrow from ideas from both the CarHMM and the HHMM here.
 
-Although they are useful in many situations, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated structures at short time scales.
+In addition, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales, including our example of both between- and within-curve temporal dependencies.
+To deal with this issue, Animal movement modellers have used hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}.
+In particular, HHMMs treat both scales as HMMs where the behaviour of the fine-scale process depends on the state of the coarse-scale process.
+
+Finally, although they are useful in many situations, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated structures at short time scales.
 The HHMM of \citet{Adam:2019} fails to capture the fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}).
@@
 
+While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together.
 In this paper, we combine several existing methods from the ecology literature in novel ways to account for complex temporally dependent functional data.
 The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models and thus extend the methods available for the analysis of functional data.
-%This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. 
-%We describe a general procedure that can be used to extract features from highly structured fine-scale behaviours that otherwise could not be modelled with existing HMM models. 
-%In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. 
-%We then perform a simulation study to compare the performance of each existing model with ours in a controlled setting.
 We first describe HMMs as well as to two variants (CarHMMs and HHMMs) and discuss how Fourier analysis can handle fine-scale dependence structures.
 We also show how these methods can be combined to analyze increasingly complex data.
-In section 3, we construct several candidate models and fit them to the data from Figure \ref{fig:data}.
-Section 4 details a simulation study based on these candidate models, and in section 5 we discuss our results.
+In Section 3, we construct several candidate models and fit them to the data from Figure \ref{fig:data}.
+Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.

### 2020-09-04 Report/Sections/models.tex
Update report- Evan Edits

@@
 
-Consider a sequence of $T$ curves (or dives, in our context). Any particular dive $t$, $t = 1,
-\ldots,T$, has a possibly high-dimensional associated curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of within-curve (or fine-scale) observations $Y^*_{t} \equiv \Big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\Big\}$.
+Consider a sequence of $T$ curves (or dives, in our context). For any particular $t = 1,
+\ldots,T$, dive $t$ is characterized by a possibly high-dimensional curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$.
 The sequence $Y^*_{t}$ is made up of function values from curve $t$.
-We call the sequence of coarse-scale observations $Y \equiv \Big\{Y_1, \ldots, Y_T\Big\}$ and the collection of all fine-scale observations $Y^* \equiv \Big\{Y^*_1,\ldots,Y^*_T \Big\}$.
+We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.
 
@@
 To simultaneously model $Y$ and $Y^*$, we use the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
-%which is comprised of two distinct, but related, HMMs that jointly model these time series. 
+%which is comprised of two distinct,x but related, HMMs that jointly model these time series. 
 Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics.
@@
 
-An HMM is comprised of a sequence of unobserved states $X = \big\{X_1, \ldots, X_T\big\}$ where $X_t$ is associated with the observation $Y_t$.
-Here we focus on the coarse scale sequence of observations $Y_1,\ldots, Y_T$, but an HMM can also be used as a sub-model for the fine-scale sequence for a particular dive.
+An HMM is comprised of a sequence of unobserved states $X \equiv \big\{X_1, \ldots, X_T\big\}$ where $X_t$ is associated with the observation $Y_t$.
+Here we focus on the coarse scale sequence of observations $Y_1,\ldots, Y_T$, but an HMM can also be used as a sub-model of the fine-scale sequence for a particular dive.
 The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time.
@@
 
-Following \citet{Barajas:2017}, we parameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows to sum to 1:
+Following \citet{Barajas:2017}, we reparameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows to sum to 1:
 %
@@
 %
-If $Y_t$ is a vector, equation (\ref{eqn:carhmm}) above can model either a subset of the dimensions of $Y_t$ if only a subset of those dimensions exhibit significant auto-correlation.
+If $Y_t$ is a vector, Equation (\ref{eqn:carhmm}) above can model either all dimensions of $Y_t$ simultaneously or just a subset of the dimensions, depending upon considerations such as subject mater expertise or expository data analysis.
 Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.
 
-The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then
+The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then:
 \begin{equation*}
-    \calL_{\text{CarHMM}}(\theta,\Gamma;y) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\theta) \mathbf{1}_N
-    \label{CarHMM_likelihood}
+    \calL_{\text{CarHMM}}(\theta,\Gamma;y) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\theta) \mathbf{1}_N,
+    \label{eqn:CarHMM_likelihood}
 \end{equation*}
@@
 
-An HHMM accounts for different levels of correlation by modelling both the coarse-scale process and fine-scale process with distinct HMMs \citep{Barajas:2017,Adam:2019}.
-The coarse-scale process is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations.
+An HHMM accounts for different scales of correlation by modelling both the coarse-scale process and fine-scale process with distinct HMMs \citep{Barajas:2017,Adam:2019}.
+The coarse-scale model is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations.
 %
-In the hierarchical setting, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv (X_{t,1}^*,\ldots, X_{t,T_t^*})$.
-The fine-scale state sequence in turn emits a sequence of fine-scale observations $Y_t^* \equiv (Y_{t,1}^*,\ldots, Y_{t,T_t^*})$.
-For each $t$, the fine-scale process ($X_t^*, Y_t^*$) then follows another HMM whose parameters depend on the value of $X_t$.
+In the hierarchical setting, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$.
+For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM whose parameters depend on the value of $X_t$.
 Specifically, if $X_t=i$, then the components of $X_t^*$ are a Markov chain with an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
@@
 We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^*)}\right)$.
-Given the coarse-scale hidden states, the $T$ fine-scale HMMs $(X_1^*, Y_1^*), \ldots, (X_T^*, Y_T^*)$, are all assumed to be independent of one another.
+Given the coarse-scale hidden states, the $T$ fine-scale HMMs $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$, are all assumed to be independent of one another.
 
 Depending upon the process being modelled, it is possible to force certain parameters to be shared to reduce the complexity of the HHMM.
-For example, in the killer whale case study (section \ref{sec:data}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( \text{i.e.
+For example, in the killer whale case study (see Section \ref{sec:data}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( \text{i.e.
 } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
@@
 %
-\[
+\begin{equation}
     \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\theta,\theta^*,\Gamma^*) \mathbf{1}_N
-\]
+    \label{eqn:HHMM_likelihood}
+\end{equation}
 %
@@
 Subject-area experts are often required to determine the optimal window size and the specific summary statistics used.
-For our case study, we use the discrete Fourier transform (DFT) of a moving window of width $h$ across $Y^*_t$:
+For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ across $Y^*_t$:
 %
@@
 %
-In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared 2-norm of the component of $\left(Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\right)$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window (by the Plancherel theorem).
+In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared two-norm of the component of the window $\left(Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\right)$ that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window (by the Plancherel theorem).
 More intuitively, $\Ztwo_{t,t^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within the window starting at time stamp $t^*$.
@@
 Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude.
-To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \left(\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\right)$.
-The down-sampling avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
+
+To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \left\{\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\right\}$ and $X^*_t = \left\{X^*_{t,1},X^*_{t,h+1},X^*_{t,2h+1},\ldots\right\}$.
+When down-sampling, we redefine the fine-scale probability transition matrices such that $\Gamma^{*(i)}_{jk} = \Pr(X^*_{t,t^*+h} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1,h+1,2h+1,\ldots$.
+%In addition, we assume that the hidden state within a window does not change, (i.e.
+$X^*_{t,t^*+s} = X^*_{t,t^*}$ for all $t$, $t^* = 1,h+1,2h+1,\ldots$ and $s = 0,1,\ldots,h-1$.
+This down-sampling both avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
 By summarizing the data using a relevant transformation, the method overall also reduces correlation in the raw observations due to the complex dependence structure.
@@
 where $P(y_t,y^*_t|y_{t-1};\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\theta^{(i)}\right)\calL_{\text{HMM}}\left(\theta^{*(i)};y^*_t\right)$.
-On the fine scale, $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods (or penalized likelihoods).
-The likelihood $\calL_{\text{HMM}}$ is then replaced by the likelihood of this general fine-scale model, $\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
+
+On the fine scale, $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods.
+In this case, the fine-scale HMM likelihood term $\calL_{\text{HMM}}$ in Equation (\ref{eqn:HHMM_likelihood}) is replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
 \[
@@
 where $P(y_t,y^*_t;\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\theta^{(i)}\right)\calL_{\text{fine}}\left(\theta^{*(i)};y^*_t\right)$.
-
+%
 Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT) in addition to many others not described here. 
 For example, a continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}.
-In particular, \citet{Xu:2018} model biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours.
+as an example, \citet{Xu:2018} explain biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours.
 However, they take a longitudinal approach whereas we use a coarse-scale HMM in our case study.
-Further, \citet{Bebbington:2007} and \citet{Borchers:2013} model data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model.
-If the fine-scale model is a simple Poisson process, this model reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
-In addition, \citep{Langrock:2018} use B-splines to model several aspects of an HMM, and this modified model admits a penalized likelihood which can be incorporated into the hierarchical structure here.
+Further, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model.
+If the fine-scale model is a simple Poisson process, then the hierarchical structure described here reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
+As a final example, \citep{Langrock:2018} use B-splines to model the emission distribution of an HMM as well as incorporate covariates.
+This non-parametric approach admits a penalized likelihood which can easily be incorporated into the hierarchical structure here.
 
@@
 Alternatively, one may construct an \textbf{HHHMM}, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$) and traditional HMMs are used to model all three levels.
-It is clear that a a myriad of possible models can be built using this framework.
+It is clear that are a myriad of possible models can be built using this framework.
 However, these models quickly become complicated and should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.

### 2020-09-04 Report/Sections/results.tex
Update report- Evan Edits

@@
 
-Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact behaviour \citep{Sutherland:1998}.
+Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact animal behaviour \citep{Sutherland:1998}.
 HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g.
@@
 NRKWs prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKWs for predation \citep{Williams:2009,Noren:2011}.
-It is therefore of interest to determine the energetic expenditure of NRKWs to forage for Chinook, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
+It is therefore of interest to determine the energetic expenditure of NRKWs foraging for Chinook as well as the frequency at which NRKWs catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
 Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the behavioural state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
@@
 The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions.
-Observations were collected at a rate of 50 Hz using Customizable Animal Solutions Tracking (CATs) tags (model number ###).
+Observations were collected at a rate of 50 Hz using Customizable Animal Solutions Tracking (CATs) tags (model number \#\#\#).
 The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left).
@@
 Tagging the killer whale caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored.
-We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm, when the tagging technology malfunctioned.
-We smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second.
-We then define a killer whale ``dive" as any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
+We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm due to a malfunction in the tagging technology.
+To process the data, we smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second.
+We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
 Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}.
@@
 Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
-Specifically, the vector of observed coarse-scale observations $y = \Big\{y_1,\ldots,y_{267}\Big\}$ is a time series of dive durations in seconds ($s$) with corresponding vector of unobserved dive types $x = \Big\{x_1,\ldots,x_{267}\Big\}$.
-For dive $t$, the fine-scale observations are contained in $y^*_{t}$, which is a vector of the within-dive acceleration data in meters per second squared ($m/s^2$).
-The corresponding vector of subdive states is $x^*_{t}$.
-Naturally, $y^* = \Big\{y^*_1,\ldots,y^*_{267}\Big\}$ and $x^* = \Big\{x^*_1,\ldots,x^*_{267}\Big\}$ is the corresponding set of subdive states.
+Specifically, the vector of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds ($s$).
+For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots \right\}$, which is a vector of the within-dive acceleration data in meters per second squared ($m/s^2$).
+Naturally, the collection of all acceleration data is $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.
 
@@
 
-Building a suitable model for these data involves selecting appropriate model structures, emission distributions, and numbers of states on both the coarse and fine scales.
+Defining a suitable model for these data involves selecting an appropriate model structure, emission distributions, and the correct number of hidden states for both the coarse and fine scale observations.
 
-To choose how to model the coarse-scale observations (dive duration), we begin by selecting the model structure.
-Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations.
-Visual inspection also shows no obvious intricate structure.
+We first select an appropriate model structure for the coarse-scale observations of dive duration.
+Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Section 1.1 of the supplementary material), and visual inspection also shows no obvious intricate structure.
 Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor transformation is called for.
@@
 %
-This is consistent with previous studies, including \citet{Barajas:2017}. 
+This is consistent with previous studies, including that of \citet{Barajas:2017}. 
 %
@@
 This is unfortunately not the case for our killer whale data, but we still consult lag plots together with a heuristic approach based on visual inspection to choose $N = 2$ dive types.
-The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).
+The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
 
-To choose how to model the fine-scale observations (acceleration), we again begin with determining model structure.
+To model the fine-scale observations of acceleration, we again begin with determining model structure.
 In contrast to the coarse-scale, the fine-scale acceleration data exhibit significant sinusoidal behaviour.
@@
 %
-We calculate $\Ztwo_{t,t^*}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to maximum frequencies of 5 Hz for each axis of acceleration.
-We also down-sample each observation $z^*_t$ such that $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\Big\}$ to avoid overlapping windows.
-The resulting sequence of observations is $z^* = \Big\{z^*_1,\ldots,z^*_{267}\Big\}$, where $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},\ldots\Big\}$ and $z^*_{t,t^*} = \{a^*_{t,t^*},w^*_{t,t^*}\}$.
+We then calculate $\Ztwo_{t,t^*}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to maximum frequencies of 5 Hz for each axis of acceleration.
+We also down-sample each observation $z^*_t$ such that $z^*_t = \left\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\right\}$ and $x^*_t = \left\{x^*_{t,1},x^*_{t,101},x^*_{t,201},\ldots\right\}$ to avoid overlapping windows.
+The resulting sequence of observations is $z^* = \left\{z^*_1,\ldots,z^*_{267}\right\}$, where $z^*_t = \left\{z^*_{t,1},z^*_{t,101},\ldots\right\}$ and $z^*_{t,t^*} = \left\{a^*_{t,t^*},w^*_{t,t^*}\right\}$.
 We have that $a^*_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $w^*_{t,t^*}$ is a scalar describing the ``wiggliness" of a given window.
 %
-After transforming the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (see section 1.1 of the supplementary material).
-Therefore, we assume that all three components of $\Zone_{t,t^*}$, $t^* = 1,101,\ldots$ follow the CarHMM as defined in section 2.2.
+After transforming the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material).
+Therefore, we assume that all three components of $\Zone_{t,t^*}$, $t^* = 1,101,\ldots$ follow the CarHMM as defined in Section 2.2.
+Due to down-sampling we redefine the probability density function $f$ such that $f^{*(i,i^*)}(\cdot|z^*_{t,t^*-100}; \theta^{*(i,i^*)})$ describes the density $\Z_{t,t^*}$ given that $X_t = i$, $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.
+We also assume that observations of all three axes of acceleration are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$.
 %
@@
 %
-While the $\Ztwo_{t,t^*}$ observations also exhibit some auto-correlation prior to fitting the model, it is less strong than that of $\Zone_{t,t^*}$, and 
+While the observations of wiggliness ($\Ztwo_{t,t^*}$) also exhibit some auto-correlation prior to fitting the model, this correlation is less strong than $\Zone_{t,t^*}$, and 
 %the biological interpretation of auto-correlation between observations of $\Ztwo_{t,t^*}$ is less clear. In addition, 
-much of the auto-correlation evident from the lag plot may be explained by subsequent observations having the same subdive state.
+much of the auto-correlation evident from the lag plot may be explained by subsequent observations occurring within the same subdive state.
 Therefore, a simple HMM structure is used to model $\Ztwo_{t,t^*}$ for all $t$ and $t^*$.
+We assume that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$.
 %
@@
 %
-
-Next, we select the emission distributions for both $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$.
+%
+After defining the model structure, we select the emission distributions for both $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$.
 We assume that the fine-scale emission parameters are shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity.
-This implies that the subdive states in dive type 1 have the same interpretation as those in dive type 2.
+This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.
 %
@@
 In other words, each axis of acceleration has its own mean and variance, but all axes have the same auto-correlation parameter.
-We assume that observations of all three axes of acceleration are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$.
 %
-On the other hand, we assume that $\Ztwo_{t,t^*}$ is gamma distributed for all $t$ and $t^*$.
+We assume that $\Ztwo_{t,t^*}$ is gamma distributed for all $t$ and $t^*$.
 Further, given that $X^*_{t,t^*} = i^*$, the distribution of $\Ztwo_{t,t^*}$ is parameterized by its mean $\mu_W^{*(\cdot,i^*)} \in \mathbb{R}$ and standard deviation $\sigma_W^{*(\cdot,i^*)} \in \mathbb{R}$.
-We assume that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$.
-
+%
 Finally, to select the number of subdive states $N^*$ we again rely on a heuristic approach based on visual inspection to select $N^* = 3$.
-Although $N^*$ is selected heuristically, we test the validity of this model using techniques in section \ref{subsec:model_validation}.
+Although $N^*$ is selected heuristically, we test the validity of this model using techniques in Section \ref{subsec:model_validation}.
 
 The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale.
-On the fine scale, each fine-scale sequence $y^*_t$ is transformed to $z^*$ using Equation (\ref{eqn:z}), $\Zone_{t,t^*}$ is modelled with a CarHMM, and $\Ztwo_{t,t^*}$ is modelled with an HMM for all $t$ and $t^*$.
-We assume that $\Ztwo_{t,t^*}$ in addition to the components of $\Zone_{t,t^*}$ are all independent of one another and that the subdive types have identical emission distributions across all dive types.
+Each fine-scale sequence $y^*_t$ is transformed to $z^*$ using Equation (\ref{eqn:z}), $\Zone_{t,t^*}$ is modelled with a CarHMM, and $\Ztwo_{t,t^*}$ is modelled with an HMM for all $t$ and $t^*$.
+We model $\Zone_{t,t^*}$ to be Normally distributed and $\Ztwo_{t,t^*}$ to be Gamma distributed.
+We also assume that $\Ztwo_{t,t^*}$ in addition to the components of $\Zone_{t,t^*}$ are all independent of one another and that the subdive types have identical emission distributions across all dive types.
 In total, the parameters to estimate are
@@
     %
-    \theta = \{\{\mu^{(1)},\sigma^{(1)}\},\{\mu^{(2)},\sigma^{(2)}\}\} \qquad \text{($Y$ emission parameters), and} \\
+    \theta = \{\mu^{(1)},\sigma^{(1)},\mu^{(2)},\sigma^{(2)}\} \qquad \text{($Y$ emission parameters), and} \\
     %
@@
 We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMMs discussed so far.
-Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model for the CarHHMM-DFT.
+Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model.
 The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
@@
 \begin{enumerate}
-    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Zone_{t,t^*}$ and $\Ztwo{t,t^*}$ as emissions of a simple HMM rather than a CarHMM.
+    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ as emissions of a simple HMM rather than a CarHMM.
     \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emssions of a CarHMM.
@@
 To illustrate an application of this method and compare the candidate models, we fit all four models to the data shown in Figure \ref{fig:data}.
-We first report the results from the full CarHHMM-DFT in detail and then assess the quality of the fit.
+We first report the results from the full CarHHMM-DFT in detail and assess the quality of the fit.
 We then compare the results of the CarHHMM-DFT with those of the other candidate models.
 
-The coarse-scale emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:coarse_emis}).
+The coarse-scale emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}).
 Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, travelling, and searching for prey.
 On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as foraging and socialization \citep{Tennessen:2019b}.
-However, no dive in this data set has a maximum depth deeper than $\approx$ 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100m.
+However, no dive in this data set has a maximum depth deeper than $\approx$ 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100 meters.
 It is therefore unlikely that any of the dives used in this study are successful foraging dives.
 
-The three subdive behaviours have distinct average ``wiggliness" values, as the means corresponding to $\Ztwo_{t,t^*}$ are separated by an order of magnitude for each subdive state (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}).
+The three subdive behaviours have distinct average ``wiggliness" values, as the means corresponding to $\Ztwo_{t,t^*}$ are separated by an order of magnitude for each subdive state (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}).
 Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,t^*}$ and the smallest variance corresponding to $\Zone_{t,t^*}$.
@@
 This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
-The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 much more often when performing long deep dives compared to short shallow dives (40\% of the time for dive type 1 and 9\% of the time for dive type 2).
+The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 40\% of the time when in a dive of type 1 but only 9\% of the time when in a dive of type 2.
 Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}.
@@
 Histograms of the pseudoresiduals mostly support that this model is well-specified.
-One exception is $\Ztwo_{t,t^*}$, whose pseudoresiduals are noticeably right-skewed (Figure \ref{fig:pseudoresids}).
-This implies that the true distribution of $\Ztwo_{t,t^*}$ may follow a heavier-tailed distribution than the gamma distribution used in the case study.
-See sections 1.4 - 1.6 of the supplementary material for pseudoresidual plots for all observations.
+One exception is $\Ztwo_{t,t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:pseudoresids}).
+This implies that the true distribution of $\Ztwo_{t,t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study.
+See Sections 1.4 through 1.6 of the supplementary material for pseudoresidual plots for all observations.
 
@@
 This results in two histograms - one corresponding to dive type 1 and another corresponding to dive type 2.
-Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}).
-See sections 1.4 - 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $z^*_{t,t^*}$.
-Our results again mostly show that the CarHHMM-DFT fits the data well, but there are some exceptions.
+Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT.
+See Sections 1.4 through 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $Z^*_{t,t^*}$.
+Our results again mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions.
 In particular, $\Ztwo_{t,t^*}$ is right-skewed and $\Zone_{t,t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
@@
 
-The CarHMM-DFT, which lacks a hierarchical structure, produces similar fine-scale parameter estimates and state estimates compared to those of the CarHHMM-DFT However, its lack of a hierarchical structure means that it fails to differentiate between short and long dives.
-This model therefore does not give any information regarding the dive-level Markov chain or the relationship between the dive and subdive levels.
-For example, the CarHMM-DFT does not indicate that the whale is much more likely to be in subdive state 1 when engaged in longer dives compared to shorter one.
-
-The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3.
+The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3 (see Figures 4 and 5 of the supplementary material).
 In addition, the HHMM-DFT produces estimates of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$ which are approximately $50$ to $100$ percent larger than those of the CarHHMM-DFT for all axes of acceleration.
-The estimated uncertainties of the estimates of $\mu_A^{*(\cdot,1)}$, $\mu_A^{*(\cdot,2)}$, and $\mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT for all axes of acceleration.
-Further, the empirical auto-correlation of acceleration within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers.
+The estimated uncertainties of the estimates of $\mu_A^{*(\cdot,1)}$, $\mu_A^{*(\cdot,2)}$, and $\mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT for all axes of acceleration (see Tables 1 and 2 of the supplementary material).
+Further, the empirical auto-correlation of acceleration within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT do not follow a standard normal distribution due to several outliers (see Figure 15 of the supplementary material).
 These findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.
 
-Finally, the CarHHMM does not use the ``wiggliness" of the acceleration data as an observation, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives}.
-This fact alone essentially disqualifies the CarHHMM as a viable model for these data.
-The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution.
-See section 1 of the supplementary material for a more complete set of results for each of the candidate models.
+The CarHHMM does not model the ``wiggliness" of the acceleration data, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives}.
+This essentially disqualifies the CarHHMM as a viable model for these data.
+The pseudoresiduals corresponding to the CarHHMM also do not appear to fit a Normal distribution.
+
+Finally, the CarHMM-DFT, which lacks a hierarchical structure, produces similar fine-scale parameter estimates and state estimates compared to those of the CarHHMM-DFT.
+However, its lack of hierarchical structure means that it fails to differentiate between short and long dives.
+This model therefore does not give any information regarding the dive-level Markov chain or the relationship between the dive and subdive levels.
+In particular, the CarHMM-DFT does not indicate that the whale is more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives.
+See Section 1 of the supplementary material for a more complete set of results for each of the candidate models.

### 2020-09-04 Report/Sections/simulation.tex
Update report- Evan Edits

@@
 To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT.
-The parameters used to generate the data are loosely based on those estimated in the case study (see able \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity.
-For example, we set $\Gamma$ such that the simulated data has approximately equal numbers of short and long dives.
-This allows meaningful conclusions to be made for both dive types.
-In addition, the number subdive states is set to $N^*=2$ and $\Ztwo_{t,t^*}$ is set to a scalar.
+The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity.
+In particular, we set $\Gamma$ such that the simulated data has equal numbers of short and long dives in expectation, the number subdive states is set to $N^*=2$, and $\Ztwo_{t,t^*}$ is set to a scalar.
 Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times.
@@
 \subsection{Data simulation}
+\label{subsec:data_simulation}
 
 We generate 500 independent data sets from the CarHHMM-DFT.
-Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives.
+Each data set consists of a sequence of 100 curves which each represent a simulated killer whale dives.
 Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
@@
 %
-Dive duration is gamma distributed, and the coarse-scale emission parameters $\Theta$ are
+Dive duration is gamma distributed, and the coarse-scale emission parameters $\theta$ are
 $$
@@
 %
-After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second segments which make up the fine-scale hidden Markov model.
-Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left(X^*_{t,1}, \ldots, X^*_{t,T^*_t} \right)$ with probability transition matrices
+After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows.
+Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
 %
@@
 %
-The observations $Z^*$ are generated using the following parameters:
+For each dive and subdive segment, we simulate the fine-scale transformed observations $Z^*_t \equiv \left\{Z^*_{t,1}, \ldots, \right\}$, where $Z^*_{t,t^*} = \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$, using the following parameters:
 %
@@
 Recall that $\Zone_{t,t^*}$ follows a CarHMM and is Normally distributed while $\Ztwo_{t,t^*}$ follows an HMM and is Gamma distributed.
-It is not possible to uniquely recreate the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix.
+It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping in the appendix.
 Figure \ref{fig:sim_data} shows this reconstructed sequence along with one realization of $Z^*$ for five dives of one simulated data set.
 
-The two dive types differ in that dives of type 1 are much shorter on average than dives of type 2.
+The two simulated dive types differ in that dives of type 1 are much shorter on average (20 seconds) than dives of type 2 (80 seconds).
 The two subdive types differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher in subdive state 2 than subdive state 1.
@@
 
-As expected, since it is the generating model, the full CarHHMM-DFT is the best-performing model.
-Using the forward-backward algorithm, the CarHHMM-DFT decodes the correct subdive type given the fine-scale data with an average confidence of $1 - 10^{-12}$ across all simulations.
-In other words, it predicts the \textit{incorrect} subdive state with an average probability of $10^{-12}$.
-For dive type, the average confidence was above 0.9 (see Table \ref{table:accuracy}).
-All parameter estimates have statistically insignificant biases except for $\hat \sigma$ and $\hat \phi$, whose MLEs are known to be biased (see section 2 of the supplementary material).
-Even still, these biases are the least severe for the CarHHMM-DFT among the four models.
-The empirical standard error of all parameter estimates ($\hat \Theta$, $\hat \Gamma$, $\hat \Theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slight underestimates of the empirical standard error.
+The full CarHHMM-DFT is the best-performing model of the four candidates, which is to be expected as the generating model.
+If $X^*_{t,t^*}$ is the unknown subdive state and $x^*_{t,t^*}$ is the known true subdive state for dive $t$ and window $t^*$, then using the forward-backward algorithm we find that $\Pr(X^*_{t,t^*} = x^*_{t,t^*}|Y,Y^*)$ is $(1 - 10^{-12})$ on average across all simulations, dives, and windows.
+We refer to this quantity as the average accuracy of the model.
+For dive type, this average accuracy was above 0.9 (see Table \ref{table:accuracy}).
+All parameter estimates of mean values and fine-scale probability transition matrices, $\hat \mu$ and $\hat \Gamma^*$, respectively, have statistically insignificant biases.
+In addition, biases for $\hat \sigma$, $\hat \phi$, and $\hat \Gamma$ are no more severe for the CarHHMM-DFT than the other three models.
+The empirical standard error of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard error.
 See Tables 6-9 of the supplementary material for more information.
 
-The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its decoding accuracy and parameter bias.
-However, the HHMM-DFT differed from the CarHHMM-DFT in its estimate of the standard deviation of acceleration, $\hat \sigma_A^*$.
-For both subdive types, $\hat \sigma_A^*$ greatly overestimates the true standard deviation.
-The estimated standard errors of $\hat \mu_A^*$ and $\hat \sigma_A^*$ using the observed Fisher information are both much smaller than the empirical standard error (see Table 7 of the supplementary material).
-This implies that the variance of acceleration can be overestimated and the standard error of the parameter estimates can be underestimated when auto-correlation is ignored.
-These findings are consistent with the results from the case study, where the HHMM-DFT produced higher variance estimates and narrower standard error estimates than the CarHHMM-DFT.
+The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its average accuracy and parameter bias.
+However, the HHMM-DFT differs from the CarHHMM-DFT in its estimates of $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$, both of which are greatly overestimated.
+In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material).
+These results imply that estimated standard errors of parameter estimates are too small when using the HHMM-DFT since auto-correlation is ignored.
+This finding is consistent with the results from the case study, where the HHMM-DFT produced higher estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.
 
-The CarHHMM is the worst-performing model in terms of decoding accuracy, with an average confidence below $0.9$ when determining both dive and subdive types.
-This result is consistent with expectations because the CarHHMM does not have access to the ``wiggles" of the fine-scale process, which are the most informative observations when determining subdive type.
+The CarHHMM is the worst-performing model in terms of decoding accuracy, with an average accuracy below $0.9$ for both dive and subdive types.
+This result is consistent with expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct observation when determining subdive type.
 The CarHHMM is also much more likely to misclassify subdive type 1 than subdive type 2 (see Table \ref{table:accuracy}).
-The CarHHMM is also the worst of the four at estimating parameters in part due to its relatively poor decoding accuracy.
-The fine-scale probability transition matrices associated with dive type 2 ($\hat \Theta^{*(2)}$ and $\hat \Gamma^{*(2)}$) are especially poor estimates (see Table 9 and Figure 33 of the supplementary material).
+In addition to its poor average accuracy, the CarHHMM is also the worst of the four candidate models at estimating parameters, with the fine-scale probability transition matrices associated with dive type 2 ($\hat \theta^{*(2)}$ and $\hat \Gamma^{*(2)}$) being especially poor estimates (see Table 9 and Figure 33 of the supplementary material).
 
@@
 In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}).
-However, it lacks any hierarchical structure, so it is unable to estimate dive types.
-The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of \textit{all} dives.
-The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated and $\hat \sigma$ underestimates the true standard deviation of all dive types by $\approx$ 5 seconds.
-See Figures \ref{fig:acc_coarse} \& \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive/subdive type probabilities as determined by each model.
+However, this model is unable to estimate dive types as it lacks any hierarchical structure.
+The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of all dives.
+The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by approximately five seconds.
+See Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive types and subdive states for each model.

### 2020-09-04 Report/main.tex
Commit Before Andrew Access
Updates from Overleaf
Clean up report and delete old files
Update report- Evan Edits

@@
 \documentclass[12pt]{TD-CJS}
+%\documentclass{TD-CJS}
 
@@
 \newcommand{\newblock}{}
-\bibliographystyle{abbrvnat}
+\bibliographystyle{apalike}
 
@@
 Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields.
-These data often exhibit complicated temporal dependence structure which cannot be modelled using common methods of Functional Data Analysis (FDA).
-Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs), but many intricate fine-scale processes violate basic assumptions of HMMs.
+These data often exhibit complicated temporal dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA).
+Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs) instead, but many intricate fine-scale processes violate basic assumptions of HMMs.
 We detail a generalization of the hierarchical HMM approach that incorporates a wider class of parametric models at fine scales. 
@@
 As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada. 
-Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing models.
+Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing baselines.
 
@@
 %
-\section{Introduction}
+\section{INTRODUCTION}
 \input Sections/intro.tex
 %
-\section{Models and parameter estimation}
+\section{MODELS AND PARAMETER ESTIMATION}
 \label{sec:models}
@@
 %
-\section{Killer Whale Case Study}
+\section{KILLER WHALE CASE STUDY}
 \label{sec:data}
@@
 %
-\section{Simulation Study}
+\section{SIMULATION STUDY}
 \label{sec:sim_study}
@@
 %%
-\section{Discussion}
+\section{DISCUSSION}
 \input Sections/discussion.tex
+%
+%\newpage
+%\input Sections/bibliography.tex
 %
@@
 \newpage
-\section{Figures and Tables}
+\section{FIGURES AND TABLES}
 \input Sections/figures.tex

### 2020-10-30 Report/CJS_Sample.tex
commit before adding t_tilde

@@
 \KWDtitle{Key words and phrases}
-Association parameters\sep clustered data\sep mean parameters\sep missing data\sep pairwise
+Hidden Markov Models\sep biologging\sep multi-state model\sep missing data\sep pairwise
 likelihood\sep repeated measurements.
 % MSC 2010 subject classification codes
-\KWDtitle{MSC 2010}Primary 62???\sep secondary 62???}
+\KWDtitle{MSC 2010}Primary 62M05 \sep secondary 62P12}
 \begin{abstract}

### 2020-10-30 Report/Sections/appendix.tex
commit before adding t_tilde

@@
 % !TeX root = ../main.tex
-
-\subsection{Equivalence of state-switching OU process and CarHMM}
-
-\citet{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
-%
-\begin{equation}
-    \label{eqn:SDE}
-    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t,
-\end{equation}
-%
-where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process in state $X_t$, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
-They refer to this as a state-switching Ornstein-Uhlenbeck process.
-Unlike an HMM, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer in this case.
-If the behavioural state $X_t$ is known and does not change between observations $\Big($i.e $X_s = X_t$ for all $s \in [t,t+\Delta t)\Big)$, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
-%
-\begin{align}
-    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
-    \label{eqn:OU_sol}
-\end{align}
-%
-where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$.
-In fact, under certain conditions, the output of a state-switching OU process is identical to that of a CarHMM, as described by Theorem 1 below.
-
-\begin{theorem}{Theorem 1.}{}%
-If the following conditions are met:
-\begin{enumerate}
-    \item The hidden behavioural process $X$ from Equation (\ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$,
-
-    \item Observations of the SDE from Equation (\ref{eqn:OU_sol}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
-\end{enumerate}
-then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
-
-\begin{align}
-\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{(i)} = \sqrt{\frac{\omega^{(i)^2}}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t})}, \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
-\end{align}
-
-\end{theorem}
-
-\begin{proof}{Proof of Theorem 1}{}
-Combining Equation (\ref{eqn:OU_sol}) with Equation (\ref{eqn:CarHMM_to_OU}) gives
-%
-\begin{align*}
-    \left(Y_{(s+1) \Delta t} | X_{s\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{s \Delta t}, \enspace \left(\sigma^{(i)}\right)^2 \right),\\
-    s = 0, \ldots, T-2.
-\end{align*}
-%
-If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:
-
-\begin{align*}
-    \left(Y'_s| X'_s = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y'_{s-1}, \enspace \left(\sigma^{(i)}\right)^2 \right)\\
-    s = 2, \ldots, T.
-\end{align*}
-%
-$X'$ is assumed to follow a (possibly unobserved) Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s-1})$ is consistent with that of a CarHMM with Normal emission distributions.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$. 
-
 To simulate $X^*_t$ for each dive $t$, note that we simulate accelerometer readings taken at 50 hertz and have two-second window lengths.
@@
 Recall that $z^*_t = \left\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\right\}$ and that $f^{*(\cdot,i^*)}(\cdot|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\iffalse
+
+\subsection{Equivalence of state-switching OU process and CarHMM}
+
+\citet{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
+%
+\begin{equation}
+    \label{eqn:SDE}
+    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t,
+\end{equation}
+%
+where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process in state $X_t$, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
+They refer to this as a state-switching Ornstein-Uhlenbeck process.
+Unlike an HMM, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer in this case.
+If the behavioural state $X_t$ is known and does not change between observations $\Big(X_s = X_t$ for all $s \in [t,t+\Delta t)\Big)$, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
+%
+\begin{align}
+    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right),
+    \label{eqn:OU_sol}
+\end{align}
+%
+where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$.
+In fact, under certain conditions, the output of a state-switching OU process is identical to that of a CarHMM, as described by Theorem 1 below.
+
+\begin{theorem}{Theorem 1.}{}%
+If the following conditions are met:
+\begin{enumerate}
+    \item The hidden behavioural process $X$ from Equation (\ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$,
+
+    \item Observations of the SDE from Equation (\ref{eqn:OU_sol}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
+\end{enumerate}
+then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\theta = \{\theta^{(1)}, \ldots, \theta^{(N)}\}; \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
+
+\begin{align}
+\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{(i)} = \sqrt{\frac{\omega^{(i)^2}}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t})}, \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
+\end{align}
+
+\end{theorem}
+
+\begin{proof}{Proof of Theorem 1}{}
+Combining Equation (\ref{eqn:OU_sol}) with Equation (\ref{eqn:CarHMM_to_OU}) gives
+%
+\begin{align*}
+    \left(Y_{(s+1) \Delta t} | X_{s\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{s \Delta t}, \enspace \left(\sigma^{(i)}\right)^2 \right),\\
+    s = 0, \ldots, T-2.
+\end{align*}
+%
+If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:
+
+\begin{align*}
+    \left(Y'_s| X'_s = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y'_{s-1}, \enspace \left(\sigma^{(i)}\right)^2 \right)\\
+    s = 2, \ldots, T.
+\end{align*}
+%
+$X'$ is assumed to follow a (possibly unobserved) Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s-1})$ is consistent with that of a CarHMM with Normal emission distributions.
+\end{proof}
+
+\fi

### 2020-10-30 Report/Sections/models.tex
commit before adding t_tilde

@@
 The first emission $Y_1$ is assumed to be fixed and known as an initial value which does not depend upon $X_1$.
-%
-We denote the conditional density of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
-For example, if $Y_t$ is a scalar, one could assume that its conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
+We denote the conditional density of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$.
+This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$.
+As a concrete example, if $Y_t$ is a $K$-dimensional vector with components $Y_t(1),\ldots,Y_t(K)$, consider the component-wise sequences $Y_1(k), \ldots, Y_T(k)$ for $k = 1,\ldots K$.
+One may assume that these sequences are independent given $X_1,\ldots,X_T$ with $Y_t(k)|X_t = i$ and that $Y_t(k)$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)}_k = \{\mu^{(i)}_k,\sigma^{(i)}_k,\phi^{(i)}_k\}$, where:
 %
@@
 \begin{split}
-\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) &= \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}, \text{ and } \\
-\mathbb{V}(Y_t| Y_{t-1} = y_{t-1}, X_t = i) &= (\sigma^{(i)})^2.
+\mathbb{E}(Y_{t}(k)|Y_{t-1}(k) = y_{t-1}(k),X_t=i) &= \phi^{(i)}_k ~ y_{t-1}(k) ~+ ~(1-\phi^{(i)}_k)  ~\mu^{(i)}_k, \\
+\mathbb{V}(Y_t(k)| Y_{t-1}(k) = y_{t-1}(k), X_t = i) &= (\sigma^{(i)}_k)^2.
 \end{split}
 \end{align}
-%
-If $Y_t$ is a vector, Equation (\ref{eqn:carhmm}) above can model either all dimensions of $Y_t$ simultaneously or just a subset of the dimensions, depending upon considerations such as subject mater expertise or expository data analysis.
-Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.
 
-The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then:
+A CarHMM which follows Equation (\ref{eqn:carhmm}) can be viewed as a discrete time version of the state-switching Ornstein-Uhlenbeck process of \citet{Michelot:2019} if observation error is negligible.
+This follows in the same way that an AR(1) process is the discrete-time version of a traditional Ornstein-Uhlenbeck process.
+Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a general CarHMM.
+The likelihood corresponding to a general CarHMM can be easily calculated using the forward algorithm.
+As previously, let $y$ be the vector of observed emissions.
+Then:
 \begin{equation*}
@@
 
-An HHMM accounts for different scales of correlation by modelling both the coarse-scale process and fine-scale process with distinct HMMs \citep{Barajas:2017,Adam:2019}.
-The coarse-scale model is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations.
+An HHMM accounts for different scales of correlation by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs.
+The coarse-scale model is as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations with state-dependent parameters $\theta^{(i)}, i = 1,\ldots,N$.
 %
 In the hierarchical setting, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$.
-For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM whose parameters depend on the value of $X_t$.
-Specifically, if $X_t=i$, then the components of $X_t^*$ are a Markov chain with an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
-For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary in general.
-The distribution of $Y_{t,t^*}$ given $X_{t,t^*}=i^*$ and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$.
-We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^*)}\right)$.
-Given the coarse-scale hidden states, the $T$ fine-scale HMMs $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$, are all assumed to be independent of one another.
+For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
+If $X_t = i$, then the components of $X_t^*$ are a Markov chain with an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
+The distribution of $Y_{t,t^*}$ given $Y_{t,t^*-1} = y_{t,t^*-1}$, $X_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$.
+We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^{*(i)})}\right)$.
+In summary:
+%
+% this is new
+%
+\begin{gather*}
+    \{Y, X\} \text{ follows (Car)HMM with } \Gamma \in \bbR^{N \times N}, \\
+    %
+    \{Y_t   | Y_{t-1} = y_{t-1}, X_t = i\} \sim f^{(i)}(\cdot|y_{t-1};\theta^{(i)}), \\
+    %
+    \{Y^*_t,X^*_t | X_t = i\} \text{ follows (Car)HMM with } \Gamma^{*(i)} \in \bbR^{N^{*(i)} \times N^{*(i)}}, \\
+    %
+    \{Y^*_{t,t^*} | Y^*_{t,t^*-1} = y^*_{t,t^*-1}, X^*_{t,t^*} = i^*, X_t = i\} \sim f^{*(i,i^*)}(\cdot|y^{*}_{t,t^*-1};\theta^{(i,i^*)}).
+\end{gather*}
+Given the coarse-scale hidden states $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$ and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
+Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.
+%
+% end new
+%
 
-Depending upon the process being modelled, it is possible to force certain parameters to be shared to reduce the complexity of the HHMM.
-For example, in the killer whale case study (see Section \ref{sec:data}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( \text{i.e.
-} \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
-Figure \ref{fig:models}c represents the dependence structure for an HHMM.
+Forcing certain parameters to be shared can reduce the complexity of the HHMM.
+For example, in the killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$.
+We also share the fine-scale emission parameters across coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
 
@@
 For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour.
-To handle these additional fine-scale structures, we recommend replacing the fine-scale observations $Y_t^*$ with relevant statistics that summarize this non-Markovian behaviour.
-To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over $Y_t^*$.
+To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize this non-Markovian behaviour.
+To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over the elements $Y_t^*$.
 Subject-area experts are often required to determine the optimal window size and the specific summary statistics used.
-For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ across $Y^*_t$:
+For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ across $Y^*_t$.
+For each $t^* = 1,2,\ldots,T^*_t-h+1$:
 %
 \begin{align*}
-    DFT\{Y^*_{t,t^*}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right),
+    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right) \qquad 
 \end{align*}
 %
-for $k = 0, 1, \ldots, h-1$.
-If $Y^*_{t,t^*}$ is high-dimensional, then the DFT is taken component-wise.
-Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
+for $k = 0, 1, \ldots, h-1$. If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise. Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
 %
@@
     \label{eqn:z}
-    \Zone_{t,t^*} \equiv \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n} \quad \text{and} \quad \Ztwo_{t,t^*} \equiv \sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
+    \Zone_{t,t^*} \equiv \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n}, \quad \Ztwo_{t,t^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k)\Big|\Big|^2.
 \end{equation}
 %
-In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared two-norm of the component of the window $\left(Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\right)$ that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window (by the Plancherel theorem).
+In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared two-norm of the component of the window $\big\{Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\big\}$ that can be attributed to frequencies between one and $\tilde{\omega} \leq h-1$ periods per window (by the Plancherel theorem).
 More intuitively, $\Ztwo_{t,t^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within the window starting at time stamp $t^*$.
@@
 
-To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \left\{\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\right\}$ and $X^*_t = \left\{X^*_{t,1},X^*_{t,h+1},X^*_{t,2h+1},\ldots\right\}$.
+To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \big\{\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\big\}$ and $X^*_t = \big\{X^*_{t,1},X^*_{t,h+1},X^*_{t,2h+1},\ldots\big\}$.
 When down-sampling, we redefine the fine-scale probability transition matrices such that $\Gamma^{*(i)}_{jk} = \Pr(X^*_{t,t^*+h} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1,h+1,2h+1,\ldots$.
-%In addition, we assume that the hidden state within a window does not change, (i.e.
-$X^*_{t,t^*+s} = X^*_{t,t^*}$ for all $t$, $t^* = 1,h+1,2h+1,\ldots$ and $s = 0,1,\ldots,h-1$.
+If the fine-scale model is a CarHMM, we also redefine the probability density function $f$ such that $f^{*(i,i^*)}(\cdot|\z_{t,t^*-100}; \theta^{*(i,i^*)})$ describes the density of $\Z_{t,t^*}$ given $X_t = i$, $X^*_{t,t^*} = i^*$, and $\Z_{t,t^*-100} = \z_{t,t^*-100}$.
 This down-sampling both avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
@@
 Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT) in addition to many others not described here.
-For example, a continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}.
-as an example, \citet{Xu:2018} explain biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours.
+
+For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model.
+If the fine-scale model is a simple Poisson process, then the hierarchical structure described here reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
+Further, \citep{Langrock:2018} use B-splines to model the emission distribution of an HMM as well as incorporate covariates.
+This non-parametric approach admits a penalized likelihood which can easily be incorporated into the hierarchical structure here.
+One additional class of fine-scale models is the set of continuous time methods such as the continuous-time HMM (CTHMM) \citep{Liu:2015} or the state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019}.
+A continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}.
+As an example, \citet{Xu:2018} explain biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours.
 However, they take a longitudinal approach whereas we use a coarse-scale HMM in our case study.
-Further, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model.
-If the fine-scale model is a simple Poisson process, then the hierarchical structure described here reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
-As a final example, \citep{Langrock:2018} use B-splines to model the emission distribution of an HMM as well as incorporate covariates.
-This non-parametric approach admits a penalized likelihood which can easily be incorporated into the hierarchical structure here.
+We also show the appendix that the state-switching OU process of \citet{Michelot:2019} is equivalent to a CarHMM under certain conditions.
 
 These examples are a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models.
-As an example, consider a \textbf{CarHHMM}, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM.
-Alternatively, one may construct an \textbf{HHHMM}, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$) and traditional HMMs are used to model all three levels.
+As an example, consider a CarHHMM, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM.
+Alternatively, one may construct an HHHMM, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$) and traditional HMMs are used to model all three levels.
 It is clear that are a myriad of possible models can be built using this framework.

### 2020-10-30 Report/Sections/simulation.tex
commit before adding t_tilde

@@
 %\section{Simulation Study}
-
-To evaluate the performance of each candidate model when the ground-truth is known, we perform a simulation study based on data generated from the full CarHHMM-DFT.
+We perform a simulation study based on data generated from the full CarHHMM-DFT as defined in Section \ref{subsec:model_selection} to evaluate each candidate model when the ground-truth is known.
 The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity.
-In particular, we set $\Gamma$ such that the simulated data has equal numbers of short and long dives in expectation, the number subdive states is set to $N^*=2$, and $\Ztwo_{t,t^*}$ is set to a scalar.
-Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, standard error of parameter estimates, and fitting times.
-To assess the accuracy of uncertainty estimates, we also compare the empirical standard error of each model's parameter estimates with the estimated standard error using the inverse of the observed Fisher information under that model.
+In particular, we set $\Gamma$ such that the simulated data have equal numbers of short and long dives in expectation, the number of subdive states is set to $N^*=2$, and $\Zone_{t,t^*}$ is set to a scalar rather than a three dimensional vector.
+We then fit the data with four models: the full CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT.
+Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times.
+To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors a given model's parameter estimates with the standard errors estimated using the inverse of the observed Fisher information under that model.
 
@@
 We generate 500 independent data sets from the CarHHMM-DFT.
-Each data set consists of a sequence of 100 curves which each represent a simulated killer whale dives.
+Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives.
 Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
@@
 %
-Dive duration is gamma distributed, and the coarse-scale emission parameters $\theta$ are
-$$
-    \mu^{(1)} = 20s, \quad \sigma^{(1)} = 5s, \quad
-    \mu^{(2)} = 80s, \enspace \text{and} \enspace \sigma^{(2)} = 25s.
-$$
-%
-After calculating the dive durations for all 100 dives in each data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows.
+Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 25s$.
+After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2T^*_t$ seconds of each dive are ignored.
 Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
 %
-$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}.$$ 
+$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}$$ 
 %
-For each dive and subdive segment, we simulate the fine-scale transformed observations $Z^*_t \equiv \left\{Z^*_{t,1}, \ldots, \right\}$, where $Z^*_{t,t^*} = \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$, using the following parameters:
+for dive types 1 and 2, respectively.
+Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $Z^*_{t,t^*} = \left\{\Zone_{t,t^*}, \Ztwo_{t,t^*}\right\}$ as defined in Equation (\ref{eqn:z}).
+Recall that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent when conditioned on all dive types and subdive states.
+Each observation of acceleration $\Zone_{t,t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,t^*}$ is Gamma-distributed without explicit auto-correlation.
+In particular, we set the distributions of $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ with the following parameters:
 %
@@
 %
-Recall that $\Zone_{t,t^*}$ follows a CarHMM and is Normally distributed while $\Ztwo_{t,t^*}$ follows an HMM and is Gamma distributed.
-It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping in the appendix.
-Figure \ref{fig:sim_data} shows this reconstructed sequence along with one realization of $Z^*$ for five dives of one simulated data set.
+It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix.
+Figure \ref{fig:sim_data} shows one realization of $Z^*$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.
 
-The two simulated dive types differ in that dives of type 1 are much shorter on average (20 seconds) than dives of type 2 (80 seconds).
-The two subdive types differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher in subdive state 2 than subdive state 1.
-This corresponds to much more vigorous and variable periodic behaviour in the acceleration data.
-
-All four candidate models (CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT) are fit to all 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
+The two simulated dive types differ in that dives of type one are much shorter on average (20 seconds) than dives of type two (80 seconds).
+The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher for subdive state 2 than for subdive state 1.
+These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data.
+All four candidate models are fit to each of the 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
 
@@
 
-The full CarHHMM-DFT is the best-performing model of the four candidates, which is to be expected as the generating model.
-If $X^*_{t,t^*}$ is the unknown subdive state and $x^*_{t,t^*}$ is the known true subdive state for dive $t$ and window $t^*$, then using the forward-backward algorithm we find that $\Pr(X^*_{t,t^*} = x^*_{t,t^*}|Y,Y^*)$ is $(1 - 10^{-12})$ on average across all simulations, dives, and windows.
-We refer to this quantity as the average accuracy of the model.
-For dive type, this average accuracy was above 0.9 (see Table \ref{table:accuracy}).
+We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}).
+If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives.
+Analogously, if $X^*_{t,t^*}$ is the unknown subdive state associated with a particular model and $x^*_{t,t^*}$ is the true simulated subdive state for dive $t$ and window $t^*$, then we define the average subdive decoding accuracy as $\Pr(X^*_{t,t^*} = x^*_{t,t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows.
+We find both the average dive decoding accuracy and the average subdive decoding accuracy using the forward-backward algorithm.
+
+The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model.
+The average dive decoding accuracy is above 0.9 while the average subdive decoding accuracy is $1-10^{-12}$.
 All parameter estimates of mean values and fine-scale probability transition matrices, $\hat \mu$ and $\hat \Gamma^*$, respectively, have statistically insignificant biases.
-In addition, biases for $\hat \sigma$, $\hat \phi$, and $\hat \Gamma$ are no more severe for the CarHHMM-DFT than the other three models.
-The empirical standard error of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard error.
-See Tables 6-9 of the supplementary material for more information.
+In addition, biases for $\hat \sigma$, $\hat \phi$, and $\hat \Gamma$ are as severe or less severe than the other three models.
+The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to slightly underestimate the empirical standard errors.
+See Tables 6 through 9 of the supplementary material for detailed results.
 
-The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its average accuracy and parameter bias.
-However, the HHMM-DFT differs from the CarHHMM-DFT in its estimates of $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$, both of which are greatly overestimated.
+The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its average dive and subdive decoding accuracy as well as its parameter estimate biases.
+However, unlike the CarHHMM-DFT, the HHMM-DFT greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$.
 In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material).
-These results imply that estimated standard errors of parameter estimates are too small when using the HHMM-DFT since auto-correlation is ignored.
-This finding is consistent with the results from the case study, where the HHMM-DFT produced higher estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.
+These results suggest that the estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored.
+This finding is consistent with the results of the case study, where the HHMM-DFT produced larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.
 
-The CarHHMM is the worst-performing model in terms of decoding accuracy, with an average accuracy below $0.9$ for both dive and subdive types.
-This result is consistent with expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct observation when determining subdive type.
-The CarHHMM is also much more likely to misclassify subdive type 1 than subdive type 2 (see Table \ref{table:accuracy}).
-In addition to its poor average accuracy, the CarHHMM is also the worst of the four candidate models at estimating parameters, with the fine-scale probability transition matrices associated with dive type 2 ($\hat \theta^{*(2)}$ and $\hat \Gamma^{*(2)}$) being especially poor estimates (see Table 9 and Figure 33 of the supplementary material).
+The CarHHMM is the worst-performing model in terms of accuracy, as its average dive decoding accuracy and average subdive decoding accuracy are both below $0.9$.
+This result is consistent with expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct difference between the subdive states.
+The CarHHMM is much more likely to misclassify subdive state 1 as subdive state 2 than vice-versa (see Table \ref{table:accuracy}).
+In addition to its poor average subdive and dive decoding accuracy, the CarHHMM is also the worst of the four candidate models at estimating parameters.
+Estimates associated with dive type 2 ($\theta^{(2)}$ and $\hat \Gamma^{*(2)}$) and subdive type 2 ($\theta^{*(\cdot,2)}$) are especially poor, likely due to the fact that the CarHHMM often classifies dive type 1 as dive type 2 and subdive state 1 as subdive state 2.
+See Section 2 of the supplementary material for more detailed results.
 
-Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of decoding accuracy, parameter bias, and parameter standard error for the fine-scale observations.
+Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of average subdive decoding accuracy, parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates.
 In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}).
-However, this model is unable to estimate dive types as it lacks any hierarchical structure.
-The CarHMM-DFT nonetheless fits a gamma distribution to the dive duration of all dives.
-The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated, and $\hat \sigma$ underestimates the true standard deviation of all dive types by approximately five seconds.
+However, this model does not estimate dive type as it lacks any hierarchical structure.
+The CarHMM-DFT nonetheless fits a single Gamma distribution over the dive duration of all dives.
+The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated, and on average the value of $\hat \sigma$ is approximately 5 seconds less than the average sample standard deviation of dive duration, indicating significant model misspecification.
 See Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive types and subdive states for each model.

### 2020-10-30 Report/main.tex
commit before adding t_tilde

@@
 %\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
-\title[]{Modelling state-switching behaviour in functional data with hidden Markov models}
+\title[]{Modelling state-switching functional data with hidden Markov models}
 \author{BlindedA\authorref{1}\thanksref{*}}
@@
 \KWDtitle{Key words and phrases}
-Association parameters\sep clustered data\sep mean parameters\sep missing data\sep pairwise
-likelihood\sep repeated measurements.
+Time series modelling\sep multi-state modelling\sep hierarchical modelling\sep biologging\sep statistical ecology\sep animal movement modelling.
 % MSC 2010 subject classification codes
-\KWDtitle{MSC 2010}Primary 62???\sep secondary 62???}
+\KWDtitle{MSC 2010}Primary 62M05 \sep secondary 62P12}
 \begin{abstract}
@@
 
+\iffalse
 Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields.
@@
 Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs) instead, but many intricate fine-scale processes violate basic assumptions of HMMs.
-We detail a generalization of the hierarchical HMM approach that incorporates a wider class of parametric models at fine scales. 
-%We also provide several examples of viable fine-scale models.
+We detail a general hierarchical approach based on HMMs which incorporates a wider class of models at fine scales.
 As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada.
 Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing baselines.
+\fi
+
+Sequences of curves sampled at high-frequency in time are increasingly common in practice, but they can exhibit complicated dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA).
+We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM).
+The distribution of each curve is then defined by another general fine-scale model which may involve a variety of methods such as moving-window transformations, Fourier analysis, and auto-correlation.
+As a case study, we use this framework to model the movement of a killer whale off the western coast of Canada.
+Through simulations we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines.
+This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.
 
@@
 %\correspondingauthor[*]{\\\email{}} % only put this in after acceptance
-
-%%%%%%%%%%%%
-%\vfill\eject
-%{\centerline{\Large{Notes from Nancy}}}
-%\input{Sections/0627NH}
-%\vfill\eject
-%%%%%%%%%%%%%
 %
@@
 %
-%\section{Case Study Results}
-%\label{sec:case_study}
-%\input Sections/results.tex
-%%
 \section{DISCUSSION}

### 2020-10-30 Report/Sections/discussion.tex

@@
 
+Current FDA literature addresses dependence between curves either with multilevel models \citep{Chen:2012,Di:2009}, which lack a time component, or with functional time series, which overlook the possibility that curves have several distinct ``types" \citep{Kokoszka:2018}.
+Our work addresses these issues and introduces a flexible framework to model functional time series data using HMMs.
+We deal with temporal dependence between curves with distinct ``types" by using either an HMM or a CarHMM to model the curve sequence.
+We then view each individual curve as an HMM emission whose distribution is described by yet another fine-scale model.
+Here we use a CarHMM as the fine-scale model, but there are a wide range of possible fine-scale models, including a Poisson process or continuous time approach similar to that of \citet{Michelot:2019}.
+We also incorporate a moving-window transformation into the fine-scale model to deal with intricate dependence structures on short time scales.
+Together, the coarse- and fine-scale models make up a hierarchical structure which can account for simultaneous processes taking place at different time scales.
+Provided the construction is not overly complex, a hierarchical model created using our method can be both flexible and easy to fit using maximum likelihood estimation.
+
+We demonstrate the usefulness of this framework using an ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a Northern Resident Killer Whale off the coast British Columbia, Canada.
+This gives a deeper understanding of the animal's tri-axial movement and thus its energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both Northern Resident Killer Whales and their prey \citep{Noren:2011}.
+This model is likely applicable to many other diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.
+In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time series data in a variety of fields.
+
+\iffalse
+
 Current FDA literature addresses dependence between curves either with multilevel models with random effects and no time component at all \citep{Chen:2012,Di:2009}, or with functional time series which model the curves as an evolving process \citep{Kokoszka:2018} which overlook the possibility that curves have several distinct ``types".
 However, many curve sequences exhibit both temporal dependence and a discrete number of types.
-%One obvious example is the case study examined here, where one marine mammal dive serves a distinct purpose such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}.
+One obvious example is the case study examined here, where one marine mammal dive serves a distinct purpose such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}.
 Another example is accelerometer data in school children, where bouts of activity can be associated with a behaviour such as running or walking.
@@
 
-%When used traditionally, each of these models require that practitioners take summary statistics of curves as emissions which may lack important information about the structure of the curve.
+When used traditionally, each of these models require that practitioners take summary statistics of curves as emissions which may lack important information about the structure of the curve.
 However, HMM emissions can also be viewed as the entire curve, where the emission distribution of that curve is described by yet another fine-scale model.
@@
 
-%One interesting class of possible fine-scale models is the set of continuous time methods such as the state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019} and the continuous-time HMM (CTHMM) \citep{Liu:2015}, which are often based on stochastic differential equations.
+One interesting class of possible fine-scale models is the set of continuous time methods such as the state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019} and the continuous-time HMM (CTHMM) \citep{Liu:2015}, which are often based on stochastic differential equations.
 Unfortunately, most continuous-time models require MCMC algorithms to perform inference and can be slow to fit.
@@
 
-%Another fine-scale model that we investigate in detail here is yet another HMM.
+Another fine-scale model that we investigate in detail here is yet another HMM.
 If the fine-scale model is an HMM, then the resulting overall model is just the HHMM as introduced by \citet{Barajas:2017}.
@@
 
-%When using HMMs to build complex hierarchical models, it is important to build a model that is complex enough to fit the data but not so complex that it over-fits the data or is computationally inefficient.
+When using HMMs to build complex hierarchical models, it is important to build a model that is complex enough to fit the data but not so complex that it over-fits the data or is computationally inefficient.
 Model complexity inevitably grows rapidly as more layers are added to the hierarchical structure.
@@
 
-%Our work provides a flexible framework to model functional time-series data using HMMs.
-We deal with temporal dependence between curves and distinct curve ``types" by using either an HMM or CarHMM.
+Our work provides a flexible framework to model functional time-series data using HMMs.
 Provided the model is not overly complex, a hierarchical model constructed using the proposed building blocks can be both flexible and easy to fit using maximum likelihood methods.
@@
 In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time-series data in a variety of fields.
+
+\fi

### 2020-10-30 Report/Sections/results.tex

@@
 \subsection{Model definition and selection}
+\label{subsec:model_selection}
 
@@
 We first select an appropriate model structure for the coarse-scale observations of dive duration.
-Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Section 1.1 of the supplementary material), and visual inspection also shows no obvious intricate structure.
-Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor transformation is called for.
 %
-Next, we select the emission distribution of dive duration conditioned on dive type. Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows
-a gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:
+We do not use information criteria to select the number of dive types $N$ since these metrics tend to overestimate the number of states in biological processes \citep{Pohle:2017}.
+If the hidden states are well-separated, a lag plot should reveal $N$ distinct patterns, where each pattern corresponds to one dive type \citep{Lawler:2019}.
+This is unfortunately not the case for our killer whale data, but we still consult lag plots together with visual inspection to choose $N = 2$ dive types.
+The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
+%
+Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Section 1.1 of the supplementary material), and visual inspection shows no obvious intricate structure.
+Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor a moving-window transformation is called for.
+%
+Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows a Gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:
 %
@@
 %
-Finally, we select the number of dive types for the coarse scale.
-We do not use information criteria since these metrics tend to overestimate the number of states in biological processes \citep{Pohle:2017}.
-If the hidden states are well-separated, the lag plot should reveal $N$ distinct patterns where each pattern corresponds to one dive type \citep{Lawler:2019}.
-This is unfortunately not the case for our killer whale data, but we still consult lag plots together with a heuristic approach based on visual inspection to choose $N = 2$ dive types.
-The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
 
-To model the fine-scale observations of acceleration, we again begin with determining model structure.
-In contrast to the coarse-scale, the fine-scale acceleration data exhibit significant sinusoidal behaviour.
-Thus, we transform each fine-scale observation sequence $y_t^*$ into $z_t^*$ using Equation (\ref{eqn:z}).
-We select a window size of two seconds since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data.
+We then select a model corresponding to the fine-scale observations of acceleration.
+Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$.
+Although $N^*$ is selected heuristically, we test the validity of this model using techniques in Section \ref{subsec:model_validation}.
 %
-We then calculate $\Ztwo_{t,t^*}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to maximum frequencies of 5 Hz for each axis of acceleration.
-We also down-sample each observation $z^*_t$ such that $z^*_t = \left\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\right\}$ and $x^*_t = \left\{x^*_{t,1},x^*_{t,101},x^*_{t,201},\ldots\right\}$ to avoid overlapping windows.
-The resulting sequence of observations is $z^* = \left\{z^*_1,\ldots,z^*_{267}\right\}$, where $z^*_t = \left\{z^*_{t,1},z^*_{t,101},\ldots\right\}$ and $z^*_{t,t^*} = \left\{a^*_{t,t^*},w^*_{t,t^*}\right\}$.
-We have that $a^*_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $w^*_{t,t^*}$ is a scalar describing the ``wiggliness" of a given window.
+In contrast to the coarse-scale observations, the fine-scale acceleration data exhibit significant sinusoidal behaviour.
+Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with window size $h=100$ (two seconds) and maximum frequency $\tilde{\omega}=10$ (5 Hz).
+%We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
+We then have that $\z_{t,t^*} = \left\{\zone_{t,t^*},\ztwo_{t,t^*}\right\}$, where $\zone_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,t^*}$ is a scalar describing the ``wiggliness" of a particular window.
+We also down-sample each observation $\z_t$ such that $\z_t = \left\{\z_{t,1},\z_{t,101},\z_{t,201},\ldots\right\}$ and $x^*_t = \left\{x^*_{t,1},x^*_{t,101},x^*_{t,201},\ldots\right\}$ to avoid overlapping windows.
+Even after transforming and down-sampling the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material).
+Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.
+
+We then select the specific emission distribution of $\Z_{t,t^*}$.
+Given the dive types and subdive states, we assume that $\Ztwo_{t,t^*}$ and all components of $\Zone_{t,t^*}$ are independent of one another.
+To reduce model complexity, we also assume that the three fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$.
+This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.
+
+All components of $\Zone_{t,t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $t^*$.
+Each component is assumed to have its own mean and variance parameters, but all components have the same auto-correlation parameter.
+%Given that $X^*_{t,t^*} = i^*$, the corresponding unknown parameters are then $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}_{>0}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
+Given that $X^*_{t,t^*} = i^*$, $\Ztwo_{t,t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
+We do not include $\Ztwo_{t,t^*-100}$ in the distribution of $\Ztwo_{t,t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
 %
-After transforming the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material).
-Therefore, we assume that all three components of $\Zone_{t,t^*}$, $t^* = 1,101,\ldots$ follow the CarHMM as defined in Section 2.2.
-Due to down-sampling we redefine the probability density function $f$ such that $f^{*(i,i^*)}(\cdot|z^*_{t,t^*-100}; \theta^{*(i,i^*)})$ describes the density $\Z_{t,t^*}$ given that $X_t = i$, $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.
-We also assume that observations of all three axes of acceleration are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$.
-%
-%$$\mathbb{E}(\Zone_{t,t^*}|\\Zone_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_A^{*(i)} \mathbf{z} + (1-\phi_A^{*(i)}) \mathbf{\mu}_A^{*(i)},$$
-%$$\mathbb{V}(\Zone_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_A^{*(i)}\right)^2\right].$$
-%
-While the observations of wiggliness ($\Ztwo_{t,t^*}$) also exhibit some auto-correlation prior to fitting the model, this correlation is less strong than $\Zone_{t,t^*}$, and 
-%the biological interpretation of auto-correlation between observations of $\Ztwo_{t,t^*}$ is less clear. In addition, 
-much of the auto-correlation evident from the lag plot may be explained by subsequent observations occurring within the same subdive state.
-Therefore, a simple HMM structure is used to model $\Ztwo_{t,t^*}$ for all $t$ and $t^*$.
-We assume that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent of one another when conditioned on the subdive state $X^*_{t,t^*}$.
-%
-%$$\mathbb{E}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_W^{*(i)}$$
-%$$\mathbb{V}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_W^{*(i)}\right)^2.$$
-%
-%
-After defining the model structure, we select the emission distributions for both $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$.
-We assume that the fine-scale emission parameters are shared across coarse-scale hidden states $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$ to reduce model complexity.
-This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.
-%
-For $\Zone_{t,t^*}$, we assume that each component is Normally distributed with mean and variance defined as in Equation (\ref{eqn:carhmm}).
-Given that $X^*_{t,t^*} = i^*$, the corresponding unknown parameters are $\phi_A^{*(\cdot,i^*)} \in \mathbb{R}$, $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$.
-In other words, each axis of acceleration has its own mean and variance, but all axes have the same auto-correlation parameter.
-%
-We assume that $\Ztwo_{t,t^*}$ is gamma distributed for all $t$ and $t^*$.
-Further, given that $X^*_{t,t^*} = i^*$, the distribution of $\Ztwo_{t,t^*}$ is parameterized by its mean $\mu_W^{*(\cdot,i^*)} \in \mathbb{R}$ and standard deviation $\sigma_W^{*(\cdot,i^*)} \in \mathbb{R}$.
-%
-Finally, to select the number of subdive states $N^*$ we again rely on a heuristic approach based on visual inspection to select $N^* = 3$.
-Although $N^*$ is selected heuristically, we test the validity of this model using techniques in Section \ref{subsec:model_validation}.
-
-The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale.
-Each fine-scale sequence $y^*_t$ is transformed to $z^*$ using Equation (\ref{eqn:z}), $\Zone_{t,t^*}$ is modelled with a CarHMM, and $\Ztwo_{t,t^*}$ is modelled with an HMM for all $t$ and $t^*$.
-We model $\Zone_{t,t^*}$ to be Normally distributed and $\Ztwo_{t,t^*}$ to be Gamma distributed.
-We also assume that $\Ztwo_{t,t^*}$ in addition to the components of $\Zone_{t,t^*}$ are all independent of one another and that the subdive types have identical emission distributions across all dive types.
+%The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale.
+Each fine-scale sequence $y^*_t$ is transformed to $\z$ using Equation (\ref{eqn:z}) and $\Z_{t,t^*}$ is modelled with a CarHMM.
+%We assume $\Zone_{t,t^*}$ is Normally distributed with auto-correlation while $\Ztwo_{t,t^*}$ is Gamma distributed without auto-correlation.
+We also assume that $\Ztwo_{t,t^*}$ and all components of $\Zone_{t,t^*}$ are each independent of one another when conditioned on the subdive state.
+The emission distribution of each subdive state does not depend upon the dive type.
 In total, the parameters to estimate are
@@
     %
-    \theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{(fine-scale emission parameters), where} \\
+    \theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{($\Z$ emission parameters), where} \\
     %
@@
 Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $Z_{t,t^*}$ when conditioned on $X^*_{t,t^*} = i^*$. 
-
-We refer to this final model as the \textbf{CarHHMM-DFT} since it has elements of all HMMs discussed so far.
+%
+We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation.
 Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model.
@@
 See the appendix for details of likelihood evaluation.
-In addition to the CarHHMM-DFT, we consider three similar variations for comparison:
+In addition to the CarHHMM-DFT, we consider three similar variations for comparison.
+As in the full model, all components of $\Z_{t,t^*}$ are assumed to be independent when conditioned on dive type and subdive state for each of the following models:
 \begin{enumerate}
-    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ as emissions of a simple HMM rather than a CarHMM.
-    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emssions of a CarHMM.
-However, the Fourier sums $\Ztwo_{t,t^*}$ are omitted from this model altogether.
-The CarHHMM therefore does not incorporate the ``wiggliness" of the accelerometer data.
-    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emissions of a CarHMM and $\Ztwo_{t,t^*}$ as emissions an HMM.
+    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Z_{t,t^*}$ and as emissions of a simple HMM rather than a CarHMM.
+    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emissions of a CarHMM.
+However, the ``wiggliness"  $\Ztwo_{t,t^*}$ is omitted from this model altogether.
+    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Z_{t,t^*}$ as emissions of a CarHMM.
 This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive.
@@
 On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as foraging and socialization \citep{Tennessen:2019b}.
-However, no dive in this data set has a maximum depth deeper than $\approx$ 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100 meters.
+However, no dive in this data set has a maximum depth deeper than 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100 meters.
 It is therefore unlikely that any of the dives used in this study are successful foraging dives.
@@
 The auto-correlation of acceleration is also slightly lower than in subdive state 1.
-We therefore hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
+We therefore hypothesize that subdive state 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
 Finally, the mean of $\Ztwo_{t,t^*}$ and variance of $\Zone_{t,t^*}$ in subdive state 3 are both much higher than in the other two states.
@@
 We use two visual tools to evaluate the CarHHMM-DFT: pseudoresidual plots and empirical histograms.
-Mathematically, the pseudoresidual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
+The pseudoresidual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
 If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution.
@@
 Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT.
-See Sections 1.4 through 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $Z^*_{t,t^*}$.
-Our results again mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions.
+See Sections 1.4 through 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $\z_{t,t^*}$.
+Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions.
 In particular, $\Ztwo_{t,t^*}$ is right-skewed and $\Zone_{t,t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
@@
 
-The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3 (see Figures 4 and 5 of the supplementary material).
+The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive state 3 (see Figures 4 and 5 of the supplementary material).
 In addition, the HHMM-DFT produces estimates of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$ which are approximately $50$ to $100$ percent larger than those of the CarHHMM-DFT for all axes of acceleration.

### 2020-11-05 Report/Sections/discussion.tex
Commit before adding t_tilde take 2

@@
 
-We demonstrate the usefulness of this framework using an ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a Northern Resident Killer Whale off the coast British Columbia, Canada.
-This gives a deeper understanding of the animal's tri-axial movement and thus its energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both Northern Resident Killer Whales and their prey \citep{Noren:2011}.
+We demonstrate the usefulness of this framework using an ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale off the coast British Columbia, Canada.
+This gives a deeper understanding of the animal's tri-axial movement and thus its energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both northern resident killer whales and their prey \citep{Noren:2011}.
 This model is likely applicable to many other diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.

### 2020-11-05 Report/Sections/simulation.tex
Commit before adding t_tilde take 2

@@
 Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times.
-To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors a given model's parameter estimates with the standard errors estimated using the inverse of the observed Fisher information under that model.
+To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors of a given model's parameter estimates with the standard errors estimated from the inverse of the observed Fisher information under that model.
 
@@
 
-The two simulated dive types differ in that dives of type one are much shorter on average (20 seconds) than dives of type two (80 seconds).
+The two simulated dive types differ in that dives of type 1 are much shorter on average (20 seconds) than dives of type 2 (80 seconds).
 The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher for subdive state 2 than for subdive state 1.
@@
 However, this model does not estimate dive type as it lacks any hierarchical structure.
-The CarHMM-DFT nonetheless fits a single Gamma distribution over the dive duration of all dives.
-The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated, and on average the value of $\hat \sigma$ is approximately 5 seconds less than the average sample standard deviation of dive duration, indicating significant model misspecification.
-See Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive types and subdive states for each model.
+The CarHMM-DFT nonetheless fits a single Gamma distribution to the dive duration of all dives.
+The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated, and on average the value of $\hat \sigma$ is approximately 5 seconds less than the true sample standard deviation, indicating significant model misspecification.
+Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} visualize the simulated data set as well as the decoded dive types and subdive states for each model.

### 2020-11-06 Report/Sections/models.tex
Fix Overleaf issues

@@
 Subject-area experts are often required to determine the optimal window size and the specific summary statistics used.
-For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ across $Y^*_t$.
-For each $t^* = 1,2,\ldots,T^*_t-h+1$:
+For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ and stride $h$ across $Y^*_t$.
+Namely:
 %
 \begin{align*}
-    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right) \qquad 
+    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right)
 \end{align*}
 %
-for $k = 0, 1, \ldots, h-1$. If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise. Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
+for $t^* = 1,h+1,2h+1,\ldots$ and for $k = 0, 1, \ldots, h-1$.
+If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise.
+Next, we index each window with $\tilde{t}^*$ and calculate the transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$, where
 %
@@
     \label{eqn:z}
-    \Zone_{t,t^*} \equiv \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n}, \quad \Ztwo_{t,t^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k)\Big|\Big|^2.
+    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|^2.
 \end{equation}
 %
-In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared two-norm of the component of the window $\big\{Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\big\}$ that can be attributed to frequencies between one and $\tilde{\omega} \leq h-1$ periods per window (by the Plancherel theorem).
-More intuitively, $\Ztwo_{t,t^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within the window starting at time stamp $t^*$.
+In words, $\Zone_{t,\tilde{t}^*}$ is the average acceleration vector over window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega} \leq h-1$ periods per window (by the Plancherel theorem).
+More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within window $\tilde{t}^*$.
 The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert.
 Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude.
+Setting the stride length equal to the window size $h$ avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
+By summarizing the data using a relevant transformation, this method reduces correlation in the raw observations due to any complex dependence structure.
 
-To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \big\{\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\big\}$ and $X^*_t = \big\{X^*_{t,1},X^*_{t,h+1},X^*_{t,2h+1},\ldots\big\}$.
-When down-sampling, we redefine the fine-scale probability transition matrices such that $\Gamma^{*(i)}_{jk} = \Pr(X^*_{t,t^*+h} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1,h+1,2h+1,\ldots$.
-If the fine-scale model is a CarHMM, we also redefine the probability density function $f$ such that $f^{*(i,i^*)}(\cdot|\z_{t,t^*-100}; \theta^{*(i,i^*)})$ describes the density of $\Z_{t,t^*}$ given $X_t = i$, $X^*_{t,t^*} = i^*$, and $\Z_{t,t^*-100} = \z_{t,t^*-100}$.
-This down-sampling both avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
-By summarizing the data using a relevant transformation, the method overall also reduces correlation in the raw observations due to the complex dependence structure.
-Figure \ref{fig:models}d represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM, including down-sampling.
-
-Once $\Z_t$ is calculated, it simply replaces the original fine-scale observations $Y^*_t$ in the fine-scale HMM structure.
-As such, the emission distributions and the likelihood remain identical to the original HMM when transforming $Y^*$ to $\Z$.
-In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations as an HMM-DFT.
+The fine-scale HMM must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself.
+In particular, the length of $\Z_t$ is denoted as $\tilde{T}_t^*$ and there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, denoted as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
+The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$.
+It then follows that the likelihood of this model is identical to that of the original HMM but with $Y^*$ replaced with $\Z$ and $X^*$ replaced with $\tilde{X^*}$.
+In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X}$ as hidden states as an HMM-DFT.
 

### 2020-11-06 Report/Sections/simulation.tex
Fix Overleaf issues

--- 2020-10-30 Report/Sections/simulation.tex
@@
 The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity.
-In particular, we set $\Gamma$ such that the simulated data have equal numbers of short and long dives in expectation, the number of subdive states is set to $N^*=2$, and $\Zone_{t,t^*}$ is set to a scalar rather than a three dimensional vector.
+In particular, we set $\Gamma$ such that the simulated data have equal numbers of short and long dives in expectation, the number of subdive states is set to $N^*=2$, and $\Zone_{t,\tilde t^*}$ is set to a scalar rather than a three dimensional vector.
 We then fit the data with four models: the full CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT.
@@
 Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 25s$.
-After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2T^*_t$ seconds of each dive are ignored.
-Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
+After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each dive are ignored.
+Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
 %
@@
 for dive types 1 and 2, respectively.
-Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $Z^*_{t,t^*} = \left\{\Zone_{t,t^*}, \Ztwo_{t,t^*}\right\}$ as defined in Equation (\ref{eqn:z}).
-Recall that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent when conditioned on all dive types and subdive states.
-Each observation of acceleration $\Zone_{t,t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,t^*}$ is Gamma-distributed without explicit auto-correlation.
-In particular, we set the distributions of $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ with the following parameters:
+Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z^*_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$ as defined in Equation (\ref{eqn:z}).
+Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states.
+Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation.
+In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
 %
@@
 We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}).
-If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives.
-Analogously, if $X^*_{t,t^*}$ is the unknown subdive state associated with a particular model and $x^*_{t,t^*}$ is the true simulated subdive state for dive $t$ and window $t^*$, then we define the average subdive decoding accuracy as $\Pr(X^*_{t,t^*} = x^*_{t,t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows.
+If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = \tilde x_{t}|Y,Y^*)$ across all simulated data sets and dives.
+Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows.
 We find both the average dive decoding accuracy and the average subdive decoding accuracy using the forward-backward algorithm.

### 2020-11-06 Report/Sections/appendix.tex
Updates from Overleaf

@@
 It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
-To simulate $X^*_t$ for each dive $t$, note that we simulate accelerometer readings taken at 50 hertz and have two-second window lengths.
-Therefore, each window contains 100 accelerometer readings and $X^*_t = \left\{X^*_{t,1}, X^*_{t,101}, \ldots, X^*_{t,100T^*_t-99}\right\}$.
-In addition, the length of each fine-scale Markov chain is $T^*_t = \lfloor Y_t/2 \rfloor$.
-Then, given the dive types $X_t$, dive durations $Y_t$, and the fine-scale probability transition matrices $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$, it is straightforward to simulate $X^*_t$ for each dive $t = 1,2,\ldots,100$.
-Recall that $\Gamma^{*(i)}_{jk} = Pr(X^*_{t,t^*+100} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1, 101, \ldots$.
+We then simulate simulate $X^*_t$ for a fixed dive $t$.
+Since accelerometer readings are taken at 50 hertz and this model has two-second window lengths, each window contains 100 accelerometer readings.
+In addition, the length of each fine-scale Markov chain is $\tilde{T}^*_t = \lfloor Y_t/2 \rfloor$.
+Then, given the dive type $X_t = i$, dive duration $Y_t = y_t$, and the fine-scale probability transition matrix $\Gamma^{*(i)}$, it is straightforward to simulate the subdive states $\tilde{X}^*_t$.
 It is much more difficult to simulate the raw acceleration data $Y^*$, and we focus on this task for the remainder of this section.
 
-Assume that $X$, $Y$, and $X^*$ are simulated using the procedure described above and are therefore known.
-Additionally, let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for convenience.
-We first simulate $\hat Y^{*(k)}_{t,t^*}$ for all dives $t = 1,2,\ldots,100$, all subdive windows $t^* = 1,101,\ldots$, and all Fourier coefficients $k = 0,1,\ldots,99$.
-We then take the inverse DFT of $\hat{Y}^*_{t,t^*}$ for all dives $t = 1,2,\ldots,100$ and subdive windows $t^* = 1,101,\ldots$ to find $Y^*$.
+Let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for convenience.
+We first simulate $\hat Y^{*(k)}_{t,t^*}$ for all subdive windows $t^* = 1,101,\ldots$, and all Fourier coefficients $k = 0,1,\ldots,99$.
+We then take the inverse DFT of $\hat{Y}^*_{t,t^*}$ for all subdive windows $t^* = 1,101,\ldots$ to find $Y^*$.
 The details of this procedure are described below:

### 2020-11-06 Report/Sections/results.tex
Updates from Overleaf

@@
 %We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
-We then have that $\z_{t,t^*} = \left\{\zone_{t,t^*},\ztwo_{t,t^*}\right\}$, where $\zone_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,t^*}$ is a scalar describing the ``wiggliness" of a particular window.
-We also down-sample each observation $\z_t$ such that $\z_t = \left\{\z_{t,1},\z_{t,101},\z_{t,201},\ldots\right\}$ and $x^*_t = \left\{x^*_{t,1},x^*_{t,101},x^*_{t,201},\ldots\right\}$ to avoid overlapping windows.
-Even after transforming and down-sampling the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material).
+We then have that $\z_{t,\tilde t^*} = \left\{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\right\}$, where $\zone_{t,\tilde t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window.
+Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material).
 Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.
 
-We then select the specific emission distribution of $\Z_{t,t^*}$.
-Given the dive types and subdive states, we assume that $\Ztwo_{t,t^*}$ and all components of $\Zone_{t,t^*}$ are independent of one another.
+We then select the specific emission distribution of $\Z_{t,\tilde t^*}$.
+Given the dive types and subdive states, we assume that $\Ztwo_{t,\tilde t^*}$ and all components of $\Zone_{t,\tilde t^*}$ are independent of one another.
 To reduce model complexity, we also assume that the three fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$.
@@
 
-All components of $\Zone_{t,t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $t^*$.
+All components of $\Zone_{t,\tilde t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $\tilde t^*$.
 Each component is assumed to have its own mean and variance parameters, but all components have the same auto-correlation parameter.
-%Given that $X^*_{t,t^*} = i^*$, the corresponding unknown parameters are then $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}_{>0}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
-Given that $X^*_{t,t^*} = i^*$, $\Ztwo_{t,t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
-We do not include $\Ztwo_{t,t^*-100}$ in the distribution of $\Ztwo_{t,t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
+%Given that $X^*_{t,\tilde t^*} = i^*$, the corresponding unknown parameters are then $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}_{>0}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
+Given that $\tilde X^*_{t,\tilde t^*} = i^*$, $\tilde \Ztwo_{t,\tilde t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
+We do not include $\tilde \Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
 %
 %The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale.
-Each fine-scale sequence $y^*_t$ is transformed to $\z$ using Equation (\ref{eqn:z}) and $\Z_{t,t^*}$ is modelled with a CarHMM.
-%We assume $\Zone_{t,t^*}$ is Normally distributed with auto-correlation while $\Ztwo_{t,t^*}$ is Gamma distributed without auto-correlation.
-We also assume that $\Ztwo_{t,t^*}$ and all components of $\Zone_{t,t^*}$ are each independent of one another when conditioned on the subdive state.
+Each fine-scale sequence $y^*_t$ is transformed to $\z$ using Equation (\ref{eqn:z}) and $\Z_{t,\tilde t^*}$ is modelled with a CarHMM.
+%We assume $\Zone_{t,\tilde t^*}$ is Normally distributed with auto-correlation while $\Ztwo_{t,\tilde t^*}$ is Gamma distributed without auto-correlation.
+We also assume that $\Ztwo_{t,\tilde t^*}$ and all components of $\Zone_{t,\tilde t^*}$ are each independent of one another when conditioned on the subdive state.
 The emission distribution of each subdive state does not depend upon the dive type.
@@
 %
-Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $Z_{t,t^*}$ when conditioned on $X^*_{t,t^*} = i^*$. 
+Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $\Z_{t,\tilde t^*}$ conditioned on $\tilde X^*_{t,\tilde t^*} = i^*$. 
 %
@@
 In addition to the CarHHMM-DFT, we consider three similar variations for comparison.
-As in the full model, all components of $\Z_{t,t^*}$ are assumed to be independent when conditioned on dive type and subdive state for each of the following models:
+As in the full model, all components of $\Z_{t,\tilde t^*}$ are assumed to be independent when conditioned on dive type and subdive state for each of the following models:
 \begin{enumerate}
-    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Z_{t,t^*}$ and as emissions of a simple HMM rather than a CarHMM.
-    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,t^*}$ as emissions of a CarHMM.
-However, the ``wiggliness"  $\Ztwo_{t,t^*}$ is omitted from this model altogether.
-    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Z_{t,t^*}$ as emissions of a CarHMM.
+    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Z_{t,\tilde t^*}$ and as emissions of a simple HMM rather than a CarHMM.
+    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,\tilde t^*}$ as emissions of a CarHMM.
+However, the ``wiggliness"  $\Ztwo_{t,\tilde t^*}$ is omitted from this model altogether.
+    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Z_{t,\tilde t^*}$ as emissions of a CarHMM.
 This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive.
@@
 %
-Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes no auto-correlation within the fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.
+Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes no auto-correlation within the fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.
 
@@
 
-The three subdive behaviours have distinct average ``wiggliness" values, as the means corresponding to $\Ztwo_{t,t^*}$ are separated by an order of magnitude for each subdive state (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}).
-Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,t^*}$ and the smallest variance corresponding to $\Zone_{t,t^*}$.
-It also has the highest auto-correlation in $\Zone_{t,t^*}$.
+The three subdive behaviours have distinct average ``wiggliness" values, as the means corresponding to $\Ztwo_{t,\tilde t^*}$ are separated by an order of magnitude for each subdive state (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}).
+Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest variance corresponding to $\Zone_{t,\tilde t^*}$.
+It also has the highest auto-correlation in $\Zone_{t,\tilde t^*}$.
 This implies less overall activity and more consistent acceleration compared to the other subdive states.
-Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,t^*}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
+Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
 The auto-correlation of acceleration is also slightly lower than in subdive state 1.
 We therefore hypothesize that subdive state 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
-Finally, the mean of $\Ztwo_{t,t^*}$ and variance of $\Zone_{t,t^*}$ in subdive state 3 are both much higher than in the other two states.
-The auto-correlation of $\Zone_{t,t^*}$ is also much lower.
+Finally, the mean of $\Ztwo_{t,\tilde t^*}$ and variance of $\Zone_{t,\tilde t^*}$ in subdive state 3 are both much higher than in the other two states.
+The auto-correlation of $\Zone_{t,\tilde t^*}$ is also much lower.
 This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).
@@
 Histograms of the pseudoresiduals mostly support that this model is well-specified.
-One exception is $\Ztwo_{t,t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:pseudoresids}).
-This implies that the true distribution of $\Ztwo_{t,t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study.
+One exception is $\Ztwo_{t,\tilde t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:pseudoresids}).
+This implies that the true distribution of $\Ztwo_{t,\tilde t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study.
 See Sections 1.4 through 1.6 of the supplementary material for pseudoresidual plots for all observations.
@@
 Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT.
-See Sections 1.4 through 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $\z_{t,t^*}$.
+See Sections 1.4 through 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $\z_{t,\tilde t^*}$.
 Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions.
-In particular, $\Ztwo_{t,t^*}$ is right-skewed and $\Zone_{t,t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
+In particular, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
 These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

### 2020-11-06 Report/Sections/simulation.tex
test

@@
 for dive types 1 and 2, respectively.
-Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z^*_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$ as defined in Equation (\ref{eqn:z}).
+Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$ as defined in Equation (\ref{eqn:z}).
 Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states.

### 2020-11-06 Report/Sections/appendix.tex
add t_tilde

@@
 It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
-We then simulate simulate $X^*_t$ for a fixed dive $t$.
 Since accelerometer readings are taken at 50 hertz and this model has two-second window lengths, each window contains 100 accelerometer readings.
-In addition, the length of each fine-scale Markov chain is $\tilde{T}^*_t = \lfloor Y_t/2 \rfloor$.
-Then, given the dive type $X_t = i$, dive duration $Y_t = y_t$, and the fine-scale probability transition matrix $\Gamma^{*(i)}$, it is straightforward to simulate the subdive states $\tilde{X}^*_t$.
-It is much more difficult to simulate the raw acceleration data $Y^*$, and we focus on this task for the remainder of this section.
+In addition, the length of each fine-scale Markov chain is $\tilde T ^*_t = \lfloor Y_t/2 \rfloor$.
+Then, given the dive type $X_t = i$, dive duration $Y_t = y_t$, and the fine-scale probability transition matrix $\Gamma^{*(i)}$, it is straightforward to simulate the subdive states $\tilde{X}^*_t$ for a particular dive $t$.
+It is much more difficult to simulate the raw acceleration data $Y_t^*$, and we focus on this task for the remainder of the section.
 
-Let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for convenience.
-We first simulate $\hat Y^{*(k)}_{t,t^*}$ for all subdive windows $t^* = 1,101,\ldots$, and all Fourier coefficients $k = 0,1,\ldots,99$.
-We then take the inverse DFT of $\hat{Y}^*_{t,t^*}$ for all subdive windows $t^* = 1,101,\ldots$ to find $Y^*$.
+Let $DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,\tilde t^*}$ and $DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$ be denoted as $\hat{Y}^*_{t,\tilde t^*}$ for convenience.
+We simulate $Y_t^*$ in three steps: 1) simulate the average acceleration within each window $\left(Y^{*(0)}_{t,\tilde t^*}\right)$, 2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and 3) take the inverse $DFT$ of $\hat{Y}^*_{t,\tilde t^*}$ to retrieve $\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$.
 The details of this procedure are described below:
@@
 \begin{enumerate}
-    \item We begin by simulating $\hat{Y}^{*(0)}_{t,1}$ for each dive $t = 1,2,\ldots,100$:
+\item Simulate the average acceleration $Y^{*(0)}_{t,\tilde t^*}$ for each window $\tilde t^* = 1,2,\ldots,\tilde T^*_t$ using the following procedure:
+\begin{enumerate}
+    \item Simulate $\hat{Y}^{*(0)}_{t,1}$:
     $$
-    	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right),
+    	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
     $$
@@
     %
-    \item Next, for each dive $t$, we simulate $\hat{Y}^{*(0)}_{t,101},\enspace \hat{Y}^{*(0)}_{t,201}, \enspace \ldots ~ $ using Equation (\ref{eqn:yhat_0}) below:
+    \item Simulate $\hat{Y}^{*(0)}_{t,2}, \ldots,\enspace \hat{Y}^{*(0)}_{t,\tilde T^*_t}$ using Equation (\ref{eqn:yhat_0}) below:
     \begin{align}	
-        \Big(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*,\hat{Y}^{*(0)}_{t,t^*-100}\Big) &\sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,t^*-100}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \nonumber \\
+        \Big(\hat{Y}^{*(0)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = i^*,\hat{Y}^{*(0)}_{t,\tilde t^*-1}\Big) &\sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,\tilde t^*-1}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \nonumber \\
         %
-        &t^* = 101,201,\ldots
+        &t^* = 2,\ldots, \tilde T^*_t
     	\label{eqn:yhat_0}
@@
     where $\phi_A^{*(\cdot,1)} = 0.99$ and $\phi_A^{*(\cdot,2)} = 0.95$. 
+    \end{enumerate}
     After this step we have simulated the average acceleration within each two-second window for all dives (see Equation (\ref{eqn:avg_val}) below).
+    \item Simulate the other Fourier coefficients $\left(Y^{*(k)}_{t,\tilde t^*}\right)$ for each window $\tilde t^* = 1,2,\ldots,\tilde T^*_t$ and $k = 1,\ldots,99$ using the following procedure:
+    \begin{enumerate}
     %
-    \item We then simulate Fourier coefficients 1 through 49 within each window:
+    \item Simulate Fourier coefficients 1 through 49:
     \begin{equation}
-        \hat{Y}^{*(k)}_{t,t^*} = a_{t,t^*}^{(k)} i\sqrt{b^{(k)}_{t,t^*}}, \quad k = 1,\ldots,49;
+        \hat{Y}^{*(k)}_{t,\tilde t^*} = a_{t,\tilde t^*}^{(k)} i\sqrt{b^{(k)}_{t,\tilde t^*}}, \quad k = 1,\ldots,49;
         \label{eqn:abYhat}
     \end{equation}
-    where $a^{(k)}_{t,t^*}$ is equal to either 1 or -1, each with probability 1/2. In addition, $b^{(k)}_{t,t^*}$ has the following distribution:
+    where $a^{(k)}_{t,\tilde t^*}$ is equal to either 1 or -1 each with probability 1/2, and $b^{(k)}_{t,\tilde t^*}$ has the following distribution:
     \begin{align}
     \begin{split}
-    	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 1) &\sim {\rm{Gamma}}(5/k^2, 1) \\
+    	(b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1) &\sim {\rm{Gamma}}(5/k^2, 1) \\
     	%
-    	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
+    	(b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2) &\sim \left\{\begin{array}{lr}
     	{\rm{Gamma}}(250,1), & k = 1 \\
@@
     The first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the scale parameter and the second is the shape parameter.
-It is important that $\mathbb{E}\left(|b^{(k)}_{t,t^*}|\right)$ is proportional to $1/k^2$, as smaller expected values for larger Fourier coefficients ``smooths out" the raw acceleration data.
+The magnitude of the $k^{th}$ Fourier coefficient is equal to $b^{(k)}_{t,\tilde t^*}$, which decays like $1/k^2$ to ``smooth out" the raw acceleration data.
     %
-    \item We set Fourier coefficients 50 through 99 strategically to ensure that the inverse DFT is real-valued:
+    \item Set Fourier coefficients 50 through 99 so the inverse DFT is real-valued:
     $$
-    \hat{Y}^{*(50)}_{t,t^*} = 0; \qquad
-	\hat{Y}^{*(k)}_{t,t^*} = -\hat{Y}^{*(100-k)}_{t,t^*} \enspace \text{for} \enspace k = 51,\ldots,99.
+    \hat{Y}^{*(50)}_{t,\tilde t^*} = 0, \qquad
+	\hat{Y}^{*(k)}_{t,\tilde t^*} = -\hat{Y}^{*(100-k)}_{t,\tilde t^*} \enspace \text{for} \enspace k = 51,\ldots,99.
     $$
+    \end{enumerate}
+    \item Finally, set the raw acceleration data within each window $\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$ with the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$:
     %
-    \item Finally, for each dive $t = 1,2,\ldots,100$ and each window $t^*=1,101,\ldots$, the raw acceleration data $\left\{Y^*_{t,t^*},\ldots,Y^*_{t,t^*+99}\right\}$ is set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
-    %
-    $$Y^*_{t,t^*+n} = IDFT\left\{\hat{Y}^*_{t,t^*}\right\}(n); \quad n = 0,\ldots,99; \quad t^* = 1,101,\ldots$$
+    $$\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\} \equiv IDFT\left\{\hat{Y}^*_{t,\tilde t^*}\right\}$$
+    for $\tilde t^* = 1,\ldots,\tilde T^*_t.$
 \end{enumerate}
 
-We now show that this construction of $Y^*$ results in the correct distributions as listed in Section \ref{subsec:data_simulation}.
-First, note that $\hat{Y}^{(0)}_{t,t^*}$ is just the sum of the raw acceleration values within the corresponding window:
+We now show that this construction of $Y^*_t$ results in the distributions listed in Section \ref{subsec:data_simulation}.
+First, note that $\hat{Y}^{(0)}_{t,\tilde t^*}$ is just the sum of the raw acceleration values within window $\tilde t$:
 %
 \begin{align}
-    \hat{Y}^{(0)}_{t,t^*} = DFT\{Y^*_{t,t^*}\}(0) &= \sum_{n=0}^{99} Y^*_{t,t^*+n} \exp\left(-\frac{i2\pi}{99}*0*n\right) \nonumber \\
-    &= \sum_{n=0}^{99} Y^*_{t,t^*+n} \nonumber \\
-    &= 100A^*_{t,t^*}.
+    \hat{Y}^{(0)}_{t,\tilde t^*} &=
+    DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}(0) \nonumber \\
+    &= \sum_{n=0}^{99} Y^*_{t,100(\tilde t^* - 1) + 1 + n} \exp\left(-\frac{i2\pi}{99}*0*n\right) \nonumber \\
+    &= \sum_{n=1}^{100} Y^*_{t,100(\tilde t^* - 1) + n} \nonumber \\
+    &= 100A^*_{t,\tilde t^*}.
     \label{eqn:avg_val}
@@
 %
-Combining Equation (\ref{eqn:avg_val}) above with Equation (\ref{eqn:yhat_0}) gives the distribution of $\Zone_{t,t^*}$: 
+Combining Equation (\ref{eqn:avg_val}) above with Equation (\ref{eqn:yhat_0}) gives the distribution of $\Zone_{t,\tilde t^*}$: 
 %
-$$\left(100\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(100\phi_A^{*(\cdot,i^*)} \Zone_{t,t^*-100}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)$$
+$$\left(100\Zone_{t,\tilde t^*} | \tilde X^*_{t,\tilde t^*} = i^*, \Zone_{t,\tilde t^*-1} \right) \sim \mathcal{N} \left(100\phi_A^{*(\cdot,i^*)} \Zone_{t,\tilde t^*-1}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)$$
 %
-$$\implies \left(\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \Zone_{t,t^*-100}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right).$$
+$$\implies \left(\Zone_{t,\tilde t^*} | \tilde X^*_{t,\tilde t^*} = i^*, \Zone_{t,\tilde t^*-1} \right) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \Zone_{t,\tilde t^*-1}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right).$$
 %
-This distribution is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(\cdot,i^*)} = 0$.
+This distribution is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(\cdot,1)} = \mu_A^{*(\cdot,2)} = \mu_A^{*(\cdot,i^*)} = 0$.
 We also have that $\sigma_A^{*(\cdot,1)} = 0.05s$, $\phi_A^{*(\cdot,1)} = 0.99$, $\sigma_A^{*(\cdot,2)} = 0.1s$, and $\phi_A^{*(\cdot,2)} = 0.95$.
 
-To calculate the wiggliness $\Ztwo_{t,t^*}$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz).
-Recall that $\Ztwo_{t,t^*}$ is the sum of Gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also Gamma-distributed.
-From Equation (\ref{eqn:abYhat}), we have:
+To calculate the wiggliness $\Ztwo_{t,\tilde t^*}$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz). From Equation (\ref{eqn:abYhat}):
 %
 \begin{equation}
-    \Ztwo_{t,t^*} = \sum_{k=1}^{10} \big|\big|\hat{Y}^{(k)}_{t,t^*}\big|\big|^2 = \sum_{k=1}^{10} \big|\big|a^{(k)}_{t,t^*}i\sqrt{b^{(k)}_{t,t^*}}\big|\big|^2 = \sum_{k=1}^{10} \big|b^{(k)}_{t,t^*}\big|
+    \Ztwo_{t,\tilde t^*} = \sum_{k=1}^{10} \big|\big|\hat{Y}^{(k)}_{t,\tilde t^*}\big|\big|^2 = \sum_{k=1}^{10} \big|\big|a^{(k)}_{t,\tilde t^*}i\sqrt{b^{(k)}_{t,\tilde t^*}}\big|\big|^2 = \sum_{k=1}^{10} b^{(k)}_{t,\tilde t^*}
     \label{eqn:Wb}
@@
 %
-since $a^{(k)}_{t,t^*} \in \{-1,1\}$. Combining Equation (\ref{eqn:Wb}) above with Equation block (\ref{eqn:bdist}) gives
+since $a^{(k)}_{t,\tilde t^*} \in \{-1,1\}$ and $b^{(k)}_{t,\tilde t^*} \geq 0$.
+Since $\Ztwo_{t,\tilde t^*}$ is the sum of Gamma-distributed random variables with identical scale parameters, the distribution of $\Ztwo_{t,\tilde t^*}$ is also Gamma.
+Combining Equation (\ref{eqn:Wb}) above with Equation (\ref{eqn:bdist}) gives
 %
-$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 5/k^2 = 10.10, 1.00 \right) \text { and }$$
+$$\left(\Ztwo_{t,\tilde t^*}|X^*_{t,\tilde t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 5/k^2 = 10.10, 1.00 \right) \text { and }$$
 %
-$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , 1.00 \right),$$
+$$\left(\Ztwo_{t,\tilde t^*}|X^*_{t,\tilde t^*} = 2\right) \sim {\rm{Gamma}}\left(300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , 1.00 \right),$$
 %
 where the first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the scale parameter and the second is the shape parameter.
-Simple calculation of the mean and variance of a Gamma distribution shows that $\mu_W^{*(\cdot,1)} = 10.10$, $\sigma_W^{*(\cdot,1)} = 3.18$, $\mu_W^{*(\cdot,2)} = 305.94$, and $\sigma_W^{*(\cdot,2)} = 17.49$.
+Simple calculation of the mean and variance of a Gamma distribution reveals $\mu_W^{*(\cdot,1)} = 10.10$, $\sigma_W^{*(\cdot,1)} = 3.18$, $\mu_W^{*(\cdot,2)} = 305.94$, and $\sigma_W^{*(\cdot,2)} = 17.49$.
 
@@
 %
-$$\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,z^*) = \delta P(y_1,z_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\theta,\theta^*,\Gamma^*) \mathbf{1}_N.$$
+$$\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,\z) = \delta P(y_1,\z_1;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,\z_t;\theta,\theta^*,\Gamma^*) \mathbf{1}_N.$$
 %
@@
 \begin{align*}
-P(y_t,z_t^*;\theta,\theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\theta^{(1)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(1)};z_t^*\right), \ldots , \\
-&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(N)};z_t^*\right) \Big],
+P(y_t,\z_t;\theta,\theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\theta^{(1)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(1)};\z_t\right), \ldots , \\
+&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(N)};\z_t\right) \Big],
 \end{align*}
@@
 %
-$$\calL_{\text{CarHMM-DFT}}\left(\theta^{*},\Gamma^{*(i)};z_t^*\right) = \delta^{*(i)} \prod_{z^*_{t,t^*} \in z^*_t} \Gamma^{*(i)} P(z^*_{t,t^*}|z^*_{t,t^*-100};\theta^*) \mathbf{1}_N,$$
+$$\calL_{\text{CarHMM-DFT}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_N,$$
 %
-where $P(z^*_{t,t^*}|z^*_{t,t^*-100};\theta^*)$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{*(\cdot,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$.
+where $P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*)$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{*(\cdot,i^*)}(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1}; \theta^{*(\cdot,i^*)})$.
 %
-Recall that $z^*_t = \left\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\right\}$ and that $f^{*(\cdot,i^*)}(\cdot|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.
+Recall that $f^{*(\cdot,i^*)}(\cdot|\z_{t,t^*-1}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,\tilde t^*}$ when $\tilde X^*_{t,\tilde t^*} = i^*$ and $\Z_{t,\tilde t^*-1} = \z_{t,\tilde t^*-1}$.
 

### 2020-11-13 Report/Sections/appendix.tex
shorten intro

@@
 It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
-Since accelerometer readings are taken at 50 hertz and this model has two-second window lengths, each window contains 100 accelerometer readings.
-In addition, the length of each fine-scale Markov chain is $\tilde T ^*_t = \lfloor Y_t/2 \rfloor$.
-Then, given the dive type $X_t = i$, dive duration $Y_t = y_t$, and the fine-scale probability transition matrix $\Gamma^{*(i)}$, it is straightforward to simulate the subdive states $\tilde{X}^*_t$ for a particular dive $t$.
-It is much more difficult to simulate the raw acceleration data $Y_t^*$, and we focus on this task for the remainder of the section.
+It is also straightforward to simulate $\tilde X^*_t$ given the length of each fine-scale Markov chain ($\tilde T^*_t = \lfloor Y_t/2 \rfloor$), the dive types $X_t$, dive durations $Y_t$, and fine-scale probability transition matrices $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$.
+We assume that $X^*_{t,t^*}$ is equal to $\tilde X^*_{t,\tilde t^*}$ within the window corresponding to time-stamp $t^*$ (i.e., $X^*_{t,t^*} = \tilde X_{t,\lceil t^*/100 \rceil}$).
+It is much more difficult to simulate the raw acceleration data $Y^*_t$, and we focus on this task for the remainder of this section.
 
-Let $DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,\tilde t^*}$ and $DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$ be denoted as $\hat{Y}^*_{t,\tilde t^*}$ for convenience.
-We simulate $Y_t^*$ in three steps: 1) simulate the average acceleration within each window $\left(Y^{*(0)}_{t,\tilde t^*}\right)$, 2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and 3) take the inverse $DFT$ of $\hat{Y}^*_{t,\tilde t^*}$ to retrieve $\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$.
-The details of this procedure are described below:
+Let $DFT\{Y^*_{t,100(\tilde t^*-1) + 1},\ldots,Y^*_{t,100t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,\tilde t^*}$ and $DFT\{Y^*_{t,100(\tilde t^*-1) + 1},\ldots,Y^*_{t,100t^*}\}$ be denoted as $\hat{Y}^{*}_{t,\tilde t^*}$, where $\{Y^*_{t,100(\tilde t^*-1) + 1},\ldots,Y^*_{t,100t^*}\}$ represents window $\tilde t^*$.
+
+We simulate $Y_t^*$ in three steps: 1) simulate the average acceleration within each window $\left(\hat{Y}^{*(0)}_{t,\tilde t^*}\right)$, 2) simulate the rest of the Fourier modes within that window $\left(\hat{Y}^{*(k)}_{t,\tilde t^*}, \enspace k = 1,\ldots,99\right)$, and 3) take the inverse Fourier transform of each set of Fourier coefficients $IDFT\left
+\{\hat{Y}^{*}_{t,\tilde t^*}\right\}$. We go over each step in detail below:
 
 \begin{enumerate}
-\item Simulate the average acceleration $Y^{*(0)}_{t,\tilde t^*}$ for each window $\tilde t^* = 1,2,\ldots,\tilde T^*_t$ using the following procedure:
+\item First simulate the average acceleration within each window $\left(\hat{Y}^{*(0)}_{t,\tilde t^*}\right)$:
 \begin{enumerate}
@@
     $$
-    	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
+    	(\hat{Y}^{*(0)}_{t,1}|\tilde X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \quad i^* \in \{1,2\}
     $$
@@
     %
-    \item Simulate $\hat{Y}^{*(0)}_{t,2}, \ldots,\enspace \hat{Y}^{*(0)}_{t,\tilde T^*_t}$ using Equation (\ref{eqn:yhat_0}) below:
+    \item Simulate $\hat{Y}^{*(0)}_{t,2}, \enspace \ldots \enspace \hat{Y}^{*(0)}_{t,\tilde{T}_t^*} $ using Equation (\ref{eqn:yhat_0}) below:
     \begin{align}	
-        \Big(\hat{Y}^{*(0)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = i^*,\hat{Y}^{*(0)}_{t,\tilde t^*-1}\Big) &\sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,\tilde t^*-1}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \nonumber \\
+        \Big(\hat{Y}^{*(0)}_{t,\tilde t^*}|\tilde X^*_{t, \tilde t^*} = i^*,\hat{Y}^{*(0)}_{t,\tilde t^*-1}\Big) &\sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,\tilde t^*-1}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \nonumber \\
         %
-        &t^* = 2,\ldots, \tilde T^*_t
+        &t^* = 1,2,\ldots,\tilde{T}_t^*
     	\label{eqn:yhat_0}
@@
 \end{enumerate}
-    After this step we have simulated the average acceleration within each two-second window for all dives (see Equation (\ref{eqn:avg_val}) below).
-    \item Simulate the other Fourier coefficients $\left(Y^{*(k)}_{t,\tilde t^*}\right)$ for each window $\tilde t^* = 1,2,\ldots,\tilde T^*_t$ and $k = 1,\ldots,99$ using the following procedure:
+%
+This process simulates the average acceleration within each two-second window $\tilde t^* = 1, \ldots, \tilde T^*_t$ (see Equation (\ref{eqn:avg_val}) below). 
+\item Next simulate the rest of the Fourier modes within each window ($\hat{Y}^{*(k)}_{t,\tilde t^*}, \enspace k = 1,\ldots,99$):
+%
 \begin{enumerate}
-    %
-    \item Simulate Fourier coefficients 1 through 49:
+    \item Simulate Fourier coefficients 1 through 49 within each window:
     \begin{equation}
-        \hat{Y}^{*(k)}_{t,\tilde t^*} = a_{t,\tilde t^*}^{(k)} i\sqrt{b^{(k)}_{t,\tilde t^*}}, \quad k = 1,\ldots,49;
+        \hat{Y}^{*(k)}_{t,\tilde t^*} = a_{t,\tilde t^*}^{(k)} i\sqrt{b^{(k)}_{t,\tilde t^*}}, \quad k = 1,\ldots,49
         \label{eqn:abYhat}
     \end{equation}
-    where $a^{(k)}_{t,\tilde t^*}$ is equal to either 1 or -1 each with probability 1/2, and $b^{(k)}_{t,\tilde t^*}$ has the following distribution:
+    where $a^{(k)}_{t,\tilde t^*}$ is equal to either 1 or -1 with probability 1/2 each, and $b^{(k)}_{t,\tilde t^*}$ has the following distribution:
     \begin{align}
@@
     The first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the scale parameter and the second is the shape parameter.
-The magnitude of the $k^{th}$ Fourier coefficient is equal to $b^{(k)}_{t,\tilde t^*}$, which decays like $1/k^2$ to ``smooth out" the raw acceleration data.
+The expected value $\mathbb{E}\left(|b^{(k)}_{t,\tilde t^*}|\right)$ decays like $1/k^2$ to ``smooth out" the raw acceleration data.
     %
-    \item Set Fourier coefficients 50 through 99 so the inverse DFT is real-valued:
+    \item Set Fourier coefficients 50 through 99 to ensure that the inverse DFT is real-valued:
     $$
@@
 \end{enumerate}
-    \item Finally, set the raw acceleration data within each window $\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$ with the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$:
+This completes the process of simulating $\hat{Y}^{*(k)}_{t,\tilde t^*}$ for all windows $\tilde t^* = 1,\ldots,\tilde T_t^*$ and Fourier coefficients $k = 0,\ldots,99$. 
+\item Finally, set each window of raw acceleration $\left\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\right\}$ using the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$:
 %
-    $$\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\} \equiv IDFT\left\{\hat{Y}^*_{t,\tilde t^*}\right\}$$
-    for $\tilde t^* = 1,\ldots,\tilde T^*_t.$
+$$Y^*_{t,100(\tilde t^* - 1)+ n+1} = IDFT\left\{\hat{Y}^*_{t,\tilde t^*}\right\}(n), \quad n = 0,\ldots,99, \quad \tilde t^* = 1,2,\ldots,\tilde T_t^*.$$
+%
+This completes the construction of $Y^*_t$.
 \end{enumerate}
 
-We now show that this construction of $Y^*_t$ results in the distributions listed in Section \ref{subsec:data_simulation}.
-First, note that $\hat{Y}^{(0)}_{t,\tilde t^*}$ is just the sum of the raw acceleration values within window $\tilde t$:
+We now show that this construction results in the distributions defined in Section \ref{subsec:data_simulation}.
+First, note that $\hat{Y}^{(0)}_{t,\tilde t^*}$ is just the sum of the raw acceleration values within the corresponding window:
 %
 \begin{align}
-    \hat{Y}^{(0)}_{t,\tilde t^*} &=
-    DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}(0) \nonumber \\
-    &= \sum_{n=0}^{99} Y^*_{t,100(\tilde t^* - 1) + 1 + n} \exp\left(-\frac{i2\pi}{99}*0*n\right) \nonumber \\
-    &= \sum_{n=1}^{100} Y^*_{t,100(\tilde t^* - 1) + n} \nonumber \\
-    &= 100A^*_{t,\tilde t^*}.
+    \hat{Y}^{(0)}_{t,\tilde t^*} &= DFT\{Y^*_{t,100(\tilde t^*-1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}(0) \nonumber \\
+    &= \sum_{n=0}^{99} Y^*_{t,100(\tilde t^*-1) + 1 + n} \exp\left(-\frac{i2\pi}{99}*0*n\right) \nonumber \\
+    &= \sum_{n=1}^{100} Y^*_{t,100(\tilde t^*-1) + n} \nonumber \\
+    &= 100\Zone_{t,\tilde t^*}.
     \label{eqn:avg_val}
@@
 %
-Combining Equation (\ref{eqn:avg_val}) above with Equation (\ref{eqn:yhat_0}) gives the distribution of $\Zone_{t,\tilde t^*}$: 
+Combining Equation (\ref{eqn:avg_val}) above with Equation (\ref{eqn:yhat_0}) gives the distribution of $\Zone_{t,t^*}$: 
 %
@@
 %
-$$\implies \left(\Zone_{t,\tilde t^*} | \tilde X^*_{t,\tilde t^*} = i^*, \Zone_{t,\tilde t^*-1} \right) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \Zone_{t,\tilde t^*-1}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right).$$
+$$\left(\Zone_{t,t^*} | \tilde X^*_{t,t^*} = i^*, \Zone_{t,\tilde t^*-1} \right) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \Zone_{t,\tilde t^*-1}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right).$$
 %
@@
 
-To calculate the wiggliness $\Ztwo_{t,\tilde t^*}$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz). From Equation (\ref{eqn:abYhat}):
+To calculate the wiggliness $\Ztwo_{t,\tilde t^*}$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz). Then, from Equation (\ref{eqn:abYhat}):
 %
@@
 %
-since $a^{(k)}_{t,\tilde t^*} \in \{-1,1\}$ and $b^{(k)}_{t,\tilde t^*} \geq 0$.
-Since $\Ztwo_{t,\tilde t^*}$ is the sum of Gamma-distributed random variables with identical scale parameters, the distribution of $\Ztwo_{t,\tilde t^*}$ is also Gamma.
+since $a^{(k)}_{t,t^*} \in \{-1,1\}$ and $b^{(k)}_{t,\tilde t^*} \geq 0$.
+Equation (\ref{eqn:Wb}) shows that $\Ztwo_{t,\tilde t^*}$ is the sum of Gamma-distributed random variables with identical scale parameters, so $\Ztwo_{t,\tilde t^*}$ is also Gamma-distributed.
 Combining Equation (\ref{eqn:Wb}) above with Equation (\ref{eqn:bdist}) gives
 %
-$$\left(\Ztwo_{t,\tilde t^*}|X^*_{t,\tilde t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 5/k^2 = 10.10, 1.00 \right) \text { and }$$
+$$\left(\Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 5/k^2 = 10.10, 1.00 \right) \text { and }$$
 %
-$$\left(\Ztwo_{t,\tilde t^*}|X^*_{t,\tilde t^*} = 2\right) \sim {\rm{Gamma}}\left(300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , 1.00 \right),$$
+$$\left(\Ztwo_{t,t^*}|\tilde X^*_{t,\tilde t^*} = 2\right) \sim {\rm{Gamma}}\left(300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , 1.00 \right),$$
 %
 where the first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the scale parameter and the second is the shape parameter.
-Simple calculation of the mean and variance of a Gamma distribution reveals $\mu_W^{*(\cdot,1)} = 10.10$, $\sigma_W^{*(\cdot,1)} = 3.18$, $\mu_W^{*(\cdot,2)} = 305.94$, and $\sigma_W^{*(\cdot,2)} = 17.49$.
+Simple calculation of the mean and variance of a Gamma distribution shows that $\mu_W^{*(\cdot,1)} = 10.10$, $\sigma_W^{*(\cdot,1)} = 3.18$, $\mu_W^{*(\cdot,2)} = 305.94$, and $\sigma_W^{*(\cdot,2)} = 17.49$.
 
@@
 
-The overall likelihood of the CarHHMM-DFT model is as follows:
+Define $\z = \{\z_1,\ldots,\z_{T}\}$. The overall likelihood of the CarHHMM-DFT model is as follows:
 %
@@
 %
-$$\calL_{\text{CarHMM-DFT}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_N,$$
+$$\calL_{\text{CarHMM-DFT}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 1}^{\tilde T_t^*} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_N,$$
 %
@@
 %
-Recall that $f^{*(\cdot,i^*)}(\cdot|\z_{t,t^*-1}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,\tilde t^*}$ when $\tilde X^*_{t,\tilde t^*} = i^*$ and $\Z_{t,\tilde t^*-1} = \z_{t,\tilde t^*-1}$.
+Recall that $f^{*(\cdot,i^*)}(\cdot|\z_{t,\tilde t^*-1}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,\tilde t^*}$ conditioned on $\tilde X^*_{t,\tilde t^*} = i^*$ and $\Z_{t,\tilde t^*-1} = \z_{t,\tilde t^*-1}$.
 

### 2020-11-13 Report/Sections/models.tex
shorten intro

@@
+% !TeX root = ../main.tex
+
 %\section{Models and parameter estimation}
 
-Consider a sequence of $T$ curves (or dives, in our context). For any particular $t = 1,
-\ldots,T$, dive $t$ is characterized by a possibly high-dimensional curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$.
-The sequence $Y^*_{t}$ is made up of function values from curve $t$.
+Consider a sequence of $T$ curves where for any particular $t = 1,
+\ldots,T$, curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$.
+Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of function values from curve $t$.
 We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.
+We first detail the structure of a traditional HMM followed by three variations which generalize its base structure.
+We then show how each of these generalized models can be synthesized to form a wide variety of more complicated models.
 
-Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$.
-We first detail the structure of a traditional HMM.
-We then review a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
-Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$.
-To simultaneously model $Y$ and $Y^*$, we use the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
-%which is comprised of two distinct,x but related, HMMs that jointly model these time series. 
-Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics.
-In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour.
+%Hidden Markov models describe state-switching Markovian processes in discrete time and is the core structure we use to model $Y$ and $Y^*$.
+We first detail the structure of a traditional HMM, followed by the conditionally auto-regressive hidden Markov model (CarHMM) \citep{Lawler:2019}, a variation which explicitly models auto-correlation between observations.
+%Either the HMM or the CarHMM can model either the coarse scale sequence $Y$ or any of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$. 
+%We then describe a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019} to simultaneously model $Y$ and $Y^*$. 
+%, which is comprised of two distinct, but related, HMMs that jointly model these time series. 
+%Next, we outline a method to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics.
+In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to capture its periodic behaviour.
 Finally, we show how each of these models can be altered and combined to form a wide variety of new models.
@@
 
-An HMM is comprised of a sequence of unobserved states $X \equiv \big\{X_1, \ldots, X_T\big\}$ where $X_t$ is associated with the observation $Y_t$.
-Here we focus on the coarse scale sequence of observations $Y_1,\ldots, Y_T$, but an HMM can also be used as a sub-model of the fine-scale sequence for a particular dive.
+Hidden Markov models describe state-switching Markovian processes in discrete time and are the core structure we use to model both $Y$ and $Y^*$.
+For simplicity we focus on $Y$ to introduce the model.
+An HMM is comprised of a sequence of unobserved states $X \equiv \big\{X_1, \ldots, X_T\big\}$ together with an observation sequence $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$, where $X_t$ is associated with the observation $Y_t$.
 The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time.
-The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$.
+The $X_t$'s form a Markov chain and can take values $1, \ldots, N$.
 Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$.
 %
-We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i^{th}$ component
+We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$ with $i^{th}$ component
 $\delta_i = \Pr(X_1 = i).$
@@
 For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour.
-To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize this non-Markovian behaviour.
+We recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant summary statistics to describe any fine-scale non-Markovian behaviour.
 To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over the elements $Y_t^*$.
 Subject-area experts are often required to determine the optimal window size and the specific summary statistics used.
-For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ and stride $h$ across $Y^*_t$.
-Namely:
+For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width and stride length $h$ across $Y^*_t$:
 %
@@
 %
-for $t^* = 1,h+1,2h+1,\ldots$ and for $k = 0, 1, \ldots, h-1$.
+for each $t^* = 1,h+1,2h+1,\ldots$ and $k = 0,1,\ldots, h-1$.
 If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise.
-Next, we index each window with $\tilde{t}^*$ and calculate the transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$, where
+We then index each window with $\tilde t^*$ and calculate $\Z_{t,t^*} \equiv \{\Zone_{t,\tilde t^*},\Ztwo_{t,\tilde t^*}\}$, where
 %
 \begin{equation}
+\begin{split}
     \label{eqn:z}
-    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|^2.
+    \Zone_{t,\tilde t^*} &\equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde t^*-1)+n} \\
+    %
+    \Ztwo_{t,\tilde t^*} &\equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde t^*-1) + 1},\ldots, Y^*_{t,h\tilde t^*}\}(k)\Big|\Big|^2.
+\end{split}
 \end{equation}
 %
-In words, $\Zone_{t,\tilde{t}^*}$ is the average acceleration vector over window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega} \leq h-1$ periods per window (by the Plancherel theorem).
-More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within window $\tilde{t}^*$.
+The collection of $\Z_{t,\tilde t^*}$ for all windows is $\Z_t \equiv \{\Z_{t,1},\ldots,\Z_{t,\tilde T_t^*}\}$, where $\tilde T_t^* = \lfloor T^*_t/h \rfloor$.
+The hidden states corresponding to $\Z_t$ are distinct from those corresponding to $Y^*_t$ and are defined as $\tilde X^*_t \equiv \{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T_t^*}\}$.
+
+In words, $\Zone_{t,\tilde t^*}$ is the average value over window $\tilde t^*$ and $\Ztwo_{t,\tilde t^*}$ is the squared two-norm of the component of window $\tilde t^*$ that can be attributed to frequencies between one and $\tilde{\omega} \leq h-1$ periods per window (by the Plancherel theorem).
+More intuitively, $\Ztwo_{t,\tilde t^*}$ corresponds to the ``wiggliness" of the acceleration data during dive $t$ and window $\tilde t^*$.
 The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert.
 Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude.
-Setting the stride length equal to the window size $h$ avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
-By summarizing the data using a relevant transformation, this method reduces correlation in the raw observations due to any complex dependence structure.
+Transforming $Y^*_t$ to $\Z_t$ using a non-overlapping window transformation both avoids introducing artificial correlation and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
+By summarizing the data using a relevant transformation, the method overall also reduces correlation in the raw observations by summarizing complex dependence structures within each window.
+Figure \ref{fig:models}d represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM, including down-sampling.
 
-The fine-scale HMM must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself.
-In particular, the length of $\Z_t$ is denoted as $\tilde{T}_t^*$ and there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, denoted as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
-The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$.
-It then follows that the likelihood of this model is identical to that of the original HMM but with $Y^*$ replaced with $\Z$ and $X^*$ replaced with $\tilde{X^*}$.
-In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X}$ as hidden states as an HMM-DFT.
+%When down-sampling, we redefine the fine-scale probability transition matrices such that $\Gamma^{*(i)}_{jk} = \Pr(X^*_{t,t^*+h} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1,h+1,2h+1,\ldots$.
+If the fine-scale model is a CarHMM, we also redefine the probability density function $f$ such that $f^{*(i,i^*)}(\cdot|\z_{t,t^*-100}; \theta^{*(i,i^*)})$ describes the density of $\Z_{t,t^*}$ given $X_t = i$, $X^*_{t,t^*} = i^*$, and $\Z_{t,t^*-100} = \z_{t,t^*-100}$.
+%
+%Once $\Z_t$ is calculated, it simply replaces the original fine-scale observations $Y^*_t$ in the fine-scale HMM structure. As such, the emission distributions and the 
+The likelihood of an HMM with transformed fine-scale observations is identical to the original HMM but with $\Z_t$ and $\tilde X^*_t$ replacing $Y^*_t$ and $X^*_t$, respectively.
+In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations as an HMM-DFT.
 

### 2020-11-13 Report/Sections/results.tex
shorten intro

--- 2020-10-30 Report/Sections/results.tex
@@
 %We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
-We then have that $\z_{t,t^*} = \left\{\zone_{t,t^*},\ztwo_{t,t^*}\right\}$, where $\zone_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,t^*}$ is a scalar describing the ``wiggliness" of a particular window.
-We also down-sample each observation $\z_t$ such that $\z_t = \left\{\z_{t,1},\z_{t,101},\z_{t,201},\ldots\right\}$ and $x^*_t = \left\{x^*_{t,1},x^*_{t,101},x^*_{t,201},\ldots\right\}$ to avoid overlapping windows.
-Even after transforming and down-sampling the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material).
+We then have that $\z_{t,\tilde t^*} = \left\{\zone_{t,\tilde t^*},\ztwo_{t,t^*}\right\}$, where $\zone_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,t^*}$ is a scalar describing the ``wiggliness" of window $\tilde t^*$.
+Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material).
 Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.
@@
 
-All components of $\Zone_{t,t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $t^*$.
+All components of $\Zone_{t,\tilde t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $\tilde t^*$.
 Each component is assumed to have its own mean and variance parameters, but all components have the same auto-correlation parameter.
 %Given that $X^*_{t,t^*} = i^*$, the corresponding unknown parameters are then $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}_{>0}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
-Given that $X^*_{t,t^*} = i^*$, $\Ztwo_{t,t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
-We do not include $\Ztwo_{t,t^*-100}$ in the distribution of $\Ztwo_{t,t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
+Given that $X^*_{t,t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
+We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
 %
@@
 %
-Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $Z_{t,t^*}$ when conditioned on $X^*_{t,t^*} = i^*$. 
+Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $\Z_{t,t^*}$ when $X^*_{t,t^*} = i^*$. 
 %
 We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation.
-Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model.
+Figure (\ref{fig:CarHHMM-DFT}) shows the corresponding graphical model.
 The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
@@
 One exception is $\Ztwo_{t,t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:pseudoresids}).
-This implies that the true distribution of $\Ztwo_{t,t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study.
+This implies that the true distribution of $\Ztwo_{t,\tilde t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study.
 See Sections 1.4 through 1.6 of the supplementary material for pseudoresidual plots for all observations.

### 2020-11-13 Report/Sections/simulation.tex
shorten intro

--- 2020-11-05 Report/Sections/simulation.tex
@@
 Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 25s$.
-After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2T^*_t$ seconds of each dive are ignored.
+After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each dive are ignored.
 Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
@@
 for dive types 1 and 2, respectively.
-Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $Z^*_{t,t^*} = \left\{\Zone_{t,t^*}, \Ztwo_{t,t^*}\right\}$ as defined in Equation (\ref{eqn:z}).
-Recall that $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ are independent when conditioned on all dive types and subdive states.
-Each observation of acceleration $\Zone_{t,t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,t^*}$ is Gamma-distributed without explicit auto-correlation.
-In particular, we set the distributions of $\Zone_{t,t^*}$ and $\Ztwo_{t,t^*}$ with the following parameters:
+Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$ as defined in Equation (\ref{eqn:z}).
+Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states.
+Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation.
+In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
 %
@@
 If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives.
-Analogously, if $X^*_{t,t^*}$ is the unknown subdive state associated with a particular model and $x^*_{t,t^*}$ is the true simulated subdive state for dive $t$ and window $t^*$, then we define the average subdive decoding accuracy as $\Pr(X^*_{t,t^*} = x^*_{t,t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows.
+Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,\tilde t^*} = \tilde x^*_{t,\tilde t^*}|Y,\tilde Y^*)$ averaged across all simulated data sets, dives, and windows.
 We find both the average dive decoding accuracy and the average subdive decoding accuracy using the forward-backward algorithm.

### 2020-11-13 Report/Sections/intro.tex

@@
+% !TeX root = ../main.tex
+
 Biologging technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
 The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}.
-In particular, the study of animal movement has been transformed by tracking devices that collect biologging data \citep{Borger:2020}.
+In particular, the study of animal movement has been transformed by advanced tracking devices which collect biologging data \citep{Borger:2020}.
 Tags can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
-These data contain a wealth of knowledge about human and animal behaviour, but parsing these large data sets is a challenge for statisticians.
+These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a new challenge for statisticians.
 
-Biologging data are sometimes modelled as a set of curves that can be analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citealt{Ramsay:2005}).
-For example, \cite{Morris:2007} view a child's activity level throughout the day as a curve where metabolic activity is a function of time.
-\citet{Fu:2017} also show that estimating the amplitude and phase variation of elephant seal (\textit{Mirounga leonina}) dive curves is useful to classify dive types, although they ignore sequential dependence between dives.
-Similarly, our case study views the dive profile of a killer whale (\textit{Orcinus orca}) as a sequence of curves where three orthogonal axes of acceleration are each functions of time.
+Biologging data are sometimes modelled as a set of curves analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citet{Ramsay:2005}).
+For example, \citet{Morris:2007} view a child's activity level throughout the day as a curve where metabolic activity is a function of time.
+Similarly, \citet{Fu:2017} view the dive profile of elephant seal (\textit{Mirounga leonina}) as a sequence of dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
+%, although they ignore sequential dependence between dives. 
+%Similarly, our case study views the dive profile of a killer whale (\textit{Orcinus orca}) as a sequence of curves where three orthogonal axes of acceleration are each functions of time.
+%
+FDA was originally developed to process curves assumed to be independent replicates, but time series data of this type are often complicated by dependencies which violate this assumption.
+In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017}.
+%, as is the case for our case study data.
+Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time.
+For example, marine mammal dive profiles show distinct dive ``types" that appear to cluster together in time  \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}.
+Fine-scale periodicity within a larger process in particular is common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
 
-FDA was originally developed for curves that can be considered independent replicates, but the structure of time-series data is often complicated by additional dependencies.
-In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017,Adam:2019}, as is the case for our case study data.
-Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time.
-Previous studies have categorized these dives into different types where each dive type is associated with a different behaviour, including resting, travelling, and foraging \citep{Tennessen:2019b}.
-Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence within dives, as the accelerometer data within each dive indicate bouts of short-term periodicity.
-Behaviourally dependent periodicity is not unique to the biologging data- in the field of machine health, \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks.
+%In addition, \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks.
+The raw acceleration data exhibit extreme periodicity (in the form of vibration) that depends upon the behaviour of the train as the load and speed of the train change.
+%To model between-curve dependence in dive profiles, previous studies have categorized killer whale (\textit{Orcinus orca}) dives into different types associated with specified behaviours such as resting, travelling, and foraging \citep{Tennessen:2019b}.
+%Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence within dives, as the accelerometer data within each dive indicate bouts of short-term periodicity. 
+%Behaviourally dependent periodicity is not unique to the biologging data- in the field of machine health, \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks.
 The raw acceleration data exhibit extreme periodicity (in the form of vibration) that depends upon the behaviour of the train as the load and speed of the train change.
 
-FDA literature provides models that account for some between-curve dependence, but these methods are not adequate when modelling certain types of time dependencies, including that of Figure \ref{fig:data}.
-For example, multilevel models incorporate dependence via random effects.
-In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors use a multilevel approach with a random effects to model variation between children.
-In a classic paper that introduced this multilevel structure,  \cite{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles.
+Some FDA models account for between-curve dependence but are inadequate when modelling certain types of time dependencies. %, including that of Figure \ref{fig:data}.
+For example, 
+%multilevel models incorporate dependence via random effects. 
+Previous studies have used multilevel models with random effects to capture variation between activity levels in children \citep{Morris:2007} and study differences in menstrual cycles \citet{Bromback:1998}.
+More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012} and \citet{Crainiceanu:2009}.
+However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types.
+%In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors used a multilevel approach with a random effects to model variation between children.
+In a classic paper that introduced this multilevel structure,  \citet{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles.
 More recent work on multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012} and \citet{Crainiceanu:2009}.
 However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types.
-In addition to multilevel models, FDA researchers have used functional time series to model dependence in sequences of curves.
+In addition to multilevel models, FDA researchers have also used functional time series to model dependence in sequences of curves.
 Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}.
-While functional time series methods can in theory be applied to biologging data such as ours, they do not account for sequences of curves that switch between a discrete number of types over time.
+%While functional time series methods can in theory be applied to biologging data such as ours,
+However, functional time series do not account for sequences of time series curves that switch between a discrete number of well defined states.
+%
+Traditional FDA techniques similarly fall short when modelling complicated within-curve %acceleration 
+data. In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function \citep{Yao:2005} 
+%use this framework to construct a method of sparse longitudinal data analysis using functional principal component analysis. Alternatively, within-curve structure is sometimes generated 
+or via random regression \citep{Rice:2001}. 
+%For example, \citet{Rice:2001} use random regression to construct non-parametric mixed effects models for unequally sampled noisy curves.
+However, the combination of apparent behavioural changes combined with periodic behaviour makes accelerometer data such as ours difficult to model with these classic FDA techniques.
+However, time series data exhibiting both sharp behavioural changes and periodic fine-scale structure is difficult to model with these classic FDA techniques.
 
-Traditional FDA techniques similarly fall short when modelling complicated within-curve acceleration data.
-In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function.
-\citet{Yao:2005} use this framework to construct a method of sparse longitudinal data analysis using functional principal component analysis.
-Alternatively, within-curve structure is sometimes generated via random regression.
-For example, \citet{Rice:2001} use random regression to construct non-parametric mixed effects models for unequally sampled noisy curves.
-However, The combination of apparent behavioural changes combined with periodic behaviour makes accelerometer data such as ours difficult to model with these classic FDA techniques.
+In light of these challenges, we turn to the field of
+%methodology for the analysis of animal movement 
+%statistical ecology
+animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
+HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from observations of its position.
+While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling, with a few notable exceptions.
+In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines,
+%This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments. 
+but this approach does not account for certain types of time correlation within or between individual time series.
 
-In light of these challenges, we turn to vast field of methodology for the analysis of animal movement \citep{Hooten:2017}, where one of the most prevalent techniques in the literature of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
-HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from movement observations.
-While ubiquitous in the ecology literature, HMMs have seen little use in non-parametric functional modelling, with a few notable exceptions.
-In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations in terms of B-splines.
-%This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments. 
-However, this approach does not account for certain types of time correlation within or between the time series themselves, which is the focus of our efforts here.
+While useful, HMMs alone are also not sufficient to model intricate fine-scale time series data for three primary reasons.
+%
+Firstly, HMMs assume that subsequent observations are independent given an underlying hidden state process, but this is often not the case when observations are taken at extremely high frequencies. 
+Several solutions have been proposed in the ecology literature such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
+The CarHMM in particular explicitly models auto-correlation into an HMM while largely maintaining its base structure.
+%
+Secondly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales, (i.e., between and within curves). 
+To deal with this issue animal movement modellers have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model processes at both scales with conditionally dependent HMMs.
+%
+Finally, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures at short time scales.
+For example, \citet{Adam:2019} fail to capture fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}) using a traditional HHMM.
+\citet{Heerah:2017} use Fourier analysis within an HMM to account for %periodic behaviour in the context of describing
+daily behavioural cycles in marine mammals.
+In addition, Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
+Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for fine-scale periodic structures.
 
-While useful, traditional HMMs are also not sufficient to model accelerometer data such as ours for three primary reasons.
-Firstly, subsequent observations are assumed to be independent given the underlying hidden state process, but this is often not the case when observations are taken at high frequencies.
-Several solutions have been proposed in the ecology literature, such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
-The CarHMM in particular explicitly models auto-correlation into an HMM while adding minimal complexity to the base structure of an HMM.
-
-In addition, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales, including our example of both between- and within-curve temporal dependencies.
-To deal with this issue, Animal movement modellers have used hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}.
-In particular, HHMMs treat both scales as HMMs where the behaviour of the fine-scale process depends on the state of the coarse-scale process.
-
-Finally, although they are useful in many situations, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated structures at short time scales.
-The HHMM of \citet{Adam:2019} fails to capture the fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}).
-\citet{Heerah:2017} have demonstrated the advantages of using Fourier analysis within an HMM to account for periodic behaviour in the context of describing daily behavioural cycles of marine mammals.
-In addition, Fourier analysis has previously been used on accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
-Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for periodic structures in fine-scale accelerometer data.
-
-While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together.
-In this paper, we combine several existing methods from the ecology literature in novel ways to account for complex temporally dependent functional data.
-The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models and thus extend the methods available for the analysis of functional data.
-We first describe HMMs as well as to two variants (CarHMMs and HHMMs) and discuss how Fourier analysis can handle fine-scale dependence structures.
+%While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together. 
+In this paper, we combine existing methods from statistical ecology literature in novel ways to account for complex temporally dependent functional data.
+The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models to model multi-scale functional and time-series data with intricate dependence structure.
+%and thus extend the methods available for the analysis of functional data. 
+We begin by describing HMMs as well as to two variants (CarHMMs and HHMMs) and discuss how Fourier analysis can handle fine-scale dependence structures.
 We also show how these methods can be combined to analyze increasingly complex data.
-In Section 3, we construct several candidate models and fit them to the data from Figure \ref{fig:data}.
+In Section 3 we present a case study in which we fit several candidate models to data from a killer whale from the threatened northern resident population off the coast of British Columbia, Canada.
 Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.

### 2020-11-13 Report/main.tex

@@
 % Initial and revised submissions should be 12 point; this will be removed in the final version.
-\documentclass[12pt]{TD-CJS}
-%\documentclass{TD-CJS}
+%\documentclass[12pt]{TD-CJS}
+\documentclass{TD-CJS}
 
 % Initial and revised submissions should also be double spaced. This command will be removed in the final version.
-\renewcommand{\baselinestretch}{2}
+%\renewcommand{\baselinestretch}{2}
 
@@
 We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM).
-The distribution of each curve is then defined by another general fine-scale model which may involve a variety of methods such as moving-window transformations, Fourier analysis, and auto-correlation.
+The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods such as moving-window transformations, Fourier analysis, and auto-regression.
 As a case study, we use this framework to model the movement of a killer whale off the western coast of Canada.

### 2020-11-15 Report/Sections/intro.tex
small updates

@@
 In particular, the study of animal movement has been transformed by advanced tracking devices which collect biologging data \citep{Borger:2020}.
-Tags can record observations at rates of tens of hertz, resulting in time series containing millions of observations over the course of several hours.
+Tags can record observations at over 50 times per second, resulting in time series containing millions of observations over the course of several hours.
 These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a new challenge for statisticians.
 
-Biologging data are sometimes modelled as a set of curves analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citet{Ramsay:2005}).
+Biologging data are frequently modelled as a set of curves analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citet{Ramsay:2005}).
 For example, \citet{Morris:2007} view a child's activity level throughout the day as a curve where metabolic activity is a function of time.
@@
 For example, marine mammal dive profiles show distinct dive ``types" that appear to cluster together in time  \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}.
-Fine-scale periodicity within a larger process in particular is common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
+Fine-scale periodicity within a larger process in particular is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
 

### 2020-11-15 Report/Sections/models.tex
small updates

--- 2020-11-06 Report/Sections/models.tex
@@
 Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$.
-We first detail the structure of a traditional HMM.
-We then review a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
+We first detail the structure of a traditional HMM and a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM) which explicitly models auto-correlation between observations.
 Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$.

### 2020-11-15 Report/main.tex
small updates

@@
 
-Sequences of curves sampled at high-frequency in time are increasingly common in practice, but they can exhibit complicated dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA).
+Data sets comprised of sequences of curves sampled at high-frequency in time are increasingly common in practice, but they can exhibit complicated dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA).
 We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM).
 The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods such as moving-window transformations, Fourier analysis, and auto-regression.
+This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.
 As a case study, we use this framework to model the movement of a killer whale off the western coast of Canada.
 Through simulations we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines.
-This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.
 

### 2020-11-16 Report/Sections/models.tex
update models section

@@
 Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$.
-We first detail the structure of a traditional HMM and a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM) which explicitly models auto-correlation between observations.
+We first detail the structure of a traditional HMM and a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
 Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$.
@@
 In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour.
-Finally, we show how each of these models can be altered and combined to form a wide variety of new models.
+Finally, we show how each of these models can be synthesized to form a wide variety of new models.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The base structure of HMMs}
+\subsection{HMMs as a base structure}
 
-An HMM is comprised of a sequence of unobserved states $X \equiv \big\{X_1, \ldots, X_T\big\}$ where $X_t$ is associated with the observation $Y_t$.
-Here we focus on the coarse scale sequence of observations $Y_1,\ldots, Y_T$, but an HMM can also be used as a sub-model of the fine-scale sequence for a particular dive.
+An HMM is comprised of a sequence of unobserved states $X \equiv \big\{X_1, \ldots, X_T\big\}$ where the hidden state $X_t$ is associated with the observation $Y_t$.
+Here we focus on the coarse scale sequence of observations, but an HMM can also be used as a sub-model of the fine-scale sequence for a particular dive.
 The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time.
@@
 %
-We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$, with $i^{th}$ component
+We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$ with $i^{th}$ component
 $\delta_i = \Pr(X_1 = i).$
@@
 %
-The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
+The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: the conditional probability density function has the form $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
 %
@@
 
-Following \citet{Barajas:2017}, we reparameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows to sum to 1:
+Following \citet{Barajas:2017}, we reparameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows sum to 1:
 %
@@
 
-\subsection{Relaxing the conditional independence assumption with a CarHMM}
+\subsection{The CarHMM and relaxing conditional independence}
 
-The CarHMM, described by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process.
-Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
+A conditionally auto-regressive hidden Markov model, or CarHMM  \citep{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process.
+Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots,X_T$ that can take on integer values between $1$ and $N$.
+It also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
 Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$.
-The first emission $Y_1$ is assumed to be fixed and known as an initial value which does not depend upon $X_1$.
+The first emission $Y_1$ is treated as fixed as an initial value which does not depend upon $X_1$.
 We denote the conditional density of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$.
@@
 As a concrete example, if $Y_t$ is a $K$-dimensional vector with components $Y_t(1),\ldots,Y_t(K)$, consider the component-wise sequences $Y_1(k), \ldots, Y_T(k)$ for $k = 1,\ldots K$.
-One may assume that these sequences are independent given $X_1,\ldots,X_T$ with $Y_t(k)|X_t = i$ and that $Y_t(k)$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)}_k = \{\mu^{(i)}_k,\sigma^{(i)}_k,\phi^{(i)}_k\}$, where:
+One may assume that these sequences are independent given $X_1,\ldots,X_T$ and that $Y_t(k)$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)}_k = \{\mu^{(i)}_k,\sigma^{(i)}_k,\phi^{(i)}_k\}$, where:
 %
@@
 \end{align}
-
-A CarHMM which follows Equation (\ref{eqn:carhmm}) can be viewed as a discrete time version of the state-switching Ornstein-Uhlenbeck process of \citet{Michelot:2019} if observation error is negligible.
+%
+A CarHMM which follows Equation (\ref{eqn:carhmm}) can be viewed as a discrete time version of a state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}.
 This follows in the same way that an AR(1) process is the discrete-time version of a traditional Ornstein-Uhlenbeck process.
 Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a general CarHMM.
-The likelihood corresponding to a general CarHMM can be easily calculated using the forward algorithm.
-As previously, let $y$ be the vector of observed emissions.
-Then:
+As previously, the likelihood corresponding to a general CarHMM can be easily calculated using the forward algorithm.
+If $y$ is a vector of observed emissions, then
 \begin{equation*}
@@
 
-\subsection{Incorporating multiple scales with an HHMM}
+\subsection{The HHMM and incorporating multiple scales}
 
-An HHMM accounts for different scales of correlation by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs.
+A hierarchical hidden Markov model, or HHMM, accounts for different scales of correlation by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs.
 The coarse-scale model is as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations with state-dependent parameters $\theta^{(i)}, i = 1,\ldots,N$.
 %
-In the hierarchical setting, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$.
+In the hierarchical setting, however, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$.
 For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
@@
 Forcing certain parameters to be shared can reduce the complexity of the HHMM.
-For example, in the killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$.
+For example, in our killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$.
 We also share the fine-scale emission parameters across coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
+Coarse-scale hidden states therefore differ only in coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$.
 
@@
 \begin{equation}
-    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*;\theta,\theta^*,\Gamma^*) \mathbf{1}_N
+    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N
     \label{eqn:HHMM_likelihood}
@@
 %
-where $P(y_t,y_t^*;\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to 
-$f^{(i)}(y_t)\calL_{\text{HMM}}\left(\theta^{*(i)},
-\Gamma^{*(i)};y_t^*\right)$.
-
+where $P(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_1)\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
+\Gamma^{*(i)};y_1^*\right)$ for an HMM and $\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
+\Gamma^{*(i)};y_1^*\right)$ for a CarHMM.
+The fine-scale likelihood $\calL_{\text{(Car)HMM}}$ corresponds to the desired fine-scale model.
+In addition, $P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to
+$f^{(i)}(y_t|y_{t-1})\calL_{\text{HMM}}\left(\theta^{*(i)},
+\Gamma^{*(i)};y_t^*\right)$ for both a CarHMM and an HMM.
+%
 For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citet{Adam:2019}.
 
-\subsection{Transforming observations to account for fine-scale behaviour}
+\subsection{The HMM-DFT and transforming fine-scale observations}
 \label{subsec:STFT}
@@
 In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far.
-For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour.
-To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize this non-Markovian behaviour.
+%For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. 
+To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize any non-Markovian behaviour.
 To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over the elements $Y_t^*$.
-Subject-area experts are often required to determine the optimal window size and the specific summary statistics used.
+Subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window.
+Larger stride lengths result in a loss of information but also reduce the dimension of the fine-scale process.
+In addition, setting stride length and window size equal to one another avoids artificial residual correlation arising from overlapping windows.
 For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ and stride $h$ across $Y^*_t$.
@@
 for $t^* = 1,h+1,2h+1,\ldots$ and for $k = 0, 1, \ldots, h-1$.
-If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise.
-Next, we index each window with $\tilde{t}^*$ and calculate the transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$, where
+We omit the final window if $t^*+h-1$ exceeds $T^*_t$, and if $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise.
+Next, we index each window with $\tilde{t}^*$ and calculate transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$:
 %
@@
     \label{eqn:z}
-    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|^2.
+    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|^2,
 \end{equation}
 %
-In words, $\Zone_{t,\tilde{t}^*}$ is the average acceleration vector over window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega} \leq h-1$ periods per window (by the Plancherel theorem).
-More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within window $\tilde{t}^*$.
-The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert.
-Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude.
-Setting the stride length equal to the window size $h$ avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting.
-By summarizing the data using a relevant transformation, this method reduces correlation in the raw observations due to any complex dependence structure.
+where $\tilde{\omega} \leq h-1$ is a problem-specific tuning parameter corresponding to the maximum recorded frequency within each window.
+In words, $\Zone_{t,\tilde{t}^*}$ is the average acceleration vector during window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window.
+%(by the Plancherel theorem).
+More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the acceleration data within dive $t$ and window $\tilde{t}^*$.
 
 The fine-scale HMM must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself.
-In particular, the length of $\Z_t$ is denoted as $\tilde{T}_t^*$ and there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, denoted as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
-The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$.
-It then follows that the likelihood of this model is identical to that of the original HMM but with $Y^*$ replaced with $\Z$ and $X^*$ replaced with $\tilde{X^*}$.
-In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X}$ as hidden states as an HMM-DFT.
+In particular, if the length of $\Z_t$ is denoted as $\tilde{T}_t^*$ then there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
+The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ instead of $X^*_t$ and $Y^*_t$.
+The likelihood of this model is therefore identical to that of the original HMM, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$.
+In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT.
 
@@
 
-By definition, an HHMM treats both the coarse-scale and the fine-scale processes as a traditional HMM.
-However, it is straightforward to generalize the coarse-scale HMM by replacing it with a CarHMM, yielding the following likelihood:
-\[
-\calL_{\text{coarse}}(\theta,\theta^*,\Gamma;y,y^*) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\theta,\theta^*) \mathbf{1}_N
-\]
-where $P(y_t,y^*_t|y_{t-1};\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\theta^{(i)}\right)\calL_{\text{HMM}}\left(\theta^{*(i)};y^*_t\right)$.
-
-On the fine scale, $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods.
-In this case, the fine-scale HMM likelihood term $\calL_{\text{HMM}}$ in Equation (\ref{eqn:HHMM_likelihood}) is replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
+%By definition, an 
+Traditional HHMMs treat both the coarse-scale and the fine-scale processes as realizations of an HMM or CarHMM. 
+%However, it is straightforward to generalize the coarse-scale HMM by replacing it with a CarHMM, yielding the following likelihood:
+%\[
+%\calL_{\text{coarse}}(\theta,\theta^*,\Gamma;y,y^*) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\theta,\theta^*) \mathbf{1}_N
+%\]
+%where $P(y_t,y^*_t|y_{t-1};\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\theta^{(i)}\right)\calL_{\text{HMM}}\left(\theta^{*(i)};y^*_t\right)$.
+However, the fine-scale observations for a particular dive $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods.
+In this case, the fine-scale HMM likelihood term $\calL_{\text{(Car)HMM}}$ in Equation (\ref{eqn:HHMM_likelihood}) is replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
 \[
@@
 Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT) in addition to many others not described here.
-
 For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model.
@@
 Further, \citep{Langrock:2018} use B-splines to model the emission distribution of an HMM as well as incorporate covariates.
-This non-parametric approach admits a penalized likelihood which can easily be incorporated into the hierarchical structure here.
-One additional class of fine-scale models is the set of continuous time methods such as the continuous-time HMM (CTHMM) \citep{Liu:2015} or the state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019}.
+This non-parametric approach admits a penalized likelihood which can easily be incorporated into this hierarchical structure.
+Another class of fine-scale models is the set of continuous time methods such as a continuous-time HMM (CTHMM) \citep{Liu:2015} or a state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019}.
 A continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}.
-As an example, \citet{Xu:2018} explain biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours.
+\citet{Xu:2018} explain biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours.
 However, they take a longitudinal approach whereas we use a coarse-scale HMM in our case study.
-We also show the appendix that the state-switching OU process of \citet{Michelot:2019} is equivalent to a CarHMM under certain conditions.
 
-These examples are a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models.
-As an example, consider a CarHHMM, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM.
-Alternatively, one may construct an HHHMM, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$) and traditional HMMs are used to model all three levels.
-It is clear that are a myriad of possible models can be built using this framework.
-However, these models quickly become complicated and should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.
+These examples include a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models.
+For example, one may construct a three-level HHMM, where three scales of observations are recorded (fine-, medium-, and coarse-scale) and traditional HMMs are used to model all three levels.
+A myriad of possible models can be built using this framework, but these models quickly become complicated and computationally expensive to fit.
+Therefore, models should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.

### 2020-11-16 Report/Sections/intro.tex
address co-author edits thru 3.0

@@
 
-Biologging technology now provides researchers with data collected almost continuously in time \citep{Hooten:2017}.
+Biologging technology now provides researchers with kinematic data collected almost continuously in time \citep{Hooten:2017}.
 The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}.
-In particular, the study of animal movement has been transformed by advanced tracking devices which collect biologging data \citep{Borger:2020}.
+In particular, the study of animal movement has been transformed by short-term, high-resolution suction-cup attached tags that record kinematic information about animal movement underwater \citep{Borger:2020,Dot:2016b}.
 Tags can record observations at over 50 times per second, resulting in time series containing millions of observations over the course of several hours.
@@
 %use this framework to construct a method of sparse longitudinal data analysis using functional principal component analysis. Alternatively, within-curve structure is sometimes generated 
-or via random regression \citep{Rice:2001}. 
+or with random regression \citep{Rice:2001}. 
 %For example, \citet{Rice:2001} use random regression to construct non-parametric mixed effects models for unequally sampled noisy curves.
@@
 Secondly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales, (i.e., between and within curves). 
-To deal with this issue animal movement modellers have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model processes at both scales with conditionally dependent HMMs.
+To address this issue, animal movement modellers have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model both scales with conditionally dependent HMMs.
 %
@@
 %While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together. 
-In this paper, we combine existing methods from statistical ecology literature in novel ways to account for complex temporally dependent functional data.
+In this paper, we combine existing methods from statistical ecology literature in novel ways to account for complex, temporally dependent functional data.
 The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models to model multi-scale functional and time-series data with intricate dependence structure.

### 2020-11-16 Report/Sections/models.tex
address co-author edits thru 3.0

@@
 
-Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easy to calculate via the forward algorithm.
+Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easily calculated using the forward algorithm.
 %
@@
 
-\subsection{General structure for building complex models}
+\subsection{Generalized hierarchical Markov models}
 
@@
 These examples include a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models.
-For example, one may construct a three-level HHMM, where three scales of observations are recorded (fine-, medium-, and coarse-scale) and traditional HMMs are used to model all three levels.
+For example, one may construct a three-level HHMM, where three scales of observations are recorded (fine, medium, and coarse-scale) and dependent traditional HMMs are used to model all three levels.
 A myriad of possible models can be built using this framework, but these models quickly become complicated and computationally expensive to fit.

### 2020-11-16 Report/Sections/results.tex
address co-author edits thru 3.0

--- 2020-11-06 Report/Sections/results.tex
@@
 
-To illustrate the process of constructing a model using these HMM building blocks, we investigate the dive behaviour of a northern resident killer whale (NRKW) off the coast of British Columbia, Canada, and construct several candidate models to describe the associated data.
+To illustrate the process of constructing a model using these HMM building blocks, we investigate the dive behaviour of a northern resident killer whale off the coast of British Columbia, Canada, and construct several candidate models to categorize and describe its diving behaviour.
 
-Understanding animal behaviour is important for conservation efforts, as environmental changes often directly impact animal behaviour \citep{Sutherland:1998}.
+Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}.
 HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g.
 \citet{DeRuiter:2017} and \citet{Isojunno:2017}).
-For NRKWs, we are interested in categorizing different diving behaviours, with a particular focus on differentiating potential foraging dives.
-NRKWs prefer to feed on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, but Chinook occur deeper than other prey, requiring significant energy expenditure from NRKWs for predation \citep{Williams:2009,Noren:2011}.
-It is therefore of interest to determine the energetic expenditure of NRKWs foraging for Chinook as well as the frequency at which NRKWs catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
+For killer whales, we are interested in categorizing different diving behaviours to identify potential foraging dives.
+Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller prey \citep{Ford:2009}.
+Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook \citep{Williams:2009,Noren:2011,Wright:2017}.
+It is therefore of interest to determine the energetic expenditure of killer whales hunting for Chinook as well as the frequency at which they catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
 Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the behavioural state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
-Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is important to understand the true energetic requirements of a NRKW.
+Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of a killer whale.
 
@@
 
-The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions.
+The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration in three orthogonal directions.
 Observations were collected at a rate of 50 Hz using Customizable Animal Solutions Tracking (CATs) tags (model number \#\#\#).
@@
 Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, travelling, and searching for prey.
-On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as foraging and socialization \citep{Tennessen:2019b}.
+On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as hunting and socialization \citep{Tennessen:2019b}.
 However, no dive in this data set has a maximum depth deeper than 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100 meters.

### 2020-11-16 Report/main.tex
address co-author edits thru 3.0

--- 2020-11-13 Report/main.tex
@@
 %\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
-\title[]{Modelling state-switching functional data with hidden Markov models}
+\title[]{Modelling state-switching functional biologging data with hidden Markov models}
 \author{BlindedA\authorref{1}\thanksref{*}}
@@
 Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields.
-These data often exhibit complicated temporal dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA).
+These data often exhibit complicated temporal dependence structures that cannot be modelled using common methods of Functional Data Analysis (FDA).
 Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs) instead, but many intricate fine-scale processes violate basic assumptions of HMMs.
@@
 
-Sequences of curves sampled at high-frequency in time are increasingly common in practice, but they can exhibit complicated dependence structures which cannot be modelled using common methods of Functional Data Analysis (FDA).
+Data sets comprised of sequences of curves sampled at high-frequency in time are increasingly common in practice, but they can exhibit complicated dependence structures that cannot be modelled using common methods of Functional Data Analysis (FDA).
 We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM).
 The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods such as moving-window transformations, Fourier analysis, and auto-regression.
-As a case study, we use this framework to model the movement of a killer whale off the western coast of Canada.
+This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.
+As a case study, we use this framework to model the fine-scale kinematic movement of a northern resident killer whale off the western coast of Canada.
 Through simulations we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines.
-This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.
 

### 2020-11-21 Report/Sections/discussion.tex
Address comments (still need to clean figures)

@@
 
-Current FDA literature addresses dependence between curves either with multilevel models \citep{Chen:2012,Di:2009}, which lack a time component, or with functional time series, which overlook the possibility that curves have several distinct ``types" \citep{Kokoszka:2018}.
+Current Functional Data Analysis literature addresses dependence between curves either with multilevel models \citep{Chen:2012,Di:2009}, which lack a time component, or with functional time series, which overlook the possibility that curves have several distinct ``types" \citep{Kokoszka:2018}.
 Our work addresses these issues and introduces a flexible framework to model functional time series data using HMMs.
@@
 
-We demonstrate the usefulness of this framework using an ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale off the coast British Columbia, Canada.
-This gives a deeper understanding of the animal's tri-axial movement and thus its energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both northern resident killer whales and their prey \citep{Noren:2011}.
-This model is likely applicable to many other diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.
+We demonstrate the usefulness of this framework using a biomechanical/ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale in Queen Charlotte Sound off the coast British Columbia, Canada.
+This gives a deeper understanding of a killer whale's tri-axial movement and thus its behaviour and energy expenditure \citep{Gleiss:2011,Qasem:2012}, both of which are important for understanding the foraging ecology and nutritional status of the Northern Resident killer whales \citep{Noren:2011}.
+This model is applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.
 In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time series data in a variety of fields.
@@
 However, many curve sequences exhibit both temporal dependence and a discrete number of types.
-One obvious example is the case study examined here, where one marine mammal dive serves a distinct purpose such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}.
+One obvious example is the case study of killer whale behaviour examined here, where one dive can serve different purposes such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}.
 Another example is accelerometer data in school children, where bouts of activity can be associated with a behaviour such as running or walking.

### 2020-11-21 Report/Sections/results.tex
Address comments (still need to clean figures)

@@
 
-To illustrate the process of constructing a model using these HMM building blocks, we investigate the dive behaviour of a northern resident killer whale off the coast of British Columbia, Canada, and construct several candidate models to categorize and describe its diving behaviour.
+To illustrate the process of constructing a model using these building blocks, we analyze the dive behaviour of a northern resident killer whale in Queen Charlotte Sound off the coast of British Columbia, Canada, and construct several candidate models to categorize and describe its diving behaviour.
 
@@
 For killer whales, we are interested in categorizing different diving behaviours to identify potential foraging dives.
-Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller prey \citep{Ford:2009}.
+Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller types of salmon \citep{Ford:2009}.
 Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook \citep{Williams:2009,Noren:2011,Wright:2017}.
-It is therefore of interest to determine the energetic expenditure of killer whales hunting for Chinook as well as the frequency at which they catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
-Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but studies suggest that the behavioural state must be taken into account to obtain accurate estimates \citep{Dot:2016}.
+%It is therefore of interest to determine the energetic expenditure of killer whales hunting for Chinook as well as the frequency at which they catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
+Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}.
 Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of a killer whale.
@@
 
-The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration in three orthogonal directions.
-Observations were collected at a rate of 50 Hz using Customizable Animal Solutions Tracking (CATs) tags (model number \#\#\#).
+The data we use were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration in three orthogonal directions.
+Observations were collected at a rate of 50 Hz using CATs biologger (Customizable Animal Tracking Solutions, www.cats.is).
 The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left).
-Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
-Tagging the killer whale caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored.
-We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm due to a malfunction in the tagging technology.
-To process the data, we smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second.
+Tri-axial acceleration readings are common in these types of tags and are often used to infer animal behaviour such as foraging \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
+The act of attaching and detaching the tag caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations taken during these time periods are ignored.
+There were also periods of time when the tag failed to record observations, resulting in data gaps between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm.
+To process the data, we smooth the accelerometer and time-depth recorder data by taking a moving average within a window of $1/10^{th}$ of a second.
 We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
@@
 Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
-Specifically, the vector of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds ($s$).
-For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots \right\}$, which is a vector of the within-dive acceleration data in meters per second squared ($m/s^2$).
-Naturally, the collection of all acceleration data is $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.
+Specifically, the vector of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds.
+For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots \right\}$, which is a vector of the within-dive acceleration data in units of meters per second squared.
+Naturally, the collection of all acceleration data is denoted as $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.
 
@@
 
-Defining a suitable model for these data involves selecting an appropriate model structure, emission distributions, and the correct number of hidden states for both the coarse and fine scale observations.
+Defining a suitable model to describe this killer whale kinematic data set involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse and fine scale observations.
 
-We first select an appropriate model structure for the coarse-scale observations of dive duration.
-%
 We do not use information criteria to select the number of dive types $N$ since these metrics tend to overestimate the number of states in biological processes \citep{Pohle:2017}.
+We instead plot the duration of each dive versus the duration of the dive preceding it (i.e., $y_t$ vs $y_{t-1}$ for $t \geq 2$).
+This type of visualization is known as a lag plot.
 If the hidden states are well-separated, a lag plot should reveal $N$ distinct patterns, where each pattern corresponds to one dive type \citep{Lawler:2019}.
-This is unfortunately not the case for our killer whale data, but we still consult lag plots together with visual inspection to choose $N = 2$ dive types.
-The absence of a more principled method highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
+This is unfortunately not the case for our killer whale data, as there is one cluster of data centred at approximately $30$ seconds.
+However, long dives appear to be characterized by bouts of less ``wiggly" behaviour in the acceleration data compared to shorter dives, so we choose $N = 2$ to differentiate these dive types.
+The absence of a more principled method to select $N$ highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
 %
@@
 Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$.
-Although $N^*$ is selected heuristically, we test the validity of this model using techniques in Section \ref{subsec:model_validation}.
+Although $N^*$ is selected heuristically, we test the validity of this model in Section \ref{subsec:model_validation}.
 %
@@
 %
-We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation.
-Figure \ref{fig:CarHHMM-DFT} shows the corresponding graphical model.
-The likelihood of this model is easy to calculate using the forward algorithm, and it can be maximized with respect to the parameters above.
-See the appendix for details of likelihood evaluation.
+We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation (see Figure \ref{fig:CarHHMM-DFT}).
+The likelihood of this model is easily calculated using the forward algorithm and can be maximized with respect to the parameters above (see appendix for details).
 In addition to the CarHHMM-DFT, we consider three similar variations for comparison.
@@
 %
-Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes no auto-correlation within the fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.
+Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes there is no auto-correlation between fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.
 
@@
 The coarse-scale emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}).
-Dive type 1 corresponds to shorter, shallower dives which may serve a variety of purposes including resting, travelling, and searching for prey.
-On the other hand, dive type 2 is longer and deeper, and these dives have been associated with behaviours such as hunting and socialization \citep{Tennessen:2019b}.
-However, no dive in this data set has a maximum depth deeper than 30 meters, and \citet{Wright:2017} note that most prey captures occur at depths deeper than 100 meters.
-It is therefore unlikely that any of the dives used in this study are successful foraging dives.
+Dive type 1 corresponds to shorter, shallower dives which likely reflects resting, travelling, and (to a lesser extent) searching for prey.
+On the other hand, dive type 2 is longer, deeper, and may be associated with behaviours such as hunting and socialization \citep{Tennessen:2019b}.
+No dive in this data set has a maximum depth deeper than 30 meters, and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters.
+However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and significant numbers of fish have been observed to be caught near the surface (Fortune and Trites, unpublished data).
+It is therefore unclear whether any of the dives in this study are successful foraging dives.
 
-The three subdive behaviours have distinct average ``wiggliness" values, as the means corresponding to $\Ztwo_{t,\tilde t^*}$ are separated by an order of magnitude for each subdive state (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}).
+The means of the ``wiggliness" $\Ztwo_{t,\tilde t^*}$ are each separated by an order of magnitude between subdive states (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}). 
 Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest variance corresponding to $\Zone_{t,\tilde t^*}$.
@@
 This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
-The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 40\% of the time when in a dive of type 1 but only 9\% of the time when in a dive of type 2.
-Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}.
+The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 40\% of the time during a dive of type 1 but only 9\% of the time during a dive of type 2.
+Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}.
 Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives.
@@
 Each observed dive duration is weighted by the decoded probability that it corresponds to a particular dive type.
-This results in two histograms - one corresponding to dive type 1 and another corresponding to dive type 2.
-Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT.
-See Sections 1.4 through 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $\z_{t,\tilde t^*}$.
+This results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2.
+Each histogram is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT.
+Analogous histograms corresponding to the fine-scale observations $\z_{t,\tilde t^*}$ are contained in Sections 1.4 through 1.6 of the supplementary material.
 Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions.
-In particular, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale.
+In particular, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale.
 These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.
@@
 In particular, the CarHMM-DFT does not indicate that the whale is more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives.
-See Section 1 of the supplementary material for a more complete set of results for each of the candidate models.
+A more complete set of results for each of the candidate models is in Section 1 of the supplementary material.

### 2020-11-22 Report/Sections/models.tex
End of day

@@
 where $P(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_1)\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
-\Gamma^{*(i)};y_1^*\right)$ for an HMM and $\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
-\Gamma^{*(i)};y_1^*\right)$ for a CarHMM.
-The fine-scale likelihood $\calL_{\text{(Car)HMM}}$ corresponds to the desired fine-scale model.
+\Gamma^{*(i)};y_1^*\right)$ if the coarse-scale model is an HMM or $\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
+\Gamma^{*(i)};y_1^*\right)$ if the coarse-scale model is a CarHMM.
+The fine-scale likelihood $\calL_{\text{(Car)HMM}}$ corresponds to the likelihood of the fine-scale model, which can be either a CarHMM or an HMM.
 In addition, $P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to

### 2020-11-22 Code/run_all_sims.sh
update Code

@@ -10,9 +10,21 @@
 
 #SBATCH --time=96:00:00
 #SBATCH --mem-per-cpu=4G
-#SBATCH --array=1-500
+#SBATCH --array=1-2000
 
-python Run_Sim.py $SLURM_ARRAY_TASK_ID CarHMM
-python Run_Sim.py $SLURM_ARRAY_TASK_ID HHMM
-python Run_Sim.py $SLURM_ARRAY_TASK_ID CarHHMM1
-python Run_Sim.py $SLURM_ARRAY_TASK_ID CarHHMM2
+if [ $SLURM_ARRAY_TASK_ID -lt 501 ]
+then
+  python Run_Sim.py $SLURM_ARRAY_TASK_ID CarHMM
+fi
+if [ $SLURM_ARRAY_TASK_ID -gt 500 ] && [ $SLURM_ARRAY_TASK_ID -lt 1001 ]
+then
+  python Run_Sim.py $SLURM_ARRAY_TASK_ID HHMM
+fi
+if [ $SLURM_ARRAY_TASK_ID -gt 1000 ] && [ $SLURM_ARRAY_TASK_ID -lt 1501 ]
+then
+  python Run_Sim.py $SLURM_ARRAY_TASK_ID CarHHMM1
+fi
+if [ $SLURM_ARRAY_TASK_ID -gt 1500 ]
+then
+  python Run_Sim.py $SLURM_ARRAY_TASK_ID CarHHMM2
+fi

### 2020-11-22 Code/Parameters.py
Update Code

@@ -26,8 +26,8 @@ class Parameters:
     def __init__(self):
 
         # loading parameters
-        self.cvc_file = '../Data.nosync/20190902-182840-CATs_OB_1.cvc'
-        self.csv_file = '../Data.nosync/20190902-182840-CATs_OB_1_001.csv'
+        self.cvc_file = '../Data/20190902-182840-CATs_OB_1.cvc'
+        self.csv_file = '../Data/20190902-182840-CATs_OB_1_001.csv'
 
         self.cvc_cols = ['date','time',
                          'GPS lat','GPS long','GPS alt','GPS speed',

### 2020-11-23 Code/Visualisor.py
Update plot notation and simulation parameters

@@ -51,15 +51,15 @@ class Visualisor:
 
         features = list(self.pars.features[0].keys()) + list(self.pars.features[1].keys())
         xlabs = ['$Y_t$',
-                 r'$\left(A^*_{t,t^*}\right)_x$',
-                 r'$\left(A^*_{t,t^*}\right)_y$',
-                 r'$\left(A^*_{t,t^*}\right)_z$',
-                 r'$W^*_{t,t^*}$']
+                 r'$\left(\tilde A^*_{t,\tilde t^*}\right)_x$',
+                 r'$\left(\tilde A^*_{t,\tilde t^*}\right)_y$',
+                 r'$\left(\tilde A^*_{t,\tilde t^*}\right)_z$',
+                 r'$\tilde W^*_{t,\tilde t^*}$']
         ylabs = ['$Y_{t+1}$',
-                 r'$\left(A^*_{t,t^*+1}\right)_x$',
-                 r'$\left(A^*_{t,t^*+1}\right)_y$',
-                 r'$\left(A^*_{t,t^*+1}\right)_z$',
-                 r'$W^*_{t,t^*+1}$']
+                 r'$\left(\tilde A^*_{t,\tilde t^*+1}\right)_x$',
+                 r'$\left(\tilde A^*_{t,\tilde t^*+1}\right)_y$',
+                 r'$\left(\tilde A^*_{t,\tilde t^*+1}\right)_z$',
+                 r'$\tilde W^*_{t,\tilde t^*+1}$']
 
         # lag plots of dive-level data
         for i,feature in enumerate(features):
@@ -190,10 +190,10 @@ class Visualisor:
                     #ax[row_num,col_num].set_title(title,fontsize=12)
                     plt.legend(['Dive Type 1','Dive Type 2'],fontsize=10)
                 else:
-                    titles = [r'$\left(A^*_{t,t^*}\right)_x$ $(m/s^2)$',
-                              r'$\left(A^*_{t,t^*}\right)_y$ $(m/s^2)$',
-                              r'$\left(A^*_{t,t^*}\right)_z$ $(m/s^2)$',
-                              r'$W^*_{t,t^*}$']
+                    titles = [r'$\left(\tilde A^*_{t,\tilde t^*}\right)_x$ $(m/s^2)$',
+                              r'$\left(\tilde A^*_{t,\tilde t^*}\right)_y$ $(m/s^2)$',
[…]

### 2020-11-23 Report/Sections/simulation.tex
Address comments (still need to clean figures)
Updates from Overleaf

--- 2020-11-06 Report/Sections/simulation.tex
@@
 %
-Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 25s$.
+Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 50s$.
 After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each dive are ignored.
@@
 The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model.
-The average dive decoding accuracy is above 0.9 while the average subdive decoding accuracy is $1-10^{-12}$.
+The average dive decoding accuracy is greater than 0.9 while the average subdive decoding accuracy is $1-10^{-12}$.
 All parameter estimates of mean values and fine-scale probability transition matrices, $\hat \mu$ and $\hat \Gamma^*$, respectively, have statistically insignificant biases.

### 2020-11-23 Report/Sections/simulation.tex
Update Simulation Study

--- 2020-11-06 Report/Sections/simulation.tex
@@
 We perform a simulation study based on data generated from the full CarHHMM-DFT as defined in Section \ref{subsec:model_selection} to evaluate each candidate model when the ground-truth is known.
-The parameters used to generate the data are loosely based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity.
-In particular, we set $\Gamma$ such that the simulated data have equal numbers of short and long dives in expectation, the number of subdive states is set to $N^*=2$, and $\Zone_{t,\tilde t^*}$ is set to a scalar rather than a three dimensional vector.
+The parameters used to generate the data are based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity.
+In particular, we set the number of subdive states to $N^*=2$ and $\Zone_{t,\tilde t^*}$ as a scalar rather than a three dimensional vector.
 We then fit the data with four models: the full CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT.
@@
 %
-$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
+$$\Gamma = \begin{pmatrix} 0.79 & 0.21 \\ 0.81 & 0.19 \end{pmatrix}.$$
 %
-Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 25s$.
+Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 25.7s$, $\sigma^{(1)} = 9.6s$, $\mu^{(2)} = 104.6s$, and $\sigma^{(2)} = 64.7s$.
 After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each dive are ignored.
-Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
+Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots \right\}$ with probability transition matrices
 %
-$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}$$ 
+$$\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}$$ 
 %
@@
 \begin{gather*}
-    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
+    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.034s, \enspace \phi_A^{*(\cdot,1)} = 0.98, \\
     %
-    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
+    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.079s, \enspace \phi_A^{*(\cdot,2)} = 0.87, \\
     %
-    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
+    \mu_W^{*(\cdot,1)} = 23.3, \quad \sigma_W^{*(\cdot,1)} = 13.0, \\
     %
-    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
+    \mu_W^{*(\cdot,2)} = 301.2, \quad \sigma_W^{*(\cdot,2)} = 330.1.
 \end{gather*}
@@
 
-The two simulated dive types differ in that dives of type one are much shorter on average (20 seconds) than dives of type two (80 seconds).
+The two simulated dive types differ in that dives of type one are much shorter on average (26 seconds) than dives of type two (105 seconds).
 The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher for subdive state 2 than for subdive state 1.
@@
 We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}).
-If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = \tilde x_{t}|Y,Y^*)$ across all simulated data sets and dives.
-Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows.
+If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives.
+Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows.
 We find both the average dive decoding accuracy and the average subdive decoding accuracy using the forward-backward algorithm.
@@
 The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model.
-The average dive decoding accuracy is above 0.9 while the average subdive decoding accuracy is $1-10^{-12}$.
-All parameter estimates of mean values and fine-scale probability transition matrices, $\hat \mu$ and $\hat \Gamma^*$, respectively, have statistically insignificant biases.
-In addition, biases for $\hat \sigma$, $\hat \phi$, and $\hat \Gamma$ are as severe or less severe than the other three models.
-The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to slightly underestimate the empirical standard errors.
-See Tables 6 through 9 of the supplementary material for detailed results.
+The average dive decoding accuracy and the average subdive decoding accuracy are both greater than 0.9.
+All parameter estimates of fine-scale mean values and probability transition matrices, $\hat \mu^*$, $\hat \Gamma^*$ and $\hat \Gamma$, respectively, have statistically insignificant biases.
+In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models.
+The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to underestimate the empirical standard errors.
+This is especially true for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error could be up to twice as large as the estimated standard error.
+See Tables 5 through 9 of the supplementary material for detailed results.
 
-The HHMM-DFT performs almost identically to the CarHHMM-DFT in most respects, including its average dive and subdive decoding accuracy as well as its parameter estimate biases.
-However, unlike the CarHHMM-DFT, the HHMM-DFT greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$.
+The HHMM-DFT performs similarly to the CarHHMM-DFT in most respects.
+Its average dive decoding accuracy is comparable to the CarHHMM-DFT while its average subdive decoding accuracy is approximately $5 \%$ worse.
+Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$.
 In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material).
@@
 
-The CarHHMM is the worst-performing model in terms of accuracy, as its average dive decoding accuracy and average subdive decoding accuracy are both below $0.9$.
+The CarHHMM is the worst-performing model in terms of accuracy, as its average dive decoding accuracy is below $0.9$ and its average subdive decoding accuracy is below $0.8$.
 This result is consistent with expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct difference between the subdive states.
-The CarHHMM is much more likely to misclassify subdive state 1 as subdive state 2 than vice-versa (see Table \ref{table:accuracy}).
-In addition to its poor average subdive and dive decoding accuracy, the CarHHMM is also the worst of the four candidate models at estimating parameters.
-Estimates associated with dive type 2 ($\theta^{(2)}$ and $\hat \Gamma^{*(2)}$) and subdive type 2 ($\theta^{*(\cdot,2)}$) are especially poor, likely due to the fact that the CarHHMM often classifies dive type 1 as dive type 2 and subdive state 1 as subdive state 2.
+In addition to its poor average decoding accuracy, the CarHHMM is also the worst of the four candidate models at estimating parameters.
+Parameter estimates associated with subdive type 2 ($\theta^{*(\cdot,2)}$) are especially poor.
 See Section 2 of the supplementary material for more detailed results.
 
-Finally, the CarHMM-DFT is comparable to the CarHHMM-DFT in terms of average subdive decoding accuracy, parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates.
+Finally, the CarHMM-DFT is nearly identical to the CarHHMM-DFT in terms of average subdive decoding accuracy, fine-scale parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates.
 In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}).
-However, this model does not estimate dive type as it lacks any hierarchical structure.
+However, this model cannot estimate dive type as it lacks any hierarchical structure.
 The CarHMM-DFT nonetheless fits a single Gamma distribution over the dive duration of all dives.
-The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated, and on average the value of $\hat \sigma$ is approximately 5 seconds less than the average sample standard deviation of dive duration, indicating significant model misspecification.
-See Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} for a visualization of the simulated data set as well as the decoded dive types and subdive states for each model.
+The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 24 of the supplementary material).
+
+Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} display one simulated data set as well as the decoded dive types and subdive states associated with each model.
+The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state while the HHMM-DFT is slightly more likely to predict that a given window corresponds to subdive state 2 (higher activity).
+The CarHHMM is the least accurate model when predicting subdive state.
+The CarHHMM-DFT and HHMM-DFT output similar estimates of dive type while the CarHHMM is less accurate and even misclassifies the second dive.
+The CarHMM-DFT does not estimate dive type.

### 2020-11-24 Code/Visualisor.py
Update results plots

@@ -311,7 +311,6 @@ class Visualisor:
 
         #title = 'Hidden State Probability Estimates'
         title = 'Decoded Dive Profile / Accelerometer Data'
-        print(df_cols)
         for i,col in enumerate(df_cols):
 
             # dive-level coloring
@@ -339,6 +338,12 @@ class Visualisor:
                     pass
                 if col == 'depth':
                     plt.gca().invert_yaxis()
+
+                # plot vlines
+                for dive_num in range(sdive+1,edive+1):
+                    vline = min(dive[dive['dive_num'] == dive_num]['sec_from_start']/60)
+                    plt.axvline(vline,color='k',linewidth=2)
+
                 fignum += 1#2
 
             # subdive-level coloring
@@ -368,6 +373,12 @@ class Visualisor:
                 #plt.title(col + ', dives %d-%d'%(sdive,edive),fontsize=24)
                 if col == 'depth':
                     plt.gca().invert_yaxis()
+
+                # plot vlines
+                for dive_num in range(sdive+1,edive+1):
+                    vline = min(dive[dive['dive_num'] == dive_num]['sec_from_start']/60)
+                    plt.axvline(vline,color='k',linewidth=2)
+
                 fignum += 1#-1
 
         t_start = dive['time'].min()
@@ -397,7 +408,7 @@ class Visualisor:
                         time.append(avg_time)
                         features[ML_state].append(dive[col])
                         feature.append(dive[col])
-                    plt.axvline(max(time))
+                    plt.axvline(max(time),color='k',linewidth=2)
                 for state in range(self.pars.K[0]):
                     plt.plot(times[state],features[state],
                              '.',color=colors[state],markersize=4)
@@ -440,7 +451,7 @@ class Visualisor:
                 plt.xlabel('Time (s)',fontsize=10)
                 plt.legend(legend,prop={'size': 10})
                 for vline in vlines:
-                    plt.axvline(vline)
+                    plt.axvline(vline,color='k',linewidth=2)
                 fignum += 1
 
 
@@ -474,7 +485,7 @@ class Visualisor:
                 plt.legend(legend,prop={'size': 10})
                 plt.title('Depth Data',fontsize=10)
                 for vline in vlines:
-                    plt.axvline(vline)
+                    plt.axvline(vline,color='k',linewidth=2)
                 fignum += 1
 
         legend_elements = [Line2D([0], [0], marker='o', color='w', label='Subdive 1',

### 2020-11-24 Report/Sections/appendix.tex
update models in Report

--- 2020-11-06 Report/Sections/appendix.tex
@@
 It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
-Since accelerometer readings are taken at 50 hertz and this model has two-second window lengths, each window contains 100 accelerometer readings.
-In addition, the length of each fine-scale Markov chain is $\tilde T ^*_t = \lfloor Y_t/2 \rfloor$.
-Then, given the dive type $X_t = i$, dive duration $Y_t = y_t$, and the fine-scale probability transition matrix $\Gamma^{*(i)}$, it is straightforward to simulate the subdive states $\tilde{X}^*_t$ for a particular dive $t$.
+Then, given $X_t = i$, $Y_t = y_t$, and $\Gamma^{*(i)}$, it is straightforward to simulate the subdive states $\tilde{X}^*_t$ for a particular dive $t$.
+Recall that the length of each fine-scale Markov chain is $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ since each window is two seconds long.
 It is much more difficult to simulate the raw acceleration data $Y_t^*$, and we focus on this task for the remainder of the section.
 
+Recall that each window is two seconds long and therefore contains $h = 100$ observations taken at 50 observations per second.
 Let $DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,\tilde t^*}$ and $DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$ be denoted as $\hat{Y}^*_{t,\tilde t^*}$ for convenience.
-We simulate $Y_t^*$ in three steps: 1) simulate the average acceleration within each window $\left(Y^{*(0)}_{t,\tilde t^*}\right)$, 2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and 3) take the inverse $DFT$ of $\hat{Y}^*_{t,\tilde t^*}$ to retrieve $\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$.
+We simulate $Y_t^*$ in three steps: (1) simulate the average acceleration within each window $\left(\hat Y^{*(0)}_{t,\tilde t^*}\right)$, (2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and (3) take the inverse $DFT$ of $\hat{Y}^*_{t,\tilde t^*}$ to retrieve $\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$.
 The details of this procedure are described below:
@@
     $$
-    	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
+    	(\hat{Y}^{*(0)}_{t,1}|\tilde X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
     $$
-    where $\sigma_A^{*(\cdot,1)} = 0.05s$ and $\sigma_A^{*(\cdot,2)} = 0.1s$. 
+    where $\sigma_A^{*(\cdot,1)} = 0.034s$ and $\sigma_A^{*(\cdot,2)} = 0.079s$. 
     %
@@
     \end{align}
-    where $\phi_A^{*(\cdot,1)} = 0.99$ and $\phi_A^{*(\cdot,2)} = 0.95$. 
+    where $\phi_A^{*(\cdot,1)} = 0.98$ and $\phi_A^{*(\cdot,2)} = 0.87$. 
     \end{enumerate}
@@
     \begin{split}
-    	(b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1) &\sim {\rm{Gamma}}(5/k^2, 1) \\
+    	(b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1) &\sim {\rm{Gamma}}(16.38/k^3, 36.23) \\
     	%
-    	(b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2) &\sim \left\{\begin{array}{lr}
-    	{\rm{Gamma}}(250,1), & k = 1 \\
-    	{\rm{Gamma}}(50,1), & k = 2 \\
-    	{\rm{Gamma}}(5/k^2, 1), \quad & k \notin \{1,2\}.
-    	\end{array}\right. 
+    	(b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2) &\sim {\rm{Gamma}}(4.20/k^3, 1825.53). \\ 
     \end{split}
@@
     \end{align}
-    The first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the scale parameter and the second is the shape parameter.
-The magnitude of the $k^{th}$ Fourier coefficient is equal to $b^{(k)}_{t,\tilde t^*}$, which decays like $1/k^2$ to ``smooth out" the raw acceleration data.
+    The first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the shape parameter and the second is the scale parameter.
+The squared magnitude of the $k^{th}$ Fourier coefficient is equal to $b^{(k)}_{t,\tilde t^*}$, which decays like $1/k^3$ to ``smooth out" the raw acceleration data.
     %
@@
 This distribution is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(\cdot,1)} = \mu_A^{*(\cdot,2)} = \mu_A^{*(\cdot,i^*)} = 0$.
-We also have that $\sigma_A^{*(\cdot,1)} = 0.05s$, $\phi_A^{*(\cdot,1)} = 0.99$, $\sigma_A^{*(\cdot,2)} = 0.1s$, and $\phi_A^{*(\cdot,2)} = 0.95$.
+We also have that $\sigma_A^{*(\cdot,1)} = 0.034s$, $\phi_A^{*(\cdot,1)} = 0.98$, $\sigma_A^{*(\cdot,2)} = 0.079s$, and $\phi_A^{*(\cdot,2)} = 0.87$.
 
@@
 %
-$$\left(\Ztwo_{t,\tilde t^*}|X^*_{t,\tilde t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 5/k^2 = 10.10, 1.00 \right) \text { and }$$
+$$\left(\Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 16.38/k^3 = 3.25 , 36.23 \right) \text { and }$$
 %
-$$\left(\Ztwo_{t,\tilde t^*}|X^*_{t,\tilde t^*} = 2\right) \sim {\rm{Gamma}}\left(300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , 1.00 \right),$$
+$$\left(\Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 4.20/k^3 = 0.83 , 1825.53 \right),$$
 %
-where the first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the scale parameter and the second is the shape parameter.
-Simple calculation of the mean and variance of a Gamma distribution reveals $\mu_W^{*(\cdot,1)} = 10.10$, $\sigma_W^{*(\cdot,1)} = 3.18$, $\mu_W^{*(\cdot,2)} = 305.94$, and $\sigma_W^{*(\cdot,2)} = 17.49$.
+where the first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the shape parameter and the second is the scale parameter.
+Simple calculation of the mean and variance of a Gamma distribution reveals $\mu_W^{*(\cdot,1)} = 23.3$, $\sigma_W^{*(\cdot,1)} = 13.0$, $\mu_W^{*(\cdot,2)} = 301.2$, and $\sigma_W^{*(\cdot,2)} = 330.1$.
 
@@
 
-\subsection{Likelihood of simulation study model}
+\subsection{Likelihood of CarHHMM-DFT}
 
@@
 %
-$$\calL_{\text{CarHMM-DFT}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_N,$$
+$$\calL_{\text{CarHMM-DFT}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},$$
 %

### 2020-11-24 Report/Sections/intro.tex
update models in Report

@@
 The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}.
-In particular, the study of animal movement has been transformed by short-term, high-resolution suction-cup attached tags that record kinematic information about animal movement underwater \citep{Borger:2020,Dot:2016b}.
+The study of animal movement in particular has been transformed by tracking devices that record kinematic information in a variety of environments \citep{Borger:2020,Dot:2016b}.
 Tags can record observations at over 50 times per second, resulting in time series containing millions of observations over the course of several hours.
-These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a new challenge for statisticians.
+These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a challenge for statisticians.
 
@@
 For example, \citet{Morris:2007} view a child's activity level throughout the day as a curve where metabolic activity is a function of time.
-Similarly, \citet{Fu:2017} view the dive profile of elephant seal (\textit{Mirounga leonina}) as a sequence of dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
+Similarly, \citet{Fu:2017} view the dive profile of an elephant seal (\textit{Mirounga leonina}) as a sequence of dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
 %, although they ignore sequential dependence between dives. 
@@
 %
-FDA was originally developed to process curves assumed to be independent replicates, but time series data of this type are often complicated by dependencies which violate this assumption.
-In particular, biologging data often display significant temporal dependence both between curves (coarse-scale) and within curves (fine-scale) \citep{Barajas:2017}.
+FDA was originally developed to process curves assumed to be independent replicates, but time series data are often complicated by sequential dependencies.
+In particular, biologging data often display significant temporal dependence both between curves on a coarse scale and within curves on fine scale \citep{Barajas:2017}.
 %, as is the case for our case study data.
 Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time.
-For example, marine mammal dive profiles show distinct dive ``types" that appear to cluster together in time  \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}.
-Fine-scale periodicity within a larger process in particular is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
+For example, marine mammal dive profiles show distinct dive ``types" that cluster together in time  \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}.
+Fine-scale periodicity within a larger process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
 
@@
 
-Some FDA models account for between-curve dependence but are inadequate when modelling certain types of time dependencies. %, including that of Figure \ref{fig:data}.
+Some FDA models account for between-curve dependence, but they are inadequate when modelling certain types of time dependencies. %, including that of Figure \ref{fig:data}.
 For example, 
 %multilevel models incorporate dependence via random effects. 
-Previous studies have used multilevel models with random effects to capture variation between activity levels in children \citep{Morris:2007} and study differences in menstrual cycles \citet{Bromback:1998}.
-More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012} and \citet{Crainiceanu:2009}.
+previous studies have used multilevel models with random effects to capture variation between activity levels in children \citep{Morris:2007} and study differences in menstrual cycles \citep{Bromback:1998}.
+More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012}, and \citet{Crainiceanu:2009}.
 However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types.
@@
 However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types.
-In addition to multilevel models, FDA researchers have also used functional time series to model dependence in sequences of curves.
+FDA researchers have also used functional time series to model dependence in sequences of curves.
 Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}.
 %While functional time series methods can in theory be applied to biologging data such as ours,
-However, functional time series do not account for sequences of time series curves that switch between a discrete number of well defined states.
+However, functional time series do not account for sequences of time series curves that switch between a discrete number of well-defined states.
 %
@@
 animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
-HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from observations of its position.
-While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling, with a few notable exceptions.
+HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position.
+While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions.
 In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines,
@@
 Several solutions have been proposed in the ecology literature such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
-The CarHMM in particular explicitly models auto-correlation into an HMM while largely maintaining its base structure.
+%The CarHMM in particular explicitly models auto-correlation into an HMM while largely maintaining its base structure. 
 %
-Secondly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales, (i.e., between and within curves). 
-To address this issue, animal movement modellers have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model both scales with conditionally dependent HMMs.
+Secondly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales (i.e., both between and within curves). 
+To address this issue, statistical ecologists have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model both scales with conditionally dependent HMMs.
 %
-Finally, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures at short time scales.
+Thirdly, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures at short time scales.
 For example, \citet{Adam:2019} fail to capture fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}) using a traditional HHMM.
 \citet{Heerah:2017} use Fourier analysis within an HMM to account for %periodic behaviour in the context of describing
-daily behavioural cycles in marine mammals.
-In addition, Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
+daily behavioural cycles in marine mammals, and Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
 Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for fine-scale periodic structures.
@@
 In this paper, we combine existing methods from statistical ecology literature in novel ways to account for complex, temporally dependent functional data.
-The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models to model multi-scale functional and time-series data with intricate dependence structure.
+The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.
 %and thus extend the methods available for the analysis of functional data. 
-We begin by describing HMMs as well as to two variants (CarHMMs and HHMMs) and discuss how Fourier analysis can handle fine-scale dependence structures.
+We begin by describing HMMs as well as to two variants, CarHMMs and HHMMs, and discuss how Fourier analysis can handle fine-scale dependence structures.
 We also show how these methods can be combined to analyze increasingly complex data.
-In Section 3 we present a case study in which we fit several candidate models to data from a killer whale from the threatened northern resident population off the coast of British Columbia, Canada.
+In Section 3 we fit several candidate models to data from a killer whale from the threatened northern resident population off the coast of British Columbia, Canada.
 Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.

### 2020-11-24 Report/Sections/models.tex
update models in Report

@@
 
-Consider a sequence of $T$ curves (or dives, in our context). For any particular $t = 1,
-\ldots,T$, dive $t$ is characterized by a possibly high-dimensional curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$.
-The sequence $Y^*_{t}$ is made up of function values from curve $t$.
+Consider a sequence of $T$ curves, where any particular curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$.
+Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of function values from curve $t$.
 We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.
+We first detail the structure of a traditional HMM followed by three variations which generalize its base structure.
+We then show how each of these generalized HMMs can be synthesized to form a wide variety of more complicated models.
 
-Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$.
+%Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$.
 We first detail the structure of a traditional HMM and a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
@@
 %which is comprised of two distinct,x but related, HMMs that jointly model these time series. 
-Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics.
+%Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics.
 In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour.
@@
 In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT.
+Figure \ref{fig:models}d displays the dependence structure for an HMM-DFT.
 

### 2020-11-24 Report/Sections/results.tex
update models in Report

@@
 %Given that $X^*_{t,\tilde t^*} = i^*$, the corresponding unknown parameters are then $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}_{>0}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
-Given that $\tilde X^*_{t,\tilde t^*} = i^*$, $\tilde \Ztwo_{t,\tilde t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
-We do not include $\tilde \Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
+Given that $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
+We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
 %

### 2020-11-24 Report/main.tex
Address comments (still need to clean figures)
update models in Report

@@
 %\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
-\title[]{Modelling state-switching functional biologging data with hidden Markov models}
+\title[]{Modelling state-switching functional data with hidden Markov models}
 \author{BlindedA\authorref{1}\thanksref{*}}
@@
 \KWDtitle{Key words and phrases}
-Time series modelling\sep multi-state modelling\sep hierarchical modelling\sep biologging\sep statistical ecology\sep animal movement modelling.
+Multi-state modelling\sep hierarchical modelling\sep time series\sep  biologging\sep statistical ecology\sep animal movement.
 % MSC 2010 subject classification codes
@@
 
-Data sets comprised of sequences of curves sampled at high-frequency in time are increasingly common in practice, but they can exhibit complicated dependence structures that cannot be modelled using common methods of Functional Data Analysis (FDA).
+Data sets comprised of sequences of curves sampled at high frequencies in time are increasingly common in practice, but they can exhibit complicated dependence structures that cannot be modelled using common methods of Functional Data Analysis (FDA).
 We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM).
-The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods such as moving-window transformations, Fourier analysis, and auto-regression.
+The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods including moving-window transformations, Fourier analysis, and auto-regression.
 This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.
@@
 %
-%\newpage
-%\input Sections/bibliography.tex
+\begin{ack}{ACKNOWLEDGEMENTS}
+This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
+\end{ack}
 %

### 2020-11-24 Report/Sections/models.tex
read over through 2.2

@@
 
-An HMM is comprised of a sequence of unobserved states $X \equiv \big\{X_1, \ldots, X_T\big\}$ where the hidden state $X_t$ is associated with the observation $Y_t$.
-Here we focus on the coarse scale sequence of observations, but an HMM can also be used as a sub-model of the fine-scale sequence for a particular dive.
+Hidden Markov models describe state-switching Markovian processes in discrete time and are the core structure we use to model both $Y$ and $Y^*$.
+For simplicity we focus on $Y$ to introduce the model.
+An HMM is comprised of a sequence of unobserved states $X \equiv \big\{X_1, \ldots, X_T\big\}$ together with an observation sequence $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$, where $X_t$ is associated with the observation $Y_t$.
 The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time.
-The $X_t$'s form a Markov chain and take possible values $1, \ldots, N$.
+The $X_t$'s form a Markov chain and can take integer values between $1$ and $N$.
 Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$.
@@
 %
-The distribution of an emission $Y_t$ depends only on the corresponding state $X_t$ and no other observations or hidden states: the conditional probability density function has the form $p\left(y_t|\{X_1,\ldots, X_T\},\{Y_1,\ldots, Y_T\}/ \{Y_t\}\right) = p(y_t|X_t)$.
+The distribution of an emission $Y_t$ conditioned on the corresponding hidden state $X_t$ does not depend upon any other observation or hidden state.
 %
-These conditional distributions are governed by state-dependent parameters.
-Namely, if $X_t = i$ then the state-dependent parameter is $\theta^{(i)}$ and we denote the conditional distribution of $Y_t$ given $X_t=i$ by its conditional density or probability mass function, denoted $f^{(i)}(\cdot ; \theta^{(i)})$, or sometimes simply $f^{(i)}(\cdot)$.
+If $X_t=i$ then we denote the conditional density of $Y_t$ as $f^{(i)}(\cdot ; \theta^{(i)})$ or simply $f^{(i)}(\cdot)$, where $\theta^{(i)}$ is a state-dependent parameter describing the emission distribution.
 %
@@
 
-Using observed emissions, here denoted $y = \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$.
-We write the likelihood $\calL_{\text{HMM}}$ using the  well-known \textit{forward algorithm} \citep{Zucchini:2016}:
+Using observation emissions, denoted here as $y \equiv \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$.
+The likelihood $\calL_{\text{HMM}}$ can be evaluated using well-known \textit{forward algorithm} \citep{Zucchini:2016}:
 %
@@
 This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$.
-As a concrete example, if $Y_t$ is a $K$-dimensional vector with components $Y_t(1),\ldots,Y_t(K)$, consider the component-wise sequences $Y_1(k), \ldots, Y_T(k)$ for $k = 1,\ldots K$.
-One may assume that these sequences are independent given $X_1,\ldots,X_T$ and that $Y_t(k)$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)}_k = \{\mu^{(i)}_k,\sigma^{(i)}_k,\phi^{(i)}_k\}$, where:
+As a concrete example, if $Y_t$ is a $K$-dimensional vector with components $Y_t(1),\ldots,Y_t(K)$, consider the component-wise sequences $\{Y_1(k), \ldots, Y_T(k)\}$ for $k = 1,\ldots K$.
+One may assume that these sequences are independent given $\{X_1,\ldots,X_T\}$ and that $Y_t(k)$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)}_k = \{\mu^{(i)}_k,\sigma^{(i)}_k,\phi^{(i)}_k\}$, where:
 %
@@
 This follows in the same way that an AR(1) process is the discrete-time version of a traditional Ornstein-Uhlenbeck process.
-Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a general CarHMM.
+Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.
 As previously, the likelihood corresponding to a general CarHMM can be easily calculated using the forward algorithm.

### 2020-11-24 Report/Sections/models.tex
read over to section 3.0

@@
 %
-If $X_t=i$ then we denote the conditional density of $Y_t$ as $f^{(i)}(\cdot ; \theta^{(i)})$ or simply $f^{(i)}(\cdot)$, where $\theta^{(i)}$ is a state-dependent parameter describing the emission distribution.
+If $X_t=i$ then we denote the conditional density or probability mass function of $Y_t$ as $f^{(i)}(\cdot ; \theta^{(i)})$ or simply $f^{(i)}(\cdot)$, where $\theta^{(i)}$ is a state-dependent parameter describing the emission distribution.
 %
-Figure \ref{fig:models}a represents the dependence structure of an HMM.
 
@@
 For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$.
+Figure \ref{fig:models}a represents the dependence structure of an HMM.
 
@@
 The first emission $Y_1$ is treated as fixed as an initial value which does not depend upon $X_1$.
-We denote the conditional density of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$.
-This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$.
+We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$.
+This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density or probability mass function that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$.
 As a concrete example, if $Y_t$ is a $K$-dimensional vector with components $Y_t(1),\ldots,Y_t(K)$, consider the component-wise sequences $\{Y_1(k), \ldots, Y_T(k)\}$ for $k = 1,\ldots K$.
@@
 This follows in the same way that an AR(1) process is the discrete-time version of a traditional Ornstein-Uhlenbeck process.
-Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.
-As previously, the likelihood corresponding to a general CarHMM can be easily calculated using the forward algorithm.
-If $y$ is a vector of observed emissions, then
+
+As previously, the likelihood corresponding to a general CarHMM can be easily calculated using the forward algorithm. If $y$ is the sequence of observed emissions, then
 \begin{equation*}
@@
 $P(y_t|y_{t-1};\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
+Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.
 
@@
 
-A hierarchical hidden Markov model, or HHMM, accounts for different scales of correlation by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs.
-The coarse-scale model is as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations with state-dependent parameters $\theta^{(i)}, i = 1,\ldots,N$.
+A hierarchical hidden Markov model, or HHMM, accounts for processes occurring simultaneously at different scales by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs.
+The coarse-scale model is as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations with state-dependent parameters $\theta^{(i)}$ for $i = 1,\ldots,N$.
 %
@@
 For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
-If $X_t = i$, then the components of $X_t^*$ are a Markov chain with an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
+If $X_t = i$, then the components of $X_t^*$ are a Markov chain with $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
 The distribution of $Y_{t,t^*}$ given $Y_{t,t^*-1} = y_{t,t^*-1}$, $X_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$.
-We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^{*(i)})}\right)$.
+We denote the set of fine-scale emission parameters corresponding to $X_t=i$ as $\theta^{*(i)}=\left\{\theta^{*(i,1)}, \ldots, \theta^{*(i,N^{*(i)})}\right\}$.
 In summary:
@@
 \begin{gather*}
-    \{Y, X\} \text{ follows (Car)HMM with } \Gamma \in \bbR^{N \times N}, \\
+    \{Y, X\} \text{ follows a (Car)HMM with } \Gamma \in \bbR^{N \times N}, \\
     %
@@
     %
-    \{Y^*_t,X^*_t | X_t = i\} \text{ follows (Car)HMM with } \Gamma^{*(i)} \in \bbR^{N^{*(i)} \times N^{*(i)}}, \\
+    \{Y^*_t,X^*_t | X_t = i\} \text{ follows a (Car)HMM with } \Gamma^{*(i)} \in \bbR^{N^{*(i)} \times N^{*(i)}}, \\
     %
@@
 Given the coarse-scale hidden states $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$ and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
-Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.
 %
@@
 We also share the fine-scale emission parameters across coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
-Coarse-scale hidden states therefore differ only in coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$.
+Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$.
 
@@
 %
-Let $y$ be the $T$-vector of the observed coarse-scale emissions and
+Let $y$ be the sequence of observed coarse-scale emissions and
 $y^* \equiv \left\{y^*_1, \ldots,y^*_T\right\}$ be the collection of $T$ observed fine-scale emission vectors.
@@
 The fine-scale likelihood $\calL_{\text{(Car)HMM}}$ corresponds to the likelihood of the fine-scale model, which can be either a CarHMM or an HMM.
-In addition, $P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to
-$f^{(i)}(y_t|y_{t-1})\calL_{\text{HMM}}\left(\theta^{*(i)},
-\Gamma^{*(i)};y_t^*\right)$ for both a CarHMM and an HMM.
+For $t \geq 2$, $P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to
+$f^{(i)}(y_t|y_{t-1})\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
+\Gamma^{*(i)};y_t^*\right)$. Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.
 %
-For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citet{Adam:2019}.
+%For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citet{Adam:2019}.
 
@@
 
-In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far.
+In many applications where data are collected at high frequencies, intricate dependency structures often arise within the fine-scale process which cannot be adequately modelled with any HMM-based model described thus far.
 %For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. 
@@
 Subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window.
-Larger stride lengths result in a loss of information but also reduce the dimension of the fine-scale process.
+Larger stride lengths result in a loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting.
 In addition, setting stride length and window size equal to one another avoids artificial residual correlation arising from overlapping windows.
@@
 We omit the final window if $t^*+h-1$ exceeds $T^*_t$, and if $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise.
-Next, we index each window with $\tilde{t}^*$ and calculate transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$:
+We index each window with $\tilde{t}^*$ and calculate transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$:
 %
@@
 where $\tilde{\omega} \leq h-1$ is a problem-specific tuning parameter corresponding to the maximum recorded frequency within each window.
-In words, $\Zone_{t,\tilde{t}^*}$ is the average acceleration vector during window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window.
+In words, $\Zone_{t,\tilde{t}^*}$ is the average acceleration vector within window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window.
 %(by the Plancherel theorem).
@@
 %By definition, an 
-Traditional HHMMs treat both the coarse-scale and the fine-scale processes as realizations of an HMM or CarHMM. 
+Traditional HHMMs treat both the coarse-scale and the fine-scale processes of interest as realizations of an HMM or CarHMM. 
 %However, it is straightforward to generalize the coarse-scale HMM by replacing it with a CarHMM, yielding the following likelihood:
@@
 However, the fine-scale observations for a particular dive $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods.
-In this case, the fine-scale HMM likelihood term $\calL_{\text{(Car)HMM}}$ in Equation (\ref{eqn:HHMM_likelihood}) is replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
+In other words, the fine-scale HMM likelihood term $\calL_{\text{(Car)HMM}}$ in Equation (\ref{eqn:HHMM_likelihood}) can be replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
 \[
@@
 %
-Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT) in addition to many others not described here.
-For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model.
+Possible candidates for the fine-scale model include any of the models described in the previous subsections in addition to many others not described here.
+For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations on a Poisson process as a fine-scale model.
 If the fine-scale model is a simple Poisson process, then the hierarchical structure described here reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
@@
 A continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}.
-\citet{Xu:2018} explain biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours.
-However, they take a longitudinal approach whereas we use a coarse-scale HMM in our case study.
+\citet{Xu:2018} explain biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours, but they take a longitudinal approach which ignores temporal dependence on the coarse scale.
 
 These examples include a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models. 
-For example, one may construct a three-level HHMM, where three scales of observations are recorded (fine, medium, and coarse-scale) and dependent traditional HMMs are used to model all three levels.
-A myriad of possible models can be built using this framework, but these models quickly become complicated and computationally expensive to fit.
+%For example, one may construct a three-level HHMM, where three scales of observations are recorded (fine, medium, and coarse-scale) and dependent traditional HMMs are used to model all three levels. 
+A myriad of possible models can be built using this framework, but these models can quickly become complicated and computationally expensive to fit.
 Therefore, models should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.

### 2020-11-24 Report/Sections/results.tex
Read through section 3.2

@@
 %It is therefore of interest to determine the energetic expenditure of killer whales hunting for Chinook as well as the frequency at which they catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
-Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}.
+Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}.
 Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of a killer whale.
@@
 
-The data we use were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration in three orthogonal directions.
-Observations were collected at a rate of 50 Hz using CATs biologger (Customizable Animal Tracking Solutions, www.cats.is).
-The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left).
+The data we use were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration over time.
+Observations were collected at a rate of 50 Hz using a CATs biologger (Customizable Animal Tracking Solutions, www.cats.is).
+The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, and right/left).
 Tri-axial acceleration readings are common in these types of tags and are often used to infer animal behaviour such as foraging \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
@@
 There were also periods of time when the tag failed to record observations, resulting in data gaps between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm.
-To process the data, we smooth the accelerometer and time-depth recorder data by taking a moving average within a window of $1/10^{th}$ of a second.
+To preprocess the data, we smooth the accelerometer and time-depth recorder data by taking a moving average within a window of $1/10^{th}$ of a second.
 We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
@@
 Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
-Specifically, the vector of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds.
-For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots \right\}$, which is a vector of the within-dive acceleration data in units of meters per second squared.
-Naturally, the collection of all acceleration data is denoted as $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.
+Specifically, the sequence of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds.
+For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots \right\}$, which is a sequence containing the within-dive acceleration data in units of meters per second squared.
+The collection of all acceleration data is denoted as $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.
 
@@
 
-Defining a suitable model to describe this killer whale kinematic data set involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse and fine scale observations.
+Defining a suitable model to describe this killer whale kinematic data involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse and fine scale observations.
 
-We do not use information criteria to select the number of dive types $N$ since these metrics tend to overestimate the number of states in biological processes \citep{Pohle:2017}.
-We instead plot the duration of each dive versus the duration of the dive preceding it (i.e., $y_t$ vs $y_{t-1}$ for $t \geq 2$).
+We do not use information criteria to select the number of dive types $N$ since these metrics tend to overestimate the number of behavioural states in biological processes \citep{Pohle:2017}.
+We instead plot the duration of each dive versus the duration of the dive preceding it ($y_t$ vs $y_{t-1}$ for $t \geq 2$).
 This type of visualization is known as a lag plot.
 If the hidden states are well-separated, a lag plot should reveal $N$ distinct patterns, where each pattern corresponds to one dive type \citep{Lawler:2019}.
-This is unfortunately not the case for our killer whale data, as there is one cluster of data centred at approximately $30$ seconds.
-However, long dives appear to be characterized by bouts of less ``wiggly" behaviour in the acceleration data compared to shorter dives, so we choose $N = 2$ to differentiate these dive types.
+This is unfortunately not the case for our killer whale data, as there is one cluster of data centred at approximately $y_t = y_{t-1} = 30$ seconds.
+However, longer dives appear to be characterized by bouts of less ``wiggly" behaviour in the acceleration data compared to shorter dives, so we choose $N = 2$ to differentiate these dive types.
 The absence of a more principled method to select $N$ highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
 %
-Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Section 1.1 of the supplementary material), and visual inspection shows no obvious intricate structure.
+Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Section 1.1 of the supplementary material), and visual inspection shows no obvious complicated dependence.
 Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor a moving-window transformation is called for.
@@
 We then select a model corresponding to the fine-scale observations of acceleration.
-Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$.
+Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$ Subdive States.
 Although $N^*$ is selected heuristically, we test the validity of this model in Section \ref{subsec:model_validation}.
@@
 In contrast to the coarse-scale observations, the fine-scale acceleration data exhibit significant sinusoidal behaviour.
-Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with window size $h=100$ (two seconds) and maximum frequency $\tilde{\omega}=10$ (5 Hz).
+Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with a window size of $h=100$ (two seconds) and a maximum frequency of $\tilde{\omega}=10$ (5 Hz).
 %We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
-We then have that $\z_{t,\tilde t^*} = \left\{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\right\}$, where $\zone_{t,\tilde t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window.
+We then have that $\z_{t,\tilde t^*} = \{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\}$, where $\zone_{t,\tilde t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window.
 Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material).
@@
 All components of $\Zone_{t,\tilde t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $\tilde t^*$.
-Each component is assumed to have its own mean and variance parameters, but all components have the same auto-correlation parameter.
+Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter.
 %Given that $X^*_{t,\tilde t^*} = i^*$, the corresponding unknown parameters are then $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}_{>0}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
@@
 %
-We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation (see Figure \ref{fig:CarHHMM-DFT}).
+We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation.
 The likelihood of this model is easily calculated using the forward algorithm and can be maximized with respect to the parameters above (see appendix for details).
-In addition to the CarHHMM-DFT, we consider three similar variations for comparison.
-As in the full model, all components of $\Z_{t,\tilde t^*}$ are assumed to be independent when conditioned on dive type and subdive state for each of the following models:
+Figure \ref{fig:CarHHMM-DFT} shows the dependence structure of the full CarHHMM-DFT.
+
+In addition to the CarHHMM-DFT, we consider three variations for comparison.
+As in the full model, all components of $\Z_{t,\tilde t^*}$ are assumed to be independent of one another when conditioned on dive type and subdive state for each of the following models:
 \begin{enumerate}
-    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models both $\Z_{t,\tilde t^*}$ and as emissions of a simple HMM rather than a CarHMM.
+    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models $\Z_{t,\tilde t^*}$ as emissions from a simple HMM rather than a CarHMM.
     \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,\tilde t^*}$ as emissions of a CarHMM.
@@
     \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Z_{t,\tilde t^*}$ as emissions of a CarHMM.
-This model assumes that there is only one dive type, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive.
+This model assumes that there is only one dive type.
+    %, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
 \end{enumerate}
 %
-Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes there is no auto-correlation between fine-scale observations, the CarHHMM does not incorporate the ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.
+Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes there is no auto-correlation between fine-scale observations, the CarHHMM does not incorporate ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.
 

### 2020-11-24 Report/Sections/results.tex
Read through 4.1

@@
 We first report the results from the full CarHHMM-DFT in detail and assess the quality of the fit.
-We then compare the results of the CarHHMM-DFT with those of the other candidate models.
+We then compare these results with those from the other candidate models.
 
-The coarse-scale emission distribution parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}).
-Dive type 1 corresponds to shorter, shallower dives which likely reflects resting, travelling, and (to a lesser extent) searching for prey.
+The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
+Dive type 1 corresponds to shorter, shallower dives which likely reflects resting, travelling, and to a lesser extent searching for prey.
 On the other hand, dive type 2 is longer, deeper, and may be associated with behaviours such as hunting and socialization \citep{Tennessen:2019b}.
@@
 
-The means of the ``wiggliness" $\Ztwo_{t,\tilde t^*}$ are each separated by an order of magnitude between subdive states (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}). 
+The means of ``wiggliness" $\Ztwo_{t,\tilde t^*}$ associated with each subdive state are separated by an order of magnitude (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}).
 Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest variance corresponding to $\Zone_{t,\tilde t^*}$.
@@
 This implies less overall activity and more consistent acceleration compared to the other subdive states.
-Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
-The auto-correlation of acceleration is also slightly lower than in subdive state 1.
+Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
+The auto-correlation of acceleration is also slightly lower than subdive state 1.
 We therefore hypothesize that subdive state 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
-Finally, the mean of $\Ztwo_{t,\tilde t^*}$ and variance of $\Zone_{t,\tilde t^*}$ in subdive state 3 are both much higher than in the other two states.
-The auto-correlation of $\Zone_{t,\tilde t^*}$ is also much lower.
+Finally, the mean of $\Ztwo_{t,\tilde t^*}$ and variance of $\Zone_{t,\tilde t^*}$ in subdive state 3 are both much higher than in the other two states, and the auto-correlation of $\Zone_{t,\tilde t^*}$ is also much lower.
 This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).
 
-The estimated probability transition matrices and associated stationary distributions are
+The estimated probability transition matrices and associated stationary distributions on the coarse scale are
 %
@@
 %
-for the transitions between dives. Within dives, we have
+for the transitions between dives. The estimated probability transition matrices and stationary distributions on the fine scale are 
 $$\hat \Gamma^{*(1)} = \begin{pmatrix} 
@@
 for dive types 1 and 2.
-The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that about 79\% of dives are short dives of type 1.
-The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1.
+%The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that 
+In summary, about 79\% of dives are short dives of type 1, and
+%The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that 
+the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1.
 This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
-The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 40\% of the time during a dive of type 1 but only 9\% of the time during a dive of type 2.
+%The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that
+Further, this killer whale is in the less active subdive state 1 40\% of the time during a dive of type 1 compared to only only 9\% of the time during a dive of type 2.
 Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}.
-Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives.
-Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.
+Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.
 
@@
 If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution.
-Histograms of the pseudoresiduals mostly support that this model is well-specified.
+Histograms of the pseudoresiduals mostly support that the CarHHMM-DFT is well-specified.
 One exception is $\Ztwo_{t,\tilde t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:pseudoresids}).
 This implies that the true distribution of $\Ztwo_{t,\tilde t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study.
-See Sections 1.4 through 1.6 of the supplementary material for pseudoresidual plots for all observations.
+See Sections 1.4 through 1.6 of the supplementary material for pseudoresidual plots corresponding to all observations and models.
 
 We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}.
-Each observed dive duration is weighted by the decoded probability that it corresponds to a particular dive type.
+Each observation of dive duration is weighted by the probability that it corresponds to a particular dive type as decoded by the \textit{forward-backward algorithm} \citep{Zucchini:2016}.
 This results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2.
-Each histogram is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT.
-Analogous histograms corresponding to the fine-scale observations $\z_{t,\tilde t^*}$ are contained in Sections 1.4 through 1.6 of the supplementary material.
+Each histogram is then plotted together with the corresponding emission distribution learned by the CarHHMM-DFT.
+Analogous histograms corresponding to the fine-scale observations are contained in Section 1 of the supplementary material.
 Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions.
@@
 In addition, the HHMM-DFT produces estimates of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$ which are approximately $50$ to $100$ percent larger than those of the CarHHMM-DFT for all axes of acceleration.
-The estimated uncertainties of the estimates of $\mu_A^{*(\cdot,1)}$, $\mu_A^{*(\cdot,2)}$, and $\mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT for all axes of acceleration (see Tables 1 and 2 of the supplementary material).
-Further, the empirical auto-correlation of acceleration within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT do not follow a standard normal distribution due to several outliers (see Figure 15 of the supplementary material).
+The estimated uncertainties of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT for all axes of acceleration (see Tables 1 and 2 of the supplementary material).
+Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of the supplementary material).
 These findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.
 
-The CarHHMM does not model the ``wiggliness" of the acceleration data, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives}.
-This essentially disqualifies the CarHHMM as a viable model for these data.
-The pseudoresiduals corresponding to the CarHHMM also do not appear to fit a Normal distribution.
+The CarHHMM does not model the ``wiggliness" of the acceleration data, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives} (see Figure 6 of the supplementary material).
+This essentially disqualifies the CarHHMM as a viable model for this data set.
+The pseudoresiduals corresponding to the CarHHMM are also light-tailed and do not fit a Normal distribution (see Figure 16 of the supplementary material).
 
@@
 However, its lack of hierarchical structure means that it fails to differentiate between short and long dives.
-This model therefore does not give any information regarding the dive-level Markov chain or the relationship between the dive and subdive levels.
+This model therefore does not infer the dive-level Markov chain or the relationship between the dive and subdive levels.
 In particular, the CarHMM-DFT does not indicate that the whale is more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives.
-A more complete set of results for each of the candidate models is in Section 1 of the supplementary material.
+
+A more complete set of results for each of the candidate models is presented in Section 1 of the supplementary material.

### 2020-11-24 Report/Sections/simulation.tex
Read through 4.1

@@
 The parameters used to generate the data are based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity.
-In particular, we set the number of subdive states to $N^*=2$ and $\Zone_{t,\tilde t^*}$ as a scalar rather than a three dimensional vector.
-We then fit the data with four models: the full CarHHMM-DFT, HHMM-DFT, CarHHMM, and CarHMM-DFT.
+In particular, we set the number of subdive states to $N^*=2$ and $\Zone_{t,\tilde t^*}$ to a scalar instead of a three dimensional vector.
+We then fit all four models to the simulated data.
 Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times.
-To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors a given model's parameter estimates with the standard errors estimated using the inverse of the observed Fisher information under that model.
+To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors of a given model's parameter estimates with the standard errors estimated using the inverse of the observed Fisher information.
 

### 2020-11-24 Report/Sections/discussion.tex
Read through main text

@@
 Our work addresses these issues and introduces a flexible framework to model functional time series data using HMMs.
-We deal with temporal dependence between curves with distinct ``types" by using either an HMM or a CarHMM to model the curve sequence.
+We deal with temporal dependence between curves by using either an HMM or a CarHMM to model the curve sequence.
 We then view each individual curve as an HMM emission whose distribution is described by yet another fine-scale model.
@@
 We demonstrate the usefulness of this framework using a biomechanical/ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale in Queen Charlotte Sound off the coast British Columbia, Canada.
-This gives a deeper understanding of a killer whale's tri-axial movement and thus its behaviour and energy expenditure \citep{Gleiss:2011,Qasem:2012}, both of which are important for understanding the foraging ecology and nutritional status of the Northern Resident killer whales \citep{Noren:2011}.
+This gives a deeper understanding of a killer whale's tri-axial movement and thus its behaviour and energy expenditure \citep{Gleiss:2011,Qasem:2012}, both of which are important for understanding the foraging ecology and nutritional status of northern resident killer whales \citep{Noren:2011}.
 This model is applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.

### 2020-11-24 Report/Sections/simulation.tex
Read through main text

@@
 
-We generate 500 independent data sets from the CarHHMM-DFT.
+We generate 500 independent data sets using the CarHHMM-DFT as a generative model.
 Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives.
@@
 %
-Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 25.7s$, $\sigma^{(1)} = 9.6s$, $\mu^{(2)} = 104.6s$, and $\sigma^{(2)} = 64.7s$.
-After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each dive are ignored.
-Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots \right\}$ with probability transition matrices
+Dive duration is Gamma distributed, and the coarse-scale emission parameters are $\mu^{(1)} = 25.7s$, $\sigma^{(1)} = 9.6s$, $\mu^{(2)} = 104.6s$, and $\sigma^{(2)} = 64.7s$.
+After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored.
+Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \right\}$ with probability transition matrices
 %
@@
 for dive types 1 and 2, respectively.
-Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$ as defined in Equation (\ref{eqn:z}).
+Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$.
 Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states.
@@
 %
-It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix.
-Figure \ref{fig:sim_data} shows one realization of $Z^*$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.
+It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix.
+Figure \ref{fig:sim_data} shows one realization of $\Z$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.
 
 The two simulated dive types differ in that dives of type one are much shorter on average (26 seconds) than dives of type two (105 seconds).
-The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$, both of which are much higher for subdive state 2 than for subdive state 1.
+The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$ since both are much higher for subdive state 2 than for subdive state 1.
 These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data.
@@
 Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows.
-We find both the average dive decoding accuracy and the average subdive decoding accuracy using the forward-backward algorithm.
+We find both the average dive decoding accuracy and the average subdive decoding accuracy using the \textit{forward-backward} algorithm \citep{Zucchini:2016}.
 
 The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model.
-The average dive decoding accuracy and the average subdive decoding accuracy are both greater than 0.9.
+Its average dive decoding accuracy and the average subdive decoding accuracy are both greater than 0.9.
 All parameter estimates of fine-scale mean values and probability transition matrices, $\hat \mu^*$, $\hat \Gamma^*$ and $\hat \Gamma$, respectively, have statistically insignificant biases.
 In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models.
-The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to underestimate the empirical standard errors.
-This is especially true for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error could be up to twice as large as the estimated standard error.
+The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors.
+This is especially true for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to twice as large as the estimated standard error.
 See Tables 5 through 9 of the supplementary material for detailed results.
@@
 The HHMM-DFT performs similarly to the CarHHMM-DFT in most respects.
-Its average dive decoding accuracy is comparable to the CarHHMM-DFT while its average subdive decoding accuracy is approximately $5 \%$ worse.
+Its average dive decoding accuracy is comparable to the CarHHMM-DFT while its average subdive decoding accuracy is worse by approximately 5 percentage points.
 Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$.
@@
 However, this model cannot estimate dive type as it lacks any hierarchical structure.
-The CarHMM-DFT nonetheless fits a single Gamma distribution over the dive duration of all dives.
+The CarHMM-DFT nonetheless fits a (misspecified) single Gamma distribution over the dive duration of all dives.
 The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 24 of the supplementary material).
@@
 Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} display one simulated data set as well as the decoded dive types and subdive states associated with each model.
-The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state while the HHMM-DFT is slightly more likely to predict that a given window corresponds to subdive state 2 (higher activity).
-The CarHHMM is the least accurate model when predicting subdive state.
-The CarHHMM-DFT and HHMM-DFT output similar estimates of dive type while the CarHHMM is less accurate and even misclassifies the second dive.
+The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state while the HHMM-DFT is slightly more likely to predict that a given window corresponds to subdive state 2.
+As expected, the CarHHMM is the least accurate model when predicting subdive state.
+The CarHHMM-DFT and HHMM-DFT yield similar estimates of dive type while the CarHHMM is less accurate and even misclassifies the second dive.
 The CarHMM-DFT does not estimate dive type.

### 2020-11-24 Report/Sections/appendix.tex
read figures and appendix

@@
 It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
-Then, given $X_t = i$, $Y_t = y_t$, and $\Gamma^{*(i)}$, it is straightforward to simulate the subdive states $\tilde{X}^*_t$ for a particular dive $t$.
-Recall that the length of each fine-scale Markov chain is $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ since each window is two seconds long.
+In addition, given $X_t = i$, $Y_t = y_t$, and $\Gamma^{*(i)}$, it is also straightforward to simulate the subdive states $\tilde{X}^*_t$ for a particular dive $t$.
+Recall that the length of each fine-scale Markov chain $\tilde{X}^*_t$ is $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ since each window is two seconds long.
 It is much more difficult to simulate the raw acceleration data $Y_t^*$, and we focus on this task for the remainder of the section.
 
-Recall that each window is two seconds long and therefore contains $h = 100$ observations taken at 50 observations per second.
+Recall that each window contains $h = 100$ observations taken at 50 observations per second.
 Let $DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,\tilde t^*}$ and $DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$ be denoted as $\hat{Y}^*_{t,\tilde t^*}$ for convenience.
@@
 \begin{enumerate}
-\item Simulate the average acceleration $Y^{*(0)}_{t,\tilde t^*}$ for each window $\tilde t^* = 1,2,\ldots,\tilde T^*_t$ using the following procedure:
+\item Simulate the average acceleration $\hat Y^{*(0)}_{t,\tilde t^*}$ for each window $\tilde t^* = 1,2,\ldots,\tilde T^*_t$ using the following procedure:
 \begin{enumerate}
@@
     %
-    \item Set Fourier coefficients 50 through 99 so the inverse DFT is real-valued:
+    \item Set Fourier coefficients 50 through 99 such that the inverse DFT is real-valued:
     $$
@@
     \end{enumerate}
-    \item Finally, set the raw acceleration data within each window $\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$ with the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$:
+    \item Finally, set the raw acceleration data within each window $\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$ as the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$:
     %
@@
 since $a^{(k)}_{t,\tilde t^*} \in \{-1,1\}$ and $b^{(k)}_{t,\tilde t^*} \geq 0$.
-Since $\Ztwo_{t,\tilde t^*}$ is the sum of Gamma-distributed random variables with identical scale parameters, the distribution of $\Ztwo_{t,\tilde t^*}$ is also Gamma.
+Because $\Ztwo_{t,\tilde t^*}$ is the sum of Gamma-distributed random variables with identical scale parameters, the distribution of $\Ztwo_{t,\tilde t^*}$ is also Gamma.
 Combining Equation (\ref{eqn:Wb}) above with Equation (\ref{eqn:bdist}) gives

### 2020-12-17 Code/Visualisor.py
Combine emission distribution figures

@@ -6,6 +6,7 @@ from matplotlib.patches import Patch
 from matplotlib.transforms import Bbox
 from matplotlib.lines import Line2D
 from matplotlib.ticker import FormatStrFormatter
+from matplotlib.gridspec import GridSpec
 from mpl_toolkits.axes_grid1 import make_axes_locatable
 
 from scipy.stats import gamma
@@ -141,7 +142,6 @@ class Visualisor:
             nrows = 2 #self.pars.K[0]
             ncols = int((len(features)+1)/2)
 
-        print(nrows,ncols)
         fig, ax = plt.subplots(nrows,ncols,figsize=(max(5,2.5*ncols),max(5,2.5*nrows)))
         ax = np.reshape(ax,(nrows,ncols))
 
@@ -219,6 +219,8 @@ class Visualisor:
 
         if file is None:
             plt.show()
+        elif file == 'noshow':
+            return (fig, ax)
         else:
             plt.tight_layout()
             bbox = Bbox([[0,0],[5,5.5]])

### 2020-12-17 Report/Sections/discussion.tex
Updates from Overleaf

@@
 Current Functional Data Analysis literature addresses dependence between curves either with multilevel models \citep{Chen:2012,Di:2009}, which lack a time component, or with functional time series, which overlook the possibility that curves have several distinct ``types" \citep{Kokoszka:2018}.
-Our work addresses these issues and introduces a flexible framework to model functional time series data using HMMs.
-We deal with temporal dependence between curves by using either an HMM or a CarHMM to model the curve sequence.
+Our work addresses these issues and introduces a flexible framework to model functional time-series data using HMMs.
+We handle temporal dependence between curves by using either an HMM or a CarHMM to model the curve sequence.
 We then view each individual curve as an HMM emission whose distribution is described by yet another fine-scale model.

### 2020-12-17 Report/Sections/intro.tex
Updates from Overleaf

@@
 For example, \citet{Morris:2007} view a child's activity level throughout the day as a curve where metabolic activity is a function of time.
-Similarly, \citet{Fu:2017} view the dive profile of an elephant seal (\textit{Mirounga leonina}) as a sequence of dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
+Similarly, \citet{Fu:2017} view the dive profile of a southern elephant seal (\textit{Mirounga leonina}) as dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
 %, although they ignore sequential dependence between dives. 
@@
 Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time.
-For example, marine mammal dive profiles show distinct dive ``types" that cluster together in time  \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}.
+For example, dive profiles of marine animals show distinct dive ``types" that cluster together in time  \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}.
 Fine-scale periodicity within a larger process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
@@
 
-Some FDA models account for between-curve dependence, but they are inadequate when modelling certain types of time dependencies. %, including that of Figure \ref{fig:data}.
+Some FDA models account for between-curve dependence, but these models are inadequate when modelling certain types of time dependencies. %, including that of Figure \ref{fig:data}.
 For example, 
@@
 %While functional time series methods can in theory be applied to biologging data such as ours,
-However, functional time series do not account for sequences of time series curves that switch between a discrete number of well-defined states.
+However, functional time series do not account for sequences of time-series curves that switch between a discrete number of well-defined states.
 %
@@
 However, the combination of apparent behavioural changes combined with periodic behaviour makes accelerometer data such as ours difficult to model with these classic FDA techniques.
-However, time series data exhibiting both sharp behavioural changes and periodic fine-scale structure is difficult to model with these classic FDA techniques.
+However, time-series data exhibiting both sharp behavioural changes and periodic fine-scale structure is difficult to model with these classic FDA techniques.
 
@@
 
-While useful, HMMs alone are also not sufficient to model intricate fine-scale time series data for three primary reasons.
+While useful, HMMs alone are also not sufficient to model intricate fine-scale time-series data for three primary reasons.
 %
@@
 We also show how these methods can be combined to analyze increasingly complex data.
-In Section 3 we fit several candidate models to data from a killer whale from the threatened northern resident population off the coast of British Columbia, Canada.
+In Section 3 we fit several candidate models to data from a killer whale (\textit{Orcinus orca}) from the threatened northern resident population off the coast of British Columbia, Canada.
 Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.

### 2020-12-17 Report/Sections/models.tex
Updates from Overleaf

@@
 Consider a sequence of $T$ curves, where any particular curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$.
-Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of function values from curve $t$.
+Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of function values from curve $t$ and indexed using $t^*$.
+Both $Y_t$ and $Y_{t,t^*}$ can be either vectors or scalars.
 We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.
@@
 Using observation emissions, denoted here as $y \equiv \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$.
-The likelihood $\calL_{\text{HMM}}$ can be evaluated using well-known \textit{forward algorithm} \citep{Zucchini:2016}:
+The likelihood $\calL_{\text{HMM}}$ can be evaluated using the well-known \textit{forward algorithm} \citep{Zucchini:2016}:
 %
-$$\calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N$$
+$$\calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N,$$
 %
@@
 
-\subsection{The CarHMM and relaxing conditional independence}
+\subsection{Relaxing conditional independence with the CarHMM}
 
@@
 Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots,X_T$ that can take on integer values between $1$ and $N$.
-It also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
+A CarHMM also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
 Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$.
-The first emission $Y_1$ is treated as fixed as an initial value which does not depend upon $X_1$.
+The first emission $Y_1$ is treated as a fixed initial value which does not depend upon $X_1$.
 We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$.
 This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density or probability mass function that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$.
-As a concrete example, if $Y_t$ is a $K$-dimensional vector with components $Y_t(1),\ldots,Y_t(K)$, consider the component-wise sequences $\{Y_1(k), \ldots, Y_T(k)\}$ for $k = 1,\ldots K$.
-One may assume that these sequences are independent given $\{X_1,\ldots,X_T\}$ and that $Y_t(k)$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)}_k = \{\mu^{(i)}_k,\sigma^{(i)}_k,\phi^{(i)}_k\}$, where:
+As a concrete example, if $Y_t$ is a scalar, then one may assume that $Y_t$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
 %
@@
 \begin{split}
-\mathbb{E}(Y_{t}(k)|Y_{t-1}(k) = y_{t-1}(k),X_t=i) &= \phi^{(i)}_k ~ y_{t-1}(k) ~+ ~(1-\phi^{(i)}_k)  ~\mu^{(i)}_k, \\
-\mathbb{V}(Y_t(k)| Y_{t-1}(k) = y_{t-1}(k), X_t = i) &= (\sigma^{(i)}_k)^2.
+\mathbb{E}(Y_{t}|Y_{t-1} = y_{t-1},X_t=i) &= \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}, \\
+\mathbb{V}(Y_t| Y_{t-1} = y_{t-1},X_t=i) &= (\sigma^{(i)})^2.
 \end{split}
@@
 
-\subsection{The HHMM and incorporating multiple scales}
+\subsection{Incorporating multiple scales with the HHMM}
 
@@
 The distribution of $Y_{t,t^*}$ given $Y_{t,t^*-1} = y_{t,t^*-1}$, $X_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$.
-We denote the set of fine-scale emission parameters corresponding to $X_t=i$ as $\theta^{*(i)}=\left\{\theta^{*(i,1)}, \ldots, \theta^{*(i,N^{*(i)})}\right\}$.
+We denote the set of fine-scale emission parameters corresponding to $X_t=i$ as $\theta^{*(i)}=\left\{\theta^{*(i,1)}, \ldots, \theta^{*\left(i,N^{*(i)}\right)}\right\}$.
 In summary:
@@
 
-Due to the nested structure of the hierarchical hidden Markov model, the likelihood is easily calculated using the forward algorithm.
+Due to the nested structure of the HHMM, the likelihood is easily calculated using the forward algorithm.
 %
@@
 
-\subsection{The HMM-DFT and transforming fine-scale observations}
+\subsection{Transforming fine-scale observations with the HMM-DFT}
 \label{subsec:STFT}
@@
 %
-for $t^* = 1,h+1,2h+1,\ldots$ and for $k = 0, 1, \ldots, h-1$.
+for $t^* = 1,h+1,2h+1,\ldots$ and $k = 0, 1, \ldots, h-1$ with $i^2 = -1$.
 We omit the final window if $t^*+h-1$ exceeds $T^*_t$, and if $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise.
@@
 where $\tilde{\omega} \leq h-1$ is a problem-specific tuning parameter corresponding to the maximum recorded frequency within each window.
-In words, $\Zone_{t,\tilde{t}^*}$ is the average acceleration vector within window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window.
+In words, $\Zone_{t,\tilde{t}^*}$ is the average value of $Y^*_t$ within window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window.
 %(by the Plancherel theorem).
-More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the acceleration data within dive $t$ and window $\tilde{t}^*$.
+More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the fine-scale data within curve $t$ and window $\tilde{t}^*$.
 
@@
 The likelihood of this model is therefore identical to that of the original HMM, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$.
-In order to clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT.
+To clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT.
 Figure \ref{fig:models}d displays the dependence structure for an HMM-DFT.
@@
 Therefore, models should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.
+
+%the fine scale behaviour follows a continuous time HMM.
+The parameters of the fine scale distributions depend on hidden states.
+However the states are modelled as components in a mixture distribution, not on an HMM.

### 2020-12-17 Report/Sections/results.tex

@@
 We then select a model corresponding to the fine-scale observations of acceleration.
-Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$ Subdive States.
+Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$ subdive states.
 Although $N^*$ is selected heuristically, we test the validity of this model in Section \ref{subsec:model_validation}.
@@
 Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with a window size of $h=100$ (two seconds) and a maximum frequency of $\tilde{\omega}=10$ (5 Hz).
-%We select a window size of two seconds since it appears be the smallest window size that captures at least one full period of each ``wiggle" in the acceleration data. 
 We then have that $\z_{t,\tilde t^*} = \{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\}$, where $\zone_{t,\tilde t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window.
@@
 
-All components of $\Zone_{t,\tilde t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) for all $t$ and $\tilde t^*$.
+All components of $\Zone_{t,\tilde t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) and independent of one another when conditioned on the dive types and subdive states.
 Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter.
@@
 Dive type 1 corresponds to shorter, shallower dives which likely reflects resting, travelling, and to a lesser extent searching for prey.
-On the other hand, dive type 2 is longer, deeper, and may be associated with behaviours such as hunting and socialization \citep{Tennessen:2019b}.
-No dive in this data set has a maximum depth deeper than 30 meters, and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters.
+Dive type 2 is longer, deeper, and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives.
+No dive in this data set has a maximum depth deeper than 30 meters and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters.
 However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and significant numbers of fish have been observed to be caught near the surface (Fortune and Trites, unpublished data).
-It is therefore unclear whether any of the dives in this study are successful foraging dives.
+Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.
 
@@
 Each observation of dive duration is weighted by the probability that it corresponds to a particular dive type as decoded by the \textit{forward-backward algorithm} \citep{Zucchini:2016}.
-This results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2.
+This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2.
 Each histogram is then plotted together with the corresponding emission distribution learned by the CarHHMM-DFT.
@@
 The CarHHMM does not model the ``wiggliness" of the acceleration data, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives} (see Figure 6 of the supplementary material).
-This essentially disqualifies the CarHHMM as a viable model for this data set.
-The pseudoresiduals corresponding to the CarHHMM are also light-tailed and do not fit a Normal distribution (see Figure 16 of the supplementary material).
+These results essentially disqualify the CarHHMM as a viable model for this data set.
+The pseudoresiduals of acceleration are also light-tailed relative to a Normal distribution (see Figure 16 of the supplementary material).
 

### 2020-12-17 Report/Sections/simulation.tex

@@
 It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix.
-Figure \ref{fig:sim_data} shows one realization of $\Z$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.
+Figure 21 of the supplementary material shows one realization of $\Z$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.
 
@@
 The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors.
-This is especially true for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to twice as large as the estimated standard error.
+This small bias is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to twice as large as the estimated standard error.
 See Tables 5 through 9 of the supplementary material for detailed results.
@@
 This finding is consistent with the results of the case study, where the HHMM-DFT produced larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.
+When standard errors are underestimated, the associated confidence intervals are too narrow, implying that researchers may be overconfident in their parameter estimates.
 
@@
 The CarHMM-DFT nonetheless fits a (misspecified) single Gamma distribution over the dive duration of all dives.
-The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 24 of the supplementary material).
+The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 25 of the supplementary material).
 

### 2020-12-17 Report/main.tex

@@
 \KWDtitle{Key words and phrases}
-Multi-state modelling\sep hierarchical modelling\sep time series\sep  biologging\sep statistical ecology\sep animal movement.
+Biologging\sep statistical ecology\sep animal movement\sep accelerometer data\sep diving behaviour\sep killer whales \sep multi-state modelling\sep hierarchical modelling\sep time series.
 % MSC 2010 subject classification codes
@@
 As a case study, we use this framework to model the fine-scale kinematic movement of a northern resident killer whale off the western coast of Canada.
-Through simulations we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines.
+Through simulations, we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines.
 
@@
 \begin{ack}{ACKNOWLEDGEMENTS}
-This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
+%This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
 \end{ack}

### 2020-12-17 Code/Visualisor.py
Add plot and sandbox for new data

@@ -142,7 +142,10 @@ class Visualisor:
             nrows = 2 #self.pars.K[0]
             ncols = int((len(features)+1)/2)
 
-        fig, ax = plt.subplots(nrows,ncols,figsize=(max(5,2.5*ncols),max(5,2.5*nrows)))
+        if level == 0:
+            fig, ax = plt.subplots(nrows,ncols,figsize=(max(5,2.5*ncols),2.5*nrows))
+        else:
+            fig, ax = plt.subplots(nrows,ncols,figsize=(max(5,2.5*ncols),max(5,2.5*nrows)))
         ax = np.reshape(ax,(nrows,ncols))
 
         for feature_num,feature in enumerate(features):

### 2020-12-17 Code/Visualisor.py
update plots

@@ -144,8 +144,10 @@ class Visualisor:
 
         if level == 0:
             fig, ax = plt.subplots(nrows,ncols,figsize=(max(5,2.5*ncols),2.5*nrows))
+            bbox = Bbox([[0,0],[max(5,2.5*ncols),2.5*nrows]])
         else:
             fig, ax = plt.subplots(nrows,ncols,figsize=(max(5,2.5*ncols),max(5,2.5*nrows)))
+            bbox = Bbox([[0,0],[max(5,2.5*ncols),2.5*nrows + 0.5]])
         ax = np.reshape(ax,(nrows,ncols))
 
         for feature_num,feature in enumerate(features):
@@ -226,7 +228,6 @@ class Visualisor:
             return (fig, ax)
         else:
             plt.tight_layout()
-            bbox = Bbox([[0,0],[5,5.5]])
             plt.savefig(file,bbox_inches=bbox,dpi=500)
 
         return

### 2020-12-22 Report/Sections/intro.tex
Fix Nancy Intro Comments

@@
 Biologging technology now provides researchers with kinematic data collected almost continuously in time \citep{Hooten:2017}.
-The collection and analysis of data from devices such as accelerometers has brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}.
+The collection and analysis of data from devices such as accelerometers have brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}.
 The study of animal movement in particular has been transformed by tracking devices that record kinematic information in a variety of environments \citep{Borger:2020,Dot:2016b}.
@@
 
-Biologging data are frequently modelled as a set of curves analyzed by methods of Functional Data Analysis (FDA, see, e.g., \citet{Ramsay:2005}).
-For example, \citet{Morris:2007} view a child's activity level throughout the day as a curve where metabolic activity is a function of time.
-Similarly, \citet{Fu:2017} view the dive profile of a southern elephant seal (\textit{Mirounga leonina}) as dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
+Biologging data are frequently modelled as a set of curves analyzed by methods of Functional Data Analysis, or FDA \citep{Ramsay:2005}.
+For example, \citet{Morris:2007} view a child's activity level as a set of daily curves where metabolic activity is a function of time.
+Similarly, \citet{Fu:2017} view the dive profile of a southern elephant seal (\textit{Mirounga leonina}) as a set of dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
 %, although they ignore sequential dependence between dives. 
@@
 %
-FDA was originally developed to process curves assumed to be independent replicates, but time series data are often complicated by sequential dependencies.
-In particular, biologging data often display significant temporal dependence both between curves on a coarse scale and within curves on fine scale \citep{Barajas:2017}.
+FDA was originally developed to process curves assumed to be independent replicates, and within-curve fine-scale structure is not usually incorporated in FDA models.
+However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
 %, as is the case for our case study data.
@@
 
-Some FDA models account for between-curve dependence, but these models are inadequate when modelling certain types of time dependencies. %, including that of Figure \ref{fig:data}.
-For example, 
-%multilevel models incorporate dependence via random effects. 
-previous studies have used multilevel models with random effects to capture variation between activity levels in children \citep{Morris:2007} and study differences in menstrual cycles \citep{Bromback:1998}.
-More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012}, and \citet{Crainiceanu:2009}.
-However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types.
-%In the aforementioned study of children's activity levels \citep{Morris:2007}, the authors used a multilevel approach with a random effects to model variation between children.
-In a classic paper that introduced this multilevel structure,  \citet{Bromback:1998} used cubic splines in a mixed effects model to study menstrual cycles.
-More recent work on multilevel models includes that of \citet{Di:2009}, \citet{Chen:2012} and \citet{Crainiceanu:2009}.
-However, multilevel models are not appropriate in many biologging applications since they account for neither temporal dependence nor different curve types.
-FDA researchers have also used functional time series to model dependence in sequences of curves.
-Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}.
-%While functional time series methods can in theory be applied to biologging data such as ours,
-However, functional time series do not account for sequences of time-series curves that switch between a discrete number of well-defined states.
+Some FDA models account for between-curve dependence occurring when multiple curves arise from separate groups of individuals. 
+Previous studies have used multilevel models with random effects to model variation between and within individuals, for example, in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}.
+More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}.
+However, these multilevel models are not appropriate in many biologging applications since they do not account for different curve types.
+In addition, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence is not appropriate in most biologging applications.
+
+FDA researchers have used functional time series to model dependence in a sequence of curves.
+Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by well-defined hidden states.
 %
@@
 However, the combination of apparent behavioural changes combined with periodic behaviour makes accelerometer data such as ours difficult to model with these classic FDA techniques.
-However, time-series data exhibiting both sharp behavioural changes and periodic fine-scale structure is difficult to model with these classic FDA techniques.
+However, time-series data exhibiting both sharp behavioural changes and periodic fine-scale structure are difficult to model with these classic FDA techniques.
 
-In light of these challenges, we turn to the field of
+To accommodate temporal dependence and changes in curve type, we turn to the field of
 %methodology for the analysis of animal movement 
@@
 animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
-HMMs interpret animal movement data as arising from different types of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position.
+HMMs interpret animal movement data as arising from a sequence of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position.
 While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions.
@@
 %This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments. 
-but this approach does not account for certain types of time correlation within or between individual time series.
+but this approach does not account for certain types of temporal correlation.
 
@@
 For example, \citet{Adam:2019} fail to capture fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}) using a traditional HHMM.
-\citet{Heerah:2017} use Fourier analysis within an HMM to account for %periodic behaviour in the context of describing
+\citet{Heerah:2017} successfully use Fourier analysis within an HMM to account for %periodic behaviour in the context of describing
 daily behavioural cycles in marine mammals, and Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
@@
 In this paper, we combine existing methods from statistical ecology literature in novel ways to account for complex, temporally dependent functional data.
-The resulting suite of methods make up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.
+The resulting suite of methods makes up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.
 %and thus extend the methods available for the analysis of functional data. 
-We begin by describing HMMs as well as to two variants, CarHMMs and HHMMs, and discuss how Fourier analysis can handle fine-scale dependence structures.
+We begin by describing HMMs as well as two variants, CarHMMs and HHMMs, and discuss how Fourier analysis can handle fine-scale dependence structures.
 We also show how these methods can be combined to analyze increasingly complex data.

### 2020-12-22 Report/main.tex
Fix Nancy Intro Comments

@@
 We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM).
-The distribution of each curve is then defined by another fine-scale model which may involve a variety of methods including moving-window transformations, Fourier analysis, and auto-regression.
+The distribution of each curve is then defined by another fine-scale model which may involve auto-regression and require data transformations using moving-window summary statistics or Fourier analysis.
 This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.
 As a case study, we use this framework to model the fine-scale kinematic movement of a northern resident killer whale off the western coast of Canada.
-Through simulations, we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing baselines.
+Through simulations, we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing methods.
 

### 2020-12-23 Report/Sections/models.tex
Nancy comments up to 2.5

@@
 Consider a sequence of $T$ curves, where any particular curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$.
-Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of function values from curve $t$ and indexed using $t^*$.
+Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up fine-scale quantities derived from curve $t$ and indexed using $t^*$.
 Both $Y_t$ and $Y_{t,t^*}$ can be either vectors or scalars.
 We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.
-We first detail the structure of a traditional HMM followed by three variations which generalize its base structure.
+To develop our model for this data, we detail the structure of a traditional HMM followed by three variations which generalize its base structure.
 We then show how each of these generalized HMMs can be synthesized to form a wide variety of more complicated models.
@@
 \subsection{HMMs as a base structure}
+\label{subsec:HMM}
 
@@
 %
-We assume that $X_1$ follows the chain's stationary distribution, which is denoted by $\delta \in \bbR^N$ with $i^{th}$ component
+We assume that $X_1$ follows the chain's stationary distribution, which is denoted by an N-dimensional row vector $\delta$, where
 $\delta_i = \Pr(X_1 = i).$
@@
 \subsection{Relaxing conditional independence with the CarHMM}
+\label{subsec:CarHMM}
 
@@
 \subsection{Incorporating multiple scales with the HHMM}
+\label{subsec:HHMM}
 
 A hierarchical hidden Markov model, or HHMM, accounts for processes occurring simultaneously at different scales by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs.
-The coarse-scale model is as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations with state-dependent parameters $\theta^{(i)}$ for $i = 1,\ldots,N$.
+The coarse-scale model is either an HMM and CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations with state-dependent parameters $\theta^{(i)}$ for $i = 1,\ldots,N$.
 %
@@
 For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
-If $X_t = i$, then the components of $X_t^*$ are a Markov chain with $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
+If $X_t = i$, then the components of $X_t^*$ follow a Markov chain with $N^{*(i)}$ possible states, an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$, and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
 The distribution of $Y_{t,t^*}$ given $Y_{t,t^*-1} = y_{t,t^*-1}$, $X_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$.
@@
     %
-    \{Y_t | Y_{t-1} = y_{t-1}, X_t = i\} \sim f^{(i)}(\cdot|y_{t-1};\theta^{(i)}), \\
+    (Y_t | Y_{t-1} = y_{t-1}, X_t = i) \text{ has density } f^{(i)}(\cdot|y_{t-1};\theta^{(i)}), \\
     %
@@
     %
-    \{Y^*_{t,t^*} | Y^*_{t,t^*-1} = y^*_{t,t^*-1}, X^*_{t,t^*} = i^*, X_t = i\} \sim f^{*(i,i^*)}(\cdot|y^{*}_{t,t^*-1};\theta^{(i,i^*)}).
+    (Y^*_{t,t^*} | Y^*_{t,t^*-1} = y^*_{t,t^*-1}, X^*_{t,t^*} = i^*, X_t = i) \text{ has density } f^{*(i,i^*)}(\cdot|y^{*}_{t,t^*-1};\theta^{(i,i^*)}).
 \end{gather*}
-Given the coarse-scale hidden states $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$ and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
+Given the coarse-scale hidden state sequence $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$ and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
 %
@@
 
-Forcing certain parameters to be shared can reduce the complexity of the HHMM.
+Forcing certain parameters to be shared can reduce complexity and increase interpretability of the HHMM.
 For example, in our killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$.
-We also share the fine-scale emission parameters across coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
+We also share the fine-scale emission parameters across the $N$ coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
 Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$.
@@
 \begin{equation}
-    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N
+    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N
     \label{eqn:HHMM_likelihood}
@@
 %
-where $P(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_1)\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
-\Gamma^{*(i)};y_1^*\right)$ if the coarse-scale model is an HMM or $\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
-\Gamma^{*(i)};y_1^*\right)$ if the coarse-scale model is a CarHMM.
-The fine-scale likelihood $\calL_{\text{(Car)HMM}}$ corresponds to the likelihood of the fine-scale model, which can be either a CarHMM or an HMM.
-For $t \geq 2$, $P(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to
-$f^{(i)}(y_t|y_{t-1})\calL_{\text{(Car)HMM}}\left(\theta^{*(i)},
-\Gamma^{*(i)};y_t^*\right)$. Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.
+where $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix whose exact structure depends upon the coarse- and fine-scale models.
+If the coarse-scale model is an HMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ both have $(i,i)^{th}$ entries equal to
+$f^{(i)}(y_t)\calL_{\text{fine}}\left(\theta^{*(i)},
+\Gamma^{*(i)};y_t^*\right)$. 
+If the coarse-scale model is a CarHMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ has $(i,i)^{th}$ entry equal to $\calL_{\text{fine}}\left(\theta^{*(i)},
+\Gamma^{*(i)};y_1^*\right)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ has $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1})\calL_{\text{fine}}\left(\theta^{*(i)},
+\Gamma^{*(i)};y_t^*\right)$.
+The fine-scale likelihood $\calL_{\text{fine}}$ corresponds to the likelihood of the fine-scale model, which can be either a CarHMM or an HMM.
+Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.
 %
@@
 
-In many applications where data are collected at high frequencies, intricate dependency structures often arise within the fine-scale process which cannot be adequately modelled with any HMM-based model described thus far.
+In many applications where data are collected at high frequencies, intricate dependency structures arise within the fine-scale process that cannot be adequately modelled with any HMM-based model described thus far.
 %For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. 
 To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize any non-Markovian behaviour.
-To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over the elements $Y_t^*$.
+To maintain the temporal structure of the fine-scale process, local summary statistics can be calculated from a moving window over the elements of $Y_t^*$.
 Subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window.
+Stride length refers to the distance between the first element of consecutive windows, so a stride length of $h$ implies that the first window starts at $Y^*_{t^*,1}$, the second at $Y^*_{t^*,h+1}$, and so on.
 Larger stride lengths result in a loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting.
@@
 %
-for $t^* = 1,h+1,2h+1,\ldots$ and $k = 0, 1, \ldots, h-1$ with $i^2 = -1$.
-We omit the final window if $t^*+h-1$ exceeds $T^*_t$, and if $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise.
-We index each window with $\tilde{t}^*$ and calculate transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$:
+for $t^* = 1,h+1,2h+1,\ldots$ and $k = 0, 1, \ldots, h-1$ with $i = \sqrt{-1}$.
+If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise.
+We omit the final window if $t^*+h-1$ exceeds $T^*_t$, denote the total number of windows as $\tilde T^*_t$, and index each window with $\tilde{t}^* = 1,\ldots,\tilde T^*_t$.
+Next, we calculate transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$:
 %
@@
 
-The fine-scale HMM must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself.
-In particular, if the length of $\Z_t$ is denoted as $\tilde{T}_t^*$ then there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
+If the stride length $h$ is greater than one, then the fine-scale HMM (or CarHMM) must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself.
+In particular, there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
 The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ instead of $X^*_t$ and $Y^*_t$.
-The likelihood of this model is therefore identical to that of the original HMM, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$.
+The likelihood of this model is therefore identical to that of the original HMM or CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$.
 To clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT.
@@
 %By definition, an 
-Traditional HHMMs treat both the coarse-scale and the fine-scale processes of interest as realizations of an HMM or CarHMM. 
+Traditional HHMMs treat both the coarse-scale and the fine-scale processes as realizations of an HMM or CarHMM. 
 %However, it is straightforward to generalize the coarse-scale HMM by replacing it with a CarHMM, yielding the following likelihood:
@@
 However, the fine-scale observations for a particular dive $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods.
-In other words, the fine-scale HMM likelihood term $\calL_{\text{(Car)HMM}}$ in Equation (\ref{eqn:HHMM_likelihood}) can be replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
-\[
-\calL_{\text{coarse}}(\theta,\theta^*,\Gamma;y,y^*) = \delta P(y_1,y^*_1;\theta,\theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\theta,\theta^*) \mathbf{1}_N
-\]
-where $P(y_t,y^*_t;\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\theta^{(i)}\right)\calL_{\text{fine}}\left(\theta^{*(i)};y^*_t\right)$.
+To this end, the fine-scale HMM likelihood term $\calL_{\text{fine}}$ in Equation (\ref{eqn:HHMM_likelihood}) can be replaced by the likelihood of a general fine-scale model whose parameters depend upon the coarse-scale hidden state.
+%$\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
+%$$
+%\calL_{\text{coarse}}(\theta,\theta^*,\Gamma;y,y^*) = \delta P(y_1,y^*_1;\theta,\theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\theta,\theta^*) \mathbf{1}_N,
+%$$
+%where $P(y_t,y^*_t;\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\theta^{(i)}\right)\calL_{\text{fine}}\left(\theta^{*(i)};y^*_t\right)$.
 %
 Possible candidates for the fine-scale model include any of the models described in the previous subsections in addition to many others not described here. 
-For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations on a Poisson process as a fine-scale model.
-If the fine-scale model is a simple Poisson process, then the hierarchical structure described here reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
-Further, \citep{Langrock:2018} use B-splines to model the emission distribution of an HMM as well as incorporate covariates.
-This non-parametric approach admits a penalized likelihood which can easily be incorporated into this hierarchical structure.
+This general For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations of a Poisson process as their fine-scale model.
+If the fine-scale model is a simple Poisson process, then this is equivalent to a Markov-modulated Poisson process \citep{Fischer:1993}.
+The fine-scale structure can also be modelled as in \citep{Langrock:2018}, who use B-splines to model the emission distribution of an HMM as well as incorporate covariates.
+This non-parametric approach uses a penalized likelihood term which can easily replace the usual fine-scale likelihood term in (\ref{eqn:HHMM_likelihood}).
 Another class of fine-scale models is the set of continuous time methods such as a continuous-time HMM (CTHMM) \citep{Liu:2015} or a state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019}.
-A continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}.
-\citet{Xu:2018} explain biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours, but they take a longitudinal approach which ignores temporal dependence on the coarse scale.
+A continuous-time HMM may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}.
+\citet{Xu:2018} model biologging accelerometer data for a collection of individuals by incorporating a CTHMM into a hierarchical model similar to ours.
+However, they use a categorical distribution as their coarse-scale model rather than a Hidden Markov Model
 

### 2020-12-25 Report/Sections/appendixnh.tex
Nancy edits through 4.1

\subsection{Detailed description of data simulation from Section \ref{subsec:data_simulation}}
%\numberwithin{equation}{section}

\setcounter{equation}{4}   %  set this to the last equation number used

{\bf{I didn't like the equation numbering restarting at (1), so I used ``setcounter".}}

We easily simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$. 
To simulate data at the fine scale 
for dive $t$, recall that the fine scale model is based on a sequence of $\tilde T_t^*$ two-second windows, each containing 100 values of raw acceleration.  
 We easily generate the fine-scale hidden Markov process,  $\tilde X ^* _{t,\tilde t^*}$,
$\tilde t^*=1,\ldots,\tilde T_t^*$, that determines the sequence of underlying states in each window according to the specified parameters in $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$.
 However, generating the raw acceleration data is not straightforward, since our model is formulated in terms of quantities derived from the raw data in each window, namely the average acceleration, $A_{t,\tilde t^*}$, and the wiggliness, $W_{ t,\tilde t^*}$. In our simulation study, we generate raw acceleration data so that we can visualize our results in terms of the underlying data curves for each dive.  Here, we explain how we generate the acceleration curves  so that the derived quantities, $A_{t,\tilde t^*}$ and $W_{ t,\tilde t^*}$, follow the specified model. A key component is the discrete Fourier transformation of the  100 raw acceleration values in the $\tilde{t}^*$th window of dive $t$:
\[
\hat{Y}^{*(k)}_{t,\tilde{t}^*}  \equiv DFT(Y^*_{t,100 (\tilde{t}^*-1) + 1 },\ldots, Y^*_{t,100 \tilde{t}^*})[k]
\]
for $k \ge 0$, as defined in Equation \ref{eq:DFTdef}.

 We simulate the raw acceleration data for dive $t$ in three steps: (1) simulate the average acceleration within each window $\left(\hat Y^{*(0)}_{t,\tilde t^*}\right)$, (2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and (3) take the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$:
 \[
 \{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\} \equiv IDFT\left\{\hat{Y}^{*(0)}_{t,\tilde t^*},
 \ldots \hat{Y}^{*(99)}_{t,\tilde t^*}\right\}
    {\rm{~for~}} \tilde t^* = 1,\ldots,\tilde T^*_t.
 \]
The details of steps (1) and (2) are described below.

For step (1), we generate $\hat Y^{*(0)}_{t,1}, \ldots, \hat Y^{*(0)}_{t,\tilde T_t^*}$  as a CARHMM with underlying Markov state sequence $X^{*}_{t,1}, \ldots, X^{*}_{t,\tilde T_t^*}$ and random first emission.  Specifically, we let
   \[
    	\hat{Y}^{*(0)}_{t,1}|\tilde X^*_{t,1} = i^* ~~\sim~~ \mathcal{N} \left(0, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
    	~~{\rm{and}}
   \]
   %
     \begin{align}	
       \hat{Y}^{*(0)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = i^*,\hat{Y}^{*(0)}_{t,\tilde t^*-1}
       &~~\sim ~~\mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,\tilde t^*-1}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \label{eqn:yhat_0} \\
        %
        &\tilde t^* = 2,\ldots, \tilde T^*_t
    	\nonumber
    \end{align}
    where $\sigma_A^{*(\cdot,1)} = 0.034s$,  $\sigma_A^{*(\cdot,2)} = 0.079s$. $\phi_A^{*(\cdot,1)} = 0.98$ and $\phi_A^{*(\cdot,2)} = 0.87$.

 For step (2),
 we first construct $\hat{Y}^{*(k)}_{t,\tilde t^*}$, $k=1,\ldots, 49$, as
    \begin{equation}
        \hat{Y}^{*(k)}_{t,\tilde t^*} = a_{t,\tilde t^*}^{(k)} i\sqrt{b^{(k)}_{t,\tilde t^*}}
        \label{eqn:abYhat}
    \end{equation}
    {\bf{WHY IS THERE AN $i$ IN THE ABOVE?}}
     where the $a^{(k)}_{t,\tilde t^*}$s are independent and equal to either 1 or -1 each with probability 1/2.
    Given the fine scale states, the $b^{(k)}_{t,\tilde t^*}$s are independent  and independent of the  $a^{(k)}_{t,\tilde t^*}$s.  
   The distribution of $b^{(k)}_{t,\tilde t^*}$ is
    \begin{align}
    \begin{split}
    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1 &~~\sim~~ {\rm{Gamma}}(16.38/k^3, 36.23) \\
    	%
    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2  &~~\sim ~~{\rm{Gamma}}(4.20/k^3, 1825.53). \\ 
    \end{split}
    \label{eqn:bdist}
    \end{align}
    The first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the shape parameter and the second is the scale parameter. The squared magnitude of the $k^{th}$ Fourier coefficient is equal to $b^{(k)}_{t,\tilde t^*}$, which decays like $1/k^3$ to ``smooth out" the raw acceleration data.

   We then define the remaining 50 Fourier coefficients:
    \[
    \hat{Y}^{*(50)}_{t,\tilde t^*} = 0 {\rm{~~and ~~}}
	\hat{Y}^{*(k)}_{t,\tilde t^*} = -\hat{Y}^{*(100-k)}_{t,\tilde t^*} , k = 51,\ldots,99.
\]
   This construction guarantees that the inverse discrete Fourier transform is real-valued.

We now show that this construction of the raw acceleration data  results in the distributions listed in Section \ref{subsec:data_simulation}. It suffices to show that the construction of the discrete Fourier transformations, the $\hat{Y}^{(k)}_{t,\tilde t^*}$s, yields the desired distributions.

First, since $\hat{Y}^{(0)}_{t,\tilde t^*} = \sum_{n=1}^{100} Y^*_{t,100(\tilde t^* - 1) + n} $ $=\Zone_{t,\tilde{t}^*}$, Equation (\ref{eqn:yhat_0}) implies that 
$\Zone_{t,1}, \ldots, \Zone_{t,\tilde{T}^*}$ 
follows a CARHMM with Normal emissions distributions and parameters 
 $\mu_A^{*(\cdot,1)} = \mu_A^{*(\cdot,2)} = \mu_A^{*(\cdot,i^*)} = 0$,  $\sigma_A^{*(\cdot,1)} = 0.034s$, $\phi_A^{*(\cdot,1)} = 0.98$, $\sigma_A^{*(\cdot,2)} = 0.079s$, and $\phi_A^{*(\cdot,2)} = 0.87$.

 From Equations \ref{eqn:z} and \ref{eqn:abYhat}, 
 the  wiggliness in window $\tilde t^*$  of dive $t$ is 
\[
    \Ztwo_{t,\tilde t^*} = \sum_{k=1}^{\tilde \omega} \big|\big|\hat{Y}^{(k)}_{t,\tilde t^*}\big|\big|^2 =  \sum_{k=1}^{\tilde \omega} b^{(k)}_{t,\tilde t^*}
%    \label{eqn:Wb}
\]
%
If $\tilde \omega < 50$, then $\Ztwo_{t,\tilde t^*}$ is the sum of independent Gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,\tilde t^*}$ is also Gamma.  Thus, by Equation  \ref{eqn:bdist}
%
\[
\Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 16.38/k^3 , 36.23 \right) 
\text { and }
\]
\[
\Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.20/k^3 , 1825.53 \right).
\]
Setting $\tilde \omega$ to 10 and carrying out simple calculation of the mean and variance of a Gamma distribution yield $\mu_W^{*(\cdot,1)} = 23.3$, $\sigma_W^{*(\cdot,1)} = 13.0$, $\mu_W^{*(\cdot,2)} = 301.2$, and $\sigma_W^{*(\cdot,2)} = 330.1$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Likelihood of CarHHMM-DFT}

The overall likelihood of the CarHHMM-DFT model is as follows:
%
$$\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,\z) = \delta P(y_1,\z_1;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,\z_t;\theta,\theta^*,\Gamma^*) \mathbf{1}_N.$$
%
In particular,
%
\begin{align*}
P(y_t,\z_t;\theta,\theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\theta^{(1)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(1)};\z_t\right), \ldots , \\
&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(N)};\z_t\right) \Big],
\end{align*}
%
where $f^{(i)}(y_t;\theta^{(i)})$ is the emission distribution of dive duration given that $X_t = i$. The likelihood $\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale model and is equal to the following:
%
$$\calL_{\text{CarHMM-DFT}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},$$
%
where $P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*)$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{*(\cdot,i^*)}(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1}; \theta^{*(\cdot,i^*)})$.
%
Recall that $f^{*(\cdot,i^*)}(\cdot|\z_{t,t^*-1}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,\tilde t^*}$ when $\tilde X^*_{t,\tilde t^*} = i^*$ and $\Z_{t,\tilde t^*-1} = \z_{t,\tilde t^*-1}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2020-12-25 Report/Sections/models.tex
Nancy edits through 4.1

@@
 %
-\begin{align*}
+\begin{align}
     DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right)
-\end{align*}
+    \label{eq:DFTdef}
+\end{align}
 %
@@
 However, the fine-scale observations for a particular dive $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods.
-To this end, the fine-scale HMM likelihood term $\calL_{\text{fine}}$ in Equation (\ref{eqn:HHMM_likelihood}) can be replaced by the likelihood of a general fine-scale model whose parameters depend upon the coarse-scale hidden state.
+As such, the fine-scale HMM likelihood term $\calL_{\text{fine}}$ in Equation (\ref{eqn:HHMM_likelihood}) can be replaced by the likelihood of a general fine-scale model whose parameters depend upon the coarse-scale hidden state.
 %$\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
@@
 Possible candidates for the fine-scale model include any of the models described in the previous subsections in addition to many others not described here. 
-This general For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations of a Poisson process as their fine-scale model.
+For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations of a Poisson process as their fine-scale model.
 If the fine-scale model is a simple Poisson process, then this is equivalent to a Markov-modulated Poisson process \citep{Fischer:1993}.
@@
 A continuous-time HMM may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}.
-\citet{Xu:2018} model biologging accelerometer data for a collection of individuals by incorporating a CTHMM into a hierarchical model similar to ours.
-However, they use a categorical distribution as their coarse-scale model rather than a Hidden Markov Model
+\citet{Xu:2018} model high-frequency biologging accelerometer data of individuals by incorporating a CTHMM into a hierarchical model similar to ours.
+However, they assume that individuals are partitioned into subgroups a priori whereas we use an HMM to infer the coarse-scale hidden states.
 

### 2020-12-25 Report/Sections/results.tex
Nancy edits through 4.1

@@
 %It is therefore of interest to determine the energetic expenditure of killer whales hunting for Chinook as well as the frequency at which they catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
-Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}.
+Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but the animal's behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}.
 Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of a killer whale.
@@
 There were also periods of time when the tag failed to record observations, resulting in data gaps between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm.
-To preprocess the data, we smooth the accelerometer and time-depth recorder data by taking a moving average within a window of $1/10^{th}$ of a second.
+To preprocess the data, we smooth the depth and acceleration curves by taking a moving average within a window of $1/10^{th}$ of a second.
 We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
@@
 Specifically, the sequence of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds.
-For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots \right\}$, which is a sequence containing the within-dive acceleration data in units of meters per second squared.
+For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots,y^*_{t,T^*_t} \right\}$, which is a sequence containing the within-dive acceleration data in units of meters per second squared.
 The collection of all acceleration data is denoted as $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.
@@
 We do not use information criteria to select the number of dive types $N$ since these metrics tend to overestimate the number of behavioural states in biological processes \citep{Pohle:2017}.
-We instead plot the duration of each dive versus the duration of the dive preceding it ($y_t$ vs $y_{t-1}$ for $t \geq 2$).
+We instead plot the duration of each dive versus the duration of the dive preceding it ($y_t$ versus $y_{t-1}$ for $t \geq 2$).
 This type of visualization is known as a lag plot.
-If the hidden states are well-separated, a lag plot should reveal $N$ distinct patterns, where each pattern corresponds to one dive type \citep{Lawler:2019}.
+If the emission distributions of the hidden states are well-separated, a lag plot should reveal $N$ distinct patterns, where each pattern corresponds to one dive type \citep{Lawler:2019}.
 This is unfortunately not the case for our killer whale data, as there is one cluster of data centred at approximately $y_t = y_{t-1} = 30$ seconds.
@@
 Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material).
-Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.
+Therefore, we choose a CarHMM as defined in Section 2.2 for the fine-scale model.
 
 We then select the specific emission distribution of $\Z_{t,\tilde t^*}$.
-Given the dive types and subdive states, we assume that $\Ztwo_{t,\tilde t^*}$ and all components of $\Zone_{t,\tilde t^*}$ are independent of one another.
+First, we assume that $\Ztwo_{t,\tilde t^*}$ and all three components of $\Zone_{t,\tilde t^*}$ are independent of one another when conditioned on the dive types and subdive states.
 To reduce model complexity, we also assume that the three fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$.
@@
 
-All components of $\Zone_{t,\tilde t^*}$ are assumed to be Normally distributed as in Equation (\ref{eqn:carhmm}) and independent of one another when conditioned on the dive types and subdive states.
+To specify the emission distribution of $\Zone_{t,\tilde t^*}$, consider the sequence $\Zone_{t,1},\Zone_{t,2},\ldots,\Zone_{t,T^*_t}$ for a particular dive $t$.
+We assume that each of the three components of this sequence are Normally distributed as in Equation (\ref{eqn:carhmm}) and independent of one another when conditioned on the subdive states $\tilde X_{t,1},\ldots,\tilde X_{t,\tilde T^*_t}$.
 Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter.
-%Given that $X^*_{t,\tilde t^*} = i^*$, the corresponding unknown parameters are then $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}_{>0}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
-Given that $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ is assumed to follow a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
+Thus the distribution of $\Zone_{t,\tilde t^*}$ given $X^*_{t,\tilde t^*} = i^*$ has parameters $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
+To specify the emission distribution of $\Ztwo_{t,\tilde t^*}$, we assume that given $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ follows a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
+In addition, $\Ztwo_{t,1},\ldots,\Ztwo_{t,T^*_t}$ are assumed to be independent of one another given the subdive state sequence $\tilde X_{t,1}, \ldots, \tilde X_{t,T_t^*}$.
 We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
@@
 We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation.
-The likelihood of this model is easily calculated using the forward algorithm and can be maximized with respect to the parameters above (see appendix for details).
+The likelihood of this model is easily calculated using the forward algorithm and can be maximized with respect to the parameters above (see the appendix for details).
 Figure \ref{fig:CarHHMM-DFT} shows the dependence structure of the full CarHHMM-DFT.
@@
 In addition to the CarHHMM-DFT, we consider three variations for comparison.
-As in the full model, all components of $\Z_{t,\tilde t^*}$ are assumed to be independent of one another when conditioned on dive type and subdive state for each of the following models:
+As in the full model, each of the following models assume that all components of $\Z_{t,\tilde t^*}$ are assumed to be conditionally independent of one another given the dive types and subdive states:
 \begin{enumerate}
@@
 %
-Each of the three candidate models above leave out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes there is no auto-correlation between fine-scale observations, the CarHHMM does not incorporate ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.
+Each of the three candidate models above leaves out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes there is no auto-correlation between fine-scale observations, the CarHHMM does not incorporate ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.
 
@@
 
-The means of ``wiggliness" $\Ztwo_{t,\tilde t^*}$ associated with each subdive state are separated by an order of magnitude (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}).
+The means of ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ associated with each subdive state are separated by an order of magnitude (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}).
 Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest variance corresponding to $\Zone_{t,\tilde t^*}$.
@@
 This implies less overall activity and more consistent acceleration compared to the other subdive states.
-Subdive state 2 has a mean ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
+Subdive state 2 has a mean ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
 The auto-correlation of acceleration is also slightly lower than subdive state 1.
@@
 We use two visual tools to evaluate the CarHHMM-DFT: pseudoresidual plots and empirical histograms.
-The pseudoresidual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
+The pseudoresidual of a coarse-scale observation $y_t$ is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T,\Z_1,\ldots,\Z_T\}/\{Y_t\}) \right)$ and the pseudoresidual of a fine-scale observation $\z_{t,\tilde t^*}$ is $\Phi^{-1} \left(Pr(\Z_{t,t^*} < \z_{t,\tilde t^*}|\{Y_1,\ldots,Y_T,\Z_1,\ldots,\Z_T\}/\{Y^*_{t,\tilde t}\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
 If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution.
@@
 This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2.
-Each histogram is then plotted together with the corresponding emission distribution learned by the CarHHMM-DFT.
+Each histogram is then plotted together with the corresponding emission distribution estimated by the CarHHMM-DFT.
 Analogous histograms corresponding to the fine-scale observations are contained in Section 1 of the supplementary material.
 Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions.
-In particular, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale.
+In histograms corresponding to subdive state 3, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavier tails compared to a normal distribution, indicating the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale.
 These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.
@@
 The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive state 3 (see Figures 4 and 5 of the supplementary material).
-In addition, the HHMM-DFT produces estimates of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$ which are approximately $50$ to $100$ percent larger than those of the CarHHMM-DFT for all axes of acceleration.
-The estimated uncertainties of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT for all axes of acceleration (see Tables 1 and 2 of the supplementary material).
+In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are approximately $50$ to $100$ percent larger than those of the CarHHMM-DFT.
+The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of the supplementary material).
 Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of the supplementary material).
-These findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.
+These findings suggest that the HHMM-DFT is a significantly worse fit to these data than the full CarHHMM-DFT.
 
@@
 
-Finally, the CarHMM-DFT, which lacks a hierarchical structure, produces similar fine-scale parameter estimates and state estimates compared to those of the CarHHMM-DFT.
+Finally, the CarHMM-DFT, which lacks a hierarchical structure, produces fine-scale parameter estimates and state estimates similar to those of the CarHHMM-DFT.
 However, its lack of hierarchical structure means that it fails to differentiate between short and long dives.

### 2020-12-25 Report/Sections/simulation.tex
Nancy edits through 4.1

@@
 %
-Dive duration is Gamma distributed, and the coarse-scale emission parameters are $\mu^{(1)} = 25.7s$, $\sigma^{(1)} = 9.6s$, $\mu^{(2)} = 104.6s$, and $\sigma^{(2)} = 64.7s$.
+Dive duration is Gamma distributed and the coarse-scale emission parameters are 
+%
+\[\mu^{(1)} = 25.7s, \enspace \sigma^{(1)} = 9.6s, \enspace \mu^{(2)} = 104.6s, \enspace \sigma^{(2)} = 64.7s.\]
+%
 After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored.
@@
 %
-$$\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}$$ 
+\[\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}\]
 %
@@
 Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$.
-Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states.
+Recall from Section \ref{subsec:model_selection} that we must specify the mean, standard deviation, and auto-correlation parameters corresponding to $\left\{\Zone_{t,1},\ldots,\Zone_{t,\tilde T_t^*}\right\}$ as well as the mean and standard deviation parameters corresponding to $\left\{\Ztwo_{t,1},\ldots,\Ztwo_{t,\tilde T_t^*}\right\}$.
+We select the following parameters in line with the results from the case study:
+%$\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states.
 Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation.
@@
 
-The two simulated dive types differ in that dives of type one are much shorter on average (26 seconds) than dives of type two (105 seconds).
+The two simulated dive types differ in that dives of type 1 are much shorter on average (26 seconds) than dives of type 2 (105 seconds).
 The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$ since both are much higher for subdive state 2 than for subdive state 1.
@@
 
-We use several metrics to compare the candidate models, including its accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}).
+We use several metrics to compare the candidate models, including accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}).
 If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives.

### 2020-12-28 Report/Sections/discussion.tex
Nancy comments through main text

@@
 Our work addresses these issues and introduces a flexible framework to model functional time-series data using HMMs.
-We handle temporal dependence between curves by using either an HMM or a CarHMM to model the curve sequence.
-We then view each individual curve as an HMM emission whose distribution is described by yet another fine-scale model.
+We suggest handling temporal dependence between curves by using either an HMM or a CarHMM to model the curve sequence.
+We then suggest viewing each individual curve as an HMM emission whose distribution is described by a fine-scale model.
 Here we use a CarHMM as the fine-scale model, but there are a wide range of possible fine-scale models, including a Poisson process or continuous time approach similar to that of \citet{Michelot:2019}.
-We also incorporate a moving-window transformation into the fine-scale model to deal with intricate dependence structures on short time scales.
+We also incorporate a moving-window transformation at the fine-scale model to capture intricate dependence structures on short time scales.
 Together, the coarse- and fine-scale models make up a hierarchical structure which can account for simultaneous processes taking place at different time scales.
@@
 
-We demonstrate the usefulness of this framework using a biomechanical/ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale in Queen Charlotte Sound off the coast British Columbia, Canada.
-This gives a deeper understanding of a killer whale's tri-axial movement and thus its behaviour and energy expenditure \citep{Gleiss:2011,Qasem:2012}, both of which are important for understanding the foraging ecology and nutritional status of northern resident killer whales \citep{Noren:2011}.
-This model is applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.
-In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time series data in a variety of fields.
+We demonstrate the usefulness of this framework using a biomechanical/ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale in Queen Charlotte Sound off the coast of British Columbia, Canada.
+Our analysis gives a deeper understanding of a killer whale's tri-axial movement and thus its behaviour and energy expenditure \citep{Gleiss:2011,Qasem:2012}, both of which are important for understanding the foraging ecology and nutritional status of northern resident killer whales \citep{Noren:2011}.
+Our model is applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.
+In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt our methodology for the analysis of a wide range of time series data in a variety of fields.
 

### 2020-12-28 Report/Sections/simulation.tex
Nancy comments through main text

@@
 These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data.
-All four candidate models are fit to each of the 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
+
+We calculate maximum likelihood estimates $\hat \theta$ for all four candidate models for each of the 500 training data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
+We then generate a corresponding independent test data set consisting of an additional 100 dives from the generative model with the true parameters $\theta$.
+
+If the parameters $\theta$, coarse-scale observations $y$, and fine-scale observations $\z$ are known, then the \textit{forward-backward} algorithm can be used to calculate the hidden state probabilities $\Pr(X_{t} = i|Y=y,\Z=\z)$ and $\Pr(\tilde X^*_{t,t^*} = i^*|Y=y,\Z=\z)$ for all $t$, $t^*$, $i = 1,\ldots,N$, and $i^* = 1,\ldots,N^*$ \citep{Zucchini:2016}.
+However, since $\theta$ is rarely known in practice, the maximum likelihood estimate $\hat \theta$ is often used instead to estimate the hidden state probabilities.
+As such, if $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define the \textit{average dive decoding accuracy} as the average value of the estimates of $\Pr(X_{t} = x_t|Y=y,\Z=\z)$ across all simulated data sets and dives.
+Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the \textit{average subdive decoding accuracy} as the average value of the estimates of $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y=y,\Z=\z)$ across all simulated data sets, dives, and windows.
+We find both the average dive decoding accuracy and the average subdive decoding accuracy using the maximum likelihood estimates $\hat \theta$ for each model and training set, the observations $y$ and $\z$ of each test set, and the forward-backward algorithm.
 
 \subsection{Simulation results}
-
-We use several metrics to compare the candidate models, including accuracy when decoding dive types and subdive states (see Table \ref{table:accuracy}).
-If $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define average dive decoding accuracy as the average value of $\Pr(X_{t} = x_{t}|Y,Y^*)$ across all simulated data sets and dives.
-Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the average subdive decoding accuracy as $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y,Y^*)$ averaged across all simulated data sets, dives, and windows.
-We find both the average dive decoding accuracy and the average subdive decoding accuracy using the \textit{forward-backward} algorithm \citep{Zucchini:2016}.
 

### 2020-12-29 Report/Sections/discussion.tex
Nancy Comments Bibliography

@@
 We demonstrate the usefulness of such a model using an ecological example, where we use a complex HMM to classify the coarse- and fine-scale diving behaviour of a killer whale.
-This gives a deeper understanding of the animal's tri-axial movement and thus its fine-scale energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both NRKWs and their prey \citep{Noren:2011}.
+Our analysis gives a deeper understanding of the animal's tri-axial movement and thus its fine-scale energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both NRKWs and their prey \citep{Noren:2011}.
 Although our case study focuses on northern resident killer whales, this model is likely applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.

### 2020-12-29 Report/Sections/appendixnh.tex
Nancy comments- Appendix

@@
 
-{\bf{I didn't like the equation numbering restarting at (1), so I used ``setcounter".}}
+%{\bf{I didn't like the equation numbering restarting at (1), so I used ``setcounter".}}
 
 We easily simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
-To simulate data at the fine scale 
-for dive $t$, recall that the fine scale model is based on a sequence of $\tilde T_t^*$ two-second windows, each containing 100 values of raw acceleration.  
- We easily generate the fine-scale hidden Markov process,  $\tilde X ^* _{t,\tilde t^*}$,
-$\tilde t^*=1,\ldots,\tilde T_t^*$, that determines the sequence of underlying states in each window according to the specified parameters in $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$.
- However, generating the raw acceleration data is not straightforward, since our model is formulated in terms of quantities derived from the raw data in each window, namely the average acceleration, $A_{t,\tilde t^*}$, and the wiggliness, $W_{ t,\tilde t^*}$.
+To simulate data at the fine scale for dive $t$, recall that the fine scale model is based on a sequence of $\tilde T_t^*$ two-second windows, each containing 100 values of raw acceleration.
+We easily generate the fine-scale hidden Markov process, $\tilde X ^* _{t,\tilde t^*}$, for $\tilde t^*=1,\ldots,\tilde T_t^*$, that determines the sequence of underlying states corresponding to each window according to the specified parameters in $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$.
+However, generating the raw acceleration data is not straightforward, since our model is formulated in terms of quantities derived from the raw data in each window, namely the average acceleration, $\Zone_{t,\tilde t^*}$, and the wiggliness, $\Ztwo_{ t,\tilde t^*}$.
 In our simulation study, we generate raw acceleration data so that we can visualize our results in terms of the underlying data curves for each dive.
-Here, we explain how we generate the acceleration curves  so that the derived quantities, $A_{t,\tilde t^*}$ and $W_{ t,\tilde t^*}$, follow the specified model.
-A key component is the discrete Fourier transformation of the  100 raw acceleration values in the $\tilde{t}^*$th window of dive $t$:
+Here, we explain how we generate the acceleration curves so that the derived quantities, $\Zone_{t,\tilde t^*}$ and $\Ztwo_{ t,\tilde t^*}$, follow the specified model.
+A key component is the discrete Fourier transformation of the 100 raw acceleration values in window $\tilde{t}^*$ of dive $t$:
 \[
-\hat{Y}^{*(k)}_{t,\tilde{t}^*}  \equiv DFT(Y^*_{t,100 (\tilde{t}^*-1) + 1 },\ldots, Y^*_{t,100 \tilde{t}^*})[k]
+    \hat{Y}^{*(k)}_{t,\tilde{t}^*}  \equiv DFT\left\{Y^*_{t,100 (\tilde{t}^*-1) + 1 },\ldots, Y^*_{t,100 \tilde{t}^*}\right\}(k)
 \]
-for $k \ge 0$, as defined in Equation \ref{eq:DFTdef}.
+for $k \geq 0$, as defined in Equation (\ref{eq:DFTdef}).
 
@@
 
-For step (1), we generate $\hat Y^{*(0)}_{t,1}, \ldots, \hat Y^{*(0)}_{t,\tilde T_t^*}$  as a CARHMM with underlying Markov state sequence $X^{*}_{t,1}, \ldots, X^{*}_{t,\tilde T_t^*}$ and random first emission.
+For step (1), we generate $\hat Y^{*(0)}_{t,1}, \ldots, \hat Y^{*(0)}_{t,\tilde T_t^*}$ as a CarHMM with underlying Markov state sequence $\tilde X^{*}_{t,1}, \ldots, \tilde X^{*}_{t,\tilde T_t^*}$ and random first emission.
 Specifically, we let
@@
     \end{equation}
-    {\bf{WHY IS THERE AN $i$ IN THE ABOVE?}}
-     where the $a^{(k)}_{t,\tilde t^*}$s are independent and equal to either 1 or -1 each with probability 1/2.
+    where the $a^{(k)}_{t,\tilde t^*}$s are independent and equal to either 1 or -1 each with probability 1/2 and $i$ is the imaginary number.
+We include $i$ in Equation (\ref{eqn:abYhat}) to force all variation within a window to take the form of a sine wave, which reduces the variation between the endpoints of windows compared to a cosine wave.
     Given the fine scale states, the $b^{(k)}_{t,\tilde t^*}$s are independent  and independent of the  $a^{(k)}_{t,\tilde t^*}$s.  
@@
 
-We now show that this construction of the raw acceleration data  results in the distributions listed in Section \ref{subsec:data_simulation}.
+We now show that this setup of the raw acceleration data results in the distributions listed in Section \ref{subsec:data_simulation}.
 It suffices to show that the construction of the discrete Fourier transformations, the $\hat{Y}^{(k)}_{t,\tilde t^*}$s, yields the desired distributions.
@@
 $\Zone_{t,1}, \ldots, \Zone_{t,\tilde{T}^*}$ 
-follows a CARHMM with Normal emissions distributions and parameters 
+follows a CarHMM with Normal emissions distributions and parameters 
  $\mu_A^{*(\cdot,1)} = \mu_A^{*(\cdot,2)} = \mu_A^{*(\cdot,i^*)} = 0$,  $\sigma_A^{*(\cdot,1)} = 0.034s$, $\phi_A^{*(\cdot,1)} = 0.98$, $\sigma_A^{*(\cdot,2)} = 0.079s$, and $\phi_A^{*(\cdot,2)} = 0.87$.
 
- From Equations \ref{eqn:z} and \ref{eqn:abYhat}, 
+ From Equations (\ref{eqn:z}) and (\ref{eqn:abYhat}), 
  the  wiggliness in window $\tilde t^*$  of dive $t$ is 
 \[
-    \Ztwo_{t,\tilde t^*} = \sum_{k=1}^{\tilde \omega} \big|\big|\hat{Y}^{(k)}_{t,\tilde t^*}\big|\big|^2 =  \sum_{k=1}^{\tilde \omega} b^{(k)}_{t,\tilde t^*}
+    \Ztwo_{t,\tilde t^*} = \sum_{k=1}^{\tilde \omega} \big|\big|\hat{Y}^{(k)}_{t,\tilde t^*}\big|\big|^2 =  \sum_{k=1}^{\tilde \omega} b^{(k)}_{t,\tilde t^*}.
 %    \label{eqn:Wb}
@@
 If $\tilde \omega < 50$, then $\Ztwo_{t,\tilde t^*}$ is the sum of independent Gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,\tilde t^*}$ is also Gamma.
-Thus, by Equation  \ref{eqn:bdist}
+Thus, by Equation (\ref{eqn:bdist})
 %
@@
 \]
-Setting $\tilde \omega$ to 10 and carrying out simple calculation of the mean and variance of a Gamma distribution yield $\mu_W^{*(\cdot,1)} = 23.3$, $\sigma_W^{*(\cdot,1)} = 13.0$, $\mu_W^{*(\cdot,2)} = 301.2$, and $\sigma_W^{*(\cdot,2)} = 330.1$.
+Setting $\tilde \omega$ to 10 and carrying out simple calculation of the mean and variance of a Gamma distribution yields $\mu_W^{*(\cdot,1)} = 23.3$, $\sigma_W^{*(\cdot,1)} = 13.0$, $\mu_W^{*(\cdot,2)} = 301.2$, and $\sigma_W^{*(\cdot,2)} = 330.1$.
 

### 2020-12-29 Report/Sections/intro.tex
Read over through 3.5

@@
 The study of animal movement in particular has been transformed by tracking devices that record kinematic information in a variety of environments \citep{Borger:2020,Dot:2016b}.
-Tags can record observations at over 50 times per second, resulting in time series containing millions of observations over the course of several hours.
+Tags can record over 50 observations per second, resulting in time series containing millions of observations over the course of several hours.
 These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a challenge for statisticians.
 
-Biologging data are frequently modelled as a set of curves analyzed by methods of Functional Data Analysis, or FDA \citep{Ramsay:2005}.
+Biologging data are frequently modelled as a set of curves analyzed by methods of Functional Data Analysis, or FDA -- see, e.g., \citet{Ramsay:2005}.
 For example, \citet{Morris:2007} view a child's activity level as a set of daily curves where metabolic activity is a function of time.
@@
 FDA was originally developed to process curves assumed to be independent replicates, and within-curve fine-scale structure is not usually incorporated in FDA models.
+
 However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
@@
 
-Some FDA models account for between-curve dependence occurring when multiple curves arise from separate groups of individuals. 
-Previous studies have used multilevel models with random effects to model variation between and within individuals, for example, in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}.
+Some FDA models account for between-curve dependence occurring when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain types of dependence in time.
+For example, previous studies have used multilevel models with random effects to model variation between and within individuals in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}.
 More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}.
 However, these multilevel models are not appropriate in many biologging applications since they do not account for different curve types.
-In addition, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence is not appropriate in most biologging applications.
+Further, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence is not appropriate in most biologging applications.
+In addition to multi-level models, FDA researchers have also used functional time series to model dependence in a sequence of curves.
+Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by well-defined hidden states.
 
-FDA researchers have used functional time series to model dependence in a sequence of curves.
-Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by well-defined hidden states.
-%
 Traditional FDA techniques similarly fall short when modelling complicated within-curve %acceleration 
@@
 %and thus extend the methods available for the analysis of functional data. 
-We begin by describing HMMs as well as two variants, CarHMMs and HHMMs, and discuss how Fourier analysis can handle fine-scale dependence structures.
+We begin in Section 2 by describing HMMs as well as two variants, CarHMMs and HHMMs, and discuss how Fourier analysis can handle fine-scale dependence structures.
 We also show how these methods can be combined to analyze increasingly complex data.

### 2020-12-29 Report/Sections/models.tex
Read over through 3.5

@@
 Consider a sequence of $T$ curves, where any particular curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$.
-Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up fine-scale quantities derived from curve $t$ and indexed using $t^*$.
-Both $Y_t$ and $Y_{t,t^*}$ can be either vectors or scalars.
+Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up fine-scale quantities derived from curve $t$ indexed using $t^*$.
+Both $Y_t$ and $Y^*_{t,t^*}$ can be either vectors or scalars.
 We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.
@@
 %
-We assume that $X_1$ follows the chain's stationary distribution, which is denoted by an N-dimensional row vector $\delta$, where
+We assume that $X_1$ follows the chain's stationary distribution, which is denoted by an $N$-dimensional row vector $\delta$, where
 $\delta_i = \Pr(X_1 = i).$
@@
 A conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process.
-Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots,X_T$ that can take on integer values between $1$ and $N$.
+Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots,X_T$ that can take integer values between $1$ and $N$.
 A CarHMM also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
@@
 For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
-If $X_t = i$, then the components of $X_t^*$ follow a Markov chain with $N^{*(i)}$ possible states, an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$, and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
-The distribution of $Y_{t,t^*}$ given $Y_{t,t^*-1} = y_{t,t^*-1}$, $X_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$.
+If $X_t = i$, then the components of $X_t^*$ make up a Markov chain with $N^{*(i)}$ possible states, an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$, and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
+The distribution of $Y^*_{t,t^*}$ given $Y^*_{t,t^*-1} = y^*_{t,t^*-1}$, $X^*_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$.
 We denote the set of fine-scale emission parameters corresponding to $X_t=i$ as $\theta^{*(i)}=\left\{\theta^{*(i,1)}, \ldots, \theta^{*\left(i,N^{*(i)}\right)}\right\}$.
@@
 where $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix whose exact structure depends upon the coarse- and fine-scale models.
-If the coarse-scale model is an HMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ both have $(i,i)^{th}$ entries equal to
+If the coarse-scale model is an HMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ for $t \geq 2$ both have $(i,i)^{th}$ entries equal to
 $f^{(i)}(y_t)\calL_{\text{fine}}\left(\theta^{*(i)},
@@
 If the coarse-scale model is a CarHMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ has $(i,i)^{th}$ entry equal to $\calL_{\text{fine}}\left(\theta^{*(i)},
-\Gamma^{*(i)};y_1^*\right)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ has $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1})\calL_{\text{fine}}\left(\theta^{*(i)},
+\Gamma^{*(i)};y_1^*\right)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ for $t \geq 2$ has $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1})\calL_{\text{fine}}\left(\theta^{*(i)},
 \Gamma^{*(i)};y_t^*\right)$.
@@
 Subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window.
-Stride length refers to the distance between the first element of consecutive windows, so a stride length of $h$ implies that the first window starts at $Y^*_{t^*,1}$, the second at $Y^*_{t^*,h+1}$, and so on.
+Stride length refers to the distance between the first element of consecutive windows, so a stride length of $h$ implies that the first window starts at $Y^*_{t,1}$, the second at $Y^*_{t,h+1}$, and so on.
 Larger stride lengths result in a loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting.
@@
 
-If the stride length $h$ is greater than one, then the fine-scale HMM (or CarHMM) must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself.
+When performing this transformation, the fine-scale HMM (or CarHMM) must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself.
 In particular, there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
@@
 To clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT.
-Figure \ref{fig:models}d displays the dependence structure for an HMM-DFT.
+Figure \ref{fig:models}d displays the dependence structure of an HMM-DFT.
 
@@
 For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations of a Poisson process as their fine-scale model.
-If the fine-scale model is a simple Poisson process, then this is equivalent to a Markov-modulated Poisson process \citep{Fischer:1993}.
-The fine-scale structure can also be modelled as in \citep{Langrock:2018}, who use B-splines to model the emission distribution of an HMM as well as incorporate covariates.
-This non-parametric approach uses a penalized likelihood term which can easily replace the usual fine-scale likelihood term in (\ref{eqn:HHMM_likelihood}).
-Another class of fine-scale models is the set of continuous time methods such as a continuous-time HMM (CTHMM) \citep{Liu:2015} or a state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019}.
+If the fine-scale model is a simple Poisson process, then this approach is equivalent to a Markov-modulated Poisson process \citep{Fischer:1993}.
+The fine-scale process can also be modelled as in \citep{Langrock:2018}, who use B-splines to model the emission distribution of an HMM as well as incorporate covariates.
+This non-parametric approach uses a penalized likelihood term which can easily replace the usual fine-scale likelihood term in Equation (\ref{eqn:HHMM_likelihood}).
+Another class of fine-scale models is the set of continuous time methods such as a continuous-time HMM (CTHMM) \citep{Liu:2015} or a state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}. 
 A continuous-time HMM may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}.

### 2020-12-29 Report/Sections/results.tex
Read over through 3.5

@@
 Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}.
-HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g.
-\citet{DeRuiter:2017} and \citet{Isojunno:2017}).
+HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., \citet{DeRuiter:2017} and \citet{Isojunno:2017}).
 For killer whales, we are interested in categorizing different diving behaviours to identify potential foraging dives.
@@
 Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but the animal's behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}.
-Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of a killer whale.
+Therefore, understanding both the behavioural state of the killer whale and the distribution of acceleration within each behavioural state is needed to determine the true energetic requirements of the animal.
 
@@
 Observations were collected at a rate of 50 Hz using a CATs biologger (Customizable Animal Tracking Solutions, www.cats.is).
-The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, and right/left).
+Acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, and right/left).
 Tri-axial acceleration readings are common in these types of tags and are often used to infer animal behaviour such as foraging \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
@@
 
-Defining a suitable model to describe this killer whale kinematic data involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse and fine scale observations.
+Defining a suitable model to describe this killer whale kinematic data involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse- and fine-scale observations.
 
@@
 Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material).
-Therefore, we choose a CarHMM as defined in Section 2.2 for the fine-scale model.
+Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.
 
@@
 This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.
-
 To specify the emission distribution of $\Zone_{t,\tilde t^*}$, consider the sequence $\Zone_{t,1},\Zone_{t,2},\ldots,\Zone_{t,T^*_t}$ for a particular dive $t$.
@@
 The emission distribution of each subdive state does not depend upon the dive type.
+
 In total, the parameters to estimate are
@@
 The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
-Dive type 1 corresponds to shorter, shallower dives which likely reflects resting, travelling, and to a lesser extent searching for prey.
+Dive type 1 corresponds to shorter, shallower dives which likely reflect resting, travelling, and to a lesser extent searching for prey.
 Dive type 2 is longer, deeper, and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives.
@@
 We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}.
-Each observation of dive duration is weighted by the probability that it corresponds to a particular dive type as decoded by the \textit{forward-backward algorithm} \citep{Zucchini:2016}.
+Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the \textit{forward-backward algorithm} \citep{Zucchini:2016}.
 This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2.

### 2020-12-30 Report/Sections/results.tex
read through main text

@@
 The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of the supplementary material).
+This suggests that including auto-correlation in the model significantly affects parameter estimates.
 Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of the supplementary material).
@@
 
-Finally, the CarHMM-DFT, which lacks a hierarchical structure, produces fine-scale parameter estimates and state estimates similar to those of the CarHHMM-DFT.
+Finally, the CarHMM-DFT, which lacks a hierarchical structure, produces fine-scale parameter estimates and subdive state estimates similar to those of the CarHHMM-DFT.
 However, its lack of hierarchical structure means that it fails to differentiate between short and long dives.

### 2020-12-30 Report/Sections/simulation.tex
read through main text

@@
 
-\subsection{Data simulation}
+\subsection{Simulation procedure}
 \label{subsec:data_simulation}
@@
 It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix.
-Figure 21 of the supplementary material shows one realization of $\Z$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.
+Figure 21 of the supplementary material shows one realization of $\Z$ for five dives of one simulated data set along with the corresponding reconstructed realization of $Y^*$.
 
@@
 
-We calculate maximum likelihood estimates $\hat \theta$ for all four candidate models for each of the 500 training data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
-We then generate a corresponding independent test data set consisting of an additional 100 dives from the generative model with the true parameters $\theta$.
+We calculate maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 training data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
+We then generate a corresponding independent test data set consisting of an additional 100 dives from the generative model with the true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
 
-If the parameters $\theta$, coarse-scale observations $y$, and fine-scale observations $\z$ are known, then the \textit{forward-backward} algorithm can be used to calculate the hidden state probabilities $\Pr(X_{t} = i|Y=y,\Z=\z)$ and $\Pr(\tilde X^*_{t,t^*} = i^*|Y=y,\Z=\z)$ for all $t$, $t^*$, $i = 1,\ldots,N$, and $i^* = 1,\ldots,N^*$ \citep{Zucchini:2016}.
-However, since $\theta$ is rarely known in practice, the maximum likelihood estimate $\hat \theta$ is often used instead to estimate the hidden state probabilities.
+If the parameters, coarse-scale observations $y$, and fine-scale observations $\z$ are known, then the \textit{forward-backward} algorithm can be used to calculate the hidden state probabilities $\Pr(X_{t} = i|Y=y,\Z=\z)$ and $\Pr(\tilde X^*_{t,t^*} = i^*|Y=y,\Z=\z)$ for all $t$, $t^*$, $i = 1,\ldots,N$, and $i^* = 1,\ldots,N^*$ \citep{Zucchini:2016}.
+However, since the true parameters rarely known in practice, the maximum likelihood estimates are often used instead to estimate the hidden state probabilities.
 As such, if $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define the \textit{average dive decoding accuracy} as the average value of the estimates of $\Pr(X_{t} = x_t|Y=y,\Z=\z)$ across all simulated data sets and dives.
 Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the \textit{average subdive decoding accuracy} as the average value of the estimates of $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y=y,\Z=\z)$ across all simulated data sets, dives, and windows.
-We find both the average dive decoding accuracy and the average subdive decoding accuracy using the maximum likelihood estimates $\hat \theta$ for each model and training set, the observations $y$ and $\z$ of each test set, and the forward-backward algorithm.
+We find both the average dive decoding accuracy and the average subdive decoding accuracy using the maximum likelihood estimates for each model and training set, the observations $y$ and $\z$ of each test set, and the forward-backward algorithm.
 
@@
 The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model.
-Its average dive decoding accuracy and the average subdive decoding accuracy are both greater than 0.9.
+Its average dive decoding accuracy and average subdive decoding accuracy are both greater than 0.9.
 All parameter estimates of fine-scale mean values and probability transition matrices, $\hat \mu^*$, $\hat \Gamma^*$ and $\hat \Gamma$, respectively, have statistically insignificant biases.
@@
 The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors.
-This small bias is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to twice as large as the estimated standard error.
+This underestimation is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to twice as large as the estimated standard error.
 See Tables 5 through 9 of the supplementary material for detailed results.
@@
 As expected, the CarHHMM is the least accurate model when predicting subdive state.
-The CarHHMM-DFT and HHMM-DFT yield similar estimates of dive type while the CarHHMM is less accurate and even misclassifies the second dive.
+The CarHHMM-DFT and HHMM-DFT yield similar estimates of dive type while the CarHHMM is less accurate and even misclassifies the dive type of the second dive.
 The CarHMM-DFT does not estimate dive type.

### 2020-12-30 Report/Sections/appendixnh.tex
Read over Appendix

@@
 We easily simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
-To simulate data at the fine scale for dive $t$, recall that the fine scale model is based on a sequence of $\tilde T_t^*$ two-second windows, each containing 100 values of raw acceleration.
-We easily generate the fine-scale hidden Markov process, $\tilde X ^* _{t,\tilde t^*}$, for $\tilde t^*=1,\ldots,\tilde T_t^*$, that determines the sequence of underlying states corresponding to each window according to the specified parameters in $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$.
-However, generating the raw acceleration data is not straightforward, since our model is formulated in terms of quantities derived from the raw data in each window, namely the average acceleration, $\Zone_{t,\tilde t^*}$, and the wiggliness, $\Ztwo_{ t,\tilde t^*}$.
+For each dive $t$, we also easily generate the fine-scale hidden Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T_t^*}\right\}$ according to one of the probability transition matrices $\Gamma^{*(1)}$ or $\Gamma^{*(2)}$, depending upon the value of $X_t$.
+This determines the sequence of fine-scale hidden states corresponding to each window.
+Recall that the fine-scale model is based on a sequence of $\tilde T_t^*$ two-second windows, each containing 100 observations, and our model is formulated in terms of quantities derived from the raw data within each window (namely, the average acceleration, $\Zone_{t,\tilde t^*}$, and wiggliness, $\Ztwo_{ t,\tilde t^*}$).
+Generating the raw acceleration data from $\Zone_{t,\tilde t^*}$ and $\Ztwo_{ t,\tilde t^*}$ is not straightforward.
 In our simulation study, we generate raw acceleration data so that we can visualize our results in terms of the underlying data curves for each dive.
-Here, we explain how we generate the acceleration curves so that the derived quantities, $\Zone_{t,\tilde t^*}$ and $\Ztwo_{ t,\tilde t^*}$, follow the specified model.
+Here, we explain how we generate the acceleration curves so that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{ t,\tilde t^*}$ both follow the specified model.
 A key component is the discrete Fourier transformation of the 100 raw acceleration values in window $\tilde{t}^*$ of dive $t$:
@@
     \{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\} \equiv IDFT\left\{\hat{Y}^{*(0)}_{t,\tilde t^*},
-    \ldots \hat{Y}^{*(99)}_{t,\tilde t^*}\right\}
+    \ldots, \hat{Y}^{*(99)}_{t,\tilde t^*}\right\}
     {\rm{~for~}} \tilde t^* = 1,\ldots,\tilde T^*_t.
@@
     \end{equation}
-    where the $a^{(k)}_{t,\tilde t^*}$s are independent and equal to either 1 or -1 each with probability 1/2 and $i$ is the imaginary number.
+    where the $a^{(k)}_{t,\tilde t^*}$'s are independent and equal to either 1 or -1 each with probability 1/2 and $i$ is the imaginary number.
 We include $i$ in Equation (\ref{eqn:abYhat}) to force all variation within a window to take the form of a sine wave, which reduces the variation between the endpoints of windows compared to a cosine wave.
-    Given the fine scale states, the $b^{(k)}_{t,\tilde t^*}$s are independent  and independent of the  $a^{(k)}_{t,\tilde t^*}$s.  
+    Given the fine scale states, the $b^{(k)}_{t,\tilde t^*}$'s are independent and independent of the  $a^{(k)}_{t,\tilde t^*}$'s.  
    The distribution of $b^{(k)}_{t,\tilde t^*}$ is
@@
 \]
-   This construction guarantees that the inverse discrete Fourier transform is real-valued.
+   This guarantees that the inverse discrete Fourier transform is real-valued.
 
-We now show that this setup of the raw acceleration data results in the distributions listed in Section \ref{subsec:data_simulation}.
-It suffices to show that the construction of the discrete Fourier transformations, the $\hat{Y}^{(k)}_{t,\tilde t^*}$s, yields the desired distributions.
+We now show that this construction of the raw acceleration data results in the distributions listed in Section \ref{subsec:data_simulation}.
+It suffices to show that the construction of the discrete Fourier transformations, the $\hat{Y}^{(k)}_{t,\tilde t^*}$'s, yields the desired distributions.
 
-First, since $\hat{Y}^{(0)}_{t,\tilde t^*} = \sum_{n=1}^{100} Y^*_{t,100(\tilde t^* - 1) + n} $ $=\Zone_{t,\tilde{t}^*}$, Equation (\ref{eqn:yhat_0}) implies that 
+First, since $\hat{Y}^{(0)}_{t,\tilde t^*} = \sum_{n=1}^{100} Y^*_{t,100(\tilde t^* - 1) + n} $ $=100\Zone_{t,\tilde{t}^*}$, Equation (\ref{eqn:yhat_0}) implies that 
 $\Zone_{t,1}, \ldots, \Zone_{t,\tilde{T}^*}$ 
@@
  From Equations (\ref{eqn:z}) and (\ref{eqn:abYhat}), 
- the  wiggliness in window $\tilde t^*$  of dive $t$ is 
+ the wiggliness within window $\tilde t^*$ of dive $t$ is 
 \[
@@
 \begin{align*}
-P(y_t,\z_t;\theta,\theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\theta^{(1)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(1)};\z_t\right), \ldots , \\
-&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(N)};\z_t\right) \Big],
+P(y_t,\z_t;\theta,\theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\theta^{(1)})\calL_{\text{fine}}\left(\theta^*,\Gamma^{*(1)};\z_t\right), \ldots , \\
+&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{fine}}\left(\theta^*,\Gamma^{*(N)};\z_t\right) \Big],
 \end{align*}
@@
 where $f^{(i)}(y_t;\theta^{(i)})$ is the emission distribution of dive duration given that $X_t = i$.
-The likelihood $\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale model and is equal to the following:
+The likelihood $\calL_{\text{fine}}$ corresponds to the fine-scale model and is equal to the following:
 %
-$$\calL_{\text{CarHMM-DFT}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},$$
+$$\calL_{\text{fine}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},$$
 %

### 2020-12-31 Report/Sections/simulation.tex
Fix wording in section 4.1

@@
 
-We generate 500 independent data sets using the CarHHMM-DFT as a generative model.
-Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives.
+We generate 500 independent training data sets using the CarHHMM-DFT as a generative model.
+Each training data set consists of a sequence of 100 curves which we call a sequence of killer whale dives.
 Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
@@
 
-We calculate maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 training data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per model.
-We then generate a corresponding independent test data set consisting of an additional 100 dives from the generative model with the true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
+We calculate maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per data set.
 
-If the parameters, coarse-scale observations $y$, and fine-scale observations $\z$ are known, then the \textit{forward-backward} algorithm can be used to calculate the hidden state probabilities $\Pr(X_{t} = i|Y=y,\Z=\z)$ and $\Pr(\tilde X^*_{t,t^*} = i^*|Y=y,\Z=\z)$ for all $t$, $t^*$, $i = 1,\ldots,N$, and $i^* = 1,\ldots,N^*$ \citep{Zucchini:2016}.
+For each of the 500 training data sets, we also simulate a test data set which consists of 100 dives and is created from the generative model with true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
+If the parameters of the generative model, coarse-scale test-set observations $y$, and fine-scale test-set observations $\z$ are known, then the \textit{forward-backward} algorithm can be used to calculate the hidden state probabilities $\Pr(X_{t} = i|Y=y,\Z=\z)$ and $\Pr(\tilde X^*_{t,t^*} = i^*|Y=y,\Z=\z)$ for all $t$, $t^*$, $i = 1,\ldots,N$, and $i^* = 1,\ldots,N^*$ \citep{Zucchini:2016}.
 However, since the true parameters rarely known in practice, the maximum likelihood estimates are often used instead to estimate the hidden state probabilities.
@@
 Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the \textit{average subdive decoding accuracy} as the average value of the estimates of $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y=y,\Z=\z)$ across all simulated data sets, dives, and windows.
-We find both the average dive decoding accuracy and the average subdive decoding accuracy using the maximum likelihood estimates for each model and training set, the observations $y$ and $\z$ of each test set, and the forward-backward algorithm.
+We find both the average dive decoding accuracy and the average subdive decoding accuracy using the maximum likelihood estimates from each training set as well as the observations from each test set in conjunction with the forward-backward algorithm.
 

### 2020-12-31 Report/main.tex
Nancy edits through 4.1
Nancy Comments Bibliography
Read over through 3.5
End of day Dec 30

@@
 %\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
-\title[]{Modelling state-switching functional data with hidden Markov models}
+\title[]{Modelling state-switching functional and biologging data with hidden Markov models}
 \author{BlindedA\authorref{1}\thanksref{*}}
@@
 \KWDtitle{Key words and phrases}
-Biologging\sep statistical ecology\sep animal movement\sep accelerometer data\sep diving behaviour\sep killer whales \sep multi-state modelling\sep hierarchical modelling\sep time series.
+Accelerometer data\sep animal movement\sep biologging\sep diving behaviour\sep hierarchical modelling\sep killer whales\sep multi-state modelling\sep statistical ecology\sep time series.
 % MSC 2010 subject classification codes
@@
 \begin{ack}{ACKNOWLEDGEMENTS}
-%This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
+% This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
+% We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC). 
+% Marie Auger M'eth'e thanks the NSERC Discovery and the Canadian Research Chair programs.
+% Evan Sidrow thanks the University of British Columbia and the Four-Year Doctoral Fellowship program.
 \end{ack}
@@
 \begin{appendix}
-\input Sections/appendix.tex
+%\input Sections/appendix.tex
+\input Sections/appendixnh.tex
 \end{appendix}

### 2021-01-04 Report/Sections/simulation.tex
Nancy edits 4.1

@@
 
-For each of the 500 training data sets, we also simulate a test data set which consists of 100 dives and is created from the generative model with true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
+For each of the 500 training data sets, we simulate a test data set to assess how well each model predicts the hidden states, as follows.
+Each test data set consists of a sequence of 100 dives and is created from the generative model with true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
+To assess the coarse-scale prediction, we estimate $p_t(i|y,\z) \equiv \Pr(X_t=i|Y=y,\Z=\z)$, $i=1,2$, $t=1,\ldots,100$ using the test-set observations $(y,\z)$ and training-set maximum likelihood estimates.
+These estimates are found using the \em{forward-backward} algorithm \citep{Zucchini:2016}.
+We compare these estimated conditional probabilities to $x_1,\ldots,x_{100}$, the true coarse-scale state realizations in the test data, by calculating the {\em{average dive decoding accuracy}} for a single training/test data set pair, $\sum_{t=1}^{100} \hat{p}(x_t|y,\z)/100$.
+We then report the average of these over the 500 training/test data set pairs.
+Analogously, to assess prediction of the fine-scale states, we estimate $p^*_{t,\tilde t^*}(i^*|y,\z) \equiv \Pr(\tilde X^*_{t,\tilde t^*}=i|Y=y,\Z=\z)$, $i^*=1,2$, $\tilde t^* = 1,\ldots,\tilde T^*_t$, $t=1,\ldots,100$, using the test-set observations, the training-set maximum likelihood estimates, and the forward-backward algorithm.
+Denoting the true fine-scale state realizations from the test data set as $\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}$, we define the overall \textit{average subdive decoding accuracy} as the average value of $p^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*}|y,\z)$ across all simulated test data sets, dives, and windows.
+The conditional probabilities are estimated according to one of the four models under study, using the maximum likelihood estimates from the training data set in conjunction with the forward-backward algorithm.
+
+%For each of the 500 training data sets, we also simulate a test data set which consists of 100 dives and is created from the generative model with true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
 If the parameters of the generative model, coarse-scale test-set observations $y$, and fine-scale test-set observations $\z$ are known, then the \textit{forward-backward} algorithm can be used to calculate the hidden state probabilities $\Pr(X_{t} = i|Y=y,\Z=\z)$ and $\Pr(\tilde X^*_{t,t^*} = i^*|Y=y,\Z=\z)$ for all $t$, $t^*$, $i = 1,\ldots,N$, and $i^* = 1,\ldots,N^*$ \citep{Zucchini:2016}.

### 2021-01-04 Report/main.tex
Nancy edits 4.1

@@
 %\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
-\title[]{Modelling state-switching functional and biologging data with hidden Markov models}
+\title[]{Modelling multi-scale functional and biologging data with hidden Markov models}
 \author{BlindedA\authorref{1}\thanksref{*}}
@@
 \KWDtitle{Key words and phrases}
-Accelerometer data\sep animal movement\sep biologging\sep diving behaviour\sep hierarchical modelling\sep killer whales\sep multi-state modelling\sep statistical ecology\sep time series.
+Accelerometer data\sep animal movement\sep biologging\sep diving behaviour\sep hierarchical modelling\sep killer whales\sep state-switching\sep statistical ecology\sep time series.
 % MSC 2010 subject classification codes

### 2021-01-06 Report/Sections/models.tex
add authors

@@
 The distribution of $Y^*_{t,t^*}$ given $Y^*_{t,t^*-1} = y^*_{t,t^*-1}$, $X^*_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$.
-We denote the set of fine-scale emission parameters corresponding to $X_t=i$ as $\theta^{*(i)}=\left\{\theta^{*(i,1)}, \ldots, \theta^{*\left(i,N^{*(i)}\right)}\right\}$.
+We denote the set of fine-scale emission parameters corresponding to $X_t=i$ as $\theta^{*(i)}=\big\{\theta^{*(i,1)}, \ldots, \theta^{*\left(i,N^{*(i)}\right)}\big\}$.
 In summary:
@@
 Let $y$ be the sequence of observed coarse-scale emissions and
-$y^* \equiv \left\{y^*_1, \ldots,y^*_T\right\}$ be the collection of $T$ observed fine-scale emission vectors.
+$y^* \equiv \{y^*_1, \ldots,y^*_T\}$ be the collection of $T$ observed fine-scale emission vectors.
 %

### 2021-01-06 Report/Sections/results.tex
add authors

@@
 Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
-Specifically, the sequence of observed coarse-scale observations $y = \left\{y_1,\ldots,y_{267}\right\}$ is a time series of dive durations in seconds.
-For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \left\{y^*_{t,1},\ldots,y^*_{t,T^*_t} \right\}$, which is a sequence containing the within-dive acceleration data in units of meters per second squared.
-The collection of all acceleration data is denoted as $y^* = \left\{y^*_1,\ldots,y^*_{267}\right\}$.
+Specifically, the sequence of observed coarse-scale observations $y = \big\{y_1,\ldots,y_{267}\big\}$ is a time series of dive durations in seconds.
+For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \big\{y^*_{t,1},\ldots,y^*_{t,T^*_t} \big\}$, which is a sequence containing the within-dive acceleration data in units of meters per second squared.
+The collection of all acceleration data is denoted as $y^* = \big\{y^*_1,\ldots,y^*_{267}\big\}$.
 

### 2021-01-06 Report/Sections/simulation.tex
small edits
add authors

@@
 After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored.
-Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \right\}$ with probability transition matrices
+Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \big\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \big\}$ with probability transition matrices
 %
@@
 for dive types 1 and 2, respectively.
-Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$.
-Recall from Section \ref{subsec:model_selection} that we must specify the mean, standard deviation, and auto-correlation parameters corresponding to $\left\{\Zone_{t,1},\ldots,\Zone_{t,\tilde T_t^*}\right\}$ as well as the mean and standard deviation parameters corresponding to $\left\{\Ztwo_{t,1},\ldots,\Ztwo_{t,\tilde T_t^*}\right\}$.
+Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \big\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\big\}$.
+Recall from Section \ref{subsec:model_selection} that we must specify the mean, standard deviation, and auto-correlation parameters corresponding to $\big\{\Zone_{t,1},\ldots,\Zone_{t,\tilde T_t^*}\big\}$ as well as the mean and standard deviation parameters corresponding to $\big\{\Ztwo_{t,1},\ldots,\Ztwo_{t,\tilde T_t^*}\big\}$.
 We select the following parameters in line with the results from the case study:
@@
 Each test data set consists of a sequence of 100 dives and is created from the generative model with true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
-To assess the coarse-scale prediction, we estimate $p_t(i|y,\z) \equiv \Pr(X_t=i|Y=y,\Z=\z)$, $i=1,2$, $t=1,\ldots,100$ using the test-set observations $(y,\z)$ and training-set maximum likelihood estimates.
-These estimates are found using the \em{forward-backward} algorithm \citep{Zucchini:2016}.
-We compare these estimated conditional probabilities to $x_1,\ldots,x_{100}$, the true coarse-scale state realizations in the test data, by calculating the {\em{average dive decoding accuracy}} for a single training/test data set pair, $\sum_{t=1}^{100} \hat{p}(x_t|y,\z)/100$.
+To assess the coarse-scale hidden state prediction, we estimate $p_t(i|y,\z) \equiv \Pr(X_t=i|Y=y,\Z=\z)$, $i=1,2$, $t=1,\ldots,100$ using the test-set observations $(y,\z)$ and training-set maximum likelihood estimates.
+These estimates are found using the {\em{forward-backward}} algorithm \citep{Zucchini:2016}.
+We compare these estimated conditional probabilities to $x_1,\ldots,x_{100}$, the true coarse-scale state realizations in the test data, by calculating the {\em{average dive decoding accuracy}} for a single training/test data set pair, $\sum_{t=1}^{100} \hat{p}_t(x_t|y,\z)/100$.
 We then report the average of these over the 500 training/test data set pairs.
 Analogously, to assess prediction of the fine-scale states, we estimate $p^*_{t,\tilde t^*}(i^*|y,\z) \equiv \Pr(\tilde X^*_{t,\tilde t^*}=i|Y=y,\Z=\z)$, $i^*=1,2$, $\tilde t^* = 1,\ldots,\tilde T^*_t$, $t=1,\ldots,100$, using the test-set observations, the training-set maximum likelihood estimates, and the forward-backward algorithm.
-Denoting the true fine-scale state realizations from the test data set as $\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}$, we define the overall \textit{average subdive decoding accuracy} as the average value of $p^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*}|y,\z)$ across all simulated test data sets, dives, and windows.
+Denoting the true fine-scale state realizations from the test data set as $\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}$, we define the overall \textit{average subdive decoding accuracy} as the average value of $\hat{p}^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*}|y,\z)$ across all simulated test data sets, dives, and windows.
 The conditional probabilities are estimated according to one of the four models under study, using the maximum likelihood estimates from the training data set in conjunction with the forward-backward algorithm.

### 2021-01-06 Report/main.tex
add authors

@@
 %\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
-\title[]{Modelling multi-scale functional and biologging data with hidden Markov models}
+\title[]{Modelling multi-scale state-switching functional data with hidden Markov models}
+%\author{Evan Sidrow\authorref{1}}
+%\author{Nancy Heckman\authorref{1}}
+%\author{Sarah Fortune\authorref{2}}
+%\author{Andrew Trites\authorref{3,4}}
+%\author{Ian Murphy\authorref{1}}
+%\author{Marie Auger-M\'eth\'e\authorref{1,4}}
+%\affiliation[1]{Department of Statistics, University of British Columbia, Vancouver, Canada}
+%\affiliation[2]{Marine Mammal Research Unit, University of British Columbia, Vancouver, Canada}
+%\affiliation[3]{Department of Zoology, University of British Columbia, Vancouver, Canada}
+%\affiliation[4]{Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada}
 \author{BlindedA\authorref{1}\thanksref{*}}
 \author{BlindedB\authorref{2}}
-\affiliation[1]{Author affiliations will go here in the accepted manuscript, 
-but do NOT include them in your initial submission because it must be anonymous.}
+\affiliation[1]{Author affiliations will go here in the accepted manuscript, but do NOT include them in your initial submission because it must be anonymous.}
 \affiliation[2]{Second Affiliation}
@@
 This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.
-As a case study, we use this framework to model the fine-scale kinematic movement of a northern resident killer whale off the western coast of Canada.
+As a case study, we use this framework to model the fine-scale kinematic movement of a northern resident killer whale ({\em{Orcinus orca}}) off the western coast of Canada.
 Through simulations, we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing methods.
@@
 \begin{ack}{ACKNOWLEDGEMENTS}
-% This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
-% We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC). 
-% Marie Auger M'eth'e thanks the NSERC Discovery and the Canadian Research Chair programs.
-% Evan Sidrow thanks the University of British Columbia and the Four-Year Doctoral Fellowship program.
+This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
+We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC). 
+Marie Auger-M\'eth\'e thanks the NSERC Discovery and the Canadian Research Chair programs.
+Evan Sidrow thanks the University of British Columbia and the Four-Year Doctoral Fellowship program.
 \end{ack}

### 2021-01-07 Report/Sections/intro.tex
Incorporate Final Comments

@@
 Tags can record over 50 observations per second, resulting in time series containing millions of observations over the course of several hours.
-These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a challenge for statisticians.
+These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a challenge for statisticians and biologists.
 
@@
 %
-FDA was originally developed to process curves assumed to be independent replicates, and within-curve fine-scale structure is not usually incorporated in FDA models.
 
+FDA was originally developed to process curves assumed to be independent replicates (i.e., there is no between-curve dependence), and within-curve fine-scale structure is not usually incorporated in FDA models.
 However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.

### 2021-01-07 Report/main.tex
Incorporate Final Comments

@@
 %\author{Sarah Fortune\authorref{2}}
-%\author{Andrew Trites\authorref{3,4}}
+%\author{Andrew W. Trites\authorref{3,4}}
 %\author{Ian Murphy\authorref{1}}
@@
 %\affiliation[4]{Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada}
-\author{BlindedA\authorref{1}\thanksref{*}}
-\author{BlindedB\authorref{2}}
-\affiliation[1]{Author affiliations will go here in the accepted manuscript, but do NOT include them in your initial submission because it must be anonymous.}
-\affiliation[2]{Second Affiliation}
+\author{BlindedA}%\authorref{1}\thanksref{*}}
+\author{BlindedB}%\authorref{2}}
+%\affiliation[1]{Author affiliations will go here in the accepted manuscript, but do NOT include them in your initial submission because it must be anonymous.}
+%\affiliation[2]{Second Affiliation}
 
@@
 %
-\begin{ack}{ACKNOWLEDGEMENTS}
-This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
-We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC). 
-Marie Auger-M\'eth\'e thanks the NSERC Discovery and the Canadian Research Chair programs.
-Evan Sidrow thanks the University of British Columbia and the Four-Year Doctoral Fellowship program.
-\end{ack}
+%\begin{ack}{ACKNOWLEDGEMENTS}
+%This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
+%We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC) as well as the support of Fisheries and Oceans Canada (DFO). 
+%Marie Auger-M\'eth\'e and Nancy Heckman thank the NSERC Discovery program, and Marie Auger-M\'eth\'e additionally thanks the Canadian Research Chair program.
+%Evan Sidrow thanks the University of British Columbia and the Four-Year Doctoral Fellowship program.
+%\end{ack}
 %

### 2021-01-09 Report/Sections/models.tex
small changes

@@
 In the hierarchical setting, however, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$.
-For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
+For each curve $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
 If $X_t = i$, then the components of $X_t^*$ make up a Markov chain with $N^{*(i)}$ possible states, an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$, and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
@@
 \end{gather*}
-Given the coarse-scale hidden state sequence $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$ and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
+Given the coarse-scale hidden state sequence $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$, and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
 %
@@
 
-In many applications where data are collected at high frequencies, intricate dependency structures arise within the fine-scale process that cannot be adequately modelled with any HMM-based model described thus far.
+In many applications where data are collected at high frequencies, intricate dependency structures arise within the fine-scale process that cannot be adequately modelled with the HMM-based models described thus far.
 %For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. 
@@
 Larger stride lengths result in a loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting.
-In addition, setting stride length and window size equal to one another avoids artificial residual correlation arising from overlapping windows.
+In addition, setting the stride length equal to the window size avoids artificial residual correlation arising from overlapping windows.
 For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ and stride $h$ across $Y^*_t$.
@@
 When performing this transformation, the fine-scale HMM (or CarHMM) must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself.
-In particular, there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
+As a result, there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
 The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ instead of $X^*_t$ and $Y^*_t$.
@@
 If the fine-scale model is a simple Poisson process, then this approach is equivalent to a Markov-modulated Poisson process \citep{Fischer:1993}.
-The fine-scale process can also be modelled as in \citep{Langrock:2018}, who use B-splines to model the emission distribution of an HMM as well as incorporate covariates.
+The fine-scale process can also be modelled as in \citep{Langrock:2018}, who use B-splines to model the emission distribution of an HMM.
 This non-parametric approach uses a penalized likelihood term which can easily replace the usual fine-scale likelihood term in Equation (\ref{eqn:HHMM_likelihood}).

### 2021-01-09 Report/Sections/results.tex
small changes

@@
 The data we use were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration over time.
-Observations were collected at a rate of 50 Hz using a CATs biologger (Customizable Animal Tracking Solutions, www.cats.is).
+Observations were collected at a rate of 50 Hz using a CATs biologger (Customizable Animal Tracking Solutions, {\em{www.cats.is}}).
 Acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, and right/left).
@@
 %
-Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Section 1.1 of the supplementary material), and visual inspection shows no obvious complicated dependence.
+Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Figure 1 of the supplementary material), and visual inspection shows no obvious complicated dependence.
 Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor a moving-window transformation is called for.
@@
 We then have that $\z_{t,\tilde t^*} = \{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\}$, where $\zone_{t,\tilde t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window.
-Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Section 1.1 of the supplementary material).
+Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Figure 1 of the supplementary material).
 Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.
@@
 First, we assume that $\Ztwo_{t,\tilde t^*}$ and all three components of $\Zone_{t,\tilde t^*}$ are independent of one another when conditioned on the dive types and subdive states.
-To reduce model complexity, we also assume that the three fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$.
+To reduce model complexity, we also assume that the three sets of fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$.
 This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.
 To specify the emission distribution of $\Zone_{t,\tilde t^*}$, consider the sequence $\Zone_{t,1},\Zone_{t,2},\ldots,\Zone_{t,T^*_t}$ for a particular dive $t$.
-We assume that each of the three components of this sequence are Normally distributed as in Equation (\ref{eqn:carhmm}) and independent of one another when conditioned on the subdive states $\tilde X_{t,1},\ldots,\tilde X_{t,\tilde T^*_t}$.
+We assume that each of the three components of this sequence are Normally distributed as in Equation (\ref{eqn:carhmm}) and all components are independent of one another when conditioned on the subdive states $\tilde X_{t,1},\ldots,\tilde X_{t,\tilde T^*_t}$.
 Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter.
@@
 To specify the emission distribution of $\Ztwo_{t,\tilde t^*}$, we assume that given $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ follows a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
-In addition, $\Ztwo_{t,1},\ldots,\Ztwo_{t,T^*_t}$ are assumed to be independent of one another given the subdive state sequence $\tilde X_{t,1}, \ldots, \tilde X_{t,T_t^*}$.
+In addition, $\Ztwo_{t,1},\ldots,\Ztwo_{t,T^*_t}$ are assumed to be independent of one another given the subdive state sequence $\big\{\tilde X_{t,1}, \ldots, \tilde X_{t,T_t^*}\big\}$.
 We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
@@
 In addition to the CarHHMM-DFT, we consider three variations for comparison.
-As in the full model, each of the following models assume that all components of $\Z_{t,\tilde t^*}$ are assumed to be conditionally independent of one another given the dive types and subdive states:
+As in the full model, each of the following models assume that all components of $\Z_{t,\tilde t^*}$ are conditionally independent of one another given the dive types and subdive states:
 \begin{enumerate}

### 2021-01-09 Report/Sections/simulation.tex
small changes

@@
 
-Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} display one simulated data set as well as the decoded dive types and subdive states associated with each model.
+Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} display five dives of one simulated data set as well as the decoded dive types and subdive states associated with each model.
 The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state while the HHMM-DFT is slightly more likely to predict that a given window corresponds to subdive state 2.

### 2021-01-14 Report/Sections/appendixnh.tex
Finalize Before Submission

@@
 We easily simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
-For each dive $t$, we also easily generate the fine-scale hidden Markov chain $\tilde X^*_t \equiv \left\{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T_t^*}\right\}$ according to one of the probability transition matrices $\Gamma^{*(1)}$ or $\Gamma^{*(2)}$, depending upon the value of $X_t$.
+For each dive $t$, we also easily generate the fine-scale hidden Markov chain $\tilde X^*_t \equiv \big\{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T_t^*}\big\}$ according to one of the probability transition matrices $\Gamma^{*(1)}$ or $\Gamma^{*(2)}$, depending upon the value of $X_t$.
 This determines the sequence of fine-scale hidden states corresponding to each window.

### 2021-01-14 Report/Sections/intro.tex
small changes
Finalize Before Submission

@@
 
-To accommodate temporal dependence and changes in curve type, we turn to the field of
+To accommodate both temporal dependence and changes in curve type, we turn to the field of
 %methodology for the analysis of animal movement 
@@
 
-While useful, HMMs alone are also not sufficient to model intricate fine-scale time-series data for three primary reasons.
+While useful, HMMs alone are also not sufficient to model high-frequency time-series data for three primary reasons.
 %
@@
 %
-Thirdly, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures at short time scales.
+Thirdly, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures on short time scales.
 For example, \citet{Adam:2019} fail to capture fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}) using a traditional HHMM.

### 2021-01-14 Report/Sections/simulation.tex
Finalize Before Submission

@@
 The two simulated dive types differ in that dives of type 1 are much shorter on average (26 seconds) than dives of type 2 (105 seconds).
-The two simulated subdive states differ primarily due to $\mu_W^*$ and $\sigma_W^*$ since both are much higher for subdive state 2 than for subdive state 1.
+The two simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 2 than for subdive state 1.
 These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data.
@@
 Its average dive decoding accuracy and average subdive decoding accuracy are both greater than 0.9.
-All parameter estimates of fine-scale mean values and probability transition matrices, $\hat \mu^*$, $\hat \Gamma^*$ and $\hat \Gamma$, respectively, have statistically insignificant biases.
+All parameter estimates of fine-scale mean values ($\hat \mu^*$) and probability transition matrices ($\hat \Gamma^*$ and $\hat \Gamma$) have statistically insignificant biases.
 In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models.
@@
 In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material).
-These results suggest that the estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored.
+These results suggest that estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored.
 This finding is consistent with the results of the case study, where the HHMM-DFT produced larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.

### 2021-05-18 Code/HHMM.py
add random initialization and Compute Canada bash script

@@ -65,7 +65,7 @@ class HHMM:
         self.train_time = None
         self.CM = None
 
-        eta_crude = -1.0 + 0.1*np.random.normal(size=(self.pars.K[0],self.pars.K[0]))
+        eta_crude = -2.0 + np.random.normal(size=(self.pars.K[0],self.pars.K[0]))
         ptm_crude = np.exp(eta_crude)
         ptm_crude = (ptm_crude.T/np.sum(ptm_crude,1)).T
 
@@ -75,7 +75,7 @@ class HHMM:
         eta_fine = []
         ptm_fine = []
         for _ in range(self.pars.K[0]):
-            eta_fine_k = -1.0 + 0.1*np.random.normal(size=(self.pars.K[1],self.pars.K[1]))
+            eta_fine_k = -2.0 + np.random.normal(size=(self.pars.K[1],self.pars.K[1]))
             ptm_fine_k = np.exp(eta_fine_k)
             ptm_fine_k = (ptm_fine_k.T/np.sum(ptm_fine_k,1)).T
             eta_fine.append(eta_fine_k)
@@ -137,10 +137,20 @@ class HHMM:
                     else:
                         theta[0][feature]['sig'][k0] = 1.0
 
-                # finally update correlations randomly
-                theta[0][feature]['corr'] = np.random.random(size=K) - 2.0
+                # initialize correlations
+                theta[0][feature]['corr'] = -1.0 * np.ones(K)
 
-
+                # add randomness in initialization
+                if settings['f'] == 'normal':
+                    theta[0][feature]['mu'] += norm.rvs(np.zeros(K),np.abs(theta[0][feature]['mu']))
+                    theta[0][feature]['sig'] *= np.exp(norm.rvs(0.0,1.0,size=K))
+                    theta[0][feature]['corr'] += norm.rvs(0.0,1.0,size=K)
+                elif settings['f'] == 'gamma':
+                    theta[0][feature]['mu'] *= np.exp(norm.rvs(0.0,1.0,size=K))
+                    theta[0][feature]['sig'] *= np.exp(norm.rvs(0.0,1.0,size=K))
+                    theta[0][feature]['corr'] += norm.rvs(0.0,1.0,size=K)
+                else:
+                    pass
 
         # then fill in the subdive level values
         theta.append([{} for _ in range(K)])
@@ -166,7 +176,6 @@ class HHMM:
                         theta[1][k0][feature]['mu'] = np.mean(feature_data)*np.ones(K)
                     else:
[…]

### 2021-05-18 Code/Parameters.py

@@ -67,7 +67,9 @@ class Parameters:
                           'bottom_variance':{'corr':False,'f':'gamma'},
                           'dive_duration':{'corr':False,'f':'gamma'}},
 
-                         {'A':{'corr':True,'f':'multivariate_normal'},
+                         {'Ax':{'corr':True,'f':'normal'},
+                          'Ay':{'corr':True,'f':'normal'},
+                          'Az':{'corr':True,'f':'normal'},
                           'FoVeDBA_low':{'thresh':5,'corr':True,'f':'gamma'},
                           'FoVeDBA_high': {'thresh':5,'corr':True,'f':'gamma'},
                           'Vz': {'corr':True,'f':'normal'},

### 2021-05-18 Code/Train_Model.py

@@ -27,12 +27,17 @@ import Parameters
 import HHMM
 import Visualisor
 
-np.random.seed(0)
+rand_seed = int(sys.argv[1])
+model = str(sys.argv[2])
 
-model = str(sys.argv[1])
+np.random.seed(rand_seed)
 
 # set parameters
 pars = Parameters.Parameters()
+
+pars.cvc_file = '../Data/2019/20190902-182840-CATs_OB_1.cvc'
+pars.csv_file = '../Data/2019/20190902-182840-CATs_OB_1_001.csv'
+
 pars.features = [{'dive_duration':{'corr':False,'f':'gamma'}},
                  {'Ahat_low':{'thresh':5,'corr':False,'f':'gamma'},
                   'Ax':{'corr':True,'f':'normal'},
@@ -53,8 +58,8 @@ elif model == 'CarHHMM1':
                         'Az':{'corr':True,'f':'normal'}}
 
 # define files
-HHMM_file = '../Params/%s_k_%s_%s_dives_same_fine_states_long' % (model,pars.K[0],pars.K[1])
-data_outfile = '../Params/Data_%s_k_%s_%s_dives_same_fine_states_long' % (model,pars.K[0],pars.K[1])
+HHMM_file = '../Params/%s_k_%s_%s_%s' % (model,pars.K[0],pars.K[1],rand_seed)
+data_outfile = '../Params/data_%s_k_%s_%s_%s' % (model,pars.K[0],pars.K[1],rand_seed)
 
 # preprocess data
 prep = Preprocessor.Preprocessor(pars)
@@ -78,7 +83,7 @@ data = prep.get_all_features(df,dive_df)
 # train_model
 print('training model')
 hhmm = HHMM.HHMM(pars,data)
-hhmm.train_DM(data,max_iters=10,max_steps=5)
+hhmm.train_DM(data,max_iters=10,max_steps=10)
 
 # get SEs
 h = 0.01

### 2021-05-18 Code/Train_Model.sh

@@ -8,24 +8,24 @@
 #SBATCH --mail-type=REQUEUE
 #SBATCH --mail-type=ALL
 
-#SBATCH --time=72:00:00
-#SBATCH --array=1-4
+#SBATCH --time=96:00:00
+#SBATCH --array=1-400
 #SBATCH --mem-per-cpu=8G
 #SBATCH --cpus-per-task=8
 
-if [ $SLURM_ARRAY_TASK_ID -eq 1 ]
+if [ $SLURM_ARRAY_TASK_ID -lt 101 ]
 then
-  python Train_Model.py CarHMM
+  python Train_Model.py $SLURM_ARRAY_TASK_ID CarHMM
 fi
-if [ $SLURM_ARRAY_TASK_ID -eq 2 ]
+if [ $SLURM_ARRAY_TASK_ID -gt 100 ] && [ $SLURM_ARRAY_TASK_ID -lt 201 ]
 then
-  python Train_Model.py HHMM
+  python Train_Model.py $SLURM_ARRAY_TASK_ID HHMM
 fi
-if [ $SLURM_ARRAY_TASK_ID -eq 3 ]
+if [ $SLURM_ARRAY_TASK_ID -gt 200 ] && [ $SLURM_ARRAY_TASK_ID -lt 301 ]
 then
-  python Train_Model.py CarHHMM1
+  python Train_Model.py $SLURM_ARRAY_TASK_ID CarHHMM1
 fi
-if [ $SLURM_ARRAY_TASK_ID -eq 4 ]
+if [ $SLURM_ARRAY_TASK_ID -gt 300 ]
 then
-  python Train_Model.py CarHHMM2
+  python Train_Model.py $SLURM_ARRAY_TASK_ID CarHHMM2
 fi

### 2021-05-18 Report/Sections/appendix.tex

@@
 
-\subsection{Detailed description of data simulation}
+\subsection{Detailed description of data simulation from Section \ref{}}
 
@@
 %
-where the first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the shape parameter and the second is the scale parameter.
+where again the first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the shape parameter and the second is the scale parameter.
 Simple calculation of the mean and variance of a Gamma distribution reveals $\mu_W^{*(\cdot,1)} = 23.3$, $\sigma_W^{*(\cdot,1)} = 13.0$, $\mu_W^{*(\cdot,2)} = 301.2$, and $\sigma_W^{*(\cdot,2)} = 330.1$.

### 2021-05-18 Report/Sections/bibliography.tex
Clean up report and delete old files

@@
 
-% !TeX root = main.tex
+%%%%%%%%%%%%%%% new references %%%%%%%%%%%%%%%
 
@@
 	pages = {271-274},
-	title = {Some theoretical results on markov-switching autoregressive models with gamma innovations},
+	title = {Some theoretical results on Markov-switching autoregressive models with gamma innovations},
 	volume = {343},
@@
 \bibitem[]{Fischer:1993,
-  title={The Markov-Modulated Poisson Process (MMPP) Cookbook},
+  title={The Markov-Modulated {P}oisson Process (MMPP) Cookbook},
   author={Wolfgang Fischer and Kathleen S. Meier-Hellstern},

### 2021-05-18 Report/Sections/discussion.tex

@@
 Here we use a CarHMM as the fine-scale model, but there are a wide range of possible fine-scale models, including a Poisson process or continuous time approach similar to that of \citet{Michelot:2019}.
-We also incorporate a moving-window transformation at the fine-scale model to capture intricate dependence structures on short time scales.
+We also incorporate a moving-window transformation at the fine scale to capture intricate dependence structures on short time scales.
 Together, the coarse- and fine-scale models make up a hierarchical structure which can account for simultaneous processes taking place at different time scales.
@@
 Our analysis gives a deeper understanding of a killer whale's tri-axial movement and thus its behaviour and energy expenditure \citep{Gleiss:2011,Qasem:2012}, both of which are important for understanding the foraging ecology and nutritional status of northern resident killer whales \citep{Noren:2011}.
-Our model is applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.
+Our model is also applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.
 In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt our methodology for the analysis of a wide range of time series data in a variety of fields.

### 2021-05-18 Report/Sections/models.tex

@@
 
-Consider a sequence of $T$ curves, where any particular curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$.
-Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up fine-scale quantities derived from curve $t$ indexed using $t^*$.
+Consider a sequence of $T$ curves, where curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$.
+Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of fine-scale quantities derived from curve $t$ indexed by $t^*$.
 Both $Y_t$ and $Y^*_{t,t^*}$ can be either vectors or scalars.
@@
 The $X_t$'s form a Markov chain and can take integer values between $1$ and $N$.
-Their distribution is governed by the distribution of the initial state $X_1$ and the $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$.
+Their distribution is governed by the distribution of the initial state $X_1$ and an $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$.
 %
@@
 A conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process.
-Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots,X_T$ that can take integer values between $1$ and $N$.
+Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $\{X_1,\ldots,X_T\}$ each taking integer values between $1$ and $N$.
 A CarHMM also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
@@
 A hierarchical hidden Markov model, or HHMM, accounts for processes occurring simultaneously at different scales by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs.
-The coarse-scale model is either an HMM and CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations with state-dependent parameters $\theta^{(i)}$ for $i = 1,\ldots,N$.
+The coarse-scale model is either an HMM and CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, where $\{X_1, \ldots, X_T\}$ is an unobserved Markov chain with $N$ possible states and $\{Y_1, \ldots, Y_T\}$ is the set of corresponding observations with state-dependent parameters $\theta^{(i)}$ for $i = 1,\ldots,N$.
 %
 In the hierarchical setting, however, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$.
-For each curve $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
-If $X_t = i$, then the components of $X_t^*$ make up a Markov chain with $N^{*(i)}$ possible states, an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$, and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
+For each curve $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
+If $X_t = i$, then the components of $X_t^*$ make up a Markov chain that has $N^{*(i)}$ possible states, an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$, and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
 The distribution of $Y^*_{t,t^*}$ given $Y^*_{t,t^*-1} = y^*_{t,t^*-1}$, $X^*_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$.
@@
 
-Forcing certain parameters to be shared can reduce complexity and increase interpretability of the HHMM.
+Forcing certain parameters to be shared can reduce the complexity and increase the interpretability of an HHMM.
 For example, in our killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$.
@@
 \begin{equation}
-    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N
+    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N,
     \label{eqn:HHMM_likelihood}
@@
     \label{eqn:z}
-    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|^2,
+    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|_2^2,
 \end{equation}
@@
 When performing this transformation, the fine-scale HMM (or CarHMM) must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself.
-As a result, there are only $\tilde{T}_t^*$ hidden states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
+As a result, there are only $\tilde{T}_t^*$ unobserved states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
 The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ instead of $X^*_t$ and $Y^*_t$.
@@
 %where $P(y_t,y^*_t|y_{t-1};\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\theta^{(i)}\right)\calL_{\text{HMM}}\left(\theta^{*(i)};y^*_t\right)$.
-However, the fine-scale observations for a particular dive $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods.
+However, the fine-scale observations of a particular dive $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods.
 As such, the fine-scale HMM likelihood term $\calL_{\text{fine}}$ in Equation (\ref{eqn:HHMM_likelihood}) can be replaced by the likelihood of a general fine-scale model whose parameters depend upon the coarse-scale hidden state.
@@
 If the fine-scale model is a simple Poisson process, then this approach is equivalent to a Markov-modulated Poisson process \citep{Fischer:1993}.
-The fine-scale process can also be modelled as in \citep{Langrock:2018}, who use B-splines to model the emission distribution of an HMM.
+The fine-scale process can also be modelled similarly to \citet{Langrock:2018}, who use B-splines to model the emission distribution of an HMM.
 This non-parametric approach uses a penalized likelihood term which can easily replace the usual fine-scale likelihood term in Equation (\ref{eqn:HHMM_likelihood}).

### 2021-05-18 Report/Sections/results.tex

@@
 Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
-Specifically, the sequence of observed coarse-scale observations $y = \big\{y_1,\ldots,y_{267}\big\}$ is a time series of dive durations in seconds.
-For dive $t$, the fine-scale observations are contained in $y^*_{t} \equiv \big\{y^*_{t,1},\ldots,y^*_{t,T^*_t} \big\}$, which is a sequence containing the within-dive acceleration data in units of meters per second squared.
+Specifically, the observed coarse-scale observations $y = \big\{y_1,\ldots,y_{267}\big\}$ make up a sequence of dive durations in seconds.
+For dive $t$, the fine-scale observations are contained in $y^*_{t} = \big\{y^*_{t,1},\ldots,y^*_{t,T^*_t} \big\}$, which is a sequence containing the within-dive acceleration data in units of meters per second squared.
 The collection of all acceleration data is denoted as $y^* = \big\{y^*_1,\ldots,y^*_{267}\big\}$.
@@
 Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with a window size of $h=100$ (two seconds) and a maximum frequency of $\tilde{\omega}=10$ (5 Hz).
-We then have that $\z_{t,\tilde t^*} = \{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\}$, where $\zone_{t,\tilde t^*}$ is a three dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window.
+We then have that $\z_{t,\tilde t^*} = \{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\}$, where $\zone_{t,\tilde t^*}$ is a three-dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window.
 Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Figure 1 of the supplementary material).
@@
 This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.
-To specify the emission distribution of $\Zone_{t,\tilde t^*}$, consider the sequence $\Zone_{t,1},\Zone_{t,2},\ldots,\Zone_{t,T^*_t}$ for a particular dive $t$.
+To specify the emission distribution of $\Zone_{t,\tilde t^*}$, consider the sequence $\{\Zone_{t,1},\ldots,\Zone_{t,T^*_t}\}$ for a particular dive $t$.
 We assume that each of the three components of this sequence are Normally distributed as in Equation (\ref{eqn:carhmm}) and all components are independent of one another when conditioned on the subdive states $\tilde X_{t,1},\ldots,\tilde X_{t,\tilde T^*_t}$.
@@
 This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2.
-Each histogram is then plotted together with the corresponding emission distribution estimated by the CarHHMM-DFT.
+Each histogram is plotted together with the corresponding emission distribution estimated by the CarHHMM-DFT.
 Analogous histograms corresponding to the fine-scale observations are contained in Section 1 of the supplementary material.

### 2021-05-18 Report/Sections/intro.tex
Updates from Overleaf

@@
 Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time.
-For example, dive profiles of marine animals show distinct dive ``types" that cluster together in time \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}.
-Fine-scale periodicity within a larger process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
+For example, dive profiles of marine animals show a discrete number of distinct dive ``types" that aggregate together in time \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}.
+Fine-scale periodicity within a larger state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
 
@@
 
-Some FDA models account for between-curve dependence occurring when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain types of dependence in time.
+Some FDA models account for between-curve dependence occurring when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain kinds of dependence in time.
 For example, previous studies have used multilevel models with random effects to model variation between and within individuals in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}.
 More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}.
-However, these multilevel models are not appropriate in many biologging applications since they do not account for different curve types.
-Further, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence is not appropriate in most biologging applications.
+%However, these multilevel models are not appropriate in many biologging applications since they rely on  do not account for a discrete number curve types.
+However, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence assumes that curves evolve smoothly in time, which is not appropriate in most biologging applications where behaviours often change suddenly between a discrete number of types.
 In addition to multi-level models, FDA researchers have also used functional time series to model dependence in a sequence of curves.
@@
 
-To accommodate both temporal dependence and changes in curve type, we turn to the field of
+We aim to identify and describe distinct curve types within sequential functional data while simultaneously modelling the complicated temporal dependence both between and within those curves.
+To accommodate both temporal dependence and sudden changes in behaviour, we turn to the field of
 %methodology for the analysis of animal movement 
@@
 animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
-HMMs interpret animal movement data as arising from a sequence of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position.
+HMMs interpret animal movement data as arising from a Markov chain of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position.
 While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions.
@@
 %While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together. 
-In this paper, we combine existing methods from statistical ecology literature in novel ways to account for complex, temporally dependent functional data.
+In this paper, we combine existing methods from statistical ecology literature in novel ways to classify and describe state-switching behaviours in functional data while accounting for complex temporal dependence.
 The resulting suite of methods makes up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.

### 2021-05-18 Report/Sections/results.tex
Updates from Overleaf

@@
 Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows a Gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:
-%
-$$\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.$$
+
+\[\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.\]
 %
@@
 In total, the parameters to estimate are
-%
+
 \begin{gather*}
@@
 No dive in this data set has a maximum depth deeper than 30 meters and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters.
-However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and significant numbers of fish have been observed to be caught near the surface (Fortune and Trites, unpublished data).
+However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and significant numbers of fish have been observed to be caught near the surface \citep{Fortune:2020}.
 Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.
@@
 The estimated probability transition matrices and associated stationary distributions on the coarse scale are
-%
-$$\hat \Gamma = \begin{pmatrix} 
+
+\[\hat \Gamma = \begin{pmatrix} 
 0.788 & 0.212 \\
 0.809 & 0.191
-\end{pmatrix} \text{ and }$$
-$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
+\end{pmatrix} \text{ and }\]
+
+\[\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}\]
 %
 for the transitions between dives. The estimated probability transition matrices and stationary distributions on the fine scale are
-$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
+
+\[\hat \Gamma^{*(1)} = \begin{pmatrix} 
 0.679 & 0.321 & 0.000 \\
@@
 0.000 & 0.216 & 0.784
-\end{pmatrix},$$
-$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
+\end{pmatrix},\]
+
+\[\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}\]
 %

### 2021-05-22 Report/Sections/intro.tex
update reviewer 1, comments 1-4a

@@
 Similarly, \citet{Fu:2017} view the dive profile of a southern elephant seal (\textit{Mirounga leonina}) as a set of dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
-%, although they ignore sequential dependence between dives. 
-%Similarly, our case study views the dive profile of a killer whale (\textit{Orcinus orca}) as a sequence of curves where three orthogonal axes of acceleration are each functions of time.
-%
 
@@
 However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
-%, as is the case for our case study data.
-Namely, the dive profile (see Figure \ref{fig:data}) of a killer whale from the threatened northern resident population off the coast of British Columbia, Canada, clearly shows temporal dependence between dives, as dives of various time durations and depths appear to be clustered together in time.
 For example, dive profiles of marine animals show a discrete number of distinct dive ``types" that aggregate together in time \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}.
 Fine-scale periodicity within a larger state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
-
-%In addition, \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks.
-The raw acceleration data exhibit extreme periodicity (in the form of vibration) that depends upon the behaviour of the train as the load and speed of the train change.
-%To model between-curve dependence in dive profiles, previous studies have categorized killer whale (\textit{Orcinus orca}) dives into different types associated with specified behaviours such as resting, travelling, and foraging \citep{Tennessen:2019b}.
-%Figure {\ref{fig:labeled_dives}} also shows clear temporal dependence within dives, as the accelerometer data within each dive indicate bouts of short-term periodicity. 
-%Behaviourally dependent periodicity is not unique to the biologging data- in the field of machine health, \citet{Lucero:2019} attach accelerometers to railway axles to detect cracks.
-The raw acceleration data exhibit extreme periodicity (in the form of vibration) that depends upon the behaviour of the train as the load and speed of the train change.
 
@@
 More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}.
-%However, these multilevel models are not appropriate in many biologging applications since they rely on  do not account for a discrete number curve types.
-However, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence assumes that curves evolve smoothly in time, which is not appropriate in most biologging applications where behaviours often change suddenly between a discrete number of types.
+%However, these multilevel models are not appropriate in many biologging applications since they do not account for a discrete number curve types.
+However, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence assumes that curves evolve smoothly in time.
+This is not appropriate in most biologging applications where behaviours often change suddenly between a discrete number of types.
 In addition to multi-level models, FDA researchers have also used functional time series to model dependence in a sequence of curves.
@@
 
-Traditional FDA techniques similarly fall short when modelling complicated within-curve %acceleration 
-data. In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function \citep{Yao:2005} 
-%use this framework to construct a method of sparse longitudinal data analysis using functional principal component analysis. Alternatively, within-curve structure is sometimes generated 
-or with random regression \citep{Rice:2001}. 
-%For example, \citet{Rice:2001} use random regression to construct non-parametric mixed effects models for unequally sampled noisy curves.
-However, the combination of apparent behavioural changes combined with periodic behaviour makes accelerometer data such as ours difficult to model with these classic FDA techniques.
+Traditional FDA techniques similarly fall short when modelling complicated within-curve data.
+In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function \citep{Yao:2005} or with random regression \citep{Rice:2001}.
 However, time-series data exhibiting both sharp behavioural changes and periodic fine-scale structure are difficult to model with these classic FDA techniques.
 
-We aim to identify and describe distinct curve types within sequential functional data while simultaneously modelling the complicated temporal dependence both between and within those curves.
-To accommodate both temporal dependence and sudden changes in behaviour, we turn to the field of
-%methodology for the analysis of animal movement 
-%statistical ecology
-animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
+The goal of this paper is to identify and describe a finite number of distinct curve types within sequential functional data while simultaneously modelling the complicated temporal dependence both between and within those curves.
+To accommodate both temporal dependence and sudden changes in behaviour, we turn to the field of animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
 HMMs interpret animal movement data as arising from a Markov chain of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position.
 While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions.
-In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines,
-%This is useful for both incorporating covariates and flexibly describing complex data, since these functions are non-parametric and take covariates as arguments. 
-but this approach does not account for certain types of temporal correlation.
+In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, but this approach does not account for certain types of temporal correlation.
 
 While useful, HMMs alone are also not sufficient to model high-frequency time-series data for three primary reasons.
-%
-Firstly, HMMs assume that subsequent observations are independent given an underlying hidden state process, but this is often not the case when observations are taken at extremely high frequencies. 
+Firstly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales (i.e., both between and within curves).
+The fine-scale, within-curve process is of particular interest in this work.
+To address this issue, statistical ecologists have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model both scales with conditionally dependent HMMs.
+Secondly, HMMs assume that subsequent observations are independent given an underlying hidden state process, but this is often not the case when observations are taken at extremely high frequencies.
 Several solutions have been proposed in the ecology literature such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
-%The CarHMM in particular explicitly models auto-correlation into an HMM while largely maintaining its base structure. 
-%
-Secondly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales (i.e., both between and within curves). 
-To address this issue, statistical ecologists have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model both scales with conditionally dependent HMMs.
-%
 Thirdly, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures on short time scales.
 For example, \citet{Adam:2019} fail to capture fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}) using a traditional HHMM.
-\citet{Heerah:2017} successfully use Fourier analysis within an HMM to account for %periodic behaviour in the context of describing
-daily behavioural cycles in marine mammals, and Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
+\citet{Heerah:2017} successfully use Fourier analysis within an HMM to account for daily behavioural cycles in marine mammals, and Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
 Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for fine-scale periodic structures.
 
-%While animal movement modellers have taken steps to solve each of these issues independently, to our knowledge there is no unifying framework combining the three together. 
 In this paper, we combine existing methods from statistical ecology literature in novel ways to classify and describe state-switching behaviours in functional data while accounting for complex temporal dependence.
 The resulting suite of methods makes up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.
-%and thus extend the methods available for the analysis of functional data. 
 We begin in Section 2 by describing HMMs as well as two variants, CarHMMs and HHMMs, and discuss how Fourier analysis can handle fine-scale dependence structures.

### 2021-05-22 Report/Sections/models.tex
update reviewer 1, comments 1-4a

@@
 
-%Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$.
-We first detail the structure of a traditional HMM and a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations.
-Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$.
-To simultaneously model $Y$ and $Y^*$, we use the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
-%which is comprised of two distinct,x but related, HMMs that jointly model these time series. 
-%Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics.
-In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour.
-Finally, we show how each of these models can be synthesized to form a wide variety of new models.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{HMMs as a base structure}
@@
 The likelihood $\calL_{\text{HMM}}$ can be evaluated using the well-known \textit{forward algorithm} \citep{Zucchini:2016}:
-%
-$$\calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N,$$
+
+\[\calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N,\]
 %
@@
 Following \citet{Barajas:2017}, we reparameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows sum to 1:
-%
+
 \[
@@
 As a concrete example, if $Y_t$ is a scalar, then one may assume that $Y_t$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
-%
+
 \begin{align}
@@
 As previously, the likelihood corresponding to a general CarHMM can be easily calculated using the forward algorithm. If $y$ is the sequence of observed emissions, then
+
 \begin{equation*}
@@
 %
-In the hierarchical setting, however, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$.
-For each curve $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
-If $X_t = i$, then the components of $X_t^*$ make up a Markov chain that has $N^{*(i)}$ possible states, an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$, and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
+In the hierarchical setting, however, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X^*_t \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$.
+If $X_t = i$, then $X^*_{t,t^*} \in \{1,\ldots,N^{*(i)}\}$ represents one of $N^{*(i)}$ possible within-curve behaviours associated with curve type $i$.
+In turn, $X^*_t$ emits a sequence of fine-scale observations $Y^*_t \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$.
+For each curve $t$, the fine-scale process $\{X^*_t, Y^*_t\}$ follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
+If $X_t = i$, then the components of $X^*_t$ make up a Markov chain that has $N^{*(i)}$ possible states, an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$, and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
 The distribution of $Y^*_{t,t^*}$ given $Y^*_{t,t^*-1} = y^*_{t,t^*-1}$, $X^*_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$.
@@
 In summary:
-%
-% this is new
-%
+
 \begin{gather*}
@@
 \end{gather*}
-Given the coarse-scale hidden state sequence $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$, and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
-%
-% end new
-%
+Given the coarse-scale hidden state sequence $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X^*_T, Y^*_T\}$, and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
 
@@
 We also share the fine-scale emission parameters across the $N$ coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
-Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$.
+Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$, \textit{not} their fine-scale emission distributions.
 
-Due to the nested structure of the HHMM, the likelihood is easily calculated using the forward algorithm.
+While at first it seems difficult to jointly infer the paramtersDue to the nested structure of the HHMM, its likelihood is easily calculated using the forward algorithm.
 %
@@
 The likelihood of the observed data is then
-%
+
 \begin{equation}
@@
 Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.
-%
-%For more information on specific considerations for HHMMs, such as incorporating covariates into the probability transition matrix, state decoding, model selection and model checking, see \citet{Adam:2019}.
 
@@
 In many applications where data are collected at high frequencies, intricate dependency structures arise within the fine-scale process that cannot be adequately modelled with the HMM-based models described thus far.
-%For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. 
-To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize any non-Markovian behaviour.
-To maintain the temporal structure of the fine-scale process, local summary statistics can be calculated from a moving window over the elements of $Y_t^*$.
-Subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window.
-Stride length refers to the distance between the first element of consecutive windows, so a stride length of $h$ implies that the first window starts at $Y^*_{t,1}$, the second at $Y^*_{t,h+1}$, and so on.
-Larger stride lengths result in a loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting.
+To handle these additional fine-scale structures, we recommend replacing $Y^*_t = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize any non-Markovian behaviour.
+To maintain the temporal structure of the fine-scale process, local summary statistics can be calculated from a moving window with stride length $h$ over the elements of $Y^*_t$.
+Stride length refers to the distance between the first element of consecutive windows, so a stride length of $h$ implies that the first window starts at $Y^*_{t,1}$, the second at $Y^*_{t,1+h}$, and so on.
+Replacing $Y^*_{t,t^*}$ with relevant summary statistics results in a loss of information since we are substituting infinite-dimensional curves with a low-dimensional representation.
+\citet{Aue:2015} and \citet{Gao:2019} use functional principle component analysis (FPCA) for dimension reduction to minimize error when predicting functional data.
+However, these papers do not account for state-switching behaviour between curves and do not guarantee that the resulting FPCA scores will be interpretable for practitioners.
+The goal of this paper is to classify and describe curve types and within-curve behaviour, not predict future functional data.
+We are thus primarily interested in the fine-scale summary statistics themselves rather than the curves from which they are calculated, and these summary statistics are known to have consistent maximum likelihood estimates \cite{Douc:2004}.
+As a result, we choose a dimension reduction technique which can effectively separate behavioural states and has an intuitive interpretation for practitioners.
+In general, subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window.
+Larger stride lengths result in a larger loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting.
 In addition, setting the stride length equal to the window size avoids artificial residual correlation arising from overlapping windows.
-For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ and stride $h$ across $Y^*_t$.
-Namely:
-%
+For our particular case study, we use the discrete Fourier transform (DFT) of a moving forward window with width $h$ and stride $h$ across $Y^*_t$:
+
 \begin{align}
@@
 Next, we calculate transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$:
-%
+
 \begin{equation}
@@
 In words, $\Zone_{t,\tilde{t}^*}$ is the average value of $Y^*_t$ within window $\tilde{t}^*$ and $\Ztwo_{t,\tilde{t}^*}$ is the squared two-norm of the component of the window that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window.
-%(by the Plancherel theorem).
 More intuitively, $\Ztwo_{t,\tilde{t}^*}$ corresponds to the ``wiggliness" of the fine-scale data within curve $t$ and window $\tilde{t}^*$.
@@
 To clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT.
-Figure \ref{fig:models}d displays the dependence structure of an HMM-DFT.
+Figure \ref{fig:models}d displays the dependence structure of a fine-scale HMM-DFT.
 
@@
 
-%By definition, an 
 Traditional HHMMs treat both the coarse-scale and the fine-scale processes as realizations of an HMM or CarHMM. 
-%However, it is straightforward to generalize the coarse-scale HMM by replacing it with a CarHMM, yielding the following likelihood:
-%\[
-%\calL_{\text{coarse}}(\theta,\theta^*,\Gamma;y,y^*) = \delta \prod_{t=2}^T \Gamma P(y_t,y^*_t|y_{t-1};\theta,\theta^*) \mathbf{1}_N
-%\]
-%where $P(y_t,y^*_t|y_{t-1};\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t|y_{t-1};\theta^{(i)}\right)\calL_{\text{HMM}}\left(\theta^{*(i)};y^*_t\right)$.
 However, the fine-scale observations of a particular dive $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods.
 As such, the fine-scale HMM likelihood term $\calL_{\text{fine}}$ in Equation (\ref{eqn:HHMM_likelihood}) can be replaced by the likelihood of a general fine-scale model whose parameters depend upon the coarse-scale hidden state.
-%$\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
-%$$
-%\calL_{\text{coarse}}(\theta,\theta^*,\Gamma;y,y^*) = \delta P(y_1,y^*_1;\theta,\theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\theta,\theta^*) \mathbf{1}_N,
-%$$
-%where $P(y_t,y^*_t;\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\theta^{(i)}\right)\calL_{\text{fine}}\left(\theta^{*(i)};y^*_t\right)$.
 %
@@
 These examples include a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models. 
-%For example, one may construct a three-level HHMM, where three scales of observations are recorded (fine, medium, and coarse-scale) and dependent traditional HMMs are used to model all three levels. 
 A myriad of possible models can be built using this framework, but these models can quickly become complicated and computationally expensive to fit.
 Therefore, models should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.
-
-%the fine scale behaviour follows a continuous time HMM.
-The parameters of the fine scale distributions depend on hidden states.
-However the states are modelled as components in a mixture distribution, not on an HMM.

### 2021-05-22 Report/Sections/results.tex
update reviewer 1, comments 1-4a

@@
 Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook \citep{Williams:2009,Noren:2011,Wright:2017}.
-%It is therefore of interest to determine the energetic expenditure of killer whales hunting for Chinook as well as the frequency at which they catch prey, especially since many Chinook salmon in the area are either threatened or endangered \citep{Ford:2015}.
 Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but the animal's behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}.
@@
 Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
-Specifically, the observed coarse-scale observations $y = \big\{y_1,\ldots,y_{267}\big\}$ make up a sequence of dive durations in seconds.
+Specifically, the observed coarse-scale observations $y = \big\{y_1,\ldots,y_{267}\big\}$ make up a sequence of dive durations in seconds, and the coarse-scale hidden states $x = \big\{x_1,\ldots,x_{267}\big\}$ represent the corresponding dive types.
 For dive $t$, the fine-scale observations are contained in $y^*_{t} = \big\{y^*_{t,1},\ldots,y^*_{t,T^*_t} \big\}$, which is a sequence containing the within-dive acceleration data in units of meters per second squared.
-The collection of all acceleration data is denoted as $y^* = \big\{y^*_1,\ldots,y^*_{267}\big\}$.
+The fine-scale hidden states of dive $t$, $x^*_{t} = \big\{x^*_{t,1},\ldots,x^*_{t,T^*_t} \big\}$, represent the subdive behavioural states of the killer whale.
+The collection of all acceleration data is denoted as $y^* = \big\{y^*_1,\ldots,y^*_{267}\big\}$ and the collection of all unknown subdive behaviours is denoted as $x^* = \big\{x^*_1,\ldots,x^*_{267}\big\}$.
 
@@
 
-Defining a suitable model to describe this killer whale kinematic data involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse- and fine-scale observations.
+The primary goal of this case study is to jointly estimate the dive types and (more importantly) the subdive behaviours of this killer whale, so only hierarchical models were considered to describe the kinematic data.
+Defining a suitable hierarchical model involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse- and fine-scale observations.
 
@@
 
-We then select the specific emission distribution of $\Z_{t,\tilde t^*}$.
+We then select the specific emission distributions of $\Z_{t,\tilde t^*}$ for all dive types and subdive states.
 First, we assume that $\Ztwo_{t,\tilde t^*}$ and all three components of $\Zone_{t,\tilde t^*}$ are independent of one another when conditioned on the dive types and subdive states.
@@
 We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
-%
-%The final model is a hierarchical hidden Markov model with a traditional HMM on the coarse scale.
-Each fine-scale sequence $y^*_t$ is transformed to $\z$ using Equation (\ref{eqn:z}) and $\Z_{t,\tilde t^*}$ is modelled with a CarHMM.
-%We assume $\Zone_{t,\tilde t^*}$ is Normally distributed with auto-correlation while $\Ztwo_{t,\tilde t^*}$ is Gamma distributed without auto-correlation.
-We also assume that $\Ztwo_{t,\tilde t^*}$ and all components of $\Zone_{t,\tilde t^*}$ are each independent of one another when conditioned on the subdive state.
-The emission distribution of each subdive state does not depend upon the dive type.
 
@@
 This model assumes that there is only one dive type.
-    %, so the fine-scale probability transition matrix $\Gamma^*$ is the same for every dive. 
 \end{enumerate}
@@
 for dive types 1 and 2.
-%The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that 
-In summary, about 79\% of dives are short dives of type 1, and
-%The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that 
-the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1.
+In summary, about 79\% of dives are short dives of type 1, and the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1.
 This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
-%The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that
 Further, this killer whale is in the less active subdive state 1 40\% of the time during a dive of type 1 compared to only only 9\% of the time during a dive of type 2.

### 2021-06-03 Report/Sections/models.tex

@@
 We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.
+We assume that $\{Y_1,\ldots,Y_T\}$ is ordered sequentially but the curves need not be equi-spaced in time.
 To develop our model for this data, we detail the structure of a traditional HMM followed by three variations which generalize its base structure.
@@
 Their distribution is governed by the distribution of the initial state $X_1$ and an $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$.
+We consider only time-homogeneous Markov chains, meaning that $\Gamma$ does not depend on time.
 %
@@
 
-Using observation emissions, denoted here as $y \equiv \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$.
-The likelihood $\calL_{\text{HMM}}$ can be evaluated using the well-known \textit{forward algorithm} \citep{Zucchini:2016}:
+%Using observation emissions, denoted here as $y \equiv \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$.
+The joint likelihood of both the parameters and the hidden states is given by the following:
+
+\[\calL_{\text{HMM}}(x,\theta,\Gamma;y) = \delta_{x_1} f^{(x_1)}(y_1; \theta^{(x_1)}) \prod_{t=2}^T \Gamma_{x_{t-1} x_t} f^{(x_t)}(y_t; \theta^{(x_t)}),\]
+%
+which can be used to obtain estimates for both the hidden states $\hat X$ and the parameters $\hat \theta$ and $\hat \Gamma$.
+Using a Bayesian approach, $\calL_{\text{HMM}}$ can be combined with a prior distribution to obtain a posterior distribution over $\hat X$, $\hat \theta$, and $\hat \Gamma$ which can be sampled from using methods such as Sequential Monte Carlo \citep{Douc:2011b}, Markov-Chain Monte Carlo \citep{Scott:2002}, or variational inference \citep{Foti:2014}.
+For the purposes of this paper, however, we focus on the frequentest approach, where it is common to marginalize out $x$ to obtain
 
@@
 %
-where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and
-%
-$P(y_t;\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry $f^{(i)}(y_t; \theta^{(i)})$.
+where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and $P(y_t;\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry $f^{(i)}(y_t; \theta^{(i)})$.
+This equation can be evaluated in $\mathcal{O}(T)$ time using the well-known \textit{forward algorithm} \citep{Zucchini:2016}.
+After obtaining maximum likelihood estimates $\{\hat \theta, \hat \Gamma\}$, estimated probabilities of each hidden state $\big(\hat{\Pr}(X_t = i|Y = y)\big)$ can be calculated from $\{\hat \theta,\hat \Gamma\}$ using the \textit{forward-backward algorithm} \citep{Zucchini:2016}.
 %
@@
 where $i,j = 1,\ldots,N$ and $\eta_{ii}$ is set to zero for identifiability.
-This simplifies likelihood maximization by removing constraints in the optimization problem.
+This formulation simplifies likelihood maximization by removing constraints in the optimization problem.
+We assume there are no covariate effects in this paper, \cite{DeSouza:2017} and \citet{Adam:2019} both incorporate covariates into $\Gamma$ by setting $\eta_{ij}(z_t) = \beta_{ij}^T z_t$ for $i \neq j$, where $z_t$ is a column vector of known covariates at time $t$ and $\beta_{ij}$ is a column vector of unknown regression coefficients.
 For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$.
@@
 
-As previously, the likelihood corresponding to a general CarHMM can be easily calculated using the forward algorithm. If $y$ is the sequence of observed emissions, then
+As previously, the likelihood of $\theta$ and $\Gamma$ corresponding to a CarHMM can be calculated using the forward algorithm. If $y$ is the sequence of observations, then
 
@@
 Forcing certain parameters to be shared can reduce the complexity and increase the interpretability of an HHMM.
-For example, in our killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$.
-We also share the fine-scale emission parameters across the $N$ coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
-Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$, \textit{not} their fine-scale emission distributions.
+For example, since $f^{*(i,i^*)}$ depends on both $X_t = i$ and $X^*_{t,t^*} = i^*$, a within-curve behaviour is defined by the pair $(X_t,X^*_{t,t^*})$.
+However, in our killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$ and share the fine-scale emission parameters across the $N$ coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
+As a result, within-curve behaviours are shared between curve types and are defined exclusively by $X^*_{t,t^*}$ (or $(\cdot,X^*_{t,t^*})$ for notational consistency).
+Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$, not their fine-scale emission parameters $\theta^{*(\cdot,i)}$.
 
-While at first it seems difficult to jointly infer the paramtersDue to the nested structure of the HHMM, its likelihood is easily calculated using the forward algorithm.
+Due to the nested structure of the HHMM, %its likelihood is easily calculated using the forward algorithm.
+it is straightforward to extend the forward algorithm and marginalize out both the coarse-scale hidden states $X$ and the fine-scale hidden states $X^*$ from the likelihood.
 %
@@
 
-When performing this transformation, the fine-scale HMM (or CarHMM) must be redefined since $\Z_t$ exists on a coarser scale than $Y^*_t$ itself.
-As a result, there are only $\tilde{T}_t^*$ unobserved states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$, which we denote as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
-The fine-scale probability transition matrices  $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ are then applied directly to $\tilde{X}^*_t$ and $\Z_t$ instead of $X^*_t$ and $Y^*_t$.
+After performing this transformation, the entire model must be redefined since the fine-scale HMM (or CarHMM) directly models the distribution of the summary statistics $\Z_t$ rather than the discretized curve $Y^*_t$.
+Because $\Z_t$ likely exists on a coarser scale than $Y^*_t$, there are only $\tilde{T}_t^*$ unobserved states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$.
+We denote these unobserved states as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
+The fine-scale probability transition matrices $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ correspond to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$, as do all fine-scale model assumptions (e.g.
+conditional independence, auto-regressive structure between observations, etc.).
+Fine-scale model selection should be done accordingly.
+%instead of $X^*_t$ and $Y^*_t$.
 The likelihood of this model is therefore identical to that of the original HMM or CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$.

### 2021-06-03 Report/Sections/intro.tex
Updates from Overleaf

@@
 Tags can record over 50 observations per second, resulting in time series containing millions of observations over the course of several hours.
-These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a challenge for statisticians and biologists.
+%These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a challenge for statisticians and biologists.
+These data contain a wealth of information about human and animal behaviour, but modelling these large data sets poses a challenge for statisticians and biologists.
+One particular difficulty addressed in this paper is that many coarse- and fine-scale processes often arise simultaneously in high-frequency data, and each of these can exhibit unique behaviours and dependence structures.
 
-Biologging data are frequently modelled as a set of curves analyzed by methods of Functional Data Analysis, or FDA -- see, e.g., \citet{Ramsay:2005}.
+Biologging data are frequently modelled as a set of curves and analyzed using methods of Functional Data Analysis, or FDA -- see, e.g., \citet{Ramsay:2005}.
 For example, \citet{Morris:2007} view a child's activity level as a set of daily curves where metabolic activity is a function of time.
@@
 While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions.
-In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, but this approach does not account for certain types of temporal correlation.
+In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, while \citet{DeSouza:2014} and \citet{DeSouza:2017} use HMMs to model state-switching behaviour within functional data.
+However, the first paper does not account for certain types of temporal correlation while the later two focus on state-switching behaviour within functions rather than between them.
 

### 2021-06-03 Report/Sections/results.tex
Updates from Overleaf

@@
 
-To illustrate an application of this method and compare the candidate models, we fit all four models to the data shown in Figure \ref{fig:data}.
+%To illustrate an application of this method and compare the candidate models, 
+We estimated the parameters of all four models to the data shown in Figure \ref{fig:data} with direct likelihood maximization using the \textit{scipy} package in Python \citep{Virtanen:2019}.
+Each model was fit 100 times using random initializations, and we kept the parameters corresponding to the maximum likelihood of those 100 optimization routines.
 We first report the results from the full CarHHMM-DFT in detail and assess the quality of the fit.
@@
 We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}.
-Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the \textit{forward-backward algorithm} \citep{Zucchini:2016}.
+Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the %\textit{forward-backward algorithm} \citep{Zucchini:2016}.
+forward-backward algorithm.
 This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2.

### 2021-06-07 Code/HHMM.py
update optimization procedure results

@@ -14,11 +14,12 @@ from scipy.special import expit
 from scipy.special import logit
 from scipy.special import logsumexp
 from scipy.optimize import minimize
+from scipy.optimize import minimize_scalar
 from scipy.optimize import LinearConstraint
 from scipy.signal import convolve
 from scipy.interpolate import interp1d
 
-from skopt import gp_minimize
+#from skopt import gp_minimize
 
 from copy import deepcopy
 
@@ -194,6 +195,26 @@ class HHMM:
         self.theta = theta
 
         return
+
+
+    def reorder_params(self):
+
+        state_order = np.argsort(self.theta[1][0]['Ax']['sig'])
+
+        # first reorder eta
+        for k0 in range(self.pars.K[0]):
+            eta_fine = deepcopy(self.eta[1][k0])
+            for i in range(k0):
+                for j in range(k0):
+                    eta_fine[i,j] = eta_fine[state_order[i],state_order[j]]
+            self.eta[1][k0] = eta_fine
+
+        # then reorder theta
+        for feature,settings in self.pars.features[1].items():
+            for k0 in range(self.pars.K[0]):
+                for param in ['mu','sig','corr']:
+                    self.theta[1][k0][feature][param] = np.array([self.theta[1][k0][feature][param][state_order[i]] for i in range(3)])
+
 
 
     def find_log_p_yt_given_xt(self,level,feature,data,data_tm1,mu,sig,corr,sample=0):
@@ -598,7 +619,7 @@ class HHMM:
 
         return (self.theta,self.eta)
 
-    def train_DM(self,data,max_iters=10,max_steps=10,tol=0.01,eps=10e-6):
+    def train_DM(self,data,max_iters=10,max_steps=50,tol=0.01,eps=10e-6):
 
         stime = time.time()
         options = {'maxiter':max_iters,'disp':False}
@@ -632,7 +653,9 @@ class HHMM:
 
                 # optimize
                 if len(x0) > 0:
-                    res = minimize(loss_fn, x0, method='BFGS',options=options)
+                    stime = time.time()
+                    res = minimize(loss_fn, x0, method='Nelder-Mead',options=options)
+                    etime = time.time()
 
                     # update final values
                     x = np.copy(res['x'])
@@ -648,8 +671,9 @@ class HHMM:
 
                     print('updated coarse eta, row %d' % i)
                     print('original: ', backup)
-                    print('new: ', x)
-                    print(self.likelihood(data))
+                    print(res)
[…]

### 2021-06-07 Code/Run_Sim.py

@@ -88,19 +88,19 @@ ptm_crude = np.array([[0.85,0.15],
 eta_crude = np.array([[ 0.    , -1.6984],
                       [ 2.3471,  0.    ]])
 
-ptm_fine = [np.array([[0.75,0.25,0.00],
+ptm_fine = [np.array([[0.74,0.25,0.01],
                       [0.08,0.87,0.05],
-                      [0.00,0.23,0.77]]),
-            np.array([[0.89,0.11,0.00],
+                      [0.01,0.23,0.76]]),
+            np.array([[0.88,0.11,0.01],
                       [0.15,0.81,0.04],
-                      [0.00,0.23,0.77]])]
+                      [0.01,0.23,0.76]])]
 
-eta_fine =  [np.array([[  0.    ,  -1.0873, -22.6511],
-                    [ -2.3615,   0.    ,  -2.823 ],
-                    [-17.1832,  -1.1937,   0.    ]]),
-              np.array([[  0.    ,  -2.0691, -13.2605],
-                     [ -1.6902,   0.    ,  -3.1311],
-                     [-16.9396,  -1.2303,   0.    ]])]
+eta_fine =  [np.array([[  0.    ,  -1.0873, -10.00 ],
+                       [ -2.3615,   0.    , -2.823 ],
+                       [ -10.000,  -1.1937,  0.    ]]),
+              np.array([[  0.    ,  -2.0691,  -10.000],
+                        [ -1.6902,   0.    ,  -3.1311],
+                        [ -10.000,  -1.2303,   0.    ]])]
 
 # train new models?
 train_new = True
@@ -242,8 +242,7 @@ hmm_FV_theta = [{'dive_duration': {'mu': np.array([np.mean(dd_mu)]),
                          'sig': acc_sig,
                          'corr': logit(corr_fine)}}]]
 
-ptm_fine_temp = ptm_fine[0]/2 + ptm_fine[1]/2
-eta_fine_temp = HHMM.ptm_2_eta(ptm_fine_temp)
+eta_fine_temp = eta_fine[0]/2 + eta_fine[1]/2
 
 hmm_FV_eta = [np.array([[0]]),
              [eta_fine_temp]]

### 2021-06-07 Code/Train_Model.py

@@ -83,7 +83,7 @@ data = prep.get_all_features(df,dive_df)
 # train_model
 print('training model')
 hhmm = HHMM.HHMM(pars,data)
-hhmm.train_DM(data,max_iters=10,max_steps=10)
+hhmm.train_DM(data,max_iters=10,max_steps=50)
 
 # get SEs
 h = 0.01

### 2021-06-07 Code/Train_Model.sh

@@ -8,10 +8,10 @@
 #SBATCH --mail-type=REQUEUE
 #SBATCH --mail-type=ALL
 
-#SBATCH --time=96:00:00
+#SBATCH --time=72:00:00
 #SBATCH --array=1-400
 #SBATCH --mem-per-cpu=8G
-#SBATCH --cpus-per-task=8
+#SBATCH --cpus-per-task=1
 
 if [ $SLURM_ARRAY_TASK_ID -lt 101 ]
 then

### 2021-06-07 Code/run_all_sims.sh

@@ -9,7 +9,7 @@
 #SBATCH --mail-type=ALL
 
 #SBATCH --time=96:00:00
-#SBATCH --mem-per-cpu=4G
+#SBATCH --mem-per-cpu=8G
 #SBATCH --array=1-2000
 
 if [ $SLURM_ARRAY_TASK_ID -lt 501 ]

### 2021-06-07 Report/Sections/results.tex
Updates from Overleaf

@@
 \[\hat \Gamma = \begin{pmatrix} 
-0.788 & 0.212 \\
-0.809 & 0.191
+0.845 & 0.155 \\
+0.913 & 0.087
 \end{pmatrix} \text{ and }\]
 
-\[\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}\]
+\[\hat \delta = \begin{pmatrix} 0.855 & 0.145 \end{pmatrix}\]
 %
@@
 \[\hat \Gamma^{*(1)} = \begin{pmatrix} 
-0.679 & 0.321 & 0.000 \\
-0.038 & 0.904 & 0.058 \\
-0.000 & 0.232 & 0.768
+0.748 & 0.252 & 0.000 \\
+0.082 & 0.867 & 0.052 \\
+0.000 & 0.233 & 0.767
 \end{pmatrix}, \qquad 
 \hat \Gamma^{*(2)} = \begin{pmatrix} 
-0.859 & 0.141 & 0.000 \\
-0.114 & 0.841 & 0.045 \\
-0.000 & 0.216 & 0.784
+0.888 & 0.112 & 0.000 \\
+0.150 & 0.814 & 0.036 \\
+0.000 & 0.226 & 0.774
 \end{pmatrix},\]
 
-\[\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}\]
+\[\hat \delta^{*(1)} = \begin{pmatrix} 0.210 & 0.647 & 0.143 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.537 & 0.401 & 0.063 \end{pmatrix}\]
 %
 for dive types 1 and 2.
-In summary, about 79\% of dives are short dives of type 1, and the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1.
+In summary, about 86\% of dives are short dives of type 1, and the whale performs an average of 5.45 short type 1 dives before switching to dive type 2 and an average of 1.10 longer type 2 dives before switching back to dive type 1.
 This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
-Further, this killer whale is in the less active subdive state 1 40\% of the time during a dive of type 1 compared to only only 9\% of the time during a dive of type 2.
+Further, this killer whale is in the less active subdive state 1 54\% of the time during a dive of type 1 compared to only only 21\% of the time during a dive of type 2.
 Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}.

### 2021-06-08 Code/Run_Sim.py
update phi -> logit(phi)

@@ -80,20 +80,13 @@ acc_sig = np.array([0.05,0.1,0.3])
 corr_crude = [0.0,0.0,0.0]
 corr_fine = [0.97,0.83,0.61]
 
-# randomly initialize a probablity transition matrix
-
-ptm_crude = np.array([[0.85,0.15],
-                      [0.91,0.09]])
+# initialize a probablity transition matrix
 
 eta_crude = np.array([[ 0.    , -1.6984],
                       [ 2.3471,  0.    ]])
 
-ptm_fine = [np.array([[0.74,0.25,0.01],
-                      [0.08,0.87,0.05],
-                      [0.01,0.23,0.76]]),
-            np.array([[0.88,0.11,0.01],
-                      [0.15,0.81,0.04],
-                      [0.01,0.23,0.76]])]
+ptm_crude = HHMM.eta_2_ptm(ptm_crude)
+print(ptm_crude)
 
 eta_fine =  [np.array([[  0.    ,  -1.0873, -10.00 ],
                        [ -2.3615,   0.    , -2.823 ],
@@ -101,6 +94,11 @@ eta_fine =  [np.array([[  0.    ,  -1.0873, -10.00 ],
               np.array([[  0.    ,  -2.0691,  -10.000],
                         [ -1.6902,   0.    ,  -3.1311],
                         [ -10.000,  -1.2303,   0.    ]])]
+
+ptm_fine = [HHMM.eta_2_ptm(eta_fine[0]),
+            HHMM.eta_2_ptm(eta_fine[1])]
+
+print(ptm_fine)
 
 # train new models?
 train_new = True

### 2021-06-08 Report/Supp_Case.tex
Updates from Overleaf

% Initial and revised submissions should be 12 point; this will be removed in the final version.
%\documentclass[12pt]{TD-CJS}
\documentclass{article}

% Initial and revised submissions should also be double spaced.  This command will be removed in the final version.
%\renewcommand{\baselinestretch}{2}

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{psfrag}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{url}
\usepackage{float}
\usepackage[margin=1in]{geometry}
\input{header}
\input{defs}

\begin{document}
%\firstpage{1}
%\lastpage{25}
%\jvol{xx}
%\issue{yy}
%\jyear{2020}
%\jid{CJS}
%\aid{???}
% The running head contains the author names
%\rhauthor{BlindedA and BlindedB}
%\copyrightline{Statistical Society of Canada}
%\Frenchcopyrightline{Soci\'et\'e statistique du Canada}
% History: received and accepted dates
%\received{\rec{9}{July}{2009}}
%\accepted{\acc{8}{July}{2010}}

% User-defined commands go here
%\renewcommand{\eqref}[1]{(\ref{#1})}
%\newcommand{\mb}[1]{\mathbf{#1}}
%\newcommand{\mbb}[1]{\mathbb{#1}}
%\newcommand{\mt}[1]{\mathrm{#1}}
%\newcommand{\rv}{random variable}
%\newcommand{\newblock}{}
\bibliographystyle{abbrvnat}

% Title, authors, affiliations
\title{Modelling state-switching functional data with hidden Markov models: supplementary material (case study)}%\query{Q1}
\date{}
\author{BlindedA, BlindedB}%\authorref{1}\thanksref{*}}
%\author{BlindedB}%\authorref{2}}
%\affiliation[1]{Author affiliations will go here in the accepted manuscript, but do NOT include them in your initial submission because it must be anonymous.}
%\affiliation[2]{Second Affiliation}

% Abstract, keywords, and classification codes

\maketitle

\newcounter{tablenum}
\addtocounter{tablenum}{1}
\newcounter{fignum}
\addtocounter{fignum}{1}

    \section{Lag plots}

        \begin{center}
        \includegraphics[height=5in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_lagplot.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis. Each plot corresponds to a feature in the killer whale dive data. Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data. They can also be used to determine the number of hidden states if multiple distinct patterns are visible. 
        \addtocounter{fignum}{1}

        \newpage

    \section{Likelihood Surface - CarHHMM-DFT}

        \begin{center}
        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-1.png}
        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-2.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis. Each plot corresponds to a feature in the killer whale dive data. Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data. They can also be used to determine the number of hidden states if multiple distinct patterns are visible. 
        \addtocounter{fignum}{1}

        \begin{center}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-eta-likelihood.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis. Each plot corresponds to a feature in the killer whale dive data. Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data. They can also be used to determine the number of hidden states if multiple distinct patterns are visible. 
        \addtocounter{fignum}{1}

        % Ahat_low
        \begin{center}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-0.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-1.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-2.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis. Each plot corresponds to a feature in the killer whale dive data. Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data. They can also be used to determine the number of hidden states if multiple distinct patterns are visible. 
        \addtocounter{fignum}{1}

        % Ax
        \begin{center}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ax-0.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ax-1.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ax-2.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis. Each plot corresponds to a feature in the killer whale dive data. Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data. They can also be used to determine the number of hidden states if multiple distinct patterns are visible. 
        \addtocounter{fignum}{1}

        % Ay
        \begin{center}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ay-0.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ay-1.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ay-2.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis. Each plot corresponds to a feature in the killer whale dive data. Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data. They can also be used to determine the number of hidden states if multiple distinct patterns are visible. 
        \addtocounter{fignum}{1}

        % Az
        \begin{center}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Az-0.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Az-1.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Az-2.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis. Each plot corresponds to a feature in the killer whale dive data. Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data. They can also be used to determine the number of hidden states if multiple distinct patterns are visible. 
        \addtocounter{fignum}{1}

        % eta1
        \begin{center}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_0.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_1.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_2.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis. Each plot corresponds to a feature in the killer whale dive data. Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data. They can also be used to determine the number of hidden states if multiple distinct patterns are visible. 
        \addtocounter{fignum}{1}

        % eta1
        \begin{center}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_0.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_1.png}
        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_2.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis. Each plot corresponds to a feature in the killer whale dive data. Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data. They can also be used to determine the number of hidden states if multiple distinct patterns are visible. 
        \addtocounter{fignum}{1}

        \newpage

    \section{Parameter Estimates}

        \subsection{CarHHMM-DFT}

            %\begin{table}[ht]
            %
            %\caption{Estimates and empirical standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the full CarHHMM-DFT.}
            \begin{center}
            \scalebox{0.75}{
            \bgroup
            \centering
            \def\arraystretch{1.5}
            \begin{tabular}{ccccc}
            \multirow{2}{*}{Feature}                                                       & \multirow{2}{*}{Dive / Subdive Type} & \multicolumn{3}{c}{Parameter Estimate}        \\
                                                                                           &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
            \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.27 (0.63)$ & $10.91 (0.56)$ & ---           \\
                                                                                           & 2                                    & $127.1 (11.3)$ & $64.1 (9.0)$ & ---           \\ \hline
            \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.423 (0.028)$ & $0.040 (0.001)$ & $0.966 (0.002)$ \\
                                                                                           & 2                                    & $0.210 (0.012)$ & $0.097 (0.002)$ & $0.827 (0.010)$ \\
                                                                                           & 3                                    & $0.232 (0.035)$ & $0.298 (0.010)$ & $0.605 (0.043)$ \\ \hline
            \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.445 (0.036)$ & $0.051 (0.001)$ & $0.966 (0.009)$ \\
                                                                                           & 2                                    & $0.438 (0.012)$ & $0.095 (0.002)$ & $0.827 (0.016)$ \\
                                                                                           & 3                                    & $0.364 (0.043)$ & $0.368 (0.012)$ & $0.605 (0.041)$ \\ \hline
            \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.680 (0.041)$ & $0.058 (0.001)$ & $0.966 (0.005)$ \\
                                                                                           & 2                                    & $-0.573 (0.014)$ & $0.112 (0.002)$ & $0.827 (0.012)$ \\
                                                                                           & 3                                    & $-0.296 (0.042)$ & $0.357 (0.012)$ & $0.605 (0.039)$ \\ \hline
            \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $34.98 (0.38)$ & $23.85 (0.39)$ & ---           \\
                                                                                           & 2                                    & $505.8 (5.8)$ & $516.7 (7.0)$ & ---           \\
                                                                                           & 3                                    & $9770 (240)$ & $14460 (380)$ & ---           \\ \hline
            \end{tabular}
            \egroup
            }
            \end{center}

            \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the full CarHHMM-DFT. The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
            \addtocounter{tablenum}{1}

        \subsection{HHMM-DFT}
            \begin{center}
            \scalebox{0.75}{
                \bgroup
                \centering
                \def\arraystretch{1.5}
                \begin{tabular}{ccccc}
                \multirow{2}{*}{Feature}                                                       & \multirow{2}{*}{Dive / Subdive Type} & \multicolumn{3}{c}{Parameter Estimate}          \\
                                                                                               &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$     \\ \hline
                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $26.565 (0.607)$ & $10.118 (0.518)$ & ---             \\
                                                                                               & 2                                    & $116.134 (10.584)$ & $64.778 (8.543)$ & ---             \\ \hline
                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $-0.001 (0.002)$ & $0.064 (0.001)$ & ---             \\
                                                                                               & 2                                    & $0.084 (0.004)$ & $0.173 (0.003)$ & ---             \\
                                                                                               & 3                                    & $0.287 (0.020)$ & $0.481 (0.014)$ & ---             \\ \hline
                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.365 (0.002)$ & $0.067 (0.001)$ & ---             \\
                                                                                               & 2                                    & $0.408 (0.002)$ & $0.107 (0.002)$ & ---             \\
                                                                                               & 3                                    & $0.278 (0.016)$ & $0.403 (0.012)$ & ---             \\ \hline
                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.441 (0.003)$ & $0.101 (0.002)$ & ---             \\
                                                                                               & 2                                    & $-0.471 (0.003)$ & $0.157 (0.002)$ & ---             \\
                                                                                               & 3                                    & $-0.300 (0.017)$ & $0.412 (0.012)$ & ---             \\ \hline
                \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $42.149 (0.507)$ & $32.091 (0.554)$ & ---             \\
                                                                                               & 2                                    & $584.178 (6.491)$ & $687.082 (8.805)$ & ---             \\
                                                                                               & 3                                    & $6841.607 (148.113)$ & $13060.895 (293.086)$ & ---             \\ \hline
                \end{tabular}
                \egroup
            }
            \end{center}

            \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the HHMM-DFT. The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
            \addtocounter{tablenum}{1}

        \subsection{CarHHMM}

            \begin{center}
            \scalebox{0.75}{
                \bgroup
                \centering
                \def\arraystretch{1.5}
                \begin{tabular}{ccccc}
                \multirow{2}{*}{Feature}                                                       & \multirow{2}{*}{Dive / Subdive Type} & \multicolumn{3}{c}{Parameter Estimate}        \\
                                                                                               &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.170 (0.629)$ & $10.821 (0.552)$ & ---           \\
                                                                                               & 2                                    & $125.090 (11.003)$ & $64.488 (8.842)$ & ---           \\ \hline
                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.066 (0.017)$ & $0.035 (0.002)$ & $0.876 (0.006)$ \\
                                                                                               & 2                                    & $0.339 (0.037)$ & $0.095 (0.011)$ & $0.878 (0.037)$ \\
                                                                                               & 3                                    & $0.258 (0.007)$ & $0.305 (0.001)$ & $0.605 (0.010)$ \\ \hline
                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.377 (0.017)$ & $0.044 (0.002)$ & $0.876 (0.014)$ \\
                                                                                               & 2                                    & $0.469 (0.045)$ & $0.097 (0.013)$ & $0.878 (0.042)$ \\
                                                                                               & 3                                    & $0.349 (0.009)$ & $0.378 (0.001)$ & $0.605 (0.013)$ \\ \hline
                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.464 (0.020)$ & $0.052 (0.002)$ & $0.876 (0.010)$ \\
                                                                                               & 2                                    & $-0.658 (0.045)$ & $0.113 (0.013)$ & $0.878 (0.040)$ \\
                                                                                               & 3                                    & $-0.305 (0.011)$ & $0.370 (0.001)$ & $0.605 (0.012)$ \\ \hline
                \end{tabular}
                \egroup
            }
            \end{center}

                \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$) and acceleration ($\Zone_{t,\tilde t^*}$) of the killer whale case study data under the CarHHMM. The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
                \addtocounter{tablenum}{1}

        \subsection{CarHMM-DFT}

            \begin{center}
            \scalebox{0.75}{
                \bgroup
                \centering
                \def\arraystretch{1.5}
                \begin{tabular}{ccccc}
                \multirow{2}{*}{Feature}                                                       & \multirow{2}{*}{Dive / Subdive Type} & \multicolumn{3}{c}{Parameter Estimate}        \\
                                                                                               &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
                \multirow{1}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $41.875 (1.249)$ & $30.633 (1.230)$ & ---           \\ \hline
                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.455 (0.030)$ & $0.039 (0.001)$ & $0.968 (0.002)$ \\
                                                                                               & 2                                    & $0.214 (0.012)$ & $0.096 (0.002)$ & $0.832 (0.009)$ \\
                                                                                               & 3                                    & $0.222 (0.034)$ & $0.295 (0.010)$ & $0.604 (0.039)$ \\ \hline
                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.456 (0.039)$ & $0.051 (0.001)$ & $0.968 (0.009)$ \\
                                                                                               & 2                                    & $0.437 (0.012)$ & $0.094 (0.002)$ & $0.832 (0.016)$ \\
                                                                                               & 3                                    & $0.366 (0.042)$ & $0.364 (0.012)$ & $0.604 (0.040)$ \\ \hline
                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.700 (0.044)$ & $0.058 (0.001)$ & $0.968 (0.005)$ \\
                                                                                               & 2                                    & $-0.574 (0.014)$ & $0.111 (0.002)$ & $0.832 (0.012)$ \\
                                                                                               & 3                                    & $-0.302 (0.041)$ & $0.354 (0.012)$ & $0.604 (0.039)$ \\ \hline
                \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $33.883 (0.367)$ & $22.847 (0.376)$ & ---           \\
                                                                                               & 2                                    & $480.622 (5.479)$ & $492.201 (6.598)$ & ---           \\
                                                                                               & 3                                    & $9987.824 (239.804)$ & $14754.912 (375.450)$ & ---           \\ \hline
                \end{tabular}
                \egroup
            }
            \end{center}

                \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the CarHMM-DFT. The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix. Note that $\hat \mu_A^{*(\cdot,1)}$ is undefined because $\hat \phi_A^{*(\cdot,1)} = 1$, indicating that the mean of $\Zone_{t,\tilde t^*}$ is equal to $\Zone_{t,\tilde t^*-1}$ when in subdive state 1.
                \addtocounter{tablenum}{1}

    \newpage
    \section{Decoded dive states}

        \subsection{CarHHMM-DFT}

        \begin{center}
        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_decoded_dives.png}
        \end{center}

        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $\left(y^*_{t,t^*}\right)_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives. Each panel is partitioned into dives by vertical black lines. The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states. Both the dive types and subdive states are estimated by fitting the CarHHMM-DFT to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
        \addtocounter{fignum}{1}

        \begin{center}
        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_decoded_states.png}
        \end{center}

        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model. Each panel is partitioned into dives by vertical black lines. The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states. Both the dive types and subdive states are estimated by fitting the CarHHMM-DFT to the data and performing the forward-backward algorithm.
        \addtocounter{fignum}{1}

        \subsection{HHMM-DFT}

        \begin{center}
        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_decoded_dives.png}
        \end{center}

        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $\left(y^*_{t,t^*}\right)_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives. Each panel is partitioned into dives by vertical black lines. The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states. Both the dive types and subdive states are estimated by fitting the HHMM-DFT to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
        \addtocounter{fignum}{1}

        \begin{center}
        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_decoded_states.png}
        \end{center}

        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model. Each panel is partitioned into dives by vertical black lines. The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states. Both the dive types and subdive states are estimated by fitting the HHMM-DFT to the data and performing the forward-backward algorithm.
        \addtocounter{fignum}{1}

        \subsection{CarHHMM}

        \begin{center}
        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_decoded_dives.png}
        \end{center}

        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $\left(y^*_{t,t^*}\right)_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives. Each panel is partitioned into dives by vertical black lines. The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states. Both the dive types and subdive states are estimated by fitting the CarHHMM to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
        \addtocounter{fignum}{1}

        \begin{center}
        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_decoded_states.png}
        \end{center}

        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model. Each panel is partitioned into dives by vertical black lines. The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states. Both the dive types and subdive states are estimated by fitting the CarHHMM to the data and performing the forward-backward algorithm.
        \addtocounter{fignum}{1}

        \subsection{CarHMM-DFT}

        \begin{center}
        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_decoded_dives.png}
        \end{center}

        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $\left(y^*_{t,t^*}\right)_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives. Each panel is partitioned into dives by vertical black lines. The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states. Both the dive types and subdive states are estimated by fitting the CarHMM-DFT to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
        \addtocounter{fignum}{1}

        \begin{center}
        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_decoded_states.png}
        \end{center}

        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model. Each panel is partitioned into dives by vertical black lines. The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states. Both the dive types and subdive states are estimated by fitting the CarHMM-DFT to the data and performing the forward-backward algorithm.
        \addtocounter{fignum}{1}

    \newpage
    \section{Model checking - dive duration ($Y_t$)}

        \subsection{CarHHMM-DFT}

        \begin{center}
        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_dive_duration.png}
        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_pseudresids_Dive_Duration.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Empirical histogram (left) and psuedoresiduals (right) of dive duration ($Y_{t}$) plotted over the estimated emission distributions and a standard normal density, respectively. Both plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
        \addtocounter{fignum}{1}

        \subsection{HHMM-DFT}

        \begin{center}
        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_dive_duration.png}
        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_pseudresids_Dive_Duration.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Empirical histogram (left) and psuedoresiduals (right) of dive duration ($Y_{t}$) plotted over the estimated emission distributions and a standard normal density, respectively. Both plots are generated using the fitted HHMM-DFT and the killer whale case study data.
        \addtocounter{fignum}{1}

        \subsection{CarHHMM}

        \begin{center}
        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_dive_duration.png}
        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_pseudresids_Dive_Duration.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Empirical histogram (left) and psuedoresiduals (right) of dive duration ($Y_{t}$) plotted over the estimated emission distributions and a standard normal density, respectively. Both plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
        \addtocounter{fignum}{1}

        \subsection{CarHMM-DFT}

        \begin{center}
        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_dive_duration.png}
        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_pseudresids_Dive_Duration.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Empirical histogram (left) and psuedoresiduals (right) of dive duration ($Y_{t}$) plotted over the estimated emission distribution (left) and a standard normal density (right), respectively. Both plots are generated using the fitted CarHMM-DFT and the killer whale case study data.
        \addtocounter{fignum}{1}

    \newpage
    \section{Model checking - acceleration ($\Zone_{t,\tilde t^*}$)}

        \subsection{CarHHMM-DFT}

        \begin{center}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ax_0.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ay_0.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Az_0.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ax_1.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ay_1.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Az_1.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ax_2.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ay_2.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Az_2.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_pseudresids_Ax.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_pseudresids_Ay.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_pseudresids_Az.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively. Note that the mean of acceleration at time $\tilde t^*$ depends upon acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted. All plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
        \addtocounter{fignum}{1}

        \newpage

        \subsection{HHMM-DFT}

        \begin{center}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ax_0.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ay_0.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Az_0.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ax_1.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ay_1.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Az_1.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ax_2.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ay_2.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Az_2.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_pseudresids_Ax.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_pseudresids_Ay.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_pseudresids_Az.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively. All plots are generated using the fitted HHMM-DFT and the killer whale case study data.
        \addtocounter{fignum}{1}

        \newpage

        \subsection{CarHHMM}

        \begin{center}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ax_0.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ay_0.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Az_0.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ax_1.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ay_1.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Az_1.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ax_2.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ay_2.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Az_2.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_pseudresids_Ax.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_pseudresids_Ay.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_pseudresids_Az.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively. Note that the mean of acceleration at time $\tilde t^*$ depends upon acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted. All plots are generated using the fitted CarHHMM and the killer whale case study data.
        \addtocounter{fignum}{1}

        \newpage

        \subsection{CarHMM-DFT}

        \begin{center}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ax_0.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ay_0.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Az_0.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ax_1.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ay_1.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Az_1.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ax_2.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ay_2.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Az_2.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_pseudresids_Ax.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_pseudresids_Ay.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_pseudresids_Az.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively. Note that the mean of acceleration at time $\tilde t^*$ depends upon acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted. All plots are generated using the fitted CarHMM-DFT and the killer whale case study data.
        \addtocounter{fignum}{1}

    \section{Model checking - wiggliness ($\Ztwo_{t,\tilde t^*}$)}

        \subsection{CarHHMM-DFT}

        \begin{center}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_ahat_0.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_ahat_1.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_ahat_2.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_pseudresids_ahat.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and psuedoresiduals (bottom row) of wiggliness ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively. All plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
        \addtocounter{fignum}{1}

        \subsection{HHMM-DFT}

        \begin{center}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_ahat_0.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_ahat_1.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_ahat_2.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_pseudresids_ahat.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and psuedoresiduals (bottom row) of wiggliness ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively. All plots are generated using the fitted HHMM-DFT and the killer whale case study data.
        \addtocounter{fignum}{1}

        \subsection{CarHHMM}

        The CarHHMM does not model wiggliness.

        \subsection{CarHMM-DFT}

        \begin{center}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_ahat_0.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_ahat_1.png}
        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_ahat_2.png}

        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_pseudresids_ahat.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and psuedoresiduals (bottom row) of wiggliness ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively. All plots are generated using the fitted CarHMM-DFT and the killer whale case study data.
        \addtocounter{fignum}{1}

\iffalse
\newpage
%\section{Simulation Study Results}

    \section{Simulated Data}

        \begin{center}
    	\includegraphics[width=5in]{../Plots/sim_data.png}
    	\end{center}

    	\noindent Figure \arabic{fignum}: Acceleration data for a five dives of a selected simulated data set. The raw accelerometer data is denoted as $Y^*_{t,\tilde t^*}$ which is recorded at 50 hertz while $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are recorded at 0.5 hertz. Dives are separated by vertical black lines. The colour of the curve corresponds to the true fine-scale state, while the colour of the background corresponds to the true dive type. Subdive state 1 can plausibly be interpreted as ``gliding" while subdive state 2 can plausibly be interpreted as active swimming, or ``fluking".
    	\addtocounter{fignum}{1}

    \newpage
    \section{Accuracy comparisons}

        \begin{center}
        \scalebox{0.75}{
        \bgroup
        \def\arraystretch{1.5}
        \begin{tabular}{ccccccc}
Model                       & \multicolumn{1}{c}{Train Time (m)} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Dive Accuracy} & \multicolumn{1}{c}{Subdive Accuracy}  \\ \hline
\multirow{5}{*}{CarHMM-DFT} & \multirow{5}{*}{$70 \pm 11$}   & Both                          & Both                             & -------------                     & $0.934 \pm 0.005$                       \\
                            &                                    & 1                             & 1                                & \multirow{2}{*}{-------------}    & $0.794 \pm 0.037$                       \\ 
                            &                                    & 1                             & 2                                &                                   & $0.958 \pm 0.008$                       \\ 
                            &                                    & 2                             & 1                                & \multirow{2}{*}{-------------}    & $0.924 \pm 0.012$                       \\ 
                            &                                    & 2                             & 2                                &                                   & $0.941 \pm 0.011$                       \\ \hline 
\multirow{5}{*}{HHMM-DFT}   & \multirow{5}{*}{$209 \pm 51$}   & Both                          & Both                             & $0.942 \pm 0.036$                   & $0.881 \pm 0.041$                       \\
                            &                                    & 1                             & 1                                & \multirow{2}{*}{$0.966\pm0.030$}    & $0.628 \pm 0.120$                       \\ 
                            &                                    & 1                             & 2                                &                                   & $0.956 \pm 0.014$                       \\ 
                            &                                    & 2                             & 1                                & \multirow{2}{*}{$0.852\pm0.095$}    & $0.789 \pm 0.158$                       \\ 
                            &                                    & 2                             & 2                                &                                   & $0.918 \pm 0.028$                       \\ \hline
\multirow{5}{*}{CarHHMM}    & \multirow{5}{*}{$236 \pm 52$}   & Both                          & Both                             & $0.873 \pm 0.165$                   & $0.740 \pm 0.051$                       \\
                            &                                    & 1                             & 1                                & \multirow{2}{*}{$0.886\pm0.214$}    & $0.765 \pm 0.205$                       \\ 
                            &                                    & 1                             & 2                                &                                   & $0.707 \pm 0.090$                       \\ 
                            &                                    & 2                             & 1                                & \multirow{2}{*}{$0.823\pm0.136$}    & $0.873 \pm 0.225$                       \\ 
                            &                                    & 2                             & 2                                &                                   & $0.657 \pm 0.099$                       \\ \hline
\multirow{5}{*}{CarHHMM-DFT}& \multirow{5}{*}{$132 \pm 40$}   & Both                          & Both                             & $0.944 \pm 0.036$                   & $0.934 \pm 0.005$                       \\
                            &                                    & 1                             & 1                                & \multirow{2}{*}{$0.963\pm0.031$}    & $0.759 \pm 0.042$                       \\ 
                            &                                    & 1                             & 2                                &                                   & $0.962 \pm 0.008$                       \\ 
                            &                                    & 2                             & 1                                & \multirow{2}{*}{$0.872\pm0.086$}    & $0.934 \pm 0.010$                       \\ 
                            &                                    & 2                             & 2                                &                                   & $0.932 \pm 0.011$                       \\ \hline
\end{tabular}
        \egroup
        }
        \end{center}

        \noindent Table \arabic{tablenum}: Average decoding accuracies and training times for all models used to categorize dive type and subdive state in the simulation study. Each of the four models was fit to 500 training data sets comprised of 100 simulated dives and tested on test data sets also comprised of 100 simulated dives. Reported values are averages, and $\pm$ refers to the sample standard deviation across the 500 data sets. Rows labelled Both/Both correspond to overall average decoding accuracy.
        \addtocounter{tablenum}{1}

    \newpage
    \section{Empirical joint distributions - dive duration ($Y_t$)}

        \begin{center}
        \scalebox{0.75}{
        \bgroup
        \centering
        \def\arraystretch{1.5}
\begin{tabular}{ccccccc}
Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}   & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
\multirow{4}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu$}        & 1                             & $42.161$                         & ---          & $4.255$                             & $2.036 \pm 0.369$                             \\
                            &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\
                            & \multirow{2}{*}{$\sigma$}     & 1                             & $31.119$                         & ---          & $5.393$                             & $2.014 \pm 0.452$                             \\
                            &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\ \hline
\multirow{4}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu$}        & 1                             & $25.776$                         & $0.096$ $(p=0.071)$          & $1.181$                             & $0.959 \pm 0.111$                             \\
                            &                               & 2                             & $108.589$                         & $3.989$ $(p=0.000)$          & $18.726$                             & $13.083 \pm 4.546$                             \\
                            & \multirow{2}{*}{$\sigma$}     & 1                             & $9.489$                         & $-0.081$ $(p=0.058)$          & $0.952$                             & $0.779 \pm 0.102$                             \\
                            &                               & 2                             & $59.514$                         & $-5.186$ $(p=0.000)$          & $14.388$                             & $11.080 \pm 3.737$                             \\ \hline
\multirow{4}{*}{CarHHMM}    & \multirow{2}{*}{$\mu$}        & 1                             & $27.877$                         & $2.197$ $(p=0.000)$          & $11.693$                             & $0.966 \pm 0.150$                             \\
                            &                               & 2                             & $96.423$                         & $-8.177$ $(p=0.000)$          & $33.095$                             & $12.341 \pm 6.893$                             \\
                            & \multirow{2}{*}{$\sigma$}     & 1                             & $9.954$                         & $0.384$ $(p=0.347)$          & $8.950$                             & $0.793 \pm 0.142$                             \\
                            &                               & 2                             & $54.291$                         & $-10.409$ $(p=0.000)$          & $18.432$                             & $10.142 \pm 5.496$                             \\ \hline
\multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu$}        & 1                             & $25.719$                         & $0.039$ $(p=0.455)$          & $1.170$                             & $0.958 \pm 0.108$                             \\
                            &                               & 2                             & $107.404$                         & $2.804$ $(p=0.001)$          & $18.535$                             & $12.830 \pm 4.471$                             \\
                            & \multirow{2}{*}{$\sigma$}     & 1                             & $9.436$                         & $-0.134$ $(p=0.001)$          & $0.930$                             & $0.777 \pm 0.097$                             \\
                            &                               & 2                             & $59.635$                         & $-5.065$ $(p=0.000)$          & $14.225$                             & $10.945 \pm 3.548$                             
\end{tabular}
        \egroup
        }
        \end{center}

        \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of dive duration across all four models in the simulation study. The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations. The $\pm$ refers to the inter-quartile range. SE stands for standard error. Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
        \addtocounter{tablenum}{1}

        \subsection{CarHHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_dive_duration_-1_0.png}
        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_dive_duration_-1_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$) for each dive type. The red star represents the true values of $\mu$ and $\sigma$ for both dive types. These estimates are for the CarHHMM-DFT.
        \addtocounter{fignum}{1}

        \subsection{HHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_dive_duration_-1_0.png}
        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_dive_duration_-1_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$) for each dive type. The red star represents the true values of $\mu$ and $\sigma$ for both dive types. These estimates are for the HHMM-DFT.
        \addtocounter{fignum}{1}

        \subsection{CarHHMM}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hhmm_V_MLE_density_dive_duration_-1_0.png}
        \includegraphics[width=3in]{../Plots/hhmm_V_MLE_density_dive_duration_-1_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$) for each dive type. The red star represents the true values of $\mu$ and $\sigma$ for both dive types. These estimates are for the CarHHMM.
        \addtocounter{fignum}{1}

        \subsection{CarHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_dive_duration_-1_0.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$). The red star represents the true values of $\mu$ and $\sigma$ for both dive types. These estimates are for the CarHMM-DFT. The CarHMM-DFT assumes that there is only one dive type. There is no red star which represents the true values because the CarHMM-DFT assumes that there is only one dive type.
        \addtocounter{fignum}{1}

    \newpage
    \section{Empirical joint distributions - acceleration ($\Zone_{t,\tilde t^*}$)}

        \begin{center}
        \scalebox{0.75}{
        \bgroup
        \centering
        \def\arraystretch{1.5}
\begin{tabular}{ccccccc}
Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}   & \multicolumn{1}{c}{Observed Fischer SE}        \\ \hline
\multirow{6}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu_A^*$}    & 1                                & $-0.001$                         & $-0.001$ $(p=0.934)$          & $0.142$                             & $0.066 \pm 0.045$                             \\
                            &                               & 2                                & $-0.000$                         & $-0.000$ $(p=0.612)$          & $0.018$                             & $0.017 \pm 0.003$                             \\
                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.034$                         & $-0.016$ $(p=0.000)$          & $0.001$                             & $0.001 \pm 0.000$                             \\
                            &                               & 2                                & $0.079$                         & $-0.021$ $(p=0.000)$          & $0.002$                             & $0.002 \pm 0.000$                             \\ 
                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & $0.978$                         & $-0.012$ $(p=0.000)$          & $0.010$                             & $0.009 \pm 0.003$                             \\
                            &                               & 2                                & $0.868$                         & $-0.082$ $(p=0.000)$          & $0.016$                             & $0.016 \pm 0.002$                             \\ \hline
\multirow{6}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu_A^*$}    & 1                                & $-0.001$                         & $-0.001$ $(p=0.302)$          & $0.027$                             & $0.006 \pm 0.001$                             \\
                            &                               & 2                                & $0.000$                         & $0.000$ $(p=0.921)$          & $0.012$                             & $0.004 \pm 0.000$                             \\
                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.136$                         & $0.086$ $(p=0.000)$          & $0.022$                             & $0.004 \pm 0.001$                             \\
                            &                               & 2                                & $0.144$                         & $0.044$ $(p=0.000)$          & $0.007$                             & $0.003 \pm 0.000$                             \\ 
                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & ------                         & ------                     & ------                             & ------                                      \\
                            &                               & 2                                & ------                         & ------                     & ------                             & ------                                      \\ \hline
\multirow{6}{*}{CarHHMM}    & \multirow{2}{*}{$\mu_A^*$}    & 1                                & $0.021$                         & $0.021$ $(p=0.436)$          & $0.591$                             & $0.016 \pm 0.004$                             \\
                            &                               & 2                                & $0.000$                         & $0.000$ $(p=0.669)$          & $0.013$                             & $0.006 \pm 0.001$                             \\
                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.065$                         & $0.015$ $(p=0.000)$          & $0.022$                             & $0.001 \pm 0.000$                             \\
                            &                               & 2                                & $0.187$                         & $0.087$ $(p=0.000)$          & $0.018$                             & $0.004 \pm 0.000$                             \\ 
                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & $0.891$                         & $-0.099$ $(p=0.000)$          & $0.155$                             & $0.010 \pm 0.002$                             \\
                            &                               & 2                                & $0.182$                         & $-0.768$ $(p=0.000)$          & $0.137$                             & $0.027 \pm 0.002$                             \\ \hline
\multirow{6}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_A^*$}    & 1                                & $-0.001$                         & $-0.001$ $(p=0.875)$          & $0.111$                             & $0.066 \pm 0.045$                             \\
                            &                               & 2                                & $-0.000$                         & $-0.000$ $(p=0.564)$          & $0.018$                             & $0.017 \pm 0.003$                             \\
                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.034$                         & $-0.016$ $(p=0.000)$          & $0.001$                             & $0.001 \pm 0.000$                             \\
                            &                               & 2                                & $0.079$                         & $-0.021$ $(p=0.000)$          & $0.002$                             & $0.002 \pm 0.000$                             \\ 
                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & $0.978$                         & $-0.012$ $(p=0.000)$          & $0.010$                             & $0.009 \pm 0.003$                             \\
                            &                               & 2                                & $0.867$                         & $-0.083$ $(p=0.000)$          & $0.016$                             & $0.016 \pm 0.002$                             \\ \hline
\end{tabular}
        \egroup
        }
        \end{center}

        \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of acceleration across all four models in the simulation study. The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations. The $\pm$ refers to the inter-quartile range. SE stands for standard error. Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
        \addtocounter{tablenum}{1}

        \subsection{CarHHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_A_0_0.png}
        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_A_0_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of acceleration parameters ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state. The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states. These estimates are for the CarHHMM-DFT.
        \addtocounter{fignum}{1}

        \subsection{HHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_A_0_0.png}
        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_A_0_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of acceleration parameters ($\hat \mu^*_A$ and $\hat \sigma^*_A$) for each subdive state. The red star represents the true values of $\mu^*_A$ and $\sigma^*_A$ for both subdive states. These estimates are for the HHMM-DFT. Note that the HHMM-DFT assumes no auto-correlation in $\Zone_{t,\tilde t^*}$, so $\hat \phi_A^{*(\cdot,i^*)}$ is absent from these plots.
        \addtocounter{fignum}{1}

        \subsubsection{CarHHMM}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hhmm_V_MLE_density_A_0_0.png}
        \includegraphics[width=3in]{../Plots/hhmm_V_MLE_density_A_0_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of acceleration parameters ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state. The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states. These estimates are for the CarHHMM. 
        \addtocounter{fignum}{1}

        \subsubsection{CarHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_A_0_0.png}
        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_A_0_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of acceleration parameters ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state. The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states. These estimates are for the CarHMM-DFT.
        \addtocounter{fignum}{1}

    \newpage
    \subsection{Empirical joint distributions - wiggliness ($\Ztwo_{t,\tilde t^*}$)}

        \begin{center}
        \scalebox{0.75}{
        \bgroup
        \centering
        \def\arraystretch{1.5}
\begin{tabular}{ccccccc}
Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}       \\ \hline
\multirow{4}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu_W^*$}    & 1                                & $23.362$                         & $0.022$ $(p=0.408)$          & $0.582$                             & $0.419 \pm 0.061$                             \\
                            &                               & 2                                & $300.689$                         & $-0.511$ $(p=0.213)$          & $9.144$                             & $4.679 \pm 0.336$                             \\
                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & $12.949$                         & $-0.001$ $(p=0.974)$          & $0.539$                             & $0.388 \pm 0.061$                             \\
                            &                               & 2                                & $329.585$                         & $-0.515$ $(p=0.280)$          & $10.622$                             & $5.748 \pm 0.396$                             \\ \hline
\multirow{4}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu_W^*$}    & 1                                & $23.353$                         & $0.013$ $(p=0.632)$          & $0.620$                             & $0.431 \pm 0.067$                             \\
                            &                               & 2                                & $298.441$                         & $-2.759$ $(p=0.000)$          & $10.438$                             & $4.788 \pm 0.432$                             \\
                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & $12.938$                         & $-0.012$ $(p=0.656)$          & $0.583$                             & $0.404 \pm 0.064$                             \\
                            &                               & 2                                & $328.496$                         & $-1.604$ $(p=0.001)$          & $10.482$                             & $5.822 \pm 0.460$                             \\ \hline
\multirow{4}{*}{CarHHMM}    & \multirow{2}{*}{$\mu_W^*$}    & 1                                & ------                         & ------                     & ------                             & ------                                      \\
                            &                               & 2                                & ------                         & ------                     & ------                             & ------                                      \\
                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & ------                         & ------                     & ------                             & ------                                      \\
                            &                               & 2                                & ------                         & ------                     & ------                             & ------                                      \\ \hline
\multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_W^*$}    & 1                                & $23.341$                         & $0.001$ $(p=0.957)$          & $0.568$                             & $0.419 \pm 0.060$                             \\
                            &                               & 2                                & $301.209$                         & $0.009$ $(p=0.982)$          & $8.869$                             & $4.697 \pm 0.353$                             \\
                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & $12.918$                         & $-0.032$ $(p=0.183)$          & $0.526$                             & $0.387 \pm 0.059$                             \\
                            &                               & 2                                & $329.684$                         & $-0.416$ $(p=0.365)$          & $10.225$                             & $5.773 \pm 0.396$                             
\end{tabular}
        \egroup
        }
        \end{center}

        \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of wiggliness across all four models in the simulation study. The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations. The $\pm$ refers to the inter-quartile range. SE stands for standard error. Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
        \addtocounter{tablenum}{1}

        \subsection{CarHHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_FoVeDBA_0_0.png}
        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_FoVeDBA_0_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of wiggliness parameters ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state. The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states. These estimates are for the CarHHMM-DFT.
        \addtocounter{fignum}{1}

        \subsection{HHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_FoVeDBA_0_0.png}
        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_FoVeDBA_0_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of wiggliness parameters ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state. The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states. These estimates are for the HHMM-DFT.
        \addtocounter{fignum}{1}

        \subsection{CarHHMM}
        The CarHHMM did not use wiggliness to model the whale's behaviour.

        \subsection{CarHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_FoVeDBA_0_0.png}
        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_FoVeDBA_0_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of wiggliness parameters ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state. The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states. These estimates are for the CarHMM-DFT.
        \addtocounter{fignum}{1}

    \newpage
    \section{Empirical joint distributions - probability transition matrices ($\Gamma, \Gamma^*$)}

        \begin{center}
        \scalebox{0.75}{
        \bgroup
        \centering
        \def\arraystretch{1.5}
\begin{tabular}{ccccccc}
Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}    & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}  & \multicolumn{1}{c}{Observed Fischer SE}       \\ \hline
\multirow{6}{*}{CarHMM-DFT}  & $\Gamma_{12}$                 & ------                          & ------                     & ------                            & ------                                        \\
                             & $\Gamma_{21}$                 & ------                          & ------                     & ------                            & ------                                        \\
                             & $\Gamma^{*(1)}_{12}$          & $0.179$                         & ------                     & $0.018$                           & $0.015 \pm 0.002$                             \\
                             & $\Gamma^{*(1)}_{21}$          & $0.079$                         & ------                     & $0.011$                           & $0.008 \pm 0.001$                             \\
                             & $\Gamma^{*(2)}_{12}$          & ------                          & ------                     & ------                            & ------                                        \\
                             & $\Gamma^{*(2)}_{21}$          & ------                          & ------                     & ------                            & ------                                        \\ \hline
\multirow{6}{*}{HHMM-DFT}    & $\Gamma_{12}$                 & $0.206$                         & $-0.006$ $(p=0.033)$       & $0.063$                           & $0.049 \pm 0.011$                             \\
                             & $\Gamma_{21}$                 & $0.822$                         & $0.013$ $(p=0.014)$        & $0.119$                           & $0.098 \pm 0.029$                             \\
                             & $\Gamma^{*(1)}_{12}$          & $0.312$                         & $-0.008$ $(p=0.011)$       & $0.070$                           & $0.048 \pm 0.019$                             \\
                             & $\Gamma^{*(1)}_{21}$          & $0.047$                         & $-0.003$ $(p=0.000)$       & $0.014$                           & $0.010 \pm 0.002$                             \\
                             & $\Gamma^{*(2)}_{12}$          & $0.139$                         & $-0.001$ $(p=0.158)$       & $0.018$                           & $0.017 \pm 0.003$                             \\
                             & $\Gamma^{*(2)}_{21}$          & $0.146$                         & $-0.004$ $(p=0.001)$       & $0.025$                           & $0.019 \pm 0.004$                             \\ \hline
\multirow{6}{*}{CarHHMM}     & $\Gamma_{12}$                 & $0.307$                         & $0.095$ $(p=0.000)$        & $0.257$                           & $0.052 \pm 0.021$                             \\
                             & $\Gamma_{21}$                 & $0.711$                         & $-0.098$ $(p=0.000)$       & $0.280$                           & $0.102 \pm 0.040$                             \\
                             & $\Gamma^{*(1)}_{12}$          & $0.269$                         & $-0.051$ $(p=0.000)$       & $0.252$                           & $0.019 \pm 0.006$                             \\
                             & $\Gamma^{*(1)}_{21}$          & $0.134$                         & $0.084$ $(p=0.000)$        & $0.081$                           & $0.016 \pm 0.003$                             \\
                             & $\Gamma^{*(2)}_{12}$          & $0.120$                         & $-0.020$ $(p=0.025)$       & $0.196$                           & $0.008 \pm 0.002$                             \\
                             & $\Gamma^{*(2)}_{21}$          & $0.144$                         & $-0.006$ $(p=0.001)$       & $0.042$                           & $0.019 \pm 0.006$                             \\ \hline
\multirow{6}{*}{CarHHMM-DFT} & $\Gamma_{12}$                 & $0.210$                         & $-0.002$ $(p=0.473)$       & $0.064$                           & $0.050 \pm 0.011$                             \\
                             & $\Gamma_{21}$                 & $0.818$                         & $0.009$ $(p=0.076)$        & $0.117$                           & $0.097 \pm 0.027$                             \\
                             & $\Gamma^{*(1)}_{12}$          & $0.325$                         & $0.005$ $(p=0.132)$        & $0.069$                           & $0.048 \pm 0.018$                             \\
                             & $\Gamma^{*(1)}_{21}$          & $0.050$                         & $0.000$ $(p=0.713)$        & $0.012$                           & $0.010 \pm 0.002$                             \\
                             & $\Gamma^{*(2)}_{12}$          & $0.141$                         & $0.001$ $(p=0.336)$        & $0.017$                           & $0.016 \pm 0.003$                             \\
                             & $\Gamma^{*(2)}_{21}$          & $0.152$                         & $0.002$ $(p=0.032)$        & $0.020$                           & $0.018 \pm 0.004$                             \\ \hline

\end{tabular}
        \egroup
        }
        \end{center}

        \noindent Table \arabic{tablenum}: Estimates and standard errors for all probability transition matrices across all four models in the simulation study. The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations. The $\pm$ refers to the inter-quartile range. SE stands for standard error. Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
        \addtocounter{tablenum}{1}

        \newpage
        \subsection{CarHHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hhmm_FV_Gamma_density_-1.png}

        \includegraphics[width=3in]{../Plots/hhmm_FV_Gamma_density_0.png}
        \includegraphics[width=3in]{../Plots/hhmm_FV_Gamma_density_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM-DFT. There are two fine-scale probability transition matrices- one corresponding to each dive type. We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix. The red star represents the true values.
        \addtocounter{fignum}{1}

        \newpage
        \subsection{HHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_Gamma_density_-1.png}

        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_Gamma_density_0.png}
        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_Gamma_density_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the HHMM-DFT. There are two fine-scale probability transition matrices- one corresponding to each dive type. We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix. The red star represents the true values.
        \addtocounter{fignum}{1}

        \newpage
        \subsection{CarHHMM}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hhmm_V_Gamma_density_-1.png}

        \includegraphics[width=3in]{../Plots/hhmm_V_Gamma_density_0.png}
        \includegraphics[width=3in]{../Plots/hhmm_V_Gamma_density_1.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM. There are two fine-scale probability transition matrices- one corresponding to each dive type. We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix. The red star represents the true values.
        \addtocounter{fignum}{1}

        \newpage
        \subsection{CarHMM-DFT}
        \begin{center}
        \includegraphics[width=3in]{../Plots/hmm_FV_Gamma_density_0.png}
        \end{center}

        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution $\hat \Gamma^{*(1)}$ for the CarHMM-DFT. There is only one fine-scale probability transition matrix and no coarse-scale probability transition matrix since we assume that there is only one dive type for the CarHMM-DFT. We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix. There is no red star which represents the true values because the CarHMM-DFT assumes that there is only one dive type.
        \addtocounter{fignum}{1}
\fi
\end{document}

### 2021-06-17 Code/HHMM.py
Add plots

@@ -123,21 +123,10 @@ class HHMM:
 
                 feature_data = [datum[feature] for datum in data]
 
-                # first find mu
-                theta[0][feature]['mu'] = np.mean(feature_data)*np.ones(K)
-
-                # then get varaince of each quantile set of data
-                data_sorted = np.sort(feature_data)
-                n = len(data_sorted)
-                for k0 in range(K):
-                    if settings['f'] != 'vonmises':
-                        std = np.std(data_sorted[int(k0*n/K):int((k0+1)*n/K)])
-                        theta[0][feature]['sig'][k0] = max(0.1,std)
-                    else:
-                        theta[0][feature]['sig'][k0] = 1.0
-
-                # initialize correlations
-                theta[0][feature]['corr'] = -1.0 * np.ones(K)
+                # first find basic parameter estimates
+                theta[0][feature]['mu'] = np.ones(K)*np.mean(feature_data)
+                theta[0][feature]['sig'] = np.ones(K)*max(0.1,np.std(feature_data))
+                theta[0][feature]['corr'] = np.ones(K)*-1.0
 
                 # add randomness in initialization
                 if settings['f'] == 'normal':
@@ -145,7 +134,8 @@ class HHMM:
                     theta[0][feature]['sig'] *= np.exp(norm.rvs(0.0,1.0,size=K))
                     theta[0][feature]['corr'] += norm.rvs(0.0,1.0,size=K)
                 elif settings['f'] == 'gamma':
-                    theta[0][feature]['mu'] *= np.exp(norm.rvs(0.0,1.0,size=K))
+                    log_std = np.std(np.log(feature_data))
+                    theta[0][feature]['mu'] *= np.exp(norm.rvs(0.0,log_std,size=K))
                     theta[0][feature]['sig'] *= np.exp(norm.rvs(0.0,1.0,size=K))
                     theta[0][feature]['corr'] += norm.rvs(0.0,1.0,size=K)
                 else:
@@ -199,22 +189,67 @@ class HHMM:
 
     def reorder_params(self):
 
-        state_order = np.argsort(self.theta[1][0]['Ax']['sig'])
+        # do coarse-scale states
+        state_order = np.argsort(self.theta[0]['dive_duration']['mu'])
+
+        # first reorder eta
+        eta_coarse = deepcopy(self.eta[0])
[…]

### 2021-06-17 Code/Run_Sim.py

@@ -48,8 +48,8 @@ K0 = 2
 K1 = 3
 
 # dive duration parameters
-dd_mu = np.array([27.27,127.07])
-dd_sig = np.array([10.91,64.06])
+dd_mu = np.array([27.34,127.55])
+dd_sig = np.array([10.96,63.89])
 
 dd_shape = np.square(dd_mu)/np.square(dd_sig)
 dd_scale = np.square(dd_sig)/np.array(dd_mu)
@@ -65,8 +65,8 @@ for i in range(K1):
 FoVeDBA_sin_scale = np.ones((K1,51))
 
 # set mean and variance
-FoVeDBA_mu = np.array([[34.98],[505.79],[9769.98]])
-FoVeDBA_sig = np.array([[23.85],[516.68],[14462.55]])
+FoVeDBA_mu = np.array([[34.01],[490.06],[9154.16]])
+FoVeDBA_sig = np.array([[22.99],[502.56],[13538.75]])
 
 # adjust shape and scale accordingly
 FoVeDBA_sin_shape *= np.square(FoVeDBA_mu)/np.square(FoVeDBA_sig)
@@ -82,18 +82,18 @@ corr_fine = [0.97,0.83,0.61]
 
 # initialize a probablity transition matrix
 
-eta_crude = np.array([[ 0.    , -1.6984],
-                      [ 2.3471,  0.    ]])
+eta_crude = np.array([[ 0.    , -1.7127],
+                      [ 2.3629,  0.    ]])
 
-ptm_crude = HHMM.eta_2_ptm(ptm_crude)
+ptm_crude = HHMM.eta_2_ptm(eta_crude)
 print(ptm_crude)
 
-eta_fine =  [np.array([[  0.    ,  -1.0873, -10.00 ],
-                       [ -2.3615,   0.    , -2.823 ],
-                       [ -10.000,  -1.1937,  0.    ]]),
-              np.array([[  0.    ,  -2.0691,  -10.000],
-                        [ -1.6902,   0.    ,  -3.1311],
-                        [ -10.000,  -1.2303,   0.    ]])]
+eta_fine =  [np.array([[  0.    ,  -1.078 , -5.9046],
+                       [ -2.3897,   0.    , -2.819 ],
+                       [ -10.000,  -1.2166,  0.    ]]),
+              np.array([[  0.    ,  -2.0510,  -10.000],
[…]

### 2021-06-17 Code/Train_Model.py

@@ -83,7 +83,7 @@ data = prep.get_all_features(df,dive_df)
 # train_model
 print('training model')
 hhmm = HHMM.HHMM(pars,data)
-hhmm.train_DM(data,max_iters=10,max_steps=50)
+hhmm.train_DM(data,max_iters=25,max_steps=25,max_time=90)
 
 # get SEs
 h = 0.01
@@ -95,6 +95,7 @@ print('labelling df')
 data,df = hhmm.label_df(data,df)
 
 # save everything
+hhmm.data = None
 hhmm.save(HHMM_file)
-with open(data_outfile, 'wb') as f:
-    pickle.dump(data, f)
+#with open(data_outfile, 'wb') as f:
+#    pickle.dump(data, f)

### 2021-06-17 Code/Train_Model.sh

@@ -8,9 +8,9 @@
 #SBATCH --mail-type=REQUEUE
 #SBATCH --mail-type=ALL
 
-#SBATCH --time=72:00:00
-#SBATCH --array=1-400
+#SBATCH --time=96:00:00
 #SBATCH --mem-per-cpu=8G
+#SBATCH --array=1-400
 #SBATCH --cpus-per-task=1
 
 if [ $SLURM_ARRAY_TASK_ID -lt 101 ]

### 2021-06-17 Code/run_all_sims.sh

@@ -8,9 +8,9 @@
 #SBATCH --mail-type=REQUEUE
 #SBATCH --mail-type=ALL
 
-#SBATCH --time=96:00:00
+#SBATCH --time=12:00:00
 #SBATCH --mem-per-cpu=8G
-#SBATCH --array=1-2000
+#SBATCH --array=501-2000
 
 if [ $SLURM_ARRAY_TASK_ID -lt 501 ]
 then

### 2021-06-17 Report/Supp_Case.tex
Update Report

@@
         \begin{center}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-1.png}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-2.png}
+        %\includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-1.png}
+        %\includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-2.png}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-eta-likelihood.png}
+        %\includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-eta-likelihood.png}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-0.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-1.png}
+        %\includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-0.png}
+        %\includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-1.png}
         \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-2.png}
@@
                                                                                                &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
-            \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.27 (0.63)$ & $10.91 (0.56)$ & ---           \\
-                                                                                           & 2                                    & $127.1 (11.3)$ & $64.1 (9.0)$ & ---           \\ \hline
-            \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.423 (0.028)$ & $0.040 (0.001)$ & $0.966 (0.002)$ \\
-                                                                                           & 2                                    & $0.210 (0.012)$ & $0.097 (0.002)$ & $0.827 (0.010)$ \\
-                                                                                           & 3                                    & $0.232 (0.035)$ & $0.298 (0.010)$ & $0.605 (0.043)$ \\ \hline
-            \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.445 (0.036)$ & $0.051 (0.001)$ & $0.966 (0.009)$ \\
-                                                                                           & 2                                    & $0.438 (0.012)$ & $0.095 (0.002)$ & $0.827 (0.016)$ \\
-                                                                                           & 3                                    & $0.364 (0.043)$ & $0.368 (0.012)$ & $0.605 (0.041)$ \\ \hline
-            \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.680 (0.041)$ & $0.058 (0.001)$ & $0.966 (0.005)$ \\
-                                                                                           & 2                                    & $-0.573 (0.014)$ & $0.112 (0.002)$ & $0.827 (0.012)$ \\
-                                                                                           & 3                                    & $-0.296 (0.042)$ & $0.357 (0.012)$ & $0.605 (0.039)$ \\ \hline
-            \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $34.98 (0.38)$ & $23.85 (0.39)$ & ---           \\
-                                                                                           & 2                                    & $505.8 (5.8)$ & $516.7 (7.0)$ & ---           \\
-                                                                                           & 3                                    & $9770 (240)$ & $14460 (380)$ & ---           \\ \hline
+                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.34 (0.63)$ & $10.96 (0.56)$ & ---           \\
+                                                                                               & 2                                    & $127.5 (11.3)$ & $63.9 (9.0)$ & ---           \\ \hline
+                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.449 (0.030)$ & $0.039 (0.001)$ & $0.968 (0.002)$ \\
+                                                                                               & 2                                    & $0.210 (0.012)$ & $0.096 (0.002)$ & $0.829 (0.010)$ \\
+                                                                                               & 3                                    & $0.232 (0.035)$ & $0.296 (0.010)$ & $0.607 (0.042)$ \\ \hline
+                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.450 (0.038)$ & $0.051 (0.001)$ & $0.968 (0.009)$ \\
+                                                                                               & 2                                    & $0.437 (0.012)$ & $0.094 (0.002)$ & $0.829 (0.016)$ \\
+                                                                                               & 3                                    & $0.366 (0.042)$ & $0.365 (0.012)$ & $0.607 (0.040)$ \\ \hline
+                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.691 (0.043)$ & $0.058 (0.001)$ & $0.968 (0.005)$ \\
+                                                                                               & 2                                    & $-0.573 (0.014)$ & $0.111 (0.002)$ & $0.829 (0.012)$ \\
+                                                                                               & 3                                    & $-0.303 (0.041)$ & $0.354 (0.012)$ & $0.607 (0.039)$ \\ \hline
+                \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $34.02 (0.36)$ & $22.99 (0.38)$ & ---           \\
+                                                                                               & 2                                    & $490.1 (5.6)$ & $502.6 (6.8)$ & ---           \\
+                                                                                               & 3                                    & $9150 (220)$ & $13540 (350)$ & ---           \\ \hline
             \end{tabular}
@@
                                                                                                &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$     \\ \hline
-                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $26.565 (0.607)$ & $10.118 (0.518)$ & ---             \\
-                                                                                               & 2                                    & $116.134 (10.584)$ & $64.778 (8.543)$ & ---             \\ \hline
-                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $-0.001 (0.002)$ & $0.064 (0.001)$ & ---             \\
-                                                                                               & 2                                    & $0.084 (0.004)$ & $0.173 (0.003)$ & ---             \\
-                                                                                               & 3                                    & $0.287 (0.020)$ & $0.481 (0.014)$ & ---             \\ \hline
-                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.365 (0.002)$ & $0.067 (0.001)$ & ---             \\
-                                                                                               & 2                                    & $0.408 (0.002)$ & $0.107 (0.002)$ & ---             \\
-                                                                                               & 3                                    & $0.278 (0.016)$ & $0.403 (0.012)$ & ---             \\ \hline
-                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.441 (0.003)$ & $0.101 (0.002)$ & ---             \\
-                                                                                               & 2                                    & $-0.471 (0.003)$ & $0.157 (0.002)$ & ---             \\
-                                                                                               & 3                                    & $-0.300 (0.017)$ & $0.412 (0.012)$ & ---             \\ \hline
-                \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $42.149 (0.507)$ & $32.091 (0.554)$ & ---             \\
-                                                                                               & 2                                    & $584.178 (6.491)$ & $687.082 (8.805)$ & ---             \\
-                                                                                               & 3                                    & $6841.607 (148.113)$ & $13060.895 (293.086)$ & ---             \\ \hline
+                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $26.61 (0.61)$ & $10.16 (0.52)$ & ---             \\
+                                                                                               & 2                                    & $116.8 (10.6)$ & $64.7 (8.6)$ & ---             \\ \hline
+                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.000 (0.002)$ & $0.065 (0.001)$ & ---             \\
+                                                                                               & 2                                    & $0.085 (0.004)$ & $0.175 (0.003)$ & ---             \\
+                                                                                               & 3                                    & $0.291 (0.020)$ & $0.484 (0.014)$ & ---             \\ \hline
+                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.365 (0.002)$ & $0.068 (0.001)$ & ---             \\
+                                                                                               & 2                                    & $0.408 (0.002)$ & $0.108 (0.002)$ & ---             \\
+                                                                                               & 3                                    & $0.275 (0.017)$ & $0.407 (0.012)$ & ---             \\ \hline
+                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.442 (0.003)$ & $0.101 (0.002)$ & ---             \\
+                                                                                               & 2                                    & $-0.471 (0.003)$ & $0.158 (0.002)$ & ---             \\
+                                                                                               & 3                                    & $-0.298 (0.017)$ & $0.415 (0.012)$ & ---             \\ \hline
+                \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $43.65 (0.52)$ & $33.66 (0.58)$ & ---             \\
+                                                                                               & 2                                    & $608.0 (6.8)$ & $717.3 (9.2)$ & ---             \\
+                                                                                               & 3                                    & $7100 (160)$ & $13620 (310)$ & ---             \\ \hline
                 \end{tabular}
@@
                                                                                                &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
-                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.170 (0.629)$ & $10.821 (0.552)$ & ---           \\
-                                                                                               & 2                                    & $125.090 (11.003)$ & $64.488 (8.842)$ & ---           \\ \hline
-                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.066 (0.017)$ & $0.035 (0.002)$ & $0.876 (0.006)$ \\
-                                                                                               & 2                                    & $0.339 (0.037)$ & $0.095 (0.011)$ & $0.878 (0.037)$ \\
-                                                                                               & 3                                    & $0.258 (0.007)$ & $0.305 (0.001)$ & $0.605 (0.010)$ \\ \hline
-                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.377 (0.017)$ & $0.044 (0.002)$ & $0.876 (0.014)$ \\
-                                                                                               & 2                                    & $0.469 (0.045)$ & $0.097 (0.013)$ & $0.878 (0.042)$ \\
-                                                                                               & 3                                    & $0.349 (0.009)$ & $0.378 (0.001)$ & $0.605 (0.013)$ \\ \hline
-                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.464 (0.020)$ & $0.052 (0.002)$ & $0.876 (0.010)$ \\
-                                                                                               & 2                                    & $-0.658 (0.045)$ & $0.113 (0.013)$ & $0.878 (0.040)$ \\
-                                                                                               & 3                                    & $-0.305 (0.011)$ & $0.370 (0.001)$ & $0.605 (0.012)$ \\ \hline
+                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.16 (0.63)$ & $10.78 (0.55)$ & ---           \\
+                                                                                               & 2                                    & $124.2 (11.0)$ & $64.6 (8.8)$ & ---           \\ \hline
+                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.064 (0.007)$ & $0.035 (0.001)$ & $0.874 (0.008)$ \\
+                                                                                               & 2                                    & $0.342 (0.017)$ & $0.095 (0.002)$ & $0.879 (0.005)$ \\
+                                                                                               & 3                                    & $0.257 (0.037)$ & $0.305 (0.010)$ & $0.605 (0.023)$ \\ \hline
+                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.376 (0.009)$ & $0.044 (0.001)$ & $0.874 (0.008)$ \\
+                                                                                               & 2                                    & $0.470 (0.017)$ & $0.097 (0.002)$ & $0.879 (0.005)$ \\
+                                                                                               & 3                                    & $0.350 (0.045)$ & $0.377 (0.013)$ & $0.605 (0.023)$ \\ \hline
+                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.462 (0.011)$ & $0.052 (0.001)$ & $0.874 (0.008)$ \\
+                                                                                               & 2                                    & $-0.660 (0.020)$ & $0.113 (0.002)$ & $0.879 (0.005)$ \\
+                                                                                               & 3                                    & $-0.306 (0.045)$ & $0.369 (0.013)$ & $0.605 (0.023)$ \\ \hline
                 \end{tabular}
@@
         \subsection{CarHMM-DFT}
-
             \begin{center}
@@
                                                                                            &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
-                \multirow{1}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $41.875 (1.249)$ & $30.633 (1.230)$ & ---           \\ \hline
-                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.455 (0.030)$ & $0.039 (0.001)$ & $0.968 (0.002)$ \\
-                                                                                               & 2                                    & $0.214 (0.012)$ & $0.096 (0.002)$ & $0.832 (0.009)$ \\
-                                                                                               & 3                                    & $0.222 (0.034)$ & $0.295 (0.010)$ & $0.604 (0.039)$ \\ \hline
-                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.456 (0.039)$ & $0.051 (0.001)$ & $0.968 (0.009)$ \\
-                                                                                               & 2                                    & $0.437 (0.012)$ & $0.094 (0.002)$ & $0.832 (0.016)$ \\
-                                                                                               & 3                                    & $0.366 (0.042)$ & $0.364 (0.012)$ & $0.604 (0.040)$ \\ \hline
-                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.700 (0.044)$ & $0.058 (0.001)$ & $0.968 (0.005)$ \\
-                                                                                               & 2                                    & $-0.574 (0.014)$ & $0.111 (0.002)$ & $0.832 (0.012)$ \\
-                                                                                               & 3                                    & $-0.302 (0.041)$ & $0.354 (0.012)$ & $0.604 (0.039)$ \\ \hline
-                \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $33.883 (0.367)$ & $22.847 (0.376)$ & ---           \\
-                                                                                               & 2                                    & $480.622 (5.479)$ & $492.201 (6.598)$ & ---           \\
-                                                                                               & 3                                    & $9987.824 (239.804)$ & $14754.912 (375.450)$ & ---           \\ \hline
+            \multirow{1}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $41.70 (1.24)$ & $30.50 (1.22)$ & ---           \\ \hline
+            \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.460 (0.030)$ & $0.039 (0.001)$ & $0.968 (0.002)$ \\
+                                                                                           & 2                                    & $0.209 (0.012)$ & $0.096 (0.002)$ & $0.828 (0.007)$ \\
+                                                                                           & 3                                    & $0.232 (0.035)$ & $0.296 (0.010)$ & $0.607 (0.023)$ \\ \hline
+            \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.457 (0.038)$ & $0.052 (0.001)$ & $0.968 (0.002)$ \\
+                                                                                           & 2                                    & $0.436 (0.012)$ & $0.094 (0.002)$ & $0.828 (0.007)$ \\
+                                                                                           & 3                                    & $0.365 (0.042)$ & $0.366 (0.012)$ & $0.607 (0.023)$ \\ \hline
+            \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.704 (0.044)$ & $0.059 (0.001)$ & $0.968 (0.002)$ \\
+                                                                                           & 2                                    & $-0.571 (0.014)$ & $0.111 (0.002)$ & $0.828 (0.007)$ \\
+                                                                                           & 3                                    & $-0.301 (0.042)$ & $0.355 (0.012)$ & $0.607 (0.023)$ \\ \hline
+            \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $34.40 (0.37)$ & $23.33 (0.38)$ & ---           \\
+                                                                                           & 2                                    & $495.0 (5.7)$ & $505.4 (6.9)$ & ---           \\
+                                                                                           & 3                                    & $9190 (220)$ & $13590 (360)$ & ---           \\ \hline
             \end{tabular}

### 2021-06-17 Report/Supp_Case.tex

@@
 
-    \section{Likelihood Surface - CarHHMM-DFT}
-
-        \begin{center}
-        %\includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-1.png}
-        %\includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-2.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
-
-        \begin{center}
-        %\includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-eta-likelihood.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
+    \section{Likelihood Surfaces - CarHHMM-DFT}
+
+        \begin{center}
+        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-1.png}
+        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-2.png}
+        \end{center}
+
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the dive duration parameters $\mu_Y^{(i)}$ and $\sigma_Y^{(i)}$ for dive types $i = 1,2$.
+All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
+The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $(\hat \mu_Y,\hat \sigma_Y)$, which is denoted by a red star.
+        \addtocounter{fignum}{1}
+
+        \begin{center}
+        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-eta-likelihood.png}
+        \end{center}
+
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the coarse-scale probability transition matrix parameters $\eta_{1,2}$ and $\eta_{2,1}$.
+Recall that $\Gamma_{i,j} = \exp(\eta_{i,j}) / \sum_k \exp(\eta_{i,k})$.
+All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
+The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $(\hat \eta_{1,2},\hat \eta_{2,1})$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
         \begin{center}
-        %\includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-0.png}
-        %\includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-1.png}
+        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-0.png}
+        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-1.png}
         \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-2.png}
@@
 
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the wiggliness parameters $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
+The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $(\hat \mu_W^{*(\cdot,i^*)}, \hat \sigma_W^{*(\cdot,i^*)})$, which is denoted by a red star.
+        \addtocounter{fignum}{1}
+
+        \newpage
 
@@
 
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $x$-acceleration parameters $\mu_{A_x}^{*(\cdot,i^*)}, \sigma_{A_x}^{*(\cdot,i^*)}$ and $\text{logit}\left(\phi_{A}^{*(\cdot,i^*)}\right)$ for subdive types $i^* = 1,2,3$.
+All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
+The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_x}^{*(\cdot,i^*)}, \hat \sigma_{A_x}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $y$-acceleration parameters $\mu_{A_y}^{*(\cdot,i^*)}, \sigma_{A_y}^{*(\cdot,i^*)}$ and $\text{logit}\left(\phi_{A}^{*(\cdot,i^*)}\right)$ for subdive types $i^* = 1,2,3$.
+All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
+The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_y}^{*(\cdot,i^*)}, \hat \sigma_{A_y}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $z$-acceleration parameters $\mu_{A_z}^{*(\cdot,i^*)}, \sigma_{A_z}^{*(\cdot,i^*)}$ and $\text{logit}\left(\phi_{A}^{*(\cdot,i^*)}\right)$ for subdive types $i^* = 1,2,3$.
+All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
+The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_z}^{*(\cdot,i^*)}, \hat \sigma_{A_z}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
+        \addtocounter{fignum}{1}
+
+        \newpage
 
@@
         \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_0.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_1.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_2.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
-
-        % eta1
-        \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_0.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_1.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_2.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_0.png}
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_0.png}
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_1.png}
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_1.png}
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_2.png}
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_2.png}
+        \end{center}
+
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the fine-scale probability transition matrix parameters associated both dive types ($\eta^{(i)*}_{i^*,j^*}$ for $i = 1,2$ and $i^*,j^* = 1,2,3$).
+Recall that $\eta^{(i)*}_{i^*,i^*} = 0$ and that $\Gamma^{(i)*}_{i^*,j^*} = \exp(\eta^{(i)*}_{i^*,j^*}) / \sum_{k^*} \exp(\eta^{(i)*}_{i^*,k^*})$.
+All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
+The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $\hat \Gamma^{(i)*}$, which is denoted by a red star.
+Note that the long ridges of approximately equal likelihood corresponds to $\hat \Gamma^{(i)*}_{1,3} \approx \hat \Gamma^{(i)*}_{3,1} \approx 0$.
+This indicates that the whale never switches directly between subdive states 1 and 3.
         \addtocounter{fignum}{1}

### 2021-06-17 Report/Supp_Sim.tex
Updates from Overleaf

--- 2021-06-08 Report/Supp_Case.tex
@@
 % Title, authors, affiliations
-\title{Modelling state-switching functional data with hidden Markov models: supplementary material (case study)}%\query{Q1}
+\title{Modelling state-switching functional data with hidden Markov models: supplementary material (simulation study)}%\query{Q1}
 \date{}
@@
 
-    \section{Lag plots}
-
-        \begin{center}
-        \includegraphics[height=5in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_lagplot.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
-
-        \newpage
-
-    \section{Likelihood Surface - CarHHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-1.png}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-theta-likelihood-dive_duration-2.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
-
-        \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-eta-likelihood.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
-
-        % Ahat_low
-        \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-0.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-1.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ahat_low-2.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
-
-        % Ax
-        \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ax-0.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ax-1.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ax-2.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
-
-        % Ay
-        \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ay-0.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ay-1.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Ay-2.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
-
-        % Az
-        \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Az-0.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Az-1.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-theta-likelihood-Az-2.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
-
-        % eta1
-        \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_0.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_1.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_2.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
-
-        % eta1
-        \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_0.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_1.png}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_2.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
-Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
-        \addtocounter{fignum}{1}
-
-        \newpage
-
-    \section{Parameter Estimates}
-
-        \subsection{CarHHMM-DFT}
-
-            %\begin{table}[ht]
-            %
-            %\caption{Estimates and empirical standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the full CarHHMM-DFT.}
-            \begin{center}
-            \scalebox{0.75}{
-            \bgroup
-            \centering
-            \def\arraystretch{1.5}
-            \begin{tabular}{ccccc}
-            \multirow{2}{*}{Feature}                                                       & \multirow{2}{*}{Dive / Subdive Type} & \multicolumn{3}{c}{Parameter Estimate}        \\
-                                                                                           &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
-            \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.27 (0.63)$ & $10.91 (0.56)$ & ---           \\
-                                                                                           & 2                                    & $127.1 (11.3)$ & $64.1 (9.0)$ & ---           \\ \hline
-            \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.423 (0.028)$ & $0.040 (0.001)$ & $0.966 (0.002)$ \\
-                                                                                           & 2                                    & $0.210 (0.012)$ & $0.097 (0.002)$ & $0.827 (0.010)$ \\
-                                                                                           & 3                                    & $0.232 (0.035)$ & $0.298 (0.010)$ & $0.605 (0.043)$ \\ \hline
-            \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.445 (0.036)$ & $0.051 (0.001)$ & $0.966 (0.009)$ \\
-                                                                                           & 2                                    & $0.438 (0.012)$ & $0.095 (0.002)$ & $0.827 (0.016)$ \\
-                                                                                           & 3                                    & $0.364 (0.043)$ & $0.368 (0.012)$ & $0.605 (0.041)$ \\ \hline
-            \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.680 (0.041)$ & $0.058 (0.001)$ & $0.966 (0.005)$ \\
-                                                                                           & 2                                    & $-0.573 (0.014)$ & $0.112 (0.002)$ & $0.827 (0.012)$ \\
-                                                                                           & 3                                    & $-0.296 (0.042)$ & $0.357 (0.012)$ & $0.605 (0.039)$ \\ \hline
-            \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $34.98 (0.38)$ & $23.85 (0.39)$ & ---           \\
-                                                                                           & 2                                    & $505.8 (5.8)$ & $516.7 (7.0)$ & ---           \\
-                                                                                           & 3                                    & $9770 (240)$ & $14460 (380)$ & ---           \\ \hline
-            \end{tabular}
-            \egroup
-            }
-            \end{center}
-
-            \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the full CarHHMM-DFT.
-The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
-            \addtocounter{tablenum}{1}
-
-        \subsection{HHMM-DFT}
-            \begin{center}
-            \scalebox{0.75}{
-                \bgroup
-                \centering
-                \def\arraystretch{1.5}
-                \begin{tabular}{ccccc}
-                \multirow{2}{*}{Feature}                                                       & \multirow{2}{*}{Dive / Subdive Type} & \multicolumn{3}{c}{Parameter Estimate}          \\
-                                                                                               &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$     \\ \hline
-                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $26.565 (0.607)$ & $10.118 (0.518)$ & ---             \\
-                                                                                               & 2                                    & $116.134 (10.584)$ & $64.778 (8.543)$ & ---             \\ \hline
-                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $-0.001 (0.002)$ & $0.064 (0.001)$ & ---             \\
-                                                                                               & 2                                    & $0.084 (0.004)$ & $0.173 (0.003)$ & ---             \\
-                                                                                               & 3                                    & $0.287 (0.020)$ & $0.481 (0.014)$ & ---             \\ \hline
-                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.365 (0.002)$ & $0.067 (0.001)$ & ---             \\
-                                                                                               & 2                                    & $0.408 (0.002)$ & $0.107 (0.002)$ & ---             \\
-                                                                                               & 3                                    & $0.278 (0.016)$ & $0.403 (0.012)$ & ---             \\ \hline
-                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.441 (0.003)$ & $0.101 (0.002)$ & ---             \\
-                                                                                               & 2                                    & $-0.471 (0.003)$ & $0.157 (0.002)$ & ---             \\
-                                                                                               & 3                                    & $-0.300 (0.017)$ & $0.412 (0.012)$ & ---             \\ \hline
-                \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $42.149 (0.507)$ & $32.091 (0.554)$ & ---             \\
-                                                                                               & 2                                    & $584.178 (6.491)$ & $687.082 (8.805)$ & ---             \\
-                                                                                               & 3                                    & $6841.607 (148.113)$ & $13060.895 (293.086)$ & ---             \\ \hline
-                \end{tabular}
-                \egroup
-            }
-            \end{center}
-
-            \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the HHMM-DFT.
-The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
-            \addtocounter{tablenum}{1}
-
-        \subsection{CarHHMM}
-
-            \begin{center}
-            \scalebox{0.75}{
-                \bgroup
-                \centering
-                \def\arraystretch{1.5}
-                \begin{tabular}{ccccc}
-                \multirow{2}{*}{Feature}                                                       & \multirow{2}{*}{Dive / Subdive Type} & \multicolumn{3}{c}{Parameter Estimate}        \\
-                                                                                               &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
-                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.170 (0.629)$ & $10.821 (0.552)$ & ---           \\
-                                                                                               & 2                                    & $125.090 (11.003)$ & $64.488 (8.842)$ & ---           \\ \hline
-                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.066 (0.017)$ & $0.035 (0.002)$ & $0.876 (0.006)$ \\
-                                                                                               & 2                                    & $0.339 (0.037)$ & $0.095 (0.011)$ & $0.878 (0.037)$ \\
-                                                                                               & 3                                    & $0.258 (0.007)$ & $0.305 (0.001)$ & $0.605 (0.010)$ \\ \hline
-                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.377 (0.017)$ & $0.044 (0.002)$ & $0.876 (0.014)$ \\
-                                                                                               & 2                                    & $0.469 (0.045)$ & $0.097 (0.013)$ & $0.878 (0.042)$ \\
-                                                                                               & 3                                    & $0.349 (0.009)$ & $0.378 (0.001)$ & $0.605 (0.013)$ \\ \hline
-                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.464 (0.020)$ & $0.052 (0.002)$ & $0.876 (0.010)$ \\
-                                                                                               & 2                                    & $-0.658 (0.045)$ & $0.113 (0.013)$ & $0.878 (0.040)$ \\
-                                                                                               & 3                                    & $-0.305 (0.011)$ & $0.370 (0.001)$ & $0.605 (0.012)$ \\ \hline
-                \end{tabular}
-                \egroup
-            }
-            \end{center}
-
-                \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$) and acceleration ($\Zone_{t,\tilde t^*}$) of the killer whale case study data under the CarHHMM.
-The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
-                \addtocounter{tablenum}{1}
-
-        \subsection{CarHMM-DFT}
-
-            \begin{center}
-            \scalebox{0.75}{
-                \bgroup
-                \centering
-                \def\arraystretch{1.5}
-                \begin{tabular}{ccccc}
-                \multirow{2}{*}{Feature}                                                       & \multirow{2}{*}{Dive / Subdive Type} & \multicolumn{3}{c}{Parameter Estimate}        \\
-                                                                                               &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
-                \multirow{1}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $41.875 (1.249)$ & $30.633 (1.230)$ & ---           \\ \hline
-                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.455 (0.030)$ & $0.039 (0.001)$ & $0.968 (0.002)$ \\
-                                                                                               & 2                                    & $0.214 (0.012)$ & $0.096 (0.002)$ & $0.832 (0.009)$ \\
-                                                                                               & 3                                    & $0.222 (0.034)$ & $0.295 (0.010)$ & $0.604 (0.039)$ \\ \hline
-                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.456 (0.039)$ & $0.051 (0.001)$ & $0.968 (0.009)$ \\
-                                                                                               & 2                                    & $0.437 (0.012)$ & $0.094 (0.002)$ & $0.832 (0.016)$ \\
-                                                                                               & 3                                    & $0.366 (0.042)$ & $0.364 (0.012)$ & $0.604 (0.040)$ \\ \hline
-                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.700 (0.044)$ & $0.058 (0.001)$ & $0.968 (0.005)$ \\
-                                                                                               & 2                                    & $-0.574 (0.014)$ & $0.111 (0.002)$ & $0.832 (0.012)$ \\
-                                                                                               & 3                                    & $-0.302 (0.041)$ & $0.354 (0.012)$ & $0.604 (0.039)$ \\ \hline
-                \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $33.883 (0.367)$ & $22.847 (0.376)$ & ---           \\
-                                                                                               & 2                                    & $480.622 (5.479)$ & $492.201 (6.598)$ & ---           \\
-                                                                                               & 3                                    & $9987.824 (239.804)$ & $14754.912 (375.450)$ & ---           \\ \hline
-                \end{tabular}
-                \egroup
-            }
-            \end{center}
-
-                \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the CarHMM-DFT.
-The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
-Note that $\hat \mu_A^{*(\cdot,1)}$ is undefined because $\hat \phi_A^{*(\cdot,1)} = 1$, indicating that the mean of $\Zone_{t,\tilde t^*}$ is equal to $\Zone_{t,\tilde t^*-1}$ when in subdive state 1.
-                \addtocounter{tablenum}{1}
-
-    \newpage
-    \section{Decoded dive states}
-
-        \subsection{CarHHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_decoded_dives.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $\left(y^*_{t,t^*}\right)_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
-Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHHMM-DFT to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
-        \addtocounter{fignum}{1}
-
-        \begin{center}
-        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_decoded_states.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model.
-Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHHMM-DFT to the data and performing the forward-backward algorithm.
-        \addtocounter{fignum}{1}
-
-        \subsection{HHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_decoded_dives.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $\left(y^*_{t,t^*}\right)_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
-Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the HHMM-DFT to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
-        \addtocounter{fignum}{1}
-
-        \begin{center}
-        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_decoded_states.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model.
-Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the HHMM-DFT to the data and performing the forward-backward algorithm.
-        \addtocounter{fignum}{1}
-
-        \subsection{CarHHMM}
-
-        \begin{center}
-        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_decoded_dives.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $\left(y^*_{t,t^*}\right)_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
-Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHHMM to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
-        \addtocounter{fignum}{1}
-
-        \begin{center}
-        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_decoded_states.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model.
-Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHHMM to the data and performing the forward-backward algorithm.
-        \addtocounter{fignum}{1}
-
-        \subsection{CarHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_decoded_dives.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $\left(y^*_{t,t^*}\right)_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
-Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHMM-DFT to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
-        \addtocounter{fignum}{1}
-
-        \begin{center}
-        \includegraphics[width=6in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_decoded_states.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model.
-Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHMM-DFT to the data and performing the forward-backward algorithm.
-        \addtocounter{fignum}{1}
-
-    \newpage
-    \section{Model checking - dive duration ($Y_t$)}
-
-        \subsection{CarHHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_dive_duration.png}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_pseudresids_Dive_Duration.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Empirical histogram (left) and psuedoresiduals (right) of dive duration ($Y_{t}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-Both plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
-        \addtocounter{fignum}{1}
-
-        \subsection{HHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_dive_duration.png}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_pseudresids_Dive_Duration.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Empirical histogram (left) and psuedoresiduals (right) of dive duration ($Y_{t}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-Both plots are generated using the fitted HHMM-DFT and the killer whale case study data.
-        \addtocounter{fignum}{1}
-
-        \subsection{CarHHMM}
-
-        \begin{center}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_dive_duration.png}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_pseudresids_Dive_Duration.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Empirical histogram (left) and psuedoresiduals (right) of dive duration ($Y_{t}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-Both plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
-        \addtocounter{fignum}{1}
-
-        \subsection{CarHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_dive_duration.png}
-        \includegraphics[width=2.25in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_pseudresids_Dive_Duration.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Empirical histogram (left) and psuedoresiduals (right) of dive duration ($Y_{t}$) plotted over the estimated emission distribution (left) and a standard normal density (right), respectively.
-Both plots are generated using the fitted CarHMM-DFT and the killer whale case study data.
-        \addtocounter{fignum}{1}
-
-    \newpage
-    \section{Model checking - acceleration ($\Zone_{t,\tilde t^*}$)}
-
-        \subsection{CarHHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ax_0.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ay_0.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Az_0.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ax_1.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ay_1.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Az_1.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ax_2.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Ay_2.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_Az_2.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_pseudresids_Ax.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_pseudresids_Ay.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_pseudresids_Az.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-Note that the mean of acceleration at time $\tilde t^*$ depends upon acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted.
-All plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
-        \addtocounter{fignum}{1}
-
-        \newpage
-
-        \subsection{HHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ax_0.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ay_0.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Az_0.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ax_1.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ay_1.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Az_1.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ax_2.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Ay_2.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_Az_2.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_pseudresids_Ax.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_pseudresids_Ay.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_pseudresids_Az.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-All plots are generated using the fitted HHMM-DFT and the killer whale case study data.
-        \addtocounter{fignum}{1}
-
-        \newpage
-
-        \subsection{CarHHMM}
-
-        \begin{center}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ax_0.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ay_0.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Az_0.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ax_1.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ay_1.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Az_1.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ax_2.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Ay_2.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_empirical_hist_Az_2.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_pseudresids_Ax.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_pseudresids_Ay.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM1_pseudresids_Az.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-Note that the mean of acceleration at time $\tilde t^*$ depends upon acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted.
-All plots are generated using the fitted CarHHMM and the killer whale case study data.
-        \addtocounter{fignum}{1}
-
-        \newpage
-
-        \subsection{CarHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ax_0.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ay_0.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Az_0.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ax_1.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ay_1.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Az_1.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ax_2.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Ay_2.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_Az_2.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_pseudresids_Ax.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_pseudresids_Ay.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_pseudresids_Az.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-Note that the mean of acceleration at time $\tilde t^*$ depends upon acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted.
-All plots are generated using the fitted CarHMM-DFT and the killer whale case study data.
-        \addtocounter{fignum}{1}
-
-    \section{Model checking - wiggliness ($\Ztwo_{t,\tilde t^*}$)}
-
-        \subsection{CarHHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_ahat_0.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_ahat_1.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_empirical_hist_ahat_2.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_pseudresids_ahat.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and psuedoresiduals (bottom row) of wiggliness ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-All plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
-        \addtocounter{fignum}{1}
-
-        \subsection{HHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_ahat_0.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_ahat_1.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_empirical_hist_ahat_2.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_HHMM_pseudresids_ahat.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and psuedoresiduals (bottom row) of wiggliness ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-All plots are generated using the fitted HHMM-DFT and the killer whale case study data.
-        \addtocounter{fignum}{1}
-
-        \subsection{CarHHMM}
-
-        The CarHHMM does not model wiggliness.
-
-        \subsection{CarHMM-DFT}
-
-        \begin{center}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_ahat_0.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_ahat_1.png}
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_empirical_hist_ahat_2.png}
-
-        \includegraphics[width=1.75in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHMM_pseudresids_ahat.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and psuedoresiduals (bottom row) of wiggliness ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-All plots are generated using the fitted CarHMM-DFT and the killer whale case study data.
-        \addtocounter{fignum}{1}
-
-\iffalse
-\newpage
 %\section{Simulation Study Results}
@@
 Model                       & \multicolumn{1}{c}{Train Time (m)} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Dive Accuracy} & \multicolumn{1}{c}{Subdive Accuracy}  \\ \hline
-\multirow{5}{*}{CarHMM-DFT} & \multirow{5}{*}{$70 \pm 11$}   & Both                          & Both                             & -------------                     & $0.934 \pm 0.005$                       \\
-                            &                                    & 1                             & 1                                & \multirow{2}{*}{-------------}    & $0.794 \pm 0.037$                       \\
-                            &                                    & 1                             & 2                                &                                   & $0.958 \pm 0.008$                       \\
-                            &                                    & 2                             & 1                                & \multirow{2}{*}{-------------}    & $0.924 \pm 0.012$                       \\
-                            &                                    & 2                             & 2                                &                                   & $0.941 \pm 0.011$                       \\ \hline
-\multirow{5}{*}{HHMM-DFT}   & \multirow{5}{*}{$209 \pm 51$}   & Both                          & Both                             & $0.942 \pm 0.036$                   & $0.881 \pm 0.041$                       \\
-                            &                                    & 1                             & 1                                & \multirow{2}{*}{$0.966\pm0.030$}    & $0.628 \pm 0.120$                       \\
-                            &                                    & 1                             & 2                                &                                   & $0.956 \pm 0.014$                       \\
-                            &                                    & 2                             & 1                                & \multirow{2}{*}{$0.852\pm0.095$}    & $0.789 \pm 0.158$                       \\
-                            &                                    & 2                             & 2                                &                                   & $0.918 \pm 0.028$                       \\ \hline
-\multirow{5}{*}{CarHHMM}    & \multirow{5}{*}{$236 \pm 52$}   & Both                          & Both                             & $0.873 \pm 0.165$                   & $0.740 \pm 0.051$                       \\
-                            &                                    & 1                             & 1                                & \multirow{2}{*}{$0.886\pm0.214$}    & $0.765 \pm 0.205$                       \\
-                            &                                    & 1                             & 2                                &                                   & $0.707 \pm 0.090$                       \\
-                            &                                    & 2                             & 1                                & \multirow{2}{*}{$0.823\pm0.136$}    & $0.873 \pm 0.225$                       \\
-                            &                                    & 2                             & 2                                &                                   & $0.657 \pm 0.099$                       \\ \hline
-\multirow{5}{*}{CarHHMM-DFT}& \multirow{5}{*}{$132 \pm 40$}   & Both                          & Both                             & $0.944 \pm 0.036$                   & $0.934 \pm 0.005$                       \\
-                            &                                    & 1                             & 1                                & \multirow{2}{*}{$0.963\pm0.031$}    & $0.759 \pm 0.042$                       \\
-                            &                                    & 1                             & 2                                &                                   & $0.962 \pm 0.008$                       \\
-                            &                                    & 2                             & 1                                & \multirow{2}{*}{$0.872\pm0.086$}    & $0.934 \pm 0.010$                       \\
-                            &                                    & 2                             & 2                                &                                   & $0.932 \pm 0.011$                       \\ \hline
+\multirow{7}{*}{CarHHMM-DFT}& \multirow{7}{*}{$156 \pm 67$}   & All                           & All                              & $0.956 \pm 0.028$                   & $0.911 \pm 0.006$                       \\
+                            &                                    & 1                             & 1                                & \multirow{3}{*}{$0.973\pm0.032$}    & $0.846 \pm 0.027$                       \\
+                            &                                    & 1                             & 2                                &                                   & $0.918 \pm 0.011$                       \\
+                            &                                    & 1                             & 3                                &                                   & $0.860 \pm 0.030$                       \\
+                            &                                    & 2                             & 1                                & \multirow{3}{*}{$0.856\pm0.097$}    & $0.949 \pm 0.010$                       \\
+                            &                                    & 2                             & 2                                &                                   & $0.915 \pm 0.015$                       \\
+                            &                                    & 2                             & 3                                &                                   & $0.871 \pm 0.057$                       \\ \hline
+\multirow{7}{*}{HHMM-DFT}   & \multirow{7}{*}{$162 \pm 64$}   & All                           & All                              & $0.959 \pm 0.025$                   & $0.844 \pm 0.024$                       \\
+                            &                                    & 1                             & 1                                & \multirow{3}{*}{$0.977\pm0.028$}    & $0.761 \pm 0.063$                       \\
+                            &                                    & 1                             & 2                                &                                   & $0.845 \pm 0.029$                       \\
+                            &                                    & 1                             & 3                                &                                   & $0.858 \pm 0.034$                       \\
+                            &                                    & 2                             & 1                                & \multirow{3}{*}{$0.851\pm0.097$}    & $0.883 \pm 0.066$                       \\
+                            &                                    & 2                             & 2                                &                                   & $0.829 \pm 0.041$                       \\
+                            &                                    & 2                             & 3                                &                                   & $0.876 \pm 0.061$                       \\ \hline
+\multirow{7}{*}{CarHHMM}    & \multirow{7}{*}{$258 \pm 106$}   & All                           & All                              & $0.924 \pm 0.127$                   & $0.845 \pm 0.018$                       \\
+                            &                                    & 1                             & 1                                & \multirow{3}{*}{$0.933\pm0.153$}    & $0.689 \pm 0.050$                       \\
+                            &                                    & 1                             & 2                                &                                   & $0.900 \pm 0.031$                       \\
+                            &                                    & 1                             & 3                                &                                   & $0.673 \pm 0.059$                       \\
+                            &                                    & 2                             & 1                                & \multirow{3}{*}{$0.864\pm0.134$}    & $0.906 \pm 0.022$                       \\
+                            &                                    & 2                             & 2                                &                                   & $0.848 \pm 0.033$                       \\
+                            &                                    & 2                             & 3                                &                                   & $0.687 \pm 0.105$                       \\ \hline
+\multirow{7}{*}{CarHMM-DFT} & \multirow{7}{*}{$45 \pm 22$}   & All                           & All                              & ---                                   & $0.912 \pm 0.006$                       \\
+                            &                                    & 1                             & 1                                & \multirow{3}{*}{---          }    & $0.854 \pm 0.026$                       \\
+                            &                                    & 1                             & 2                                &                                   & $0.921 \pm 0.011$                       \\
+                            &                                    & 1                             & 3                                &                                   & $0.860 \pm 0.032$                       \\
+                            &                                    & 2                             & 1                                & \multirow{3}{*}{---          }    & $0.943 \pm 0.011$                       \\
+                            &                                    & 2                             & 2                                &                                   & $0.919 \pm 0.014$                       \\
+                            &                                    & 2                             & 3                                &                                   & $0.878 \pm 0.053$                       \\ \hline
 \end{tabular}
@@
 Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}   & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
-\multirow{4}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu$}        & 1                             & $42.161$                         & ---          & $4.255$                             & $2.036 \pm 0.369$                             \\
+\multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.337$                         & $0.067$ $(p=0.287)$          & $1.393$                             & $1.025 \pm 0.124$                             \\
+                            &                               & 2                             & $127.140$                         & $0.070$ $(p=0.945)$          & $22.562$                             & $15.710 \pm 6.726$                             \\
+                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.801$                         & $-0.109$ $(p=0.035)$          & $1.154$                             & $0.838 \pm 0.110$                             \\
+                            &                               & 2                             & $59.520$                         & $-4.540$ $(p=0.000)$          & $17.678$                             & $13.040 \pm 5.530$                             \\ \hline
+\multirow{4}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.373$                         & $0.103$ $(p=0.084)$          & $1.327$                             & $1.030 \pm 0.113$                             \\
+                            &                               & 2                             & $126.843$                         & $-0.227$ $(p=0.818)$          & $22.004$                             & $15.367 \pm 6.105$                             \\
+                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.860$                         & $-0.050$ $(p=0.351)$          & $1.187$                             & $0.845 \pm 0.110$                             \\
+                            &                               & 2                             & $57.136$                         & $-6.924$ $(p=0.000)$          & $16.118$                             & $12.624 \pm 5.165$                             \\ \hline
+\multirow{4}{*}{CarHHMM}    & \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.248$                         & $-0.022$ $(p=0.826)$          & $2.196$                             & $1.042 \pm 0.138$                             \\
+                            &                               & 2                             & $117.624$                         & $-9.446$ $(p=0.000)$          & $34.336$                             & $15.251 \pm 7.111$                             \\
+                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.523$                         & $-0.387$ $(p=0.001)$          & $2.594$                             & $0.858 \pm 0.148$                             \\
+                            &                               & 2                             & $56.981$                         & $-7.079$ $(p=0.000)$          & $16.325$                             & $12.530 \pm 6.095$                             \\ \hline
+\multirow{4}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu_Y$}      & 1                             & $41.685$                         & ---                        & $4.268$                             & $2.001 \pm 0.347$                             \\
                             &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\
-                            & \multirow{2}{*}{$\sigma$}     & 1                             & $31.119$                         & ---          & $5.393$                             & $2.014 \pm 0.452$                             \\
-                            &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\ \hline
-\multirow{4}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu$}        & 1                             & $25.776$                         & $0.096$ $(p=0.071)$          & $1.181$                             & $0.959 \pm 0.111$                             \\
-                            &                               & 2                             & $108.589$                         & $3.989$ $(p=0.000)$          & $18.726$                             & $13.083 \pm 4.546$                             \\
-                            & \multirow{2}{*}{$\sigma$}     & 1                             & $9.489$                         & $-0.081$ $(p=0.058)$          & $0.952$                             & $0.779 \pm 0.102$                             \\
-                            &                               & 2                             & $59.514$                         & $-5.186$ $(p=0.000)$          & $14.388$                             & $11.080 \pm 3.737$                             \\ \hline
-\multirow{4}{*}{CarHHMM}    & \multirow{2}{*}{$\mu$}        & 1                             & $27.877$                         & $2.197$ $(p=0.000)$          & $11.693$                             & $0.966 \pm 0.150$                             \\
-                            &                               & 2                             & $96.423$                         & $-8.177$ $(p=0.000)$          & $33.095$                             & $12.341 \pm 6.893$                             \\
-                            & \multirow{2}{*}{$\sigma$}     & 1                             & $9.954$                         & $0.384$ $(p=0.347)$          & $8.950$                             & $0.793 \pm 0.142$                             \\
-                            &                               & 2                             & $54.291$                         & $-10.409$ $(p=0.000)$          & $18.432$                             & $10.142 \pm 5.496$                             \\ \hline
-\multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu$}        & 1                             & $25.719$                         & $0.039$ $(p=0.455)$          & $1.170$                             & $0.958 \pm 0.108$                             \\
-                            &                               & 2                             & $107.404$                         & $2.804$ $(p=0.001)$          & $18.535$                             & $12.830 \pm 4.471$                             \\
-                            & \multirow{2}{*}{$\sigma$}     & 1                             & $9.436$                         & $-0.134$ $(p=0.001)$          & $0.930$                             & $0.777 \pm 0.097$                             \\
-                            &                               & 2                             & $59.635$                         & $-5.065$ $(p=0.000)$          & $14.225$                             & $10.945 \pm 3.548$
+                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $30.299$                         & ---                        & $5.439$                             & $1.948 \pm 0.443$                             \\
+                            &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\
 \end{tabular}
@@
 Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}   & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
-\multirow{6}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu_A^*$}    & 1                                & $-0.001$                         & $-0.001$ $(p=0.934)$          & $0.142$                             & $0.066 \pm 0.045$                             \\
-                            &                               & 2                                & $-0.000$                         & $-0.000$ $(p=0.612)$          & $0.018$                             & $0.017 \pm 0.003$                             \\
-                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.034$                         & $-0.016$ $(p=0.000)$          & $0.001$                             & $0.001 \pm 0.000$                             \\
-                            &                               & 2                                & $0.079$                         & $-0.021$ $(p=0.000)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
-                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & $0.978$                         & $-0.012$ $(p=0.000)$          & $0.010$                             & $0.009 \pm 0.003$                             \\
-                            &                               & 2                                & $0.868$                         & $-0.082$ $(p=0.000)$          & $0.016$                             & $0.016 \pm 0.002$                             \\ \hline
-\multirow{6}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu_A^*$}    & 1                                & $-0.001$                         & $-0.001$ $(p=0.302)$          & $0.027$                             & $0.006 \pm 0.001$                             \\
-                            &                               & 2                                & $0.000$                         & $0.000$ $(p=0.921)$          & $0.012$                             & $0.004 \pm 0.000$                             \\
-                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.136$                         & $0.086$ $(p=0.000)$          & $0.022$                             & $0.004 \pm 0.001$                             \\
-                            &                               & 2                                & $0.144$                         & $0.044$ $(p=0.000)$          & $0.007$                             & $0.003 \pm 0.000$                             \\
-                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & ------                         & ------                     & ------                             & ------                                      \\
-                            &                               & 2                                & ------                         & ------                     & ------                             & ------                                      \\ \hline
-\multirow{6}{*}{CarHHMM}    & \multirow{2}{*}{$\mu_A^*$}    & 1                                & $0.021$                         & $0.021$ $(p=0.436)$          & $0.591$                             & $0.016 \pm 0.004$                             \\
-                            &                               & 2                                & $0.000$                         & $0.000$ $(p=0.669)$          & $0.013$                             & $0.006 \pm 0.001$                             \\
-                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.065$                         & $0.015$ $(p=0.000)$          & $0.022$                             & $0.001 \pm 0.000$                             \\
-                            &                               & 2                                & $0.187$                         & $0.087$ $(p=0.000)$          & $0.018$                             & $0.004 \pm 0.000$                             \\
-                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & $0.891$                         & $-0.099$ $(p=0.000)$          & $0.155$                             & $0.010 \pm 0.002$                             \\
-                            &                               & 2                                & $0.182$                         & $-0.768$ $(p=0.000)$          & $0.137$                             & $0.027 \pm 0.002$                             \\ \hline
-\multirow{6}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_A^*$}    & 1                                & $-0.001$                         & $-0.001$ $(p=0.875)$          & $0.111$                             & $0.066 \pm 0.045$                             \\
-                            &                               & 2                                & $-0.000$                         & $-0.000$ $(p=0.564)$          & $0.018$                             & $0.017 \pm 0.003$                             \\
-                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.034$                         & $-0.016$ $(p=0.000)$          & $0.001$                             & $0.001 \pm 0.000$                             \\
-                            &                               & 2                                & $0.079$                         & $-0.021$ $(p=0.000)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
-                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & $0.978$                         & $-0.012$ $(p=0.000)$          & $0.010$                             & $0.009 \pm 0.003$                             \\
-                            &                               & 2                                & $0.867$                         & $-0.083$ $(p=0.000)$          & $0.016$                             & $0.016 \pm 0.002$                             \\ \hline
+\multirow{9}{*}{CarHHMM-DFT}& \multirow{3}{*}{$\mu_A^*$}    & 1                                & $-0.012$                         & $-0.012$ $(p=0.004)$          & $0.093$                             & $0.066 \pm 0.038$                             \\
+                            &                               & 2                                & $0.099$                         & $-0.001$ $(p=0.436)$          & $0.019$                             & $0.019 \pm 0.003$                             \\
+                            &                               & 3                                & $0.200$                         & $0.000$ $(p=0.965)$          & $0.060$                             & $0.053 \pm 0.013$                             \\
+                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.050$                         & $0.000$ $(p=0.376)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
+                            &                               & 2                                & $0.100$                         & $-0.000$ $(p=0.672)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
+                            &                               & 3                                & $0.298$                         & $-0.002$ $(p=0.024)$          & $0.016$                             & $0.015 \pm 0.002$                             \\
+                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & $0.968$                         & $-0.002$ $(p=0.001)$          & $0.012$                             & $0.010 \pm 0.002$                             \\
+                            &                               & 2                                & $0.828$                         & $-0.002$ $(p=0.007)$          & $0.017$                             & $0.017 \pm 0.002$                             \\
+                            &                               & 3                                & $0.600$                         & $-0.010$ $(p=0.001)$          & $0.064$                             & $0.065 \pm 0.011$                             \\ \hline
+\multirow{9}{*}{HHMM-DFT}   & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $0.048$                         & $0.048$ $(p=0.000)$          & $0.033$                             & $0.007 \pm 0.001$                             \\
+                            &                               & 2                                & $0.087$                         & $-0.013$ $(p=0.000)$          & $0.018$                             & $0.006 \pm 0.001$                             \\
+                            &                               & 3                                & $0.215$                         & $0.015$ $(p=0.000)$          & $0.065$                             & $0.022 \pm 0.004$                             \\
+                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.160$                         & $0.110$ $(p=0.000)$          & $0.023$                             & $0.005 \pm 0.001$                             \\
+                            &                               & 2                                & $0.158$                         & $0.058$ $(p=0.000)$          & $0.010$                             & $0.004 \pm 0.000$                             \\
+                            &                               & 3                                & $0.356$                         & $0.056$ $(p=0.000)$          & $0.035$                             & $0.015 \pm 0.002$                             \\
+                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & ------                         & ------                     & ------                             & ------                                      \\
+                            &                               & 2                                & ------                         & ------                     & ------                             & ------                                      \\
+                            &                               & 3                                & ------                         & ------                     & ------                             & ------                                      \\ \hline
+\multirow{9}{*}{CarHHMM}    & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $0.060$                         & $0.060$ $(p=0.000)$          & $0.034$                             & $0.024 \pm 0.054$                             \\
+                            &                               & 2                                & $0.097$                         & $-0.003$ $(p=0.000)$          & $0.016$                             & $0.052 \pm 0.055$                             \\
+                            &                               & 3                                & $0.201$                         & $0.001$ $(p=0.634)$          & $0.058$                             & $0.018 \pm 0.007$                             \\
+                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.111$                         & $0.061$ $(p=0.000)$          & $0.004$                             & $0.006 \pm 0.042$                             \\
+                            &                               & 2                                & $0.285$                         & $0.185$ $(p=0.000)$          & $0.009$                             & $0.040 \pm 0.043$                             \\
+                            &                               & 3                                & $1.150$                         & $0.850$ $(p=0.000)$          & $0.080$                             & $0.003 \pm 0.001$                             \\
+                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & $0.830$                         & $-0.140$ $(p=0.000)$          & $0.048$                             & $0.016  $                             \\
+                            &                               & 2                                & $0.004$                         & $-0.826$ $(p=0.000)$          & $0.012$                             & $0.023  $                             \\
+                            &                               & 3                                & $0.000$                         & $-0.610$ $(p=0.000)$          & $0.000$                             & $0.016 \pm 0.004$                             \\ \hline
+\multirow{9}{*}{CarHMM-DFT} & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $-0.008$                         & $-0.008$ $(p=0.142)$          & $0.114$                             & $0.064 \pm 0.037$                             \\
+                            &                               & 2                                & $0.101$                         & $0.001$ $(p=0.438)$          & $0.019$                             & $0.019 \pm 0.003$                             \\
+                            &                               & 3                                & $0.200$                         & $0.000$ $(p=0.926)$          & $0.057$                             & $0.054 \pm 0.014$                             \\
+                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.050$                         & $-0.000$ $(p=0.005)$          & $0.001$                             & $0.002 \pm 0.000$                             \\
+                            &                               & 2                                & $0.100$                         & $-0.000$ $(p=0.000)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
+                            &                               & 3                                & $0.299$                         & $-0.001$ $(p=0.117)$          & $0.015$                             & $0.015 \pm 0.002$                             \\
+                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & $0.968$                         & $-0.002$ $(p=0.001)$          & $0.012$                             & $0.010 \pm 0.002$                             \\
+                            &                               & 2                                & $0.830$                         & $-0.000$ $(p=0.552)$          & $0.016$                             & $0.017 \pm 0.002$                             \\
+                            &                               & 3                                & $0.595$                         & $-0.015$ $(p=0.000)$          & $0.071$                             & $0.065 \pm 0.010$                             \\ \hline
 \end{tabular}
@@
         \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_A_0_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_A_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_MLE_density_A_0_0.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_MLE_density_A_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_MLE_density_A_0_2.png}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_A_0_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_A_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_uncorr_MLE_density_A_0_0.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_uncorr_MLE_density_A_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_uncorr_MLE_density_A_0_2.png}
         \end{center}
@@
 
-        \subsubsection{CarHHMM}
+        \subsection{CarHHMM}
         \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_V_MLE_density_A_0_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_V_MLE_density_A_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_V_MLE_density_A_0_0.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_V_MLE_density_A_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_V_MLE_density_A_0_2.png}
         \end{center}
@@
 
-        \subsubsection{CarHMM-DFT}
+        \subsection{CarHMM-DFT}
         \begin{center}
-        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_A_0_0.png}
-        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_A_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hmm_FV_MLE_density_A_0_0.png}
+        \includegraphics[width=2.1in]{../Plots/hmm_FV_MLE_density_A_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hmm_FV_MLE_density_A_0_2.png}
         \end{center}
@@
     \newpage
-    \subsection{Empirical joint distributions - wiggliness ($\Ztwo_{t,\tilde t^*}$)}
+    \section{Empirical joint distributions - wiggliness ($\Ztwo_{t,\tilde t^*}$)}
 
@@
 Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}       \\ \hline
-\multirow{4}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu_W^*$}    & 1                                & $23.362$                         & $0.022$ $(p=0.408)$          & $0.582$                             & $0.419 \pm 0.061$                             \\
-                            &                               & 2                                & $300.689$                         & $-0.511$ $(p=0.213)$          & $9.144$                             & $4.679 \pm 0.336$                             \\
-                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & $12.949$                         & $-0.001$ $(p=0.974)$          & $0.539$                             & $0.388 \pm 0.061$                             \\
-                            &                               & 2                                & $329.585$                         & $-0.515$ $(p=0.280)$          & $10.622$                             & $5.748 \pm 0.396$                             \\ \hline
-\multirow{4}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu_W^*$}    & 1                                & $23.353$                         & $0.013$ $(p=0.632)$          & $0.620$                             & $0.431 \pm 0.067$                             \\
-                            &                               & 2                                & $298.441$                         & $-2.759$ $(p=0.000)$          & $10.438$                             & $4.788 \pm 0.432$                             \\
-                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & $12.938$                         & $-0.012$ $(p=0.656)$          & $0.583$                             & $0.404 \pm 0.064$                             \\
-                            &                               & 2                                & $328.496$                         & $-1.604$ $(p=0.001)$          & $10.482$                             & $5.822 \pm 0.460$                             \\ \hline
-\multirow{4}{*}{CarHHMM}    & \multirow{2}{*}{$\mu_W^*$}    & 1                                & ------                         & ------                     & ------                             & ------                                      \\
+\multirow{6}{*}{CarHHMM-DFT}& \multirow{3}{*}{$\mu_W^*$}    & 1                                & $34.981$                         & $0.001$ $(p=0.978)$          & $1.026$                             & $0.658 \pm 0.088$                             \\
+                            &                               & 2                                & $505.104$                         & $-0.686$ $(p=0.394)$          & $17.936$                             & $9.740 \pm 0.814$                             \\
+                            &                               & 3                                & $9787.146$                         & $17.166$ $(p=0.718)$          & $1060.450$                             & $280.244 \pm 97.120$                             \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.849$                         & $-0.001$ $(p=0.991)$          & $0.989$                             & $0.652 \pm 0.094$                             \\
+                            &                               & 2                                & $514.207$                         & $-2.473$ $(p=0.010)$          & $21.427$                             & $11.410 \pm 0.947$                             \\
+                            &                               & 3                                & $14429.732$                         & $-32.818$ $(p=0.647)$          & $1596.020$                             & $302.698 \pm 165.935$                             \\ \hline
+\multirow{6}{*}{HHMM-DFT}   & \multirow{3}{*}{$\mu_W^*$}    & 1                                & $35.067$                         & $0.087$ $(p=0.097)$          & $1.170$                             & $0.679 \pm 0.100$                             \\
+                            &                               & 2                                & $500.928$                         & $-4.862$ $(p=0.000)$          & $24.024$                             & $10.754 \pm 1.205$                             \\
+                            &                               & 3                                & $6840.995$                         & $-2928.985$ $(p=0.000)$          & $1164.288$                             & $198.162 \pm 74.720$                             \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.800$                         & $-0.050$ $(p=0.351)$          & $1.192$                             & $0.676 \pm 0.104$                             \\
+                            &                               & 2                                & $513.263$                         & $-3.417$ $(p=0.001)$          & $22.596$                             & $12.497 \pm 1.301$                             \\
+                            &                               & 3                                & $11458.795$                         & $-3003.755$ $(p=0.000)$          & $1738.793$                             & $302.698 \pm 193.611$                             \\ \hline
+\multirow{6}{*}{CarHHMM}    & \multirow{3}{*}{$\mu_W^*$}    & 1                                & ------                         & ------                     & ------                             & ------                                      \\
                             &                               & 2                                & ------                         & ------                     & ------                             & ------                                      \\
-                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & ------                         & ------                     & ------                             & ------                                      \\
-                            &                               & 2                                & ------                         & ------                     & ------                             & ------                                      \\ \hline
-\multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_W^*$}    & 1                                & $23.341$                         & $0.001$ $(p=0.957)$          & $0.568$                             & $0.419 \pm 0.060$                             \\
-                            &                               & 2                                & $301.209$                         & $0.009$ $(p=0.982)$          & $8.869$                             & $4.697 \pm 0.353$                             \\
-                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & $12.918$                         & $-0.032$ $(p=0.183)$          & $0.526$                             & $0.387 \pm 0.059$                             \\
-                            &                               & 2                                & $329.684$                         & $-0.416$ $(p=0.365)$          & $10.225$                             & $5.773 \pm 0.396$
+                            &                               & 3                                & ------                         & ------                     & ------                             & ------                                      \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & ------                         & ------                     & ------                             & ------                                      \\
+                            &                               & 2                                & ------                         & ------                     & ------                             & ------                                      \\
+                            &                               & 3                                & ------                         & ------                     & ------                             & ------                                      \\ \hline
+\multirow{6}{*}{CarHMM-DFT} & \multirow{3}{*}{$\mu_W^*$}    & 1                                & $35.034$                         & $0.054$ $(p=0.264)$          & $1.076$                             & $0.661 \pm 0.093$                             \\
+                            &                               & 2                                & $503.762$                         & $-2.028$ $(p=0.013)$          & $18.258$                             & $9.641 \pm 0.916$                             \\
+                            &                               & 3                                & $9817.785$                         & $47.805$ $(p=0.328)$          & $1089.110$                             & $280.244 \pm 136.259$                             \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.890$                         & $0.040$ $(p=0.401)$          & $1.058$                             & $0.656 \pm 0.101$                             \\
+                            &                               & 2                                & $515.228$                         & $-1.452$ $(p=0.125)$          & $21.087$                             & $11.371 \pm 0.990$                             \\
+                            &                               & 3                                & $14437.867$                         & $-24.683$ $(p=0.740)$          & $1660.764$                             & $331.589 \pm 165.935$                             \\
 \end{tabular}
@@
         \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_FoVeDBA_0_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_FoVeDBA_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_MLE_density_FoVeDBA_0_0.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_MLE_density_FoVeDBA_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_MLE_density_FoVeDBA_0_2.png}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_FoVeDBA_0_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_FoVeDBA_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_uncorr_MLE_density_FoVeDBA_0_0.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_uncorr_MLE_density_FoVeDBA_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hhmm_FV_uncorr_MLE_density_FoVeDBA_0_2.png}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_FoVeDBA_0_0.png}
-        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_FoVeDBA_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hmm_FV_MLE_density_FoVeDBA_0_0.png}
+        \includegraphics[width=2.1in]{../Plots/hmm_FV_MLE_density_FoVeDBA_0_1.png}
+        \includegraphics[width=2.1in]{../Plots/hmm_FV_MLE_density_FoVeDBA_0_2.png}
         \end{center}
@@
         \begin{center}
+        \scalebox{0.7}{
+        \bgroup
+        \centering
+        \def\arraystretch{1.5}
+\begin{tabular}{ccccccc}
+Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
+\multirow{12}{*}{CarHHMM-DFT}& $\Gamma^{*(1)}_{12}$          & $0.258$                         & $0.006$ $(p=0.004)$        & $0.043$                           & $0.032 \pm 0.006$                             \\
+                             & $\Gamma^{*(1)}_{13}$          & $0.002$                         & $0.002$ $(p=0.000)$        & $0.004$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{21}$          & $0.083$                         & $0.001$ $(p=0.243)$        & $0.015$                           & $0.012 \pm 0.002$                             \\
+                             & $\Gamma^{*(1)}_{23}$          & $0.052$                         & $-0.000$ $(p=0.984)$        & $0.010$                           & $0.009 \pm 0.001$                             \\
+                             & $\Gamma^{*(1)}_{31}$          & $0.003$                         & $0.003$ $(p=0.000)$        & $0.008$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{32}$          & $0.234$                         & $0.002$ $(p=0.322)$        & $0.042$                           & $0.036 \pm 0.008$                             \\
+                             & $\Gamma^{*(2)}_{12}$          & $0.111$                         & $0.146$ $(p=0.349)$        & $0.017$                           & $0.015 \pm 0.004$                             \\
+                             & $\Gamma^{*(2)}_{13}$          & $0.001$                         & $0.002$ $(p=0.000)$        & $0.002$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.151$                         & $-0.068$ $(p=0.584)$        & $0.023$                           & $0.021 \pm 0.005$                             \\
+                             & $\Gamma^{*(2)}_{23}$          & $0.036$                         & $0.016$ $(p=0.944)$        & $0.013$                           & $0.010 \pm 0.003$                             \\
+                             & $\Gamma^{*(2)}_{31}$          & $0.007$                         & $0.003$ $(p=0.000)$        & $0.024$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(2)}_{32}$          & $0.252$                         & $0.008$ $(p=0.000)$        & $0.096$                           & $0.063 \pm 0.028$                             \\ \hline
+\multirow{12}{*}{HHMM-DFT}   & $\Gamma^{*(1)}_{12}$          & $0.240$                         & $-0.012$ $(p=0.000)$        & $0.042$                           & $0.032 \pm 0.006$                             \\
+                             & $\Gamma^{*(1)}_{13}$          & $0.003$                         & $0.003$ $(p=0.000)$        & $0.006$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{21}$          & $0.077$                         & $-0.005$ $(p=0.000)$        & $0.019$                           & $0.012 \pm 0.002$                             \\
+                             & $\Gamma^{*(1)}_{23}$          & $0.047$                         & $-0.004$ $(p=0.000)$        & $0.009$                           & $0.008 \pm 0.001$                             \\
+                             & $\Gamma^{*(1)}_{31}$          & $0.005$                         & $0.005$ $(p=0.000)$        & $0.009$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{32}$          & $0.153$                         & $-0.079$ $(p=0.000)$        & $0.029$                           & $0.026 \pm 0.005$                             \\
+                             & $\Gamma^{*(2)}_{12}$          & $0.109$                         & $0.128$ $(p=0.000)$        & $0.018$                           & $0.016 \pm 0.004$                             \\
+                             & $\Gamma^{*(2)}_{13}$          & $0.002$                         & $0.003$ $(p=0.000)$        & $0.004$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.151$                         & $-0.073$ $(p=0.712)$        & $0.031$                           & $0.023 \pm 0.006$                             \\
+                             & $\Gamma^{*(2)}_{23}$          & $0.037$                         & $0.012$ $(p=0.009)$        & $0.013$                           & $0.011 \pm 0.004$                             \\
+                             & $\Gamma^{*(2)}_{31}$          & $0.021$                         & $0.005$ $(p=0.000)$        & $0.026$                           & $0.034  $                             \\
+                             & $\Gamma^{*(2)}_{32}$          & $0.135$                         & $-0.073$ $(p=0.000)$        & $0.057$                           & $0.038 \pm 0.018$                             \\ \hline
+\multirow{12}{*}{CarHHMM}    & $\Gamma^{*(1)}_{12}$          & $0.138$                         & $-0.114$ $(p=0.000)$        & $0.131$                           & $0.018 \pm 0.006$                             \\
+                             & $\Gamma^{*(1)}_{13}$          & $0.023$                         & $0.022$ $(p=0.000)$        & $0.107$                           & $0.039  $                             \\
+                             & $\Gamma^{*(1)}_{21}$          & $0.037$                         & $-0.044$ $(p=0.000)$        & $0.054$                           & $0.007 \pm 0.001$                             \\
+                             & $\Gamma^{*(1)}_{23}$          & $0.027$                         & $-0.025$ $(p=0.000)$        & $0.050$                           & $0.004 \pm 0.001$                             \\
+                             & $\Gamma^{*(1)}_{31}$          & $0.013$                         & $0.013$ $(p=0.000)$        & $0.022$                           & $0.013  $                             \\
+                             & $\Gamma^{*(1)}_{32}$          & $0.134$                         & $-0.099$ $(p=0.000)$        & $0.043$                           & $0.023 \pm 0.006$                             \\
+                             & $\Gamma^{*(2)}_{12}$          & $0.056$                         & $0.026$ $(p=0.000)$        & $0.035$                           & $0.008 \pm 0.002$                             \\
+                             & $\Gamma^{*(2)}_{13}$          & $0.002$                         & $0.022$ $(p=0.004)$        & $0.017$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.069$                         & $-0.113$ $(p=0.000)$        & $0.032$                           & $0.011 \pm 0.003$                             \\
+                             & $\Gamma^{*(2)}_{23}$          & $0.017$                         & $-0.009$ $(p=0.000)$        & $0.012$                           & $0.005 \pm 0.002$                             \\
+                             & $\Gamma^{*(2)}_{31}$          & $0.011$                         & $0.013$ $(p=0.000)$        & $0.016$                           & $0.029  $                             \\
+                             & $\Gamma^{*(2)}_{32}$          & $0.138$                         & $-0.093$ $(p=0.000)$        & $0.079$                           & $0.037 \pm 0.018$                             \\ \hline
+\multirow{6}{*}{CarHMM-DFT}  & $\Gamma^{*(1)}_{12}$          & $0.158$                         & ------                   & $0.019$                           & $0.014 \pm 0.002$                             \\
+                             & $\Gamma^{*(1)}_{13}$          & $0.001$                         & ------                   & $0.002$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{21}$          & $0.096$                         & ------                   & $0.012$                           & $0.010 \pm 0.001$                             \\
+                             & $\Gamma^{*(1)}_{23}$          & $0.046$                         & ------                   & $0.007$                           & $0.007 \pm 0.001$                             \\
+                             & $\Gamma^{*(1)}_{31}$          & $0.002$                         & ------                   & $0.004$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{32}$          & $0.227$                         & ------                   & $0.035$                           & $0.030 \pm 0.006$                             \\
+\end{tabular}
+        \egroup
+        }
+        \end{center}
+
+        \noindent Table \arabic{tablenum}: Estimates and standard errors for fine-scale probability transition matrices across all four models in the simulation study.
+The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
+The $\pm$ refers to the inter-quartile range.
+SE stands for standard error.
+Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
+        \addtocounter{tablenum}{1}
+
+        \begin{center}
         \scalebox{0.75}{
@@
 Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
-\multirow{6}{*}{CarHMM-DFT}  & $\Gamma_{12}$                 & ------                          & ------                     & ------                            & ------                                        \\
+\multirow{2}{*}{CarHHMM-DFT}& $\Gamma_{12}$                 & $0.160$                         & $0.005$ $(p=0.030)$        & $0.056$                           & $0.042 \pm 0.012$                             \\
+                             & $\Gamma_{21}$                 & $0.907$                         & $-0.006$ $(p=0.214)$        & $0.102$                           & $0.116$                             \\ \hline
+\multirow{2}{*}{HHMM-DFT}   & $\Gamma_{12}$                 & $0.161$                         & $0.007$ $(p=0.007)$        & $0.055$                           & $0.041 \pm 0.011$                             \\
+                             & $\Gamma_{21}$                 & $0.916$                         & $0.004$ $(p=0.372)$        & $0.088$                           & $0.112$                             \\ \hline
+\multirow{2}{*}{CarHHMM}    & $\Gamma_{12}$                 & $0.198$                         & $0.044$ $(p=0.000)$        & $0.152$                           & $0.044 \pm 0.014$                             \\
+                             & $\Gamma_{21}$                 & $0.832$                         & $-0.081$ $(p=0.000)$        & $0.195$                           & $0.103 \pm 0.073$                             \\ \hline
+\multirow{2}{*}{CarHMM-DFT} & $\Gamma_{12}$                 & ------                         & ------                   & ------                           & ------                                      \\
                              & $\Gamma_{21}$                 & ------                         & ------                   & ------                           & ------                                      \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.179$                         & ------                     & $0.018$                           & $0.015 \pm 0.002$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.079$                         & ------                     & $0.011$                           & $0.008 \pm 0.001$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & ------                          & ------                     & ------                            & ------                                        \\
-                             & $\Gamma^{*(2)}_{21}$          & ------                          & ------                     & ------                            & ------                                        \\ \hline
-\multirow{6}{*}{HHMM-DFT}    & $\Gamma_{12}$                 & $0.206$                         & $-0.006$ $(p=0.033)$       & $0.063$                           & $0.049 \pm 0.011$                             \\
-                             & $\Gamma_{21}$                 & $0.822$                         & $0.013$ $(p=0.014)$        & $0.119$                           & $0.098 \pm 0.029$                             \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.312$                         & $-0.008$ $(p=0.011)$       & $0.070$                           & $0.048 \pm 0.019$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.047$                         & $-0.003$ $(p=0.000)$       & $0.014$                           & $0.010 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.139$                         & $-0.001$ $(p=0.158)$       & $0.018$                           & $0.017 \pm 0.003$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.146$                         & $-0.004$ $(p=0.001)$       & $0.025$                           & $0.019 \pm 0.004$                             \\ \hline
-\multirow{6}{*}{CarHHMM}     & $\Gamma_{12}$                 & $0.307$                         & $0.095$ $(p=0.000)$        & $0.257$                           & $0.052 \pm 0.021$                             \\
-                             & $\Gamma_{21}$                 & $0.711$                         & $-0.098$ $(p=0.000)$       & $0.280$                           & $0.102 \pm 0.040$                             \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.269$                         & $-0.051$ $(p=0.000)$       & $0.252$                           & $0.019 \pm 0.006$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.134$                         & $0.084$ $(p=0.000)$        & $0.081$                           & $0.016 \pm 0.003$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.120$                         & $-0.020$ $(p=0.025)$       & $0.196$                           & $0.008 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.144$                         & $-0.006$ $(p=0.001)$       & $0.042$                           & $0.019 \pm 0.006$                             \\ \hline
-\multirow{6}{*}{CarHHMM-DFT} & $\Gamma_{12}$                 & $0.210$                         & $-0.002$ $(p=0.473)$       & $0.064$                           & $0.050 \pm 0.011$                             \\
-                             & $\Gamma_{21}$                 & $0.818$                         & $0.009$ $(p=0.076)$        & $0.117$                           & $0.097 \pm 0.027$                             \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.325$                         & $0.005$ $(p=0.132)$        & $0.069$                           & $0.048 \pm 0.018$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.050$                         & $0.000$ $(p=0.713)$        & $0.012$                           & $0.010 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.141$                         & $0.001$ $(p=0.336)$        & $0.017$                           & $0.016 \pm 0.003$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.152$                         & $0.002$ $(p=0.032)$        & $0.020$                           & $0.018 \pm 0.004$                             \\ \hline
-
 \end{tabular}
@@
 
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for all probability transition matrices across all four models in the simulation study.
+        \noindent Table \arabic{tablenum}: Estimates and standard errors for coarse-scale probability transition matrices across all four models in the simulation study.
 The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
@@
         \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_Gamma_density_-1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_Gamma_density_-1_row_-1.png}
 
-        \includegraphics[width=3in]{../Plots/hhmm_FV_Gamma_density_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_Gamma_density_1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_Gamma_density_0_row_0.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_Gamma_density_1_row_0.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_Gamma_density_0_row_1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_Gamma_density_1_row_1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_Gamma_density_0_row_2.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_Gamma_density_1_row_2.png}
         \end{center}
@@
 We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix.
-The red star represents the true values.
+The red star represents the true values and the white star represents the mean of all maximum likelihood estimates.
         \addtocounter{fignum}{1}
@@
         \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_Gamma_density_-1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_uncorr_Gamma_density_-1_row_-1.png}
 
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_Gamma_density_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_Gamma_density_1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_uncorr_Gamma_density_0_row_0.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_uncorr_Gamma_density_1_row_0.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_uncorr_Gamma_density_0_row_1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_uncorr_Gamma_density_1_row_1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_uncorr_Gamma_density_0_row_2.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_FV_uncorr_Gamma_density_1_row_2.png}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_V_Gamma_density_-1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_V_Gamma_density_-1_row_-1.png}
 
-        \includegraphics[width=3in]{../Plots/hhmm_V_Gamma_density_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_V_Gamma_density_1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_V_Gamma_density_0_row_0.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_V_Gamma_density_1_row_0.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_V_Gamma_density_0_row_1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_V_Gamma_density_1_row_1.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_V_Gamma_density_0_row_2.png}
+        \includegraphics[width=2.5in]{../Plots/hhmm_V_Gamma_density_1_row_2.png}
         \end{center}
@@
         \begin{center}
-        \includegraphics[width=3in]{../Plots/hmm_FV_Gamma_density_0.png}
+        \includegraphics[width=2.5in]{../Plots/hmm_FV_Gamma_density_0_row_0.png} \\
+        \includegraphics[width=2.5in]{../Plots/hmm_FV_Gamma_density_0_row_1.png} \\
+        \includegraphics[width=2.5in]{../Plots/hmm_FV_Gamma_density_0_row_2.png} \\
         \end{center}
@@
         \addtocounter{fignum}{1}
-\fi
 \end{document}

### 2021-06-18 Report/Supp_Sim.tex

@@
     	\noindent Figure \arabic{fignum}: Acceleration data for a five dives of a selected simulated data set.
-The raw accelerometer data is denoted as $Y^*_{t,\tilde t^*}$ which is recorded at 50 hertz while $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are recorded at 0.5 hertz.
+The raw accelerometer data is denoted as $Y^*_{t,\tilde t^*}$ and recorded at 50 times per second while $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are recorded at once every two seconds.
 Dives are separated by vertical black lines.
 The colour of the curve corresponds to the true fine-scale state, while the colour of the background corresponds to the true dive type.
-Subdive state 1 can plausibly be interpreted as ``gliding" while subdive state 2 can plausibly be interpreted as active swimming, or ``fluking".
+Subdive state 1 can plausibly be interpreted as ``gliding", subdive state 2 can plausibly be interpreted as active swimming, and subdive state 3 can be interpretted as sudden or aggressive movement.
     	\addtocounter{fignum}{1}
@@
 
+        \newpage
+
         \subsection{CarHHMM-DFT}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$) for each dive type.
-The red star represents the true values of $\mu$ and $\sigma$ for both dive types.
-These estimates are for the CarHHMM-DFT.
+        \noindent Figure \arabic{fignum}: Kernel density estimates of the distributions of estimates of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
+These estimates are associated with the CarHHMM-DFT and correspond to both dive types.
+The red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$) for each dive type.
-The red star represents the true values of $\mu$ and $\sigma$ for both dive types.
-These estimates are for the HHMM-DFT.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of estimates of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
+These estimates are associated with the HHMM-DFT and correspond to both dive types.
+The red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$) for each dive type.
-The red star represents the true values of $\mu$ and $\sigma$ for both dive types.
-These estimates are for the CarHHMM.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
+These estimates are associated with the CarHHMM and correspond to both dive types.
+The red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$).
-The red star represents the true values of $\mu$ and $\sigma$ for both dive types.
-These estimates are for the CarHMM-DFT.
-The CarHMM-DFT assumes that there is only one dive type.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
+These estimates are associated with the CarHMM-DFT.
 There is no red star which represents the true values because the CarHMM-DFT assumes that there is only one dive type.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of acceleration parameters ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of estimates of acceleration parameters ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
 The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of acceleration parameters ($\hat \mu^*_A$ and $\hat \sigma^*_A$) for each subdive state.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of estimates of acceleration parameters ($\hat \mu^*_A$ and $\hat \sigma^*_A$) for each subdive state.
 The red star represents the true values of $\mu^*_A$ and $\sigma^*_A$ for both subdive states.
@@
 \multirow{6}{*}{CarHHMM-DFT}& \multirow{3}{*}{$\mu_W^*$}    & 1                                & $34.981$                         & $0.001$ $(p=0.978)$          & $1.026$                             & $0.658 \pm 0.088$                             \\
-                            &                               & 2                                & $505.104$                         & $-0.686$ $(p=0.394)$          & $17.936$                             & $9.740 \pm 0.814$                             \\
-                            &                               & 3                                & $9787.146$                         & $17.166$ $(p=0.718)$          & $1060.450$                             & $280.244 \pm 97.120$                             \\
+                            &                               & 2                                & $505.10$                         & $-0.69$ $(p=0.394)$          & $17.94$                             & $9.74 \pm 0.81$                             \\
+                            &                               & 3                                & $9787.1$                         & $17.2$ $(p=0.718)$           & $1060.5$                            & $280.2 \pm 97.1$                             \\
                             & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.849$                         & $-0.001$ $(p=0.991)$         & $0.989$                             & $0.652 \pm 0.094$                             \\
-                            &                               & 2                                & $514.207$                         & $-2.473$ $(p=0.010)$          & $21.427$                             & $11.410 \pm 0.947$                             \\
-                            &                               & 3                                & $14429.732$                         & $-32.818$ $(p=0.647)$          & $1596.020$                             & $302.698 \pm 165.935$                             \\ \hline
+                            &                               & 2                                & $514.21$                         & $-2.47$ $(p=0.010)$          & $21.43$                             & $11.41 \pm 0.95$                             \\
+                            &                               & 3                                & $14430$                          & $-33$ $(p=0.647)$            & $1596$                              & $303 \pm 166$                             \\ \hline
 \multirow{6}{*}{HHMM-DFT}   & \multirow{3}{*}{$\mu_W^*$}    & 1                                & $35.067$                         & $0.087$ $(p=0.097)$          & $1.170$                             & $0.679 \pm 0.100$                             \\
-                            &                               & 2                                & $500.928$                         & $-4.862$ $(p=0.000)$          & $24.024$                             & $10.754 \pm 1.205$                             \\
-                            &                               & 3                                & $6840.995$                         & $-2928.985$ $(p=0.000)$          & $1164.288$                             & $198.162 \pm 74.720$                             \\
+                            &                               & 2                                & $500.93$                         & $-4.86$ $(p=0.000)$          & $24.02$                             & $10.75 \pm 1.21$                             \\
+                            &                               & 3                                & $6841.0$                         & $-2929.0$ $(p=0.000)$        & $1164.3$                            & $198.2 \pm 74.7$                             \\
                             & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.800$                         & $-0.050$ $(p=0.351)$         & $1.192$                             & $0.676 \pm 0.104$                             \\
-                            &                               & 2                                & $513.263$                         & $-3.417$ $(p=0.001)$          & $22.596$                             & $12.497 \pm 1.301$                             \\
-                            &                               & 3                                & $11458.795$                         & $-3003.755$ $(p=0.000)$          & $1738.793$                             & $302.698 \pm 193.611$                             \\ \hline
+                            &                               & 2                                & $513.26$                         & $-3.42$ $(p=0.001)$          & $22.60$                             & $12.50 \pm 1.30$                             \\
+                            &                               & 3                                & $11459$                          & $-3004$ $(p=0.000)$          & $1739$                              & $303 \pm 194$                             \\ \hline
 \multirow{6}{*}{CarHHMM}    & \multirow{3}{*}{$\mu_W^*$}    & 1                                & ------                         & ------                     & ------                             & ------                                      \\
@@
 \multirow{6}{*}{CarHMM-DFT} & \multirow{3}{*}{$\mu_W^*$}    & 1                                & $35.034$                         & $0.054$ $(p=0.264)$          & $1.076$                             & $0.661 \pm 0.093$                             \\
-                            &                               & 2                                & $503.762$                         & $-2.028$ $(p=0.013)$          & $18.258$                             & $9.641 \pm 0.916$                             \\
-                            &                               & 3                                & $9817.785$                         & $47.805$ $(p=0.328)$          & $1089.110$                             & $280.244 \pm 136.259$                             \\
+                            &                               & 2                                & $503.76$                         & $-2.03$ $(p=0.013)$          & $18.26$                             & $9.64 \pm 0.92$                             \\
+                            &                               & 3                                & $9817.8$                         & $47.8$ $(p=0.328)$          & $1089.1$                             & $280.2 \pm 136.3$                             \\
                             & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.890$                         & $0.040$ $(p=0.401)$          & $1.058$                             & $0.656 \pm 0.101$                             \\
-                            &                               & 2                                & $515.228$                         & $-1.452$ $(p=0.125)$          & $21.087$                             & $11.371 \pm 0.990$                             \\
-                            &                               & 3                                & $14437.867$                         & $-24.683$ $(p=0.740)$          & $1660.764$                             & $331.589 \pm 165.935$                             \\
+                            &                               & 2                                & $515.23$                         & $-1.45$ $(p=0.125)$          & $21.09$                             & $11.37 \pm 0.99$                             \\
+                            &                               & 3                                & $14437$                          & $-25$ $(p=0.740)$          & $1661$                             & $332 \pm 166$                             \\
 \end{tabular}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of wiggliness parameters ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of estimates of wiggliness parameters ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
 The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
@@
     \section{Empirical joint distributions - probability transition matrices ($\Gamma, \Gamma^*$)}
+
+        \begin{center}
+        \scalebox{0.75}{
+        \bgroup
+        \centering
+        \def\arraystretch{1.5}
+\begin{tabular}{ccccccc}
+Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
+\multirow{2}{*}{CarHHMM-DFT}& $\Gamma_{12}$                 & $0.160$                         & $0.005$ $(p=0.030)$        & $0.056$                           & $0.042 \pm 0.012$                             \\
+                             & $\Gamma_{21}$                 & $0.907$                         & $-0.006$ $(p=0.214)$        & $0.102$                           & $0.116$                             \\ \hline
+\multirow{2}{*}{HHMM-DFT}   & $\Gamma_{12}$                 & $0.161$                         & $0.007$ $(p=0.007)$        & $0.055$                           & $0.041 \pm 0.011$                             \\
+                             & $\Gamma_{21}$                 & $0.916$                         & $0.004$ $(p=0.372)$        & $0.088$                           & $0.112$                             \\ \hline
+\multirow{2}{*}{CarHHMM}    & $\Gamma_{12}$                 & $0.198$                         & $0.044$ $(p=0.000)$        & $0.152$                           & $0.044 \pm 0.014$                             \\
+                             & $\Gamma_{21}$                 & $0.832$                         & $-0.081$ $(p=0.000)$        & $0.195$                           & $0.103 \pm 0.073$                             \\ \hline
+\multirow{2}{*}{CarHMM-DFT} & $\Gamma_{12}$                 & ------                         & ------                   & ------                           & ------                                      \\
+                             & $\Gamma_{21}$                 & ------                         & ------                   & ------                           & ------                                      \\
+\end{tabular}
+        \egroup
+        }
+        \end{center}
+
+        \noindent Table \arabic{tablenum}: Estimates and standard errors for coarse-scale probability transition matrices across all four models in the simulation study.
+The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
+The $\pm$ refers to the inter-quartile range.
+SE stands for standard error.
+Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
+        \addtocounter{tablenum}{1}
+
+        \newpage
 
@@
 
-        \begin{center}
-        \scalebox{0.75}{
-        \bgroup
-        \centering
-        \def\arraystretch{1.5}
-\begin{tabular}{ccccccc}
-Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
-\multirow{2}{*}{CarHHMM-DFT}& $\Gamma_{12}$                 & $0.160$                         & $0.005$ $(p=0.030)$        & $0.056$                           & $0.042 \pm 0.012$                             \\
-                             & $\Gamma_{21}$                 & $0.907$                         & $-0.006$ $(p=0.214)$        & $0.102$                           & $0.116$                             \\ \hline
-\multirow{2}{*}{HHMM-DFT}   & $\Gamma_{12}$                 & $0.161$                         & $0.007$ $(p=0.007)$        & $0.055$                           & $0.041 \pm 0.011$                             \\
-                             & $\Gamma_{21}$                 & $0.916$                         & $0.004$ $(p=0.372)$        & $0.088$                           & $0.112$                             \\ \hline
-\multirow{2}{*}{CarHHMM}    & $\Gamma_{12}$                 & $0.198$                         & $0.044$ $(p=0.000)$        & $0.152$                           & $0.044 \pm 0.014$                             \\
-                             & $\Gamma_{21}$                 & $0.832$                         & $-0.081$ $(p=0.000)$        & $0.195$                           & $0.103 \pm 0.073$                             \\ \hline
-\multirow{2}{*}{CarHMM-DFT} & $\Gamma_{12}$                 & ------                         & ------                   & ------                           & ------                                      \\
-                             & $\Gamma_{21}$                 & ------                         & ------                   & ------                           & ------                                      \\
-\end{tabular}
-        \egroup
-        }
-        \end{center}
+        \newpage
 
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for coarse-scale probability transition matrices across all four models in the simulation study.
-The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
-The $\pm$ refers to the inter-quartile range.
-SE stands for standard error.
-Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
-        \addtocounter{tablenum}{1}
-
-        \newpage
         \subsection{CarHHMM-DFT}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM-DFT.
+        \noindent Figure \arabic{fignum}: Kernel density  estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM-DFT.
 There are two fine-scale probability transition matrices- one corresponding to each dive type.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the HHMM-DFT.
+        \noindent Figure \arabic{fignum}: Kernel density  estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the HHMM-DFT.
 There are two fine-scale probability transition matrices- one corresponding to each dive type.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM.
+        \noindent Figure \arabic{fignum}: Kernel density  estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM.
 There are two fine-scale probability transition matrices- one corresponding to each dive type.

### 2021-06-18 Code/HHMM.py
Fix update_params

@@ -189,6 +189,10 @@ class HHMM:
 
     def reorder_params(self):
 
+        if self.CM is None:
+            self.CM = [np.zeros((self.pars.K[0],self.pars.K[0])),
+                       [np.zeros((self.pars.K[1],self.pars.K[1])) for _ in range(self.pars.K[0])]]
+
         # do coarse-scale states
         state_order = np.argsort(self.theta[0]['dive_duration']['mu'])
 
@@ -215,20 +219,24 @@ class HHMM:
         self.CM[0] = CM_coarse
 
         # then reorder theta
+        theta_coarse = deepcopy(self.theta[0])
+        SEs = deepcopy(self.SEs)
+
         for feature,settings in self.pars.features[0].items():
             for param in ['mu','sig','corr']:
-                self.theta[0][feature][param] = np.array([self.theta[0][feature][param][state_order[i]] for i in range(self.pars.K[0])])
-                self.SEs[feature][param] = np.array([self.SEs[feature][param][state_order[i]] for i in range(self.pars.K[0])])
+                self.theta[0][feature][param] = np.array([theta_coarse[feature][param][state_order[i]] for i in range(self.pars.K[0])])
+                self.SEs[feature][param] = np.array([SEs[feature][param][state_order[i]] for i in range(self.pars.K[0])])
 
         # do fine-scale states
+        theta_fine = deepcopy(self.theta[1][0])
         if 'Ahat_low' in self.theta[1][0]:
-            state_order = np.argsort(self.theta[1][0]['Ahat_low']['mu'])
+            state_order = np.argsort(theta_fine['Ahat_low']['mu'])
         elif 'FoVeDBA' in self.theta[1][0]:
-            state_order = np.argsort(self.theta[1][0]['FoVeDBA']['mu'])
+            state_order = np.argsort(theta_fine['FoVeDBA']['mu'])
         elif 'Ax' in self.theta[1][0]:
-            state_order = np.argsort(self.theta[1][0]['Ax']['sig'])
+            state_order = np.argsort(theta_fine['Ax']['sig'])
         else:
-            state_order = np.argsort(self.theta[1][0]['A']['sig'])
+            state_order = np.argsort(theta_fine['A']['sig'])
 
         # first reorder eta
         for k0 in range(self.pars.K[0]):
@@ -248,8 +256,8 @@ class HHMM:
         for feature,settings in self.pars.features[1].items():
             for k0 in range(self.pars.K[0]):
[…]

### 2021-06-18 Code/Run_Sim.py

@@ -310,7 +310,7 @@ hhmm_Vs = []
 hhmm_FV_uncorrs = []
 hhmm_FVs = []
 
-h = 0.001
+h = 0.01
 
 for dataset_num in range(ndatasets):
 

### 2021-06-21 Report/Supp_Case.tex
Updates from Overleaf

@@
 % Title, authors, affiliations
-\title{Modelling state-switching functional data with hidden Markov models: supplementary material (case study)}%\query{Q1}
+\title{Modelling state-switching functional data with hidden Markov models: supplement A  (case study)}%\query{Q1}
 \date{}
@@
 All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $(\hat \mu_Y,\hat \sigma_Y)$, which is denoted by a red star.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_Y,\hat \sigma_Y)$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
 All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $(\hat \eta_{1,2},\hat \eta_{2,1})$, which is denoted by a red star.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \eta_{1,2},\hat \eta_{2,1})$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
 All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $(\hat \mu_W^{*(\cdot,i^*)}, \hat \sigma_W^{*(\cdot,i^*)})$, which is denoted by a red star.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_W^{*(\cdot,i^*)}, \hat \sigma_W^{*(\cdot,i^*)})$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
 All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_x}^{*(\cdot,i^*)}, \hat \sigma_{A_x}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_x}^{*(\cdot,i^*)}, \hat \sigma_{A_x}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
 All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_y}^{*(\cdot,i^*)}, \hat \sigma_{A_y}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_y}^{*(\cdot,i^*)}, \hat \sigma_{A_y}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
 All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_z}^{*(\cdot,i^*)}, \hat \sigma_{A_z}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_z}^{*(\cdot,i^*)}, \hat \sigma_{A_z}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the fine-scale probability transition matrix parameters associated both dive types ($\eta^{(i)*}_{i^*,j^*}$ for $i = 1,2$ and $i^*,j^* = 1,2,3$).
-Recall that $\eta^{(i)*}_{i^*,i^*} = 0$ and that $\Gamma^{(i)*}_{i^*,j^*} = \exp(\eta^{(i)*}_{i^*,j^*}) / \sum_{k^*} \exp(\eta^{(i)*}_{i^*,k^*})$.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the fine-scale probability transition matrix parameters associated both dive types ($\eta^{*(i)}_{i^*,j^*}$ for dive types $i = 1,2$ and subdive states $i^*,j^* = 1,2,3$).
+Recall that $\eta^{*(i)}_{i^*,i^*} = 0$ and that $\Gamma^{*(i)}_{i^*,j^*} = \exp(\eta^{*(i)}_{i^*,j^*}) / \sum_{k^*} \exp(\eta^{*(i)}_{i^*,k^*})$.
 All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is scaled by the log-likelihood of the maximum likelihood estimate $\hat \Gamma^{(i)*}$, which is denoted by a red star.
-Note that the long ridges of approximately equal likelihood corresponds to $\hat \Gamma^{(i)*}_{1,3} \approx \hat \Gamma^{(i)*}_{3,1} \approx 0$.
-This indicates that the whale never switches directly between subdive states 1 and 3.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\hat \Gamma^{*(i)}$, which is denoted by a red star.
+Note that the long ridges of approximately equal likelihood are due to the fact that $\hat \Gamma^{*(i)}_{1,3} \approx \hat \Gamma^{*(i)}_{3,1} \approx 0$.
+This indicates that the whale rarely switches directly between subdive states 1 and 3.
         \addtocounter{fignum}{1}
@@
                 \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.449 (0.030)$ & $0.039 (0.001)$ & $0.968 (0.002)$ \\
-                                                                                               & 2                                    & $0.210 (0.012)$ & $0.096 (0.002)$ & $0.829 (0.010)$ \\
-                                                                                               & 3                                    & $0.232 (0.035)$ & $0.296 (0.010)$ & $0.607 (0.042)$ \\ \hline
-                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.450 (0.038)$ & $0.051 (0.001)$ & $0.968 (0.009)$ \\
-                                                                                               & 2                                    & $0.437 (0.012)$ & $0.094 (0.002)$ & $0.829 (0.016)$ \\
-                                                                                               & 3                                    & $0.366 (0.042)$ & $0.365 (0.012)$ & $0.607 (0.040)$ \\ \hline
-                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.691 (0.043)$ & $0.058 (0.001)$ & $0.968 (0.005)$ \\
-                                                                                               & 2                                    & $-0.573 (0.014)$ & $0.111 (0.002)$ & $0.829 (0.012)$ \\
-                                                                                               & 3                                    & $-0.303 (0.041)$ & $0.354 (0.012)$ & $0.607 (0.039)$ \\ \hline
+                                                                                               & 2                                    & $0.210 (0.012)$ & $0.096 (0.002)$ & $0.829 (0.007)$ \\
+                                                                                               & 3                                    & $0.232 (0.035)$ & $0.296 (0.010)$ & $0.607 (0.023)$ \\ \hline
+                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.450 (0.038)$ & $0.051 (0.001)$ & $0.968 (0.002)$ \\
+                                                                                               & 2                                    & $0.437 (0.012)$ & $0.094 (0.002)$ & $0.829 (0.007)$ \\
+                                                                                               & 3                                    & $0.366 (0.042)$ & $0.365 (0.012)$ & $0.607 (0.023)$ \\ \hline
+                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.691 (0.043)$ & $0.058 (0.001)$ & $0.968 (0.002)$ \\
+                                                                                               & 2                                    & $-0.573 (0.014)$ & $0.111 (0.002)$ & $0.829 (0.007)$ \\
+                                                                                               & 3                                    & $-0.303 (0.041)$ & $0.354 (0.012)$ & $0.607 (0.023)$ \\ \hline
                 \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $34.02 (0.36)$ & $22.99 (0.38)$ & ---           \\
@@
         \subsection{HHMM-DFT}
+
             \begin{center}
@@
         \subsection{CarHMM-DFT}
+
             \begin{center}
@@
 The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
-Note that $\hat \mu_A^{*(\cdot,1)}$ is undefined because $\hat \phi_A^{*(\cdot,1)} = 1$, indicating that the mean of $\Zone_{t,\tilde t^*}$ is equal to $\Zone_{t,\tilde t^*-1}$ when in subdive state 1.
                 \addtocounter{tablenum}{1}
@@
 
-\iffalse
-\newpage
-%\section{Simulation Study Results}
-
-    \section{Simulated Data}
-
-        \begin{center}
-    	\includegraphics[width=5in]{../Plots/sim_data.png}
-    	\end{center}
-
-    	\noindent Figure \arabic{fignum}: Acceleration data for a five dives of a selected simulated data set.
-The raw accelerometer data is denoted as $Y^*_{t,\tilde t^*}$ which is recorded at 50 hertz while $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are recorded at 0.5 hertz.
-Dives are separated by vertical black lines.
-The colour of the curve corresponds to the true fine-scale state, while the colour of the background corresponds to the true dive type.
-Subdive state 1 can plausibly be interpreted as ``gliding" while subdive state 2 can plausibly be interpreted as active swimming, or ``fluking".
-    	\addtocounter{fignum}{1}
-
-    \newpage
-    \section{Accuracy comparisons}
-
-        \begin{center}
-        \scalebox{0.75}{
-        \bgroup
-        \def\arraystretch{1.5}
-        \begin{tabular}{ccccccc}
-Model                       & \multicolumn{1}{c}{Train Time (m)} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Dive Accuracy} & \multicolumn{1}{c}{Subdive Accuracy}  \\ \hline
-\multirow{5}{*}{CarHMM-DFT} & \multirow{5}{*}{$70 \pm 11$}   & Both                          & Both                             & -------------                     & $0.934 \pm 0.005$                       \\
-                            &                                    & 1                             & 1                                & \multirow{2}{*}{-------------}    & $0.794 \pm 0.037$                       \\
-                            &                                    & 1                             & 2                                &                                   & $0.958 \pm 0.008$                       \\
-                            &                                    & 2                             & 1                                & \multirow{2}{*}{-------------}    & $0.924 \pm 0.012$                       \\
-                            &                                    & 2                             & 2                                &                                   & $0.941 \pm 0.011$                       \\ \hline
-\multirow{5}{*}{HHMM-DFT}   & \multirow{5}{*}{$209 \pm 51$}   & Both                          & Both                             & $0.942 \pm 0.036$                   & $0.881 \pm 0.041$                       \\
-                            &                                    & 1                             & 1                                & \multirow{2}{*}{$0.966\pm0.030$}    & $0.628 \pm 0.120$                       \\
-                            &                                    & 1                             & 2                                &                                   & $0.956 \pm 0.014$                       \\
-                            &                                    & 2                             & 1                                & \multirow{2}{*}{$0.852\pm0.095$}    & $0.789 \pm 0.158$                       \\
-                            &                                    & 2                             & 2                                &                                   & $0.918 \pm 0.028$                       \\ \hline
-\multirow{5}{*}{CarHHMM}    & \multirow{5}{*}{$236 \pm 52$}   & Both                          & Both                             & $0.873 \pm 0.165$                   & $0.740 \pm 0.051$                       \\
-                            &                                    & 1                             & 1                                & \multirow{2}{*}{$0.886\pm0.214$}    & $0.765 \pm 0.205$                       \\
-                            &                                    & 1                             & 2                                &                                   & $0.707 \pm 0.090$                       \\
-                            &                                    & 2                             & 1                                & \multirow{2}{*}{$0.823\pm0.136$}    & $0.873 \pm 0.225$                       \\
-                            &                                    & 2                             & 2                                &                                   & $0.657 \pm 0.099$                       \\ \hline
-\multirow{5}{*}{CarHHMM-DFT}& \multirow{5}{*}{$132 \pm 40$}   & Both                          & Both                             & $0.944 \pm 0.036$                   & $0.934 \pm 0.005$                       \\
-                            &                                    & 1                             & 1                                & \multirow{2}{*}{$0.963\pm0.031$}    & $0.759 \pm 0.042$                       \\
-                            &                                    & 1                             & 2                                &                                   & $0.962 \pm 0.008$                       \\
-                            &                                    & 2                             & 1                                & \multirow{2}{*}{$0.872\pm0.086$}    & $0.934 \pm 0.010$                       \\
-                            &                                    & 2                             & 2                                &                                   & $0.932 \pm 0.011$                       \\ \hline
-\end{tabular}
-        \egroup
-        }
-        \end{center}
-
-        \noindent Table \arabic{tablenum}: Average decoding accuracies and training times for all models used to categorize dive type and subdive state in the simulation study.
-Each of the four models was fit to 500 training data sets comprised of 100 simulated dives and tested on test data sets also comprised of 100 simulated dives.
-Reported values are averages, and $\pm$ refers to the sample standard deviation across the 500 data sets.
-Rows labelled Both/Both correspond to overall average decoding accuracy.
-        \addtocounter{tablenum}{1}
-
-    \newpage
-    \section{Empirical joint distributions - dive duration ($Y_t$)}
-
-        \begin{center}
-        \scalebox{0.75}{
-        \bgroup
-        \centering
-        \def\arraystretch{1.5}
-\begin{tabular}{ccccccc}
-Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}   & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
-\multirow{4}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu$}        & 1                             & $42.161$                         & ---          & $4.255$                             & $2.036 \pm 0.369$                             \\
-                            &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\
-                            & \multirow{2}{*}{$\sigma$}     & 1                             & $31.119$                         & ---          & $5.393$                             & $2.014 \pm 0.452$                             \\
-                            &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\ \hline
-\multirow{4}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu$}        & 1                             & $25.776$                         & $0.096$ $(p=0.071)$          & $1.181$                             & $0.959 \pm 0.111$                             \\
-                            &                               & 2                             & $108.589$                         & $3.989$ $(p=0.000)$          & $18.726$                             & $13.083 \pm 4.546$                             \\
-                            & \multirow{2}{*}{$\sigma$}     & 1                             & $9.489$                         & $-0.081$ $(p=0.058)$          & $0.952$                             & $0.779 \pm 0.102$                             \\
-                            &                               & 2                             & $59.514$                         & $-5.186$ $(p=0.000)$          & $14.388$                             & $11.080 \pm 3.737$                             \\ \hline
-\multirow{4}{*}{CarHHMM}    & \multirow{2}{*}{$\mu$}        & 1                             & $27.877$                         & $2.197$ $(p=0.000)$          & $11.693$                             & $0.966 \pm 0.150$                             \\
-                            &                               & 2                             & $96.423$                         & $-8.177$ $(p=0.000)$          & $33.095$                             & $12.341 \pm 6.893$                             \\
-                            & \multirow{2}{*}{$\sigma$}     & 1                             & $9.954$                         & $0.384$ $(p=0.347)$          & $8.950$                             & $0.793 \pm 0.142$                             \\
-                            &                               & 2                             & $54.291$                         & $-10.409$ $(p=0.000)$          & $18.432$                             & $10.142 \pm 5.496$                             \\ \hline
-\multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu$}        & 1                             & $25.719$                         & $0.039$ $(p=0.455)$          & $1.170$                             & $0.958 \pm 0.108$                             \\
-                            &                               & 2                             & $107.404$                         & $2.804$ $(p=0.001)$          & $18.535$                             & $12.830 \pm 4.471$                             \\
-                            & \multirow{2}{*}{$\sigma$}     & 1                             & $9.436$                         & $-0.134$ $(p=0.001)$          & $0.930$                             & $0.777 \pm 0.097$                             \\
-                            &                               & 2                             & $59.635$                         & $-5.065$ $(p=0.000)$          & $14.225$                             & $10.945 \pm 3.548$
-\end{tabular}
-        \egroup
-        }
-        \end{center}
-
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of dive duration across all four models in the simulation study.
-The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
-The $\pm$ refers to the inter-quartile range.
-SE stands for standard error.
-Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
-        \addtocounter{tablenum}{1}
-
-        \subsection{CarHHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_dive_duration_-1_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_dive_duration_-1_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$) for each dive type.
-The red star represents the true values of $\mu$ and $\sigma$ for both dive types.
-These estimates are for the CarHHMM-DFT.
-        \addtocounter{fignum}{1}
-
-        \subsection{HHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_dive_duration_-1_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_dive_duration_-1_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$) for each dive type.
-The red star represents the true values of $\mu$ and $\sigma$ for both dive types.
-These estimates are for the HHMM-DFT.
-        \addtocounter{fignum}{1}
-
-        \subsection{CarHHMM}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_V_MLE_density_dive_duration_-1_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_V_MLE_density_dive_duration_-1_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$) for each dive type.
-The red star represents the true values of $\mu$ and $\sigma$ for both dive types.
-These estimates are for the CarHHMM.
-        \addtocounter{fignum}{1}
-
-        \subsection{CarHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_dive_duration_-1_0.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of dive duration parameters ($\hat \mu$ and $\hat \sigma$).
-The red star represents the true values of $\mu$ and $\sigma$ for both dive types.
-These estimates are for the CarHMM-DFT.
-The CarHMM-DFT assumes that there is only one dive type.
-There is no red star which represents the true values because the CarHMM-DFT assumes that there is only one dive type.
-        \addtocounter{fignum}{1}
-
-    \newpage
-    \section{Empirical joint distributions - acceleration ($\Zone_{t,\tilde t^*}$)}
-
-        \begin{center}
-        \scalebox{0.75}{
-        \bgroup
-        \centering
-        \def\arraystretch{1.5}
-\begin{tabular}{ccccccc}
-Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}   & \multicolumn{1}{c}{Observed Fischer SE}        \\ \hline
-\multirow{6}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu_A^*$}    & 1                                & $-0.001$                         & $-0.001$ $(p=0.934)$          & $0.142$                             & $0.066 \pm 0.045$                             \\
-                            &                               & 2                                & $-0.000$                         & $-0.000$ $(p=0.612)$          & $0.018$                             & $0.017 \pm 0.003$                             \\
-                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.034$                         & $-0.016$ $(p=0.000)$          & $0.001$                             & $0.001 \pm 0.000$                             \\
-                            &                               & 2                                & $0.079$                         & $-0.021$ $(p=0.000)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
-                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & $0.978$                         & $-0.012$ $(p=0.000)$          & $0.010$                             & $0.009 \pm 0.003$                             \\
-                            &                               & 2                                & $0.868$                         & $-0.082$ $(p=0.000)$          & $0.016$                             & $0.016 \pm 0.002$                             \\ \hline
-\multirow{6}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu_A^*$}    & 1                                & $-0.001$                         & $-0.001$ $(p=0.302)$          & $0.027$                             & $0.006 \pm 0.001$                             \\
-                            &                               & 2                                & $0.000$                         & $0.000$ $(p=0.921)$          & $0.012$                             & $0.004 \pm 0.000$                             \\
-                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.136$                         & $0.086$ $(p=0.000)$          & $0.022$                             & $0.004 \pm 0.001$                             \\
-                            &                               & 2                                & $0.144$                         & $0.044$ $(p=0.000)$          & $0.007$                             & $0.003 \pm 0.000$                             \\
-                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & ------                         & ------                     & ------                             & ------                                      \\
-                            &                               & 2                                & ------                         & ------                     & ------                             & ------                                      \\ \hline
-\multirow{6}{*}{CarHHMM}    & \multirow{2}{*}{$\mu_A^*$}    & 1                                & $0.021$                         & $0.021$ $(p=0.436)$          & $0.591$                             & $0.016 \pm 0.004$                             \\
-                            &                               & 2                                & $0.000$                         & $0.000$ $(p=0.669)$          & $0.013$                             & $0.006 \pm 0.001$                             \\
-                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.065$                         & $0.015$ $(p=0.000)$          & $0.022$                             & $0.001 \pm 0.000$                             \\
-                            &                               & 2                                & $0.187$                         & $0.087$ $(p=0.000)$          & $0.018$                             & $0.004 \pm 0.000$                             \\
-                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & $0.891$                         & $-0.099$ $(p=0.000)$          & $0.155$                             & $0.010 \pm 0.002$                             \\
-                            &                               & 2                                & $0.182$                         & $-0.768$ $(p=0.000)$          & $0.137$                             & $0.027 \pm 0.002$                             \\ \hline
-\multirow{6}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_A^*$}    & 1                                & $-0.001$                         & $-0.001$ $(p=0.875)$          & $0.111$                             & $0.066 \pm 0.045$                             \\
-                            &                               & 2                                & $-0.000$                         & $-0.000$ $(p=0.564)$          & $0.018$                             & $0.017 \pm 0.003$                             \\
-                            & \multirow{2}{*}{$\sigma_A^*$} & 1                                & $0.034$                         & $-0.016$ $(p=0.000)$          & $0.001$                             & $0.001 \pm 0.000$                             \\
-                            &                               & 2                                & $0.079$                         & $-0.021$ $(p=0.000)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
-                            & \multirow{2}{*}{$\phi_A^*$}   & 1                                & $0.978$                         & $-0.012$ $(p=0.000)$          & $0.010$                             & $0.009 \pm 0.003$                             \\
-                            &                               & 2                                & $0.867$                         & $-0.083$ $(p=0.000)$          & $0.016$                             & $0.016 \pm 0.002$                             \\ \hline
-\end{tabular}
-        \egroup
-        }
-        \end{center}
-
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of acceleration across all four models in the simulation study.
-The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
-The $\pm$ refers to the inter-quartile range.
-SE stands for standard error.
-Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
-        \addtocounter{tablenum}{1}
-
-        \subsection{CarHHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_A_0_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_A_0_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of acceleration parameters ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
-These estimates are for the CarHHMM-DFT.
-        \addtocounter{fignum}{1}
-
-        \subsection{HHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_A_0_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_A_0_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of acceleration parameters ($\hat \mu^*_A$ and $\hat \sigma^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$ and $\sigma^*_A$ for both subdive states.
-These estimates are for the HHMM-DFT.
-Note that the HHMM-DFT assumes no auto-correlation in $\Zone_{t,\tilde t^*}$, so $\hat \phi_A^{*(\cdot,i^*)}$ is absent from these plots.
-        \addtocounter{fignum}{1}
-
-        \subsubsection{CarHHMM}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_V_MLE_density_A_0_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_V_MLE_density_A_0_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of acceleration parameters ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
-These estimates are for the CarHHMM.
-        \addtocounter{fignum}{1}
-
-        \subsubsection{CarHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_A_0_0.png}
-        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_A_0_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of acceleration parameters ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
-These estimates are for the CarHMM-DFT.
-        \addtocounter{fignum}{1}
-
-    \newpage
-    \subsection{Empirical joint distributions - wiggliness ($\Ztwo_{t,\tilde t^*}$)}
-
-        \begin{center}
-        \scalebox{0.75}{
-        \bgroup
-        \centering
-        \def\arraystretch{1.5}
-\begin{tabular}{ccccccc}
-Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}       \\ \hline
-\multirow{4}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu_W^*$}    & 1                                & $23.362$                         & $0.022$ $(p=0.408)$          & $0.582$                             & $0.419 \pm 0.061$                             \\
-                            &                               & 2                                & $300.689$                         & $-0.511$ $(p=0.213)$          & $9.144$                             & $4.679 \pm 0.336$                             \\
-                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & $12.949$                         & $-0.001$ $(p=0.974)$          & $0.539$                             & $0.388 \pm 0.061$                             \\
-                            &                               & 2                                & $329.585$                         & $-0.515$ $(p=0.280)$          & $10.622$                             & $5.748 \pm 0.396$                             \\ \hline
-\multirow{4}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu_W^*$}    & 1                                & $23.353$                         & $0.013$ $(p=0.632)$          & $0.620$                             & $0.431 \pm 0.067$                             \\
-                            &                               & 2                                & $298.441$                         & $-2.759$ $(p=0.000)$          & $10.438$                             & $4.788 \pm 0.432$                             \\
-                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & $12.938$                         & $-0.012$ $(p=0.656)$          & $0.583$                             & $0.404 \pm 0.064$                             \\
-                            &                               & 2                                & $328.496$                         & $-1.604$ $(p=0.001)$          & $10.482$                             & $5.822 \pm 0.460$                             \\ \hline
-\multirow{4}{*}{CarHHMM}    & \multirow{2}{*}{$\mu_W^*$}    & 1                                & ------                         & ------                     & ------                             & ------                                      \\
-                            &                               & 2                                & ------                         & ------                     & ------                             & ------                                      \\
-                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & ------                         & ------                     & ------                             & ------                                      \\
-                            &                               & 2                                & ------                         & ------                     & ------                             & ------                                      \\ \hline
-\multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_W^*$}    & 1                                & $23.341$                         & $0.001$ $(p=0.957)$          & $0.568$                             & $0.419 \pm 0.060$                             \\
-                            &                               & 2                                & $301.209$                         & $0.009$ $(p=0.982)$          & $8.869$                             & $4.697 \pm 0.353$                             \\
-                            & \multirow{2}{*}{$\sigma_W^*$} & 1                                & $12.918$                         & $-0.032$ $(p=0.183)$          & $0.526$                             & $0.387 \pm 0.059$                             \\
-                            &                               & 2                                & $329.684$                         & $-0.416$ $(p=0.365)$          & $10.225$                             & $5.773 \pm 0.396$
-\end{tabular}
-        \egroup
-        }
-        \end{center}
-
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of wiggliness across all four models in the simulation study.
-The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
-The $\pm$ refers to the inter-quartile range.
-SE stands for standard error.
-Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
-        \addtocounter{tablenum}{1}
-
-        \subsection{CarHHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_FoVeDBA_0_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_MLE_density_FoVeDBA_0_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of wiggliness parameters ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
-The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
-These estimates are for the CarHHMM-DFT.
-        \addtocounter{fignum}{1}
-
-        \subsection{HHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_FoVeDBA_0_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_MLE_density_FoVeDBA_0_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of wiggliness parameters ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
-The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
-These estimates are for the HHMM-DFT.
-        \addtocounter{fignum}{1}
-
-        \subsection{CarHHMM}
-        The CarHHMM did not use wiggliness to model the whale's behaviour.
-
-        \subsection{CarHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_FoVeDBA_0_0.png}
-        \includegraphics[width=3in]{../Plots/hmm_FV_MLE_density_FoVeDBA_0_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of estimates of wiggliness parameters ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
-The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
-These estimates are for the CarHMM-DFT.
-        \addtocounter{fignum}{1}
-
-    \newpage
-    \section{Empirical joint distributions - probability transition matrices ($\Gamma, \Gamma^*$)}
-
-        \begin{center}
-        \scalebox{0.75}{
-        \bgroup
-        \centering
-        \def\arraystretch{1.5}
-\begin{tabular}{ccccccc}
-Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}    & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}  & \multicolumn{1}{c}{Observed Fischer SE}       \\ \hline
-\multirow{6}{*}{CarHMM-DFT}  & $\Gamma_{12}$                 & ------                          & ------                     & ------                            & ------                                        \\
-                             & $\Gamma_{21}$                 & ------                          & ------                     & ------                            & ------                                        \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.179$                         & ------                     & $0.018$                           & $0.015 \pm 0.002$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.079$                         & ------                     & $0.011$                           & $0.008 \pm 0.001$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & ------                          & ------                     & ------                            & ------                                        \\
-                             & $\Gamma^{*(2)}_{21}$          & ------                          & ------                     & ------                            & ------                                        \\ \hline
-\multirow{6}{*}{HHMM-DFT}    & $\Gamma_{12}$                 & $0.206$                         & $-0.006$ $(p=0.033)$       & $0.063$                           & $0.049 \pm 0.011$                             \\
-                             & $\Gamma_{21}$                 & $0.822$                         & $0.013$ $(p=0.014)$        & $0.119$                           & $0.098 \pm 0.029$                             \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.312$                         & $-0.008$ $(p=0.011)$       & $0.070$                           & $0.048 \pm 0.019$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.047$                         & $-0.003$ $(p=0.000)$       & $0.014$                           & $0.010 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.139$                         & $-0.001$ $(p=0.158)$       & $0.018$                           & $0.017 \pm 0.003$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.146$                         & $-0.004$ $(p=0.001)$       & $0.025$                           & $0.019 \pm 0.004$                             \\ \hline
-\multirow{6}{*}{CarHHMM}     & $\Gamma_{12}$                 & $0.307$                         & $0.095$ $(p=0.000)$        & $0.257$                           & $0.052 \pm 0.021$                             \\
-                             & $\Gamma_{21}$                 & $0.711$                         & $-0.098$ $(p=0.000)$       & $0.280$                           & $0.102 \pm 0.040$                             \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.269$                         & $-0.051$ $(p=0.000)$       & $0.252$                           & $0.019 \pm 0.006$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.134$                         & $0.084$ $(p=0.000)$        & $0.081$                           & $0.016 \pm 0.003$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.120$                         & $-0.020$ $(p=0.025)$       & $0.196$                           & $0.008 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.144$                         & $-0.006$ $(p=0.001)$       & $0.042$                           & $0.019 \pm 0.006$                             \\ \hline
-\multirow{6}{*}{CarHHMM-DFT} & $\Gamma_{12}$                 & $0.210$                         & $-0.002$ $(p=0.473)$       & $0.064$                           & $0.050 \pm 0.011$                             \\
-                             & $\Gamma_{21}$                 & $0.818$                         & $0.009$ $(p=0.076)$        & $0.117$                           & $0.097 \pm 0.027$                             \\
-                             & $\Gamma^{*(1)}_{12}$          & $0.325$                         & $0.005$ $(p=0.132)$        & $0.069$                           & $0.048 \pm 0.018$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.050$                         & $0.000$ $(p=0.713)$        & $0.012$                           & $0.010 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.141$                         & $0.001$ $(p=0.336)$        & $0.017$                           & $0.016 \pm 0.003$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.152$                         & $0.002$ $(p=0.032)$        & $0.020$                           & $0.018 \pm 0.004$                             \\ \hline
-
-\end{tabular}
-        \egroup
-        }
-        \end{center}
-
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for all probability transition matrices across all four models in the simulation study.
-The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
-The $\pm$ refers to the inter-quartile range.
-SE stands for standard error.
-Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
-        \addtocounter{tablenum}{1}
-
-        \newpage
-        \subsection{CarHHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_Gamma_density_-1.png}
-
-        \includegraphics[width=3in]{../Plots/hhmm_FV_Gamma_density_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_Gamma_density_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM-DFT.
-There are two fine-scale probability transition matrices- one corresponding to each dive type.
-We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix.
-The red star represents the true values.
-        \addtocounter{fignum}{1}
-
-        \newpage
-        \subsection{HHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_Gamma_density_-1.png}
-
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_Gamma_density_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_FV_uncorr_Gamma_density_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the HHMM-DFT.
-There are two fine-scale probability transition matrices- one corresponding to each dive type.
-We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix.
-The red star represents the true values.
-        \addtocounter{fignum}{1}
-
-        \newpage
-        \subsection{CarHHMM}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hhmm_V_Gamma_density_-1.png}
-
-        \includegraphics[width=3in]{../Plots/hhmm_V_Gamma_density_0.png}
-        \includegraphics[width=3in]{../Plots/hhmm_V_Gamma_density_1.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM.
-There are two fine-scale probability transition matrices- one corresponding to each dive type.
-We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix.
-The red star represents the true values.
-        \addtocounter{fignum}{1}
-
-        \newpage
-        \subsection{CarHMM-DFT}
-        \begin{center}
-        \includegraphics[width=3in]{../Plots/hmm_FV_Gamma_density_0.png}
-        \end{center}
-
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution $\hat \Gamma^{*(1)}$ for the CarHMM-DFT.
-There is only one fine-scale probability transition matrix and no coarse-scale probability transition matrix since we assume that there is only one dive type for the CarHMM-DFT.
-We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix.
-There is no red star which represents the true values because the CarHMM-DFT assumes that there is only one dive type.
-        \addtocounter{fignum}{1}
-\fi
 \end{document}

### 2021-06-21 Report/Supp_Sim.tex
Updates from Overleaf

@@
 Dives are separated by vertical black lines.
-The colour of the curve corresponds to the true fine-scale state, while the colour of the background corresponds to the true dive type.
+The colour of the curve corresponds to the true subdive state, while the colour of the background corresponds to the true dive type.
 Subdive state 1 can plausibly be interpreted as ``gliding", subdive state 2 can plausibly be interpreted as active swimming, and subdive state 3 can be interpretted as sudden or aggressive movement.
@@
 Reported values are averages, and $\pm$ refers to the sample standard deviation across the 500 data sets.
-Rows labelled Both/Both correspond to overall average decoding accuracy.
+Rows labelled All/All correspond to overall average decoding accuracy.
         \addtocounter{tablenum}{1}
@@
 Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}   & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
-\multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.337$                         & $0.067$ $(p=0.287)$          & $1.393$                             & $1.025 \pm 0.124$                             \\
-                            &                               & 2                             & $127.140$                         & $0.070$ $(p=0.945)$          & $22.562$                             & $15.710 \pm 6.726$                             \\
-                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.801$                         & $-0.109$ $(p=0.035)$          & $1.154$                             & $0.838 \pm 0.110$                             \\
-                            &                               & 2                             & $59.520$                         & $-4.540$ $(p=0.000)$          & $17.678$                             & $13.040 \pm 5.530$                             \\ \hline
-\multirow{4}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.373$                         & $0.103$ $(p=0.084)$          & $1.327$                             & $1.030 \pm 0.113$                             \\
-                            &                               & 2                             & $126.843$                         & $-0.227$ $(p=0.818)$          & $22.004$                             & $15.367 \pm 6.105$                             \\
-                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.860$                         & $-0.050$ $(p=0.351)$          & $1.187$                             & $0.845 \pm 0.110$                             \\
-                            &                               & 2                             & $57.136$                         & $-6.924$ $(p=0.000)$          & $16.118$                             & $12.624 \pm 5.165$                             \\ \hline
-\multirow{4}{*}{CarHHMM}    & \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.248$                         & $-0.022$ $(p=0.826)$          & $2.196$                             & $1.042 \pm 0.138$                             \\
-                            &                               & 2                             & $117.624$                         & $-9.446$ $(p=0.000)$          & $34.336$                             & $15.251 \pm 7.111$                             \\
-                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.523$                         & $-0.387$ $(p=0.001)$          & $2.594$                             & $0.858 \pm 0.148$                             \\
-                            &                               & 2                             & $56.981$                         & $-7.079$ $(p=0.000)$          & $16.325$                             & $12.530 \pm 6.095$                             \\ \hline
-\multirow{4}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu_Y$}      & 1                             & $41.685$                         & ---                        & $4.268$                             & $2.001 \pm 0.347$                             \\
+\multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.33$                         & $0.07$ $(p=0.287)$          & $1.39$                             & $1.03 \pm 0.12$                             \\
+                            &                               & 2                             & $127.14$                        & $0.07$ $(p=0.945)$          & $22.56$                             & $15.71 \pm 6.726$                             \\
+                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.80$                         & $-0.11$ $(p=0.035)$          & $1.15$                             & $0.84 \pm 0.11$                             \\
+                            &                               & 2                             & $59.52$                         & $-4.54$ $(p=0.000)$          & $17.68$                             & $13.04 \pm 5.53$                             \\ \hline
+\multirow{4}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.37$                         & $0.10$ $(p=0.084)$          & $1.33$                             & $1.03 \pm 0.11$                             \\
+                            &                               & 2                             & $126.84$                         & $-0.23$ $(p=0.818)$          & $22.00$                             & $15.37 \pm 6.11$                             \\
+                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.86$                         & $-0.05$ $(p=0.351)$          & $1.19$                             & $0.85 \pm 0.11$                             \\
+                            &                               & 2                             & $57.14$                         & $-6.92$ $(p=0.000)$          & $16.12$                             & $12.62 \pm 5.17$                             \\ \hline
+\multirow{4}{*}{CarHHMM}    & \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.25$                         & $-0.02$ $(p=0.826)$          & $2.20$                             & $1.04 \pm 0.14$                             \\
+                            &                               & 2                             & $117.62$                         & $-9.45$ $(p=0.000)$          & $34.34$                             & $15.13 \pm 7.10$                             \\
+                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.52$                         & $-0.39$ $(p=0.001)$          & $2.59$                             & $0.86 \pm 0.15$                             \\
+                            &                               & 2                             & $56.98$                         & $-7.08$ $(p=0.000)$          & $16.33$                             & $12.53 \pm 5.95$                             \\ \hline
+\multirow{4}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu_Y$}      & 1                             & $41.69$                         & ---                        & $4.27$                             & $2.00 \pm 0.35$                             \\
                             &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\
-                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $30.299$                         & ---                        & $5.439$                             & $1.948 \pm 0.443$                             \\
+                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $30.30$                         & ---                        & $5.44$                             & $1.95 \pm 0.44$                             \\
                             &                               & ---                           & ---                            & ---                        & ---                                & ---                                         \\
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimates of the distributions of estimates of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
 These estimates are associated with the CarHHMM-DFT and correspond to both dive types.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of estimates of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
 These estimates are associated with the HHMM-DFT and correspond to both dive types.
@@
                             &                               & 3                                & ------                         & ------                     & ------                             & ------                                      \\ \hline
-\multirow{9}{*}{CarHHMM}    & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $0.060$                         & $0.060$ $(p=0.000)$          & $0.034$                             & $0.024 \pm 0.054$                             \\
-                            &                               & 2                                & $0.097$                         & $-0.003$ $(p=0.000)$          & $0.016$                             & $0.052 \pm 0.055$                             \\
-                            &                               & 3                                & $0.201$                         & $0.001$ $(p=0.634)$          & $0.058$                             & $0.018 \pm 0.007$                             \\
-                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.111$                         & $0.061$ $(p=0.000)$          & $0.004$                             & $0.006 \pm 0.042$                             \\
-                            &                               & 2                                & $0.285$                         & $0.185$ $(p=0.000)$          & $0.009$                             & $0.040 \pm 0.043$                             \\
-                            &                               & 3                                & $1.150$                         & $0.850$ $(p=0.000)$          & $0.080$                             & $0.003 \pm 0.001$                             \\
-                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & $0.830$                         & $-0.140$ $(p=0.000)$          & $0.048$                             & $0.016  $                             \\
-                            &                               & 2                                & $0.004$                         & $-0.826$ $(p=0.000)$          & $0.012$                             & $0.023  $                             \\
-                            &                               & 3                                & $0.000$                         & $-0.610$ $(p=0.000)$          & $0.000$                             & $0.016 \pm 0.004$                             \\ \hline
+\multirow{9}{*}{CarHHMM}    & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $0.060$                         & $0.060$ $(p=0.000)$          & $0.034$                             & $0.019 \pm 0.006$                             \\
+                            &                               & 2                                & $0.097$                         & $-0.003$ $(p=0.000)$          & $0.016$                             & $0.007 \pm 0.001$                             \\
+                            &                               & 3                                & $0.201$                         & $0.001$ $(p=0.634)$          & $0.058$                             & $0.062 \pm 0.011$                             \\
+                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.111$                         & $0.061$ $(p=0.000)$          & $0.004$                             & $0.003 \pm 0.000$                             \\
+                            &                               & 2                                & $0.285$                         & $0.185$ $(p=0.000)$          & $0.009$                             & $0.005 \pm 0.000$                             \\
+                            &                               & 3                                & $1.150$                         & $0.850$ $(p=0.000)$          & $0.080$                             & $0.048 \pm 0.009$                             \\
+                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & $0.830$                         & $-0.140$ $(p=0.000)$          & $0.048$                             & $0.016 \pm 0.003$                             \\
+                            &                               & 2                                & $0.004$                         & $-0.826$ $(p=0.000)$          & $0.012$                             & $0.001 \pm 0.023$                             \\
+                            &                               & 3                                & $0.000$                         & $-0.610$ $(p=0.000)$          & $0.000$                             & $0.000 \pm 0.000$                             \\ \hline
 \multirow{9}{*}{CarHMM-DFT} & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $-0.008$                         & $-0.008$ $(p=0.142)$          & $0.114$                             & $0.064 \pm 0.037$                             \\
@@
 SE stands for standard error.
+Note that standard error cannot be estimated when $\phi^*_A \approx 0$ or $\phi^*_A \approx 1$.
 Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of estimates of acceleration parameters ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of acceleration parameter estimates ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
 The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of estimates of acceleration parameters ($\hat \mu^*_A$ and $\hat \sigma^*_A$) for each subdive state.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of acceleration parameter estimates ($\hat \mu^*_A$ and $\hat \sigma^*_A$) for each subdive state.
 The red star represents the true values of $\mu^*_A$ and $\sigma^*_A$ for both subdive states.
@@
                             &                               & 2                                & $505.10$                         & $-0.69$ $(p=0.394)$          & $17.94$                             & $9.74 \pm 0.81$                             \\
-                            &                               & 3                                & $9787.1$                         & $17.2$ $(p=0.718)$           & $1060.5$                            & $280.2 \pm 97.1$                             \\
-                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.849$                         & $-0.001$ $(p=0.991)$         & $0.989$                             & $0.652 \pm 0.094$                             \\
-                            &                               & 2                                & $514.21$                         & $-2.47$ $(p=0.010)$          & $21.43$                             & $11.41 \pm 0.95$                             \\
-                            &                               & 3                                & $14430$                          & $-33$ $(p=0.647)$            & $1596$                              & $303 \pm 166$                             \\ \hline
+                            &                               & 3                                & $9787.1$                         & $17.2$ $(p=0.718)$           & $1060.5$                            & $262.1 \pm 97.120$                             \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.849$                         & $-0.001$ $(p=0.991)$         & $0.989$                             & $0.658 \pm 0.088$                             \\
+                            &                               & 2                                & $514.21$                         & $-2.47$ $(p=0.010)$          & $21.43$                             & $9.74 \pm 0.81$                             \\
+                            &                               & 3                                & $14430$                          & $-33$ $(p=0.647)$            & $1596$                              & $262 \pm 97$                             \\ \hline
 \multirow{6}{*}{HHMM-DFT}   & \multirow{3}{*}{$\mu_W^*$}    & 1                                & $35.067$                         & $0.087$ $(p=0.097)$          & $1.170$                             & $0.679 \pm 0.100$                             \\
                             &                               & 2                                & $500.93$                         & $-4.86$ $(p=0.000)$          & $24.02$                             & $10.75 \pm 1.21$                             \\
-                            &                               & 3                                & $6841.0$                         & $-2929.0$ $(p=0.000)$        & $1164.3$                            & $198.2 \pm 74.7$                             \\
-                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.800$                         & $-0.050$ $(p=0.351)$         & $1.192$                             & $0.676 \pm 0.104$                             \\
-                            &                               & 2                                & $513.26$                         & $-3.42$ $(p=0.001)$          & $22.60$                             & $12.50 \pm 1.30$                             \\
-                            &                               & 3                                & $11459$                          & $-3004$ $(p=0.000)$          & $1739$                              & $303 \pm 194$                             \\ \hline
+                            &                               & 3                                & $6841.0$                         & $-2929.0$ $(p=0.000)$        & $1164.3$                             & $198.2 \pm 77.1$                             \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.800$                         & $-0.050$ $(p=0.351)$         & $1.192$                             & $0.679 \pm 0.100$                             \\
+                            &                               & 2                                & $513.26$                         & $-3.42$ $(p=0.001)$          & $22.60$                             & $10.75 \pm 1.21$                             \\
+                            &                               & 3                                & $11459$                          & $-3004$ $(p=0.000)$          & $1739$                              & $198 \pm 77$                             \\ \hline
 \multirow{6}{*}{CarHHMM}    & \multirow{3}{*}{$\mu_W^*$}    & 1                                & ------                         & ------                     & ------                             & ------                                      \\
@@
                             &                               & 3                                & $9817.8$                         & $47.8$ $(p=0.328)$           & $1089.1$                            & $280.2 \pm 136.3$                             \\
-                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.890$                         & $0.040$ $(p=0.401)$          & $1.058$                             & $0.656 \pm 0.101$                             \\
-                            &                               & 2                                & $515.23$                         & $-1.45$ $(p=0.125)$          & $21.09$                             & $11.37 \pm 0.99$                             \\
-                            &                               & 3                                & $14437$                          & $-25$ $(p=0.740)$          & $1661$                             & $332 \pm 166$                             \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.890$                         & $0.040$ $(p=0.401)$          & $1.058$                             & $0.661 \pm 0.093$                             \\
+                            &                               & 2                                & $515.23$                         & $-1.45$ $(p=0.125)$          & $21.09$                             & $9.64 \pm 0.92$                             \\
+                            &                               & 3                                & $14438$                          & $-25$ $(p=0.740)$            & $1661$                              & $280 \pm 136$                             \\
 \end{tabular}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of estimates of wiggliness parameters ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of wiggliness parameter estimates ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
 The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
@@
 \multirow{2}{*}{CarHHMM-DFT}& $\Gamma_{12}$                 & $0.160$                         & $0.005$ $(p=0.030)$        & $0.056$                           & $0.042 \pm 0.012$                             \\
-                             & $\Gamma_{21}$                 & $0.907$                         & $-0.006$ $(p=0.214)$        & $0.102$                           & $0.116$                             \\ \hline
+                             & $\Gamma_{21}$                 & $0.907$                         & $-0.006$ $(p=0.214)$        & $0.102$                           & $0.094 \pm 0.032$                             \\ \hline
 \multirow{2}{*}{HHMM-DFT}   & $\Gamma_{12}$                 & $0.161$                         & $0.007$ $(p=0.007)$        & $0.055$                           & $0.041 \pm 0.011$                             \\
-                             & $\Gamma_{21}$                 & $0.916$                         & $0.004$ $(p=0.372)$        & $0.088$                           & $0.112$                             \\ \hline
-\multirow{2}{*}{CarHHMM}    & $\Gamma_{12}$                 & $0.198$                         & $0.044$ $(p=0.000)$        & $0.152$                           & $0.044 \pm 0.014$                             \\
-                             & $\Gamma_{21}$                 & $0.832$                         & $-0.081$ $(p=0.000)$        & $0.195$                           & $0.103 \pm 0.073$                             \\ \hline
+                             & $\Gamma_{21}$                 & $0.916$                         & $0.004$ $(p=0.372)$        & $0.088$                           & $0.091 \pm 0.028$                             \\ \hline
+\multirow{2}{*}{CarHHMM}    & $\Gamma_{12}$                 & $0.198$                         & $0.044$ $(p=0.000)$        & $0.152$                           & $0.043 \pm 0.013$                             \\
+                             & $\Gamma_{21}$                 & $0.832$                         & $-0.081$ $(p=0.000)$        & $0.195$                           & $0.093 \pm 0.034$                             \\ \hline
 \multirow{2}{*}{CarHMM-DFT} & $\Gamma_{12}$                 & ------                         & ------                   & ------                           & ------                                      \\
@@
 SE stands for standard error.
+Note that standard error cannot be estimated when $\Gamma_{i,j} \approx 0$.
 Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
@@
 \multirow{12}{*}{CarHHMM-DFT}& $\Gamma^{*(1)}_{12}$          & $0.258$                         & $0.006$ $(p=0.004)$        & $0.043$                           & $0.032 \pm 0.006$                             \\
-                             & $\Gamma^{*(1)}_{13}$          & $0.002$                         & $0.002$ $(p=0.000)$        & $0.004$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{13}$          & $0.002$                         & $0.002$ $(p=0.000)$        & $0.004$                           & $0.009 \pm 0.004$                             \\
                              & $\Gamma^{*(1)}_{21}$          & $0.083$                         & $0.001$ $(p=0.243)$        & $0.015$                           & $0.012 \pm 0.002$                             \\
                              & $\Gamma^{*(1)}_{23}$          & $0.052$                         & $-0.000$ $(p=0.984)$        & $0.010$                           & $0.009 \pm 0.001$                             \\
-                             & $\Gamma^{*(1)}_{31}$          & $0.003$                         & $0.003$ $(p=0.000)$        & $0.008$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{31}$          & $0.003$                         & $0.003$ $(p=0.000)$        & $0.008$                           & $0.014 \pm 0.005$                             \\
                              & $\Gamma^{*(1)}_{32}$          & $0.234$                         & $0.002$ $(p=0.322)$        & $0.042$                           & $0.036 \pm 0.008$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.111$                         & $0.146$ $(p=0.349)$        & $0.017$                           & $0.015 \pm 0.004$                             \\
-                             & $\Gamma^{*(2)}_{13}$          & $0.001$                         & $0.002$ $(p=0.000)$        & $0.002$                           & $\text{N/A}$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.151$                         & $-0.068$ $(p=0.584)$        & $0.023$                           & $0.021 \pm 0.005$                             \\
-                             & $\Gamma^{*(2)}_{23}$          & $0.036$                         & $0.016$ $(p=0.944)$        & $0.013$                           & $0.010 \pm 0.003$                             \\
-                             & $\Gamma^{*(2)}_{31}$          & $0.007$                         & $0.003$ $(p=0.000)$        & $0.024$                           & $\text{N/A}$                             \\
-                             & $\Gamma^{*(2)}_{32}$          & $0.252$                         & $0.008$ $(p=0.000)$        & $0.096$                           & $0.063 \pm 0.028$                             \\ \hline
+                             & $\Gamma^{*(2)}_{12}$          & $0.111$                         & $-0.001$ $(p=0.349)$        & $0.017$                           & $0.015 \pm 0.004$                             \\
+                             & $\Gamma^{*(2)}_{13}$          & $0.001$                         & $0.000$ $(p=0.000)$        & $0.002$                           & $0.003 \pm 0.002$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.151$                         & $0.001$ $(p=0.584)$        & $0.023$                           & $0.021 \pm 0.005$                             \\
+                             & $\Gamma^{*(2)}_{23}$          & $0.036$                         & $-0.000$ $(p=0.944)$        & $0.013$                           & $0.010 \pm 0.003$                             \\
+                             & $\Gamma^{*(2)}_{31}$          & $0.007$                         & $0.007$ $(p=0.000)$        & $0.024$                           & $0.025 \pm 0.015$                             \\
+                             & $\Gamma^{*(2)}_{32}$          & $0.252$                         & $0.026$ $(p=0.000)$        & $0.096$                           & $0.063 \pm 0.028$                             \\ \hline
 \multirow{12}{*}{HHMM-DFT}   & $\Gamma^{*(1)}_{12}$          & $0.240$                         & $-0.012$ $(p=0.000)$        & $0.042$                           & $0.032 \pm 0.006$                             \\
-                             & $\Gamma^{*(1)}_{13}$          & $0.003$                         & $0.003$ $(p=0.000)$        & $0.006$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{13}$          & $0.003$                         & $0.003$ $(p=0.000)$        & $0.006$                           & $0.009 \pm 0.004$                             \\
                              & $\Gamma^{*(1)}_{21}$          & $0.077$                         & $-0.005$ $(p=0.000)$        & $0.019$                           & $0.012 \pm 0.002$                             \\
                              & $\Gamma^{*(1)}_{23}$          & $0.047$                         & $-0.004$ $(p=0.000)$        & $0.009$                           & $0.008 \pm 0.001$                             \\
-                             & $\Gamma^{*(1)}_{31}$          & $0.005$                         & $0.005$ $(p=0.000)$        & $0.009$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{31}$          & $0.005$                         & $0.005$ $(p=0.000)$        & $0.009$                           & $0.014 \pm 0.004$                             \\
                              & $\Gamma^{*(1)}_{32}$          & $0.153$                         & $-0.079$ $(p=0.000)$        & $0.029$                           & $0.026 \pm 0.005$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.109$                         & $0.128$ $(p=0.000)$        & $0.018$                           & $0.016 \pm 0.004$                             \\
-                             & $\Gamma^{*(2)}_{13}$          & $0.002$                         & $0.003$ $(p=0.000)$        & $0.004$                           & $\text{N/A}$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.151$                         & $-0.073$ $(p=0.712)$        & $0.031$                           & $0.023 \pm 0.006$                             \\
-                             & $\Gamma^{*(2)}_{23}$          & $0.037$                         & $0.012$ $(p=0.009)$        & $0.013$                           & $0.011 \pm 0.004$                             \\
-                             & $\Gamma^{*(2)}_{31}$          & $0.021$                         & $0.005$ $(p=0.000)$        & $0.026$                           & $0.034  $                             \\
-                             & $\Gamma^{*(2)}_{32}$          & $0.135$                         & $-0.073$ $(p=0.000)$        & $0.057$                           & $0.038 \pm 0.018$                             \\ \hline
-\multirow{12}{*}{CarHHMM}    & $\Gamma^{*(1)}_{12}$          & $0.138$                         & $-0.114$ $(p=0.000)$        & $0.131$                           & $0.018 \pm 0.006$                             \\
-                             & $\Gamma^{*(1)}_{13}$          & $0.023$                         & $0.022$ $(p=0.000)$        & $0.107$                           & $0.039  $                             \\
+                             & $\Gamma^{*(2)}_{12}$          & $0.109$                         & $-0.004$ $(p=0.000)$        & $0.018$                           & $0.016 \pm 0.004$                             \\
+                             & $\Gamma^{*(2)}_{13}$          & $0.002$                         & $0.002$ $(p=0.000)$        & $0.004$                           & $0.005 \pm 0.002$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.151$                         & $0.001$ $(p=0.712)$        & $0.031$                           & $0.023 \pm 0.006$                             \\
+                             & $\Gamma^{*(2)}_{23}$          & $0.037$                         & $0.002$ $(p=0.009)$        & $0.013$                           & $0.011 \pm 0.004$                             \\
+                             & $\Gamma^{*(2)}_{31}$          & $0.021$                         & $0.021$ $(p=0.000)$        & $0.026$                           & $0.023 \pm 0.012$                             \\
+                             & $\Gamma^{*(2)}_{32}$          & $0.135$                         & $-0.091$ $(p=0.000)$        & $0.057$                           & $0.038 \pm 0.018$                             \\ \hline
+\multirow{12}{*}{CarHHMM}    & $\Gamma^{*(1)}_{12}$          & $0.138$                         & $-0.114$ $(p=0.000)$        & $0.131$                           & $0.018 \pm 0.005$                             \\
+                             & $\Gamma^{*(1)}_{13}$          & $0.023$                         & $0.022$ $(p=0.000)$        & $0.107$                           & $0.007 \pm 0.004$                             \\
                              & $\Gamma^{*(1)}_{21}$          & $0.037$                         & $-0.044$ $(p=0.000)$        & $0.054$                           & $0.007 \pm 0.001$                             \\
                              & $\Gamma^{*(1)}_{23}$          & $0.027$                         & $-0.025$ $(p=0.000)$        & $0.050$                           & $0.004 \pm 0.001$                             \\
-                             & $\Gamma^{*(1)}_{31}$          & $0.013$                         & $0.013$ $(p=0.000)$        & $0.022$                           & $0.013  $                             \\
+                             & $\Gamma^{*(1)}_{31}$          & $0.013$                         & $0.013$ $(p=0.000)$        & $0.022$                           & $0.011 \pm 0.005$                             \\
                              & $\Gamma^{*(1)}_{32}$          & $0.134$                         & $-0.099$ $(p=0.000)$        & $0.043$                           & $0.023 \pm 0.006$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.056$                         & $0.026$ $(p=0.000)$        & $0.035$                           & $0.008 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{13}$          & $0.002$                         & $0.022$ $(p=0.004)$        & $0.017$                           & $\text{N/A}$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.069$                         & $-0.113$ $(p=0.000)$        & $0.032$                           & $0.011 \pm 0.003$                             \\
-                             & $\Gamma^{*(2)}_{23}$          & $0.017$                         & $-0.009$ $(p=0.000)$        & $0.012$                           & $0.005 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{31}$          & $0.011$                         & $0.013$ $(p=0.000)$        & $0.016$                           & $0.029  $                             \\
-                             & $\Gamma^{*(2)}_{32}$          & $0.138$                         & $-0.093$ $(p=0.000)$        & $0.079$                           & $0.037 \pm 0.018$                             \\ \hline
+                             & $\Gamma^{*(2)}_{12}$          & $0.056$                         & $-0.056$ $(p=0.000)$        & $0.035$                           & $0.008 \pm 0.002$                             \\
+                             & $\Gamma^{*(2)}_{13}$          & $0.002$                         & $0.002$ $(p=0.004)$        & $0.017$                           & $0.002 \pm 0.001$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.069$                         & $-0.081$ $(p=0.000)$        & $0.032$                           & $0.011 \pm 0.003$                             \\
+                             & $\Gamma^{*(2)}_{23}$          & $0.017$                         & $-0.019$ $(p=0.000)$        & $0.012$                           & $0.005 \pm 0.002$                             \\
+                             & $\Gamma^{*(2)}_{31}$          & $0.011$                         & $0.011$ $(p=0.000)$        & $0.016$                           & $0.018 \pm 0.009$                             \\
+                             & $\Gamma^{*(2)}_{32}$          & $0.138$                         & $-0.088$ $(p=0.000)$        & $0.079$                           & $0.037 \pm 0.016$                             \\ \hline
 \multirow{6}{*}{CarHMM-DFT}  & $\Gamma^{*(1)}_{12}$          & $0.158$                         & ------                   & $0.019$                           & $0.014 \pm 0.002$                             \\
-                             & $\Gamma^{*(1)}_{13}$          & $0.001$                         & ------                   & $0.002$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{13}$          & $0.001$                         & ------                   & $0.002$                           & $0.003 \pm 0.001$                             \\
                              & $\Gamma^{*(1)}_{21}$          & $0.096$                         & ------                   & $0.012$                           & $0.010 \pm 0.001$                             \\
                              & $\Gamma^{*(1)}_{23}$          & $0.046$                         & ------                   & $0.007$                           & $0.007 \pm 0.001$                             \\
-                             & $\Gamma^{*(1)}_{31}$          & $0.002$                         & ------                   & $0.004$                           & $\text{N/A}$                             \\
+                             & $\Gamma^{*(1)}_{31}$          & $0.002$                         & ------                   & $0.004$                           & $0.008 \pm 0.003$                             \\
                              & $\Gamma^{*(1)}_{32}$          & $0.227$                         & ------                   & $0.035$                           & $0.030 \pm 0.006$                             \\
@@
 SE stands for standard error.
+Note that standard error cannot be estimated when $\Gamma^{*(i)}_{i^*,j^*} \approx 0$.
 Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution $\hat \Gamma^{*(1)}$ for the CarHMM-DFT.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of $\hat \Gamma^{*(1)}$ for the CarHMM-DFT.
 There is only one fine-scale probability transition matrix and no coarse-scale probability transition matrix since we assume that there is only one dive type for the CarHMM-DFT.

### 2021-06-21 Report/Sections/results.tex

--- 2021-06-03 Report/Sections/results.tex
@@
 %
-Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Figure 1 of the supplementary material), and visual inspection shows no obvious complicated dependence.
+Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Figure 1 of of supplement A), and visual inspection shows no obvious complicated dependence.
 Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor a moving-window transformation is called for.
@@
 We then have that $\z_{t,\tilde t^*} = \{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\}$, where $\zone_{t,\tilde t^*}$ is a three-dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window.
-Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Figure 1 of the supplementary material).
+Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Figure 1 of of supplement A).
 Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.
@@
 
+%To illustrate an application of this method and compare the candidate models, 
+We estimated the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the \textit{scipy} package in Python \citep{Virtanen:2019}.
+Each model was fit 100 times using random initializations, and
+%Let $\mu_0,\sigma_0$, and $\phi_0$ represent the initial mean, standard deviation, and auto-correlation parameters for a given observation.
+If the observation is strictly non-negative, then we set $\log(\mu_0) \sim \mathcal{N}(\log(\bar Y),s_{\log})$, $\log(\sigma) \sim \mathcal{N}(s_{\log},1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean of the observations and $s_{\log}$ represents the sample standard deviation the logarithm of the observations values.
+If the observation can be any real number, then we set $\mu_0 \sim \mathcal{N}(\bar Y,s)$, $\log(\sigma) \sim \mathcal{N}(s,1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean and $s$ represents the sample standard deviation of the observations values
+we kept the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines.
+
 \subsection{Case study results}
 
-%To illustrate an application of this method and compare the candidate models, 
-We estimated the parameters of all four models to the data shown in Figure \ref{fig:data} with direct likelihood maximization using the \textit{scipy} package in Python \citep{Virtanen:2019}.
-Each model was fit 100 times using random initializations, and we kept the parameters corresponding to the maximum likelihood of those 100 optimization routines.
-We first report the results from the full CarHHMM-DFT in detail and assess the quality of the fit.
-We then compare these results with those from the other candidate models.
+We first report the results from the full CarHHMM-DFT in detail and assess the quality of the fit. We then compare these results with those from the other candidate models.
 
@@
 No dive in this data set has a maximum depth deeper than 30 meters and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters.
-However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and significant numbers of fish have been observed to be caught near the surface \citep{Fortune:2020}.
+However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and researchers have observed significant numbers of prey-capture events near the surface \citep{Fortune:2020}.
 Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.
@@
 The means of ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ associated with each subdive state are separated by an order of magnitude (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}).
-Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest variance corresponding to $\Zone_{t,\tilde t^*}$.
+Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest standard deviation corresponding to $\Zone_{t,\tilde t^*}$.
 It also has the highest auto-correlation in $\Zone_{t,\tilde t^*}$.
 This implies less overall activity and more consistent acceleration compared to the other subdive states.
-Subdive state 2 has a mean ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1.
+Subdive state 2 has a mean ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ one order of magnitude higher than subdive state 1 and its acceleration has about twice the standard deviation compared to subdive state 1.
 The auto-correlation of acceleration is also slightly lower than subdive state 1.
@@
 \[\hat \Gamma = \begin{pmatrix} 
-0.788 & 0.212 \\
-0.809 & 0.191
+0.847 & 0.153 \\
+0.914 & 0.086
 \end{pmatrix} \text{ and }\]
 
-\[\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}\]
+\[\hat \delta = \begin{pmatrix} 0.857 & 0.143 \end{pmatrix}\]
 %
@@
 \[\hat \Gamma^{*(1)} = \begin{pmatrix} 
-0.679 & 0.321 & 0.000 \\
-0.038 & 0.904 & 0.058 \\
-0.000 & 0.232 & 0.768
+0.745 & 0.253 & 0.002 \\
+0.080 & 0.869 & 0.052 \\
+0.000 & 0.229 & 0.771
 \end{pmatrix}, \qquad 
 \hat \Gamma^{*(2)} = \begin{pmatrix} 
-0.859 & 0.141 & 0.000 \\
-0.114 & 0.841 & 0.045 \\
-0.000 & 0.216 & 0.784
+0.886 & 0.114 & 0.000 \\
+0.150 & 0.815 & 0.036 \\
+0.000 & 0.225 & 0.775
 \end{pmatrix},\]
 
-\[\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}\]
+\[\hat \delta^{*(1)} = \begin{pmatrix} 0.202 & 0.649 & 0.149 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.531 & 0.405 & 0.064 \end{pmatrix}\]
 %
 for dive types 1 and 2.
-In summary, about 79\% of dives are short dives of type 1, and the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1.
+In summary, about 86\% of dives are short dives of type 1, and the whale performs an average of 6.54 short type 1 dives before switching to dive type 2 and an average of 1.09 longer type 2 dives before switching back to dive type 1.
 This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
-Further, this killer whale is in the less active subdive state 1 40\% of the time during a dive of type 1 compared to only only 9\% of the time during a dive of type 2.
+Further, this killer whale is in the less active subdive state 1 only 20\% of the time during a dive of type 1 compared to 53\% of the time during a dive of type 2.
 Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}.
-Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.
+Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 4 of of supplement A also shows the probability of each dive type and subdive state given the data and the fitted model.
 
@@
 This implies that the true distribution of $\Ztwo_{t,\tilde t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study.
-See Sections 1.4 through 1.6 of the supplementary material for pseudoresidual plots corresponding to all observations and models.
+See Sections 5 through 7 of supplement A for pseudoresidual plots corresponding to all observations and models.
 
@@
 Each histogram is plotted together with the corresponding emission distribution estimated by the CarHHMM-DFT.
-Analogous histograms corresponding to the fine-scale observations are contained in Section 1 of the supplementary material.
+Analogous histograms corresponding to the fine-scale observations are contained in Section 6 and 7 of of supplement A.
 Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions.
@@
 
-The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive state 3 (see Figures 4 and 5 of the supplementary material).
-In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are approximately $50$ to $100$ percent larger than those of the CarHHMM-DFT.
-The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of the supplementary material).
+The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize 
+any given behaviour as subdive state 3
+%the behaviour at the beginning and end of dives as subdive state 3 
+(see Figures 11 and 12 of supplement A).
+In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are %approximately $50$ to $100$ percent
+significantly larger than those of the CarHHMM-DFT.
+The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of of supplement A).
 This suggests that including auto-correlation in the model significantly affects parameter estimates.
-Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of the supplementary material).
-These findings suggest that the HHMM-DFT is a significantly worse fit to these data than the full CarHHMM-DFT.
+%Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of of supplement A).
+These findings suggest
+and that the HHMM-DFT is a significantly worse fit to these data than the full CarHHMM-DFT.
 
-The CarHHMM does not model the ``wiggliness" of the acceleration data, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives} (see Figure 6 of the supplementary material).
+The CarHHMM does not model the ``wiggliness" of the acceleration data, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives} (also see Figure 13 of of supplement A).
 These results essentially disqualify the CarHHMM as a viable model for this data set.
-The pseudoresiduals of acceleration are also light-tailed relative to a Normal distribution (see Figure 16 of the supplementary material).
+The pseudoresiduals of acceleration are also light-tailed relative to a Normal distribution (see Figure 23 of of supplement A).
 
@@
 However, its lack of hierarchical structure means that it fails to differentiate between short and long dives.
-This model therefore does not infer the dive-level Markov chain or the relationship between the dive and subdive levels.
-In particular, the CarHMM-DFT does not indicate that the whale is more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives.
+This model therefore does not infer the dive-level Markov chain or the relationship between the dive types and subdive states.
+For example, the CarHMM-DFT does not indicate that the whale is more likely to be in subdive state 1 when engaged in longer dives compared to shorter dives.
 
-A more complete set of results for each of the candidate models is presented in Section 1 of the supplementary material.
+For a more complete set of results for each of the candidate models, see Supplement A.

### 2021-06-21 Report/Sections/results.tex
Add sentence on likelihood surfaces
Update Simulation Study in Report

@@
 
-The primary goal of this case study is to jointly estimate the dive types and (more importantly) the subdive behaviours of this killer whale, so only hierarchical models were considered to describe the kinematic data.
+The primary goal of this case study is to jointly estimate the dive types and (more importantly) the subdive behaviours of this killer whale, so we only consider models with a fine-scale component to describe the kinematic data.
 Defining a suitable hierarchical model involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse- and fine-scale observations.
@@
 %To illustrate an application of this method and compare the candidate models, 
-We estimated the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the \textit{scipy} package in Python \citep{Virtanen:2019}.
-Each model was fit 100 times using random initializations, and
+We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the \textit{scipy} package in Python \citep{Virtanen:2019}.
+Each model is fit 100 times using random initializations, and
 %Let $\mu_0,\sigma_0$, and $\phi_0$ represent the initial mean, standard deviation, and auto-correlation parameters for a given observation.
 If the observation is strictly non-negative, then we set $\log(\mu_0) \sim \mathcal{N}(\log(\bar Y),s_{\log})$, $\log(\sigma) \sim \mathcal{N}(s_{\log},1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean of the observations and $s_{\log}$ represents the sample standard deviation the logarithm of the observations values.
-If the observation can be any real number, then we set $\mu_0 \sim \mathcal{N}(\bar Y,s)$, $\log(\sigma) \sim \mathcal{N}(s,1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean and $s$ represents the sample standard deviation of the observations values
-we kept the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines.
+If the observation can be any real number, then we set $\mu_0 \sim \mathcal{N}(\bar Y,s)$, $\log(\sigma) \sim \mathcal{N}(s,1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean and $s$ represents the sample standard deviation of the observations values.
+we keep the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines (see Section 3.4 of \citet{Zucchini:2016} for details).
+Plots of the likelihood surface near the maximum likelihood parameter estimates are shown in Section 2 of Supplement A.
 

### 2021-06-21 Report/Sections/simulation.tex
Update Simulation Study in Report

@@
 The parameters used to generate the data are based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity.
-In particular, we set the number of subdive states to $N^*=2$ and $\Zone_{t,\tilde t^*}$ to a scalar instead of a three dimensional vector.
+In particular, we set $\Zone_{t,\tilde t^*}$ to a scalar instead of a three dimensional vector.
 We then fit all four models to the simulated data.
@@
 Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
+
+\[\Gamma = \begin{pmatrix} 0.847 & 0.153 \\ 0.914 & 0.086 \end{pmatrix}.\]
 %
-$$\Gamma = \begin{pmatrix} 0.79 & 0.21 \\ 0.81 & 0.19 \end{pmatrix}.$$
-%
-Dive duration is Gamma distributed and the coarse-scale emission parameters are 
-%
-\[\mu^{(1)} = 25.7s, \enspace \sigma^{(1)} = 9.6s, \enspace \mu^{(2)} = 104.6s, \enspace \sigma^{(2)} = 64.7s.\]
+Dive duration is Gamma-distributed and the coarse-scale emission parameters are
+
+\[\mu^{(1)} = 27.34s, \enspace \sigma^{(1)} = 10.96s, \enspace \mu^{(2)} = 127.55s, \enspace \sigma^{(2)} = 63.89s.\]
 %
@@
 Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \big\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \big\}$ with probability transition matrices
-%
-\[\Gamma^{*(1)} = \begin{pmatrix} 0.68 & 0.32 \\ 0.05 & 0.95 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.86 & 0.14 \\ 0.15 & 0.85 \end{pmatrix}\]
+
+\[\Gamma^{*(1)} = \begin{pmatrix} 0.745 & 0.253 & 0.002 \\ 0.080 & 0.868 & 0.052 \\ 0.000 & 0.229 & 0.771 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.886 & 0.139 & 0.000 \\ 0.150 & 0.815 & 0.035 \\ 0.000 & 0.225 & 0.775 \end{pmatrix}\]
 %
@@
 We select the following parameters in line with the results from the case study:
-%$\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states.
-Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation.
-In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
+
+\begin{gather*}
+    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.97, \\
     %
-\begin{gather*}
-    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.034s, \enspace \phi_A^{*(\cdot,1)} = 0.98, \\
+    \mu_W^{*(\cdot,1)} = 34.01, \quad \sigma_W^{*(\cdot,1)} = 22.99, \\
     %
-    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.079s, \enspace \phi_A^{*(\cdot,2)} = 0.87, \\
+    \mu_A^{*(\cdot,2)} = 0.1 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.83, \\
     %
-    \mu_W^{*(\cdot,1)} = 23.3, \quad \sigma_W^{*(\cdot,1)} = 13.0, \\
+    \mu_W^{*(\cdot,2)} = 490.06, \quad \sigma_W^{*(\cdot,2)} = 502.56.
     %
-    \mu_W^{*(\cdot,2)} = 301.2, \quad \sigma_W^{*(\cdot,2)} = 330.1.
+    \mu_A^{*(\cdot,3)} = 0.2 s, \enspace \sigma_A^{*(\cdot,3)} = 0.3 s, \enspace \phi_A^{*(\cdot,3)} = 0.97, \\
+    %
+    \mu_W^{*(\cdot,3)} = 9154.16, \quad \sigma_W^{*(\cdot,3)} = 13538.75.
 \end{gather*}
@@
 It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix.
-Figure 21 of the supplementary material shows one realization of $\Z$ for five dives of one simulated data set along with the corresponding reconstructed realization of $Y^*$.
+Figure 1 of supplement B shows one realization of $\Z$ for five dives of one simulated data set along with the corresponding reconstructed realization of $Y^*$.
 
-The two simulated dive types differ in that dives of type 1 are much shorter on average (26 seconds) than dives of type 2 (105 seconds).
-The two simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 2 than for subdive state 1.
+The two simulated dive types differ in that dives of type 1 are much shorter on average (27 seconds) than dives of type 2 (128 seconds).
+The two simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 3 than for subdive state 2, which in turn is much higher than for subdive state 1.
 These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data.
 
-We calculate maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using the Cedar Compute Canada cluster with 1 CPU and 4 GB of dedicated memory per data set.
+We calculate maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using a similar procedure as the one from the case study.
+Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory per data set.
 
-For each of the 500 training data sets, we simulate a test data set to assess how well each model predicts the hidden states, as follows.
+For each of the 500 training data sets, we simulate a test data set to assess how well each model predicts the hidden states as follows.
 Each test data set consists of a sequence of 100 dives and is created from the generative model with true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
@@
 We then report the average of these over the 500 training/test data set pairs.
-Analogously, to assess prediction of the fine-scale states, we estimate $p^*_{t,\tilde t^*}(i^*|y,\z) \equiv \Pr(\tilde X^*_{t,\tilde t^*}=i|Y=y,\Z=\z)$, $i^*=1,2$, $\tilde t^* = 1,\ldots,\tilde T^*_t$, $t=1,\ldots,100$, using the test-set observations, the training-set maximum likelihood estimates, and the forward-backward algorithm.
+Analogously, to assess prediction of the fine-scale states, we estimate $p^*_{t,\tilde t^*}(i^*|y,\z) \equiv \Pr(\tilde X^*_{t,\tilde t^*}=i|Y=y,\Z=\z)$, $i^*=1,2,3$, $\tilde t^* = 1,\ldots,\tilde T^*_t$, $t=1,\ldots,100$, using the test-set observations, the training-set maximum likelihood estimates, and the forward-backward algorithm.
 Denoting the true fine-scale state realizations from the test data set as $\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}$, we define the overall \textit{average subdive decoding accuracy} as the average value of $\hat{p}^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*}|y,\z)$ across all simulated test data sets, dives, and windows.
-The conditional probabilities are estimated according to one of the four models under study, using the maximum likelihood estimates from the training data set in conjunction with the forward-backward algorithm.
-
-%For each of the 500 training data sets, we also simulate a test data set which consists of 100 dives and is created from the generative model with true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
-If the parameters of the generative model, coarse-scale test-set observations $y$, and fine-scale test-set observations $\z$ are known, then the \textit{forward-backward} algorithm can be used to calculate the hidden state probabilities $\Pr(X_{t} = i|Y=y,\Z=\z)$ and $\Pr(\tilde X^*_{t,t^*} = i^*|Y=y,\Z=\z)$ for all $t$, $t^*$, $i = 1,\ldots,N$, and $i^* = 1,\ldots,N^*$ \citep{Zucchini:2016}.
-However, since the true parameters rarely known in practice, the maximum likelihood estimates are often used instead to estimate the hidden state probabilities.
-As such, if $X_t$ is the unknown dive type associated with a given model and $x_t$ is the true simulated dive type of dive $t$, then we define the \textit{average dive decoding accuracy} as the average value of the estimates of $\Pr(X_{t} = x_t|Y=y,\Z=\z)$ across all simulated data sets and dives.
-Analogously, if $\tilde X^*_{t,\tilde t^*}$ is the unknown subdive state associated with a particular model and $\tilde x^*_{t,\tilde t^*}$ is the true simulated subdive state for dive $t$ and window $\tilde t^*$, then we define the \textit{average subdive decoding accuracy} as the average value of the estimates of $\Pr(\tilde X^*_{t,t^*} = \tilde x^*_{t,\tilde t^*}|Y=y,\Z=\z)$ across all simulated data sets, dives, and windows.
-We find both the average dive decoding accuracy and the average subdive decoding accuracy using the maximum likelihood estimates from each training set as well as the observations from each test set in conjunction with the forward-backward algorithm.
+The conditional probabilities are estimated according to each of the four models under study using the maximum likelihood estimates from the training data set in conjunction with the forward-backward algorithm.
 
@@
 The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model.
-Its average dive decoding accuracy and average subdive decoding accuracy are both greater than 0.9.
-All parameter estimates of fine-scale mean values ($\hat \mu^*$) and probability transition matrices ($\hat \Gamma^*$ and $\hat \Gamma$) have statistically insignificant biases.
+Its average dive decoding accuracy is approximately $0.96$ and its average subdive decoding accuracy is approximately $0.91$.
+All parameter estimates of emission distributions ($\theta$ and $\theta^*$) and probability transition matrices ($\Gamma$,$\Gamma^*$) on both the coarse scale and fine scale are either comparable or favourable to all other models.
+The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors.
+This underestimation is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to four times as large as the estimated standard error.
+See Tables 2 through 6 of supplement B for detailed results.
+%All parameter estimates of fine-scale mean values ($\hat \mu^*$) and probability transition matrices ($\hat \Gamma^*$ and $\hat \Gamma$) have statistically insignificant biases.
 In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models.
-The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors.
-This underestimation is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to twice as large as the estimated standard error.
-See Tables 5 through 9 of the supplementary material for detailed results.
 
-The HHMM-DFT performs similarly to the CarHHMM-DFT in most respects.
-Its average dive decoding accuracy is comparable to the CarHHMM-DFT while its average subdive decoding accuracy is worse by approximately 5 percentage points.
-Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$.
-In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \sigma_A^{*(\cdot,1)}$, and $\hat \sigma_A^{*(\cdot,2)}$ are much smaller than the associated empirical standard errors (see Table 7 of the supplementary material).
+The HHMM-DFT has a average dive decoding accuracy is comparable to the CarHHMM-DFT ($0.96$), but its average subdive decoding accuracy is worse by approximately 7 percentage points ($0.84$).
+Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$ and slightly overestimates $\sigma_A^{*(\cdot,3)}$.
+In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \mu_A^{*(\cdot,23)}$, $\hat \sigma_A^{*(\cdot,1)}$, $\hat \sigma_A^{*(\cdot,2)}$ and $\hat \sigma_A^{*(\cdot,3)}$ are much smaller than the associated empirical standard errors (see Table 3 of supplement B).
 These results suggest that estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored.
@@
 
-The CarHHMM is the worst-performing model in terms of accuracy, as its average dive decoding accuracy is below $0.9$ and its average subdive decoding accuracy is below $0.8$.
+The CarHHMM is the worst-performing model in terms of dive decoding accuracy, which is approximately 4 percentage points worse than the CarHHMM-DFT ($0.92$).
+Its average subdive decoding accuracy is approximately 6 percentage points worse than the CarHHMM-DFT ($0.85$).
 This result is consistent with expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct difference between the subdive states.
-In addition to its poor average decoding accuracy, the CarHHMM is also the worst of the four candidate models at estimating parameters.
-Parameter estimates associated with subdive type 2 ($\theta^{*(\cdot,2)}$) are especially poor.
-See Section 2 of the supplementary material for more detailed results.
+In addition to its relatively poor average decoding accuracy, the CarHHMM is also the worst of the four candidate models at estimating emission parameters and probability transition matrices.
+Estimates associated with subdive states 2 and 3 ($\theta^{*(\cdot,2)}, , \hat \mu_A^{*(\cdot,3)}$) are especially poor.
+See supplement B for more detailed results.
 
 Finally, the CarHMM-DFT is nearly identical to the CarHHMM-DFT in terms of average subdive decoding accuracy, fine-scale parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates.
-In addition, the time required to fit the CarHMM-DFT is less than half of that of the other models (see Table \ref{table:accuracy}).
+In addition, the time required to fit the CarHMM-DFT is less than a third of that of the other models (see Table \ref{table:accuracy}).
 However, this model cannot estimate dive type as it lacks any hierarchical structure.
 The CarHMM-DFT nonetheless fits a (misspecified) single Gamma distribution over the dive duration of all dives.
-The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 25 of the supplementary material).
+The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 5 of supplement B).
 
 Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} display five dives of one simulated data set as well as the decoded dive types and subdive states associated with each model.
-The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state while the HHMM-DFT is slightly more likely to predict that a given window corresponds to subdive state 2.
-As expected, the CarHHMM is the least accurate model when predicting subdive state.
-The CarHHMM-DFT and HHMM-DFT yield similar estimates of dive type while the CarHHMM is less accurate and even misclassifies the dive type of the second dive.
-The CarHMM-DFT does not estimate dive type.
+The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state, the HHMM-DFT is slightly more likely to misclassify subdive state 1, and the CarHHMM is most likely to misclassify subdive state 3.
+All models classify dive type with high accuracy with the exception of the CarHMM-DFT, which does not estimate dive type.

### 2021-06-24 Report/Reviewer_Response.tex
Update Report

\documentclass{article}

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{psfrag}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{url}
\usepackage{float}
\usepackage[margin=1in]{geometry}
\input{header}
\input{defs}

\begin{document}
\title{Modelling state-switching functional data with hidden Markov models: reviewer response}
\date{}
\author{Blinded}

\maketitle

We sincerely thank the reviewers for their comments, and we believe that the paper is better today than it was prior to peer review. Before addressing the reviewer comments one-by-one, there are several changes to the paper that we would like to highlight.

\begin{itemize}
    \item The case study has been re-run and the results have changed very slightly.
    \item The simulation study has been re-run with 3 behavioural states instead of 2.
    \item The supplemental material has been broken into two: supplement A contains material related to the case study while supplement B contains material related to the simulation study. 
\end{itemize}

\section{Reviewing: 1}

\begin{enumerate}
    \item 
    \begin{enumerate}
        \item \textit{What is the difference between `different curve types' and `between-curve dependence'?}

        The different curve types and between-curve dependence are closely related to one another. We assume that each curve is labelled with one of a discrete number of curve types, where each curve type corresponds to a distribution over curves of that type. We also assume that the sequence of curve types makes up an unobserved Markov chain, which is what induces the between-curve dependence. We have edited page 3, lines 49-53 as well as page 5, lines 83-84 to make this distinction clearer. In addition, we have changed ``different curve types" to ``a discrete number of curve types" where applicable to highlight that we are using a Markov chain with a discrete number of states to model the between-curve dependence.

        \item \textit{Is the paper trying to cluster the curves or identify the progression of curves at a given time t?}

        ``Our goal is to identify and describe a finite number of behaviours at multiple scales within sequential functional data. We simultaneously aim to model complicated temporal dependence both between and within curves." We have added this excerpt to page 5, lines 81-83 to clarify the goal of this paper. Using our case study as an example, we simultaneously identify and describe two different dive types and three different swimming behaviours. The distribution of the swimming behaviours can be used to estimate killer whale energy expenditure, and state-decoding algorithms can be used to identify possible foraging dives and feeding events.

        \item \textit{If the dependence between curves is purely based on the hidden variables, then there should be more justifications in 2.4 as curves are transformed to lower-dimensional variables. For functional time series, please see "On the prediction of stationary functional time series", "High-dimensional functional time series forecasting: An application to age-specific mortality rates".}

        We do transform the curves to lower-dimensional variables, which causes a significant amount of reconstruction error. We do not bound this reconstruction error, and only talk about reconstruction in appendix for the purpose of the simulation study. This is an admitted weakness of our paper. However, our goal is not to reconstruct the curves, but rather to identify the coarse-scale and fine-scale curve types as well as describe the dependence between those types. We added an explanation of this on pages 15-16, lines 281-294. Nonetheless, bounding prediction and reconstruction error is a promising area of future study. In particular, it would be useful to prove an analogue of Theorem 3.2 in "On the prediction of stationary functional time series" for state-switching functional time series. We added a sentence about this to the discussion on page 40, lines 754-757.

    \end{enumerate}

    \item \textit{What is the within-curve structure? There are numerous ways to deal with periodic functions. fPCA and smoothing with a Fourier basis will give you a better transformation of curves than the proposed in 2.4.}

    This is an excellent point- fPCA is a better way to reduce reconstruction error, and we added a discussion of this on pages 15-16, lines 281-294. However, our goal is to provide a means of summarizing these curves in a way that is interpretable for practitioners, not to reconstruct the curves.

    \item \begin{enumerate}
        \item \textit{There is no discussion about $x^*$ other than the transitional matrix. What's the significance of this hidden variable and in the hierarchy?}

        If $x_t = i$, then $x_{t,t*}^*$ represents one of $N^{*(i)}$ possible within-curve hidden behaviours associated with observation $y_{t,t*}^*$ and dive type $i$. We have added a sentence on page 12, lines 225-227 to make this more clear. For example, in the case study we have that there are 3 possible subdive states. If the whale is in subdive state 2 for dive t and window $\tilde t^*$, then $\tilde x^*_{t,\tilde t^*} = 2$.  In addition, we added verbiage on page 21, lines 405-411 to make clear what $x$ and $x^*$ are in the context of the case study.

        \item \textit{Have you tried to fit a simple HMM with only $x_t$ and $y_t$?}

        This is a good point- we suspect that a simple HMM would likely have comparable performance to our model to determine the dive types ($x_t$) and the conditional distributions of dive duration ($y_t$). However, we are primarily interested in jointly estimating the dive types and the subdive behaviours for the killer whale case study. If anything, we are personally more interested in the fine-scale behaviours. To make this more clear to the reader, we have added a sentence on page 22, lines 413-418.

    \end{enumerate}

    \item \begin{enumerate}
        \item \textit{The `between-curve dependence' or the hidden variable $x_t$ is significantly influenced by $y_t$ not the fine-scale observations as shown in Figure 4.}

        We have added some verbiage to the caption of Figure 1(d) to indicate that we are omitting the coarse-scale process and are only looking at the HMM-DFT as a fine-scale model of the within-curve behaviour of curve $t$.

        \item \textit{Where is the relationship between $x_t$ and $x^*_t$? The likelihood function does not include this relation.}

        $x_t$ corresponds to the curve type of curve $t$, and the curve type determines the parameters of the fine-scale model used to model the fine-scale observations $y^*_t$. If that fine-scale model is again an HMM, then $x^*_t$ is a Markov chain whose probability transition matrix is determined by the curve type ($x_t$). In addition, the distribution of $y^*_{t,t^*}$ is determined by both the curve type ($x_t$) and the fine-scale hidden state ($x^*_{t,t^*}$). We have edited page 13, lines 241-244 to make this fact more clear. The likelihood function does not include either $x_t$ or $x^*_t$ because both hidden variables can be marginalized out of the likelihood. We have made this more clear by adding verbiage to pages 8-9, lines 163-172 as well as page 14, lines 253-255.

    \end{enumerate}

    \item \textit{It would be better to see more than 2 hidden states where emission distributions are similar in the simulation studies.}

    It is confusing that we switch from 3 fine-scale hidden states in the case study to 2 fine-scale hidden states in the simulation study. As such, we re-ran our simulation study with 3 fine-scale hidden states and parameters that more closely represent the case-study. See section 4 of the paper for the updated results.  We have based the simulation study on the case study in order to simulate data that is as realistic as possible. We believe that this gives a good balance between separated and overlapping hidden states so the simulation study is informative.

    \item \textit{It is interesting to see the posterior distributions when the model is fitted with a sampling approach (sequential Monte Carlo).}

    This is an interesting approach that has been looked at in a couple of papers. This is beyond the scope of our paper, but we have added references to several sampling methods for HMMs (and relevant citations) on page 9, lines 165-170.

    \item \textit{After the transformation, does the autoregressive pattern hold? Is the likelihood function of $\tilde y^*$ in Equation (4) valid?}

    After the transformation we are redefining our model such that all modelling assumptions apply to the transformed variables rather than the original fine-scale variables. Therefore, the model selection and model validation techniques in Sections 3.2 and 3.4 should be  on the transformed observations, not the original observations. For example, Figure 1 of Supplement A shows lag plots for the transformed fine-scale observations, and there appears to be strong auto-correlation. We have added several sentences on page 17, lines 315-324 to clarify this.

    \item \begin{enumerate}
        \item \textit{The problem is overcomplicated, what is the optimization routine? }

        For each model, we ran 100 direct maximization routines using random initializations using the Nelder-Mead method of the optimize package of spicy. In particular, we optimized one row at a time of the probability transition matrices and one set of parameters associated with a (feature,hidden state) pair at a time. We found that most, but not all, of the optimization routines converged to similar parameter estimates. We have added an explanation of our optimization routine on pages 26-17, lines 502-510.

        \item \textit{Can you provide some surface plots of the likelihood?}

        We thank the reviewer for this suggestion, as plotting the likelihood gave us evidence that our optimization routine did in fact converge. We added plots of the likelihood surface associated with the CarHHMM-DFT in supplement A. Unfortunately the parameter space is high-dimensional, so we plotted the likelihood of the parameters associated with each feature / hidden state pair while keeping all other parameters fixed at their maximum likelihood estimates.

    \end{enumerate}

\end{enumerate}

\section{Reviewing: 2}

\textit{The authors of CJS-21-0003 present a novel and promising approach based on Hidden Markov Models to analyze functional data with intricate dependence structures. I would like to congratulate the authors on the very well-written paper. Below are just a few comments I would like the authors to address.}

\begin{enumerate}
    \item \textit{Introduction, page 5, around line 15. Before the work of Langrock et al. (2018), there has been the work of De Souza and Heckman (2014), which was the first paper proposing this new class of nonparametric regression models, called switching nonparametric regression models, allowing the hidden states to be independent or to follow a Markov structure. Later De Souza, Heckman, and Xu (2017) extended their approach to multiple curves allowing for different variance structures within each curve. Together with Langrock et. al. (2018), I believe these two references should be included as important examples of HMMs in nonparametric functional modelling:}

    \textit{de Souza, C. P. E., Heckman, N. E. and Xu, F., (2017) ``Switching nonparametric regression models for multi-curve data", The Canadian Journal of Statistics}

    \textit{de Souza, C. P. E. and Heckman, N. E. (2014) ``Switching nonparametric regression models", Journal of Nonparametric Statistics}

    We thank the reviewer for the references and have added them to page 5, lines 92-96.

    \item \textit{Section 2, end of page 6: the authors start the description of their model by presenting this idea of coarse-scale versus fine-scale, but I missed a connection between these and what was said in the introduction regarding the problem of analyzing biologging data. It would be great if the authors could make this connection clear. There is something written later in Section 2.4, maybe some of that can be said earlier.}

    The introduction primarily uses the adjectives "between-curve" and "within-curve" to focus on FDA, while the subsequent sections switch to "coarse-scale" and "fine-scale" to be more general. The connection between these adjectives was not clear before, so we have added a sentence to page 2, lines 34-36 to introduce "coarse-scale" vs "fine-scale" earlier. We have also added a sentence to page 5, lines 81-85 to make the connection between coarse-scale / fine-scale and curve types / behavioural states.

    \item \textit{Section 2, end of page 6: is there a special reason for denoting the number of curves by $T$? I find it less confusing to use, for example, $N$ or $M$ for that purpose so that each curve can be denoted by $i$ or $j$ because $t$ reminds me of evaluation time points which later you denote by $t^*$.}

    It is confusing that the number of curves is denoted by $T$, which usually indicates time points. However, we currently use $N$ and $N^*$ to represent the number of coarse-scale and fine-scale hidden states, respectively. In addition, the curves are assumed to be sequential in time, and there is a connection between the coarse scale index $t$ and the fine-scale index $t^*$. We have added a sentence on page 7, lines 138-139 to clarify that these curves are sequential but not necessarily equi-spaced in time. In addition, there is some precedence for indexing sequential curves with $t$- see "Cam\'er-Karhunen-Lo\`eve representation and harmonic principal component analysis of functional time series" by Panaretos and Tavakoli and "Dynamic functional principal components" by Hormann and Kidinski for examples.

    \item \textit{In Section 2.1, page 7 around line 50: it is important to explicitly say that a time-homogeneous chain is being considered, that is, transition probabilities do not depend on time.}

    We agree and have added a sentence to clarify this on page 8, lines 152-153.

    \item \textit{Page 8 around line 25: when presenting the likelihood $\mathcal{L}_{\text{HMM}}$, I would first present it and then mention the forward algorithm otherwise the reader may think that the equation has to do with the algorithm, but it is just the likelihood. Another thing is, I believe the algorithm is the backward-forward algorithm because it involves calculating both the so-called backward and forward quantities.}

    We have edited page 9, lines 172-177 to clarify the difference between the likelihood and the forward algorithm. We have also described the forward algorithm, which is used to evaluate the likelihood, as well as the forward-backward algorithm, which is used for state decoding and maximizing the likelihood via the Baum-Welch algorithm. See Zucchini et al (2016) for details.

    \item \textit{Could the author clarify how the maximum likelihood estimation is conducted? Is it via direct maximization of the likelihood using numerical methods or via the EM algorithm? I missed a section or paragraph describing exactly how inference is conducted. }

    For each model, we ran 100 direct maximization routines using random initializations using the Nelder-Mead method of the optimize package of spicy. In particular, we optimized one row at a time of the probability transition matrices and one set of parameters associated with a (feature,hidden state) pair at a time. We found that most, but not all, of the optimization routines converged to similar parameter estimates. We have added an explanation of our optimization routine on pages 26-17, lines 502-510.

    \item \textit{de Souza, C. P. E. et al. (2017) also consider the case where the distribution of the hidden states depends on some covariates such as weather conditions. In this case, for biologging data, would it be useful or would it make sense to potentially have transition probabilities depending on covariates via the same, logit link, representation that the authors already used.}

    This is a very useful and common practice in the ecology literature. While it is beyond the scope of our case study, we have added a brief discussion of covariates on page 10, lines 183-186.

\end{enumerate}

\end{document}

### 2021-06-24 Report/Sections/appendixnh.tex
Update Report

@@
 A key component is the discrete Fourier transformation of the 100 raw acceleration values in window $\tilde{t}^*$ of dive $t$:
+
 \[
@@
 We simulate the raw acceleration data for dive $t$ in three steps: (1) simulate the average acceleration within each window $\left(\hat Y^{*(0)}_{t,\tilde t^*}\right)$, (2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and (3) take the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$:
+
 \[
@@
 Specifically, we let
+
    \[
-    	\hat{Y}^{*(0)}_{t,1}|\tilde X^*_{t,1} = i^* ~~\sim~~ \mathcal{N} \left(0, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
+    	\hat{Y}^{*(0)}_{t,1}|\tilde X^*_{t,1} = i^* ~~\sim~~ \mathcal{N} \left(100 \mu^{*(\cdot,i^*)}_A, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
     	~~{\rm{and}}
    \]
-   %
+
     \begin{align}	
        \hat{Y}^{*(0)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = i^*,\hat{Y}^{*(0)}_{t,\tilde t^*-1}
-       &~~\sim ~~\mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,\tilde t^*-1}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \label{eqn:yhat_0} \\
+       &~~\sim ~~\mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,\tilde t^*-1} + 100(1-\phi_A^{*(\cdot,i^*)}) \mu^{*(\cdot,i^*)}_A, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \label{eqn:yhat_0} \\
         %
@@
     \end{align}
-    where $\sigma_A^{*(\cdot,1)} = 0.034s$,  $\sigma_A^{*(\cdot,2)} = 0.079s$. $\phi_A^{*(\cdot,1)} = 0.98$ and $\phi_A^{*(\cdot,2)} = 0.87$.
+    where $\mu_A^{*(\cdot,i^*)}$, $\sigma_A^{*(\cdot,i^*)}$, and $\phi_A^{*(\cdot,i^*)}$ are as defined in the simulation study in Section 4.
 
@@
  we first construct $\hat{Y}^{*(k)}_{t,\tilde t^*}$, $k=1,\ldots, 49$, as
+
     \begin{equation}
@@
 We include $i$ in Equation (\ref{eqn:abYhat}) to force all variation within a window to take the form of a sine wave, which reduces the variation between the endpoints of windows compared to a cosine wave.
-    Given the fine scale states, the $b^{(k)}_{t,\tilde t^*}$'s are independent and independent of the  $a^{(k)}_{t,\tilde t^*}$'s.  
+    Given the fine scale states, the $b^{(k)}_{t,\tilde t^*}$'s are independent of one another and independent of the $a^{(k)}_{t,\tilde t^*}$'s.  
    The distribution of $b^{(k)}_{t,\tilde t^*}$ is
+
     \begin{align}
     \begin{split}
-    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1 &~~\sim~~ {\rm{Gamma}}(16.38/k^3, 36.23) \\
+    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1 &~~\sim~~ {\rm{Gamma}}(11.03/(k+1)^3, 15.54) \\
     	%
-    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2  &~~\sim ~~{\rm{Gamma}}(4.20/k^3, 1825.53). \\ 
+    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2  &~~\sim ~~{\rm{Gamma}}(4.80/(k+1)^3, 515.38). \\ 
+    	%
+    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 3  &~~\sim ~~{\rm{Gamma}}(2.31/(k+1)^3, 20023.44). 
     \end{split}
@@
    We then define the remaining 50 Fourier coefficients:
+
     \[
@@
 $\Zone_{t,1}, \ldots, \Zone_{t,\tilde{T}^*}$ 
-follows a CarHMM with Normal emissions distributions and parameters 
- $\mu_A^{*(\cdot,1)} = \mu_A^{*(\cdot,2)} = \mu_A^{*(\cdot,i^*)} = 0$,  $\sigma_A^{*(\cdot,1)} = 0.034s$, $\phi_A^{*(\cdot,1)} = 0.98$, $\sigma_A^{*(\cdot,2)} = 0.079s$, and $\phi_A^{*(\cdot,2)} = 0.87$.
+follows a CarHMM with Normal emissions distributions and parameters as defined in the simulation study in Section 4.
+%$\mu_A^{*(\cdot,1)} = \mu_A^{*(\cdot,2)} = \mu_A^{*(\cdot,i^*)} = 0$,  $\sigma_A^{*(\cdot,1)} = 0.034s$, $\phi_A^{*(\cdot,1)} = 0.98$, $\sigma_A^{*(\cdot,2)} = 0.079s$, and $\phi_A^{*(\cdot,2)} = 0.87$.
 
@@
 the wiggliness within window $\tilde t^*$ of dive $t$ is
+
 \[
@@
 If $\tilde \omega < 50$, then $\Ztwo_{t,\tilde t^*}$ is the sum of independent Gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,\tilde t^*}$ is also Gamma.
-Thus, by Equation (\ref{eqn:bdist})
+Thus, by Equation (\ref{eqn:bdist}):
+
+\[
+    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 11.03/(k+1)^3 , 15.54 \right),
+\]
 %
 \[
-\Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 16.38/k^3 , 36.23 \right) 
-\text { and }
+    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.80/(k+1)^3 , 515.38 \right), \text{and}
 \]
+%
 \[
-\Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.20/k^3 , 1825.53 \right).
+    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 3  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 2.31/(k+1)^3 , 20023.44 \right).
 \]
-Setting $\tilde \omega$ to 10 and carrying out simple calculation of the mean and variance of a Gamma distribution yields $\mu_W^{*(\cdot,1)} = 23.3$, $\sigma_W^{*(\cdot,1)} = 13.0$, $\mu_W^{*(\cdot,2)} = 301.2$, and $\sigma_W^{*(\cdot,2)} = 330.1$.
+Setting $\tilde \omega$ to 10 and carrying out simple calculation of the mean and variance of a Gamma distribution gives the proper values for $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$.
 
@@
 The overall likelihood of the CarHHMM-DFT model is as follows:
-%
-$$\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,\z) = \delta P(y_1,\z_1;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,\z_t;\theta,\theta^*,\Gamma^*) \mathbf{1}_N.$$
+
+\[\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,\z) = \delta P(y_1,\z_1;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,\z_t;\theta,\theta^*,\Gamma^*) \mathbf{1}_N.\]
 %
 In particular,
-%
+
 \begin{align*}
@@
 The likelihood $\calL_{\text{fine}}$ corresponds to the fine-scale model and is equal to the following:
-%
-$$\calL_{\text{fine}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},$$
+
+\[\calL_{\text{fine}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},\]
 %

### 2021-06-24 Report/Sections/discussion.tex
update reviewer 1, comments 1-4a
Update Report

@@
 Here we use a CarHMM as the fine-scale model, but there are a wide range of possible fine-scale models, including a Poisson process or continuous time approach similar to that of \citet{Michelot:2019}.
-We also incorporate a moving-window transformation at the fine scale to capture intricate dependence structures on short time scales.
+We also incorporate a moving-window transformation at the fine scale to capture intricate dependence structures.
 Together, the coarse- and fine-scale models make up a hierarchical structure which can account for simultaneous processes taking place at different time scales.
 Provided the construction is not overly complex, a hierarchical model created using our method can be both flexible and easy to fit using maximum likelihood estimation.
+Our method is not intended to minimize prediction error for functional data, but incorporating HMMs into auto-regressive predictive models such as those described by \citet{Aue:2015} and \citet{Gao:2019} is a promising and natural direction for future study.
 
 We demonstrate the usefulness of this framework using a biomechanical/ecological example, where we use HMMs to classify the coarse- and fine-scale diving behaviour of a northern resident killer whale in Queen Charlotte Sound off the coast of British Columbia, Canada.
-Our analysis gives a deeper understanding of a killer whale's tri-axial movement and thus its behaviour and energy expenditure \citep{Gleiss:2011,Qasem:2012}, both of which are important for understanding the foraging ecology and nutritional status of northern resident killer whales \citep{Noren:2011}.
+Our analysis gives a deeper understanding of the killer whale's tri-axial movement and thus its behaviour and energy expenditure \citep{Gleiss:2011,Qasem:2012}, both of which are important for understanding the foraging ecology and nutritional status of northern resident killer whales \citep{Noren:2011}.
 Our model is also applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.

### 2021-06-24 Report/Sections/intro.tex
Update Report

@@
 These data contain a wealth of information about human and animal behaviour, but modelling these large data sets poses a challenge for statisticians and biologists.
-One particular difficulty addressed in this paper is that many coarse- and fine-scale processes often arise simultaneously in high-frequency data, and each of these can exhibit unique behaviours and dependence structures.
+One particular difficulty is that simultaneous coarse- and fine-scale processes are often reflected in high-frequency data, and each scale can exhibit unique and complicated dependence structures.
 
@@
 However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
-For example, dive profiles of marine animals show a discrete number of distinct dive ``types" that aggregate together in time \citep{Tennessen:2019b} while simultaneously displaying bouts of short-term periodicity within dives \citep{Adam:2019}.
-Fine-scale periodicity within a larger state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
+For example, dive profiles of marine animals show a discrete number of distinct dive ``types"
+%that aggregate together in time 
+where the sequence of dive types exhibits state-switching behaviour \citep{Tennessen:2019b}.
+These profiles simultaneously display bouts of short-term periodicity within each dive \citep{Adam:2019}.
+Fine-scale periodicity nested within a coarser state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
 
-Some FDA models account for between-curve dependence occurring when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain kinds of dependence in time.
+Some FDA models account for between-curve dependence that occurs when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain kinds of dependence in time.
 For example, previous studies have used multilevel models with random effects to model variation between and within individuals in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}.
@@
 This is not appropriate in most biologging applications where behaviours often change suddenly between a discrete number of types.
-In addition to multi-level models, FDA researchers have also used functional time series to model dependence in a sequence of curves.
-Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by well-defined hidden states.
+In addition to multi-level models, FDA researchers have used functional time series to model dependence in a sequence of curves.
+Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by a discrete number of well-defined hidden states.
 
@@
 
-The goal of this paper is to identify and describe a finite number of distinct curve types within sequential functional data while simultaneously modelling the complicated temporal dependence both between and within those curves.
-To accommodate both temporal dependence and sudden changes in behaviour, we turn to the field of animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
-HMMs interpret animal movement data as arising from a Markov chain of unobserved behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position.
+Our goal is to identify and describe discrete behavioural states at multiple scales within functional data (e.g., coarse-scale curve types and fine-scale behavioural states) and to model the dependence structure between those states.
+To accomplish this, we turn to the field of animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
+HMMs interpret animal movement data as arising from a Markov chain of a discrete number of behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position.
 While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions.
-In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, while \citet{DeSouza:2014} and \citet{DeSouza:2017} use HMMs to model state-switching behaviour within functional data.
-However, the first paper does not account for certain types of temporal correlation while the later two focus on state-switching behaviour within functions rather than between them.
+In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, and \citet{DeSouza:2014,DeSouza:2017} use HMMs to model state-switching behaviour within functional data.
+However, none of these papers account for certain types of temporal correlation taking place on multiple scales.
 
 While useful, HMMs alone are also not sufficient to model high-frequency time-series data for three primary reasons.
-Firstly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales (i.e., both between and within curves).
-The fine-scale, within-curve process is of particular interest in this work.
+Firstly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales (e.g., both between and within curves).
 To address this issue, statistical ecologists have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model both scales with conditionally dependent HMMs.
 Secondly, HMMs assume that subsequent observations are independent given an underlying hidden state process, but this is often not the case when observations are taken at extremely high frequencies.
-Several solutions have been proposed in the ecology literature such as the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
+Several solutions have been proposed in the ecology literature, including the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
 Thirdly, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures on short time scales.
@@
 
-In this paper, we combine existing methods from statistical ecology literature in novel ways to classify and describe state-switching behaviours in functional data while accounting for complex temporal dependence.
+We combine these existing methods from statistical ecology literature in novel ways to classify and describe state-switching behaviours in functional data while accounting for complex temporal dependence.
 The resulting suite of methods makes up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.

### 2021-06-24 Report/Sections/models.tex
Update Report

@@
 Consider a sequence of $T$ curves, where curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$.
-Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of fine-scale quantities derived from curve $t$ indexed by $t^*$.
+Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of fine-scale quantities derived from curve $t$ and indexed by $t^*$.
 Both $Y_t$ and $Y^*_{t,t^*}$ can be either vectors or scalars.
 We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.
-We assume that $\{Y_1,\ldots,Y_T\}$ is ordered sequentially but the curves need not be equi-spaced in time.
+We assume that $\{Y_1,\ldots,Y_T\}$ is ordered sequentially in time, but the curves need not be equally spaced.
 To develop our model for this data, we detail the structure of a traditional HMM followed by three variations which generalize its base structure.
@@
 which can be used to obtain estimates for both the hidden states $\hat X$ and the parameters $\hat \theta$ and $\hat \Gamma$.
-Using a Bayesian approach, $\calL_{\text{HMM}}$ can be combined with a prior distribution to obtain a posterior distribution over $\hat X$, $\hat \theta$, and $\hat \Gamma$ which can be sampled from using methods such as Sequential Monte Carlo \citep{Douc:2011b}, Markov-Chain Monte Carlo \citep{Scott:2002}, or variational inference \citep{Foti:2014}.
-For the purposes of this paper, however, we focus on the frequentest approach, where it is common to marginalize out $x$ to obtain
+Using a Bayesian approach, $\calL_{\text{HMM}}$ can be combined with a prior distribution to obtain a posterior distribution over $\hat X$, $\hat \theta$, and $\hat \Gamma$.
+This posterior distribution can be sampled from using methods such as sequential Monte Carlo \citep{Douc:2011b}, Markov-chain Monte Carlo \citep{Scott:2002}, or variational inference \citep{Foti:2014}.
+For the purposes of this paper, however, we focus on the frequentest approach, where it is common to marginalize out $x$ from the likelihood:
 
@@
 This equation can be evaluated in $\mathcal{O}(T)$ time using the well-known \textit{forward algorithm} \citep{Zucchini:2016}.
-After obtaining maximum likelihood estimates $\{\hat \theta, \hat \Gamma\}$, estimated probabilities of each hidden state $\big(\hat{\Pr}(X_t = i|Y = y)\big)$ can be calculated from $\{\hat \theta,\hat \Gamma\}$ using the \textit{forward-backward algorithm} \citep{Zucchini:2016}.
+After obtaining maximum likelihood estimates $\{\hat \theta, \hat \Gamma\}$, estimated probabilities of each hidden state $\big(\hat{\Pr}(X_t = i|Y = y)\big)$ can be calculated from $\{\hat \theta,\hat \Gamma\}$ using the similarly-named \textit{forward-backward algorithm} \citep{Zucchini:2016}.
 %
@@
 This formulation simplifies likelihood maximization by removing constraints in the optimization problem.
-We assume there are no covariate effects in this paper, \cite{DeSouza:2017} and \citet{Adam:2019} both incorporate covariates into $\Gamma$ by setting $\eta_{ij}(z_t) = \beta_{ij}^T z_t$ for $i \neq j$, where $z_t$ is a column vector of known covariates at time $t$ and $\beta_{ij}$ is a column vector of unknown regression coefficients.
+We assume there are no covariate effects, but \cite{DeSouza:2017} and \citet{Adam:2019} both incorporate covariates into $\Gamma$ by setting $\eta_{ij}(z_t) = \beta_{ij}^T z_t$ for $i \neq j$, where $z_t$ is a column vector of known covariates and $\beta_{ij}$ is a column vector of unknown regression coefficients.
 For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$.
@@
 A CarHMM also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
-Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$.
+Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $\{X_1,\ldots, X_T\}$ and $\{Y_1,\ldots, Y_{t-1}\}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$.
 The first emission $Y_1$ is treated as a fixed initial value which does not depend upon $X_1$.
 We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$.
-This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density or probability mass function that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$.
+%This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density or probability mass function that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$.
 As a concrete example, if $Y_t$ is a scalar, then one may assume that $Y_t$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
@@
 In the hierarchical setting, however, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X^*_t \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$.
-If $X_t = i$, then $X^*_{t,t^*} \in \{1,\ldots,N^{*(i)}\}$ represents one of $N^{*(i)}$ possible within-curve behaviours associated with curve type $i$.
 In turn, $X^*_t$ emits a sequence of fine-scale observations $Y^*_t \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$.
 For each curve $t$, the fine-scale process $\{X^*_t, Y^*_t\}$ follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
-If $X_t = i$, then the components of $X^*_t$ make up a Markov chain that has $N^{*(i)}$ possible states, an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$, and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
+If $X_t = i$, then $X^*_{t,t^*}$ represents one of $N^{*(i)}$ possible fine-scale behaviours associated with coarse-scale hidden state $i$.
+Further, the components of $X^*_t$ make up a Markov chain with an $N^{*(i)} \times N^{*(i)}$ probability transition matrix $\Gamma^{*(i)}$ and an initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
 The distribution of $Y^*_{t,t^*}$ given $Y^*_{t,t^*-1} = y^*_{t,t^*-1}$, $X^*_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$.
@@
 Forcing certain parameters to be shared can reduce the complexity and increase the interpretability of an HHMM.
-For example, since $f^{*(i,i^*)}$ depends on both $X_t = i$ and $X^*_{t,t^*} = i^*$, a within-curve behaviour is defined by the pair $(X_t,X^*_{t,t^*})$.
+For example, since $f^{*(i,i^*)}$ depends on both $X_t = i$ and $X^*_{t,t^*} = i^*$, the distribution of $Y^*_{t,t^*}$ is defined by the pair $(X_t,X^*_{t,t^*})$.
 However, in our killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$ and share the fine-scale emission parameters across the $N$ coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
-As a result, within-curve behaviours are shared between curve types and are defined exclusively by $X^*_{t,t^*}$ (or $(\cdot,X^*_{t,t^*})$ for notational consistency).
-Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$, not their fine-scale emission parameters $\theta^{*(\cdot,i)}$.
+As a result, the distribution of the fine-scale observation $Y^*_{t,t^*}$ is defined exclusively by $X^*_{t,t^*}$ (or $(\cdot,X^*_{t,t^*})$ for notational consistency).
+%Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$, not their fine-scale emission parameters $\theta^{*(\cdot,i)}$.
 
@@
 The fine-scale likelihood $\calL_{\text{fine}}$ corresponds to the likelihood of the fine-scale model, which can be either a CarHMM or an HMM.
-Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.
+Figure \ref{fig:models}c displays the dependence structure of an HHMM.
 
@@
 
-In many applications where data are collected at high frequencies, intricate dependency structures arise within the fine-scale process that cannot be adequately modelled with the HMM-based models described thus far.
+In many applications where data are collected at high frequencies, intricate dependency structures arise within the fine-scale process that cannot be adequately modelled with the HMM variations described thus far.
 To handle these additional fine-scale structures, we recommend replacing $Y^*_t = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize any non-Markovian behaviour.
+
+Replacing $Y^*_t$ with summary statistics results in a loss of information since it involves substituting infinite-dimensional curves with a finite-dimensional representation.
+To minimize prediction error for sequential functional data, we direct the reader to \citet{Aue:2015} and \citet{Gao:2019}, who use functional principle component analysis (FPCA) and derive bounds on the error caused by dimension reduction.
+However, these papers do not account for state-switching behaviour between curves or windows, and do not guarantee that the resulting FPCA scores will be interpretable for practitioners.
+Our goal is not to predict future functional data, but instead to classify and describe coarse- and fine-scale behaviour.
+We are thus primarily interested in the fine-scale summary statistics themselves rather than the curves from which they are calculated, and these summary statistics are known to have consistent maximum likelihood estimates \citep{Douc:2004}.
+As a result, we choose a dimension reduction technique which can effectively separate behavioural states and has an intuitive interpretation for practitioners.
+
 To maintain the temporal structure of the fine-scale process, local summary statistics can be calculated from a moving window with stride length $h$ over the elements of $Y^*_t$.
 Stride length refers to the distance between the first element of consecutive windows, so a stride length of $h$ implies that the first window starts at $Y^*_{t,1}$, the second at $Y^*_{t,1+h}$, and so on.
-Replacing $Y^*_{t,t^*}$ with relevant summary statistics results in a loss of information since we are substituting infinite-dimensional curves with a low-dimensional representation.
-\citet{Aue:2015} and \citet{Gao:2019} use functional principle component analysis (FPCA) for dimension reduction to minimize error when predicting functional data.
-However, these papers do not account for state-switching behaviour between curves and do not guarantee that the resulting FPCA scores will be interpretable for practitioners.
-The goal of this paper is to classify and describe curve types and within-curve behaviour, not predict future functional data.
-We are thus primarily interested in the fine-scale summary statistics themselves rather than the curves from which they are calculated, and these summary statistics are known to have consistent maximum likelihood estimates \cite{Douc:2004}.
-As a result, we choose a dimension reduction technique which can effectively separate behavioural states and has an intuitive interpretation for practitioners.
-In general, subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window.
+Subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window.
 Larger stride lengths result in a larger loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting.
@@
 %
-for $t^* = 1,h+1,2h+1,\ldots$ and $k = 0, 1, \ldots, h-1$ with $i = \sqrt{-1}$.
+for $t^* = 1,h+1,2h+1,\ldots$ and $k = 0, 1, \ldots, h-1$, where $i = \sqrt{-1}$.
 If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise.
@@
 After performing this transformation, the entire model must be redefined since the fine-scale HMM (or CarHMM) directly models the distribution of the summary statistics $\Z_t$ rather than the discretized curve $Y^*_t$.
-Because $\Z_t$ likely exists on a coarser scale than $Y^*_t$, there are only $\tilde{T}_t^*$ unobserved states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$.
+Because $\Z_t$ exists on a coarser scale than $Y^*_t$, there are only $\tilde{T}_t^*$ unobserved states associated with $\Z_t = \big\{\Z_{t,1},\ldots, \Z_{t,\tilde{T}^*_t}\big\}$.
 We denote these unobserved states as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
-The fine-scale probability transition matrices $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ correspond to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$, as do all fine-scale model assumptions (e.g.
-conditional independence, auto-regressive structure between observations, etc.).
-Fine-scale model selection should be done accordingly.
+The fine-scale probability transition matrices $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ correspond to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$, as do all fine-scale model assumptions (e.g., conditional independence, auto-regressive structure between observations, etc.) Fine-scale model selection and validation should therefore be adjusted accordingly.
 %instead of $X^*_t$ and $Y^*_t$.
-The likelihood of this model is therefore identical to that of the original HMM or CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$.
+The likelihood of this model is identical to that of the original HMM or CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$.
 To clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT.
@@
 %
-Possible candidates for the fine-scale model include any of the models described in the previous subsections in addition to many others not described here. 
+%Possible candidates for the fine-scale model include any of the models described in the previous subsections in addition to many others not described here. 
 For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations of a Poisson process as their fine-scale model.

### 2021-06-24 Report/Sections/results.tex
Update Report

@@
 HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., \citet{DeRuiter:2017} and \citet{Isojunno:2017}).
-For killer whales, we are interested in categorizing different diving behaviours to identify potential foraging dives.
+For killer whales, we are interested in categorizing different diving behaviours and identifying potential foraging dives.
 Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller types of salmon \citep{Ford:2009}.
@@
 
-The primary goal of this case study is to jointly estimate the dive types and (more importantly) the subdive behaviours of this killer whale, so we only consider models with a fine-scale component to describe the kinematic data.
+The primary goal of this case study is to jointly estimate the dive types and subdive behaviours of this killer whale, so we only consider models with both a coarse-scale and fine-scale component to describe the kinematic data.
 Defining a suitable hierarchical model involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse- and fine-scale observations.
@@
 To specify the emission distribution of $\Zone_{t,\tilde t^*}$, consider the sequence $\{\Zone_{t,1},\ldots,\Zone_{t,T^*_t}\}$ for a particular dive $t$.
-We assume that each of the three components of this sequence are Normally distributed as in Equation (\ref{eqn:carhmm}) and all components are independent of one another when conditioned on the subdive states $\tilde X_{t,1},\ldots,\tilde X_{t,\tilde T^*_t}$.
+We assume that each of the three components of this sequence are Normally distributed as in Equation (\ref{eqn:carhmm}) and all components are independent of one another when conditioned on the subdive states $\tilde X^*_t = \{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T^*_t}\}$.
 Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter.
@@
 we keep the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines (see Section 3.4 of \citet{Zucchini:2016} for details).
-Plots of the likelihood surface near the maximum likelihood parameter estimates are shown in Section 2 of Supplement A.
+Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory.
+Most, but not all, initializations converge to the same parameter estimates, highlighting the need to perform multiple optimizations with a variety of initial guesses.
+Plots of the likelihood surface near the maximum likelihood estimates are shown in Section 2 of supplement A.
 
@@
 Finally, the mean of $\Ztwo_{t,\tilde t^*}$ and variance of $\Zone_{t,\tilde t^*}$ in subdive state 3 are both much higher than in the other two states, and the auto-correlation of $\Zone_{t,\tilde t^*}$ is also much lower.
-This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).
+This corresponds to vigorous swimming activity, especially as the killer whale ends a dive (see Figure \ref{fig:labeled_dives}).
 
@@
 Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}.
-Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 4 of of supplement A also shows the probability of each dive type and subdive state given the data and the fitted model.
+Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 4 of supplement A also shows the probability of each dive type and subdive state given the data and the fitted model.
 
@@
 The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of of supplement A).
-This suggests that including auto-correlation in the model significantly affects parameter estimates.
-%Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of of supplement A).
+This suggests %that including auto-correlation in the model significantly affects parameter estimates, %Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of of supplement A).
 These findings suggest
-and that the HHMM-DFT is a significantly worse fit to these data than the full CarHHMM-DFT.
+that the HHMM-DFT overlooks some auto-correlation in the data, and that the HHMM-DFT may be overconfident in its parameters estimates compared to the full CarHHMM-DFT.
 
@@
 
-For a more complete set of results for each of the candidate models, see Supplement A.
+For a more complete set of results for each of the candidate models, see supplement A.

### 2021-06-24 Report/Sections/simulation.tex
Update Report

@@
     %
-    \mu_W^{*(\cdot,2)} = 490.06, \quad \sigma_W^{*(\cdot,2)} = 502.56.
+    \mu_W^{*(\cdot,2)} = 490.06, \quad \sigma_W^{*(\cdot,2)} = 502.56, \\
     %
@@
 The two simulated dive types differ in that dives of type 1 are much shorter on average (27 seconds) than dives of type 2 (128 seconds).
-The two simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 3 than for subdive state 2, which in turn is much higher than for subdive state 1.
+The three simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 3 than for subdive state 2, which in turn is much higher than for subdive state 1.
 These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data.
 
-We calculate maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using a similar procedure as the one from the case study.
-Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory per data set.
+We calculate maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using a similar optimization procedure as the one from the case study.
+%Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory per data set.
 
@@
 These estimates are found using the {\em{forward-backward}} algorithm \citep{Zucchini:2016}.
-We compare these estimated conditional probabilities to $x_1,\ldots,x_{100}$, the true coarse-scale state realizations in the test data, by calculating the {\em{average dive decoding accuracy}} for a single training/test data set pair, $\sum_{t=1}^{100} \hat{p}_t(x_t|y,\z)/100$.
+We compare these estimated conditional probabilities to $\{x_1,\ldots,x_{100}\}$, the true coarse-scale state realizations in the test data, by calculating the {\em{average dive decoding accuracy}} for a single training/test data set pair, $\sum_{t=1}^{100} \hat{p}_t(x_t|y,\z)/100$.
 We then report the average of these over the 500 training/test data set pairs.
 Analogously, to assess prediction of the fine-scale states, we estimate $p^*_{t,\tilde t^*}(i^*|y,\z) \equiv \Pr(\tilde X^*_{t,\tilde t^*}=i|Y=y,\Z=\z)$, $i^*=1,2,3$, $\tilde t^* = 1,\ldots,\tilde T^*_t$, $t=1,\ldots,100$, using the test-set observations, the training-set maximum likelihood estimates, and the forward-backward algorithm.
-Denoting the true fine-scale state realizations from the test data set as $\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}$, we define the overall \textit{average subdive decoding accuracy} as the average value of $\hat{p}^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*}|y,\z)$ across all simulated test data sets, dives, and windows.
+Denoting the true fine-scale state realizations from the test data set as $\{\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}\}$, we define the overall \textit{average subdive decoding accuracy} as the average value of $\hat{p}^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*}|y,\z)$ across all simulated test data sets, dives, and windows.
 The conditional probabilities are estimated according to each of the four models under study using the maximum likelihood estimates from the training data set in conjunction with the forward-backward algorithm.
@@
 The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors.
-This underestimation is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to four times as large as the estimated standard error.
+This underestimation is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to five times as large as the estimated standard error.
 See Tables 2 through 6 of supplement B for detailed results.
@@
 Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$ and slightly overestimates $\sigma_A^{*(\cdot,3)}$.
-In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \mu_A^{*(\cdot,23)}$, $\hat \sigma_A^{*(\cdot,1)}$, $\hat \sigma_A^{*(\cdot,2)}$ and $\hat \sigma_A^{*(\cdot,3)}$ are much smaller than the associated empirical standard errors (see Table 3 of supplement B).
+In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \mu_A^{*(\cdot,3)}$, $\hat \sigma_A^{*(\cdot,1)}$, $\hat \sigma_A^{*(\cdot,2)}$ and $\hat \sigma_A^{*(\cdot,3)}$ are much smaller than the associated empirical standard errors (see Table 3 of supplement B).
 These results suggest that estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored.
@@
 
-The CarHHMM is the worst-performing model in terms of dive decoding accuracy, which is approximately 4 percentage points worse than the CarHHMM-DFT ($0.92$).
+The CarHHMM is the worst-performing model in terms of dive decoding accuracy and is approximately 4 percentage points worse than the CarHHMM-DFT ($0.92$).
 Its average subdive decoding accuracy is approximately 6 percentage points worse than the CarHHMM-DFT ($0.85$).
@@
 In addition to its relatively poor average decoding accuracy, the CarHHMM is also the worst of the four candidate models at estimating emission parameters and probability transition matrices.
-Estimates associated with subdive states 2 and 3 ($\theta^{*(\cdot,2)}, , \hat \mu_A^{*(\cdot,3)}$) are especially poor.
+Estimates associated with subdive states 2 and 3 ($\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}$) are especially poor.
 See supplement B for more detailed results.
@@
 In addition, the time required to fit the CarHMM-DFT is less than a third of that of the other models (see Table \ref{table:accuracy}).
-However, this model cannot estimate dive type as it lacks any hierarchical structure.
+However, this model cannot differentiate between dive types as it assumes that there is only one.
 The CarHMM-DFT nonetheless fits a (misspecified) single Gamma distribution over the dive duration of all dives.
-The resulting parameter estimates ($\hat \mu$ and $\hat \sigma$) are highly correlated (See Figure 5 of supplement B).
+The resulting parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$) are highly correlated (See Figure 5 of supplement B).
 

### 2021-06-24 Report/Supp_Sim.tex
Update Simulation Study in Report
Update Report

@@
 % Title, authors, affiliations
-\title{Modelling state-switching functional data with hidden Markov models: supplementary material (simulation study)}%\query{Q1}
+\title{Modelling state-switching functional data with hidden Markov models: supplement B (simulation study)}%\query{Q1}
 \date{}
@@
 These estimates are associated with the CarHHMM-DFT and correspond to both dive types.
-The red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
+The white star represents the mean of all maximum likelihood estimates and the red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
         \addtocounter{fignum}{1}
@@
 These estimates are associated with the HHMM-DFT and correspond to both dive types.
-The red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
+The white star represents the mean of all maximum likelihood estimates and the red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
         \addtocounter{fignum}{1}
@@
 These estimates are associated with the CarHHMM and correspond to both dive types.
-The red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
+The white star represents the mean of all maximum likelihood estimates and the red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
         \addtocounter{fignum}{1}
@@
         \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of acceleration parameter estimates ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
+The white star represents the mean of all maximum likelihood estimates and the red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
 These estimates are for the CarHHMM-DFT.
@@
         \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of acceleration parameter estimates ($\hat \mu^*_A$ and $\hat \sigma^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$ and $\sigma^*_A$ for both subdive states.
+The white star represents the mean of all maximum likelihood estimates and the red star represents the true values of $\mu^*_A$ and $\sigma^*_A$ for both subdive states.
 These estimates are for the HHMM-DFT.
@@
         \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of acceleration parameter estimates ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
+The white star represents the mean of all maximum likelihood estimates and the red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
 These estimates are for the CarHHMM.
@@
         \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of acceleration parameter estimates ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
+The white star represents the mean of all maximum likelihood estimates and the red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
 These estimates are for the CarHMM-DFT.
@@
         \def\arraystretch{1.5}
+
 \begin{tabular}{ccccccc}
@@
                             &                               & 2                                & $505.10$                         & $-0.69$ $(p=0.394)$          & $17.94$                             & $9.74 \pm 0.81$                             \\
-                            &                               & 3                                & $9787.1$                         & $17.2$ $(p=0.718)$           & $1060.5$                            & $262.1 \pm 97.120$                             \\
-                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.849$                         & $-0.001$ $(p=0.991)$         & $0.989$                             & $0.658 \pm 0.088$                             \\
-                            &                               & 2                                & $514.21$                         & $-2.47$ $(p=0.010)$          & $21.43$                             & $9.74 \pm 0.81$                             \\
-                            &                               & 3                                & $14430$                          & $-33$ $(p=0.647)$            & $1596$                              & $262 \pm 97$                             \\ \hline
+                            &                               & 3                                & $9787.1$                         & $17.2$ $(p=0.718)$          & $1060.5$                             & $262.1 \pm 97.1$                             \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.849$                         & $-0.001$ $(p=0.991)$          & $0.989$                             & $0.652 \pm 0.094$                             \\
+                            &                               & 2                                & $514.21$                         & $-2.47$ $(p=0.010)$          & $21.43$                             & $11.41 \pm 0.95$                             \\
+                            &                               & 3                                & $14430$                         & $-33$ $(p=0.647)$          & $1596$                             & $303 \pm 166$                             \\ \hline
 \multirow{6}{*}{HHMM-DFT}   & \multirow{3}{*}{$\mu_W^*$}    & 1                                & $35.067$                         & $0.087$ $(p=0.097)$          & $1.170$                             & $0.679 \pm 0.100$                             \\
@@
                             &                               & 3                                & $6841.0$                         & $-2929.0$ $(p=0.000)$          & $1164.3$                             & $198.2 \pm 77.1$                             \\
-                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.800$                         & $-0.050$ $(p=0.351)$         & $1.192$                             & $0.679 \pm 0.100$                             \\
-                            &                               & 2                                & $513.26$                         & $-3.42$ $(p=0.001)$          & $22.60$                             & $10.75 \pm 1.21$                             \\
-                            &                               & 3                                & $11459$                          & $-3004$ $(p=0.000)$          & $1739$                              & $198 \pm 77$                             \\ \hline
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.800$                         & $-0.050$ $(p=0.351)$          & $1.192$                             & $0.676 \pm 0.104$                             \\
+                            &                               & 2                                & $513.26$                         & $-3.42$ $(p=0.001)$          & $22.60$                             & $12.50 \pm 1.30$                             \\
+                            &                               & 3                                & $11459$                         & $-3004$ $(p=0.000)$          & $1739$                             & $280 \pm 136$                             \\ \hline
 \multirow{6}{*}{CarHHMM}    & \multirow{3}{*}{$\mu_W^*$}    & 1                                & ------                         & ------                     & ------                             & ------                                      \\
@@
                             &                               & 3                                & $9817.8$                         & $47.8$ $(p=0.328)$          & $1089.1$                             & $280.2 \pm 136.3$                             \\
-                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.890$                         & $0.040$ $(p=0.401)$          & $1.058$                             & $0.661 \pm 0.093$                             \\
-                            &                               & 2                                & $515.23$                         & $-1.45$ $(p=0.125)$          & $21.09$                             & $9.64 \pm 0.92$                             \\
-                            &                               & 3                                & $14438$                          & $-25$ $(p=0.740)$            & $1661$                              & $280 \pm 136$                             \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.890$                         & $0.040$ $(p=0.401)$          & $1.058$                             & $0.656 \pm 0.101$                             \\
+                            &                               & 2                                & $515.23$                         & $-1.45$ $(p=0.125)$          & $21.09$                             & $11.37 \pm 0.99$                             \\
+                            &                               & 3                                & $14438$                         & $-25$ $(p=0.740)$          & $1661$                             & $303 \pm 166$                             \\
 \end{tabular}
@@
         \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of wiggliness parameter estimates ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
-The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
+The white star represents the mean of all maximum likelihood estimates and the red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
 These estimates are for the CarHHMM-DFT.
@@
         \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of wiggliness parameter estimates ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
-The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
+The white star represents the mean of all maximum likelihood estimates and the red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
 These estimates are for the HHMM-DFT.
@@
         \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of wiggliness parameter estimates ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
-The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
+The white star represents the mean of all maximum likelihood estimates and the red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
 These estimates are for the CarHMM-DFT.

### 2021-06-24 Report/Reviewer_Response.tex
Update Report- ready for editing by MAM and NH

@@
 \begin{itemize}
-    \item The case study has been re-run and the results have changed very slightly.
+    \item The case study has been re-run with more random initializations and the results have changed slightly.
     \item The simulation study has been re-run with 3 behavioural states instead of 2.
     \item The supplemental material has been broken into two: supplement A contains material related to the case study while supplement B contains material related to the simulation study. 
+    \item We have added line numbers for convenience
 \end{itemize}
@@
 We also assume that the sequence of curve types makes up an unobserved Markov chain, which is what induces the between-curve dependence.
-We have edited page 3, lines 49-53 as well as page 5, lines 83-84 to make this distinction clearer.
+We have edited page 3, lines 49-51 as well as page 5, lines 80-82 to make this distinction clearer.
 In addition, we have changed ``different curve types" to ``a discrete number of curve types" where applicable to highlight that we are using a Markov chain with a discrete number of states to model the between-curve dependence.
@@
 
-        ``Our goal is to identify and describe a finite number of behaviours at multiple scales within sequential functional data.
-We simultaneously aim to model complicated temporal dependence both between and within curves." We have added this excerpt to page 5, lines 81-83 to clarify the goal of this paper.
+        ``Our goal is to identify and describe discrete behavioural states at multiple scales within functional data (e.g., coarse-scale curve types and fine-scale behavioural states) and to model the dependence structure between those states." We have added this excerpt to page 5, lines 80-82 to clarify the goal of this paper.
 Using our case study as an example, we simultaneously identify and describe two different dive types and three different swimming behaviours.
@@
         \item \textit{If the dependence between curves is purely based on the hidden variables, then there should be more justifications in 2.4 as curves are transformed to lower-dimensional variables.
-For functional time series, please see "On the prediction of stationary functional time series", "High-dimensional functional time series forecasting: An application to age-specific mortality rates".}
+For functional time series, please see ``On the prediction of stationary functional time series", ``High-dimensional functional time series forecasting: An application to age-specific mortality rates".}
 
@@
 However, our goal is not to reconstruct the curves, but rather to identify the coarse-scale and fine-scale curve types as well as describe the dependence between those types.
-We added an explanation of this on pages 15-16, lines 281-294.
+We added an explanation of this on pages 15, lines 268-281.
 Nonetheless, bounding prediction and reconstruction error is a promising area of future study.
 In particular, it would be useful to prove an analogue of Theorem 3.2 in "On the prediction of stationary functional time series" for state-switching functional time series.
-We added a sentence about this to the discussion on page 40, lines 754-757.
+We added a sentence about this to the discussion on page 40, lines 743-746.
 
@@
 
-    This is an excellent point- fPCA is a better way to reduce reconstruction error, and we added a discussion of this on pages 15-16, lines 281-294.
+    This is an excellent point- fPCA is a better way to reduce reconstruction error, and we added a discussion of this on pages 15, lines 268-281.
 However, our goal is to provide a means of summarizing these curves in a way that is interpretable for practitioners, not to reconstruct the curves.
@@
         If $x_t = i$, then $x_{t,t*}^*$ represents one of $N^{*(i)}$ possible within-curve hidden behaviours associated with observation $y_{t,t*}^*$ and dive type $i$.
-We have added a sentence on page 12, lines 225-227 to make this more clear.
+We have added a sentence on page 12, lines 224-225 to make this more clear.
 For example, in the case study we have that there are 3 possible subdive states.
 If the whale is in subdive state 2 for dive t and window $\tilde t^*$, then $\tilde x^*_{t,\tilde t^*} = 2$.
-In addition, we added verbiage on page 21, lines 405-411 to make clear what $x$ and $x^*$ are in the context of the case study.
+In addition, we added verbiage on page 21, lines 394-400 to make clear what $x$ and $x^*$ are in the context of the case study.
 
@@
 If anything, we are personally more interested in the fine-scale behaviours.
-To make this more clear to the reader, we have added a sentence on page 22, lines 413-418.
+To make this more clear to the reader, we have added a sentence on page 22, lines 402-404.
 
@@
 
-        We have added some verbiage to the caption of Figure 1(d) to indicate that we are omitting the coarse-scale process and are only looking at the HMM-DFT as a fine-scale model of the within-curve behaviour of curve $t$.
+        We have added some verbiage to the caption of figure 1 (d) to indicate that we are omitting the coarse-scale process and are only looking at the HMM-DFT as a fine-scale model of the within-curve behaviour of curve $t$.
+In figure 4, note that the fine-scale emission distributions are shared across all dive types.
+We have included an explanation of which emission distributions go with dive types and which emission distributions go with subdive states in the caption of figure 4.
 
@@
 The likelihood function does not include either $x_t$ or $x^*_t$ because both hidden variables can be marginalized out of the likelihood.
-We have made this more clear by adding verbiage to pages 8-9, lines 163-172 as well as page 14, lines 253-255.
+We have made this more clear by adding verbiage to pages 8-9, lines 161-170 as well as page 14, lines 245-247.
 
@@
     This is an interesting approach that has been looked at in a couple of papers.
-This is beyond the scope of our paper, but we have added references to several sampling methods for HMMs (and relevant citations) on page 9, lines 165-170.
+This is beyond the scope of our paper, but we have added references to several sampling methods for HMMs (and relevant citations) on page 9, lines 164-168.
 
@@
 For example, Figure 1 of Supplement A shows lag plots for the transformed fine-scale observations, and there appears to be strong auto-correlation.
-We have added several sentences on page 17, lines 315-324 to clarify this.
+We have added several sentences on page 17, lines 305-314 to clarify this.
 
@@
 We found that most, but not all, of the optimization routines converged to similar parameter estimates.
-We have added an explanation of our optimization routine on pages 26-17, lines 502-510.
+We have added an explanation of our optimization routine on page 26, lines 491-500.
 
@@
 
-        We thank the reviewer for this suggestion, as plotting the likelihood gave us evidence that our optimization routine did in fact converge.
-We added plots of the likelihood surface associated with the CarHHMM-DFT in supplement A.
+        We have added plots of the likelihood surface associated with the CarHHMM-DFT in supplement A.
 Unfortunately the parameter space is high-dimensional, so we plotted the likelihood of the parameters associated with each feature / hidden state pair while keeping all other parameters fixed at their maximum likelihood estimates.
@@
 
-    The introduction primarily uses the adjectives "between-curve" and "within-curve" to focus on FDA, while the subsequent sections switch to "coarse-scale" and "fine-scale" to be more general.
-The connection between these adjectives was not clear before, so we have added a sentence to page 2, lines 34-36 to introduce "coarse-scale" vs "fine-scale" earlier.
-We have also added a sentence to page 5, lines 81-85 to make the connection between coarse-scale / fine-scale and curve types / behavioural states.
+    The introduction primarily uses the adjectives ``between-curve" and ``within-curve" to focus on FDA, while the subsequent sections switch to ``coarse-scale" and ``fine-scale" to be more general.
+The connection between these adjectives was not clear before, so we have added a sentence to page 2, lines 34-36 to introduce ``coarse-scale" vs ``fine-scale" earlier.
+We have also added a sentence to page 5, lines 80-82 to make the connection between coarse-scale / fine-scale and curve types / behavioural states.
 
@@
 In addition, the curves are assumed to be sequential in time, and there is a connection between the coarse scale index $t$ and the fine-scale index $t^*$.
-We have added a sentence on page 7, lines 138-139 to clarify that these curves are sequential but not necessarily equi-spaced in time.
+We have added a sentence on page 7, lines 137-138 to clarify that these curves are sequential but not necessarily equi-spaced in time.
 In addition, there is some precedence for indexing sequential curves with $t$- see "Cam\'er-Karhunen-Lo\`eve representation and harmonic principal component analysis of functional time series" by Panaretos and Tavakoli and "Dynamic functional principal components" by Hormann and Kidinski for examples.
@@
 
-    We agree and have added a sentence to clarify this on page 8, lines 152-153.
+    We agree and have added a sentence to clarify this on page 8, lines 151-152.
 
@@
 
-    We have edited page 9, lines 172-177 to clarify the difference between the likelihood and the forward algorithm.
-We have also described the forward algorithm, which is used to evaluate the likelihood, as well as the forward-backward algorithm, which is used for state decoding and maximizing the likelihood via the Baum-Welch algorithm.
-See Zucchini et al (2016) for details.
+    We have edited page 9, lines 171-176 to clarify the difference between the likelihood and the forward algorithm.
+We have also described the forward algorithm, which is used to evaluate the likelihood, as well as the forward-backward algorithm, which is used for state decoding and maximizing the likelihood via the Baum-Welch algorithm (Zucchini et al, 2016).
 
@@
     For each model, we ran 100 direct maximization routines using random initializations using the Nelder-Mead method of the optimize package of spicy.
-In particular, we optimized one row at a time of the probability transition matrices and one set of parameters associated with a (feature,hidden state) pair at a time.
+In particular, we optimized one row at a time of the probability transition matrices and one set of parameters associated with a (feature, hidden state) pair at a time.
 We found that most, but not all, of the optimization routines converged to similar parameter estimates.
-We have added an explanation of our optimization routine on pages 26-17, lines 502-510.
+We have added an explanation of our optimization routine on page 26, lines 491-500.
 
@@
 
-    This is a very useful and common practice in the ecology literature. While it is beyond the scope of our case study, we have added a brief discussion of covariates on page 10, lines 183-186.
+    This is a very useful and common practice in the ecology literature. While it is beyond the scope of our case study, we have added a brief discussion of covariates on page 10, lines 182-185.
 

### 2021-07-06 Code/HHMM.py
Update Gamma Plots

@@ -189,6 +189,9 @@ class HHMM:
 
     def reorder_params(self):
 
+        if self.SEs is None:
+            self.get_SEs(self.data,0.01)
+
         if self.CM is None:
             self.CM = [np.zeros((self.pars.K[0],self.pars.K[0])),
                        [np.zeros((self.pars.K[1],self.pars.K[1])) for _ in range(self.pars.K[0])]]
@@ -723,7 +726,7 @@ class HHMM:
                         res = minimize(loss_fn, x0, method='Nelder-Mead', options=options)
                         etime = time.time()
 
-                        print('optimized fine theta %s, dive_state %d subdive state %d'\
+                        print('optimized fine theta %s, dive type %d subdive state %d'\
                               % (feature,k0,k1))
                         print(res)
                         print('original: ', x0)
@@ -996,6 +999,8 @@ class HHMM:
 
         SEs = {}
 
+        orig_theta = deepcopy(self.theta)
+
         # coarse-scale theta
         for feature in self.theta[0]:
             print('')
@@ -1030,10 +1035,7 @@ class HHMM:
                     th_tm1 = self.likelihood(data)
 
                     # return theta
-                    if param == 'corr':
-                        self.theta[0][feature][param][state_num] = corr_2_rho(theta)
-                    else:
-                        self.theta[0][feature][param][state_num] += h
+                    self.theta = deepcopy(orig_theta)
 
                     # get estimate
                     if param =='corr':
@@ -1106,18 +1108,7 @@ class HHMM:
                     print(th_tp1)
 
                     # return theta
-                    if param == 'corr':
-                        for k0 in range(self.pars.K[0]):
-                            if feature in ['Ax','Ay','Az']:
-                                self.theta[1][k0]['Ax'][param][state_num] = corr_2_rho(theta)
-                                self.theta[1][k0]['Ay'][param][state_num] = corr_2_rho(theta)
-                                self.theta[1][k0]['Az'][param][state_num] = corr_2_rho(theta)
-                            else:
-                                self.theta[1][k0][feature][param][state_num] = corr_2_rho(theta)
-                    else:
-                        self.theta[1][0][feature][param][state_num] += h
-                        for i in range(1,self.pars.K[0]):
-                            self.theta[1][i][feature][param][state_num] = self.theta[1][0][feature][param][state_num]
+                    self.theta = deepcopy(orig_theta)
 
                     # get estimate
                     if param =='corr':
[…]

### 2021-07-06 Code/Parameters.py

@@ -60,7 +60,7 @@ class Parameters:
 
         # parameters for HHMM features
         self.timescales = [(10,'min'),(2,'sec')]
-        self.K = [2,2] # number of states per level
+        self.K = [2,3] # number of states per level
         self.share_fine_states = True # are fine states same for all crude states?
 
         self.features = [{'max_depth':{'corr':False,'f':'gamma'},

### 2021-07-06 Code/Train_Model.py

@@ -35,15 +35,23 @@ np.random.seed(rand_seed)
 # set parameters
 pars = Parameters.Parameters()
 
-pars.cvc_file = '../Data/2019/20190902-182840-CATs_OB_1.cvc'
-pars.csv_file = '../Data/2019/20190902-182840-CATs_OB_1_001.csv'
+#pars.cvc_file = '../Data/2019/20190902-182840-CATs_OB_1.cvc'
+#pars.csv_file = '../Data/2019/20190902-182840-CATs_OB_1_001.csv'
+
+# parameters for 20200830-180909-Gikumi_CC_136
+pars.cvc_file = '../Data/2020/20200830-180909-Gikumi_CC_136/20200830-180909-Gikumi_CC_136.cvc'
+pars.csv_file = '../Data/2020/20200830-180909-Gikumi_CC_136/20200830-180909-Gikumi_CC_136.csv'
+pars.stime = '2020-08-30 10:50:00'
+pars.etime = '2020-08-30 16:00:00'
+pars.drop_times = []
 
 pars.features = [{'dive_duration':{'corr':False,'f':'gamma'}},
                  {'Ahat_low':{'thresh':5,'corr':False,'f':'gamma'},
                   'Ax':{'corr':True,'f':'normal'},
                   'Ay':{'corr':True,'f':'normal'},
                   'Az':{'corr':True,'f':'normal'}}]
-pars.K = [2,3]
+
+pars.K = [3,3]
 pars.share_fine_states = True
 
 if model == 'CarHMM':
@@ -83,19 +91,21 @@ data = prep.get_all_features(df,dive_df)
 # train_model
 print('training model')
 hhmm = HHMM.HHMM(pars,data)
-hhmm.train_DM(data,max_iters=25,max_steps=25,max_time=90)
+hhmm.train_DM(data,max_iters=25,max_steps=25,max_time=48)
+hhmm.save(HHMM_file)
 
 # get SEs
 h = 0.01
 print('finding standard errors')
 hhmm.get_SEs(data,h)
+hhmm.save(HHMM_file)
 
 # label data
-print('labelling df')
-data,df = hhmm.label_df(data,df)
+#print('labelling df')
+#data,df = hhmm.label_df(data,df)
 
 # save everything
-hhmm.data = None
-hhmm.save(HHMM_file)
+#hhmm.data = None
+#hhmm.save(HHMM_file)
 #with open(data_outfile, 'wb') as f:
 #    pickle.dump(data, f)

### 2021-07-06 Code/Train_Model.sh

@@ -8,14 +8,14 @@
 #SBATCH --mail-type=REQUEUE
 #SBATCH --mail-type=ALL
 
-#SBATCH --time=96:00:00
+#SBATCH --time=60:00:00
 #SBATCH --mem-per-cpu=8G
-#SBATCH --array=1-400
+#SBATCH --array=1-100
 #SBATCH --cpus-per-task=1
 
 if [ $SLURM_ARRAY_TASK_ID -lt 101 ]
 then
-  python Train_Model.py $SLURM_ARRAY_TASK_ID CarHMM
+  python Train_Model.py $SLURM_ARRAY_TASK_ID CarHHMM2
 fi
 if [ $SLURM_ARRAY_TASK_ID -gt 100 ] && [ $SLURM_ARRAY_TASK_ID -lt 201 ]
 then
@@ -27,5 +27,5 @@ then
 fi
 if [ $SLURM_ARRAY_TASK_ID -gt 300 ]
 then
-  python Train_Model.py $SLURM_ARRAY_TASK_ID CarHHMM2
+  python Train_Model.py $SLURM_ARRAY_TASK_ID CarHMM
 fi

### 2021-07-06 Code/Visualisor.py

@@ -161,7 +161,7 @@ class Visualisor:
 
                 dist = self.pars.features[0][feature]['f']
                 K = self.pars.K[0]
-                colors = [cm.get_cmap('tab10')(i) for i in [0,1]]
+                colors = [cm.get_cmap('tab10')(i) for i in range(K)]
                 legend = ['Dive Type %d'%(x+1) for x in range(K)]
             else:
                 mu = self.hhmm.theta[1][0][feature]['mu']
@@ -169,7 +169,7 @@ class Visualisor:
 
                 dist = self.pars.features[1][feature]['f']
                 K = self.pars.K[1]
-                colors = [cm.get_cmap('viridis')(i) for i in [0.,0.5,1.]]
+                colors = [cm.get_cmap('viridis')(i/(K-1.0)) for i in range(K)]
                 legend = ['Subdive Behavior %d'%(x+1) for x in range(K)]
 
             for state in range(K):
@@ -323,7 +323,7 @@ class Visualisor:
             if 'subdive_state_' not in col:
 
                 plt.subplot(nrows,1,fignum)
-                colors = [cm.get_cmap('tab10')(i) for i in [0,1]]
+                colors = [cm.get_cmap('tab10')(i) for i in range(self.pars.K[0])]
                 legend = ['Dive Type %d' % (i+1) for i in range(self.pars.K[0])]
                 for state,dive_df in enumerate(dives):
                     plt.plot(dive_df['sec_from_start']/60,dive_df[col],
@@ -346,16 +346,17 @@ class Visualisor:
                     plt.gca().invert_yaxis()
 
                 # plot vlines
-                for dive_num in range(sdive+1,edive+1):
-                    vline = min(dive[dive['dive_num'] == dive_num]['sec_from_start']/60)
-                    plt.axvline(vline,color='k',linewidth=2)
-
+                #for dive_num in range(sdive+1,edive+1):
+                    #vline = min(dive[dive['dive_num'] == dive_num]['sec_from_start']/60)
+                    #plt.axvline(vline,color='k',linewidth=2)
+                if col == 'Ax':
+                    plt.ylim([-2,2])
                 fignum += 1#2
 
             # subdive-level coloring
             if 'dive_state_' not in col or 'subdive_state' in col:
                 plt.subplot(nrows,1,fignum)
-                colors = [cm.get_cmap('viridis')(i) for i in [0.,0.5,1.]]
[…]

### 2021-07-07 Report/Sections/results.tex
Updates from Overleaf

@@
 %To illustrate an application of this method and compare the candidate models, 
-We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the \textit{scipy} package in Python \citep{Virtanen:2019}.
+We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the \textit{SciPy} package in Python \citep{Virtanen:2019}.
 Each model is fit 100 times using random initializations, and
@@
 If the observation can be any real number, then we set $\mu_0 \sim \mathcal{N}(\bar Y,s)$, $\log(\sigma) \sim \mathcal{N}(s,1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean and $s$ represents the sample standard deviation of the observations values.
-we keep the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines (see Section 3.4 of \citet{Zucchini:2016} for details).
+we keep the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines \citep{Zucchini:2016}.
 Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory.
@@
 
-The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:coarse_emis}). 
+The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:emis}). 
 Dive type 1 corresponds to shorter, shallower dives which likely reflect resting, travelling, and to a lesser extent searching for prey.
@@
 No dive in this data set has a maximum depth deeper than 30 meters and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters.
-However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and researchers have observed significant numbers of prey-capture events near the surface \citep{Fortune:2020}.
+However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and %researchers
+unpublished data collected by the authors suggest that %significant numbers of 
+prey-capture events can occur near the surface. %\citep{Fortune:2020}. 
 Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.
 
-The means of ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ associated with each subdive state are separated by an order of magnitude (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:fine_emis}).
+The means of ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ associated with each subdive state are separated by an order of magnitude (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:emis}).
 Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest standard deviation corresponding to $\Zone_{t,\tilde t^*}$.
@@
 Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions.
-In histograms corresponding to subdive state 3, $\Ztwo_{t,\tilde t^*}$ is right-skewed and $\Zone_{t,\tilde t^*}$ has heavier tails compared to a normal distribution, indicating the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale.
+In particular, histograms corresponding to subdive state 3 show that $\Zone_{t,\tilde t^*}$ has heavier tails compared to a normal distribution.
+This indicates the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale.
 These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

### 2021-07-07 Report/Sections/simulation.tex
Updates from Overleaf

@@
 After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored.
-Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \big\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \big\}$ with probability transition matrices
+Each two-second segment is assigned one of $N^*=3$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \big\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \big\}$ with probability transition matrices
 
@@
 
-The HHMM-DFT has a average dive decoding accuracy is comparable to the CarHHMM-DFT ($0.96$), but its average subdive decoding accuracy is worse by approximately 7 percentage points ($0.84$).
+The HHMM-DFT has an average dive decoding accuracy is comparable to the CarHHMM-DFT ($0.96$), but its average subdive decoding accuracy is worse by approximately 7 percentage points ($0.84$).
 Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$ and slightly overestimates $\sigma_A^{*(\cdot,3)}$.

### 2021-07-07 Report/Sections/intro.tex

@@
 However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
-For example, dive profiles of marine animals show a discrete number of distinct dive ``types"
+On a coarse scale, dive profiles of marine animals show a discrete number of distinct dive ``types"
 %that aggregate together in time 
 where the sequence of dive types exhibits state-switching behaviour \citep{Tennessen:2019b}.
-These profiles simultaneously display bouts of short-term periodicity within each dive \citep{Adam:2019}.
+On a fine scale, these profiles display bouts of short-term periodicity within each dive \citep{Adam:2019}.
 Fine-scale periodicity nested within a coarser state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
@@
 In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, and \citet{DeSouza:2014,DeSouza:2017} use HMMs to model state-switching behaviour within functional data.
-However, none of these papers account for certain types of temporal correlation taking place on multiple scales.
+However, none of these papers account for temporal correlation taking place on multiple scales.
 
@@
 \citet{Heerah:2017} successfully use Fourier analysis within an HMM to account for daily behavioural cycles in marine mammals, and Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
-Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for fine-scale periodic structures.
+Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for fine-scale periodic structures in biologging data.
 
@@
 The resulting suite of methods makes up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.
-We begin in Section 2 by describing HMMs as well as two variants, CarHMMs and HHMMs, and discuss how Fourier analysis can handle fine-scale dependence structures.
+We begin in Section 2 by describing HMMs as well as two variants, CarHMMs and HHMMs, and discuss how summary statistics over moving windows can handle fine-scale dependence structures.
 We also show how these methods can be combined to analyze increasingly complex data.

### 2021-07-07 Report/Sections/models.tex
Update Report- ready for editing by MAM and NH

@@
 We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.
-We assume that $\{Y_1,\ldots,Y_T\}$ is ordered sequentially in time, but the curves need not be equally spaced.
+We assume that the curves $Y_1,\ldots,Y_T$ are indexed according to the temporal order in which they are observed, but the $T$ observations need not be equally spaced in time.
 To develop our model for this data, we detail the structure of a traditional HMM followed by three variations which generalize its base structure.
@@
 
-\[\calL_{\text{HMM}}(x,\theta,\Gamma;y) = \delta_{x_1} f^{(x_1)}(y_1; \theta^{(x_1)}) \prod_{t=2}^T \Gamma_{x_{t-1} x_t} f^{(x_t)}(y_t; \theta^{(x_t)}),\]
+\begin{equation}
+    \calL_{\text{HMM}}(x,\theta,\Gamma;y) = \delta_{x_1} f^{(x_1)}(y_1; \theta^{(x_1)}) \prod_{t=2}^T \Gamma_{x_{t-1} x_t} f^{(x_t)}(y_t; \theta^{(x_t)}),
+    \label{eqn:like}
+\end{equation}
 %
-which can be used to obtain estimates for both the hidden states $\hat X$ and the parameters $\hat \theta$ and $\hat \Gamma$.
-Using a Bayesian approach, $\calL_{\text{HMM}}$ can be combined with a prior distribution to obtain a posterior distribution over $\hat X$, $\hat \theta$, and $\hat \Gamma$.
-This posterior distribution can be sampled from using methods such as sequential Monte Carlo \citep{Douc:2011b}, Markov-chain Monte Carlo \citep{Scott:2002}, or variational inference \citep{Foti:2014}.
-For the purposes of this paper, however, we focus on the frequentest approach, where it is common to marginalize out $x$ from the likelihood:
+but it is also computationally tractable to sum over all possible values of $x$ to obtain the following:
 
-\[\calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N,\]
+\begin{equation}
+    \calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N,
+    \label{eqn:like_marginal}
+\end{equation}
 %
@@
 This equation can be evaluated in $\mathcal{O}(T)$ time using the well-known \textit{forward algorithm} \citep{Zucchini:2016}.
-After obtaining maximum likelihood estimates $\{\hat \theta, \hat \Gamma\}$, estimated probabilities of each hidden state $\big(\hat{\Pr}(X_t = i|Y = y)\big)$ can be calculated from $\{\hat \theta,\hat \Gamma\}$ using the similarly-named \textit{forward-backward algorithm} \citep{Zucchini:2016}.
+
+We take a frequentest approach in this paper, which involves obtaining maximum likelihood estimates $\{\hat \theta, \hat \Gamma\}$ from Equation (\ref{eqn:like_marginal}).
+The estimated probability of each hidden state $\hat{\Pr}(X_t = i|Y = y)$ can then be calculated using $\{\hat \theta, \hat \Gamma\}$ and the \textit{forward-backward algorithm} \citep{Zucchini:2016}.
+Alternatively, it is reasonable to use a Bayesian approach, where Equation (\ref{eqn:like}) is combined with a prior to obtain a posterior distribution over $X$, $\theta$, and $\Gamma$.
+This posterior distribution can be sampled from using methods such as sequential Monte Carlo \citep{Douc:2011b}, Markov-chain Monte Carlo \citep{Scott:2002}, or variational inference \citep{Foti:2014}.
 %
@@
 For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$.
-Figure \ref{fig:models}a represents the dependence structure of an HMM.
+Figure \ref{fig:models} shows the dependence structure of an HMM.
 
@@
 $P(y_t|y_{t-1};\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
-Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.
+Figure \ref{fig:models} shows a graphical representation of the dependence structure of a CarHMM.
 
@@
 However, in our killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$ and share the fine-scale emission parameters across the $N$ coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.
-As a result, the distribution of the fine-scale observation $Y^*_{t,t^*}$ is defined exclusively by $X^*_{t,t^*}$ (or $(\cdot,X^*_{t,t^*})$ for notational consistency).
+As a result, the distribution of the fine-scale observation $Y^*_{t,t^*}$ does not depend on $X_t$, but instead is defined exclusively by $X^*_{t,t^*}$ (or $(\cdot,X^*_{t,t^*})$ for notational consistency).
 %Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$, not their fine-scale emission parameters $\theta^{*(\cdot,i)}$.
@@
 Due to the nested structure of the HHMM, %its likelihood is easily calculated using the forward algorithm.
-it is straightforward to extend the forward algorithm and marginalize out both the coarse-scale hidden states $X$ and the fine-scale hidden states $X^*$ from the likelihood.
+it is straightforward to extend the forward algorithm and sum over all possible coarse-scale hidden states $X$ and all possible fine-scale hidden states $X^*$ in the likelihood.
 %
@@
 The fine-scale likelihood $\calL_{\text{fine}}$ corresponds to the likelihood of the fine-scale model, which can be either a CarHMM or an HMM.
-Figure \ref{fig:models}c displays the dependence structure of an HHMM.
+Figure \ref{fig:models} displays the dependence structure of an HHMM.
 
@@
 To handle these additional fine-scale structures, we recommend replacing $Y^*_t = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize any non-Markovian behaviour.
-
-Replacing $Y^*_t$ with summary statistics results in a loss of information since it involves substituting infinite-dimensional curves with a finite-dimensional representation.
-To minimize prediction error for sequential functional data, we direct the reader to \citet{Aue:2015} and \citet{Gao:2019}, who use functional principle component analysis (FPCA) and derive bounds on the error caused by dimension reduction.
-However, these papers do not account for state-switching behaviour between curves or windows, and do not guarantee that the resulting FPCA scores will be interpretable for practitioners.
-Our goal is not to predict future functional data, but instead to classify and describe coarse- and fine-scale behaviour.
-We are thus primarily interested in the fine-scale summary statistics themselves rather than the curves from which they are calculated, and these summary statistics are known to have consistent maximum likelihood estimates \citep{Douc:2004}.
-As a result, we choose a dimension reduction technique which can effectively separate behavioural states and has an intuitive interpretation for practitioners.
-
 To maintain the temporal structure of the fine-scale process, local summary statistics can be calculated from a moving window with stride length $h$ over the elements of $Y^*_t$.
@@
 In addition, setting the stride length equal to the window size avoids artificial residual correlation arising from overlapping windows.
-For our particular case study, we use the discrete Fourier transform (DFT) of a moving forward window with width $h$ and stride $h$ across $Y^*_t$:
+
+We are interested in fine-scale summary statistics that are interpretable for practitioners and can effectively differentiate between fine-scale behavioural states.
+As a result, we use the discrete Fourier transform (DFT) of a moving forward window with width $h$ and stride $h$ across $Y^*_t$:
 
@@
 %instead of $X^*_t$ and $Y^*_t$.
+
 The likelihood of this model is identical to that of the original HMM or CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$.
 To clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT.
-Figure \ref{fig:models}d displays the dependence structure of a fine-scale HMM-DFT.
+Figure \ref{fig:models} displays the dependence structure of a fine-scale HMM-DFT.
+
+It is important to note that replacing $Y^*_t$ with summary statistics results in a loss of information since it involves substituting infinite-dimensional curves with a finite-dimensional representation.
+Readers interested in summary statistics that prediction error for sequential functional data can read \citet{Aue:2015} and \citet{Gao:2019}, who both use functional principle component analysis (FPCA) and derive bounds on the error caused by dimension reduction.
+While these papers do not account for state-switching behaviour between curves or windows, deriving similar error bounds for FPCA within an HMM framework appears to be a promising direction for future research.
 

### 2021-07-07 Report/Supp_Case.tex
Update Simulation Study in Report
Update Report

@@
 
-    \section{Likelihood Surfaces - CarHHMM-DFT}
+    \section{Likelihood Surface of the CarHHMM-DFT}
 
@@
         \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the dive duration parameters $\mu_Y^{(i)}$ and $\sigma_Y^{(i)}$ for dive types $i = 1,2$.
-All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
+All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_Y,\hat \sigma_Y)$, which is denoted by a red star.
@@
 Recall that $\Gamma_{i,j} = \exp(\eta_{i,j}) / \sum_k \exp(\eta_{i,k})$.
-All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
+All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \eta_{1,2},\hat \eta_{2,1})$, which is denoted by a red star.
@@
         \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the wiggliness parameters $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
+All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_W^{*(\cdot,i^*)}, \hat \sigma_W^{*(\cdot,i^*)})$, which is denoted by a red star.
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $x$-acceleration parameters $\mu_{A_x}^{*(\cdot,i^*)}, \sigma_{A_x}^{*(\cdot,i^*)}$ and $\text{logit}\left(\phi_{A}^{*(\cdot,i^*)}\right)$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_x}^{*(\cdot,i^*)}, \hat \sigma_{A_x}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $x$-acceleration parameters $\mu_{A_x}^{*(\cdot,i^*)}, \sigma_{A_x}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_x}^{*(\cdot,i^*)}, \hat \sigma_{A_x}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $y$-acceleration parameters $\mu_{A_y}^{*(\cdot,i^*)}, \sigma_{A_y}^{*(\cdot,i^*)}$ and $\text{logit}\left(\phi_{A}^{*(\cdot,i^*)}\right)$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_y}^{*(\cdot,i^*)}, \hat \sigma_{A_y}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $y$-acceleration parameters $\mu_{A_y}^{*(\cdot,i^*)}, \sigma_{A_y}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_y}^{*(\cdot,i^*)}, \hat \sigma_{A_y}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $z$-acceleration parameters $\mu_{A_z}^{*(\cdot,i^*)}, \sigma_{A_z}^{*(\cdot,i^*)}$ and $\text{logit}\left(\phi_{A}^{*(\cdot,i^*)}\right)$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\left(\hat \mu_{A_z}^{*(\cdot,i^*)}, \hat \sigma_{A_z}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)}\right)$, which is denoted by a red star.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $z$-acceleration parameters $\mu_{A_z}^{*(\cdot,i^*)}, \sigma_{A_z}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_z}^{*(\cdot,i^*)}, \hat \sigma_{A_z}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
 Recall that $\eta^{*(i)}_{i^*,i^*} = 0$ and that $\Gamma^{*(i)}_{i^*,j^*} = \exp(\eta^{*(i)}_{i^*,j^*}) / \sum_{k^*} \exp(\eta^{*(i)}_{i^*,k^*})$.
-All other parameters and probability transition matrices are assumed to be the maximum likelihood estimates shown in Table (1) of Section 3.1.
+All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\hat \Gamma^{*(i)}$, which is denoted by a red star.
@@
 
-        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $\left(y^*_{t,t^*}\right)_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
+        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $(y^*_{t,t^*})_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
 Each panel is partitioned into dives by vertical black lines.

### 2021-07-09 Report/Reviewer_1.tex
Address MAM and NH comments, update page/line numbers for reviewer 1

\documentclass{article}

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{psfrag}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{url}
\usepackage{float}
\usepackage[margin=1in]{geometry}
\input{header}
\input{defs}

\begin{document}
\title{Modelling state-switching functional data with hidden Markov models: reviewer 1 response}
\date{}
\author{Blinded}

\maketitle

We sincerely thank the reviewers for their comments, and we believe that the paper is better today than it was prior to peer review. Before addressing the reviewer comments one-by-one, there are several changes to the paper that we would like to highlight.

\begin{itemize}
    \item To address the potential optimization issues raised by the reviewers, we have re-run the case with more random initializations. The numerical results have changed slightly, but the qualitative results remain the same.
    \item Reviewer 1 commented that it would be better to see more than 2 hidden states in the simulation study. As such, we have re-run the simulation study with 3 fine-scale behavioural states instead of 2.
    \item Reviewer 1 requested that we add the likelihood surface for the CarHHMM-DFT, and we have done so in the supplemental material. Additionally, the supplemental material has been broken into two: supplement A contains material related to the case study while supplement B contains material related to the simulation study. 
    \item We have added line numbers for convenience. All page and line references refer to the current manuscript.
\end{itemize}

\section{Reviewer 1}

\begin{enumerate}
    \item 
    \begin{enumerate}
        \item \textit{What is the difference between `different curve types' and `between-curve dependence'?}

        The different curve types and between-curve dependence are closely related to one another according to a hidden Markov model. We assume that each curve is labelled with one of a discrete number of curve types, and each curve is randomly generated from a probability distribution that depends upon its type. Conditional on type, the curves are independent. We also assume that the sequence of curve types makes up an unobserved Markov chain, which is what induces the between-curve dependence. We have edited page 3, lines 49-51 as well as pages 4-5, lines 80-82 to make this distinction clearer. In addition, we have changed ``different curve types" to ``a discrete number of curve types" where applicable to highlight that we are using a Markov chain with a discrete number of states to model the between-curve dependence.

        \item \textit{Is the paper trying to cluster the curves or identify the progression of curves at a given time t?}

        We are identifying curve types and behavioural states, which is a sort of clustering. We are also studying the progression of curves types and behavioural states by inferring the probability transition matrix of the hidden Markov chain. We have added an explanation to pages 4-5, lines 80-82 to clarify the goal of this paper. Using our case study as an example, we identify and describe two different dive types and three different swimming behaviours while describing how the whale transitions between these dive types and behavioural states. The distribution of the swimming behaviours can be used to estimate killer whale energy expenditure, and state-decoding algorithms can be used to identify possible foraging dives and feeding events.

        \item \textit{If the dependence between curves is purely based on the hidden variables, then there should be more justifications in 2.4 as curves are transformed to lower-dimensional variables. For functional time series, please see ``On the prediction of stationary functional time series", ``High-dimensional functional time series forecasting: An application to age-specific mortality rates".}

        We do transform the curves to lower-dimensional variables, which causes a significant amount of reconstruction error. We do not bound this reconstruction error, and only talk about reconstruction in appendix for the purpose of the simulation study. However, our goal is not to reconstruct the curves, but rather to identify the coarse-scale and fine-scale curve types as well as describe the dependence between those types. We have clarified our goal on page 15, lines 280-282 and have discussed and cited the aforementioned papers on page 17, lines 310-319. Nonetheless, bounding prediction and reconstruction error is a promising area of future study. In particular, it would be useful to prove an analogue of Theorem 3.2 in ``On the prediction of stationary functional time series" for state-switching functional time series. We added a sentence about this to the discussion on page 40, lines 743-746.

    \end{enumerate}

    \item \textit{What is the within-curve structure? There are numerous ways to deal with periodic functions. fPCA and smoothing with a Fourier basis will give you a better transformation of curves than the proposed in 2.4.}

    This is an excellent point -- fPCA is an effective way to reduce reconstruction error, and we added a discussion of this on pages 17, lines 310-319. We now direct the reader to a variety of alternative summary statistics over the moving windows, including fPCA scores. However, the goal of our case study is to provide a means of summarizing these curves in a way that is interpretable for practitioners, not to reconstruct the curves themselves. We therefore chose Fourier analysis because it is familiar to practitioners in ecology/ biologging.

    \item \begin{enumerate}
        \item \textit{There is no discussion about $x^*$ other than the transitional matrix. What's the significance of this hidden variable and in the hierarchy?}

        If $x_t = i$, then $x_{t,t*}^*$ is a state that represents one of $N^{*(i)}$ possible within-curve hidden behaviours associated with dive type $i$. The value of $x_{t,t*}^*$ determines the distribution of the observation $y^*_{t,t^*}$. We have added a sentence on page 12, lines 225-226 to make this more clear. For example, in the case study we have that there are 3 possible subdive states. If the whale is in subdive state $i =2$ for dive $t$ and window $\tilde t^*$, then $\tilde x^*_{t,\tilde t^*} = 2$.  In addition, we added an explanation on page 21, lines 392-400 to make clear what $x$ and $x^*$ are in the context of the case study.

        \item \textit{Have you tried to fit a simple HMM with only $x_t$ and $y_t$?}

        This is a good point. However, we suspect that a simple HMM would likely have comparable performance to our model to determine the dive types ($x_t$) and the conditional distributions of dive duration ($y_t$). However, we are primarily interested in jointly estimating the dive types and the subdive behaviours for the killer whale case study. If anything, biologists are usually more interested in the fine-scale behaviours. To make this more clear to the reader, we have added a sentence on page 22, lines 402-404.

    \end{enumerate}

    \item \begin{enumerate}
        \item \textit{The `between-curve dependence' or the hidden variable $x_t$ is significantly influenced by $y_t$ not the fine-scale observations as shown in Figure 4.}

        We are assuming that the reviewer is pointing out that it is confusing that the fourth panel of Figure 1 only models the fine-scale process. We have edited the caption of Figure 1 to indicate that we are omitting the coarse-scale process and only use the HMM-DFT as a fine-scale model of the fine-scale process. In addition, we assume that the reviewer may be concerned about the fact that Figure 4 only shows the distribution of \textit{three} fine-scale behavioural states rather than \textit{six} dive-type / behavioural state pairs. There are only three fine-scale emission distributions because the three fine-scale behavioural states are shared across both dive types. We have included an explanation of which emission distributions go with dive types and which emission distributions go with subdive states in the caption of Figure 4.

        \item \textit{Where is the relationship between $x_t$ and $x^*_t$? The likelihood function does not include this relation.}

        $x_t$ corresponds to the curve type of curve $t$, and the curve type determines the parameters of the fine-scale model used to model the fine-scale observations $y^*_t$. If that fine-scale model is again an HMM, then $x^*_t$ is a Markov chain whose probability transition matrix is determined by the curve type ($x_t$). In addition, the distribution of $y^*_{t,t^*}$ is determined by both the curve type ($x_t$) and the fine-scale hidden state ($x^*_{t,t^*}$). We have edited page 13, lines 238-240 to make this fact more clear. The likelihood function does not include either $x_t$ or $x^*_t$ because both hidden variables can be marginalized out of the likelihood. We have made this more clear by adding an explanation to page 9, lines 163-166 as well as page 13, lines 246-248.

    \end{enumerate}

    \item \textit{It would be better to see more than 2 hidden states where emission distributions are similar in the simulation studies.}

    We agree that it would be better to have 3 rather than 2 fine-scale hidden states and have emission distributions that more closely match the case study. As such, we re-ran our simulation study with 3 fine-scale hidden states and parameters that more closely represent the case-study. Using 3 fine-scale states in the simulation study means that the structure of the model in the simulation study and case study match. See section 4 of the paper for the updated results. We have based the simulation study on the case study in order to simulate data that is as realistic as possible. We believe that this gives a good balance between separated and overlapping hidden states so the simulation study is informative.

    \item \textit{It is interesting to see the posterior distributions when the model is fitted with a sampling approach (sequential Monte Carlo).}

    This is an interesting approach that has been studied in several other papers. This is beyond the scope of our paper, but we have added references to several sampling methods for HMMs (and relevant citations) on page 9, lines 169-177.

    \item \textit{After the transformation, does the autoregressive pattern hold? Is the likelihood function of $\tilde y^*$ in Equation (4) valid?}

    After the transformation we are redefining our model such that all modelling assumptions apply to the transformed variables rather than the original fine-scale variables. Therefore, the model selection and model validation techniques in Sections 3.2 and 3.4 should be on the transformed observations, not the original observations. For example, Figure 1 of Supplement A shows lag plots for the transformed fine-scale observations, and there appears to be strong auto-correlation. We have added several sentences on page 16, lines 295-304 to clarify this. Thank you for prompting us to correct this omission.

    \item \begin{enumerate}
        \item \textit{The problem is overcomplicated, what is the optimization routine? }

        For each model, we ran 100 direct maximization routines using random initializations using the Nelder-Mead method of the optimize package of spicy. In particular, we optimized one row at a time of the probability transition matrices and one set of parameters associated with a (feature,hidden state) pair at a time. We found that most, but not all, of the optimization routines converged to similar parameter estimates. We have added an explanation of our optimization routine on page 26, lines 491-500. Thank you for prompting us to correct this omission.

        \item \textit{Can you provide some surface plots of the likelihood?}

        We have added plots of the likelihood surface associated with the CarHHMM-DFT in supplement A. Unfortunately the parameter space is high-dimensional, so we plotted the likelihood of the parameters associated with each feature / hidden state pair while keeping all other parameters fixed at their maximum likelihood estimates.

    \end{enumerate}

\end{enumerate}

We are grateful for the comments from reviewer 1. We believe the modifications we have made in response to their comments have greatly improved our manuscript.

\end{document}

### 2021-07-09 Report/Reviewer_2.tex
Address MAM and NH comments, update page/line numbers for reviewer 1

\documentclass{article}

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{psfrag}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{url}
\usepackage{float}
\usepackage[margin=1in]{geometry}
\input{header}
\input{defs}

\begin{document}
\title{Modelling state-switching functional data with hidden Markov models: reviewer 2 response}
\date{}
\author{Blinded}

\maketitle

We sincerely thank the reviewers for their comments, and we believe that the paper is better today than it was prior to peer review. Before addressing the reviewer comments one-by-one, there are several changes to the paper that we would like to highlight.

\begin{itemize}
    \item To address the potential optimization issues raised by the reviewers, we have re-run the case with more random initializations. The numerical results have changed slightly, but the qualitative results remain the same.
    \item Reviewer 1 commented that it would be better to see more than 2 hidden states in the simulation study. As such, we have re-run the simulation study with 3 fine-scale behavioural states instead of 2.
    \item Reviewer 1 requested that we add the likelihood surface for the CarHHMM-DFT, and we have done so in the supplemental material. Additionally, the supplemental material has been broken into two: supplement A contains material related to the case study while supplement B contains material related to the simulation study. 
    \item We have added line numbers for convenience. All page and line references refer to the current manuscript.
\end{itemize}

\section{Reviewer 2}

\textit{The authors of CJS-21-0003 present a novel and promising approach based on Hidden Markov Models to analyze functional data with intricate dependence structures. I would like to congratulate the authors on the very well-written paper. Below are just a few comments I would like the authors to address.}

We appreciate the kind words and are glad that the reviewer has enjoyed the paper.

\begin{enumerate}
    \item \textit{Introduction, page 5, around line 15. Before the work of Langrock et al. (2018), there has been the work of De Souza and Heckman (2014), which was the first paper proposing this new class of nonparametric regression models, called switching nonparametric regression models, allowing the hidden states to be independent or to follow a Markov structure. Later De Souza, Heckman, and Xu (2017) extended their approach to multiple curves allowing for different variance structures within each curve. Together with Langrock et. al. (2018), I believe these two references should be included as important examples of HMMs in nonparametric functional modelling:}

    \textit{de Souza, C. P. E., Heckman, N. E. and Xu, F., (2017) ``Switching nonparametric regression models for multi-curve data", The Canadian Journal of Statistics}

    \textit{de Souza, C. P. E. and Heckman, N. E. (2014) ``Switching nonparametric regression models", Journal of Nonparametric Statistics}

    We thank the reviewer for the references and have added them to page 5, lines 92-96.

    \item \textit{Section 2, end of page 6: the authors start the description of their model by presenting this idea of coarse-scale versus fine-scale, but I missed a connection between these and what was said in the introduction regarding the problem of analyzing biologging data. It would be great if the authors could make this connection clear. There is something written later in Section 2.4, maybe some of that can be said earlier.}

    The introduction primarily uses the adjectives ``between-curve" and ``within-curve" to focus on FDA, while the subsequent sections switch to ``coarse-scale" and ``fine-scale" to be more general. The connection between these adjectives was not clear before, so we have added a sentence to page 2, lines 34-36 to introduce ``coarse-scale" vs ``fine-scale" earlier. We have also added a sentence to page 5, lines 80-82 to make the connection between coarse-scale / fine-scale and curve types / behavioural states.

    \item \textit{Section 2, end of page 6: is there a special reason for denoting the number of curves by $T$? I find it less confusing to use, for example, $N$ or $M$ for that purpose so that each curve can be denoted by $i$ or $j$ because $t$ reminds me of evaluation time points which later you denote by $t^*$.}

    It is confusing that the number of curves is denoted by $T$, which usually indicates time points. However, the curves are assumed to be sequential in time, and there is a connection between the coarse scale index $t$ and the fine-scale index $t^*$. We have added a sentence on page 7, lines 137-138 to clarify that these curves are sequential but not necessarily equi-spaced in time. In addition, there is some precedence for indexing sequential curves with $t$. See ``Cam\'er-Karhunen-Lo\`eve representation and harmonic principal component analysis of functional time series" (2013) by Panaretos and Tavakoli and ``Dynamic functional principal components" by Hormann and Kidinski (2014) for examples.

    \item \textit{In Section 2.1, page 7 around line 50: it is important to explicitly say that a time-homogeneous chain is being considered, that is, transition probabilities do not depend on time.}

    We agree and have added a sentence to clarify this on page 8, lines 151-152. Thank you for noting the omission.

    \item \textit{Page 8 around line 25: when presenting the likelihood $\mathcal{L}_{\text{HMM}}$, I would first present it and then mention the forward algorithm otherwise the reader may think that the equation has to do with the algorithm, but it is just the likelihood. Another thing is, I believe the algorithm is the backward-forward algorithm because it involves calculating both the so-called backward and forward quantities.}

    We have edited page 9, lines 171-176 to clarify the difference between the likelihood and the forward algorithm. We have also described the forward algorithm, which is used to evaluate the likelihood, as well as the forward-backward algorithm, which is used for state decoding and maximizing the likelihood via the Baum-Welch algorithm (Zucchini et al, 2016).

    \item \textit{Could the author clarify how the maximum likelihood estimation is conducted? Is it via direct maximization of the likelihood using numerical methods or via the EM algorithm? I missed a section or paragraph describing exactly how inference is conducted. }

    We personally maximized the likelihood directly. For each model, we ran 100 direct maximization routines using random initializations using the Nelder-Mead method of the optimize package of SciPy. In particular, we optimized one row at a time of the probability transition matrices and one set of parameters associated with a (feature, hidden state) pair at a time. We found that most, but not all, of the optimization routines converged to similar parameter estimates. We have added an explanation of our optimization routine on page 26, lines 491-500. Thank you for calling attention to this omission.

    \item \textit{de Souza, C. P. E. et al. (2017) also consider the case where the distribution of the hidden states depends on some covariates such as weather conditions. In this case, for biologging data, would it be useful or would it make sense to potentially have transition probabilities depending on covariates via the same, logit link, representation that the authors already used.}

    This is a very useful and common practice in the ecology literature. While it is beyond the scope of our case study, we have added a brief discussion of covariates on page 10, lines 182-185.

\end{enumerate}

We are grateful for the kind words and recommendations from reviewer 2. Their suggestions have been incorporated, and the paper is more complete as a result.

\end{document}

### 2021-07-09 Report/Sections/intro.tex
Address MAM and NH comments, update page/line numbers for reviewer 1

@@
 These data contain a wealth of information about human and animal behaviour, but modelling these large data sets poses a challenge for statisticians and biologists.
-One particular difficulty is that simultaneous coarse- and fine-scale processes are often reflected in high-frequency data, and each scale can exhibit unique and complicated dependence structures.
+One particular difficulty is that simultaneous coarse- and fine-scale processes are often reflected in high-frequency data, and each scale can exhibit unique and complicated structure.
 
@@
 However, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence assumes that curves evolve smoothly in time.
-This is not appropriate in most biologging applications where behaviours often change suddenly between a discrete number of types.
+This is not appropriate in many biologging applications where behaviours often change suddenly between a discrete number of types.
 In addition to multi-level models, FDA researchers have used functional time series to model dependence in a sequence of curves.

### 2021-07-09 Report/Sections/models.tex
Address MAM and NH comments, update page/line numbers for reviewer 1

@@
 We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.
-We assume that the curves $Y_1,\ldots,Y_T$ are indexed according to the temporal order in which they are observed, but the $T$ observations need not be equally spaced in time.
+We assume that the curves $Y_1,\ldots,Y_T$ are indexed according to the order in which they are observed in time, but the $T$ observations need not be equally spaced apart.
 To develop our model for this data, we detail the structure of a traditional HMM followed by three variations which generalize its base structure.
@@
 %
-but it is also computationally tractable to sum over all possible values of $x$ to obtain the following:
+but it is also tractable to sum over all possible values of $x$ to obtain the following:
 
@@
 
-It is important to note that replacing $Y^*_t$ with summary statistics results in a loss of information since it involves substituting infinite-dimensional curves with a finite-dimensional representation.
-Readers interested in summary statistics that prediction error for sequential functional data can read \citet{Aue:2015} and \citet{Gao:2019}, who both use functional principle component analysis (FPCA) and derive bounds on the error caused by dimension reduction.
+Replacing $Y^*_t$ with summary statistics results in a loss of information since it involves substituting infinite-dimensional curves with a finite-dimensional representation.
+While we are primarily interested in classifying and describing coarse- and fine-scale behaviour, some researchers may be less interested in interpretability and more interested in predicting future functional data.
+To this end, \citet{Aue:2015} and \citet{Gao:2019} both use functional principle component analysis (FPCA) and derive bounds on the error caused by dimension reduction.
 While these papers do not account for state-switching behaviour between curves or windows, deriving similar error bounds for FPCA within an HMM framework appears to be a promising direction for future research.

### 2021-07-09 Report/Sections/results.tex
Address MAM and NH comments, update page/line numbers for reviewer 1

@@
 %To illustrate an application of this method and compare the candidate models, 
-We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the \textit{SciPy} package in Python \citep{Virtanen:2019}.
+We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the SciPy package in Python \citep{Virtanen:2019}.
 Each model is fit 100 times using random initializations, and

### 2021-07-09 Report/Supp_Case.tex
Address MAM and NH comments, update page/line numbers for reviewer 1

@@
         \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the dive duration parameters $\mu_Y^{(i)}$ and $\sigma_Y^{(i)}$ for dive types $i = 1,2$.
-All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
+All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_Y,\hat \sigma_Y)$, which is denoted by a red star.
@@
         \begin{center}
-        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-eta-likelihood.png}
+        \includegraphics[width=2.1in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_coarse-gamma-likelihood.png}
         \end{center}
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the coarse-scale probability transition matrix parameters $\eta_{1,2}$ and $\eta_{2,1}$.
-Recall that $\Gamma_{i,j} = \exp(\eta_{i,j}) / \sum_k \exp(\eta_{i,k})$.
-All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \eta_{1,2},\hat \eta_{2,1})$, which is denoted by a red star.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the coarse-scale probability transition matrix parameters $\Gamma_{1,2}$ and $\Gamma_{2,1}$.
+All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \Gamma_{1,2},\hat \Gamma_{2,1})$, which is denoted by a red star.
         \addtocounter{fignum}{1}
@@
         \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the wiggliness parameters $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
+All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_W^{*(\cdot,i^*)}, \hat \sigma_W^{*(\cdot,i^*)})$, which is denoted by a red star.
@@
         \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $x$-acceleration parameters $\mu_{A_x}^{*(\cdot,i^*)}, \sigma_{A_x}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
+All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_x}^{*(\cdot,i^*)}, \hat \sigma_{A_x}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
@@
         \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $y$-acceleration parameters $\mu_{A_y}^{*(\cdot,i^*)}, \sigma_{A_y}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
+All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_y}^{*(\cdot,i^*)}, \hat \sigma_{A_y}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
@@
         \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $z$-acceleration parameters $\mu_{A_z}^{*(\cdot,i^*)}, \sigma_{A_z}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
+All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_z}^{*(\cdot,i^*)}, \hat \sigma_{A_z}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
@@
         \begin{center}
-        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_0.png}
-        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_0.png}
-        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_1.png}
-        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_1.png}
-        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-0-row_2.png}
-        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-eta-likelihood-1-row_2.png}
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-0-row_0.png}
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-1-row_0.png}
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-0-row_1.png}
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-1-row_1.png}
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-0-row_2.png}
+        \includegraphics[width=3in]{../Plots/2019/20190902-182840-CATs_OB_1_0_267_CarHHMM2_fine-gamma-likelihood-1-row_2.png}
         \end{center}
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the fine-scale probability transition matrix parameters associated both dive types ($\eta^{*(i)}_{i^*,j^*}$ for dive types $i = 1,2$ and subdive states $i^*,j^* = 1,2,3$).
-Recall that $\eta^{*(i)}_{i^*,i^*} = 0$ and that $\Gamma^{*(i)}_{i^*,j^*} = \exp(\eta^{*(i)}_{i^*,j^*}) / \sum_{k^*} \exp(\eta^{*(i)}_{i^*,k^*})$.
-All other parameters and probability transition matrices are set equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the fine-scale probability transition matrix parameters associated both dive types ($\Gamma^{*(i)}_{i^*,j^*}$ for dive types $i = 1,2$ and subdive states $i^*,j^* = 1,2,3$).
+All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\hat \Gamma^{*(i)}$, which is denoted by a red star.
-Note that the long ridges of approximately equal likelihood are due to the fact that $\hat \Gamma^{*(i)}_{1,3} \approx \hat \Gamma^{*(i)}_{3,1} \approx 0$.
+Note that $\hat \Gamma^{*(i)}_{1,3} \approx \hat \Gamma^{*(i)}_{3,1} \approx 0$.
 This indicates that the whale rarely switches directly between subdive states 1 and 3.

### 2021-07-09 Report/Reviewer_2.tex
Update page / line numbers, reviewer 2

@@
 
-    We thank the reviewer for the references and have added them to page 5, lines 92-96.
+    We thank the reviewer for the references and have added them to page 5, lines 92-95.
 
@@
 However, the curves are assumed to be sequential in time, and there is a connection between the coarse scale index $t$ and the fine-scale index $t^*$.
-We have added a sentence on page 7, lines 137-138 to clarify that these curves are sequential but not necessarily equi-spaced in time.
+We have added a sentence on page 7, lines 138-140 to clarify that these curves are sequential but not necessarily equi-spaced in time.
 In addition, there is some precedence for indexing sequential curves with $t$.
@@
 
-    We agree and have added a sentence to clarify this on page 8, lines 151-152. Thank you for noting the omission.
+    We agree and have added a sentence to clarify this on page 8, lines 153-154. Thank you for noting the omission.
 
@@
 
-    We have edited page 9, lines 171-176 to clarify the difference between the likelihood and the forward algorithm.
+    We have edited pages 9, lines 165-172 to clarify the difference between the likelihood and the forward algorithm.
 We have also described the forward algorithm, which is used to evaluate the likelihood, as well as the forward-backward algorithm, which is used for state decoding and maximizing the likelihood via the Baum-Welch algorithm (Zucchini et al, 2016).
@@
 We have added an explanation of our optimization routine on page 26, lines 491-500.
-Thank you for calling attention to this omission.
+Thank you for prompting us to correct this omission.
 
@@
 
-    This is a very useful and common practice in the ecology literature. While it is beyond the scope of our case study, we have added a brief discussion of covariates on page 10, lines 182-185.
+    This is a very useful and common practice in the ecology literature. While it is beyond the scope of our case study, we have added a brief discussion of covariates on page 10, lines 183-186.
 

### 2021-07-09 Report/Sections/intro.tex
Final edits

@@
 where the sequence of dive types exhibits state-switching behaviour \citep{Tennessen:2019b}.
-On a fine scale, these profiles display bouts of short-term periodicity within each dive \citep{Adam:2019}.
+On a fine scale, these profiles can display bouts of short-term periodicity within each dive \citep{Adam:2019}.
 Fine-scale periodicity nested within a coarser state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
@@
 While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions.
-In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, and \citet{DeSouza:2014,DeSouza:2017} use HMMs to model state-switching behaviour within functional data.
+In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, while \citet{DeSouza:2014} and \citet{DeSouza:2017} use HMMs to model state-switching behaviour within functional data.
 However, none of these papers account for temporal correlation taking place on multiple scales.

### 2021-07-09 Report/Sections/models.tex
Final edits

@@
 The estimated probability of each hidden state $\hat{\Pr}(X_t = i|Y = y)$ can then be calculated using $\{\hat \theta, \hat \Gamma\}$ and the \textit{forward-backward algorithm} \citep{Zucchini:2016}.
-Alternatively, it is reasonable to use a Bayesian approach, where Equation (\ref{eqn:like}) is combined with a prior to obtain a posterior distribution over $X$, $\theta$, and $\Gamma$.
+Alternatively, it is reasonable to use a Bayesian approach, where Equation (\ref{eqn:like}) is combined with a prior distribution to obtain a posterior distribution over $X$, $\theta$, and $\Gamma$.
 This posterior distribution can be sampled from using methods such as sequential Monte Carlo \citep{Douc:2011b}, Markov-chain Monte Carlo \citep{Scott:2002}, or variational inference \citep{Foti:2014}.
@@
 While we are primarily interested in classifying and describing coarse- and fine-scale behaviour, some researchers may be less interested in interpretability and more interested in predicting future functional data.
-To this end, \citet{Aue:2015} and \citet{Gao:2019} both use functional principle component analysis (FPCA) and derive bounds on the error caused by dimension reduction.
+To this end, \citet{Aue:2015} and \citet{Gao:2019} both use functional principle component analysis (FPCA) and derive bounds on the reconstruction error caused by dimension reduction.
 While these papers do not account for state-switching behaviour between curves or windows, deriving similar error bounds for FPCA within an HMM framework appears to be a promising direction for future research.

### 2021-07-09 Report/Sections/results.tex
Final edits

@@
 Dive type 2 is longer, deeper, and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives.
-No dive in this data set has a maximum depth deeper than 30 meters and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters.
+No dive in this data set has a maximum depth deeper than 30 meters, and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters.
 However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and %researchers
@@
 Histograms of the pseudoresiduals mostly support that the CarHHMM-DFT is well-specified.
-One exception is $\Ztwo_{t,\tilde t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:pseudoresids}).
+One exception is $\Ztwo_{t,\tilde t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:model_checking}).
 This implies that the true distribution of $\Ztwo_{t,\tilde t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study.
@@
 
-We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}.
+We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:model_checking}.
 Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the %\textit{forward-backward algorithm} \citep{Zucchini:2016}.

### 2021-07-09 Report/Sections/simulation.tex
Final edits

@@
 The two simulated dive types differ in that dives of type 1 are much shorter on average (27 seconds) than dives of type 2 (128 seconds).
-The three simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 3 than for subdive state 2, which in turn is much higher than for subdive state 1.
+The three simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 3 than for subdive state 2, which in turn are much higher than for subdive state 1.
 These larger parameter values correspond to much more vigorous and variable periodic behaviour in the acceleration data.

### 2021-09-16 Report/Sections/appendix.tex
Updates from Overleaf
Update Final Version

--- 2020-11-24 Report/Sections/appendix.tex
@@
-% !TeX root = ../main.tex
+\subsection{Detailed description of data simulation from Section \ref{subsec:data_simulation}}
 
-\subsection{Detailed description of data simulation}
+\setcounter{equation}{6}   %  set this to the last equation number used
 
-It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
-Then, given $X_t = i$, $Y_t = y_t$, and $\Gamma^{*(i)}$, it is straightforward to simulate the subdive states $\tilde{X}^*_t$ for a particular dive $t$.
-Recall that the length of each fine-scale Markov chain is $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ since each window is two seconds long.
-It is much more difficult to simulate the raw acceleration data $Y_t^*$, and we focus on this task for the remainder of the section.
+We easily simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
+For each dive $t$, we also easily generate the fine-scale hidden Markov chain $\tilde X^*_t \equiv \big\{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T_t^*}\big\}$ according to one of the probability transition matrices $\Gamma^{*(1)}$ or $\Gamma^{*(2)}$, depending upon the value of $X_t$.
+This determines the sequence of fine-scale hidden states corresponding to each window.
+Recall that the fine-scale model is based on a sequence of $\tilde T_t^*$ two-second windows, each containing 100 observations, and our model is formulated in terms of quantities derived from the raw data within each window (namely, the average acceleration, $\Zone_{t,\tilde t^*}$, and wiggliness, $\Ztwo_{ t,\tilde t^*}$).
+Generating the raw acceleration data from $\Zone_{t,\tilde t^*}$ and $\Ztwo_{ t,\tilde t^*}$ is not straightforward.
+In our simulation study, we generate raw acceleration data so that we can visualize our results in terms of the underlying data curves for each dive.
+Here, we explain how we generate the acceleration curves so that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{ t,\tilde t^*}$ both follow the specified model.
+A key component is the discrete Fourier transformation of the 100 raw acceleration values in window $\tilde{t}^*$ of dive $t$:
 
-Recall that each window is two seconds long and therefore contains $h = 100$ observations taken at 50 observations per second.
-Let $DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,\tilde t^*}$ and $DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$ be denoted as $\hat{Y}^*_{t,\tilde t^*}$ for convenience.
-We simulate $Y_t^*$ in three steps: (1) simulate the average acceleration within each window $\left(\hat Y^{*(0)}_{t,\tilde t^*}\right)$, (2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and (3) take the inverse $DFT$ of $\hat{Y}^*_{t,\tilde t^*}$ to retrieve $\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$.
-The details of this procedure are described below:
+\[
+    \hat{Y}^{*(k)}_{t,\tilde{t}^*}  \equiv DFT\left\{Y^*_{t,100 (\tilde{t}^*-1) + 1 },\ldots, Y^*_{t,100 \tilde{t}^*}\right\}(k)
+\]
+for $k \geq 0$, as defined in Equation (\ref{eq:DFTdef}).
 
-\begin{enumerate}
-\item Simulate the average acceleration $Y^{*(0)}_{t,\tilde t^*}$ for each window $\tilde t^* = 1,2,\ldots,\tilde T^*_t$ using the following procedure:
-\begin{enumerate}
-    \item Simulate $\hat{Y}^{*(0)}_{t,1}$:
-    $$
-    	(\hat{Y}^{*(0)}_{t,1}|\tilde X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
-    $$
-    where $\sigma_A^{*(\cdot,1)} = 0.034s$ and $\sigma_A^{*(\cdot,2)} = 0.079s$. 
+We simulate the raw acceleration data for dive $t$ in three steps: (1) simulate the average acceleration within each window $\left(\hat Y^{*(0)}_{t,\tilde t^*}\right)$, (2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and (3) take the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$:
+
+\[
+    \{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\} \equiv IDFT\left\{\hat{Y}^{*(0)}_{t,\tilde t^*},
+    \ldots, \hat{Y}^{*(99)}_{t,\tilde t^*}\right\}
+    {\rm{~for~}} \tilde t^* = 1,\ldots,\tilde T^*_t.
+\]
+The details of steps (1) and (2) are described below.
+
+For step (1), we generate $\hat Y^{*(0)}_{t,1}, \ldots, \hat Y^{*(0)}_{t,\tilde T_t^*}$ as a CarHMM with underlying Markov state sequence $\tilde X^{*}_{t,1}, \ldots, \tilde X^{*}_{t,\tilde T_t^*}$ and random first emission.
+Specifically, we let
+
+   \[
+    	\hat{Y}^{*(0)}_{t,1}|\tilde X^*_{t,1} = i^* ~~\sim~~ \mathcal{N} \left(100 \mu^{*(\cdot,i^*)}_A, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
+    	~~{\rm{and}}
+   \]
+
+    \begin{align}	
+       \hat{Y}^{*(0)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = i^*,\hat{Y}^{*(0)}_{t,\tilde t^*-1}
+       &~~\sim ~~\mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,\tilde t^*-1} + 100(1-\phi_A^{*(\cdot,i^*)}) \mu^{*(\cdot,i^*)}_A, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \label{eqn:yhat_0} \\
         %
-    \item Simulate $\hat{Y}^{*(0)}_{t,2}, \ldots,\enspace \hat{Y}^{*(0)}_{t,\tilde T^*_t}$ using Equation (\ref{eqn:yhat_0}) below:
-    \begin{align}	
-        \Big(\hat{Y}^{*(0)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = i^*,\hat{Y}^{*(0)}_{t,\tilde t^*-1}\Big) &\sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,\tilde t^*-1}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \nonumber \\
-        %
-        &t^* = 2,\ldots, \tilde T^*_t
-    	\label{eqn:yhat_0}
+        &\tilde t^* = 2,\ldots, \tilde T^*_t
+    	\nonumber
     \end{align}
-    where $\phi_A^{*(\cdot,1)} = 0.98$ and $\phi_A^{*(\cdot,2)} = 0.87$. 
-    \end{enumerate}
-    After this step we have simulated the average acceleration within each two-second window for all dives (see Equation (\ref{eqn:avg_val}) below).
-    \item Simulate the other Fourier coefficients $\left(Y^{*(k)}_{t,\tilde t^*}\right)$ for each window $\tilde t^* = 1,2,\ldots,\tilde T^*_t$ and $k = 1,\ldots,99$ using the following procedure:
-    \begin{enumerate}
-    %
-    \item Simulate Fourier coefficients 1 through 49:
+    where $\mu_A^{*(\cdot,i^*)}$, $\sigma_A^{*(\cdot,i^*)}$, and $\phi_A^{*(\cdot,i^*)}$ are as defined in the simulation study in Section 4.
+
+ For step (2),
+ we first construct $\hat{Y}^{*(k)}_{t,\tilde t^*}$, $k=1,\ldots, 49$, as
+
     \begin{equation}
-        \hat{Y}^{*(k)}_{t,\tilde t^*} = a_{t,\tilde t^*}^{(k)} i\sqrt{b^{(k)}_{t,\tilde t^*}}, \quad k = 1,\ldots,49;
+        \hat{Y}^{*(k)}_{t,\tilde t^*} = a_{t,\tilde t^*}^{(k)} i\sqrt{b^{(k)}_{t,\tilde t^*}}
         \label{eqn:abYhat}
     \end{equation}
-    where $a^{(k)}_{t,\tilde t^*}$ is equal to either 1 or -1 each with probability 1/2, and $b^{(k)}_{t,\tilde t^*}$ has the following distribution:
+    where the $a^{(k)}_{t,\tilde t^*}$'s are independent and equal to either 1 or -1 each with probability 1/2 and $i$ is the imaginary number.
+We include $i$ in Equation (\ref{eqn:abYhat}) to force all variation within a window to take the form of a sine wave, which reduces the variation between the endpoints of windows compared to a cosine wave.
+    Given the fine scale states, the $b^{(k)}_{t,\tilde t^*}$'s are independent of one another and independent of the $a^{(k)}_{t,\tilde t^*}$'s.  
+   The distribution of $b^{(k)}_{t,\tilde t^*}$ is
+
     \begin{align}
     \begin{split}
-    	(b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1) &\sim {\rm{Gamma}}(16.38/k^3, 36.23) \\
+    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1 &~~\sim~~ {\rm{Gamma}}(11.03/(k+1)^3, 15.54) \\
     	%
-    	(b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2) &\sim {\rm{Gamma}}(4.20/k^3, 1825.53). \\ 
+    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2  &~~\sim ~~{\rm{Gamma}}(4.80/(k+1)^3, 515.38). \\ 
+    	%
+    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 3  &~~\sim ~~{\rm{Gamma}}(2.31/(k+1)^3, 20023.44). 
     \end{split}
@@
 The squared magnitude of the $k^{th}$ Fourier coefficient is equal to $b^{(k)}_{t,\tilde t^*}$, which decays like $1/k^3$ to ``smooth out" the raw acceleration data.
+
+   We then define the remaining 50 Fourier coefficients:
+
+    \[
+        \hat{Y}^{*(50)}_{t,\tilde t^*} = 0 {\rm{~~and ~~}}
+	    \hat{Y}^{*(k)}_{t,\tilde t^*} = -\hat{Y}^{*(100-k)}_{t,\tilde t^*} , k = 51,\ldots,99.
+    \]
+   This guarantees that the inverse discrete Fourier transform is real-valued.
+
+We now show that this construction of the raw acceleration data results in the distributions listed in Section \ref{subsec:data_simulation}.
+It suffices to show that the construction of the discrete Fourier transformations, the $\hat{Y}^{(k)}_{t,\tilde t^*}$'s, yields the desired distributions.
+
+First, since $\hat{Y}^{(0)}_{t,\tilde t^*} = \sum_{n=1}^{100} Y^*_{t,100(\tilde t^* - 1) + n} $ $=100\Zone_{t,\tilde{t}^*}$, Equation (\ref{eqn:yhat_0}) implies that 
+$\Zone_{t,1}, \ldots, \Zone_{t,\tilde{T}^*}$ 
+follows a CarHMM with Normal emissions distributions and parameters as defined in the simulation study in Section 4.
+
+From Equations (\ref{eqn:z}) and (\ref{eqn:abYhat}), 
+the wiggliness within window $\tilde t^*$ of dive $t$ is
+
+\[
+    \Ztwo_{t,\tilde t^*} = \sum_{k=1}^{\tilde \omega} \big|\big|\hat{Y}^{(k)}_{t,\tilde t^*}\big|\big|^2 =  \sum_{k=1}^{\tilde \omega} b^{(k)}_{t,\tilde t^*}.
+\]
 %
-    \item Set Fourier coefficients 50 through 99 so the inverse DFT is real-valued:
-    $$
-    \hat{Y}^{*(50)}_{t,\tilde t^*} = 0, \qquad
-	\hat{Y}^{*(k)}_{t,\tilde t^*} = -\hat{Y}^{*(100-k)}_{t,\tilde t^*} \enspace \text{for} \enspace k = 51,\ldots,99.
-    $$
-    \end{enumerate}
-    \item Finally, set the raw acceleration data within each window $\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}$ with the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$:
+If $\tilde \omega < 50$, then $\Ztwo_{t,\tilde t^*}$ is the sum of independent Gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,\tilde t^*}$ is also Gamma.
+Thus, by Equation (\ref{eqn:bdist}):
+
+\[
+    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 11.03/(k+1)^3 , 15.54 \right),
+\]
 %
-    $$\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\} \equiv IDFT\left\{\hat{Y}^*_{t,\tilde t^*}\right\}$$
-    for $\tilde t^* = 1,\ldots,\tilde T^*_t.$
-\end{enumerate}
-
-We now show that this construction of $Y^*_t$ results in the distributions listed in Section \ref{subsec:data_simulation}.
-First, note that $\hat{Y}^{(0)}_{t,\tilde t^*}$ is just the sum of the raw acceleration values within window $\tilde t$:
+\[
+    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.80/(k+1)^3 , 515.38 \right), \text{and}
+\]
 %
-\begin{align}
-    \hat{Y}^{(0)}_{t,\tilde t^*} &=
-    DFT\{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\}(0) \nonumber \\
-    &= \sum_{n=0}^{99} Y^*_{t,100(\tilde t^* - 1) + 1 + n} \exp\left(-\frac{i2\pi}{99}*0*n\right) \nonumber \\
-    &= \sum_{n=1}^{100} Y^*_{t,100(\tilde t^* - 1) + n} \nonumber \\
-    &= 100A^*_{t,\tilde t^*}.
-    \label{eqn:avg_val}
-\end{align}
-%
-Combining Equation (\ref{eqn:avg_val}) above with Equation (\ref{eqn:yhat_0}) gives the distribution of $\Zone_{t,\tilde t^*}$: 
-%
-$$\left(100\Zone_{t,\tilde t^*} | \tilde X^*_{t,\tilde t^*} = i^*, \Zone_{t,\tilde t^*-1} \right) \sim \mathcal{N} \left(100\phi_A^{*(\cdot,i^*)} \Zone_{t,\tilde t^*-1}, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)$$
-%
-$$\implies \left(\Zone_{t,\tilde t^*} | \tilde X^*_{t,\tilde t^*} = i^*, \Zone_{t,\tilde t^*-1} \right) \sim \mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \Zone_{t,\tilde t^*-1}, \left(\sigma_A^{*(\cdot,i^*)}\right)^2 \right).$$
-%
-This distribution is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(\cdot,1)} = \mu_A^{*(\cdot,2)} = \mu_A^{*(\cdot,i^*)} = 0$.
-We also have that $\sigma_A^{*(\cdot,1)} = 0.034s$, $\phi_A^{*(\cdot,1)} = 0.98$, $\sigma_A^{*(\cdot,2)} = 0.079s$, and $\phi_A^{*(\cdot,2)} = 0.87$.
-
-To calculate the wiggliness $\Ztwo_{t,\tilde t^*}$, we pick $\tilde{\omega} = 10$ periods per window (or 5 Hz). From Equation (\ref{eqn:abYhat}):
-%
-\begin{equation}
-    \Ztwo_{t,\tilde t^*} = \sum_{k=1}^{10} \big|\big|\hat{Y}^{(k)}_{t,\tilde t^*}\big|\big|^2 = \sum_{k=1}^{10} \big|\big|a^{(k)}_{t,\tilde t^*}i\sqrt{b^{(k)}_{t,\tilde t^*}}\big|\big|^2 = \sum_{k=1}^{10} b^{(k)}_{t,\tilde t^*}
-    \label{eqn:Wb}
-\end{equation}
-%
-since $a^{(k)}_{t,\tilde t^*} \in \{-1,1\}$ and $b^{(k)}_{t,\tilde t^*} \geq 0$.
-Since $\Ztwo_{t,\tilde t^*}$ is the sum of Gamma-distributed random variables with identical scale parameters, the distribution of $\Ztwo_{t,\tilde t^*}$ is also Gamma.
-Combining Equation (\ref{eqn:Wb}) above with Equation (\ref{eqn:bdist}) gives
-%
-$$\left(\Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 16.38/k^3 = 3.25 , 36.23 \right) \text { and }$$
-%
-$$\left(\Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2\right) \sim {\rm{Gamma}}\left(\sum_{k=1}^{10} 4.20/k^3 = 0.83 , 1825.53 \right),$$
-%
-where the first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the shape parameter and the second is the scale parameter.
-Simple calculation of the mean and variance of a Gamma distribution reveals $\mu_W^{*(\cdot,1)} = 23.3$, $\sigma_W^{*(\cdot,1)} = 13.0$, $\mu_W^{*(\cdot,2)} = 301.2$, and $\sigma_W^{*(\cdot,2)} = 330.1$.
+\[
+    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 3  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 2.31/(k+1)^3 , 20023.44 \right).
+\]
+Setting $\tilde \omega$ to 10 and carrying out simple calculation of the mean and variance of a Gamma distribution gives the proper values for $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$.
 
@@
 The overall likelihood of the CarHHMM-DFT model is as follows:
-%
-$$\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,\z) = \delta P(y_1,\z_1;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,\z_t;\theta,\theta^*,\Gamma^*) \mathbf{1}_N.$$
+
+\[\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,\z) = \delta P(y_1,\z_1;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,\z_t;\theta,\theta^*,\Gamma^*) \mathbf{1}_N.\]
 %
 In particular,
-%
+
 \begin{align*}
-P(y_t,\z_t;\theta,\theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\theta^{(1)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(1)};\z_t\right), \ldots , \\
-&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{CarHMM-DFT}}\left(\theta^*,\Gamma^{*(N)};\z_t\right) \Big],
+P(y_t,\z_t;\theta,\theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\theta^{(1)})\calL_{\text{fine}}\left(\theta^*,\Gamma^{*(1)};\z_t\right), \ldots , \\
+&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{fine}}\left(\theta^*,\Gamma^{*(N)};\z_t\right) \Big],
 \end{align*}
@@
 where $f^{(i)}(y_t;\theta^{(i)})$ is the emission distribution of dive duration given that $X_t = i$.
-The likelihood $\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale model and is equal to the following:
-%
-$$\calL_{\text{CarHMM-DFT}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},$$
+The likelihood $\calL_{\text{fine}}$ corresponds to the fine-scale model and is equal to the following:
+
+\[\calL_{\text{fine}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},\]
 %
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\iffalse
-
-\subsection{Equivalence of state-switching OU process and CarHMM}
-
-\citet{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
-%
-\begin{equation}
-    \label{eqn:SDE}
-    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t,
-\end{equation}
-%
-where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process in state $X_t$, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process.
-They refer to this as a state-switching Ornstein-Uhlenbeck process.
-Unlike an HMM, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer in this case.
-If the behavioural state $X_t$ is known and does not change between observations $\Big(X_s = X_t$ for all $s \in [t,t+\Delta t)\Big)$, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
-%
-\begin{align}
-    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right),
-    \label{eqn:OU_sol}
-\end{align}
-%
-where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$.
-In fact, under certain conditions, the output of a state-switching OU process is identical to that of a CarHMM, as described by Theorem 1 below.
-
-\begin{theorem}{Theorem 1.}{}%
-If the following conditions are met:
-\begin{enumerate}
-    \item The hidden behavioural process $X$ from Equation (\ref{eqn:SDE}) follows a Markov chain with $N$ possible states and transitions occur at equi-spaced time stamps $\left(\Delta t, \ldots, (T-1)\Delta t\right)$,
-
-    \item Observations of the SDE from Equation (\ref{eqn:OU_sol}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
-\end{enumerate}
-then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\theta = \{\theta^{(1)}, \ldots, \theta^{(N)}\}; \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
-
-\begin{align}
-\mu^{(i)} = \gamma^{(i)}, \qquad \sigma^{(i)} = \sqrt{\frac{\omega^{(i)^2}}{2\beta^{(i)}} (1-e^{-2\beta^{(i)}\Delta t})}, \qquad \phi^{(i)} = e^{-\beta^{(i)}\Delta t} \label{eqn:CarHMM_to_OU}
-\end{align}
-
-\end{theorem}
-
-\begin{proof}{Proof of Theorem 1}{}
-Combining Equation (\ref{eqn:OU_sol}) with Equation (\ref{eqn:CarHMM_to_OU}) gives
-%
-\begin{align*}
-    \left(Y_{(s+1) \Delta t} | X_{s\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{s \Delta t}, \enspace \left(\sigma^{(i)}\right)^2 \right),\\
-    s = 0, \ldots, T-2.
-\end{align*}
-%
-If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:
-
-\begin{align*}
-    \left(Y'_s| X'_s = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y'_{s-1}, \enspace \left(\sigma^{(i)}\right)^2 \right)\\
-    s = 2, \ldots, T.
-\end{align*}
-%
-$X'$ is assumed to follow a (possibly unobserved) Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s-1})$ is consistent with that of a CarHMM with Normal emission distributions.
-\end{proof}
-
-\fi
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-09-16 Report/Sections/discussion.tex
Update Final Version

@@
 In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt our methodology for the analysis of a wide range of time series data in a variety of fields.
-
-\iffalse
-
-Current FDA literature addresses dependence between curves either with multilevel models with random effects and no time component at all \citep{Chen:2012,Di:2009}, or with functional time series which model the curves as an evolving process \citep{Kokoszka:2018} which overlook the possibility that curves have several distinct ``types".
-However, many curve sequences exhibit both temporal dependence and a discrete number of types.
-One obvious example is the case study of killer whale behaviour examined here, where one dive can serve different purposes such as travelling, socializing, resting, or foraging \citep{Tennessen:2019a}.
-Another example is accelerometer data in school children, where bouts of activity can be associated with a behaviour such as running or walking.
-\citep{Morris:2007}.
-In this work, we deal with both temporal dependence between curves and distinct curve ``types" by using either an HMM or CarHMM.
-To help select the appropriate model between these two, lag plots can be used to test if there is significant auto-correlation between curves.
-
-When used traditionally, each of these models require that practitioners take summary statistics of curves as emissions which may lack important information about the structure of the curve.
-However, HMM emissions can also be viewed as the entire curve, where the emission distribution of that curve is described by yet another fine-scale model.
-There are a wide range of possible fine-scale models, including, but not limited to, another HMM, a CarHMM, a Poisson process, or a spline-based approach similar to that of \citet{Langrock:2018}.
-The parameters of this fine-scale model then depend upon the curve type as determined by the coarse-scale model.
-Together, these coarse- and fine-scale models make up a hierarchical model which can account for simultaneous processes taking place at different time-scales.
-
-One interesting class of possible fine-scale models is the set of continuous time methods such as the state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019} and the continuous-time HMM (CTHMM) \citep{Liu:2015}, which are often based on stochastic differential equations.
-Unfortunately, most continuous-time models require MCMC algorithms to perform inference and can be slow to fit.
-However, we show the appendix that the state-switching OU process of \citet{Michelot:2019} is equivalent to a CarHMM under certain conditions.
-This equivalence implies that a CarHMM is interpretable as a continuous-time model, but has the added advantage of relative computational efficiency.
-
-Another fine-scale model that we investigate in detail here is yet another HMM.
-If the fine-scale model is an HMM, then the resulting overall model is just the HHMM as introduced by \citet{Barajas:2017}.
-\citet{Adam:2019} in particular use the HHMM to analyze 25 Hz accelerometer data from horn sharks off the coast of southern California.
-This model can be inappropriate at very fine scales, as \citet{Adam:2019} find that subsequent fine-scale observations are not independent when conditioned on the hidden states due to periodic behaviour in the acceleration data.
-This indicates the need to account for complicated dependence structures within fine-scale processes, especially those with high sampling frequencies.
-To deal with these problematic fine-scale phenomena, we explore a moving-window transformation over the fine-scale observations.
-Since our case study data clearly exhibits periodic behaviour, we specifically use a DFT-based transformation.
-The effectiveness of this approach is evidenced by both our case study, where the CarHHMM without a DFT transformation visually fails to recognize fluking behaviour in a killer whale, and our simulation study, where the average accuracy associated with the subdive state is by far the worst when omitting $\Ztwo_{t,t^*}$ (``wiggliness'') from the model.
-After transforming the fine-scale observations, either subject area experts or exploratory analysis may suggest a large degree of temporal correlation.
-In this case, we suggest using a CarHMM instead of a traditional HMM to account for the auto-correlation in the transformed fine-scale observations.
-Our simulation study shows how ignoring this auto-correlation can result in underestimating standard errors and overestimating the variance of the observations.
-
-When using HMMs to build complex hierarchical models, it is important to build a model that is complex enough to fit the data but not so complex that it over-fits the data or is computationally inefficient.
-Model complexity inevitably grows rapidly as more layers are added to the hierarchical structure.
-For example, by far the fastest model to train in the simulation study is the CarHMM-DFT, as it has no hierarchical component and is the simplest of the four candidate models.
-The CarHMM-DFT also has near-perfect accuracy when decoding the simulated subdive behavioural states.
-However, this model does not allow ecologists to study the joint relationship between the dive type and intra-dive behaviour.
-One way to temper model complexity is to reduce the number of parameters within a given model structure.
-In particular, we assume that the fine-scale state parameters are the same for all coarse-scale states for all models in both our case study and simulation study.
-
-%To illustrate the potential applications of this method, we model the behaviour of a northern resident killer whale animal behaviour using data collected from time depth recorders and accelerometers.
-Understanding the diving behaviour and energetic requirements of NRKWs is essential to understand their foraging behaviour and survival \citep{Williams:2009,Noren:2011}.
-However, it is necessary to classify the underlying activity \citep{Dot:2016} or dive type \citep{Hastie:2006} of an animal in order to effectively understand and estimate its energy expenditure from biologging data.
-The CarHMMM-DFT model uses coarse-scale data to separate dives into different types while also describing the distribution of acceleration within each subdive state on the fine scale.
-
-Our work provides a flexible framework to model functional time-series data using HMMs.
-Provided the model is not overly complex, a hierarchical model constructed using the proposed building blocks can be both flexible and easy to fit using maximum likelihood methods.
-We demonstrate the usefulness of such a model using an ecological example, where we use a complex HMM to classify the coarse- and fine-scale diving behaviour of a killer whale.
-Our analysis gives a deeper understanding of the animal's tri-axial movement and thus its fine-scale energy expenditure \citep{Gleiss:2011,Qasem:2012}, which is essential when discerning the survival of both NRKWs and their prey \citep{Noren:2011}.
-Although our case study focuses on northern resident killer whales, this model is likely applicable to many diving animals such as sharks \citep{Adam:2019}, seals \citep{Dot:2016}, and porpoises \citep{Barajas:2017}.
-In addition, since complicated state-switching processes with temporal dependence are common in settings ranging from speech recognition \citep{Juang:1991} and neuroscience \citep{Langrock:2013} to oceanography \citep{Bulla:2012} and ecology \citep{Adam:2019}, we believe that researchers can adapt this methodology for the analysis of a wide range of time-series data in a variety of fields.
-
-\fi

### 2021-09-16 Report/Sections/intro.tex
Update Final Version

@@
 Tags can record over 50 observations per second, resulting in time series containing millions of observations over the course of several hours.
-%These data contain a wealth of information about human and animal behaviour, but parsing these large data sets poses a challenge for statisticians and biologists.
 These data contain a wealth of information about human and animal behaviour, but modelling these large data sets poses a challenge for statisticians and biologists.
@@
 However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
-On a coarse scale, dive profiles of marine animals show a discrete number of distinct dive ``types"
-%that aggregate together in time 
-where the sequence of dive types exhibits state-switching behaviour \citep{Tennessen:2019b}.
+On a coarse scale, dive profiles of marine animals show a discrete number of distinct dive ``types" where the sequence of dive types exhibits state-switching behaviour \citep{Tennessen:2019b}.
 On a fine scale, these profiles can display bouts of short-term periodicity within each dive \citep{Adam:2019}.
@@
 More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}.
-%However, these multilevel models are not appropriate in many biologging applications since they do not account for a discrete number curve types.
 However, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence assumes that curves evolve smoothly in time.

### 2021-09-16 Report/Sections/models.tex
Update Final Version

@@
 
-%Using observation emissions, denoted here as $y \equiv \{y_1,\ldots,y_T\}$, we can find the maximum likelihood estimates of the parameters $\Gamma$ and $\theta \equiv \{\theta^{(1)},\ldots,\theta^{(N)}\}$.
 The joint likelihood of both the parameters and the hidden states is given by the following:
@@
 This formulation simplifies likelihood maximization by removing constraints in the optimization problem.
-We assume there are no covariate effects, but \cite{DeSouza:2017} and \citet{Adam:2019} both incorporate covariates into $\Gamma$ by setting $\eta_{ij}(z_t) = \beta_{ij}^T z_t$ for $i \neq j$, where $z_t$ is a column vector of known covariates and $\beta_{ij}$ is a column vector of unknown regression coefficients.
+We assume there are no covariate effects, but similarly to the work of \citet{DeSouza:2017} for independent states and \citet{Adam:2019} for Markovian states, one could incorporate covariates into $\Gamma$ by setting $\eta_{ij}(z_t) = \beta_{ij}^T z_t$ for $i \neq j$, where $z_t$ is a column vector of known covariates and $\beta_{ij}$ is a column vector of unknown regression coefficients.
 For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$.
@@
 We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$.
-%This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density or probability mass function that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$.
+%
 As a concrete example, if $Y_t$ is a scalar, then one may assume that $Y_t$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
@@
 As a result, the distribution of the fine-scale observation $Y^*_{t,t^*}$ does not depend on $X_t$, but instead is defined exclusively by $X^*_{t,t^*}$ (or $(\cdot,X^*_{t,t^*})$ for notational consistency).
-%Coarse-scale hidden states therefore differ only in their coarse-scale emission parameters $\theta^{(i)}$ and fine-scale probability transition matrices $\Gamma^{*(i)}$, not their fine-scale emission parameters $\theta^{*(\cdot,i)}$.
 
-Due to the nested structure of the HHMM, %its likelihood is easily calculated using the forward algorithm.
-it is straightforward to extend the forward algorithm and sum over all possible coarse-scale hidden states $X$ and all possible fine-scale hidden states $X^*$ in the likelihood.
+Due to the nested structure of the HHMM, it is straightforward to extend the forward algorithm and sum over all possible coarse-scale hidden states $X$ and all possible fine-scale hidden states $X^*$ in the likelihood.
 %
@@
 The fine-scale probability transition matrices $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ correspond to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$, as do all fine-scale model assumptions (e.g., conditional independence, auto-regressive structure between observations, etc.) Fine-scale model selection and validation should therefore be adjusted accordingly.
-%instead of $X^*_t$ and $Y^*_t$.
 
@@
 %
-%Possible candidates for the fine-scale model include any of the models described in the previous subsections in addition to many others not described here. 
 For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations of a Poisson process as their fine-scale model.

### 2021-09-16 Report/Sections/results.tex
Update Final Version

@@
 
-%To illustrate an application of this method and compare the candidate models, 
 We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the SciPy package in Python \citep{Virtanen:2019}.
-Each model is fit 100 times using random initializations, and
-%Let $\mu_0,\sigma_0$, and $\phi_0$ represent the initial mean, standard deviation, and auto-correlation parameters for a given observation.
-If the observation is strictly non-negative, then we set $\log(\mu_0) \sim \mathcal{N}(\log(\bar Y),s_{\log})$, $\log(\sigma) \sim \mathcal{N}(s_{\log},1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean of the observations and $s_{\log}$ represents the sample standard deviation the logarithm of the observations values.
-If the observation can be any real number, then we set $\mu_0 \sim \mathcal{N}(\bar Y,s)$, $\log(\sigma) \sim \mathcal{N}(s,1)$, and $\logit(\phi) \sim \mathcal{N}(-1,1)$ where $\bar Y$ represents the sample mean and $s$ represents the sample standard deviation of the observations values.
-we keep the parameters estimates corresponding to the maximum likelihood of those 100 optimization routines \citep{Zucchini:2016}.
+Each model is fit 100 times using random initializations, and we keep the parameter estimates corresponding to the maximum likelihood of those 100 optimization routines \citep{Zucchini:2016}.
 Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory.
@@
 No dive in this data set has a maximum depth deeper than 30 meters, and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters.
-However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and %researchers
-unpublished data collected by the authors suggest that %significant numbers of 
-prey-capture events can occur near the surface. %\citep{Fortune:2020}. 
+However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and unpublished data collected by the authors suggest that prey-capture events can occur near the surface.
 Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.
@@
 We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:model_checking}.
-Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the %\textit{forward-backward algorithm} \citep{Zucchini:2016}.
-forward-backward algorithm.
+Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the forward-backward algorithm.
 This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2.
@@
 The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize 
-any given behaviour as subdive state 3
-%the behaviour at the beginning and end of dives as subdive state 3 
-(see Figures 11 and 12 of supplement A).
-In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are %approximately $50$ to $100$ percent
-significantly larger than those of the CarHHMM-DFT.
+any given behaviour as subdive state 3 (see Figures 11 and 12 of supplement A).
+In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are significantly larger than those of the CarHHMM-DFT.
 The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of of supplement A).
-This suggests %that including auto-correlation in the model significantly affects parameter estimates, %Further, the pseudoresiduals of the HHMM-DFT are noticeably light-tailed and do not follow a standard normal distribution (see Figure 15 of of supplement A).
-These findings suggest
-that the HHMM-DFT overlooks some auto-correlation in the data, and that the HHMM-DFT may be overconfident in its parameters estimates compared to the full CarHHMM-DFT.
+This suggests that the HHMM-DFT overlooks some auto-correlation in the data, and that the HHMM-DFT may be overconfident in its parameters estimates compared to the full CarHHMM-DFT.
 

### 2021-09-16 Report/main.tex
small changes
End of day Jan 6 2021
Updates from Overleaf
Update Final Version

--- 2021-01-06 Report/main.tex
@@
 \usepackage{url}
+%\usepackage{lineno}
 \input{header}
@@
 
+%\linenumbers
+
 \begin{document}
 \firstpage{1}
-\lastpage{25}
+\lastpage{34}
 \jvol{xx}
 \issue{yy}
-\jyear{2020}
+\jyear{2021}
 \jid{CJS}
@@
 % The running head contains the author names
-\rhauthor{BlindedA and BlindedB}
+\rhauthor{Sidrow et al.}
 \copyrightline{Statistical Society of Canada}
@@
 % History: received and accepted dates
-\received{\rec{9}{July}{2009}}
-\accepted{\acc{8}{July}{2010}}
+\received{\rec{14}{January}{2021}}
+\accepted{\acc{14}{September}{2021}}
 
@@
 \title[]{Modelling multi-scale state-switching functional data with hidden Markov models}
-%\author{Evan Sidrow\authorref{1}}
-%\author{Nancy Heckman\authorref{1}}
-%\author{Sarah Fortune\authorref{2}}
-%\author{Andrew Trites\authorref{3,4}}
-%\author{Ian Murphy\authorref{1}}
-%\author{Marie Auger-M\'eth\'e\authorref{1,4}}
-%\affiliation[1]{Department of Statistics, University of British Columbia, Vancouver, Canada}
-%\affiliation[2]{Marine Mammal Research Unit, University of British Columbia, Vancouver, Canada}
-%\affiliation[3]{Department of Zoology, University of British Columbia, Vancouver, Canada}
-%\affiliation[4]{Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada}
-\author{BlindedA\authorref{1}\thanksref{*}}
-\author{BlindedB\authorref{2}}
-\affiliation[1]{Author affiliations will go here in the accepted manuscript, but do NOT include them in your initial submission because it must be anonymous.}
-\affiliation[2]{Second Affiliation}
+\author{Evan Sidrow\authorref{*1}}
+\author{Nancy Heckman\authorref{1}}
+\author{Sarah M. E. Fortune\authorref{2}}
+\author{Andrew W. Trites\authorref{3,4}}
+\author{Ian Murphy\authorref{5}}
+\author{Marie Auger-M\'eth\'e\authorref{1,4}}
+\affiliation[1]{Department of Statistics, University of British Columbia, Vancouver, Canada}
+\affiliation[2]{Marine Mammal Research Unit, University of British Columbia, Vancouver, Canada}
+\affiliation[3]{Department of Zoology, University of British Columbia, Vancouver, Canada}
+\affiliation[4]{Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada}
+\affiliation[5]{Department of Biostatistics, University of Florida, Gainesville, USA}
 
@@
 \abstractsection{Abstract}
-%Insert your abstract here; it should
-%typically be up to ten lines long. Avoid symbols as much as possible. Formulas
-%are strongly discouraged, and citations should be avoided. 
-%The title and the abstract should be concise and descriptive;
-%list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
-%http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.
-
-\iffalse
-Advances in biologging technology have made a vast amount of high-frequency time series data available in a variety of fields.
-These data often exhibit complicated temporal dependence structures that cannot be modelled using common methods of Functional Data Analysis (FDA).
-Animal movement modellers often employ a hierarchical approach using hidden Markov models (HMMs) instead, but many intricate fine-scale processes violate basic assumptions of HMMs.
-We detail a general hierarchical approach based on HMMs which incorporates a wider class of models at fine scales.
-As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada.
-Our model produces more interpretable results as well as more accurate parameter and standard error estimates compared to existing baselines.
-\fi
 
@@
 % Email address for corresponding author
-%\correspondingauthor[*]{\\\email{}} % only put this in after acceptance
+\correspondingauthor[*]{\\\email{evan.sidrow@stat.ubc.ca}} % only put this in after acceptance
 %
@@
 \begin{ack}{ACKNOWLEDGEMENTS}
+All killer whale data was collected under University of British Columbia Animal Care Permit no.
+A19-0053 and Fisheries and Oceans Canada Marine Mammal Scientific License for Whale Research no.
+XMMS 6 2019.
+This project was supported in part by a financial contribution from Fisheries and Oceans Canada and the Natural Sciences and Engineering Research Council of Canada (Whale Science for Tomorrow).
 This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
-We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC). 
-Marie Auger-M\'eth\'e thanks the NSERC Discovery and the Canadian Research Chair programs.
+We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC) as well as the support of Fisheries and Oceans Canada (DFO). 
+Marie Auger-M\'eth\'e and Nancy Heckman thank the NSERC Discovery program, and Marie Auger-M\'eth\'e additionally thanks the Canadian Research Chair program.
 Evan Sidrow thanks the University of British Columbia and the Four-Year Doctoral Fellowship program.
+We are grateful to Dr. Joe Watson for his constructive suggestions.
 \end{ack}
@@
 \begin{appendix}
-%\input Sections/appendix.tex
-\input Sections/appendixnh.tex
+\input Sections/appendix.tex
 \end{appendix}

### 2021-10-25 Report/Sections/appendix.tex
address copy editor's comments (main text)

@@
 We easily simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
-For each dive $t$, we also easily generate the fine-scale hidden Markov chain $\tilde X^*_t \equiv \big\{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T_t^*}\big\}$ according to one of the probability transition matrices $\Gamma^{*(1)}$ or $\Gamma^{*(2)}$, depending upon the value of $X_t$.
+For each dive $t$, we also easily generate the fine-scale hidden Markov chain $\tilde X^*_t = \big\{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T_t^*}\big\}$ according to one of the probability transition matrices $\Gamma^{*(1)}$ or $\Gamma^{*(2)}$, depending upon the value of $X_t$.
 This determines the sequence of fine-scale hidden states corresponding to each window.
@@
 \[
-    \hat{Y}^{*(k)}_{t,\tilde{t}^*}  \equiv DFT\left\{Y^*_{t,100 (\tilde{t}^*-1) + 1 },\ldots, Y^*_{t,100 \tilde{t}^*}\right\}(k)
+    \hat{Y}^{*(k)}_{t,\tilde{t}^*}  = DFT\left\{Y^*_{t,100 (\tilde{t}^*-1) + 1 },\ldots, Y^*_{t,100 \tilde{t}^*}\right\}(k)
 \]
@@
 \[
-    \{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\} \equiv IDFT\left\{\hat{Y}^{*(0)}_{t,\tilde t^*},
+    \{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\} = IDFT\left\{\hat{Y}^{*(0)}_{t,\tilde t^*},
     \ldots, \hat{Y}^{*(99)}_{t,\tilde t^*}\right\}
@@
    \[
-    	\hat{Y}^{*(0)}_{t,1}|\tilde X^*_{t,1} = i^* ~~\sim~~ \mathcal{N} \left(100 \mu^{*(\cdot,i^*)}_A, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
+    	\hat{Y}^{*(0)}_{t,1}\mid\tilde X^*_{t,1} = i^* ~~\sim~~ \mathcal{N} \left(100 \mu^{*(\cdot,i^*)}_A, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
     	~~{\rm{and}}
@@
     \begin{align}	
-       \hat{Y}^{*(0)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = i^*,\hat{Y}^{*(0)}_{t,\tilde t^*-1}
+       \hat{Y}^{*(0)}_{t,\tilde t^*}\mid\tilde X^*_{t,\tilde t^*} = i^*,\hat{Y}^{*(0)}_{t,\tilde t^*-1}
        &~~\sim ~~\mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,\tilde t^*-1} + 100(1-\phi_A^{*(\cdot,i^*)}) \mu^{*(\cdot,i^*)}_A, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \label{eqn:yhat_0} \\
@@
     \begin{split}
-    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1 &~~\sim~~ {\rm{Gamma}}(11.03/(k+1)^3, 15.54) \\
+    	b^{(k)}_{t,\tilde t^*}\mid\tilde X^*_{t,\tilde t^*} = 1 &~~\sim~~ {\rm{Gamma}}(11.03/(k+1)^3, 15.54) \\
     	%
-    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2  &~~\sim ~~{\rm{Gamma}}(4.80/(k+1)^3, 515.38). \\ 
+    	b^{(k)}_{t,\tilde t^*}\mid\tilde X^*_{t,\tilde t^*} = 2  &~~\sim ~~{\rm{Gamma}}(4.80/(k+1)^3, 515.38). \\ 
     	%
-    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 3  &~~\sim ~~{\rm{Gamma}}(2.31/(k+1)^3, 20023.44). 
+    	b^{(k)}_{t,\tilde t^*}\mid\tilde X^*_{t,\tilde t^*} = 3  &~~\sim ~~{\rm{Gamma}}(2.31/(k+1)^3, 20023.44). 
     \end{split}
@@
 \[
-    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 11.03/(k+1)^3 , 15.54 \right),
+    \Ztwo_{t,\tilde t^*}\mid\tilde X^*_{t,\tilde t^*} = 1  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 11.03/(k+1)^3 , 15.54 \right),
 \]
@@
 \[
-    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.80/(k+1)^3 , 515.38 \right), \text{and}
+    \Ztwo_{t,\tilde t^*}\mid\tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.80/(k+1)^3 , 515.38 \right), \text{and}
 \]
@@
 \[
-    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 3  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 2.31/(k+1)^3 , 20023.44 \right).
+    \Ztwo_{t,\tilde t^*}\mid\tilde X^*_{t,\tilde t^*} = 3  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 2.31/(k+1)^3 , 20023.44 \right).
 \]
@@
 
-\[\calL_{\text{fine}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},\]
+\[\calL_{\text{fine}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}\mid\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},\]
 %
-where $P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*)$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{*(\cdot,i^*)}(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1}; \theta^{*(\cdot,i^*)})$.
+where $P(\z_{t,\tilde t^*}\mid\z_{t,\tilde t^*-1};\theta^*)$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{*(\cdot,i^*)}(\z_{t,\tilde t^*}\mid\z_{t,\tilde t^*-1}; \theta^{*(\cdot,i^*)})$.
 %
-Recall that $f^{*(\cdot,i^*)}(\cdot|\z_{t,t^*-1}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,\tilde t^*}$ when $\tilde X^*_{t,\tilde t^*} = i^*$ and $\Z_{t,\tilde t^*-1} = \z_{t,\tilde t^*-1}$.
+Recall that $f^{*(\cdot,i^*)}(\cdot\mid\z_{t,t^*-1}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,\tilde t^*}$ when $\tilde X^*_{t,\tilde t^*} = i^*$ and $\Z_{t,\tilde t^*-1} = \z_{t,\tilde t^*-1}$.
 

### 2021-10-25 Report/Sections/discussion.tex
address copy editor's comments (main text)

@@
 
-Current Functional Data Analysis literature addresses dependence between curves either with multilevel models \citep{Chen:2012,Di:2009}, which lack a time component, or with functional time series, which overlook the possibility that curves have several distinct ``types" \citep{Kokoszka:2018}.
-Our work addresses these issues and introduces a flexible framework to model functional time-series data using HMMs.
+Current functional data analysis literature addresses dependence between curves either with multi-level models \citep{Di:2009,Chen:2012}, which lack a time component, or with functional time series, which overlook the possibility that curves have several distinct ``types" \citep{Kokoszka:2018}.
+Our work addresses these issues and introduces a flexible framework to model functional time series data using HMMs.
 We suggest handling temporal dependence between curves by using either an HMM or a CarHMM to model the curve sequence.
 We then suggest viewing each individual curve as an HMM emission whose distribution is described by a fine-scale model.
-Here we use a CarHMM as the fine-scale model, but there are a wide range of possible fine-scale models, including a Poisson process or continuous time approach similar to that of \citet{Michelot:2019}.
+Here we use a CarHMM as the fine-scale model, but there are a wide range of possible fine-scale models, including a Poisson process or a continuous time approach similar to that of \citet{Michelot:2019}.
 We also incorporate a moving-window transformation at the fine scale to capture intricate dependence structures.
-Together, the coarse- and fine-scale models make up a hierarchical structure which can account for simultaneous processes taking place at different time scales.
-Provided the construction is not overly complex, a hierarchical model created using our method can be both flexible and easy to fit using maximum likelihood estimation.
-Our method is not intended to minimize prediction error for functional data, but incorporating HMMs into auto-regressive predictive models such as those described by \citet{Aue:2015} and \citet{Gao:2019} is a promising and natural direction for future study.
+Together, the coarse- and fine-scale models make up a hierarchical structure that can account for simultaneous processes taking place at different time scales.
+Provided that the construction is not overly complex, a hierarchical model created using our method can be both flexible and easy to fit using maximum likelihood estimation.
+Our method is not intended to minimize prediction error for functional data, but incorporating HMMs into autoregressive predictive models such as those described in \citet{Aue:2015} and \citet{Gao:2019} is a promising and natural direction for future study.
 

### 2021-10-25 Report/Sections/intro.tex
address copy editor's comments (main text)

@@
 Biologging technology now provides researchers with kinematic data collected almost continuously in time \citep{Hooten:2017}.
-The collection and analysis of data from devices such as accelerometers have brought new insights to areas ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}.
-The study of animal movement in particular has been transformed by tracking devices that record kinematic information in a variety of environments \citep{Borger:2020,Dot:2016b}.
+The collection and analysis of data from devices such as accelerometers have brought new insights to research tasks ranging from monitoring machine health \citep{Getman:2009} to understanding physical activity levels in children \citep{Morris:2007}.
+The study of animal movement in particular has been transformed by tracking devices that record kinematic information in a variety of environments \citep{Dot:2016b,Borger:2020}.
 Tags can record over 50 observations per second, resulting in time series containing millions of observations over the course of several hours.
 These data contain a wealth of information about human and animal behaviour, but modelling these large data sets poses a challenge for statisticians and biologists.
-One particular difficulty is that simultaneous coarse- and fine-scale processes are often reflected in high-frequency data, and each scale can exhibit unique and complicated structure.
+One particular difficulty is that simultaneous coarse- and fine-scale processes are often reflected in high-frequency data, and each scale can exhibit a unique and complicated structure.
 
-Biologging data are frequently modelled as a set of curves and analyzed using methods of Functional Data Analysis, or FDA -- see, e.g., \citet{Ramsay:2005}.
-For example, \citet{Morris:2007} view a child's activity level as a set of daily curves where metabolic activity is a function of time.
-Similarly, \citet{Fu:2017} view the dive profile of a southern elephant seal (\textit{Mirounga leonina}) as a set of dive curves whereby the amplitude and phase variation of the dive is used to classify dive types.
+Biologging data are frequently modelled as a set of curves and analyzed using methods for functional data analysis, or FDA \nocite{Ramsay:2005}(e.g., Ramsay and Silverman, 2005).
+For example, \citet{Morris:2007} views a child's activity level as a set of daily curves where metabolic activity is a function of time.
+Similarly, \citet{Fu:2017} views the dive profile of a southern elephant seal (\textit{Mirounga leonina}) as a set of dive curves.
+Dive amplitude and phase variation are used for classification into ``dive types".
 
-FDA was originally developed to process curves assumed to be independent replicates (i.e., there is no between-curve dependence), and within-curve fine-scale structure is not usually incorporated in FDA models.
-However, sets of curves often exhibit complex sequential dependencies both between curves and within curves, especially in the case of biologging data \citep{Barajas:2017}.
-On a coarse scale, dive profiles of marine animals show a discrete number of distinct dive ``types" where the sequence of dive types exhibits state-switching behaviour \citep{Tennessen:2019b}.
+FDA was originally developed to process curves assumed to be independent replicates (i.e., there is no between-curve dependence).
+Within-curve, fine-scale structure is not usually incorporated in FDA models.
+However, sets of curves often exhibit complex sequential dependencies, both between and within curves.
+This is especially the case for biologging data \citep{Barajas:2017}.
+On a coarse scale, the dive profiles of marine animals show a discrete number of distinct dive types, where the sequence of dive types exhibits state-switching behaviour \citep{Tennessen:2019b}.
 On a fine scale, these profiles can display bouts of short-term periodicity within each dive \citep{Adam:2019}.
-Fine-scale periodicity nested within a coarser state-switching process is also common in fields ranging from machine health \citep{Xin:2018,Lucero:2019} to speech recognition \citep{Juang:1991}.
+Fine-scale periodicity nested within a coarser state-switching process is also common in fields such as machine health \citep{Xin:2018,Lucero:2019} and speech recognition \citep{Juang:1991}.
 
-Some FDA models account for between-curve dependence that occurs when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain kinds of dependence in time.
-For example, previous studies have used multilevel models with random effects to model variation between and within individuals in daily activity levels of children \citep{Morris:2007} or in menstrual cycles of adults \citep{Bromback:1998}.
-More recent work involving multilevel models includes that of \citet{Di:2009}, \citet{Crainiceanu:2009}, and \citet{Chen:2012}.
-However, the first two papers do not account for temporal between-curve dependence and the third paper's model of temporal dependence assumes that curves evolve smoothly in time.
-This is not appropriate in many biologging applications where behaviours often change suddenly between a discrete number of types.
+Some FDA models account for between-curve dependence that occurs when multiple curves arise from separate groups of individuals, but these models are inadequate when modelling certain kinds of time dependencies.
+For example, previous studies have used multi-level models with random effects to model variation between and within individuals in the daily activity levels of children \citep{Morris:2007} or in the menstrual cycles of adults \citep{Bromback:1998}.
+More recent work involving multi-level models includes that of \citet{Crainiceanu:2009}, \citet{Di:2009}, and \citet{Chen:2012}.
+However, the first two papers do not account for temporal between-curve dependence and the third article's model of temporal dependence assumes that curves evolve smoothly in time.
+This is not appropriate in many biologging applications, where behaviours often change suddenly between a discrete number of types.
 In addition to multi-level models, FDA researchers have used functional time series to model dependence in a sequence of curves.
-Functional time series extend the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but they do not account for sequences of time-series curves whose distributions are determined by a discrete number of well-defined hidden states.
+Functional time series extends the ideas of classic time series to model the evolution of one curve into the next \citep{Kokoszka:2018}, but does not account for sequences of time series curves whose distributions are determined by a discrete number of well-defined hidden states.
 
@@
 In particular, within-curve structure is usually modelled by a generic smooth mean function and a covariance function \citep{Yao:2005} or with random regression \citep{Rice:2001}.
-However, time-series data exhibiting both sharp behavioural changes and periodic fine-scale structure are difficult to model with these classic FDA techniques.
+However, time series data exhibiting both sharp behavioural changes and periodic fine-scale structure are difficult to model with these classical FDA techniques.
 
@@
 To accomplish this, we turn to the field of animal movement modelling \citep{Hooten:2017}, where one of the most prevalent techniques of late is the hidden Markov model, or HMM \citep{Patterson:2017,McClintock:2020}.
-HMMs interpret animal movement data as arising from a Markov chain of a discrete number of behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position.
-While ubiquitous in ecology literature, HMMs have seen little use in non-parametric functional modelling with a few notable exceptions.
-In particular, \citet{Langrock:2018} take a non-parametric approach to model the distributions of HMM observations with B-splines, while \citet{DeSouza:2014} and \citet{DeSouza:2017} use HMMs to model state-switching behaviour within functional data.
+HMMs interpret animal movement data as arising from a Markov chain on a discrete number of behavioural states, allowing biologists to infer the underlying behaviour of an animal from sequential observations of its position.
+While ubiquitous in ecology literature, HMMs have seen little use in nonparametric functional modelling, with a few notable exceptions.
+In particular, \citet{Langrock:2018} takes a nonparametric approach to model the distributions of HMM observations with B-splines, while \citet{DeSouza:2014} and \citet{DeSouza:2017} use HMMs to model state-switching behaviour within functional data.
 However, none of these papers account for temporal correlation taking place on multiple scales.
 
-While useful, HMMs alone are also not sufficient to model high-frequency time-series data for three primary reasons.
-Firstly, classic HMMs fail to model simultaneous behavioural processes that occur at different time scales (e.g., both between and within curves).
-To address this issue, statistical ecologists have employed hierarchical hidden Markov models (HHMMs) \citep{Barajas:2017,Adam:2019}, which model both scales with conditionally dependent HMMs.
-Secondly, HMMs assume that subsequent observations are independent given an underlying hidden state process, but this is often not the case when observations are taken at extremely high frequencies.
-Several solutions have been proposed in the ecology literature, including the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}.
-Thirdly, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures on short time scales.
-For example, \citet{Adam:2019} fail to capture fine-scale periodic swimming pattern of horn sharks (\textit{Heterodontus francisci}) using a traditional HHMM.
-\citet{Heerah:2017} successfully use Fourier analysis within an HMM to account for daily behavioural cycles in marine mammals, and Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
-Thus, incorporating Fourier analysis within the structure of an HMM appears to be a promising approach to account for fine-scale periodic structures in biologging data.
+While useful, HMMs alone are also not sufficient to model high-frequency time series data for three primary reasons.
+First, classical HMMs fail to model simultaneous behavioural processes that occur at different time scales (e.g., both between and within curves).
+To address this issue, statistical ecologists have employed hierarchical HMMs (HHMMs) \citep{Barajas:2017,Adam:2019}, which model both scales with conditionally dependent HMMs.
+Second, HMMs assume that subsequent observations are independent given an underlying hidden state process, but this is often not the case when observations are taken at extremely high frequencies.
+Several solutions have been proposed in the ecology literature, including the hidden movement Markov model \citep{Whoriskey:2016} and the conditionally autoregressive HMM, or CarHMM \citep{Lawler:2019}.
+Third, traditional HMMs, CarHMMs, and HHMMs cannot easily capture complicated dependence structures on short time scales.
+For example, \citet{Adam:2019} fails to capture the fine-scale periodic swimming patterns of horn sharks (\textit{Heterodontus francisci}) using a traditional HHMM.
+\citet{Heerah:2017} successfully use Fourier analysis within an HMM to account for daily behavioural cycles in marine mammals.
+Fourier analysis has previously been used with accelerometer data to explain animal behaviour \citep{Fehlmann:2017,Shorter:2017}.
+Thus, incorporating Fourier analysis into the structure of an HMM appears to be a promising approach to account for fine-scale periodic structures in biologging data.
 
 We combine these existing methods from statistical ecology literature in novel ways to classify and describe state-switching behaviours in functional data while accounting for complex temporal dependence.
-The resulting suite of methods makes up a ``tool box" that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time-series data with intricate dependence structure.
+The resulting methods make up a tool box that can be used to build arbitrarily complex hierarchical models to explain multi-scale functional and time series data with intricate dependence structures.
 We begin in Section 2 by describing HMMs as well as two variants, CarHMMs and HHMMs, and discuss how summary statistics over moving windows can handle fine-scale dependence structures.
@@
 In Section 3 we fit several candidate models to data from a killer whale (\textit{Orcinus orca}) from the threatened northern resident population off the coast of British Columbia, Canada.
-Section 4 details a simulation study based on these candidate models, and in Section 5 we discuss our results.
+Section 4 details a simulation study based on these candidate models, and in Section 5, we discuss our results.

### 2021-10-25 Report/Sections/models.tex
address copy editor's comments (main text)

@@
 Consider a sequence of $T$ curves, where curve $t$ is characterized by a curve-level (or coarse-scale) observation $Y_t$ as well as a sequence of $T^*_t$ within-curve (or fine-scale) observations $Y^*_{t}$.
-Namely, $Y^*_{t} \equiv \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of fine-scale quantities derived from curve $t$ and indexed by $t^*$.
+Namely, $Y^*_{t} = \big\{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\big\}$ is made up of fine-scale quantities derived from curve $t$ and indexed by $t^*$.
 Both $Y_t$ and $Y^*_{t,t^*}$ can be either vectors or scalars.
-We call the sequence of coarse-scale observations $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* \equiv \big\{Y^*_1,\ldots,Y^*_T \big\}$.
-We assume that the curves $Y_1,\ldots,Y_T$ are indexed according to the order in which they are observed in time, but the $T$ observations need not be equally spaced apart.
+We call the sequence of coarse-scale observations $Y = \big\{Y_1, \ldots, Y_T\big\}$ and the collection of all fine-scale observations $Y^* = \big\{Y^*_1,\ldots,Y^*_T \big\}$.
+We assume that the curves $Y_1,\ldots,Y_T$ are indexed according to the order in which they are observed in time, but the $T$ observations need not be equally spaced in time.
 To develop our model for this data, we detail the structure of a traditional HMM followed by three variations which generalize its base structure.
-We then show how each of these generalized HMMs can be synthesized to form a wide variety of more complicated models.
+We then show how each of these generalized HMMs can be synthesized to form a wide variety of more-complicated models.
 
@@
 
-Hidden Markov models describe state-switching Markovian processes in discrete time and are the core structure we use to model both $Y$ and $Y^*$.
+HMMs describe state-switching Markovian processes in discrete time and are the core structure we use to model both $Y$ and $Y^*$.
 For simplicity we focus on $Y$ to introduce the model.
-An HMM is comprised of a sequence of unobserved states $X \equiv \big\{X_1, \ldots, X_T\big\}$ together with an observation sequence $Y \equiv \big\{Y_1, \ldots, Y_T\big\}$, where $X_t$ is associated with the observation $Y_t$.
-The $Y_t$'s are often referred to as ``emissions'' and the index $t$ typically refers to time.
-The $X_t$'s form a Markov chain and can take integer values between $1$ and $N$.
-Their distribution is governed by the distribution of the initial state $X_1$ and an $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j | X_t = i)$.
+An HMM is composed of a sequence of unobserved states $X = \big\{X_1, \ldots, X_T\big\}$ together with an observation sequence $Y = \big\{Y_1, \ldots, Y_T\big\}$, where $X_t$ is associated with the observation $Y_t$.
+The $Y_t$s are often referred to as ``emissions'' and the index $t$ typically refers to time.
+The $X_t$s form a Markov chain and can take integer values between $1$ and $N$.
+Their distribution is governed by the distribution of the initial state $X_1$ and an $N \times N$ transition probability matrix $\Gamma$, where $\Gamma_{ij} = \Pr(X_{t+1} = j \mid X_t = i)$.
 We consider only time-homogeneous Markov chains, meaning that $\Gamma$ does not depend on time.
@@
 $\delta_i = \Pr(X_1 = i).$
-A Markov chain's stationary distribution is determined by its probability transition matrix via $\delta = \delta \Gamma$ and $\sum_{i=1}^N \delta_i = 1$.
+A Markov chain's stationary distribution is determined by its transition probability matrix via $\delta = \delta \Gamma$ where $\sum_{i=1}^N \delta_i = 1$.
 %
@@
 %
-If $X_t=i$ then we denote the conditional density or probability mass function of $Y_t$ as $f^{(i)}(\cdot ; \theta^{(i)})$ or simply $f^{(i)}(\cdot)$, where $\theta^{(i)}$ is a state-dependent parameter describing the emission distribution.
+If $X_t=i$, then we denote the conditional density or probability mass function of $Y_t$ as $f^{(i)}(\cdot ; \theta^{(i)})$ or simply $f^{(i)}(\cdot)$, where $\theta^{(i)}$ is a state-dependent parameter describing the emission distribution.
 %
 
-The joint likelihood of both the parameters and the hidden states is given by the following:
+The joint likelihood of both the parameters and the hidden states is given by
 
 \begin{equation}
-    \calL_{\text{HMM}}(x,\theta,\Gamma;y) = \delta_{x_1} f^{(x_1)}(y_1; \theta^{(x_1)}) \prod_{t=2}^T \Gamma_{x_{t-1} x_t} f^{(x_t)}(y_t; \theta^{(x_t)}),
+    \calL_{\text{HMM}}(x,\theta,\Gamma;y) = \delta_{x_1} f^{(x_1)}(y_1; \theta^{(x_1)}) \prod_{t=2}^\top \Gamma_{x_{t-1} x_t} f^{(x_t)}(y_t; \theta^{(x_t)}),
     \label{eqn:like}
@@
 %
-but it is also tractable to sum over all possible values of $x$ to obtain the following:
+but it is also tractable to sum over all possible values of $x$ to obtain
 
 \begin{equation}
-    \calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^T \Gamma P(y_t;\theta) \mathbf{1}_N,
+    \calL_{\text{HMM}}(\theta,\Gamma;y) = \delta P(y_1;\theta) \prod_{t=2}^\top \Gamma P(y_t;\theta) \mathbf{1}_N,
     \label{eqn:like_marginal}
@@
 %
-where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and $P(y_t;\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry $f^{(i)}(y_t; \theta^{(i)})$.
-This equation can be evaluated in $\mathcal{O}(T)$ time using the well-known \textit{forward algorithm} \citep{Zucchini:2016}.
+where $\mathbf{1}_N$ is an $N$-dimensional column vector of ones and $P(y_t;\theta)$ is an $N \times N$ diagonal matrix with the $(i,i)^{th}$ entry $f^{(i)}(y_t; \theta^{(i)})$.
+This expression can be evaluated in $\mathcal{O}(T)$ time using the well-known forward algorithm \citep{Zucchini:2016}.
 
-We take a frequentest approach in this paper, which involves obtaining maximum likelihood estimates $\{\hat \theta, \hat \Gamma\}$ from Equation (\ref{eqn:like_marginal}).
-The estimated probability of each hidden state $\hat{\Pr}(X_t = i|Y = y)$ can then be calculated using $\{\hat \theta, \hat \Gamma\}$ and the \textit{forward-backward algorithm} \citep{Zucchini:2016}.
+We take a frequentist approach in this article, which involves obtaining maximum likelihood estimates $\{\hat \theta, \hat \Gamma\}$ from Equation (\ref{eqn:like_marginal}).
+The estimated probability of each hidden state $\hat{\Pr}(X_t = i \mid Y = y)$ can then be calculated using $\{\hat \theta, \hat \Gamma\}$ and the forward--backward algorithm \citep{Zucchini:2016}.
 Alternatively, it is reasonable to use a Bayesian approach, where Equation (\ref{eqn:like}) is combined with a prior distribution to obtain a posterior distribution over $X$, $\theta$, and $\Gamma$.
@@
 
-Following \citet{Barajas:2017}, we reparameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows sum to 1:
+Following \citet{Barajas:2017}, we reparameterize the $N \times N$ transition probability matrix $\Gamma$ such that the entries of the matrix are forced to be non-negative and the rows sum to one:
 
-\[
+\begin{equation*}
     \Gamma_{ij} = \frac{\exp(\eta_{ij})}{\sum_{k=1}^N \exp(\eta_{ik})}, 
-\]
+\end{equation*}
 %
@@
 This formulation simplifies likelihood maximization by removing constraints in the optimization problem.
-We assume there are no covariate effects, but similarly to the work of \citet{DeSouza:2017} for independent states and \citet{Adam:2019} for Markovian states, one could incorporate covariates into $\Gamma$ by setting $\eta_{ij}(z_t) = \beta_{ij}^T z_t$ for $i \neq j$, where $z_t$ is a column vector of known covariates and $\beta_{ij}$ is a column vector of unknown regression coefficients.
+We assume there are no covariate effects, but similar to the work of \citet{DeSouza:2017} for independent states and \citet{Adam:2019} for Markovian states, one could incorporate covariates into $\Gamma$ by setting $\eta_{ij}(z_t) = \beta_{ij}^\top z_t$ for $i \neq j$, where $z_t$ is a column vector of known covariates and $\beta_{ij}$ is a column vector of unknown regression coefficients.
 For simplicity, we will continue to use $\Gamma$ in our notation, suppressing the reparameterization in terms of $\eta$.
@@
 
-A conditionally auto-regressive hidden Markov model, or CarHMM \citep{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process.
-Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $\{X_1,\ldots,X_T\}$ each taking integer values between $1$ and $N$.
-A CarHMM also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
-Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $\{X_1,\ldots, X_T\}$ and $\{Y_1,\ldots, Y_{t-1}\}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$.
-The first emission $Y_1$ is treated as a fixed initial value which does not depend upon $X_1$.
-We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$.
+The CarHMM \citep{Lawler:2019} is a generalization of the HMM, which explicitly models autocorrelation in the observation sequence beyond the correlation induced by the hidden state process.
+Like the traditional HMM, the CarHMM is made up of a Markov chain of unobserved states $\{X_1,\ldots,X_T\}$, each taking integer values between $1$ and $N$.
+The CarHMM also has a transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$.
+Unlike the traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $\{X_1,\ldots, X_T\}$ and $\{Y_1,\ldots, Y_{t-1}\}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$.
+The first emission $Y_1$ is treated as a fixed initial value that does not depend upon $X_1$.
+We denote the conditional density or probability mass function of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ by $f^{(i)}(\cdot \mid y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot \mid y_{t-1})$.
 %
-As a concrete example, if $Y_t$ is a scalar, then one may assume that $Y_t$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:
+As a concrete example, if $Y_t$ is a scalar, then one may assume that $Y_t$ given $X_t = i$ is normally distributed with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where
 
-\begin{align}
-\label{eqn:carhmm}
-    \begin{split}
-    \mathbb{E}(Y_{t}|Y_{t-1} = y_{t-1},X_t=i) &= \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}, \\
-    \mathbb{V}(Y_t| Y_{t-1} = y_{t-1},X_t=i) &= (\sigma^{(i)})^2.
-    \end{split}
-\end{align}
+\begin{equation}
+    \mathbb{E}(Y_{t} \mid Y_{t-1} = y_{t-1},X_t=i) = \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)},
+    \label{eqn:carhmm1}
+\end{equation}
+and
+\begin{equation}
+    \mathbb{V}(Y_t \mid Y_{t-1} = y_{t-1},X_t=i) = (\sigma^{(i)})^2.
+    \label{eqn:carhmm2}
+\end{equation}
 %
-A CarHMM which follows Equation (\ref{eqn:carhmm}) can be viewed as a discrete time version of a state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}.
+A CarHMM following Equations (\ref{eqn:carhmm1}) and (\ref{eqn:carhmm2}) can be viewed as a discrete time version of a state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}.
 This follows in the same way that an AR(1) process is the discrete-time version of a traditional Ornstein-Uhlenbeck process.
 
-As previously, the likelihood of $\theta$ and $\Gamma$ corresponding to a CarHMM can be calculated using the forward algorithm. If $y$ is the sequence of observations, then
+As above, the likelihood of $\theta$ and $\Gamma$ corresponding to the CarHMM can be calculated using the forward algorithm. If $y$ is the sequence of observations, then
 
 \begin{equation*}
-    \calL_{\text{CarHMM}}(\theta,\Gamma;y) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\theta) \mathbf{1}_N,
+    \calL_{\text{CarHMM}}(\theta,\Gamma;y) = \delta \prod_{t=2}^\top \Gamma P(y_t \mid y_{t-1};\theta) \mathbf{1}_N,
     \label{eqn:CarHMM_likelihood}
@@
 %
-$P(y_t|y_{t-1};\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.
-Figure \ref{fig:models} shows a graphical representation of the dependence structure of a CarHMM.
+$P(y_t \mid y_{t-1};\theta)$ is an $N \times N$ diagonal matrix with the $(i,i)^{th}$ entry equal to $f^{(i)}(y_t \mid y_{t-1}; \theta^{(i)})$.
+Figure \ref{fig:models} shows a graphical representation of the dependence structure of the CarHMM.
 
@@
 
-A hierarchical hidden Markov model, or HHMM, accounts for processes occurring simultaneously at different scales by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs.
-The coarse-scale model is either an HMM and CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, where $\{X_1, \ldots, X_T\}$ is an unobserved Markov chain with $N$ possible states and $\{Y_1, \ldots, Y_T\}$ is the set of corresponding observations with state-dependent parameters $\theta^{(i)}$ for $i = 1,\ldots,N$.
+An HHMM accounts for processes occurring simultaneously at different scales by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs, as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}.
+Recall that $\{X_1, \ldots, X_T\}$ is an unobserved Markov chain with $N$ possible states, and $\{Y_1, \ldots, Y_T\}$ is the set of corresponding observations with state-dependent parameters $\theta^{(i)}$ for $i = 1,\ldots,N$.
 %
-In the hierarchical setting, however, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X^*_t \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$.
-In turn, $X^*_t$ emits a sequence of fine-scale observations $Y^*_t \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$.
-For each curve $t$, the fine-scale process $\{X^*_t, Y^*_t\}$ follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$.
-If $X_t = i$, then $X^*_{t,t^*}$ represents one of $N^{*(i)}$ possible fine-scale behaviours associated with coarse-scale hidden state $i$.
-Further, the components of $X^*_t$ make up a Markov chain with an $N^{*(i)} \times N^{*(i)}$ probability transition matrix $\Gamma^{*(i)}$ and an initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain.
-The distribution of $Y^*_{t,t^*}$ given $Y^*_{t,t^*-1} = y^*_{t,t^*-1}$, $X^*_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$.
-We denote the set of fine-scale emission parameters corresponding to $X_t=i$ as $\theta^{*(i)}=\big\{\theta^{*(i,1)}, \ldots, \theta^{*\left(i,N^{*(i)}\right)}\big\}$.
-In summary:
+In the hierarchical setting, each state $X_t$ also emits a sequence of fine-scale unobserved states, $X^*_t = \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$.
+In turn, $X^*_t$ emits a sequence of fine-scale observations $Y^*_t = \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$.
+For each curve $t$, the parameters of the HMM (or CarHMM) describing the fine-scale process $\{X^*_t, Y^*_t\}$ depend on the value of $X_t$.
+If $X_t = i$, then $X^*_{t,t^*}$ represents one of $N^{*(i)}$ possible fine-scale behaviours associated with the coarse-scale hidden state $i$.
+Further, the components of $X^*_t$ make up a Markov chain with an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$ and an initial distribution $\delta^{*(i)}$ which we assume is equal to the stationary distribution of the chain.
+The distribution of $Y^*_{t,t^*}$ given $Y^*_{t,t^*-1} = y^*_{t,t^*-1}$, $X^*_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has a density or probability mass function denoted by $f^{*(i,i^*)}\left(\cdot \mid y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot \mid y^*_{t,t^*-1})$.
+We denote the set of fine-scale emission parameters corresponding to $X_t=i$ by $\theta^{*(i)}=\big\{\theta^{*(i,1)}, \ldots, \theta^{*\left(i,N^{*(i)}\right)}\big\}$.
+In summary,
 
-\begin{gather*}
-    \{Y, X\} \text{ follows a (Car)HMM with } \Gamma \in \bbR^{N \times N}, \\
+\begin{enumerate}
+    \item $\{Y, X\}$ follows a (Car)HMM with $\Gamma \in \bbR^{N \times N}$,
     %
-    (Y_t | Y_{t-1} = y_{t-1}, X_t = i) \text{ has density } f^{(i)}(\cdot|y_{t-1};\theta^{(i)}), \\
+    \item $(Y_t  \mid  Y_{t-1} = y_{t-1}, X_t = i)$ has the density $f^{(i)}(\cdot \mid y_{t-1};\theta^{(i)})$,
     %
-    \{Y^*_t,X^*_t | X_t = i\} \text{ follows a (Car)HMM with } \Gamma^{*(i)} \in \bbR^{N^{*(i)} \times N^{*(i)}}, \\
+    \item $\{Y^*_t,X^*_t  \mid  X_t = i\}$ follows a (Car)HMM with $\Gamma^{*(i)} \in \bbR^{N^{*(i)} \times N^{*(i)}}$, and
     %
-    (Y^*_{t,t^*} | Y^*_{t,t^*-1} = y^*_{t,t^*-1}, X^*_{t,t^*} = i^*, X_t = i) \text{ has density } f^{*(i,i^*)}(\cdot|y^{*}_{t,t^*-1};\theta^{(i,i^*)}).
-\end{gather*}
+    \item $(Y^*_{t,t^*}  \mid  Y^*_{t,t^*-1} = y^*_{t,t^*-1}, X^*_{t,t^*} = i^*, X_t = i)$ has the density $f^{*(i,i^*)}(\cdot \mid y^{*}_{t,t^*-1};\theta^{(i,i^*)})$.    
+\end{enumerate}
+%
 Given the coarse-scale hidden state sequence $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X^*_T, Y^*_T\}$, and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another.
@@
 %
-Let $y$ be the sequence of observed coarse-scale emissions and
-$y^* \equiv \{y^*_1, \ldots,y^*_T\}$ be the collection of $T$ observed fine-scale emission vectors.
+Let $y = \{y_1, \ldots,y_T\}$ be the sequence of observed coarse-scale emissions and
+$y^* = \{y^*_1, \ldots,y^*_T\}$ be the collection of $T$ observed fine-scale emission vectors.
 %
-In addition, let $\theta^* \equiv \{\theta^{*(1)}, \ldots, \theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* \equiv \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices.
+In addition, let $\theta^* = \{\theta^{*(1)}, \ldots, \theta^{*(N)}\}$ denote the collection of all fine-scale emission parameters and $\Gamma^* = \{\Gamma^{*(1)}, \ldots, \Gamma^{*(N)}\}$ denote the collection of all fine-scale transition probability matrices.
 The likelihood of the observed data is then
@@
 \begin{equation}
-    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N,
+    \calL_{\text{HHMM}}(\theta,\theta^*,\Gamma,\Gamma^*;y,y^*) = \delta P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*) \prod_{t=2}^\top \Gamma P_m(y_t,y_t^* \mid y_{t-1};\theta,\theta^*,\Gamma^*) \mathbf{1}_N,
     \label{eqn:HHMM_likelihood}
@@
 %
-where $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix whose exact structure depends upon the coarse- and fine-scale models.
-If the coarse-scale model is an HMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ for $t \geq 2$ both have $(i,i)^{th}$ entries equal to
+where $P_m(y_t,y_t^* \mid y_{t-1};\theta,\theta^*,\Gamma^*)$ is an $N \times N$ diagonal matrix whose exact structure depends upon the coarse- and fine-scale models.
+If the coarse-scale model is an HMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ and $P_m(y_t,y_t^* \mid y_{t-1};\theta,\theta^*,\Gamma^*)$, for $t \geq 2$, both have $(i,i)^{th}$ entries equal to
 $f^{(i)}(y_t)\calL_{\text{fine}}\left(\theta^{*(i)},
 \Gamma^{*(i)};y_t^*\right)$. 
-If the coarse-scale model is a CarHMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ has $(i,i)^{th}$ entry equal to $\calL_{\text{fine}}\left(\theta^{*(i)},
-\Gamma^{*(i)};y_1^*\right)$ and $P_m(y_t,y_t^*|y_{t-1};\theta,\theta^*,\Gamma^*)$ for $t \geq 2$ has $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1})\calL_{\text{fine}}\left(\theta^{*(i)},
+If the coarse-scale model is a CarHMM, $P_m(y_1,y_1^*;\theta,\theta^*,\Gamma^*)$ has its $(i,i)^{th}$ entry equal to $\calL_{\text{fine}}\left(\theta^{*(i)},
+\Gamma^{*(i)};y_1^*\right)$ and $P_m(y_t,y_t^* \mid y_{t-1};\theta,\theta^*,\Gamma^*)$, for $t \geq 2$, has its $(i,i)^{th}$ entry equal to $f^{(i)}(y_t \mid y_{t-1})\calL_{\text{fine}}\left(\theta^{*(i)},
 \Gamma^{*(i)};y_t^*\right)$.
@@
 To handle these additional fine-scale structures, we recommend replacing $Y^*_t = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize any non-Markovian behaviour.
-To maintain the temporal structure of the fine-scale process, local summary statistics can be calculated from a moving window with stride length $h$ over the elements of $Y^*_t$.
+To maintain the temporal structure of the fine-scale process, local summary statistics can be calculated from a moving window with the stride length $h$ over the elements of $Y^*_t$.
 Stride length refers to the distance between the first element of consecutive windows, so a stride length of $h$ implies that the first window starts at $Y^*_{t,1}$, the second at $Y^*_{t,1+h}$, and so on.
 Subject matter experts are often required to determine the specific summary statistics employed as well as the optimal window size and stride length of the moving window.
-Larger stride lengths result in a larger loss of information but also reduce the dimension of the fine-scale process, which allows for faster model fitting.
+Larger stride lengths result in a larger loss of information but also reduce the dimensionality of the fine-scale process, which allows for faster model fitting.
 In addition, setting the stride length equal to the window size avoids artificial residual correlation arising from overlapping windows.
 
-We are interested in fine-scale summary statistics that are interpretable for practitioners and can effectively differentiate between fine-scale behavioural states.
-As a result, we use the discrete Fourier transform (DFT) of a moving forward window with width $h$ and stride $h$ across $Y^*_t$:
+We are interested in fine-scale summary statistics that are interpretable to practitioners and can effectively differentiate between fine-scale behavioural states.
+As a result, we use the discrete Fourier transform (DFT) of a moving forward window with a width of $h$ and a stride of $h$ across $Y^*_t$:
 
 \begin{align}
-    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right)
+    \mathrm{DFT}\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right),
     \label{eq:DFTdef}
@@
 for $t^* = 1,h+1,2h+1,\ldots$ and $k = 0, 1, \ldots, h-1$, where $i = \sqrt{-1}$.
-If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise.
-We omit the final window if $t^*+h-1$ exceeds $T^*_t$, denote the total number of windows as $\tilde T^*_t$, and index each window with $\tilde{t}^* = 1,\ldots,\tilde T^*_t$.
-Next, we calculate transformed observations $\Z_{t,\tilde{t}^*} \equiv \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$:
+If $Y^*_{t,t^*}$ is a vector, then the DFT is taken component-wise.
+We omit the final window if $t^*+h-1$ exceeds $T^*_t$, denote the total number of windows by $\tilde T^*_t$, and index the windows with $\tilde{t}^* = 1,\ldots,\tilde T^*_t$.
+Next, we calculate the transformed observations $\Z_{t,\tilde{t}^*} = \{\Zone_{t,\tilde{t}^*},\Ztwo_{t,\tilde{t}^*}\}$ as
 
@@
     \label{eqn:z}
-    \Zone_{t,\tilde{t}^*} \equiv \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n}, \quad \Ztwo_{t,\tilde{t}^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|_2^2,
+    \Zone_{t,\tilde{t}^*} = \frac{1}{h}\sum_{n=1}^{h}Y^*_{t,h(\tilde{t}^*-1)+n} \enspace \text{and} \enspace \Ztwo_{t,\tilde{t}^*} = \sum_{k=1}^{\tilde{\omega}}\Big|\Big|\mathrm{DFT}\{Y^*_{t,h(\tilde{t}^*-1)+1},\ldots, Y^*_{t,h\tilde{t}^*}\}(k)\Big|\Big|_2^2,
 \end{equation}
@@
 We denote these unobserved states as $\tilde{X}^*_t = \big\{\tilde{X}^*_{t,1},\ldots, \tilde{X}^*_{t,\tilde{T}^*_t}\big\}$.
-The fine-scale probability transition matrices $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ correspond to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$, as do all fine-scale model assumptions (e.g., conditional independence, auto-regressive structure between observations, etc.) Fine-scale model selection and validation should therefore be adjusted accordingly.
+The fine-scale transition probability matrices $\Gamma^{*(i)}$ and probability density functions $f^{*(i,i^*)}$ correspond to $\tilde{X}^*_t$ and $\Z_t$ rather than $X^*_t$ and $Y^*_t$, as do all fine-scale model assumptions (e.g., conditional independence, autoregressive structure between observations, etc.) Fine-scale model selection and validation should therefore be adjusted accordingly.
 
-The likelihood of this model is identical to that of the original HMM or CarHMM as defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$.
-To clearly differentiate models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT.
+The likelihood of this model is identical to that of the original HMM or CarHMM defined in Sections \ref{subsec:HMM} and \ref{subsec:CarHMM}, but $Y^*$ is replaced with $\Z$ and $X^*$ is replaced with $\tilde{X^*}$.
+To clearly differentiate the models, we refer to an HMM with $\Z$ as observations and $\tilde{X^*}$ as hidden states as an HMM-DFT.
 Figure \ref{fig:models} displays the dependence structure of a fine-scale HMM-DFT.
@@
 While we are primarily interested in classifying and describing coarse- and fine-scale behaviour, some researchers may be less interested in interpretability and more interested in predicting future functional data.
-To this end, \citet{Aue:2015} and \citet{Gao:2019} both use functional principle component analysis (FPCA) and derive bounds on the reconstruction error caused by dimension reduction.
+To this end, \citet{Aue:2015} and \citet{Gao:2019} both use functional principal component analysis (FPCA) and derive bounds on the reconstruction error caused by dimension reduction.
 While these papers do not account for state-switching behaviour between curves or windows, deriving similar error bounds for FPCA within an HMM framework appears to be a promising direction for future research.
@@
 Traditional HHMMs treat both the coarse-scale and the fine-scale processes as realizations of an HMM or CarHMM. 
-However, the fine-scale observations of a particular dive $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods.
+However, the fine-scale observations of a particular dive $Y^*_t$ can be modelled using a large variety of parametric models that admit easy-to-compute likelihoods or penalized likelihoods.
 As such, the fine-scale HMM likelihood term $\calL_{\text{fine}}$ in Equation (\ref{eqn:HHMM_likelihood}) can be replaced by the likelihood of a general fine-scale model whose parameters depend upon the coarse-scale hidden state.
@@
 If the fine-scale model is a simple Poisson process, then this approach is equivalent to a Markov-modulated Poisson process \citep{Fischer:1993}.
-The fine-scale process can also be modelled similarly to \citet{Langrock:2018}, who use B-splines to model the emission distribution of an HMM.
-This non-parametric approach uses a penalized likelihood term which can easily replace the usual fine-scale likelihood term in Equation (\ref{eqn:HHMM_likelihood}).
-Another class of fine-scale models is the set of continuous time methods such as a continuous-time HMM (CTHMM) \citep{Liu:2015} or a state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}. 
-A continuous-time HMM may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}.
-\citet{Xu:2018} model high-frequency biologging accelerometer data of individuals by incorporating a CTHMM into a hierarchical model similar to ours.
-However, they assume that individuals are partitioned into subgroups a priori whereas we use an HMM to infer the coarse-scale hidden states.
+The fine-scale process can also be modelled similarly to \citet{Langrock:2018}, which uses B-splines to model the emission distribution of an HMM.
+This nonparametric approach uses a penalized likelihood term that can easily replace the usual fine-scale likelihood term in Equation (\ref{eqn:HHMM_likelihood}).
+Another class of fine-scale models is the set of continuous time methods such as the continuous-time HMM (CTHMM) \citep{Liu:2015} and the state-switching Ornstein-Uhlenbeck process \citep{Michelot:2019}.
+A  CTHMM may be appropriate if observations are not equally spaced in time \citep{Liu:2015}.
+\citet{Xu:2018} models high-frequency biologging accelerometer data from individuals by incorporating a CTHMM into a hierarchical model similar to ours.
+However, they assume that individuals are partitioned into subgroups a priori whereas we use an HMM to infer coarse-scale hidden states.
 
@@
 A myriad of possible models can be built using this framework, but these models can quickly become complicated and computationally expensive to fit.
-Therefore, models should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.
+Therefore, models should be constructed with care to achieve an adequate fit to the data while avoiding overfitting and high computational costs.

### 2021-10-25 Report/Sections/results.tex
address copy editor's comments (main text)

@@
 Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}.
-HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., \citet{DeRuiter:2017} and \citet{Isojunno:2017}).
+HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., DeRuiter et al., 2017; Isojunno et al.,2017)\nocite{DeRuiter:2017,Isojunno:2017}.
 For killer whales, we are interested in categorizing different diving behaviours and identifying potential foraging dives.
 Northern resident killer whales feed almost exclusively on calorie-rich Chinook salmon (\textit{Oncorhynchus tshawytscha}) \citep{Ford:2006}, which typically occur deeper and are less numerous than smaller types of salmon \citep{Ford:2009}.
-Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook \citep{Williams:2009,Noren:2011,Wright:2017}.
+Northern resident killer whales therefore must expend significant amounts of energy to capture Chinook salmon \citep{Williams:2009,Noren:2011,Wright:2017}.
 Acceleration data can be used to estimate an animal's energy expenditure \citep{Green:2009,Wilson:2019}, but the animal's behavioural state must be accounted for in order to obtain accurate estimates \citep{Dot:2016}.
@@
 
-The data we use were collected on September 2, 2019 from 12:49 pm to 6:06 pm PDT and consist of depth and acceleration over time.
-Observations were collected at a rate of 50 Hz using a CATs biologger (Customizable Animal Tracking Solutions, {\em{www.cats.is}}).
+The data we use were collected on September 2, 2019 from 12:49 p.m.
+to 6:06 p.m.
+PDT and consist of depth and acceleration over time.
+Observations were collected at a rate of 50 Hz using a CATS biologger (Customizable Animal Tracking Solutions, {\em{www.cats.is}}).
 Acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, and right/left).
-Tri-axial acceleration readings are common in these types of tags and are often used to infer animal behaviour such as foraging \citep{Cade:2017,Fehlmann:2017,Wright:2017}.
-The act of attaching and detaching the tag caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations taken during these time periods are ignored.
-There were also periods of time when the tag failed to record observations, resulting in data gaps between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm.
-To preprocess the data, we smooth the depth and acceleration curves by taking a moving average within a window of $1/10^{th}$ of a second.
-We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds.
+Tri-axial acceleration readings are common in these types of tags and are often used to infer animal behaviour such as foraging \citep{Fehlmann:2017,Wright:2017,Cade:2017}.
+The act of attaching and detaching the tag caused anomalous behaviour before 1:20 p.m.
+and after 6:00 p.m., so observations taken during these time periods are ignored.
+There were also periods of time when the tag failed to record observations, resulting in data gaps between 2:25 p.m.
+and 2:37 p.m.
+and between 4:07 p.m.
+and 5:07 p.m.
+To preprocess the data, we smooth the depth and acceleration curves by taking a moving average within a window of one-tenth of a second.
+We then define a killer whale ``dive" as any continuous interval of data that occurs below 0.5 $m$ in depth and lasts for at least 10 $s$.
 Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}.
@@
 Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process.
-Specifically, the observed coarse-scale observations $y = \big\{y_1,\ldots,y_{267}\big\}$ make up a sequence of dive durations in seconds, and the coarse-scale hidden states $x = \big\{x_1,\ldots,x_{267}\big\}$ represent the corresponding dive types.
-For dive $t$, the fine-scale observations are contained in $y^*_{t} = \big\{y^*_{t,1},\ldots,y^*_{t,T^*_t} \big\}$, which is a sequence containing the within-dive acceleration data in units of meters per second squared.
+Specifically, the observed coarse-scale observations $y = \big\{y_1,\ldots,y_{267}\big\}$ make up a sequence of dive durations in $s$, and the coarse-scale hidden states $x = \big\{x_1,\ldots,x_{267}\big\}$ represent the corresponding dive types.
+The fine-scale observations for dive $t$ are measured in units of $m/s^2$ and are contained in $y^*_{t} = \big\{y^*_{t,1},\ldots,y^*_{t,T^*_t} \big\}$.
 The fine-scale hidden states of dive $t$, $x^*_{t} = \big\{x^*_{t,1},\ldots,x^*_{t,T^*_t} \big\}$, represent the subdive behavioural states of the killer whale.
-The collection of all acceleration data is denoted as $y^* = \big\{y^*_1,\ldots,y^*_{267}\big\}$ and the collection of all unknown subdive behaviours is denoted as $x^* = \big\{x^*_1,\ldots,x^*_{267}\big\}$.
+The collection of all acceleration data is denoted by $y^* = \big\{y^*_1,\ldots,y^*_{267}\big\}$ and the collection of all unknown subdive behaviours is denoted by $x^* = \big\{x^*_1,\ldots,x^*_{267}\big\}$.
 
@@
 The primary goal of this case study is to jointly estimate the dive types and subdive behaviours of this killer whale, so we only consider models with both a coarse-scale and fine-scale component to describe the kinematic data.
-Defining a suitable hierarchical model involves selecting an appropriate number of hidden states, model structure, and emission distributions for both the coarse- and fine-scale observations.
+Defining a suitable hierarchical model involves selecting an appropriate number of hidden states, model structure, and set of emission distributions for both the coarse- and fine-scale observations.
 
@@
 This type of visualization is known as a lag plot.
-If the emission distributions of the hidden states are well-separated, a lag plot should reveal $N$ distinct patterns, where each pattern corresponds to one dive type \citep{Lawler:2019}.
-This is unfortunately not the case for our killer whale data, as there is one cluster of data centred at approximately $y_t = y_{t-1} = 30$ seconds.
-However, longer dives appear to be characterized by bouts of less ``wiggly" behaviour in the acceleration data compared to shorter dives, so we choose $N = 2$ to differentiate these dive types.
-The absence of a more principled method to select $N$ highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
+If the emission distributions of the hidden states are well separated, a lag plot should reveal $N$ distinct patterns, where each pattern corresponds to one dive type \citep{Lawler:2019}.
+This is unfortunately not the case for our killer whale data, as there is one cluster of data centred at approximately $y_t = y_{t-1} = 30$ $s$.
+However, longer dives appear to be characterized by bouts of less-``wiggly" behaviour in the acceleration data compared to shorter dives, so we choose $N = 2$ to differentiate these dive types.
+The absence of a more-principled method to select $N$ highlights the importance of model validation techniques in lieu of information criteria (see Section \ref{subsec:model_validation}).
 %
-Prior to fitting the model, lag plots reveal no significant auto-correlation between dive duration observations (see Figure 1 of of supplement A), and visual inspection shows no obvious complicated dependence.
+Lag plots reveal no significant autocorrelation between dive duration observations (see Figure 1 of Supplement A) and visual inspection shows no obvious complicated dependence.
 Therefore, we select a simple HMM to model the coarse-scale process since neither a CarHMM nor a moving-window transformation is called for.
 %
-Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows a Gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$:
-
-\[\mathbb{E}(Y_t|X_t = i) = \mu^{(i)}, \qquad \mathbb{V}(Y_t|X_t = i) = \left(\sigma^{(i)}\right)^2.\]
-%
-This is consistent with previous studies, including that of \citet{Barajas:2017}. 
+Given that dive $t$ is of type $i$, we assume that the dive duration $Y_t$ follows a gamma distribution with unknown parameters $\mu^{(i)}$ and $\sigma^{(i)}$: $\mathbb{E}(Y_t \mid X_t = i) = \mu^{(i)}$ and $\mathbb{V}(Y_t \mid X_t = i) = \left(\sigma^{(i)}\right)^2$.
+This is consistent with previous studies, including \citet{Barajas:2017}.
 %
@@
 We then select a model corresponding to the fine-scale observations of acceleration.
-Similarly to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$ subdive states.
+Similar to the coarse model, we rely on lag plots and visual inspection to select $N^*=3$ subdive states.
 Although $N^*$ is selected heuristically, we test the validity of this model in Section \ref{subsec:model_validation}.
@@
 In contrast to the coarse-scale observations, the fine-scale acceleration data exhibit significant sinusoidal behaviour.
-Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with a window size of $h=100$ (two seconds) and a maximum frequency of $\tilde{\omega}=10$ (5 Hz).
+Thus, we transform each fine-scale observation sequence $y_t^*$ into $\z_t$ using Equation (\ref{eqn:z}) with a window size of $h=100$ (2 $s$) and a maximum frequency of $\tilde{\omega}=10$ (5 $Hz$).
 We then have that $\z_{t,\tilde t^*} = \{\zone_{t,\tilde t^*},\ztwo_{t,\tilde t^*}\}$, where $\zone_{t,\tilde t^*}$ is a three-dimensional vector of component-wise average acceleration and $\ztwo_{t,\tilde t^*}$ is a scalar describing the ``wiggliness" of a particular window.
-Even after transforming the raw acceleration data, there is still strong auto-correlation within each component of $\zone_{t,\tilde t^*}$ prior to fitting the model (see Figure 1 of of supplement A).
-Therefore, we choose a CarHMM as defined in Section 2.2 as the fine-scale model.
+Even after transforming the raw acceleration data, there is still strong autocorrelation within each component of $\zone_{t,\tilde t^*}$ (see Figure 1 of Supplement A).
+Therefore, we choose a CarHMM as the fine-scale model.
 
-We then select the specific emission distributions of $\Z_{t,\tilde t^*}$ for all dive types and subdive states.
+We then select the specific emission distribution of $\Z_{t,\tilde t^*}$ for all dive types and subdive states.
 First, we assume that $\Ztwo_{t,\tilde t^*}$ and all three components of $\Zone_{t,\tilde t^*}$ are independent of one another when conditioned on the dive types and subdive states.
-To reduce model complexity, we also assume that the three sets of fine-scale emission parameters are shared across the two dive types $\left(\theta^{*(1,i^*)} = \theta^{*(2,i^*)} \equiv \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$.
+To reduce model complexity, we also assume that the three sets of fine-scale emission parameters are shared across the two dive types $\left(\text{i.e., }\theta^{*(1,i^*)} = \theta^{*(2,i^*)} = \theta^{*(\cdot,i^*)} \text{ for } i^* = 1,2,3\right)$.
 This implies that the subdive states within dive type 1 have the same interpretation as those within dive type 2.
 To specify the emission distribution of $\Zone_{t,\tilde t^*}$, consider the sequence $\{\Zone_{t,1},\ldots,\Zone_{t,T^*_t}\}$ for a particular dive $t$.
-We assume that each of the three components of this sequence are Normally distributed as in Equation (\ref{eqn:carhmm}) and all components are independent of one another when conditioned on the subdive states $\tilde X^*_t = \{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T^*_t}\}$.
-Each component is assumed to have its own mean and variance parameters, but all components share the same auto-correlation parameter.
-Thus the distribution of $\Zone_{t,\tilde t^*}$ given $X^*_{t,\tilde t^*} = i^*$ has parameters $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
-To specify the emission distribution of $\Ztwo_{t,\tilde t^*}$, we assume that given $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ follows a Gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
+We assume that each of the three components of this sequence are normally distributed as in Equations (\ref{eqn:carhmm1}) and (\ref{eqn:carhmm2}), and we assume that all components are independent of one another when conditioned on the subdive states $\tilde X^*_t = \{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T^*_t}\}$.
+Each component is assumed to have its own mean and variance parameters, but all components share the same autocorrelation parameter.
+Thus, the distribution of $\Zone_{t,\tilde t^*}$ given $X^*_{t,\tilde t^*} = i^*$ has the parameters $\mathbf{\mu}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, $\mathbf{\sigma}_A^{*(\cdot,i^*)} \in \mathbb{R}^3$, and $\phi_A^{*(\cdot,i^*)} \in [0,1]$.
+To specify the emission distribution of $\Ztwo_{t,\tilde t^*}$, we assume that given $\tilde X^*_{t,\tilde t^*} = i^*$, $\Ztwo_{t,\tilde t^*}$ follows a gamma distribution parameterized by its mean $\mu_W^{*(\cdot,i^*)}$ and standard deviation $\sigma_W^{*(\cdot,i^*)}$.
 In addition, $\Ztwo_{t,1},\ldots,\Ztwo_{t,T^*_t}$ are assumed to be independent of one another given the subdive state sequence $\big\{\tilde X_{t,1}, \ldots, \tilde X_{t,T_t^*}\big\}$.
-We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the auto-correlation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
+We do not include $\Ztwo_{t,\tilde t^*-1}$ in the distribution of $\Ztwo_{t,\tilde t^*}$ because the autocorrelation evident from the lag plot is not severe and may be explained by subsequent observations occurring within the same subdive state.
 
-In total, the parameters to estimate are
-
-\begin{gather*}
-    \Gamma, \qquad \Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\} \qquad \text{(probability transition matrices)}, \\
-    %
-    \theta = \{\mu^{(1)},\sigma^{(1)},\mu^{(2)},\sigma^{(2)}\} \qquad \text{($Y$ emission parameters), and} \\
-    %
-    \theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}  \qquad \text{($\Z$ emission parameters), where} \\
-    %
-    \theta^{*(\cdot,i^*)} =  \{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)},\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}.
-\end{gather*}
-%
+In total, the parameters to be estimated are the transition probability matrices $\Gamma$ and $\Gamma^{*} = \{\Gamma^{*(1)},\Gamma^{*(2)}\}$, the coarse-scale emission parameters $\theta = \{\mu^{(1)},\sigma^{(1)},\mu^{(2)},\sigma^{(2)}\}$, and the fine-scale emission parameters $\theta^* = \{\theta^{*(\cdot,1)},\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}\}$, where $\theta^{*(\cdot,i^*)} = \{\mu_A^{*(\cdot,i^*)},\sigma_A^{*(\cdot,i^*)},\phi_A^{*(\cdot,i^*)},\mu_W^{*(\cdot,i^*)},\sigma_W^{*(\cdot,i^*)}\}$.
 Recall that $\theta^{*(\cdot,i^*)}$ is the set of parameters describing the distribution of $\Z_{t,\tilde t^*}$ conditioned on $\tilde X^*_{t,\tilde t^*} = i^*$.
 %
-We refer to this final model as the \textbf{CarHHMM-DFT} since it includes a CarHMM, hierarchical HMM, and DFT-based transformation.
+We refer to this final model as the CarHHMM-DFT since it includes a CarHMM, HHMM, and DFT-based transformation.
 The likelihood of this model is easily calculated using the forward algorithm and can be maximized with respect to the parameters above (see the appendix for details).
@@
 In addition to the CarHHMM-DFT, we consider three variations for comparison.
-As in the full model, each of the following models assume that all components of $\Z_{t,\tilde t^*}$ are conditionally independent of one another given the dive types and subdive states:
+As in the full model, each of the following models assume that all components of $\Z_{t,\tilde t^*}$ are conditionally independent of one another given the dive types and subdive states.
 \begin{enumerate}
-    \item An \textbf{HHMM-DFT}, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}), but models $\Z_{t,\tilde t^*}$ as emissions from a simple HMM rather than a CarHMM.
-    \item A \textbf{CarHHMM}, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,\tilde t^*}$ as emissions of a CarHMM.
-However, the ``wiggliness"  $\Ztwo_{t,\tilde t^*}$ is omitted from this model altogether.
-    \item A \textbf{CarHMM-DFT}, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Z_{t,\tilde t^*}$ as emissions of a CarHMM.
-This model assumes that there is only one dive type.
+    \item An HHMM-DFT, which models the coarse-scale observations with an HMM and transforms the fine-scale observations using Equation (\ref{eqn:z}) but models $\Z_{t,\tilde t^*}$ as emissions from a simple HMM rather than a CarHMM.
+    \item A CarHHMM, which models the coarse-scale observations with an HMM, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Zone_{t,\tilde t^*}$ as emissions of a CarHMM.
+    \item A CarHMM-DFT, which models the coarse-scale observations as an independent and identically distributed sequence of dives, transforms the fine-scale observations using Equation (\ref{eqn:z}), and models $\Z_{t,\tilde t^*}$ as emissions of a CarHMM.
 \end{enumerate}
 %
-Each of the three candidate models above leaves out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes there is no auto-correlation between fine-scale observations, the CarHHMM does not incorporate ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and thus does not distinguish between dive types.
+Each of the three candidate models above leaves out one important aspect of the full CarHHMM-DFT: the HHMM-DFT assumes there is no autocorrelation between fine-scale observations, the CarHHMM does not incorporate ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$, and the CarHMM-DFT lacks a hierarchical structure and assumes that there is only one dive type.
 
 We estimate the parameters of all four models using the data shown in Figure \ref{fig:data} and direct likelihood maximization using the SciPy package in Python \citep{Virtanen:2019}.
-Each model is fit 100 times using random initializations, and we keep the parameter estimates corresponding to the maximum likelihood of those 100 optimization routines \citep{Zucchini:2016}.
-Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory.
-Most, but not all, initializations converge to the same parameter estimates, highlighting the need to perform multiple optimizations with a variety of initial guesses.
-Plots of the likelihood surface near the maximum likelihood estimates are shown in Section 2 of supplement A.
+Each model is fit 100 times using random initializations, and we keep the parameter estimates corresponding to the maximum likelihood of each optimization routine \citep{Zucchini:2016}.
+Likelihood maximization is performed on the Cedar Compute Canada cluster with one CPU and 8 GB of dedicated memory.
+Most but not all initializations converge to the same parameter estimates, highlighting the need to perform multiple optimizations with a variety of initial guesses.
+Plots of the likelihood surface near the maximum likelihood estimates are shown in Section 2 of Supplement A.
 
@@
 
-We first report the results from the full CarHHMM-DFT in detail and assess the quality of the fit. We then compare these results with those from the other candidate models.
+We first report results from the full CarHHMM-DFT in detail and assess the quality of the fit. We then compare these results with those from the other candidate models.
 
 The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:emis}). 
-Dive type 1 corresponds to shorter, shallower dives which likely reflect resting, travelling, and to a lesser extent searching for prey.
-Dive type 2 is longer, deeper, and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives.
-No dive in this data set has a maximum depth deeper than 30 meters, and a study by \citet{Wright:2017} of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 meters.
+Dive type 1 corresponds to shorter and shallower dives that likely reflect resting, travelling, and to a lesser extent searching for prey.
+Dive type 2 is longer and deeper and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives.
+No dive in this data set has a maximum depth greater than 30 $m$, while \citet{Wright:2017}, in a study of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 $m$.
 However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and unpublished data collected by the authors suggest that prey-capture events can occur near the surface.
-Dive type 2 could also be associated with behaviours such as socializing that can take place several meters below the surface \citep{Tennessen:2019b}.
+Dive type 2 could also be associated with behaviours such as socializing, which can take place several meters below the surface \citep{Tennessen:2019b}.
 
@@
 Subdive state 1 has the smallest mean corresponding to $\Ztwo_{t,\tilde t^*}$ and the smallest standard deviation corresponding to $\Zone_{t,\tilde t^*}$.
-It also has the highest auto-correlation in $\Zone_{t,\tilde t^*}$.
+It also has the highest autocorrelation in $\Zone_{t,\tilde t^*}$.
 This implies less overall activity and more consistent acceleration compared to the other subdive states.
-Subdive state 2 has a mean ``wiggliness" $\big(\Ztwo_{t,\tilde t^*}\big)$ one order of magnitude higher than subdive state 1 and its acceleration has about twice the standard deviation compared to subdive state 1.
-The auto-correlation of acceleration is also slightly lower than subdive state 1.
-We therefore hypothesize that subdive state 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
-Finally, the mean of $\Ztwo_{t,\tilde t^*}$ and variance of $\Zone_{t,\tilde t^*}$ in subdive state 3 are both much higher than in the other two states, and the auto-correlation of $\Zone_{t,\tilde t^*}$ is also much lower.
+Subdive state 2 has a mean ``wiggliness" one order of magnitude higher than that of subdive state 1 and its acceleration has about twice the standard deviation compared to subdive state 1.
+The autocorrelation of acceleration is also slightly lower than in subdive state 1.
+We therefore hypothesize that subdive state 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of fluking in marine mammals \citep{Simon:2012}.
+Finally, the mean of $\Ztwo_{t,\tilde t^*}$ and variance of $\Zone_{t,\tilde t^*}$ in subdive state 3 are both much higher than in the other two states, and the autocorrelation of $\Zone_{t,\tilde t^*}$ is also much lower.
 This corresponds to vigorous swimming activity, especially as the killer whale ends a dive (see Figure \ref{fig:labeled_dives}).
 
-The estimated probability transition matrices and associated stationary distributions on the coarse scale are
+The estimated transition probability matrix and associated stationary distribution on the coarse scale (i.e., between dives) are
 
@@
 0.914 & 0.086
-\end{pmatrix} \text{ and }\]
-
-\[\hat \delta = \begin{pmatrix} 0.857 & 0.143 \end{pmatrix}\]
+\end{pmatrix}\]
 %
-for the transitions between dives. The estimated probability transition matrices and stationary distributions on the fine scale are
+and $\hat \delta = \left(0.857, 0.143 \right)$. The estimated transition probability matrices and stationary distributions on the fine scale are
 
@@
 
-\[\hat \delta^{*(1)} = \begin{pmatrix} 0.202 & 0.649 & 0.149 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.531 & 0.405 & 0.064 \end{pmatrix}\]
-%
-for dive types 1 and 2.
-In summary, about 86\% of dives are short dives of type 1, and the whale performs an average of 6.54 short type 1 dives before switching to dive type 2 and an average of 1.09 longer type 2 dives before switching back to dive type 1.
-This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive.
-Further, this killer whale is in the less active subdive state 1 only 20\% of the time during a dive of type 1 compared to 53\% of the time during a dive of type 2.
-Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to depth and holding their breath for long periods \citep{Williams:1999,Hastie:2006}.
-Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives, and Section 4 of supplement A also shows the probability of each dive type and subdive state given the data and the fitted model.
+$\hat \delta^{*(1)} = \left(0.202, 0.649, 0.149\right)$, and $\hat \delta^{*(2)} = \left( 0.531, 0.405, 0.064 \right)$ for dive types 1 and 2.
+In summary, about 86\% of dives are short dives of type 1.
+The whale performs an average of 6.54 short type 1 dives before switching to dive type 2 and an average of 1.09 longer type 2 dives before switching back to dive type 1.
+This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of which describe common bouts of short resting dives before a longer, more-energy-intensive deep dive.
+Further, this killer whale is in the less-active subdive state 1 only 20\% of the time during a dive of type 1, compared to 53\% of the time during a dive of type 2.
+Less-active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving to greater depths and holding their breath for long periods \citep{Williams:1999,Hastie:2006}.
+Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives.
+Section 4 of Supplement A also shows the probability of each dive type and subdive state given the data and the fitted model.
 
@@
 We use two visual tools to evaluate the CarHHMM-DFT: pseudoresidual plots and empirical histograms.
-The pseudoresidual of a coarse-scale observation $y_t$ is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T,\Z_1,\ldots,\Z_T\}/\{Y_t\}) \right)$ and the pseudoresidual of a fine-scale observation $\z_{t,\tilde t^*}$ is $\Phi^{-1} \left(Pr(\Z_{t,t^*} < \z_{t,\tilde t^*}|\{Y_1,\ldots,Y_T,\Z_1,\ldots,\Z_T\}/\{Y^*_{t,\tilde t}\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution.
-If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution.
-Histograms of the pseudoresiduals mostly support that the CarHHMM-DFT is well-specified.
+The pseudoresidual of a coarse-scale observation $y_t$ is $\Phi^{-1} \left(\mathrm{Pr}(Y_t < y_t \mid \{Y_1,\ldots,Y_T,\Z_1,\ldots,\Z_T\} \setminus \{Y_t\}) \right)$ and the pseudoresidual of a fine-scale observation $\z_{t,\tilde t^*}$ is $\Phi^{-1} \left(\mathrm{Pr}(\Z_{t,t^*} < \z_{t,\tilde t^*} \mid \{Y_1,\ldots,Y_T,\Z_1,\ldots,\Z_T\} \setminus \{Y^*_{t,\tilde t}\}) \right)$, where $\Phi$ is the cumulative distribution function of the standard normal distribution.
+If the model is correct, then all pseudoresiduals are independent and follow the standard normal distribution.
+Histograms of the pseudoresiduals mostly support that the CarHHMM-DFT is well specified.
 One exception is $\Ztwo_{t,\tilde t^*}$, whose pseudoresiduals are noticeably right-skewed (see Figure \ref{fig:model_checking}).
-This implies that the true distribution of $\Ztwo_{t,\tilde t^*}$ may follow a heavier-tailed distribution than the Gamma distribution used in the case study.
-See Sections 5 through 7 of supplement A for pseudoresidual plots corresponding to all observations and models.
+This implies that the true distribution of $\Ztwo_{t,\tilde t^*}$ may be heavier-tailed than the gamma distribution used in the case study.
+See Sections 5--7 of Supplement A for pseudoresidual plots corresponding to all observations and models.
 
 We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:model_checking}.
-Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the forward-backward algorithm.
-This procedure results in two histograms -- one corresponding to dive type 1 and another corresponding to dive type 2.
+Each observation of dive duration is weighted by the estimated probability that it corresponds to a particular dive type as decoded by the forward--backward algorithm.
+This procedure results in two histograms---one corresponding to dive type 1 and another corresponding to dive type 2.
 Each histogram is plotted together with the corresponding emission distribution estimated by the CarHHMM-DFT.
-Analogous histograms corresponding to the fine-scale observations are contained in Section 6 and 7 of of supplement A.
+Analogous histograms corresponding to the fine-scale observations are contained in Sections 6 and 7 of Supplement A.
 Our results mostly show that the CarHHMM-DFT explains the data well, but there are some exceptions.
 In particular, histograms corresponding to subdive state 3 show that $\Zone_{t,\tilde t^*}$ has heavier tails compared to a normal distribution.
-This indicates the existence of rare events corresponding to exceptionally sudden changes in acceleration of the killer whale.
+This indicates the existence of rare events corresponding to exceptionally sudden changes in the acceleration of the killer whale.
 These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.
@@
 
-The HHMM-DFT, which ignores auto-correlation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize 
-any given behaviour as subdive state 3 (see Figures 11 and 12 of supplement A).
-In addition, for all three components of each of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are significantly larger than those of the CarHHMM-DFT.
-The estimated uncertainties of the three components of each of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of of supplement A).
-This suggests that the HHMM-DFT overlooks some auto-correlation in the data, and that the HHMM-DFT may be overconfident in its parameters estimates compared to the full CarHHMM-DFT.
+The HHMM-DFT, which ignores autocorrelation in acceleration, decodes dive types and subdive states similarly to the CarHHMM-DFT, but is less likely to categorize 
+any given behaviour as subdive state 3 (see Figures 11 and 12 of Supplement A).
+In addition, for all three components of $\sigma_A^{*(\cdot,1)}$, $\sigma_A^{*(\cdot,2)}$, and $\sigma_A^{*(\cdot,3)}$, the HHMM-DFT produces estimates which are significantly larger than those of the CarHHMM-DFT.
+The estimated uncertainties of the three components of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, and $\hat \mu_A^{*(\cdot,3)}$ are also less than half of those for the CarHHMM-DFT (see Tables 1 and 2 of Supplement A).
+This suggests that the HHMM-DFT overlooks some autocorrelation in the data, and that the HHMM-DFT may be overconfident in its parameter estimates compared to the full CarHHMM-DFT.
 
-The CarHHMM does not model the ``wiggliness" of the acceleration data, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives} (also see Figure 13 of of supplement A).
+The CarHHMM does not model the ``wiggliness" of the acceleration data, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives} (see Figure 13 of Supplement A).
 These results essentially disqualify the CarHHMM as a viable model for this data set.
-The pseudoresiduals of acceleration are also light-tailed relative to a Normal distribution (see Figure 23 of of supplement A).
+The acceleration pseudoresiduals are also light-tailed relative to a normal distribution (see Figure 23 of Supplement A).
 
 Finally, the CarHMM-DFT, which lacks a hierarchical structure, produces fine-scale parameter estimates and subdive state estimates similar to those of the CarHHMM-DFT.
-However, its lack of hierarchical structure means that it fails to differentiate between short and long dives.
+However, the former's lack of hierarchical structure means that it fails to differentiate between short and long dives.
 This model therefore does not infer the dive-level Markov chain or the relationship between the dive types and subdive states.
@@
 
-For a more complete set of results for each of the candidate models, see supplement A.
+For a more complete set of results for each of the candidate models, see Supplement A.

### 2021-10-25 Report/Sections/simulation.tex
Update Final Version
address copy editor's comments (main text)

@@
 %\section{Simulation Study}
-We perform a simulation study based on data generated from the full CarHHMM-DFT as defined in Section \ref{subsec:model_selection} to evaluate each candidate model when the ground-truth is known.
+We perform a simulation study based on data generated from the full CarHHMM-DFT, as defined in Section \ref{subsec:model_selection}, to evaluate each candidate model when the ground truth is known.
 The parameters used to generate the data are based on those estimated in the case study (see Table \ref{table:emis_dists_CarHHMM-DFT}), with slight modifications made for simplicity.
-In particular, we set $\Zone_{t,\tilde t^*}$ to a scalar instead of a three dimensional vector.
+In particular, we set $\Zone_{t,\tilde t^*}$ to a scalar instead of a three-dimensional vector.
 We then fit all four models to the simulated data.
-Metrics used to evaluate each model include decoding accuracy of hidden states, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times.
+Metrics used to evaluate each model include hidden state decoding accuracy, bias in parameter estimates, empirical standard errors of parameter estimates, and fitting times.
 To assess the accuracy of uncertainty estimates, we also compare the empirical standard errors of a given model's parameter estimates with the standard errors estimated using the inverse of the observed Fisher information.
@@
 Each training data set consists of a sequence of 100 curves which we call a sequence of killer whale dives.
-Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
+Each dive can be one of $N=2$ dive types based on a Markov chain with transition probability matrix
 
@@
 %
-Dive duration is Gamma-distributed and the coarse-scale emission parameters are
-
-\[\mu^{(1)} = 27.34s, \enspace \sigma^{(1)} = 10.96s, \enspace \mu^{(2)} = 127.55s, \enspace \sigma^{(2)} = 63.89s.\]
+Dive duration is gamma distributed and the coarse-scale emission parameters are $\mu^{(1)} = 27.34$ $s$, $\sigma^{(1)} = 10.96$ $s$, $\mu^{(2)} = 127.55$ $s$, and $\sigma^{(2)} = 63.89$ $s$.
 %
 After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored.
-Each two-second segment is assigned one of $N^*=3$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \big\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \big\}$ with probability transition matrices
+Each two-second segment is assigned one of $N^*=3$ behaviours according to a fine-scale Markov chain $\tilde X^*_t = \big\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \big\}$ with transition probability matrices
 
@@
 Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \big\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\big\}$.
-Recall from Section \ref{subsec:model_selection} that we must specify the mean, standard deviation, and auto-correlation parameters corresponding to $\big\{\Zone_{t,1},\ldots,\Zone_{t,\tilde T_t^*}\big\}$ as well as the mean and standard deviation parameters corresponding to $\big\{\Ztwo_{t,1},\ldots,\Ztwo_{t,\tilde T_t^*}\big\}$.
+Recall from Section \ref{subsec:model_selection} that we must specify the mean, standard deviation, and autocorrelation parameters corresponding to $\big\{\Zone_{t,1},\ldots,\Zone_{t,\tilde T_t^*}\big\}$ as well as the mean and standard deviation parameters corresponding to $\big\{\Ztwo_{t,1},\ldots,\Ztwo_{t,\tilde T_t^*}\big\}$.
 We select the following parameters in line with the results from the case study:
 
-\begin{gather*}
-    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.97, \\
+\begin{enumerate}
+    \item $\mu_A^{*(\cdot,1)} = 0.0$ $s$, $\sigma_A^{*(\cdot,1)} = 0.05$ $s$, $\phi_A^{*(\cdot,1)} = 0.97$, $\mu_W^{*(\cdot,1)} = 34.01$, and $\sigma_W^{*(\cdot,1)} = 22.99$.
     %
-    \mu_W^{*(\cdot,1)} = 34.01, \quad \sigma_W^{*(\cdot,1)} = 22.99, \\
+    \item $\mu_A^{*(\cdot,2)} = 0.1$ $s$, $\sigma_A^{*(\cdot,2)} = 0.1$ $s$, $\phi_A^{*(\cdot,2)} = 0.83$, $\mu_W^{*(\cdot,2)} = 490.06$, and $\sigma_W^{*(\cdot,2)} = 502.56$.
     %
-    \mu_A^{*(\cdot,2)} = 0.1 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.83, \\
-    %
-    \mu_W^{*(\cdot,2)} = 490.06, \quad \sigma_W^{*(\cdot,2)} = 502.56, \\
-    %
-    \mu_A^{*(\cdot,3)} = 0.2 s, \enspace \sigma_A^{*(\cdot,3)} = 0.3 s, \enspace \phi_A^{*(\cdot,3)} = 0.97, \\
-    %
-    \mu_W^{*(\cdot,3)} = 9154.16, \quad \sigma_W^{*(\cdot,3)} = 13538.75.
-\end{gather*}
+    \item $\mu_A^{*(\cdot,3)} = 0.2$ $s$, $\sigma_A^{*(\cdot,3)} = 0.3$ $s$, $\phi_A^{*(\cdot,3)} = 0.61$, $\mu_W^{*(\cdot,3)} = 9154.16$, and $\sigma_W^{*(\cdot,3)} = 13538.75$.
+\end{enumerate}
 %
 It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $\Z$ alone, but we describe one possible mapping from $\Z$ to $Y^*$ in the appendix.
-Figure 1 of supplement B shows one realization of $\Z$ for five dives of one simulated data set along with the corresponding reconstructed realization of $Y^*$.
+Figure 1 of Supplement B shows one realization of $\Z$ for five dives of one simulated data set along with the corresponding reconstructed realizations of $Y^*$.
 
-The two simulated dive types differ in that dives of type 1 are much shorter on average (27 seconds) than dives of type 2 (128 seconds).
+The two simulated dive types differ in that dives of type 1 are much shorter on average (27 $s$) than dives of type 2 (128 $s$).
 The three simulated subdive states differ primarily because $\mu_W^*$ and $\sigma_W^*$ are much higher for subdive state 3 than for subdive state 2, which in turn are much higher than for subdive state 1.
@@
 
-We calculate maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using a similar optimization procedure as the one from the case study.
-%Likelihood maximization is performed on the Cedar Compute Canada cluster with 1 CPU and 8 GB of dedicated memory per data set.
+We calculate the maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using an optimization procedure similar to that in the case study.
 
 For each of the 500 training data sets, we simulate a test data set to assess how well each model predicts the hidden states as follows.
-Each test data set consists of a sequence of 100 dives and is created from the generative model with true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
-To assess the coarse-scale hidden state prediction, we estimate $p_t(i|y,\z) \equiv \Pr(X_t=i|Y=y,\Z=\z)$, $i=1,2$, $t=1,\ldots,100$ using the test-set observations $(y,\z)$ and training-set maximum likelihood estimates.
-These estimates are found using the {\em{forward-backward}} algorithm \citep{Zucchini:2016}.
-We compare these estimated conditional probabilities to $\{x_1,\ldots,x_{100}\}$, the true coarse-scale state realizations in the test data, by calculating the {\em{average dive decoding accuracy}} for a single training/test data set pair, $\sum_{t=1}^{100} \hat{p}_t(x_t|y,\z)/100$.
-We then report the average of these over the 500 training/test data set pairs.
-Analogously, to assess prediction of the fine-scale states, we estimate $p^*_{t,\tilde t^*}(i^*|y,\z) \equiv \Pr(\tilde X^*_{t,\tilde t^*}=i|Y=y,\Z=\z)$, $i^*=1,2,3$, $\tilde t^* = 1,\ldots,\tilde T^*_t$, $t=1,\ldots,100$, using the test-set observations, the training-set maximum likelihood estimates, and the forward-backward algorithm.
-Denoting the true fine-scale state realizations from the test data set as $\{\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}\}$, we define the overall \textit{average subdive decoding accuracy} as the average value of $\hat{p}^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*}|y,\z)$ across all simulated test data sets, dives, and windows.
-The conditional probabilities are estimated according to each of the four models under study using the maximum likelihood estimates from the training data set in conjunction with the forward-backward algorithm.
+Each test data set consists of a sequence of 100 dives and is created from the generative model with the true parameters $\{\theta, \Gamma, \theta^*, \Gamma^*\}$.
+To assess the coarse-scale hidden state prediction, we estimate $p_t(i \mid y,\z) = \Pr(X_t=i \mid Y=y,\Z=\z)$ for $i=1,2$ and $t=1,\ldots,100$ using the test set observations $(y,\z)$ and the training set maximum likelihood estimates.
+These estimates are found using the forward--backward algorithm \citep{Zucchini:2016}.
+We compare these estimated conditional probabilities to $\{x_1,\ldots,x_{100}\}$, the true coarse-scale state realizations in the test data, by calculating the average dive decoding accuracy for a single training/test data set pair, $\sum_{t=1}^{100} \hat{p}_t(x_t \mid y,\z)/100$.
+We then report the average of this measure over the 500 training/test data set pairs.
+Analogously, to assess fine-scale state prediction, we estimate $p^*_{t,\tilde t^*}(i^* \mid y,\z) = \Pr(\tilde X^*_{t,\tilde t^*}=i \mid Y=y,\Z=\z)$ for $i^*=1,2,3$, $\tilde t^* = 1,\ldots,\tilde T^*_t$, and $t=1,\ldots,100$ using the test set observations, the training set maximum likelihood estimates, and the forward--backward algorithm.
+Denoting the true fine-scale state realizations from the test data set by $\{\tilde x^*_{t,1},\ldots,\tilde x^*_{t,\tilde T^*_t}\}$, we define the overall average subdive decoding accuracy as the average value of $\hat{p}^*_{t,\tilde t^*}(\tilde x^*_{t,\tilde t^*} \mid y,\z)$ across all simulated test data sets, dives, and windows.
+The conditional probabilities are estimated according to each of the four models under study using the maximum likelihood estimates from the training data set in conjunction with the forward--backward algorithm.
 
@@
 
-The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model.
+The full CarHHMM-DFT is the best-performing model of the four candidates since it is the generating model.
 Its average dive decoding accuracy is approximately $0.96$ and its average subdive decoding accuracy is approximately $0.91$.
-All parameter estimates of emission distributions ($\theta$ and $\theta^*$) and probability transition matrices ($\Gamma$,$\Gamma^*$) on both the coarse scale and fine scale are either comparable or favourable to all other models.
-The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors.
-This underestimation is especially noticeable for parameters associated with the wiggliness $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to five times as large as the estimated standard error.
-See Tables 2 through 6 of supplement B for detailed results.
-%All parameter estimates of fine-scale mean values ($\hat \mu^*$) and probability transition matrices ($\hat \Gamma^*$ and $\hat \Gamma$) have statistically insignificant biases.
-In addition, the biases of $\hat \sigma$, $\hat \phi$, and $\hat \mu$ are either as small or smaller than the other three candidate models.
+All parameter estimates of emission distributions ($\theta$ and $\theta^*$) and transition probability matrices ($\Gamma$ and $\Gamma^*$) on both the coarse scale and fine scale are either comparable or favourable relative to all other models.
+The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, and $\hat \Gamma^*$) are well approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to be slightly smaller than the empirical standard errors.
+This underestimation is especially noticeable for parameters associated with the ``wiggliness" $\Ztwo_{t,\tilde t^*}$, where the empirical standard error can be up to five times as large as the estimated standard error.
+See Tables 2--6 of Supplement B for detailed results.
 
-The HHMM-DFT has an average dive decoding accuracy is comparable to the CarHHMM-DFT ($0.96$), but its average subdive decoding accuracy is worse by approximately 7 percentage points ($0.84$).
-Its parameter estimates are comparable to the CarHHMM-DFT with the notable exception that it greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$ and slightly overestimates $\sigma_A^{*(\cdot,3)}$.
-In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \mu_A^{*(\cdot,3)}$, $\hat \sigma_A^{*(\cdot,1)}$, $\hat \sigma_A^{*(\cdot,2)}$ and $\hat \sigma_A^{*(\cdot,3)}$ are much smaller than the associated empirical standard errors (see Table 3 of supplement B).
-These results suggest that estimates of standard deviation can be too large and estimates of standard errors can be too small when auto-correlation is ignored.
-This finding is consistent with the results of the case study, where the HHMM-DFT produced larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.
-When standard errors are underestimated, the associated confidence intervals are too narrow, implying that researchers may be overconfident in their parameter estimates.
+The HHMM-DFT has an average dive decoding accuracy comparable to that of the CarHHMM-DFT ($0.96$), but its average subdive decoding accuracy is worse by approximately seven percentage points ($0.84$).
+The HHMM-DFT's parameter estimates are comparable to the CarHHMM-DFT with the notable exception that the former greatly overestimates $\sigma_A^{*(\cdot,1)}$ and $\sigma_A^{*(\cdot,2)}$ and slightly overestimates $\sigma_A^{*(\cdot,3)}$.
+In addition, the estimated standard errors of $\hat \mu_A^{*(\cdot,1)}$, $\hat \mu_A^{*(\cdot,2)}$, $\hat \mu_A^{*(\cdot,3)}$, $\hat \sigma_A^{*(\cdot,1)}$, $\hat \sigma_A^{*(\cdot,2)}$, and $\hat \sigma_A^{*(\cdot,3)}$ are much smaller than the associated empirical standard errors (see Table 3 of Supplement B).
+These results suggest that estimates of standard deviation can be too large and estimates of standard errors can be too small when autocorrelation is ignored.
+This finding is consistent with the results of the case study, where the HHMM-DFT produce larger estimates of $\sigma_A^*$ and smaller estimates of standard error compared to the CarHHMM-DFT.
+When standard errors are underestimated, the associated confidence intervals are too narrow, and so researchers may be overconfident in their parameter estimates.
 
-The CarHHMM is the worst-performing model in terms of dive decoding accuracy and is approximately 4 percentage points worse than the CarHHMM-DFT ($0.92$).
-Its average subdive decoding accuracy is approximately 6 percentage points worse than the CarHHMM-DFT ($0.85$).
-This result is consistent with expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct difference between the subdive states.
-In addition to its relatively poor average decoding accuracy, the CarHHMM is also the worst of the four candidate models at estimating emission parameters and probability transition matrices.
-Estimates associated with subdive states 2 and 3 ($\theta^{*(\cdot,2)},\theta^{*(\cdot,3)}$) are especially poor.
-See supplement B for more detailed results.
+The CarHHMM is the worst-performing model in terms of its dive decoding accuracy, which is approximately four percentage points worse than that of the CarHHMM-DFT ($0.92$).
+The former's average subdive decoding accuracy is approximately six percentage points worse than the CarHHMM-DFT ($0.85$).
+This result is consistent with our expectations because the CarHHMM does not model the ``wiggliness" of the fine-scale process, which is the most distinct difference between the subdive states.
+In addition to its relatively poor average decoding accuracy, the CarHHMM is also the worst of the four candidate models at estimating emission parameters and transition probability matrices.
+Estimates associated with subdive states 2 and 3 ($\theta^{*(\cdot,2)}$ and $\theta^{*(\cdot,3)}$) are especially poor.
+See Supplement B for more-detailed results.
 
 Finally, the CarHMM-DFT is nearly identical to the CarHHMM-DFT in terms of average subdive decoding accuracy, fine-scale parameter biases, and both estimated and empirical standard error for the fine-scale parameter estimates.
-In addition, the time required to fit the CarHMM-DFT is less than a third of that of the other models (see Table \ref{table:accuracy}).
+In addition, the time required to fit the CarHMM-DFT is less than one-third that of the other models (see Table \ref{table:accuracy}).
 However, this model cannot differentiate between dive types as it assumes that there is only one.
-The CarHMM-DFT nonetheless fits a (misspecified) single Gamma distribution over the dive duration of all dives.
-The resulting parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$) are highly correlated (See Figure 5 of supplement B).
+The CarHMM-DFT nonetheless fits a (misspecified) single gamma distribution over the dive duration of all dives.
+The resulting parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$) are highly correlated (See Figure 5 of Supplement B).
 
 Figures \ref{fig:acc_coarse} and \ref{fig:acc_fine} display five dives of one simulated data set as well as the decoded dive types and subdive states associated with each model.
-The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state, the HHMM-DFT is slightly more likely to misclassify subdive state 1, and the CarHHMM is most likely to misclassify subdive state 3.
-All models classify dive type with high accuracy with the exception of the CarHMM-DFT, which does not estimate dive type.
+The CarHHMM-DFT and CarHMM-DFT produce similar estimates of subdive state, the HHMM-DFT is slightly more likely to misclassify subdive state 1, and the CarHHMM is more likely to misclassify subdive state 3.
+All models classify dive type with high accuracy, with the exception of the CarHMM-DFT, which does not estimate dive type.

### 2021-10-25 Report/main.tex
address copy editor's comments (main text)

@@
 %\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
-\title[]{Modelling multi-scale state-switching functional data with hidden Markov models}
+\title[]{Modelling multi-scale, state-switching functional data with hidden Markov models}
 \author{Evan Sidrow\authorref{*1}}
@@
 \author{Marie Auger-M\'eth\'e\authorref{1,4}}
-\affiliation[1]{Department of Statistics, University of British Columbia, Vancouver, Canada}
-\affiliation[2]{Marine Mammal Research Unit, University of British Columbia, Vancouver, Canada}
-\affiliation[3]{Department of Zoology, University of British Columbia, Vancouver, Canada}
-\affiliation[4]{Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada}
-\affiliation[5]{Department of Biostatistics, University of Florida, Gainesville, USA}
+\affiliation[1]{Department of Statistics, University of British Columbia, Vancouver, Canada, V6T 1Z4}
+\affiliation[2]{Marine Mammal Research Unit, University of British Columbia, Vancouver, Canada, V6T 1Z4}
+\affiliation[3]{Department of Zoology, University of British Columbia, Vancouver, Canada, V6T 1Z4}
+\affiliation[4]{Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada, V6T 1Z4}
+\affiliation[5]{Department of Biostatistics, University of Florida, Gainesville, U.S.A., 32611}
 
@@
 \KWDtitle{Key words and phrases}
-Accelerometer data\sep animal movement\sep biologging\sep diving behaviour\sep hierarchical modelling\sep killer whales\sep state-switching\sep statistical ecology\sep time series.
+Accelerometer data\sep animal movement\sep biologging\sep diving behaviour\sep hierarchical modelling\sep killer whales\sep state switching\sep statistical ecology\sep time series.
 % MSC 2010 subject classification codes
-\KWDtitle{MSC 2010}Primary 62M05 \sep secondary 62P12}
+\KWDtitle{MSC 2010}Primary 62M05\sep secondary 62P12}
 \begin{abstract}
@@
 
-Data sets comprised of sequences of curves sampled at high frequencies in time are increasingly common in practice, but they can exhibit complicated dependence structures that cannot be modelled using common methods of Functional Data Analysis (FDA).
-We detail a hierarchical approach which treats the curves as observations from a hidden Markov model (HMM).
-The distribution of each curve is then defined by another fine-scale model which may involve auto-regression and require data transformations using moving-window summary statistics or Fourier analysis.
+Data sets composed of sequences of curves sampled at high frequencies in time are increasingly common in practice, but they can exhibit complicated dependence structures that cannot be modelled using common methods in functional data analysis.
+We detail a hierarchical approach that treats the curves as observations from a hidden Markov model.
+The distribution of each curve is then defined by another fine-scale model that may involve autoregression and require data transformations using moving-window summary statistics or Fourier analysis.
 This approach is broadly applicable to sequences of curves exhibiting intricate dependence structures.
-As a case study, we use this framework to model the fine-scale kinematic movement of a northern resident killer whale ({\em{Orcinus orca}}) off the western coast of Canada.
+As a case study, we use this framework to model the fine-scale kinematic movements of a northern resident killer whale ({\em{Orcinus orca}}) off the western coast of Canada.
 Through simulations, we show that our model produces more interpretable state estimation and more accurate parameter estimates compared to existing methods.
@@
 XMMS 6 2019.
-This project was supported in part by a financial contribution from Fisheries and Oceans Canada and the Natural Sciences and Engineering Research Council of Canada (Whale Science for Tomorrow).
+We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC) as well as the support of Fisheries and Oceans Canada (DFO).
+This project was supported in part by a financial contribution from the DFO and NSERC (Whale Science for Tomorrow).
 This research was enabled in part by support provided by WestGrid (www.westgrid.ca) and Compute Canada (www.computecanada.ca).
-We acknowledge the support of the Natural Sciences and Engineering Research Council of Canada (NSERC) as well as the support of Fisheries and Oceans Canada (DFO). 
 Marie Auger-M\'eth\'e and Nancy Heckman thank the NSERC Discovery program, and Marie Auger-M\'eth\'e additionally thanks the Canadian Research Chair program.
-Evan Sidrow thanks the University of British Columbia and the Four-Year Doctoral Fellowship program.
+Evan Sidrow thanks the University of British Columbia for funding provided via the Four-Year Doctoral Fellowship program.
 We are grateful to Dr. Joe Watson for his constructive suggestions.

### 2021-10-25 Report/Sections/appendix.tex
add copy editor edits for bibliography, appendix, figures, and tables

--- 2021-09-16 Report/Sections/appendix.tex
@@
-\subsection{Detailed description of data simulation from Section \ref{subsec:data_simulation}}
+\subsection{Detailed description of the data simulation in Section \ref{subsec:data_simulation}}
 
-\setcounter{equation}{6}   %  set this to the last equation number used
+\setcounter{equation}{7}   %  set this to the last equation number used
 
-We easily simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
-For each dive $t$, we also easily generate the fine-scale hidden Markov chain $\tilde X^*_t \equiv \big\{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T_t^*}\big\}$ according to one of the probability transition matrices $\Gamma^{*(1)}$ or $\Gamma^{*(2)}$, depending upon the value of $X_t$.
+We can easily simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.
+For each dive $t$, we can also easily generate the fine-scale hidden Markov chain $\tilde X^*_t = \big\{\tilde X^*_{t,1},\ldots,\tilde X^*_{t,\tilde T_t^*}\big\}$ according to one of the transition probability matrices $\Gamma^{*(1)}$ or $\Gamma^{*(2)}$ depending upon the value of $X_t$.
 This determines the sequence of fine-scale hidden states corresponding to each window.
-Recall that the fine-scale model is based on a sequence of $\tilde T_t^*$ two-second windows, each containing 100 observations, and our model is formulated in terms of quantities derived from the raw data within each window (namely, the average acceleration, $\Zone_{t,\tilde t^*}$, and wiggliness, $\Ztwo_{ t,\tilde t^*}$).
+Recall that the fine-scale model is based on a sequence of $\tilde T_t^*$ two-second windows, each containing 100 observations, and that our model is formulated in terms of quantities derived from the raw data within each window (namely, the average acceleration, $\Zone_{t,\tilde t^*}$, and wiggliness, $\Ztwo_{ t,\tilde t^*}$).
 Generating the raw acceleration data from $\Zone_{t,\tilde t^*}$ and $\Ztwo_{ t,\tilde t^*}$ is not straightforward.
@@
 Here, we explain how we generate the acceleration curves so that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{ t,\tilde t^*}$ both follow the specified model.
-A key component is the discrete Fourier transformation of the 100 raw acceleration values in window $\tilde{t}^*$ of dive $t$:
+A key component is the DFT of the 100 raw acceleration values in window $\tilde{t}^*$ of dive $t$,
 
 \[
-    \hat{Y}^{*(k)}_{t,\tilde{t}^*}  \equiv DFT\left\{Y^*_{t,100 (\tilde{t}^*-1) + 1 },\ldots, Y^*_{t,100 \tilde{t}^*}\right\}(k)
+    \hat{Y}^{*(k)}_{t,\tilde{t}^*}  = \mathrm{DFT}\left\{Y^*_{t,100 (\tilde{t}^*-1) + 1 },\ldots, Y^*_{t,100 \tilde{t}^*}\right\}(k)
 \]
@@
 
-We simulate the raw acceleration data for dive $t$ in three steps: (1) simulate the average acceleration within each window $\left(\hat Y^{*(0)}_{t,\tilde t^*}\right)$, (2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and (3) take the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$:
+We simulate the raw acceleration data for dive $t$ in three steps: (1) simulate the average acceleration within each window $\left(\hat Y^{*(0)}_{t,\tilde t^*}\right)$, (2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and (3) take the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$,
 
 \[
-    \{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\} \equiv IDFT\left\{\hat{Y}^{*(0)}_{t,\tilde t^*},
+    \{Y^*_{t,100(\tilde t^* - 1) + 1},\ldots,Y^*_{t,100\tilde t^*}\} = \mathrm{IDFT}\left\{\hat{Y}^{*(0)}_{t,\tilde t^*},
     \ldots, \hat{Y}^{*(99)}_{t,\tilde t^*}\right\}
-    {\rm{~for~}} \tilde t^* = 1,\ldots,\tilde T^*_t.
 \]
-The details of steps (1) and (2) are described below.
+for $\tilde t^* = 1,\ldots,\tilde T^*_t$. The details of steps (1) and (2) are below.
 
-For step (1), we generate $\hat Y^{*(0)}_{t,1}, \ldots, \hat Y^{*(0)}_{t,\tilde T_t^*}$ as a CarHMM with underlying Markov state sequence $\tilde X^{*}_{t,1}, \ldots, \tilde X^{*}_{t,\tilde T_t^*}$ and random first emission.
+For step (1), we generate $\hat Y^{*(0)}_{t,1}, \ldots, \hat Y^{*(0)}_{t,\tilde T_t^*}$ as a CarHMM with the underlying Markov state sequence $\tilde X^{*}_{t,1}, \ldots, \tilde X^{*}_{t,\tilde T_t^*}$ and a random first emission.
 Specifically, we let
@@
 \[
-    	\hat{Y}^{*(0)}_{t,1}|\tilde X^*_{t,1} = i^* ~~\sim~~ \mathcal{N} \left(100 \mu^{*(\cdot,i^*)}_A, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
-    	~~{\rm{and}}
+	\hat{Y}^{*(0)}_{t,1} \mid \tilde X^*_{t,1} = i^* ~~\sim~~ \mathcal{N} \left(100 \mu^{*(\cdot,i^*)}_A, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right)
 \]
+and
+\begin{align}	
+   \hat{Y}^{*(0)}_{t,\tilde t^*}\mid\tilde X^*_{t,\tilde t^*} = i^*,\hat{Y}^{*(0)}_{t,\tilde t^*-1}
+   &~~\sim ~~\mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,\tilde t^*-1} + 100(1-\phi_A^{*(\cdot,i^*)}) \mu^{*(\cdot,i^*)}_A, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \label{eqn:yhat_0}
+\end{align}
+for $\tilde t^* = 2,\ldots, \tilde T^*_t$, where $\mu_A^{*(\cdot,i^*)}$, $\sigma_A^{*(\cdot,i^*)}$, and $\phi_A^{*(\cdot,i^*)}$ are as defined in the simulation study in Section 4.
 
-    \begin{align}	
-       \hat{Y}^{*(0)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = i^*,\hat{Y}^{*(0)}_{t,\tilde t^*-1}
-       &~~\sim ~~\mathcal{N} \left(\phi_A^{*(\cdot,i^*)} \hat{Y}^{*(0)}_{t,\tilde t^*-1} + 100(1-\phi_A^{*(\cdot,i^*)}) \mu^{*(\cdot,i^*)}_A, \left(100\sigma_A^{*(\cdot,i^*)}\right)^2 \right), \label{eqn:yhat_0} \\
+For step (2), we first construct $\hat{Y}^{*(k)}_{t,\tilde t^*}$ for $k=1,\ldots, 49$, as
 %
-        &\tilde t^* = 2,\ldots, \tilde T^*_t
-    	\nonumber
-    \end{align}
-    where $\mu_A^{*(\cdot,i^*)}$, $\sigma_A^{*(\cdot,i^*)}$, and $\phi_A^{*(\cdot,i^*)}$ are as defined in the simulation study in Section 4.
-
- For step (2),
- we first construct $\hat{Y}^{*(k)}_{t,\tilde t^*}$, $k=1,\ldots, 49$, as
-
 \begin{equation}
-        \hat{Y}^{*(k)}_{t,\tilde t^*} = a_{t,\tilde t^*}^{(k)} i\sqrt{b^{(k)}_{t,\tilde t^*}}
+    \hat{Y}^{*(k)}_{t,\tilde t^*} = a_{t,\tilde t^*}^{(k)} i\sqrt{b^{(k)}_{t,\tilde t^*}},
     \label{eqn:abYhat}
 \end{equation}
-    where the $a^{(k)}_{t,\tilde t^*}$'s are independent and equal to either 1 or -1 each with probability 1/2 and $i$ is the imaginary number.
-We include $i$ in Equation (\ref{eqn:abYhat}) to force all variation within a window to take the form of a sine wave, which reduces the variation between the endpoints of windows compared to a cosine wave.
-    Given the fine scale states, the $b^{(k)}_{t,\tilde t^*}$'s are independent of one another and independent of the $a^{(k)}_{t,\tilde t^*}$'s.  
+%
+where the $a^{(k)}_{t,\tilde t^*}$s are independent and equal to either 1 or -1 each with probability 1/2 and $i$ is the imaginary unit.
+We include $i$ in Equation (\ref{eqn:abYhat}) to force all variation within a window to take the form of a sine wave, which reduces the variation between the endpoints of windows relative to a cosine wave.
+Given the fine scale states, the $b^{(k)}_{t,\tilde t^*}$s are independent of one another and independent of the $a^{(k)}_{t,\tilde t^*}$s.
 The distribution of $b^{(k)}_{t,\tilde t^*}$ is
@@
 \begin{split}
-    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1 &~~\sim~~ {\rm{Gamma}}(11.03/(k+1)^3, 15.54) \\
+	b^{(k)}_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 1 &~~\sim~~ {\rm{Gamma}}(11.03/(k+1)^3, 15.54) \\
 	%
-    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2  &~~\sim ~~{\rm{Gamma}}(4.80/(k+1)^3, 515.38). \\ 
+	b^{(k)}_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 2  &~~\sim ~~{\rm{Gamma}}(4.80/(k+1)^3, 515.38) \\ 
 	%
-    	b^{(k)}_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 3  &~~\sim ~~{\rm{Gamma}}(2.31/(k+1)^3, 20023.44). 
+	b^{(k)}_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 3  &~~\sim ~~{\rm{Gamma}}(2.31/(k+1)^3, 20023.44). 
 \end{split}
@@
 \end{align}
-    The first argument of ${\rm{Gamma}}\left(\cdot,\cdot\right)$ is the shape parameter and the second is the scale parameter.
-The squared magnitude of the $k^{th}$ Fourier coefficient is equal to $b^{(k)}_{t,\tilde t^*}$, which decays like $1/k^3$ to ``smooth out" the raw acceleration data.
 
-   We then define the remaining 50 Fourier coefficients:
+The first argument of $\mathrm{Gamma}(\cdot,\cdot)$ is the shape parameter and the second is the scale parameter.
+The squared magnitude of the $k^{th}$ Fourier coefficient is equal to $b^{(k)}_{t,\tilde t^*}$, which decays on the order of $1/k^3$ to ``smooth out" the raw acceleration data.
 
-    \[
-        \hat{Y}^{*(50)}_{t,\tilde t^*} = 0 {\rm{~~and ~~}}
-	    \hat{Y}^{*(k)}_{t,\tilde t^*} = -\hat{Y}^{*(100-k)}_{t,\tilde t^*} , k = 51,\ldots,99.
-    \]
-   This guarantees that the inverse discrete Fourier transform is real-valued.
+We then define the remaining 50 Fourier coefficients as $\hat{Y}^{*(50)}_{t,\tilde t^*} = 0$ and $\hat{Y}^{*(k)}_{t,\tilde t^*} = -\hat{Y}^{*(100-k)}_{t,\tilde t^*}$ for $k = 51,\ldots,99$.
+This guarantees that the inverse DFT is real-valued.
 
 We now show that this construction of the raw acceleration data results in the distributions listed in Section \ref{subsec:data_simulation}.
-It suffices to show that the construction of the discrete Fourier transformations, the $\hat{Y}^{(k)}_{t,\tilde t^*}$'s, yields the desired distributions.
+It suffices to show that the construction of the DFTs, the $\hat{Y}^{(k)}_{t,\tilde t^*}$s, yields the desired distributions.
 
@@
 $\Zone_{t,1}, \ldots, \Zone_{t,\tilde{T}^*}$ 
-follows a CarHMM with Normal emissions distributions and parameters as defined in the simulation study in Section 4.
+follows a CarHMM with normal emission distributions and parameters as defined in the simulation study in Section 4.
 
 From Equations (\ref{eqn:z}) and (\ref{eqn:abYhat}), 
-the wiggliness within window $\tilde t^*$ of dive $t$ is
+the ``wiggliness" within window $\tilde t^*$ of dive $t$ is
 
@@
 %
-If $\tilde \omega < 50$, then $\Ztwo_{t,\tilde t^*}$ is the sum of independent Gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,\tilde t^*}$ is also Gamma.
-Thus, by Equation (\ref{eqn:bdist}):
+If $\tilde \omega < 50$, then $\Ztwo_{t,\tilde t^*}$ is the sum of independent gamma distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,\tilde t^*}$ is also gamma distributed.
+Thus, by Equation (\ref{eqn:bdist}),
 
 \[
-    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 1  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 11.03/(k+1)^3 , 15.54 \right),
+    \Ztwo_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 1  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 11.03/(k+1)^3 , 15.54 \right),
 \]
@@
 \[
-    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.80/(k+1)^3 , 515.38 \right), \text{and}
+    \Ztwo_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.80/(k+1)^3 , 515.38 \right), \text{and}
 \]
-%
+and
 \[
-    \Ztwo_{t,\tilde t^*}|\tilde X^*_{t,\tilde t^*} = 3  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 2.31/(k+1)^3 , 20023.44 \right).
+    \Ztwo_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 3  ~~ \sim ~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 2.31/(k+1)^3 , 20023.44 \right).
 \]
-Setting $\tilde \omega$ to 10 and carrying out simple calculation of the mean and variance of a Gamma distribution gives the proper values for $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$.
+Setting $\tilde \omega$ to 10 and carrying out simple calculation of the mean and variance of a gamma distribution gives the desired values for $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$.
 
@@
 
-\subsection{Likelihood of CarHHMM-DFT}
+\subsection{Likelihood of the CarHHMM-DFT}
 
-The overall likelihood of the CarHHMM-DFT model is as follows:
+The overall likelihood of the CarHHMM-DFT is
 
@@
 %
-In particular,
+where
 
@@
 %
-where $f^{(i)}(y_t;\theta^{(i)})$ is the emission distribution of dive duration given that $X_t = i$.
-The likelihood $\calL_{\text{fine}}$ corresponds to the fine-scale model and is equal to the following:
+and $f^{(i)}(y_t;\theta^{(i)})$ is the emission distribution of dive duration given that $X_t = i$. The likelihood corresponding to the fine-scale model is
 
-\[\calL_{\text{fine}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},\]
+\[\calL_{\text{fine}}\left(\theta^{*},\Gamma^{*(i)};\z_t\right) = \delta^{*(i)} \prod_{\tilde t^* = 2}^{\tilde T^*_t} \Gamma^{*(i)} P(\z_{t,\tilde t^*} \mid \z_{t,\tilde t^*-1};\theta^*) \mathbf{1}_{N^*},\]
 %
-where $P(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1};\theta^*)$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{*(\cdot,i^*)}(\z_{t,\tilde t^*}|\z_{t,\tilde t^*-1}; \theta^{*(\cdot,i^*)})$.
+where $P(\z_{t,\tilde t^*}\mid\z_{t,\tilde t^*-1};\theta^*)$ is an $N^* \times N^*$ diagonal matrix with the $(i^*,i^*)^{th}$ entry $f^{*(\cdot,i^*)}(\z_{t,\tilde t^*}\mid\z_{t,\tilde t^*-1}; \theta^{*(\cdot,i^*)})$.
 %
-Recall that $f^{*(\cdot,i^*)}(\cdot|\z_{t,t^*-1}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,\tilde t^*}$ when $\tilde X^*_{t,\tilde t^*} = i^*$ and $\Z_{t,\tilde t^*-1} = \z_{t,\tilde t^*-1}$.
+Recall that $f^{*(\cdot,i^*)}(\cdot \mid \z_{t,t^*-1}; \theta^{*(\cdot,i^*)})$ is the probability density function of $\Z_{t,\tilde t^*}$ when $\tilde X^*_{t,\tilde t^*} = i^*$ and $\Z_{t,\tilde t^*-1} = \z_{t,\tilde t^*-1}$.
 

### 2021-10-25 Code/Visualisor.py
Update copy editor edits to plots

@@ -94,23 +94,23 @@ class Visualisor:
             positions = np.vstack([Xtemp.ravel(), Ytemp.ravel()])
             Ztemp = np.reshape(kernel.pdf(positions).T, Xtemp.shape)
 
-            np.set_printoptions(suppress=True,precision=4)
+            #np.set_printoptions(suppress=True,precision=4)
             im = plt.imshow(np.rot90(Ztemp),extent = xlim + ylim)
-            plt.xlabel(xlabs[i],fontsize=12)
-            plt.ylabel(ylabs[i],fontsize=12)
-            plt.xticks(fontsize=10)
-            plt.yticks(fontsize=10)
-            plt.gca().ticklabel_format(style='sci',scilimits = (-3,3))
+            plt.xlabel(xlabs[i],fontsize=10)
+            plt.ylabel(ylabs[i],fontsize=10)
+            plt.xticks(fontsize=8)
+            plt.yticks(fontsize=8)
+            plt.gca().ticklabel_format(style='sci',scilimits = (-10,10),useMathText=True)
             plt.gca().yaxis.get_offset_text().set_fontsize(10)
             plt.gca().xaxis.get_offset_text().set_fontsize(10)
             divider = make_axes_locatable(plt.gca())
             cax = divider.append_axes("right", size="5%", pad=0.05)
             c = fig.colorbar(im,cax=cax)#im, orientation='horizontal',pad=0.125)
-            c.ax.tick_params(labelsize=10)
-            c.ax.ticklabel_format(style='sci',scilimits = (-3,3))
+            c.ax.tick_params(labelsize=8)
+            c.ax.ticklabel_format(style='sci',scilimits = (-10,10))
             c.ax.yaxis.get_offset_text().set_fontsize(10)
             if i%2 == 1 or i==4:
-                c.set_label('Density',fontsize = 10)
+                c.set_label('Density',fontsize = 8)
 
             fig_num += 1
 
@@ -162,7 +162,7 @@ class Visualisor:
                 dist = self.pars.features[0][feature]['f']
                 K = self.pars.K[0]
                 colors = [cm.get_cmap('tab10')(i) for i in range(K)]
-                legend = ['Dive Type %d'%(x+1) for x in range(K)]
+                legend = ['Dive type %d'%(x+1) for x in range(K)]
             else:
                 mu = self.hhmm.theta[1][0][feature]['mu']
                 sig = self.hhmm.theta[1][0][feature]['sig']
[…]

### 2021-10-26 Report/Supp_Case.tex
Update Final Version

@@
 % Title, authors, affiliations
-\title{Modelling state-switching functional data with hidden Markov models: supplement A (case study)}%\query{Q1}
+\title{Modelling multi-scale, state-switching functional data with hidden Markov models: Supplement A (case study)}
 \date{}
-\author{BlindedA, BlindedB}%\authorref{1}\thanksref{*}}
-%\author{BlindedB}%\authorref{2}}
-%\affiliation[1]{Author affiliations will go here in the accepted manuscript, but do NOT include them in your initial submission because it must be anonymous.}
-%\affiliation[2]{Second Affiliation}
+\author{Evan Sidrow, Nancy Heckman, Sarah M. E. Fortune, \\ Andrew W. Trites, Ian Murphy, and Marie Auger-M\'eth\'e}
 
@@
 
-    \section{Lag plots}
+    \section{Lag Plots}
 
@@
 
-        \noindent Figure \arabic{fignum}: Collection of 5 lag plots where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
+        \noindent Figure \arabic{fignum}: Collection of five lag plots, where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
 Each plot corresponds to a feature in the killer whale dive data.
-Lag plots are used to decide whether to include auto-correlation in the HMMs used to model the killer whale's kinematic data.
-They can also be used to determine the number of hidden states if multiple distinct patterns are visible.
+Lag plots are used to decide whether to include autocorrelation in the HMMs for modelling the killer whale's kinematic data.
+These plots can also be used to determine the number of hidden states if multiple distinct patterns are visible.
         \addtocounter{fignum}{1}
@@
         \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the dive duration parameters $\mu_Y^{(i)}$ and $\sigma_Y^{(i)}$ for dive types $i = 1,2$.
-All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_Y,\hat \sigma_Y)$, which is denoted by a red star.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_Y,\hat \sigma_Y)$, which is denoted by red stars.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the coarse-scale probability transition matrix parameters $\Gamma_{1,2}$ and $\Gamma_{2,1}$.
-All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the coarse-scale transition probability matrix parameters $\Gamma_{1,2}$ and $\Gamma_{2,1}$.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \Gamma_{1,2},\hat \Gamma_{2,1})$, which is denoted by a red star.
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the wiggliness parameters $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_W^{*(\cdot,i^*)}, \hat \sigma_W^{*(\cdot,i^*)})$, which is denoted by a red star.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the ``wiggliness" parameters $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_W^{*(\cdot,i^*)}, \hat \sigma_W^{*(\cdot,i^*)})$, which is denoted by red stars.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $x$-acceleration parameters $\mu_{A_x}^{*(\cdot,i^*)}, \sigma_{A_x}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_x}^{*(\cdot,i^*)}, \hat \sigma_{A_x}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $x$-acceleration parameters $\mu_{A_x}^{*(\cdot,i^*)}, \sigma_{A_x}^{*(\cdot,i^*)}$ and the logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_x}^{*(\cdot,i^*)}, \hat \sigma_{A_x}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by red stars.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $y$-acceleration parameters $\mu_{A_y}^{*(\cdot,i^*)}, \sigma_{A_y}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_y}^{*(\cdot,i^*)}, \hat \sigma_{A_y}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $y$-acceleration parameters $\mu_{A_y}^{*(\cdot,i^*)}, \sigma_{A_y}^{*(\cdot,i^*)}$ and the logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_y}^{*(\cdot,i^*)}, \hat \sigma_{A_y}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by red stars.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $z$-acceleration parameters $\mu_{A_z}^{*(\cdot,i^*)}, \sigma_{A_z}^{*(\cdot,i^*)}$ and logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_z}^{*(\cdot,i^*)}, \hat \sigma_{A_z}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by a red star.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $z$-acceleration parameters $\mu_{A_z}^{*(\cdot,i^*)}, \sigma_{A_z}^{*(\cdot,i^*)}$ and the logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_z}^{*(\cdot,i^*)}, \hat \sigma_{A_z}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by red stars.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the fine-scale probability transition matrix parameters associated both dive types ($\Gamma^{*(i)}_{i^*,j^*}$ for dive types $i = 1,2$ and subdive states $i^*,j^* = 1,2,3$).
-All other parameters and probability transition matrices are set to be equal to the maximum likelihood estimates shown in Table (1) of Section 3.1.
-The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\hat \Gamma^{*(i)}$, which is denoted by a red star.
-Note that $\hat \Gamma^{*(i)}_{1,3} \approx \hat \Gamma^{*(i)}_{3,1} \approx 0$.
-This indicates that the whale rarely switches directly between subdive states 1 and 3.
+        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the fine-scale transition probability matrix parameters associated with both dive types ($\Gamma^{*(i)}_{i^*,j^*}$ for dive types $i = 1,2$ and subdive states $i^*,j^* = 1,2,3$).
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\hat \Gamma^{*(i)}$, which is denoted by red stars.
+Note that $\hat \Gamma^{*(i)}_{1,3} \approx \hat \Gamma^{*(i)}_{3,1} \approx 0$: T=this indicates that the whale rarely switches directly between subdive states 1 and 3.
         \addtocounter{fignum}{1}
@@
             %
-            %\caption{Estimates and empirical standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the full CarHHMM-DFT.}
+            %\caption{Estimates and empirical standard errors of emission parameters for the distributions of dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the full CarHHMM-DFT.}
             \begin{center}
@@
             \begin{tabular}{ccccc}
-                \multirow{2}{*}{Feature}                                                       & \multirow{2}{*}{Dive / Subdive Type} & \multicolumn{3}{c}{Parameter Estimate}        \\
+                \multirow{2}{*}{Feature}                                                       & \multirow{2}{*}{Dive type / subdive state} & \multicolumn{3}{c}{Parameter estimate}  \\
                                                                                                &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
-                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.34 (0.63)$ & $10.96 (0.56)$ & ---           \\
-                                                                                               & 2                                    & $127.5 (11.3)$ & $63.9 (9.0)$ & ---           \\ \hline
-                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.449 (0.030)$ & $0.039 (0.001)$ & $0.968 (0.002)$ \\
-                                                                                               & 2                                    & $0.210 (0.012)$ & $0.096 (0.002)$ & $0.829 (0.007)$ \\
-                                                                                               & 3                                    & $0.232 (0.035)$ & $0.296 (0.010)$ & $0.607 (0.023)$ \\ \hline
-                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.450 (0.038)$ & $0.051 (0.001)$ & $0.968 (0.002)$ \\
-                                                                                               & 2                                    & $0.437 (0.012)$ & $0.094 (0.002)$ & $0.829 (0.007)$ \\
-                                                                                               & 3                                    & $0.366 (0.042)$ & $0.365 (0.012)$ & $0.607 (0.023)$ \\ \hline
-                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.691 (0.043)$ & $0.058 (0.001)$ & $0.968 (0.002)$ \\
-                                                                                               & 2                                    & $-0.573 (0.014)$ & $0.111 (0.002)$ & $0.829 (0.007)$ \\
-                                                                                               & 3                                    & $-0.303 (0.041)$ & $0.354 (0.012)$ & $0.607 (0.023)$ \\ \hline
-                \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $34.02 (0.36)$ & $22.99 (0.38)$ & ---           \\
-                                                                                               & 2                                    & $490.1 (5.6)$ & $502.6 (6.8)$ & ---           \\
-                                                                                               & 3                                    & $9150 (220)$ & $13540 (350)$ & ---           \\ \hline
+                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.342 \pm 0.633$ & $10.961 \pm 0.560$ & ---           \\
+                                                                                               & 2                                    & $127.548 \pm 11.341$ & $63.888 \pm 9.032$ & ---           \\ \hline
+                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.449 \pm 0.030$ & $0.039 \pm 0.001$ & $0.968 \pm 0.002$ \\
+                                                                                               & 2                                    & $0.210 \pm 0.012$ & $0.096 \pm 0.002$ & $0.829 \pm 0.007$ \\
+                                                                                               & 3                                    & $0.232 \pm 0.035$ & $0.296 \pm 0.010$ & $0.607 \pm 0.023$ \\ \hline
+                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.450 \pm 0.038$ & $0.051 \pm 0.001$ & $0.968 \pm 0.002$ \\
+                                                                                               & 2                                    & $0.437 \pm 0.012$ & $0.094 \pm 0.002$ & $0.829 \pm 0.007$ \\
+                                                                                               & 3                                    & $0.366 \pm 0.042$ & $0.365 \pm 0.012$ & $0.607 \pm 0.023$ \\ \hline
+                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.691 \pm 0.043$ & $0.058 \pm 0.001$ & $0.968 \pm 0.002$ \\
+                                                                                               & 2                                    & $-0.573 \pm 0.014$ & $0.111 \pm 0.002$ & $0.829 \pm 0.007$ \\
+                                                                                               & 3                                    & $-0.303 \pm 0.041$ & $0.354 \pm 0.012$ & $0.607 \pm 0.023$ \\ \hline
+                \multirow{3}{*}{``Wiggliness" -- $\Ztwo_{t,\tilde t^*}$}                          & 1                                    & $34.015 \pm 0.368$ & $22.986 \pm 0.378$ & ---           \\
+                                                                                               & 2                                    & $490.068 \pm 5.584$ & $502.558 \pm 6.776$ & ---           \\
+                                                                                               & 3                                    & $9154.156 \pm 220.765$ & $13538.747 \pm 354.281$ & ---           \\ \hline
             \end{tabular}
@@
 
-            \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the full CarHHMM-DFT.
-The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
+            \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the full CarHHMM-DFT.
+Figures following $\pm$ refer to estimated standard errors derived from the inverse of the observed information matrix.
             \addtocounter{tablenum}{1}
@@
                                                                                                &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$     \\ \hline
-                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $26.61 (0.61)$ & $10.16 (0.52)$ & ---             \\
-                                                                                               & 2                                    & $116.8 (10.6)$ & $64.7 (8.6)$ & ---             \\ \hline
-                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.000 (0.002)$ & $0.065 (0.001)$ & ---             \\
-                                                                                               & 2                                    & $0.085 (0.004)$ & $0.175 (0.003)$ & ---             \\
-                                                                                               & 3                                    & $0.291 (0.020)$ & $0.484 (0.014)$ & ---             \\ \hline
-                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.365 (0.002)$ & $0.068 (0.001)$ & ---             \\
-                                                                                               & 2                                    & $0.408 (0.002)$ & $0.108 (0.002)$ & ---             \\
-                                                                                               & 3                                    & $0.275 (0.017)$ & $0.407 (0.012)$ & ---             \\ \hline
-                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.442 (0.003)$ & $0.101 (0.002)$ & ---             \\
-                                                                                               & 2                                    & $-0.471 (0.003)$ & $0.158 (0.002)$ & ---             \\
-                                                                                               & 3                                    & $-0.298 (0.017)$ & $0.415 (0.012)$ & ---             \\ \hline
-                \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $43.65 (0.52)$ & $33.66 (0.58)$ & ---             \\
-                                                                                               & 2                                    & $608.0 (6.8)$ & $717.3 (9.2)$ & ---             \\
-                                                                                               & 3                                    & $7100 (160)$ & $13620 (310)$ & ---             \\ \hline
+                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $26.613 \pm 0.608$ & $10.161 \pm 0.519$ & ---             \\
+                                                                                               & 2                                    & $116.785 \pm 10.626$ & $64.712 \pm 8.574$ & ---             \\ \hline
+                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.000 \pm 0.002$ & $0.065 \pm 0.001$ & ---             \\
+                                                                                               & 2                                    & $0.085 \pm 0.004$ & $0.175 \pm 0.003$ & ---             \\
+                                                                                               & 3                                    & $0.291 \pm 0.020$ & $0.484 \pm 0.014$ & ---             \\ \hline
+                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.365 \pm 0.002$ & $0.068 \pm 0.001$ & ---             \\
+                                                                                               & 2                                    & $0.408 \pm 0.002$ & $0.108 \pm 0.002$ & ---             \\
+                                                                                               & 3                                    & $0.275 \pm 0.017$ & $0.407 \pm 0.012$ & ---             \\ \hline
+                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.442 \pm 0.003$ & $0.101 \pm 0.002$ & ---             \\
+                                                                                               & 2                                    & $-0.471 \pm 0.003$ & $0.158 \pm 0.002$ & ---             \\
+                                                                                               & 3                                    & $-0.298 \pm 0.017$ & $0.415 \pm 0.012$ & ---             \\ \hline
+                \multirow{3}{*}{``Wiggliness" -- $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $43.652 \pm 0.521$ & $33.662 \pm 0.576$ & ---             \\
+                                                                                               & 2                                    & $608.019 \pm 6.764$ & $717.285 \pm 9.194$ & ---             \\
+                                                                                               & 3                                    & $7101.915 \pm 155.145$ & $13617.196 \pm 312.209$ & ---             \\ \hline
                 \end{tabular}
@@
 
-            \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the HHMM-DFT.
-The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
+            \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the HHMM-DFT.
+Figures following $\pm$ refer to estimated standard errors derived from the inverse of the observed information matrix.
             \addtocounter{tablenum}{1}
@@
                                                                                                    &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
-                \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.16 (0.63)$ & $10.78 (0.55)$ & ---           \\
-                                                                                               & 2                                    & $124.2 (11.0)$ & $64.6 (8.8)$ & ---           \\ \hline
-                \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.064 (0.007)$ & $0.035 (0.001)$ & $0.874 (0.008)$ \\
-                                                                                               & 2                                    & $0.342 (0.017)$ & $0.095 (0.002)$ & $0.879 (0.005)$ \\
-                                                                                               & 3                                    & $0.257 (0.037)$ & $0.305 (0.010)$ & $0.605 (0.023)$ \\ \hline
-                \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.376 (0.009)$ & $0.044 (0.001)$ & $0.874 (0.008)$ \\
-                                                                                               & 2                                    & $0.470 (0.017)$ & $0.097 (0.002)$ & $0.879 (0.005)$ \\
-                                                                                               & 3                                    & $0.350 (0.045)$ & $0.377 (0.013)$ & $0.605 (0.023)$ \\ \hline
-                \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.462 (0.011)$ & $0.052 (0.001)$ & $0.874 (0.008)$ \\
-                                                                                               & 2                                    & $-0.660 (0.020)$ & $0.113 (0.002)$ & $0.879 (0.005)$ \\
-                                                                                               & 3                                    & $-0.306 (0.045)$ & $0.369 (0.013)$ & $0.605 (0.023)$ \\ \hline
+                    \multirow{2}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $27.163 \pm 0.627$ & $10.780 \pm 0.549$ & ---           \\
+                                                                                                   & 2                                    & $124.235 \pm 10.965$ & $64.590 \pm 8.816$ & ---           \\ \hline
+                    \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.064 \pm 0.007$ & $0.035 \pm 0.001$ & $0.874 \pm 0.008$ \\
+                                                                                                   & 2                                    & $0.342 \pm 0.017$ & $0.095 \pm 0.002$ & $0.879 \pm 0.005$ \\
+                                                                                                   & 3                                    & $0.257 \pm 0.037$ & $0.305 \pm 0.010$ & $0.605 \pm 0.023$ \\ \hline
+                    \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.376 \pm 0.009$ & $0.044 \pm 0.001$ & $0.874 \pm 0.008$ \\
+                                                                                                   & 2                                    & $0.470 \pm 0.017$ & $0.097 \pm 0.002$ & $0.879 \pm 0.005$ \\
+                                                                                                   & 3                                    & $0.350 \pm 0.045$ & $0.377 \pm 0.013$ & $0.605 \pm 0.023$ \\ \hline
+                    \multirow{3}{*}{$z$-Acc.
+$(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.462 \pm 0.011$ & $0.052 \pm 0.001$ & $0.874 \pm 0.008$ \\
+                                                                                                   & 2                                    & $-0.660 \pm 0.020$ & $0.113 \pm 0.002$ & $0.879 \pm 0.005$ \\
+                                                                                                   & 3                                    & $-0.306 \pm 0.045$ & $0.369 \pm 0.013$ & $0.605 \pm 0.023$ \\ \hline
                 \end{tabular}
@@
 
-                \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$) and acceleration ($\Zone_{t,\tilde t^*}$) of the killer whale case study data under the CarHHMM.
-The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
+                \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$) and acceleration ($\Zone_{t,\tilde t^*}$) of the killer whale case study data under the CarHHMM.
+Figures following $\pm$ refer to estimated standard errors derived from the inverse of the observed information matrix.
                 \addtocounter{tablenum}{1}
@@
                                                                                                    &                                      & $\hat \mu$    & $\hat \sigma$ & $\hat \phi$   \\ \hline
-            \multirow{1}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $41.70 (1.24)$ & $30.50 (1.22)$ & ---           \\ \hline
-            \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.460 (0.030)$ & $0.039 (0.001)$ & $0.968 (0.002)$ \\
-                                                                                           & 2                                    & $0.209 (0.012)$ & $0.096 (0.002)$ & $0.828 (0.007)$ \\
-                                                                                           & 3                                    & $0.232 (0.035)$ & $0.296 (0.010)$ & $0.607 (0.023)$ \\ \hline
-            \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.457 (0.038)$ & $0.052 (0.001)$ & $0.968 (0.002)$ \\
-                                                                                           & 2                                    & $0.436 (0.012)$ & $0.094 (0.002)$ & $0.828 (0.007)$ \\
-                                                                                           & 3                                    & $0.365 (0.042)$ & $0.366 (0.012)$ & $0.607 (0.023)$ \\ \hline
-            \multirow{3}{*}{$z$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.704 (0.044)$ & $0.059 (0.001)$ & $0.968 (0.002)$ \\
-                                                                                           & 2                                    & $-0.571 (0.014)$ & $0.111 (0.002)$ & $0.828 (0.007)$ \\
-                                                                                           & 3                                    & $-0.301 (0.042)$ & $0.355 (0.012)$ & $0.607 (0.023)$ \\ \hline
-            \multirow{3}{*}{Wiggliness - $\Ztwo_{t,\tilde t^*}$}                           & 1                                    & $34.40 (0.37)$ & $23.33 (0.38)$ & ---           \\
-                                                                                           & 2                                    & $495.0 (5.7)$ & $505.4 (6.9)$ & ---           \\
-                                                                                           & 3                                    & $9190 (220)$ & $13590 (360)$ & ---           \\ \hline
+                    \multirow{1}{*}{Dive Duration $(s)$ -- $Y_t$}                                  & 1                                    & $41.695 \pm 1.244$ & $30.499 \pm 1.220$ & ---           \\ \hline
+                    \multirow{3}{*}{$x$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_x$}   & 1                                    & $0.460 \pm 0.030$ & $0.039 \pm 0.001$ & $0.968 \pm 0.002$ \\
+                                                                                                   & 2                                    & $0.209 \pm 0.012$ & $0.096 \pm 0.002$ & $0.828 \pm 0.007$ \\
+                                                                                                   & 3                                    & $0.232 \pm 0.035$ & $0.296 \pm 0.010$ & $0.607 \pm 0.023$ \\ \hline
+                    \multirow{3}{*}{$y$-Acc. $(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_y$}   & 1                                    & $0.457 \pm 0.038$ & $0.052 \pm 0.001$ & $0.968 \pm 0.002$ \\
+                                                                                                   & 2                                    & $0.436 \pm 0.012$ & $0.094 \pm 0.002$ & $0.828 \pm 0.007$ \\
+                                                                                                   & 3                                    & $0.365 \pm 0.042$ & $0.366 \pm 0.012$ & $0.607 \pm 0.023$ \\ \hline
+                    \multirow{3}{*}{$z$-Acc.
+$(m/s^2)$ -- $\left(\Zone_{t,\tilde t^*}\right)_z$}   & 1                                    & $-0.704 \pm 0.044$ & $0.059 \pm 0.001$ & $0.968 \pm 0.002$ \\
+                                                                                                   & 2                                    & $-0.571 \pm 0.014$ & $0.111 \pm 0.002$ & $0.828 \pm 0.007$ \\
+                                                                                                   & 3                                    & $-0.301 \pm 0.042$ & $0.355 \pm 0.012$ & $0.607 \pm 0.023$ \\ \hline
+                    \multirow{3}{*}{``Wiggliness" -- $\Ztwo_{t,\tilde t^*}$}                          & 1                                    & $34.403 \pm 0.371$ & $23.331 \pm 0.381$ & ---           \\
+                                                                                                   & 2                                    & $495.033 \pm 5.691$ & $505.423 \pm 6.879$ & ---           \\
+                                                                                                   & 3                                    & $9189.109 \pm 223.965$ & $13588.427 \pm 360.082$ & ---           \\ \hline
                 \end{tabular}
@@
 
-                \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and wiggliness ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the CarHMM-DFT.
-The parentheses refer to the estimated standard error of the parameter estimates derived from the inverse of the observed information matrix.
+                \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the CarHMM-DFT.
+Figures following $\pm$ refer to estimated standard errors derived from the inverse of the observed information matrix.
                 \addtocounter{tablenum}{1}
@@
     \newpage
-    \section{Decoded dive states}
+    \section{Decoded Dive States}
 
@@
 
-        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $(y^*_{t,t^*})_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
+        \noindent Figure \arabic{fignum}: The $x$-component of acceleration, $(y^*_{t,t^*})_x$ (top two panels), and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
 Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHHMM-DFT to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
+Curve colour in the first and third panels corresponds to estimated dive type while curve colour in the second and fourth panels corresponds to estimated subdive state.
+Both dive type and subdive state are estimated by fitting the CarHHMM-DFT to the data and applying the forward--backward algorithm to determine the hidden state with the highest probability.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model.
+        \noindent Figure \arabic{fignum}: The probability that each dive is of a particular type or that the whale is in a particular subdive state, given the fitted model.
 Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHHMM-DFT to the data and performing the forward-backward algorithm.
+Curve colour in the first and third panels corresponds to estimated dive type while curve colour in the second and fourth panels corresponds to estimated subdive state.
+Both dive type and subdive state are estimated by fitting the CarHHMM-DFT to the data and applying the forward--backward algorithm.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $(y^*_{t,t^*})_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
+        \noindent Figure \arabic{fignum}: The $x$-component of acceleration, $(y^*_{t,t^*})_x$ (top two panels), and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
 Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the HHMM-DFT to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
+Curve colour in the first and third panels corresponds to estimated dive type while curve colour in the second and fourth panels corresponds to estimated subdive state.
+Both dive type and subdive state are estimated by fitting the HHMM-DFT to the data and applying the forward--backward algorithm to determine the hidden state with the highest probability.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model.
+        \noindent Figure \arabic{fignum}: The probability that each dive is of a particular type or that the whale is in a particular subdive state, given the fitted model.
 Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the HHMM-DFT to the data and performing the forward-backward algorithm.
+Curve colour in the first and third panels corresponds to estimated dive type while curve colour in the second and fourth panels corresponds to estimated subdive state.
+Both dive type and subdive state are estimated by fitting the HHMM-DFT to the data and applying the forward--backward algorithm.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $(y^*_{t,t^*})_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
+        \noindent Figure \arabic{fignum}: The $x$-component of acceleration, $(y^*_{t,t^*})_x$ (top two panels), and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
 Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHHMM to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
+Curve colour in the first and third panels corresponds to estimated dive type while curve colour in the second and fourth panels corresponds to estimated subdive state.
+Both dive type and subdive state are estimated by fitting the CarHHMM to the data and applying the forward--backward algorithm to determine the hidden state with the highest probability.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model.
+        \noindent Figure \arabic{fignum}: The probability that each dive is of a particular type or that the whale is in a particular subdive state, given the fitted model.
 Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHHMM to the data and performing the forward-backward algorithm.
+Curve colour in the first and third panels corresponds to estimated dive type while curve colour in the second and fourth panels corresponds to estimated subdive state.
+Both dive type and subdive state are estimated by fitting the CarHHMM to the data and applying the forward--backward algorithm.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: The $x$--component of acceleration $(y^*_{t,t^*})_x$ (top two panels) and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
+        \noindent Figure \arabic{fignum}: The $x$-component of acceleration, $(y^*_{t,t^*})_x$ (top two panels), and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
 Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHMM-DFT to the data and performing the forward-backward algorithm to determine the hidden state with the highest probability.
+Curve colour in the first and third panels corresponds to estimated dive type while curve colour in the second and fourth panels corresponds to estimated subdive state.
+Both dive type and subdive state are estimated by fitting the CarHMM-DFT to the data and applying the forward--backward algorithm to determine the hidden state with the highest probability.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: The probability that each dive is a particular type or that the whale is in a particular subdive state, given the fitted model.
+        \noindent Figure \arabic{fignum}: The probability that each dive is of a particular type or that the whale is in a particular subdive state, given the fitted model.
 Each panel is partitioned into dives by vertical black lines.
-The curve colours in the first and third panels correspond to the estimated dive types while the curve colours of the second and fourth panels correspond the estimated subdive states.
-Both the dive types and subdive states are estimated by fitting the CarHMM-DFT to the data and performing the forward-backward algorithm.
+Curve colour in the first and third panels corresponds to estimated dive type while curve colour in the second and fourth panels corresponds to estimated subdive state.
+Both dive type and subdive state are estimated by fitting the CarHMM-DFT to the data and applying the forward--backward algorithm.
         \addtocounter{fignum}{1}
@@
     \newpage
-    \section{Model checking - dive duration ($Y_t$)}
+    \section{Model Checking: Dive Duration ($Y_t$)}
 
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histogram (left) and psuedoresiduals (right) of dive duration ($Y_{t}$) plotted over the estimated emission distributions and a standard normal density, respectively.
+        \noindent Figure \arabic{fignum}: Empirical histogram (left) and pseudoresidual histogram (right) for dive duration ($Y_{t}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 Both plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histogram (left) and psuedoresiduals (right) of dive duration ($Y_{t}$) plotted over the estimated emission distributions and a standard normal density, respectively.
+        \noindent Figure \arabic{fignum}: Empirical histogram (left) and pseudoresidual histogram (right) for dive duration ($Y_{t}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 Both plots are generated using the fitted HHMM-DFT and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histogram (left) and psuedoresiduals (right) of dive duration ($Y_{t}$) plotted over the estimated emission distribution (left) and a standard normal density (right), respectively.
+        \noindent Figure \arabic{fignum}: Empirical histogram (left) and pseudoresidual histogram (right) for dive duration ($Y_{t}$) plotted over the estimated emission distribution (left) and a standard normal density (right), respectively.
 Both plots are generated using the fitted CarHMM-DFT and the killer whale case study data.
@@
     \newpage
-    \section{Model checking - acceleration ($\Zone_{t,\tilde t^*}$)}
+    \section{Model Checking: Acceleration ($\Zone_{t,\tilde t^*}$)}
 
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-Note that the mean of acceleration at time $\tilde t^*$ depends upon acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted.
+        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and pseudoresidual histograms (bottom row) for acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
+Note that the mean acceleration at time $\tilde t^*$ depends on acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted.
 All plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
+        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and pseudoresidual histograms (bottom row) for acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 All plots are generated using the fitted HHMM-DFT and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-Note that the mean of acceleration at time $\tilde t^*$ depends upon acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted.
+        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and pseudoresidual histograms (bottom row) for acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
+Note that the mean acceleration at time $\tilde t^*$ depends on acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted.
 All plots are generated using the fitted CarHHMM and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and psuedoresiduals (bottom row) of acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
-Note that the mean of acceleration at time $\tilde t^*$ depends upon acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted.
+        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and pseudoresidual histograms (bottom row) for acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
+Note that the mean acceleration at time $\tilde t^*$ depends on acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted.
 All plots are generated using the fitted CarHMM-DFT and the killer whale case study data.
@@
 
-    \section{Model checking - wiggliness ($\Ztwo_{t,\tilde t^*}$)}
+    \section{Model Checking: ``Wiggliness" ($\Ztwo_{t,\tilde t^*}$)}
 
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and psuedoresiduals (bottom row) of wiggliness ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
+        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and pseudoresidual histograms (bottom row) for ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 All plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and psuedoresiduals (bottom row) of wiggliness ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
+        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and pseudoresidual histograms (bottom row) for ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 All plots are generated using the fitted HHMM-DFT and the killer whale case study data.
@@
 
-        The CarHHMM does not model wiggliness.
+        The CarHHMM does not model ``wiggliness".
 
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and psuedoresiduals (bottom row) of wiggliness ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distributions and a standard normal density, respectively.
+        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and pseudoresidual histograms (bottom row) for ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 All plots are generated using the fitted CarHMM-DFT and the killer whale case study data.

### 2021-10-26 Report/Supp_Sim.tex
Update Final Version

--- 2021-06-21 Report/Supp_Sim.tex
@@
 % Title, authors, affiliations
-\title{Modelling state-switching functional data with hidden Markov models: supplementary material (simulation study)}%\query{Q1}
+\title{Modelling multi-scale, state-switching functional data with hidden Markov models: Supplement B (simulation study)}
 \date{}
-\author{BlindedA, BlindedB}%\authorref{1}\thanksref{*}}
-%\author{BlindedB}%\authorref{2}}
-%\affiliation[1]{Author affiliations will go here in the accepted manuscript, but do NOT include them in your initial submission because it must be anonymous.}
-%\affiliation[2]{Second Affiliation}
+\author{Evan Sidrow, Nancy Heckman, Sarah M. E. Fortune, \\ Andrew W. Trites, Ian Murphy, and Marie Auger-M\'eth\'e}
 
@@
 
-    	\noindent Figure \arabic{fignum}: Acceleration data for a five dives of a selected simulated data set.
-The raw accelerometer data is denoted as $Y^*_{t,\tilde t^*}$ and recorded at 50 times per second while $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are recorded at once every two seconds.
+    	\noindent Figure \arabic{fignum}: Acceleration data for five dives of a selected simulated data set.
+The raw accelerometer data is denoted by $Y^*_{t,\tilde t^*}$ and recorded at 50 $Hz$ while $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are recorded at $0.5 Hz$.
 Dives are separated by vertical black lines.
-The colour of the curve corresponds to the true subdive state, while the colour of the background corresponds to the true dive type.
-Subdive state 1 can plausibly be interpreted as ``gliding", subdive state 2 can plausibly be interpreted as active swimming, and subdive state 3 can be interpretted as sudden or aggressive movement.
+Curve colour corresponds to the true subdive state, while background colour corresponds to true dive type.
+Subdive state 1 can plausibly be interpreted as ``gliding", subdive state 2 can plausibly be interpreted as active swimming, and subdive state 3 can be interpreted as sudden or aggressive movement.
     	\addtocounter{fignum}{1}
@@
     \newpage
-    \section{Accuracy comparisons}
+    \section{Accuracy Comparisons}
 
@@
 \begin{tabular}{ccccccc}
-Model                       & \multicolumn{1}{c}{Train Time (m)} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Dive Accuracy} & \multicolumn{1}{c}{Subdive Accuracy}  \\ \hline
+Model                       & \multicolumn{1}{c}{Training time ($min$)} & \multicolumn{1}{c}{Dive type} & \multicolumn{1}{c}{Subdive state} & \multicolumn{1}{c}{Dive accuracy} & \multicolumn{1}{c}{Subdive accuracy}  \\ \hline
 \multirow{7}{*}{CarHHMM-DFT}& \multirow{7}{*}{$156 \pm 67$}   & All                           & All                              & $0.956 \pm 0.028$                   & $0.911 \pm 0.006$                       \\
@@
         \noindent Table \arabic{tablenum}: Average decoding accuracies and training times for all models used to categorize dive type and subdive state in the simulation study.
-Each of the four models was fit to 500 training data sets comprised of 100 simulated dives and tested on test data sets also comprised of 100 simulated dives.
-Reported values are averages, and $\pm$ refers to the sample standard deviation across the 500 data sets.
-Rows labelled All/All correspond to overall average decoding accuracy.
+Each of the four models was fit to 500 training data sets composed of 100 simulated dives and tested on test data sets also composed of 100 simulated dives.
+Reported values are averages, and figures following $\pm$ refer to sample standard deviation across the 500 data sets.
+``All" denotes overall average decoding accuracy.
         \addtocounter{tablenum}{1}
@@
     \newpage
-    \section{Empirical joint distributions - dive duration ($Y_t$)}
+    \section{Empirical Joint Distributions: Dive Duration ($Y_t$)}
 
@@
 \begin{tabular}{ccccccc}
-Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Dive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}   & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
-\multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.33$                         & $0.07$ $(p=0.287)$          & $1.39$                             & $1.03 \pm 0.12$                             \\
-                            &                               & 2                             & $127.14$                        & $0.07$ $(p=0.945)$          & $22.56$                             & $15.71 \pm 6.726$                             \\
-                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.80$                         & $-0.11$ $(p=0.035)$          & $1.15$                             & $0.84 \pm 0.11$                             \\
-                            &                               & 2                             & $59.52$                         & $-4.54$ $(p=0.000)$          & $17.68$                             & $13.04 \pm 5.53$                             \\ \hline
-\multirow{4}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.37$                         & $0.10$ $(p=0.084)$          & $1.33$                             & $1.03 \pm 0.11$                             \\
-                            &                               & 2                             & $126.84$                         & $-0.23$ $(p=0.818)$          & $22.00$                             & $15.37 \pm 6.11$                             \\
-                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.86$                         & $-0.05$ $(p=0.351)$          & $1.19$                             & $0.85 \pm 0.11$                             \\
-                            &                               & 2                             & $57.14$                         & $-6.92$ $(p=0.000)$          & $16.12$                             & $12.62 \pm 5.17$                             \\ \hline
-\multirow{4}{*}{CarHHMM}    & \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.25$                         & $-0.02$ $(p=0.826)$          & $2.20$                             & $1.04 \pm 0.14$                             \\
-                            &                               & 2                             & $117.62$                         & $-9.45$ $(p=0.000)$          & $34.34$                             & $15.13 \pm 7.10$                             \\
-                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.52$                         & $-0.39$ $(p=0.001)$          & $2.59$                             & $0.86 \pm 0.15$                             \\
-                            &                               & 2                             & $56.98$                         & $-7.08$ $(p=0.000)$          & $16.33$                             & $12.53 \pm 5.95$                             \\ \hline
+Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Dive type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}   & \multicolumn{1}{c}{Observed Fisher SE}     \\ \hline
+\multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.33$                         & $0.07$ $(P=0.287)$          & $1.39$                             & $1.03 \pm 0.12$                             \\
+                            &                               & 2                             & $127.14$                        & $0.07$ $(P=0.945)$          & $22.56$                             & $15.71 \pm 6.726$                             \\
+                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.80$                         & $-0.11$ $(P=0.035)$          & $1.15$                             & $0.84 \pm 0.11$                             \\
+                            &                               & 2                             & $59.52$                         & $-4.54$ $(P=0.000)$          & $17.68$                             & $13.04 \pm 5.53$                             \\ \hline
+\multirow{4}{*}{HHMM-DFT}   & \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.37$                         & $0.10$ $(P=0.084)$          & $1.33$                             & $1.03 \pm 0.11$                             \\
+                            &                               & 2                             & $126.84$                         & $-0.23$ $(P=0.818)$          & $22.00$                             & $15.37 \pm 6.11$                             \\
+                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.86$                         & $-0.05$ $(P=0.351)$          & $1.19$                             & $0.85 \pm 0.11$                             \\
+                            &                               & 2                             & $57.14$                         & $-6.92$ $(P=0.000)$          & $16.12$                             & $12.62 \pm 5.17$                             \\ \hline
+\multirow{4}{*}{CarHHMM}    & \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.25$                         & $-0.02$ $(P=0.826)$          & $2.20$                             & $1.04 \pm 0.14$                             \\
+                            &                               & 2                             & $117.62$                         & $-9.45$ $(P=0.000)$          & $34.34$                             & $15.13 \pm 7.10$                             \\
+                            & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.52$                         & $-0.39$ $(P=0.001)$          & $2.59$                             & $0.86 \pm 0.15$                             \\
+                            &                               & 2                             & $56.98$                         & $-7.08$ $(P=0.000)$          & $16.33$                             & $12.53 \pm 5.95$                             \\ \hline
 \multirow{4}{*}{CarHMM-DFT} & \multirow{2}{*}{$\mu_Y$}      & 1                             & $41.69$                         & ---                        & $4.27$                             & $2.00 \pm 0.35$                             \\
@@
         \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of dive duration across all four models in the simulation study.
-The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
-The $\pm$ refers to the inter-quartile range.
-SE stands for standard error.
-Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
+The ``Estimate" column gives averages across 500 simulations, and the ``Observed Fisher SE" column gives medians across 500 simulations.
+Figures following $\pm$ refer to inter quartile range.
+``SE" stands for standard error.
+Reported $P$-values correspond to one-sample $t$-tests of whether bias is statistically significant.
         \addtocounter{tablenum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
 These estimates are associated with the CarHHMM-DFT and correspond to both dive types.
-The red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
+White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
 These estimates are associated with the HHMM-DFT and correspond to both dive types.
-The red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
+White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
 These estimates are associated with the CarHHMM and correspond to both dive types.
-The red star represents the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
+White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu_Y$ and $\sigma_Y$ for both dive types.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of dive duration parameter estimates ($\hat \mu_Y$ and $\hat \sigma_Y$).
-These estimates are associated with the CarHMM-DFT.
-There is no red star which represents the true values because the CarHMM-DFT assumes that there is only one dive type.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
+This estimate is associated with the CarHMM-DFT.
+The White star represents the mean of all maximum likelihood estimates.
+Unlike in other figures, there is no red star representing true values because the CarHMM-DFT assumes that there is only one dive type.
         \addtocounter{fignum}{1}
@@
     \newpage
-    \section{Empirical joint distributions - acceleration ($\Zone_{t,\tilde t^*}$)}
+    \section{Empirical Joint Distributions: Acceleration ($\Zone_{t,\tilde t^*}$)}
 
@@
 \begin{tabular}{ccccccc}
-Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}   & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
-\multirow{9}{*}{CarHHMM-DFT}& \multirow{3}{*}{$\mu_A^*$}    & 1                                & $-0.012$                         & $-0.012$ $(p=0.004)$          & $0.093$                             & $0.066 \pm 0.038$                             \\
-                            &                               & 2                                & $0.099$                         & $-0.001$ $(p=0.436)$          & $0.019$                             & $0.019 \pm 0.003$                             \\
-                            &                               & 3                                & $0.200$                         & $0.000$ $(p=0.965)$          & $0.060$                             & $0.053 \pm 0.013$                             \\
-                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.050$                         & $0.000$ $(p=0.376)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
-                            &                               & 2                                & $0.100$                         & $-0.000$ $(p=0.672)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
-                            &                               & 3                                & $0.298$                         & $-0.002$ $(p=0.024)$          & $0.016$                             & $0.015 \pm 0.002$                             \\
-                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & $0.968$                         & $-0.002$ $(p=0.001)$          & $0.012$                             & $0.010 \pm 0.002$                             \\
-                            &                               & 2                                & $0.828$                         & $-0.002$ $(p=0.007)$          & $0.017$                             & $0.017 \pm 0.002$                             \\
-                            &                               & 3                                & $0.600$                         & $-0.010$ $(p=0.001)$          & $0.064$                             & $0.065 \pm 0.011$                             \\ \hline
-\multirow{9}{*}{HHMM-DFT}   & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $0.048$                         & $0.048$ $(p=0.000)$          & $0.033$                             & $0.007 \pm 0.001$                             \\
-                            &                               & 2                                & $0.087$                         & $-0.013$ $(p=0.000)$          & $0.018$                             & $0.006 \pm 0.001$                             \\
-                            &                               & 3                                & $0.215$                         & $0.015$ $(p=0.000)$          & $0.065$                             & $0.022 \pm 0.004$                             \\
-                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.160$                         & $0.110$ $(p=0.000)$          & $0.023$                             & $0.005 \pm 0.001$                             \\
-                            &                               & 2                                & $0.158$                         & $0.058$ $(p=0.000)$          & $0.010$                             & $0.004 \pm 0.000$                             \\
-                            &                               & 3                                & $0.356$                         & $0.056$ $(p=0.000)$          & $0.035$                             & $0.015 \pm 0.002$                             \\
+Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive state} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE}   & \multicolumn{1}{c}{Observed Fisher SE}     \\ \hline
+\multirow{9}{*}{CarHHMM-DFT}& \multirow{3}{*}{$\mu_A^*$}    & 1                                & $-0.012$                         & $-0.012$ $(P=0.004)$          & $0.093$                             & $0.066 \pm 0.038$                             \\
+                            &                               & 2                                & $0.099$                         & $-0.001$ $(P=0.436)$          & $0.019$                             & $0.019 \pm 0.003$                             \\
+                            &                               & 3                                & $0.200$                         & $0.000$ $(P=0.965)$          & $0.060$                             & $0.053 \pm 0.013$                             \\
+                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.050$                         & $0.000$ $(P=0.376)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
+                            &                               & 2                                & $0.100$                         & $-0.000$ $(P=0.672)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
+                            &                               & 3                                & $0.298$                         & $-0.002$ $(P=0.024)$          & $0.016$                             & $0.015 \pm 0.002$                             \\
+                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & $0.968$                         & $-0.002$ $(P=0.001)$          & $0.012$                             & $0.010 \pm 0.002$                             \\
+                            &                               & 2                                & $0.828$                         & $-0.002$ $(P=0.007)$          & $0.017$                             & $0.017 \pm 0.002$                             \\
+                            &                               & 3                                & $0.600$                         & $-0.010$ $(P=0.001)$          & $0.064$                             & $0.065 \pm 0.011$                             \\ \hline
+\multirow{9}{*}{HHMM-DFT}   & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $0.048$                         & $0.048$ $(P=0.000)$          & $0.033$                             & $0.007 \pm 0.001$                             \\
+                            &                               & 2                                & $0.087$                         & $-0.013$ $(P=0.000)$          & $0.018$                             & $0.006 \pm 0.001$                             \\
+                            &                               & 3                                & $0.215$                         & $0.015$ $(P=0.000)$          & $0.065$                             & $0.022 \pm 0.004$                             \\
+                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.160$                         & $0.110$ $(P=0.000)$          & $0.023$                             & $0.005 \pm 0.001$                             \\
+                            &                               & 2                                & $0.158$                         & $0.058$ $(P=0.000)$          & $0.010$                             & $0.004 \pm 0.000$                             \\
+                            &                               & 3                                & $0.356$                         & $0.056$ $(P=0.000)$          & $0.035$                             & $0.015 \pm 0.002$                             \\
                             & \multirow{3}{*}{$\phi_A^*$}   & 1                                & ------                         & ------                     & ------                             & ------                                      \\
@@
                             &                               & 3                                & ------                         & ------                     & ------                             & ------                                      \\ \hline
-\multirow{9}{*}{CarHHMM}    & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $0.060$                         & $0.060$ $(p=0.000)$          & $0.034$                             & $0.019 \pm 0.006$                             \\
-                            &                               & 2                                & $0.097$                         & $-0.003$ $(p=0.000)$          & $0.016$                             & $0.007 \pm 0.001$                             \\
-                            &                               & 3                                & $0.201$                         & $0.001$ $(p=0.634)$          & $0.058$                             & $0.062 \pm 0.011$                             \\
-                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.111$                         & $0.061$ $(p=0.000)$          & $0.004$                             & $0.003 \pm 0.000$                             \\
-                            &                               & 2                                & $0.285$                         & $0.185$ $(p=0.000)$          & $0.009$                             & $0.005 \pm 0.000$                             \\
-                            &                               & 3                                & $1.150$                         & $0.850$ $(p=0.000)$          & $0.080$                             & $0.048 \pm 0.009$                             \\
-                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & $0.830$                         & $-0.140$ $(p=0.000)$          & $0.048$                             & $0.016 \pm 0.003$                             \\
-                            &                               & 2                                & $0.004$                         & $-0.826$ $(p=0.000)$          & $0.012$                             & $0.001 \pm 0.023$                             \\
-                            &                               & 3                                & $0.000$                         & $-0.610$ $(p=0.000)$          & $0.000$                             & $0.000 \pm 0.000$                             \\ \hline
-\multirow{9}{*}{CarHMM-DFT} & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $-0.008$                         & $-0.008$ $(p=0.142)$          & $0.114$                             & $0.064 \pm 0.037$                             \\
-                            &                               & 2                                & $0.101$                         & $0.001$ $(p=0.438)$          & $0.019$                             & $0.019 \pm 0.003$                             \\
-                            &                               & 3                                & $0.200$                         & $0.000$ $(p=0.926)$          & $0.057$                             & $0.054 \pm 0.014$                             \\
-                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.050$                         & $-0.000$ $(p=0.005)$          & $0.001$                             & $0.002 \pm 0.000$                             \\
-                            &                               & 2                                & $0.100$                         & $-0.000$ $(p=0.000)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
-                            &                               & 3                                & $0.299$                         & $-0.001$ $(p=0.117)$          & $0.015$                             & $0.015 \pm 0.002$                             \\
-                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & $0.968$                         & $-0.002$ $(p=0.001)$          & $0.012$                             & $0.010 \pm 0.002$                             \\
-                            &                               & 2                                & $0.830$                         & $-0.000$ $(p=0.552)$          & $0.016$                             & $0.017 \pm 0.002$                             \\
-                            &                               & 3                                & $0.595$                         & $-0.015$ $(p=0.000)$          & $0.071$                             & $0.065 \pm 0.010$                             \\ \hline
+\multirow{9}{*}{CarHHMM}    & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $0.060$                         & $0.060$ $(P=0.000)$          & $0.034$                             & $0.019 \pm 0.006$                             \\
+                            &                               & 2                                & $0.097$                         & $-0.003$ $(P=0.000)$          & $0.016$                             & $0.007 \pm 0.001$                             \\
+                            &                               & 3                                & $0.201$                         & $0.001$ $(P=0.634)$          & $0.058$                             & $0.062 \pm 0.011$                             \\
+                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.111$                         & $0.061$ $(P=0.000)$          & $0.004$                             & $0.003 \pm 0.000$                             \\
+                            &                               & 2                                & $0.285$                         & $0.185$ $(P=0.000)$          & $0.009$                             & $0.005 \pm 0.000$                             \\
+                            &                               & 3                                & $1.150$                         & $0.850$ $(P=0.000)$          & $0.080$                             & $0.048 \pm 0.009$                             \\
+                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & $0.830$                         & $-0.140$ $(P=0.000)$          & $0.048$                             & $0.016 \pm 0.003$                             \\
+                            &                               & 2                                & $0.004$                         & $-0.826$ $(P=0.000)$          & $0.012$                             & $0.001 \pm 0.023$                             \\
+                            &                               & 3                                & $0.000$                         & $-0.610$ $(P=0.000)$          & $0.000$                             & $0.000 \pm 0.000$                             \\ \hline
+\multirow{9}{*}{CarHMM-DFT} & \multirow{3}{*}{$\mu_A^*$}    & 1                                & $-0.008$                         & $-0.008$ $(P=0.142)$          & $0.114$                             & $0.064 \pm 0.037$                             \\
+                            &                               & 2                                & $0.101$                         & $0.001$ $(P=0.438)$          & $0.019$                             & $0.019 \pm 0.003$                             \\
+                            &                               & 3                                & $0.200$                         & $0.000$ $(P=0.926)$          & $0.057$                             & $0.054 \pm 0.014$                             \\
+                            & \multirow{3}{*}{$\sigma_A^*$} & 1                                & $0.050$                         & $-0.000$ $(P=0.005)$          & $0.001$                             & $0.002 \pm 0.000$                             \\
+                            &                               & 2                                & $0.100$                         & $-0.000$ $(P=0.000)$          & $0.002$                             & $0.002 \pm 0.000$                             \\
+                            &                               & 3                                & $0.299$                         & $-0.001$ $(P=0.117)$          & $0.015$                             & $0.015 \pm 0.002$                             \\
+                            & \multirow{3}{*}{$\phi_A^*$}   & 1                                & $0.968$                         & $-0.002$ $(P=0.001)$          & $0.012$                             & $0.010 \pm 0.002$                             \\
+                            &                               & 2                                & $0.830$                         & $-0.000$ $(P=0.552)$          & $0.016$                             & $0.017 \pm 0.002$                             \\
+                            &                               & 3                                & $0.595$                         & $-0.015$ $(P=0.000)$          & $0.071$                             & $0.065 \pm 0.010$                             \\ \hline
 \end{tabular}
@@
         \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of acceleration across all four models in the simulation study.
-The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
-The $\pm$ refers to the inter-quartile range.
-SE stands for standard error.
+The ``Estimate" column gives averages across 500 simulations, and the ``Observed Fisher SE" column gives medians across 500 simulations.
+Figures following $\pm$ refer to inter quartile range.
+``SE" stands for standard error.
 Note that standard error cannot be estimated when $\phi^*_A \approx 0$ or $\phi^*_A \approx 1$.
-Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
+Reported $P$-values correspond to one-sample $t$-tests of whether bias is statistically significant.
         \addtocounter{tablenum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of acceleration parameter estimates ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
+        \noindent Figure \arabic{fignum}: Kernel density estimates of the distribution of acceleration parameter estimators ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
+White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for each subdive state.
 These estimates are for the CarHHMM-DFT.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of acceleration parameter estimates ($\hat \mu^*_A$ and $\hat \sigma^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$ and $\sigma^*_A$ for both subdive states.
+        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of acceleration parameter estimators ($\hat \mu^*_A$ and $\hat \sigma^*_A$) for each subdive state.
+White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu^*_A$ and $\sigma^*_A$ for each subdive state.
 These estimates are for the HHMM-DFT.
-Note that the HHMM-DFT assumes no auto-correlation in $\Zone_{t,\tilde t^*}$, so $\hat \phi_A^{*(\cdot,i^*)}$ is absent from these plots.
+Note that the HHMM-DFT assumes no autocorrelation in $\Zone_{t,\tilde t^*}$, so $\hat \phi_A^{*(\cdot,i^*)}$ is absent from these plots.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of acceleration parameter estimates ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
+        \noindent Figure \arabic{fignum}: Kernel density estimates of the distribution of acceleration parameter estimators ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
+White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for each subdive state.
 These estimates are for the CarHHMM.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of acceleration parameter estimates ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
-The red star represents the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for both subdive states.
+        \noindent Figure \arabic{fignum}: Kernel density estimates of the distribution of acceleration parameter estimators ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
+White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for each subdive state.
 These estimates are for the CarHMM-DFT.
@@
     \newpage
-    \section{Empirical joint distributions - wiggliness ($\Ztwo_{t,\tilde t^*}$)}
+    \section{Empirical Joint Distributions: ``Wiggliness" ($\Ztwo_{t,\tilde t^*}$)}
 
@@
         \def\arraystretch{1.5}
+
 \begin{tabular}{ccccccc}
-Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive Type} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}       \\ \hline
-\multirow{6}{*}{CarHHMM-DFT}& \multirow{3}{*}{$\mu_W^*$}    & 1                                & $34.981$                         & $0.001$ $(p=0.978)$          & $1.026$                             & $0.658 \pm 0.088$                             \\
-                            &                               & 2                                & $505.10$                         & $-0.69$ $(p=0.394)$          & $17.94$                             & $9.74 \pm 0.81$                             \\
-                            &                               & 3                                & $9787.1$                         & $17.2$ $(p=0.718)$           & $1060.5$                            & $262.1 \pm 97.120$                             \\
-                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.849$                         & $-0.001$ $(p=0.991)$         & $0.989$                             & $0.658 \pm 0.088$                             \\
-                            &                               & 2                                & $514.21$                         & $-2.47$ $(p=0.010)$          & $21.43$                             & $9.74 \pm 0.81$                             \\
-                            &                               & 3                                & $14430$                          & $-33$ $(p=0.647)$            & $1596$                              & $262 \pm 97$                             \\ \hline
-\multirow{6}{*}{HHMM-DFT}   & \multirow{3}{*}{$\mu_W^*$}    & 1                                & $35.067$                         & $0.087$ $(p=0.097)$          & $1.170$                             & $0.679 \pm 0.100$                             \\
-                            &                               & 2                                & $500.93$                         & $-4.86$ $(p=0.000)$          & $24.02$                             & $10.75 \pm 1.21$                             \\
-                            &                               & 3                                & $6841.0$                         & $-2929.0$ $(p=0.000)$        & $1164.3$                             & $198.2 \pm 77.1$                             \\
-                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.800$                         & $-0.050$ $(p=0.351)$         & $1.192$                             & $0.679 \pm 0.100$                             \\
-                            &                               & 2                                & $513.26$                         & $-3.42$ $(p=0.001)$          & $22.60$                             & $10.75 \pm 1.21$                             \\
-                            &                               & 3                                & $11459$                          & $-3004$ $(p=0.000)$          & $1739$                              & $198 \pm 77$                             \\ \hline
+Model                       & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Subdive state} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias}   & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fisher SE}       \\ \hline
+\multirow{6}{*}{CarHHMM-DFT}& \multirow{3}{*}{$\mu_W^*$}    & 1                                & $34.981$                         & $0.001$ $(P=0.978)$          & $1.026$                             & $0.658 \pm 0.088$                             \\
+                            &                               & 2                                & $505.104$                         & $-0.686$ $(P=0.394)$          & $17.936$                             & $9.740 \pm 0.814$                             \\
+                            &                               & 3                                & $9787.146$                         & $17.166$ $(P=0.718)$          & $1060.450$                             & $262.144 \pm 97.120$                             \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.849$                         & $-0.001$ $(P=0.991)$          & $0.989$                             & $0.652 \pm 0.094$                             \\
+                            &                               & 2                                & $514.207$                         & $-2.473$ $(P=0.010)$          & $21.427$                             & $11.410 \pm 0.947$                             \\
+                            &                               & 3                                & $14429.732$                         & $-32.818$ $(P=0.647)$          & $1596.020$                             & $302.698 \pm 165.935$                             \\ \hline
+\multirow{6}{*}{HHMM-DFT}   & \multirow{3}{*}{$\mu_W^*$}    & 1                                & $35.067$                         & $0.087$ $(P=0.097)$          & $1.170$                             & $0.679 \pm 0.100$                             \\
+                            &                               & 2                                & $500.928$                         & $-4.862$ $(P=0.000)$          & $24.024$                             & $10.754 \pm 1.205$                             \\
+                            &                               & 3                                & $6840.995$                         & $-2928.985$ $(P=0.000)$          & $1164.288$                             & $198.162 \pm 77.050$                             \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.800$                         & $-0.050$ $(P=0.351)$          & $1.192$                             & $0.676 \pm 0.104$                             \\
+                            &                               & 2                                & $513.263$                         & $-3.417$ $(P=0.001)$          & $22.596$                             & $12.497 \pm 1.301$                             \\
+                            &                               & 3                                & $11458.795$                         & $-3003.755$ $(P=0.000)$          & $1738.793$                             & $280.244 \pm 136.259$                             \\ \hline
 \multirow{6}{*}{CarHHMM}    & \multirow{3}{*}{$\mu_W^*$}    & 1                                & ------                         & ------                     & ------                             & ------                                      \\
@@
                             &                               & 3                                & ------                         & ------                     & ------                             & ------                                      \\ \hline
-\multirow{6}{*}{CarHMM-DFT} & \multirow{3}{*}{$\mu_W^*$}    & 1                                & $35.034$                         & $0.054$ $(p=0.264)$          & $1.076$                             & $0.661 \pm 0.093$                             \\
-                            &                               & 2                                & $503.76$                         & $-2.03$ $(p=0.013)$          & $18.26$                             & $9.64 \pm 0.92$                             \\
-                            &                               & 3                                & $9817.8$                         & $47.8$ $(p=0.328)$           & $1089.1$                            & $280.2 \pm 136.3$                             \\
-                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.890$                         & $0.040$ $(p=0.401)$          & $1.058$                             & $0.661 \pm 0.093$                             \\
-                            &                               & 2                                & $515.23$                         & $-1.45$ $(p=0.125)$          & $21.09$                             & $9.64 \pm 0.92$                             \\
-                            &                               & 3                                & $14438$                          & $-25$ $(p=0.740)$            & $1661$                              & $280 \pm 136$                             \\
+\multirow{6}{*}{CarHMM-DFT} & \multirow{3}{*}{$\mu_W^*$}    & 1                                & $35.034$                         & $0.054$ $(P=0.264)$          & $1.076$                             & $0.661 \pm 0.093$                             \\
+                            &                               & 2                                & $503.762$                         & $-2.028$ $(P=0.013)$          & $18.258$                             & $9.641 \pm 0.916$                             \\
+                            &                               & 3                                & $9817.785$                         & $47.805$ $(P=0.328)$          & $1089.110$                             & $280.244 \pm 136.259$                             \\
+                            & \multirow{3}{*}{$\sigma_W^*$} & 1                                & $23.890$                         & $0.040$ $(P=0.401)$          & $1.058$                             & $0.656 \pm 0.101$                             \\
+                            &                               & 2                                & $515.228$                         & $-1.452$ $(P=0.125)$          & $21.087$                             & $11.371 \pm 0.990$                             \\
+                            &                               & 3                                & $14437.867$                         & $-24.683$ $(P=0.740)$          & $1660.764$                             & $302.698 \pm 165.935$                             \\
 \end{tabular}
@@
         \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of wiggliness across all four models in the simulation study.
-The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
-The $\pm$ refers to the inter-quartile range.
-SE stands for standard error.
-Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
+The ``Estimate" column gives averages across 500 simulations, and the ``Observed Fisher SE" column gives medians across 500 simulations.
+Figures following $\pm$ refer to inter quartile range.
+``SE" stands for standard error.
+Reported $P$-values correspond to one-sample $t$-tests of whether bias is statistically significant.
         \addtocounter{tablenum}{1}
@@
         \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of wiggliness parameter estimates ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
-The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
+White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu^*_W$ and $\sigma^*_W$ for each subdive state.
 These estimates are for the CarHHMM-DFT.
@@
         \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of wiggliness parameter estimates ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
-The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
+White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu^*_W$ and $\sigma^*_W$ for each subdive state.
 These estimates are for the HHMM-DFT.
@@
         \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of wiggliness parameter estimates ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
-The red star represents the true values of $\mu^*_W$ and $\sigma^*_W$ for both subdive states.
+White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu^*_W$ and $\sigma^*_W$ for each subdive state.
 These estimates are for the CarHMM-DFT.
@@
     \newpage
-    \section{Empirical joint distributions - probability transition matrices ($\Gamma, \Gamma^*$)}
+    \section{Empirical Joint Distributions - Transition Probability Matrices ($\Gamma$ and $\Gamma^*$)}
 
@@
 \begin{tabular}{ccccccc}
-Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
-\multirow{2}{*}{CarHHMM-DFT}& $\Gamma_{12}$                 & $0.160$                         & $0.005$ $(p=0.030)$        & $0.056$                           & $0.042 \pm 0.012$                             \\
-                             & $\Gamma_{21}$                 & $0.907$                         & $-0.006$ $(p=0.214)$        & $0.102$                           & $0.094 \pm 0.032$                             \\ \hline
-\multirow{2}{*}{HHMM-DFT}   & $\Gamma_{12}$                 & $0.161$                         & $0.007$ $(p=0.007)$        & $0.055$                           & $0.041 \pm 0.011$                             \\
-                             & $\Gamma_{21}$                 & $0.916$                         & $0.004$ $(p=0.372)$        & $0.088$                           & $0.091 \pm 0.028$                             \\ \hline
-\multirow{2}{*}{CarHHMM}    & $\Gamma_{12}$                 & $0.198$                         & $0.044$ $(p=0.000)$        & $0.152$                           & $0.043 \pm 0.013$                             \\
-                             & $\Gamma_{21}$                 & $0.832$                         & $-0.081$ $(p=0.000)$        & $0.195$                           & $0.093 \pm 0.034$                             \\ \hline
+Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fisher SE}     \\ \hline
+\multirow{2}{*}{CarHHMM-DFT}& $\Gamma_{12}$                 & $0.160$                         & $0.005$ $(P=0.030)$        & $0.056$                           & $0.042 \pm 0.012$                             \\
+                             & $\Gamma_{21}$                 & $0.907$                         & $-0.006$ $(P=0.214)$        & $0.102$                           & $0.094 \pm 0.032$                             \\ \hline
+\multirow{2}{*}{HHMM-DFT}   & $\Gamma_{12}$                 & $0.161$                         & $0.007$ $(P=0.007)$        & $0.055$                           & $0.041 \pm 0.011$                             \\
+                             & $\Gamma_{21}$                 & $0.916$                         & $0.004$ $(P=0.372)$        & $0.088$                           & $0.091 \pm 0.028$                             \\ \hline
+\multirow{2}{*}{CarHHMM}    & $\Gamma_{12}$                 & $0.198$                         & $0.044$ $(P=0.000)$        & $0.152$                           & $0.043 \pm 0.013$                             \\
+                             & $\Gamma_{21}$                 & $0.832$                         & $-0.081$ $(P=0.000)$        & $0.195$                           & $0.093 \pm 0.034$                             \\ \hline
 \multirow{2}{*}{CarHMM-DFT} & $\Gamma_{12}$                 & ------                         & ------                   & ------                           & ------                                      \\
@@
 
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for coarse-scale probability transition matrices across all four models in the simulation study.
-The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
-The $\pm$ refers to the inter-quartile range.
-SE stands for standard error.
+        \noindent Table \arabic{tablenum}: Estimates and standard errors for the coarse-scale transition probability matrices across all four models in the simulation study.
+The ``Estimate" column gives averages across 500 simulations, and the ``Observed Fisher SE" column gives medians across 500 simulations.
+Figures following $\pm$ refer to inter quartile range.
+``SE" stands for standard error.
 Note that standard error cannot be estimated when $\Gamma_{i,j} \approx 0$.
-Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
+Reported $P$-values correspond to one-sample $t$-tests of whether bias is statistically significant.
         \addtocounter{tablenum}{1}
@@
 \begin{tabular}{ccccccc}
-Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fischer SE}     \\ \hline
-\multirow{12}{*}{CarHHMM-DFT}& $\Gamma^{*(1)}_{12}$          & $0.258$                         & $0.006$ $(p=0.004)$        & $0.043$                           & $0.032 \pm 0.006$                             \\
-                             & $\Gamma^{*(1)}_{13}$          & $0.002$                         & $0.002$ $(p=0.000)$        & $0.004$                           & $0.009 \pm 0.004$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.083$                         & $0.001$ $(p=0.243)$        & $0.015$                           & $0.012 \pm 0.002$                             \\
-                             & $\Gamma^{*(1)}_{23}$          & $0.052$                         & $-0.000$ $(p=0.984)$        & $0.010$                           & $0.009 \pm 0.001$                             \\
-                             & $\Gamma^{*(1)}_{31}$          & $0.003$                         & $0.003$ $(p=0.000)$        & $0.008$                           & $0.014 \pm 0.005$                             \\
-                             & $\Gamma^{*(1)}_{32}$          & $0.234$                         & $0.002$ $(p=0.322)$        & $0.042$                           & $0.036 \pm 0.008$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.111$                         & $-0.001$ $(p=0.349)$        & $0.017$                           & $0.015 \pm 0.004$                             \\
-                             & $\Gamma^{*(2)}_{13}$          & $0.001$                         & $0.000$ $(p=0.000)$        & $0.002$                           & $0.003 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.151$                         & $0.001$ $(p=0.584)$        & $0.023$                           & $0.021 \pm 0.005$                             \\
-                             & $\Gamma^{*(2)}_{23}$          & $0.036$                         & $-0.000$ $(p=0.944)$        & $0.013$                           & $0.010 \pm 0.003$                             \\
-                             & $\Gamma^{*(2)}_{31}$          & $0.007$                         & $0.007$ $(p=0.000)$        & $0.024$                           & $0.025 \pm 0.015$                             \\
-                             & $\Gamma^{*(2)}_{32}$          & $0.252$                         & $0.026$ $(p=0.000)$        & $0.096$                           & $0.063 \pm 0.028$                             \\ \hline
-\multirow{12}{*}{HHMM-DFT}   & $\Gamma^{*(1)}_{12}$          & $0.240$                         & $-0.012$ $(p=0.000)$        & $0.042$                           & $0.032 \pm 0.006$                             \\
-                             & $\Gamma^{*(1)}_{13}$          & $0.003$                         & $0.003$ $(p=0.000)$        & $0.006$                           & $0.009 \pm 0.004$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.077$                         & $-0.005$ $(p=0.000)$        & $0.019$                           & $0.012 \pm 0.002$                             \\
-                             & $\Gamma^{*(1)}_{23}$          & $0.047$                         & $-0.004$ $(p=0.000)$        & $0.009$                           & $0.008 \pm 0.001$                             \\
-                             & $\Gamma^{*(1)}_{31}$          & $0.005$                         & $0.005$ $(p=0.000)$        & $0.009$                           & $0.014 \pm 0.004$                             \\
-                             & $\Gamma^{*(1)}_{32}$          & $0.153$                         & $-0.079$ $(p=0.000)$        & $0.029$                           & $0.026 \pm 0.005$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.109$                         & $-0.004$ $(p=0.000)$        & $0.018$                           & $0.016 \pm 0.004$                             \\
-                             & $\Gamma^{*(2)}_{13}$          & $0.002$                         & $0.002$ $(p=0.000)$        & $0.004$                           & $0.005 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.151$                         & $0.001$ $(p=0.712)$        & $0.031$                           & $0.023 \pm 0.006$                             \\
-                             & $\Gamma^{*(2)}_{23}$          & $0.037$                         & $0.002$ $(p=0.009)$        & $0.013$                           & $0.011 \pm 0.004$                             \\
-                             & $\Gamma^{*(2)}_{31}$          & $0.021$                         & $0.021$ $(p=0.000)$        & $0.026$                           & $0.023 \pm 0.012$                             \\
-                             & $\Gamma^{*(2)}_{32}$          & $0.135$                         & $-0.091$ $(p=0.000)$        & $0.057$                           & $0.038 \pm 0.018$                             \\ \hline
-\multirow{12}{*}{CarHHMM}    & $\Gamma^{*(1)}_{12}$          & $0.138$                         & $-0.114$ $(p=0.000)$        & $0.131$                           & $0.018 \pm 0.005$                             \\
-                             & $\Gamma^{*(1)}_{13}$          & $0.023$                         & $0.022$ $(p=0.000)$        & $0.107$                           & $0.007 \pm 0.004$                             \\
-                             & $\Gamma^{*(1)}_{21}$          & $0.037$                         & $-0.044$ $(p=0.000)$        & $0.054$                           & $0.007 \pm 0.001$                             \\
-                             & $\Gamma^{*(1)}_{23}$          & $0.027$                         & $-0.025$ $(p=0.000)$        & $0.050$                           & $0.004 \pm 0.001$                             \\
-                             & $\Gamma^{*(1)}_{31}$          & $0.013$                         & $0.013$ $(p=0.000)$        & $0.022$                           & $0.011 \pm 0.005$                             \\
-                             & $\Gamma^{*(1)}_{32}$          & $0.134$                         & $-0.099$ $(p=0.000)$        & $0.043$                           & $0.023 \pm 0.006$                             \\
-                             & $\Gamma^{*(2)}_{12}$          & $0.056$                         & $-0.056$ $(p=0.000)$        & $0.035$                           & $0.008 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{13}$          & $0.002$                         & $0.002$ $(p=0.004)$        & $0.017$                           & $0.002 \pm 0.001$                             \\
-                             & $\Gamma^{*(2)}_{21}$          & $0.069$                         & $-0.081$ $(p=0.000)$        & $0.032$                           & $0.011 \pm 0.003$                             \\
-                             & $\Gamma^{*(2)}_{23}$          & $0.017$                         & $-0.019$ $(p=0.000)$        & $0.012$                           & $0.005 \pm 0.002$                             \\
-                             & $\Gamma^{*(2)}_{31}$          & $0.011$                         & $0.011$ $(p=0.000)$        & $0.016$                           & $0.018 \pm 0.009$                             \\
-                             & $\Gamma^{*(2)}_{32}$          & $0.138$                         & $-0.088$ $(p=0.000)$        & $0.079$                           & $0.037 \pm 0.016$                             \\ \hline
+Model                        & \multicolumn{1}{c}{Parameter} & \multicolumn{1}{c}{Estimate}   & \multicolumn{1}{c}{Bias} & \multicolumn{1}{c}{Empirical SE} & \multicolumn{1}{c}{Observed Fisher SE}     \\ \hline
+\multirow{12}{*}{CarHHMM-DFT}& $\Gamma^{*(1)}_{12}$          & $0.258$                         & $0.006$ $(P=0.004)$        & $0.043$                           & $0.032 \pm 0.006$                             \\
+                             & $\Gamma^{*(1)}_{13}$          & $0.002$                         & $0.002$ $(P=0.000)$        & $0.004$                           & $0.009 \pm 0.004$                             \\
+                             & $\Gamma^{*(1)}_{21}$          & $0.083$                         & $0.001$ $(P=0.243)$        & $0.015$                           & $0.012 \pm 0.002$                             \\
+                             & $\Gamma^{*(1)}_{23}$          & $0.052$                         & $-0.000$ $(P=0.984)$        & $0.010$                           & $0.009 \pm 0.001$                             \\
+                             & $\Gamma^{*(1)}_{31}$          & $0.003$                         & $0.003$ $(P=0.000)$        & $0.008$                           & $0.014 \pm 0.005$                             \\
+                             & $\Gamma^{*(1)}_{32}$          & $0.234$                         & $0.002$ $(P=0.322)$        & $0.042$                           & $0.036 \pm 0.008$                             \\
+                             & $\Gamma^{*(2)}_{12}$          & $0.111$                         & $-0.001$ $(P=0.349)$        & $0.017$                           & $0.015 \pm 0.004$                             \\
+                             & $\Gamma^{*(2)}_{13}$          & $0.001$                         & $0.000$ $(P=0.000)$        & $0.002$                           & $0.003 \pm 0.002$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.151$                         & $0.001$ $(P=0.584)$        & $0.023$                           & $0.021 \pm 0.005$                             \\
+                             & $\Gamma^{*(2)}_{23}$          & $0.036$                         & $-0.000$ $(P=0.944)$        & $0.013$                           & $0.010 \pm 0.003$                             \\
+                             & $\Gamma^{*(2)}_{31}$          & $0.007$                         & $0.007$ $(P=0.000)$        & $0.024$                           & $0.025 \pm 0.015$                             \\
+                             & $\Gamma^{*(2)}_{32}$          & $0.252$                         & $0.026$ $(P=0.000)$        & $0.096$                           & $0.063 \pm 0.028$                             \\ \hline
+\multirow{12}{*}{HHMM-DFT}   & $\Gamma^{*(1)}_{12}$          & $0.240$                         & $-0.012$ $(P=0.000)$        & $0.042$                           & $0.032 \pm 0.006$                             \\
+                             & $\Gamma^{*(1)}_{13}$          & $0.003$                         & $0.003$ $(P=0.000)$        & $0.006$                           & $0.009 \pm 0.004$                             \\
+                             & $\Gamma^{*(1)}_{21}$          & $0.077$                         & $-0.005$ $(P=0.000)$        & $0.019$                           & $0.012 \pm 0.002$                             \\
+                             & $\Gamma^{*(1)}_{23}$          & $0.047$                         & $-0.004$ $(P=0.000)$        & $0.009$                           & $0.008 \pm 0.001$                             \\
+                             & $\Gamma^{*(1)}_{31}$          & $0.005$                         & $0.005$ $(P=0.000)$        & $0.009$                           & $0.014 \pm 0.004$                             \\
+                             & $\Gamma^{*(1)}_{32}$          & $0.153$                         & $-0.079$ $(P=0.000)$        & $0.029$                           & $0.026 \pm 0.005$                             \\
+                             & $\Gamma^{*(2)}_{12}$          & $0.109$                         & $-0.004$ $(P=0.000)$        & $0.018$                           & $0.016 \pm 0.004$                             \\
+                             & $\Gamma^{*(2)}_{13}$          & $0.002$                         & $0.002$ $(P=0.000)$        & $0.004$                           & $0.005 \pm 0.002$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.151$                         & $0.001$ $(P=0.712)$        & $0.031$                           & $0.023 \pm 0.006$                             \\
+                             & $\Gamma^{*(2)}_{23}$          & $0.037$                         & $0.002$ $(P=0.009)$        & $0.013$                           & $0.011 \pm 0.004$                             \\
+                             & $\Gamma^{*(2)}_{31}$          & $0.021$                         & $0.021$ $(P=0.000)$        & $0.026$                           & $0.023 \pm 0.012$                             \\
+                             & $\Gamma^{*(2)}_{32}$          & $0.135$                         & $-0.091$ $(P=0.000)$        & $0.057$                           & $0.038 \pm 0.018$                             \\ \hline
+\multirow{12}{*}{CarHHMM}    & $\Gamma^{*(1)}_{12}$          & $0.138$                         & $-0.114$ $(P=0.000)$        & $0.131$                           & $0.018 \pm 0.005$                             \\
+                             & $\Gamma^{*(1)}_{13}$          & $0.023$                         & $0.022$ $(P=0.000)$        & $0.107$                           & $0.007 \pm 0.004$                             \\
+                             & $\Gamma^{*(1)}_{21}$          & $0.037$                         & $-0.044$ $(P=0.000)$        & $0.054$                           & $0.007 \pm 0.001$                             \\
+                             & $\Gamma^{*(1)}_{23}$          & $0.027$                         & $-0.025$ $(P=0.000)$        & $0.050$                           & $0.004 \pm 0.001$                             \\
+                             & $\Gamma^{*(1)}_{31}$          & $0.013$                         & $0.013$ $(P=0.000)$        & $0.022$                           & $0.011 \pm 0.005$                             \\
+                             & $\Gamma^{*(1)}_{32}$          & $0.134$                         & $-0.099$ $(P=0.000)$        & $0.043$                           & $0.023 \pm 0.006$                             \\
+                             & $\Gamma^{*(2)}_{12}$          & $0.056$                         & $-0.056$ $(P=0.000)$        & $0.035$                           & $0.008 \pm 0.002$                             \\
+                             & $\Gamma^{*(2)}_{13}$          & $0.002$                         & $0.002$ $(P=0.004)$        & $0.017$                           & $0.002 \pm 0.001$                             \\
+                             & $\Gamma^{*(2)}_{21}$          & $0.069$                         & $-0.081$ $(P=0.000)$        & $0.032$                           & $0.011 \pm 0.003$                             \\
+                             & $\Gamma^{*(2)}_{23}$          & $0.017$                         & $-0.019$ $(P=0.000)$        & $0.012$                           & $0.005 \pm 0.002$                             \\
+                             & $\Gamma^{*(2)}_{31}$          & $0.011$                         & $0.011$ $(P=0.000)$        & $0.016$                           & $0.018 \pm 0.009$                             \\
+                             & $\Gamma^{*(2)}_{32}$          & $0.138$                         & $-0.088$ $(P=0.000)$        & $0.079$                           & $0.037 \pm 0.016$                             \\ \hline
 \multirow{6}{*}{CarHMM-DFT}  & $\Gamma^{*(1)}_{12}$          & $0.158$                         & ------                   & $0.019$                           & $0.014 \pm 0.002$                             \\
@@
 
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for fine-scale probability transition matrices across all four models in the simulation study.
-The ``Estimate" column is the average across 500 simulations, and the ``Observed Fisher SE" column is the median across 500 simulations.
-The $\pm$ refers to the inter-quartile range.
-SE stands for standard error.
+        \noindent Table \arabic{tablenum}: Estimates and standard errors for fine-scale transition probability matrices across all four models in the simulation study.
+The ``Estimate" column gives averages across 500 simulations, and the ``Observed Fisher SE" column gives medians across 500 simulations.
+Figures following $\pm$ refer to inter quartile range.
+``SE" stands for standard error.
 Note that standard error cannot be estimated when $\Gamma^{*(i)}_{i^*,j^*} \approx 0$.
-Reported $p$-values test if the observed bias is statistically significant using a one-sample $t$-test.
+Reported $P$-values correspond to one-sample $t$-tests of whether bias is statistically significant.
         \addtocounter{tablenum}{1}
@@
         \noindent Figure \arabic{fignum}: Kernel density estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM-DFT.
-There are two fine-scale probability transition matrices- one corresponding to each dive type.
-We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix.
-The red star represents the true values and the white star represents the mean of all maximum likelihood estimates.
+There are two fine-scale transition probability matrices---one corresponding to each dive type.
+We plot only the off-diagonal elements of the transition probability matrices since these elements uniquely define a stochastic matrix.
+Red stars represent the true values and white stars represent the mean of all maximum likelihood estimates.
         \addtocounter{fignum}{1}
@@
         \noindent Figure \arabic{fignum}: Kernel density estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the HHMM-DFT.
-There are two fine-scale probability transition matrices- one corresponding to each dive type.
-We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix.
-The red star represents the true values and the white star represents the mean of all maximum likelihood estimates.
+There are two fine-scale transition probability matrices---one corresponding to each dive type.
+We plot only the off-diagonal elements of the transition probability matrices since these elements uniquely define a stochastic matrix.
+Red stars represent the true values and white stars represent the mean of all maximum likelihood estimates.
         \addtocounter{fignum}{1}
@@
         \noindent Figure \arabic{fignum}: Kernel density estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM.
-There are two fine-scale probability transition matrices- one corresponding to each dive type.
-We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix.
-The red star represents the true values and the white star represents the mean of all maximum likelihood estimates.
+There are two fine-scale transition probability matrices---one corresponding to each dive type.
+We plot only the off-diagonal elements of the transition probability matrices since these elements uniquely define a stochastic matrix.
+Red stars represent the true values and white stars represent the mean of all maximum likelihood estimates.
         \addtocounter{fignum}{1}
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of $\hat \Gamma^{*(1)}$ for the CarHMM-DFT.
-There is only one fine-scale probability transition matrix and no coarse-scale probability transition matrix since we assume that there is only one dive type for the CarHMM-DFT.
-We only plot the off-diagonal elements of the probability transition matrices since this uniquely defines a stochastic matrix.
-There is no red star which represents the true values because the CarHMM-DFT assumes that there is only one dive type.
+        \noindent Figure \arabic{fignum}: Kernel density estimates of the distribution of $\hat \Gamma^{*(1)}$ for the CarHMM-DFT.
+There is only one fine-scale transition probability matrix and no coarse-scale transition probability matrix since the CarHMM-DFT assumes that there is only one dive type.
+We plot only the off-diagonal elements of the transition probability matrix since these elements uniquely define a stochastic matrix.
+White stars represent the mean of all maximum likelihood estimates.
+Unlike other figures, there are no red stars which represents the true values because the CarHMM-DFT assumes that there is only one dive type.
         \addtocounter{fignum}{1}

### 2021-12-07 Report/Sections/appendix.tex
Update Report

@@
 
-We simulate the raw acceleration data for dive $t$ in three steps: (1) simulate the average acceleration within each window $\left(\hat Y^{*(0)}_{t,\tilde t^*}\right)$, (2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and (3) take the inverse discrete Fourier transform of $\hat{Y}^*_{t,\tilde t^*}$,
+We simulate the raw acceleration data for dive $t$ in three steps: (1) simulate the average acceleration within each window $\left(\hat Y^{*(0)}_{t,\tilde t^*}\right)$, (2) simulate all other Fourier coefficients within each window $\left(\hat Y^{*(k)}_{t,\tilde t^*}, k = 1,\ldots,99\right)$, and (3) take the inverse DFT of $\hat{Y}^*_{t,\tilde t^*}$,
 
@@
 \[
-    \Ztwo_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.80/(k+1)^3 , 515.38 \right), \text{and}
+    \Ztwo_{t,\tilde t^*} \mid \tilde X^*_{t,\tilde t^*} = 2 ~~\sim~~ {\rm{Gamma}}\left(\sum_{k=1}^{\tilde \omega} 4.80/(k+1)^3 , 515.38 \right),
 \]
@@
 
-\[\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,\z) = \delta P(y_1,\z_1;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,\z_t;\theta,\theta^*,\Gamma^*) \mathbf{1}_N.\]
+\[\calL_{\text{CarHHMM-DFT}}(\theta,\theta^*,\Gamma,\Gamma^*;y,\z) = \delta P(y_1,\z_1;\theta,\theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,\z_t;\theta,\theta^*,\Gamma^*) \mathbf{1}_N\]
 %
@@
 P(y_t,\z_t;\theta,\theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\theta^{(1)})\calL_{\text{fine}}\left(\theta^*,\Gamma^{*(1)};\z_t\right), \ldots , \\
-&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{fine}}\left(\theta^*,\Gamma^{*(N)};\z_t\right) \Big],
+&f^{(N)}(y_t;\theta^{(N)})\calL_{\text{fine}}\left(\theta^*,\Gamma^{*(N)};\z_t\right) \Big]
 \end{align*}

### 2021-12-07 Report/Sections/results.tex
Update Report

@@
 Understanding animal behaviour is important for conservation efforts, as environmental changes caused by anthropogenic activity can directly impact animal behaviour \citep{Sutherland:1998}.
-HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., DeRuiter et al., 2017; Isojunno et al.,2017)\nocite{DeRuiter:2017,Isojunno:2017}.
+HMMs have been used to understand how diving behaviours of various species are affected by disturbances (e.g., DeRuiter et al., 2017; Isojunno et al., 2017)\nocite{DeRuiter:2017,Isojunno:2017}.
 For killer whales, we are interested in categorizing different diving behaviours and identifying potential foraging dives.
@@
 The coarse-scale parameter estimates suggest that the killer whale has at least two distinct dive behaviours (see Table \ref{table:emis_dists_CarHHMM-DFT} and Figure \ref{fig:emis}). 
-Dive type 1 corresponds to shorter and shallower dives that likely reflect resting, travelling, and to a lesser extent searching for prey.
+Dive type 1 corresponds to shorter and shallower dives that likely reflect resting, travelling, and, to a lesser extent, searching for prey.
 Dive type 2 is longer and deeper and may be associated with behaviours such as hunting  \citep{Tennessen:2019b}, but it is unclear whether any of the dives in this study are successful foraging dives.
-No dive in this data set has a maximum depth greater than 30 $m$, while \citet{Wright:2017}, in a study of killer whales in Johnstone Strait found that most prey captures occur at depths deeper than 100 $m$.
+No dive in this data set has a maximum depth greater than 30 $m$, while \citet{Wright:2017}, in a study of killer whales in Johnstone Strait, found that most prey captures occur at depths deeper than 100 $m$.
 However, the killer whale studied here was tagged north of Johnstone Strait in Queen Charlotte Sound, and unpublished data collected by the authors suggest that prey-capture events can occur near the surface.
@@
 \end{pmatrix},\]
-
+%
 $\hat \delta^{*(1)} = \left(0.202, 0.649, 0.149\right)$, and $\hat \delta^{*(2)} = \left( 0.531, 0.405, 0.064 \right)$ for dive types 1 and 2.

### 2021-12-07 Report/Sections/simulation.tex
Update Report

@@
 After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each simulated dive are ignored.
-Each two-second segment is assigned one of $N^*=3$ behaviours according to a fine-scale Markov chain $\tilde X^*_t = \big\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \big\}$ with transition probability matrices
+Each 2 $s$ segment is assigned one of $N^*=3$ behaviours according to a fine-scale Markov chain $\tilde X^*_t = \big\{\tilde X^*_{t,1}, \ldots, \tilde X^*_{t,\tilde T^*_t} \big\}$ with the transition probability matrices
 
@@
 We calculate the maximum likelihood estimates $\{\hat \theta, \hat \Gamma, \hat \theta^*, \hat \Gamma^*\}$ for all four candidate models for each of the 500 data sets using an optimization procedure similar to that in the case study.
-
+%
 For each of the 500 training data sets, we simulate a test data set to assess how well each model predicts the hidden states as follows.

### 2021-12-07 Report/Supp_Case.tex
Finish copy editor edits
Update Report

@@
 
-        \noindent Figure \arabic{fignum}: Collection of five lag plots, where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
+        \noindent Figure S\arabic{fignum}: Collection of five lag plots, where a given observation is on the $x$-axis and the subsequent observation is on the $y$-axis.
 Each plot corresponds to a feature in the killer whale dive data.
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the dive duration parameters $\mu_Y^{(i)}$ and $\sigma_Y^{(i)}$ for dive types $i = 1,2$.
-All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+        \noindent Figure S\arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the dive duration parameters $\mu_Y^{(i)}$ and $\sigma_Y^{(i)}$ for dive types $i = 1,2$.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of the main text.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_Y,\hat \sigma_Y)$, which is denoted by red stars.
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the coarse-scale transition probability matrix parameters $\Gamma_{1,2}$ and $\Gamma_{2,1}$.
-All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+        \noindent Figure S\arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the coarse-scale transition probability matrix parameters $\Gamma_{1,2}$ and $\Gamma_{2,1}$.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of the main text.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \Gamma_{1,2},\hat \Gamma_{2,1})$, which is denoted by a red star.
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the ``wiggliness" parameters $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+        \noindent Figure S\arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the ``wiggliness" parameters $\mu_W^{*(\cdot,i^*)}$ and $\sigma_W^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of the main text.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_W^{*(\cdot,i^*)}, \hat \sigma_W^{*(\cdot,i^*)})$, which is denoted by red stars.
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $x$-acceleration parameters $\mu_{A_x}^{*(\cdot,i^*)}, \sigma_{A_x}^{*(\cdot,i^*)}$ and the logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+        \noindent Figure S\arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $x$-acceleration parameters $\mu_{A_x}^{*(\cdot,i^*)}, \sigma_{A_x}^{*(\cdot,i^*)}$ and the logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of the main text.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_x}^{*(\cdot,i^*)}, \hat \sigma_{A_x}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by red stars.
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $y$-acceleration parameters $\mu_{A_y}^{*(\cdot,i^*)}, \sigma_{A_y}^{*(\cdot,i^*)}$ and the logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+        \noindent Figure S\arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $y$-acceleration parameters $\mu_{A_y}^{*(\cdot,i^*)}, \sigma_{A_y}^{*(\cdot,i^*)}$ and the logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of the main text.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_y}^{*(\cdot,i^*)}, \hat \sigma_{A_y}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by red stars.
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $z$-acceleration parameters $\mu_{A_z}^{*(\cdot,i^*)}, \sigma_{A_z}^{*(\cdot,i^*)}$ and the logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
-All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+        \noindent Figure S\arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the $z$-acceleration parameters $\mu_{A_z}^{*(\cdot,i^*)}, \sigma_{A_z}^{*(\cdot,i^*)}$ and the logit of $\phi_{A}^{*(\cdot,i^*)}$ for subdive types $i^* = 1,2,3$.
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of the main text.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $(\hat \mu_{A_z}^{*(\cdot,i^*)}, \hat \sigma_{A_z}^{*(\cdot,i^*)}, \hat \phi_{A}^{*(\cdot,i^*)})$, which is denoted by red stars.
@@
 
-        \noindent Figure \arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the fine-scale transition probability matrix parameters associated with both dive types ($\Gamma^{*(i)}_{i^*,j^*}$ for dive types $i = 1,2$ and subdive states $i^*,j^* = 1,2,3$).
-All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of Section 3.1.
+        \noindent Figure S\arabic{fignum}: Log-likelihood of the CarHHMM-DFT as a function of the fine-scale transition probability matrix parameters associated with both dive types ($\Gamma^{*(i)}_{i^*,j^*}$ for dive types $i = 1,2$ and subdive states $i^*,j^* = 1,2,3$).
+All other parameters and transition probability matrices are set to be equal to the maximum likelihood estimates shown in Table 1 of the main text.
 The log-likelihood is added to the negative log-likelihood of the maximum likelihood estimate $\hat \Gamma^{*(i)}$, which is denoted by red stars.
@@
 
-            \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the full CarHHMM-DFT.
+            \noindent Table S\arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the full CarHHMM-DFT.
 Figures following $\pm$ refer to estimated standard errors derived from the inverse of the observed information matrix.
@@
 
-            \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the HHMM-DFT.
+            \noindent Table S\arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the HHMM-DFT.
 Figures following $\pm$ refer to estimated standard errors derived from the inverse of the observed information matrix.
@@
 
-                \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$) and acceleration ($\Zone_{t,\tilde t^*}$) of the killer whale case study data under the CarHHMM.
+                \noindent Table S\arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$) and acceleration ($\Zone_{t,\tilde t^*}$) of the killer whale case study data under the CarHHMM.
 Figures following $\pm$ refer to estimated standard errors derived from the inverse of the observed information matrix.
@@
 
-                \noindent Table \arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the CarHMM-DFT.
+                \noindent Table S\arabic{tablenum}: Estimates and predicted standard errors of emission parameters for the distributions of dive duration ($Y_t$), acceleration ($\Zone_{t,\tilde t^*}$), and ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) of the killer whale case study data under the CarHMM-DFT.
 Figures following $\pm$ refer to estimated standard errors derived from the inverse of the observed information matrix.
@@
 
-        \noindent Figure \arabic{fignum}: The $x$-component of acceleration, $(y^*_{t,t^*})_x$ (top two panels), and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
+        \noindent Figure S\arabic{fignum}: The $x$-component of acceleration, $(y^*_{t,t^*})_x$ (top two panels), and dive depth (bottom two panels) of a northern resident killer whale for a sequence of six selected dives.
 Each panel is partitioned into dives by vertical black lines.
@@
 
-        \noindent Figure \arabic{fignum}: The probability that each dive is of a particular type or that the whale is in a particular subdive state, given the fitted model.
+        \noindent Figure S\arabic{fignum}: The probability that each dive is of a particular type or that the whale is in a particular subdive state, given the fitted model.
 Each panel is partitioned into dives by vertical black lines.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histogram (left) and pseudoresidual histogram (right) for dive duration ($Y_{t}$) plotted over the estimated emission distribution and a standard normal density, respectively.
+        \noindent Figure S\arabic{fignum}: Empirical histogram (left) and pseudoresidual histogram (right) for dive duration ($Y_{t}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 Both plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histogram (left) and pseudoresidual histogram (right) for dive duration ($Y_{t}$) plotted over the estimated emission distribution and a standard normal density, respectively.
+        \noindent Figure S\arabic{fignum}: Empirical histogram (left) and pseudoresidual histogram (right) for dive duration ($Y_{t}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 Both plots are generated using the fitted HHMM-DFT and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histogram (left) and pseudoresidual histogram (right) for dive duration ($Y_{t}$) plotted over the estimated emission distribution (left) and a standard normal density (right), respectively.
+        \noindent Figure S\arabic{fignum}: Empirical histogram (left) and pseudoresidual histogram (right) for dive duration ($Y_{t}$) plotted over the estimated emission distribution (left) and a standard normal density (right), respectively.
 Both plots are generated using the fitted CarHMM-DFT and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and pseudoresidual histograms (bottom row) for acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
+        \noindent Figure S\arabic{fignum}: Empirical histograms (top three rows) and pseudoresidual histograms (bottom row) for acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 Note that the mean acceleration at time $\tilde t^*$ depends on acceleration at time $\tilde t^*-1$, so only the deviation from the conditional mean at each particular time step is plotted.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top three rows) and pseudoresidual histograms (bottom row) for acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
+        \noindent Figure S\arabic{fignum}: Empirical histograms (top three rows) and pseudoresidual histograms (bottom row) for acceleration ($\Zone_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 All plots are generated using the fitted HHMM-DFT and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and pseudoresidual histograms (bottom row) for ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
+        \noindent Figure S\arabic{fignum}: Empirical histograms (top row) and pseudoresidual histograms (bottom row) for ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 All plots are generated using the fitted CarHHMM-DFT and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and pseudoresidual histograms (bottom row) for ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
+        \noindent Figure S\arabic{fignum}: Empirical histograms (top row) and pseudoresidual histograms (bottom row) for ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 All plots are generated using the fitted HHMM-DFT and the killer whale case study data.
@@
 
-        \noindent Figure \arabic{fignum}: Empirical histograms (top row) and pseudoresidual histograms (bottom row) for ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
+        \noindent Figure S\arabic{fignum}: Empirical histograms (top row) and pseudoresidual histograms (bottom row) for ``wiggliness" ($\Ztwo_{t,\tilde t^*}$) plotted over the estimated emission distribution and a standard normal density, respectively.
 All plots are generated using the fitted CarHMM-DFT and the killer whale case study data.

### 2021-12-07 Report/Supp_Sim.tex
Finish copy editor edits
Update Report

@@
 
-    	\noindent Figure \arabic{fignum}: Acceleration data for five dives of a selected simulated data set.
+    	\noindent Figure S\arabic{fignum}: Acceleration data for five dives of a selected simulated data set.
 The raw accelerometer data is denoted by $Y^*_{t,\tilde t^*}$ and recorded at 50 $Hz$ while $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are recorded at $0.5 Hz$.
@@
 
-        \noindent Table \arabic{tablenum}: Average decoding accuracies and training times for all models used to categorize dive type and subdive state in the simulation study.
+        \noindent Table S\arabic{tablenum}: Average decoding accuracies and training times for all models used to categorize dive type and subdive state in the simulation study.
 Each of the four models was fit to 500 training data sets composed of 100 simulated dives and tested on test data sets also composed of 100 simulated dives.
@@
 \multirow{4}{*}{CarHHMM-DFT}& \multirow{2}{*}{$\mu_Y$}      & 1                             & $27.33$                         & $0.07$ $(P=0.287)$          & $1.39$                             & $1.03 \pm 0.12$                             \\
-                            &                               & 2                             & $127.14$                        & $0.07$ $(P=0.945)$          & $22.56$                             & $15.71 \pm 6.726$                             \\
+                            &                               & 2                             & $127.14$                        & $0.07$ $(P=0.945)$          & $22.56$                             & $15.71 \pm 6.73$                             \\
                             & \multirow{2}{*}{$\sigma_Y$}   & 1                             & $10.80$                         & $-0.11$ $(P=0.035)$          & $1.15$                             & $0.84 \pm 0.11$                             \\
@@
 
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of dive duration across all four models in the simulation study.
+        \noindent Table S\arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of dive duration across all four models in the simulation study.
 The ``Estimate" column gives averages across 500 simulations, and the ``Observed Fisher SE" column gives medians across 500 simulations.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
+        \noindent Figure S\arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
 These estimates are associated with the CarHHMM-DFT and correspond to both dive types.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
+        \noindent Figure S\arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
 These estimates are associated with the HHMM-DFT and correspond to both dive types.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
+        \noindent Figure S\arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
 These estimates are associated with the CarHHMM and correspond to both dive types.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
+        \noindent Figure S\arabic{fignum}: Kernel density estimate of the distribution of the dive duration parameter estimators ($\hat \mu_Y$ and $\hat \sigma_Y$).
 This estimate is associated with the CarHMM-DFT.
@@
 
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of acceleration across all four models in the simulation study.
+        \noindent Table S\arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of acceleration across all four models in the simulation study.
 The ``Estimate" column gives averages across 500 simulations, and the ``Observed Fisher SE" column gives medians across 500 simulations.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimates of the distribution of acceleration parameter estimators ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
+        \noindent Figure S\arabic{fignum}: Kernel density estimates of the distribution of acceleration parameter estimators ($\hat \mu^*_A$, $\hat \sigma^*_A$, and $\hat \phi^*_A$) for each subdive state.
 White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu^*_A$, $\sigma^*_A$, and $\phi^*_A$ for each subdive state.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distribution of acceleration parameter estimators ($\hat \mu^*_A$ and $\hat \sigma^*_A$) for each subdive state.
+        \noindent Figure S\arabic{fignum}: Kernel density estimate of the distribution of acceleration parameter estimators ($\hat \mu^*_A$ and $\hat \sigma^*_A$) for each subdive state.
 White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu^*_A$ and $\sigma^*_A$ for each subdive state.
@@
 
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of wiggliness across all four models in the simulation study.
+        \noindent Table S\arabic{tablenum}: Estimates and standard errors for the parameters of the distribution of wiggliness across all four models in the simulation study.
 The ``Estimate" column gives averages across 500 simulations, and the ``Observed Fisher SE" column gives medians across 500 simulations.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimate of the distributions of wiggliness parameter estimates ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
+        \noindent Figure S\arabic{fignum}: Kernel density estimate of the distributions of wiggliness parameter estimates ($\hat \mu^*_W$ and $\hat \sigma^*_W$) for each subdive state.
 White stars represent the mean of all maximum likelihood estimates and red stars represent the true values of $\mu^*_W$ and $\sigma^*_W$ for each subdive state.
@@
 
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for the coarse-scale transition probability matrices across all four models in the simulation study.
+        \noindent Table S\arabic{tablenum}: Estimates and standard errors for the coarse-scale transition probability matrices across all four models in the simulation study.
 The ``Estimate" column gives averages across 500 simulations, and the ``Observed Fisher SE" column gives medians across 500 simulations.
@@
 
-        \noindent Table \arabic{tablenum}: Estimates and standard errors for fine-scale transition probability matrices across all four models in the simulation study.
+        \noindent Table S\arabic{tablenum}: Estimates and standard errors for fine-scale transition probability matrices across all four models in the simulation study.
 The ``Estimate" column gives averages across 500 simulations, and the ``Observed Fisher SE" column gives medians across 500 simulations.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM-DFT.
+        \noindent Figure S\arabic{fignum}: Kernel density estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM-DFT.
 There are two fine-scale transition probability matrices---one corresponding to each dive type.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the HHMM-DFT.
+        \noindent Figure S\arabic{fignum}: Kernel density estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the HHMM-DFT.
 There are two fine-scale transition probability matrices---one corresponding to each dive type.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM.
+        \noindent Figure S\arabic{fignum}: Kernel density estimates of the distributions of $\hat \Gamma$, $\hat \Gamma^{*(1)}$, and $\hat \Gamma^{*(2)}$ for the CarHHMM.
 There are two fine-scale transition probability matrices---one corresponding to each dive type.
@@
 
-        \noindent Figure \arabic{fignum}: Kernel density estimates of the distribution of $\hat \Gamma^{*(1)}$ for the CarHMM-DFT.
+        \noindent Figure S\arabic{fignum}: Kernel density estimates of the distribution of $\hat \Gamma^{*(1)}$ for the CarHMM-DFT.
 There is only one fine-scale transition probability matrix and no coarse-scale transition probability matrix since the CarHMM-DFT assumes that there is only one dive type.
@@
 White stars represent the mean of all maximum likelihood estimates.
-Unlike other figures, there are no red stars which represents the true values because the CarHMM-DFT assumes that there is only one dive type.
+Unlike other figures, there are no red stars representing the true values because the CarHMM-DFT assumes that there is only one dive type.
         \addtocounter{fignum}{1}
